### 2012-10-30 gpi-debug/debugging.tex
add gpi debug lecture (skeleton)

\documentclass[xcolor=dvipsnames]{beamer}
\usepackage{lmodern}
\usepackage[T1]{fontenc}
\usepackage[english]{babel}
\usepackage[utf8]{inputenc}

\usepackage{manfnt}
\usepackage{wasysym}
\usepackage{listings}
\usepackage{graphicx}
\usepackage{url}
\usepackage{ulem}
\usepackage{marvosym}
\usepackage{proof}
\usepackage{array}
\setbeamertemplate{navigation symbols}{}

\title[Demystifying Debugging]{{\bf Demystifying Debugging}}
\subtitle[]{{\em because you can't get it right the first time every time}}
\author[Ben Blum]{Ben Blum \texttt{(bblum@andrew.cmu.edu)}}

\institute[98-172]{Great Practical Ideas for Computer Scientists}
\date[]{2012, November 1}

\setbeamertemplate{footline}{\hspace*{.5cm}\scriptsize{\insertauthor\hspace*{50pt} \hfill\insertframenumber\hspace*{.5cm}}}

\usecolortheme{seahorse}
\usecolortheme{rose}
\useoutertheme{infolines}

\usecolortheme[named=RoyalBlue]{structure}

\newcommand\noob{\mathsf{noob}}
\newcommand\gibs{\mathsf{gibs}}
\newcommand\dps{\mathsf{dps}}
\newcommand\squig\rightsquigarrow
\newcommand\Coloneqq{\mathrel{\mathop{::}}=}
\newcommand\dmg{\text{\Laserbeam}}
\newcommand\delter\delta
\newcommand\alpher\alpha
\newcommand\defnor{\text{ }|\text{ }}

\newcommand\pimp{\mathop{\supset}}
\newcommand\pand{\mathop{\wedge}}
\newcommand\por{\mathop{\vee}}
\newcommand\ptrue{\top}
\newcommand\pfalse{\bot}

\begin{document}
% \renewcommand{\inserttotalframenumber}{28} % If you want to hide bonus slides
\normalem
\begin{frame}
	\titlepage
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\newcommand\linegap{\vspace{0.2in}}
\newcommand\breakslide[1]{\begin{frame}{} \begin{center} \Large #1 \end{center} \end{frame}}
\newcommand\related[1]{\textsuperscript{\em [#1]}}
\newcommand\hilight[2]{\color{#1}#2\color{black}}

\begin{frame}{Outline}
	\textbf{Why teach debugging?}
	\linegap

	\textbf{``Tell Me a Story''}
	\begin{itemize}
		\item The good debugger's attitude
		\item Binary Search
	\end{itemize}
	\linegap

	{\bf Two Main Approaches}
	\begin{itemize}
		\item Time Travel
		\item Space Travel
	\end{itemize}
	\linegap

	{\bf Assorted Tricks}
	\begin{itemize}
		\item Question assumptions
		\item Collect data
	\end{itemize}
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Storytelling}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Main Approach}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{Travelling Through Time}

\breakslide{When things go wrong\dots}

\begin{frame}{foo}
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% vim: foldmethod=indent

### 2012-11-01 landslide-defence/defence.tex
add landslide lecture and defence

\documentclass[xcolor=dvipsnames]{beamer}
\usepackage{lmodern}
\usepackage[T1]{fontenc}
\usepackage[english]{babel}
\usepackage[utf8]{inputenc}

\usepackage{manfnt}
\usepackage{wasysym}
\usepackage{listings}
\usepackage{graphicx}
\usepackage{url}
\usepackage{ulem}
\usepackage{marvosym}
\usepackage{skull}
\usepackage{proof}
\usepackage{array}
\setbeamertemplate{navigation symbols}{}

\title[Landslide]{{\bf Landslide: Systematic Dynamic Race Detection in Kernel-space}}
\subtitle[]{ {\em showing that Pebbles are less stable than you thought since 2011.}}
\author[Ben Blum]{Ben Blum \texttt{(bblum@andrew.cmu.edu)}}

\institute[CMU]{Carnegie Mellon University}
\date[]{2012, May 10}

\setbeamertemplate{footline}{\hspace*{.5cm}\scriptsize{\insertauthor\hspace*{50pt} \hfill\insertframenumber\hspace*{.5cm}}}

\usecolortheme{seahorse}
\usecolortheme{rose}
\useoutertheme{infolines}

\usecolortheme[named=RoyalBlue]{structure}

\newcommand\noob{\mathsf{noob}}
\newcommand\gibs{\mathsf{gibs}}
\newcommand\dps{\mathsf{dps}}
\newcommand\squig\rightsquigarrow
\newcommand\Coloneqq{\mathrel{\mathop{::}}=}
\newcommand\dmg{\text{\Laserbeam}}
\newcommand\delter\delta
\newcommand\alpher\alpha
\newcommand\defnor{\text{ }|\text{ }}

\newcommand\pimp{\mathop{\supset}}
\newcommand\pand{\mathop{\wedge}}
\newcommand\por{\mathop{\vee}}
\newcommand\ptrue{\top}
\newcommand\pfalse{\bot}

\begin{document}
\normalem
\begin{frame}
	\titlepage
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\newcommand\linegap{\vspace{0.2in}}
\newcommand\breakslide[1]{\begin{frame}{} \begin{center} #1 \end{center} \end{frame}}
\newcommand\related[1]{\textsuperscript{\em [#1]}}

\begin{frame}{Outline}
	\textbf{Motivation: Concurrency debugging}
	\begin{itemize}
		\item Systematic testing versus stress testing
		\item Challenges of kernel-space
	\end{itemize}
	\linegap

	{\bf Tool: Landslide}
	\begin{itemize}
		\item Design and interface
		\item Addressing challenges
	\end{itemize}
	\linegap

	{\bf Evaluation: Finding Races}
	\begin{itemize}
		\item Student user study
		\item Case study
	\end{itemize}
	\linegap

	{\bf Future Work and Conclusion}
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Motivation}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Race Conditions and the Finding Thereof}

\begin{frame}[fragile]{Case Study}
	\begin{center}
	\begin{verbatim}
	        int thread_fork()
	        {
	            thread_t *child = spawn_new_thread();
	            add_to_runqueue(child);
	            return child->tid;
	        }
	\end{verbatim}
	\end{center}

	\begin{itemize}
		\item On exit, child's state is freed
		\item Forking thread does use-after-free
		\item Might return garbage instead of thread ID
	\end{itemize}
\end{frame}

\begin{frame}{Testing Techniques} % slide: sys ex vs stress testing
	\textbf{Stress Testing:} Common testing approach
	\begin{itemize}
		\item Attempting to exercise as many interleavings as possible
		\item Exposes race conditions at random
		\begin{itemize}
			\item ``If a preemption occurs at {\em just the right time}\dots''
		\end{itemize}
		\item Cryptic panic messages or machine reboots
	\end{itemize}
	\linegap

	{\bf Systematic Testing} \related{Musuvathi '08}
	\begin{itemize}
		\item Make educated guesses about when to preempt
		\item Run {\em every single} interleaving
		\item Provide better debugging information, reproducibility
	\end{itemize}
\end{frame}

\begin{frame}{Execution Tree} % slide: decision tree
	\begin{center}
		\includegraphics[width=\textwidth]{tree.png}
	\end{center}
\end{frame}

\begin{frame}{Finding Races in Kernel-space} % slide: 15-410 6-week project why we need it in kernels
	\textbf{15-410:} Undergraduate Operating Systems
	\begin{itemize}
		\item Project 3: students write a kernel in 6 weeks.
		\item ``Pebbles'' is a UNIX-like, preemptible kernel specification.
	\end{itemize}
	\linegap

	Previous work in kernel-space has not used systematic testing.
	%\begin{itemize}
	%	\item Data race detection\related{Erickson '10}
	%	\item Failure resiliency\related{Kadav '09}
	%\end{itemize}
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Challenges of Kernel-space}

\breakslide{Why is this environment different from all other environments?}

\begin{frame}{Causes of Concurrency} % slide: causes of concurrency, thread scheduling
	\textbf{Kernels contain their own concurrency implementation.}

	\linegap
	Context switching
	\begin{itemize}
		\item Non-deterministic timer-driven thread scheduling
	\end{itemize}
	Runqueue tracking
	\begin{itemize}
		\item Which threads are runnable?
	\end{itemize}
	Thread lifecycle tracking
	\begin{itemize}
		\item When are threads created/destroyed?
	\end{itemize}
\end{frame}

% slide: ad-hoc communication, memory conflicts (freeing, runqueue)
%\begin{frame}{Ad-Hoc Thread Communication}
%	No message-passing % TODO this slide sucks
%	
%	User-space systematic exploration relies on it\related{Simsa '10}; we cannot.
%	
%	Kernel threads share state via globals, heap.
%\end{frame}

\begin{frame}{The Kernel as One Program} % slide: the kernel is too big to test all at once
	% say: it's also important to understand the kernel's internals instead of just viewing it as one ``black box''
	\begin{center}
	\includegraphics[width=\textwidth]{pebbles.png}
	\end{center}
	% now that we see these internal interfaces, how do we apply systematic testing to it?
\end{frame}
\begin{frame}{The Kernel as One Program} % slide: the kernel is too big to test all at once
	``Everything interleaves with everything else''?
	\begin{itemize}
		% Say: "we don't want to have to explore every possible interaction between each of these components"
		\item Impractical to explore resulting state spaces
	\end{itemize}
	\linegap

	What is a ``bug'', anyway?\related{Engler '01}
	\begin{itemize}
		\item False-negative-oriented approach
	\end{itemize}
\end{frame}

\begin{frame}{Contribution} % slide: summary: "sys ex is a tool \ldots``
	Systematic testing is a {\em tool} for testing concurrent systems.
	\begin{itemize}
		\item How can it be applied in kernel-space?
		\item What simple things can we ask of the user to make it possible?
	\end{itemize}
	\linegap

	{\bf Landslide} is an effort to answer these questions.
	% say: landslide is an exploratory work into finding out what lies behind these questions.
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Landslide}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\newcommand\done[1]{\textcolor{gray}{\em \small #1}}

%\begin{frame}{Outline}
%	\done{Motivation: Concurrency debugging}
%	\begin{itemize}
%		\item \done{Systematic exploration versus stress testing}
%		\item \done{Challenges of kernel-space}
%	\end{itemize}
%	\linegap
%
%	{\bf Tool: Landslide}
%	\begin{itemize}
%		\item {\bf Design and interface}
%		\item {\bf Addressing challenges}
%	\end{itemize}
%	\linegap
%
%	Evaluation: Finding Races
%	\begin{itemize}
%		\item Student user study
%		\item Case study
%	\end{itemize}
%	\linegap
%
%	Future Work and Conclusion
%\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Design}

\begin{frame}{Environment}
	\textbf{Simics:} a full-system x86 simulator
	\begin{itemize}
		\item Students in 15-410 use Simics to test their kernels.
		\item Landslide is a transparently-embedded Simics module.
	\end{itemize}
\end{frame}

\begin{frame}{Anatomy of Landslide}
	\begin{center}
	\includegraphics[width=0.7\textwidth]{landslide.png}
	\end{center}
	% Remember to talk about solving the thread scheduling challenge
\end{frame}

\subsection{Interface}

\begin{frame}{Using Landslide}
	\begin{center}
	\includegraphics[width=0.9\textwidth]{tell_landslide.png}
	\end{center}
	User must tell Landslide when certain kernel events happen:
	\begin{itemize}
		\item When do threads become runnable / descheduled?
		\item When does the scheduler switch threads?
	\end{itemize}
	\linegap
	% Make sure to expound on what these are
	Some complicated conditions require in-Landslide instrumentation.
	\begin{itemize}
		\item Scheduler design details
	\end{itemize}
\end{frame}

\begin{frame}{Decision Trace}
	% FIXME make the stack trace coloured
	\includegraphics[width=\textwidth]{found_a_bug.png}
	% Say: So landslide had to explore 5 different interleavings to find the bug. But this was a pretty simple test case, and 
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Addressing Challenges} % rename

\begin{frame}{State Space Reduction}
	% Say: ``combinatorial explosion''
	\begin{center}
	\includegraphics[width=\textwidth]{undiamond1.png}
	\end{center}
\end{frame}
\begin{frame}{State Space Reduction}
	State spaces grow exponentially.
	\begin{itemize}
		\item Fortunately, some sequences result in identical states.
		\item Dynamic Partial Order Reduction\related{Flanagan '05}
		\begin{itemize}
			\item Requires ``memory independence relation'' between transitions.
		\end{itemize}
	\end{itemize}
\end{frame}

\begin{frame}{Shared Memory Conflicts} % slide: ignoring shared memory accesses - venn diagram
	\begin{center}
	\includegraphics[width=\textwidth]{shm-conflict0.png}
	\end{center}
\end{frame}
\begin{frame}{Shared Memory Conflicts}
	\begin{center}
	\includegraphics[width=\textwidth]{shm-conflict1.png}
	\end{center}
\end{frame}
\begin{frame}{Shared Memory Conflicts}
	\begin{center}
	\includegraphics[width=\textwidth]{shm-conflict2.png}
	\end{center}
\end{frame}
\begin{frame}{Shared Memory Conflicts}
	%\textbf{Problem:} Many conflicts we don't care about
	%\begin{itemize}
	%	\item Every transition accesses scheduler state.
	%	\item Many transitions access ``unrelated'' global objects.
	%\end{itemize}
	%\linegap
	\textbf{Solution:} Only consider ``relevant'' memory conflicts.
	\begin{itemize}
		\item Ignore scheduler, global objects assumed to be correct
		\begin{itemize}
			\item Sacrifice ability to test these to efficiently test everything else.
		\end{itemize}
		\item User must specify what to ignore.
	\end{itemize}
\end{frame}

\begin{frame}[fragile]{Identifying Bugs}
	\begin{columns}[l]
		\column{0.75\textwidth}
		\textbf{How do we know we've found a bug?}
		\linegap

		Definite bugs
		\begin{itemize}
			\item Kernel panics / assertion failures
			\item Use-after-free
			\item Deadlock
		\end{itemize}
		Probable bugs
		\begin{itemize}
			\item Infinite loop
			\begin{itemize}
				\item Use structure of execution tree to judge progress.
			\end{itemize}
		\end{itemize}
		\column{0.25\textwidth}
		\includegraphics[width=\textwidth]{blank-livelock.png}
	\end{columns}
\end{frame}
\begin{frame}[fragile]{Identifying Bugs}
	\begin{columns}[l]
		\column{0.75\textwidth}
		\textbf{How do we know we've found a bug?}
		\linegap

		Definite bugs
		\begin{itemize}
			\item Kernel panics / assertion failures
			\item Use-after-free
			\item Deadlock
		\end{itemize}
		Probable bugs
		\begin{itemize}
			\item Infinite loop
			\begin{itemize}
				\item Use structure of execution tree to judge progress.
			\end{itemize}
		\end{itemize}
		\column{0.25\textwidth}
		\includegraphics[width=\textwidth]{livelock.png}
	\end{columns}
\end{frame}
% slide: other_prob: decide on mutex_lock as opposed to everywhere, but...
% slide: manual shaping - have *user* specify what to focus on (automating intuition)
\begin{frame}{Focusing the Search Space}
	Recommended decision points: \texttt{mutex\_lock()} and \texttt{mutex\_unlock()}
	\linegap

	{\bf Problem:} Kernel uses mutexes {\em everywhere}.
	\begin{itemize}
		\item Good: \texttt{exit()} calls \texttt{mutex\_lock()}
		\item Bad: \texttt{exit()} calls \texttt{destroy\_address\_space()} calls \texttt{mutex\_lock()}
	\end{itemize}
	\linegap
	{\bf Solution:} User specifies what modules of the kernel to pay attention to.
	\begin{itemize}
		\item \texttt{within\_function exit}
		\item \texttt{without\_function destroy\_address\_space}
	\end{itemize}
\end{frame}
% slide: exciting teaser "i think this is not too far off!"
\begin{frame}{Teaser: (Semi-)Automated Search Focusing}
	Searching for bugs given a stack trace
	\begin{itemize}
		\item User knows {\em where} the bug is, but not {\em what}.
		\item Testing framework can use stack trace to conduct a more focused search.
	\end{itemize}
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Evaluation}

\breakslide{Evaluating Landslide}

%\begin{frame}{Outline}
%	\done{Motivation: Concurrency debugging}
%	\begin{itemize}
%		\item \done{Systematic exploration versus stress testing}
%		\item \done{Challenges of kernel-space}
%	\end{itemize}
%	\linegap
%
%	\done{Tool: Landslide}
%	\begin{itemize}
%		\item \done{Design and interface}
%		\item \done{Addressing challenges}
%	\end{itemize}
%	\linegap
%
%	{\bf Evaluation: Finding Races}
%	\begin{itemize}
%		\item {\bf Student user study}
%		\item {\bf Case study}
%	\end{itemize}
%	\linegap
%
%	Future Work and Conclusion
%\end{frame}

\subsection{User Study}

\begin{frame}{Working with Students} % slide: worked with students - give some time breakdown
	Solicited students in 15-410 to use Landslide on their kernels.
	\begin{itemize}
		\item Can Landslide find bugs ``in the wild''?
		\item How much time does it take to get it working?
	\end{itemize}
	\linegap

	Five groups participated; four got it to work.
	\begin{itemize}
		\item Average instrumentation time 100 minutes (55 to 158)
		%\item Average customisation time 25 minutes (0 to 60)
	\end{itemize}
\end{frame}

\begin{frame}{Landslide Victories} % slide: anecdotes: two deterministic bugs & two races
	All four groups who got it to work found bugs.
	\linegap

	Race conditions
	\begin{itemize}
		\item Too many waiters allowed
		\begin{itemize}
			\item One parent thread might block forever
		\end{itemize}
		\item Accidental return to an exited thread
		% Also talk about their vanish/vanish bug
	\end{itemize}
	\linegap

	Deterministic bugs
	\begin{itemize}
		\item Inappropriate idling
		\begin{itemize}
			\item Waiting for a timer tick with no useful work to do
		\end{itemize}
		\item Use-after-free
		\begin{itemize}
			\item Required memory tracking to detect
		\end{itemize}
	\end{itemize}
\end{frame}

\subsection{Case Study}

\begin{frame}{Studying Bugs In-Depth} % slide: "I have built a software smarter than me!" 30-40sec to find
	Worked with two kernels: my own, and one I graded
	\linegap

	``I have made something smarter than me.''
	% Heh, if it were really smarter than me, I wouldn't have had to debug it.
	\begin{itemize}
		\item \texttt{double\_wait} - two parent threads \texttt{wait()} on one exiting child
		\begin{itemize}
			\item Expected: One thread reaps child, the other returns failure
			\item Got: \texttt{failed assertion `nobe->cond\_queue\_next == NULL'}
		\end{itemize}
		\item Previously unknown, not found by stress testing or by TA
	\end{itemize}
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusion}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%\begin{frame}{Outline}
%	\done{Motivation: Concurrency debugging}
%	\begin{itemize}
%		\item \done{Systematic exploration versus stress testing}
%		\item \done{Challenges of kernel-space}
%	\end{itemize}
%	\linegap
%
%	\done{Tool: Landslide}
%	\begin{itemize}
%		\item \done{Design and interface}
%		\item \done{Addressing challenges}
%	\end{itemize}
%	\linegap
%
%	\done{Evaluation: Finding Races}
%	\begin{itemize}
%		\item \done{Student user study}
%		\item \done{Case study}
%	\end{itemize}
%	\linegap
%
%	{\bf Future Work and Conclusion}
%\end{frame}

%\subsection{Future Work}
\begin{frame}{Promising Directions}
	Top 15-410 students, effectively supervised, can find hard bugs faster.
	\begin{itemize}
		\item Use as a debugging tool may help students learn more.
	\end{itemize}
	\linegap

	After Pebbles\dots
	\begin{itemize}
		\item Potential application to production kernels
		\begin{itemize}
			\item General purpose OSes, device drivers
			% Be ready to field questions about SMP.
		\end{itemize}
	\end{itemize}
	\linegap

	Emphasis on users ``steering'' test by changing parameters
	\begin{itemize}
		\item Start with coarse granularity, iterate refining decision points
		\item Perhaps we can automate this steering?
	\end{itemize}
	\linegap
\end{frame}

%\subsection{Summary}
\begin{frame}{Summary}
	\textbf{Systematic testing in kernel-space}
	\begin{itemize}
		\item Use internal kernel abstractions to understand concurrency behaviour.
		\item Relying on user's knowledge makes testing easier.
	\end{itemize}
	\linegap
	\textbf{Landslide}
	\begin{itemize}
		\item Has potential to be a useful tool for 15-410.
		\item A first step towards sophisticated kernel debugging techniques.
	\end{itemize}

	\linegap
	\linegap
	\begin{center}
	{\scriptsize \em Please hold on. This transit is approaching the Landslide terminal, \\baggage claim, ground transportation, and ticketing.}
	\end{center}
\end{frame}

% slide: bonus slides

\end{document}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Introduction}
\subsection{Introduction}

\subsection{Race Conditions}

% Say: Since I'm no longer a member of course staff, I'm allowed to give away a
% race condition that some of you might have, for the sake of example.

\begin{frame}{Decision Points (``good'' case)}
	\begin{tabular}{|l|l|l}
		\cline{1-2}
		{\bf Thread 1} & {\bf Thread 2} & \\
		\cline{1-2}
		\texttt{spawn\_new\_thread} && \\
		\cline{1-2}
		\texttt{add\_to\_runqueue} && (new thread) \\
		\cline{1-2}
		\texttt{return child->tid} &&  \\
		\cline{1-2}
		& \texttt{vanish} & \\
		\cline{1-2}
		& (TCB gets freed) & (voluntary reschedule) \\
		\cline{1-2}
	\end{tabular}
\end{frame}

\begin{frame}{Decision Points (race condition)}
	\begin{tabular}{|l|l|l}
		\cline{1-2}
		{\bf Thread 1} & {\bf Thread 2} & \\
		\cline{1-2}
		\texttt{spawn\_new\_thread} && \\
		\cline{1-2}
		\texttt{add\_to\_runqueue} && (new thread + preempted) \\
		\cline{1-2}
		& \texttt{vanish} & \\
		\cline{1-2}
		& (TCB gets freed) & (voluntary reschedule) \\
		\cline{1-2}
		\texttt{return child->tid} && (bad!) \\
		\cline{1-2}
	\end{tabular}
\end{frame}

\breakslide{\Large A different way of looking at race conditions\ldots}

\begin{frame}{Execution Tree}
	\begin{center}
		%\includegraphics[width=0.9\textwidth]{threadfork0.png}
	\end{center}
\end{frame}
\begin{frame}{Execution Tree}
	\begin{center}
		%\includegraphics[width=0.9\textwidth]{threadfork05.png}
	\end{center}
\end{frame}
\begin{frame}{Execution Tree}
	\begin{center}
		%\includegraphics[width=0.9\textwidth]{threadfork1.png}
	\end{center}
\end{frame}
\begin{frame}{Execution Tree}
	\begin{center}
		%\includegraphics[width=0.9\textwidth]{threadfork2.png}
	\end{center}
\end{frame}

\begin{frame}{Decision Points} % : duplicate slide with later
	A {\bf decision point} is\ldots

	\linegap
	A code location where being preempted causes different behaviour.
	\begin{itemize}
		\item Intuitively: Somewhere that interesting interleavings can happen around.
	\end{itemize}
	Examples:
	\begin{itemize}
		\item A new thread becomes runnable
		\item Voluntary reschedule (e.g. \texttt{yield}, \texttt{cond\_wait})
		\item Synchronization primitives
	\end{itemize}
	% Say: We'll revisit this later.
\end{frame}

% Say: Obviously we want to be able to always run the third path. But knowing
% that ahead of time is impossible.
% Say: Okay, now what if we could build a tool that could ``systematically''
% explore this tree, and find the branch with the bug in it? What would such a
% tool need to be able to do?

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Systematic Testing}

\breakslide{\Large Systematic Testing}

\begin{frame}{Systematic Testing}
	\textbf{Systematic testing} is:
	\begin{itemize}
		\item Systematically enumerating different interleavings
		\begin{itemize}
			\item Intuitively: Generate many ``tabular execution traces''
		\end{itemize}
		\item Exploring all branches in these trees
		\begin{itemize}
			\item (by controlling scheduling decisions at decision points)
		\end{itemize}
		\item In practice: Depth-first search of branches
	\end{itemize}
\end{frame}

\subsection{Requirements}

\begin{frame}{Execution Tree Exploration}
	Important point: When does a branch end?
	\begin{itemize}
		\item All threads run to completion, or
		\item A bug is detected % Say: more on this later.
	\end{itemize}
	{\bf Backtracking}:
	\begin{itemize}
		\item Identify a decision point to choose differently
		\item Reset machine state and start over
		\item Replay test from the beginning, with different choices
	\end{itemize}
\end{frame}

\begin{frame}{More on Decision Points}
	Important point: What does ``all possible interleavings'' mean?

	\linegap
	One extreme: Decide at every instruction
	\begin{itemize}
		\item Good news: Will find every possible race condition.
		\item Bad news: Runtime of test will be impossibly large.
	\end{itemize}
	\linegap

	Other extreme: Nothing is a decision point
	\begin{itemize}
		\item Good news: Test will finish quickly.
		\item Bad news: Only one execution was checked for bugginess.
		\item Bad news: No alternative interleavings explored.
		\begin{itemize}
			\item Makes ``no race found'' a weak claim.
		\end{itemize}
	\end{itemize}
\end{frame}

\begin{frame}{More on Decision Points}
	Sweet spot: Insert a thread switch everywhere it ``might matter''.

	\linegap
	When do we fear being preempted?
	\begin{itemize}
		\item New threads becoming runnable (\texttt{fork}, \texttt{cond\_signal}, etc)
			\begin{itemize}
				\item Preemptions may cause it to run before we're ready
			\end{itemize}
		\item Synchronization primitives (\texttt{mutex\_lock}, etc)
			\begin{itemize}
				\item If used improperly\ldots
			\end{itemize}
		\item Unprotected shared memory accesses % Say: more on this later.
			\begin{itemize}
				\item May result in data structure corruption
			\end{itemize}
	\end{itemize}
	\linegap

	Finding the sweet spot is a joint effort between programmer and tool. (More on this later.)
\end{frame}

\subsection{Challenges}

\begin{frame}{Controlling Scheduling Decisions}
	Control over sources of nondeterminism
	\begin{itemize}
		\item Device interrupts/input
			\begin{itemize}
				\item Disk drivers: when disk reads finish
				\item Ethernet drivers: when packets arrive
			\end{itemize}
		\item To control thread switches in a 410 kernel, vary when clock ticks happen.
	\end{itemize}
\end{frame}

\begin{frame}{Memory Interposition}
	In order to find use-after-free, need to know:
	\begin{itemize}
		\item When objects are \texttt{free()}d
		\item When threads access shared memory in the heap
	\end{itemize}
	\linegap

	Solution: Keep track of all memory events
	\begin{itemize}
		\item All calls to \texttt{malloc}/\texttt{free}
		\item All shared memory reads/writes
	\end{itemize}
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{State Space Explosion}

\begin{frame}{State Space Explosion}
	State spaces grow exponentially
	\begin{itemize}
		\item With $d$ decision points, $k$ runnable threads, size $d^k$.
		\item Threatens our ability to explore everything.
		\item Fortunately, some sequences result in identical states.
	\end{itemize}
	\linegap

	{\bf Partial Order Reduction} can help.
	\begin{itemize}
		\item Complicated algorithm; ask me later for details.
		\item Intuitive explanation follows.
	\end{itemize}
\end{frame}

\begin{frame}{State Space Explosion}
	\begin{center}
	%\includegraphics[width=\textwidth]{undiamond0.png}
	\end{center}
\end{frame}
\begin{frame}{State Space Explosion}
	\begin{center}
	%\includegraphics[width=\textwidth]{undiamond1.png}
	\end{center}
\end{frame}
\begin{frame}{State Space Explosion}
	\begin{center}
	%\includegraphics[width=\textwidth]{diamond1.png}
	\end{center}
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Landslide}

\breakslide{\Huge Landslide}

\begin{frame}{About The Project}
	5th year MS since June 2011 % Say: planning to finish this may, hopefully

	\linegap
	Working with Garth Gibson, Jiri Simsa

	\linegap
	{\bf Landslide}: Shows that your Pebbles are not as stable as you thought.
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{How Landslide Works}

\begin{frame}{Landslide in Simics}
	As a Simics module, Landslide knows:
	\begin{itemize}
		\item Every instruction the kernel executes
		\item Every memory address the kernel reads/writes
	\end{itemize}
	\linegap
	Artificially causes timer interrupts

	\linegap
	Checkpointing/backtracking via Simics bookmarks
\end{frame}

\begin{frame}{Anatomy}
	\begin{center}
	%\includegraphics[width=0.6\textwidth]{landslide.png}
	\end{center}
\end{frame}

\begin{frame}{Identifying Bugs}
	Landslide can {\em definitely discover}:
	\begin{itemize}
		\item Kernel panics % Say: note: the better your asserts are..!
		\item Deadlock
		\item Use-after-free / double-free
	\end{itemize}
	Landslide can {\em reasonably suspect}:
	\begin{itemize}
		\item Memory leak
		\item Progress sense (halting problem)
	\end{itemize}
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{Using Landslide}

\breakslide{\Large Using Landslide}

\begin{frame}{In Which Ben Offers Help}
	% Say: So, this wasn't just a research talk
	This is something you can try!
	\linegap

	Mutual benefit
	\begin{itemize}
		\item Landslide may help you find bugs
		\item You may help Ben evaluate his thesis project
	\end{itemize}
\end{frame}

\begin{frame}{Keeping It Real}
	Finding race conditions is hard for humans.

	\linegap
	It is hard for computer programs too.

	\linegap
	Landslide is not an oracle.
\end{frame}

\begin{frame}{Annotating Your Kernel}
	Step 1
	\begin{center}
		%\includegraphics[width=0.9\textwidth]{tell_landslide.png}
	\end{center}
	Your kernel needs to say when certain events happen:
	\begin{itemize}
		\item When do threads become runnable / descheduled?
		\item When does the scheduler switch threads?
	\end{itemize}
	\linegap

	\framebox{Time estimate: 40 minutes}
\end{frame}

\begin{frame}{Configuring Landslide}
	Step 2
	\begin{center}
		%\includegraphics[width=0.8\textwidth]{config-landslide.png}
	\end{center}
	Edit \texttt{config.landslide} with some details and tweaks

	Fill in two implementation-dependent C functions in Landslide ($\le$10 lines)
	% Say: Which depend on your kernel's specific implementation
	\linegap

	\framebox{Time estimate: 60 minutes}
\end{frame}

\begin{frame}{Configuring Decision Points}
	Landslide automatically identifies a minimal set of decision points.
	\begin{itemize}
		\item It might find bugs.
		\item It might overlook more fine-grained interleavings.
	\end{itemize}
	With help from you, it could find more.
	\begin{itemize}
		\item Optional annotation: \texttt{tell\_landslide\_decide()}
		\item Hints to where a context switch should be forced.
		\item Inside every call to \texttt{mutex\_lock}\ldots
	\end{itemize}
\end{frame}

\breakslide{\Large Quick Demo}

%\begin{frame}{Decision Points (noticing a theme here?)}
%	Landslide is {\bf false negative}-oriented.
%	\begin{itemize}
%		\item If Landslide reports a bug, there is a bug for sure.
%			% Say: Well, except progress sense might be wrong.
%		\item If not,
%			\begin{itemize}
%				\item Maybe there was no race, or
%				\item Maybe your decision points were not granular enough.
%			\end{itemize}
%	\end{itemize}
%\end{frame}
%\begin{frame}{Decision Points (noticing a theme here?)}
%	Sometimes the default set will work.
%
%	Sometimes you need to tweak\ldots
%	\begin{itemize}
%		\item Ignore certain global mutexes
%		\item Whitelist or blacklist entire syscalls
%		\item (get creative)
%	\end{itemize}
%	Intuitively: focusing Landslide on relevant parts of the kernel.
%\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{User Study}

\begin{frame}{In Which Ben Offers Help - Warning}
	\textbf{If you are already struggling, this will not ``save'' you.}
	\begin{itemize}
		\item False-negatives: not guaranteed to find races at all
		\item Research-quality: possibly difficult to integrate with your kernel
		\item Finishing the kernel project is more important.
	\end{itemize}
	You should:
	\begin{itemize}
		\item Be expecting an A or B
		\item Be able to turn in without late days (if you really had to)
		\item Be looking for\ldots
			\begin{itemize}
				\item That ``one pesky race''
				\item A race that stress-testing missed
				\item Or just familiarity with a new technique
			\end{itemize}
	\end{itemize}
\end{frame}

\begin{frame}{In Which Ben Offers Help}
	Your kernel
	\begin{itemize}
		\item Must load the shell and run programs
		\item \texttt{fork},
			\texttt{exec},
			\texttt{vanish},
			\texttt{wait},
			\texttt{readline}
		\item Must never spin-wait (see hurdle form!)
		\item Should \texttt{assert()} important invariants
		\begin{itemize}
			\item Think of \texttt{panic()} as \texttt{tell\_landslide\_bug()}
		\end{itemize}
	\end{itemize}
\end{frame}

\begin{frame}{In Which Ben Offers Help}
	User study next week, starting Monday

	\linegap
	Expect to spend:
	\begin{itemize}
		\item Up to 4 hours, just to try it out.
		\item 6-8 hours, if you find a bug and track it down.
		\item More, for multiple bugs or the truly dedicated\ldots
	\end{itemize}
	\linegap

	Give feedback (intuitive? frustrating? found bugs?)

	\linegap
	{\bf Watch \texttt{410.announce} for details!}
\end{frame}

\subsection{End}
\breakslide{\Large Questions?}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% \section{Conclusion}
% \subsection{Conclusion}
% 
% \begin{frame}{Summary}
% 	\begin{itemize}
% 		\item
% 	\end{itemize}
% \end{frame}

%\begin{frame}{Partial Order Reduction}
%	\begin{center}
%	The next 6 slides won't be on the exam.
%
%		\Huge \dbend
%	\end{center}
%\end{frame}
%
%\begin{frame}{Partial Order Reduction}
%	A {\bf transition} is\ldots
%	\begin{itemize}
%		\item The sequence of instructions executed between choice points.
%	\end{itemize}
%	To express concurrency between transitions:
%	% Say: As a prerequisite for this algorithm, we need to build two sets to express the concurrency relations between the transitions
%	\begin{itemize}
%		\item Independence relation
%		\item Happens-before relation
%	\end{itemize}
%	% Say: This is hard, and I don't expect you to understand it during this lecture; you won't be tested on it or anything. My goal in presenting it is to convey how intensely cool the algorithm is, and hopefully you will keep thinking about it later. I'd be happy to answer questions about it later.
%\end{frame}
%
%\begin{frame}[fragile]{Partial Order Reduction - Independence Relation}
%	\begin{columns}[l]
%		\column{0.5\textwidth}
%		\begin{center}
%		\begin{verbatim}
%		void thread_A() {
%		    foo = 42;
%		}
%		void thread_B() {
%		    if (bar == 1337)
%		        print "Hello"
%		}
%		\end{verbatim}
%		\end{center}
%		\linegap
%
%		$A$ and $B$ are {\bf independent}.
%		\column{0.5\textwidth}
%		\begin{center}
%		\begin{verbatim}
%		void thread_A() {
%		    foo = 42;
%		}
%		void thread_C() {
%		    if (foo == 1337)
%		        print "Hello"
%		}
%		\end{verbatim}
%		\end{center}
%		\linegap
%
%		$A$ and $C$ {\bf conflict}.
%	\end{columns}
%\end{frame}
%
%\begin{frame}{Partial Order Reduction - Independence Relation}
%	Consider a trace of all shared memory reads/writes
%
%	\linegap
%	Two transitions (instruction sequences) $A$ and $B$ are independent iff:
%	\begin{itemize}
%		\item $Writes(A) \cap Writes(B) = \emptyset$
%		\item $Writes(A) \cap Reads(B) = \emptyset$
%		\item $Reads(A) \cap Writes(B) = \emptyset$
%	\end{itemize}
%\end{frame}
%
%\begin{frame}{Partial Order Reduction - Happens-before Relation}
%	\begin{columns}[l]
%		\column{0.5\textwidth}
%		%\includegraphics[width=\textwidth]{hb.png}
%		\column{0.5\textwidth}
%		Is a transition ``enabled'' by a previous one?
%		\begin{itemize}
%			\item $A$ enables $B$
%			\item $B$ enables $C$ (same thread)
%			\item $B$ enables $D$
%		\end{itemize}
%%		Transitive closure
%%		\begin{itemize}
%%			\item $HB^* = \{AB, AC, AD, BC, BD\}$
%%		\end{itemize}
%		$C$ and $D$ are {\bf concurrent}!
%	\end{columns}
%\end{frame}
%
%\begin{frame}{Partial Order Reduction - Tagging}
%	Run POR at the end of each branch
%	\begin{itemize}
%		\item Identify which branches to explore next
%		\item Tag only a subset of all branches (hopefully)
%	\end{itemize}
%\end{frame}
%
%\begin{frame}{Partial Order Reduction - Tagging}
%	For each transition $T$, find an {\bf evil ancestor} $E$:
%	\begin{itemize}
%		\item $E$ and $T$ are concurrent.
%		\item $E$ and $T$ are not independent.
%		\item Intuitively: $T$ and $E$ could be reordered, and might behave differently.
%	\end{itemize}
%	Find a {\bf good sibling} $G$ of $E$:
%	\begin{itemize}
%		\item $G$ is the same thread as $T$
%		% Say: (if no such sibling, all siblings)
%		\item Intuitively: Find a branch in which $T$ runs before $E$.
%	\end{itemize}
%	Explore tagged branches depth-first.
%\end{frame}
%
%\begin{frame}{Partial Order Reduction}
%	\begin{center}
%		{\Huge \dbend}
%		\linegap
%		\linegap
%		\linegap
%
%		Feel free to ask me about this later.
%	\end{center}
%\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%\subsection{Other Dynamic Techniques}
%
%\begin{frame}{Heuristic-Based Searching}
%	% Say: Note that this is beyond the scope of my personal project.
%	Maybe the tree is too big to search anyway?
%
%	\linegap
%	Search {\em most likely} branches first.
%	\begin{itemize}
%		\item ``Iterative Context Bounding''
%		\footnote{Microsoft Research - CHESS - Musuvathi et al.}
%		\begin{itemize}
%			\item Insight: Races tend to show up with few forced context switches.
%			\item Search branches with fewer preemptions first.
%		\end{itemize}
%		\item Find behaviours that are typical of buggy software.
%		\footnote{Stanford University - Dawson Engler}
%		\begin{itemize}
%			\item Different behaviour in different branches.
%				% Say: different all the time: no. different rarely: yes.
%			\item Use suspicious memory accesses as future choice points.
%		\end{itemize}
%	\end{itemize}
%	\linegap
%	Open research field, possibly my future work.
%\end{frame}
%
%% Say: I'm going to take a break from talking about systematic exploration to talk about other types of dynamic race detection - FIXME: if these slides fit in 60min.
%\begin{frame}{Data Race Detection}
%	Requires happens-before set and ``lock set''
%	\linegap
%
%	Two memory accesses $m_1$ and $m_2$ are a data race iff:
%	\begin{itemize}
%		\item Read/write or write/write to the same address, and
%		\item $(m_1,m_2) \not\in HB$, and
%		\item $Lock(m_1) \cap Lock(m_2) = \emptyset$
%	\end{itemize}
%\end{frame}
%\begin{frame}{Data Race Detection}
%	Data race detection vs systematic exploration
%	\begin{itemize}
%		\item More false positives
%			\begin{itemize}
%				\item (but still good style-checking!)
%			\end{itemize}
%		\item No false negatives
%		\item Still unable to find some types of bugs
%	\end{itemize}
%	\linegap
%	Hybrid approach? (future research)
%\end{frame}

% vim: foldmethod=indent

### 2012-11-01 landslide-lecture/lecture.tex
add landslide lecture and defence

--- 2012-11-01 landslide-defence/defence.tex
@@
 
-\title[Landslide]{{\bf Landslide: Systematic Dynamic Race Detection in Kernel-space}}
-\subtitle[]{ {\em showing that Pebbles are less stable than you thought since 2011.}}
+\title[Landslide]{{\bf Systematic Dynamic Race Condition Detection}}
+\subtitle[]{ {\em more clever than } ``\texttt{slaughter cho2}'' {\em since 2011.}}
 \author[Ben Blum]{Ben Blum \texttt{(bblum@andrew.cmu.edu)}}
 
-\institute[CMU]{Carnegie Mellon University}
-\date[]{2012, May 10}
+\institute[CMU 15-410]{Carnegie Mellon University - 15-410}
+\date[]{2012, April 6}
 
@@
 
-\usecolortheme[named=RoyalBlue]{structure}
+\usecolortheme[named=ForestGreen]{structure}
 
@@
 \newcommand\breakslide[1]{\begin{frame}{} \begin{center} #1 \end{center} \end{frame}}
-\newcommand\related[1]{\textsuperscript{\em [#1]}}
+
+\section{Introduction}
+\subsection{Introduction}
 
 \begin{frame}{Outline}
-	\textbf{Motivation: Concurrency debugging}
+	\textbf{Theory: Seeing race conditions in a new way}
 	\begin{itemize}
-		\item Systematic testing versus stress testing
-		\item Challenges of kernel-space
+		\item Case study (example)
+		\item The execution tree
+		\item Decision points
 	\end{itemize}
-	\linegap
-
+	{\bf Technique: Systematic testing}
+	% Say: A way of enumerating all possible interleavings of a concurrent test case.
+	\begin{itemize}
+		\item Requirements
+		\item Challenges and feasibility
+	\end{itemize}
 	{\bf Tool: Landslide}
 	\begin{itemize}
-		\item Design and interface
-		\item Addressing challenges
+		\item How it works
+		\item The user-tool relationship
+		\item User study (that's you!)
+			% !!! SAY !!! : If you are struggling, this is not for you, etc.
 	\end{itemize}
-	\linegap
-
-	{\bf Evaluation: Finding Races}
-	\begin{itemize}
-		\item Student user study
-		\item Case study
-	\end{itemize}
-	\linegap
-
-	{\bf Future Work and Conclusion}
 \end{frame}
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Motivation}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Race Conditions}
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Race Conditions and the Finding Thereof}
+% Say: Since I'm no longer a member of course staff, I'm allowed to give away a
+% race condition that some of you might have, for the sake of example.
 
@@
 
-	\begin{itemize}
-		\item On exit, child's state is freed
-		\item Forking thread does use-after-free
-		\item Might return garbage instead of thread ID
-	\end{itemize}
+%	\begin{itemize}
+%		\item On vanish, child's TCB is freed
+%		\item Forking thread does use-after-free
+%		\item Might return garbage instead of TID
+%	\end{itemize}
 \end{frame}
-
-\begin{frame}{Testing Techniques} % slide: sys ex vs stress testing
-	\textbf{Stress Testing:} Common testing approach
-	\begin{itemize}
-		\item Attempting to exercise as many interleavings as possible
-		\item Exposes race conditions at random
-		\begin{itemize}
-			\item ``If a preemption occurs at {\em just the right time}\dots''
-		\end{itemize}
-		\item Cryptic panic messages or machine reboots
-	\end{itemize}
-	\linegap
-
-	{\bf Systematic Testing} \related{Musuvathi '08}
-	\begin{itemize}
-		\item Make educated guesses about when to preempt
-		\item Run {\em every single} interleaving
-		\item Provide better debugging information, reproducibility
-	\end{itemize}
-\end{frame}
-
-\begin{frame}{Execution Tree} % slide: decision tree
-	\begin{center}
-		\includegraphics[width=\textwidth]{tree.png}
-	\end{center}
-\end{frame}
-
-\begin{frame}{Finding Races in Kernel-space} % slide: 15-410 6-week project why we need it in kernels
-	\textbf{15-410:} Undergraduate Operating Systems
-	\begin{itemize}
-		\item Project 3: students write a kernel in 6 weeks.
-		\item ``Pebbles'' is a UNIX-like, preemptible kernel specification.
-	\end{itemize}
-	\linegap
-
-	Previous work in kernel-space has not used systematic testing.
-	%\begin{itemize}
-	%	\item Data race detection\related{Erickson '10}
-	%	\item Failure resiliency\related{Kadav '09}
-	%\end{itemize}
-\end{frame}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Challenges of Kernel-space}
-
-\breakslide{Why is this environment different from all other environments?}
-
-\begin{frame}{Causes of Concurrency} % slide: causes of concurrency, thread scheduling
-	\textbf{Kernels contain their own concurrency implementation.}
-
-	\linegap
-	Context switching
-	\begin{itemize}
-		\item Non-deterministic timer-driven thread scheduling
-	\end{itemize}
-	Runqueue tracking
-	\begin{itemize}
-		\item Which threads are runnable?
-	\end{itemize}
-	Thread lifecycle tracking
-	\begin{itemize}
-		\item When are threads created/destroyed?
-	\end{itemize}
-\end{frame}
-
-% slide: ad-hoc communication, memory conflicts (freeing, runqueue)
-%\begin{frame}{Ad-Hoc Thread Communication}
-%	No message-passing % TODO this slide sucks
-%	
-%	User-space systematic exploration relies on it\related{Simsa '10}; we cannot.
-%	
-%	Kernel threads share state via globals, heap.
-%\end{frame}
-
-\begin{frame}{The Kernel as One Program} % slide: the kernel is too big to test all at once
-	% say: it's also important to understand the kernel's internals instead of just viewing it as one ``black box''
-	\begin{center}
-	\includegraphics[width=\textwidth]{pebbles.png}
-	\end{center}
-	% now that we see these internal interfaces, how do we apply systematic testing to it?
-\end{frame}
-\begin{frame}{The Kernel as One Program} % slide: the kernel is too big to test all at once
-	``Everything interleaves with everything else''?
-	\begin{itemize}
-		% Say: "we don't want to have to explore every possible interaction between each of these components"
-		\item Impractical to explore resulting state spaces
-	\end{itemize}
-	\linegap
-
-	What is a ``bug'', anyway?\related{Engler '01}
-	\begin{itemize}
-		\item False-negative-oriented approach
-	\end{itemize}
-\end{frame}
-
-\begin{frame}{Contribution} % slide: summary: "sys ex is a tool \ldots``
-	Systematic testing is a {\em tool} for testing concurrent systems.
-	\begin{itemize}
-		\item How can it be applied in kernel-space?
-		\item What simple things can we ask of the user to make it possible?
-	\end{itemize}
-	\linegap
-
-	{\bf Landslide} is an effort to answer these questions.
-	% say: landslide is an exploratory work into finding out what lies behind these questions.
-\end{frame}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Landslide}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\newcommand\done[1]{\textcolor{gray}{\em \small #1}}
-
-%\begin{frame}{Outline}
-%	\done{Motivation: Concurrency debugging}
-%	\begin{itemize}
-%		\item \done{Systematic exploration versus stress testing}
-%		\item \done{Challenges of kernel-space}
-%	\end{itemize}
-%	\linegap
-%
-%	{\bf Tool: Landslide}
-%	\begin{itemize}
-%		\item {\bf Design and interface}
-%		\item {\bf Addressing challenges}
-%	\end{itemize}
-%	\linegap
-%
-%	Evaluation: Finding Races
-%	\begin{itemize}
-%		\item Student user study
-%		\item Case study
-%	\end{itemize}
-%	\linegap
-%
-%	Future Work and Conclusion
-%\end{frame}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Design}
-
-\begin{frame}{Environment}
-	\textbf{Simics:} a full-system x86 simulator
-	\begin{itemize}
-		\item Students in 15-410 use Simics to test their kernels.
-		\item Landslide is a transparently-embedded Simics module.
-	\end{itemize}
-\end{frame}
-
-\begin{frame}{Anatomy of Landslide}
-	\begin{center}
-	\includegraphics[width=0.7\textwidth]{landslide.png}
-	\end{center}
-	% Remember to talk about solving the thread scheduling challenge
-\end{frame}
-
-\subsection{Interface}
-
-\begin{frame}{Using Landslide}
-	\begin{center}
-	\includegraphics[width=0.9\textwidth]{tell_landslide.png}
-	\end{center}
-	User must tell Landslide when certain kernel events happen:
-	\begin{itemize}
-		\item When do threads become runnable / descheduled?
-		\item When does the scheduler switch threads?
-	\end{itemize}
-	\linegap
-	% Make sure to expound on what these are
-	Some complicated conditions require in-Landslide instrumentation.
-	\begin{itemize}
-		\item Scheduler design details
-	\end{itemize}
-\end{frame}
-
-\begin{frame}{Decision Trace}
-	% FIXME make the stack trace coloured
-	\includegraphics[width=\textwidth]{found_a_bug.png}
-	% Say: So landslide had to explore 5 different interleavings to find the bug. But this was a pretty simple test case, and 
-\end{frame}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Addressing Challenges} % rename
-
-\begin{frame}{State Space Reduction}
-	% Say: ``combinatorial explosion''
-	\begin{center}
-	\includegraphics[width=\textwidth]{undiamond1.png}
-	\end{center}
-\end{frame}
-\begin{frame}{State Space Reduction}
-	State spaces grow exponentially.
-	\begin{itemize}
-		\item Fortunately, some sequences result in identical states.
-		\item Dynamic Partial Order Reduction\related{Flanagan '05}
-		\begin{itemize}
-			\item Requires ``memory independence relation'' between transitions.
-		\end{itemize}
-	\end{itemize}
-\end{frame}
-
-\begin{frame}{Shared Memory Conflicts} % slide: ignoring shared memory accesses - venn diagram
-	\begin{center}
-	\includegraphics[width=\textwidth]{shm-conflict0.png}
-	\end{center}
-\end{frame}
-\begin{frame}{Shared Memory Conflicts}
-	\begin{center}
-	\includegraphics[width=\textwidth]{shm-conflict1.png}
-	\end{center}
-\end{frame}
-\begin{frame}{Shared Memory Conflicts}
-	\begin{center}
-	\includegraphics[width=\textwidth]{shm-conflict2.png}
-	\end{center}
-\end{frame}
-\begin{frame}{Shared Memory Conflicts}
-	%\textbf{Problem:} Many conflicts we don't care about
-	%\begin{itemize}
-	%	\item Every transition accesses scheduler state.
-	%	\item Many transitions access ``unrelated'' global objects.
-	%\end{itemize}
-	%\linegap
-	\textbf{Solution:} Only consider ``relevant'' memory conflicts.
-	\begin{itemize}
-		\item Ignore scheduler, global objects assumed to be correct
-		\begin{itemize}
-			\item Sacrifice ability to test these to efficiently test everything else.
-		\end{itemize}
-		\item User must specify what to ignore.
-	\end{itemize}
-\end{frame}
-
-\begin{frame}[fragile]{Identifying Bugs}
-	\begin{columns}[l]
-		\column{0.75\textwidth}
-		\textbf{How do we know we've found a bug?}
-		\linegap
-
-		Definite bugs
-		\begin{itemize}
-			\item Kernel panics / assertion failures
-			\item Use-after-free
-			\item Deadlock
-		\end{itemize}
-		Probable bugs
-		\begin{itemize}
-			\item Infinite loop
-			\begin{itemize}
-				\item Use structure of execution tree to judge progress.
-			\end{itemize}
-		\end{itemize}
-		\column{0.25\textwidth}
-		\includegraphics[width=\textwidth]{blank-livelock.png}
-	\end{columns}
-\end{frame}
-\begin{frame}[fragile]{Identifying Bugs}
-	\begin{columns}[l]
-		\column{0.75\textwidth}
-		\textbf{How do we know we've found a bug?}
-		\linegap
-
-		Definite bugs
-		\begin{itemize}
-			\item Kernel panics / assertion failures
-			\item Use-after-free
-			\item Deadlock
-		\end{itemize}
-		Probable bugs
-		\begin{itemize}
-			\item Infinite loop
-			\begin{itemize}
-				\item Use structure of execution tree to judge progress.
-			\end{itemize}
-		\end{itemize}
-		\column{0.25\textwidth}
-		\includegraphics[width=\textwidth]{livelock.png}
-	\end{columns}
-\end{frame}
-% slide: other_prob: decide on mutex_lock as opposed to everywhere, but...
-% slide: manual shaping - have *user* specify what to focus on (automating intuition)
-\begin{frame}{Focusing the Search Space}
-	Recommended decision points: \texttt{mutex\_lock()} and \texttt{mutex\_unlock()}
-	\linegap
-
-	{\bf Problem:} Kernel uses mutexes {\em everywhere}.
-	\begin{itemize}
-		\item Good: \texttt{exit()} calls \texttt{mutex\_lock()}
-		\item Bad: \texttt{exit()} calls \texttt{destroy\_address\_space()} calls \texttt{mutex\_lock()}
-	\end{itemize}
-	\linegap
-	{\bf Solution:} User specifies what modules of the kernel to pay attention to.
-	\begin{itemize}
-		\item \texttt{within\_function exit}
-		\item \texttt{without\_function destroy\_address\_space}
-	\end{itemize}
-\end{frame}
-% slide: exciting teaser "i think this is not too far off!"
-\begin{frame}{Teaser: (Semi-)Automated Search Focusing}
-	Searching for bugs given a stack trace
-	\begin{itemize}
-		\item User knows {\em where} the bug is, but not {\em what}.
-		\item Testing framework can use stack trace to conduct a more focused search.
-	\end{itemize}
-\end{frame}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Evaluation}
-
-\breakslide{Evaluating Landslide}
-
-%\begin{frame}{Outline}
-%	\done{Motivation: Concurrency debugging}
-%	\begin{itemize}
-%		\item \done{Systematic exploration versus stress testing}
-%		\item \done{Challenges of kernel-space}
-%	\end{itemize}
-%	\linegap
-%
-%	\done{Tool: Landslide}
-%	\begin{itemize}
-%		\item \done{Design and interface}
-%		\item \done{Addressing challenges}
-%	\end{itemize}
-%	\linegap
-%
-%	{\bf Evaluation: Finding Races}
-%	\begin{itemize}
-%		\item {\bf Student user study}
-%		\item {\bf Case study}
-%	\end{itemize}
-%	\linegap
-%
-%	Future Work and Conclusion
-%\end{frame}
-
-\subsection{User Study}
-
-\begin{frame}{Working with Students} % slide: worked with students - give some time breakdown
-	Solicited students in 15-410 to use Landslide on their kernels.
-	\begin{itemize}
-		\item Can Landslide find bugs ``in the wild''?
-		\item How much time does it take to get it working?
-	\end{itemize}
-	\linegap
-
-	Five groups participated; four got it to work.
-	\begin{itemize}
-		\item Average instrumentation time 100 minutes (55 to 158)
-		%\item Average customisation time 25 minutes (0 to 60)
-	\end{itemize}
-\end{frame}
-
-\begin{frame}{Landslide Victories} % slide: anecdotes: two deterministic bugs & two races
-	All four groups who got it to work found bugs.
-	\linegap
-
-	Race conditions
-	\begin{itemize}
-		\item Too many waiters allowed
-		\begin{itemize}
-			\item One parent thread might block forever
-		\end{itemize}
-		\item Accidental return to an exited thread
-		% Also talk about their vanish/vanish bug
-	\end{itemize}
-	\linegap
-
-	Deterministic bugs
-	\begin{itemize}
-		\item Inappropriate idling
-		\begin{itemize}
-			\item Waiting for a timer tick with no useful work to do
-		\end{itemize}
-		\item Use-after-free
-		\begin{itemize}
-			\item Required memory tracking to detect
-		\end{itemize}
-	\end{itemize}
-\end{frame}
-
-\subsection{Case Study}
-
-\begin{frame}{Studying Bugs In-Depth} % slide: "I have built a software smarter than me!" 30-40sec to find
-	Worked with two kernels: my own, and one I graded
-	\linegap
-
-	``I have made something smarter than me.''
-	% Heh, if it were really smarter than me, I wouldn't have had to debug it.
-	\begin{itemize}
-		\item \texttt{double\_wait} - two parent threads \texttt{wait()} on one exiting child
-		\begin{itemize}
-			\item Expected: One thread reaps child, the other returns failure
-			\item Got: \texttt{failed assertion `nobe->cond\_queue\_next == NULL'}
-		\end{itemize}
-		\item Previously unknown, not found by stress testing or by TA
-	\end{itemize}
-\end{frame}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Conclusion}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%\begin{frame}{Outline}
-%	\done{Motivation: Concurrency debugging}
-%	\begin{itemize}
-%		\item \done{Systematic exploration versus stress testing}
-%		\item \done{Challenges of kernel-space}
-%	\end{itemize}
-%	\linegap
-%
-%	\done{Tool: Landslide}
-%	\begin{itemize}
-%		\item \done{Design and interface}
-%		\item \done{Addressing challenges}
-%	\end{itemize}
-%	\linegap
-%
-%	\done{Evaluation: Finding Races}
-%	\begin{itemize}
-%		\item \done{Student user study}
-%		\item \done{Case study}
-%	\end{itemize}
-%	\linegap
-%
-%	{\bf Future Work and Conclusion}
-%\end{frame}
-
-%\subsection{Future Work}
-\begin{frame}{Promising Directions}
-	Top 15-410 students, effectively supervised, can find hard bugs faster.
-	\begin{itemize}
-		\item Use as a debugging tool may help students learn more.
-	\end{itemize}
-	\linegap
-
-	After Pebbles\dots
-	\begin{itemize}
-		\item Potential application to production kernels
-		\begin{itemize}
-			\item General purpose OSes, device drivers
-			% Be ready to field questions about SMP.
-		\end{itemize}
-	\end{itemize}
-	\linegap
-
-	Emphasis on users ``steering'' test by changing parameters
-	\begin{itemize}
-		\item Start with coarse granularity, iterate refining decision points
-		\item Perhaps we can automate this steering?
-	\end{itemize}
-	\linegap
-\end{frame}
-
-%\subsection{Summary}
-\begin{frame}{Summary}
-	\textbf{Systematic testing in kernel-space}
-	\begin{itemize}
-		\item Use internal kernel abstractions to understand concurrency behaviour.
-		\item Relying on user's knowledge makes testing easier.
-	\end{itemize}
-	\linegap
-	\textbf{Landslide}
-	\begin{itemize}
-		\item Has potential to be a useful tool for 15-410.
-		\item A first step towards sophisticated kernel debugging techniques.
-	\end{itemize}
-
-	\linegap
-	\linegap
-	\begin{center}
-	{\scriptsize \em Please hold on. This transit is approaching the Landslide terminal, \\baggage claim, ground transportation, and ticketing.}
-	\end{center}
-\end{frame}
-
-% slide: bonus slides
-
-\end{document}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{Introduction}
-\subsection{Introduction}
-
-\subsection{Race Conditions}
-
-% Say: Since I'm no longer a member of course staff, I'm allowed to give away a
-% race condition that some of you might have, for the sake of example.
 
@@
 
+\begin{frame}{Testing Mechanisms}
+	\textbf{Stress testing}: \texttt{slaughter cho2} and friends
+	\begin{itemize}
+		\item Attempting to exercise as many interleavings as practical
+		\item Exposes race conditions at random
+		\begin{itemize}
+			\item ``If a preemption occurs at just the right time\ldots''
+		\end{itemize}
+		\item Cryptic panic messages or machine reboots
+	\end{itemize}
+	What if\ldots
+	\begin{itemize}
+		\item Make educated guesses about when to preempt
+		\item Preempt enough times to run {\em every single} interleaving
+		\item Tell the story of what {\em actually happened}.
+		\item Overlook fewer bugs!
+	\end{itemize}
+	% Say: At the end of this lecture, I'm going to be looking for some among you...
+\end{frame}
+
 \breakslide{\Large A different way of looking at race conditions\ldots}
@@
 	\begin{center}
-		%\includegraphics[width=0.9\textwidth]{threadfork0.png}
+		\includegraphics[width=0.9\textwidth]{threadfork0.png}
 	\end{center}
@@
 	\begin{center}
-		%\includegraphics[width=0.9\textwidth]{threadfork05.png}
+		\includegraphics[width=0.9\textwidth]{threadfork05.png}
 	\end{center}
@@
 	\begin{center}
-		%\includegraphics[width=0.9\textwidth]{threadfork1.png}
+		\includegraphics[width=0.9\textwidth]{threadfork1.png}
 	\end{center}
@@
 	\begin{center}
-		%\includegraphics[width=0.9\textwidth]{threadfork2.png}
+		\includegraphics[width=0.9\textwidth]{threadfork2.png}
 	\end{center}
@@
 
-\begin{frame}{Decision Points} % : duplicate slide with later
+\begin{frame}{Decision Points} % TODO: duplicate slide with later
 	A {\bf decision point} is\ldots
@@
 	\begin{center}
-	%\includegraphics[width=\textwidth]{undiamond0.png}
+	\includegraphics[width=\textwidth]{undiamond0.png}
 	\end{center}
@@
 	\begin{center}
-	%\includegraphics[width=\textwidth]{undiamond1.png}
+	\includegraphics[width=\textwidth]{undiamond1.png}
 	\end{center}
@@
 	\begin{center}
-	%\includegraphics[width=\textwidth]{diamond1.png}
+	\includegraphics[width=\textwidth]{diamond1.png}
 	\end{center}
@@
 	\begin{center}
-	%\includegraphics[width=0.6\textwidth]{landslide.png}
+	\includegraphics[width=0.6\textwidth]{landslide.png}
 	\end{center}
@@
 	\begin{center}
-		%\includegraphics[width=0.9\textwidth]{tell_landslide.png}
+		\includegraphics[width=0.9\textwidth]{tell_landslide.png}
 	\end{center}
@@
 	\begin{center}
-		%\includegraphics[width=0.8\textwidth]{config-landslide.png}
+		\includegraphics[width=0.8\textwidth]{config-landslide.png}
 	\end{center}
@@
 \breakslide{\Large Quick Demo}
+
+% TODO more of these with screenshots
 
@@
 
+\end{document}
 %\begin{frame}{Partial Order Reduction}
@@
 %		\column{0.5\textwidth}
-%		%\includegraphics[width=\textwidth]{hb.png}
+%		\includegraphics[width=\textwidth]{hb.png}
 %		\column{0.5\textwidth}
@@
 %\end{frame}
+
+% TODO: run through an example
 
@@
 %\end{frame}
-
-% vim: foldmethod=indent

### 2012-11-06 gpi-debug/debugging.tex
write some debugging slides

@@
 \title[Demystifying Debugging]{{\bf Demystifying Debugging}}
-\subtitle[]{{\em because you can't get it right the first time every time}}
+\subtitle[]{{\em when getting it right the first time every time fails}}
 \author[Ben Blum]{Ben Blum \texttt{(bblum@andrew.cmu.edu)}}
@@
 	\textbf{Why teach debugging?}
+		% brief motivation
+		% toolbox
 	\linegap
@@
 		\item The good debugger's attitude
+			% "get more operators"
 		\item Binary Search
@@
 		\item Time Travel
+			% small test cases / minimize the test case
 		\item Space Travel
+			% printf debugging
+			% invariant checkers and asserts
 	\end{itemize}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Motivation}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{frame}{``The Universal Backup Plan''}
+	\textbf{15-112, 15-122, 15-150, ...}
+	\begin{itemize}
+		\item What do these classes all have in common?
+		\pause
+		\item How to write code that {\em works}
+	\end{itemize}
+	\pause
+	{\bf 98-172}
+	\begin{itemize}
+		\item What to do with code that {\em doesn't}
+	\end{itemize}
+	% TODO: a picture goes here -- blank slate vs equipped/prepared technician
+\end{frame}
+
+\begin{frame}{Take-away}
+	\textbf{Debugging toolbox}
+	\begin{itemize}
+		\item Two major {\em strategies} for approaching bugs
+		\item Assorted {\em tactics} for applying to broken programs
+	\end{itemize}
+	% Say: "It's not, e.g., a checklist of bugs that you can just run down and test one-by-one; it's more abstract than that.
+You have to use your brain to figure out how to apply each tool.
+But the hope is that you won't feel lost staring at a blank slate of possible approaches.
+\end{frame}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Storytelling}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{frame}{The Essence of Debugging}
+	...is {\bf telling a good story.}
+	\pause
+	\linegap
+	Two stories, actually:
+	\begin{itemize}
+		\item What you wanted to happen
+		\item What the computer actually did
+	\end{itemize}
+	\pause
+	Your bug: The point where the stories diverge
+	\begin{itemize}
+		\item Can you describe that point in as much detail as possible?
+	\end{itemize}
+\end{frame}
+
+\begin{frame}{The Essence of Debugging}
+	Because you will write unique bugs, you will be telling a story nobody else knows.
+
+	\linegap
+	``Plot summaries'' are often not enough: My program crashes...
+	\pause
+	\begin{itemize}
+		\item with a segmentation fault...
+		\pause
+		\begin{itemize}
+			\item in \texttt{take\_final\_exam()}...
+			\pause
+			\begin{itemize}
+				\item because \texttt{study\_for\_final()} passed \texttt{NULL} to it.
+				\item There we go!
+			\end{itemize}
+		\end{itemize}
+	\end{itemize}
+\end{frame}
+
+\begin{frame}{Asking for Help}
+	This DOESN'T mean...
+	\begin{itemize}
+		\item That TAs can't help you without a detailed story
+		\item (That you shouldn't seek help until you've already found your bug?!)
+	\end{itemize}
+	\pause
+	This does mean...
+	\begin{itemize}
+		\item You need a detailed story to understand your bug.
+		\item If the TA doesn't know the story, how can they help?
+		\pause
+		\begin{itemize}
+			\item They know what you should do to make your story better.
+		\end{itemize}
+	\end{itemize}
+\end{frame}
+
+\begin{frame}{Running Example}
+	% TODO
+\end{frame}
 
@@
 
-\subsection{Travelling Through Time}
+\subsection{Time Travel}
 
-\breakslide{When things go wrong\dots}
+\breakslide{{\bf When} things go wrong\dots}
 
-\begin{frame}{foo}
+\begin{frame}{Telling a Story of Time}
+	Remember, two stories: Expected outcome vs Actual outcome
+	\linegap
+	What does this mean 
+	% Text: "First I ... then I ... but then it ...!"
 \end{frame}
 
+\begin{frame}{Print debugging}
+	% Insert some code: the test case with print statements inserted in it
+	% Insert a picture: the time-cloud with some printed-out things in it
+	Should usually be the first thing you try!
+\end{frame}
+
+\begin{frame}{Asserts: Stealthier, more concealable prints}
+\end{frame}
+
+\subsection{Space Travel}
+
+\breakslide{{\bf Where} things go wrong\dots}
+
+\begin{frame}{Telling a Story of Space}
+	Remember, two stories: Expected outcome vs Actual outcome
+	\linegap
+\end{frame}
+
+\end{document}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2012-11-06 gpi-debug/debugging.tex
more debug

@@
 \newcommand\hilight[2]{\color{#1}#2\color{black}}
+\definecolor{grey}{RGB}{127,127,127}
+\definecolor{darkcyan}{RGB}{0,127,127}
+\definecolor{olivegreen}{RGB}{0,127,0}
+\definecolor{violet}{RGB}{127,0,127}
+\definecolor{brickred}{RGB}{127,0,0}
+\definecolor{brown}{RGB}{127,63,0}
 
@@
 	\linegap
+
 	Two stories, actually:
@@
 \begin{frame}{Running Example}
-	% TODO
+	\textbf{Broken \texttt{insert\_sorted}}
+
+	\linegap
+	Sorts OK into the middle and front of list:
+	\begin{itemize}
+		\item \texttt{insert\_sorted} $\quad \hilight{olivegreen}{10} \quad  [20,30] \quad \Rightarrow \quad [\hilight{olivegreen}{10},20,30]$
+		\item \texttt{insert\_sorted} $\quad \hilight{olivegreen}{32} \quad  [16,64] \quad \Rightarrow \quad [16,\hilight{olivegreen}{32},64]$
+	\end{itemize}
+
+	\linegap
+	But not so great at the end:
+	\begin{itemize}
+	\item \texttt{insert\_sorted} $\quad \hilight{red}{100}  \quad [97,98,99] \quad \Rightarrow \quad [97,98,\hilight{red}{100},99]$
+	\end{itemize}
+\end{frame}
+
+\begin{frame}{Running Example}
+	\textbf{Broken \texttt{insert\_sorted}}
+
+	\linegap
+	In case you're curious, here's the bug:
+
+	\linegap
+		\texttt{~~~~insert\_sorted(list,~element)~\{} \\
+			\texttt{~~~~~~~~\hilight{olivegreen}{int}~i~=~\hilight{brickred}{0};} \\
+		\texttt{~~~~~~~~\hilight{brown}{while}~(i~<~length(list))~\{~\hilight{darkcyan}{//~Should~be~"<="!}} \\
+		\texttt{~~~~~~~~~~~~\hilight{brown}{if}~(element~<~list[i])~\{} \\
+		\texttt{~~~~~~~~~~~~~~~~\hilight{brown}{break};} \\
+		\texttt{~~~~~~~~~~~~\}} \\
+		\texttt{~~~~~~~~\}} \\
+		\texttt{~~~~~~~~insert\_at\_index(list,~element);} \\
+		\texttt{~~~~\}} \\
+\end{frame}
+
+\begin{frame}{Running Example}
+	\textbf{Broken \texttt{insert\_sorted}}
+
+	\linegap
+	Here is a bad test case. How can we improve it?
+
+	\linegap
+		\texttt{~~~~test\_insert\_sorted()~\{} \\
+		\texttt{~~~~~~~~input~=~\hilight{brickred}{[99,~42,~3,~100,~87,~56,~22]};} \\
+		\texttt{~~~~~~~~output~=~\hilight{brickred}{[]};} \\
+		\texttt{~~~~~~~~\hilight{brown}{foreach}~element~\hilight{brown}{in}~input~\{} \\
+		\texttt{~~~~~~~~~~~~insert\_sorted(output,~element);} \\
+		\texttt{~~~~~~~~\}} \\
+		\texttt{~~~~~~~~\hilight{brown}{return}~is\_sorted(output);} \\
+		\texttt{~~~~\}} \\
 \end{frame}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Main Approach}
+\section{Strategies}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 	Remember, two stories: Expected outcome vs Actual outcome
+
 	\linegap
@@
 	Remember, two stories: Expected outcome vs Actual outcome
+
 	\linegap
 \end{frame}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Main Approach}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 

### 2012-11-06 gpi-debug/debugging.tex
add time diagrams and 213 disclaimer

@@
 \institute[98-172]{Great Practical Ideas for Computer Scientists}
-\date[]{2012, November 1}
+\date[]{2012, November 8}
 
@@
 
-\begin{frame}{Asking for Help}
+\begin{frame}{Aside: Asking for Help}
 	This DOESN'T mean...
@@
 	\linegap
-	Here is a bad test case. How can we improve it?
+	Consider this test case:
 
 	\linegap
-		\texttt{~~~~test\_insert\_sorted()~\{} \\
-		\texttt{~~~~~~~~input~=~\hilight{brickred}{[99,~42,~3,~100,~87,~56,~22]};} \\
-		\texttt{~~~~~~~~output~=~\hilight{brickred}{[]};} \\
-		\texttt{~~~~~~~~\hilight{brown}{foreach}~element~\hilight{brown}{in}~input~\{} \\
-		\texttt{~~~~~~~~~~~~insert\_sorted(output,~element);} \\
-		\texttt{~~~~~~~~\}} \\
-		\texttt{~~~~~~~~\hilight{brown}{return}~is\_sorted(output);} \\
+		\texttt{~~~~input~=~\hilight{brickred}{[99,~42,~3,~100,~87,~56,~22]};} \\
+		\texttt{~~~~output~=~\hilight{brickred}{[]};} \\
+		\texttt{~~~~\hilight{brown}{foreach}~element~\hilight{brown}{in}~input~\{} \\
+		\texttt{~~~~~~~~insert\_sorted(output,~element);} \\
 		\texttt{~~~~\}} \\
+		\texttt{~~~~\hilight{brown}{return}~is\_sorted(output);~\hilight{darkcyan}{//~returns~false~:(}} \\
+	\linegap
+
+	\pause
+	\begin{itemize}
+		\item Says there is a bug
+		\item Doesn't say where
+	\end{itemize}
+	\pause
+	{\bf How can we improve it?}
 \end{frame}
@@
 	\linegap
-	What does this mean 
-	% Text: "First I ... then I ... but then it ...!"
+	{\em Time-wise}, this means ``It did X, then should have done Y but did Z''
+	\begin{itemize}
+		\item Are X,Y,Z in enough detail?
+		\item Did something important happen in between?
+	\end{itemize}
 \end{frame}
 
-\begin{frame}{Print debugging}
-	% Insert some code: the test case with print statements inserted in it
-	% Insert a picture: the time-cloud with some printed-out things in it
-	Should usually be the first thing you try!
+\begin{frame}{Telling a Story of Time}
+	Our test case tells this story:
+
+	\linegap
+	\includegraphics[width=\textwidth]{time0.png}
+\end{frame}
+
+\begin{frame}{Print Debugging}
+	Make your program print its state during the big pink ``???''.
+	\begin{itemize}
+		\item Reduce how much of the story the ``???'' obscures.
+		\begin{itemize}
+			\item Find the {\em last} print {\em before} the bug
+			\item Find the {\em first} print {\em after} the bug
+		\end{itemize}
+	\end{itemize}
+	\pause
+
+	\linegap
+		\texttt{~~~~\hilight{brown}{foreach}~element~\hilight{brown}{in}~input~\{} \\
+		\texttt{~~~~~~~~\hilight{olivegreen}{print(output);}} \\
+		\texttt{~~~~~~~~insert\_sorted(output,~element);} \\
+		\texttt{~~~~\}} \\
+\end{frame}
+
+\begin{frame}{Print Debugging}
+	New story:
+
+	\linegap
+	\includegraphics[width=\textwidth]{time1.png}
 \end{frame}
@@
 \begin{frame}{Asserts: Stealthier, more concealable prints}
+\end{frame}
+
+\begin{frame}{Aside: 15-213 ``malloc lab''}
+	At the end of 15-213, you will implement \texttt{malloc()}.
+	\begin{itemize}
+		\item Complex, low-level systems code
+		\item Very easy to end up with difficult, subtle bugs
+		\begin{itemize}
+			\item Your data winds up in the wrong place, gets clobbered by client code
+			\item Stale pointers that point to nonsense memory
+		\end{itemize}
+		\item Your code won't even misbehave immediately, only ``later''.
+	\end{itemize}
+	\pause
+	\linegap
+
+	The assignment will ask you to write a {\bf heap checker}.
+	\begin{itemize}
+		\item In essence, a very powerful \texttt{assert} for your code.
+		\item Debugging and asserts are friends!
+	\end{itemize}
+\end{frame}
+
+\begin{frame}{Aside: 15-213 ``malloc lab''}
+	Many students think: ``I just want to write \texttt{malloc()}. I can do the heap checker later.''
+	\pause
+	\begin{itemize}
+		\item Think of it like a C0 contract that must hold forever.
+		\item \textbf{Write the heap checker simultaneously!}
+			\pause
+			\begin{itemize}
+				\item Save time/effort debugging everything else
+				\item Your debugging skill will level up
+				\item Ben jumps up and down for emphasis
+			\end{itemize}
+	\end{itemize}
 \end{frame}

### 2012-11-06 gpi-debug/debugging.tex
tell a story of space and conclusions

@@
 \begin{frame}{Outline}
+	\begin{columns}[l]
+	\column{0.05\textwidth}
+	\column{0.5\textwidth}
 	\textbf{Why teach debugging?}
@@
 	\end{itemize}
+	\column{0.45\textwidth}
+	%\begin{center}
+	%	\includegraphics[width=0.9\textwidth]{vip1066720.jpg}
+	%\end{center}
+	\end{columns}
 \end{frame}
@@
 
-\breakslide{{\bf When} things go wrong\dots}
+\breakslide{``Time Travel'': {\bf When} things go wrong\dots}
 
@@
 
-	\linegap
 	\includegraphics[width=\textwidth]{time0.png}
@@
 		\texttt{~~~~\}} \\
+	\pause
+	\linegap
+
+	Creating {\em bounds} on when the problem happened
+	\begin{itemize}
+		\item Adding more prints = making bounds smaller
+		\item (Lots and lots of time? Binary search!)
+	\end{itemize}
 \end{frame}
@@
 
-	\linegap
 	\includegraphics[width=\textwidth]{time1.png}
@@
 \begin{frame}{Asserts: Stealthier, more concealable prints}
+	Adding print statements everywhere is very verbose
+	\begin{itemize}
+		\item Pro: Get to see everything
+		\item Con: Have to look at everything
+	\end{itemize}
+	\pause
+	\linegap
+
+	% TODO
 \end{frame}
@@
 
-\breakslide{{\bf Where} things go wrong\dots}
+\breakslide{``Space Travel'': {\bf Where} things go wrong\dots}
 
@@
 	\linegap
+	{\em Space-wise}, this means ``It supports doing X and Y operations, but operation Z is broken.''
+	\begin{itemize}
+		\item Now you can ignore X and Y
+	\end{itemize}
+\end{frame}
+
+\begin{frame}{Minimizing the Test Case}
+	Our test case tells this story:
+
+	\begin{center}
+	\includegraphics[width=0.65\textwidth]{space0.png}
+	\end{center}
+\end{frame}
+
+\begin{frame}{Minimizing the Test Case}
+	Which parts of your code are right?
+	\begin{itemize}
+		\item Write {\bf small tests that pass}
+	\end{itemize}
+	\linegap
+
+	Which parts of your code are wrong?
+	\begin{itemize}
+		\item {\bf Minimize the failing test}
+		\begin{itemize}
+			\item Make it as small as possible, such that it still fails
+		\end{itemize}
+	\end{itemize}
+	\pause
+	\linegap
+
+	\texttt{~~~~input~=~\hilight{brickred}{[99]};~~~~~\hilight{darkcyan}{//~OK}} \\
+	\texttt{~~~~input~=~\hilight{brickred}{[100,99]};~\hilight{darkcyan}{//~OK}} \\
+	\pause
+	\texttt{~~~~input~=~\hilight{brickred}{[99,100]};~\hilight{darkcyan}{//~Fails}} \\
+\end{frame}
+
+\begin{frame}{Minimizing the Test Case}
+	New story:
+
+	\begin{center}
+	\includegraphics[width=0.65\textwidth]{space1.png}
+	\end{center}
 \end{frame}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Main Approach}
+\section{Assorted Tricks}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\breakslide{Assorted Tricks \\ {\normalsize (Useful stuff that isn't really about ``space'' or ``time'')}}
+
+% TODO
+%\begin{frame}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Conclusion}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\breakslide{What's in this ``toolbox'', again? \\ {\normalsize (Conclusion)}}
+
+\begin{frame}{The Toolbox}
+	\begin{columns}[l]
+	\column{0.05\textwidth}
+	\column{0.5\textwidth}
+	\textbf{``Tell Me a Story''}
+	\begin{itemize}
+		\item Get more detail
+		\item Rinse \& repeat
+		\pause
+		\item {\bf Tell a Story of Time}
+		\begin{itemize}
+			\item Print debugging
+			\item Asserts
+		\end{itemize}
+		\pause
+		\item {\bf Tell a Story of Space}
+		\begin{itemize}
+			\item Minimize the failing test case
+			\item Make small tests that work
+		\end{itemize}
+		\pause
+		\item Draw pictures, collect data, question assumptions
+	\end{itemize}
+	\column{0.45\textwidth}
+	\pause
+	\begin{center}
+		\includegraphics[width=0.9\textwidth]{vip1066720.jpg}
+	\end{center}
+	\end{columns}
+\end{frame}
 

### 2012-11-07 gpi-debug/debugging.tex
code-complete lecture slides

@@
 	\pause
+	\linegap
+
 	{\bf 98-172}
@@
 \begin{frame}{Take-away}
+	This lecture's goal: To give you a {\em framework} for thinking when faced with buggy code.
+	\linegap
+
 	\textbf{Debugging toolbox}
@@
 	\pause
+	\linegap
+
 	This does mean...
@@
 \begin{frame}{Print Debugging}
+	Old story:
+
+	\includegraphics[width=\textwidth]{time0.png}
+\end{frame}
+\begin{frame}{Print Debugging}
 	New story:
@@
 	\pause
+	15-122's C0 ``contracts'' are assertions
+	\begin{itemize}
+		\item You don't need to use C0 to write assert-ful code
+		\item Asserts can go anywhere, not just around functions/loops
+	\end{itemize}
+	\pause
 	\linegap
 
-	% TODO
+	\texttt{~~~~\hilight{brown}{if}~(something\_that\_should\_never\_happen)~\{} \\
+	\texttt{~~~~~~~~print(\hilight{brickred}{"it~happened!"});} \\
+	\texttt{~~~~~~~~crash();} \\
+	\texttt{~~~~\}} \\
+\end{frame}
+\begin{frame}{Asserts: Stealthier, more concealable prints}
+	Adding print statements everywhere is very verbose
+	\begin{itemize}
+		\item Pro: Get to see everything
+		\item Con: Have to look at everything
+	\end{itemize}
+	15-122's C0 ``contracts'' are assertions
+	\begin{itemize}
+		\item You don't need to use C0 to write assert-ful code
+		\item Asserts can go anywhere, not just around functions/loops
+	\end{itemize}
+	\linegap
+
+		\texttt{~~~~\hilight{brown}{foreach}~element~\hilight{brown}{in}~input~\{} \\
+		\texttt{~~~~~~~~insert\_sorted(output,~element);} \\
+		\texttt{~~~~~~~~\hilight{olivegreen}{assert(is\_sorted(output));}} \\
+		\texttt{~~~~\}} \\
 \end{frame}
@@
 \begin{frame}{Minimizing the Test Case}
+	Old story:
+
+	\begin{center}
+	\includegraphics[width=0.65\textwidth]{space0.png}
+	\end{center}
+\end{frame}
+\begin{frame}{Minimizing the Test Case}
 	New story:
@@
 % TODO
-%\begin{frame}
+\begin{frame}{Rubber Duck / Cardboard Dog Debugging}
+	\begin{columns}[l]
+	\column{0.75\textwidth}
+	Hacker folklore says: describe your problem to a duck!
+	\begin{itemize}
+		\item ``What could I possibly get out of talking to an inanimate object?''
+	\end{itemize}
+	\linegap
+
+	\onslide<2->{
+	Hearing yourself speak makes you {\bf question your assumptions}.
+	\begin{itemize}
+		\item ``The bug can't be in X.''
+		\item ``Why not? Are you sure X works?''
+	\end{itemize}
+	\linegap
+
+	Helps figure out what part of the story to start searching at
+	}
+	\column{0.25\textwidth}
+	\includegraphics[width=\textwidth]{rubberduck.jpg}
+	\end{columns}
+\end{frame}
+
+\begin{frame}{Draw Pictures}
+	Draw what you think should happen
+
+	Label relevant variables and their values
+	\linegap
+
+	\begin{center}
+		\includegraphics[width=0.8\textwidth]{draw-a-picture.png}
+	\end{center}
+\end{frame}
 

### 2012-11-08 gpi-debug/debugging.tex
minor revise outline
finish debugging slides

@@
 			% "get more operators"
-		\item Binary Search
 	\end{itemize}
@@
 		\item Question assumptions
-		\item Collect data
+		\item Draw pictures
+		\item Binary search
 	\end{itemize}
@@
 		\texttt{~~~~~~~~\}} \\
-		\texttt{~~~~~~~~insert\_at\_index(list,~element);} \\
+		\texttt{~~~~~~~~insert\_at\_index(list,~element,~i);} \\
 		\texttt{~~~~\}} \\
@@
 	\linegap
-		\texttt{~~~~input~=~\hilight{brickred}{[99,~42,~3,~100,~87,~56,~22]};} \\
+		\texttt{~~~~input~=~\hilight{brickred}{[99,~42,~3,~100,~87,~56,~22,~38]};} \\
 		\texttt{~~~~output~=~\hilight{brickred}{[]};} \\
@@
 		\texttt{~~~~\}} \\
-	\pause
-	\linegap
 
-	Creating {\em bounds} on when the problem happened
-	\begin{itemize}
-		\item Adding more prints = making bounds smaller
-		\item (Lots and lots of time? Binary search!)
-	\end{itemize}
 \end{frame}
@@
 	\texttt{~~~~\hilight{brown}{if}~(something\_that\_should\_never\_happen)~\{} \\
-	\texttt{~~~~~~~~print(\hilight{brickred}{"it~happened!"});} \\
+	\texttt{~~~~~~~~print(\hilight{brickred}{"oh~no,~it~happened!"});} \\
 	\texttt{~~~~~~~~crash();} \\
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Assorted Tricks}
+\section{Assorted Tips}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-\breakslide{Assorted Tricks \\ {\normalsize (Useful stuff that isn't really about ``space'' or ``time'')}}
+\breakslide{Assorted Tips \\ {\normalsize (Useful stuff that isn't specifically about ``space'' or ``time'')}}
 
-% TODO
 \begin{frame}{Rubber Duck / Cardboard Dog Debugging}
@@
 \end{frame}
+\begin{frame}{Rubber Duck / Cardboard Dog Debugging}
+	\begin{columns}[l]
+	\column{0.75\textwidth}
+	{\em Question assumptions} also means {\em question explanations}.
+	\begin{itemize}
+		\item You: ``I think X is wrong''
+		\item Duck: ``How can you be sure?''
+	\end{itemize}
+	\onslide<2->{
+	\linegap
+	The {\bf Hypothesis/Test} workflow
+	\begin{itemize}
+		\item Guess what the root cause is
+		\item Write a small test that fails {\em iff} that's right.
+		\onslide<3->{
+		\begin{itemize}
+			\item Not ``change code blindly, see if that fixes it''
+			\item Until you know what's wrong, write {\em tests}
+		\end{itemize}
+		}
+	\end{itemize}
+	}
+	\column{0.25\textwidth}
+	\vspace{0.035in}
+	\includegraphics[width=\textwidth]{rubberduck.jpg}
+	\end{columns}
+\end{frame}
 
@@
 	\end{center}
+\end{frame}
+
+\begin{frame}{Binary Search}
+	Print statements and unit tests create {\em bounds} on when the problem happened
+	\begin{itemize}
+		\item Iteratively making bounds smaller helps you hone in on the bug.
+		\begin{itemize}
+			\item Add more print statements
+			\item Write unit tests to test many (small!) parts of the code
+		\end{itemize}
+		\item Hey wait, this sounds like a certain search algorithm...
+	\end{itemize}
+	\pause
+	Binary search: ``$O(\mathsf{log}(n))$'' is not just talking about code!
+	\linegap
+	\pause
+
+	\includegraphics[width=\textwidth]{binary1.png}
+\end{frame}
+\begin{frame}{Binary Search}
+	Print statements and unit tests create {\em bounds} on when the problem happened
+	\begin{itemize}
+		\item Iteratively making bounds smaller helps you hone in on the bug.
+		\begin{itemize}
+			\item Add more print statements
+			\item Write unit tests to test many (small!) parts of the code
+		\end{itemize}
+		\item Hey wait, this sounds like a certain search algorithm...
+	\end{itemize}
+	Binary search: ``$O(\mathsf{log}(n))$'' is not just talking about code!
+	\linegap
+
+	\includegraphics[width=\textwidth]{binary2.png}
+\end{frame}
+\begin{frame}{Binary Search}
+	Print statements and unit tests create {\em bounds} on when the problem happened
+	\begin{itemize}
+		\item Iteratively making bounds smaller helps you hone in on the bug.
+		\begin{itemize}
+			\item Add more print statements
+			\item Write unit tests to test many (small!) parts of the code
+		\end{itemize}
+		\item Hey wait, this sounds like a certain search algorithm...
+	\end{itemize}
+	Binary search: ``$O(\mathsf{log}(n))$'' is not just talking about code!
+	\linegap
+
+	\includegraphics[width=\textwidth]{binary3.png}
+\end{frame}
+\begin{frame}{Binary Search}
+	Print statements and unit tests create {\em bounds} on when the problem happened
+	\begin{itemize}
+		\item Iteratively making bounds smaller helps you hone in on the bug.
+		\begin{itemize}
+			\item Add more print statements
+			\item Write unit tests to test many (small!) parts of the code
+		\end{itemize}
+		\item Hey wait, this sounds like a certain search algorithm...
+	\end{itemize}
+	Binary search: ``$O(\mathsf{log}(n))$'' is not just talking about code!
+	\linegap
+
+	\includegraphics[width=\textwidth]{binary4.png}
+\end{frame}
+\begin{frame}{Binary Search}
+	This also applies to ``space'', not just to ''time''
+	\begin{itemize}
+		\item Sometimes the only thing that makes your code crash is a 1000-line test case
+		\begin{itemize}
+			\item Binary search on ``which parts of the test need to be there''
+		\end{itemize}
+	\end{itemize}
+	\pause
+	\linegap
+	Still don't believe me? \texttt{git bisect}
+	\begin{itemize}
+		\item Helps find which {\em git revision} introduced a bug
+		\item Start with one ``known good'' revision, one ``known bad'' one
+		\item \texttt{bisect} binary-searches on revisions in-between, asking you at each if it's good or bad.
+	\end{itemize}
+\end{frame}
+
+\begin{frame}{One Last Aside: Repeatability}
+	All the tests and bugs here have been {\em deterministic}.
+	\begin{itemize}
+		\item If you see the failure once, you can keep making it happen until it's fixed.
+	\end{itemize}
+	\pause
+	\linegap
+
+	In the ``real world'', and in 400-level classes, failures are not so easy to reproduce.
+	\begin{itemize}
+		\item Consider: \texttt{input = [rand(), rand(), rand(), rand()]}
+		\item Might have failed last week, but never since
+	\end{itemize}
+	\pause
+	\linegap
+	Make sure your test cases are {\bf repeatable}
+	\begin{itemize}
+		\item Getting help (from a TA, from a company) easier when you say ``Here is a failing test'' than ''I once saw a test fail''!
+		\item Nonrepeatable tests are often complex
+		\begin{itemize}
+			\item Use the rest of these techniques to produce a relevant, minimal failing test case.
+		\end{itemize}
+	\end{itemize}
 \end{frame}
@@
 		\pause
-		\item Draw pictures, collect data, question assumptions
+		\item Draw pictures, collect data, question assumptions, binary search
 	\end{itemize}

### 2012-11-08 gpi-debug/debugging.tex
final revisions i think

@@
 
-\begin{frame}{Running Example}
-	\textbf{Broken \texttt{insert\_sorted}}
-
-	\linegap
-	Sorts OK into the middle and front of list:
-	\begin{itemize}
-		\item \texttt{insert\_sorted} $\quad \hilight{olivegreen}{10} \quad  [20,30] \quad \Rightarrow \quad [\hilight{olivegreen}{10},20,30]$
-		\item \texttt{insert\_sorted} $\quad \hilight{olivegreen}{32} \quad  [16,64] \quad \Rightarrow \quad [16,\hilight{olivegreen}{32},64]$
-	\end{itemize}
-
-	\linegap
-	But not so great at the end:
-	\begin{itemize}
-	\item \texttt{insert\_sorted} $\quad \hilight{red}{100}  \quad [97,98,99] \quad \Rightarrow \quad [97,98,\hilight{red}{100},99]$
-	\end{itemize}
-\end{frame}
+% \begin{frame}{Running Example}
+%
+%	\linegap
+%	Sorts OK into the middle and front of list:
+%	\begin{itemize}
+%		\item \texttt{insert\_sorted} $\quad \hilight{olivegreen}{10} \quad  [20,30] \quad \Rightarrow \quad [\hilight{olivegreen}{10},20,30]$
+%		\item \texttt{insert\_sorted} $\quad \hilight{olivegreen}{32} \quad  [16,64] \quad \Rightarrow \quad [16,\hilight{olivegreen}{32},64]$
+%	\end{itemize}
+%
+%	\linegap
+%	But not so great at the end:
+%	\begin{itemize}
+%	\item \texttt{insert\_sorted} $\quad \hilight{red}{100}  \quad [97,98,99] \quad \Rightarrow \quad [97,98,\hilight{red}{100},99]$
+%	\end{itemize}
+% \end{frame}
 
@@
 	\linegap
-	In case you're curious, here's the bug:
-
-	\linegap
 		\texttt{~~~~insert\_sorted(list,~element)~\{} \\
 			\texttt{~~~~~~~~\hilight{olivegreen}{int}~i~=~\hilight{brickred}{0};} \\
-		\texttt{~~~~~~~~\hilight{brown}{while}~(i~<~length(list))~\{~\hilight{darkcyan}{//~Should~be~"<="!}} \\
+		\texttt{~~~~~~~~\hilight{brown}{while}~(i~<~length(list))~\{} \\
 		\texttt{~~~~~~~~~~~~\hilight{brown}{if}~(element~<~list[i])~\{} \\
@@
 		\texttt{~~~~~~~~~~~~\}} \\
+		\texttt{~~~~~~~~~~~~i = i + 1;} \\
 		\texttt{~~~~~~~~\}} \\
@@
 		\texttt{~~~~\}} \\
+	\pause
+	\linegap
+	\begin{itemize}
+		\item Staring at the code until you see it sometimes works, but is a bad habit.
+	\end{itemize}
 \end{frame}
@@
 
-\subsection{Time Travel}
+\subsection{A Story of Time}
 
@@
 	\begin{itemize}
-		\item You don't need to use C0 to write assert-ful code
+		\item You don't need to be writing C0 to write assert-ful code
 		\item Asserts can go anywhere, not just around functions/loops
@@
 		\item Think of it like a C0 contract that must hold forever.
-		\item \textbf{Write the heap checker simultaneously!}
+		\item \textbf{Write the heap checker simultaneously (or even first)!}
 			\pause
@@
 
-\subsection{Space Travel}
+\subsection{A Story of Space}
 
@@
 
-	\texttt{~~~~input~=~\hilight{brickred}{[99]};~~~~~\hilight{darkcyan}{//~OK}} \\
-	\texttt{~~~~input~=~\hilight{brickred}{[100,99]};~\hilight{darkcyan}{//~OK}} \\
+	%\texttt{~~~~input~=~\hilight{brickred}{[99]};~~~~~\hilight{darkcyan}{//~OK}} \\
+	%\texttt{~~~~input~=~\hilight{brickred}{[100,99]};~\hilight{darkcyan}{//~OK}} \\
+	%\pause
+	%\texttt{~~~~input~=~\hilight{brickred}{[99,100]};~\hilight{darkcyan}{//~Fails}} \\
+
+	\texttt{insert\_sorted} $\quad \hilight{olivegreen}{99}  \quad [~]{\small~~~~} \quad \Rightarrow \quad [\hilight{olivegreen}{99}] \quad$ ~~~~~~(empty list OK) \\
+	\texttt{insert\_sorted} $\quad \hilight{olivegreen}{99}  \quad [100] \quad \Rightarrow \quad [\hilight{olivegreen}{99},100] \quad$ (front of list OK) \\
 	\pause
-	\texttt{~~~~input~=~\hilight{brickred}{[99,100]};~\hilight{darkcyan}{//~Fails}} \\
+	\texttt{insert\_sorted} $\quad \hilight{red}{100}  \quad [99] \quad \Rightarrow \quad [\hilight{red}{100},99] \quad$ (end of list BAD)
 \end{frame}
@@
 
+\begin{frame}{\texttt{insert\_sorted}, revisited}
+	Now that we know {\em where} and {\em when} the program breaks, let's fix it.
+	\begin{itemize}
+		\item Raise your hand (don't say anything) when you see what's wrong.
+		\item Remember: only fails at the {\em end of the list}.
+	\end{itemize}
+
+	\linegap
+		\texttt{~~~~insert\_sorted(list,~element)~\{} \\
+			\texttt{~~~~~~~~\hilight{olivegreen}{int}~i~=~\hilight{brickred}{0};} \\
+		\texttt{~~~~~~~~\hilight{brown}{while}~(i~<~length(list))~\{} \\
+		\texttt{~~~~~~~~~~~~\hilight{brown}{if}~(element~<~list[i])~\{} \\
+		\texttt{~~~~~~~~~~~~~~~~\hilight{brown}{break};} \\
+		\texttt{~~~~~~~~~~~~\}} \\
+		\texttt{~~~~~~~~~~~~i = i + 1;} \\
+		\texttt{~~~~~~~~\}} \\
+		\texttt{~~~~~~~~insert\_at\_index(list,~element,~i);} \\
+		\texttt{~~~~\}} \\
+\end{frame}
+\begin{frame}{\texttt{insert\_sorted}, revisited}
+	Now that we know {\em where} and {\em when} the program breaks, let's fix it.
+	\begin{itemize}
+		\item Raise your hand (don't say anything) when you see what's wrong.
+		\item Remember: only fails at the {\em end of the list}.
+	\end{itemize}
+
+	\linegap
+		\texttt{~~~~insert\_sorted(list,~element)~\{} \\
+			\texttt{~~~~~~~~\hilight{olivegreen}{int}~i~=~\hilight{brickred}{0};} \\
+			\texttt{~~~~~~~~\hilight{brown}{while}~(\hilight{red}{i~<~length(list)})~\{~\hilight{darkcyan}{//~Should~be~"<="!}} \\
+		\texttt{~~~~~~~~~~~~\hilight{brown}{if}~(element~<~list[i])~\{} \\
+		\texttt{~~~~~~~~~~~~~~~~\hilight{brown}{break};} \\
+		\texttt{~~~~~~~~~~~~\}} \\
+		\texttt{~~~~~~~~~~~~i = i + 1;} \\
+		\texttt{~~~~~~~~\}} \\
+		\texttt{~~~~~~~~insert\_at\_index(list,~element,~i);} \\
+		\texttt{~~~~\}} \\
+\end{frame}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 		\item ``What could I possibly get out of talking to an inanimate object?''
+		\pause
+		\begin{itemize}
+			\item The duck only knows what you tell it
+			\item When it doesn't know, it responds with ``WHY?''
 		\end{itemize}
+	\end{itemize}
+	\pause
 	\linegap
@@
 	\column{0.25\textwidth}
-	\vspace{0.035in}
+	\vspace{0.095in}
 	\includegraphics[width=\textwidth]{rubberduck.jpg}
@@
 	\begin{itemize}
-		\item Getting help (from a TA, from a company) easier when you say ``Here is a failing test'' than ''I once saw a test fail''!
+		\item Getting help (from a TA) easier when you say ``Here is a failing test'', not ''I once saw a test fail''!
 		\item Nonrepeatable tests are often complex
 		\begin{itemize}
-			\item Use the rest of these techniques to produce a relevant, minimal failing test case.
+			\item Use the rest of these techniques to produce a relevant, minimal, {\em reliable} failing test case.
 		\end{itemize}
@@
 		\begin{itemize}
+			\item Make small tests that work
 			\item Minimize the failing test case
-			\item Make small tests that work
 		\end{itemize}

### 2012-11-08 gpi-debug/debugging.tex
fix the accidentally correct code to be wrong

@@
 			\texttt{~~~~~~~~\hilight{olivegreen}{int}~i~=~\hilight{brickred}{0};} \\
-		\texttt{~~~~~~~~\hilight{brown}{while}~(i~<~length(list))~\{} \\
+			\texttt{~~~~~~~~\hilight{brown}{while}~(i~<~length(list)~-~\hilight{brickred}{1})~\{} \\
 		\texttt{~~~~~~~~~~~~\hilight{brown}{if}~(element~<~list[i])~\{} \\
@@
 		\texttt{~~~~~~~~~~~~\}} \\
-		\texttt{~~~~~~~~~~~~i = i + 1;} \\
+		\texttt{~~~~~~~~~~~~i = i + \hilight{brickred}{1};} \\
 		\texttt{~~~~~~~~\}} \\
@@
 			\texttt{~~~~~~~~\hilight{olivegreen}{int}~i~=~\hilight{brickred}{0};} \\
-			\texttt{~~~~~~~~\hilight{brown}{while}~(\hilight{red}{i~<~length(list)})~\{~\hilight{darkcyan}{//~Should~be~"<="!}} \\
-		\texttt{~~~~~~~~~~~~\hilight{brown}{if}~(element~<~list[i])~\{} \\
+			\texttt{~~~~~~~~\hilight{brown}{while}~(\hilight{red}{i~<~length(list)~-~1})~\{~\hilight{darkcyan}{//~Should~be~"<="!}} \\
+				\texttt{~~~~~~~~~~~~\hilight{brown}{if}~(element~<~list[i])~\{~~~\hilight{darkcyan}{~~~(or~not~"-1")}} \\
 		\texttt{~~~~~~~~~~~~~~~~\hilight{brown}{break};} \\

### 2013-04-15 landslide-lecture/lecture.tex
Revise slide deck for 2013

@@
 \institute[CMU 15-410]{Carnegie Mellon University - 15-410}
-\date[]{2012, April 6}
+\date[]{2013, April 14}
 
@@
 	    {
-	    thread_t *child = spawn_new_thread();
+	        thread_t *child = construct_new_thread();
 	        add_to_runqueue(child);
@@
 	\end{itemize}
+	\linegap
 	What if\ldots
@@
 	\end{itemize}
+	\linegap
 	{\bf Backtracking}:
@@
 		\item Reset machine state and start over
-		\item Replay test from the beginning, with different choices
+		\item Replay test from the beginning, with different ``decisions''
 	\end{itemize}
@@
 	\begin{itemize}
-		\item Good news: Will find every possible race condition.
-		\item Bad news: Runtime of test will be impossibly large.
+		\item {\em Good news}: Will find every possible race condition.
+		\item {\em Bad news}: Runtime of test will be impossibly large.
 	\end{itemize}
@@
 	\begin{itemize}
-		\item Good news: Test will finish quickly.
-		\item Bad news: Only one execution was checked for bugginess.
-		\item Bad news: No alternative interleavings explored.
+		\item {\em Good news}: Test will finish quickly.
+		\item {\em Bad news}: Only one execution was checked for bugginess.
+		\item {\em Bad news}: No alternative interleavings explored.
 		\begin{itemize}
@@
 	5th year MS since June 2011 % Say: planning to finish this may, hopefully
+	My MS thesis project (June 2011-2012)
+
+	\linegap
+	Now I'm a Ph.D. student\dots
 
@@
 	\linegap
-	Checkpointing/backtracking via Simics bookmarks
+	Checkpointing/backtracking via Simics bookmarks (reverse execution)
 \end{frame}
@@
 	\end{itemize}
+	\linegap
 	Landslide can {\em reasonably suspect}:
@@
 		\item Memory leak
-		\item Progress sense (halting problem)
+		\item Infinite loop (halting problem)
 	\end{itemize}
@@
 		\item Landslide may help you find bugs
-		\item You may help Ben evaluate his thesis project
+		\item You help Ben evaluate his research
 	\end{itemize}
@@
 	\end{itemize}
-	\linegap
-
-	\framebox{Time estimate: 40 minutes}
 \end{frame}
@@
 	% Say: Which depend on your kernel's specific implementation
-	\linegap
-
-	\framebox{Time estimate: 60 minutes}
 \end{frame}
@@
 	\end{itemize}
+	\linegap
 	With help from you, it could find more.
@@
 	\end{itemize}
+\end{frame}
+
+\begin{frame}{And Hopefully\dots}
+	\begin{center}
+		\includegraphics[width=\textwidth]{found_a_bug.png}
+	\end{center}
 \end{frame}
@@
 
+\begin{frame}{Last Year's Results}
+	\begin{itemize}
+		\item Worked with 5 groups
+		\begin{itemize}
+			\item \dots{}of which 4 put in enough time to make Landslide work
+		\end{itemize}
+		\linegap
+		\item {\bf Investment}: about 2 to 3 hours
+		\begin{itemize}
+			\item Average 112 minutes doing required instrumentation
+			\item Average 35 minutes configuring extra decision points \& interpreting ``found a bug'' output
+		\end{itemize}
+		\pause
+		\linegap
+		\item {\bf Return}: All 4 groups found bugs in their kernels.
+		\begin{itemize}
+			\item All groups found deterministic bugs (e.g., use-after-free)
+			\item 2 groups found race conditions (i.e., Landslide needed to backtrack)
+		\end{itemize}
+	\end{itemize}
+\end{frame}
+
 \begin{frame}{In Which Ben Offers Help - Warning}
@@
 	\end{itemize}
-	You should:
+	\pause
+	\linegap
+	This is for you if:
 	\begin{itemize}
-		\item Be expecting an A or B
-		\item Be able to turn in without late days (if you really had to)
-		\item Be looking for\ldots
+		%\item Be expecting an A or a B
+		\item You have already submitted
+		\item You are using late days, but willing to work over Carnival
+		%\item Be able to turn in without late days (if you really had to)
+		\item You are taking p3extra, but you already pass the hurdle
+		\item You are looking for\ldots
 			\begin{itemize}
@@
 \begin{frame}{In Which Ben Offers Help}
-	User study next week, starting Monday
+	User study this week, by appointment (\texttt{bblum@cs.cmu.edu})
 
@@
 	\begin{itemize}
-		\item Up to 4 hours, just to try it out.
-		\item 6-8 hours, if you find a bug and track it down.
+		\item Up to 3 hours, just to try it out.
+		\item 4-6 hours, if you find a bug and track it down.
 		\item More, for multiple bugs or the truly dedicated\ldots

### 2013-08-13 mozilla-internship2/talk.tex
Add 2013 moz intern talk

\documentclass[xcolor=dvipsnames]{beamer}
\usepackage{lmodern}
\usepackage[T1]{fontenc}
\usepackage[english]{babel}
\usepackage[utf8]{inputenc}

\usepackage{manfnt}
\usepackage{wasysym}
\usepackage{listings}
\usepackage{graphicx}
\usepackage{url}
\usepackage{ulem}
\usepackage{marvosym}
\usepackage{skull}
\usepackage{proof}
\usepackage{array}
\setbeamertemplate{navigation symbols}{}

\title[Kinds for Parallelism]{{\bf Types of Types in Rust}}
\subtitle[]{ {\em Better parallelism through traits and kinds}}
\author[Ben Blum]{Ben Blum \texttt{(bblum@andrew.cmu.edu)}}

\institute[Mozilla Research]{Mozilla Research}
\date[]{2013, August 13}

\setbeamertemplate{footline}{\hspace*{.5cm}\scriptsize{\insertauthor\hspace*{50pt} \hfill\insertframenumber\hspace*{.5cm}}}

\usecolortheme{seahorse}
\usecolortheme{rose}
\useoutertheme{infolines}

\usecolortheme[named=RoyalBlue]{structure}

\newcommand\noob{\mathsf{noob}}
\newcommand\gibs{\mathsf{gibs}}
\newcommand\dps{\mathsf{dps}}
\newcommand\squig\rightsquigarrow
\newcommand\Coloneqq{\mathrel{\mathop{::}}=}
\newcommand\dmg{\text{\Laserbeam}}
\newcommand\delter\delta
\newcommand\alpher\alpha
\newcommand\defnor{\text{ }|\text{ }}

\newcommand\pimp{\mathop{\supset}}
\newcommand\pand{\mathop{\wedge}}
\newcommand\por{\mathop{\vee}}
\newcommand\ptrue{\top}
\newcommand\pfalse{\bot}

\begin{document}
\renewcommand{\inserttotalframenumber}{28}
\normalem
\begin{frame}
	\titlepage
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\newcommand\linegap{\vspace{0.2in}}
\newcommand\breakslide[1]{\begin{frame}{} \begin{center} \Large #1 \end{center} \end{frame}}
\newcommand\related[1]{\textsuperscript{\em [#1]}}
\newcommand\hilight[2]{\color{#1}#2\color{black}}

\begin{frame}{Outline}
	\begin{columns}
	\column{0.05\textwidth}
	\column{0.5\textwidth}
	\textbf{Overview}
	\begin{itemize}
		\item Kinds Overview
		\item Traits Overview
		\item Existential Types Overview
	\end{itemize}
	\linegap

	{\bf Kind Bounds}
	\begin{itemize}
		\item Reasoning about Captures
		\item New Expressivity
	\end{itemize}
	\linegap

	{\bf Use Case: Concurrent Data Types}
	\linegap

	\column{0.45\textwidth}
	\begin{center}
	\includegraphics[width=0.9\textwidth]{rust.png}
	\end{center}
	\end{columns}
\end{frame}

\begin{frame}{We have a new runtime!}
	I also worked on it, but... \\
	\pause
	\begin{center}
		\includegraphics[width=0.8\textwidth]{last-year.png}
	\end{center}
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Overview}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\newcommand\code[1]{{\begin{center}\fbox{\begin{tabular}{l} #1 \end{tabular}} \end{center}}}

\definecolor{grey}{RGB}{127,127,127}
\definecolor{darkcyan}{RGB}{0,127,127}
\definecolor{olivegreen}{RGB}{0,127,0}
\definecolor{violet}{RGB}{127,0,127}
\definecolor{brickred}{RGB}{127,0,0}
\definecolor{brown}{RGB}{127,63,0}

\breakslide{Kinds}

\subsection{Kinds Overview}

\begin{frame}{What are Kinds?}
	\begin{itemize}
		\item Type theory lingo: ``kinds'' == ``types of types''
		\pause
		\item In Rust, kinds (aka ``builtin bounds'') describe {\em capabilities} of types:
		\begin{itemize}
			\item {\tt \hilight{olivegreen}{Send}} - Data can be sent to other tasks
			\item {\tt \hilight{olivegreen}{Freeze}} - Data has no mutable interior
			\item {\tt \hilight{olivegreen}{Static}} - Data never goes out of scope
			\item {\tt \hilight{olivegreen}{Copy}} - Data can be implicitly copied
				\pause
				\begin{itemize}
					\item $\skull$
				\end{itemize}
		\end{itemize}
		\pause
		\item Kind bounds are automatically derived: \\
		\begin{itemize}
			\item \begin{tabular}{l}
				\texttt{\hilight{brown}{enum}~\hilight{olivegreen}{Option}<T>~\{} \\
				\texttt{~~~~\hilight{brickred}{Some}(T),} \\
				\texttt{~~~~\hilight{brickred}{None}} \\
				\texttt{\}} \\
			\end{tabular}
		\item {\tt \hilight{olivegreen}{Option}<T>} has kind {\tt K} iff {\tt T} has kind {\tt K}.
		\end{itemize}
	\end{itemize}
\end{frame}
\begin{frame}{What are kind bounds good for?}
	\begin{itemize}
		\item Message passing:
		\begin{itemize}
			\item \begin{tabular}{l}
					\texttt{\hilight{brown}{fn}~stream<T:~\hilight{olivegreen}{Send}>()~->~(\hilight{olivegreen}{Port}<T>,~\hilight{olivegreen}{Chan}<T>)} \\
					\texttt{} \\
					\texttt{\hilight{brown}{fn}~recv<T:~\hilight{olivegreen}{Send}>(port:~\&\hilight{olivegreen}{Port}<T>)~->~T} \\ 
					\texttt{} \\
					\texttt{\hilight{brown}{fn}~send<T:~\hilight{olivegreen}{Send}>(chan:~\&\hilight{olivegreen}{Chan}<T>,~data:~T)} \\

			\end{tabular}
			\item \texttt{\hilight{olivegreen}{Send}} bound ensures once data is sent, sender can no longer access it.
		\end{itemize}
		\pause
		\linegap
		\item Shared state across tasks:
		\begin{itemize}
			\item \begin{tabular}{l}
					\texttt{\hilight{brown}{fn}~\hilight{olivegreen}{Arc}<T:~\hilight{olivegreen}{Send}+\hilight{olivegreen}{Freeze}>(data:~T)~->~\hilight{olivegreen}{Arc}<T>} \\
					\texttt{} \\
					\texttt{\hilight{brown}{fn}~get<'a,~T:~\hilight{olivegreen}{Send}+\hilight{olivegreen}{Freeze}>(arc:~\&'a~\hilight{olivegreen}{Arc}<T>)~->~\&'a~T} \\
					\\
					\texttt{\hilight{brown}{fn}~clone<T:~\hilight{olivegreen}{Send}+\hilight{olivegreen}{Freeze}>(arc:~\&\hilight{olivegreen}{Arc}<T>)~->~\hilight{olivegreen}{Arc}<T>} \\

			\end{tabular}
		\end{itemize}
	\end{itemize}
\end{frame}

\subsection{Traits Overview}
\breakslide{Traits and Objects}

\begin{frame}{What are Traits?}
	\begin{itemize}
		\item Typeclasses, interfaces, ...
			\begin{itemize}
				\item \begin{tabular}{l}
						\texttt{\hilight{brown}{trait}~\hilight{olivegreen}{Pet}~\{} \\

					\texttt{~~~~\hilight{brown}{fn}~name(\&self)~->~\textasciitilde{}str} \\
					\texttt{~~~~\hilight{brown}{fn}~num\_legs(\&self)~->~int} \\
					\texttt{\}} \\

				\end{tabular}
			\end{itemize}
		\pause
		\item Also objects (existential types):
			\begin{itemize}
				\item \begin{tabular}{l}
						\texttt{struct~\hilight{olivegreen}{Dog}~\{~name:~\textasciitilde{}str,~bark\_decibels:~int~\}} \\
					\texttt{} \\
					\texttt{\hilight{brown}{impl}~\hilight{olivegreen}{Pet}~\hilight{brown}{for}~\hilight{olivegreen}{Dog}~\{~...~\}} \\
					\texttt{} \\
					\texttt{\hilight{brown}{let}~alan\_turing:~\hilight{olivegreen}{Dog}~=~\hilight{olivegreen}{Dog}~\{~...~\}} \\
					\texttt{} \\
					\texttt{\hilight{brown}{let}~generic\_pet:~\&\hilight{olivegreen}{Pet}~=~\&alan\_turing~\hilight{brown}{as}~\&\hilight{olivegreen}{Pet};} \\
					\texttt{~~~~\hilight{darkcyan}{//~\^~value~contains~a~vtable}} \\

				\end{tabular}
			\end{itemize}
	\end{itemize}
\end{frame}

\begin{frame}{Existential Types}
	\begin{itemize}
		\item ``Opaque type that presents a generic interface''
		\begin{itemize}
			\item $\exists x:T \rightarrow \mathsf{SomeInterface}(x)$
		\end{itemize}
		\item OOP: Interfaces-as-objects
		\pause
		\item Closures are a ``special case'' of trait objects:
		\begin{itemize}
			\item \begin{tabular}{l}
					\texttt{\hilight{brown}{trait}~\hilight{olivegreen}{Closure}<Args,Ret>~\{} \\
			\texttt{~~~~\hilight{brown}{fn}~call(self,~Args)~->~Ret;} \\
			\texttt{\}} \\
			\end{tabular}
		\end{itemize}
		\pause
		\item In Rust, existential objects are {\em dynamically-sized}
			\begin{itemize}
				\item Have to be hidden behind a {\tt \&}, {\tt \textasciitilde}, or {\tt @} pointer.
			\end{itemize}

	\end{itemize}
\end{frame}

\begin{frame}{What's in an Object?}
	\begin{itemize}
		\item (In terms of kind bounds?)
		\pause
		\item Some inflexible defaults:
			\begin{itemize}
				\item {\tt \textasciitilde{}Trait} can only contain {\tt \hilight{olivegreen}{Send}}able values
				\item {\tt @Trait} can only contain {\tt \hilight{olivegreen}{Static}} values
				\item {\tt \&Trait} can contain any values
				\pause
				\item Traits are not {\tt \hilight{olivegreen}{Freeze}}able or {\tt \hilight{olivegreen}{Copy}}able.
			\end{itemize}
			\pause
		\item Enter {\em kind bounds}.
	\end{itemize}
\end{frame}

\section{Kind Bounds}
\breakslide{Kind Bounds}

\subsection{Reasoning about Captured Values}

\begin{frame}{Kind Bounds on Traits}
	\begin{itemize}
		\item Today, \texttt{\textasciitilde{}\hilight{olivegreen}{Pet}:\hilight{olivegreen}{Copy}} is also a type.

		\begin{itemize}
			\item Restriction: ``Only copyable values can be captured in this``
			\item Capability: ``...so this whole thing can be copied!''
		\end{itemize}
		\pause
		Example:
		\begin{itemize}
			\item \begin{tabular}{l}
				\texttt{\hilight{brown}{let}~noncopyable\_alan\_turing:~\hilight{olivegreen}{Dog}~=~\hilight{olivegreen}{Dog}~\{~...~\}} \\
				\pause
				\texttt{} \\
				\texttt{\hilight{brown}{let}~generic\_pet:~\&\hilight{olivegreen}{Pet}~=~\&noncopyable\_alan\_turing~\hilight{brown}{as}~\&\hilight{olivegreen}{Pet};} \\
				\texttt{~~~~\hilight{darkcyan}{//~\^~OK}} \\
				\texttt{} \\
			\texttt{\hilight{brown}{let}~copyable\_pet:~\&\hilight{olivegreen}{Pet}:\hilight{olivegreen}{Copy}~=} \\
				\texttt{~~~~~~~~\&noncopyable\_alan\_turing~\hilight{brown}{as}~\&\hilight{olivegreen}{Pet}:\hilight{olivegreen}{Copy};} \\
				\texttt{~~~~\hilight{darkcyan}{//~\^~{error}:~cannot~capture~noncopyable}} \\
				\texttt{~~~~\hilight{darkcyan}{//~~~~~~~~~~~value~in~{trait}~bounded~by~{Copy}}} \\
			\end{tabular}
		\end{itemize}
	\end{itemize}
\end{frame}

\begin{frame}{What is this good for?}
	\begin{itemize}
		\item {\bf Sharing lists of objects between threads}
		\pause
		\begin{itemize}
			\item ``Heterogeneous'' objects, i.e., different implementations of the same trait
		\end{itemize}
		\pause
		Example:
		\begin{itemize}
			\item \begin{tabular}{l}
					\texttt{\hilight{brown}{let}~alan\_turing~=~\hilight{olivegreen}{Dog}~\{~...~\}} \\
					\texttt{\hilight{brown}{let}~ada\_lovelace~=~\hilight{olivegreen}{Cat}~\{~...~\}} \\
					\texttt{\hilight{brown}{let}~albert\_einstein~=~\hilight{olivegreen}{Bird}~\{~...~\}} \\
					\pause
					\texttt{}\\
					\texttt{\hilight{brown}{let}~pets~=~\textasciitilde{}[alan\_turing~~~~~\hilight{brown}{as}~\hilight{olivegreen}{Dog}:~~\hilight{olivegreen}{Send}+\hilight{olivegreen}{Freeze},} \\
						\texttt{~~~~~~~~~~~~~ada\_lovelace~~~~\hilight{brown}{as}~\hilight{olivegreen}{Cat}:~~\hilight{olivegreen}{Send}+\hilight{olivegreen}{Freeze},} \\
					\texttt{~~~~~~~~~~~~~albert\_einstein~\hilight{brown}{as}~\hilight{olivegreen}{Bird}:~\hilight{olivegreen}{Send}+\hilight{olivegreen}{Freeze}];} \\
					\texttt{} \\
					\texttt{\hilight{brown}{let}~shared\_pets~=~\hilight{olivegreen}{Arc}(pets);} \\
					\texttt{\hilight{darkcyan}{//~proceed~to~share~pets~among~many~parallel~tasks}} \\

			\end{tabular}
		\end{itemize}
	\end{itemize}
\end{frame}

\begin{frame}{Perspective}
	Can't C++ already do this?
	\begin{itemize}
		\item Q: ``IDGI, why is it so great that you used to not be able to do something C++ could always do?''
		\pause
		\item A: Doing the same thing while not compromising on safety guarantees (no data races)
	\end{itemize}
\end{frame}

\breakslide{Thanks to\dots}

\begin{frame}{Questions?}
	% rust logo
	\begin{center}
		\includegraphics[width=0.6\textwidth]{rust.png}
	\end{center}
\end{frame}

\end{document}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% vim: foldmethod=indent

### 2013-08-13 mozilla-internship2/talk.tex
finish talk hopefully

@@
 
-	{\bf Use Case: Concurrent Data Types}
+	{\bf Use Case: Concurrent Data Structures}
 	\linegap
@@
 		\begin{itemize}
-			\item {\tt \hilight{olivegreen}{Send}} - Data can be sent to other tasks
-			\item {\tt \hilight{olivegreen}{Freeze}} - Data has no mutable interior
-			\item {\tt \hilight{olivegreen}{Static}} - Data never goes out of scope
-			\item {\tt \hilight{olivegreen}{Copy}} - Data can be implicitly copied
+			\item {\tt \hilight{violet}{Send}} - Data can be sent to other tasks
+			\item {\tt \hilight{violet}{Freeze}} - Data has no mutable interior
+			\item {\tt \hilight{violet}{Static}} - Data never goes out of scope
+			\item {\tt \hilight{violet}{Copy}} - Data can be implicitly copied
 				\pause
@@
 			\end{tabular}
-		\item {\tt \hilight{olivegreen}{Option}<T>} has kind {\tt K} iff {\tt T} has kind {\tt K}.
+		\item {\tt \hilight{olivegreen}{Option}<T>} has kind {\tt \hilight{violet}{K}} iff {\tt T} has kind {\tt \hilight{violet}{K}}.
 		\end{itemize}
@@
 \end{frame}
+
+\subsection{Motivation}
 \begin{frame}{What are kind bounds good for?}
-	\begin{itemize}
-		\item Message passing:
+		Message passing:
 		\begin{itemize}
 			\item \begin{tabular}{l}
-					\texttt{\hilight{brown}{fn}~stream<T:~\hilight{olivegreen}{Send}>()~->~(\hilight{olivegreen}{Port}<T>,~\hilight{olivegreen}{Chan}<T>)} \\
+					\texttt{\hilight{brown}{fn}~stream<T:~\hilight{violet}{Send}>()~->~(\hilight{olivegreen}{Port}<T>,~\hilight{olivegreen}{Chan}<T>)} \\
 					\texttt{} \\
-					\texttt{\hilight{brown}{fn}~recv<T:~\hilight{olivegreen}{Send}>(port:~\&\hilight{olivegreen}{Port}<T>)~->~T} \\ 
+					\texttt{\hilight{brown}{fn}~recv<T:~\hilight{violet}{Send}>(port:~\&\hilight{olivegreen}{Port}<T>)~->~T} \\ 
 					\texttt{} \\
-					\texttt{\hilight{brown}{fn}~send<T:~\hilight{olivegreen}{Send}>(chan:~\&\hilight{olivegreen}{Chan}<T>,~data:~T)} \\
+					\texttt{\hilight{brown}{fn}~send<T:~\hilight{violet}{Send}>(chan:~\&\hilight{olivegreen}{Chan}<T>,~data:~T)} \\
 
 			\end{tabular}
-			\item \texttt{\hilight{olivegreen}{Send}} bound ensures once data is sent, sender can no longer access it.
-		\end{itemize}
 			\pause
 			\linegap
-		\item Shared state across tasks:
+			\item \texttt{\hilight{violet}{Send}} bound ensures once data is sent, sender can no longer access it.
+		\end{itemize}
+\end{frame}
+\begin{frame}{What are kind bounds good for?}
+		Shared state across tasks:
 		\begin{itemize}
 			\item \begin{tabular}{l}
-					\texttt{\hilight{brown}{fn}~\hilight{olivegreen}{Arc}<T:~\hilight{olivegreen}{Send}+\hilight{olivegreen}{Freeze}>(data:~T)~->~\hilight{olivegreen}{Arc}<T>} \\
+					\texttt{\hilight{brown}{fn}~arc<T:~\hilight{violet}{Send}+\hilight{violet}{Freeze}>(data:~T)~->~\hilight{olivegreen}{Arc}<T>} \\
+					\\
+					\texttt{\hilight{brown}{fn}~clone<T:~\hilight{violet}{Send}+\hilight{violet}{Freeze}>(arc:~\&\hilight{olivegreen}{Arc}<T>)~->~\hilight{olivegreen}{Arc}<T>} \\
 					\texttt{} \\
-					\texttt{\hilight{brown}{fn}~get<'a,~T:~\hilight{olivegreen}{Send}+\hilight{olivegreen}{Freeze}>(arc:~\&'a~\hilight{olivegreen}{Arc}<T>)~->~\&'a~T} \\
-					\\
-					\texttt{\hilight{brown}{fn}~clone<T:~\hilight{olivegreen}{Send}+\hilight{olivegreen}{Freeze}>(arc:~\&\hilight{olivegreen}{Arc}<T>)~->~\hilight{olivegreen}{Arc}<T>} \\
+					\texttt{\hilight{darkcyan}{//~note~immutable~pointer~to~T}} \\
+					\texttt{\hilight{brown}{fn}~get<'a,~T:~\hilight{violet}{Send}+\hilight{violet}{Freeze}>(arc:~\&'a~\hilight{olivegreen}{Arc}<T>)~->~\&'a~T} \\
 
 			\end{tabular}
-		\end{itemize}
+			\pause
+			\linegap
+		\item \texttt{\hilight{violet}{Send}} bound ensures only way to access data is through the \texttt{\hilight{olivegreen}{Arc}}
+		\item \texttt{\hilight{violet}{Freeze}} bound ensures data is immutable
 		\end{itemize}
@@
 
-					\texttt{~~~~\hilight{brown}{fn}~name(\&self)~->~\textasciitilde{}str} \\
-					\texttt{~~~~\hilight{brown}{fn}~num\_legs(\&self)~->~int} \\
+							\texttt{~~~~\hilight{brown}{fn}~name(\&self)~->~\textasciitilde{}\hilight{olivegreen}{str}} \\
+							\texttt{~~~~\hilight{brown}{fn}~num\_legs(\&self)~->~\hilight{olivegreen}{int}} \\
+							\texttt{~~~~\hilight{brown}{fn}~is\_well\_behaved(\&self)~->~\hilight{olivegreen}{bool}} \\
 					\texttt{\}} \\
@@
 				\item \begin{tabular}{l}
-						\texttt{struct~\hilight{olivegreen}{Dog}~\{~name:~\textasciitilde{}str,~bark\_decibels:~int~\}} \\
+						\texttt{struct~\hilight{olivegreen}{Dog}~\{~name:~\textasciitilde{}\hilight{olivegreen}{str},~bark\_decibels:~\hilight{olivegreen}{int}~\}} \\
 					\texttt{} \\
@@
 			\begin{itemize}
-				\item {\tt \textasciitilde{}Trait} can only contain {\tt \hilight{olivegreen}{Send}}able values
-				\item {\tt @Trait} can only contain {\tt \hilight{olivegreen}{Static}} values
-				\item {\tt \&Trait} can contain any values
+				\item {\tt \textasciitilde{}\hilight{olivegreen}{Trait}} can only contain {\tt \hilight{violet}{Send}}able values
+				\item {\tt @\hilight{olivegreen}{Trait}} can only contain {\tt \hilight{violet}{Static}} values
+				\item {\tt \&\hilight{olivegreen}{Trait}} can contain any values
 				\pause
-				\item Traits are not {\tt \hilight{olivegreen}{Freeze}}able or {\tt \hilight{olivegreen}{Copy}}able.
+				\item Traits are not {\tt \hilight{violet}{Freeze}}able or {\tt \hilight{violet}{Copy}}able.
 			\end{itemize}
@@
 	\begin{itemize}
-		\item Today, \texttt{\textasciitilde{}\hilight{olivegreen}{Pet}:\hilight{olivegreen}{Copy}} is also a type.
+		\item Today, \texttt{\textasciitilde{}\hilight{olivegreen}{Pet}:\hilight{violet}{Copy}} is also a type.
 
@@
 				\texttt{} \\
-			\texttt{\hilight{brown}{let}~copyable\_pet:~\&\hilight{olivegreen}{Pet}:\hilight{olivegreen}{Copy}~=} \\
-				\texttt{~~~~~~~~\&noncopyable\_alan\_turing~\hilight{brown}{as}~\&\hilight{olivegreen}{Pet}:\hilight{olivegreen}{Copy};} \\
+			\texttt{\hilight{brown}{let}~copyable\_pet:~\&\hilight{olivegreen}{Pet}:\hilight{violet}{Copy}~=} \\
+				\texttt{~~~~~~~~\&noncopyable\_alan\_turing~\hilight{brown}{as}~\&\hilight{olivegreen}{Pet}:\hilight{violet}{Copy};} \\
 				\texttt{~~~~\hilight{darkcyan}{//~\^~{error}:~cannot~capture~noncopyable}} \\
@@
 			\item \begin{tabular}{l}
-					\texttt{\hilight{brown}{let}~alan\_turing~=~\hilight{olivegreen}{Dog}~\{~...~\}} \\
-					\texttt{\hilight{brown}{let}~ada\_lovelace~=~\hilight{olivegreen}{Cat}~\{~...~\}} \\
-					\texttt{\hilight{brown}{let}~albert\_einstein~=~\hilight{olivegreen}{Bird}~\{~...~\}} \\
+					\texttt{\hilight{brown}{let}~alan\_turing~=~\textasciitilde{}\hilight{olivegreen}{Dog}~\{~...~\}} \\
+					\texttt{\hilight{brown}{let}~ada\_lovelace~=~\textasciitilde{}\hilight{olivegreen}{Cat}~\{~...~\}} \\
+					\texttt{\hilight{brown}{let}~albert\_einstein~=~\textasciitilde{}\hilight{olivegreen}{Bird}~\{~...~\}} \\
 					\pause
 					\texttt{}\\
-					\texttt{\hilight{brown}{let}~pets~=~\textasciitilde{}[alan\_turing~~~~~\hilight{brown}{as}~\hilight{olivegreen}{Dog}:~~\hilight{olivegreen}{Send}+\hilight{olivegreen}{Freeze},} \\
-						\texttt{~~~~~~~~~~~~~ada\_lovelace~~~~\hilight{brown}{as}~\hilight{olivegreen}{Cat}:~~\hilight{olivegreen}{Send}+\hilight{olivegreen}{Freeze},} \\
-					\texttt{~~~~~~~~~~~~~albert\_einstein~\hilight{brown}{as}~\hilight{olivegreen}{Bird}:~\hilight{olivegreen}{Send}+\hilight{olivegreen}{Freeze}];} \\
+					\texttt{\hilight{brown}{let}~pets~=~\textasciitilde{}[alan\_turing~~~~~\hilight{brown}{as}~\textasciitilde{}\hilight{olivegreen}{Pet}:~\hilight{violet}{Send}+\hilight{violet}{Freeze},} \\
+						\texttt{~~~~~~~~~~~~~ada\_lovelace~~~~\hilight{brown}{as}~\textasciitilde{}\hilight{olivegreen}{Pet}:~\hilight{violet}{Send}+\hilight{violet}{Freeze},} \\
+					\texttt{~~~~~~~~~~~~~albert\_einstein~\hilight{brown}{as}~\textasciitilde{}\hilight{olivegreen}{Pet}:~\hilight{violet}{Send}+\hilight{violet}{Freeze}];} \\
 					\texttt{} \\
@@
 
+\begin{frame}{Bonus slides}
+	\begin{center}
+	\large \textit{(I'm glad you asked that\dots)}
+	\end{center}
+\end{frame}
+
+\begin{frame}{Sharing Mutable State}
+	\textbf{``RW-Arc'':} reader-/writer-locked atomically-refcounted shared state.
+	\code {
+		\texttt{\hilight{brown}{impl}~<T:~\hilight{violet}{Freeze}+\hilight{violet}{Send}>~\&\hilight{olivegreen}{RWArc}<T>~\{} \\
+\texttt{~~~~\hilight{brown}{fn}~\hilight{olivegreen}{read}~~(blk:~\hilight{brown}{fn}(\&T))~...} \\
+\texttt{~~~~\hilight{brown}{fn}~\hilight{olivegreen}{write}~(blk:~\hilight{brown}{fn}(\&\hilight{brown}{mut}~T))~...} \\
+%\texttt{~~~~\hilight{brown}{fn}~\hilight{olivegreen}{write\_cond}(blk:~\hilight{brown}{fn}(\&\hilight{brown}{mut}~T,~\&condvar));} \\
+\texttt{\}} \\
+	}
+%		impl <T: const send> &rw_arc<T> {
+%		    fn read      (blk: fn(&T));
+%		    fn write     (blk: fn(&mut T));
+%		    fn write_cond(blk: fn(&mut T, &condvar));
+%		}
+
+	\begin{center}
+	\begin{tabular}{cc}
+		\begin{tabular}{l}
+%\texttt{\hilight{brown}{do}~arc.write\_cond~|state,c|~\{} \\
+\texttt{\hilight{brown}{do}~arc.write~|state:~\&\hilight{brown}{mut}~T|~\{} \\
+\texttt{~~~~\hilight{darkcyan}{//~exclusive~access;}} \\
+%\texttt{~~~~c.wait();} \\
+\texttt{~~~~\hilight{darkcyan}{//~can~modify~the~data}} \\
+\texttt{~~~~*state~=~\hilight{brickred}{17};} \\
+%\texttt{~~~~c.signal();} \\
+\texttt{\}} \\
+		\end{tabular}
+	&
+%		do arc.write_cond |state, cond| {
+%		    // exclusive access
+%		    cond.wait();    // can block...
+%		    *state = 17;    // modify the data...
+%		    cond.signal();  // and wake others
+%		}
+		\begin{tabular}{l}
+\texttt{\hilight{brown}{do}~arc.read~|state:~\&T|~\{} \\
+\texttt{~~~~\hilight{darkcyan}{//~concurrent~access}} \\
+\texttt{~~~~\hilight{darkcyan}{//~immutable~state}} \\
+\texttt{~~~~\hilight{red}{assert}~*state~==~\hilight{brickred}{17};} \\
+\texttt{~~~~\hilight{darkcyan}{//~won't~compile}} \\
+\texttt{~~~~\hilight{darkcyan}{//~*state += 1;}} \\
+\texttt{\}} \\
+		\end{tabular}
+	\end{tabular}
+	\end{center}
+%		do arc.read |state| {
+%		    // shared access - immutable state
+%		    assert *state == 17;  // OK
+%		    *state += 1;  // Won't compile
+%		}
+\end{frame}
+
 \end{document}

### 2013-12-16 15869/paper.tex
Add paper

\documentclass{article}
\usepackage{lmodern}
\usepackage[charter]{mathdesign}
\usepackage[english]{babel}
\usepackage[utf8]{inputenc}

\usepackage{fullpage}
\usepackage{manfnt}
\usepackage{wasysym}
\usepackage{listings}
\usepackage{graphicx}
\usepackage{url}
%\usepackage{ulem}
\usepackage{marvosym}
%\usepackage{skull}
\usepackage{proof}
\usepackage{array}

\title{{\bf Dynamic Scheduling in Halide}}
\author{Ben Blum \texttt{(bblum@andrew.cmu.edu)}}

\date{2013, December 16}

\newcommand\noob{\mathsf{noob}}
\newcommand\gibs{\mathsf{gibs}}
\newcommand\dps{\mathsf{dps}}
\newcommand\squig\rightsquigarrow
\newcommand\Coloneqq{\mathrel{\mathop{::}}=}
\newcommand\dmg{\text{\Laserbeam}}
\newcommand\delter\delta
\newcommand\alpher\alpha
\newcommand\defnor{\text{ }|\text{ }}

\newcommand\pimp{\mathop{\supset}}
\newcommand\pand{\mathop{\wedge}}
\newcommand\por{\mathop{\vee}}
\newcommand\ptrue{\top}
\newcommand\pfalse{\bot}

\begin{document}
\maketitle

\begin{figure}[h]
	\begin{center}
	\begin{tabular}{llllll}
		\begin{tabular}{l}
                \texttt{blur\_y(x,y) =} \\
                \texttt{~~~~(input(x, y-1) +}\\
                \texttt{~~~~~input(x, y~~) +}\\
                \texttt{~~~~~input(x, y+1))/3;}\\
                \\
                \texttt{blur\_x(x,y) =} \\
                \texttt{~~~~(blur\_y(x-1, y) +}\\
                \texttt{~~~~~blur\_y(x,~~~y) +}\\
                \texttt{~~~~~blur\_y(x+1, y))/3;}\\
		\end{tabular}
		& & & & &
		\begin{tabular}{l}
                \texttt{blur\_y\_far\_away(x,y) =} \\
                \texttt{~~~~(input(x, {\bf y-rand()}) +}\\
                \texttt{~~~~~input(x, y~~~~~~~) +}\\
                \texttt{~~~~~input(x, {\bf y+rand()}))/3;}\\
                \\
                \texttt{blur\_x\_far\_away(x,y) =} \\
                \texttt{~~~~(blur\_y({\bf x-rand()}, y) +}\\
                \texttt{~~~~~blur\_y(x,~~~~~~~~y) +}\\
                \texttt{~~~~~blur\_y({\bf x+rand()}, y))/3;}\\
		\end{tabular}

	\end{tabular}
	\end{center}
	\caption{Static and dynamic data dependencies in Halide. On the left, a 3x3 blur (the canonical example Halide program) need only reference input data up to 1 pixel away in each direction for each output pixel. On the right, the algorithm is modified such that the compiler cannot statically know which input pixels each output pixel may reference.}
	\label{fig:intro}
\end{figure}

\section{Introduction}

Halide~\cite{halide} is a programming language for image processing pipelines designed to separate the concerns of performance and expressivity. The programming model is functional, representing images as functions from pixel coordinates to colour values, and representing each pass of an image processing algorithm as a function from images to images. Once implemented, the programmer's algorithm is then automatically parallelized by the Halide runtime, taking into account considerations such as how to split each pass into parallel tiles and the data dependencies between two consecutive passes.
The structure of the generated loop nest and the function calls therein is known as the {\em function schedule}.

One weakness of Halide is its inability to express image processing passes that don't operate uniformly across their domain; i.e., algorithms whose data dependencies on portions of previous passes cannot be known statically. 
Figure~\ref{fig:intro} above contrasts two simple example algorithms, showcasing the difference between static and dynamic data dependencies.

In this project I investigate and implement {\em dynamic function scheduling}, a new language feature for Halide to support such algorithms.
My approach allows the computations for early pipeline stages to be cached across loop iterations of later stages, so that the results may be reused when data dependencies happen to coincide on the same location.
The principal contribution of this work is the implementation and performance evaluation of {\em dynamic scheduling at pixel granularity}. I also investigated dynamic scheduling at coarser (e.g. tile) granularities, but the problem was more difficult than anticipated; for this I provide some discussion and a design sketch for future work.

\section{Motivation}
% TODO

\section{Design and Interface}
% TODO

\section{Implementation}

\subsection{Code Generation}
% TODO

\subsection{Legality Checking}
% TODO

\subsection{Code}

The reader should feel free to find and peruse my code at \texttt{https://github.com/bblum/Halide}. During the weeks following this submission I will polish it and submit a pull request upstream.

\section{Evaluation}
% TODO

\section{Open Questions}
% TODO

\section{Conclusion}
% TODO

\section*{Acknowledgements}

Thanks to Jonathan Ragan-Kelley and Andrew Adams, developers of Halide, for their guidance, insights, and patience. Thanks to Kayvon Fatahalian for supervising this project, and for a great semester in 15-869.

\bibliography{citations}{}
\bibliographystyle{alpha}
\end{document}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% vim: foldmethod=indent

### 2013-12-17 15869/paper.tex
finish talk and start paper

@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Introduction}
@@
 In this project I investigate and implement {\em dynamic function scheduling}, a new language feature for Halide to support such algorithms.
-My approach allows the computations for early pipeline stages to be cached across loop iterations of later stages, so that the results may be reused when data dependencies happen to coincide on the same location.
+My approach allows the computations for early pipeline stages to be cached across loop iterations of later stages, so that the results may be reused when data dependencies happen to coincide on the same location, saving redundant computation of a possibly expensive intermediate function.
 The principal contribution of this work is the implementation and performance evaluation of {\em dynamic scheduling at pixel granularity}.
 I also investigated dynamic scheduling at coarser (e.g.
-tile) granularities, but the problem was more difficult than anticipated; for this I provide some discussion and a design sketch for future work.
+tile) granularities, with which entire loop iterations may be skipped.
+However, the problem was more difficult than anticipated, so I instead provide some discussion and a design sketch for future work.
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Motivation}
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Design and Interface}
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Implementation}
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Evaluation}
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Open Questions}
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Conclusion}
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section*{Acknowledgements}
@@
 Thanks to Jonathan Ragan-Kelley and Andrew Adams, developers of Halide, for their guidance, insights, and patience.
-Thanks to Kayvon Fatahalian for supervising this project, and for a great semester in 15-869.
+Thanks to Kayvon Fatahalian for supervising this project and for a great semester in 15-869.
 

### 2013-12-17 15869/paper.tex
motivation

@@
 
+\usepackage{caption,subcaption}
 \usepackage{fullpage}
+\usepackage{color}
 \usepackage{manfnt}
@@
 \section{Motivation}
-% TODO
+
+\definecolor{grey}{RGB}{127,127,127}
+\definecolor{darkcyan}{RGB}{0,127,127}
+\definecolor{olivegreen}{RGB}{0,127,0}
+\definecolor{violet}{RGB}{127,0,127}
+\definecolor{brickred}{RGB}{127,0,0}
+\definecolor{brown}{RGB}{127,63,0}
+\definecolor{orange}{RGB}{192,96,0}
+\newcommand\hilight[2]{\color{#1}#2\color{black}}
+
+\begin{figure}[t]
+	\begin{center}
+	\begin{subfigure}[b]{0.45\textwidth}
+		\begin{center}
+		\begin{tabular}{l}
+		{\em User-provided schedule:} \\
+		\\
+		\hilight{blue}{\texttt{blur\_x /* nothing special */;}} \\
+		\hilight{olivegreen}{\texttt{blur\_y.store\_root().compute\_root();}} \\
+		\\
+		{\em Compiler-generated code:} \\
+		\\
+                \texttt{\hilight{olivegreen}{float~blur\_y[HEIGHT][WIDTH];}} \\
+                \texttt{\hilight{olivegreen}{for~(row~=~0~to~HEIGHT)~\{}} \\
+                \texttt{\hilight{olivegreen}{~~for~(col~=~0~to~WIDTH)~\{}} \\
+                \texttt{\hilight{olivegreen}{~~~~blur\_y[row][col] = ...;}}\\
+                \texttt{\hilight{olivegreen}{~~\}}} \\
+                \texttt{\hilight{olivegreen}{\}}} \\
+		\\
+                \texttt{\hilight{blue}{float~blur\_x[HEIGHT][WIDTH];}} \\
+                \texttt{\hilight{blue}{for~(row~=~0~to~HEIGHT)~\{}} \\
+                \texttt{\hilight{blue}{~~for~(col~=~0~to~WIDTH)~\{}} \\
+                \texttt{\hilight{blue}{~~~~blur\_x[row][col] = ...;}}\\
+                \texttt{\hilight{blue}{~~\}}} \\
+                \texttt{\hilight{blue}{\}}} \\
+		\end{tabular}
+		\end{center}
+		\caption{\texttt{blur\_x} and \texttt{blur\_y} are scheduled separately.
+This schedule, while easy to understand, offers poor locality (values computed are not used until a full image traversal later) and poor parallelism (creating one thread per row or column of pixels both floods the system with too many threads and creates cache contention).}
+		\label{fig:scheds-a}
+	\end{subfigure}
+	\qquad
+	\begin{subfigure}[b]{0.5\textwidth}
+		\begin{center}
+		\begin{tabular}{l}
+		{\em User-provided schedule:} \\
+		\\
+                \hilight{blue}{\texttt{blur\_x.tile(x, y, x2, y2, 8, 8).parallel(y);}}\\
+                \hilight{olivegreen}{\texttt{blur\_y.store\_at(blur\_x, x).compute\_at(blur\_x, y2);}}\\
+                \\
+		{\em Compiler-generated code:} \\
+		\\
+                \texttt{\hilight{blue}{float~blur\_x[HEIGHT][WIDTH];}} \\
+                \texttt{\hilight{blue}{PARALLEL for~(row~=~0~to~8)~\{}} \\
+                \texttt{\hilight{blue}{~~for~(col~=~0~to~8)~\{}} \\
+                \texttt{\hilight{olivegreen}{~~~~float~blur\_y[HEIGHT/8 + 2][WIDTH/8 + 2];}} \\
+                \texttt{\hilight{blue}{~~~~for~(row2~=~0~to~WIDTH/8 + 2)~\{}} \\
+                \texttt{\hilight{olivegreen}{~~~~~~for~(col2~=~0~to~WIDTH/8 + 2)~\{}} \\
+                \texttt{\hilight{olivegreen}{~~~~~~~~blur\_y[row2][col2]~=~...;}} \\
+                \texttt{\hilight{olivegreen}{~~~~~~\}}} \\
+                \texttt{\hilight{blue}{~~~~~~for~(col2~=~0~to~WIDTH/8)~\{}} \\
+                \texttt{\hilight{blue}{~~~~~~~~blur\_x[row*8~+~row2][col*8~+~col2]~=~...;}} \\
+                \texttt{\hilight{blue}{~~~~~~\}}} \\
+                \texttt{\hilight{blue}{~~~~\}}} \\
+                \texttt{\hilight{blue}{~~\}}} \\
+                \texttt{\hilight{blue}{\}}} \\
+		\end{tabular}
+		\end{center}
+		\caption{\texttt{blur\_x} is tiled in 8-fold splits in each dimension, and \texttt{blur\_y} is scheduled at the tile level within \texttt{blur\_x}.
+This schedule offers good locality (values are used soon after being computed) and good parallelism (one thread per tile; tile size can be customized).}
+		\label{fig:scheds-b}
+	\end{subfigure}
+	\end{center}
+	\caption{Two example function schedules for the two-stage blur algorithm shown in Figure~\ref{fig:intro}.}
+	\label{fig:scheds}
+\end{figure}
+
+The high concept of Halide is to separate the logic of an image processing algorithm from the order in which the functions for each stage are scheduled.
+Figure~\ref{fig:scheds} shows two possible schedules for our example algorithm, and discusses the performance characteristics of each.
+Though Figure~\ref{fig:scheds-b} is more cache-friendly and offers superior paralellism opportunity, a notable disadvantage is that some computation is duplicated across tiles.
+This is shown in the extra 2 pixels computed for \texttt{blur\_y} in each dimension.
+In this example, the overhead is minor, as Halide can statically reason about the maximum extra ``padding'' that \texttt{blur\_y} must compute for each tile of \texttt{blur\_x}.
+
+However, this extra computation can become a significant problem when the dependencies between the algorithm's stages are computed at runtime and hence are not available for compile-time reasoning, as in the PatchMatch algorithm~\cite{patchmatch} and in edge detection.
+If the user wishes to use a cache-friendly, highly-parallel function schedule, as in Figure~\ref{fig:scheds-b}, Halide must make a conservative estimate regarding how much extra padding is required, resulting in extra work duplication.
+
+To get a sense for how much extra unnecessary computation can arise, we introduce a sample two-stage Halide program in which the second stage (\texttt{flip()}) dynamically computes which pixels it requires from the first stage (\texttt{invert()}).
+This program is described in Figure~\ref{fig:flip}.
+We then ran this program on a test image using the simple schedule from Figure~\ref{fig:scheds-a} and using several nested schedules as in Figure~\ref{fig:scheds-b}.
+We found that in the former situation, 13,068 calls to \texttt{invert()} were required, while in the latter, \texttt{invert()} was invoked anywhere between 252,648 to 7,847,400 times depending on the schedule used.
+Hence, the motivation for {\em dynamic scheduling} is to extend Halide's ability to avoid duplicate work to apply to this wider range of algorithms.
+
+\begin{figure}[t]
+	\begin{tabular}{cc}
+	\begin{subfigure}[b]{0.45\textwidth}
+		\begin{center}
+		\includegraphics[width=0.85\textwidth]{canoe-0.png}
+		\end{center}
+		\caption{Original input image.}
+	\end{subfigure} &
+	\begin{subfigure}[b]{0.45\textwidth}
+		\begin{center}
+		\includegraphics[width=0.85\textwidth]{canoe-1.png}
+		\end{center}
+		\caption{Stage 1 is a simple value-invert of each input pixel.}
+		\label{fig:invert}
+	\end{subfigure} \\
+	& \\
+	\begin{subfigure}[b]{0.45\textwidth}
+		\begin{center}
+		\includegraphics[width=0.85\textwidth]{canoe-2.png}
+		\end{center}
+		\caption{The core of stage 2 is a transformation across a vector field, which maps points near the center to the corners and points near the corners to the center, followed by a 3x3 blur.
+The blur causes each output pixel to depend on multiple input pixels, which allows for the possibility of work duplication.}
+		\label{fig:flip-blur-1}
+	\end{subfigure} &
+	\begin{subfigure}[b]{0.45\textwidth}
+		\begin{center}
+		\includegraphics[width=0.85\textwidth]{canoe-3.png}
+		\end{center}
+		\caption{Final output.
+Stage 2 is altered to only apply the vector field to pixels when their colour value is darker than a certain threshold.
+This condition, which depends on a nondeterministic property of the input image, prevents Halide's optimizations from reasoning about the data dependencies between the two stages.}
+		\label{fig:flip-blur-2}
+	\end{subfigure}
+	\end{tabular}
+	\caption{Our sample image processing program consists of two stages, \texttt{invert()} (shown in \ref{fig:invert}) and \texttt{flip()} (shown in \ref{fig:flip-blur-1} and \ref{fig:flip-blur-2}).
+We designed this program expressly to create dynamic data dependencies with wide ranges across the image, to ensure a static function schedule could not avoid duplicating work.}
+
+	\label{fig:flip}
+\end{figure}
 

### 2013-12-17 15869/paper.tex
write about design

@@
 
-To get a sense for how much extra unnecessary computation can arise, we introduce a sample two-stage Halide program in which the second stage (\texttt{flip()}) dynamically computes which pixels it requires from the first stage (\texttt{invert()}).
-This program is described in Figure~\ref{fig:flip}.
-We then ran this program on a test image using the simple schedule from Figure~\ref{fig:scheds-a} and using several nested schedules as in Figure~\ref{fig:scheds-b}.
-We found that in the former situation, 13,068 calls to \texttt{invert()} were required, while in the latter, \texttt{invert()} was invoked anywhere between 252,648 to 7,847,400 times depending on the schedule used.
+To get a sense for how much extra unnecessary computation can arise, I introduce a sample two-stage Halide program in which the second stage (\texttt{flip()}) dynamically computes which pixels it requires from the first stage (\texttt{invert()}).
+This program is described in Figures~\ref{fig:flip} and~\ref{fig:flip-code}.
+I then ran this program on a test image using the simple schedule from Figure~\ref{fig:scheds-a} and using several nested schedules as in Figure~\ref{fig:scheds-b}.
+I found that in the former situation, 13,068 calls to \texttt{invert()} were required, while in the latter, \texttt{invert()} was invoked anywhere between 252,648 to 7,847,400 times depending on the schedule used.
 Hence, the motivation for {\em dynamic scheduling} is to extend Halide's ability to avoid duplicate work to apply to this wider range of algorithms.
@@
 	\caption{Our sample image processing program consists of two stages, \texttt{invert()} (shown in \ref{fig:invert}) and \texttt{flip()} (shown in \ref{fig:flip-blur-1} and \ref{fig:flip-blur-2}).
-We designed this program expressly to create dynamic data dependencies with wide ranges across the image, to ensure a static function schedule could not avoid duplicating work.}
+I designed this program expressly to create dynamic data dependencies with wide ranges across the image, to ensure a static function schedule could not avoid duplicating work.}
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Design and Interface}
-% TODO
+\begin{figure}[t]
+	\begin{center}
+	\begin{tabular}{l}
+\texttt{\hilight{olivegreen}{Var}~x(\hilight{brickred}{"column"}), y(\hilight{brickred}{"row"}), c(\hilight{brickred}{"colour"});} \\
+\\
+\texttt{\hilight{darkcyan}{/* Stage 1 */}} \\
+\texttt{\hilight{olivegreen}{Func}~invert(\hilight{olivegreen}{Func}~input)~\{} \\
+\texttt{~~~~\hilight{olivegreen}{Func}~f(\hilight{brickred}{"invert"});} \\
+\texttt{~~~~f(x,~y,~c)~=~\hilight{brickred}{1}~-~input(x,~y,~c);~} \\
+\texttt{~~~~\hilight{brown}{return}~f;} \\
+\texttt{\}} \\
+\\
+\texttt{\hilight{darkcyan}{/* The vector field used in stage 2 */}} \\
+\texttt{\#define~vf\_x(image,~do\_flip,~x)~(\hilight{brown}{select}(do\_flip,~x,~((\hilight{brickred}{3}~*~image.width()~~/~\hilight{brickred}{2})~-~x)))} \\
+\texttt{\#define~vf\_y(image,~do\_flip,~y)~(\hilight{brown}{select}(do\_flip,~y,~((\hilight{brickred}{3}~*~image.height()~/~\hilight{brickred}{2})~-~y)))} \\
+\texttt{} \\
+\texttt{\hilight{darkcyan}{/* Stage 2 */}} \\
+\texttt{\hilight{olivegreen}{Func}~flip(\hilight{olivegreen}{Image}<\hilight{olivegreen}{float}>~\&i,~\hilight{olivegreen}{Func}~input)~\{} \\
+\texttt{~~~~\hilight{olivegreen}{Expr}~do\_flip~=~input(x,~y,~c)~>~\hilight{brickred}{0.5f};} \\
+\texttt{~~~~\hilight{olivegreen}{Func}~f(\hilight{brickred}{"flip"});} \\
+\texttt{~~~~f(x,~y,~c)~=} \\
+\texttt{~~~~~~~~(input(vf\_x(i,~do\_flip,~x-\hilight{brickred}{1}),~vf\_y(i,~do\_flip,~y-\hilight{brickred}{1}),~c)~+} \\
+\texttt{~~~~~~~~~\hilight{darkcyan}{/* ... intermediate coordinates of 3x3 blur omitted ... */}} \\
+\texttt{~~~~~~~~~input(vf\_x(i,~do\_flip,~x+\hilight{brickred}{1}),~vf\_y(i,~do\_flip,~y+\hilight{brickred}{1}),~c))~/~\hilight{brickred}{9};} \\
+\texttt{~~~~\hilight{brown}{return}~f;} \\
+\texttt{\}} \\
+
+	\end{tabular}
+	\end{center}
+	\caption{Selected code from the implementation of our sample image processing program.}
+	\label{fig:flip-code}
+\end{figure}
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Implementation}
+\section{Design and Implementation}
+
+\subsection{Overview}
+
+Conceptually, my goal for dynamic function scheduling as a Halide language feature was as follows.
+In addition to the existing directives \texttt{compute\_at()}, \texttt{store\_at()}, and so on which users may call on their \texttt{Func} objects, supplying appropriate loop variable arguments, I would add a new directive \texttt{compute\_lazy()}.
+
+{\bf Pixel-granularity dynamic scheduling.} Used with no arguments, this directive would instruct Halide to allocate a ``bitmask'', an array of booleans with the same dimensions as the array used to store the function's results themselves.
+Halide would then surround any expression which computed a value to store in the array at a certain index with a check to this bitmask at the same index, which would cause the expression to be bypassed if a value for that index had been computed before.
+I refer to this as dynamic scheduling at {\em pixel granularity} because there is one bitmask check per expression that would compute a value for a single pixel.
+In Figure~\ref{fig:dynamic-pixel} I show sample code that Halide would generate for a program that used pixel-level dynamic scheduling.
+
+% TODO
+{\bf Tile-granularity dynamic scheduling.} DISCUSSION FORTHCOMING.
+
+\begin{figure}[t]
+	\begin{center}
+	\begin{subfigure}[b]{0.475\textwidth}
+		\begin{center}
+		\begin{tabular}{l}
+		{\em User-provided schedule:} \\
+		\\
+                \hilight{blue}{\texttt{flip.tile(col, row, col2, row2, 8, 8);}} \\
+                \hilight{olivegreen}{\texttt{invert.store\_at(flip, col);}} \\
+		\hilight{olivegreen}{\texttt{invert.compute\_at(invert, row2);}}\\
+                {\bf \hilight{orange}{\texttt{invert.compute\_lazy();}}} \\
+                \\
+		{\em Compiler-generated code:} \\
+		\\
+                \texttt{\hilight{blue}{float~flip[HEIGHT][WIDTH];}} \\
+                \texttt{\hilight{blue}{for~(row~=~0~to~8)~\{}} \\
+                \texttt{\hilight{blue}{~~for~(col~=~0~to~8)~\{}} \\
+                \texttt{\hilight{olivegreen}{~~~~float~invert[...][...];}} \\
+		{\bf \texttt{\hilight{orange}{~~~~bool~result\_computed[...][...];}}} \\
+                \texttt{\hilight{blue}{~~~~for~(row2~=~0~to~WIDTH/8)~\{}} \\
+                \texttt{\hilight{olivegreen}{~~~~~~let min\_row = ...;}} \\
+                \texttt{\hilight{olivegreen}{~~~~~~let max\_row = ...;}} \\
+                \texttt{\hilight{olivegreen}{~~~~~~let min\_col = ...;}} \\
+                \texttt{\hilight{olivegreen}{~~~~~~let max\_col = ...;}} \\
+                \texttt{\hilight{olivegreen}{~~~~~~for~(row\_i~=~min\_row~to~max\_row)~\{}} \\
+                \texttt{\hilight{olivegreen}{~~~~~~~~for~(col\_i~=~min\_col~to~max\_col)~\{}} \\
+		{\bf \texttt{\hilight{orange}{~~~~~~~~~~if (!result\_computed[...][...]) \{}}} \\
+                \texttt{\hilight{olivegreen}{~~~~~~~~~~~~invert[row\_i][col\_i]~=~...;}} \\
+		{\bf \texttt{\hilight{orange}{~~~~~~~~~~~~result\_computed[...][...]~=~TRUE;}}} \\
+		{\bf \texttt{\hilight{orange}{~~~~~~~~~~\}}}} \\
+                \texttt{\hilight{olivegreen}{~~~~~~~~\}}} \\
+                \texttt{\hilight{olivegreen}{~~~~~~\}}} \\
+                \texttt{\hilight{blue}{~~~~~~for~(col2~=~0~to~WIDTH/8)~\{}} \\
+                \texttt{\hilight{blue}{~~~~~~~~flip[...][...]~=~...;}} \\
+                \texttt{\hilight{blue}{~~~~~~\}}} \\
+                \texttt{\hilight{blue}{~~~~\}}} \\
+                \texttt{\hilight{blue}{~~\}}} \\
+                \texttt{\hilight{blue}{\}}} \\
+		\end{tabular}
+		\end{center}
+		\caption{Dynamic scheduling at pixel granularity.
+The directive \texttt{invert.compute\_lazy()} with no arguments is shorthand for supplying \texttt{invert}'s innermost loop variable (in this case, \texttt{col\_i}).}
+		\label{fig:dynamic-pixel}
+	\end{subfigure}
+	\qquad
+	\begin{subfigure}[b]{0.475\textwidth}
+		\begin{center}
+		\begin{tabular}{l}
+		{\em User-provided schedule:} \\
+		\\
+                \hilight{blue}{\texttt{flip.tile(col, row, col2, row2, 8, 8);}} \\
+                \hilight{olivegreen}{\texttt{invert.store\_at(flip, col);}} \\
+		\hilight{olivegreen}{\texttt{invert.compute\_at(invert, row2);}}\\
+                {\bf \hilight{orange}{\texttt{invert.compute\_lazy(row\_i);}}} \\
+                \\
+		{\em Compiler-generated code:} \\
+		\\
+                \texttt{\hilight{blue}{float~flip[HEIGHT][WIDTH];}} \\
+                \texttt{\hilight{blue}{for~(row~=~0~to~8)~\{}} \\
+                \texttt{\hilight{blue}{~~for~(col~=~0~to~8)~\{}} \\
+                \texttt{\hilight{olivegreen}{~~~~float~invert[...][...];}} \\
+		{\bf \texttt{\hilight{orange}{~~~~bool~result\_computed[...][...];}}} \\
+                \texttt{\hilight{blue}{~~~~for~(row2~=~0~to~WIDTH/8)~\{}} \\
+                \texttt{\hilight{olivegreen}{~~~~~~let min\_row = ...;}} \\
+                \texttt{\hilight{olivegreen}{~~~~~~let max\_row = ...;}} \\
+                \texttt{\hilight{olivegreen}{~~~~~~let min\_col = ...;}} \\
+                \texttt{\hilight{olivegreen}{~~~~~~let max\_col = ...;}} \\
+                \texttt{\hilight{olivegreen}{~~~~~~for~(row\_i~=~min\_row~to~max\_row)~\{}} \\
+		{\bf \texttt{\hilight{orange}{~~~~~~~~if (!result\_computed[...][...]) \{}}} \\
+                \texttt{\hilight{olivegreen}{~~~~~~~~~~for~(col\_i~=~min\_col~to~max\_col)~\{}} \\
+                \texttt{\hilight{olivegreen}{~~~~~~~~~~~~invert[row\_i][col\_i]~=~...;}} \\
+                \texttt{\hilight{olivegreen}{~~~~~~~~~~\}}} \\
+		{\bf \texttt{\hilight{orange}{~~~~~~~~~~result\_computed[...][...]~=~TRUE;}}} \\
+		{\bf \texttt{\hilight{orange}{~~~~~~~~\}}}} \\
+                \texttt{\hilight{olivegreen}{~~~~~~\}}} \\
+                \texttt{\hilight{blue}{~~~~~~for~(col2~=~0~to~WIDTH/8)~\{}} \\
+		\texttt{\hilight{blue}{~~~~~~~~flip[...][...]~=~...;}} \\
+                \texttt{\hilight{blue}{~~~~~~\}}} \\
+                \texttt{\hilight{blue}{~~~~\}}} \\
+                \texttt{\hilight{blue}{~~\}}} \\
+                \texttt{\hilight{blue}{\}}} \\
+		\end{tabular}
+		\end{center}
+		\caption{Dynamic scheduling at tile granularity.
+Here, the ``tile'' is simply the row denited by the variable \texttt{row\_i}.
+Simply using a single boolean for each tile can be unsound in some cases, as discussed in Section~\ref{sec:future-tiles}.}
+		\label{fig:dynamic-tile}
+	\end{subfigure}
+	\end{center}
+	\caption{Generated code listings for dynamic function schedules, at both pixel and tile granularity.
+In addition to highlighting the code associated with the \hilight{blue}{flip}~and \hilight{olivegreen}{invert}~stages of the algorithm, I also {\bf \hilight{orange}{emphasize}} the additional code generated for dynamic scheduling.}
+	\label{fig:dynamic}
+\end{figure}
 
 \subsection{Code Generation}
-% TODO
+
+Inside the compiler, the dynamic schedule configuration is stored in a \texttt{Func}'s \texttt{Schedule} object as a \texttt{LoopLevel} named \texttt{lazy\_level}, alongside the existing configurations \texttt{store\_level} and \texttt{compute\_level}.
+Unlike the latter two, however, the \texttt{lazy\_level} is not defined in terms of the loop variable of a different function, but rather for one of the function's own loop variables.
+Hence, \texttt{lazy\_level.func} always refers to the containing function.
+
+The code generation itself occurs during the \texttt{Lower} phase of compilation, between the \texttt{schedule\_functions()} phase (found in \texttt{Lower.cpp}) and the \texttt{storage\_flattening()} phase (found in \texttt{StorageFlattening.cpp}).
+The former processes the \texttt{lazy\_level} configuration and generates IR to represent it, and the latter converts those IR nodes into conventional IR that can be code-generated.
+For this purpose I added a new type of IR node, \texttt{DynamicStmt} (found in \texttt{IR.h}), which represents the ``if not-in-cache then compute; mark-cache; end'' pattern.
+
+\begin{itemize}
+	\item {\bf Function Scheduling.} This phase converts the dynamic schedule configuration into IR in two places.
+The first is when it generates a \texttt{Realize} node (in \texttt{build\_realize()}); it simply sets the node's \texttt{lazy} flag to \texttt{true} if the associated function's \texttt{lazy\_level} was set.\footnote{
+	I experimented here with computing and storing an expression representing the necessary size for the bitmask, rather than just a boolean. Future work may find this approach worth doing after all.}
+	The second is when building a function's loop nest (in \texttt{build\_provide\_loop\_nest()}); when the approprate loop level is encountered, it inserts a \texttt{DynamicStmt} node between the \texttt{For} loop and its body.
+	\item {\bf Storage Flattening.} This phase visits the IR described above and ``flattens'' it into arrays more reminiscent of the C memory model.
+In \texttt{visit(Realize *)}, where it makes the \texttt{Allocate} node for the function's computed results, it now also will make another \texttt{Allocate} and a brief initialization loop if \texttt{realize->lazy} is set.
+Second, in \texttt{visit(DynamicStmt *)}, it converts the \texttt{DynamicStmt} node into a conventional \texttt{IfThenElse} node using the provided index for the allocated bitmask.
+\end{itemize}
 
 \subsection{Legality Checking}
-% TODO
+
+During the function scheduling in \texttt{Lower.cpp} Halide must also check that a specified dynamic schedule is legal. I currently check for the following cases:
+
+\begin{itemize}
+	\item {\bf Vectorized loop.} If any of the loops between the \texttt{store\_level} and the \texttt{lazy\_level} are vectorized, it may not make sense to emit a check which conditionally skips much of the computation.
+Though it would make sense to generate code for a single boolean check around a single pixel computation, with more sophisticated tile-level dynamic scheduling the vectorized execution might behave pathologically (especially if the conditional check involves traversing a sophisticated data structure as suggested in Section~\ref{sec:future-tiles}).
+	\item {\bf Parallel loop.} If any of the loops between the \texttt{store\_level} and the \texttt{lazy\_level} are parallelized, then one thread attempting to record the result of a computation might race with another thread attempting to use that result.
+With x86's sequentially-consistent memory model and the simple boolean check, this is not an issue, but when compiling for ARM or using a more sophisticated data structure for memoizing Halide would need to emit memory barriers at best and locks at worst.
+This is not infeasible to implement, but raises research questions of its own and so is left to future work.
+	\item {\bf Update phase.} Dynamic scheduling is not (yet) compatible with functions which have an update phase in their \texttt{Pipeline} node, which can arise when the user provides a reduction definition.
+This is merely a case I did not consider in detail, and may well be possible to handle.
+\end{itemize}
 
@@
 % TODO
+
+\label{sec:future-tiles}
 

### 2013-12-17 15869/paper.tex
almost finish paper

@@
 The high concept of Halide is to separate the logic of an image processing algorithm from the order in which the functions for each stage are scheduled.
-Figure~\ref{fig:scheds} shows two possible schedules for our example algorithm, and discusses the performance characteristics of each.
+Figure~\ref{fig:scheds} shows two possible schedules for my example algorithm, and discusses the performance characteristics of each.
 Though Figure~\ref{fig:scheds-b} is more cache-friendly and offers superior paralellism opportunity, a notable disadvantage is that some computation is duplicated across tiles.
@@
 	\end{center}
-	\caption{Selected code from the implementation of our sample image processing program.}
+	\caption{Selected code from the implementation of the sample image processing program.}
 	\label{fig:flip-code}
@@
 \section{Design and Implementation}
+
+% This section discusses how I organized my code to implement this project.
+The reader seeking a research perspective will likely be disinterested, while the reader seeking to extend or maintain my work may well find this the most useful section in the report.
 
@@
 {\bf Tile-granularity dynamic scheduling.} DISCUSSION FORTHCOMING.
+
+\subsection{Code Generation}
+
+Inside the compiler, the dynamic schedule configuration is stored in a \texttt{Func}'s \texttt{Schedule} object as a \texttt{LoopLevel} named \texttt{lazy\_level}, alongside the existing configurations \texttt{store\_level} and \texttt{compute\_level}.
+Unlike the latter two, however, the \texttt{lazy\_level} is not defined in terms of the loop variable of a different function, but rather for one of the function's own loop variables.
+Hence, \texttt{lazy\_level.func} always refers to the containing function.
+
+The code generation itself occurs during the \texttt{Lower} phase of compilation, between the \texttt{schedule\_functions()} phase (found in \texttt{Lower.cpp}) and the \texttt{storage\_flattening()} phase (found in \texttt{StorageFlattening.cpp}).
+The former processes the \texttt{lazy\_level} configuration and generates IR to represent it, and the latter converts those IR nodes into conventional IR that can be code-generated.
+For this purpose I added a new type of IR node, \texttt{DynamicStmt} (found in \texttt{IR.h}), which represents the ``if not-in-cache then compute; mark-cache; end'' pattern.
+
+\begin{itemize}
+	\item {\bf Function Scheduling.} This phase converts the dynamic schedule configuration into IR in two places.
+The first is when it generates a \texttt{Realize} node (in \texttt{build\_realize()}); it simply sets the node's \texttt{lazy} flag to \texttt{true} if the associated function's \texttt{lazy\_level} was set.\footnote{
+	I experimented here with computing and storing an expression representing the necessary size for the bitmask, rather than just a boolean. Future work may find this approach worth doing after all.}
+	The second is when building a function's loop nest (in \texttt{build\_provide\_loop\_nest()}); when the approprate loop level is encountered, it inserts a \texttt{DynamicStmt} node between the \texttt{For} loop and its body.
+	\item {\bf Storage Flattening.} This phase visits the IR described above and ``flattens'' it into arrays more reminiscent of the C memory model.
+In \texttt{visit(Realize *)}, where it makes the \texttt{Allocate} node for the function's computed results, it now also will make another \texttt{Allocate} and a brief initialization loop if \texttt{realize->lazy} is set.
+Second, in \texttt{visit(DynamicStmt *)}, it converts the \texttt{DynamicStmt} node into a conventional \texttt{IfThenElse} node using the provided index for the allocated bitmask.
+\end{itemize}
+
+\subsection{Legality Checking}
+
+During the function scheduling in \texttt{Lower.cpp} Halide must also check that a specified dynamic schedule is legal. I currently check for the following cases:
+
+\begin{itemize}
+	\item {\bf Vectorized loop.} If any of the loops between the \texttt{store\_level} and the \texttt{lazy\_level} are vectorized, it may not make sense to emit a check which conditionally skips much of the computation.
+Though it would make sense to generate code for a single boolean check around a single pixel computation, with more sophisticated tile-level dynamic scheduling the vectorized execution might behave pathologically (especially if the conditional check involves traversing a sophisticated data structure as suggested in Section~\ref{sec:future-tiles}).
+	\item {\bf Parallel loop.} If any of the loops between the \texttt{store\_level} and the \texttt{lazy\_level} are parallelized, then one thread attempting to record the result of a computation might race with another thread attempting to use that result.
+With x86's sequentially-consistent memory model and the simple boolean check, this is not an issue, but when compiling for ARM or using a more sophisticated data structure for memoizing Halide would need to emit memory barriers at best and locks at worst.
+This is not infeasible to implement, but raises research questions of its own and so is left to future work.
+	\item {\bf Update phase.} Dynamic scheduling is not (yet) compatible with functions which have an update phase in their \texttt{Pipeline} node, which can arise when the user provides a reduction definition.
+This is merely a case I did not consider in detail, and may well be possible to handle.
+\end{itemize}
+
+\subsection{Code}
+
+The reader should feel free to find and peruse my code at \texttt{https://github.com/bblum/Halide}. During the weeks following this submission I will polish it and submit a pull request upstream.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Evaluation}
+
+To evaluate dynamic scheduling I sought to answer the following questions.
+
+\begin{itemize}
+	\item How much duplicate work does dynamic scheduling let us avoid?
+	\item How does the amount of duplicate work avoided and the execution time vary when different function schedules are used?
+\end{itemize}
+
+To approach these questions I ran experiments with input images of three different sizes using 8 different function schedules.
+I list the schedules used in Figure~\ref{fig:eval-scheds}, and show the images in Figure~\ref{fig:eval-images}.
+
+\begin{figure}[h]
+	\begin{center}
+	\begin{tabular}{c|c}
+		x-axis label & schedule used \\
+		\hline
+                1 & \texttt{invert.store\_at(flip, c).compute\_at(flip, yi);~} \\
+                2 & \texttt{invert.store\_at(flip, c).compute\_at(flip, xi);~} \\
+                3 & \texttt{invert.store\_at(flip, y).compute\_at(flip, x);~~} \\
+                4 & \texttt{invert.store\_at(flip, y).compute\_at(flip, yi);~} \\
+                5 & \texttt{invert.store\_at(flip, y).compute\_at(flip, xi);~} \\
+                6 & \texttt{invert.store\_at(flip, x).compute\_at(flip, yi);~} \\
+                7 & \texttt{invert.store\_at(flip, x).compute\_at(flip, xi);~} \\
+                8 & \texttt{invert.store\_at(flip, yi).compute\_at(flip, xi);} 
+	\end{tabular}
+	\end{center}
+	\caption{All experiments used the schedule \texttt{flipped.tile(x, y, xi, yi, 2, 2);}.
+On the subsequent graphs in this evaluation, the x-axis labels correspond to the corresponding schedule for \texttt{invert} here.}
+	\label{fig:eval-scheds}
+\end{figure}
+\begin{figure}[h]
+        \begin{center}
+        \begin{tabular}{ccc}
+	\begin{subfigure}[b]{0.22\textwidth}
+        \includegraphics[width=\textwidth]{gradient.png} \\
+        \\
+        \includegraphics[width=\textwidth]{gradient-output.png} \\
+        \caption{Small image (64x64)}
+	\end{subfigure}
+	&
+	\begin{subfigure}[b]{0.34\textwidth}
+        \includegraphics[width=\textwidth]{readers-writers.png} \\
+        \\
+        \includegraphics[width=\textwidth]{rw-output.png} \\
+        \caption{Medium image (400x261)}
+	\end{subfigure}
+	&
+	\begin{subfigure}[b]{0.335\textwidth}
+        \includegraphics[width=\textwidth]{canoe-0.png} \\
+        \\
+        \includegraphics[width=\textwidth]{canoe-3.png} \\
+        \caption{Large image (963x642)}
+	\end{subfigure}
+        \end{tabular}
+        \caption{Images used for evaluation (top), and the outputs given by the invert/flip algorithm (bottom).}
+        \label{fig:eval-images}
+        \end{center}
+\end{figure}
+
+\begin{figure}[h]
+	\begin{center}
+	\includegraphics[width=0.4\textwidth]{graphs/small.pdf}
+	\end{center}
+	\caption{Analysis of how much duplicate work was saved for the small image.}
+	\label{fig:graph-trace}
+\end{figure}
+\begin{figure}[h]
+	\begin{center}
+	\includegraphics[width=0.4\textwidth]{graphs/medium.pdf} \quad
+	\includegraphics[width=0.4\textwidth]{graphs/large.pdf}
+	\end{center}
+	\caption{Performance results for the medium and large images.}
+	\label{fig:graph-perf}
+\end{figure}
+
+\subsection{Work Saved}
+
+For the small input image, I used Halide's tracing facility to measure the number of pixels computed by the \texttt{invert()} operation (\texttt{invert.trace\_stores()}), comparing the numbers obtained from dynamic and from static scheduling.
+For each schedule, the difference in this number represents the amount of computation saved.
+
+Figure~\ref{fig:graph-trace} shows this experiment's results.
+We can see that in some cases, several orders of magnitude of work are saved; in some, a modest amount of work is saved; and in some it makes no difference.
+Naturally dynamic scheduling never creates additional \texttt{invert()} calls.
+
+\subsection{Peformance Discussion}
+
+For the medium and large input image I measured the elapsed wall clock time during execution, running both statically and dynamically.
+Figure~\ref{fig:graph-perf} shows this experiment's results.
+The performance patterns are similar between the medium and large images, apart from the obvious difference in number of total seconds taken.
+
+One notable anomaly is that in schedules 6, 7, and 8, performance is actually worse with dynamic scheduling than with static scheduling.
+I attribute this to the fact that in these schedules the \texttt{invert()} code is confined to the innermost loop levels of \texttt{flip()}, which causes \texttt{invert()}'s buffer to be freed and re-allocated many times, thereby hurting the potential for saving computation across loop iterations.
+Though these schedules still reduce the number of \texttt{invert()} calls (as shown in the previous experiment), it fails to compensate for the incidental overhead of using dynamic scheduling to begin with -- that is, the cost of allocating an extra buffer and branching on a read from it inside of a tight loop.
+From this we conclude that dynamic scheduling achieves the greatest wins when the store level is closer to the root of the loop nest.
+
+Also note the significant performance difference when using \texttt{xi}, the innermost loop variable, as the compute level (schedules \#2 and \#5) as compared to computing at more outer loops (schedules \#1, \#3, \#4).
+This trend is independent of static vs dynamic scheduling, but suggests that this algorithm benefits from computing \texttt{invert()} in tiles or rows at a time (with respect to \texttt{flip()}).
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Open Questions}
+\label{sec:future-tiles}
+% TODO
+
+% kayvon's idea: an analysis tool for algs that says whether they should be lazy
+
+DISCUSSION FORTHCOMING
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Conclusion}
+
+Halide provides a novel way of implementing image processing algorithms: expressing the {\em algorithm logic} and the {\em function schedule} independently from each other.
+The programmer supplies a functional-style algorithm description and a set of scheduling directives, and the Halide compiler and runtime translates it into a structured loop nest that can be flexibly and portably parallelized, vectorized, and so on.
+However, Halide's ability to generate performant code is limited by its reliance on {\em static knowledge} of data dependencies between an algorithm's stages.
+In situations where it lacks such information, it must make conservative approximations that often cause duplicate work across outer loop iterations.
+
+I have extended Halide with a new feature, called {\em dynamic scheduling}, which enables it to avoid a majority of the duplicate work that static schedules would incur in such situations.
+Users may now direct the Halide compiler to generate code which decides at runtime, not compile-time, when to perform computation associated with an algorithm's early stages.
+I have also supplied an evaluation of dynamic scheduling which proves the technique's usefulness and shows some performance patterns which suggest how best to use it.
+Finally, I have discussed some possible directions for extending this work which will hopefully make the technique even more broadly applicable and useful.
 
@@
 
-\subsection{Code Generation}
-
-Inside the compiler, the dynamic schedule configuration is stored in a \texttt{Func}'s \texttt{Schedule} object as a \texttt{LoopLevel} named \texttt{lazy\_level}, alongside the existing configurations \texttt{store\_level} and \texttt{compute\_level}.
-Unlike the latter two, however, the \texttt{lazy\_level} is not defined in terms of the loop variable of a different function, but rather for one of the function's own loop variables.
-Hence, \texttt{lazy\_level.func} always refers to the containing function.
-
-The code generation itself occurs during the \texttt{Lower} phase of compilation, between the \texttt{schedule\_functions()} phase (found in \texttt{Lower.cpp}) and the \texttt{storage\_flattening()} phase (found in \texttt{StorageFlattening.cpp}).
-The former processes the \texttt{lazy\_level} configuration and generates IR to represent it, and the latter converts those IR nodes into conventional IR that can be code-generated.
-For this purpose I added a new type of IR node, \texttt{DynamicStmt} (found in \texttt{IR.h}), which represents the ``if not-in-cache then compute; mark-cache; end'' pattern.
-
-\begin{itemize}
-	\item {\bf Function Scheduling.} This phase converts the dynamic schedule configuration into IR in two places.
-The first is when it generates a \texttt{Realize} node (in \texttt{build\_realize()}); it simply sets the node's \texttt{lazy} flag to \texttt{true} if the associated function's \texttt{lazy\_level} was set.\footnote{
-	I experimented here with computing and storing an expression representing the necessary size for the bitmask, rather than just a boolean. Future work may find this approach worth doing after all.}
-	The second is when building a function's loop nest (in \texttt{build\_provide\_loop\_nest()}); when the approprate loop level is encountered, it inserts a \texttt{DynamicStmt} node between the \texttt{For} loop and its body.
-	\item {\bf Storage Flattening.} This phase visits the IR described above and ``flattens'' it into arrays more reminiscent of the C memory model.
-In \texttt{visit(Realize *)}, where it makes the \texttt{Allocate} node for the function's computed results, it now also will make another \texttt{Allocate} and a brief initialization loop if \texttt{realize->lazy} is set.
-Second, in \texttt{visit(DynamicStmt *)}, it converts the \texttt{DynamicStmt} node into a conventional \texttt{IfThenElse} node using the provided index for the allocated bitmask.
-\end{itemize}
-
-\subsection{Legality Checking}
-
-During the function scheduling in \texttt{Lower.cpp} Halide must also check that a specified dynamic schedule is legal. I currently check for the following cases:
-
-\begin{itemize}
-	\item {\bf Vectorized loop.} If any of the loops between the \texttt{store\_level} and the \texttt{lazy\_level} are vectorized, it may not make sense to emit a check which conditionally skips much of the computation.
-Though it would make sense to generate code for a single boolean check around a single pixel computation, with more sophisticated tile-level dynamic scheduling the vectorized execution might behave pathologically (especially if the conditional check involves traversing a sophisticated data structure as suggested in Section~\ref{sec:future-tiles}).
-	\item {\bf Parallel loop.} If any of the loops between the \texttt{store\_level} and the \texttt{lazy\_level} are parallelized, then one thread attempting to record the result of a computation might race with another thread attempting to use that result.
-With x86's sequentially-consistent memory model and the simple boolean check, this is not an issue, but when compiling for ARM or using a more sophisticated data structure for memoizing Halide would need to emit memory barriers at best and locks at worst.
-This is not infeasible to implement, but raises research questions of its own and so is left to future work.
-	\item {\bf Update phase.} Dynamic scheduling is not (yet) compatible with functions which have an update phase in their \texttt{Pipeline} node, which can arise when the user provides a reduction definition.
-This is merely a case I did not consider in detail, and may well be possible to handle.
-\end{itemize}
-
-\subsection{Code}
-
-The reader should feel free to find and peruse my code at \texttt{https://github.com/bblum/Halide}. During the weeks following this submission I will polish it and submit a pull request upstream.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Evaluation}
-% TODO
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Open Questions}
-% TODO
-
-\label{sec:future-tiles}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Conclusion}
-% TODO
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2013-12-17 15869/paper.tex
finish writeup

@@
 \usepackage{proof}
+\usepackage{wrapfig}
 \usepackage{array}
@@
 
-% TODO
-{\bf Tile-granularity dynamic scheduling.} DISCUSSION FORTHCOMING.
+{\bf Tile-granularity dynamic scheduling.} If the user supplies a loop variable as the argument to \texttt{compute\_lazy()}, Halide will place the conditional check just inside the loop corresponding to that variable, instead of immediately around the store operation inside the innermost loop.
+This results in caching of already-computed tiles (or rows or columns as the case may be).
+This can provide several benefits: performing the conditional check only once per group of pixels saves the overhead of a branch, allows vectorizing the innermost loops, and allows skipping entire iterations across loop ranges in addition to just skipping a function call.
+In Figure~\ref{fig:dynamic-tile} I show sample code that Halide would generate for a program that used tile-level dynamic scheduling.
 
@@
 	\item {\bf Vectorized loop.} If any of the loops between the \texttt{store\_level} and the \texttt{lazy\_level} are vectorized, it may not make sense to emit a check which conditionally skips much of the computation.
-Though it would make sense to generate code for a single boolean check around a single pixel computation, with more sophisticated tile-level dynamic scheduling the vectorized execution might behave pathologically (especially if the conditional check involves traversing a sophisticated data structure as suggested in Section~\ref{sec:future-tiles}).
+Though it would make sense to generate code for a single boolean check around a single pixel computation, with more sophisticated tile-level dynamic scheduling the vectorized execution might behave pathologically (especially if the conditional check involves traversing a sophisticated data structure as suggested in Section~\ref{sec:future}).
 	\item {\bf Parallel loop.} If any of the loops between the \texttt{store\_level} and the \texttt{lazy\_level} are parallelized, then one thread attempting to record the result of a computation might race with another thread attempting to use that result.
@@
 
+\subsection{Problems with Tile-Level Dynamic Scheduling}
+\label{sec:problem}
+
+\begin{wrapfigure}{r}{0.28\textwidth}
+	\begin{center}
+	\includegraphics[width=0.2\textwidth]{bloopers/01.png}
+	\end{center}
+	\caption{Dynamic scheduling at tile granularity can fail to compute results for some pixels, using uninitialized values instead. (Compare to the expected output in Figure~\ref{fig:correct-output}.)}
+	\label{fig:uh-oh}
+\end{wrapfigure}
+
+To cache computed results at coarser-than-pixel granularity, I used the coordinate of the ``top left'' pixel within the block as the index into the bitmask.
+Unfortunately, this proved inadequate in some cases, leading to the program erroneously believing it had aleady computed a block of pixels and using uninitialized values for some pixels.
+
+In Figure~\ref{fig:dynamic-tile}, note the two lines of code marked \texttt{XXX}.
+These lines compute the bounds of the \texttt{col\_i} dimension, and their values depend on the value of the outer \texttt{col} loop variable, which varies across iterations.
+Hence, on one iteration we might compute \texttt{invert} values for the range (0,0) to (0,14), store that we had already computed the row starting at (0,0), then later skip computing values for the range (0,0) to (0,15), leaving the last pixel undefined.
+This behaviour is shown in Figure~\ref{fig:uh-oh}, where the rightmost column of the image has been generated with undefined values.
+
+Note that this problem only arises with some schedules; for example, changing \texttt{invert.store\_at(flip, row);} to \texttt{store\_at(flip, col);} would prevent tiles of different sizes from arising.
+Nevertheless, a correct implementation in the general case must memoize the dimensions of each tile computed in addition to the location.
+Section~\ref{sec:future} discusses potential solutions.
+
 \subsection{Code}
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Evaluation}
-
-To evaluate dynamic scheduling I sought to answer the following questions.
-
-\begin{itemize}
-	\item How much duplicate work does dynamic scheduling let us avoid?
-	\item How does the amount of duplicate work avoided and the execution time vary when different function schedules are used?
-\end{itemize}
-
-To approach these questions I ran experiments with input images of three different sizes using 8 different function schedules.
-I list the schedules used in Figure~\ref{fig:eval-scheds}, and show the images in Figure~\ref{fig:eval-images}.
-
-\begin{figure}[h]
-	\begin{center}
-	\begin{tabular}{c|c}
-		x-axis label & schedule used \\
-		\hline
-                1 & \texttt{invert.store\_at(flip, c).compute\_at(flip, yi);~} \\
-                2 & \texttt{invert.store\_at(flip, c).compute\_at(flip, xi);~} \\
-                3 & \texttt{invert.store\_at(flip, y).compute\_at(flip, x);~~} \\
-                4 & \texttt{invert.store\_at(flip, y).compute\_at(flip, yi);~} \\
-                5 & \texttt{invert.store\_at(flip, y).compute\_at(flip, xi);~} \\
-                6 & \texttt{invert.store\_at(flip, x).compute\_at(flip, yi);~} \\
-                7 & \texttt{invert.store\_at(flip, x).compute\_at(flip, xi);~} \\
-                8 & \texttt{invert.store\_at(flip, yi).compute\_at(flip, xi);} 
-	\end{tabular}
-	\end{center}
-	\caption{All experiments used the schedule \texttt{flipped.tile(x, y, xi, yi, 2, 2);}.
-On the subsequent graphs in this evaluation, the x-axis labels correspond to the corresponding schedule for \texttt{invert} here.}
-	\label{fig:eval-scheds}
-\end{figure}
-\begin{figure}[h]
-        \begin{center}
-        \begin{tabular}{ccc}
-	\begin{subfigure}[b]{0.22\textwidth}
-        \includegraphics[width=\textwidth]{gradient.png} \\
-        \\
-        \includegraphics[width=\textwidth]{gradient-output.png} \\
-        \caption{Small image (64x64)}
-	\end{subfigure}
-	&
-	\begin{subfigure}[b]{0.34\textwidth}
-        \includegraphics[width=\textwidth]{readers-writers.png} \\
-        \\
-        \includegraphics[width=\textwidth]{rw-output.png} \\
-        \caption{Medium image (400x261)}
-	\end{subfigure}
-	&
-	\begin{subfigure}[b]{0.335\textwidth}
-        \includegraphics[width=\textwidth]{canoe-0.png} \\
-        \\
-        \includegraphics[width=\textwidth]{canoe-3.png} \\
-        \caption{Large image (963x642)}
-	\end{subfigure}
-        \end{tabular}
-        \caption{Images used for evaluation (top), and the outputs given by the invert/flip algorithm (bottom).}
-        \label{fig:eval-images}
-        \end{center}
-\end{figure}
-
-\begin{figure}[h]
-	\begin{center}
-	\includegraphics[width=0.4\textwidth]{graphs/small.pdf}
-	\end{center}
-	\caption{Analysis of how much duplicate work was saved for the small image.}
-	\label{fig:graph-trace}
-\end{figure}
-\begin{figure}[h]
-	\begin{center}
-	\includegraphics[width=0.4\textwidth]{graphs/medium.pdf} \quad
-	\includegraphics[width=0.4\textwidth]{graphs/large.pdf}
-	\end{center}
-	\caption{Performance results for the medium and large images.}
-	\label{fig:graph-perf}
-\end{figure}
-
-\subsection{Work Saved}
-
-For the small input image, I used Halide's tracing facility to measure the number of pixels computed by the \texttt{invert()} operation (\texttt{invert.trace\_stores()}), comparing the numbers obtained from dynamic and from static scheduling.
-For each schedule, the difference in this number represents the amount of computation saved.
-
-Figure~\ref{fig:graph-trace} shows this experiment's results.
-We can see that in some cases, several orders of magnitude of work are saved; in some, a modest amount of work is saved; and in some it makes no difference.
-Naturally dynamic scheduling never creates additional \texttt{invert()} calls.
-
-\subsection{Peformance Discussion}
-
-For the medium and large input image I measured the elapsed wall clock time during execution, running both statically and dynamically.
-Figure~\ref{fig:graph-perf} shows this experiment's results.
-The performance patterns are similar between the medium and large images, apart from the obvious difference in number of total seconds taken.
-
-One notable anomaly is that in schedules 6, 7, and 8, performance is actually worse with dynamic scheduling than with static scheduling.
-I attribute this to the fact that in these schedules the \texttt{invert()} code is confined to the innermost loop levels of \texttt{flip()}, which causes \texttt{invert()}'s buffer to be freed and re-allocated many times, thereby hurting the potential for saving computation across loop iterations.
-Though these schedules still reduce the number of \texttt{invert()} calls (as shown in the previous experiment), it fails to compensate for the incidental overhead of using dynamic scheduling to begin with -- that is, the cost of allocating an extra buffer and branching on a read from it inside of a tight loop.
-From this we conclude that dynamic scheduling achieves the greatest wins when the store level is closer to the root of the loop nest.
-
-Also note the significant performance difference when using \texttt{xi}, the innermost loop variable, as the compute level (schedules \#2 and \#5) as compared to computing at more outer loops (schedules \#1, \#3, \#4).
-This trend is independent of static vs dynamic scheduling, but suggests that this algorithm benefits from computing \texttt{invert()} in tiles or rows at a time (with respect to \texttt{flip()}).
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Open Questions}
-\label{sec:future-tiles}
-% TODO
-
-% kayvon's idea: an analysis tool for algs that says whether they should be lazy
-
-DISCUSSION FORTHCOMING
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Conclusion}
-
-Halide provides a novel way of implementing image processing algorithms: expressing the {\em algorithm logic} and the {\em function schedule} independently from each other.
-The programmer supplies a functional-style algorithm description and a set of scheduling directives, and the Halide compiler and runtime translates it into a structured loop nest that can be flexibly and portably parallelized, vectorized, and so on.
-However, Halide's ability to generate performant code is limited by its reliance on {\em static knowledge} of data dependencies between an algorithm's stages.
-In situations where it lacks such information, it must make conservative approximations that often cause duplicate work across outer loop iterations.
-
-I have extended Halide with a new feature, called {\em dynamic scheduling}, which enables it to avoid a majority of the duplicate work that static schedules would incur in such situations.
-Users may now direct the Halide compiler to generate code which decides at runtime, not compile-time, when to perform computation associated with an algorithm's early stages.
-I have also supplied an evaluation of dynamic scheduling which proves the technique's usefulness and shows some performance patterns which suggest how best to use it.
-Finally, I have discussed some possible directions for extending this work which will hopefully make the technique even more broadly applicable and useful.
-
-\begin{figure}[t]
+\pagebreak
+\begin{figure}[h!]
 	\begin{center}
@@
                 \hilight{blue}{\texttt{flip.tile(col, row, col2, row2, 8, 8);}} \\
-                \hilight{olivegreen}{\texttt{invert.store\_at(flip, col);}} \\
+                \hilight{olivegreen}{\texttt{invert.store\_at(flip, row);}} \\
 		\hilight{olivegreen}{\texttt{invert.compute\_at(invert, row2);}}\\
@@
                 \texttt{\hilight{blue}{for~(row~=~0~to~8)~\{}} \\
+                \texttt{\hilight{olivegreen}{~~float~invert[...][...];}} \\
+		{\bf \texttt{\hilight{orange}{~~bool~result\_computed[...][...];}}} \\
                 \texttt{\hilight{blue}{~~for~(col~=~0~to~8)~\{}} \\
-                \texttt{\hilight{olivegreen}{~~~~float~invert[...][...];}} \\
-		{\bf \texttt{\hilight{orange}{~~~~bool~result\_computed[...][...];}}} \\
                 \texttt{\hilight{blue}{~~~~for~(row2~=~0~to~WIDTH/8)~\{}} \\
@@
                 \texttt{\hilight{olivegreen}{~~~~~~let max\_row = ...;}} \\
-                \texttt{\hilight{olivegreen}{~~~~~~let min\_col = ...;}} \\
-                \texttt{\hilight{olivegreen}{~~~~~~let max\_col = ...;}} \\
+                \texttt{\hilight{olivegreen}{~~~~~~let min\_col = ...; // XXX}} \\
+                \texttt{\hilight{olivegreen}{~~~~~~let max\_col = ...; // XXX}} \\
                 \texttt{\hilight{olivegreen}{~~~~~~for~(row\_i~=~min\_row~to~max\_row)~\{}} \\
@@
 Here, the ``tile'' is simply the row denited by the variable \texttt{row\_i}.
-Simply using a single boolean for each tile can be unsound in some cases, as discussed in Section~\ref{sec:future-tiles}.}
+Simply using a single boolean for each tile can be unsound in some cases, as discussed in Section~\ref{sec:problem}.}
 		\label{fig:dynamic-tile}
@@
 \end{figure}
+
+\pagebreak
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Evaluation}
+
+To evaluate dynamic scheduling I sought to answer the following questions.
+
+\begin{itemize}
+	\item How much duplicate work does dynamic scheduling let us avoid?
+	\item How does the amount of duplicate work avoided and the execution time vary when different function schedules are used?
+\end{itemize}
+
+To approach these questions I ran experiments with input images of three different sizes using 8 different function schedules.
+I list the schedules used in Figure~\ref{fig:eval-scheds}, and show the images in Figure~\ref{fig:eval-images}.
+
+\begin{figure}[h]
+	\begin{center}
+	\begin{tabular}{c|c}
+		x-axis label & schedule used for \texttt{invert()} \\
+		\hline
+                1 & \texttt{invert.store\_at(flip, c).compute\_at(flip, yi);~} \\
+                2 & \texttt{invert.store\_at(flip, c).compute\_at(flip, xi);~} \\
+                3 & \texttt{invert.store\_at(flip, y).compute\_at(flip, x);~~} \\
+                4 & \texttt{invert.store\_at(flip, y).compute\_at(flip, yi);~} \\
+                5 & \texttt{invert.store\_at(flip, y).compute\_at(flip, xi);~} \\
+                6 & \texttt{invert.store\_at(flip, x).compute\_at(flip, yi);~} \\
+                7 & \texttt{invert.store\_at(flip, x).compute\_at(flip, xi);~} \\
+                8 & \texttt{invert.store\_at(flip, yi).compute\_at(flip, xi);} 
+	\end{tabular}
+	\end{center}
+	\caption{All experiments used the schedule \texttt{flipped.tile(x, y, xi, yi, 2, 2);}.
+On the subsequent graphs in this evaluation, the x-axis labels correspond to the corresponding schedule for \texttt{invert} here.}
+	\label{fig:eval-scheds}
+\end{figure}
+\begin{figure}[h]
+        \begin{center}
+        \begin{tabular}{ccc}
+	\begin{subfigure}[b]{0.22\textwidth}
+        \includegraphics[width=\textwidth]{gradient.png} \\
+        \\
+        \includegraphics[width=\textwidth]{gradient-output.png} \\
+        \caption{Small image (64x64)}
+        \label{fig:correct-output}
+	\end{subfigure}
+	&
+	\begin{subfigure}[b]{0.34\textwidth}
+        \includegraphics[width=\textwidth]{readers-writers.png} \\
+        \\
+        \includegraphics[width=\textwidth]{rw-output.png} \\
+        \caption{Medium image (400x261)}
+	\end{subfigure}
+	&
+	\begin{subfigure}[b]{0.335\textwidth}
+        \includegraphics[width=\textwidth]{canoe-0.png} \\
+        \\
+        \includegraphics[width=\textwidth]{canoe-3.png} \\
+        \caption{Large image (963x642)}
+	\end{subfigure}
+        \end{tabular}
+        \caption{Images used for evaluation (top), and the outputs given by the invert/flip algorithm (bottom).}
+        \label{fig:eval-images}
+        \end{center}
+\end{figure}
+
+\begin{figure}[h]
+	\begin{center}
+	\includegraphics[width=0.47\textwidth]{graphs/small.pdf}
+	\end{center}
+	\caption{Analysis of how much duplicate work was saved for the small image. Lower is better.}
+	\label{fig:graph-trace}
+\end{figure}
+\begin{figure}[h]
+	\begin{center}
+	\includegraphics[width=0.47\textwidth]{graphs/medium.pdf} \quad
+	\includegraphics[width=0.47\textwidth]{graphs/large.pdf}
+	\end{center}
+	\caption{Performance results for the medium and large images. Lower is better.}
+	\label{fig:graph-perf}
+\end{figure}
+
+\subsection{Work Saved}
+
+For the small input image, I used Halide's tracing facility to measure the number of pixels computed by the \texttt{invert()} operation (\texttt{invert.trace\_stores()}), comparing the numbers obtained from dynamic and from static scheduling.
+For each schedule, the difference in this number represents the amount of computation saved.
+
+Figure~\ref{fig:graph-trace} shows this experiment's results.
+We can see that in some cases, several orders of magnitude of work are saved; in some, a modest amount of work is saved; and in some it makes no difference.
+Naturally dynamic scheduling never creates additional \texttt{invert()} calls.
+
+\subsection{Peformance Discussion}
+
+For the medium and large input image I measured the elapsed wall clock time during execution, running both statically and dynamically.
+Figure~\ref{fig:graph-perf} shows this experiment's results.
+The performance patterns are similar between the medium and large images, apart from the obvious difference in number of total seconds taken.
+
+One notable anomaly is that in schedules 6, 7, and 8, performance is actually worse with dynamic scheduling than with static scheduling.
+I attribute this to the fact that in these schedules the \texttt{invert()} code is confined to the innermost loop levels of \texttt{flip()}, which causes \texttt{invert()}'s buffer to be freed and re-allocated many times, thereby hurting the potential for saving computation across loop iterations.
+Though these schedules still reduce the number of \texttt{invert()} calls (as shown in the previous experiment), they fail to compensate for the incidental overhead of using dynamic scheduling to begin with -- that is, the cost of allocating an extra buffer and branching on a read from it inside of a tight loop.
+From this I conclude that dynamic scheduling achieves the greatest wins when the store level is closer to the root of the loop nest.
+Considering that the \texttt{invert()} operation is not very expensive, I also predict that performance savings would improve for an algorithm with a more expensive first stage (or several stages) to be dynamically scheduled.
+
+Also note the significant performance difference when using \texttt{xi}, the innermost loop variable, as the compute level (schedules \#2 and \#5) as compared to computing at more outer loops (schedules \#1, \#3, \#4).
+This trend is independent of static vs dynamic scheduling, but suggests that this algorithm benefits from computing \texttt{invert()} in tiles or rows at a time (with respect to \texttt{flip()}).
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Open Questions}
+\label{sec:future}
+
+\begin{itemize}
+\item \textbf{How can tile-level dynamic scheduling provide correct output when tiles overlap across loop iterations?} I know of two possible approaches to this question.
+In one approach, the cache is indexed by tile location, as before, but also stores the extent of the tile computed in each dimension.
+Then, querying the cache will return true if the tile we are about to compute has already been covered by one or more previously computed tiles.
+This may be implemented either by maintaining a pixel-granularity bitmask, as before, and setting bits for all pixels contained in a tile (which may be optimized with SSE), or by using a more sophisticated data structure maintained by the Halide runtime such as a K-D tree.
+
+In the other approach, rather than caching computed results for each tile we compute, we partition the cache in its own tiling pattern, and split the loops during code generation according to the cache's tile boundaries.
+Thus, rather than skipping entire tiles, some or all of each tile may be skipped in ``cache entry''-sized blocks.
+In each of these approaches, how much work can be deduplicated will be dependent on the workload.
+
+\item \textbf{How can dynamic scheduling be thread-safe within a parallel loop?}
+If a loop between the \texttt{store\_level} and the \texttt{lazy\_level} is paralell, multiple threads may simultaneously read and write to the bitmask and make decisions about whether or not to skip work.
+On an architecture with sequentially-consistent memory semantics and pixel-level dynamic scheduling, my implementation already generates thread-safe code.
+For relaxed-memory architectures such as ARM, Halide will need to emit memory fences between the produce and consume phase of each pipeline node.
+If the data structure used becomes more complicated than a simple array of booleans (as suggested in the previous question), more complex synchronization such as locks may be needed.
+
+\item \textbf{How can a user know when to use dynamic scheduling?}
+Dynamic scheduling is far from being an intuitive language feature, and it is difficult to reason about when it will provide a performance boost.
+A clever user might trace executions of their algorithm with different function schedules and measure the amount of duplicated work compared to the optimal case, and decide whether to use dynamic scheduling based on that difference and the computational expense of their algorithm's early stages.
+To assist more inexperienced users, this process might be automated by a tool which searches through different possible function schedules and heuristically makes a recommendation about how to use dynamic scheduling.
+
+\end{itemize}
+
+% kayvon's idea: an analysis tool for algs that says whether they should be lazy
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Conclusion}
+
+Halide provides a novel way of implementing image processing algorithms: expressing the {\em algorithm logic} and the {\em function schedule} independently from each other.
+The programmer supplies a functional-style algorithm description and a set of scheduling directives, and the Halide compiler and runtime translates it into a structured loop nest that can be flexibly and portably parallelized, vectorized, and so on.
+However, Halide's ability to generate performant code is limited by its reliance on {\em static knowledge} of data dependencies between an algorithm's stages.
+In situations where it lacks such information, it must make conservative approximations that often cause duplicate work across outer loop iterations.
+
+I have extended Halide with a new feature, called {\em dynamic scheduling}, which enables it to avoid a majority of the duplicate work that static schedules would incur in such situations.
+Users may now direct the Halide compiler to generate code which decides at runtime, not compile-time, when to perform computation associated with an algorithm's early stages.
+I have also supplied an evaluation of dynamic scheduling which proves the technique's usefulness and shows some performance patterns which suggest how best to use it.
+Finally, I have discussed some possible directions for extending this work which will hopefully make the technique even more broadly applicable and useful.
 

### 2014-04-02 landslide-lecture/lecture.tex
fix about-the-project slide and add citation
cleanup other things i guess
update landslide lecture for s2014

@@
 
-\title[Landslide]{{\bf Systematic Dynamic Race Condition Detection}}
+\title[Landslide]{{\bf Systematic Testing: A New Perspective on Races}}
 \subtitle[]{ {\em more clever than } ``\texttt{slaughter cho2}'' {\em since 2011.}}
@@
 \institute[CMU 15-410]{Carnegie Mellon University - 15-410}
-\date[]{2013, April 14}
+\date[]{2014, April 2}
 
@@
 
+\newcommand\hilight[2]{\color{#1}#2\color{black}}
+\definecolor{olivegreen}{RGB}{0,127,0}
+
 \begin{document}
@@
 		\item The execution tree
-		\item Decision points
+		\item Preemption points
 	\end{itemize}
@@
 
-\begin{frame}{Decision Points (``good'' case)}
+\begin{frame}{Thread Interleavings (``good'' case)}
 	\begin{tabular}{|l|l|l}
@@
 
-\begin{frame}{Decision Points (race condition)}
+\begin{frame}{Thread Interleavings (race condition)}
 	\begin{tabular}{|l|l|l}
@@
 	\end{tabular}
+\end{frame}
+
+\begin{frame}{Testing}
+	How can programmers be confident in their code's correctness?
+	\begin{itemize}
+		\item Theorem proving
+		\item Unit tests
+		\item Stress tests
+		\begin{itemize}
+			\item (state of the art in 15-410)
+		\end{itemize}
+		\item Releasing to paying customers and worrying about correctness later
+	\end{itemize}
+	\linegap
+
+	{\bf Motivation}: Can we do better than stress testing?
 \end{frame}
@@
 
-\begin{frame}{Decision Points} % TODO: duplicate slide with later
-	A {\bf decision point} is\ldots
+\begin{frame}{Preemption Points} % TODO: duplicate slide with later
+	A {\bf preemption point} is\ldots
 
@@
 		\begin{itemize}
-			\item (by controlling scheduling decisions at decision points)
+			\item (by controlling scheduling decisions at preemption points)
 		\end{itemize}
@@
 	\begin{itemize}
-		\item Identify a decision point to choose differently
+		\item Identify a preemption point to replay differently
 		\item Reset machine state and start over
-		\item Replay test from the beginning, with different ``decisions''
+		\item Replay test from the beginning, with different preemptions
 	\end{itemize}
@@
 
-\begin{frame}{More on Decision Points}
+\begin{frame}{More on Preemption Points}
 	Important point: What does ``all possible interleavings'' mean?
@@
 
-	Other extreme: Nothing is a decision point
+	Other extreme: Nothing is a preemption point
 	\begin{itemize}
@@
 		\item {\em Bad news}: Only one execution was checked for bugginess.
-		\item {\em Bad news}: No alternative interleavings explored.
 		\begin{itemize}
+			\item No alternative interleavings explored.
 			\item Makes ``no race found'' a weak claim.
@@
 	\end{itemize}
+	\linegap
+
+	Is there a ``sweet spot''?
 \end{frame}
 
-\begin{frame}{More on Decision Points}
+\begin{frame}{More on Preemption Points}
 	Sweet spot: Insert a thread switch everywhere it ``might matter''.
@@
 \begin{frame}{Controlling Scheduling Decisions}
-	Control over sources of nondeterminism
+	How do we control which thread runs when?
+	\linegap
+
+	In other words, what sources of nondeterminism must we control?
+
 	\begin{itemize}
@@
 \begin{frame}{Memory Interposition}
+	How do we detect race-induced heap corruption?
+	\linegap
+
 	In order to find use-after-free, need to know:
@@
 \begin{frame}{State Space Explosion}
-	State spaces grow exponentially
+	Serious problem: State spaces grow exponentially
 	\begin{itemize}
-		\item With $d$ decision points, $k$ runnable threads, size $d^k$.
+		\item With $p$ preemption points and $k$ runnable threads, size $p^k$.
 		\item Threatens our ability to explore everything.
@@
 	\begin{itemize}
-		\item Complicated algorithm; ask me later for details.
-		\item Intuitive explanation follows.
+		\item Intuitively: After exploring each interleaving, identify ``equivalent'' alternatives and skip over them.
+		\item Example follows\dots
 	\end{itemize}
@@
 \begin{frame}{About The Project}
-	5th year MS since June 2011 % Say: planning to finish this may, hopefully
-	My MS thesis project (June 2011-2012)
+	% 5th year MS since June 2011 % Say: planning to finish this may, hopefully
+	My MS thesis project (2011-2012)
+	\begin{itemize}
+		\item {}[CMU-CS-12-118 -- \url{http://bblum.net/thesis.pdf}]
+	\end{itemize}
 
@@
 	\linegap
-	Artificially causes timer interrupts
+	Artificially causes timer interrupts to control scheduling
 
@@
 
-\begin{frame}{In Which Ben Offers Help}
+\begin{frame}{Recruitment}
 	% Say: So, this wasn't just a research talk
@@
 
-\begin{frame}{Keeping It Real}
+\begin{frame}{Words of Warning}
 	Finding race conditions is hard for humans.
@@
 	\begin{center}
-		\includegraphics[width=0.9\textwidth]{tell_landslide.png}
+		{\small
+		\begin{tabular}{l}
+		\texttt{\hilight{olivegreen}{int}~thread\_fork() \{} \\
+	        \texttt{~~~~\hilight{olivegreen}{thread\_t}~*child = construct\_new\_thread();} \\
+		\texttt{~~~~\hilight{purple}{tell\_landslide\_forking}();~\hilight{blue}{/* A new thread was created.~*/}} \\
+	        \texttt{~~~~add\_to\_runqueue(child);} \\
+	        \texttt{~~~~return child->tid;} \\
+		\texttt{\}} \\
+		\texttt{\hilight{olivegreen}{void}~add\_to\_runqueue(\hilight{olivegreen}{thread\_t}~*t) \{} \\
+		\texttt{~~~~\hilight{blue}{/* This thread is now runnable.~*/}} \\
+		\texttt{~~~~\hilight{purple}{tell\_landslide\_thread\_on\_rq}(t->tid);} \\
+		\texttt{~~~~...} \\
+		\texttt{\}} \\
+		\end{tabular}
+		}
 	\end{center}
-	Your kernel needs to say when certain events happen:
+	Your kernel needs to say when certain events happen, for example:
 	\begin{itemize}
@@
 	\begin{center}
-		\includegraphics[width=0.8\textwidth]{config-landslide.png}
+		{\small
+		\begin{tabular}{l}
+		\texttt{\hilight{blue}{\# What test case to run?}} \\
+		\texttt{\hilight{black}{TEST\_CASE}=\hilight{olivegreen}{"double\_thread\_fork"}} \\
+		\texttt{} \\
+		\texttt{\hilight{blue}{\# The names of some important functions~~~~~~~~~~~}} \\
+		\texttt{\hilight{black}{TIMER\_WRAPPER}=\hilight{olivegreen}{"timer\_handler\_wrapper"}} \\
+		\texttt{\hilight{black}{CONTEXT\_SWITCH}=\hilight{olivegreen}{"ctx\_switch"}} \\
+		\texttt{} \\
+		\texttt{\hilight{blue}{\# What functions to pay attention to?}} \\
+		\texttt{within\_function \hilight{olivegreen}{"thread\_fork"}} \\
+		\texttt{within\_function \hilight{olivegreen}{"vanish"}} \\
+		\end{tabular}
+		}
 	\end{center}
-	Edit \texttt{config.landslide} with some details and tweaks
+	\linegap
 
-	Fill in two implementation-dependent C functions in Landslide ($\le$10 lines)
+	Fill out \texttt{config.landslide} with design details and names of important functions
+
+	% Fill in two implementation-dependent C functions in Landslide ($\le$10 lines)
 	% Say: Which depend on your kernel's specific implementation
@@
 
-\begin{frame}{Configuring Decision Points}
-	Landslide automatically identifies a minimal set of decision points.
+\begin{frame}{Configuring Preemption Points}
+	Landslide automatically identifies a minimal set of preemption points.
 	\begin{itemize}
@@
 	\begin{itemize}
-		\item Optional annotation: \texttt{tell\_landslide\_decide()}
+		\item Optional annotation: \texttt{tell\_landslide\_preempt()}
 		\item Hints to where a context switch should be forced.
@@
 	\begin{center}
-		\includegraphics[width=\textwidth]{found_a_bug.png}
+	{\small
+        \begin{tabular}{l}
+        \texttt{\hilight{red}{USE AFTER FREE - read from 0x199630 at eip 0x104264}} \\
+        \texttt{Heap contents:~\{...\}} \\
+        \texttt{[0x199630 | 4140] was allocated by TID3 at (...)} \\
+        \texttt{~~~~~~~~~~~~~~~~~~~~~~and freed by TID4 at (...)} \\
+        \texttt{\hilight{red}{****~~~~~~A bug was found!~~~~~****}} \\
+        \texttt{\hilight{red}{**** Preemption trace follows.~****}} \\
+        \texttt{\hilight{olivegreen}{1:~~TID 3 -{}-> TID 4}} \\
+        \texttt{~~~~TID3 at 0x105a10 in \hilight{blue}{context\_switch},} \\
+        \texttt{~~~~~~~~~~~~0x1041f4 in \hilight{blue}{thread\_fork},} \\
+        \texttt{~~~~~~~~~~~~0x10362b in \hilight{blue}{thread\_fork\_wrapper}} \\
+        \texttt{\hilight{olivegreen}{2:~~TID 4 -{}-> TID 3}} \\
+        \texttt{~~~~TID4 at 0x105a10 in \hilight{blue}{context\_switch},} \\
+        \texttt{~~~~~~~~~~~~0x104681 in \hilight{blue}{yield},} \\
+        \texttt{~~~~~~~~~~~~0x104570 in \hilight{blue}{vanish},} \\
+        \texttt{~~~~~~~~~~~~0x103708 in \hilight{blue}{vanish\_wrapper}} \\
+        \texttt{\hilight{red}{Current stack:}}\\
+        \texttt{~~~~TID3 at 0x104209 in \hilight{blue}{thread\_fork},} \\
+        \texttt{~~~~~~~~~~~~0x10362b in \hilight{blue}{thread\_fork\_wrapper}}
+        \end{tabular}
+	}
+
 	\end{center}
@@
 
-%\begin{frame}{Decision Points (noticing a theme here?)}
+%\begin{frame}{Preemption Points (noticing a theme here?)}
 %	Landslide is {\bf false negative}-oriented.
@@
 %				\item Maybe there was no race, or
-%				\item Maybe your decision points were not granular enough.
+%				\item Maybe your preemption points were not granular enough.
 %			\end{itemize}
@@
 %\end{frame}
-%\begin{frame}{Decision Points (noticing a theme here?)}
+%\begin{frame}{Preemption Points (noticing a theme here?)}
 %	Sometimes the default set will work.
@@
 			\item Average 112 minutes doing required instrumentation
-			\item Average 35 minutes configuring extra decision points \& interpreting ``found a bug'' output
+			\item Average 35 minutes configuring extra preemption points \& interpreting ``found a bug'' output
 		\end{itemize}
@@
 
-\begin{frame}{In Which Ben Offers Help - Warning}
+\begin{frame}{Words of Warning Again}
 	\textbf{If you are already struggling, this will not ``save'' you.}
@@
 
-\begin{frame}{In Which Ben Offers Help}
+\begin{frame}{Recruitment - Kernel Requirements}
 	Your kernel
@@
 			\texttt{readline}
-		\item Must never spin-wait (see hurdle form!)
+		\item Must pass the hurdle architectural requirements
+		\begin{itemize}
+			\item especially: must never spin-wait!
+		\end{itemize}
 		\item Should \texttt{assert()} important invariants
@@
 
-\begin{frame}{In Which Ben Offers Help}
+\begin{frame}{Recruitment}
 	User study this week, by appointment (\texttt{bblum@cs.cmu.edu})

### 2014-10-07 irb/infosheet.tex
add irb app

\documentclass{article}
%\usepackage{amsmath,amsthm,amssymb,fullpage,yfonts,graphicx,proof,subfig,wrapfig,appendix,hyperref,mdwlist,wasysym}
\usepackage{amsmath,amsthm,amssymb,fullpage,yfonts,graphicx,proof,appendix,hyperref,mdwlist,wasysym}
\usepackage{upgreek}
%\usepackage{times}
\usepackage[charter]{mathdesign}
\usepackage{hyperref}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{multirow}
\usepackage[usenames,dvipsnames]{xcolor}
%\usepackage{epsfig}
\usepackage[bottom]{footmisc}
%\usepackage{mjz-titlepage}
\usepackage{framed}
\usepackage{setspace}
%\setstretch{1.05}
\usepackage{subfig}
\usepackage{changebar}
\usepackage{colortbl}
\usepackage{wrapfig}

\newcommand\true{\;\textit{true}}
\newcommand\false{\;\textit{false}}

\newcommand\alpher\alpha
\newcommand\beter\beta
\newcommand\gammer\gamma
\newcommand\delter\delta
\newcommand\zeter\zeta
\newcommand\Sigmer\Sigma

\newcommand\NN{\mathbb{N}}
\newcommand\QQ{\mathbb{Q}}
\newcommand\RR{\mathbb{R}}
\newcommand\ZZ{\mathbb{Z}}

\begin{document}
%\captionsetup{width=.75\textwidth,font=small,labelfont=bf}

\title{\bf Landslide: Systematic Testing in 15-410 \\ User Study Information Sheet}
\author{Ben Blum}
\date{}
\maketitle

\section{Purpose of this Study}

{\em Systematic testing}, a novel technique for testing concurrent programs, offers a mechanical way to exhaustively explore large state spaces of thread interleavings.
The purpose of the study is to compare systematic testing against stress testing in the context of 15-410.
State-of-the-art systematic testing tools are fundamentally limited by exponential state space explosion: the user must manually configure the set of {\em preemption points} which define the state space in order to achieve meaningful test results within a reasonable CPU budget.
This high cognitive overhead makes such tools unsuitable for use in a classroom setting.
However, we have extended our systematic testing platform, known as Landslide, with several novel automation techniques.
We hope to show these techniques can make systematic testing just as accessible to students as conventional stress testing.
% This study will evaluate whether Landslide is useful as an educational tool.

\section{Risks and Benefits}

{\bf Risks.}
We expect the risks and discomfort associated with participation in this study to be no greater than those ordinarily encountered while working on 15-410 class projects.
As always with concurrent programming, it is possible to spend a lot of time looking for a bug with no meaningful result, while that time could otherwise have been spent on project implementation.
However, as Landslide is a research tool still under development, it is also possible that a bug in Landslide itself may produce a misleading error report. We expect you to protect yourself by responsibly allocating time between using Landslide and working on P2 normally.
\\

\noindent {\bf Benefits.}
Because we are providing you with a more powerful testing tool during P2 than you would normally have access to, we hope you will be able to find and fix bugs faster, resulting in a higher grade.

\section{Confidentiality}

During this study we will collect personally identifiable information from you as follows. While you use Landslide, it will automatically record your activity in terms of commands issued, time spent running tests, bug reports, and snapshots of your code. We will also collect paper surveys which you will fill out while using Landslide.
Carnegie Mellon may be required to disclose personally identifiable information as required by law, regulation, subpoena or court order.
Otherwise, we will protect your confidentiality as follows:

\begin{itemize}
	\item Data collected by Landslide will be stored in AFS accessible only to Ben Blum and Garth Gibson.
	\item Paper surveys will be stored in a locked filing cabinet accessible only to Ben Blum and Garth Gibson.
	\item Within one semester after the study, we will destroy all personally identifying information, keeping only anonymized and aggregate results.
	\item Professor Eckhardt will have access to study data only after final semester grades are submitted. Other members of course staff will never have access to study data. Course staff may, however, use Landslide on your final submitted code on their own.
\end{itemize}

\section{Your Rights}
There is no coercion or deception associated with this study.
Your participation is voluntary.
You are free to stop your participation at any point; likewise, the Principal Investigator may at his discretion remove you from the study.
Refusal to participate or discontinued participation in the study will not result in any penalty or loss of benefits or rights to which you might otherwise be entitled.
Your participation or lack thereof will have no impact in your 15-410 class grade, except as it might (hopefully) help you write better code.
\\

\noindent If you have any questions about this study, desire additional information, or wish to withdraw
your participation please contact the Principal Investigator by mail, phone or email.
If you have questions pertaining to your rights as a research participant,
or to report concerns about this study, you should contact the
Office of Research Integrity and Compliance at Carnegie Mellon University.
Email: {\tt irb-review@andrew.cmu.edu}. Phone: 412-268-1901 or 412-268-5460.

\section{Contact Information}

\begin{itemize}
	\item {\bf Ben Blum} (Principal Investigator): Email: {\tt bblum@cs.cmu.edu}; Office: GHC 9009; Phone: 412-304-4294
	\item {\bf Garth Gibson} (Faculty Advisor): Email: {\tt garth@cs.cmu.edu}; Office: GHC 9111
	\item {\bf David Eckhardt} (Course Instructor): Email: {\tt de0u@andrew.cmu.edu}; Office: GHC 7717
\end{itemize}

\end{document}

### 2014-10-07 irb/survey.tex
add irb app

\documentclass{article}
%\usepackage{amsmath,amsthm,amssymb,fullpage,yfonts,graphicx,proof,subfig,wrapfig,appendix,hyperref,mdwlist,wasysym}
\usepackage{amsmath,amsthm,amssymb,fullpage,yfonts,graphicx,proof,appendix,hyperref,mdwlist,wasysym}
\usepackage{upgreek}
%\usepackage{times}
\usepackage[charter]{mathdesign}
\usepackage{hyperref}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{multirow}
\usepackage[usenames,dvipsnames]{xcolor}
%\usepackage{epsfig}
\usepackage[bottom]{footmisc}
%\usepackage{mjz-titlepage}
\usepackage{framed}
\usepackage{setspace}
%\setstretch{1.05}
\usepackage{subfig}
\usepackage{changebar}
\usepackage{colortbl}
\usepackage{wrapfig}

\newcommand\true{\;\textit{true}}
\newcommand\false{\;\textit{false}}

\newcommand\alpher\alpha
\newcommand\beter\beta
\newcommand\gammer\gamma
\newcommand\delter\delta
\newcommand\zeter\zeta
\newcommand\Sigmer\Sigma

\newcommand\NN{\mathbb{N}}
\newcommand\QQ{\mathbb{Q}}
\newcommand\RR{\mathbb{R}}
\newcommand\ZZ{\mathbb{Z}}

\begin{document}
%\captionsetup{width=.75\textwidth,font=small,labelfont=bf}

\title{\bf Landslide: Systematic Testing in 15-410 \\ User Guide and Survey}
\author{Ben Blum}
\date{}
\maketitle

\section{Introduction}

Normally here I would write something researchy about systematic testing and preemption points, but if you're here it probably means you know what this is about. If you don't, please refer to the lecture material.
\\

\noindent Thanks for volunteering to try out Landslide. We have worked hard to make it nice enough to give to you, and we hope you're just as excited as we are to try out some systematic testing techniques the world has never seen before.
\\

\noindent You should have received a document titled {\em User Study Information Sheet} that describes your rights as a study participant and our confidentiality procedures. If you are missing this, please contact Ben immediately at {\tt bblum@cs.cmu.edu}.
% TODO: Host them on maxi or contrib and link to them.

\section{Instructions}

The very first thing you should do is {\bf READ THIS DOCUMENT ENTIRELY FIRST!} We need you to print out several survey sheets before you dive in.
\\

\noindent We encourage you to use Landslide on the GHC cluster or UNIX pool machines ({\tt ghcXX.ghc.andrew.cmu.edu} or {\tt unix.andrew.cmu.edu}). Landslide has not been tested with other configurations and will probably fail mysteriously on your personal laptop, and we cannot promise to be of any help.

\begin{enumerate}
	\item Clone the repository -- {\tt git clone https://github.com/bblum/landslide.git} -- this will be your workspace.
	\item Run {\tt ./p2-setup.sh PATH}, where {\tt PATH} is the absolute path to your p2 implementation.
		\begin{itemize}
			\item This command will import your p2 into {\tt ./pebsim/p2-basecode/} and then attempt to build it.
			\item {\tt PATH} should be the path with {\tt 410kern}, {\tt 410user}, {user/libthread}, etc as subdirectories, not the {\tt user/libthread} directory itself.
		\end{itemize}
	\item Next, make sure you have read this document entirely. At this point you should have several survey sheets printed out, ready to write on -- see the Survey section below. You should also have some means of timing yourself.
	\item Run {\tt ./landslide OPTIONS} to run tests with Landslide. {\bf Fill out a fresh survey sheet each time you do this.} Repeat until satisfied.
		\begin{itemize}
			\item Use {\tt ./landslide -h} to view command line options. The most important ones are {\tt -t}, max testing time, and {\tt -p}, which test program to run.
				% TODO: Remove this TODO when you have implemented glue to move the html files to the toplevel directory.
			\item Landslide's bug reports will show up as HTML files. If you are using a cluster machine in person, you can view them by pointing a browser to {\tt file:///PATH/TO/landslide/FILENAME.html}.
				\footnote{Another option: move it to {\tt \textasciitilde/www} in your AFS, and view it at {\tt http://www.contrib.andrew.cmu.edu/\textasciitilde{}YOUR\_ANDREWID/FILENAME.html}. Note the {\tt www} in the URL is mandatory; note also that this will allow the world to see it, so be careful!}
			\item The available test programs are {\tt thr\_exit\_join} (the default), {\tt broadcast\_test}, {\tt paraguay}, and {\tt rwlock\_downgrade\_read\_test}.
				\footnote{If you have your own custom test case, you can put it in {\tt pebsim/p2-basecode/user/progs} and edit \\ {\tt pebsim/p2-basecode/config.mk} to build it, then run {\tt ./p2-setup.sh} again.}
		\end{itemize}
\end{enumerate}

\noindent {\bf What to do when something goes wrong:}
If you find a Landslide bug, such as an assertion/crash, or such as a bug report that you think is actually a bug in Landslide rather than a bug in your P2, or such as it getting stuck and not exiting in the given time limit, please create a tarball of your workspace, just as you would to report a reference kernel bug, but send it to Ben instead of 410 staff.
For any other technical issues getting it to work, or if you have difficulty understanding a bug report, send email.
\\

\noindent
However, we {\em cannot} provide help on how to understand {\em why} your bug happened or how to fix it. The point of the study is that Landslide's automation plus your own brain should be enough! On the other hand, if you have a design question while deciding between potential ways to fix a bug, send it to 410 course staff as usual.

\newpage

\section{Survey}
\subsection{Overall Survey Questions}

\begin{center}
{\bf Print this page out once.}
\end{center}

\begin{enumerate}
	\item How much time did you spend directly using Landslide? This includes configuring test parameters and interpreting debug output. It does not include time spent running Landslide in the background, e.g. overnight.
		\vspace{0.5in}

	\item What test configurations did you use that DIDN'T find bugs? For example, {\tt ./landslide -t 15m -p thr\_exit\_join}
\end{enumerate}

\newpage
\subsection{Per-Bug Survey Questions}
\begin{center}
{\bf \large PRINT OUT THIS PAGE MULTIPLE TIMES.
\\

Each time you find a new bug with Landslide, fill this out again.}
\end{center}

\begin{enumerate}
		% TODO: find a way to allow ze studence to condense non-bug-finding runs onto 1 page.
	\item What test configuration did you use? For example, {\tt ./landslide -t 15m -p thr\_exit\_join}
		\vspace{0.5in}
	\item Summarize the bug.
		\vspace{0.5in}
	\item How IMPORTANT was this bug? Rate on a scale from 1 to 10, where:
		\begin{itemize}
			\item 1 = Trivial; an obvious ``oops'' that does not reflect any key lessons learned during the project. I would expect to not lose any points for this bug.
			\item 4 = Minor; a logic problem not related to threads/synchronization. I would expect to lose 1 or 2 points at most for this bug.
			\item 7 = Major; a concurrency-related bug related to subject matter discussed in lecture.
			\item 10 = Severe; this bug reflects a fundamental lesson about concurrency that we learned during the project. I would expect to receive ``see course staff'' in my ink if I turned in code with this bug!
		\end{itemize}
		\vspace{0.5in}
	\item How much EFFORT/TIME did this bug require to diagnose and fix? Rate on a scale from 1 to 10, where:
		\begin{itemize}
			\item 1 = Very little; we knew immediately what was wrong and needed to change only 1 or 2 lines of code.
			\item 5 = Some; the bug took considerable thought to fix, and required nontrivial code changes.
			\item 10 = Lots; we worked for several hours to fix the bug and/or required a major refactoring of our code.
		\end{itemize}
		\vspace{0.5in}
	\item How likely would a CONVENTIONAL STRESS TEST be able to find the bug (i.e., running the program repeatedly, not using Landslide)? Rate on a scale from 1 to 10, where:
		\begin{itemize}
			\item 1 = No hope. The window for a random timer interrupt to expose the bug is so small that a conventional stress test could easily run continuously overnight without encountering it.
			\item 5 = Moderate chance. Most random timer interrupts would not expose the bug, but the window is more than just a few instructions.
			\item 10 = Very likely. A random timer interrupt during the test would expose the bug more often than not.
		\end{itemize}
\end{enumerate}

\end{document}

### 2014-10-07 irb/survey.tex
fix typoo

@@
 			\item This command will import your p2 into {\tt ./pebsim/p2-basecode/} and then attempt to build it.
-			\item {\tt PATH} should be the path with {\tt 410kern}, {\tt 410user}, {user/libthread}, etc as subdirectories, not the {\tt user/libthread} directory itself.
+			\item {\tt PATH} should be the path with {\tt 410kern}, {\tt 410user}, {\tt user/libthread}, etc as subdirectories, not the {\tt user/libthread} directory itself.
 		\end{itemize}

### 2014-12-02 landslide-lecture/2014/lecture.tex
add in progress 410 lecture for f14

--- 2014-04-02 landslide-lecture/lecture.tex
@@
 	\begin{itemize}
-		\item Theorem proving
 		\item Unit tests
+		\begin{itemize}
+			\item good for basic functionality, bad for concurrency
+		\end{itemize}
 		\item Stress tests
 		\begin{itemize}
-			\item (state of the art in 15-410)
+			\item state of the art in 15-410
+		\end{itemize}
+		\item Theorem proving
+		\begin{itemize}
+			\item heavy burden on the programmers
 		\end{itemize}
@@
 
-\begin{frame}{Preemption Points} % TODO: duplicate slide with later
-	A {\bf preemption point} is\ldots
-
-	\linegap
-	A code location where being preempted causes different behaviour.
-	\begin{itemize}
-		\item Intuitively: Somewhere that interesting interleavings can happen around.
-	\end{itemize}
-	Examples:
-	\begin{itemize}
-		\item A new thread becomes runnable
-		\item Voluntary reschedule (e.g. \texttt{yield}, \texttt{cond\_wait})
-		\item Synchronization primitives
-	\end{itemize}
-	% Say: We'll revisit this later.
-\end{frame}
-
 % Say: Obviously we want to be able to always run the third path. But knowing
@@
 
-\begin{frame}{Systematic Testing}
-	\textbf{Systematic testing} is:
+\begin{frame}{Systematic Testing - The Big Picture}
+	\textbf{Goal: Force the system to execute every possible interleaving.}
+	% Say: checking for crashes/memory errors/etc in each one.
 	\begin{itemize}
-		\item Systematically enumerating different interleavings
+		\item At end of each branch, rewind system and restart test.
+		\item Artificially add preemptions to test different thread interleavings.
+		\item Intuitively: Generate many ``tabular execution traces''.
+	\end{itemize}
+	\pause
+	\linegap
+
+	Okay, wait a sec...
 	\begin{itemize}
-			\item Intuitively: Generate many ``tabular execution traces''
-		\end{itemize}
-		\item Exploring all branches in these trees
-		\begin{itemize}
-			\item (by controlling scheduling decisions at preemption points)
-		\end{itemize}
-		\item In practice: Depth-first search of branches
+		\item How can you possibly execute {\em every possible} interleaving?
+		\item How can you know where preempting matters most?
 	\end{itemize}
@@
 
-\subsection{Requirements}
-
-\begin{frame}{Execution Tree Exploration}
-	Important point: When does a branch end?
+\begin{frame}{Preemption Points}
+	\textbf{Preemption points} (PPs) are code location where being preempted may cause different behaviour.
 	\begin{itemize}
-		\item All threads run to completion, or
-		\item A bug is detected % Say: more on this later.
+		\item IOW, somewhere that interesting interleavings can happen around.
 	\end{itemize}
 	\linegap
-	{\bf Backtracking}:
+
+	Systematic tests are {\em parameterized} by the set of PPs.
 	\begin{itemize}
-		\item Identify a preemption point to replay differently
-		\item Reset machine state and start over
-		\item Replay test from the beginning, with different preemptions
+		\item If there are $n$ PPs and $k$ threads, state space size is $n^k$.
+		\item Need to choose the set of PPs very carefully for test to be effective.
 	\end{itemize}
@@
 
-\begin{frame}{More on Preemption Points}
-	Important point: What does ``all possible interleavings'' mean?
+\begin{frame}{Preemption Points}
+	What does ``all possible interleavings'' actually mean?
 
 	\linegap
-	One extreme: Decide at every instruction
+	One extreme: Preempt at every instruction
 	\begin{itemize}
@@
 
-\begin{frame}{More on Preemption Points}
+\begin{frame}{Preemption Points}
 	Sweet spot: Insert a thread switch everywhere it ``might matter''.
@@
 	\begin{itemize}
-		\item New threads becoming runnable (\texttt{fork}, \texttt{cond\_signal}, etc)
+		\item Threads becoming runnable (\texttt{fork()}, \texttt{cond\_signal()}, etc.)
 			\begin{itemize}
@@
 			\end{itemize}
-		\item Synchronization primitives (\texttt{mutex\_lock}, etc)
+		\item Synchronization primitives (\texttt{mutex\_lock()}/\texttt{unlock()}, etc.)
 			\begin{itemize}
@@
 
-	Finding the sweet spot is a joint effort between programmer and tool. (More on this later.)
+	%%%% Not anymore! :D
+	%Finding the sweet spot is a joint effort between programmer and tool. (More on this later.)
 \end{frame}
@@
 \subsection{Challenges}
+
+%% TODO -- left off working on slides here.
+
+% TODO TODO 2014: move this slide below the "preemption point" series, to after "controlling scheduling decisions", so it can be a two-parter -- "What are the two parts necessary to navigate a path through the tree?"
+\begin{frame}{Execution Tree Exploration}
+	Important point: When does a branch end?
+	\begin{itemize}
+		\item All threads run to completion, or
+		\item A bug is detected % Say: more on this later.
+	\end{itemize}
+	\linegap
+	{\bf Backtracking}:
+	\begin{itemize}
+		\item Identify a preemption point to replay differently
+		\item Reset machine state and start over
+		\item Replay test from the beginning, with different preemptions
+	\end{itemize}
+\end{frame}
 
@@
 
+% TODO TODO 2014: this slide can be deleted
 \begin{frame}{Memory Interposition}
@@
 
+% TODO TODO 2014: move this slide later, to be neighbours with the other words of warning
 \begin{frame}{Words of Warning}
@@
 
+% TODO TODO 2014: hopefully, update this with a tabular format in a semester's time!
 \begin{frame}{And Hopefully\dots}

### 2014-12-03 landslide-lecture/2014/lecture.tex
Finish landslide lecture for f14

@@
 \usepackage{array}
+\usepackage{colortbl}
 \setbeamertemplate{navigation symbols}{}
@@
 \institute[CMU 15-410]{Carnegie Mellon University - 15-410}
-\date[]{2014, April 2}
+\date[]{2014, December 3}
 
@@
 		\item Case study (example)
+		\item Tabular execution traces
 		\item The execution tree
-		\item Preemption points
 	\end{itemize}
@@
 	\begin{itemize}
-		\item Requirements
+		\item Preemption points
 		\item Challenges and feasibility
@@
 		\item How it works
-		\item The user-tool relationship
+		\item Coping with state space explosion
+		\item Automatically choosing preemption points
 		\item User study (that's you!)
@@
 
+\definecolor{thread1}{RGB}{87,172,255}
+\definecolor{thread2}{RGB}{255,201,102}
+
 \begin{frame}{Thread Interleavings (``good'' case)}
@@
 		\cline{1-2}
-		{\bf Thread 1} & {\bf Thread 2} & \\
+		\cellcolor{thread1} {\bf Thread 1} & \cellcolor{thread2} {\bf Thread 2} & \\
 		\cline{1-2}
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{Systematic Testing}
+
 \breakslide{\Large A different way of looking at race conditions\ldots}
@@
 % that ahead of time is impossible.
-% Say: Okay, now what if we could build a tool that could ``systematically''
-% explore this tree, and find the branch with the bug in it? What would such a
-% tool need to be able to do?
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{Systematic Testing}
-
-\breakslide{\Large Systematic Testing}
+% Say: In the rest of the lecture I'm going to refer to the "state space of
+% possible interleavings", simply referring to this tree, and "size" of the state
+% space refers to the number of branches.
 
@@
 	\begin{itemize}
+		\item On 1st execution, schedule threads arbitrarily until program ends.
 		\item At end of each branch, rewind system and restart test.
@@
 		\item How can you possibly execute {\em every possible} interleaving?
-		\item How can you know where preempting matters most?
+		%\item How can you know where preempting matters most?
+		\item How did you know to draw that tree's branches where they matter?
 	\end{itemize}
@@
 \begin{frame}{Preemption Points}
-	\textbf{Preemption points} (PPs) are code location where being preempted may cause different behaviour.
+	\textbf{Preemption points} (PPs) are code locations where being preempted may cause different behaviour.
 	\begin{itemize}
@@
 	\linegap
-	One extreme: Preempt at every instruction
+	{\bf One extreme}: Preempt at {\em every instruction}
 	\begin{itemize}
-		\item {\em Good news}: Will find every possible race condition.
-		\item {\em Bad news}: Runtime of test will be impossibly large.
+		\item Good news: Will find every possible race condition.
+		\item Bad news: Runtime of test will be impossibly large.
 	\end{itemize}
@@
 
-	Other extreme: Nothing is a preemption point
+	{\bf Other extreme}: {\em Nothing} is a preemption point
 	\begin{itemize}
-		\item {\em Good news}: Test will finish quickly.
-		\item {\em Bad news}: Only one execution was checked for bugginess.
+		\item Good news: Test will finish quickly.
+		\item Bad news: Only one execution was checked for bugginess.
 		\begin{itemize}
@@
 
-	Is there a ``sweet spot''?
+	Is there a {\bf ``sweet spot''}?
 \end{frame}
@@
 \begin{frame}{Preemption Points}
-	Sweet spot: Insert a thread switch everywhere it ``might matter''.
+	\textbf{Sweet spot}: Insert a thread switch everywhere it ``might matter''.
 
@@
 			\begin{itemize}
-				\item If used improperly\ldots
+				\item If buggy or used improperly\ldots
 			\end{itemize}
-		\item Unprotected shared memory accesses % Say: more on this later.
+		\item Unprotected shared memory accesses (``data races'')% Say: more on this later.
 			\begin{itemize}
@@
 	\end{itemize}
-	\linegap
 
@@
 	%Finding the sweet spot is a joint effort between programmer and tool. (More on this later.)
+	%However, all these points together may still be too expensive...
 \end{frame}
 
-\subsection{Challenges}
-
-%% TODO -- left off working on slides here.
-
-% TODO TODO 2014: move this slide below the "preemption point" series, to after "controlling scheduling decisions", so it can be a two-parter -- "What are the two parts necessary to navigate a path through the tree?"
-\begin{frame}{Execution Tree Exploration}
-	Important point: When does a branch end?
+\begin{frame}{Challenges}
+	Parameters of systematic tests must be kept small.
 	\begin{itemize}
-		\item All threads run to completion, or
-		\item A bug is detected % Say: more on this later.
-	\end{itemize}
-	\linegap
-	{\bf Backtracking}:
-	\begin{itemize}
-		\item Identify a preemption point to replay differently
-		\item Reset machine state and start over
-		\item Replay test from the beginning, with different preemptions
-	\end{itemize}
-\end{frame}
-
-\begin{frame}{Controlling Scheduling Decisions}
-	How do we control which thread runs when?
-	\linegap
-
-	In other words, what sources of nondeterminism must we control?
-
-	\begin{itemize}
-		\item Device interrupts/input
-			\begin{itemize}
-				\item Disk drivers: when disk reads finish
-				\item Ethernet drivers: when packets arrive
-			\end{itemize}
-		\item To control thread switches in a 410 kernel, vary when clock ticks happen.
-	\end{itemize}
-\end{frame}
-
-% TODO TODO 2014: this slide can be deleted
-\begin{frame}{Memory Interposition}
-	How do we detect race-induced heap corruption?
-	\linegap
-
-	In order to find use-after-free, need to know:
-	\begin{itemize}
-		\item When objects are \texttt{free()}d
-		\item When threads access shared memory in the heap
+		\item Test program length / number of threads
+		\item Number of preemption points used
 	\end{itemize}
@@
 
-	Solution: Keep track of all memory events
-	\begin{itemize}
-		\item All calls to \texttt{malloc}/\texttt{free}
-		\item All shared memory reads/writes
-	\end{itemize}
-\end{frame}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\subsection{State Space Explosion}
-
-\begin{frame}{State Space Explosion}
-	Serious problem: State spaces grow exponentially
-	\begin{itemize}
-		\item With $p$ preemption points and $k$ runnable threads, size $p^k$.
-		\item Threatens our ability to explore everything.
-		\item Fortunately, some sequences result in identical states.
-	\end{itemize}
-	\linegap
-
-	{\bf Partial Order Reduction} can help.
-	\begin{itemize}
-		\item Intuitively: After exploring each interleaving, identify ``equivalent'' alternatives and skip over them.
-		\item Example follows\dots
-	\end{itemize}
-\end{frame}
-
-\begin{frame}{State Space Explosion}
 	\begin{center}
-	\includegraphics[width=\textwidth]{undiamond0.png}
-	\end{center}
-\end{frame}
-\begin{frame}{State Space Explosion}
-	\begin{center}
-	\includegraphics[width=\textwidth]{undiamond1.png}
-	\end{center}
-\end{frame}
-\begin{frame}{State Space Explosion}
-	\begin{center}
-	\includegraphics[width=\textwidth]{diamond1.png}
+		\hspace{-0.25in}\includegraphics[width=0.3\textwidth]{expo.pdf} \\
+		\vspace{-1.33in}\hspace{-1.75in}{\rotatebox{90}{\small state space size}} \\
+		\vspace{0.15in}\hspace{-0.25in}{\small number of PPs}
 	\end{center}
@@
 \begin{frame}{About The Project}
-	% 5th year MS since June 2011 % Say: planning to finish this may, hopefully
-	My MS thesis project (2011-2012)
+	\textbf{About me}: 4th year graduate student, advised by Garth Gibson
 	\begin{itemize}
-		\item {}[CMU-CS-12-118 -- \url{http://bblum.net/thesis.pdf}]
+		\item TAed 15-410 for 3 semesters during undergrad
+		\item 1st graduate year was 5th year MS, rest in Ph.D. program
+		\begin{itemize}
+			\item \url{http://www.contrib.andrew.cmu.edu/~bblum/thesis.pdf}
 		\end{itemize}
+	\end{itemize}
+	\pause
+	\linegap
 
-	\linegap
-	Now I'm a Ph.D. student\dots
-
-	\linegap
-	Working with Garth Gibson, Jiri Simsa
-
-	\linegap
-	{\bf Landslide}: Shows that your Pebbles are not as stable as you thought.
+	{\bf About Landslide}
+	\begin{itemize}
+		\item Simics module, which traces:
+			\begin{itemize}
+				\item Every instruction executed
+				\item Every memory access read/written
+			\end{itemize}
+			% FIXME for next year -- how much slowdown is there? 20x?
+		\item Originally supported only P3s; can now test P2s as well
+		\item {\em Landslide} shows that your {\em Pebbles} are less stable than you thought.
+	\end{itemize}
 \end{frame}
@@
 
-\subsection{How Landslide Works}
+\subsection{Technical Overview}
 
-\begin{frame}{Landslide in Simics}
-	As a Simics module, Landslide knows:
+\begin{frame}{Big Picture: Execution Tree Exploration}
+	\textbf{Backtracking}
 	\begin{itemize}
-		\item Every instruction the kernel executes
-		\item Every memory address the kernel reads/writes
+		\item At end of each branch, identify a PP to replay differently
+		\item Reset machine state and start over
+		\item Implemented using Simics bookmarks
+			\begin{itemize}
+				\item {\tt set-bookmark} and {\tt skip-to}
 			\end{itemize}
+		\item Replay test from the beginning, with a different interleaving
+	\end{itemize}
+	\pause
 	\linegap
-	Artificially causes timer interrupts to control scheduling
 
-	\linegap
-	Checkpointing/backtracking via Simics bookmarks (reverse execution)
+	{\bf Controlling scheduling decisions}
+	\begin{itemize}
+		\item Tool must control all sources of nondeterminism
+		\item In 15-410, just timer and keyboard interrupts
+			% optional - say: "in real kernels, disk/network io interrupts"
+		\item Landslide repeatedly fires timer ticks until desired thread is run.
+	\end{itemize}
 \end{frame}
@@
 	\begin{itemize}
-		\item Memory leak
 		\item Infinite loop (halting problem)
+		\item Data races
 	\end{itemize}
+\end{frame}
+
+\begin{frame}{Brief Aside: Data Races}
+	A {\bf data race} is a pair of memory accesses between two threads, where:
+	\begin{itemize}
+		\item At least one of the accesses is a write
+		\item The threads are not holding the same mutex
+		\item The threads can be reordered (e.g. no \texttt{cond\_signal()} in between)
+	\end{itemize}
+	\pause
+	\linegap
+
+	Data races are {\em not necessarily} bugs, just highly suspicious!
+	\begin{itemize}
+		\item ``If threads interleaved the wrong way here, it might crash later.''
+		\begin{itemize}
+			\item Hmmm...
+		\end{itemize}
+	\end{itemize}
+\end{frame}
+
+\breakslide{
+	\begin{center}
+		\includegraphics[width=0.85\textwidth]{trace1.png}
+	\end{center}
+}
+\breakslide{
+	\begin{center}
+		\includegraphics[width=\textwidth]{trace2.png}
+	\end{center}
+}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\subsection{State Space Reduction}
+
+\begin{frame}{State Space Explosion}
+	Serious problem: State spaces grow exponentially
+	\begin{itemize}
+		\item With $p$ preemption points and $k$ runnable threads, size $p^k$.
+		\item Threatens our ability to explore everything.
+		\item Fortunately, some sequences result in identical states.
+	\end{itemize}
+	\linegap
+
+	{\bf Partial Order Reduction} identifies and skips ``equivalent'' interleavings.
+	\begin{itemize}
+		\item After each execution, compare memory reads/writes of each thread.
+		\item Find when reordering threads couldn't possibly change behaviour.
+		\item Example follows\dots
+	\end{itemize}
+\end{frame}
+
+\begin{frame}{State Space Reduction}
+	\begin{center}
+	\includegraphics[width=\textwidth]{undiamond0.png}
+	\end{center}
+\end{frame}
+\begin{frame}{State Space Reduction}
+	\begin{center}
+	\includegraphics[width=\textwidth]{undiamond1.png}
+	\end{center}
+\end{frame}
+\begin{frame}{State Space Reduction}
+	\begin{center}
+	\includegraphics[width=\textwidth]{diamond1.png}
+	\end{center}
 \end{frame}
@@
 
-\subsection{Using Landslide}
+\subsection{Iterative Deepening}
 
-\breakslide{\Large Using Landslide}
-
-\begin{frame}{Recruitment}
-	% Say: So, this wasn't just a research talk
-	This is something you can try!
+\begin{frame}{Choosing the Right Preemption Points}
+	So far we only {\em mitigated}, not {\em solved}, exponential explosion.
+	\pause
 	\linegap
 
-	Mutual benefit
+	State of the art tools hard-code a fixed set of preemption points.
 	\begin{itemize}
-		\item Landslide may help you find bugs
-		\item You help Ben evaluate his research
+		\item E.g., ``all thread library API calls'' or ``all kernel mutex locks/unlocks''
+		\item Depending on length of test, completion time is unpredictable.
+		\item More often, a subset is better in terms of time/coverage. 
+	\end{itemize}
+	\pause
+	\linegap
+	Current systematic testing model is not user-friendly.
+	\begin{itemize}
+		\item Tool: ``I want to use these PPs, but can't predict completion time.''
+		\item User: ``I have 16 CPUs and 24 hours to test my program.''
+	\end{itemize}
+	\linegap
+	{\em Stress testing offers this -- can we offer it too?}
+\end{frame}
+
+\begin{frame}{``Iterative Deepening'' (my current project)}
+	Goal: Run the best tests for a given CPU budget.
+	\linegap
+
+	Based on experience from past 15-410 student volunteers
+	\begin{itemize}
+		\item Students worked best with an iterative process
+		\item ``Start small, then add more preemption points as time allows''
+		\item Can we automate this process?
+	\end{itemize}
+	\linegap
+
+	Named after analogous technique in chess AI.
+	\begin{itemize}
+		\item Chess search is DFS limited by max number of moves (ply).
+		\item Chess AIs repeat DFS, increasing ply, until timeout.
 	\end{itemize}
@@
 
-% TODO TODO 2014: move this slide later, to be neighbours with the other words of warning
-\begin{frame}{Words of Warning}
-	Finding race conditions is hard for humans.
-
-	\linegap
-	It is hard for computer programs too.
-
-	\linegap
-	Landslide is not an oracle.
-\end{frame}
-
-\begin{frame}{Annotating Your Kernel}
-	Step 1
+\begin{frame}{Iterative Deepening}
+	Repeat state space explorations, adding preemption points, until time is exhausted.
+	\vspace{0.29in}
 	\begin{center}
-		{\small
-		\begin{tabular}{l}
-		\texttt{\hilight{olivegreen}{int}~thread\_fork() \{} \\
-	        \texttt{~~~~\hilight{olivegreen}{thread\_t}~*child = construct\_new\_thread();} \\
-		\texttt{~~~~\hilight{purple}{tell\_landslide\_forking}();~\hilight{blue}{/* A new thread was created.~*/}} \\
-	        \texttt{~~~~add\_to\_runqueue(child);} \\
-	        \texttt{~~~~return child->tid;} \\
-		\texttt{\}} \\
-		\texttt{\hilight{olivegreen}{void}~add\_to\_runqueue(\hilight{olivegreen}{thread\_t}~*t) \{} \\
-		\texttt{~~~~\hilight{blue}{/* This thread is now runnable.~*/}} \\
-		\texttt{~~~~\hilight{purple}{tell\_landslide\_thread\_on\_rq}(t->tid);} \\
-		\texttt{~~~~...} \\
-		\texttt{\}} \\
-		\end{tabular}
-		}
-	\end{center}
-	Your kernel needs to say when certain events happen, for example:
-	\begin{itemize}
-		\item When do threads become runnable / descheduled?
-		\item When does the scheduler switch threads?
-	\end{itemize}
-\end{frame}
-
-\begin{frame}{Configuring Landslide}
-	Step 2
-	\begin{center}
-		{\small
-		\begin{tabular}{l}
-		\texttt{\hilight{blue}{\# What test case to run?}} \\
-		\texttt{\hilight{black}{TEST\_CASE}=\hilight{olivegreen}{"double\_thread\_fork"}} \\
-		\texttt{} \\
-		\texttt{\hilight{blue}{\# The names of some important functions~~~~~~~~~~~}} \\
-		\texttt{\hilight{black}{TIMER\_WRAPPER}=\hilight{olivegreen}{"timer\_handler\_wrapper"}} \\
-		\texttt{\hilight{black}{CONTEXT\_SWITCH}=\hilight{olivegreen}{"ctx\_switch"}} \\
-		\texttt{} \\
-		\texttt{\hilight{blue}{\# What functions to pay attention to?}} \\
-		\texttt{within\_function \hilight{olivegreen}{"thread\_fork"}} \\
-		\texttt{within\_function \hilight{olivegreen}{"vanish"}} \\
-		\end{tabular}
-		}
-	\end{center}
-	\linegap
-
-	Fill out \texttt{config.landslide} with design details and names of important functions
-
-	% Fill in two implementation-dependent C functions in Landslide ($\le$10 lines)
-	% Say: Which depend on your kernel's specific implementation
-\end{frame}
-
-\begin{frame}{Configuring Preemption Points}
-	Landslide automatically identifies a minimal set of preemption points.
-	\begin{itemize}
-		\item It might find bugs.
-		\item It might overlook more fine-grained interleavings.
-	\end{itemize}
-	\linegap
-	With help from you, it could find more.
-	\begin{itemize}
-		\item Optional annotation: \texttt{tell\_landslide\_preempt()}
-		\item Hints to where a context switch should be forced.
-		\item Inside every call to \texttt{mutex\_lock}\ldots
-	\end{itemize}
-\end{frame}
-
-% TODO TODO 2014: hopefully, update this with a tabular format in a semester's time!
-\begin{frame}{And Hopefully\dots}
-	\begin{center}
-	{\small
-        \begin{tabular}{l}
-        \texttt{\hilight{red}{USE AFTER FREE - read from 0x199630 at eip 0x104264}} \\
-        \texttt{Heap contents:~\{...\}} \\
-        \texttt{[0x199630 | 4140] was allocated by TID3 at (...)} \\
-        \texttt{~~~~~~~~~~~~~~~~~~~~~~and freed by TID4 at (...)} \\
-        \texttt{\hilight{red}{****~~~~~~A bug was found!~~~~~****}} \\
-        \texttt{\hilight{red}{**** Preemption trace follows.~****}} \\
-        \texttt{\hilight{olivegreen}{1:~~TID 3 -{}-> TID 4}} \\
-        \texttt{~~~~TID3 at 0x105a10 in \hilight{blue}{context\_switch},} \\
-        \texttt{~~~~~~~~~~~~0x1041f4 in \hilight{blue}{thread\_fork},} \\
-        \texttt{~~~~~~~~~~~~0x10362b in \hilight{blue}{thread\_fork\_wrapper}} \\
-        \texttt{\hilight{olivegreen}{2:~~TID 4 -{}-> TID 3}} \\
-        \texttt{~~~~TID4 at 0x105a10 in \hilight{blue}{context\_switch},} \\
-        \texttt{~~~~~~~~~~~~0x104681 in \hilight{blue}{yield},} \\
-        \texttt{~~~~~~~~~~~~0x104570 in \hilight{blue}{vanish},} \\
-        \texttt{~~~~~~~~~~~~0x103708 in \hilight{blue}{vanish\_wrapper}} \\
-        \texttt{\hilight{red}{Current stack:}}\\
-        \texttt{~~~~TID3 at 0x104209 in \hilight{blue}{thread\_fork},} \\
-        \texttt{~~~~~~~~~~~~0x10362b in \hilight{blue}{thread\_fork\_wrapper}}
-        \end{tabular}
-	}
-
+		\includegraphics[width=0.25\textwidth]{tree0.pdf}
 	\end{center}
@@
 
-\breakslide{\Large Quick Demo}
+\begin{frame}{Iterative Deepening}
+	Repeat state space explorations, adding preemption points, until time is exhausted.
+	\vspace{0.15in}
+	\begin{center}
+		\begin{tabular}{cc}
+			\includegraphics[width=0.35\textwidth]{tree1.pdf} &
+			\includegraphics[width=0.35\textwidth]{tree2.pdf}
+		\end{tabular}
+	\end{center}
+\end{frame}
 
-% TODO more of these with screenshots
+\begin{frame}{Iterative Deepening}
+	Repeat state space explorations, adding preemption points, until time is exhausted.
+	\begin{center}
+		\includegraphics[width=0.5\textwidth]{tree3.pdf}
+	\end{center}
+	% say: and depending on your time budget, you might decide that you expect this one to be feasible, or you might fall back to a smaller state space from before.
+\end{frame}
 
-%\begin{frame}{Preemption Points (noticing a theme here?)}
-%	Landslide is {\bf false negative}-oriented.
-%	\begin{itemize}
-%		\item If Landslide reports a bug, there is a bug for sure.
-%			% Say: Well, except progress sense might be wrong.
-%		\item If not,
-%			\begin{itemize}
-%				\item Maybe there was no race, or
-%				\item Maybe your preemption points were not granular enough.
-%			\end{itemize}
-%	\end{itemize}
-%\end{frame}
-%\begin{frame}{Preemption Points (noticing a theme here?)}
-%	Sometimes the default set will work.
-%
-%	Sometimes you need to tweak\ldots
-%	\begin{itemize}
-%		\item Ignore certain global mutexes
-%		\item Whitelist or blacklist entire syscalls
-%		\item (get creative)
-%	\end{itemize}
-%	Intuitively: focusing Landslide on relevant parts of the kernel.
-%\end{frame}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\subsection{User Study}
-
-\begin{frame}{Last Year's Results}
+\begin{frame}{Iterative Deepening in Landslide}
+	\textbf{Landslide-ID}: wrapper program for Landslide
 	\begin{itemize}
-		\item Worked with 5 groups
+		\item Manages work queue of jobs with different PPs
+		\item Each job represents a new state space for Landslide to explore
+		\item Prioritizes which jobs are important / likely to finish
 		\begin{itemize}
-			\item \dots{}of which 4 put in enough time to make Landslide work
-		\end{itemize}
-		\linegap
-		\item {\bf Investment}: about 2 to 3 hours
-		\begin{itemize}
-			\item Average 112 minutes doing required instrumentation
-			\item Average 35 minutes configuring extra preemption points \& interpreting ``found a bug'' output
-		\end{itemize}
-		\pause
-		\linegap
-		\item {\bf Return}: All 4 groups found bugs in their kernels.
-		\begin{itemize}
-			\item All groups found deterministic bugs (e.g., use-after-free)
-			\item 2 groups found race conditions (i.e., Landslide needed to backtrack)
+			\item Based on nature of PPs (data races? mutexes?)
+			\item Based on estimated completion time
 		\end{itemize}
 	\end{itemize}
-\end{frame}
-
-\begin{frame}{Words of Warning Again}
-	\textbf{If you are already struggling, this will not ``save'' you.}
-	\begin{itemize}
-		\item False-negatives: not guaranteed to find races at all
-		\item Research-quality: possibly difficult to integrate with your kernel
-		\item Finishing the kernel project is more important.
-	\end{itemize}
-	\pause
-	\linegap
-	This is for you if:
-	\begin{itemize}
-		%\item Be expecting an A or a B
-		\item You have already submitted
-		\item You are using late days, but willing to work over Carnival
-		%\item Be able to turn in without late days (if you really had to)
-		\item You are taking p3extra, but you already pass the hurdle
-		\item You are looking for\ldots
-			\begin{itemize}
-				\item That ``one pesky race''
-				\item A race that stress-testing missed
-				\item Or just familiarity with a new technique
-			\end{itemize}
-	\end{itemize}
-\end{frame}
-
-\begin{frame}{Recruitment - Kernel Requirements}
-	Your kernel
-	\begin{itemize}
-		\item Must load the shell and run programs
-		\item \texttt{fork},
-			\texttt{exec},
-			\texttt{vanish},
-			\texttt{wait},
-			\texttt{readline}
-		\item Must pass the hurdle architectural requirements
-		\begin{itemize}
-			\item especially: must never spin-wait!
-		\end{itemize}
-		\item Should \texttt{assert()} important invariants
-		\begin{itemize}
-			\item Think of \texttt{panic()} as \texttt{tell\_landslide\_bug()}
-		\end{itemize}
-	\end{itemize}
-\end{frame}
-
-\begin{frame}{Recruitment}
-	User study this week, by appointment (\texttt{bblum@cs.cmu.edu})
 
 	\linegap
-	Expect to spend:
-	\begin{itemize}
-		\item Up to 3 hours, just to try it out.
-		\item 4-6 hours, if you find a bug and track it down.
-		\item More, for multiple bugs or the truly dedicated\ldots
-	\end{itemize}
-	\linegap
-
-	Give feedback (intuitive? frustrating? found bugs?)
-
-	\linegap
-	{\bf Watch \texttt{410.announce} for details!}
+	{\em Only required argument is CPU budget}
 \end{frame}
-
-\subsection{End}
-\breakslide{\Large Questions?}
 
@@
 
-% \section{Conclusion}
-% \subsection{Conclusion}
+\section{Evaluation}
+\breakslide{\Large Evaluation}
+
+\begin{frame}{Preliminary Evaluation}
+	We tested 22 random submitted P2s from Spring 2014.
+	\begin{itemize}
+		\item Test cases:
+		\begin{itemize}
+			\item {\tt thr\_exit\_join}
+			\item {\tt paraguay}
+			\item {\tt rwlock\_downgrade\_read\_test}
+			\item {\tt broadcast\_test} (new)
+		\end{itemize}
+		\item Each test given CPU budget of 11 cores $\times$ 10 minutes.
+	\end{itemize}
+\end{frame}
+
+\begin{frame}{Preliminary Results}
+	88 total tests were run (9,680 total CPU-minutes)
+	\linegap
+
+	Found 7 {\bf deterministic} initialization bugs (e.g. use-after-free)
+	\linegap
+
+	56 tests ran to completion without finding bugs:
+	\begin{itemize}
+		\item 47 exhausted the CPU budget.
+		\item 9 completed all state spaces before time ran out.
+	\end{itemize}
+	\linegap
+
+	Between 20 and 33 {\bf non-deterministic} bugs were found:
+	\begin{itemize}
+		\item 20 deadlock races among 12 P2s
+		\item 10 use-after-free races among 6 P2s
+		\item 3 misc (assert fail, segfault) among 2 P2s
+	\end{itemize}
+\end{frame}
+
+\begin{frame}{Recruitment}
+	\textbf{User study}: Come try Landslide on your P2!
+	\begin{itemize}
+		\item Bare minimum effort: No more than 30 minutes
+		\begin{itemize}
+			\item Clone a github URL, run setup script, run, send results
+		\end{itemize}
+		\item Testing P2s is fully automated, but can also test P3s
+		\begin{itemize}
+			\item Landslide requires manual kernel annotations
+			\item Comes with pre-annotated reference kernel for P2
+			\item Past students take approx. 1 hour to annotate own P3s.
+		\end{itemize}
+	\end{itemize}
+	\pause
+	\linegap
+
+	{\bf Ultimate goal}: Provide Landslide to every 410 student during P2.
+	\begin{itemize}
+		\item This semester: Prep \& testing
+			\begin{itemize}
+				\item Improving user interface
+				\item Making sure Landslide itself is stable
+			\end{itemize}
+		%\item Please help us!
+		\item {\bf \large Just for fun -- this will NOT impact your grade.}
+	\end{itemize}
+\end{frame}
+
+\begin{frame}{Recruitment}
+	%User study this week, by appointment (\texttt{bblum@cs.cmu.edu})
+	User study during exam week, by appointment (\texttt{bblum@cs.cmu.edu})
+	\linegap
+
+%	Recommended to use fully-automated P2 test suite, but you can try testing your P3 with approx. 1 hour of manual annotation.
+%	\linegap
 %
-% \begin{frame}{Summary}
-% 	\begin{itemize}
-% 		\item
-% 	\end{itemize}
-% \end{frame}
+	%Expect to spend:
+	%\begin{itemize}
+	%	\item Up to 3 hours, just to try it out.
+	%	\item 4-6 hours, if you find a bug and track it down.
+	%	\item More, for multiple bugs or the truly dedicated\ldots
+	%\end{itemize}
+	%\linegap
+
+	Give feedback (intuitive? frustrating? found bugs?)
+
+	\linegap
+	Watch your email for details!
+\end{frame}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{End}
+\breakslide{\Large Questions?
+\linegap
+
+\begin{center}
+	\includegraphics[width=0.65\textwidth]{3word-questions.png}
+\end{center}}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{Bonus Slide(s)}
+
+\begin{frame}{More Results}
+	\textbf{Is dynamically adding ``data race'' PPs effective?}
+	\begin{itemize}
+		\item 3 among reported bugs (9-15\%) required data races to expose.
+	\end{itemize}
+	\linegap
+
+	{\bf Is ``iterative deepening'' better than state-of-the-art?}
+	\begin{itemize}
+		\item Control experiment: Just 1 state space, same CPU time
+		\begin{itemize}
+			\item PPs used: \texttt{mutex\_lock()}, \texttt{mutex\_unlock()}
+			\item 110 minutes on 1 CPU
+		\end{itemize}
+		\item Of 33 total bug reports, control failed to find 10 (30\%).
+		\begin{itemize}
+			\item 3 required data race PPs to expose
+			\item 1 ran out of time
+			\item 6 obscured by different bug in same state space
+		\end{itemize}
+	\end{itemize}
+\end{frame}
 
 \end{document}
-%\begin{frame}{Partial Order Reduction}
-%	\begin{center}
-%	The next 6 slides won't be on the exam.
-%
-%		\Huge \dbend
-%	\end{center}
-%\end{frame}
-%
-%\begin{frame}{Partial Order Reduction}
-%	A {\bf transition} is\ldots
-%	\begin{itemize}
-%		\item The sequence of instructions executed between choice points.
-%	\end{itemize}
-%	To express concurrency between transitions:
-%	% Say: As a prerequisite for this algorithm, we need to build two sets to express the concurrency relations between the transitions
-%	\begin{itemize}
-%		\item Independence relation
-%		\item Happens-before relation
-%	\end{itemize}
-%	% Say: This is hard, and I don't expect you to understand it during this lecture; you won't be tested on it or anything.
-My goal in presenting it is to convey how intensely cool the algorithm is, and hopefully you will keep thinking about it later.
-I'd be happy to answer questions about it later.
-%\end{frame}
-%
-%\begin{frame}[fragile]{Partial Order Reduction - Independence Relation}
-%	\begin{columns}[l]
-%		\column{0.5\textwidth}
-%		\begin{center}
-%		\begin{verbatim}
-%		void thread_A() {
-%		    foo = 42;
-%		}
-%		void thread_B() {
-%		    if (bar == 1337)
-%		        print "Hello"
-%		}
-%		\end{verbatim}
-%		\end{center}
-%		\linegap
-%
-%		$A$ and $B$ are {\bf independent}.
-%		\column{0.5\textwidth}
-%		\begin{center}
-%		\begin{verbatim}
-%		void thread_A() {
-%		    foo = 42;
-%		}
-%		void thread_C() {
-%		    if (foo == 1337)
-%		        print "Hello"
-%		}
-%		\end{verbatim}
-%		\end{center}
-%		\linegap
-%
-%		$A$ and $C$ {\bf conflict}.
-%	\end{columns}
-%\end{frame}
-%
-%\begin{frame}{Partial Order Reduction - Independence Relation}
-%	Consider a trace of all shared memory reads/writes
-%
-%	\linegap
-%	Two transitions (instruction sequences) $A$ and $B$ are independent iff:
-%	\begin{itemize}
-%		\item $Writes(A) \cap Writes(B) = \emptyset$
-%		\item $Writes(A) \cap Reads(B) = \emptyset$
-%		\item $Reads(A) \cap Writes(B) = \emptyset$
-%	\end{itemize}
-%\end{frame}
-%
-%\begin{frame}{Partial Order Reduction - Happens-before Relation}
-%	\begin{columns}[l]
-%		\column{0.5\textwidth}
-%		\includegraphics[width=\textwidth]{hb.png}
-%		\column{0.5\textwidth}
-%		Is a transition ``enabled'' by a previous one?
-%		\begin{itemize}
-%			\item $A$ enables $B$
-%			\item $B$ enables $C$ (same thread)
-%			\item $B$ enables $D$
-%		\end{itemize}
-%%		Transitive closure
-%%		\begin{itemize}
-%%			\item $HB^* = \{AB, AC, AD, BC, BD\}$
-%%		\end{itemize}
-%		$C$ and $D$ are {\bf concurrent}!
-%	\end{columns}
-%\end{frame}
-%
-%\begin{frame}{Partial Order Reduction - Tagging}
-%	Run POR at the end of each branch
-%	\begin{itemize}
-%		\item Identify which branches to explore next
-%		\item Tag only a subset of all branches (hopefully)
-%	\end{itemize}
-%\end{frame}
-%
-%\begin{frame}{Partial Order Reduction - Tagging}
-%	For each transition $T$, find an {\bf evil ancestor} $E$:
-%	\begin{itemize}
-%		\item $E$ and $T$ are concurrent.
-%		\item $E$ and $T$ are not independent.
-%		\item Intuitively: $T$ and $E$ could be reordered, and might behave differently.
-%	\end{itemize}
-%	Find a {\bf good sibling} $G$ of $E$:
-%	\begin{itemize}
-%		\item $G$ is the same thread as $T$
-%		% Say: (if no such sibling, all siblings)
-%		\item Intuitively: Find a branch in which $T$ runs before $E$.
-%	\end{itemize}
-%	Explore tagged branches depth-first.
-%\end{frame}
-%
-%\begin{frame}{Partial Order Reduction}
-%	\begin{center}
-%		{\Huge \dbend}
-%		\linegap
-%		\linegap
-%		\linegap
-%
-%		Feel free to ask me about this later.
-%	\end{center}
-%\end{frame}
-
-% TODO: run through an example
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%\subsection{Other Dynamic Techniques}
-%
-%\begin{frame}{Heuristic-Based Searching}
-%	% Say: Note that this is beyond the scope of my personal project.
-%	Maybe the tree is too big to search anyway?
-%
-%	\linegap
-%	Search {\em most likely} branches first.
-%	\begin{itemize}
-%		\item ``Iterative Context Bounding''
-%		\footnote{Microsoft Research - CHESS - Musuvathi et al.}
-%		\begin{itemize}
-%			\item Insight: Races tend to show up with few forced context switches.
-%			\item Search branches with fewer preemptions first.
-%		\end{itemize}
-%		\item Find behaviours that are typical of buggy software.
-%		\footnote{Stanford University - Dawson Engler}
-%		\begin{itemize}
-%			\item Different behaviour in different branches.
-%				% Say: different all the time: no. different rarely: yes.
-%			\item Use suspicious memory accesses as future choice points.
-%		\end{itemize}
-%	\end{itemize}
-%	\linegap
-%	Open research field, possibly my future work.
-%\end{frame}
-%
-%% Say: I'm going to take a break from talking about systematic exploration to talk about other types of dynamic race detection - FIXME: if these slides fit in 60min.
-%\begin{frame}{Data Race Detection}
-%	Requires happens-before set and ``lock set''
-%	\linegap
-%
-%	Two memory accesses $m_1$ and $m_2$ are a data race iff:
-%	\begin{itemize}
-%		\item Read/write or write/write to the same address, and
-%		\item $(m_1,m_2) \not\in HB$, and
-%		\item $Lock(m_1) \cap Lock(m_2) = \emptyset$
-%	\end{itemize}
-%\end{frame}
-%\begin{frame}{Data Race Detection}
-%	Data race detection vs systematic exploration
-%	\begin{itemize}
-%		\item More false positives
-%			\begin{itemize}
-%				\item (but still good style-checking!)
-%			\end{itemize}
-%		\item No false negatives
-%		\item Still unable to find some types of bugs
-%	\end{itemize}
-%	\linegap
-%	Hybrid approach? (future research)
-%\end{frame}

### 2014-12-10 irb/survey.tex
Update infosheet and survey for landslide study

@@
 
-Normally here I would write something researchy about systematic testing and preemption points, but if you're here it probably means you know what this is about.
+%Normally here I would write something researchy about systematic testing and preemption points, but if you're here it probably means you know what this is about.
 If you don't, please refer to the lecture material.
+%\\
+%
+%\noindent Thanks for volunteering to try out Landslide.
+We have worked hard to make it nice enough to give to you, and we hope you're just as excited as we are to try out some systematic testing techniques the world has never seen before.
+%\\
+
+\noindent You should have received a document titled {\em User Study Information Sheet} that describes your rights as a study participant and our confidentiality procedures.
+If you are missing this, please contact me immediately.
+My email address is {\tt bblum@cs.cmu.edu} and my phone number is 412-304-4294; email is preferred.
+% TODO: Host them on maxi or contrib and link to them.
 \\
 
-\noindent Thanks for volunteering to try out Landslide.
-We have worked hard to make it nice enough to give to you, and we hope you're just as excited as we are to try out some systematic testing techniques the world has never seen before.
-\\
-
-\noindent You should have received a document titled {\em User Study Information Sheet} that describes your rights as a study participant and our confidentiality procedures.
-If you are missing this, please contact Ben immediately at {\tt bblum@cs.cmu.edu}.
-% TODO: Host them on maxi or contrib and link to them.
+\noindent Likewise, please make sure you have read and understand the lecture slides on systematic testing, which are available on the 15-410 course website.
+If you have any questions about the research background I would be happy to answer them.
 
@@
 Note the {\tt www} in the URL is mandatory; note also that this will allow the world to see it, so be careful!}
+				\footnote{If you have your own custom test case, you can put it in {\tt pebsim/p2-basecode/user/progs} and edit \\ {\tt pebsim/p2-basecode/config.mk} to build it, then run {\tt ./p2-setup.sh} again.}
 			\item The available test programs are {\tt thr\_exit\_join} (the default), {\tt broadcast\_test}, {\tt paraguay}, and {\tt rwlock\_downgrade\_read\_test}.
-				\footnote{If you have your own custom test case, you can put it in {\tt pebsim/p2-basecode/user/progs} and edit \\ {\tt pebsim/p2-basecode/config.mk} to build it, then run {\tt ./p2-setup.sh} again.}
+			\item Suggested test configurations to get you started are as follows.
+			\begin{itemize}
+				\item {\tt ./landslide} --- Run the default test program ({\tt thr\_exit\_join}) for the default amount of time (one hour).
+				\item {\tt ./landslide -p broadcast\_test -t 600} --- Run {\tt broadcast\_test} for 600 seconds (10 minutes).
+				\item {\tt ./landslide -p paraguay -t 12h -c 4} --- Run {\tt paraguay} overnight (12 hours), using at most 4 CPUs at a time.
+			\end{itemize}
 		\end{itemize}
@@
 \noindent {\bf What to do when something goes wrong:}
-If you find a Landslide bug, such as an assertion/crash, or such as a bug report that you think is actually a bug in Landslide rather than a bug in your P2, or such as it getting stuck and not exiting in the given time limit, please create a tarball of your workspace, just as you would to report a reference kernel bug, but send it to Ben instead of 410 staff.
+If you find a Landslide bug, such as an assertion/crash, or such as a bug report that you think is actually a bug in Landslide rather than a bug in your P2, or such as it getting stuck and not exiting in the given time limit, please create a tarball of your workspace, just as you would to report a reference kernel bug, but send it to me instead of 410 staff.
 For any other technical issues getting it to work, or if you have difficulty understanding a bug report, send email.

### 2015-01-16 irb/survey.tex
update irb forms slightly

--- 2014-10-07 irb/survey.tex
@@
 
-Normally here I would write something researchy about systematic testing and preemption points, but if you're here it probably means you know what this is about.
+%Normally here I would write something researchy about systematic testing and preemption points, but if you're here it probably means you know what this is about.
 If you don't, please refer to the lecture material.
-\\
-
-\noindent Thanks for volunteering to try out Landslide.
+%\noindent
+Thanks for volunteering to try out Landslide.
 We have worked hard to make it nice enough to give to you, and we hope you're just as excited as we are to try out some systematic testing techniques the world has never seen before.
@@
 If you are missing this, please contact Ben immediately at {\tt bblum@cs.cmu.edu}.
+If you have not yet reviewed the lecture slides (found on the course website), please do so before proceeding.
 % TODO: Host them on maxi or contrib and link to them.

### 2015-02-04 irb/survey.tex
final updates to irb survey.tex

--- 2014-12-10 irb/survey.tex
@@
 			\item The available test programs are {\tt thr\_exit\_join} (the default), {\tt broadcast\_test}, {\tt paraguay}, and {\tt rwlock\_downgrade\_read\_test}.
-			\item Suggested test configurations to get you started are as follows.
+			\item The suggested sequence of test configurations to run is as follows.
+You can, of course, change the time limits or test programs as you see fit.
+For the longer-running configurations, it's recommended to run them inside of a {\tt screen} session.
 			\begin{itemize}
-				\item {\tt ./landslide} --- Run the default test program ({\tt thr\_exit\_join}) for the default amount of time (one hour).
-				\item {\tt ./landslide -p broadcast\_test -t 600} --- Run {\tt broadcast\_test} for 600 seconds (10 minutes).
-				\item {\tt ./landslide -p paraguay -t 12h -c 4} --- Run {\tt paraguay} overnight (12 hours), using at most 4 CPUs at a time.
+				\item {\tt ./landslide -p thr\_exit\_join -t 30m}
+				\item {\tt ./landslide -p broadcast\_test -t 30m}
+				\item {\tt ./landslide -p paraguay -t 30m}
+				\item {\tt ./landslide -p rwlock\_downgrade\_read\_test -t 30m}
+				\item {\tt ./landslide -p thr\_exit\_join -t 3h}
+				\item {\tt ./landslide -p broadcast\_test -t 3h}
+				\item {\tt ./landslide -p paraguay -t 3h}
+				\item {\tt ./landslide -p rwlock\_downgrade\_read\_test -t 3h}
+				\item {\tt ./landslide -p thr\_exit\_join -t 12h}
+				\item {\tt ./landslide -p broadcast\_test -t 12h}
+				\item {\tt ./landslide -p paraguay -t 12h}
+				\item {\tt ./landslide -p rwlock\_downgrade\_read\_test -t 12h}
 			\end{itemize}
@@
 \noindent {\bf What to do when something goes wrong:}
-If you find a Landslide bug, such as an assertion/crash, or such as a bug report that you think is actually a bug in Landslide rather than a bug in your P2, or such as it getting stuck and not exiting in the given time limit, please create a tarball of your workspace, just as you would to report a reference kernel bug, but send it to me instead of 410 staff.
-For any other technical issues getting it to work, or if you have difficulty understanding a bug report, send email.
+If you find a Landslide bug, such as an assertion/crash, or such as a bug report that you think is actually a bug in Landslide rather than a bug in your P2, or such as it getting stuck and not exiting in the given time limit, please create a tarball of your workspace, just as you would to report a reference kernel bug, but send it to Ben (\texttt{bblum@cs.cmu.edu}) instead of 410 staff.
+For any other technical issues getting it to work, or if you have difficulty understanding a bug report, send email to Ben.
 \\
@@
 	\item Summarize the bug.
-		\vspace{0.5in}
+		\vspace{2in}
 	\item How IMPORTANT was this bug? Rate on a scale from 1 to 10, where:
@@
 		\end{itemize}
-		\vspace{0.5in}
+		\vspace{0.2in}
 	\item How much EFFORT/TIME did this bug require to diagnose and fix? Rate on a scale from 1 to 10, where:
@@
 		\end{itemize}
-		\vspace{0.5in}
+		\vspace{0.2in}
 	\item How likely would a CONVENTIONAL STRESS TEST be able to find the bug (i.e., running the program repeatedly, not using Landslide)? Rate on a scale from 1 to 10, where:

### 2015-02-06 landslide-lecture/2015/lecture.tex
Add rough draft of landslide lecture for p2 for s15.

--- 2014-12-03 landslide-lecture/2014/lecture.tex
@@
 
-\title[Landslide]{{\bf Systematic Testing: A New Perspective on Races}}
-\subtitle[]{ {\em more clever than } ``\texttt{slaughter cho2}'' {\em since 2011.}}
+\title[Landslide]{{\bf Landslide: A New Race-Finding Tool for 15-410}}
+\subtitle[]{ {\em more clever than } ``\texttt{mandelbrot}'' {\em since 2011.}}
 \author[Ben Blum]{Ben Blum \texttt{(bblum@andrew.cmu.edu)}}
@@
 \begin{document}
+\renewcommand{\inserttotalframenumber}{41}
 \normalem
@@
 
-% Say: Since I'm no longer a member of course staff, I'm allowed to give away a
-% race condition that some of you might have, for the sake of example.
-
-\begin{frame}[fragile]{Case Study}
-	\begin{center}
-	\begin{verbatim}
-	    int thread_fork()
-	    {
-	        thread_t *child = construct_new_thread();
-	        add_to_runqueue(child);
-	        return child->tid;
-	    }
-	\end{verbatim}
-	\end{center}
-
-%	\begin{itemize}
-%		\item On vanish, child's TCB is freed
-%		\item Forking thread does use-after-free
-%		\item Might return garbage instead of TID
-%	\end{itemize}
-\end{frame}
-
 \definecolor{thread1}{RGB}{87,172,255}
 \definecolor{thread2}{RGB}{255,201,102}
+\definecolor{thread3}{RGB}{255,87,130}
 
+\begin{frame}{Case Study}
+	\begin{center}
+		\begin{tabular}{llll}
+			\begin{tabular}{l}
+				Consumer thread \\
+				\hline
+				\\
+				\texttt{LOCK(mx);} \\
+				\\
+				\texttt{if~(!work\_exists())} \\
+				\texttt{~~~~WAIT(cvar,~mx);} \\
+				\texttt{work~=~claim\_work();} \\
+				\\
+				\texttt{UNLOCK(mx);} \\
+				\texttt{assert(work~!=~NULL);} \\
+			\end{tabular}
+			& & &
+			\begin{tabular}{l}
+				Producer thread \\
+				\hline
+				\\
+				\texttt{LOCK(mx);} \\
+				\\
+				\texttt{give\_work(work);} \\
+				\texttt{SIGNAL(cvar);} \\
+				\\
+				\texttt{UNLOCK(mx);} \\
+				\\
+				\\
+			\end{tabular}
+		\end{tabular}
+		\linegap
+
+		\begin{itemize}
+			\item See {\bf Paradise Lost} lecture!
+			\item {\tt if} vs {\tt while}: Two consumers can race to make one fail.
+		\end{itemize}
+	\end{center}
+\end{frame}
 \begin{frame}{Thread Interleavings (``good'' case)}
-	\begin{tabular}{|l|l|l}
-		\cline{1-2}
-		\cellcolor{thread1} {\bf Thread 1} & \cellcolor{thread2} {\bf Thread 2} & \\
-		\cline{1-2}
-		\texttt{spawn\_new\_thread} && \\
-		\cline{1-2}
-		\texttt{add\_to\_runqueue} && (new thread) \\
-		\cline{1-2}
-		\texttt{return child->tid} &&  \\
-		\cline{1-2}
-		& \texttt{vanish} & \\
-		\cline{1-2}
-		& (TCB gets freed) & (voluntary reschedule) \\
-		\cline{1-2}
+	\begin{center}
+		\begin{tabular}{|l|l|l|}
+			\hline
+			\cellcolor{thread1} {\bf Thread 1} & \cellcolor{thread2} {\bf Thread 2} & \cellcolor{thread3} {\bf Thread 3} \\
+			\hline
+			\small \texttt{LOCK(mx);} & & \\
+			\small \texttt{if~(!work\_exists())} & & \\
+			\small \texttt{~~~~WAIT(cvar,~mx);} & & \\
+
+			& \small \texttt{LOCK(mx);} & \\
+			& \small \texttt{give\_work(work);} & \\
+			& \small \texttt{SIGNAL(cvar);} & \\
+			& \small \texttt{UNLOCK(mx);} & \\
+
+			\small \texttt{work~=~claim\_work();~} & & \\
+			\small \texttt{UNLOCK(mx);} & & \\
+			\small \texttt{//~got~the~work} {\large \smiley}& & \\
+
+			& & \small \texttt{LOCK(mx);} \\
+			& & \small \texttt{if~(!work\_exists())~} \\
+			& & \small \texttt{~~~~WAIT(cvar,~mx);} \\
+			\hline
 		\end{tabular}
+	\end{center}
 \end{frame}
+\begin{frame}{Thread Interleavings (``good'' case, again)}
+	\begin{center}
+		\begin{tabular}{|l|l|l|}
+			\hline
+			\cellcolor{thread1} {\bf Thread 1} & \cellcolor{thread2} {\bf Thread 2} & \cellcolor{thread3} {\bf Thread 3} \\
+			\hline
+			\small \texttt{LOCK(mx);} & & \\
+			\small \texttt{if~(!work\_exists())} & & \\
+			\small \texttt{~~~~WAIT(cvar,~mx);} & & \\
 
+			& & \small \texttt{LOCK(mx);} \\
+			& & \small \texttt{if~(!work\_exists())~} \\
+			& & \small \texttt{~~~~WAIT(cvar,~mx);} \\
+
+			& \small \texttt{LOCK(mx);} & \\
+			& \small \texttt{give\_work(work);} & \\
+			& \small \texttt{SIGNAL(cvar);} & \\
+			& \small \texttt{UNLOCK(mx);} & \\
+
+			\small \texttt{work~=~claim\_work();~} & & \\
+			\small \texttt{UNLOCK(mx);} & & \\
+			\small \texttt{//~got~the~work} {\large \smiley}& & \\
+			\hline
+		\end{tabular}
+	\end{center}
+\end{frame}
 \begin{frame}{Thread Interleavings (race condition)}
-	\begin{tabular}{|l|l|l}
-		\cline{1-2}
-		\cellcolor{thread1} {\bf Thread 1} & \cellcolor{thread2} {\bf Thread 2} & \\
-		\cline{1-2}
-		\texttt{spawn\_new\_thread} && \\
-		\cline{1-2}
-		\texttt{add\_to\_runqueue} && (new thread + preempted) \\
-		\cline{1-2}
-		& \texttt{vanish} & \\
-		\cline{1-2}
-		& (TCB gets freed) & (voluntary reschedule) \\
-		\cline{1-2}
-		\texttt{return child->tid} && (bad!) \\
-		\cline{1-2}
+	\begin{center}
+		\begin{tabular}{|l|l|l|}
+			\hline
+			\cellcolor{thread1} {\bf Thread 1} & \cellcolor{thread2} {\bf Thread 2} & \cellcolor{thread3} {\bf Thread 3} \\
+			\hline
+			\small \texttt{LOCK(mx);} & & \\
+			\small \texttt{if~(!work\_exists())} & & \\
+			\small \texttt{~~~~WAIT(cvar,~mx);} & & \\
+
+			& \small \texttt{LOCK(mx);} & \\
+			& \small \texttt{give\_work(work);} & \\
+			& \small \texttt{SIGNAL(cvar);} & \\
+			& \small \texttt{UNLOCK(mx);} & \\
+
+			& & \small \texttt{LOCK(mx);} \\
+			& & \small \texttt{work~=~claim\_work();} \\
+			& & \small \texttt{UNLOCK(mx);} \\
+
+			\small \texttt{work~=~claim\_work();~} & & \\
+			\small \texttt{UNLOCK(mx);} & & \\
+			\small \texttt{//~work~was~stolen} {\large \frownie}& & \\
+			\hline
 		\end{tabular}
+	\end{center}
 \end{frame}
@@
 \begin{frame}{Testing Mechanisms}
-	\textbf{Stress testing}: \texttt{slaughter cho2} and friends
+	\textbf{Stress testing}: \texttt{largetest}, \texttt{mandelbrot} and friends
 	\begin{itemize}
@@
 	\begin{itemize}
-		\item Threads becoming runnable (\texttt{fork()}, \texttt{cond\_signal()}, etc.)
+		\item Threads becoming runnable (\texttt{thr\_create()}, \texttt{cond\_signal()}, etc.)
 			\begin{itemize}
@@
 			% FIXME for next year -- how much slowdown is there? 20x?
-		\item Originally supported only P3s; can now test P2s as well
+		\item Originally supported only P3s; can now test P2s fully-automated
 		\item {\em Landslide} shows that your {\em Pebbles} are less stable than you thought.
@@
 
-\begin{frame}{Anatomy}
-	\begin{center}
-	\includegraphics[width=0.6\textwidth]{landslide.png}
-	\end{center}
-\end{frame}
+%\begin{frame}{Anatomy}
+%	\begin{center}
+%	\includegraphics[width=0.6\textwidth]{landslide.png}
+%	\end{center}
+%\end{frame}
 
@@
 	\begin{itemize}
-		\item Kernel panics % Say: note: the better your asserts are..!
+		\item Assertion failures % Say: note: the better your asserts are..!
+		\item Segfaults
 		\item Deadlock
@@
 		\item Infinite loop (halting problem)
-		\item Data races
+		\item Data race bugs
 	\end{itemize}
@@
 
-\begin{frame}{Brief Aside: Data Races}
+\begin{frame}{What is a Data Race?}
 	A {\bf data race} is a pair of memory accesses between two threads, where:
@@
 	\begin{itemize}
-		\item ``If threads interleaved the wrong way here, it might crash later.''
+		\item Is unprotected {\tt foobar\_statistics\_counter++} bad?
+		\item What about unprotected {\tt next\_thread\_id++}?
+		\item ``If threads interleaved the wrong way here, it {\em might} crash later.''
 		\begin{itemize}
@@
 	\end{itemize}
-\end{frame}
-
-\breakslide{
-	\begin{center}
-		\includegraphics[width=0.85\textwidth]{trace1.png}
-	\end{center}
-}
-\breakslide{
-	\begin{center}
-		\includegraphics[width=\textwidth]{trace2.png}
-	\end{center}
-}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\subsection{State Space Reduction}
-
-\begin{frame}{State Space Explosion}
-	Serious problem: State spaces grow exponentially
-	\begin{itemize}
-		\item With $p$ preemption points and $k$ runnable threads, size $p^k$.
-		\item Threatens our ability to explore everything.
-		\item Fortunately, some sequences result in identical states.
-	\end{itemize}
-	\linegap
-
-	{\bf Partial Order Reduction} identifies and skips ``equivalent'' interleavings.
-	\begin{itemize}
-		\item After each execution, compare memory reads/writes of each thread.
-		\item Find when reordering threads couldn't possibly change behaviour.
-		\item Example follows\dots
-	\end{itemize}
-\end{frame}
-
-\begin{frame}{State Space Reduction}
-	\begin{center}
-	\includegraphics[width=\textwidth]{undiamond0.png}
-	\end{center}
-\end{frame}
-\begin{frame}{State Space Reduction}
-	\begin{center}
-	\includegraphics[width=\textwidth]{undiamond1.png}
-	\end{center}
-\end{frame}
-\begin{frame}{State Space Reduction}
-	\begin{center}
-	\includegraphics[width=\textwidth]{diamond1.png}
-	\end{center}
 \end{frame}
@@
 \begin{frame}{Choosing the Right Preemption Points}
-	So far we only {\em mitigated}, not {\em solved}, exponential explosion.
+	How can we address exponential state space explosion?
 	\pause
@@
 
-\begin{frame}{``Iterative Deepening'' (my current project)}
+\begin{frame}{``Iterative Deepening'' of Preemption Points}
 	Goal: Run the best tests for a given CPU budget.
@@
 		\item ``Start small, then add more preemption points as time allows''
-		\item Can we automate this process?
+		\item Landslide now automates this process
 	\end{itemize}
@@
 
+\breakslide{\Huge Demo}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-\section{Evaluation}
-\breakslide{\Large Evaluation}
+\section{Landslide for the People}
+
+\begin{frame}{User Study}
+	\textbf{Try Landslide on your P2!}
+	\begin{itemize}
+		\item Bare minimum effort: No more than 1 hour
+		\begin{itemize}
+			\item Clone a github URL, run setup script, run, answer survey questions
+			\item Landslide will automatically report test results
+		\end{itemize}
+		\item Full study plan: 4-8 hours of active attention
+		\begin{itemize}
+			\item (Estimated, including time to diagnose and fix bugs)
+			\item However, many tests should run passively overnight -- start soon!
+		\end{itemize}
+	\end{itemize}
+	\pause
+	\linegap
+
+	Prerequisites
+	\begin{itemize}
+		\item You {\em must} pass the P2 hurdle before using Landslide.
+		\begin{itemize}
+			\item {\tt startle}, {\tt agility\_drill}, {\tt cyclone}, {\tt join\_specific\_test}, {\tt thr\_exit\_join}
+		\end{itemize}
+		\item Must have {\em attempted} several stress tests
+		\begin{itemize}
+			\item {\tt juggle}, {\tt multitest}, {\tt racer} (for 1 hour), {\tt paraguay}
+		\end{itemize}
+
+	\end{itemize}
+\end{frame}
+
+\begin{frame}{User Study - Additional Information}
+	Human Subjects Research
+	\begin{itemize}
+		\item CMU IRB has approved this study
+		\item Landslide will collect results while you use it
+		\begin{itemize}
+			\item Record commands issued, take snapshots of your P2 code
+			\item All data will be anonymized before publication
+		\end{itemize}
+		\item No coercion: {\bf \em There is no penalty for not participating.}
+		\begin{itemize}
+			\item I am not on course staff, cannot influence your grade
+			\item Course staff will not have access to study data during semester
+		\end{itemize}
+	\end{itemize}
+	\pause
+	\linegap
+
+	Risks \& Benefits
+	\begin{itemize}
+		\item Benefit: Landslide may help you find/fix bugs, improving your grade!
+		\item Risk: Landslide may find no bugs and be a waste of your time.
+	\end{itemize}
+	% Say: Accordingly, although I hope you will use landslide enough to get good results for me, you should also responsibly allocate time between testing with landslide versus actually implementing your P2.
+\end{frame}
+
+% IRB approved, plan on publishing collected data, anonymized
+% Expect to spend X time
+% Participation has no impact on grade but you will find bugs
+
+\begin{frame}{User Study - How to Participate}
+	To participate\dots
+	\begin{itemize}
+		\item Review this lecture and study information sheet
+		\item Meet prerequisites of passing P2 tests
+		\item Email me ({\tt bblum@cs.cmu.edu}) for instructions, survey
+		\item Use Landslide in addition to stress tests until P2 is due!
+	\end{itemize}
+	\linegap
+
+	Watch your email for more details!
+\end{frame}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{End}
+\breakslide{\Large Questions?
+\linegap
+
+\begin{center}
+	\includegraphics[width=0.65\textwidth]{3word-questions.png}
+\end{center}}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{Bonus Slides}
+
+\subsection{State Space Reduction}
+
+\begin{frame}{Coping with State Space Explosion}
+	Serious problem: State spaces grow exponentially
+	\begin{itemize}
+		\item With $p$ preemption points and $k$ runnable threads, size $p^k$.
+		\item Threatens our ability to explore everything.
+		\item Fortunately, some sequences result in identical states.
+	\end{itemize}
+	\linegap
+
+	{\bf Partial Order Reduction} identifies and skips ``equivalent'' interleavings.
+	\begin{itemize}
+		\item After each execution, compare memory reads/writes of each thread.
+		\item Find when reordering threads couldn't possibly change behaviour.
+		\item Example follows\dots
+	\end{itemize}
+\end{frame}
+
+\begin{frame}{State Space Reduction}
+	\begin{center}
+	\includegraphics[width=\textwidth]{undiamond0.png}
+	\end{center}
+\end{frame}
+\begin{frame}{State Space Reduction}
+	\begin{center}
+	\includegraphics[width=\textwidth]{undiamond1.png}
+	\end{center}
+\end{frame}
+\begin{frame}{State Space Reduction}
+	\begin{center}
+	\includegraphics[width=\textwidth]{diamond1.png}
+	\end{center}
+\end{frame}
+
+\subsection{Experimental Results}
 
@@
 
-\begin{frame}{Recruitment}
-	\textbf{User study}: Come try Landslide on your P2!
-	\begin{itemize}
-		\item Bare minimum effort: No more than 30 minutes
-		\begin{itemize}
-			\item Clone a github URL, run setup script, run, send results
-		\end{itemize}
-		\item Testing P2s is fully automated, but can also test P3s
-		\begin{itemize}
-			\item Landslide requires manual kernel annotations
-			\item Comes with pre-annotated reference kernel for P2
-			\item Past students take approx. 1 hour to annotate own P3s.
-		\end{itemize}
-	\end{itemize}
-	\pause
-	\linegap
-
-	{\bf Ultimate goal}: Provide Landslide to every 410 student during P2.
-	\begin{itemize}
-		\item This semester: Prep \& testing
-			\begin{itemize}
-				\item Improving user interface
-				\item Making sure Landslide itself is stable
-			\end{itemize}
-		%\item Please help us!
-		\item {\bf \large Just for fun -- this will NOT impact your grade.}
-	\end{itemize}
-\end{frame}
-
-\begin{frame}{Recruitment}
-	%User study this week, by appointment (\texttt{bblum@cs.cmu.edu})
-	User study during exam week, by appointment (\texttt{bblum@cs.cmu.edu})
-	\linegap
-
-%	Recommended to use fully-automated P2 test suite, but you can try testing your P3 with approx. 1 hour of manual annotation.
-%	\linegap
-%
-	%Expect to spend:
-	%\begin{itemize}
-	%	\item Up to 3 hours, just to try it out.
-	%	\item 4-6 hours, if you find a bug and track it down.
-	%	\item More, for multiple bugs or the truly dedicated\ldots
-	%\end{itemize}
-	%\linegap
-
-	Give feedback (intuitive? frustrating? found bugs?)
-
-	\linegap
-	Watch your email for details!
-\end{frame}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{End}
-\breakslide{\Large Questions?
-\linegap
-
-\begin{center}
-	\includegraphics[width=0.65\textwidth]{3word-questions.png}
-\end{center}}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{Bonus Slide(s)}
-
-\begin{frame}{More Results}
+\begin{frame}{More Preliminary Results}
 	\textbf{Is dynamically adding ``data race'' PPs effective?}

### 2015-02-07 landslide-lecture/2015/lecture.tex
Finish rought draft of landslide lecture for s15

@@
 \begin{document}
-\renewcommand{\inserttotalframenumber}{41}
+\renewcommand{\inserttotalframenumber}{34}
 \normalem
@@
 \definecolor{thread2}{RGB}{255,201,102}
-\definecolor{thread3}{RGB}{255,87,130}
+\definecolor{thread3}{RGB}{255,128,160}
 
@@
 				\\
-				\texttt{LOCK(mx);} \\
+				\texttt{mutex\_lock(mx);} \\
 				\\
 				\texttt{if~(!work\_exists())} \\
-				\texttt{~~~~WAIT(cvar,~mx);} \\
-				\texttt{work~=~claim\_work();} \\
+				\texttt{~~~~cond\_wait(cvar,~mx);} \\
+				\texttt{work~=~dequeue();} \\
 				\\
-				\texttt{UNLOCK(mx);} \\
-				\texttt{assert(work~!=~NULL);} \\
+				\texttt{mutex\_unlock(mx);} \\
+				\texttt{access(work->data);} \\
 			\end{tabular}
@@
 				\\
-				\texttt{LOCK(mx);} \\
+				\texttt{mutex\_lock(mx);} \\
 				\\
-				\texttt{give\_work(work);} \\
-				\texttt{SIGNAL(cvar);} \\
+				\texttt{enqueue(work);} \\
+				\texttt{signal(cvar);} \\
 				\\
-				\texttt{UNLOCK(mx);} \\
+				\texttt{mutex\_unlock(mx);} \\
 				\\
@@
 			\hline
-			\small \texttt{LOCK(mx);} & & \\
+			\small \texttt{lock(mx);} & & \\
 			\small \texttt{if~(!work\_exists())} & & \\
-			\small \texttt{~~~~WAIT(cvar,~mx);} & & \\
+			\small \texttt{~~~~wait(cvar,~mx);} & & \\
 
-			& \small \texttt{LOCK(mx);} & \\
-			& \small \texttt{give\_work(work);} & \\
-			& \small \texttt{SIGNAL(cvar);} & \\
-			& \small \texttt{UNLOCK(mx);} & \\
+			& & \small \texttt{lock(mx);} \\
+			& & \small \texttt{if~(!work\_exists())} \\
+			& & \small \texttt{~~~~wait(cvar,~mx);} \\
 
-			\small \texttt{work~=~claim\_work();~} & & \\
-			\small \texttt{UNLOCK(mx);} & & \\
-			\small \texttt{//~got~the~work} {\large \smiley}& & \\
+			& \small \texttt{lock(mx);} & \\
+			& \small \texttt{enqueue(work);} & \\
+			& \small \texttt{signal(cvar);} & \\
+			& \small \texttt{unlock(mx);} & \\
 
-			& & \small \texttt{LOCK(mx);} \\
-			& & \small \texttt{if~(!work\_exists())~} \\
-			& & \small \texttt{~~~~WAIT(cvar,~mx);} \\
+			\small \texttt{work~=~dequeue();~} & & \\
+			\small \texttt{unlock(mx);} & & \\
+			\small \texttt{access(work->data);}& & \\
 			\hline
@@
 			\hline
-			\small \texttt{LOCK(mx);} & & \\
+			\small \texttt{lock(mx);} & & \\
 			\small \texttt{if~(!work\_exists())} & & \\
-			\small \texttt{~~~~WAIT(cvar,~mx);} & & \\
+			\small \texttt{~~~~wait(cvar,~mx);} & & \\
 
-			& & \small \texttt{LOCK(mx);} \\
-			& & \small \texttt{if~(!work\_exists())~} \\
-			& & \small \texttt{~~~~WAIT(cvar,~mx);} \\
+			& \small \texttt{lock(mx);} & \\
+			& \small \texttt{enqueue(work);} & \\
+			& \small \texttt{signal(cvar);} & \\
+			& \small \texttt{unlock(mx);} & \\
 
-			& \small \texttt{LOCK(mx);} & \\
-			& \small \texttt{give\_work(work);} & \\
-			& \small \texttt{SIGNAL(cvar);} & \\
-			& \small \texttt{UNLOCK(mx);} & \\
+			\small \texttt{work~=~dequeue();~} & & \\
+			\small \texttt{unlock(mx);} & & \\
+			\small \texttt{access(work->data);}& & \\
 
-			\small \texttt{work~=~claim\_work();~} & & \\
-			\small \texttt{UNLOCK(mx);} & & \\
-			\small \texttt{//~got~the~work} {\large \smiley}& & \\
+			& & \small \texttt{lock(mx);} \\
+			& & \small \texttt{if~(!work\_exists())} \\
+			& & \small \texttt{~~~~wait(cvar,~mx);} \\
 			\hline
@@
 			\hline
-			\small \texttt{LOCK(mx);} & & \\
+			\small \texttt{lock(mx);} & & \\
 			\small \texttt{if~(!work\_exists())} & & \\
-			\small \texttt{~~~~WAIT(cvar,~mx);} & & \\
+			\small \texttt{~~~~wait(cvar,~mx);} & & \\
 
-			& \small \texttt{LOCK(mx);} & \\
-			& \small \texttt{give\_work(work);} & \\
-			& \small \texttt{SIGNAL(cvar);} & \\
-			& \small \texttt{UNLOCK(mx);} & \\
+			& \small \texttt{lock(mx);} & \\
+			& \small \texttt{enqueue(work);} & \\
+			& \small \texttt{signal(cvar);} & \\
+			& \small \texttt{unlock(mx);} & \\
 
-			& & \small \texttt{LOCK(mx);} \\
-			& & \small \texttt{work~=~claim\_work();} \\
-			& & \small \texttt{UNLOCK(mx);} \\
+			& & \small \texttt{lock(mx);} \\
+			& & \small \texttt{work~=~dequeue();~~} \\
+			& & \small \texttt{unlock(mx);} \\
 
-			\small \texttt{work~=~claim\_work();~} & & \\
-			\small \texttt{UNLOCK(mx);} & & \\
-			\small \texttt{//~work~was~stolen} {\large \frownie}& & \\
+			\small \texttt{work~=~dequeue();} & & \\
+			\small \texttt{unlock(mx);} & & \\
+			\small \texttt{//~SIGSEGV} {\large \frownie}& & \\
 			\hline
@@
 \begin{frame}{Execution Tree}
-	\begin{center}
-		\includegraphics[width=0.9\textwidth]{threadfork0.png}
-	\end{center}
+		\includegraphics[width=0.78\textwidth]{execution-tree-0.pdf} \\
+		\vspace{-3in}
+		\hspace{3in}
+		\includegraphics[width=0.34\textwidth]{table-0.png}
+		\\
+		\vspace{1.2in}
+		\hspace{3.7in}
+		\includegraphics[width=0.15\textwidth]{frownie-blank.png}
 \end{frame}
 \begin{frame}{Execution Tree}
-	\begin{center}
-		\includegraphics[width=0.9\textwidth]{threadfork05.png}
-	\end{center}
+		\includegraphics[width=0.78\textwidth]{execution-tree-1.pdf} \\
+		\vspace{-3in}
+		\hspace{3in}
+		\includegraphics[width=0.34\textwidth]{table-0.png}
+		\\
+		\vspace{1.2in}
+		\hspace{3.7in}
+		\includegraphics[width=0.15\textwidth]{frownie-blank.png}
 \end{frame}
 \begin{frame}{Execution Tree}
-	\begin{center}
-		\includegraphics[width=0.9\textwidth]{threadfork1.png}
-	\end{center}
+		\includegraphics[width=0.78\textwidth]{execution-tree-2.pdf} \\
+		\vspace{-3in}
+		\hspace{3in}
+		\includegraphics[width=0.34\textwidth]{table-1.png}
+		\\
+		\vspace{1.2in}
+		\hspace{3.7in}
+		\includegraphics[width=0.15\textwidth]{frownie-blank.png}
 \end{frame}
 \begin{frame}{Execution Tree}
-	\begin{center}
-		\includegraphics[width=0.9\textwidth]{threadfork2.png}
-	\end{center}
+		\includegraphics[width=0.78\textwidth]{execution-tree-full.pdf} \\
+		\vspace{-3in}
+		\hspace{3in}
+		\includegraphics[width=0.34\textwidth]{table-2.png}
+		\\
+		\vspace{1.2in}
+		\hspace{3.7in}
+		\includegraphics[width=0.15\textwidth]{frownie.png}
 \end{frame}
@@
 \begin{frame}{User Study - How to Participate}
+	Interested?
+	\linegap
+
 	To participate\dots

### 2015-02-10 irb/hurdle.tex
Add 'hurdle'-like sign-up sheet for students to fill out.

--- 2015-01-16 irb/survey.tex
@@
 
-\title{\bf Landslide: Systematic Testing in 15-410 \\ User Guide and Survey}
+\newcommand\checkbox[1]{\huge \Square & #1 \\}
+
+\title{\bf Landslide: Systematic Testing in 15-410 \\ Sign-Up Sheet}
 \author{Ben Blum}
@@
 
-\section{Introduction}
-
-%Normally here I would write something researchy about systematic testing and preemption points, but if you're here it probably means you know what this is about.
-If you don't, please refer to the lecture material.
-%\noindent
-Thanks for volunteering to try out Landslide.
-We have worked hard to make it nice enough to give to you, and we hope you're just as excited as we are to try out some systematic testing techniques the world has never seen before.
+Thanks for volunteering to try out Landslide!
+To participate in the study, you and your partner must meet all the prerequisites below.
+Submit this form at my office (GHC 9009), any weekday between 13:00 and 17:00, or if you can't make it during those times, scan it after signing and email to {\tt bblum@cs.cmu.edu}.
 \\
 
-\noindent You should have received a document titled {\em User Study Information Sheet} that describes your rights as a study participant and our confidentiality procedures.
-If you are missing this, please contact Ben immediately at {\tt bblum@cs.cmu.edu}.
-If you have not yet reviewed the lecture slides (found on the course website), please do so before proceeding.
-% TODO: Host them on maxi or contrib and link to them.
+\begin{tabular}{lp{5.5in}}
+	\checkbox{We have read the lecture slides, available on the 410 course website (or were present and paid attention in class on Wednesday).}
 
-\section{Instructions}
+	\\
+	\checkbox{We have read the study information sheet: \url{http://www.contrib.andrew.cmu.edu/~bblum/landslide-info-sheet.pdf}.}
 
-The very first thing you should do is {\bf READ THIS DOCUMENT ENTIRELY FIRST!} We need you to print out several survey sheets before you dive in.
 	\\
+	\checkbox{We agree that, upon receiving the instructions sheet, we will read it in its entirety before beginning any tests.}
 
-\noindent We encourage you to use Landslide on the GHC cluster or UNIX pool machines ({\tt ghcXX.ghc.andrew.cmu.edu} or {\tt unix.andrew.cmu.edu}).
-Landslide has not been tested with other configurations and will probably fail mysteriously on your personal laptop, and we cannot promise to be of any help.
+	\\
+	\checkbox{We pass the ``hurdle'' test suite listed in the P2 handout (on at least a vast majority of runs).
+This includes {\tt startle}, {\tt agility\_drill}, {\tt cyclone}, {\tt join\_specific\_test}, {\tt thr\_exit\_join}.}
 
-\begin{enumerate}
-	\item Clone the repository -- {\tt git clone https://github.com/bblum/landslide.git} -- this will be your workspace.
-	\item Run {\tt ./p2-setup.sh PATH}, where {\tt PATH} is the absolute path to your p2 implementation.
-		\begin{itemize}
-			\item This command will import your p2 into {\tt ./pebsim/p2-basecode/} and then attempt to build it.
-			\item {\tt PATH} should be the path with {\tt 410kern}, {\tt 410user}, {\tt user/libthread}, etc as subdirectories, not the {\tt user/libthread} directory itself.
-		\end{itemize}
-	\item Next, make sure you have read this document entirely.
-At this point you should have several survey sheets printed out, ready to write on -- see the Survey section below.
-You should also have some means of timing yourself.
-	\item Run {\tt ./landslide OPTIONS} to run tests with Landslide. {\bf Fill out a fresh survey sheet each time you do this.} Repeat until satisfied.
-		\begin{itemize}
-			\item Use {\tt ./landslide -h} to view command line options. The most important ones are {\tt -t}, max testing time, and {\tt -p}, which test program to run.
-				% TODO: Remove this TODO when you have implemented glue to move the html files to the toplevel directory.
-			\item Landslide's bug reports will show up as HTML files. If you are using a cluster machine in person, you can view them by pointing a browser to {\tt file:///PATH/TO/landslide/FILENAME.html}.
-				\footnote{Another option: move it to {\tt \textasciitilde/www} in your AFS, and view it at {\tt http://www.contrib.andrew.cmu.edu/\textasciitilde{}YOUR\_ANDREWID/FILENAME.html}.
-Note the {\tt www} in the URL is mandatory; note also that this will allow the world to see it, so be careful!}
-			\item The available test programs are {\tt thr\_exit\_join} (the default), {\tt broadcast\_test}, {\tt paraguay}, and {\tt rwlock\_downgrade\_read\_test}.
-				\footnote{If you have your own custom test case, you can put it in {\tt pebsim/p2-basecode/user/progs} and edit \\ {\tt pebsim/p2-basecode/config.mk} to build it, then run {\tt ./p2-setup.sh} again.}
-		\end{itemize}
-\end{enumerate}
+	\\
+	\checkbox{We have {\em attempted}, but not necessarily pass, the following stress tests; and {\em attempted} to debug and fix any observed bugs: {\tt juggle 4 3 2 0}, {\tt multitest}, {\tt racer} (for 15 minutes), {\tt paraguay}.}
 
-\noindent {\bf What to do when something goes wrong:}
-If you find a Landslide bug, such as an assertion/crash, or such as a bug report that you think is actually a bug in Landslide rather than a bug in your P2, or such as it getting stuck and not exiting in the given time limit, please create a tarball of your workspace, just as you would to report a reference kernel bug, but send it to Ben instead of 410 staff.
-For any other technical issues getting it to work, or if you have difficulty understanding a bug report, send email.
+\end{tabular}
 \\
-
-\noindent
-However, we {\em cannot} provide help on how to understand {\em why} your bug happened or how to fix it.
-The point of the study is that Landslide's automation plus your own brain should be enough!
-On the other hand, if you have a design question while deciding between potential ways to fix a bug, send it to 410 course staff as usual.
-
-\newpage
-
-\section{Survey}
-\subsection{Overall Survey Questions}
-
-\begin{center}
-{\bf Print this page out once.}
-\end{center}
-
-\begin{enumerate}
-	\item How much time did you spend directly using Landslide?
-This includes configuring test parameters and interpreting debug output.
-It does not include time spent running Landslide in the background, e.g.
-overnight.
 \vspace{0.5in}
-
-	\item What test configurations did you use that DIDN'T find bugs? For example, {\tt ./landslide -t 15m -p thr\_exit\_join}
-\end{enumerate}
-
-\newpage
-\subsection{Per-Bug Survey Questions}
-\begin{center}
-{\bf \large PRINT OUT THIS PAGE MULTIPLE TIMES.
 \\
-
-Each time you find a new bug with Landslide, fill this out again.}
-\end{center}
-
-\begin{enumerate}
-		% TODO: find a way to allow ze studence to condense non-bug-finding runs onto 1 page.
-	\item What test configuration did you use? For example, {\tt ./landslide -t 15m -p thr\_exit\_join}
-		\vspace{0.5in}
-	\item Summarize the bug.
-		\vspace{0.5in}
-	\item How IMPORTANT was this bug? Rate on a scale from 1 to 10, where:
-		\begin{itemize}
-			\item 1 = Trivial; an obvious ``oops'' that does not reflect any key lessons learned during the project. I would expect to not lose any points for this bug.
-			\item 4 = Minor; a logic problem not related to threads/synchronization. I would expect to lose 1 or 2 points at most for this bug.
-			\item 7 = Major; a concurrency-related bug related to subject matter discussed in lecture.
-			\item 10 = Severe; this bug reflects a fundamental lesson about concurrency that we learned during the project.
-I would expect to receive ``see course staff'' in my ink if I turned in code with this bug!
-		\end{itemize}
-		\vspace{0.5in}
-	\item How much EFFORT/TIME did this bug require to diagnose and fix? Rate on a scale from 1 to 10, where:
-		\begin{itemize}
-			\item 1 = Very little; we knew immediately what was wrong and needed to change only 1 or 2 lines of code.
-			\item 5 = Some; the bug took considerable thought to fix, and required nontrivial code changes.
-			\item 10 = Lots; we worked for several hours to fix the bug and/or required a major refactoring of our code.
-		\end{itemize}
-		\vspace{0.5in}
-	\item How likely would a CONVENTIONAL STRESS TEST be able to find the bug (i.e., running the program repeatedly, not using Landslide)? Rate on a scale from 1 to 10, where:
-		\begin{itemize}
-			\item 1 = No hope. The window for a random timer interrupt to expose the bug is so small that a conventional stress test could easily run continuously overnight without encountering it.
-			\item 5 = Moderate chance. Most random timer interrupts would not expose the bug, but the window is more than just a few instructions.
-			\item 10 = Very likely. A random timer interrupt during the test would expose the bug more often than not.
-		\end{itemize}
-\end{enumerate}
+Group number: \underline{\hspace{1.5in}}\\
+\vspace{0.5in}\\
+Partner 1 andrewid: \underline{\hspace{1.5in}} Signature: \underline{\hspace{1.5in}} Date: \underline{\hspace{1in}}\\
+\vspace{0.5in}\\
+Partner 2 andrewid: \underline{\hspace{1.5in}} Signature: \underline{\hspace{1.5in}} Date: \underline{\hspace{1in}}
 

### 2015-02-10 irb/hurdle.tex
Update sign-up sheet; write solicitation email; update lecture

@@
 	\checkbox{We pass the ``hurdle'' test suite listed in the P2 handout (on at least a vast majority of runs).
-This includes {\tt startle}, {\tt agility\_drill}, {\tt cyclone}, {\tt join\_specific\_test}, {\tt thr\_exit\_join}.}
+This includes {\tt startle}, {\tt agility\_drill}, {\tt cyclone}, {\tt join\_specific\_test}, {\tt thr\_exit\_join}.
+We acknowledge that {\bf a P2 that cannot pass these tests is incomplete, and Landslide will not work on it at all.}}
 

### 2015-02-10 landslide-lecture/2015/lecture.tex
update landslide lecture
Update sign-up sheet; write solicitation email; update lecture

@@
 \institute[CMU 15-410]{Carnegie Mellon University - 15-410}
-\date[]{2014, December 3}
+\date[]{2015, February 11}
 
@@
 \begin{document}
-\renewcommand{\inserttotalframenumber}{34}
+\renewcommand{\inserttotalframenumber}{36}
 \normalem
@@
 	\end{itemize}
-	{\bf Technique: Systematic testing}
+	{\bf Research Technique: ``Systematic testing''}
 	% Say: A way of enumerating all possible interleavings of a concurrent test case.
@@
 		\begin{tabular}{llll}
+			\fbox{
 			\begin{tabular}{l}
@@
 			\end{tabular}
+			}
 			& & &
+			\fbox{
 			\begin{tabular}{l}
@@
 			\end{tabular}
+			}
 		\end{tabular}
@@
 \end{frame}
-\begin{frame}{Thread Interleavings (``good'' case, again)}
+\begin{frame}{Thread Interleavings (different ``good'' case)}
 	\begin{center}
@@
 \begin{frame}{Testing}
-	How can programmers be confident in their code's correctness?
+	How can programmers be confident in the correctness of their code?
 	\begin{itemize}
@@
 		\item On 1st execution, schedule threads arbitrarily until program ends.
+			\begin{itemize}
+				\item This represents one branch of the tree.
+			\end{itemize}
 		\item At end of each branch, rewind system and restart test.
-		\item Artificially add preemptions to test different thread interleavings.
+		\item Artificially add preemptions to produce different thread interleavings.
 		\item Intuitively: Generate many ``tabular execution traces''.
@@
 		\item Need to choose the set of PPs very carefully for test to be effective.
+		\begin{itemize}
+			\item ``Effective'' = both comprehensive and feasible.
+		\end{itemize}
 	\end{itemize}
@@
 	\linegap
-	{\em Stress testing offers this -- can we offer it too?}
+	{\em Stress testing allows user to choose total run time -- can we offer this too?}
 \end{frame}
 
-\begin{frame}{``Iterative Deepening'' of Preemption Points}
+\begin{frame}{Iterative Deepening of Preemption Points}
 	Goal: Run the best tests for a given CPU budget.
+	\begin{itemize}
+		\item Technique: ``Iterative Deepening''
+	\end{itemize}
 	\linegap
@@
 
-\breakslide{\Huge Demo}
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-\section{Landslide for the People}
+\section{Evaluation}
+
+\breakslide{\Huge Demo}
+
+\begin{frame}{Preliminary Evaluation}
+	We tested 22 random submitted P2s from Spring 2014.
+	\begin{itemize}
+		\item Test cases:
+		\begin{itemize}
+			\item {\tt thr\_exit\_join}
+			\item {\tt paraguay}
+			\item {\tt rwlock\_downgrade\_read\_test}
+			\item {\tt broadcast\_test} (new)
+		\end{itemize}
+		\item Each test given CPU budget of 11 cores $\times$ 10 minutes.
+	\end{itemize}
+\end{frame}
+
+\begin{frame}{Preliminary Results}
+	88 total tests were run (9,680 total CPU-minutes)
+	\linegap
+
+	Found 7 {\bf deterministic} initialization bugs (e.g. use-after-free)
+	\linegap
+
+	56 tests ran to completion without finding bugs:
+	\begin{itemize}
+		\item 47 exhausted the CPU budget.
+		\item 9 completed all state spaces before time ran out.
+	\end{itemize}
+	\linegap
+
+	Between 20 and 33 {\bf non-deterministic} bugs were found:
+	\begin{itemize}
+		\item 20 deadlock races among 12 P2s
+		\item 10 use-after-free races among 6 P2s
+		\item 3 misc (assert fail, segfault) among 2 P2s
+	\end{itemize}
+\end{frame}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\subsection{Landslide for the People}
 
@@
 		\item Risk: Landslide may find no bugs and be a waste of your time.
+		\item Benefit: You might learn something...
 	\end{itemize}
@@
 \end{frame}
-
-% IRB approved, plan on publishing collected data, anonymized
-% Expect to spend X time
-% Participation has no impact on grade but you will find bugs
 
@@
 \section{Bonus Slides}
+
+\subsection{Experimental Results}
+
+\begin{frame}{More Preliminary Results}
+	\textbf{Is dynamically adding ``data race'' PPs effective?}
+	\begin{itemize}
+		\item 3 among reported bugs (9-15\%) required data races to expose.
+	\end{itemize}
+	\linegap
+
+	{\bf Is ``iterative deepening'' better than state-of-the-art?}
+	\begin{itemize}
+		\item Control experiment: Just 1 state space, same CPU time
+		\begin{itemize}
+			\item PPs used: \texttt{mutex\_lock()}, \texttt{mutex\_unlock()}
+			\item 110 minutes on 1 CPU
+		\end{itemize}
+		\item Of 33 total bug reports, control failed to find 10 (30\%).
+		\begin{itemize}
+			\item 3 required data race PPs to expose
+			\item 1 ran out of time
+			\item 6 obscured by different bug in same state space
+		\end{itemize}
+	\end{itemize}
+\end{frame}
 
@@
 
-\subsection{Experimental Results}
-
-\begin{frame}{Preliminary Evaluation}
-	We tested 22 random submitted P2s from Spring 2014.
-	\begin{itemize}
-		\item Test cases:
-		\begin{itemize}
-			\item {\tt thr\_exit\_join}
-			\item {\tt paraguay}
-			\item {\tt rwlock\_downgrade\_read\_test}
-			\item {\tt broadcast\_test} (new)
-		\end{itemize}
-		\item Each test given CPU budget of 11 cores $\times$ 10 minutes.
-	\end{itemize}
-\end{frame}
-
-\begin{frame}{Preliminary Results}
-	88 total tests were run (9,680 total CPU-minutes)
-	\linegap
-
-	Found 7 {\bf deterministic} initialization bugs (e.g. use-after-free)
-	\linegap
-
-	56 tests ran to completion without finding bugs:
-	\begin{itemize}
-		\item 47 exhausted the CPU budget.
-		\item 9 completed all state spaces before time ran out.
-	\end{itemize}
-	\linegap
-
-	Between 20 and 33 {\bf non-deterministic} bugs were found:
-	\begin{itemize}
-		\item 20 deadlock races among 12 P2s
-		\item 10 use-after-free races among 6 P2s
-		\item 3 misc (assert fail, segfault) among 2 P2s
-	\end{itemize}
-\end{frame}
-
-\begin{frame}{More Preliminary Results}
-	\textbf{Is dynamically adding ``data race'' PPs effective?}
-	\begin{itemize}
-		\item 3 among reported bugs (9-15\%) required data races to expose.
-	\end{itemize}
-	\linegap
-
-	{\bf Is ``iterative deepening'' better than state-of-the-art?}
-	\begin{itemize}
-		\item Control experiment: Just 1 state space, same CPU time
-		\begin{itemize}
-			\item PPs used: \texttt{mutex\_lock()}, \texttt{mutex\_unlock()}
-			\item 110 minutes on 1 CPU
-		\end{itemize}
-		\item Of 33 total bug reports, control failed to find 10 (30\%).
-		\begin{itemize}
-			\item 3 required data race PPs to expose
-			\item 1 ran out of time
-			\item 6 obscured by different bug in same state space
-		\end{itemize}
-	\end{itemize}
-\end{frame}
-
 \end{document}

### 2015-02-10 irb/hurdle.tex
update sign up sheet

--- 2015-02-10 irb/hurdle.tex
@@
 	\checkbox{We pass the ``hurdle'' test suite listed in the P2 handout (on at least a vast majority of runs).
-This includes {\tt startle}, {\tt agility\_drill}, {\tt cyclone}, {\tt join\_specific\_test}, {\tt thr\_exit\_join}.}
+This includes {\tt startle}; {\tt agility\_drill}; {\tt cyclone}; {\tt join\_specific\_test}; {\tt thr\_exit\_join}.
+We acknowledge that {\bf a P2 that cannot pass these tests is incomplete, and Landslide would not work on it at all.}}
 
 	\\
-	\checkbox{We have {\em attempted}, but not necessarily pass, the following stress tests; and {\em attempted} to debug and fix any observed bugs: {\tt juggle 4 3 2 0}, {\tt multitest}, {\tt racer} (for 15 minutes), {\tt paraguay}.}
+	\checkbox{We have {\em attempted}, but do not necessarily pass, the following stress tests; and {\em attempted} to debug and fix any observed bugs: {\tt juggle 4 3 2 0}; {\tt multitest}; {\tt racer} (for 15 minutes); {\tt paraguay}.}
 

### 2015-02-10 irb/hurdle.tex
update hurdle sign up

@@
 	\\
-	\checkbox{We have {\em attempted}, but do not necessarily pass, the following stress tests; and {\em attempted} to debug and fix any observed bugs: {\tt juggle 4 3 2 0}; {\tt multitest}; {\tt racer} (for 15 minutes); {\tt paraguay}.}
+	\checkbox{We have {\em attempted}, but do not necessarily pass, the following stress tests: {\tt juggle 4 3 2 0}; {\tt multitest}; {\tt racer} (for 15 minutes); {\tt paraguay}.}
 

### 2015-02-10 irb/hurdle.tex
More changes to signup, solicitation, lecture

@@
 To participate in the study, you and your partner must meet all the prerequisites below.
-Submit this form at my office (GHC 9009), any weekday between 13:00 and 17:00, or if you can't make it during those times, scan it after signing and email to {\tt bblum@cs.cmu.edu}.
+Submit this form at my office (GHC 9009), any weekday between 13:00 and 17:00, or if you can't make it during those times, scan it or take a picture after signing and email to {\tt bblum@cs.cmu.edu}.
 \\
@@
 \vspace{0.5in}\\
-Partner 1 andrewid: \underline{\hspace{1.5in}} Signature: \underline{\hspace{1.5in}} Date: \underline{\hspace{1in}}\\
+Partner 1 Andrew username: \underline{\hspace{2.5in}} Date: \underline{\hspace{1.5in}}
 \vspace{0.5in}\\
-Partner 2 andrewid: \underline{\hspace{1.5in}} Signature: \underline{\hspace{1.5in}} Date: \underline{\hspace{1in}}
+Partner 2 Andrew username: \underline{\hspace{2.5in}} Date: \underline{\hspace{1.5in}}
 

### 2015-02-10 landslide-lecture/2015/lecture.tex
fix a bug in lecture
More changes to signup, solicitation, lecture
Fix iterative-deepening slide sequence to be less researcher-speak-y.

@@
 		\item Originally supported only P3s; can now test P2s fully-automated
-		\item {\em Landslide} shows that your {\em Pebbles} are less stable than you thought.
+		\item {\em Landslide} shows how your {\em Pebbles} programs are not stable.
 	\end{itemize}
@@
 	\begin{itemize}
-		\item Is unprotected {\tt foobar\_statistics\_counter++} bad?
+		\item Bakery algorithm: Is {\tt number[i]=max(number[0],number[1])} bad?
+		%\item Is unprotected {\tt foobar\_statistics\_counter++} bad?
 		\item What about unprotected {\tt next\_thread\_id++}?
@@
 
-\begin{frame}{Iterative Deepening}
-	Repeat state space explorations, adding preemption points, until time is exhausted.
-	\vspace{0.29in}
-	\begin{center}
-		\includegraphics[width=0.25\textwidth]{tree0.pdf}
-	\end{center}
-\end{frame}
-
-\begin{frame}{Iterative Deepening}
-	Repeat state space explorations, adding preemption points, until time is exhausted.
-	\vspace{0.15in}
-	\begin{center}
-		\begin{tabular}{cc}
-			\includegraphics[width=0.35\textwidth]{tree1.pdf} &
-			\includegraphics[width=0.35\textwidth]{tree2.pdf}
-		\end{tabular}
-	\end{center}
-\end{frame}
-
-\begin{frame}{Iterative Deepening}
-	Repeat state space explorations, adding preemption points, until time is exhausted.
-	\begin{center}
-		\includegraphics[width=0.5\textwidth]{tree3.pdf}
-	\end{center}
-	% say: and depending on your time budget, you might decide that you expect this one to be feasible, or you might fall back to a smaller state space from before.
-\end{frame}
-
 \begin{frame}{Iterative Deepening in Landslide}
-	\textbf{Landslide-ID}: wrapper program for Landslide
+	Landslide automatically iterates through different configurations of PPs.
 	\begin{itemize}
@@
 		\end{itemize}
+		\item Repeat state space explorations, adding preemption points, until time is exhausted.
 	\end{itemize}
@@
 	{\em Only required argument is CPU budget}
+\end{frame}
+
+\begin{frame}{Iterative Deepening}
+	Minimal state space includes only ``mandatory'' context switches
+	\begin{itemize}
+		\item e.g., {\tt yield()}, {\tt cond\_wait()}.
+	\end{itemize}
+	\vspace{0.29in}
+	\begin{center}
+		\includegraphics[width=0.35\textwidth]{tree0.pdf}
+	\end{center}
+\end{frame}
+
+\begin{frame}{Iterative Deepening}
+	Adding different PPs can produce state spaces of different sizes; Landslide tries them in parallel.
+	\vspace{0.15in}
+	\begin{center}
+		\begin{tabular}{cc}
+			\includegraphics[width=0.45\textwidth]{tree1.pdf} &
+			\includegraphics[width=0.5\textwidth]{tree2.pdf}
+		\end{tabular}
+	\end{center}
+\end{frame}
+
+\begin{frame}{Iterative Deepening}
+	If time allows, Landslide will combine PPs into larger, more comprehensive state spaces.
+	\begin{center}
+		\includegraphics[width=0.8\textwidth]{tree3.pdf}
+	\end{center}
+	% say: and depending on your time budget, you might decide that you expect this one to be feasible, or you might fall back to a smaller state space from before.
 \end{frame}
@@
 		\begin{itemize}
-			\item {\tt juggle}, {\tt multitest}, {\tt racer} (for 1 hour), {\tt paraguay}
+			\item {\tt juggle 4 3 2 0}, {\tt multitest}, {\tt racer} (15 min), {\tt paraguay}
 		\end{itemize}

### 2015-06-12 fp/paper.tex
Add skeleton of FP C0 tech report

--- 2013-12-17 15869/paper.tex
@@
 \usepackage{proof}
+\usepackage{wrapfig}
 \usepackage{array}
 
-\title{{\bf Dynamic Scheduling in Halide}}
-\author{Ben Blum \texttt{(bblum@andrew.cmu.edu)}}
+\title{{\bf Systematic Testing of Concurrent C0 with Landslide}}
+\author{Ben Blum \texttt{(bblum@cs.cmu.edu)} \and Garth Gibson \texttt{(garth@cs.cmu.edu)}}
 
-\date{2013, December 16}
+\date{\today}
 
@@
 
-\begin{figure}[h]
-	\begin{center}
-	\begin{tabular}{llllll}
-		\begin{tabular}{l}
-                \texttt{blur\_y(x,y) =} \\
-                \texttt{~~~~(input(x, y-1) +}\\
-                \texttt{~~~~~input(x, y~~) +}\\
-                \texttt{~~~~~input(x, y+1))/3;}\\
-                \\
-                \texttt{blur\_x(x,y) =} \\
-                \texttt{~~~~(blur\_y(x-1, y) +}\\
-                \texttt{~~~~~blur\_y(x,~~~y) +}\\
-                \texttt{~~~~~blur\_y(x+1, y))/3;}\\
-		\end{tabular}
-		& & & & &
-		\begin{tabular}{l}
-                \texttt{blur\_y\_far\_away(x,y) =} \\
-                \texttt{~~~~(input(x, {\bf y-rand()}) +}\\
-                \texttt{~~~~~input(x, y~~~~~~~) +}\\
-                \texttt{~~~~~input(x, {\bf y+rand()}))/3;}\\
-                \\
-                \texttt{blur\_x\_far\_away(x,y) =} \\
-                \texttt{~~~~(blur\_y({\bf x-rand()}, y) +}\\
-                \texttt{~~~~~blur\_y(x,~~~~~~~~y) +}\\
-                \texttt{~~~~~blur\_y({\bf x+rand()}, y))/3;}\\
-		\end{tabular}
-
-	\end{tabular}
-	\end{center}
-	\caption{Static and dynamic data dependencies in Halide.
-On the left, a 3x3 blur (the canonical example Halide program) need only reference input data up to 1 pixel away in each direction for each output pixel.
-On the right, the algorithm is modified such that the compiler cannot statically know which input pixels each output pixel may reference.}
-	\label{fig:intro}
-\end{figure}
+\begin{abstract}
+We have recently extended Landslide, our systematic testing platform originally for kernels, with some support for userspace testing.
+While the primary focus of this support is to test student thread library implementations, it can also test userspace apps built on top of those thread libraries.
+In this tech report we describe our experience using Landslide to test Concurrent C0, a type-safe instructional language with a concurrent message-passing implementation in its runtime libraries.
+\end{abstract}
 
@@
 
-Halide~\cite{halide} is a programming language for image processing pipelines designed to separate the concerns of performance and expressivity.
-The programming model is functional, representing images as functions from pixel coordinates to colour values, and representing each pass of an image processing algorithm as a function from images to images.
-Once implemented, the programmer's algorithm is then automatically parallelized by the Halide runtime, taking into account considerations such as how to split each pass into parallel tiles and the data dependencies between two consecutive passes.
-The structure of the generated loop nest and the function calls therein is known as the {\em function schedule}.
+Landslide \cite{landslide} is
 
-One weakness of Halide is its inability to express image processing passes that don't operate uniformly across their domain; i.e., algorithms whose data dependencies on portions of previous passes cannot be known statically.
-Figure~\ref{fig:intro} above contrasts two simple example algorithms, showcasing the difference between static and dynamic data dependencies.
-
-In this project I investigate and implement {\em dynamic function scheduling}, a new language feature for Halide to support such algorithms.
-My approach allows the computations for early pipeline stages to be cached across loop iterations of later stages, so that the results may be reused when data dependencies happen to coincide on the same location, saving redundant computation of a possibly expensive intermediate function.
-The principal contribution of this work is the implementation and performance evaluation of {\em dynamic scheduling at pixel granularity}.
-I also investigated dynamic scheduling at coarser (e.g.
-tile) granularities, with which entire loop iterations may be skipped.
-However, the problem was more difficult than anticipated, so I instead provide some discussion and a design sketch for future work.
+C0 \cite{c0} is...
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Motivation}
-
-\definecolor{grey}{RGB}{127,127,127}
-\definecolor{darkcyan}{RGB}{0,127,127}
-\definecolor{olivegreen}{RGB}{0,127,0}
-\definecolor{violet}{RGB}{127,0,127}
-\definecolor{brickred}{RGB}{127,0,0}
-\definecolor{brown}{RGB}{127,63,0}
-\definecolor{orange}{RGB}{192,96,0}
-\newcommand\hilight[2]{\color{#1}#2\color{black}}
-
-\begin{figure}[t]
-	\begin{center}
-	\begin{subfigure}[b]{0.45\textwidth}
-		\begin{center}
-		\begin{tabular}{l}
-		{\em User-provided schedule:} \\
-		\\
-		\hilight{blue}{\texttt{blur\_x /* nothing special */;}} \\
-		\hilight{olivegreen}{\texttt{blur\_y.store\_root().compute\_root();}} \\
-		\\
-		{\em Compiler-generated code:} \\
-		\\
-                \texttt{\hilight{olivegreen}{float~blur\_y[HEIGHT][WIDTH];}} \\
-                \texttt{\hilight{olivegreen}{for~(row~=~0~to~HEIGHT)~\{}} \\
-                \texttt{\hilight{olivegreen}{~~for~(col~=~0~to~WIDTH)~\{}} \\
-                \texttt{\hilight{olivegreen}{~~~~blur\_y[row][col] = ...;}}\\
-                \texttt{\hilight{olivegreen}{~~\}}} \\
-                \texttt{\hilight{olivegreen}{\}}} \\
-		\\
-                \texttt{\hilight{blue}{float~blur\_x[HEIGHT][WIDTH];}} \\
-                \texttt{\hilight{blue}{for~(row~=~0~to~HEIGHT)~\{}} \\
-                \texttt{\hilight{blue}{~~for~(col~=~0~to~WIDTH)~\{}} \\
-                \texttt{\hilight{blue}{~~~~blur\_x[row][col] = ...;}}\\
-                \texttt{\hilight{blue}{~~\}}} \\
-                \texttt{\hilight{blue}{\}}} \\
-		\end{tabular}
-		\end{center}
-		\caption{\texttt{blur\_x} and \texttt{blur\_y} are scheduled separately.
-This schedule, while easy to understand, offers poor locality (values computed are not used until a full image traversal later) and poor parallelism (creating one thread per row or column of pixels both floods the system with too many threads and creates cache contention).}
-		\label{fig:scheds-a}
-	\end{subfigure}
-	\qquad
-	\begin{subfigure}[b]{0.5\textwidth}
-		\begin{center}
-		\begin{tabular}{l}
-		{\em User-provided schedule:} \\
-		\\
-                \hilight{blue}{\texttt{blur\_x.tile(x, y, x2, y2, 8, 8).parallel(y);}}\\
-                \hilight{olivegreen}{\texttt{blur\_y.store\_at(blur\_x, x).compute\_at(blur\_x, y2);}}\\
-                \\
-		{\em Compiler-generated code:} \\
-		\\
-                \texttt{\hilight{blue}{float~blur\_x[HEIGHT][WIDTH];}} \\
-                \texttt{\hilight{blue}{PARALLEL for~(row~=~0~to~8)~\{}} \\
-                \texttt{\hilight{blue}{~~for~(col~=~0~to~8)~\{}} \\
-                \texttt{\hilight{olivegreen}{~~~~float~blur\_y[HEIGHT/8 + 2][WIDTH/8 + 2];}} \\
-                \texttt{\hilight{blue}{~~~~for~(row2~=~0~to~WIDTH/8 + 2)~\{}} \\
-                \texttt{\hilight{olivegreen}{~~~~~~for~(col2~=~0~to~WIDTH/8 + 2)~\{}} \\
-                \texttt{\hilight{olivegreen}{~~~~~~~~blur\_y[row2][col2]~=~...;}} \\
-                \texttt{\hilight{olivegreen}{~~~~~~\}}} \\
-                \texttt{\hilight{blue}{~~~~~~for~(col2~=~0~to~WIDTH/8)~\{}} \\
-                \texttt{\hilight{blue}{~~~~~~~~blur\_x[row*8~+~row2][col*8~+~col2]~=~...;}} \\
-                \texttt{\hilight{blue}{~~~~~~\}}} \\
-                \texttt{\hilight{blue}{~~~~\}}} \\
-                \texttt{\hilight{blue}{~~\}}} \\
-                \texttt{\hilight{blue}{\}}} \\
-		\end{tabular}
-		\end{center}
-		\caption{\texttt{blur\_x} is tiled in 8-fold splits in each dimension, and \texttt{blur\_y} is scheduled at the tile level within \texttt{blur\_x}.
-This schedule offers good locality (values are used soon after being computed) and good parallelism (one thread per tile; tile size can be customized).}
-		\label{fig:scheds-b}
-	\end{subfigure}
-	\end{center}
-	\caption{Two example function schedules for the two-stage blur algorithm shown in Figure~\ref{fig:intro}.}
-	\label{fig:scheds}
-\end{figure}
-
-The high concept of Halide is to separate the logic of an image processing algorithm from the order in which the functions for each stage are scheduled.
-Figure~\ref{fig:scheds} shows two possible schedules for our example algorithm, and discusses the performance characteristics of each.
-Though Figure~\ref{fig:scheds-b} is more cache-friendly and offers superior paralellism opportunity, a notable disadvantage is that some computation is duplicated across tiles.
-This is shown in the extra 2 pixels computed for \texttt{blur\_y} in each dimension.
-In this example, the overhead is minor, as Halide can statically reason about the maximum extra ``padding'' that \texttt{blur\_y} must compute for each tile of \texttt{blur\_x}.
-
-However, this extra computation can become a significant problem when the dependencies between the algorithm's stages are computed at runtime and hence are not available for compile-time reasoning, as in the PatchMatch algorithm~\cite{patchmatch} and in edge detection.
-If the user wishes to use a cache-friendly, highly-parallel function schedule, as in Figure~\ref{fig:scheds-b}, Halide must make a conservative estimate regarding how much extra padding is required, resulting in extra work duplication.
-
-To get a sense for how much extra unnecessary computation can arise, I introduce a sample two-stage Halide program in which the second stage (\texttt{flip()}) dynamically computes which pixels it requires from the first stage (\texttt{invert()}).
-This program is described in Figures~\ref{fig:flip} and~\ref{fig:flip-code}.
-I then ran this program on a test image using the simple schedule from Figure~\ref{fig:scheds-a} and using several nested schedules as in Figure~\ref{fig:scheds-b}.
-I found that in the former situation, 13,068 calls to \texttt{invert()} were required, while in the latter, \texttt{invert()} was invoked anywhere between 252,648 to 7,847,400 times depending on the schedule used.
-Hence, the motivation for {\em dynamic scheduling} is to extend Halide's ability to avoid duplicate work to apply to this wider range of algorithms.
-
-\begin{figure}[t]
-	\begin{tabular}{cc}
-	\begin{subfigure}[b]{0.45\textwidth}
-		\begin{center}
-		\includegraphics[width=0.85\textwidth]{canoe-0.png}
-		\end{center}
-		\caption{Original input image.}
-	\end{subfigure} &
-	\begin{subfigure}[b]{0.45\textwidth}
-		\begin{center}
-		\includegraphics[width=0.85\textwidth]{canoe-1.png}
-		\end{center}
-		\caption{Stage 1 is a simple value-invert of each input pixel.}
-		\label{fig:invert}
-	\end{subfigure} \\
-	& \\
-	\begin{subfigure}[b]{0.45\textwidth}
-		\begin{center}
-		\includegraphics[width=0.85\textwidth]{canoe-2.png}
-		\end{center}
-		\caption{The core of stage 2 is a transformation across a vector field, which maps points near the center to the corners and points near the corners to the center, followed by a 3x3 blur.
-The blur causes each output pixel to depend on multiple input pixels, which allows for the possibility of work duplication.}
-		\label{fig:flip-blur-1}
-	\end{subfigure} &
-	\begin{subfigure}[b]{0.45\textwidth}
-		\begin{center}
-		\includegraphics[width=0.85\textwidth]{canoe-3.png}
-		\end{center}
-		\caption{Final output.
-Stage 2 is altered to only apply the vector field to pixels when their colour value is darker than a certain threshold.
-This condition, which depends on a nondeterministic property of the input image, prevents Halide's optimizations from reasoning about the data dependencies between the two stages.}
-		\label{fig:flip-blur-2}
-	\end{subfigure}
-	\end{tabular}
-	\caption{Our sample image processing program consists of two stages, \texttt{invert()} (shown in \ref{fig:invert}) and \texttt{flip()} (shown in \ref{fig:flip-blur-1} and \ref{fig:flip-blur-2}).
-I designed this program expressly to create dynamic data dependencies with wide ranges across the image, to ensure a static function schedule could not avoid duplicating work.}
-
-	\label{fig:flip}
-\end{figure}
-
-\begin{figure}[t]
-	\begin{center}
-	\begin{tabular}{l}
-\texttt{\hilight{olivegreen}{Var}~x(\hilight{brickred}{"column"}), y(\hilight{brickred}{"row"}), c(\hilight{brickred}{"colour"});} \\
-\\
-\texttt{\hilight{darkcyan}{/* Stage 1 */}} \\
-\texttt{\hilight{olivegreen}{Func}~invert(\hilight{olivegreen}{Func}~input)~\{} \\
-\texttt{~~~~\hilight{olivegreen}{Func}~f(\hilight{brickred}{"invert"});} \\
-\texttt{~~~~f(x,~y,~c)~=~\hilight{brickred}{1}~-~input(x,~y,~c);~} \\
-\texttt{~~~~\hilight{brown}{return}~f;} \\
-\texttt{\}} \\
-\\
-\texttt{\hilight{darkcyan}{/* The vector field used in stage 2 */}} \\
-\texttt{\#define~vf\_x(image,~do\_flip,~x)~(\hilight{brown}{select}(do\_flip,~x,~((\hilight{brickred}{3}~*~image.width()~~/~\hilight{brickred}{2})~-~x)))} \\
-\texttt{\#define~vf\_y(image,~do\_flip,~y)~(\hilight{brown}{select}(do\_flip,~y,~((\hilight{brickred}{3}~*~image.height()~/~\hilight{brickred}{2})~-~y)))} \\
-\texttt{} \\
-\texttt{\hilight{darkcyan}{/* Stage 2 */}} \\
-\texttt{\hilight{olivegreen}{Func}~flip(\hilight{olivegreen}{Image}<\hilight{olivegreen}{float}>~\&i,~\hilight{olivegreen}{Func}~input)~\{} \\
-\texttt{~~~~\hilight{olivegreen}{Expr}~do\_flip~=~input(x,~y,~c)~>~\hilight{brickred}{0.5f};} \\
-\texttt{~~~~\hilight{olivegreen}{Func}~f(\hilight{brickred}{"flip"});} \\
-\texttt{~~~~f(x,~y,~c)~=} \\
-\texttt{~~~~~~~~(input(vf\_x(i,~do\_flip,~x-\hilight{brickred}{1}),~vf\_y(i,~do\_flip,~y-\hilight{brickred}{1}),~c)~+} \\
-\texttt{~~~~~~~~~\hilight{darkcyan}{/* ... intermediate coordinates of 3x3 blur omitted ... */}} \\
-\texttt{~~~~~~~~~input(vf\_x(i,~do\_flip,~x+\hilight{brickred}{1}),~vf\_y(i,~do\_flip,~y+\hilight{brickred}{1}),~c))~/~\hilight{brickred}{9};} \\
-\texttt{~~~~\hilight{brown}{return}~f;} \\
-\texttt{\}} \\
-
-	\end{tabular}
-	\end{center}
-	\caption{Selected code from the implementation of our sample image processing program.}
-	\label{fig:flip-code}
-\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Design and Implementation}
-
-\subsection{Overview}
-
-Conceptually, my goal for dynamic function scheduling as a Halide language feature was as follows.
-In addition to the existing directives \texttt{compute\_at()}, \texttt{store\_at()}, and so on which users may call on their \texttt{Func} objects, supplying appropriate loop variable arguments, I would add a new directive \texttt{compute\_lazy()}.
-
-{\bf Pixel-granularity dynamic scheduling.} Used with no arguments, this directive would instruct Halide to allocate a ``bitmask'', an array of booleans with the same dimensions as the array used to store the function's results themselves.
-Halide would then surround any expression which computed a value to store in the array at a certain index with a check to this bitmask at the same index, which would cause the expression to be bypassed if a value for that index had been computed before.
-I refer to this as dynamic scheduling at {\em pixel granularity} because there is one bitmask check per expression that would compute a value for a single pixel.
-In Figure~\ref{fig:dynamic-pixel} I show sample code that Halide would generate for a program that used pixel-level dynamic scheduling.
-
-% TODO
-{\bf Tile-granularity dynamic scheduling.} DISCUSSION FORTHCOMING.
-
-\begin{figure}[t]
-	\begin{center}
-	\begin{subfigure}[b]{0.475\textwidth}
-		\begin{center}
-		\begin{tabular}{l}
-		{\em User-provided schedule:} \\
-		\\
-                \hilight{blue}{\texttt{flip.tile(col, row, col2, row2, 8, 8);}} \\
-                \hilight{olivegreen}{\texttt{invert.store\_at(flip, col);}} \\
-		\hilight{olivegreen}{\texttt{invert.compute\_at(invert, row2);}}\\
-                {\bf \hilight{orange}{\texttt{invert.compute\_lazy();}}} \\
-                \\
-		{\em Compiler-generated code:} \\
-		\\
-                \texttt{\hilight{blue}{float~flip[HEIGHT][WIDTH];}} \\
-                \texttt{\hilight{blue}{for~(row~=~0~to~8)~\{}} \\
-                \texttt{\hilight{blue}{~~for~(col~=~0~to~8)~\{}} \\
-                \texttt{\hilight{olivegreen}{~~~~float~invert[...][...];}} \\
-		{\bf \texttt{\hilight{orange}{~~~~bool~result\_computed[...][...];}}} \\
-                \texttt{\hilight{blue}{~~~~for~(row2~=~0~to~WIDTH/8)~\{}} \\
-                \texttt{\hilight{olivegreen}{~~~~~~let min\_row = ...;}} \\
-                \texttt{\hilight{olivegreen}{~~~~~~let max\_row = ...;}} \\
-                \texttt{\hilight{olivegreen}{~~~~~~let min\_col = ...;}} \\
-                \texttt{\hilight{olivegreen}{~~~~~~let max\_col = ...;}} \\
-                \texttt{\hilight{olivegreen}{~~~~~~for~(row\_i~=~min\_row~to~max\_row)~\{}} \\
-                \texttt{\hilight{olivegreen}{~~~~~~~~for~(col\_i~=~min\_col~to~max\_col)~\{}} \\
-		{\bf \texttt{\hilight{orange}{~~~~~~~~~~if (!result\_computed[...][...]) \{}}} \\
-                \texttt{\hilight{olivegreen}{~~~~~~~~~~~~invert[row\_i][col\_i]~=~...;}} \\
-		{\bf \texttt{\hilight{orange}{~~~~~~~~~~~~result\_computed[...][...]~=~TRUE;}}} \\
-		{\bf \texttt{\hilight{orange}{~~~~~~~~~~\}}}} \\
-                \texttt{\hilight{olivegreen}{~~~~~~~~\}}} \\
-                \texttt{\hilight{olivegreen}{~~~~~~\}}} \\
-                \texttt{\hilight{blue}{~~~~~~for~(col2~=~0~to~WIDTH/8)~\{}} \\
-                \texttt{\hilight{blue}{~~~~~~~~flip[...][...]~=~...;}} \\
-                \texttt{\hilight{blue}{~~~~~~\}}} \\
-                \texttt{\hilight{blue}{~~~~\}}} \\
-                \texttt{\hilight{blue}{~~\}}} \\
-                \texttt{\hilight{blue}{\}}} \\
-		\end{tabular}
-		\end{center}
-		\caption{Dynamic scheduling at pixel granularity.
-The directive \texttt{invert.compute\_lazy()} with no arguments is shorthand for supplying \texttt{invert}'s innermost loop variable (in this case, \texttt{col\_i}).}
-		\label{fig:dynamic-pixel}
-	\end{subfigure}
-	\qquad
-	\begin{subfigure}[b]{0.475\textwidth}
-		\begin{center}
-		\begin{tabular}{l}
-		{\em User-provided schedule:} \\
-		\\
-                \hilight{blue}{\texttt{flip.tile(col, row, col2, row2, 8, 8);}} \\
-                \hilight{olivegreen}{\texttt{invert.store\_at(flip, col);}} \\
-		\hilight{olivegreen}{\texttt{invert.compute\_at(invert, row2);}}\\
-                {\bf \hilight{orange}{\texttt{invert.compute\_lazy(row\_i);}}} \\
-                \\
-		{\em Compiler-generated code:} \\
-		\\
-                \texttt{\hilight{blue}{float~flip[HEIGHT][WIDTH];}} \\
-                \texttt{\hilight{blue}{for~(row~=~0~to~8)~\{}} \\
-                \texttt{\hilight{blue}{~~for~(col~=~0~to~8)~\{}} \\
-                \texttt{\hilight{olivegreen}{~~~~float~invert[...][...];}} \\
-		{\bf \texttt{\hilight{orange}{~~~~bool~result\_computed[...][...];}}} \\
-                \texttt{\hilight{blue}{~~~~for~(row2~=~0~to~WIDTH/8)~\{}} \\
-                \texttt{\hilight{olivegreen}{~~~~~~let min\_row = ...;}} \\
-                \texttt{\hilight{olivegreen}{~~~~~~let max\_row = ...;}} \\
-                \texttt{\hilight{olivegreen}{~~~~~~let min\_col = ...;}} \\
-                \texttt{\hilight{olivegreen}{~~~~~~let max\_col = ...;}} \\
-                \texttt{\hilight{olivegreen}{~~~~~~for~(row\_i~=~min\_row~to~max\_row)~\{}} \\
-		{\bf \texttt{\hilight{orange}{~~~~~~~~if (!result\_computed[...][...]) \{}}} \\
-                \texttt{\hilight{olivegreen}{~~~~~~~~~~for~(col\_i~=~min\_col~to~max\_col)~\{}} \\
-                \texttt{\hilight{olivegreen}{~~~~~~~~~~~~invert[row\_i][col\_i]~=~...;}} \\
-                \texttt{\hilight{olivegreen}{~~~~~~~~~~\}}} \\
-		{\bf \texttt{\hilight{orange}{~~~~~~~~~~result\_computed[...][...]~=~TRUE;}}} \\
-		{\bf \texttt{\hilight{orange}{~~~~~~~~\}}}} \\
-                \texttt{\hilight{olivegreen}{~~~~~~\}}} \\
-                \texttt{\hilight{blue}{~~~~~~for~(col2~=~0~to~WIDTH/8)~\{}} \\
-		\texttt{\hilight{blue}{~~~~~~~~flip[...][...]~=~...;}} \\
-                \texttt{\hilight{blue}{~~~~~~\}}} \\
-                \texttt{\hilight{blue}{~~~~\}}} \\
-                \texttt{\hilight{blue}{~~\}}} \\
-                \texttt{\hilight{blue}{\}}} \\
-		\end{tabular}
-		\end{center}
-		\caption{Dynamic scheduling at tile granularity.
-Here, the ``tile'' is simply the row denited by the variable \texttt{row\_i}.
-Simply using a single boolean for each tile can be unsound in some cases, as discussed in Section~\ref{sec:future-tiles}.}
-		\label{fig:dynamic-tile}
-	\end{subfigure}
-	\end{center}
-	\caption{Generated code listings for dynamic function schedules, at both pixel and tile granularity.
-In addition to highlighting the code associated with the \hilight{blue}{flip}~and \hilight{olivegreen}{invert}~stages of the algorithm, I also {\bf \hilight{orange}{emphasize}} the additional code generated for dynamic scheduling.}
-	\label{fig:dynamic}
-\end{figure}
-
-\subsection{Code Generation}
-
-Inside the compiler, the dynamic schedule configuration is stored in a \texttt{Func}'s \texttt{Schedule} object as a \texttt{LoopLevel} named \texttt{lazy\_level}, alongside the existing configurations \texttt{store\_level} and \texttt{compute\_level}.
-Unlike the latter two, however, the \texttt{lazy\_level} is not defined in terms of the loop variable of a different function, but rather for one of the function's own loop variables.
-Hence, \texttt{lazy\_level.func} always refers to the containing function.
-
-The code generation itself occurs during the \texttt{Lower} phase of compilation, between the \texttt{schedule\_functions()} phase (found in \texttt{Lower.cpp}) and the \texttt{storage\_flattening()} phase (found in \texttt{StorageFlattening.cpp}).
-The former processes the \texttt{lazy\_level} configuration and generates IR to represent it, and the latter converts those IR nodes into conventional IR that can be code-generated.
-For this purpose I added a new type of IR node, \texttt{DynamicStmt} (found in \texttt{IR.h}), which represents the ``if not-in-cache then compute; mark-cache; end'' pattern.
-
-\begin{itemize}
-	\item {\bf Function Scheduling.} This phase converts the dynamic schedule configuration into IR in two places.
-The first is when it generates a \texttt{Realize} node (in \texttt{build\_realize()}); it simply sets the node's \texttt{lazy} flag to \texttt{true} if the associated function's \texttt{lazy\_level} was set.\footnote{
-	I experimented here with computing and storing an expression representing the necessary size for the bitmask, rather than just a boolean. Future work may find this approach worth doing after all.}
-	The second is when building a function's loop nest (in \texttt{build\_provide\_loop\_nest()}); when the approprate loop level is encountered, it inserts a \texttt{DynamicStmt} node between the \texttt{For} loop and its body.
-	\item {\bf Storage Flattening.} This phase visits the IR described above and ``flattens'' it into arrays more reminiscent of the C memory model.
-In \texttt{visit(Realize *)}, where it makes the \texttt{Allocate} node for the function's computed results, it now also will make another \texttt{Allocate} and a brief initialization loop if \texttt{realize->lazy} is set.
-Second, in \texttt{visit(DynamicStmt *)}, it converts the \texttt{DynamicStmt} node into a conventional \texttt{IfThenElse} node using the provided index for the allocated bitmask.
-\end{itemize}
-
-\subsection{Legality Checking}
-
-During the function scheduling in \texttt{Lower.cpp} Halide must also check that a specified dynamic schedule is legal. I currently check for the following cases:
-
-\begin{itemize}
-	\item {\bf Vectorized loop.} If any of the loops between the \texttt{store\_level} and the \texttt{lazy\_level} are vectorized, it may not make sense to emit a check which conditionally skips much of the computation.
-Though it would make sense to generate code for a single boolean check around a single pixel computation, with more sophisticated tile-level dynamic scheduling the vectorized execution might behave pathologically (especially if the conditional check involves traversing a sophisticated data structure as suggested in Section~\ref{sec:future-tiles}).
-	\item {\bf Parallel loop.} If any of the loops between the \texttt{store\_level} and the \texttt{lazy\_level} are parallelized, then one thread attempting to record the result of a computation might race with another thread attempting to use that result.
-With x86's sequentially-consistent memory model and the simple boolean check, this is not an issue, but when compiling for ARM or using a more sophisticated data structure for memoizing Halide would need to emit memory barriers at best and locks at worst.
-This is not infeasible to implement, but raises research questions of its own and so is left to future work.
-	\item {\bf Update phase.} Dynamic scheduling is not (yet) compatible with functions which have an update phase in their \texttt{Pipeline} node, which can arise when the user provides a reduction definition.
-This is merely a case I did not consider in detail, and may well be possible to handle.
-\end{itemize}
-
-\subsection{Code}
-
-The reader should feel free to find and peruse my code at \texttt{https://github.com/bblum/Halide}. During the weeks following this submission I will polish it and submit a pull request upstream.
+\section{Techniques}
 
@@
 \section{Evaluation}
-% TODO
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Open Questions}
-% TODO
-
-\label{sec:future-tiles}
+evaluating stuff here
 
@@
 \section{Conclusion}
-% TODO
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section*{Acknowledgements}
-
-Thanks to Jonathan Ragan-Kelley and Andrew Adams, developers of Halide, for their guidance, insights, and patience.
-Thanks to Kayvon Fatahalian for supervising this project and for a great semester in 15-869.
 

### 2015-06-12 fp/paper.tex
write some more in fp tr

@@
 \title{{\bf Systematic Testing of Concurrent C0 with Landslide}}
-\author{Ben Blum \texttt{(bblum@cs.cmu.edu)} \and Garth Gibson \texttt{(garth@cs.cmu.edu)}}
+\author{Ben Blum \\ \texttt{bblum@cs.cmu.edu}
+	\and Frank Pfenning \\ \texttt{fp@cs.cmu.edu}
+	\and Garth Gibson \\ \texttt{garth@cs.cmu.edu}}
 
@@
 While the primary focus of this support is to test student thread library implementations, it can also test userspace apps built on top of those thread libraries.
-In this tech report we describe our experience using Landslide to test Concurrent C0, a type-safe instructional language with a concurrent message-passing implementation in its runtime libraries.
+In this tech report we describe our experience using Landslide to test Concurrent C0, a type-safe educational language with a concurrent message-passing implementation in its runtime libraries.
 \end{abstract}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Background}
+
+We assume the reader will be familiar with the background in {\em systematic testing} \cite{verisoft}, the concurrency testing technique upon which Landslide is built.
+Examples of state-of-the-art systematic testing tools from prior work include dBug \cite{dbug-ssv,dbug-retreat,parrot}, CHESS \cite{chess}, MoDist \cite{modist}, and MaceMC \cite{macemc}.
+
+The primary state-space reduction technique for systematic testing is {\em dynamic partial order reduction} (DPOR) \cite{dpor}, various enhancements of which have been proposed in \cite{distributed-dpor,sdpor,parallel-dpor,bpor,optimal-dpor}.
+Alternative and/or orthogonal systematic testing techniques include {\em iterative context bounding} (ICB) \cite{chess}, {\em dynamic interface reduction} (DIR) \cite{demeter}, {\em deterministic multithreading} (DMT) \cite{peregrine,dthreads,kendo,parrot}, {\em data race detection} \cite{eraser,tsan} and {\em state space estimation} \cite{estimation}.
+We are presently developing {\em iterative deepening} (ID), a new orthogonal technique for automatically configuring subsets of preemption points (PPs) according to a fixed CPU budget, which has yet to be published.
 
@@
 
-Landslide \cite{landslide} is
+Landslide \cite{landslide} is a systematic testing tool built with a focus on educational kernel architectures.
+It is built as a module for Wind River Simics \cite{simics}, and supports testing Pebbles kernel code \cite{kspec}, userspace thread library code running on top of Pebbles \cite{thrlib}, and any userspace programs built on top of those.
+Recently, we have also added support for testing Pintos kernels \cite{pintos} from schools such as Berkeley and U. Chicago; however in this paper we will focus on the Pebbles projects at CMU.
 
-C0 \cite{c0} is...
+C0 \cite{c0} is a C-like, type-safe educational programming language recently introduced for CMU's introductory CS curriculum.
+It has recently been extended to support concurrent programming using a message-passing implementation known as channels.
+These channels are defined by a {\tt choice}, a union-like protocol comprising multiple branches of possible message types, each with an expected payload and send/receive direction.
+Operations on these channels include sending and receiving, as well as forwarding, a way to join two channels of matching type together so messages automatically flow across them.
+
+These channels' definitions and usage are type-checked at compile-time, so no type-safety violations (i.e.
+crashes), or even deadlocks, are possible -- at least, assuming the runtime implementation is correct.
+It is this implementation which we seek to verify in this paper.
 
@@
 
+The astute follower of our work will note that the advantages Landslide offers over the state-of-the-art seem ill-matched with the testing requirements of C0.
+The C0 runtime is a userspace program built on top of pthreads, so why jump through hoops to use a testing platform as heavyweight as Landslide, when tools like dBug exist that directly target the pthreads interface?
+
+We chose Landslide because of its data-race analysis and iterative-deepening framework.
+A type-safe concurrent language's runtime must necessarily contain some type-unsafe programming patterns, i.e., some manual sharing of state between threads outside the normal mutex/condvar primitives.
+Even if all concurrent memory accesses end up being adequately protected by those primitives, we require instruction-level memory introspection to verify that property.
+Landslide, which runs in a simulated environment and runs a data-race analysis on the fly, is ideal for this task, while the library-level interposition of dBug is too coarse-grained.
+After identifying data race candidates, the iterative-deepening framework aids to manage multiple explorations to test interleavings around those accesses.
+
+\subsection{Porting C0 to Pebbles}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2015-06-25 fp/paper.tex
fp: write about DRs

@@
 It is built as a module for Wind River Simics \cite{simics}, and supports testing Pebbles kernel code \cite{kspec}, userspace thread library code running on top of Pebbles \cite{thrlib}, and any userspace programs built on top of those.
-Recently, we have also added support for testing Pintos kernels \cite{pintos} from schools such as Berkeley and U. Chicago; however in this paper we will focus on the Pebbles projects at CMU.
+Recently, we have also added support for testing Pintos kernels \cite{pintos} from schools such as Berkeley and U. Chicago; however in this project focused on the better-supported Pebbles environment.
 
@@
 \subsection{Porting C0 to Pebbles}
+
+Because the C0 concurrency runtime is implemented on top of the pthread interface, which Pebbles thread libraries, adapting it to run on top of a Pebbles kernel was a small task.
+The only missing functionality we needed to add was a {\tt thr\_detach()} implementation, requiring only about 10 lines of code.
+Apart from that, we stubbed out C0's file I/O and ncurses routines, as Pebbles does not provide system calls for those services.
+The ported C0 libraries can be found at \url{https://github.com/bblum/landslide/tree/fp/pebsim/p2-basecode/user}.
+
+\subsection{Improving Data Race Detection}
+
+\definecolor{grey}{RGB}{127,127,127}
+\definecolor{darkcyan}{RGB}{0,127,127}
+\definecolor{olivegreen}{RGB}{0,127,0}
+\definecolor{violet}{RGB}{127,0,127}
+\definecolor{brickred}{RGB}{127,0,0}
+\definecolor{brown}{RGB}{127,63,0}
+\definecolor{orange}{RGB}{192,96,0}
+\newcommand\hilight[2]{\color{#1}#2\color{black}}
+
+\begin{figure}[h]
+	\begin{subfigure}[b]{0.55\textwidth}
+		\begin{center}
+		\begin{tabular}{l|l}
+			Thread 1 & Thread 2 \\
+			\hline
+			\begin{tabular}{l}
+			\texttt{temp~=~malloc(...);} \\
+			\texttt{\hilight{brickred}{temp->foo~=~bar;}} \\
+			\\
+			\texttt{mutex\_lock(...);} \\
+			\texttt{\hilight{olivegreen}{global\_x~=~temp;}} \\
+			\texttt{mutex\_unlock(...);} \\
+			\end{tabular}
+			& \\ &
+			\begin{tabular}{l}
+			\texttt{mutex\_lock(...);} \\
+			\texttt{\hilight{olivegreen}{temp~=~global\_x;}} \\
+			\texttt{mutex\_unlock(...);} \\
+			\\
+			\texttt{\hilight{brickred}{access(temp->foo);}} \\
+			\end{tabular}
+		\end{tabular}
+		\end{center}
+		\caption{Initialization false positive.}
+		\label{fig:dra}
+	\end{subfigure}
+	\begin{subfigure}[b]{0.40\textwidth}
+		\begin{center}
+		\begin{tabular}{l|l}
+			Thread 1 & Thread 2 \\
+			\hline
+			\\
+			\begin{tabular}{l}
+			\texttt{\hilight{brickred}{access(y->foo);}} \\
+			\texttt{\hilight{olivegreen}{free(y);}} \\
+			\end{tabular}
+			& \\ &
+			\begin{tabular}{l}
+			{\em // malloc reuses y's} \\
+			{\em // old address for z} \\
+			\texttt{\hilight{olivegreen}{z~=~malloc(...);}} \\
+			\texttt{\hilight{brickred}{z->foo~=~quux;}} \\
+			\\
+			\end{tabular}
+		\end{tabular}
+		\\
+		\end{center}
+		\caption{Free/re-malloc false positive.}
+		\label{fig:drb}
+	\end{subfigure}
+	\caption{Example code patterns that produce false positive data race reports.
+In both cases, the accesses to the {\tt foo} field appear concurrent because the threads' transitions can be reordered, but in such a case, the value of the pointer dereferenced would change.}
+	\label{fig:dr}
+\end{figure}
+
+In the course of making sure the new Landslide instrumentation worked correctly, we found two old problems with the data race (DR) detector that were exposed in testing C0 programs.
+Both cases follow a "message sending" pattern, where while both threads are co-enabled, reordering the threads would cause them to access separate memory locations entirely.
+This is because, although there is no concrete happens-before relation (from DPOR's perspective), a secondary pair of accesses elsewhere creates causality between the two primary accesses.
+The two patterns are shown in Figure~\ref{fig:dr}, the primary access pair (the one falsely flagged as a DR) is coloured in \hilight{brickred}{red}, and the secondary pair is coloured in \hilight{olivegreen}{green}.
+
+Originally, we thought we could suppress these false positives by classifying DR candidates as ``suspected'' or ``confirmed'', depending on whether we've observed them reordered.
+A DR being ``confirmed'' would indicate that the racy accesses truly can be interleaved at instruction granularity.
+However, because Landslide's testing model involves allowing threads to run several shared memory accesses before being preempted, some suspected DRs may truly be bugs, yet impossible to confirm (i.e.
+reorder) without more fine-grained PPs.
+In a systematic testing model where threads are preempted between all shared memory accesses, this suspected/confirmed method would work for soundly eliminating false positives.
+Of course, part of Landslide's philosophy is to test state spaces without all PPs enabled, and the whole point of integrating DR detection with systematic testing is the figure out which subsets of memory accesses to preempt on to begin with.
+Hence the need for heuristically eliminating these false positives.
+
+{\bf Initialization False Positives.} As shown in Figure~\ref{fig:dra}, one thread may locklessly initialize the fields of a shared data structure immediately after {\tt malloc}ing it, before publishing it to be visible to other threads.
+The second thread then accesses the global pointer under a lock, but then ``takes ownership'' of the pointed-to struct (e.g., by popping it off a linked list) and so does not need a lock during the dereference.
+If the threads were reordered, the locked access to the global pointer would return either {\tt NULL} or a different pointer instead.
+
+The Eraser paper \cite{eraser} describes a heuristic for suppressing these DR reports by waiting until the memory is accessed by 2 threads, but this introduces the possibility of false negatives.
+We think the best way of dealing with this problem is to incorporate a dataflow analysis, either static or dynamic, to track {\tt malloc}ed pointers and identify when they are first stored in a globally-visible location (i.e., not on the initializing thread's stack).
+However, in Landslide we are taking a far simpler approach for now. We provide an annotation, {\tt ignore\_dr\_function}, to identify regions of code following this initialization pattern.
+
+{\bf Free/Re-Malloc False Positives.} Figure~\ref{fig:drb} shows a similar pattern.
+Here, the variables each thread accesses are not even logically connected to the same data structure, and may not even be the same type.
+However, in the course of {\tt free}ing and {\tt malloc}ing memory, the same address is used in a later co-enabled thread transition.
+If the threads were reordered, {\tt malloc} would simply return a different heap chunk and the accesses would not collide.
+We fix these false positives as a special case by introducing a generation counter into our heap tracker module.
+Whenever a memory access is recorded for later data race analysis, the containing heap chunk's generation counter (if any) is stored as metadata, and later compared against the generation counters of other candidate accesses.
+
+\subsection{Test Case Design}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2015-06-25 fp/paper.tex
fp - add test case source codez

@@
 {\bf Free/Re-Malloc False Positives.} Figure~\ref{fig:drb} shows a similar pattern.
-Here, the variables each thread accesses are not even logically connected to the same data structure, and may not even be the same type.
+Here, the variables each thread accesses are not logically connected to the same data structure, and may not even be the same type.
 However, in the course of {\tt free}ing and {\tt malloc}ing memory, the same address is used in a later co-enabled thread transition.
@@
 \bibliographystyle{alpha}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\appendix
+\section{Test Case Source Code}
+\label{sec:codez}
+\subsection{odd-even-sort4.c1}
+\small
+
+\begin{tabular}{l}
+	\texttt{\hilight{darkcyan}{//test~return~0}} \\
+	\texttt{} \\
+	\texttt{\hilight{darkcyan}{/*~}} \\
+	\texttt{\hilight{darkcyan}{~*~A~version~of~odd-even~sort~where~only~the}} \\
+	\texttt{\hilight{darkcyan}{~*~'tail'~process~counts~down~to~0~(or~1).~~Processes}} \\
+	\texttt{\hilight{darkcyan}{~*~alternate~between~elemL~and~elemR~to~enhance}} \\
+	\texttt{\hilight{darkcyan}{~*~parallelism,~even~while~the~sorting~network}} \\
+	\texttt{\hilight{darkcyan}{~*~is~constructed.}} \\
+	\texttt{\hilight{darkcyan}{~*}} \\
+	\texttt{\hilight{darkcyan}{~*~This~version~tests~on~all~permutations~of~a~given~size.}} \\
+	\texttt{\hilight{darkcyan}{~*/}} \\
+	\texttt{} \\
+	\texttt{\hilight{violet}{\#use~<conio>}} \\
+	\texttt{} \\
+	\texttt{\hilight{brown}{choice}~\hilight{olivegreen}{list}~\{} \\
+	\texttt{~~<>~~~~~~~~~~~~~~~~~~~\hilight{brickred}{Nil};} \\
+	\texttt{~~<!\hilight{olivegreen}{int};~!\hilight{brown}{choice}~\hilight{olivegreen}{list}>~\hilight{brickred}{Cons};} \\
+	\texttt{\};} \\
+	\texttt{\hilight{olivegreen}{typedef}~<!\hilight{brown}{choice}~\hilight{olivegreen}{list}>~\hilight{olivegreen}{list};} \\
+	\texttt{} \\
+	\texttt{\hilight{olivegreen}{list}~\$c~nil()~\{} \\
+	\texttt{~~\$c.\hilight{brickred}{Nil};} \\
+	\texttt{~~\hilight{blue}{close}(\$c);} \\
+	\texttt{\}} \\
+	\texttt{} \\
+	\texttt{\hilight{olivegreen}{list}~\$c~cons(\hilight{olivegreen}{int}~n,~\hilight{olivegreen}{list}~\$d)~\{} \\
+	\texttt{~~\$c.\hilight{brickred}{Cons};} \\
+	\texttt{~~\hilight{blue}{send}(\$c,~n);} \\
+	\texttt{~~\$c~=~\$d;} \\
+	\texttt{\}} \\
+	\texttt{} \\
+	\texttt{\hilight{brown}{choice}~cell~\{} \\
+	\texttt{~~<!\hilight{olivegreen}{int};~?\hilight{brown}{choice}~cell\_response>~SomeR;} \\
+	\texttt{~~<!\hilight{brown}{choice}~cell>~~~~~~~~~~~~~~~~Tail;} \\
+	\texttt{~~<!\hilight{olivegreen}{list};~>~~~~~~~~~~~~~~~~~~~~~DoneR;} \\
+	\texttt{\};} \\
+	\texttt{\hilight{brown}{choice}~cell\_response~\{} \\
+	\texttt{~~<?\hilight{olivegreen}{int};~!\hilight{brown}{choice}~cell>~~SomeL;} \\
+	\texttt{~~<!\hilight{brown}{choice}~cell>~~~~~~~~Head;} \\
+	\texttt{\};} \\
+	\texttt{\hilight{olivegreen}{typedef}~<!\hilight{brown}{choice}~cell>~cell;} \\
+	\texttt{\hilight{olivegreen}{typedef}~<?\hilight{brown}{choice}~cell\_response>~cell\_response;} \\
+	\texttt{} \\
+	\texttt{cell~\$c~~~~~elemL(\hilight{olivegreen}{int}~m,~cell~\$d);} \\
+	\texttt{cell~\$c~~~~~elemR(\hilight{olivegreen}{int}~m,~cell~\$d);} \\
+	\texttt{<!\hilight{olivegreen}{list};>~\$c~head(cell~\$d);} \\
+	\texttt{cell~\$c~~~~~tail(\hilight{olivegreen}{int}~i);} \\
+	\texttt{} \\
+	\texttt{cell~\$c~elemL(\hilight{olivegreen}{int}~m,~cell~\$d)~\{} \\
+	\texttt{~~~~\hilight{brown}{switch}(\$d)~\{} \\
+	\texttt{~~~~~~\hilight{brown}{case}~SomeR:} \\
+	\texttt{~~~~~~~~\hilight{olivegreen}{int}~n~=~\hilight{blue}{recv}(\$d);} \\
+	\texttt{~~~~~~~~\$d.SomeL;} \\
+	\texttt{~~~~~~~~\hilight{blue}{send}(\$d,~m);} \\
+\end{tabular}
+\begin{tabular}{l}
+	\texttt{~~~~~~~~\hilight{brown}{if}~(m~>~n)} \\
+	\texttt{~~~~~~~~~~\$c~=~elemR(n,~\$d);} \\
+	\texttt{~~~~~~~~else} \\
+	\texttt{~~~~~~~~~~\$c~=~elemR(m,~\$d);} \\
+	\texttt{~~~~~~\hilight{brown}{case}~Tail:} \\
+	\texttt{~~~~~~~~\$c~=~elemR(m,~\$d);} \\
+	\texttt{~~~~~~\hilight{brown}{case}~DoneR:} \\
+	\texttt{~~~~~~~~\hilight{olivegreen}{list}~\$l~=~\hilight{blue}{recv}(\$d);} \\
+	\texttt{~~~~~~~~\hilight{blue}{wait}(\$d);} \\
+	\texttt{~~~~~~~~\$c.DoneR;} \\
+	\texttt{~~~~~~~~\hilight{olivegreen}{list}~\$k~=~cons(m,~\$l);} \\
+	\texttt{~~~~~~~~\hilight{blue}{send}(\$c,~\$k);} \\
+	\texttt{~~~~~~~~\hilight{blue}{close}(\$c);} \\
+	\texttt{~~~~\}} \\
+	\texttt{\}} \\
+	\texttt{cell~\$c~elemR(\hilight{olivegreen}{int}~m,~cell~\$d)~\{} \\
+	\texttt{~~\$c.SomeR;} \\
+	\texttt{~~\hilight{blue}{send}(\$c,~m);} \\
+	\texttt{~~\hilight{brown}{switch}~(\$c)~\{} \\
+	\texttt{~~~~\hilight{brown}{case}~SomeL:} \\
+	\texttt{~~~~~~\hilight{olivegreen}{int}~n~=~\hilight{blue}{recv}(\$c);} \\
+	\texttt{~~~~~~\hilight{brown}{if}~(n~>~m)} \\
+	\texttt{~~~~~~~~\$c~=~elemL(n,~\$d);} \\
+	\texttt{~~~~~~else} \\
+	\texttt{~~~~~~~~\$c~=~elemL(m,~\$d);} \\
+	\texttt{~~~~\hilight{brown}{case}~Head:} \\
+	\texttt{~~~~~~\$c~=~elemL(m,~\$d);} \\
+	\texttt{~~\}} \\
+	\texttt{\}} \\
+	\texttt{} \\
+	\texttt{<!\hilight{olivegreen}{list};>~\$c~head(cell~\$d)~\{} \\
+	\texttt{~~\hilight{brown}{switch}~(\$d)~\{} \\
+	\texttt{~~~~\hilight{brown}{case}~SomeR:} \\
+	\texttt{~~~~~~\hilight{olivegreen}{int}~n~=~\hilight{blue}{recv}(\$d);} \\
+	\texttt{~~~~~~\$d.Head;} \\
+	\texttt{~~~~~~\$c~=~head(\$d);} \\
+	\texttt{~~~~\hilight{brown}{case}~Tail:} \\
+	\texttt{~~~~~~\$c~=~head(\$d);} \\
+	\texttt{~~~~\hilight{brown}{case}~DoneR:} \\
+	\texttt{~~~~~~\$c~=~\$d;} \\
+	\texttt{~~\}} \\
+	\texttt{\}} \\
+	\texttt{} \\
+	\texttt{cell~\$c~tail(\hilight{olivegreen}{int}~i)~\{} \\
+	\texttt{~~\hilight{brown}{if}~(i~>~1)~\{} \\
+	\texttt{~~~~\$c.Tail;} \\
+	\texttt{~~~~\$c~=~tail(i-2);} \\
+	\texttt{~~\}~else~\{~\hilight{darkcyan}{/*~i~==~0~or~1~*/}} \\
+	\texttt{~~~~\$c.DoneR;} \\
+	\texttt{~~~~\hilight{olivegreen}{list}~\$l~=~nil();} \\
+	\texttt{~~~~\hilight{blue}{send}(\$c,~\$l);} \\
+	\texttt{~~~~\hilight{blue}{close}(\$c);} \\
+	\texttt{~~\}} \\
+	\texttt{\}} \\
+	%\texttt{} \\
+\end{tabular}
+
+\begin{tabular}{l}
+	\texttt{\hilight{olivegreen}{void}~sort(\hilight{olivegreen}{int}[]~A,~\hilight{olivegreen}{int}[]~B,~\hilight{olivegreen}{int}~n)} \\
+	\texttt{\hilight{darkcyan}{//@requires~0~<=~n~\&\&~n~<=~length(A);}} \\
+	\texttt{\hilight{darkcyan}{//@requires~0~<=~n~\&\&~n~<=~length(B);}} \\
+	\texttt{\{} \\
+	\texttt{~~cell~\$c~=~tail(n);} \\
+	\texttt{~~\hilight{olivegreen}{int}~i;} \\
+	\texttt{~~\hilight{brown}{for}~(i~=~0;~i~<~n-1;~i~=~i+2)~\{} \\
+	\texttt{~~~~\hilight{darkcyan}{/*~alternating~L~and~R~enhances~parallelism~*/}} \\
+	\texttt{~~~~\$c~=~elemL(A[i],~\$c);} \\
+	\texttt{~~~~\$c~=~elemR(A[i+1],~\$c);} \\
+	\texttt{~~\}} \\
+	\texttt{~~\hilight{brown}{if}~(i~<~n)~\$c~=~elemL(A[i],~\$c);} \\
+	\texttt{~~<!\hilight{olivegreen}{list};>~\$d~=~head(\$c);} \\
+	\texttt{~~\hilight{olivegreen}{list}~\$l~=~\hilight{blue}{recv}(\$d);} \\
+	\texttt{~~\hilight{blue}{wait}(\$d);} \\
+	\texttt{~~\hilight{brown}{for}~(i~=~0;~i~<~n;~i~=~i+1)~\{} \\
+	\texttt{~~~~\hilight{brown}{switch}(\$l)~\{} \\
+	\texttt{~~~~~~\hilight{brown}{case}~\hilight{brickred}{Nil}:~\hilight{red}{error}(\hilight{brickred}{"impossible!(8)"});} \\
+	\texttt{~~~~~~\hilight{brown}{case}~\hilight{brickred}{Cons}:} \\
+	\texttt{~~~~~~~~B[i]~=~\hilight{blue}{recv}(\$l);} \\
+	\texttt{~~~~~~~~break;} \\
+	\texttt{~~~~\}} \\
+	\texttt{~~\}} \\
+	\texttt{~~\hilight{brown}{switch}(\$l)~\{} \\
+	\texttt{~~~~\hilight{brown}{case}~\hilight{brickred}{Nil}:~\hilight{blue}{wait}(\$l);~break;} \\
+	\texttt{~~~~\hilight{brown}{case}~\hilight{brickred}{Cons}:~\hilight{red}{error}(\hilight{brickred}{"impossible!(9)"});} \\
+	\texttt{~~\}} \\
+	\texttt{~~return;} \\
+	\texttt{\}} \\
+	\texttt{\hilight{olivegreen}{void}~print\_array(\hilight{olivegreen}{int}[]~A,~\hilight{olivegreen}{int}~n)} \\
+	\texttt{\hilight{darkcyan}{//@requires~0~<=~n~\&\&~n~<=~length(A);}} \\
+	\texttt{\{} \\
+	\texttt{~~\hilight{brown}{for}~(\hilight{olivegreen}{int}~i~=~0;~i~<~n;~i++)~\{} \\
+	\texttt{~~~~printint(A[i]);~print(\hilight{brickred}{"~"});} \\
+	\texttt{~~\}} \\
+	\texttt{~~print(\hilight{brickred}{"n"});} \\
+	\texttt{\}} \\
+	\texttt{} \\
+	\texttt{\hilight{olivegreen}{bool}~is\_sorted(\hilight{olivegreen}{int}[]~A,~\hilight{olivegreen}{int}~n)} \\
+	\texttt{\hilight{darkcyan}{//@requires~0~<=~n~\&\&~n~<=~length(A);}} \\
+	\texttt{\{} \\
+	\texttt{~~\hilight{brown}{for}~(\hilight{olivegreen}{int}~i~=~0;~i~<~n-1;~i++)} \\
+	\texttt{~~~~\hilight{brown}{if}~(A[i]~>~A[i+1])~return~false;} \\
+	\texttt{~~return~true;} \\
+	\texttt{\}} \\
+	\texttt{} \\
+	\texttt{\hilight{olivegreen}{void}~swap(\hilight{olivegreen}{int}[]~A,~\hilight{olivegreen}{int}~i,~\hilight{olivegreen}{int}~j)} \\
+	\texttt{\hilight{darkcyan}{//@requires~0~<=~i~\&\&~i~<=~length(A);}} \\
+	\texttt{\hilight{darkcyan}{//@requires~0~<=~j~\&\&~j~<=~length(A);}} \\
+	\texttt{\{} \\
+	\texttt{~~\hilight{olivegreen}{int}~tmp~=~A[i];} \\
+	\texttt{~~A[i]~=~A[j];} \\
+	\texttt{~~A[j]~=~tmp;} \\
+	\texttt{~~return;} \\
+	\texttt{\}} \\
+	%\texttt{} \\
+\end{tabular}
+\begin{tabular}{l}
+	\texttt{\hilight{darkcyan}{/*~enumerate~all~permutations,}} \\
+	\texttt{\hilight{darkcyan}{~*~based~on~Heap's~algorithm~*/}} \\
+	\texttt{\hilight{olivegreen}{void}~permute\_and\_sort(\hilight{olivegreen}{int}[]~A,~\hilight{olivegreen}{int}[]~B,} \\
+	\texttt{~~~~~~~~~~~~~~~~~~~~~~~\hilight{olivegreen}{int}~n,~\hilight{olivegreen}{int}~k)} \\
+	\texttt{\hilight{darkcyan}{//@requires~0~<=~n~\&\&~n~<=~length(A);}} \\
+	\texttt{\hilight{darkcyan}{//@requires~0~<=~n~\&\&~n~<=~length(B);}} \\
+	\texttt{\hilight{darkcyan}{//@requires~0~<=~k~\&\&~k~<~n;}} \\
+	\texttt{~\{} \\
+	\texttt{~~\hilight{brown}{if}~(k~==~0)~\{} \\
+	\texttt{~~~~\hilight{darkcyan}{//~print\_array(A,~n);}} \\
+	\texttt{~~~~sort(A,~B,~n);} \\
+	\texttt{~~~~\hilight{darkcyan}{//~print\_array(B,~n);}} \\
+	\texttt{~~~~\hilight{darkcyan}{//~print(\hilight{brickred}{"-----------n"});}} \\
+	\texttt{~~~~print(\hilight{brickred}{"."});~flush();} \\
+	\texttt{~~~~\hilight{red}{assert}(is\_sorted(B,~n));} \\
+	\texttt{~~\}~else~\{} \\
+	\texttt{~~~~\hilight{brown}{for}~(\hilight{olivegreen}{int}~i~=~0;~i~<=~k;~i++)~\{} \\
+	\texttt{~~~~~~permute\_and\_sort(A,~B,~n,~k-1);} \\
+	\texttt{~~~~~~swap(A,~k\%2==0?0:i,~k);} \\
+	\texttt{~~~~\}} \\
+	\texttt{~~\}} \\
+	\texttt{\}} \\
+	\texttt{\hilight{olivegreen}{int}~main()~\{} \\
+	\texttt{~~\hilight{olivegreen}{int}[]~A~=~alloc\_array(\hilight{olivegreen}{int},~6);} \\
+	\texttt{~~\hilight{olivegreen}{int}[]~A\_out~=~alloc\_array(\hilight{olivegreen}{int},~6);} \\
+	\texttt{} \\
+	\texttt{~~\hilight{darkcyan}{//\hilight{darkcyan}{/*~6~elements~*/}}} \\
+	\texttt{~~\hilight{darkcyan}{//\hilight{brown}{for}~(\hilight{olivegreen}{int}~i~=~0;~i~<~6;~i++)~A[i]~=~i;}} \\
+	\texttt{~~\hilight{darkcyan}{//print(\hilight{brickred}{"Testing~all~permutations~of~6~elements"});}} \\
+	\texttt{~~\hilight{darkcyan}{//permute\_and\_sort(A,~A\_out,~6,~5);}} \\
+	\texttt{~~\hilight{darkcyan}{//print(\hilight{brickred}{"n"});}} \\
+	\texttt{} \\
+	\texttt{~~\hilight{darkcyan}{/*~1~element~*/}} \\
+	\texttt{~~print(\hilight{brickred}{"Testing~all~permutations~of~1~element"});} \\
+	\texttt{~~A[0]~=~0;} \\
+	\texttt{~~permute\_and\_sort(A,~A\_out,~1,~0);} \\
+	\texttt{~~print(\hilight{brickred}{"n"});} \\
+	\texttt{} \\
+	\texttt{~~return~0;} \\
+	\texttt{\}} \\
+\end{tabular}
+
+\subsection{seg.c1}
+\small
+
+\begin{tabular}{l}
+	\texttt{\hilight{darkcyan}{//test~return~2}} \\
+	\texttt{} \\
+	\texttt{\hilight{brown}{choice}~\hilight{olivegreen}{list}~\{} \\
+	\texttt{~~<>~~~~~~~~~~~~~~~~~~~\hilight{brickred}{Nil};} \\
+	\texttt{~~<!\hilight{olivegreen}{int};~!\hilight{brown}{choice}~\hilight{olivegreen}{list}>~\hilight{brickred}{Cons};} \\
+	\texttt{\};} \\
+	\texttt{\hilight{brown}{typedef}~<!\hilight{brown}{choice}~\hilight{olivegreen}{list}>~\hilight{olivegreen}{list};} \\
+	\texttt{} \\
+	\texttt{\hilight{olivegreen}{list}~\$c~nil()~\{} \\
+	\texttt{~~\$c.\hilight{brickred}{Nil};} \\
+	\texttt{~~\hilight{blue}{close}(\$c);} \\
+	\texttt{\}} \\
+	\texttt{} \\
+	\texttt{\hilight{olivegreen}{list}~\$c~cons(\hilight{olivegreen}{int}~x,~\hilight{olivegreen}{list}~\$d)~\{~~~~~~~~~~~~~~~~} \\
+	\texttt{~~\$c.\hilight{brickred}{Cons};} \\
+	\texttt{~~\hilight{blue}{send}(\$c,~x);} \\
+	\texttt{~~\$c~=~\$d;} \\
+	\texttt{\}} \\
+	\texttt{} \\
+	\texttt{\hilight{brown}{typedef}~<?\hilight{olivegreen}{list};~\hilight{olivegreen}{list}>~seg;} \\
+	\texttt{} \\
+	\texttt{seg~\$c~empty()~\{} \\
+	\texttt{~~\hilight{olivegreen}{list}~\$tail~=~\hilight{blue}{recv}(\$c);} \\
+	\texttt{~~\$c~=~\$tail;} \\
+	\texttt{\}} \\
+	\texttt{} \\
+	\texttt{seg~\$c~concat(seg~\$d,~seg~\$e)~\{} \\
+	\texttt{~~\hilight{olivegreen}{list}~\$tail~=~\hilight{blue}{recv}(\$c);} \\
+	\texttt{~~\hilight{blue}{send}(\$e,~\$tail);} \\
+	\texttt{~~\hilight{blue}{send}(\$d,~\$e);} \\
+	\texttt{~~\$c~=~\$d;~} \\
+	\texttt{\}} \\
+	\texttt{} \\
+	\texttt{seg~\$c~prepend(\hilight{olivegreen}{int}~x,~seg~\$e)~\{} \\
+	\texttt{~~\hilight{olivegreen}{list}~\$tail~=~\hilight{blue}{recv}(\$c);} \\
+	\texttt{~~\hilight{blue}{send}(\$e,~\$tail);} \\
+	\texttt{~~\$c~=~cons(x,~\$e);} \\
+	\texttt{\}} \\
+	\texttt{} \\
+	\texttt{seg~\$c~append(seg~\$d,~\hilight{olivegreen}{int}~x)~\{} \\
+	\texttt{~~\hilight{olivegreen}{list}~\$tail~=~\hilight{blue}{recv}(\$c);} \\
+	\texttt{~~\hilight{olivegreen}{list}~\$e~=~cons(x,~\$tail);} \\
+	\texttt{~~\hilight{blue}{send}(\$d,~\$e);} \\
+	\texttt{~~\$c~=~\$d;~} \\
+	\texttt{\}} \\
+	%\texttt{} \\
+\end{tabular}
+\begin{tabular}{l}
+	\texttt{\hilight{olivegreen}{int}~read\_elements(seg~\$d)~\{} \\
+	\texttt{~~\hilight{olivegreen}{list}~\$nil~=~nil();} \\
+	\texttt{~~\hilight{blue}{send}(\$d,~\$nil);} \\
+	\texttt{~~\hilight{olivegreen}{int}~n~=~0;} \\
+	\texttt{~~while~(true)~\{} \\
+	\texttt{~~~~\hilight{brown}{switch}~(\$d)~\{} \\
+	\texttt{~~~~~~\hilight{brown}{case}~\hilight{brickred}{Nil}:~\hilight{blue}{wait}(\$d);~return~n;} \\
+	\texttt{~~~~~~\hilight{brown}{case}~\hilight{brickred}{Cons}:} \\
+	\texttt{~~~~~~~~\hilight{olivegreen}{int}~d~=~\hilight{blue}{recv}(\$d);} \\
+	\texttt{~~~~~~~~n~=~10*n+d;} \\
+	\texttt{~~~~~~~~break;} \\
+	\texttt{~~~~\}~~~} \\
+	\texttt{~~\}} \\
+	\texttt{~~\hilight{red}{error}(\hilight{brickred}{"exited~infinite~loop"});} \\
+	\texttt{\}} \\
+	\texttt{} \\
+	\texttt{\hilight{olivegreen}{int}~main()~\{} \\
+	\texttt{~~seg~\$d1~=~empty();} \\
+	\texttt{~~seg~\$d2~=~append(\$d1,~2);} \\
+	\texttt{~~\hilight{olivegreen}{int}~n~=~read\_elements(\$d2);} \\
+	\texttt{~~\hilight{red}{assert}(n~==~2);} \\
+	\texttt{~~return~n;} \\
+	\texttt{\}} \\
+\end{tabular}
+
+\pagebreak
+\subsection{bitstring1.c1}
+
+\small
+
+\begin{tabular}{l}
+	\texttt{\hilight{darkcyan}{//test~return~COUNT}} \\
+	\texttt{} \\
+	\texttt{\hilight{violet}{\#use~<conio>}} \\
+	\texttt{} \\
+	\texttt{\hilight{brown}{typedef}~<?\hilight{brown}{choice}~bits>~bits;} \\
+	\texttt{\hilight{brown}{choice}~bits~\{} \\
+	\texttt{~~<bits>~~~~~~Inc;} \\
+	\texttt{~~<!\hilight{olivegreen}{int};~bits>~Val;} \\
+	\texttt{~~<>~~~~~~~~~~Dealloc;} \\
+	\texttt{\};} \\
+	\texttt{} \\
+	\texttt{bits~\$lower~bit(\hilight{olivegreen}{bool}~b,~bits~\$higher)~\{} \\
+	\texttt{~~\hilight{brown}{switch}~(\$lower)~\{} \\
+	\texttt{~~~~\hilight{brown}{case}~Inc:} \\
+	\texttt{~~~~~~\hilight{brown}{if}~(!b)~\{} \\
+	\texttt{~~~~~~~~\$lower~=~bit(true,~\$higher);} \\
+	\texttt{~~~~~~\}~else~\{} \\
+	\texttt{~~~~~~~~\$higher.Inc;} \\
+	\texttt{~~~~~~~~\$lower~=~bit(false,~\$higher);} \\
+	\texttt{~~~~~~\}~~~} \\
+	\texttt{~~~~\hilight{brown}{case}~Val:} \\
+	\texttt{~~~~~~\$higher.Val;} \\
+	\texttt{~~~~~~\hilight{olivegreen}{int}~n~=~\hilight{blue}{recv}(\$higher);} \\
+	\texttt{~~~~~~\hilight{blue}{send}(\$lower,~2*n+(b?1:0));} \\
+	\texttt{~~~~~~\$lower~=~bit(b,~\$higher);} \\
+	\texttt{~~~~\hilight{brown}{case}~Dealloc:} \\
+	\texttt{~~~~~~\$higher.Dealloc;} \\
+	\texttt{~~~~~~\hilight{blue}{wait}(\$higher);} \\
+	\texttt{~~~~~~\hilight{blue}{close}(\$lower);} \\
+	\texttt{~~\}} \\
+	\texttt{\}} \\
+	\texttt{} \\
+	\texttt{bits~\$lower~eps()~\{} \\
+	\texttt{~~\hilight{brown}{switch}~(\$lower)~\{} \\
+	\texttt{~~~~\hilight{brown}{case}~Inc:} \\
+	\texttt{~~~~~~bits~\$zero~=~eps();} \\
+	\texttt{~~~~~~bits~\$one~=~bit(true,~\$zero);} \\
+	\texttt{~~~~~~\$lower~=~\$one;} \\
+	\texttt{~~~~\hilight{brown}{case}~Val:} \\
+	\texttt{~~~~~~\hilight{blue}{send}(\$lower,~0);~} \\
+	\texttt{~~~~~~\$lower~=~eps();} \\
+	\texttt{~~~~\hilight{brown}{case}~Dealloc:} \\
+	\texttt{~~~~~~\hilight{blue}{close}(\$lower);} \\
+	\texttt{~~\}} \\
+	\texttt{\}} \\
+	\texttt{} \\
+	\texttt{\hilight{olivegreen}{int}~main()~\{} \\
+	\texttt{~~bits~\$c~=~eps();} \\
+	\texttt{~~\hilight{brown}{for}~(\hilight{olivegreen}{int}~i~=~0;~i~<~COUNT;~i++)~\{} \\
+	\texttt{~~~~\$c.Inc;} \\
+	\texttt{~~\}~} \\
+	\texttt{~~\$c.Val;} \\
+	\texttt{~~\hilight{olivegreen}{int}~n~=~\hilight{blue}{recv}(\$c);} \\
+	\texttt{~~\$c.Dealloc;~\hilight{blue}{wait}(\$c);} \\
+	\texttt{~~\hilight{red}{assert}(n~==~COUNT);~} \\
+	\texttt{~~return~n;} \\
+	\texttt{\}} \\
+\end{tabular}
+
 \end{document}

### 2015-06-25 fp/paper.tex
fp: write about test case design

@@
 
+As always, the feasibility of a systematic test depends not just on the set of preemption points, but by the length and branch factor of the input test program (``test case'') used to drive the execution.
+Even though we use Iterative Deepening, a new technique under development for automatically choosing feasible PPs, a test case's stress factor itself needs to be considered so that even a minimal state space can be within range for iterative deepening to expand upon.
+
+{\bf Configuring test parameters.}
+In C0, this is an especially sensitive issue, since many individual program statements involving channels can spawn a new thread, increasing not only the program length (in terms of number of concurrency events) but also the branch factor at the same time.
+We chose the three test programs {\tt odd\_even\_sort4.c1}, {\tt seg.c1}, and {\tt bitstring1.c1}, and configured (by manual trial-and-error and with the help of Landslide's state space estimator) input parameters that would produce appropriately-sized state spaces.
+The test case source code, including the parameters we settled on, are shown in Appendix~\ref{sec:codez}.
+For the case of {\tt bitstring1.c1}, we tested 3 separate configurations, with the values 1, 2, and 4 for {\tt COUNT}.
+We abbreviate these test case names as OES, seg, BS(1), BS(2), and BS(4) respectively..
+
+\newcommand\clis{\texttt{cs()}}
+\newcommand\clir{\texttt{cr()}}
+\newcommand\pros{\texttt{ps()}}
+\newcommand\pror{\texttt{pr()}}
+\newcommand\fwd{\texttt{fwd()}}
+\newcommand\all{\texttt{ALL}}
+{\bf Whitelisting functions.}
+Landslide provides a feature to ``focus'' preemption points on specific sub-modules using the commands {\tt within\_function} and {\tt without\_function} (described in \cite{landslide}).
+We found that, even limiting the stress factor of the above test cases to reasonable ``mid-range'' numbers, the state spaces could be too unwieldy, forcing us to fall back to minimal, often degenerate, parameters (i.e., making a bitstring incremet loop only go around 1 time would miss covering lots of program logic) if we wanted to achieve 100\% completion to verify the absence of bugs.
+However, we found that using Landslide's whitelist/blacklist commands to further limit PPs to subsets of C0's concurrency implementation produced easily manageable state spaces.
+This optimization allows us to more thoroughly verify each function alone, at the expense of more thorough testing of interleavings among 2 or more of each function.
+
+The C0 concurrency runtime features five functions of interest for concurrency testing. These are
+\texttt{client\_send\_msg()},
+\texttt{client\_recv\_msg()},
+\texttt{provider\_send\_msg()},
+\texttt{provider\_recv\_msg()},
+and
+\texttt{forward()}.
+We chose to run our evaluation with six possible test configurations; five where each of these functions is whitelisted alone in turn, which we'll abbreviate respectively as \clis, \clir, \pros, \pror, and \fwd; and one where all functions are whitelisted together, denoted \all.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-evaluating stuff here
+TODO
 
@@
 \section{Conclusion}
+
+TODO
 

### 2015-09-30 landslide-lecture/2015/lecture.tex
Update landslide lecture to have better picture
update lecture again
Tweak phrasing in latest landslide lecture.
Update landslide lecture for fall 2015

@@
 \institute[CMU 15-410]{Carnegie Mellon University - 15-410}
-\date[]{2015, February 11}
+\date[]{2015, September 30}
 
@@
 		\item How it works
-		\item Coping with state space explosion
 		\item Automatically choosing preemption points
@@
 		\item Releasing to paying customers and worrying about correctness later
+			% Say... "and I'm hoping to prevent you from finding yourself in a future where this happens!"
 	\end{itemize}
@@
 		\end{itemize}
-		\item Cryptic panic messages or machine reboots
+		\item Cryptic panic messages when failure occurs
 	\end{itemize}
@@
 \begin{frame}{About The Project}
-	\textbf{About me}: 4th year graduate student, advised by Garth Gibson
+	\textbf{About me}: 5th year graduate student, advised by Garth Gibson
 	\begin{itemize}
@@
 		\item Originally supported only P3s; can now test P2s fully-automated
-		\item {\em Landslide} shows how your {\em Pebbles} programs are not stable.
+		\item {\em Landslide} shows how your {\em Pebbles} programs may not be stable.
 	\end{itemize}
@@
 
-%\begin{frame}{Anatomy}
-%	\begin{center}
-%	\includegraphics[width=0.6\textwidth]{landslide.png}
-%	\end{center}
-%\end{frame}
+\begin{frame}{Landslide \& You}
+	\begin{center}
+	\includegraphics[width=0.85\textwidth]{landslide-new-blank.pdf}
+	\end{center}
+\end{frame}
+
+\begin{frame}{Landslide \& You}
+	\begin{center}
+	\includegraphics[width=0.85\textwidth]{landslide-new.pdf}
+	\end{center}
+\end{frame}
 
@@
 
-\begin{frame}{Preliminary Evaluation}
-	We tested 22 random submitted P2s from Spring 2014.
-	\begin{itemize}
-		\item Test cases:
+\begin{frame}{Test Suite}
+	Landslide ships with 6 test cases:
 	\begin{itemize}
@@
 		\item {\tt broadcast\_test} (new)
-		\end{itemize}
-		\item Each test given CPU budget of 11 cores $\times$ 10 minutes.
+		\item {\tt paradise\_lost} (new this semester)
+		\item {\tt mutex\_test} (new this semester)
 	\end{itemize}
@@
 
-\begin{frame}{Preliminary Results}
-	88 total tests were run (9,680 total CPU-minutes)
+%\begin{frame}{Preliminary Evaluation}
+%	We tested 22 random submitted P2s from Spring 2014.
+%	\begin{itemize}
+%		\item Test cases:
+%		\begin{itemize}
+%			\item {\tt thr\_exit\_join}
+%			\item {\tt paraguay}
+%			\item {\tt rwlock\_downgrade\_read\_test}
+%			\item {\tt broadcast\_test} (new)
+%		\end{itemize}
+%		\item Each test given CPU budget of 11 cores $\times$ 10 minutes.
+%	\end{itemize}
+%\end{frame}
+%
+%\begin{frame}{Preliminary Results}
+%%	88 total tests were run (9,680 total CPU-minutes)
+%%	\linegap
+%%
+%	Found 7 {\bf deterministic} initialization bugs (e.g. use-after-free)
+%	\linegap
+%
+%	56 tests ran to completion without finding bugs:
+%	\begin{itemize}
+%		\item 47 exhausted the CPU budget.
+%		\item 9 completed all state spaces before time ran out.
+%	\end{itemize}
+%	\linegap
+%
+%	Between 20 and 33 {\bf non-deterministic} bugs were found:
+%	\begin{itemize}
+%		\item 20 deadlock races among 12 P2s
+%		\item 10 use-after-free races among 6 P2s
+%		\item 3 misc (assert fail, segfault) among 2 P2s
+%	\end{itemize}
+%\end{frame}
+
+\begin{frame}{Last Semester}
+	Spring 2015: 7 groups signed up to use Landslide; 5 found bugs
+	%\begin{itemize}
+	%	\item (that they didn't find during stress testing)
+	%\end{itemize}
+	%%%% hopefully!
 	\linegap
 
-	Found 7 {\bf deterministic} initialization bugs (e.g. use-after-free)
+	Among all groups, 38 total tests were run
+	\begin{itemize}
+		\item 21 tests ran to completion without finding bugs
+		\item 4 {\em deterministic} bugs
+		\item 13 distinct {\em non-deterministic} bugs
+			\begin{itemize}
+				\item 6 in {\tt thr\_exit\_join} \small (use-after-free, deadlock, NULL, infinite loop...)
+				\item 4 in {\tt paraguay} (deadlock, infinite loop)
+				\item 2 in {\tt broadcast\_test} (deadlock, infinite loop)
+				\item 1 in {\tt rwlock\_downgrade\_read\_test} (use-after-free)
+			\end{itemize}
+			%\begin{itemize}
+			%	\item 6 of which required data-race PPs to expose
+			%\end{itemize}
+	\end{itemize}
+\end{frame}
+
+\begin{frame}{A new treat since last semester}
+	\texttt{mutex\_test} tries to make your mutexes fail or deadlock.
 	\linegap
 
-	56 tests ran to completion without finding bugs:
+	Landslide will look for data races inside your mutex implementation.
 	\begin{itemize}
-		\item 47 exhausted the CPU budget.
-		\item 9 completed all state spaces before time ran out.
+		\item {\tt xchg}/{\tt cmpxchg}/{\tt xadd}
+		\item {\tt mutex->owner = ...}
 	\end{itemize}
@@
 
-	Between 20 and 33 {\bf non-deterministic} bugs were found:
+	Tested 77 P2s from S'14, F'14, S'15
 	\begin{itemize}
-		\item 20 deadlock races among 12 P2s
-		\item 10 use-after-free races among 6 P2s
-		\item 3 misc (assert fail, segfault) among 2 P2s
+		\item 5 unrelated deterministic bugs (use-after-free, crash, etc)
+			% TODO check these for false positives
+			% But this was run after the arbiter fp deadlock avoidance
+			% So I think it's ok
+		\item 1 ``two threads got the lock at once'' race
+			\begin{itemize}
+				\item {\em ...that the TA who graded it didn't find!}
+			\end{itemize}
+		\item 7 deadlock races
+		\item 4 other races (NULL crash, infinite loop, use-after-free)
 	\end{itemize}
@@
 		\begin{itemize}
-			\item Clone a github URL, run setup script, run, answer survey questions
+			\item Clone a github URL, run setup script, run tests %, answer survey questions
 			\item Landslide will automatically report test results
@@
 		\item Meet prerequisites of passing P2 tests
-		\item Email me ({\tt bblum@cs.cmu.edu}) for instructions, survey
+		\item Complete sign-up form online to get further instructions
+			\begin{itemize}
+				\item \url{http://tinyurl.com/landslide-p2-f15}
+			\end{itemize}
+
 		\item Use Landslide in addition to stress tests until P2 is due!
@@
 
-\begin{frame}{More Preliminary Results}
+\begin{frame}{More Evaluation}
 	\textbf{Is dynamically adding ``data race'' PPs effective?}
 	\begin{itemize}
-		\item 3 among reported bugs (9-15\%) required data races to expose.
+		\item 3 among preliminary bugs (9-15\%) required data races to expose.
+		\item 6 among S'15 student bugs (46\%)
 	\end{itemize}

### 2015-09-30 landslide-lecture/2015/lecture.tex
update lecture more
improve lecture more

@@
 \begin{document}
-\renewcommand{\inserttotalframenumber}{36}
+\renewcommand{\inserttotalframenumber}{40}
 \normalem
@@
 	\end{itemize}
+\end{frame}
+
+\begin{frame}{Preemption Point Example (remember this?)}
+	\fbox{
+	\begin{tabular}{l}
+	\texttt{boolean~want[2]~=~\{~false,~false~\};}\\
+	\\
+	\begin{tabular}{rll}
+		1 & \texttt{want[i]~=~true;} & \\
+		  & & \footnotesize (preemption point A) \\
+		2 & \texttt{while~(want[j])} & \\
+		  & & \footnotesize (preemption point B) \\
+		3 & \texttt{~~~~~~~~continue;} & \\
+		  & & \footnotesize (preemption point C) \\
+		4 & \texttt{//~...critical~section...} & \\
+		  & & \footnotesize (preemption point D) \\
+		5 & \texttt{want[i]~=~false;} & \\
+	\end{tabular}
+	\end{tabular}
+	}
 	\linegap
 
-	Is there a {\bf ``sweet spot''}?
+	{\bf Some preemption points will expose bugs. \\
+	Some preemption points don't matter.}
+
+\end{frame}
+
+\begin{frame}{Preemption Point Example (remember this?)}
+	\fbox{
+	\begin{tabular}{l}
+	\texttt{boolean~want[2]~=~\{~false,~false~\};}\\
+	\\
+	\begin{tabular}{rll}
+		1 & \texttt{want[i]~=~true;} & \\
+		  & & \footnotesize (preemption point A) \\
+		2 & \texttt{while~(want[j])} & \\
+		  & & \footnotesize ~ \\
+		3 & \texttt{~~~~~~~~continue;} & \\
+		  & & \footnotesize ~ \\
+		4 & \texttt{//~...critical~section...} & \\
+		  & & \footnotesize ~ \\
+		5 & \texttt{want[i]~=~false;} & \\
+	\end{tabular}
+	\end{tabular}
+	}
+	\linegap
+
+	Here, only preemption point A will trigger a deadlock. {\bf ~}\\
+	All other interleavings are benign.{\bf ~}
+
 \end{frame}
@@
 
-\begin{frame}{Challenges}
-	Parameters of systematic tests must be kept small.
-	\begin{itemize}
-		\item Test program length / number of threads
-		\item Number of preemption points used
-	\end{itemize}
-	\linegap
-
-	\begin{center}
-		\hspace{-0.25in}\includegraphics[width=0.3\textwidth]{expo.pdf} \\
-		\vspace{-1.33in}\hspace{-1.75in}{\rotatebox{90}{\small state space size}} \\
-		\vspace{0.15in}\hspace{-0.25in}{\small number of PPs}
-	\end{center}
-\end{frame}
+%\begin{frame}{Challenges}
+%	Parameters of systematic tests must be kept small.
+%	\begin{itemize}
+%		\item Test program length / number of threads
+%		\item Number of preemption points used
+%	\end{itemize}
+%	\linegap
+%
+%	\begin{center}
+%		\hspace{-0.25in}\includegraphics[width=0.3\textwidth]{expo.pdf} \\
+%		\vspace{-1.33in}\hspace{-1.75in}{\rotatebox{90}{\small state space size}} \\
+%		\vspace{0.15in}\hspace{-0.25in}{\small number of PPs}
+%	\end{center}
+%\end{frame}
 
@@
 		\item TAed 15-410 for 3 semesters during undergrad
-		\item 1st graduate year was 5th year MS, rest in Ph.D. program
+		\item 1st graduate year was 5th year M.S., rest in Ph.D. program
 		\begin{itemize}
@@
 \begin{frame}{Identifying Bugs}
-	Landslide can {\em definitely discover}:
+	Landslide can {\em definitely discover:}
 	\begin{itemize}
@@
 	\linegap
-	Landslide can {\em reasonably suspect}:
+	Landslide can {\em reasonably suspect:}
 	\begin{itemize}
@@
 		\item The threads are not holding the same mutex
-		\item The threads can be reordered (e.g. no \texttt{cond\_signal()} in between)
+		\item The threads can be reordered (e.g., no \texttt{cond\_signal()} in between)
 	\end{itemize}
@@
 	\begin{itemize}
-		\item Students worked best with an iterative process
+		%\item Students worked best with an iterative process
 		\item ``Start small, then add more preemption points as time allows''
@@
 		\item Each job represents a new state space for Landslide to explore
-		\item Prioritizes which jobs are important / likely to finish
-		\begin{itemize}
-			\item Based on nature of PPs (data races? mutexes?)
-			\item Based on estimated completion time
+		\item Prioritizes jobs based on estimated completion time
+		%\item Prioritizes which jobs are important / likely to finish
+		%\begin{itemize}
+		%	\item Based on nature of PPs (data races? mutexes?)
+		%	\item Based on estimated completion time
+		%\end{itemize}
 	\end{itemize}
-		\item Repeat state space explorations, adding preemption points, until time is exhausted.
-	\end{itemize}
+	\linegap
+
+	Repeat state space explorations, adding preemption points, until time is exhausted.
 
@@
 \begin{frame}{Test Suite}
-	Landslide ships with 6 test cases:
+	Landslide ships with 6 approved test cases:
+	\linegap
+
+	Standard P2 tests
 	\begin{itemize}
@@
 		\item {\tt rwlock\_downgrade\_read\_test}
-		\item {\tt broadcast\_test} (new)
+	\end{itemize}
+
+	New tests
+	\begin{itemize}
+		\item {\tt broadcast\_test}
 		\item {\tt paradise\_lost} (new this semester)
@@
 		\item {\tt xchg}/{\tt cmpxchg}/{\tt xadd}
-		\item {\tt mutex->owner = ...}
+		\item {\tt mutex->whose\_turn\_is\_it = ...;}
 	\end{itemize}
+	\pause
 	\linegap
@@
 
-	Watch your email for more details!
+	{\em Really} interested?
+	\begin{itemize}
+		\item I know of an M.S. thesis topic if you want to do post-OS research.
+	\end{itemize}
 \end{frame}
@@
 	\begin{itemize}
-		\item 3 among preliminary bugs (9-15\%) required data races to expose.
+		%\item 3 among preliminary bugs (9-15\%) required data races to expose.
 		\item 6 among S'15 student bugs (46\%)
+		\item 10 among {\tt mutex\_test} bugs (71\%)
 	\end{itemize}

### 2016-02-10 landslide-lecture/2015/lecture.tex
Update lecture for S16

@@
 \institute[CMU 15-410]{Carnegie Mellon University - 15-410}
-\date[]{2015, September 30}
+\date[]{2016, February 10}
 
@@
 \begin{document}
-\renewcommand{\inserttotalframenumber}{40}
+\renewcommand{\inserttotalframenumber}{39}
 \normalem
@@
 				\item May result in data structure corruption
+				\item More on this later...
 			\end{itemize}
@@
 	\begin{itemize}
-		\item Bakery algorithm: Is {\tt number[i]=max(number[0],number[1])} bad?
+		\item Bakery alg: Is {\tt number[i]=max(number[0],number[1])+1} bad?
 		%\item Is unprotected {\tt foobar\_statistics\_counter++} bad?
@@
 		\item e.g., {\tt yield()}, {\tt cond\_wait()}.
+		% Say: When the test invokes yield, the kernel scheduler will pick
+		% one thread to run next. With Landslide, we will pick ALL threads to
+		% run next, in one parallel universe at a time.
 	\end{itemize}
@@
 		\item {\tt broadcast\_test}
-		\item {\tt paradise\_lost} (new this semester)
-		\item {\tt mutex\_test} (new this semester)
+		\item {\tt paradise\_lost} %(new this semester)
+		\item {\tt mutex\_test} %(new this semester)
 	\end{itemize}
@@
 \begin{frame}{Last Semester}
-	Spring 2015: 7 groups signed up to use Landslide; 5 found bugs
+	%Spring 2015: 7 groups signed up to use Landslide; 5 found bugs
+	%%\begin{itemize}
+	%%	\item (that they didn't find during stress testing)
+	%%\end{itemize}
+	%%%%% hopefully!
+	%\linegap
+
+	%Among all groups, 38 total tests were run
 	%\begin{itemize}
-	%	\item (that they didn't find during stress testing)
+	%	\item 21 tests ran to completion without finding bugs
+	%	\item 4 {\em deterministic} bugs
+	%	\item 13 distinct {\em non-deterministic} bugs
+	%		\begin{itemize}
+	%			\item 6 in {\tt thr\_exit\_join} \small (use-after-free, deadlock, NULL, infinite loop...)
+	%			\item 4 in {\tt paraguay} (deadlock, infinite loop)
+	%			\item 2 in {\tt broadcast\_test} (deadlock, infinite loop)
+	%			\item 1 in {\tt rwlock\_downgrade\_read\_test} (use-after-free)
+	%		\end{itemize}
+	%		%\begin{itemize}
+	%		%	\item 6 of which required data-race PPs to expose
+	%		%\end{itemize}
 	%\end{itemize}
+	Fall 2015: 22 groups signed up to use Landslide; 21 found bugs
+	% 18 groups found races
+	\begin{itemize}
+		\item (that they didn't find during stress testing)
+	\end{itemize}
 	%%%% hopefully!
@@
 
-	Among all groups, 38 total tests were run
+	Among all groups, 185 distinct tests were run
 	\begin{itemize}
-		\item 21 tests ran to completion without finding bugs
-		\item 4 {\em deterministic} bugs
-		\item 13 distinct {\em non-deterministic} bugs
+		\item 137 tests ran to completion without finding bugs
+		\item 21 {\em deterministic} bugs
+	\end{itemize}
+	\pause % ask: "who can tell me why landslide might find a *deterministic* bug not found during past testing? A: use-after-free
+	\linegap
+
+	38 distinct {\em non-deterministic} bugs
 	\begin{itemize}
-				\item 6 in {\tt thr\_exit\_join} \small (use-after-free, deadlock, NULL, infinite loop...)
-				\item 4 in {\tt paraguay} (deadlock, infinite loop)
-				\item 2 in {\tt broadcast\_test} (deadlock, infinite loop)
-				\item 1 in {\tt rwlock\_downgrade\_read\_test} (use-after-free)
+			%\begin{itemize}
+			%	\item 7 of which required data-race PPs to expose
+			%\end{itemize}
+		\item 11 groups (50\%) fixed all their bugs
+			\begin{itemize}
+				\item (as verified by running Landslide again -- not a guarantee!)
 			\end{itemize}
-			%\begin{itemize}
-			%	\item 6 of which required data-race PPs to expose
-			%\end{itemize}
+		\item 2.7 bugs ``fixed'' (see above) on average per group
+		\item Most ambitious group: 11 distinct races found + fixed!
 	\end{itemize}
@@
 
-\begin{frame}{A new treat since last semester}
-	\texttt{mutex\_test} tries to make your mutexes fail or deadlock.
-	\linegap
-
-	Landslide will look for data races inside your mutex implementation.
-	\begin{itemize}
-		\item {\tt xchg}/{\tt cmpxchg}/{\tt xadd}
-		\item {\tt mutex->whose\_turn\_is\_it = ...;}
-	\end{itemize}
-	\pause
-	\linegap
-
-	Tested 77 P2s from S'14, F'14, S'15
-	\begin{itemize}
-		\item 5 unrelated deterministic bugs (use-after-free, crash, etc)
-			% TODO check these for false positives
-			% But this was run after the arbiter fp deadlock avoidance
-			% So I think it's ok
-		\item 1 ``two threads got the lock at once'' race
-			\begin{itemize}
-				\item {\em ...that the TA who graded it didn't find!}
-			\end{itemize}
-		\item 7 deadlock races
-		\item 4 other races (NULL crash, infinite loop, use-after-free)
-	\end{itemize}
-\end{frame}
+%\begin{frame}{A new treat} %since last semester}
+%	\texttt{mutex\_test} tries to make your mutexes fail or deadlock.
+%	\linegap
+%
+%	Landslide will look for data races inside your mutex implementation.
+%	\begin{itemize}
+%		\item {\tt xchg}/{\tt cmpxchg}/{\tt xadd}
+%		\item {\tt mutex->whose\_turn\_is\_it = ...;}
+%	\end{itemize}
+%	\pause
+%	\linegap
+%	
+%	Last semester: 12 non-determinstic bugs among 8 groups
+%	\linegap
+%
+%	%Tested 77 P2s from S'14, F'14, S'15
+%	%\begin{itemize}
+%	%	\item 5 unrelated deterministic bugs (use-after-free, crash, etc)
+%	%		% TODO check these for false positives
+%	%		% But this was run after the arbiter fp deadlock avoidance
+%	%		% So I think it's ok
+%	%	\item 7 deadlock races
+%	%	\item 4 other races (NULL crash, infinite loop, use-after-free)
+%	%	\item 1 ``two threads got the lock at once'' race
+%	%		\begin{itemize}
+%	%			\item {\em ...that the TA who graded it didn't find!}
+%	%		\end{itemize}
+%	%\end{itemize}
+%\end{frame}
 
@@
 			\begin{itemize}
-				\item \url{http://tinyurl.com/landslide-p2-f15}
+				\item \url{http://tinyurl.com/landslide-p2-s16}
 			\end{itemize}
@@
 
-\subsection{Experimental Results}
+%\subsection{Experimental Results}
+%
+%\begin{frame}{More Evaluation}
+%	\textbf{Is dynamically adding ``data race'' PPs effective?}
+%	\begin{itemize}
+%		%\item 3 among preliminary bugs (9-15\%) required data races to expose.
+%		\item 6 among S'15 student bugs (46\%)
+%		\item 10 among {\tt mutex\_test} bugs (71\%)
+%	\end{itemize}
+%	\linegap
+%
+%	{\bf Is ``iterative deepening'' better than state-of-the-art?}
+%	\begin{itemize}
+%		\item Control experiment: Just 1 state space, same CPU time
+%		\begin{itemize}
+%			\item PPs used: \texttt{mutex\_lock()}, \texttt{mutex\_unlock()}
+%			\item 110 minutes on 1 CPU
+%		\end{itemize}
+%		\item Of 33 total bug reports, control failed to find 10 (30\%).
+%		\begin{itemize}
+%			\item 3 required data race PPs to expose
+%			\item 1 ran out of time
+%			\item 6 obscured by different bug in same state space
+%		\end{itemize}
+%	\end{itemize}
+%\end{frame}
 
-\begin{frame}{More Evaluation}
-	\textbf{Is dynamically adding ``data race'' PPs effective?}
-	\begin{itemize}
-		%\item 3 among preliminary bugs (9-15\%) required data races to expose.
-		\item 6 among S'15 student bugs (46\%)
-		\item 10 among {\tt mutex\_test} bugs (71\%)
-	\end{itemize}
+\begin{frame}{Last-last Semester}
+	Spring 2015: 7 groups signed up to use Landslide; 5 found bugs
+	%\begin{itemize}
+	%	\item (that they didn't find during stress testing)
+	%\end{itemize}
+	%%%% hopefully!
 	\linegap
 
-	{\bf Is ``iterative deepening'' better than state-of-the-art?}
+	Among all groups, 38 total tests were run
 	\begin{itemize}
-		\item Control experiment: Just 1 state space, same CPU time
+		\item 21 tests ran to completion without finding bugs
+		\item 4 {\em deterministic} bugs
+		\item 13 distinct {\em non-deterministic} bugs
 			\begin{itemize}
-			\item PPs used: \texttt{mutex\_lock()}, \texttt{mutex\_unlock()}
-			\item 110 minutes on 1 CPU
+				\item 6 in {\tt thr\_exit\_join} \small (use-after-free, deadlock, NULL, infinite loop...)
+				\item 4 in {\tt paraguay} (deadlock, infinite loop)
+				\item 2 in {\tt broadcast\_test} (deadlock, infinite loop)
+				\item 1 in {\tt rwlock\_downgrade\_read\_test} (use-after-free)
 			\end{itemize}
-		\item Of 33 total bug reports, control failed to find 10 (30\%).
-		\begin{itemize}
-			\item 3 required data race PPs to expose
-			\item 1 ran out of time
-			\item 6 obscured by different bug in same state space
-		\end{itemize}
+			%\begin{itemize}
+			%	\item 6 of which required data-race PPs to expose
+			%\end{itemize}
 	\end{itemize}

### 2016-02-10 landslide-lecture/2015/lecture.tex
some more changes s16
add bullet landslide clinic

@@
 		\item How did you know to draw that tree's branches where they matter?
-	\end{itemize}
-\end{frame}
-
-\begin{frame}{Preemption Points}
-	\textbf{Preemption points} (PPs) are code locations where being preempted may cause different behaviour.
-	\begin{itemize}
-		\item IOW, somewhere that interesting interleavings can happen around.
-	\end{itemize}
-	\linegap
-
-	Systematic tests are {\em parameterized} by the set of PPs.
-	\begin{itemize}
-		\item If there are $n$ PPs and $k$ threads, state space size is $n^k$.
-		\item Need to choose the set of PPs very carefully for test to be effective.
-		\begin{itemize}
-			\item ``Effective'' = both comprehensive and feasible.
-		\end{itemize}
-	\end{itemize}
-\end{frame}
-
-\begin{frame}{Preemption Points}
-	What does ``all possible interleavings'' actually mean?
-
-	\linegap
-	{\bf One extreme}: Preempt at {\em every instruction}
-	\begin{itemize}
-		\item Good news: Will find every possible race condition.
-		\item Bad news: Runtime of test will be impossibly large.
-	\end{itemize}
-	\linegap
-
-	{\bf Other extreme}: {\em Nothing} is a preemption point
-	\begin{itemize}
-		\item Good news: Test will finish quickly.
-		\item Bad news: Only one execution was checked for bugginess.
-		\begin{itemize}
-			\item No alternative interleavings explored.
-			\item Makes ``no race found'' a weak claim.
-		\end{itemize}
 	\end{itemize}
@@
 
+\end{frame}
+
+\begin{frame}{Preemption Points}
+	\textbf{Preemption points} (PPs) are code locations where being preempted may cause different behaviour.
+	\begin{itemize}
+		\item IOW, somewhere that interesting interleavings can happen around.
+	\end{itemize}
+	\linegap
+
+	Systematic tests are {\em parameterized} by the set of PPs.
+	\begin{itemize}
+		\item If there are $n$ PPs and $k$ threads, state space size is $n^k$.
+		\item Need to choose the set of PPs very carefully for test to be effective.
+		\begin{itemize}
+			\item ``Effective'' = both comprehensive and feasible.
+		\end{itemize}
+	\end{itemize}
+\end{frame}
+
+\begin{frame}{Preemption Points}
+	What does ``all possible interleavings'' actually mean?
+
+	\linegap
+	{\bf One extreme}: Preempt at {\em every instruction}
+	\begin{itemize}
+		\item Good news: Will find every possible race condition.
+		\item Bad news: Runtime of test will be impossibly large.
+	\end{itemize}
+	\linegap
+
+	{\bf Other extreme}: {\em Nothing} is a preemption point
+	\begin{itemize}
+		\item Good news: Test will finish quickly.
+		\item Bad news: Only one execution was checked for bugginess.
+		\begin{itemize}
+			\item No alternative interleavings explored.
+			\item Makes ``no race found'' a weak claim.
+		\end{itemize}
+	\end{itemize}
 \end{frame}
@@
 	\begin{itemize}
-		\item At end of each branch, identify a PP to replay differently
+		\item Each time test completes, identify a PP to replay differently
 		\item Reset machine state and start over
@@
 			\item Clone a github URL, run setup script, run tests %, answer survey questions
-			\item Landslide will automatically report test results
+			\item Landslide will automatically report test results (as described below)
 		\end{itemize}
@@
 			\end{itemize}
-
+		\item Optional ``Landslide clinic'' for in-person tech (or moral) support
+			\begin{itemize}
+				\item Mid-next-week, room and time TBD
+			\end{itemize}
 		\item Use Landslide in addition to stress tests until P2 is due!

### 2016-03-01 oopsla/calcm.tex
Add beginnings of CALCM seminar talk

--- 2015-09-30 landslide-lecture/2015/lecture.tex
@@
 
-\title[Landslide]{{\bf Landslide: A New Race-Finding Tool for 15-410}}
-\subtitle[]{ {\em more clever than } ``\texttt{mandelbrot}'' {\em since 2011.}}
+\title[Landslide]{{\bf Stateless Model Checking \\ with Data-Race Preemption Points}}
+%\subtitle[]{CALCM group seminar}
 \author[Ben Blum]{Ben Blum \texttt{(bblum@andrew.cmu.edu)}}
 
-\institute[CMU 15-410]{Carnegie Mellon University - 15-410}
-\date[]{2015, September 30}
+\institute[]{Carnegie Mellon University - CALCM group seminar}
+\date[]{2016, March 01}
 
@@
 
-\section{Introduction}
-\subsection{Introduction}
+\begin{frame}{}
+	Working on OOPSLA submission (March 23)
+	\linegap
 
-\begin{frame}{Outline}
-	\textbf{Theory: Seeing race conditions in a new way}
+	Feedback wanted on:
 	\begin{itemize}
-		\item Case study (example)
-		\item Tabular execution traces
-		\item The execution tree
-	\end{itemize}
-	{\bf Research Technique: ``Systematic testing''}
-	% Say: A way of enumerating all possible interleavings of a concurrent test case.
-	\begin{itemize}
-		\item Preemption points
-		\item Challenges and feasibility
-	\end{itemize}
-	{\bf Tool: Landslide}
-	\begin{itemize}
-		\item How it works
-		\item Automatically choosing preemption points
-		\item User study (that's you!)
-			% !!! SAY !!! : If you are struggling, this is not for you, etc.
+		\item What do OOPSLA reviewers look for?
+		\item Collaboration and/or ideas for future work
 	\end{itemize}
@@
 
-\subsection{Race Conditions}
+\begin{frame}{Outline}
+	\textbf{Background}
+	\begin{itemize}
+		\item Preemption points
+		\item Stateless model checking (aka, systematic testing)
+		\item Data races
+	\end{itemize}
+	{\bf Quicksand \& Iterative Deepening}
+	\begin{itemize}
+		\item State space estimation
+		\item Data-race preemption points
+		\item Implementation with Landslide
+	\end{itemize}
+	{\bf Evaluation}
+	\begin{itemize}
+		\item Comparison to ``single-state-space'' approach
+		\item Nondeterministic data-race candidates
+		\item Providing total verification
+	\end{itemize}
+\end{frame}
 
+\section{Introduction}
+
+\subsection{Stateless Model Checking}
 \definecolor{thread1}{RGB}{87,172,255}
@@
 		\begin{itemize}
-			\item See {\bf Paradise Lost} lecture!
-			\item {\tt if} vs {\tt while}: Two consumers can race to make one fail.
+			%\item See {\bf Paradise Lost} lecture!
+			\item {\tt if} vs {\tt while}: Two consumers can interleave to make one fail.
 		\end{itemize}
@@
 
-\begin{frame}{Testing}
-	How can programmers be confident in the correctness of their code?
+\newcommand\related[1]{\textsuperscript{\em [#1]}}
+
+\begin{frame}{Motivating Stateless Model Checking}
+	Goal: Test as many interleavings as you can (given some CPU budget)
+	\linegap
+
+	\textbf{Stress testing}
 	\begin{itemize}
-		\item Unit tests
-		\begin{itemize}
-			\item good for basic functionality, bad for concurrency
-		\end{itemize}
-		\item Stress tests
-		\begin{itemize}
-			\item state of the art in 15-410
-		\end{itemize}
-		\item Theorem proving
-		\begin{itemize}
-			\item heavy burden on the programmers
-		\end{itemize}
-		\item Releasing to paying customers and worrying about correctness later
-			% Say... "and I'm hoping to prevent you from finding yourself in a future where this happens!"
+		\item Relies on randomly-occurring timer (or I/O) interrupts
+			% to perturb threads
+		\item Exposes nondeterministic bugs at random
+		\item Might not find bugs in any bounded time
 	\end{itemize}
@@
 
-	{\bf Motivation}: Can we do better than stress testing?
+	\textbf{Stateless Model Checking} (MC)\related{Godefroid '97}
+	\begin{itemize}
+		\item Exhaustively exercise {\em all} thread interleavings
+		\begin{itemize}
+			\item ...around a given set of {\em preemption points}
+		\end{itemize}
+		\item Identify bugs for certain (e.g. segfault, deadlock, panic)
+		\item More reliable coverage, better reproducibility
+		\item Exponentially-sized state space of interleavings
+			%depending on \# of preemption points
+	\end{itemize}
+	%Moving beyond stress testing: which {\em preemption points} are important?
 \end{frame}
 
-\begin{frame}{Testing Mechanisms}
-	\textbf{Stress testing}: \texttt{largetest}, \texttt{mandelbrot} and friends
-	\begin{itemize}
-		\item Attempting to exercise as many interleavings as practical
-		\item Exposes race conditions at random
-		\begin{itemize}
-			\item ``If a preemption occurs at just the right time\ldots''
-		\end{itemize}
-		\item Cryptic panic messages when failure occurs
-	\end{itemize}
+\begin{frame}{Preemption Points}
+	\textbf{Preemption points} (PPs) are program locations where a context switch might expose different behaviour.
 	\linegap
-	What if\ldots
-	\begin{itemize}
-		\item Make educated guesses about when to preempt
-		\item Preempt enough times to run {\em every single} interleaving
-		\item Tell the story of what {\em actually happened}.
-		\item Overlook fewer bugs!
-	\end{itemize}
-	% Say: At the end of this lecture, I'm going to be looking for some among you...
-\end{frame}
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+	Stateless MC is {\em parameterized} by which PPs can trigger thread switches.
+	\linegap
 
-\section{Systematic Testing}
-
-\breakslide{\Large A different way of looking at race conditions\ldots}
-
-\begin{frame}{Execution Tree}
-		\includegraphics[width=0.78\textwidth]{execution-tree-0.pdf} \\
-		\vspace{-3in}
-		\hspace{3in}
-		\includegraphics[width=0.34\textwidth]{table-0.png}
-		\\
-		\vspace{1.2in}
-		\hspace{3.7in}
-		\includegraphics[width=0.15\textwidth]{frownie-blank.png}
-\end{frame}
-\begin{frame}{Execution Tree}
-		\includegraphics[width=0.78\textwidth]{execution-tree-1.pdf} \\
-		\vspace{-3in}
-		\hspace{3in}
-		\includegraphics[width=0.34\textwidth]{table-0.png}
-		\\
-		\vspace{1.2in}
-		\hspace{3.7in}
-		\includegraphics[width=0.15\textwidth]{frownie-blank.png}
-\end{frame}
-\begin{frame}{Execution Tree}
-		\includegraphics[width=0.78\textwidth]{execution-tree-2.pdf} \\
-		\vspace{-3in}
-		\hspace{3in}
-		\includegraphics[width=0.34\textwidth]{table-1.png}
-		\\
-		\vspace{1.2in}
-		\hspace{3.7in}
-		\includegraphics[width=0.15\textwidth]{frownie-blank.png}
-\end{frame}
-\begin{frame}{Execution Tree}
-		\includegraphics[width=0.78\textwidth]{execution-tree-full.pdf} \\
-		\vspace{-3in}
-		\hspace{3in}
-		\includegraphics[width=0.34\textwidth]{table-2.png}
-		\\
-		\vspace{1.2in}
-		\hspace{3.7in}
-		\includegraphics[width=0.15\textwidth]{frownie.png}
-\end{frame}
-
-% Say: Obviously we want to be able to always run the third path. But knowing
-% that ahead of time is impossible.
-
-% Say: In the rest of the lecture I'm going to refer to the "state space of
-% possible interleavings", simply referring to this tree, and "size" of the state
-% space refers to the number of branches.
-
-\begin{frame}{Systematic Testing - The Big Picture}
-	\textbf{Goal: Force the system to execute every possible interleaving.}
-	% Say: checking for crashes/memory errors/etc in each one.
-	\begin{itemize}
-		\item On 1st execution, schedule threads arbitrarily until program ends.
-			\begin{itemize}
-				\item This represents one branch of the tree.
-			\end{itemize}
-		\item At end of each branch, rewind system and restart test.
-		\item Artificially add preemptions to produce different thread interleavings.
-		\item Intuitively: Generate many ``tabular execution traces''.
-	\end{itemize}
+	For case study, {\tt mutex\_lock}, {\tt mutex\_unlock}, {\tt cond\_wait} are relevant.
 	\pause
@@
 
-	Okay, wait a sec...
+	Which PPs to use?
 	\begin{itemize}
-		\item How can you possibly execute {\em every possible} interleaving?
-		%\item How can you know where preempting matters most?
-		\item How did you know to draw that tree's branches where they matter?
+		\item Every instruction (exhaustive, but intractable)
+		\item Synchronization API calls\related{Musuvathi '08, Simsa '10, Blum '12}
+		%\item Kernel mutex interface\related{Blum '12}
+		\item Global/heap memory accesses\related{Holzmann '97, Yang '08}
 	\end{itemize}
@@
 
-\begin{frame}{Preemption Points}
-	\textbf{Preemption points} (PPs) are code locations where being preempted may cause different behaviour.
+\begin{frame}{MC State of the Art}
+	Modern stateless MC tools can {\em mitigate} exponential explosion.
 	\begin{itemize}
-		\item IOW, somewhere that interesting interleavings can happen around.
+		\item Dynamic Partial Order Reduction\related{Flanagan '05}
+		\item State space estimation\related{Simsa '12}
 	\end{itemize}
@@
 
-	Systematic tests are {\em parameterized} by the set of PPs.
+	However, existing MCs ``hard-code'' their PPs in advance.
 	\begin{itemize}
-		\item If there are $n$ PPs and $k$ threads, state space size is $n^k$.
-		\item Need to choose the set of PPs very carefully for test to be effective.
-		\begin{itemize}
-			\item ``Effective'' = both comprehensive and feasible.
-		\end{itemize}
-	\end{itemize}
-\end{frame}
-
-\begin{frame}{Preemption Points}
-	What does ``all possible interleavings'' actually mean?
-
-	\linegap
-	{\bf One extreme}: Preempt at {\em every instruction}
-	\begin{itemize}
-		\item Good news: Will find every possible race condition.
-		\item Bad news: Runtime of test will be impossibly large.
+		\item With too many preemption points, tests time out.
+		\item Which PPs are ``good'' differs for each test case.
+		\item Relationship between PPs and testing time is unpredictable.
 	\end{itemize}
@@
 
-	{\bf Other extreme}: {\em Nothing} is a preemption point
+	Current MC model is not user-friendly.
 	\begin{itemize}
-		\item Good news: Test will finish quickly.
-		\item Bad news: Only one execution was checked for bugginess.
-		\begin{itemize}
-			\item No alternative interleavings explored.
-			\item Makes ``no race found'' a weak claim.
-		\end{itemize}
-	\end{itemize}
-\end{frame}
-
-\begin{frame}{Preemption Point Example (remember this?)}
-	\fbox{
-	\begin{tabular}{l}
-	\texttt{boolean~want[2]~=~\{~false,~false~\};}\\
-	\\
-	\begin{tabular}{rll}
-		1 & \texttt{want[i]~=~true;} & \\
-		  & & \footnotesize (preemption point A) \\
-		2 & \texttt{while~(want[j])} & \\
-		  & & \footnotesize (preemption point B) \\
-		3 & \texttt{~~~~~~~~continue;} & \\
-		  & & \footnotesize (preemption point C) \\
-		4 & \texttt{//~...critical~section...} & \\
-		  & & \footnotesize (preemption point D) \\
-		5 & \texttt{want[i]~=~false;} & \\
-	\end{tabular}
-	\end{tabular}
-	}
-	\linegap
-
-	{\bf Some preemption points will expose bugs. \\
-	Some preemption points don't matter.}
-
-\end{frame}
-
-\begin{frame}{Preemption Point Example (remember this?)}
-	\fbox{
-	\begin{tabular}{l}
-	\texttt{boolean~want[2]~=~\{~false,~false~\};}\\
-	\\
-	\begin{tabular}{rll}
-		1 & \texttt{want[i]~=~true;} & \\
-		  & & \footnotesize (preemption point A) \\
-		2 & \texttt{while~(want[j])} & \\
-		  & & \footnotesize ~ \\
-		3 & \texttt{~~~~~~~~continue;} & \\
-		  & & \footnotesize ~ \\
-		4 & \texttt{//~...critical~section...} & \\
-		  & & \footnotesize ~ \\
-		5 & \texttt{want[i]~=~false;} & \\
-	\end{tabular}
-	\end{tabular}
-	}
-	\linegap
-
-	Here, only preemption point A will trigger a deadlock. {\bf ~}\\
-	All other interleavings are benign.{\bf ~}
-
-\end{frame}
-
-\begin{frame}{Preemption Points}
-	\textbf{Sweet spot}: Insert a thread switch everywhere it ``might matter''.
-
-	\linegap
-	When do we fear being preempted?
-	\begin{itemize}
-		\item Threads becoming runnable (\texttt{thr\_create()}, \texttt{cond\_signal()}, etc.)
-			\begin{itemize}
-				\item Preemptions may cause it to run before we're ready
-			\end{itemize}
-		\item Synchronization primitives (\texttt{mutex\_lock()}/\texttt{unlock()}, etc.)
-			\begin{itemize}
-				\item If buggy or used improperly\ldots
-			\end{itemize}
-		\item Unprotected shared memory accesses (``data races'')% Say: more on this later.
-			\begin{itemize}
-				\item May result in data structure corruption
-			\end{itemize}
-	\end{itemize}
-
-	%%%% Not anymore! :D
-	%Finding the sweet spot is a joint effort between programmer and tool. (More on this later.)
-	%However, all these points together may still be too expensive...
-\end{frame}
-
-%\begin{frame}{Challenges}
-%	Parameters of systematic tests must be kept small.
-%	\begin{itemize}
-%		\item Test program length / number of threads
-%		\item Number of preemption points used
-%	\end{itemize}
-%	\linegap
-%
-%	\begin{center}
-%		\hspace{-0.25in}\includegraphics[width=0.3\textwidth]{expo.pdf} \\
-%		\vspace{-1.33in}\hspace{-1.75in}{\rotatebox{90}{\small state space size}} \\
-%		\vspace{0.15in}\hspace{-0.25in}{\small number of PPs}
-%	\end{center}
-%\end{frame}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{Landslide}
-
-\breakslide{\Huge Landslide}
-
-\begin{frame}{About The Project}
-	\textbf{About me}: 5th year graduate student, advised by Garth Gibson
-	\begin{itemize}
-		\item TAed 15-410 for 3 semesters during undergrad
-		\item 1st graduate year was 5th year M.S., rest in Ph.D. program
-		\begin{itemize}
-			\item \url{http://www.contrib.andrew.cmu.edu/~bblum/thesis.pdf}
-		\end{itemize}
-	\end{itemize}
-	\pause
-	\linegap
-
-	{\bf About Landslide}
-	\begin{itemize}
-		\item Simics module, which traces:
-			\begin{itemize}
-				\item Every instruction executed
-				\item Every memory access read/written
-			\end{itemize}
-			% FIXME for next year -- how much slowdown is there? 20x?
-		\item Originally supported only P3s; can now test P2s fully-automated
-		\item {\em Landslide} shows how your {\em Pebbles} programs may not be stable.
+		\item Tool: ``I want to use these PPs, but can't predict completion time.''
+		\item User: ``I have 16 CPUs and 24 hours to test my program.''
 	\end{itemize}
@@
 
-\subsection{Technical Overview}
+\subsection{Data Races}
 
-\begin{frame}{Big Picture: Execution Tree Exploration}
-	\textbf{Backtracking}
+\begin{frame}{Data Races\related{Savage '97}}
+	A {\bf data race} is a pair of memory accesses between two threads, where:
 	\begin{itemize}
-		\item At end of each branch, identify a PP to replay differently
-		\item Reset machine state and start over
-		\item Implemented using Simics bookmarks
+		\item At least one of the accesses is a write
+		\item The threads are not holding the same mutex
+		%\item The threads can be reordered (e.g., no \texttt{cond\_signal()} in between)
+		\item The threads' executions are not related by {\em Happens-Before}
+	\end{itemize}
+
+	Data races are {\em not necessarily} bugs, just highly suspicious!
 	\begin{itemize}
-				\item {\tt set-bookmark} and {\tt skip-to}
+		\item ``If threads interleaved the wrong way here, it {\em might} crash later.''
+		\item Hence, data race locations are ideal PP candidates.
 	\end{itemize}
-		\item Replay test from the beginning, with a different interleaving
+\end{frame}
+
+\begin{frame}{The Happens-Before (HB) Relation}
+	Most data-race analyses use {\em Pure HB} to determine simultaneity.
+	\begin{itemize}
+		\item HB is established by any synchronization (mutexes, cvars, xchg...)
+		\item Represents possiblity to threads' behaviour to change if reordered
+			%\begin{itemize}
+			%	\item ...that the programmers obviously intended.
+			%\end{itemize}
+		\item May exclude some races from alternate interleavings ({\em false negatives})
 	\end{itemize}
@@
 
-	{\bf Controlling scheduling decisions}
+	To avoid false negatives, we use {\em Limited HB}\related{O'Callahan '03} instead.
+	%, Serebryany '09}
 	\begin{itemize}
-		\item Tool must control all sources of nondeterminism
-		\item In 15-410, just timer and keyboard interrupts
-			% optional - say: "in real kernels, disk/network io interrupts"
-		\item Landslide repeatedly fires timer ticks until desired thread is run.
+		\item HB is established only when thread runnability changes (cvars).
+		\item Represents possibility to reorder accesses to occur simultaneously
+		\item Captures accesses separated by unrelated, synchronized, logic
+		\item Will also falsely flag some {\em false positives}
+		\item Combined with lock-set analysis to increase accuracy
 	\end{itemize}
@@
 
-\begin{frame}{Landslide \& You}
+\begin{frame}{Happens-Before Example}
 	\begin{center}
-	\includegraphics[width=0.85\textwidth]{landslide-new-blank.pdf}
+		\begin{tabular}{|l|l|}
+			\hline
+			\cellcolor{thread1} {\bf Thread 1} & \cellcolor{thread2} {\bf Thread 2} \\
+			\hline
+			& \\
+			\small \texttt{x++;} // A & \\
+			\small \texttt{lock(mx);} & \\
+			%\small \texttt{y = true;} & \\
+			\small \texttt{unlock(mx);} & \\
+
+			& \\
+			& \small \texttt{lock(mx);} \\
+			& \small \texttt{unlock(mx);} \\
+			& \small \texttt{x++;} // B \\
+			& \\
+			\hline
+		\end{tabular}
 	\end{center}
+	Pure HB: \texttt{mx} establishes $HB(A, B)$; no data race reported (false negative).
+	\linegap
+
+	Limited HB: $\neg HB(A, B)$; data race is reported (correctly).
 \end{frame}
 
-\begin{frame}{Landslide \& You}
+\begin{frame}{Happens-Before Example}
 	\begin{center}
-	\includegraphics[width=0.85\textwidth]{landslide-new.pdf}
+		\begin{tabular}{|l|l|}
+			\hline
+			\cellcolor{thread1} {\bf Thread 1} & \cellcolor{thread2} {\bf Thread 2} \\
+			\hline
+			\small \texttt{x++;} // A & \\
+			\small \texttt{lock(mx);} & \\
+			\small \texttt{y = true;} & \\
+			\small \texttt{unlock(mx);} & \\
+
+			& \small \texttt{lock(mx);} \\
+			& \small \texttt{bool z = y;} \\
+			& \small \texttt{unlock(mx);} \\
+			& \small \texttt{if (z);} \\
+			& \small \texttt{~~x++;} // B \\
+			\hline
+		\end{tabular}
 	\end{center}
-\end{frame}
-
-\begin{frame}{Identifying Bugs}
-	Landslide can {\em definitely discover:}
-	\begin{itemize}
-		\item Assertion failures % Say: note: the better your asserts are..!
-		\item Segfaults
-		\item Deadlock
-		\item Use-after-free / double-free
-	\end{itemize}
-	\linegap
-	Landslide can {\em reasonably suspect:}
-	\begin{itemize}
-		\item Infinite loop (halting problem)
-		\item Data race bugs
-	\end{itemize}
-\end{frame}
-
-\begin{frame}{What is a Data Race?}
-	A {\bf data race} is a pair of memory accesses between two threads, where:
-	\begin{itemize}
-		\item At least one of the accesses is a write
-		\item The threads are not holding the same mutex
-		\item The threads can be reordered (e.g., no \texttt{cond\_signal()} in between)
-	\end{itemize}
-	\pause
+	Pure HB: \texttt{mx} establishes $HB(A, B)$; no data race reported (correctly).
 	\linegap
 
-	Data races are {\em not necessarily} bugs, just highly suspicious!
-	\begin{itemize}
-		\item Bakery algorithm: Is {\tt number[i]=max(number[0],number[1])} bad?
-		%\item Is unprotected {\tt foobar\_statistics\_counter++} bad?
-		\item What about unprotected {\tt next\_thread\_id++}?
-		\item ``If threads interleaved the wrong way here, it {\em might} crash later.''
-		\begin{itemize}
-			\item Hmmm...
-		\end{itemize}
-	\end{itemize}
+	Limited HB: $\neg HB(A, B)$; data race is reported (false positive).
 \end{frame}
@@
 
-\subsection{Iterative Deepening}
-
-\begin{frame}{Choosing the Right Preemption Points}
-	How can we address exponential state space explosion?
-	\pause
-	\linegap
-
-	State of the art tools hard-code a fixed set of preemption points.
-	\begin{itemize}
-		\item E.g., ``all thread library API calls'' or ``all kernel mutex locks/unlocks''
-		\item Depending on length of test, completion time is unpredictable.
-		\item More often, a subset is better in terms of time/coverage. 
-	\end{itemize}
-	\pause
-	\linegap
-	Current systematic testing model is not user-friendly.
-	\begin{itemize}
-		\item Tool: ``I want to use these PPs, but can't predict completion time.''
-		\item User: ``I have 16 CPUs and 24 hours to test my program.''
-	\end{itemize}
-	\linegap
-	{\em Stress testing allows user to choose total run time -- can we offer this too?}
-\end{frame}
+\breakslide{\Huge Quicksand}
 
@@
 	Goal: Run the best tests for a given CPU budget.
-	\begin{itemize}
-		\item Technique: ``Iterative Deepening''
-	\end{itemize}
 	\linegap
 
-	Based on experience from past 15-410 student volunteers
+	Inspired by past experience with Landslide, our MC tool\related{Blum '12}
 	\begin{itemize}
-		%\item Students worked best with an iterative process
+		\item 15-410 student volunteers worked best with iterative process
 		\item ``Start small, then add more preemption points as time allows''
-		\item Landslide now automates this process
+		\item Quicksand now automates this via {\em Iterative Deepening}
 	\end{itemize}
@@
 
-\begin{frame}{Iterative Deepening in Landslide}
-	Landslide automatically iterates through different configurations of PPs.
+\begin{frame}{Iterative Deepening in Quicksand}
+	Quicksand automatically iterates through different configurations of PPs.
 	\begin{itemize}
@@
 		%\end{itemize}
+		\item New PPs added on-the-fly from data-race analysis
 	\end{itemize}
@@
 \begin{frame}{Iterative Deepening}
-	Adding different PPs can produce state spaces of different sizes; Landslide tries them in parallel.
+	Adding different PPs can produce state spaces of different sizes; Quicksand tries them in parallel, as informed by state space estimation\related{Simsa '12}.
 	\vspace{0.15in}
@@
 \begin{frame}{Iterative Deepening}
-	If time allows, Landslide will combine PPs into larger, more comprehensive state spaces.
+	If time allows, combine PPs into larger, more comprehensive state spaces.
 	\begin{center}
@@
 
+\begin{frame}{Total Verification}
+	% Let's come back to this issue of ...
+	MC can {\em totally verify} a test by testing a state space with all possible PPs.
+	\begin{itemize}
+		\item 
+	\end{itemize}
+	%One downside of hard-coding PPs in advance is timing out when the test is too large, but another downside is if the test is small enough, you might miss a possible full verification!
+\end{frame}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-\breakslide{\Huge Demo}
-
-\begin{frame}{Test Suite}
-	Landslide ships with 6 approved test cases:
-	\linegap
-
-	Standard P2 tests
-	\begin{itemize}
-		\item {\tt thr\_exit\_join}
-		\item {\tt paraguay}
-		\item {\tt rwlock\_downgrade\_read\_test}
-	\end{itemize}
-
-	New tests
-	\begin{itemize}
-		\item {\tt broadcast\_test}
-		\item {\tt paradise\_lost} (new this semester)
-		\item {\tt mutex\_test} (new this semester)
-	\end{itemize}
-\end{frame}
-
-%\begin{frame}{Preliminary Evaluation}
-%	We tested 22 random submitted P2s from Spring 2014.
-%	\begin{itemize}
-%		\item Test cases:
-%		\begin{itemize}
-%			\item {\tt thr\_exit\_join}
-%			\item {\tt paraguay}
-%			\item {\tt rwlock\_downgrade\_read\_test}
-%			\item {\tt broadcast\_test} (new)
-%		\end{itemize}
-%		\item Each test given CPU budget of 11 cores $\times$ 10 minutes.
-%	\end{itemize}
-%\end{frame}
-%
-%\begin{frame}{Preliminary Results}
-%%	88 total tests were run (9,680 total CPU-minutes)
-%%	\linegap
-%%
-%	Found 7 {\bf deterministic} initialization bugs (e.g. use-after-free)
-%	\linegap
-%
-%	56 tests ran to completion without finding bugs:
-%	\begin{itemize}
-%		\item 47 exhausted the CPU budget.
-%		\item 9 completed all state spaces before time ran out.
-%	\end{itemize}
-%	\linegap
-%
-%	Between 20 and 33 {\bf non-deterministic} bugs were found:
-%	\begin{itemize}
-%		\item 20 deadlock races among 12 P2s
-%		\item 10 use-after-free races among 6 P2s
-%		\item 3 misc (assert fail, segfault) among 2 P2s
-%	\end{itemize}
-%\end{frame}
-
-\begin{frame}{Last Semester}
-	Spring 2015: 7 groups signed up to use Landslide; 5 found bugs
-	%\begin{itemize}
-	%	\item (that they didn't find during stress testing)
-	%\end{itemize}
-	%%%% hopefully!
-	\linegap
-
-	Among all groups, 38 total tests were run
-	\begin{itemize}
-		\item 21 tests ran to completion without finding bugs
-		\item 4 {\em deterministic} bugs
-		\item 13 distinct {\em non-deterministic} bugs
-			\begin{itemize}
-				\item 6 in {\tt thr\_exit\_join} \small (use-after-free, deadlock, NULL, infinite loop...)
-				\item 4 in {\tt paraguay} (deadlock, infinite loop)
-				\item 2 in {\tt broadcast\_test} (deadlock, infinite loop)
-				\item 1 in {\tt rwlock\_downgrade\_read\_test} (use-after-free)
-			\end{itemize}
-			%\begin{itemize}
-			%	\item 6 of which required data-race PPs to expose
-			%\end{itemize}
-	\end{itemize}
-\end{frame}
-
-\begin{frame}{A new treat since last semester}
-	\texttt{mutex\_test} tries to make your mutexes fail or deadlock.
-	\linegap
-
-	Landslide will look for data races inside your mutex implementation.
-	\begin{itemize}
-		\item {\tt xchg}/{\tt cmpxchg}/{\tt xadd}
-		\item {\tt mutex->whose\_turn\_is\_it = ...;}
-	\end{itemize}
-	\pause
-	\linegap
-
-	Tested 77 P2s from S'14, F'14, S'15
-	\begin{itemize}
-		\item 5 unrelated deterministic bugs (use-after-free, crash, etc)
-			% TODO check these for false positives
-			% But this was run after the arbiter fp deadlock avoidance
-			% So I think it's ok
-		\item 1 ``two threads got the lock at once'' race
-			\begin{itemize}
-				\item {\em ...that the TA who graded it didn't find!}
-			\end{itemize}
-		\item 7 deadlock races
-		\item 4 other races (NULL crash, infinite loop, use-after-free)
-	\end{itemize}
-\end{frame}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\subsection{Landslide for the People}
-
-\begin{frame}{User Study}
-	\textbf{Try Landslide on your P2!}
-	\begin{itemize}
-		\item Bare minimum effort: No more than 1 hour
-		\begin{itemize}
-			\item Clone a github URL, run setup script, run tests %, answer survey questions
-			\item Landslide will automatically report test results
-		\end{itemize}
-		\item Full study plan: 4-8 hours of active attention
-		\begin{itemize}
-			\item (Estimated, including time to diagnose and fix bugs)
-			\item However, many tests should run passively overnight -- start soon!
-		\end{itemize}
-	\end{itemize}
-	\pause
-	\linegap
-
-	Prerequisites
-	\begin{itemize}
-		\item You {\em must} pass the P2 hurdle before using Landslide.
-		\begin{itemize}
-			\item {\tt startle}, {\tt agility\_drill}, {\tt cyclone}, {\tt join\_specific\_test}, {\tt thr\_exit\_join}
-		\end{itemize}
-		\item Must have {\em attempted} several stress tests
-		\begin{itemize}
-			\item {\tt juggle 4 3 2 0}, {\tt multitest}, {\tt racer} (15 min), {\tt paraguay}
-		\end{itemize}
-
-	\end{itemize}
-\end{frame}
-
-\begin{frame}{User Study - Additional Information}
-	Human Subjects Research
-	\begin{itemize}
-		\item CMU IRB has approved this study
-		\item Landslide will collect results while you use it
-		\begin{itemize}
-			\item Record commands issued, take snapshots of your P2 code
-			\item All data will be anonymized before publication
-		\end{itemize}
-		\item No coercion: {\bf \em There is no penalty for not participating.}
-		\begin{itemize}
-			\item I am not on course staff, cannot influence your grade
-			\item Course staff will not have access to study data during semester
-		\end{itemize}
-	\end{itemize}
-	\pause
-	\linegap
-
-	Risks \& Benefits
-	\begin{itemize}
-		\item Benefit: Landslide may help you find/fix bugs, improving your grade!
-		\item Risk: Landslide may find no bugs and be a waste of your time.
-		\item Benefit: You might learn something...
-	\end{itemize}
-	% Say: Accordingly, although I hope you will use landslide enough to get good results for me, you should also responsibly allocate time between testing with landslide versus actually implementing your P2.
-\end{frame}
-
-\begin{frame}{User Study - How to Participate}
-	Interested?
-	\linegap
-
-	To participate\dots
-	\begin{itemize}
-		\item Review this lecture and study information sheet
-		\item Meet prerequisites of passing P2 tests
-		\item Complete sign-up form online to get further instructions
-			\begin{itemize}
-				\item \url{http://tinyurl.com/landslide-p2-f15}
-			\end{itemize}
-
-		\item Use Landslide in addition to stress tests until P2 is due!
-	\end{itemize}
-	\linegap
-
-	{\em Really} interested?
-	\begin{itemize}
-		\item I know of an M.S. thesis topic if you want to do post-OS research.
-	\end{itemize}
-\end{frame}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-\subsection{Experimental Results}
-
-\begin{frame}{More Evaluation}
-	\textbf{Is dynamically adding ``data race'' PPs effective?}
-	\begin{itemize}
-		%\item 3 among preliminary bugs (9-15\%) required data races to expose.
-		\item 6 among S'15 student bugs (46\%)
-		\item 10 among {\tt mutex\_test} bugs (71\%)
-	\end{itemize}
-	\linegap
-
-	{\bf Is ``iterative deepening'' better than state-of-the-art?}
-	\begin{itemize}
-		\item Control experiment: Just 1 state space, same CPU time
-		\begin{itemize}
-			\item PPs used: \texttt{mutex\_lock()}, \texttt{mutex\_unlock()}
-			\item 110 minutes on 1 CPU
-		\end{itemize}
-		\item Of 33 total bug reports, control failed to find 10 (30\%).
-		\begin{itemize}
-			\item 3 required data race PPs to expose
-			\item 1 ran out of time
-			\item 6 obscured by different bug in same state space
-		\end{itemize}
-	\end{itemize}
-\end{frame}
-
-\subsection{State Space Reduction}
-
-\begin{frame}{Coping with State Space Explosion}
-	Serious problem: State spaces grow exponentially
-	\begin{itemize}
-		\item With $p$ preemption points and $k$ runnable threads, size $p^k$.
-		\item Threatens our ability to explore everything.
-		\item Fortunately, some sequences result in identical states.
-	\end{itemize}
-	\linegap
-
-	{\bf Partial Order Reduction} identifies and skips ``equivalent'' interleavings.
-	\begin{itemize}
-		\item After each execution, compare memory reads/writes of each thread.
-		\item Find when reordering threads couldn't possibly change behaviour.
-		\item Example follows\dots
-	\end{itemize}
-\end{frame}
-
-\begin{frame}{State Space Reduction}
-	\begin{center}
-	\includegraphics[width=\textwidth]{undiamond0.png}
-	\end{center}
-\end{frame}
-\begin{frame}{State Space Reduction}
-	\begin{center}
-	\includegraphics[width=\textwidth]{undiamond1.png}
-	\end{center}
-\end{frame}
-\begin{frame}{State Space Reduction}
-	\begin{center}
-	\includegraphics[width=\textwidth]{diamond1.png}
-	\end{center}
-\end{frame}
-
 \end{document}

### 2016-03-01 oopsla/calcm.tex
write more for calcm

@@
 	\begin{itemize}
-		\item Dynamic Partial Order Reduction\related{Flanagan '05}
+		\item Dynamic Partial Order Reduction (DPOR)\related{Flanagan '05}
 		\item State space estimation\related{Simsa '12}
@@
 	\end{center}
-	Pure HB: \texttt{mx} establishes $HB(A, B)$; no data race reported (false negative).
+	{\bf Pure HB}: \texttt{mx} establishes $HB(A, B)$; no race reported (false negative).
 	\linegap
 
-	Limited HB: $\neg HB(A, B)$; data race is reported (correctly).
+	{\bf Limited HB}: $\neg HB(A, B)$; race is reported (correctly).
 \end{frame}
@@
 	\end{center}
-	Pure HB: \texttt{mx} establishes $HB(A, B)$; no data race reported (correctly).
+	{\bf Pure HB}: \texttt{mx} establishes $HB(A, B)$; no race reported (correctly).
 	\linegap
 
-	Limited HB: $\neg HB(A, B)$; data race is reported (false positive).
+	{\bf Limited HB}: $\neg HB(A, B)$; race is reported (false positive).
 \end{frame}
@@
 \begin{frame}{Iterative Deepening}
-	If time allows, combine PPs into larger, more comprehensive state spaces.
+	If time allows, combine PPs into larger, more comprehensive state spaces, eventually converging to using all PPs at once ({\em ``maximal'' state space}).
 	\begin{center}
@@
 	\begin{itemize}
-		\item 
+		\item Ordinarily infeasible, but some tests are small enough
+		\item Whether a test can be totally verified can't be known in advance.
+		\item ``Hard-coded'' PP approach\related{Musuvathi '08} sacrifices this possibility.
+	%One downside of hard-coding PPs in advance is timing out when the test is too large, but another downside is if the test is small enough, you might miss a possible full verification!
+		\item What PP strategy works for both small and large tests?
 	\end{itemize}
-	%One downside of hard-coding PPs in advance is timing out when the test is too large, but another downside is if the test is small enough, you might miss a possible full verification!
+	\pause
+	\linegap
+
+	Synchronization API PPs \& data-race PPs are sufficient!
+	\begin{itemize}
+		\item {\bf Theorem (Convergence)}:
+			{\em If a bug can be exposed by any thread interleaving
+				possible by preempting on all instructions,
+				% during a specific test
+			Quicksand will eventually test an equivalent interleaving which exposes the same bug.}
+		\item Proof appeals to DPOR soundness\related{Flanagan '05} and Limited HB.
+	\end{itemize}
 \end{frame}
@@
 \section{Evaluation}
+
+\breakslide{\Huge Evaluation}
+
+\begin{frame}{Evaluation}
+	Evaluation questions:
+	\begin{enumerate}
+		\item Do we find more bugs than prior MC with data-race PPs?
+		\item Do we find bugs faster than prior MC, even without data-race PPs?
+			% say: "the same bugs"
+		\item Does MC uncover new, nondeterministic, data-race candidates?
+		\item Do we provide more total verifications than prior MC?
+	\end{enumerate}
+\end{frame}
+
+\begin{frame}{Experimental Setup}
+	\textbf{Test suite}: 79 ``P2''s $\times$ 6 test cases + 78 ``Pintos''s $\times$ 3 test cases
+	\begin{itemize}
+		\item P2: CMU 15-410 thread library project (\textasciitilde{}1800 LOC)
+		\item Pintos: Berkeley CS162/U. Chicago CS320 kernel project (\textasciitilde{}700 LOC)
+		\item 629 unique state spaces (i.e., project/test pairs)
+	\end{itemize}
+	\pause
+	\linegap
+
+	{\bf Quicksand experiment}: 1 hour $\times$ 10 CPUs per test
+	\begin{itemize}
+		\item {\bf QS}: Data-race PPs enabled
+		\item {\bf QS-no-DRs}: Restrict iterative deepening to sync API PPs only
+	\end{itemize}
+	\linegap
+
+	{\bf Control experiment}: 10 hours $\times$ 1 CPU per test
+	\begin{itemize}
+		\item Test only ``maximal`` state space, using all sync API PPs at once
+		\item {\bf DPOR}: Use DPOR\related{Flanagan '05} to test maximal space in single pass
+		\item {\bf ICB}: Use Iterative Context Bounding\related{Musuvathi '07, Coons '13} to repeatedly test maximal space with increasing preemption bound
+			% dont mention BPOR
+	\end{itemize}
+\end{frame}
 

### 2016-03-01 oopsla/calcm.tex
write calcm conclusions slide

@@
 
-\begin{frame}{MC State of the Art}
+\begin{frame}{MC State of the Art}{(my new DJ name?)}
 	Modern stateless MC tools can {\em mitigate} exponential explosion.
@@
 \begin{frame}{Iterative Deepening}
-	If time allows, combine PPs into larger, more comprehensive state spaces, eventually converging to using all PPs at once ({\em ``maximal'' state space}).
+	If time allows, combine PPs into larger, more comprehensive state spaces.
+	\linegap
+
+	All PPs enabled = ``maximal'' state space
+	\begin{itemize}
+		\item Prior work tools test this state space only.
+	\end{itemize}
+	%eventually converging to using all PPs at once ({\em ``maximal'' state space}).
 	\begin{center}
@@
 		\item Ordinarily infeasible, but some tests are small enough
-		\item Whether a test can be totally verified can't be known in advance.
+		%\item Whether a test can be totally verified can't be known in advance.
 		\item ``Hard-coded'' PP approach\related{Musuvathi '08} sacrifices this possibility.
@@
 	\end{itemize}
-	\pause
-	\linegap
-
-	{\bf Quicksand experiment}: 1 hour $\times$ 10 CPUs per test
+\end{frame}
+\begin{frame}{Experimental Setup}
+	\textbf{Quicksand experiment}: 1 hour $\times$ 10 CPUs per test
 	\begin{itemize}
@@
 
+\begin{frame}{1. Finding more bugs with data-race PPs}
+\end{frame}
+
+\begin{frame}{2. Finding the same bugs... slower}
+\end{frame}
+
+\begin{frame}{3. Nondeterministic data-race candidates}
+\end{frame}
+
+\begin{frame}{4. Total verification}
+\end{frame}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \section{End}
+\breakslide{\Huge Conclusion}
+
+\begin{frame}{Conclusion}
+	Stateless Model Checking depends on good {\bf preemption points} (PPs).
+	\linegap
+
+	MC State of the Art chooses a fixed PP set in advance.
+	\linegap
+
+	Quicksand schedules multiple tests, with different PPs, in parallel.
+	\begin{itemize}
+		\item Iterative Deepening incorporates new data-race PPs on-the-fly
+		\item Data race PPs uncover {\bf 80\% more bugs} than sync API PPs alone
+		\item Data race PPs enable {\bf total verification} when test is small enough
+	\end{itemize}
+	\linegap
+
+	OOPSLA submission soon; looking for feedback!
+\end{frame}
+
 \breakslide{\Large Questions?

### 2016-03-01 oopsla/calcm.tex
Finish 1st draft of calcm slides

@@
 \begin{frame}{1. Finding more bugs with data-race PPs}
+	\begin{center}
+		\includegraphics[width=0.6\textwidth]{eval1.pdf}
+	\end{center}
+	\begin{itemize}
+		\item 17-118 seconds: Quicksand loses on parallelization/startup overhead
+			%\begin{itemize}
+			%	\item (remember: Quicksand on 10 CPUs, control on 1)
+			%\end{itemize}
+		\item Ultimately, 180\% as many bugs in total!
+	\end{itemize}
 \end{frame}
@@
 \begin{frame}{2. Finding the same bugs... slower}
+	\begin{center}
+		\includegraphics[width=0.6\textwidth]{eval2.pdf}
+	\end{center}
+	\begin{itemize}
+		\item ICB outperforms Iterative Deepening when using the same PPs
+		\item Suggests Quicksand could incorporate ICB in its scheduling strategy
+	\end{itemize}
 \end{frame}
@@
 \begin{frame}{3. Nondeterministic data-race candidates}
+	Some data races come from code that executes only nondeterministically.
+	\begin{itemize}
+		\item In a single-execution data race analysis, these accesses won't be seen.
+		\item MC can uncover the racy logic, identify a new PP, and find more bugs.
+	\end{itemize}
+
+	\begin{center}
+		\includegraphics[width=0.7\textwidth]{eval3.pdf}
+	\end{center}
 \end{frame}
@@
 \begin{frame}{4. Total verification}
+	\begin{center}
+		\includegraphics[width=0.6\textwidth]{eval4.pdf}
+	\end{center}
+	\begin{itemize}
+		\item If no data-race candidates, can verify with sync API PPs alone.
+		\item Data-race PPs allow total verification of 425\% as many tests!
+		\item ICB slower to verify than single-pass DPOR due to repeated work.
+	\end{itemize}
 \end{frame}

### 2016-03-01 oopsla/calcm.tex
small edits calcm

@@
 		\item HB is established by any synchronization (mutexes, cvars, xchg...)
-		\item Represents possiblity to threads' behaviour to change if reordered
+		\item Represents possiblity of threads' behaviour to change if reordered
 			%\begin{itemize}
@@
 		\item HB is established only when thread runnability changes (cvars).
-		\item Represents possibility to reorder accesses to occur simultaneously
+		\item Represents ability to reorder accesses to occur simultaneously
 		\item Captures accesses separated by unrelated, synchronized, logic
@@
 		\item Proof appeals to DPOR soundness\related{Flanagan '05} and Limited HB.
+		\item Intuition: ``Adding data-race PPs until saturation = all possible thread communication points.''
 	\end{itemize}

### 2016-04-14 proposal/thesprop.tex
add stuff

%for a more compact document, add the option openany to avoid
%starting all chapters on odd numbered pages
\documentclass[12pt]{cmuthesis}

% This is a template for a CMU thesis.  It is 18 pages without any content :-)
% The source for this is pulled from a variety of sources and people.
% Here's a partial list of people who may or may have not contributed:
%
%        bnoble   = Brian Noble
%        caruana  = Rich Caruana
%        colohan  = Chris Colohan
%        jab      = Justin Boyan
%        josullvn = Joseph O'Sullivan
%        jrs      = Jonathan Shewchuk
%        kosak    = Corey Kosak
%        mjz      = Matt Zekauskas (mattz@cs)
%        pdinda   = Peter Dinda
%        pfr      = Patrick Riley
%        dkoes = David Koes (me)

% My main contribution is putting everything into a single class files and small
% template since I prefer this to some complicated sprawling directory tree with
% makefiles.

% some useful packages
\usepackage{times}
\usepackage{fullpage}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage[numbers,sort]{natbib}
\usepackage[backref,pageanchor=true,plainpages=false, pdfpagelabels, bookmarks,bookmarksnumbered,
%pdfborder=0 0 0,  %removes outlines around hyper links in online display
]{hyperref}
\usepackage{subfigure}

% Approximately 1" margins, more space on binding side
%\usepackage[letterpaper,twoside,vscale=.8,hscale=.75,nomarginpar]{geometry}
%for general printing (not binding)
\usepackage[letterpaper,twoside,vscale=.8,hscale=.75,nomarginpar,hmarginratio=1:1]{geometry}

% Provides a draft mark at the top of the document. 
\draftstamp{\today}{DRAFT}

\begin {document} 
\frontmatter

%initialize page style, so contents come out right (see bot) -mjz
\pagestyle{empty}

\title{ %% {\it \huge Thesis Proposal}\\
{\bf Practical Concurrency Testing}}
\author{Ben Blum}
\date{Janucember 2016}
\Year{2016}
\trnumber{}

\committee{
Garth Gibson, Chair \\
David A. Eckhardt \\
Yet another person \\
Someone from a strange and faraway land
}

\support{}
\disclaimer{}

% copyright notice generated automatically from Year and author.
% permission added if \permission{} given.

\keywords{landslide terminal, baggage claim, public transportation, ticketing}

\maketitle

\begin{dedication}
	For my dog, Louie.
\end{dedication}

\pagestyle{plain} % for toc, was empty

%% Obviously, it's probably a good idea to break the various sections of your thesis
%% into different files and input them into this file...

\begin{abstract}
A short summary.
\end{abstract}

\begin{acknowledgments}
My advisor is cool.

But I am cooler.
\end{acknowledgments}

\tableofcontents
\listoffigures
\listoftables

\mainmatter

%% Double space document for easy review:
%\renewcommand{\baselinestretch}{1.66}\normalsize

% The other requirements Catherine has:
%
%  - avoid large margins.  She wants the thesis to use fewer pages, 
%    especially if it requires colour printing.
%
%  - The thesis should be formatted for double-sided printing.  This
%    means that all chapters, acknowledgements, table of contents, etc.
%    should start on odd numbered (right facing) pages.
%
%  - You need to use the department standard tech report title page.  I
%    have tried to ensure that the title page here conforms to this
%    standard.
%
%  - Use a nice serif font, such as Times Roman.  Sans serif looks bad.
%
% Other than that, just make it look good...

\chapter{Introduction}
\chapter{Conclusion}

%\appendix
%\include{appendix}

\backmatter

%\renewcommand{\baselinestretch}{1.0}\normalsize

% By default \bibsection is \chapter*, but we really want this to show
% up in the table of contents and pdf bookmarks.
\renewcommand{\bibsection}{\chapter{\bibname}}
%\newcommand{\bibpreamble}{This text goes between the ``Bibliography''
%  header and the actual list of references}
\bibliographystyle{plainnat}
\bibliography{register} %your bib file

\end{document}

### 2016-04-14 irb/infosheet.tex
Update infosheet and survey for landslide study
Add another fall 2014 note - commented out - in irb info sheet
update irb forms slightly
update irb stuff, stale changes lying around in my git diff

@@
 We expect you to protect yourself by responsibly allocating time between using Landslide and working on P2 normally.
+It is expected that participation in this study will require from 2 to 8 hours of active work, depending on bug-finding results.
 \\
@@
 Because we are providing you with a more powerful testing tool during P2 than you would normally have access to, we hope you will be able to find and fix bugs faster, resulting in a higher grade.
+%\\
+%
+%\noindent {\bf FALL 2014 NOTE}: As this semester we are conducting a preliminary study on a ``for fun'' basis, at the end of the semester rather than during P2, there will be no impact on your grade.
+15-410 staff will not accept patches based on your findings with Landslide.
 
@@
 While you use Landslide, it will automatically record your activity in terms of commands issued, time spent running tests, bug reports, and snapshots of your code.
-We will also collect paper surveys which you will fill out while using Landslide.
+%We will also collect paper surveys which you will fill out while using Landslide.
 Carnegie Mellon may be required to disclose personally identifiable information as required by law, regulation, subpoena or court order.
@@
 	\item Data collected by Landslide will be stored in AFS accessible only to Ben Blum and Garth Gibson.
-	\item Paper surveys will be stored in a locked filing cabinet accessible only to Ben Blum and Garth Gibson.
+	%\item Paper surveys will be stored in a locked filing cabinet accessible only to Ben Blum and Garth Gibson.
 	\item Within one semester after the study, we will destroy all personally identifying information, keeping only anonymized and aggregate results.
@@
 \end{itemize}
+%
+%\noindent {\bf FALL 2014 NOTE}: This semester, Landslide's automatic data collection is not implemented.
+We will rely on you to supply commands issued and time spent in the survey section, and to submit Landslide's bug reports and snapshots of your code in a tarball of your workspace.
 
@@
 Your participation is voluntary.
+You must be 18 years of age or older to participate.
 You are free to stop your participation at any point; likewise, the Principal Investigator may at his discretion remove you from the study.

### 2016-04-14 irb/survey.tex
update irb stuff, stale changes lying around in my git diff

@@
 
-\title{\bf Landslide: Systematic Testing in 15-410 \\ User Guide and Survey}
+\title{\bf Landslide: Systematic Testing in 15-410 \\ User Guide
+%and Survey
+}
 \author{Ben Blum}
@@
 
-The very first thing you should do is {\bf READ THIS DOCUMENT ENTIRELY FIRST!} We need you to print out several survey sheets before you dive in.
-\\
+%The very first thing you should do is {\bf READ THIS DOCUMENT ENTIRELY FIRST!} We need you to print out several survey sheets before you dive in.
+%\\
 
@@
 		\end{itemize}
-	\item Next, make sure you have read this document entirely.
+	%\item Next, make sure you have read this document entirely.
 At this point you should have several survey sheets printed out, ready to write on -- see the Survey section below.
 You should also have some means of timing yourself.
-	\item Run {\tt ./landslide OPTIONS} to run tests with Landslide. {\bf Fill out a fresh survey sheet each time you do this.} Repeat until satisfied.
+	\item Run {\tt ./landslide OPTIONS} to run tests with Landslide, for various values of {\tt OPTIONS}:
+		%{\bf Fill out a fresh survey sheet each time you do this.} Repeat until satisfied.
 		\begin{itemize}
@@
 			\item Landslide's bug reports will show up as HTML files. If you are using a cluster machine in person, you can view them by pointing a browser to {\tt file:///PATH/TO/landslide/FILENAME.html}.
-				\footnote{Another option: move it to {\tt \textasciitilde/www} in your AFS, and view it at {\tt http://www.contrib.andrew.cmu.edu/\textasciitilde{}YOUR\_ANDREWID/FILENAME.html}.
+				\footnote{For remote usage: move it to {\tt \textasciitilde/www} in your AFS, and view it at {\tt http://www.contrib.andrew.cmu.edu/\textasciitilde{}YOUR\_ANDREWID/FILENAME.html}.
 Note the {\tt www} in the URL is mandatory; note also that this will allow the world to see it, so be careful!}
+			\item The available test programs are {\tt thr\_exit\_join} (the default), {\tt mutex\_test}, {\tt paradise\_lost}, \\ {\tt broadcast\_test}, {\tt paraguay}, and {\tt rwlock\_downgrade\_read\_test}.
 				\footnote{If you have your own custom test case, you can put it in {\tt pebsim/p2-basecode/user/progs} and edit \\ {\tt pebsim/p2-basecode/config.mk} to build it, then run {\tt ./p2-setup.sh} again.}
-			\item The available test programs are {\tt thr\_exit\_join} (the default), {\tt broadcast\_test}, {\tt paraguay}, and {\tt rwlock\_downgrade\_read\_test}.
 			\item The suggested sequence of test configurations to run is as follows.
@@
 				\item {\tt ./landslide -p thr\_exit\_join -t 30m}
+				\item {\tt ./landslide -p mutex\_test -t 30m}
+				\item {\tt ./landslide -p paradise\_lost -t 30m}
 				\item {\tt ./landslide -p broadcast\_test -t 30m}
 				\item {\tt ./landslide -p paraguay -t 30m}
-				\item {\tt ./landslide -p rwlock\_downgrade\_read\_test -t 30m}
+				\item {\tt ./landslide -p rwlock\_downgrade\_read\_test -t 30m} \\
 				\item {\tt ./landslide -p thr\_exit\_join -t 3h}
+				\item {\tt ./landslide -p mutex\_test -t 3h}
+				\item {\tt ./landslide -p paradise\_lost -t 3h}
 				\item {\tt ./landslide -p broadcast\_test -t 3h}
 				\item {\tt ./landslide -p paraguay -t 3h}
-				\item {\tt ./landslide -p rwlock\_downgrade\_read\_test -t 3h}
+				\item {\tt ./landslide -p rwlock\_downgrade\_read\_test -t 3h} \\
 				\item {\tt ./landslide -p thr\_exit\_join -t 12h}
+				\item {\tt ./landslide -p mutex\_test -t 12h}
+				\item {\tt ./landslide -p paradise\_lost -t 12h}
 				\item {\tt ./landslide -p broadcast\_test -t 12h}
@@
 
-\newpage
-
-\section{Survey}
-\subsection{Overall Survey Questions}
-
-\begin{center}
-{\bf Print this page out once.}
-\end{center}
-
-\begin{enumerate}
-	\item How much time did you spend directly using Landslide?
+%\newpage
+%
+%\section{Survey}
+%\subsection{Overall Survey Questions}
+%
+%\begin{center}
+%{\bf Print this page out once.}
+%\end{center}
+%
+%\begin{enumerate}
+%	\item How much time did you spend directly using Landslide?
 This includes configuring test parameters and interpreting debug output.
@@
 overnight.
-		\vspace{0.5in}
-
-	\item What test configurations did you use that DIDN'T find bugs? For example, {\tt ./landslide -t 15m -p thr\_exit\_join}
-\end{enumerate}
-
-\newpage
-\subsection{Per-Bug Survey Questions}
-\begin{center}
-{\bf \large PRINT OUT THIS PAGE MULTIPLE TIMES.
-\\
-
-Each time you find a new bug with Landslide, fill this out again.}
-\end{center}
-
-\begin{enumerate}
-		% TODO: find a way to allow ze studence to condense non-bug-finding runs onto 1 page.
-	\item What test configuration did you use? For example, {\tt ./landslide -t 15m -p thr\_exit\_join}
-		\vspace{0.5in}
-	\item Summarize the bug.
-		\vspace{2in}
-	\item How IMPORTANT was this bug? Rate on a scale from 1 to 10, where:
-		\begin{itemize}
-			\item 1 = Trivial; an obvious ``oops'' that does not reflect any key lessons learned during the project. I would expect to not lose any points for this bug.
-			\item 4 = Minor; a logic problem not related to threads/synchronization. I would expect to lose 1 or 2 points at most for this bug.
-			\item 7 = Major; a concurrency-related bug related to subject matter discussed in lecture.
-			\item 10 = Severe; this bug reflects a fundamental lesson about concurrency that we learned during the project.
+%		\vspace{0.5in}
+%
+%	\item What test configurations did you use that DIDN'T find bugs? For example, {\tt ./landslide -t 15m -p thr\_exit\_join}
+%\end{enumerate}
+%
+%\newpage
+%\subsection{Per-Bug Survey Questions}
+%\begin{center}
+%{\bf \large PRINT OUT THIS PAGE MULTIPLE TIMES.
+%\\
+%
+%Each time you find a new bug with Landslide, fill this out again.}
+%\end{center}
+%
+%\begin{enumerate}
+%		% TODO: find a way to allow ze studence to condense non-bug-finding runs onto 1 page.
+%	\item What test configuration did you use? For example, {\tt ./landslide -t 15m -p thr\_exit\_join}
+%		\vspace{0.5in}
+%	\item Summarize the bug.
+%		\vspace{2in}
+%	\item How IMPORTANT was this bug? Rate on a scale from 1 to 10, where:
+%		\begin{itemize}
+%			\item 1 = Trivial; an obvious ``oops'' that does not reflect any key lessons learned during the project. I would expect to not lose any points for this bug.
+%			\item 4 = Minor; a logic problem not related to threads/synchronization. I would expect to lose 1 or 2 points at most for this bug.
+%			\item 7 = Major; a concurrency-related bug related to subject matter discussed in lecture.
+%			\item 10 = Severe; this bug reflects a fundamental lesson about concurrency that we learned during the project.
 I would expect to receive ``see course staff'' in my ink if I turned in code with this bug!
-		\end{itemize}
-		\vspace{0.2in}
-	\item How much EFFORT/TIME did this bug require to diagnose and fix? Rate on a scale from 1 to 10, where:
-		\begin{itemize}
-			\item 1 = Very little; we knew immediately what was wrong and needed to change only 1 or 2 lines of code.
-			\item 5 = Some; the bug took considerable thought to fix, and required nontrivial code changes.
-			\item 10 = Lots; we worked for several hours to fix the bug and/or required a major refactoring of our code.
-		\end{itemize}
-		\vspace{0.2in}
-	\item How likely would a CONVENTIONAL STRESS TEST be able to find the bug (i.e., running the program repeatedly, not using Landslide)? Rate on a scale from 1 to 10, where:
-		\begin{itemize}
-			\item 1 = No hope. The window for a random timer interrupt to expose the bug is so small that a conventional stress test could easily run continuously overnight without encountering it.
-			\item 5 = Moderate chance. Most random timer interrupts would not expose the bug, but the window is more than just a few instructions.
-			\item 10 = Very likely. A random timer interrupt during the test would expose the bug more often than not.
-		\end{itemize}
-\end{enumerate}
+%		\end{itemize}
+%		\vspace{0.2in}
+%	\item How much EFFORT/TIME did this bug require to diagnose and fix? Rate on a scale from 1 to 10, where:
+%		\begin{itemize}
+%			\item 1 = Very little; we knew immediately what was wrong and needed to change only 1 or 2 lines of code.
+%			\item 5 = Some; the bug took considerable thought to fix, and required nontrivial code changes.
+%			\item 10 = Lots; we worked for several hours to fix the bug and/or required a major refactoring of our code.
+%		\end{itemize}
+%		\vspace{0.2in}
+%	\item How likely would a CONVENTIONAL STRESS TEST be able to find the bug (i.e., running the program repeatedly, not using Landslide)? Rate on a scale from 1 to 10, where:
+%		\begin{itemize}
+%			\item 1 = No hope. The window for a random timer interrupt to expose the bug is so small that a conventional stress test could easily run continuously overnight without encountering it.
+%			\item 5 = Moderate chance. Most random timer interrupts would not expose the bug, but the window is more than just a few instructions.
+%			\item 10 = Very likely. A random timer interrupt during the test would expose the bug more often than not.
+%		\end{itemize}
+%\end{enumerate}
 

### 2016-04-14 proposal/thesprop.tex
abstrax

@@
 \usepackage{amsmath}
+\usepackage[charter]{mathdesign}
 \usepackage[numbers,sort]{natbib}
@@
 \begin{abstract}
-A short summary.
+Concurrent programming presents a challenge to students and experts alike because of the complexity of multithreaded interactions and the difficulty to reproduce and reason about bugs.
+Stateless model checking is a concurrency testing approach which forces a program to interleave its threads in many different ways, checking for bugs each time.
+This technique is powerful, in principle capable of finding any nondeterministic bug in finite time,
+but suffers from exponential explosion as program size increases.
+Checking an exponential number of thread interleavings is not a practical or predictable approach for programmers to find concurrency bugs before their project deadlines.
+
+In this thesis, I propose to make stateless model checking more practical for human use by way of several new techniques.
+I have built Landslide, a stateless model checker for testing student projects for undergraduate operating systems classes.
+Landslide includes a novel algorithm for automatically managing
+%the exploration of
+multiple state spaces according to their estimated completion times,
+which I will show quickly finds bugs should they exist and also quickly verifies correctness otherwise.
+I will evaluate Landslide's suitability for inexpert use by presenting the results of a many-semester study giving it as a debugging aid to students in 15-410, CMU's Operating System Design and Implementation class.
+Finally, I will present several new techniques that allow stateless model checking to be practically employed on real-world programs.
 \end{abstract}
@@
 \tableofcontents
-\listoffigures
-\listoftables
+%\listoffigures
+%\listoftables
 

### 2016-04-28 proposal/thesprop.tex
write thesprop intro

@@
 In this thesis, I propose to make stateless model checking more practical for human use by way of several new techniques.
-I have built Landslide, a stateless model checker for testing student projects for undergraduate operating systems classes.
+I have built Landslide, a stateless model checker specializing in student projects for undergraduate operating systems classes.
 Landslide includes a novel algorithm for automatically managing
@@
 which I will show quickly finds bugs should they exist and also quickly verifies correctness otherwise.
-I will evaluate Landslide's suitability for inexpert use by presenting the results of a many-semester study giving it as a debugging aid to students in 15-410, CMU's Operating System Design and Implementation class.
-Finally, I will present several new techniques that allow stateless model checking to be practically employed on real-world programs.
+I will evaluate Landslide's suitability for inexpert use by presenting the results of many semesters providing it to students in 15-410, CMU's Operating System Design and Implementation class.
+%Finally, I will present several new techniques that allow stateless model checking to be practically employed on real-world programs.
+Finally, I will explore broader impact by extending Landslide to test real-world programs and to be used by students at other universities.
 \end{abstract}
@@
 \chapter{Introduction}
+
+Modern computer architectures have turned to increasing CPU core count, rather than clock speed, to improve processing power.
+%To take advantage of multiple cores, programs must be written {\em concurrently}
+To take advantage of multiple cores for performance, programmers must write software to execute {\em concurrently} --
+using multiple {\em threads} which execute multiple parts of a program's logic simultaneously.
+However, when threads access the same shared data, they may interleave in unexpected ways which change the outcomes of their execution.
+When an unexpected interleaving produces undesirable program behaviour,
+for example, by corrupting shared data structures,
+we call it a {\em concurrency bug}.
+Concurrency bugs are notoriously hard for programmers to find and debug
+because the specific thread interleaving required to trigger them arises at random during normal execution,
+and often with very low probability.
+%Concurrency bugs are notoriously hard to find and reproduce because they only appear in specific thread interleavings, which arise at random during normal program execution.
+% TODO(LAYMAN): give example of trying to open car door at same time as friend turns key to unlock it.
+
+Most commonly, a programmer searches for concurrency bugs in her code by running it many times (in parallel, in serial, or both),
+hoping that eventually, it will run according to the particular interleaving required to expose a hypothetical bug.
+This technique, known as {\em stress testing}, is unreliable,
+providing no guarantee of finding the failing interleaving in any finite amount of time, even when a bug does exist.
+It also provides no assurance of correctness:
+when finished, there is no way of knowing how much of the space of possible thread interleavings was actually tested.
+%it may by chance test only a single interleaving over and over again!
+Nevertheless, stress testing remains popular because of how easily a programmer can use it:
+she simply wraps her program in a loop, sets it to run overnight, and kills it if her patience runs out before it finds a bug.
+
+{\em Stateless model checking} \cite{verisoft} is an alternative way to test for concurrency bugs,
+or to verify their absence,
+which provides more reliable coverage, progress, and verification than stress testing.
+A stateless model checker tests a program by forcing it to execute a new unique thread interleaving on each iteration of the test,
+capturing and controlling the randomness in a finite state space of all possible interleavings.
+
+Unfortunately, the size of these state spaces is exponentially proportional to the size of the tested program.
+% TODO(LAYMAN): explain exponential explosion by relating the parable of grains of rice on a chessboard.
+For even moderately-sized programs, there may be more possible ways to interleave every thread's every instruction
+than particles in the universe.
+Accordingly, a programmer who wants her test to make reasonable progress through the state space must choose a subset of ways that her threads could interleave,
+focusing on fully testing that subset, while ignoring other possibilities she doesn't care about.
+%However, choosing this subset is a difficult trade-off for humans to make before even knowing whether a bug exists.
+However, it is difficult to choose a subset of thread interleavings that will produce a meaningful, yet feasible test.
+Until computers can automatically navigate this trade-off in some intelligent way,
+programmers will continue to fall back to the random approach of stress testing.
+
+Another problem stateless model checking suffers is that certain types of programs cannot be tested without the programmer putting forth some manual instrumentation effort.
+For example, operating system kernels implement their own sources of concurrency and their own synchronization primitives,
+so the checker needs to be told how to identify and control the execution of each thread.
+Some expert programmers may be willing to add manual annotations to their code,
+if they are already veteran concurrency research wizards,
+but required manual effort is a serious downside for anyone with a looming deadline,
+and especially so for students who are still learning basic concurrency principles.
+%We should not expect programmers to add effortful manual annotations to their code,
+%or they will abandon our fancy technique to instead simply run stress tests until their deadline tomorrow evening.
+
+This thesis will solve both problems discussed above.
+My thesis statement is as follows:
+
+\begin{center}
+	{\em my research is cool. please let me graduate.}
+\end{center}
+
+\vspace{1em}
+
+I have built Landslide \cite{landslide}, a stateless model checker for thread libraries and kernels,
+and I have developed some techniques for automatically choosing the best thread interleavings to test
+and for automatically instrumenting operating system kernels in an educational setting.
+This thesis will comprise three major contributions:
+
+\begin{itemize}
+	\item {\bf Meaningful state spaces (\ref{chap:pps}).}
+		I will present {\em Iterative Deepening}, a new algorithm for navigating the trade-off in how many preemption points to test at once.
+		Iterative Deepening incorporates state space estimation \cite{estimation} to decide on-the-fly whether each state space is worth pursuing, and uses data race analysis \cite{tsan} to find new preemption point candidates based on a program's dynamic behaviour.
+		This section will include a large evaluation of the technique, comparing its performance to three prior work approaches across 629 unique tests.
+		I will show that Iterative Deepening of preemption points outperforms prior work in terms both of finding bugs quickly and of completely verifying correctness when no bug exists.
+	\item {\bf Educational use (\ref{chap:410}).}
+		For the past three semesters, I have offered a fully-automated version of Landslide to students in 15-410, CMU's undergraduate Operating System Design and Implementation class \cite{kspec,thrlib}, for use as a debugging aid during the thread library project.
+		I will continue these user studies, and use the data to evaluate the suitability of stateless model checking in an educational setting.
+		%, investigating the following questions in particular:
+		%\begin{itemize}
+		%	\item Does Landslide improve project submission quality when used as a debugging aid?
+		%	\item Does Landslide teach students lessons in concurrent programming 
+		%\end{itemize}
+	\item {\bf Broader impact (\ref{chap:lipservice}).}:
+		Landslide includes several new techniques for testing kernels and thread libraries with little or no manual instrumentation required.
+		So far, a fully-automatic testing mode is available only for 15-410 thread library projects.
+		To prove these techniques are relevant beyond CMU's walls, I will extend Landslide to handle both Pintos kernel projects from other universities \cite{pintos} and ``real-world'' Linux programs.
+		I will conduct a user study in which students at another university test their Pintoses with Landslide,
+		and evaluate Landslide's ability to find known and/or new bugs in real-world programs.
+\end{itemize}
+
+% TODO(thesis): Other chapters of this thesis include...
+
+\chapter{Meaningful State Spaces}
+\label{chap:pps}
+
+\chapter{Education}
+\label{chap:410}
+
+\chapter{Broader Impact}
+\label{chap:lipservice}
+
+%Reduction
+%techniques such as Dynamic Partial Order Reduction \cite{dpor} and Maximal Causality Reduction \cite{mcr} expand the limits of feasible test completion,
+%and search ordering strategies such as Iterative Context Bounding \cite{chess-icb} \revision{encourage}~bugs to be found sooner in a given space should they exist.
+
+%Topics of thesis:
+%1. Data race preemption points / iterative deepening
+%2. Education in 410. Evaluation questions are:
+%	2a. Does P2 submission quality improve?
+%	2b. Does P3 submission quality improve correlated with students who used (got the most out of) Landslide during P2?
+%	2c. Does Landslide reach the bottom of the class? (Look at bottom-of-class P0s/P1s, see how they improve)
+%3. Broader impact
+%	3a. Real world programs (and the challenges therein?)
+%	3b. Give to pintos kids (and open-source landslide on bochs in doing so)
+
 \chapter{Conclusion}
@@
 \bibliographystyle{plainnat}
-\bibliography{register} %your bib file
+\bibliography{citations} %your bib file
 

### 2016-04-29 proposal/thesprop.tex
write some background text

@@
 
-\begin{dedication}
-	For my dog, Louie.
-\end{dedication}
+%\begin{dedication}
+%	For my dog, Louie.
+%\end{dedication}
 
@@
 This technique, known as {\em stress testing}, is unreliable,
-providing no guarantee of finding the failing interleaving in any finite amount of time, even when a bug does exist.
+providing no guarantee of finding the failing interleaving in any finite amount of time.
+%, even when a bug does exist.
 It also provides no assurance of correctness:
@@
 so the checker needs to be told how to identify and control the execution of each thread.
-Some expert programmers may be willing to add manual annotations to their code,
-if they are already veteran concurrency research wizards,
+Some expert concurrency research wizards may be willing to add manual annotations to their code,
 but required manual effort is a serious downside for anyone with a looming deadline,
@@
 
+\input{overview}
+
 \chapter{Meaningful State Spaces}

### 2016-04-30 proposal/410.tex
Add a bunch of boilerplate

\chapter{Education}
\label{chap:410}

This chapter proposes my plan to evaluate Landslide's effectiveness as a debugging aid for students in an educational setting.
This is the second of the three projects I am proposing for this thesis, and is currently ongoing work.

### 2016-04-30 proposal/lipservice.tex
Add a bunch of boilerplate

\chapter{Broader Impact}
\label{chap:lipservice}

In this chapter I propose to add several enhancements to Landslide to allow it to be used beyond CMU's walls,
with little to no manual instrumentation effort required.
This is the third of the three projects I am proposing for this thesis and so far exists only in dreams.

### 2016-04-30 proposal/quicksand.tex
Add a bunch of boilerplate

\chapter{State Spaces}
\label{chap:quicksand}

In this chapter, I will present a new algorithm for model checkers to automatically choose which preemption points to use to test reasonably-sized state spaces.
This is the first of the three projects I am proposing for this thesis, and has largely been completed and is under submission.

### 2016-04-30 proposal/thesprop.tex
Add a bunch of boilerplate

@@
 
-\begin{acknowledgments}
-My advisor is cool.
-
-But I am cooler.
-\end{acknowledgments}
+%\begin{acknowledgments}
+%My advisor is cool.
+%
+%But I am cooler.
+%\end{acknowledgments}
 
@@
 using multiple {\em threads} which execute multiple parts of a program's logic simultaneously.
-However, when threads access the same shared data, they may interleave in unexpected ways which change the outcomes of their execution.
+However, when threads access the same shared data, they may interleave in unexpected ways which change the outcome of their execution.
 When an unexpected interleaving produces undesirable program behaviour,
@@
 It also provides no assurance of correctness:
-when finished, there is no way of knowing how much of the space of possible thread interleavings was actually tested.
+when finished, there is no way of knowing how many distinct thread interleavings were actually tested.
 %it may by chance test only a single interleaving over and over again!
@@
 
+\vspace{1em}
+
 \begin{center}
@@
 \begin{itemize}
-	\item {\bf Meaningful state spaces (\ref{chap:pps}).}
+	\item {\bf Meaningful state spaces (Chapter \ref{chap:quicksand}).}
 		I will present {\em Iterative Deepening}, a new algorithm for navigating the trade-off in how many preemption points to test at once.
 		Iterative Deepening incorporates state space estimation \cite{estimation} to decide on-the-fly whether each state space is worth pursuing, and uses data race analysis \cite{tsan} to find new preemption point candidates based on a program's dynamic behaviour.
-		This section will include a large evaluation of the technique, comparing its performance to three prior work approaches across 629 unique tests.
+		This section will include a large evaluation of the technique, comparing its performance to three prior work approaches across 600+ unique tests.
 		I will show that Iterative Deepening of preemption points outperforms prior work in terms both of finding bugs quickly and of completely verifying correctness when no bug exists.
-	\item {\bf Educational use (\ref{chap:410}).}
+
+		This work is {\bf largely completed} and is currently under submission at OOPSLA 2016.
+	\item {\bf Educational use (Chapter \ref{chap:410}).}
 		For the past three semesters, I have offered a fully-automated version of Landslide to students in 15-410, CMU's undergraduate Operating System Design and Implementation class \cite{kspec,thrlib}, for use as a debugging aid during the thread library project.
@@
 		%\end{itemize}
-	\item {\bf Broader impact (\ref{chap:lipservice}).}:
+
+		This is {\bf ongoing work}; I have run the user study for 3 semesters so far and am proposing to continue them and follow them up with an evaluation.
+	\item {\bf Broader impact (Chapter \ref{chap:lipservice}).}
 		Landslide includes several new techniques for testing kernels and thread libraries with little or no manual instrumentation required.
@@
 		and evaluate Landslide's ability to find known and/or new bugs in real-world programs.
+
+		This will be {\bf mostly new work}, as Landslide presently supports testing Pintoses only with some manual effort,
+		and does not yet feature instrumentation to run on Linux programs.
 \end{itemize}
@@
 \input{overview}
-
-\chapter{Meaningful State Spaces}
-\label{chap:pps}
-
-\chapter{Education}
-\label{chap:410}
-
-\chapter{Broader Impact}
-\label{chap:lipservice}
+\input{quicksand}
+\input{410}
+\input{lipservice}
 
@@
 
-\chapter{Conclusion}
+\chapter{Conclusion and Timeline}
+
+Stateless model checking is a powerful technique for testing concurrent programs,
+capable in theory of rooting out any bug or providing total verification on any program,
+but suffers several problems which relegate that theory to fantasy land \cite{vargomax}.
+Chief among those problems is exponential explosion of state spaces,
+making it difficult to decide in advance which combinations of preemption points can be productively tested within a given time limit.
+Another major problem is the manual annotation effort required to test certain types of concurrent programs,
+which is especially relevant for operating systems students who implement their own kernels.
+In this document I have proposed a thesis which will solve both problems.
+I leave you now with a proposed timeline for bringing each project to fruition.
+
+\begin{itemize}
+	\item TODO % TODO
+\end{itemize}
+
+% TODO - make a timeline
 

### 2016-04-30 proposal/quicksand.tex
write about quicksand motiv and deisgn

@@
 This is the first of the three projects I am proposing for this thesis, and has largely been completed and is under submission.
+
+\section{Motivation}
+
+This project is motivated by my experience contributing to Parrot \cite{parrot},
+a partially-determinizing runtime for multithreaded programs which allowed some nondeterminism in performance-critical sections to achieve high performance.
+We combined Parrot with dBug \cite{dbug-ssv}, a model checker, to check the correctness of the nondeterministic sections of programs which remained.
+Unfortunately, dBug was able to finish a state space for fewer than half of these programs in our evaluation in a 24-hour time limit,
+even after the reduction Parrot provided.
+In that project, our model checker was limited to a fixed set of preemption points;
+in other words, it would always test different programs with the same granularity of thread interleavings.
+Consequently, the resulting completion time for each test would vary wildly,
+unpredictably falling on one side or the other of our arbitrary time limit.
+
+I argue that this is the wrong usage model for model checkers to present to users who wish to test for concurrency bugs.
+Users want concrete bug reports (obviously), or in absence of those, they want some assurance of the program's correctness.
+However, users do not live in a $\infty$-sized fantasy land \cite{vargomax},
+and are generally willing to wait only for some fixed finite time.
+If full verification of all possible interleavings is not possible within that time,
+a {\em partial verification} guarantee for some smaller state spaces is better than an ``I don't know'' verdict for one single state space.
+
+\section{Design}
+
+Hence, I propose a more user-friendly framework for model checking, in which many different state spaces are tested,
+and higher priority is given to state spaces which appear possible to complete before the user's patience runs out or their project is due.
+At the heart of this framework is an algorithm I call {\bf Iterative Deepening},
+by analogy with the eponymous technique in chess artificial intelligence \cite{iterative-deepening-chess-ai}.
+In both chess and model checking, Iterative Deepening makes progressively ``deeper'' searches of an exponentially-sized state space,
+repeatedly increasing the size of a subset state space to be explored,
+until a prescribed time limit is exceeded.
+While in chess, the size bound is determined by a single depth parameter,
+the size of model checking state spaces is determined by any number of different combinations of preemption points being enabled or disabled.
+
+{\bf Quicksand.}
+I have built a wrapper program, called Quicksand, which manages the execution of multiple simultaneous Landslide instances.
+Quicksand relies on state space estimation \cite{estimation} to decide at runtime how to prioritize these jobs.
+Landslide can provide these estimates to a reasonable accuracy before actually testing a large fraction of interleavings for a given state space.
+Quicksand seeks to maximize completed state spaces,
+as each one serves as a guarantee that all interleavings possible with its preemption points were tested.
+Moreover, because Iterative Deepening treats the set of preemption points as mutable,
+it can add new preemption points reactively based on any runtime analysis.
+Landslide begins with a statically-coded set of synchronization API preemption points (such as used by dBug \cite{dbug-ssv}),
+and during testing, runs a dynamic data-race analysis \cite{hybriddatarace,tsan,ifrit} to identify new candidate preemption points.
+
+\newcommand\PendingJobs{\ensuremath{\mathcal{P}}}
+\newcommand\SuspendedJobs{\ensuremath{\mathcal{S}}}
+\newcommand\GetETA[1]{ETA(#1)}
+\newcommand\GetPPSet[1]{PPSet(#1)}
+
+{\bf Iterative Deepening.}
+With a limited CPU budget, Quicksand must avoid running tests that are likely to be fruitless.
+Hence, it separates the available preemption point sets into a set of {\em suspended} jobs (partially-explored state spaces with high ETAs),
+and a set of {\em pending} jobs (untested ones with unknown ETAs).
+When Landslide reports an ETA exceeding the time limit,
+Quicksand compares with other pending and suspended jobs to find another one more likely to complete in time.
+The Iterative Deepening algorithm implements this comparison, presented formally in Algorithm~\ref{alg:shouldworkblock}.
+Its main feature is understanding that when one Landslide instance is testing a superset of preemption points compared to another, the state space is correspondingly guaranteed to contain a superset of possible thread interleavings.
+%This allows it to extend estimation information for too-large state spaces to even larger state spaces
+%that if \GetPPSet{$j_1$} $\subset$ \GetPPSet{$j_2$},
+%and $j_1$ is suspended,
+%then $j_2$'s state space is guaranteed to be strictly larger, so $j_2$ will take at least as long.
+%Hence we should avoid testing $j_2$ unless $j_1$ is later resumed and its ETA improves after further execution. %over time.
+%%reveals that it might finish in time after all.
+%Similarly, whenever a job finds a bug, we cancel all pending superset jobs, as they might find only the same bug.
+
+\begin{algorithm}[t]
+        \SetKwInOut{Input}{Input}
+        %\textbf{Function} GetBestJob($j_0$, PendingJobs, SuspendedJobs): \\
+        \Input{$j_0$, the currently-running job}
+        %\Input{$eta$, $j_0$'s predicted completion time}
+        \Input{\PendingJobs, the list of pending jobs, sorted by decreasing heuristic priority}
+        \Input{\SuspendedJobs, the list of already-suspended jobs, sorted by increasing ETA}
+        \Input{$T$, the remaining time in the CPU budget}
+        \If{\GetETA{$j_0$} $<$ HeuristicETAFactor $\times$ $T$}{
+                return $j_0$ // Common case: job is expected to finish.
+        }
+        \ForEach{job $j_P \in$ \PendingJobs}{
+                // Don't run a pending job if a subset of it is already suspended; its ETA would be at least as bad. \\
+                \If {$\forall j_S \in$ \SuspendedJobs, \GetPPSet{$j_S$} $\not\subset$ \GetPPSet{$j_P$}}{
+                        return $j_P$
+                }
+        }
+        %// no pending jobs; maybe resume a suspended job \\
+        \ForEach{job $j_S \in$ \SuspendedJobs}{
+                \If{\GetPPSet{$j_0$} $\not\subset$ \GetPPSet{$j_S$}
+                        $\land$
+                        \GetETA{$j_0$} $>$ \GetETA{$j_S$}}{
+                        // If a subset of $j_S$ is also suspended, don't run the larger one first. \\
+                        \If{$\forall j_{S2} \in$ \SuspendedJobs, \GetPPSet{$j_{S2}$} $\not\subset$ \GetPPSet{$j_S$}}{
+                                return $j_S$
+                        }
+                }
+        }
+        return $j_0$ // \GetETA{$j_0$} was bad, but no other $j$ was better.
+        \caption{Suspending exploration of a state space in favour of a potentially smaller one.}
+        \label{alg:shouldworkblock}
+\end{algorithm}
+
+\newcommand\AllJobs{\ensuremath{\mathcal{J}}}
+\begin{algorithm}[t]
+        \SetKwInOut{Input}{Input}
+        \Input{$j_0$, the currently-running job}
+        \Input{\AllJobs, the set of all existing (or completed) jobs}
+        \Input{$\alpha$, an instruction reported by the MC as part of a racing access pair}
+        \If{$\forall j \in \AllJobs,$
+        \GetPPSet{$j_0$} $\cup$ $\alpha$
+        $\not\subseteq$
+        \GetPPSet{$j$}
+        }{
+                AddNewJob(\GetPPSet{$j_0$} $\cup$ $\alpha$, HeuristicPriority($\alpha$)) \\
+        }
+        \If{$\forall j \in \AllJobs,$ \GetPPSet{$j$} $\neq \{yield, \alpha\}$}{
+                AddNewJob($\{yield, \alpha\}$, HeuristicPriority($\alpha$))
+        }
+        \caption{Adding new jobs with data-race PPs.}
+        \label{alg:handledatarace}
+\end{algorithm}
+
+{\bf Data-race preemption points.}
+As mentioned, runtime data race analysis produces new potential sites at which preempting the program may produce new behaviour.
+With Iterative Deepening, this is a simple matter of creating a new state space with an additional preemption point enabled on the racing instructions by each thread, shown formally in Algorithm~\ref{alg:handledatarace}.
+The new state spaces may expose a failure, in which case Landslide reports a data-race bug,
+or complete successfully, indicating a benign or false-positive data race.
+They may also uncover a new data-race candidate entirely, %in some alternate interleaving,
+in which case Quicksand iteratively advances to a superset state space containing PPs for both racing access pairs.
+Being constrained by a CPU budget,
+time may run out before completing a data race's associated state space,
+in which case Quicksand reports a potential false positive that the user must handle.
+
+\section{Verification}
+
+\section{Evaluation}
+
+% TODO: is related work needed?
+% i can easily talk about it here but i dont' want to set a weird precedent for super weak related work in my other sections.

### 2016-05-01 proposal/overview.tex
some overview

\chapter{Background}
\label{chap:background}

This chapter will introduce the necessary background material on concurrency, stateless model checking, data-race analysis, and the relevant undergraduate operating systems classes.

\section{Concurrency Bugs}

\section{Stateless Model Checking}

Model checking \cite{verisoft} is a testing technique for systematically exploring the possible thread interleavings of a concurrent program.
A model checker executes the program repeatedly, each time according to a new thread interleaving, until the state space (or the CPU budget) is exhausted.
During each execution, it forces threads to execute serially, thereby confining the program's nondeterminism to controlled thread switches.
Some model checkers explicitly store the set of visited program states as a means of identifying equivalent interleavings \cite{spin}.
This approach is called {\em stateful} model checking.
In this thesis, I focus on {\em stateless} model checking,
which instead analyzes the sequence of execution events to avoid a prohibitive memory footprint.
Henceforth I will abbreviate ``stateless model checking'' simply as ``model checking'' for brevity.

{\bf Static versus dynamic analysis.}
Model checking is a {\em dynamic} program analysis, meaning that it observes the operations and accesses performed by the program as its code is executed.
In contrast, {\em static} program analyses check certain properties at the source code level.
Static analyses are ideal for ensuring certain standards of code quality, which often correlates with correctness,
but cannot decide for certain whether a given program will fail during execution without actually running the code.
% TODO: cite incompleteness theorem
Static analyses face the challenge of {\em false alarms} (or {\em false positives}):
code patterns which look suspicious but are actually correct.
A debugging tool which reports too many false alarms will dissuade developers from using it \cite{racerx}.
Dynamic analysis, our approach, identifies program behaviours that are definitely wrong,
so each bug report is accompanied by concrete evidence of the violation.
Assertions, segfaults, use-after-free of heap memory, and deadlock are examples of such failures we check for,
although a checker may also include arbitrary program-specific predicates.

{\bf Preemption points.}
During execution, a model checker identifies a subset of the program's operations as ``interesting'', i.e.,
points at which interrupting the current thread to run a different one is likely to produce different behaviour.
These points, called {\em preemption points}, may be identified by any combination of human intuition and machine analysis.
Typical preemption points include the boundaries of synchronization APIs (e.g., {\tt mutex\_lock}) or accesses to shared variables.
Considering that at each preemption point multiple threads exist as options to run next, the set of possible ways to execute the program can be viewed as a tree.
% TODO: fig tree
I show an example execution tree in Figure~\ref{fig:tree}.
The number of preemption points in each execution define the depth of this tree, and the number of threads available to run define the branching factor.
Hence, in a program with $n$ preemption points and $k$ threads available to run at each, the state space size is $n^k$.
Addressing the scaling problem in this exponential relation is the central research problem for all model checkers.

{\bf Reduction techniques.}
Dynamic Partial Order Reduction \cite{dpor} (henceforth, DPOR) is the premier algorithm for mitigating the exponential explosion that arises as program size increases.
It identifies equivalent execution sequences according to Mazurkiewicz trace theory \cite{mazurkiewicz},
and tests at least one execution from each equivalence class.
Intuitively, if two thread transitions between preemption points do not conflict on any shared resource access, reordering them produces an equivalent interleaving, i.e., the same program behaviour.
% TODO: put fig dpor
Figure~\ref{fig:dpor} shows a part of an execution tree in which the operations by threads 1 and 2 are independent, and hence only one of the subtrees need be tested.
Over the years, researchers have developed many enhancements to DPOR, such as Optimal DPOR \cite{optimal-dpor}, parallelizable DPOR \cite{parallel-dpor}, SAT-directed model checking \cite{satcheck}, Maximal Causality Reduction \cite{mcr}, and DPOR for relaxed memory architectures \cite{tsopso}.

{\bf Search heuristics.}
However, even though DPOR can prune an exponential number of redundant interleavings, the state space size is still exponential in the number of {\em dependent} (conflicting) interleavings.
Developers will always want to test larger and larger programs, so no matter the quality of our reduction algorithm,
we must accept that some tests will be too large to be fully tested in a reasonable time.
Hence, recent model checking research has turned to heuristic techniques for achieving further reduction,
optimizing the search to try to uncover bugs faster (should they exist)
at the expense of possibly missing other bugs,
or missing the chance to complete a full verification.
Iterative Context Bounding \cite{chess-icb} is a popular such technique which heuristically reorders the search to prioritize interleavings with fewer preemptions first.
This heuristic is based on the insight that most bugs require few preemptions to uncover, so interleavings with a number of preemptions that exceeds a certain bound will be de-prioritized, only tested until after all the fewer-preemption interleavings are completed.
Preemption sealing \cite{sealing} is another heuristic strategy which restricts the scope of the search by limiting the model checker to use only preemption points arising from certain functions in the source code.
This allows developers to vastly reduce state space size by identifying which program modules are already trusted,
although it requires some human intuition to correctly mark those boundaries.

\section{Data Race Analsyis}

\section{Education}

### 2016-05-03 proposal/overview.tex
add overview for DRs

@@
 \section{Concurrency Bugs}
+% TODO
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \section{Data Race Analsyis}
 
+\begin{figure}[t]
+        \small
+	\begin{center}
+\begin{tabular}{c}
+\begin{tabular}{rll}
+        & \multicolumn{2}{c}{\texttt{int x = 0; bool y = false; mutex\_t mx;}} \\
+        \\
+        & {\bf Thread 1} & {\bf Thread 2} \\
+        1 & \texttt{\hilight{brickred}{x++;}~// A1} & \\
+        2 & \texttt{mutex\_lock(\&mx);} & \\
+        3 & \texttt{mutex\_unlock(\&mx);} & \\
+        4 & & \texttt{mutex\_lock(\&mx);} \\
+        5 & & \texttt{mutex\_unlock(\&mx);} \\
+        6 & & \texttt{\hilight{brickred}{x++;}~// A2} \\
+\end{tabular}
+\\
+	{\normalsize (a) True potential data race.}
+\\
+\\
+\begin{tabular}{rll}
+        %& \multicolumn{2}{c}{\texttt{int x = 0; bool y = false; mutex\_t mx;}} \\
+        & {\bf Thread 1} & {\bf Thread 2} \\
+        1 & \texttt{\hilight{brickred}{x++;}~// B1} & \\
+        2 & \texttt{mutex\_lock(\&mx);} & \\
+        3 & \texttt{y = true;} & \\
+        4 & \texttt{mutex\_unlock(\&mx);} & \\
+        5 & & \texttt{mutex\_lock(\&mx);} \\
+        6 & & \texttt{bool tmp = y;} \\
+        7 & & \texttt{mutex\_unlock(\&mx);} \\
+        8 & & \texttt{if (tmp) \hilight{brickred}{x++;}~// B2} \\
+        %8 & & \texttt{if (tmp)} \\
+        %9 & & \texttt{~~~~\hilight{brickred}{x++;}~// B2} \\
+\end{tabular}
+\\
+{\normalsize (b) No data race in any interleaving.}
+\end{tabular}
+	\end{center}
+\caption{{Data-race analyses may be prone to either {\em false negatives} or {\em false positives}.
+Applying Happens-Before to program (a) will miss the potential race possible between A1/A2 in an alternate interleaving,
+while using Limited Happens-Before on (b) will produce a false alarm on B1/B2.}}
+\label{fig:hb-example}
+\end{figure}
+
+Data race analysis \cite{eraser} identifies pairs of unsynchronized memory accesses between threads.
+Two instructions are said to race if:
+\begin{enumerate}
+	\item they both access the same memory address,
+	\item at least one is a write,
+	\item the threads do not hold the same lock,
+	\item and no synchronization enforces an order on the thread transitions (the {\em Happens-Before} relation).
+\end{enumerate}
+%In Figure~\ref{fig:example}, lines 3 and 5 each race with 2 and 6, and line 6 races with 8.
+In Figure~\ref{fig:hb-example}, the pairs of lines marked with comments (A1/A2/B1/B2) race.
+
+A data race analysis may be either {\em static} (inspecting source code) \cite{racerx} or {\em dynamic} (tracking individual accesses arising at run-time) \cite{tsan}.
+This paper focuses exclusively on dynamic analysis,
+so although our example refers to numbered source lines for ease of explanation,
+in practice we are actually classifying the individual memory access events corresponding to those lines during execution.
+Actually, each {\tt x++} statement likely compiles to two separate load and store instructions, so each of those two instructions from each of the two marked source lines pairwise will race (except for the two loads, which are both reads).
+
+{\bf Variants of Happens-Before.}
+        Most prior work focuses on {\em Happens-Before} \cite{lamport-clocks} as the order relation between accesses.
+\cite{predictive-dr} and \cite{hybriddatarace} identify a problem with this approach:
+it cannot identify access pairs separated by an unrelated lock operation which could race in an alternate interleaving,
+as shown in the example program in Figure~\ref{fig:hb-example}(a).
+We call such unreported access pairs {\em false negatives}.
+\cite{hybriddatarace} introduces the {\em Limited Happens-Before} relation,
+which will report such potential races
+by considering only blocking operations like {\tt cond\_wait} to enforce the order.
+However, consider the similar program in Figure~\ref{fig:hb-example}(b),
+in which the access pair ceases to exist in the alternate interleaving.
+Limited Happens-Before will report all potential races, avoiding false negatives \cite{tsan},
+but at the cost of necessarily reporting some such {\em false positives}.
+
+%Finally, the {\em Causally-Precedes} relation \cite{predictive-dr} %strikes a middle ground,
+%extends Happens-Before to additionally report a subset of potential races while soundly avoiding false positives.
+%It tracks conflicting accesses in intervening
+%critical sections to determine whether lock events are unrelated to a potential race.
+%Causally-Precedes will identify the potential race in Figure~\ref{fig:hb-example}(a), as the two critical sections do not conflict,
+%although it can still miss true potential races in other cases.
+
+In Chapter~\ref{chap:quicksand}, we use the Limited Happens-Before relation for our analysis.
+The justification for this is that, while stand-alone data-race analyses must avoid inundating the user with false alarms \cite{racerx},
+my work incorporates data-race analysis in an internal feedback loop, and reports only directly observed failures to the user.
+Hence, I accept some overhead from false positives for the sake of more thorough testing.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \section{Education}
+
+In this thesis I will tackle Pebbles and Pintos, two different system architectures used in educational operating systems courses.
+This section describes the projects which students implement and which Landslide tests.
+
+\subsection{Pebbles}
+
+In the course of a semester, students work on five programming assignments;
+the first two are individual, and the remaining three are the products of two-person teams.
+I will focus on the third and fourth of these, the thread library and kernel,
+called ``P2'' and ``P3'' respectively (the project numbers start at 0).
+The other three (a stack-crawling backtrace utility, a bare-metal game with device drivers, and a small extension to the P3 kernel) are not of concern in this thesis.
+Both P2 and P3 are built using the {\em Pebbles} system call specification, outlined in Table~\ref{tab:syscalls}
+
+\begin{table}
+        \center
+        \begin{tabular}{|l|p{0.75\textwidth}|}
+                \hline
+                \bf System call name & \bf Summary \\
+                \hline
+                \multicolumn{2}{c}{\em Lifecycle management} \\
+                \hline
+                \texttt{fork} & Duplicates the invoking task, including all memory regions. \\
+                \texttt{thread\_fork} & Creates a new thread in the current task.\\
+                \texttt{exec} & Replaces the program currently running in the invoking task with a new one specified. \\
+                \texttt{set\_status} & Records the exit status of the current task. \\
+                \texttt{vanish} & Terminates execution of the calling thread. \\
+                \texttt{wait} & Blocks execution until another task terminates, and collects its exit status.\\
+                \texttt{task\_vanish}* & Causes all threads of a task to \texttt{vanish}. \\
+                \hline
+                \multicolumn{2}{c}{\em Thread management} \\
+                \hline
+                \texttt{gettid} & Returns the ID of the invoking thread. \\
+                \texttt{yield} & Defers execution to a specified thread. \\
+                \texttt{deschedule} & Blocks execution of the invoking thread. \\
+                \texttt{make\_runnable} & Wakes up another \texttt{deschedule}d thread. \\
+                \texttt{get\_ticks} & Gets the number of timer ticks since bootup. \\
+                \texttt{sleep} & Blocks a thread for a given number of ticks. \\
+                \texttt{swexn} & Registers a user-space function as a software exception handler.\\
+                \hline
+                \multicolumn{2}{c}{\em Memory management} \\
+                \hline
+                \texttt{new\_pages} & Allocates a specified region of memory. \\
+                \texttt{remove\_pages} & Deallocates same. \\
+                \hline
+                \multicolumn{2}{c}{\em Console I/O} \\
+                \hline
+                \texttt{getchar}* & Reads one character from keyboard input. \\
+                \texttt{readline} & Reads the next line from keyboard input. \\
+                \texttt{print} & Prints a given memory buffer to the console. \\
+                \texttt{set\_term\_color} & Sets the color for future console output. \\
+                \texttt{set\_cursor\_pos} & Sets the console cursor location. \\
+                \texttt{get\_cursor\_pos} & Retrieves the console cursor location. \\
+                \hline
+                \multicolumn{2}{c}{\em Miscellaneous} \\
+                \hline
+                \texttt{ls} & Loads a given buffer with the names of files stored in the RAM disk ``file system.'' \\
+                \texttt{halt} & Ceases execution of the operating system. \\
+                \texttt{misbehave}* & Selects among several thread-scheduling policies. \\
+                \hline
+        \end{tabular}
+        \caption{The Pebbles specifcation defines 25 system calls. Students are not required to implement ones marked with an asterisk (*), though the reference kernel provides them. }
+        \label{tab:syscalls}
+\end{table}
+
+{\bf P2.}
+% TODO
+
+{\bf P3.}
+% TODO
+
+\subsection{Pintos}
+% TODO

### 2016-05-13 proposal/overview.tex
some minor changes, subtitle

@@
 	\item the threads do not hold the same lock,
-	\item and no synchronization enforces an order on the thread transitions (the {\em Happens-Before} relation).
+	\item and no synchronization enforces an order on the thread transitions (the {\em Happens-Before} relation, described below).
 \end{enumerate}
 %In Figure~\ref{fig:example}, lines 3 and 5 each race with 2 and 6, and line 6 races with 8.
-In Figure~\ref{fig:hb-example}, the pairs of lines marked with comments (A1/A2/B1/B2) race.
+In Figure~\ref{fig:hb-example}, the pairs of lines marked with comments (A1 and A2, B1 and B2) race.
 

### 2016-05-13 proposal/thesprop.tex
lol moores law
write about quicksand motiv and deisgn
add overview for DRs
some minor changes, subtitle

@@
 \usepackage{subfigure}
+\usepackage[linesnumbered,ruled]{algorithm2e}
+\usepackage{algpseudocode}
 
@@
 
+\newcommand\simics{\textsc{Simics}}
+\newcommand{\sect}[1]{\S #1}
+\newcommand\hilight[2]{\color{#1}#2\color{black}}
+\definecolor{orange}{RGB}{192,96,0}
+\definecolor{olivegreen}{RGB}{0,127,0}
+\definecolor{brickred}{RGB}{192,0,0}
+\definecolor{commentblue}{RGB}{0,0,192}
+
 \begin {document}
@@
 \title{ %% {\it \huge Thesis Proposal}\\
-{\bf Practical Concurrency Testing}}
+{\bf Practical Concurrency Testing} \\
+\normalsize \vspace{1em}
+{\em or, How I Learned to Stop Worrying and Love the Exponential Explosion}}
 \author{Ben Blum}
@@
 
-Modern computer architectures have turned to increasing CPU core count, rather than clock speed, to improve processing power.
+Modern computer architectures have turned to increasing CPU core count, rather than clock speed, to improve processing power \cite{mooreslaw}.
 %To take advantage of multiple cores, programs must be written {\em concurrently}

### 2016-05-13 proposal/related.tex
start writing related work

\chapter{Related Work}
\label{chap:related}

In this chapter I provide a brief tour of the related work.

\section{History of Stateless Model Checking}

{\bf Hall of Fame.}
Stateless model checking dates back to Verisoft \cite{verisoft}, the 1997 tool which first attempted to exhaustively explore the possibile ways to interleave threads.
Since then, researchers have built many tools along the same lines to test many kinds of programs.
%list of mcs follows
The most popular model checker, according to citation count, is Microsoft Research's CHESS \cite{chess},
a checker for userspace C++ programs which preempts at synchronization APIs.
%which introduced the Iterative Context Bounding technique \cite{chess-icb} for heuristically ordering the search to uncover bugs faster.
%
Other checkers include MaceMC \cite{macemc}, MoDist \cite{modist}, SAMC \cite{samc}, ETA \cite{dbug-retreat},
each of which limit thread communication to the boundaries of a message-passing API;
%
dBug \cite{dbug-ssv}, another CMU original, similar to CHESS;
%
and finally SPIN \cite{spin} and Inspect \cite{inspect} which can preempt at any shared variable access.
%
Perhaps by now, Landslide itself deserves a spot on this list: I first introduced it in my MS thesis \cite{landslide},
and it is most notable for testing kernel-level code through the use of a simulator.

{\bf Theoretical advances.}
% TODO

\section{Search Strategies}

% TODO
%etc..

\section{Educational Testing Tools}

\section{Real-World Techniques}

### 2016-09-16 proposal/related.tex
write about reduction techniques

@@
 
-{\bf Theoretical advances.}
+{\bf Search Strategies}
+To date a number of techniques have been proposed to mitigate exponential explosion, the Sisyphean rock of stateless model checkers.
+Dynamic Partial Order Reduction (DPOR) \cite{dpor}, the baseline approach, prunes redundant interleavings by identifying independent thread transitions which can commute without changing the program state.
+It is a {\em sound} reduction algorithm, meaning it will never fail to test a possible program behavior, despite skipping many execution sequences.
+DPOR has since been extended in several ways:
+a provably optimal version which guarantees to explore no more than one interleaving from each equivalence class \cite{optimal-dpor},
+% TODO: a 3rd example?
+and an extension of the algorithm to include nondeterminism arising from weak memory models \cite{tsopso}.
+Recently, SATCheck and Maximal Causality Reduction have emerged as better-performing alternatives to DPOR.
+These algorithms use SMT solvers \cite{z3} to identify additional equivalences by analyzing the values read and written by each memory access.
+
+A number of other techniques address the problem that even with DPOR, a state space may be too large.
+Iterative Context Bounding (ICB) \cite{chess-icb} is a search ordering heuristic which prioritizes interleavings with fewer preemptions,
+according to the insight that most bugs require fewer preemptions to expose.
+Bounded Partial Order Reduction \cite{bpor} adapts DPOR to be soundly compatible with ICB.
+Preemption Sealing \cite{sealing} allows programmers to exclude preemption points arising from trusted source code modules,
+reducing the state space by limiting the scope of the test.
+Probabilistic Concurrency Testing \cite{randomized-scheduler} eschews DPOR's depth-first approach entirely (as well as any potential for completing the state space), randomly sampling a broad cross-section of the state space and providing probabilistic bounds on uncovering bugs.
+
+{\bf Other theoretical advances.}
 % TODO
-
-\section{Search Strategies}
-
-% TODO
-%etc..
 

### 2016-09-20 proposal/overview.tex
write about concurrency bug basics

@@
 \section{Concurrency Bugs}
+
+\subsection{The Basics}
+
+Modern software often turns to multithreading to improve performance.
+In a multithreaded program, multiple execution units (or {\em threads}) execute the same or different sections of code simultaneously.
+
+{\bf Simultaneity.}
+This simultaneity of threads is achieved either by executing each one on a separate CPU, or by interleaving them nondeterministically (as controlled by clock interrupts) on the same CPU.
+Because clock interrupts can occur at any instruction\footnote{
+	With some exceptions in kernel-level programming, which we discuss later.
+},
+we consider single-CPU multithreading to be simultaneous at the granularity of individual instructions.
+%
+Likewise, when multiple CPUs access the same memory,
+hardware protocols generally ensure that the events of a single instruction are executed atomically from the perspective of all CPUs.
+Although there are some exceptions --
+unlocked memory-to-memory instructions,
+unaligned writes \cite{unaligned-writes},
+and weak memory consistency models \cite{memory-consistency-models} --
+we model multicore concurrency the same way as above,
+deferring these exceptions beyond the scope of this work.
+We refer to an execution trace depicting the global sequence of events as a {\em thread interleaving} or {\em schedule}.
+
+{\bf Shared state.}
+% TODO cite more languages / type theories
+When a programming language offers multithreaded parallelism but forbids access to any shared state between threads \cite{rust-language},
+the simultaneity of threads is largely irrelevant to the program's behaviour.
+However, ``thread-unsafe'' languages such as C, C++, Java, and so on remain popular,
+in which threads may access global or heap-allocated variables and data structures with no enforced access discipline.
+The behaviour of such programs is then subject to the manner in which these accesses interleave.
+
+{\bf Identifying bugs.}
+Even if a program's behaviour is nondeterministic, that does not necessarily mean it has a bug.
+After all, many programs use random number generation to intentionally generate different outputs.
+We say a {\em concurrency bug} occurs when one or more of a program's nondeterministic behaviours is both {\em unanticipated} and {\em undesired}.
+Most often, a concurrency novice who programs with shared state will consider the possible interleavings where one thread's access sequence occurs entirely before the other's, but neglect to consider intermediate outcomes in which the threads' access sequences are interleaved.
+%
+Consider the program in Figure \ref{fig:concurrency-bug}: Any output between 2 and 2000 is possible\footnote{
+	Exercise for the reader: Show why 2 is a possible output, but 1 is not!
+},
+but whether this constitutes a bug is a matter of perspective.
+Was the program written to count to 2000, or was it written to compute a randomized distribution?
+In this thesis, we make no attempt to reason about the ``intent'' of programs,
+so we further restrict {\em concurrency bug} to denote a program behaviour which is mechanically identifiable,
+according to subjective notion of what programs behaviours are always bad.
+%
+Bug conditions include assertion failures,
+memory access errors (i.e., segmentation fault or bus error),
+heap errors (i.e., use-after-free or overflow),
+deadlocks,
+and infinite loops (which must be identified heuristically \cite{entscheidungsproblem}).
+
+\begin{figure}[t]
+	\begin{tabular}{cc}
+		\begin{tabular}{p{0.45\textwidth}p{0.5\textwidth}}
+			{\footnotesize
+			\begin{tabular}{l}
+				\texttt{int x;} \\
+				\texttt{void count() \{} \\
+				\texttt{~~~~for (int i = 0; i < 1000;} \\
+				\texttt{~~~~~~~~~i++)} \\
+				\texttt{~~~~~~~~x++;} \\
+				\texttt{\}} \\
+				\texttt{void main() \{} \\
+				\texttt{~~~~tid1 = thr\_create(count);} \\
+				\texttt{~~~~tid2 = thr\_create(count);} \\
+				\texttt{~~~~thr\_join(tid1);} \\
+				\texttt{~~~~thr\_join(tid2);} \\
+				\texttt{~~~~printf("\%d\textbackslash{}n", x);} \\
+				\texttt{\}} \\
+			\end{tabular}
+			}
+			&
+			{\footnotesize \begin{tabular}{ll}
+				{\bf \normalsize Thread 1} & {\bf \normalsize Thread 2} \\
+				\hline
+				\texttt{load tmp <- x;} & \\
+				& \texttt{load tmp <- x;} \\
+				& \texttt{add tmp <- 1;} \\
+				& \texttt{store x <- tmp;} \\
+				\texttt{add tmp <- 1;} & \\
+				\texttt{store x <- tmp;} & \\
+			\end{tabular}
+			}
+			\\
+			\\
+			(a) Source listing for a multithreaded program which might count to 2000.
+			&
+			(b) Example interleaving of the compiled assembly for (a),
+			in which 2 concurrent iterations of the loop yield 1 net increment of {\tt x}.
+		\end{tabular}
+	\end{tabular}
+	\caption{Example concurrent program in which simultaneous accesses to shared state may interleave to produce unexpected results.}
+	\label{fig:concurrency-bug}
+\end{figure}
+
+\subsection{Concurrency Primitives}
+
 % TODO

### 2016-09-20 proposal/overview.tex
some more changes

@@
 Static analyses are ideal for ensuring certain standards of code quality, which often correlates with correctness,
-but cannot decide for certain whether a given program will fail during execution without actually running the code.
-% TODO: cite incompleteness theorem
+but cannot decide for certain whether a given program will fail during execution without actually running the code \cite{incompleteness}.
 Static analyses face the challenge of {\em false alarms} (or {\em false positives}):
@@
 Intuitively, if two thread transitions between preemption points do not conflict on any shared resource access, reordering them produces an equivalent interleaving, i.e., the same program behaviour.
-% TODO: put fig dpor
 Figure~\ref{fig:dpor} shows a part of an execution tree in which the operations by threads 1 and 2 are independent, and hence only one of the subtrees need be tested.
 Over the years, researchers have developed many enhancements to DPOR, such as Optimal DPOR \cite{optimal-dpor}, parallelizable DPOR \cite{parallel-dpor}, SAT-directed model checking \cite{satcheck}, Maximal Causality Reduction \cite{mcr}, and DPOR for relaxed memory architectures \cite{tsopso}.
+
+\begin{figure}[t]
+	\begin{center}
+	\includegraphics[width=0.4\textwidth]{dpor.pdf}
+	\end{center}
+	\caption{DPOR identifies independent transitions by different threads which can commute without affecting program behaviour.
+Here, if the transitions marked 1 and 2 have no shared memory conflicts, the states marked with the red arrow are guaranteed identical.
+Hence, only one of the subtrees need be explored.}
+	\label{fig:dpor}
+\end{figure}
 
@@
 \\
+\\
 	{\normalsize (a) True potential data race.}
@@
 \end{tabular}
+\\
 \\

### 2016-09-28 landslide-lecture/2016/lecture.tex
import 2015 landslide lecture to new 2016 dir
Update lecture for f16

--- 2016-02-10 landslide-lecture/2015/lecture.tex
@@
 \institute[CMU 15-410]{Carnegie Mellon University - 15-410}
-\date[]{2016, February 10}
+\date[]{2016, September 28}
 
@@
 		\item Preempt enough times to run {\em every single} interleaving
-		\item Tell the story of what {\em actually happened}.
+		%\item Tell the story of what {\em actually happened}.
 		\item Overlook fewer bugs!
@@
 		\item At end of each branch, rewind system and restart test.
-		\item Artificially add preemptions to produce different thread interleavings.
+		\item Artificially preempt to interleave threads differently.
 		\item Intuitively: Generate many ``tabular execution traces''.
@@
 	\begin{itemize}
-		\item If there are $n$ PPs and $k$ threads, state space size is $n^k$.
-		\item Need to choose the set of PPs very carefully for test to be effective.
+		\item $n$ PPs and $k$ threads $\Rightarrow$ state space size is $n^k$.
+		\item Need to choose PPs very carefully for test to be effective.
 		\begin{itemize}
@@
 	\linegap
-	When do we fear being preempted?
+	When are preemptions dangerous?
 	\begin{itemize}
@@
 \begin{frame}{About The Project}
-	\textbf{About me}: 5th year graduate student, advised by Garth Gibson
+	\textbf{About me}: 6th year graduate student, advised by Garth Gibson
 	\begin{itemize}
 		\item TAed 15-410 for 3 semesters during undergrad
-		\item 1st graduate year was 5th year M.S., rest in Ph.D. program
+		%\item 1st graduate year was 5th year M.S., rest in Ph.D. program
+		\item Landslide's publication history
+		\begin{itemize}
+			\item Master's thesis
 			\begin{itemize}
 				\item \url{http://www.contrib.andrew.cmu.edu/~bblum/thesis.pdf}
+			\end{itemize}
+			\item Conference paper (to appear in OOPSLA this november)
+			\begin{itemize}
+				\item \url{http://www.contrib.andrew.cmu.edu/~bblum/oopsla.pdf}
+			\end{itemize}
 		\end{itemize}
@@
 			% FIXME for next year -- how much slowdown is there? 20x?
-		\item Originally supported only P3s; can now test P2s fully-automated
+		%\item Originally supported only P3s; can now test P2s fully-automated
 		\item {\em Landslide} shows how your {\em Pebbles} programs may not be stable.
@@
 
-	State of the art tools hard-code a fixed set of preemption points.
-	\begin{itemize}
-		\item E.g., ``all thread library API calls'' or ``all kernel mutex locks/unlocks''
+	State of the art tools choose a fixed set of preemption points.
+	\begin{itemize}
+		\item E.g., ``all thread API calls'' or ``all kernel mutex locks/unlocks''
 		\item Depending on length of test, completion time is unpredictable.
@@
 	Goal: Run the best tests for a given CPU budget.
-	\begin{itemize}
-		\item Technique: ``Iterative Deepening''
-	\end{itemize}
-	\linegap
+	%\begin{itemize}
+	%	\item Technique: ``Iterative Deepening''
+	%\end{itemize}
+	%\linegap
 
@@
 
-\begin{frame}{Last Semester}
+\begin{frame}{Previous Semesters}
 	%Spring 2015: 7 groups signed up to use Landslide; 5 found bugs
@@
 	%\end{itemize}
-	Fall 2015: 22 groups signed up to use Landslide; 21 found bugs
+	S'15, F'15, S'16: 48 groups signed up to use Landslide; 38 found bugs
 	% 18 groups found races
-	\begin{itemize}
-		\item (that they didn't find during stress testing)
-	\end{itemize}
+	%\begin{itemize}
+	%	\item (that they didn't find during stress testing)
+	%\end{itemize}
 	%%%% hopefully!
@@
 
-	Among all groups, 185 distinct tests were run
-	\begin{itemize}
-		\item 137 tests ran to completion without finding bugs
-		\item 21 {\em deterministic} bugs
-	\end{itemize}
-	\pause % ask: "who can tell me why landslide might find a *deterministic* bug not found during past testing? A: use-after-free
-	\linegap
-
-	38 distinct {\em non-deterministic} bugs
-	\begin{itemize}
+	44 {\em deterministic} bugs (e.g. swexn, initialization)
+	\linegap
+
+	85 distinct {\em non-deterministic} bugs (among 32 groups)
+	\begin{itemize}
+		\item 24 groups (75\%) fixed $\ge 1$ such bug
+			\begin{itemize}
+				\item (as verified by running Landslide again -- not a guarantee!)
+			\end{itemize}
+		\item 15 groups (47\%) fixed {\em all} such bugs
+		\item Most ambitious group: 11 distinct races found + fixed!
+	\end{itemize}
+	\linegap
+	% 44 dets 85 races
+
+	%Among all groups, 185 distinct tests were run
 	%\begin{itemize}
-			%	\item 7 of which required data-race PPs to expose
+	%	\item 137 tests ran to completion without finding bugs
+	%	\item 21 {\em deterministic} bugs
 	%\end{itemize}
-		\item 11 groups (50\%) fixed all their bugs
-			\begin{itemize}
-				\item (as verified by running Landslide again -- not a guarantee!)
-			\end{itemize}
-		\item 2.7 bugs ``fixed'' (see above) on average per group
-		\item Most ambitious group: 11 distinct races found + fixed!
-	\end{itemize}
+	%\pause % ask: "who can tell me why landslide might find a *deterministic* bug not found during past testing? A: use-after-free
+	%\linegap
+
+	%38 distinct {\em non-deterministic} bugs
+	%\begin{itemize}
+	%		%\begin{itemize}
+	%		%	\item 7 of which required data-race PPs to expose
+	%		%\end{itemize}
+	%	\item 11 groups (50\%) fixed all their bugs
+	%		\begin{itemize}
+	%			\item (as verified by running Landslide again -- not a guarantee!)
+	%		\end{itemize}
+	%	\item 2.7 bugs ``fixed'' (see above) on average per group
+	%	\item Most ambitious group: 11 distinct races found + fixed!
+	%\end{itemize}
 \end{frame}
@@
 			\begin{itemize}
-				\item \url{http://tinyurl.com/landslide-p2-s16}
+				\item \url{http://tinyurl.com/landslide-p2-f16}
 			\end{itemize}
@@
 
-\begin{frame}{Last-last Semester}
-	Spring 2015: 7 groups signed up to use Landslide; 5 found bugs
-	%\begin{itemize}
-	%	\item (that they didn't find during stress testing)
-	%\end{itemize}
-	%%%% hopefully!
-	\linegap
-
-	Among all groups, 38 total tests were run
-	\begin{itemize}
-		\item 21 tests ran to completion without finding bugs
-		\item 4 {\em deterministic} bugs
-		\item 13 distinct {\em non-deterministic} bugs
-			\begin{itemize}
-				\item 6 in {\tt thr\_exit\_join} \small (use-after-free, deadlock, NULL, infinite loop...)
-				\item 4 in {\tt paraguay} (deadlock, infinite loop)
-				\item 2 in {\tt broadcast\_test} (deadlock, infinite loop)
-				\item 1 in {\tt rwlock\_downgrade\_read\_test} (use-after-free)
-			\end{itemize}
-			%\begin{itemize}
-			%	\item 6 of which required data-race PPs to expose
-			%\end{itemize}
-	\end{itemize}
-\end{frame}
+%\begin{frame}{Last-last Semester}
+%	Spring 2015: 7 groups signed up to use Landslide; 5 found bugs
+%	%\begin{itemize}
+%	%	\item (that they didn't find during stress testing)
+%	%\end{itemize}
+%	%%%% hopefully!
+%	\linegap
+%
+%	Among all groups, 38 total tests were run
+%	\begin{itemize}
+%		\item 21 tests ran to completion without finding bugs
+%		\item 4 {\em deterministic} bugs
+%		\item 13 distinct {\em non-deterministic} bugs
+%			\begin{itemize}
+%				\item 6 in {\tt thr\_exit\_join} \small (use-after-free, deadlock, NULL, infinite loop...)
+%				\item 4 in {\tt paraguay} (deadlock, infinite loop)
+%				\item 2 in {\tt broadcast\_test} (deadlock, infinite loop)
+%				\item 1 in {\tt rwlock\_downgrade\_read\_test} (use-after-free)
+%			\end{itemize}
+%			%\begin{itemize}
+%			%	\item 6 of which required data-race PPs to expose
+%			%\end{itemize}
+%	\end{itemize}
+%\end{frame}
 

### 2016-12-25 proposal/lipservice.tex
write htm stuff

@@
 
-In this chapter I propose to add several enhancements to Landslide to allow it to be used beyond CMU's walls,
-with little to no manual instrumentation effort required.
+In this chapter I propose to add several enhancements to Landslide to apply its techniques beyond CMU's walls.
+%with little to no manual instrumentation effort required.
 This is the third of the three projects I am proposing for this thesis and so far exists only in dreams.
+The project will have two parts: adding support for Pintos \cite{pintos}, and adding support for Transactional Memory \cite{transactional-memory}.
+
+\section{Pintos}
+
+\section{Transactional Memory}
+
+Transactional Memory (TM) is a mechanism by which programmers may avoid conventional concurrency primitives, optimizing for performance in the common case when threads do not conflict.
+A transactional program surrounds its critical section(s) with transaction begin/end statements, which ensure that no other thread can observe an intermediate state during the transaction.
+If a conflict is observed, the transaction {\em aborts}, rolling the program back to the transaction's initial state, and executing an optional back-up code path.
+The programmer may also explicitly abort the transaction using an abort statement.
+An example transactional program is shown in Figure~\ref{fig:txn-example}.
+
+\begin{figure}[t]
+	\begin{center}
+	\begin{tabular}{ll}
+		\multicolumn{2}{c}{Initially {\tt int foo, bar = 0; mutex\_t m;}} \\
+		\\
+		\begin{tabular}{l}
+		{\bf Thread 1} \\
+		\hline
+		\texttt{if (\_xbegin() ==} \\
+		\texttt{~~~~~~~~\_XBEGIN\_STARTED) \{} \\
+		\texttt{~~~~foo++;} \\
+		\texttt{~~~~\_xend();} \\
+		\texttt{\} else \{} \\
+		\texttt{~~~~mutex\_lock(\&m);} \\
+		\texttt{~~~~assert(foo > 0 ||} \\
+		\texttt{~~~~~~~~~~~bar > 0);} \\
+		\texttt{~~~~mutex\_unlock(\&m);} \\
+		\texttt{\}} \\
+		\end{tabular}
+		&
+		\begin{tabular}{l}
+		{\bf Thread 2} \\
+		\hline
+		\texttt{if (\_xbegin() ==} \\
+		\texttt{~~~~~~~~\_XBEGIN\_STARTED) \{} \\
+		\texttt{~~~~bar++;} \\
+		\texttt{~~~~\_xend();} \\
+		\texttt{\} else \{} \\
+		\texttt{~~~~mutex\_lock(\&m);} \\
+		\texttt{~~~~assert(foo > 0 ||} \\
+		\texttt{~~~~~~~~~~~bar > 0);} \\
+		\texttt{~~~~mutex\_unlock(\&m);} \\
+		\texttt{\}} \\
+		\end{tabular}
+	\end{tabular}
+	\end{center}
+	\caption{Example transactional program, written using GCC's transactional memory compiler intrinsics \cite{htm-gcc}.
+	%The x86 assembly instructions are named {\tt xbegin}, {\tt xend}, and {\tt xabort}; I have named the functions of this imaginary C interface the same.
+	Different behaviours are possible depending whether the transactions are backed by HTM or STM.}
+	\label{fig:txn-example}
+\end{figure}
+
+TM may be implemented either in hardware (HTM) \cite{htm-haswell}, or in software (STM) \cite{stm-pldi06}.
+Though their interfaces to the programmer are similar, their semantics demand a slightly different treatment from Landslide's perspective.
+The key difference is that HTM transactions may fail for any reason, beyond the scope of the program's behaviour, such as the CPU's cache being too full.
+STM transactions, on the other hand, will fail only if an actual conflict is observed from another thread.
+
+Consider again the example program: The transactions of the two threads do not conflict, so they may abort only under HTM.
+However, when they abort for a reason other than a conflict on {\tt foo} or {\tt bar}, the assertions in the backup code will fail.
+Hence, some programs which are correct under STM may contain bugs under HTM.
+Supporting TM in Landslide will consist of several steps, extending the concurrency model to incorporate failure injections and extending DPOR to determine when transaction aborts are possible depending on the underlying TM mechanism.
+
+\subsection{HTM}
+
+{\bf Mutex isomorphism.}
+When modelling TM in Landslide, we do not care about fidelity to performance characteristics or non-observable roll-back semantics.
+The goal of model checking is to exercise all observable program behaviours,
+so Landslide can model the execution of transactional programs using existing primitives if possible.
+In the first stage of the project, I will prove that a transactional program using is equivalent to one with a global mutex swapped for its {\tt xbegin}s and {\tt xend}s (assuming a retry-only policy for handling aborts).
+This will allow Landslide to test all observable TM behaviours using its existing infrastructure for mutexes,
+rather than relying on the platform providing accurate TM emulation.
+
+{\bf Abort nondeterminism.}
+Next, I will extend the concurrency model to support the nondeterminism arising from transaction aborts.
+During execution, Landslide inject a failure to force threads to branch into backup code paths.
+Failure injections add an extra ``dimension'' of non-determinism:
+at each {\tt xbegin} operation which is a preemption point, Landslide may force a normal context switch to re-interleave threads, or it may inject a transaction abort to test the backup code.
+(This also avoids the need to speculatively execute and/or roll-back failing transactions.)
+Because HTM transactions can fail for any reason, outside of the programmer's control, Landslide's DPOR implementation will consider failure injections always ``enabled''.
+
+{\bf Reduction challenge.}
+I've identified a case under abort nondeterminism in which the current implementation of DPOR will fail to achieve a possible reduction.
+Let $1A$, $1B$, $2A$, and $2B$ denote the transitions which execute the {\tt if} and {\tt else} branches by each thread, respectively.
+Note that $1A$ conflicts with $2B$, and $1B$ with $2A$, but $1A$ and $2A$ are independent, as are $1B$ and $2B$.
+Figure~\ref{fig:txn-graph} provides a visual aid.
+There are two possible reductions: pruning $2A,1A$ after testing $1A,2A$, and pruning $2B,1B$ after testing $1B,2B$;
+in other words, branches 5 and 8 can be skipped.
+However, %if we simply extended DPOR to always inject failures when possible,
+our current DPOR implementation will tag the $2B$ subtree to explore after observing the conflict in branch 2 (likewise the $2A$ subtree from branch 3),
+but then have no memory of whether it was subsequently supposed to run $1A$ or $1B$, and have to try both.
+To fix this will require an optimization analogous to ``sleep sets'' \cite{dpor}:
+the tag to explore the $2A$ or $2B$ subtree must include an annotation to remember whether or not the conflict arose after a failure injection in $1$.
+
+\begin{figure}[t]
+	\begin{center}
+		\includegraphics[width=0.8\textwidth]{htm-graph.pdf}
+	\end{center}
+	\caption{State space corresponding to the program in Figure~\ref{fig:txn-example}.
+	Conflicting transitions are marked in bold;
+	transitions which are independent from their predecessors are italicized.
+	End states colored the same (the cyan and yellow ones) are equivalent.
+	}
+	\label{fig:txn-graph}
+\end{figure}
+
+\subsection{STM}
+
+STM transactions abort only when multiple threads conflict.
+Because Landslide already computes memory conflicts among each pair of transitions, it will be natural to extend DPOR to consult the conflict set when deciding whether to exercise a failure injection.
+I will extend Landslide in this manner, and even support programs with both types of transactions, as long as the different {\tt xbegin} invocations are suitably annotated.
+
+\subsection{Hybrid HTM/STM}
+
+A recent paper \cite{hybrid-htm-stm} introduced several ways of combining HTM and STM in the same program, nesting transactions of different types.
+It presented several semantics for such transactions transactions, the most interesting of which being ``open nesting'', in which a nested transaction's state becomes visible to other threads even during a containing transaction.
+That state can then be rolled back if the latter aborts.
+I plan to develop a theoretical model for how such transactions would affect Landslide's concurrency model,
+although I expect to relegate any implementation thereof to future work.

### 2016-12-25 proposal/overview.tex
write htm stuff

@@
 % TODO
+
+\subsection{Transactional Memory}
+
+Critical sections of code must be protected from concurrent access, even when it's not known in advance whether the shared memory accesses between threads will actually conflict on the same memory addresses.
+The concurrency primitives discussed above take a pessimistic approach, imposing a uniform performance penalty (associated with the primitives' implementation logic) on all critical sections, whether or not a conflict is likely.
+Some implementations may be optimized for ``fast paths'' in the absence of contention, but must still access shared memory in which the primitive's state resides.
+
+Transactional memory \cite{transactional-memory} offers a more optimistic approach: critical sections of code are marked as ``transactions'', analogously to locking a mutex, and allowed to speculatively execute with no protection.
+If a conflict between transactions is detected, the program state is rolled back to the beginning of the transaction, and a backup code path may optionally be taken.
+Consequently, no intermediate state of a transacting thread is ever visible to other threads; all changes to memory within a transaction become globally visible ``all at once'' (or not at all).
+This method optimizes for a common no-contention case of little-to-no overhead, pushing extra both code and implementation complexity to handling conflicts.
+
+Transactional memory may be implemented either in hardware, using special instructions and existing cache coherence algorithms, or in software, via library calls and a log-based commit approach.
+Software transactions (STM) \cite{stm-pldi06} can be used on any commodity processor, but must impose some runtime overhead associated with the logs.
+Hardware transactions (HTM) \cite{htm-experience, htm-performance} achieve better performance by reusing existing cache coherence logic to detect conflicts, but require explicit support from the CPU, which is not yet widespread.
+Haswell \cite{htm-haswell} is the first x86 architecture to support HTM,
+offering three new instructions: \texttt{xbegin}, \texttt{xend}, and \texttt{xabort}, to begin, commit, and fail a transaction, respectively.
+
+In this thesis, we will focus on HTM as our platform for testing transactional programs, to highlight the importance of researching advanced testing techniques in anticipation of upcoming hardware features.
+In principle, however, our treatment is agnostic to whether a transactional interface is backed by HTM or STM, so will apply in theory to both.
 

### 2016-12-25 proposal/thesprop.tex
start writing related work
write htm stuff

@@
 %Finally, I will present several new techniques that allow stateless model checking to be practically employed on real-world programs.
-Finally, I will explore broader impact by extending Landslide to test real-world programs and to be used by students at other universities.
+Finally, I will explore broader impact by extending Landslide to test some real-world programs and to be used by students at other universities.
 \end{abstract}
@@
 		So far, a fully-automatic testing mode is available only for 15-410 thread library projects.
-		To prove these techniques are relevant beyond CMU's walls, I will extend Landslide to handle both Pintos kernel projects from other universities \cite{pintos} and ``real-world'' Linux programs.
-		I will conduct a user study in which students at another university test their Pintoses with Landslide,
-		and evaluate Landslide's ability to find known and/or new bugs in real-world programs.
+		To prove these techniques are relevant beyond CMU's walls, I will extend Landslide to handle Pintos kernel projects from other universities \cite{pintos}. %and ``real-world'' Linux programs.
+		%I will conduct a user study in which students at another university test their Pintoses with Landslide,
+		%and evaluate Landslide's ability to find known and/or new bugs in real-world programs.
+		I will also extend Landslide's concurrency model to support Hardware Transactional Memory (HTM), a new concurrent programming technique \cite{transactional-memory}, and test several ``real-world'' HTM programs.
 
 		This will be {\bf mostly new work}, as Landslide presently supports testing Pintoses only with some manual effort,
-		and does not yet feature instrumentation to run on Linux programs.
+		%and does not yet feature instrumentation to run on Linux programs.
+		and will need its concurrency model extended to incorporate the HTM programming model.
 \end{itemize}
@@
 \input{lipservice}
+\input{related}
 

### 2016-12-28 proposal/lipservice.tex
write about pintos

@@
 \section{Pintos}
+
+The Pintos architecture \cite{pintos}, outlined in Section~\ref{sec:overview-pintos}, provides an opportunity to test Landslide's pedagogical mettle beyond CMU's walls.
+For the experiments in Chapter~\ref{chap:quicksand}, I already implemented rudimentary support to test Pintos kernels.
+Because Pintos provides substantially more basecode than Pebbles, most if not all of Landslide's kernel instrumentation can be done automatically,
+using the names of the core scheduler functions already provided.
+
+{\bf Stuff I already did.}
+The existing Pintos support includes
+a new set of annotations (e.g., {\tt mutex\_lock} in Pebbles is now called {\tt sema\_down};
+handling several quirks of the basecode's scheduling behaviour (ask me if you really care to know);
+extending the heap tracker to handle Pintos's page allocator {\tt palloc} as well as the kernel {\tt malloc},
+as well as the fact that kernel memory is not direct-mapped like in Pebbles;
+new thread-liveness code to detect when a test case terminates;
+skipping some busy-wait loops in the boot sequence used for device communication,
+a complicated fall-back case in Landslide's timer injection algorithm the workings of which I don't even remember anymore,
+and extending the lock-set analysis to include Pintos's interrupt-disabling infrastructure (which was actually pretty tricky when using vector clocks for pure-HB).
+All these features apply to behaviours included with the stock base-code, so should apply generally to all but the most adventurous student implementations.
+
+{\bf Stuff I still need to do.}
+However, as you might have guessed, there are still a few cases where the existing instrumentation is not fully general.
+When preparing Chapter~\ref{chap:quicksand}'s experiments, I needed to adjust some student submissions to be compatible with Landslide's existing annotation process.
+In order to distribute Landslide for student use beyond CMU, it will need to support these cases automatically.
+These include:
+\begin{itemize}
+	\item Allowing for the priority scheduler's runqueues to be implemented as an array of queues, rather than the single queue head as provided in the basecode.
+	\item Insert the {\tt tell\_landslide\_sleeping} annotation more intelligently than looking for a {\tt list\_insert\_ordered} call (a basecode function which most, but not all, students use).
+	\item Some versions of the basecode distribute a {\tt sema\_up} implementation which {\tt yield}s unconditionally, which Landslide must bypass.
+	\item Some students use {\tt timer\_sleep} or {\tt while (!flag) continue} when they should use {\tt yield}, which can lead to false-positive deadlocks if not automatically replaced.
+	\item I'm sure I've forgotten some less-common ones, which I expect to handle on-the-fly when students email me Landslide crash reports.
+\end{itemize}
+When fixing my sample of Pintoses by hand, there were also myriad deterministic bugs, such as use-after-frees e.g.~arising from unsafe {\tt strlen} calls.
+As with Pebbles student experiments, however, I intend the students to encounter such bug reports and fix them on their own, despite not being concurrency bugs.
+It would be a Ph.D. unto itself to automatically ensure stable determinized execution of arbitrary student code.
+Actually, probably outright impossible.
+
+{\bf Experimental goals.}
+Having only one semester (ideally; or two, if my timeline ends up delayed) to test Landslide with Pintos students,
+I'll not be able to assemble as comprehensive a dataset as I have with 15-410 at CMU.
+This project will serve more as a proof-of-concept, demonstrating that model checking can realistically replace stress testing in operating systems courses worldwide.
+Nevertheless, I intend also to analyze what little data I can gather, comparing the incidences and types of concurrency bugs found between Pebbles and Pintos populations.
+This will perhaps shed light on the advantages and/or shortcomings of either project,
+leading to recommendations for improving either project's educational value (independently of using MC).
 

### 2016-12-28 proposal/lipservice.tex
oh yeah, the most important part

@@
 
+{\bf New emulation platform.}
+Most importantly, I will need to free Landslide from its dependence on Simics, which requires paid licenses for use beyond CMU's walls.
+Other candidate emulation (simulation or virtualization) platforms for Landslide include Bochs, QEMU, VMWare, and Xen.
+I am presently studying which of these will most suit Landslide's needs,
+but so far Bochs seems like the best bet, which happens to be the simulator used at Berkeley, Stanford, and U. of Chicago anyway.
+
 {\bf Experimental goals.}
 Having only one semester (ideally; or two, if my timeline ends up delayed) to test Landslide with Pintos students,
-I'll not be able to assemble as comprehensive a dataset as I have with 15-410 at CMU.
-This project will serve more as a proof-of-concept, demonstrating that model checking can realistically replace stress testing in operating systems courses worldwide.
-Nevertheless, I intend also to analyze what little data I can gather, comparing the incidences and types of concurrency bugs found between Pebbles and Pintos populations.
+I'll be unable to gather as comprehensive a dataset as I have with 15-410 at CMU.
+This will serve more as a proof-of-concept, demonstrating that MC can realistically replace stress testing in operating systems courses worldwide.
+Nevertheless, I intend also to analyze what little data I can gather, comparing the incidences and types of bugs found between Pebbles and Pintos populations.
 This will perhaps shed light on the advantages and/or shortcomings of either project,

### 2016-12-30 proposal/overview.tex
some updates, i vs we pronouns, etc

@@
 This chapter will introduce the necessary background material on concurrency, stateless model checking, data-race analysis, and the relevant undergraduate operating systems classes.
+
+Thesis proposal note: I've written this section with the intent of reusing it in the thesis.
+The other chapters, being necessarily more impermanent, are aimed directly at you (my thesis committee) as the audience.
+This section is written with the intention for programmers of any level to read.
 
@@
 Because clock interrupts can occur at any instruction\footnote{
-	With some exceptions in kernel-level programming, which we discuss later.
+	With some exceptions in kernel-level programming, which I discuss later.
 },
@@
 {\bf Shared state.}
-% TODO cite more languages / type theories
 When a programming language offers multithreaded parallelism but forbids access to any shared state between threads \cite{rust-language},
@@
 so we further restrict {\em concurrency bug} to denote a program behaviour which is mechanically identifiable,
-according to subjective notion of what programs behaviours are always bad.
+according to commonly-accepted notions of what programs behaviours are always bad.
 %
@@
 
-Transactional memory may be implemented either in hardware, using special instructions and existing cache coherence algorithms, or in software, via library calls and a log-based commit approach.
-Software transactions (STM) \cite{stm-pldi06} can be used on any commodity processor, but must impose some runtime overhead associated with the logs.
+Transactional memory (TM) may be implemented either in hardware, using special instructions and existing cache coherence algorithms,
+or in software, via library calls and a log-based commit approach.
+Software transactions (STM) \cite{stm-pldi06} can be used on any commodity processor, but must impose runtime overhead associated with logging.
 Hardware transactions (HTM) \cite{htm-experience, htm-performance} achieve better performance by reusing existing cache coherence logic to detect conflicts, but require explicit support from the CPU, which is not yet widespread.
@@
 
-In this thesis, we will focus on HTM as our platform for testing transactional programs, to highlight the importance of researching advanced testing techniques in anticipation of upcoming hardware features.
-In principle, however, our treatment is agnostic to whether a transactional interface is backed by HTM or STM, so will apply in theory to both.
+In this thesis, I will focus on HTM as my platform for testing transactional programs, to highlight the importance of researching advanced testing techniques in anticipation of upcoming hardware features.
+My treatment of TM will be largely agnostic to whether a transactional interface is backed by HTM or STM,
+and I will extend it to cover STM as well, as discussed in Section~\ref{sec:tm}.
 

### 2016-12-30 proposal/thesprop.tex
ground transportation
some updates, i vs we pronouns, etc

@@
 
-\keywords{landslide terminal, baggage claim, public transportation, ticketing}
+\keywords{landslide terminal, baggage claim, ground transportation, ticketing}
 
@@
 \begin{center}
+	% TODO: lol
 	{\em my research is cool. please let me graduate.}
@@
 
-		This work is {\bf largely completed} and is currently under submission at OOPSLA 2016.
+		This work is {\bf completed} and was published at OOPSLA 2016.
 	\item {\bf Educational use (Chapter \ref{chap:410}).}
@@
 
-		This is {\bf ongoing work}; I have run the user study for 3 semesters so far and am proposing to continue them and follow them up with an evaluation.
+		This is {\bf ongoing work}; I have run the user study for 4 semesters so far and am proposing to continue them and follow them up with an evaluation.
 	\item {\bf Broader impact (Chapter \ref{chap:lipservice}).}

### 2016-12-30 proposal/overview.tex
write overview of mutexes

@@
 
-% TODO
+To prevent unexpected interleavings such as the example in Figure~\ref{fig:concurrency-bug}(b),
+most concurrent programs use {\em concurrency primitives} to control which interleavings are possible.
+Controlling nondeterminism is not typically provided by any features of programming languages themselves;
+rather, it is achieved via special atomicity mechanisms provided by the CPU and/or operating system -- hence the term ``primitive''.
+For example, x86 CPUs provide the {\tt xchg} instruction, which performs both a read and subsequent write to some shared memory, with no possibility for other logic to interleave in between.
+Using such atomic instructions as building blocks, concurrency libraries provide abstractions for controlling nondeterminism in several commonly-desired ways.
+These include {\em locks}, {\em descheduling}, {\em condition variables}, {\em semaphores}, {\em reader-writer locks}, and {\em message-passing}.
+
+\begin{figure}[t]
+	\begin{tabular}{p{0.45\textwidth}p{0.5\textwidth}}
+		{\footnotesize
+		\begin{tabular}{l}
+			\texttt{typedef struct mutex \{} \\
+			\texttt{~~~~volatile int held;} \\
+			\texttt{~~~~int owner;} \\
+			\texttt{\} mutex\_t;} \\
+			\texttt{void mutex\_lock(mutex\_t *mp) \{}\\
+			\texttt{~~~~while (xchg(mp->held, 1))} \\
+			\texttt{~~~~~~~~yield(mp->owner);} \\
+			\texttt{~~~~mp->owner = gettid();} \\
+			\texttt{\}} \\
+			\texttt{void mutex\_unlock(mutex\_t *mp) \{}\\
+			\texttt{~~~~mp->owner = -1;} \\
+			\texttt{~~~~mp->held = 0;} \\
+			\texttt{\}} \\
+		\end{tabular}
+		}
+		&
+		{\footnotesize
+		\begin{tabular}{l}
+			\texttt{int x;} \\
+			\texttt{mutex\_t m;} \\
+			\texttt{void count() \{} \\
+			\texttt{~~~~for (int i = 0; i < 1000;} \\
+			\texttt{~~~~~~~~~i++)} \\
+			\texttt{~~~~~~~~mutex\_lock(\&m);} \\
+			\texttt{~~~~~~~~x++;} \\
+			\texttt{~~~~~~~~mutex\_unlock(\&m);} \\
+			\texttt{\}} \\
+		\end{tabular}
+		}
+		\\
+		(a) A simple mutual exclusion lock built using the {\tt xchg} instruction. %(accessed using a GCC compiler intrinsic).
+		&
+		(b) The {\tt count} function from Figure~\ref{fig:concurrency-bug}, adjusted to use a mutex to ensure each increment of {\tt x} is uninterruptible.
+	\end{tabular}
+	\caption{Using a locking primitive to protect accesses to shared state.}
+	\label{fig:mutex}
+\end{figure}
+
+Each such abstraction provides certain semantics about what thread interleavings can arise surrounding their use.
+When building a tool for testing concurrent programs,
+one may include some computational understanding of the behaviour of any, or all, of these abstractions.
+Annotating a certain abstraction's semantics treats it as a trusted concurrency primitive in its own right,
+and allows the testing tool to reduce the possible space of interleavings (or the set of false positive data-race candidates reported, etc.),
+at the cost of increasing the implementation and theoretical complexity of the analysis.
+In this thesis, I will consider locks and descheduling to be the only concurrency primitives,
+and assume the others listed above are implemented using those as building blocks (an exercise for the reader \cite{thrlib}).
+
+Locks (or {\em mutexes}, short for ``mutual exclusion locks'') are objects, shared by multiple threads, which allow the programmer to mark certain {\em critical sections} of code that must not interleave with each other.
+When one thread completes a call to {\tt mutex\_lock(mp)}, all invocations by other threads on the same {\tt mp} will wait (or ``block'') until the corresponding {\tt mutex\_unlock(mp)}.
+Figure~\ref{fig:mutex}(a) shows how a yielding mutex (not the best implementation, but the simplest) may be implemented using {\tt xchg},
+and (b) shows how a mutex may be used to fix the example from Figure~\ref{fig:concurrency-bug}.
 
@@
 \subsection{Pintos}
+\label{sec:overview-pintos}
 % TODO

### 2016-12-30 proposal/overview.tex
overview pebbles

@@
 
-{\bf P2.}
-% TODO
+\subsubsection{P2}
+The thread library project \cite{thrlib} has two main components: implementing concurrency primitives, and implementing thread lifecycle and management routines.
+The required concurrency primitives are as follows:
+\begin{itemize}
+	\item Mutexes, with the interface {\tt mutex\_lock(mp)} and {\tt mutex\_unlock(mp)}, whose functionality is described earlier this chapter.
+Students may use any x86 atomic instruction(s) they desire, such as {\tt xchg}, {\tt xadd}, or {\tt cmpxchg}, and/or the {\tt deschedule}/ {\tt make\_runnable} system calls offered by the reference kernel.
+	\item Condition variables, with the interface {\tt cond\_wait(cvp, mp)}, {\tt cond\_signal} {\tt (cvp)}, and {\tt cond\_broadcast(cvp)}.
+{\tt cond\_wait} blocks the invoking thread, ``simultaneously'' releasing a mutex which protects some associated state (atomically, with respect to other calls to signal or broadcast under that mutex).
+		{\tt cond\_signal} and {\tt cond\_broadcast} wake one or all waiting threads.
+		Students must use the {\tt deschedule} and {\tt make\_runnable} system calls to implement blocking (busy-waiting is forbidden), and typically include an internal mutex to protect the condition variable's state as well.
+		The primary challenge of this exercise is ensuring the aforementioned atomicity between {\tt cond\_wait}'s unlock and deschedule, with respect to the rest of the interface.
+	\item Semaphores, with the interface {\tt sem\_wait(sp)} and {\tt sem\_signal(sp)} (sometimes called {\em proberen} and {\em verhogen} in other literature).
+The semaphore can be initialized to any integer value; if initialized to 1, it behaves like a mutex.
+		Students typically implement semaphores using mutexes and condition variables, not using atomic instructions or system calls directly.
+	\item Reader-writer locks (rwlocks), with the interface {\tt rwlock\_lock(rwp, mode)} and {\tt rwlock\_unlock(rwp)}. {\tt mode} may be either {\tt RWLOCK\_READ} or {\tt RWLOCK\_}- {\tt WRITE}.
+		Behaves as mutexes, but multiple readers may access the critical section simultaneously.
+		Students typically implement rwlocks using mutexes and condition variables, not using atomic instructions or system calls directly.
+\end{itemize}
+The interface to each also includes an associated {\tt \_init()} and {\tt \_destory()} function.
 
-{\bf P3.}
-% TODO
+The thread lifecycle/management routines are as follows:
+\begin{itemize}
+	\item {\tt thr\_init(stack\_size)} initializes the thread library, setting a default stack size to be allocated to new threads.
+	\item {\tt thr\_create(child\_func, child\_arg)} spawns a new thread to run the specified function with the specified argument.
+There is a semantic gap between this function and the {\tt thread\_fork} system call (which takes no parameters, makes no changes to the user's address space, and cannot meaningfully be invoked from C code) which students must bridge.
+		Returns an integer thread ID of the newly created thread.
+	\item {\tt thr\_exit(status)} aborts execution of the calling thread, recording an exit status value.
+		The main challenge of this function is to allow another thread to free the memory used for the exiting thread's stack,
+		without risking any corruption as long as the exiting thread continues to run.
+	\item {\tt thr\_join(tid, statusp)} blocks the calling thread until the thread with the specified thread ID exits, then returns, collecting its exit status.
+\end{itemize}
+Other than {\tt thr\_init} (which is necessarily single-threaded), several concurrency errors between any two (or all three) of these functions are very common in student submissions.
+
+Finally, students also implement automatic stack growth using the {\tt swexn} system call, which is not relevant to this thesis.
+
+\subsubsection{P3}
+In P3, students implement a kernel which provides the same system calls shown in Table~\ref{tab:syscalls}, previously provided by the reference kernel.
+Although the internal interfaces are not mandated like they were in P2, all Pebbles kernels must necessarily contain the same abstract components. These include:
+\begin{itemize}
+	\item A round-robin scheduler, including context switching, timer handling, and runqueue management;
+	\item Some approach to locking, often analogous to P2's concurrency primitives (henceforth referred to as ``kernel mutexes''), 
+		and some approach to blocking threads indefinitely;
+	\item A virtual memory implementation, including a program loader;
+	\item Lifecycle management code for creation and destruction of kernel threads and processes;
+	\item Other miscellany such as a suite of fault handlers to ensure no user program can cause the kernel itself to crash.
+\end{itemize}
+Because any combination of system calls or fault handlers can be invoked by user programs simultaneously,
+concurrency bugs can arise from the interaction of any subset of kernel components with each other.
+The most common bugs students face arise from the interaction of some component with itself (e.g., concurrent invocations of {\tt new\_pages}/{\tt remove\_pages} in the same process),
+or from the interaction between an exiting thread and some other thread trying to communicate with it ({\tt vanish} versus, well, anything else, really).
+The most difficult concurrency problem in P3 is that of coordinating a parent and a child task that simultaneously exit:
+when a task completes, live children and exited zombies must be handed off to the task's parent or to the {\tt init} process,
+when the task's parent may itself be exiting;
+meanwhile, threads in tasks that receive new children may need to be awakened from {\tt wait}.
+Careless solutions to this problem are prone to data races or deadlocks.
 

### 2016-12-31 proposal/lipservice.tex
write overview of mutexes
write overview pintos

@@
 The Pintos architecture \cite{pintos}, outlined in Section~\ref{sec:overview-pintos}, provides an opportunity to test Landslide's pedagogical mettle beyond CMU's walls.
-For the experiments in Chapter~\ref{chap:quicksand}, I already implemented rudimentary support to test Pintos kernels.
-Because Pintos provides substantially more basecode than Pebbles, most if not all of Landslide's kernel instrumentation can be done automatically,
+However, the existing scheduler and concurrency primitives which Pintos provides somewhat limits the remaining concurrency-critical code left to write,
+hence limiting the degree to which Landslide can test student submissions.
+This is in contrast to the Pebbles assignments at CMU, in which students must implement all the functionality listed in Section~\ref{sec:overview-pintos}.
+On the other hand, the upside of Pintos providing substantially more basecode than Pebbles is that most, if not all, of Landslide's kernel instrumentation can be done automatically,
 using the names of the core scheduler functions already provided.
+For the experiments in Chapter~\ref{chap:quicksand}, I already implemented rudimentary support to test Pintos kernels, but more remains to be done.
 
@@
 \section{Transactional Memory}
+\label{sec:tm}
 

### 2016-12-31 proposal/overview.tex
write overview pintos

@@
 \label{sec:overview-pintos}
-% TODO
+
+The Pintos kernel architecture \cite{pintos} is used at several universities, including Berkeley, Stanford, and the University of Chicago.
+The Pintos basecode implements a rudimentary kernel, consisting of a context switcher, round-robin scheduler, locking primitives, and program loader.
+upon which students add more features in several projects.
+Most relevant to this thesis, the basecode provides the following functions/libraries, among others:
+\begin{itemize}
+	\item Semaphores (the basic concurrency primitive, implemented using direct scheduler calls): {\tt sema\_up}, {\tt sema\_down}, {\tt sema\_try\_down};
+	\item Locks (which simply wrap a semaphore initialized to 1), {\tt lock\_acquire}, {\tt lock\_}- {\tt release}, {\tt lock\_try\_acquire};
+	\item Condition variables (also implemented using scheduler calls): {\tt cond\_wait}, {\tt cond\_}- {\tt signal}, {\tt cond\_broadcast}, with the same semantics as Pebbles P2 condvars;
+	\item Basic round-robin scheduling facilities: {\tt thread\_block} (a kernel-level analogue to Pebbles's {\tt deschedule}), {\tt thread\_yield}
+	\item Kernel thread lifecycle management, {\tt thread\_create} and {\tt thread\_exit}, including stack space memory management;
+	\item Interrupt and fault handlers;
+	\item A page allocator, {\tt palloc\_get\_page}, {\tt palloc\_get\_multiple}, {\tt palloc\_}- {\tt free\_page}, {\tt palloc\_free\_multiple}
+\end{itemize}
+Both Pebbles and Pintos basecodes offer a standard C library including {\tt malloc}, string-formatting, printing, etc.
+
+Although there is some variety in supplemental assignments, all Pintos courses include three core projects building on the Pintos basecode:
+\begin{itemize}
+	\item {\em Threads}: Students must implement an ``alarm clock'' (analogous to Pebbles's {\tt sleep} system call),
+		a priority scheduling algorithm, and a multi-level feedback queue scheduler.
+		% TODO: Later. Talk about how concurrency testing can only test certain parts of this crap.
+	\item {\em Userprog}: Provided with rudimentary virtual memory and ELF loader implementations, students must implement argument passing and several system calls associated with userspace programs, including {\tt exec}, {\tt exit}, {\tt wait}, and file descriptor management.
+	\item {\em Filesys}: Provided with a simple ``flat'' filesystem implementation, students must extend it with a buffer cache, extensible files, and subdirectories.
+\end{itemize}
+
+Some schools further offer a virtual memory project, extending the provided VM with a frame table and supplemental page table and fault handler \cite{standford-cs140,uchicago-cs230}, or supplemental HTTP server and {\tt malloc} assignments \cite{berkeley-cs162}.
+Being largely architectural/algorithmic projects rather than concurrency-oriented ones, I am not concerned with these assignments in this thesis.
+The main concurrency challenges in Pintos projects arise from the threads and userprog assignments:
+implementing a correct {\tt alarm} routine,
+ensuring the priority scheduler remains safe in the presence of concurrent threads of the same priority,
+and designing correct interactions between the {\tt wait} and {\tt exit} system calls.

### 2017-01-03 proposal/quicksand.tex
quicksand verif

@@
 In this chapter, I will present a new algorithm for model checkers to automatically choose which preemption points to use to test reasonably-sized state spaces.
-This is the first of the three projects I am proposing for this thesis, and has largely been completed and is under submission.
+This is the first of the three projects I am proposing for this thesis, and has already been published at OOPSLA 2016 \cite{quicksand}.
+The thesis will largely restate the paper, and also go into more detail on some ideas which didn't make it into the paper, such as measuring the variance in each experimental data series.
+Here in the proposal I will simply summarize the content of the paper.
 
 \section{Motivation}
+\label{sec:qs-motivation}
 
 This project is motivated by my experience contributing to Parrot \cite{parrot},
-a partially-determinizing runtime for multithreaded programs which allowed some nondeterminism in performance-critical sections to achieve high performance.
+a (partially) determinizing runtime for multithreaded programs which allowed some nondeterminism in performance-critical sections to achieve high performance.
 We combined Parrot with dBug \cite{dbug-ssv}, a model checker, to check the correctness of the nondeterministic sections of programs which remained.
@@
 
-\begin{algorithm}[t]
+\begin{algorithm}[h]
         \SetKwInOut{Input}{Input}
@@
 \newcommand\AllJobs{\ensuremath{\mathcal{J}}}
-\begin{algorithm}[t]
+\begin{algorithm}[h]
         \SetKwInOut{Input}{Input}
@@
 
+\subsection{Convergence of Iterative Deepening}
+When the a test's state spaces are small enough that Quicksand can exhaustively check all of them within a given CPU budget,
+the resulting verification claim turns out to be equivalent to a na\"{i}model checker which preempts on every memory access.
+In this way, the Iterative Deepening algorithm provides the ``best of both worlds'' for the tradeoff mentioned in Section~\ref{sec:qs-motivation}:
+when tests are too large, Quicksand falls back on prioritization heuristics to find bugs quickly;
+when tests are tractable, Quicksand's verification is equal to the strongest from any single-state-space technique.
+
+The theorem statement is as follows:
+
+\begin{theorem}[Convergence]
+	If a bug can be exposed by any
+thread interleaving possible by preempting on all instructions during a specific test, Iterative Deepening will eventually test an equivalent interleaving which exposes the same bug.
+\end{theorem}
+
+The proof is available at \cite{quicksand-soundness}.
+
+\subsection{Pruning false-positive data-race candidates}
+I have also proved a (somewhat less remarkable) theorem concerning the soundness of a new tactic for pruning certain types of false-positive data race candidates (under Limited HB).
+A ``malloc-recycle'' false positive occurs when two threads seem to race on the same memory address, but only because the {\tt malloc}ed block containing that address was {\tt free}d and reallocated in between.
+If the transitions of the two threads were reordered, the address collision would disappear and there would be no race.
+The proof is available at \cite{quicksand-soundness}; the theorem statement is:
+
+\begin{theorem}[Soundness of eliminating malloc-recycle candidates]
+	If a malloc-recycle candidate is not a false positive,
+	DPOR will test an alternate thread interleaving in which the
+	accesses can race without fitting the malloc-recycle pattern.
+\end{theorem}
+
 \section{Evaluation}

### 2017-01-04 proposal/quicksand.tex
quicksand eval

@@
 a {\em partial verification} guarantee for some smaller state spaces is better than an ``I don't know'' verdict for one single state space.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \section{Verification}
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \section{Evaluation}
 
-% TODO: is related work needed?
-% i can easily talk about it here but i dont' want to set a weird precedent for super weak related work in my other sections.
+I evaluated Quicksand on the product of the following two questions: For an arbitrary fixed CPU budget...
+
+\begin{itemize}
+	\item ...does Quicksand find more bugs than a single-state-space (SSS-MC) approach...
+	\item ...does Quicksand provide more total verifications than SSS-MC...
+\end{itemize}
+...where the SSS-MC control is configured to preempt...
+\begin{itemize}
+	\item ...on synchronization API boundaries only?
+	\item ...on every single shared memory access?
+\end{itemize}
+
+I also conducted a separate experiment focused on the issue of {\em nondeterministic} data-races:
+candidates that require model checking to expose in the first place, which would be
+false-negatives during a single-execution analysis.
+
+I also also ran several different configurations of Quicksand in each of these experiments,
+alternately using Limited Happens-Before or Pure Happens-Before for its data-race analysis.
+Both of these cases outperformed all the other cases of SSS-MC, so this was mostly useful for comparing the two techniques against each other.
+
+\subsection{Setup and Test Suite}
+
+The test suite for all these experiments was 157 student projects: 79 P2 thread libraries from CMU, and 78 Pintos kernels from Berkeley and the University of Chicago (Section~\ref{sec:overview-edu}).
+I paired each submission with several test cases, 6 for each P2 and 3 for each Pintos.
+In total there were 629 unique pairs of a test case and a student project, henceforth called simply ``tests''.
+(Savvy mathematicians will note that doesn't quite add up -- some of the Pintoses were incomplete, and could only run 1 or 2 tests.)
+
+For each of the 629 tests, I ran Quicksand and SSS-MC for 10 CPU-hours each in each of the configurations named above.
+Quicksand, being inherently parallel, was given 10 CPUs for 1 wall-clock hour,
+while SSS-MC control experiments were given 10 wall-clock hours on 1 CPU
+(which is being charitable: comparing CPU-time gives the controls perfectly efficient hypothetical scaling from parallelization).
+
+\subsection{Results}
+
+The results are shown in Figure~\ref{fig:graphs}.
+The first 3 graphs all plot a cumulative distribution of their respective achievements,
+with elapsed time on the X-axis (as a log scale, to emphasize detail in the ``easy'' tests).
+The captions largely explain each result.
+To summarize, Quicksand outperforms the state-of-the-art on all metrics, excepting the impact of parallelization start-up overhead for very small CPU-time limits.
+The Limited HB data-race strategy generally finds bugs faster when they exist,
+but suffers on total verifications compared to Pure HB,
+because of its false positives which create excess redundant state spaces for Landslide to check.
+
+\begin{figure}[h]
+	\begin{tabular}{cc}
+		\begin{tabular}{p{0.48\textwidth}p{0.48\textwidth}}
+			\includegraphics[width=0.48\textwidth]{dowefindbugsfaster-v2.pdf}
+		&
+			\includegraphics[width=0.48\textwidth]{dowefindbugsfaster-wallclock-v2.pdf}
+		\\
+			(a)
+			Bugs found as a function of elapsed CPU time. After the break-even point at
+			\textasciitilde{}200 seconds, Quicksand outperforms all control experiments.
+		&
+			(b)
+			Bugs found by elapsed wall-clock time.
+			This presentation emphasizes that the ``break-even'' point in (a)
+			is just an artifact of parallelization start-up overhead.
+			Quicksand is parallelized tenfold; the vertical line indicates its 1 hour limit.
+		\\
+			\includegraphics[width=0.48\textwidth]{totalverifs-v2.pdf}
+		&
+			\includegraphics[width=0.48\textwidth]{nondets.pdf}
+		\\
+			(c)
+			Total verifications provided as a function of CPU time.
+			The lines for SSS-MC-ICB and SSS-MC-DPOR are artificially penalized to exclude tests with data races,
+			as preempting on sync APIs alone is not sufficient to verify those.
+			SSS-MC-Shared-Mem, however, is not penalized thus;
+			its failure here is due to the massive computational overhead from so many preemption points.
+		&
+			(d)
+			Plot of how many data-race bugs were found using ``single-pass'' data-race candidates versus ``nondeterministic'' ones.
+			With both Limited and Pure HB approaches, the MC's ability to find new data-race candidates in obscure thread interleavings led to more bugs found in all test cases.
+			Between Limited and Pure HB, Pure HB is more reliant on this phenomenon, as Limited HB can often find more potential races on the first pass.
+		\end{tabular}
+	\end{tabular}
+	\caption{Results.}
+	\label{fig:graphs}
+\end{figure}

### 2017-01-04 proposal/overview.tex
quicksand eval
easter egg

@@
 \section{Education}
+\label{sec:overview-edu}
 
@@
 concurrency bugs can arise from the interaction of any subset of kernel components with each other.
-The most common bugs students face arise from the interaction of some component with itself (e.g., concurrent invocations of {\tt new\_pages}/{\tt remove\_pages} in the same process),
+The most common bugs studence face arise from the interaction of some component with itself (e.g., concurrent invocations of {\tt new\_pages}/{\tt remove\_pages} in the same process),
 or from the interaction between an exiting thread and some other thread trying to communicate with it ({\tt vanish} versus, well, anything else, really).

### 2017-01-04 proposal/overview.tex
cuidado...

@@
 \begin{itemize}
-	\item Mutexes, with the interface {\tt mutex\_lock(mp)} and {\tt mutex\_unlock(mp)}, whose functionality is described earlier this chapter.
+	\llitem Mutexes, with the interface {\tt mutex\_lock(mp)} and {\tt mutex\_unlock(mp)}, whose functionality is described earlier this chapter.
 Students may use any x86 atomic instruction(s) they desire, such as {\tt xchg}, {\tt xadd}, or {\tt cmpxchg}, and/or the {\tt deschedule}/ {\tt make\_runnable} system calls offered by the reference kernel.
-	\item Condition variables, with the interface {\tt cond\_wait(cvp, mp)}, {\tt cond\_signal} {\tt (cvp)}, and {\tt cond\_broadcast(cvp)}.
+	\llitem Condition variables, with the interface {\tt cond\_wait(cvp, mp)}, {\tt cond\_signal} {\tt (cvp)}, and {\tt cond\_broadcast(cvp)}.
 {\tt cond\_wait} blocks the invoking thread, ``simultaneously'' releasing a mutex which protects some associated state (atomically, with respect to other calls to signal or broadcast under that mutex).
@@
 		The primary challenge of this exercise is ensuring the aforementioned atomicity between {\tt cond\_wait}'s unlock and deschedule, with respect to the rest of the interface.
-	\item Semaphores, with the interface {\tt sem\_wait(sp)} and {\tt sem\_signal(sp)} (sometimes called {\em proberen} and {\em verhogen} in other literature).
+	\llitem Semaphores, with the interface {\tt sem\_wait(sp)} and {\tt sem\_signal(sp)} (sometimes called {\em proberen} and {\em verhogen} in other literature).
 The semaphore can be initialized to any integer value; if initialized to 1, it behaves like a mutex.
 		Students typically implement semaphores using mutexes and condition variables, not using atomic instructions or system calls directly.
-	\item Reader-writer locks (rwlocks), with the interface {\tt rwlock\_lock(rwp, mode)} and {\tt rwlock\_unlock(rwp)}. {\tt mode} may be either {\tt RWLOCK\_READ} or {\tt RWLOCK\_}- {\tt WRITE}.
+	\llitem Reader-writer locks (rwlocks), with the interface {\tt rwlock\_lock(rwp, mode)} and {\tt rwlock\_unlock(rwp)}. {\tt mode} may be either {\tt RWLOCK\_READ} or {\tt RWLOCK\_}- {\tt WRITE}.
 		Behaves as mutexes, but multiple readers may access the critical section simultaneously.
@@
 \begin{itemize}
-	\item {\tt thr\_init(stack\_size)} initializes the thread library, setting a default stack size to be allocated to new threads.
-	\item {\tt thr\_create(child\_func, child\_arg)} spawns a new thread to run the specified function with the specified argument.
+	\llitem {\tt thr\_init(stack\_size)} initializes the thread library, setting a default stack size to be allocated to new threads.
+	\llitem {\tt thr\_create(child\_func, child\_arg)} spawns a new thread to run the specified function with the specified argument.
 There is a semantic gap between this function and the {\tt thread\_fork} system call (which takes no parameters, makes no changes to the user's address space, and cannot meaningfully be invoked from C code) which students must bridge.
 		Returns an integer thread ID of the newly created thread.
-	\item {\tt thr\_exit(status)} aborts execution of the calling thread, recording an exit status value.
+	\llitem {\tt thr\_exit(status)} aborts execution of the calling thread, recording an exit status value.
 		The main challenge of this function is to allow another thread to free the memory used for the exiting thread's stack,
 		without risking any corruption as long as the exiting thread continues to run.
-	\item {\tt thr\_join(tid, statusp)} blocks the calling thread until the thread with the specified thread ID exits, then returns, collecting its exit status.
+	\llitem {\tt thr\_join(tid, statusp)} blocks the calling thread until the thread with the specified thread ID exits, then returns, collecting its exit status.
 \end{itemize}
@@
 \begin{itemize}
-	\item A round-robin scheduler, including context switching, timer handling, and runqueue management;
-	\item Some approach to locking, often analogous to P2's concurrency primitives (henceforth referred to as ``kernel mutexes''), 
-		and some approach to blocking threads indefinitely;
-	\item A virtual memory implementation, including a program loader;
-	\item Lifecycle management code for creation and destruction of kernel threads and processes;
-	\item Other miscellany such as a suite of fault handlers to ensure no user program can cause the kernel itself to crash.
+	\llitem A round-robin scheduler, including context switching, timer handling, and runqueue management;
+	\llitem Some approach to locking, often analogous to P2's concurrency primitives (henceforth referred to as ``kernel mutexes''), 
+	 ll       and some approach to blocking threads indefinitely;
+	\llitem A virtual memory implementation, including a program loader;
+	\llitem Lifecycle management code for creation and destruction of kernel threads and processes;
+	\llitem Other miscellany such as a suite of fault handlers to ensure no user program can cause the kernel itself to crash.
 \end{itemize}

### 2017-01-04 proposal/thesprop.tex
dates and committe members
quicksand verif
cuidado...
timeline

@@
 \usepackage{amsmath}
+\usepackage{amsthm}
 \usepackage[charter]{mathdesign}
@@
 
+\newtheorem{theorem}{Theorem}
+
+\newcommand\llama[1]{\includegraphics[width=#1]{llama.pdf}}
+\newcommand\llitem{\item[\llama{1em}]}
+
 \begin {document}
@@
 \author{Ben Blum}
-\date{Janucember 2016}
-\Year{2016}
+\date{January 2017}
+\Year{2017}
 \trnumber{}
@@
 David A. Eckhardt \\
-Yet another person \\
-Someone from a strange and faraway land
+Brandon Lucia \\
+Haryadi Gunawi (University of Chicago)
 }
@@
 In this document I have proposed a thesis which will solve both problems.
+
 I leave you now with a proposed timeline for bringing each project to fruition.
+Bear in mind that for the first half of 2017, I'll be working with a full-time research assistant,
+David Vernet, on the transactional memory and porting Landslide projects.
 
 \begin{itemize}
-	\item TODO % TODO
+	\item 2017 January: Submit thesis proposal, begin work on TM project (Section~\ref{sec:tm}). Also submit revised IRB proposal for planned user studies.
+	\item 2017 ???uary: Present thesis proposal (whenever the committee is free).
+	\item 2017 February: Develop ``grading'' TA user study and deploy it in 15-410 (Section~\ref{sec:grading}).
+	\item 2017 March-April: Finish implementation/evaluation of TM; write a paper about it.
+	\item 2017 April-June: Begin ``port Landslide to VMware or Bochs'' project (Section~\ref{sec:pintos-todo}).
+	\item 2017 June-July: Resubmit paper about 410 user studies (Section~\ref{sec:studence}) (previously rejected from SIGCSE). Also run ``variance'' Quicksand experiments (Section~\ref{chap:quicksand}).
+	\item 2017 August-October: Polish and deploy Pintos user study (Section~\ref{sec:pintos-todo}).
+	\item 2017 September-December: Write thesis. Also deploy 410 TA grading study once more for more data.
+	\item 2017 December: Defend \& graduate.
 \end{itemize}
+
+For pretty much any reason (much like hardware transactions),
+I may need to delay the timeline by a semester, defending instead in 2018 June.
+For example, the above plan to deploy a Pintos user study within one semester's time of finishing the open-source Landslide port is quite ambitious,
+assuming among other things smooth negotiations with the OS course staff at another university.
+Anyway, this is my optimistic proposal. Thanks for reading.
 

### 2017-01-05 proposal/overview.tex
minor fixes

@@
 The number of preemption points in each execution define the depth of this tree, and the number of threads available to run define the branching factor.
-Hence, in a program with $n$ preemption points and $k$ threads available to run at each, the state space size is $n^k$.
+Hence, in a program with $n$ preemption points and $k$ threads available to run at each, the state space size is $O(n^k)$.
 Addressing the scaling problem in this exponential relation is the central research problem for all model checkers.
@@
 in practice we are actually classifying the individual memory access events corresponding to those lines during execution.
-Actually, each {\tt x++} statement likely compiles to two separate load and store instructions, so each of those two instructions from each of the two marked source lines pairwise will race (except for the two loads, which are both reads).
+Actually, each {\tt x++} statement likely compiles to two separate load or store instructions, so each of those two instructions from each of the two marked source lines pairwise will race (except for the two loads, which are both reads).
 

### 2017-01-05 proposal/410.tex
timeline
write about education implementation

@@
 This is the second of the three projects I am proposing for this thesis, and is currently ongoing work.
+
+\section{Motivation}
+
+In my MS thesis \cite{landslide}, I solicited students at CMU's Operating Systems Design and Implementation class
+(henceforth, ``15-410'')
+to volunteer at the end of the P3 project to annotate their kernels and try debugging them with Landslide.
+However, the annotation burden undermined Landslide's purpose:
+the only students willing to spend free time on manual instrumentation were biased to be those who were already doing well in the class,
+and hence least likely to benefit from Landslide's debugging potential.
+(Actually, even the best 15-410 students still have concurrency bugs,
+but in principle, an educational tool must reach the more struggling students,
+the so-called ``middle'' and/or ``bottom'' of the class.)
+%
+Requiring annotations hurts Landslide's case as a grading tool, as well:
+TAs need to understand the kernel to begin with in order to annotate correctly,
+and while achieving such understanding they may as well grade it by hand, as before.
+
+Since then, I've extended Landslide to support testing P2 thread libraries (Section~\ref{sec:pebbles}) as well.
+Because P2 mandates specific function names for the project's internal APIs
+-- most importantly, for the concurrency primitives --
+Landslide can automatically annotate arbitrary student implementations with no manual effort required of the user (whether student or TA).
+The addition of Quicksand and Iterative Deepening (Chapter~\ref{chap:quicksand}) partly fulfills this purpose,
+freeing student attention from the issue of which state spaces to test.
+This chapter will detail my further techniques and evaluation which are specific to educational use.
+
+\section{Implementation Details}
+
+Every stateless model checker must make some assumptions about the tested programs' concurrency model \cite{chess,inspect,dbug-ssv}.
+However, arbitrary programs may break conventional disciplines of concurrent programs, while still being bug-free.
+For example, thread communication via ad-hoc {\tt yield} loops may appear as an infinite loop, livelock, or deadlock.
+This is especially true of student code,
+written by people who are just learning concurrent programming discipline for the first time,
+and/or written under the time pressure of a project deadline.
+To be an effective tool for struggling students,
+a model checker must somehow coax adversarial programs to fit its concurrency model, to effectively test for real bugs,
+rather than rejecting them outright on some stylistic or disciplinary grounds.
+
+Fully-automatic instrumentation of student P2s has been no walk in the park.
+I have equipped Landslide with several powerful algorithms and heuristics for handling the most common anti-patterns in student submissions.
+
+% TODO
+\begin{itemize}
+	\item 
+	\item 
+	\item 
+\end{itemize}
+
+However, some of the less common offenses are both more difficult to handle algorithmically, and also more worrying from a pedagogical point of view.
+%After some collaboration with David Eckhardt (15-410 professor and member of this thesis committee),
+For the following cases, we configured Landslide to abort, and warn the student that they must find a better solution before it could test their code.
+
+\begin{itemize}
+	\item Busy-wait loops containing neither {\tt yield} nor {\tt xchg} (nor any other atomic instruction), such as {\tt while (!other\_thread\_ready) continue;}.
+		This blurs the line between anti-pattern and concurrency bug:
+		because it does not yield the CPU, a uni-processor machine must wait for the next timer tick (several milliseconds!) before making any progress;
+		also, because it does not use atomic instructions, an optimizing compiler may reorder or even delete the loop's accesses.
+		%
+		Landslide also cannot easily identify it as similar to message-passing,
+		appearing indistinguishable
+		% not really? there would be dpor conflix, at least
+		from a thread-local infinite computation,
+		which is of course impossible to judge for halting \cite{entscheidungsproblem}.
+
+		In such a case, Landslide will issue a bug report with the special message:
+		{\em I have run a loop in [function name] an alarming number of times.
+		This version of Landslide cannot distinguish between this loop being infinite versus merely undesirable.
+		Please refer to the ``Synchronization (2)'' lecture.}
+	\item Recursive mutex use (i.e., locking the same mutex twice in the same thread, then subsequently unlocking it twice).
+		While it would not be difficult for Landslide's lock-sets to support recursive locking (using a nesting counter instead of a boolean flag),
+		that would assume the corresponding mutex implementation provides the same semantics,
+		which is risky business with student code. % ;)
+		Furthermore, recursive locking is not an obvious solution to any of P2's challenges;
+		far more often, it arises when a student's mutex tries to {\tt malloc} some internal state,
+		which itself requires a mutex for safe allocation, which can lead to stack overflow and a crash.
+
+		In such a case, Landslide will issue a bug report with the special message:
+		{\em This version of Landslide cannot debug recursive implementations of mutex\_lock.
+		Please examine this stack trace and determine for yourself whether it indicates a bug.}
+\end{itemize}
+
+\section{Landslide as a Debugging Tool}
+\label{sec:studence}
+
+TODO
+
+\section{Landslide as a Grading Tool}
+\label{sec:grading}
+
+TODO

### 2017-01-06 proposal/overview.tex
write about education implementation
undo easter egg in overview; it'll be in yield block discussion instead

--- 2017-01-04 proposal/overview.tex
@@
 \subsection{Pebbles}
+\label{sec:pebbles}
 
@@
 	\item Some approach to locking, often analogous to P2's concurrency primitives (henceforth referred to as ``kernel mutexes''), 
-		and some approach to blocking threads indefinitely;
+	 ll       and some approach to blocking threads indefinitely;
 	\item A virtual memory implementation, including a program loader;
@@
 Careless solutions to this problem are prone to data races or deadlocks.
+
+% TODO: Talk about the hurdle (both for p2 and p3).
 

### 2017-01-06 proposal/410.tex
implementation deets

@@
 
-Every stateless model checker must make some assumptions about the tested programs' concurrency model \cite{chess,inspect,dbug-ssv}.
+Every stateless model checker must make some assumptions about the tested programs' concurrency model \cite{chess}.
 However, arbitrary programs may break conventional disciplines of concurrent programs, while still being bug-free.
@@
 
-% TODO
 \begin{itemize}
-	\item 
-	\item 
-	\item 
+	\item {\bf Yield-blocking.}
+		Landslide recognizes open-coded busy-wait loops used for ad-hoc synchronization,
+		and is able to treat threads as blocked (or ``disabled'' in model-checking parlance),
+		avoiding getting stuck in an infinitely-long interleaving (or ``cyclic state space'')
+		which should never arise during normal execution.
+		The heuristically-driven algorithm is as follows:
+		\begin{itemize}
+			\llitem Whenever a thread performs a {\tt yield}, {\tt xchg}, or other atomic instruction,
+				Landslide increments a per-thread counter to track its (supposed) busy-wait loop iterations.
+				A thread's counter is reset any time it performs some ``interesting'' activity not likely to appear in a true busy-wait loop:
+				any condvar, semaphore, or rwlock invocation (but not mutexes),
+				and the beginning or end of any thread library function (create, join, or exit).
+				\footnote{Implemented in {\tt check\_user\_yield\_activity} and {\tt check\_user\_xchg} in {\tt user\_sync.c}.}
+				% TODO: Wow, what happened to KEEP_RUNNING_YIELDING_THREADS?? I swear I had some logic in arbiter.c about that.
+			\llitem When a thread's loop counter reaches some heuristic limit (10 for {\tt yield}s, 100 for {\tt xchg}s),
+				% Or 20 for xchgs which end up being data-race PPs.
+				Landslide marks the thread blocked (or ``disabled'', in model-checking parlance),
+				just as though it had invoked {\tt deschedule}.
+				It also retroactively disables that thread at all preceding {\tt yield}s/{\tt xchg}s in that sequence,
+				which prevents DPOR from trying to use each as a preemption point,
+				and avoids a state space explosion (by a factor of the heuristic yield limit).
+				% Well, only the whole state space will explode if this yield-block *always* happens.
+				% Otherwise it's (heuristic limit) * (size of subtree) * (number of yield-block occurrences).
+				% Tbf, I guess it's very likely for that to be half, or more, of the whole space.
+				\footnote{Implemented in {\tt update\_blocked\_transition} in {\tt user\_sync.c}.}
+			\llitem When retroactively disabling a thread across all its preceding loop iterations,
+				Landslide's state space estimator must account for the ``pruning'' of duplicate subtrees at those (now disabled) preemption points.
+				% TODO: Yeah, thinking about this again, I'm pretty sure untag_blocked_branch is deadcode.
+				If in any previous thread interleaving, DPOR tagged the now-yield-blocked thread to interleave at another thread's preemption point,
+				the estimator would have included that potential subtree in its computation of how much unexplored state space exists.
+				Accordingly, in this case Landslide will invert the estimation algorithm,
+				including propagating the reduced subtree size all the way to the state space's root.
+				\footnote{Implemented in {\tt untag\_blocked\_branch} in {\tt estimate.c}.}
+			\llitem Landslide can precisely identify when another thread may trigger the yield-blocking one to fall out of its loop,
+				by analyzing the shared memory conflicts involving only accesses performed in the loop.
+				At any such memory conflict, Landslide will reenable the yielding thread.
+				(If the other thread's access does not fulfill the yielding thread's wait condition,
+				the latter will just re-trigger the heuristic and become blocked again.)
+				% Hmm.... if two threads are in a deadlock between two ad-hoc loops,
+				% and each has some spurious access which kicks the other,
+				% it will show up as an infinite loop instead. I guess that's ok?
+				\footnote{Implemented in {\tt check\_unblock\_yield\_loop} in {\tt user\_sync.c}.}
+		\end{itemize}
+		This approach is similar to the Fair-Bounded Search algorithm in \cite{bpor},
+		although it avoids a major assumption of the latter (threads {\tt yield} if and only if not making progress),
+		and also avoids the need to iteratively deepen the yield bound by fixing it as a heuristic constant.
+		% Actually, I don't think that's true. Using DPOR to unblock a yield-looping thread at any conflict covers this.
+		%The cost of this heuristic is that falsely identifying threads as blocked can lead to unsoundness (...)
+	\item {\bf False-positive deadlock avoidance.}
+		In contrast to its treatment of data races, Landslide must never report a false-positive {\em bug}.
+		If its heuristics falsely identify a thread as blocked, and all other threads are truly blocked,
+		waiting on some progress from that thread, Landslide would report a deadlock bug,
+		and confuse students horribly.
+
+		The yield-loop heuristic assumes that ``too many'' yields or atomics should not arise during normal, non-looping execution of thread library routines.
+		Though extremely rare, this assumption can be violated by an adversarial student submission.
+		More often, Landslide can falsely block threads in the special case of {\tt mutex\_test} (see \cite{quicksand}),
+		where it uses preemption points within the implementation of {\tt mutex\_lock} itself.
+		False deadlocks can also arise from the heuristic blocking of ICB \cite{chess-icb}
+		(used in Quicksand's control experiments, Section~\ref{sec:quicksand-eval}).
+
+		When a deadlock arises under conditions where one or more threads are heuristically blocked,
+		Landslide attempts to refute it as a false positive by artificially unblocking all heuristically-blocked threads%
+		\footnote{If any threads are ICB-blocked, I prioritize waking those before trying to wake any yield-blocked threads.
+		% Hmm... I don't remember exactly why.
+		Waking all threads at once here can lead to unsoundness.}
+		If the deadlock is true, each thread will immediately trigger the yield-blocking heuristic again,
+		bringing the system back into deadlocked state.
+		Landslide then repeats this process a heuristic constant number of times (128),
+		allowing the program that many chances to make progress before proclaiming deadlock.
+		(Note that this heuristic cannot miss true deadlocks as false negatives.)
+		\footnote{Implemented in {\tt try\_avoid\_fp\_deadlock} in {\tt arbiter.c}.}
+	\item {\bf Lock hand-off.}
+		A common, though discouraged, idiom for implementing thread destruction involves one thread ``handing off'' ownership of a mutex to another.
+		That thread will then release the lock with no corresponding acquire in its own execution.
+		% In this case, some other thread communication enforces an ordering ... hmm,, pure vs limited hb?
+		Although Landslide cannot easily recognize at what point the latter thread's accesses are protected by that lock for the sake of data-race analysis,
+		potentially leading to false positive races,
+		it must release the lock in its bookkeeping to avoid false {\em negatives} if any later access that should be protected by that lock isn't.
+		When releasing a lock, if absent from the current thread's lockset, Landslide searches the locksets of all other existing threads, and releases it there.
+		Landslide can instead optionally be configured to treat lock hand-off as an outright bug, as a matter of discipline.
+		\footnote{Implemented in {\tt lockset\_remove} in {\tt lockset.c}.}
+
+	% Not really research.
+	%\item Other small engineering fixes, such as: recognizing when a system call's access to user memory constitutes a ``communication back-channel'' for user threads that should be considered for DPOR;
 \end{itemize}

### 2017-01-06 proposal/410.tex
410 resultce

@@
 This is the second of the three projects I am proposing for this thesis, and is currently ongoing work.
+What remains to be done is detailed in Section~\ref{sec:grading}.
 
@@
 
-TODO
+In the Spring 2015, Fall 2015, and Spring 2016 semesters,
+I've made Landslide (with Quicksand) available to 15-410 students during the last week of P2.
+I introduced stateless model checking in a guest lecture given to the class,
+and required students to pass the thread library hurdle \cite{thrlib}
+to ensure a minimum of basic functionality necessary to use Landslide.
+
+To analyze its effectiveness as a debugging tool for students,
+I configured it to record a snapshot of the results of each use by the students:
+which options were used (test program and time limit),
+a snapshot of the student’s code, and the result of the test
+(completed, timed out, bug found, or ctrl-C’ed).
+I manually interpreted the snapshots to determine:
+\begin{itemize}
+	\item How many unique bugs the student found (discounting multiple runs that produced ``the same'' bug);
+	\item How many of those bugs were deterministic versus concurrency bugs (did Landslide find the bug on the first interleaving or did it need to test multiple);
+	\item How many of each category of bug the students fixed (determined when a subsequent run of Landslide on the test case failed to find the bug again).
+\end{itemize}
+
+{\bf Results -- bugs found.}
+Of 90 two-student groups in those 3 semesters, 47 tested their thread libraries with Landslide.
+Table~\ref{tab:this-table-sucks-but-it's-the-best-i-got} shows
+that Landslide found in total 44 deterministic bugs and 85 concurrency bugs.
+It found at least one concurrency bug for 32 groups (68\%),
+24 of whom (51\%) were able to fix at least one, verifying their update
+with a succesful re-run of the same test.
+
+The table's data presentation is somewhat confusing:
+the dependent columns count the number of {\em groups}, categorized by how many bugs they found, not the raw number of {\em bugs},
+which I instead indicate in the ``Total bugs'' row at the bottom.
+So, the number of total bugs is equal to the sum of the products between each cell in that column and its corresponding number of bugs,
+e.g., $44 = 27\times0 + 6\times1 + 9\times2 + 2\times3 + 1\times4 + 2\times5$.
+Sorry about that; I'll try to find a better way to present this data in the thesis.
+
+\begin{table}[h]
+	\begin{center}
+	\begin{tabular}{r|cc|cc}
+	%det bugs found det bugs fixed  races fixed     races found
+	& \multicolumn{4}{c}{Number of groups} \\
+	Number & \multicolumn{2}{c|}{Deterministic bugs} & \multicolumn{2}{c}{Concurrency bugs} \\
+	of bugs & found & fixed & found & fixed \\
+	\hline
+	0       & 27    & 32    & 15    & 23    \\
+	1       & 6     & 5     & 7     & 11    \\
+	2       & 9     & 6     & 12    & 8     \\
+	3       & 2     & 1     & 7     & 2     \\
+	4       & 1     & 1     & 3     & 1     \\
+	5       & 2     & 2     & 2     & 1     \\
+	11      &       &       & 1     & 1     \\
+	\hline
+	%Total groups
+	%       & 47    & 47    & 47    & 47    \\
+	Total bugs
+	        & 44    & 34    & 85    & 53    \\
+	\end{tabular}
+	\end{center}
+	\caption{Summary of how many bugs were found and/or fixed by how many groups.
+	Each column counts how many groups found/fixed the number of bugs in each row;
+	for instance, two groups found 5 concurrency bugs, one of which fixed all 5.}
+	\label{tab:this-table-sucks-but-it's-the-best-i-got}
+\end{table}
+
+{\bf Results -- impact on grades.}
+Next, I investigated the impact Landslide ultimately had on students' performance.
+I collected the ultimate P2 and P3 grades from the last 6 semesters, forming three categories:
+students from the past 3 semesters who volunteered to use Landslide,
+students from those semesters who opted out,
+and students from the other 3 semesters, for whom Landslide was not available at all.
+I hoped to measure a correlation between use of Landslide and ultimate P2 grades,
+which would show that it helps students submit more correct P2s,
+and/or a correlation with ultimate P3 grades,
+which would (tentatively) show that it teaches better ways of thinking that students retain for future projects (i.e., measuring actual learning).
+
+Unfortunately, this turned out to be a negative result, as shown in Figure~\ref{fig:grade-distribution}.
+The distribution of Landslide users' ultimate grades is indistinguishable from that of students for whom Landslide was unavailable.
+While there was a 3\% increase in P2 grades between opt-out-ers and Landslide users, it's prone to selection bias:
+perhaps those who opted out of Landslide were already the worst of the class who simply didn't have any free time for it.
+The distribution of P3 grades is indistinguishable as well.
+In the thesis, I will discuss possible factors contributing to this failure, and how they might be controlled for in future work,
+although I am not planning to run any further experiments in the limited time I have.
+
+\begin{figure}[h]
+	\begin{center}
+	\includegraphics[width=0.96\textwidth]{p2-p3-distribution.pdf}
+	\end{center}
+	\caption{Distribution of P2 (thrlib) and P3 (kernel) grades among students who did or didn't use Landslide.}
+	\label{fig:grade-distribution}
+\end{figure}
+
+{\bf Future work.}
+One evaluation question remains which I will answer in the thesis:
+what is the accuracy of Landslide's automation heuristics (see previous section)?
+In other words, were false-positive infinite loops or deadlocks ever reported
+where Landslide failed to recognize an ad-hoc synchronization?
+(I expect a near 100\% success rate,
+having tuned the heuristics by hand on an (independent) set of P2 submissions already.)
 

### 2017-01-06 proposal/410.tex
landlside for grabing

@@
 
-TODO
+Beginning in 2017, I will recruit 15-410 Teaching Assistants (TAs) for my user study rather than students.
+The modified user study will evaluate Landslide as a grading tool,
+alternative to the ``Fritz'' stress-testing infrastructure which 15-410 currently employs,
+and possibly substituting as well for the human process of reading student code looking for concurrency bugs by hand.
+The intent is not for Landslide to ultimately replace 15-410's manual grading process,
+but rather to demonstrate that comparably-rigorous grading could be deployed automatedly at other schools wishing to adopt the 15-410 curriculum.
+
+I will pursue the following evaluation questions:
+
+\begin{itemize}
+	\item Provided appropriate test cases, can Landslide find the most common ``deep'' concurrency bugs in student submissions?
+	\item How does Landslide's bug-finding compare to that of Fritz, the class's stress tester?
+	\item How does Landslide's bug-finding compare to that of TAs grading by hand?
+\end{itemize}
+
+Finally, I will conclude this chapter in the thesis by discussing how future work could adapt Pebbles's interface to support fully-automatic model checking without compromising the educational value of P3.

### 2017-01-06 proposal/410.tex
several small changes

@@
 	\begin{center}
-	\includegraphics[width=0.96\textwidth]{p2-p3-distribution.pdf}
+	\includegraphics[width=0.85\textwidth]{p2-p3-distribution.pdf}
 	\end{center}
@@
 One evaluation question remains which I will answer in the thesis:
-what is the accuracy of Landslide's automation heuristics (see previous section)?
+what is the accuracy of Landslide's automation heuristics?
 In other words, were false-positive infinite loops or deadlocks ever reported

### 2017-01-06 proposal/quicksand.tex
minor fixes
implementation deets
several small changes

@@
 
-\subsection{Pruning false-positive data-race candidates}
+\subsection{Pruning False Positives} % data-race candidates}
 I have also proved a (somewhat less remarkable) theorem concerning the soundness of a new tactic for pruning certain types of false-positive data race candidates (under Limited HB).
@@
 \section{Evaluation}
+\label{sec:quicksand-eval}
 
@@
 	\end{tabular}
-	\caption{Results.}
+	\caption{Many kernels died to bring us this information.}
 	\label{fig:graphs}

### 2017-01-06 proposal/thesprop.tex
minor fixes
several small changes
NOOOOOOOOOOO

--- 2016-12-30 proposal/thesprop.tex
@@
 
+\newtheorem{theorem}{Theorem}
+
+\newcommand\llama[1]{\includegraphics[width=#1]{llama.pdf}}
+\newcommand\llitem{\item[\llama{1.2em}]}
+
 \begin {document}
@@
 
-\keywords{landslide terminal, baggage claim, ground transportation, ticketing}
+% TODO(thesis): Save this one for the thesis itself...
+%\keywords{landslide terminal, baggage claim, ground transportation, ticketing}
 
@@
 \begin{itemize}
-	\item TODO % TODO
+	\item 2017 January: Submit thesis proposal, submit revised IRB proposal for planned user studies.
+	\item 2017 ???uary: Present thesis proposal (whenever the committee is free).
+	\item 2017 February-March: Develop ``grading'' TA user study and deploy it in 15-410 (Section~\ref{sec:grading}).
+	\item 2017 March-August: Port Landslide to Bochs for use in Pintos OS classes (Section~\ref{sec:pintos-todo}).
+	\item 2017 March-August: Implement, evaluate, and write paper about transactional memory project (Section~\ref{sec:tm}).
+	\item 2017 August-October: Polish and deploy Pintos user study (Section~\ref{sec:pintos-todo}).
+	\item 2017 ???ember: Resubmit paper about 410 user studies (Section~\ref{sec:studence}) (previously rejected from SIGCSE) (whenever an appropriate conference deadline arises.
+	\item 2017 October-November: Deploy 410 TA grading study again for more data.
+	\item 2017 December: Begin writing thesis.
+	\item 2018 January-April: Write thesis.
+Also fill in miscellaneous gaps, such as the ``variance'' Quicksand experiments (Section~\ref{chap:quicksand}) and measuring Landslide's heuristics' accuracy (Section~\ref{sec:studence}).
+	\item 2018 February-March: Deploy Pintos student and 410 TA studies one final time.
+	\item 2018 April-May: Defend \& graduate.
 \end{itemize}
 
-% TODO - make a timeline
+%For pretty much any reason (much like hardware transactions),
+%I may need to delay the timeline by a semester, defending instead in 2018 June.
+%For example, the above plan to deploy a Pintos user study within one semester's time of finishing the open-source Landslide port is quite ambitious,
+%assuming among other things smooth negotiations with the OS course staff at another university.
+%Anyway, this is my optimistic proposal. Thanks for reading.
 

### 2017-01-07 proposal/overview.tex
related work

@@
 The other three (a stack-crawling backtrace utility, a bare-metal game with device drivers, and a small extension to the P3 kernel) are not of concern in this thesis.
+The course's prerequisite is Introduction to Computer Systems \cite{sigcse01:CSaPP}.
 Both P2 and P3 are built using the {\em Pebbles} system call specification, outlined in Table~\ref{tab:syscalls}
@@
 In P3, students implement a kernel which provides the same system calls shown in Table~\ref{tab:syscalls}, previously provided by the reference kernel.
+Pebbles adopts the Mach \cite{DBLP:conf/usenix/AccettaBBGRTY86} distinction between {\em tasks}, which are resource containers, and {\em threads}, each of which executes within a single task.
+This requires less implementation complexity than the more featureful Plan 9's {\tt rfork} \cite{Pike90plan9} or Linux's {\tt clone} models.
+
 Although the internal interfaces are not mandated like they were in P2, all Pebbles kernels must necessarily contain the same abstract components. These include:

### 2017-01-07 proposal/related.tex
some more changes
related work

@@
 Perhaps by now, Landslide itself deserves a spot on this list: I first introduced it in my MS thesis \cite{landslide},
-and it is most notable for testing kernel-level code through the use of a simulator.
+and it is most notable for checking all shared memory accesses and supporting kernel-level code through the use of a simulator.
 
@@
 
+{\bf Other theoretical advances.}
 A number of other techniques address the problem that even with DPOR, a state space may be too large.
@@
 
-{\bf Other theoretical advances.}
-% TODO
+\section{Concurrency in Education}
 
-\section{Educational Testing Tools}
+I built Landslide upon the Pebbles \cite{kspec} curriculum and concurrency model, naturally, as it is closest to home. % but also, it's the best.
+Pintos \cite{pintos} has recently emerged as the most popular educational kernel
+(by count of top CS schools in the USA who teach by it);
+it trades off the prevalence of its concurrency challenges to cover various OS topics more broadly,
+especially advanced scheduling algorithms.
+Pintos is the stand-alone evolution of its predecessor, Nachos \cite{nachos}, which originally ran as a UNIX process with simulated device drivers.
+%
+Xv6 \cite{xv6}, from MIT, is another major educational kernel, which is also UNIX-like and runs in QEMU,
+and a natural target for model checking in future work.
+Recently, Columbia introduced a new Android-focused OS course \cite{teaching-android},
+which perhaps highlights the importance of related work on model-checking event-driven applications \cite{r4}.
 
-\section{Real-World Techniques}
+To my knowledge, this is the first study of model checking in an educational setting, although teaching concurrency is not itself an unstudied problem.
+\cite{towards-a-framework} presents a promising framework for testing concurrent programs,
+which can incorporate model checking as well as static analysis, resource exhaustion, data-race analysis, and coverage analysis.
+However, it lacks an evaluation, and makes mention of its educational value only in its future work remarks.
+%
+\cite{how-studence} presents a survey of how students think about concurrent program behaviour and debugging,
+while \cite{learning-concurrency} investigates how students form cognitive patterns about concurrent programming that could either aid or stunt their reasoning.
+Both of these studies could help optimize Landslide's bug reports for clarity and student enlightenment.
+
+\section{Other Techniques}
+
+Data race analysis, originating with the lockset-only analysis of Eraser \cite{eraser},
+has since grown into a mature field.
+Race detectors are largely distinguished by their particular flavour of the Happens-Before (HB) relation:
+some tools \cite{fasttrack,djit} soundly avoid false positives using ``Pure'' HB in the Lamport sense \cite{lamport-clocks};
+others \cite{tsan,hybriddatarace} introduced the ``Limited'' HB relation to find more potential races in a single pass.
+I implemented both approaches in Landslide and evaluated their trade-offs in Section~\ref{sec:quicksand-eval}.
+%
+Recently, the Causally-Precedes relation \cite{predictive-dr} emerged as a refinement of Limited HB which avoids the most common cases of false positives; it would be a welcome enhancement in Landslide.
+
+Replay analysis extends single-pass data-race analysis to apply model checking (in limited quantities) to classify data-race candidates by their impact on program behaviour.
+It was first introduced in \cite{recordreplaydrs}, which compares the program states immediately after the access pair for differences, preferring still to err on the side of false positives.
+RaceFuzzer \cite{racefuzzer} avoids false positives by requiring an actual failure be exhibited, as we do,
+although it uses random schedule fuzzing rather than model checking.
+%
+Portend \cite{portend} is closest in spirit to Quicksand:
+it tests alternate executions based on single-pass data-race candidates to classify them in a taxonomy of likely severity.
+It uses symbolic execution to test input nondeterminism as well as schedule nondeterminism,
+although its verification properties are not as strong as Quicksand's.
+
+% TODO: Devote a paragraph to symbolic execution, including mention of e.g. concolic testing. Cite klee, z3, etc.
+
+Transactional memory (TM), first introduced in \cite{transactional-memory},
+has received renewed attention in recent years with the announcement of Intel's Haswell architecture \cite{htm-haswell},
+which promises to support hardware transactions using new {\tt x86} instructions.
+%
+The combination of TM and testing in the literature is sparse so far.
+{\em Learning from Mistakes} \cite{learning-from-mistakes} is a study of the characteristics of many different types of concurrency bugs;
+it found that TM could potentially fix some, but not all, of the studied bugs,
+although in some cases it must be combined with other concurrency primitives,
+which indicates a major need for verifying transactional code with model checking.

### 2017-01-07 proposal/related.tex
more relwork on tm

@@
 
-\section{Other Techniques}
+\section{Other Concurrency Techniques}
 
@@
 which promises to support hardware transactions using new {\tt x86} instructions.
+Because hardware transactions can fail for any reason, not just on memory conflicts,
+software TM remains relevant,
+and recent works such as \cite{hybrid-htm-stm} and \cite{stm-relaxed-memory} enhance it to be nested with hardware transactions, or to be used on weak memory architectures, respectively;
+which in turn produce ever more complicated concurrency models for people like me to tackle.
 %
-The combination of TM and testing in the literature is sparse so far.
+Testing approaches to TM are sparsely represented in the literature so far.
+%The Weakest Reasonable Condition
+\cite{specifying-verifying-tm} introduces several conditions to help specify/verify the correctness of TM {\em implementations},
+but not that of client programs which use TM.
 {\em Learning from Mistakes} \cite{learning-from-mistakes} is a study of the characteristics of many different types of concurrency bugs;

### 2017-01-07 proposal/410.tex
small fixes

@@
 \begin{itemize}
-	\item Busy-wait loops containing neither {\tt yield} nor {\tt xchg} (nor any other atomic instruction), such as {\tt while (!other\_thread\_ready) continue;}.
+	\item {\bf Busy-wait loops} containing neither {\tt yield} nor {\tt xchg} (nor any other atomic instruction), such as {\tt while (!other\_thread\_ready) continue;}.
 		This blurs the line between anti-pattern and concurrency bug:
@@
 		Please refer to the ``Synchronization (2)'' lecture.}
-	\item Recursive mutex use (i.e., locking the same mutex twice in the same thread, then subsequently unlocking it twice).
+	\item {\bf Recursive mutex use} (i.e., locking the same mutex twice in the same thread, then subsequently unlocking it twice).
 		While it would not be difficult for Landslide's lock-sets to support recursive locking (using a nesting counter instead of a boolean flag),
@@
 which options were used (test program and time limit),
-a snapshot of the student’s code, and the result of the test
-(completed, timed out, bug found, or ctrl-C’ed).
+a snapshot of the student's code, and the result of the test
+(completed, timed out, bug found, or ctrl-C'ed).
 I manually interpreted the snapshots to determine:

### 2017-01-07 proposal/related.tex
small fixes

@@
 
-In this chapter I provide a brief tour of the related work.
+This chapter provides a brief tour of the related work.
 
@@
 has received renewed attention in recent years with the announcement of Intel's Haswell architecture \cite{htm-haswell},
-which promises to support hardware transactions using new {\tt x86} instructions.
+which promises to support hardware transactions using new x86 instructions.
 Because hardware transactions can fail for any reason, not just on memory conflicts,
@@
 %The Weakest Reasonable Condition
-\cite{specifying-verifying-tm} introduces several conditions to help specify/verify the correctness of TM {\em implementations},
+\cite{specifying-verifying-tm} introduces several conditions to help specify/verify the correctness of TM implementations,
 but not that of client programs which use TM.

### 2017-01-07 proposal/related.tex
do more justice to MC TM related work

@@
 %
-Testing approaches to TM are sparsely represented in the literature so far.
+Testing approaches for transactional programs are sparsely represented in the literature so far.
 %The Weakest Reasonable Condition
-\cite{specifying-verifying-tm} introduces several conditions to help specify/verify the correctness of TM implementations,
+Several related works \cite{tm-correctness,tm-completeness,specifying-verifying-tm} are building up to formal proofs of the correctness of TM {\em implementations},
 but not that of client programs which use TM.
-{\em Learning from Mistakes} \cite{learning-from-mistakes} is a study of the characteristics of many different types of concurrency bugs;
+Getting closer, McRT STM \cite{mc-tm-with-spin} uses SPIN \cite{spin} to model check an STM implementation
+up to 2 threads running 1 transaction each with up to 3 memory accesses.
+This kind of verification is an important stepping stone for trusting the results Landslide will provide.
+% TODO: make this point once you've added discussion of mutex_test to chap:quicksand.
+% This kind of verification is analogous to the {\tt mutex\_test}
+%
+{\em Learning from Mistakes} \cite{learning-from-mistakes} studies the characteristics of many types of concurrency bugs;
 it found that TM could potentially fix some, but not all, of the studied bugs,

### 2017-02-07 proposal/thesprop.tex
an actual thesis statemence

@@
 \begin{center}
-	% TODO: lol
-	{\em my research is cool. please let me graduate.}
+	{\em Thanks to the new algorithms, heuristics, and concurrency models I have developed,
+	stateless model checking is an appropriate and accessible concurrency testing technique
+	for programmers in both educational and real-world settings.}
 \end{center}
@@
 	\item 2017 August-October: Polish and deploy Pintos user study (Section~\ref{sec:pintos-todo}).
-	\item 2017 ???ember: Resubmit paper about 410 user studies (Section~\ref{sec:studence}) (previously rejected from SIGCSE) (whenever an appropriate conference deadline arises.
+	\item 2017 ???ember: Resubmit paper about 410 user studies (Section~\ref{sec:studence}) (previously rejected from SIGCSE) (whenever an appropriate conference deadline arises).
 	\item 2017 October-November: Deploy 410 TA grading study again for more data.

### 2017-02-28 landslide-lecture/2017/lecture.tex
update lecture for s17
minor lecture fix

--- 2016-09-28 landslide-lecture/2016/lecture.tex
@@
 \institute[CMU 15-410]{Carnegie Mellon University - 15-410}
-\date[]{2016, September 28}
+\date[]{2017, February 20}
 
@@
 	Okay, wait a sec...
+	\pause
 	\begin{itemize}
@@
 	\begin{itemize}
-		\item $n$ PPs and $k$ threads $\Rightarrow$ state space size is $n^k$.
+		\item $n$ PPs and $k$ threads $\Rightarrow$ state space size is $O(n^k)$.
 		\item Need to choose PPs very carefully for test to be effective.
@@
 			\end{itemize}
-			\item Conference paper (to appear in OOPSLA this november)
+			\item Conference paper (OOPSLA 2016)
 			\begin{itemize}
@@
 	%\end{itemize}
-	%\linegap
+	\linegap
 
@@
 	\begin{itemize}
-		\item Review this lecture and study information sheet
+		%\item Review this lecture and study information sheet
 		\item Meet prerequisites of passing P2 tests
@@
 			\begin{itemize}
-				\item \url{http://tinyurl.com/landslide-p2-f16}
+				\item \url{http://tinyurl.com/landslide-p2-s17}
 			\end{itemize}
@@
 			\begin{itemize}
-				\item Mid-next-week, room and time TBD
-			\end{itemize}
-		\item Use Landslide in addition to stress tests until P2 is due!
+				\item Later this week, room and time TBD
+			\end{itemize}
+		%\item Use Landslide in addition to stress tests until P2 is due!
 	\end{itemize}

### 2017-03-01 proposal/thesprop.tex
thesprop: split pintos and tm into their own chapters

@@
 		This work is {\bf completed} and was published at OOPSLA 2016.
-	\item {\bf Educational use (Chapter \ref{chap:410}).}
+	\item {\bf Educational use (Chapters \ref{chap:410} and \ref{chap:pintos}).}
 		For the past three semesters, I have offered a fully-automated version of Landslide to students in 15-410, CMU's undergraduate Operating System Design and Implementation class \cite{kspec,thrlib}, for use as a debugging aid during the thread library project.
@@
 
-		This is {\bf ongoing work}; I have run the user study for 4 semesters so far and am proposing to continue them and follow them up with an evaluation.
-	\item {\bf Broader impact (Chapter \ref{chap:lipservice}).}
-		Landslide includes several new techniques for testing kernels and thread libraries with little or no manual instrumentation required.
-		So far, a fully-automatic testing mode is available only for 15-410 thread library projects.
-		To prove these techniques are relevant beyond CMU's walls, I will extend Landslide to handle Pintos kernel projects from other universities \cite{pintos}. %and ``real-world'' Linux programs.
-		%I will conduct a user study in which students at another university test their Pintoses with Landslide,
-		%and evaluate Landslide's ability to find known and/or new bugs in real-world programs.
-		I will also extend Landslide's concurrency model to support Hardware Transactional Memory (HTM), a new concurrent programming technique \cite{transactional-memory}, and test several ``real-world'' HTM programs.
+		So far, the fully-automatic testing mode is available only for 15-410 thread library projects.
+		To prove these techniques are relevant beyond CMU's walls, I will extend Landslide to handle Pintos kernel projects from other universities \cite{pintos}.
+		I will then collaborate with those schools to deploy Landslide to their operating systems students.
 
-		This will be {\bf mostly new work}, as Landslide presently supports testing Pintoses only with some manual effort,
+		This is {\bf ongoing work}; I have run the user study for 4 semesters so far and am proposing to continue them
+		and extend the existing evaluation,
+		as well as deploy a Pintos port.
+	\item {\bf Transactional Memory (Chapter \ref{chap:tm}).}
+		Transactional Memory (TM) is a relatively new concurrent programming technique \cite{transactional-memory}
+		which is not yet addressed by modern model checkers.
+		I will extend Landslide's concurrency model to support both hardware (HTM) and software (STM) variants of TM,
+		and test several ``real-world'' TM programs.
+
+		This will be {\bf new work},
+		involving some implementation effort, some proofs, and designing a new evaluation.
+		%, as Landslide presently supports testing Pintoses only with some manual effort,
 		%and does not yet feature instrumentation to run on Linux programs.
-		and will need its concurrency model extended to incorporate the HTM programming model.
+		%and will need its concurrency model extended to incorporate the HTM programming model.
 \end{itemize}
@@
 \input{410}
-\input{lipservice}
+\input{pintos}
+\input{tm}
 \input{related}
@@
 	\item 2017 February-March: Develop ``grading'' TA user study and deploy it in 15-410 (Section~\ref{sec:grading}).
-	\item 2017 March-August: Port Landslide to Bochs for use in Pintos OS classes (Section~\ref{sec:pintos-todo}).
+	\item 2017 March-August: Port Landslide to Bochs for use in Pintos OS classes (Section~\ref{chap:pintos}).
 	\item 2017 March-August: Implement, evaluate, and write paper about transactional memory project (Section~\ref{sec:tm}).
-	\item 2017 August-October: Polish and deploy Pintos user study (Section~\ref{sec:pintos-todo}).
+	\item 2017 August-October: Polish and deploy Pintos user study (Section~\ref{chap:pintos}).
 	\item 2017 ???ember: Resubmit paper about 410 user studies (Section~\ref{sec:studence}) (previously rejected from SIGCSE) (whenever an appropriate conference deadline arises).

### 2017-03-01 proposal/tm.tex
thesprop: split pintos and tm into their own chapters

--- 2016-12-28 proposal/lipservice.tex
@@
-\chapter{Broader Impact}
-\label{chap:lipservice}
+\chapter{Transactional Memory}
+\label{chap:tm}
 
-In this chapter I propose to add several enhancements to Landslide to apply its techniques beyond CMU's walls.
-%with little to no manual instrumentation effort required.
-This is the third of the three projects I am proposing for this thesis and so far exists only in dreams.
-The project will have two parts: adding support for Pintos \cite{pintos}, and adding support for Transactional Memory \cite{transactional-memory}.
+In this chapter I propose to extend Landslide's concurrency model to support Transactional Memory (TM) \cite{transactional-memory}.
+This is the last of the four projects I am proposing for this thesis and so far exists only in dreams.
 
-\section{Pintos}
-
-The Pintos architecture \cite{pintos}, outlined in Section~\ref{sec:overview-pintos}, provides an opportunity to test Landslide's pedagogical mettle beyond CMU's walls.
-For the experiments in Chapter~\ref{chap:quicksand}, I already implemented rudimentary support to test Pintos kernels.
-Because Pintos provides substantially more basecode than Pebbles, most if not all of Landslide's kernel instrumentation can be done automatically,
-using the names of the core scheduler functions already provided.
-
-{\bf Stuff I already did.}
-The existing Pintos support includes
-a new set of annotations (e.g., {\tt mutex\_lock} in Pebbles is now called {\tt sema\_down};
-handling several quirks of the basecode's scheduling behaviour (ask me if you really care to know);
-extending the heap tracker to handle Pintos's page allocator {\tt palloc} as well as the kernel {\tt malloc},
-as well as the fact that kernel memory is not direct-mapped like in Pebbles;
-new thread-liveness code to detect when a test case terminates;
-skipping some busy-wait loops in the boot sequence used for device communication,
-a complicated fall-back case in Landslide's timer injection algorithm the workings of which I don't even remember anymore,
-and extending the lock-set analysis to include Pintos's interrupt-disabling infrastructure (which was actually pretty tricky when using vector clocks for pure-HB).
-All these features apply to behaviours included with the stock base-code, so should apply generally to all but the most adventurous student implementations.
-
-{\bf Stuff I still need to do.}
-However, as you might have guessed, there are still a few cases where the existing instrumentation is not fully general.
-When preparing Chapter~\ref{chap:quicksand}'s experiments, I needed to adjust some student submissions to be compatible with Landslide's existing annotation process.
-In order to distribute Landslide for student use beyond CMU, it will need to support these cases automatically.
-These include:
-\begin{itemize}
-	\item Allowing for the priority scheduler's runqueues to be implemented as an array of queues, rather than the single queue head as provided in the basecode.
-	\item Insert the {\tt tell\_landslide\_sleeping} annotation more intelligently than looking for a {\tt list\_insert\_ordered} call (a basecode function which most, but not all, students use).
-	\item Some versions of the basecode distribute a {\tt sema\_up} implementation which {\tt yield}s unconditionally, which Landslide must bypass.
-	\item Some students use {\tt timer\_sleep} or {\tt while (!flag) continue} when they should use {\tt yield}, which can lead to false-positive deadlocks if not automatically replaced.
-	\item I'm sure I've forgotten some less-common ones, which I expect to handle on-the-fly when students email me Landslide crash reports.
-\end{itemize}
-When fixing my sample of Pintoses by hand, there were also myriad deterministic bugs, such as use-after-frees e.g.~arising from unsafe {\tt strlen} calls.
-As with Pebbles student experiments, however, I intend the students to encounter such bug reports and fix them on their own, despite not being concurrency bugs.
-It would be a Ph.D. unto itself to automatically ensure stable determinized execution of arbitrary student code.
-Actually, probably outright impossible.
-
-{\bf Experimental goals.}
-Having only one semester (ideally; or two, if my timeline ends up delayed) to test Landslide with Pintos students,
-I'll not be able to assemble as comprehensive a dataset as I have with 15-410 at CMU.
-This project will serve more as a proof-of-concept, demonstrating that model checking can realistically replace stress testing in operating systems courses worldwide.
-Nevertheless, I intend also to analyze what little data I can gather, comparing the incidences and types of concurrency bugs found between Pebbles and Pintos populations.
-This will perhaps shed light on the advantages and/or shortcomings of either project,
-leading to recommendations for improving either project's educational value (independently of using MC).
-
-\section{Transactional Memory}
-
-Transactional Memory (TM) is a mechanism by which programmers may avoid conventional concurrency primitives, optimizing for performance in the common case when threads do not conflict.
+TM is a mechanism by which programmers may avoid conventional concurrency primitives, optimizing for performance in the common case when threads do not conflict.
 A transactional program surrounds its critical section(s) with transaction begin/end statements, which ensure that no other thread can observe an intermediate state during the transaction.
@@
 
-\subsection{HTM}
+\section{HTM}
 
@@
 
-\subsection{STM}
+\section{STM}
 
@@
 
-\subsection{Hybrid HTM/STM}
+\section{Hybrid HTM/STM}
 
@@
 although I expect to relegate any implementation thereof to future work.
+
+\section{Evaluation Plan}
+
+% TODO

### 2017-03-01 proposal/overview.tex
add HTM example in overview

@@
 offering three new instructions: \texttt{xbegin}, \texttt{xend}, and \texttt{xabort}, to begin, commit, and fail a transaction, respectively.
+The example program in Figure~\ref{fig:htm-example} demonstrates how these instructions can be used to synchronize a simple shared access without locking overhead in the common case,
+using GCC's compiler intrinsics \cite{htm-gcc}.
+
+\begin{figure}[h]
+	\begin{center}
+		\begin{tabular}{l}
+		\textbf{(All threads)} \\
+		\hline
+		\texttt{if ((status = \_xbegin()) == \_XBEGIN\_STARTED) \{} \\
+		\texttt{~~~~if (too\_much(foo)) \{} \\
+		\texttt{~~~~~~~~\_xabort();} \\
+		\texttt{~~~~\} else \{} \\
+		\texttt{~~~~~~~~foo++;} \\
+		\texttt{~~~~~~~~\_xend();} \\
+		\texttt{~~~~\}} \\
+		\texttt{\} else if (status != \_XABORT\_EXPLICIT) \{} \\
+		\texttt{~~~~mutex\_lock(\&m);} \\
+		\texttt{~~~~foo++;} \\
+		\texttt{~~~~mutex\_unlock(\&m);} \\
+		\texttt{\}} \\
+		\end{tabular}
+	\end{center}
+	\caption{An example program demonstrating the various HTM primitives.
+		If the transaction in the top branch aborts,
+		whether from a conflict or from the programmer's intention,
+		execution will revert to the return of {\tt \_xbegin},
+		and {\tt status} will be assigned an error code indicating the abort reason.
+		The programmer can then use explicit synchronization, such as a mutex, to resolve the conflict.}
+	\label{fig:htm-example}
+\end{figure}
 

### 2017-03-01 proposal/tm.tex
write brief eval plan for tm

@@
 
-{\bf Mutex isomorphism.}
+\subsection{Mutex Isomorphism}
 When modelling TM in Landslide, we do not care about fidelity to performance characteristics or non-observable roll-back semantics.
@@
 
-{\bf Abort nondeterminism.}
+\subsection{Abort Nondeterminism}
 Next, I will extend the concurrency model to support the nondeterminism arising from transaction aborts.
@@
 
-{\bf Reduction challenge.}
+\subsection{Reduction Challenge}
+\label{sec:tm-reduction-challenge}
 I've identified a case under abort nondeterminism in which the current implementation of DPOR will fail to achieve a possible reduction.
@@
 
-% TODO
+Because this will be the first model checker for testing transactional programs
+(to my knowledge; see Section \ref{sec:related-tm}),
+I don't expect any prior work will exist for a direct performance comparison.
+Nevertheless, I would pose two main questions to evaluate Landslide's proposed TM support:
+\begin{itemize}
+	\item Does Landslide model TM concurrency accurately enough to find all the bugs and verify all the correct programs in a benchmark suite?
+	\item To what degree can DPOR reduce these programs' state spaces?
+Assuming I achieve the reduction challenge (Section \ref{sec:tm-reduction-challenge}), how much benefit is conferred, and how commonly across different types of programs?
+\end{itemize}
+
+Part of this work will be assembling a benchmark suite itself.
+I will investigate the benchmarks used in the Hybrid HTM/STM paper \cite{hybrid-htm-stm} (porting them to C if necessary),
+as well as more from various personal projects found online such as \cite{tm-benchmark-github} and \cite{tm-benchmark-cmu}.

### 2017-03-01 proposal/pintos.tex
garth's minor revisions

--- 2016-12-31 proposal/lipservice.tex
@@
-\chapter{Broader Impact}
-\label{chap:lipservice}
+\chapter{Pintos}
+\label{chap:pintos}
 
-In this chapter I propose to add several enhancements to Landslide to apply its techniques beyond CMU's walls.
-%with little to no manual instrumentation effort required.
-This is the third of the three projects I am proposing for this thesis and so far exists only in dreams.
-The project will have two parts: adding support for Pintos \cite{pintos}, and adding support for Transactional Memory \cite{transactional-memory}.
+The Pintos architecture \cite{pintos}, outlined in Section~\ref{sec:overview-pintos},
+provides an opportunity to test Landslide's pedagogical mettle beyond CMU's walls.
+Supporting Pintos as well as Pebbles will broaden Landslide's impact considerably,
+%as it will prove that the previous two chapters' techniques apply generally...
+as Pintos is already used by many more universities than Pebbles.
 
-\section{Pintos}
-
-The Pintos architecture \cite{pintos}, outlined in Section~\ref{sec:overview-pintos}, provides an opportunity to test Landslide's pedagogical mettle beyond CMU's walls.
-However, the existing scheduler and concurrency primitives which Pintos provides somewhat limits the remaining concurrency-critical code left to write,
-hence limiting the degree to which Landslide can test student submissions.
+The scheduler and concurrency primitives which Pintos's basecode provides
+limit the remaining concurrency-critical code left for students to write,
+which limits the degree to which Landslide can test student submissions.
 This is in contrast to the Pebbles assignments at CMU, in which students must implement all the functionality listed in Section~\ref{sec:overview-pintos}.
-On the other hand, the upside of Pintos providing substantially more basecode than Pebbles is that most, if not all, of Landslide's kernel instrumentation can be done automatically,
+On the other hand, the upside of Pintos providing substantially more basecode than Pebbles
+is that most, if not all, of Landslide's kernel instrumentation can be done automatically,
 using the names of the core scheduler functions already provided.
@@
 
-{\bf Stuff I already did.}
+\section{Existing Progress}
+
 The existing Pintos support includes
 a new set of annotations (e.g., {\tt mutex\_lock} in Pebbles is now called {\tt sema\_down};
-handling several quirks of the basecode's scheduling behaviour (ask me if you really care to know);
+handling several quirks of the basecode's scheduling behaviour; %(ask me if you really care to know);
 extending the heap tracker to handle Pintos's page allocator {\tt palloc} as well as the kernel {\tt malloc},
@@
 skipping some busy-wait loops in the boot sequence used for device communication,
-a complicated fall-back case in Landslide's timer injection algorithm the workings of which I don't even remember anymore,
-and extending the lock-set analysis to include Pintos's interrupt-disabling infrastructure (which was actually pretty tricky when using vector clocks for pure-HB).
+an extra fall-back case in Landslide's timer injection algorithm, %the workings of which I don't even remember anymore,
+and extending the lock-set and vector-clock analyses to include disabling interrupts.
+%(which was actually pretty tricky when using vector clocks for pure-HB).
 All these features apply to behaviours included with the stock base-code, so should apply generally to all but the most adventurous student implementations.
 
-{\bf Stuff I still need to do.}
-However, as you might have guessed, there are still a few cases where the existing instrumentation is not fully general.
-When preparing Chapter~\ref{chap:quicksand}'s experiments, I needed to adjust some student submissions to be compatible with Landslide's existing annotation process.
+\section{Proposed Work}
+
+%However, as you might have guessed,
+There are still a few cases where the existing instrumentation is not fully general.
+When preparing Chapter~\ref{chap:quicksand}'s experiments, I manually adjusted some student submissions to be compatible with Landslide's existing annotation process.
 In order to distribute Landslide for student use beyond CMU, it will need to support these cases automatically.
-These include:
+Examples of such ad-hoc fixes I would need to automate include:
 \begin{itemize}
@@
 	\item Some students use {\tt timer\_sleep} or {\tt while (!flag) continue} when they should use {\tt yield}, which can lead to false-positive deadlocks if not automatically replaced.
-	\item I'm sure I've forgotten some less-common ones, which I expect to handle on-the-fly when students email me Landslide crash reports.
+	%\item I'm sure I've forgotten some less-common ones, which I expect to handle on-the-fly when students email me Landslide crash reports.
 \end{itemize}
@@
 As with Pebbles student experiments, however, I intend the students to encounter such bug reports and fix them on their own, despite not being concurrency bugs.
-It would be a Ph.D. unto itself to automatically ensure stable determinized execution of arbitrary student code.
-Actually, probably outright impossible.
+%It would be a Ph.D. unto itself to automatically ensure
+Automatically ensuring stable determinized execution of arbitrary student code is beyond the scope of this thesis (if not outright impossible).
+%Actually, probably outright impossible.
 
-{\bf New emulation platform.}
-Most importantly, I will need to free Landslide from its dependence on Simics, which requires paid licenses for use beyond CMU's walls.
+\subsection{New emulation platform}
+
+Most practically, I will need to free Landslide from its dependence on Simics,
+which requires paid licenses for its use. %beyond CMU's walls.
 Other candidate emulation (simulation or virtualization) platforms for Landslide include Bochs, QEMU, VMWare, and Xen.
 I am presently studying which of these will most suit Landslide's needs,
-but so far Bochs seems like the best bet, which happens to be the simulator used at Berkeley, Stanford, and U. of Chicago anyway.
+but so far Bochs seems like the best bet,
+which the Berkeley, Stanford, and U. of Chicago classes already use as their simulator.
 
-{\bf Experimental goals.}
-Having only one semester (ideally; or two, if my timeline ends up delayed) to test Landslide with Pintos students,
+\subsection{Experimental goals}
+
+Having only two semesters to test Landslide with Pintos students,
 I'll be unable to gather as comprehensive a dataset as I have with 15-410 at CMU.
@@
 leading to recommendations for improving either project's educational value (independently of using MC).
-
-\section{Transactional Memory}
-\label{sec:tm}
-
-Transactional Memory (TM) is a mechanism by which programmers may avoid conventional concurrency primitives, optimizing for performance in the common case when threads do not conflict.
-A transactional program surrounds its critical section(s) with transaction begin/end statements, which ensure that no other thread can observe an intermediate state during the transaction.
-If a conflict is observed, the transaction {\em aborts}, rolling the program back to the transaction's initial state, and executing an optional back-up code path.
-The programmer may also explicitly abort the transaction using an abort statement.
-An example transactional program is shown in Figure~\ref{fig:txn-example}.
-
-\begin{figure}[t]
-	\begin{center}
-	\begin{tabular}{ll}
-		\multicolumn{2}{c}{Initially {\tt int foo, bar = 0; mutex\_t m;}} \\
-		\\
-		\begin{tabular}{l}
-		{\bf Thread 1} \\
-		\hline
-		\texttt{if (\_xbegin() ==} \\
-		\texttt{~~~~~~~~\_XBEGIN\_STARTED) \{} \\
-		\texttt{~~~~foo++;} \\
-		\texttt{~~~~\_xend();} \\
-		\texttt{\} else \{} \\
-		\texttt{~~~~mutex\_lock(\&m);} \\
-		\texttt{~~~~assert(foo > 0 ||} \\
-		\texttt{~~~~~~~~~~~bar > 0);} \\
-		\texttt{~~~~mutex\_unlock(\&m);} \\
-		\texttt{\}} \\
-		\end{tabular}
-		&
-		\begin{tabular}{l}
-		{\bf Thread 2} \\
-		\hline
-		\texttt{if (\_xbegin() ==} \\
-		\texttt{~~~~~~~~\_XBEGIN\_STARTED) \{} \\
-		\texttt{~~~~bar++;} \\
-		\texttt{~~~~\_xend();} \\
-		\texttt{\} else \{} \\
-		\texttt{~~~~mutex\_lock(\&m);} \\
-		\texttt{~~~~assert(foo > 0 ||} \\
-		\texttt{~~~~~~~~~~~bar > 0);} \\
-		\texttt{~~~~mutex\_unlock(\&m);} \\
-		\texttt{\}} \\
-		\end{tabular}
-	\end{tabular}
-	\end{center}
-	\caption{Example transactional program, written using GCC's transactional memory compiler intrinsics \cite{htm-gcc}.
-	%The x86 assembly instructions are named {\tt xbegin}, {\tt xend}, and {\tt xabort}; I have named the functions of this imaginary C interface the same.
-	Different behaviours are possible depending whether the transactions are backed by HTM or STM.}
-	\label{fig:txn-example}
-\end{figure}
-
-TM may be implemented either in hardware (HTM) \cite{htm-haswell}, or in software (STM) \cite{stm-pldi06}.
-Though their interfaces to the programmer are similar, their semantics demand a slightly different treatment from Landslide's perspective.
-The key difference is that HTM transactions may fail for any reason, beyond the scope of the program's behaviour, such as the CPU's cache being too full.
-STM transactions, on the other hand, will fail only if an actual conflict is observed from another thread.
-
-Consider again the example program: The transactions of the two threads do not conflict, so they may abort only under HTM.
-However, when they abort for a reason other than a conflict on {\tt foo} or {\tt bar}, the assertions in the backup code will fail.
-Hence, some programs which are correct under STM may contain bugs under HTM.
-Supporting TM in Landslide will consist of several steps, extending the concurrency model to incorporate failure injections and extending DPOR to determine when transaction aborts are possible depending on the underlying TM mechanism.
-
-\subsection{HTM}
-
-{\bf Mutex isomorphism.}
-When modelling TM in Landslide, we do not care about fidelity to performance characteristics or non-observable roll-back semantics.
-The goal of model checking is to exercise all observable program behaviours,
-so Landslide can model the execution of transactional programs using existing primitives if possible.
-In the first stage of the project, I will prove that a transactional program using is equivalent to one with a global mutex swapped for its {\tt xbegin}s and {\tt xend}s (assuming a retry-only policy for handling aborts).
-This will allow Landslide to test all observable TM behaviours using its existing infrastructure for mutexes,
-rather than relying on the platform providing accurate TM emulation.
-
-{\bf Abort nondeterminism.}
-Next, I will extend the concurrency model to support the nondeterminism arising from transaction aborts.
-During execution, Landslide inject a failure to force threads to branch into backup code paths.
-Failure injections add an extra ``dimension'' of non-determinism:
-at each {\tt xbegin} operation which is a preemption point, Landslide may force a normal context switch to re-interleave threads, or it may inject a transaction abort to test the backup code.
-(This also avoids the need to speculatively execute and/or roll-back failing transactions.)
-Because HTM transactions can fail for any reason, outside of the programmer's control, Landslide's DPOR implementation will consider failure injections always ``enabled''.
-
-{\bf Reduction challenge.}
-I've identified a case under abort nondeterminism in which the current implementation of DPOR will fail to achieve a possible reduction.
-Let $1A$, $1B$, $2A$, and $2B$ denote the transitions which execute the {\tt if} and {\tt else} branches by each thread, respectively.
-Note that $1A$ conflicts with $2B$, and $1B$ with $2A$, but $1A$ and $2A$ are independent, as are $1B$ and $2B$.
-Figure~\ref{fig:txn-graph} provides a visual aid.
-There are two possible reductions: pruning $2A,1A$ after testing $1A,2A$, and pruning $2B,1B$ after testing $1B,2B$;
-in other words, branches 5 and 8 can be skipped.
-However, %if we simply extended DPOR to always inject failures when possible,
-our current DPOR implementation will tag the $2B$ subtree to explore after observing the conflict in branch 2 (likewise the $2A$ subtree from branch 3),
-but then have no memory of whether it was subsequently supposed to run $1A$ or $1B$, and have to try both.
-To fix this will require an optimization analogous to ``sleep sets'' \cite{dpor}:
-the tag to explore the $2A$ or $2B$ subtree must include an annotation to remember whether or not the conflict arose after a failure injection in $1$.
-
-\begin{figure}[t]
-	\begin{center}
-		\includegraphics[width=0.8\textwidth]{htm-graph.pdf}
-	\end{center}
-	\caption{State space corresponding to the program in Figure~\ref{fig:txn-example}.
-	Conflicting transitions are marked in bold;
-	transitions which are independent from their predecessors are italicized.
-	End states colored the same (the cyan and yellow ones) are equivalent.
-	}
-	\label{fig:txn-graph}
-\end{figure}
-
-\subsection{STM}
-
-STM transactions abort only when multiple threads conflict.
-Because Landslide already computes memory conflicts among each pair of transitions, it will be natural to extend DPOR to consult the conflict set when deciding whether to exercise a failure injection.
-I will extend Landslide in this manner, and even support programs with both types of transactions, as long as the different {\tt xbegin} invocations are suitably annotated.
-
-\subsection{Hybrid HTM/STM}
-
-A recent paper \cite{hybrid-htm-stm} introduced several ways of combining HTM and STM in the same program, nesting transactions of different types.
-It presented several semantics for such transactions transactions, the most interesting of which being ``open nesting'', in which a nested transaction's state becomes visible to other threads even during a containing transaction.
-That state can then be rolled back if the latter aborts.
-I plan to develop a theoretical model for how such transactions would affect Landslide's concurrency model,
-although I expect to relegate any implementation thereof to future work.

### 2017-04-03 proposal/410.tex
thesprop: split pintos and tm into their own chapters
garth revisions for thesprop

@@
-\chapter{Education}
+\chapter{Pebbles}
 \label{chap:410}
@@
 (Actually, even the best 15-410 students still have concurrency bugs,
-but in principle, an educational tool must reach the more struggling students,
+but in principle, an educational tool should reach the more struggling students,
 the so-called ``middle'' and/or ``bottom'' of the class.)
@@
 
-Since then, I've extended Landslide to support testing P2 thread libraries (Section~\ref{sec:pebbles}) as well.
+Since then, I've extended Landslide to support testing Project 2 (P2) thread libraries (Section~\ref{sec:pebbles}) as well.
 Because P2 mandates specific function names for the project's internal APIs
@@
 However, arbitrary programs may break conventional disciplines of concurrent programs, while still being bug-free.
-For example, thread communication via ad-hoc {\tt yield} loops may appear as an infinite loop, livelock, or deadlock.
+For example, thread communication via ad-hoc {\tt yield} loops may appear to automated tools as a possible
+infinite loop, livelock, or deadlock.
 This is especially true of student code,
@@
 To be an effective tool for struggling students,
-a model checker must somehow coax adversarial programs to fit its concurrency model, to effectively test for real bugs,
+a model checker should somehow coax adversarial programs to fit its concurrency model, to effectively test for real bugs,
 rather than rejecting them outright on some stylistic or disciplinary grounds.
@@
 				\footnote{Implemented in {\tt check\_user\_yield\_activity} and {\tt check\_user\_xchg} in {\tt user\_sync.c}.}
+				The consequence of any this heuristic's inaccuracy is
+				that an unusual wait-loop might be classified as an infinite loop bug anyway.
 				% TODO: Wow, what happened to KEEP_RUNNING_YIELDING_THREADS?? I swear I had some logic in arbiter.c about that.
@@
 				% Or 20 for xchgs which end up being data-race PPs.
-				Landslide marks the thread blocked (or ``disabled'', in model-checking parlance),
+				Landslide speculatively marks the thread blocked (or ``disabled'', in model-checking parlance),
 				just as though it had invoked {\tt deschedule}.
@@
 		although it avoids a major assumption of the latter (threads {\tt yield} if and only if not making progress),
-		and also avoids the need to iteratively deepen the yield bound by fixing it as a heuristic constant.
+		and also avoids the need to iteratively deepen the yield bound by instead fixing it as a heuristic constant.
 		% Actually, I don't think that's true. Using DPOR to unblock a yield-looping thread at any conflict covers this.
@@
 		When a deadlock arises under conditions where one or more threads are heuristically blocked,
-		Landslide attempts to refute it as a false positive by artificially unblocking all heuristically-blocked threads%
+		Landslide attempts to refute it as a false positive by artificially unblocking all heuristically-blocked threads.
 		\footnote{If any threads are ICB-blocked, I prioritize waking those before trying to wake any yield-blocked threads.
@@
 		Waking all threads at once here can lead to unsoundness.}
-		If the deadlock is true, each thread will immediately trigger the yield-blocking heuristic again,
-		bringing the system back into deadlocked state.
 		Landslide then repeats this process a heuristic constant number of times (128),
 		allowing the program that many chances to make progress before proclaiming deadlock.
-		(Note that this heuristic cannot miss true deadlocks as false negatives.)
 		\footnote{Implemented in {\tt try\_avoid\_fp\_deadlock} in {\tt arbiter.c}.}
+		Note that this heuristic cannot miss true deadlocks as false negatives:
+		if the deadlock is true, each thread will immediately trigger the yield-blocking heuristic again,
+		bringing the system back into deadlocked state as many times as necessary to exhaust the heuristic limit.
 	\item {\bf Lock hand-off.}
@@
 		When releasing a lock, if absent from the current thread's lockset, Landslide searches the locksets of all other existing threads, and releases it there.
-		Landslide can instead optionally be configured to treat lock hand-off as an outright bug, as a matter of discipline.
+		Landslide can instead optionally be configured to treat lock hand-off as a style warning or as an outright bug, as a matter of discipline.
 		\footnote{Implemented in {\tt lockset\_remove} in {\tt lockset.c}.}
@@
 %After some collaboration with David Eckhardt (15-410 professor and member of this thesis committee),
-For the following cases, we configured Landslide to abort, and warn the student that they must find a better solution before it could test their code.
+For the following cases, I configured Landslide to abort, and warn the student that they must find a better solution before it could test their code (in other words, I promoted these patterns to be treated as bugs).
 
@@
 To analyze its effectiveness as a debugging tool for students,
-I configured it to record a snapshot of the results of each use by the students:
+I configured Landslide to record a snapshot of the inputs and results of each use by the students:
 which options were used (test program and time limit),
@@
 It found at least one concurrency bug for 32 groups (68\%),
-24 of whom (51\%) were able to fix at least one, verifying their update
+24 of whom (51\%) were able to fix at least one bug, verifying their update
 with a succesful re-run of the same test.
-
-The table's data presentation is somewhat confusing:
+\footnote{The table's data presentation is somewhat confusing:
 the dependent columns count the number of {\em groups}, categorized by how many bugs they found, not the raw number of {\em bugs},
-which I instead indicate in the ``Total bugs'' row at the bottom.
-So, the number of total bugs is equal to the sum of the products between each cell in that column and its corresponding number of bugs,
+instead indicated in the ``Total bugs'' row at the bottom.
+So, the number of total bugs is the sum of the products between each cell in that column and its corresponding number of bugs,
 e.g., $44 = 27\times0 + 6\times1 + 9\times2 + 2\times3 + 1\times4 + 2\times5$.
-Sorry about that; I'll try to find a better way to present this data in the thesis.
+%Sorry about that;
+I'll try to find a better way to present this data in the thesis.}
+I view this as a positive result -- among other statistics, Landslide was able to help the majority of students debug, among those who found concurrency bugs at all -- although it is difficult to control for the amount of time students spent with Landslide that would otherwise have been spent on old-fashioned stress testing.
 
@@
 The distribution of P3 grades is indistinguishable as well.
-In the thesis, I will discuss possible factors contributing to this failure, and how they might be controlled for in future work,
-although I am not planning to run any further experiments in the limited time I have.
+In the thesis, I will discuss possible factors contributing to this failure, and how they might be controlled for in future work.
+%although I am not planning to run any further experiments in the limited time I have.
 
@@
 having tuned the heuristics by hand on an (independent) set of P2 submissions already.)
-
-\section{Landslide as a Grading Tool}
-\label{sec:grading}
-
-Beginning in 2017, I will recruit 15-410 Teaching Assistants (TAs) for my user study rather than students.
-The modified user study will evaluate Landslide as a grading tool,
-alternative to the ``Fritz'' stress-testing infrastructure which 15-410 currently employs,
-and possibly substituting as well for the human process of reading student code looking for concurrency bugs by hand.
-The intent is not for Landslide to ultimately replace 15-410's manual grading process,
-but rather to demonstrate that comparably-rigorous grading could be deployed automatedly at other schools wishing to adopt the 15-410 curriculum.
-
-I will pursue the following evaluation questions:
-
-\begin{itemize}
-	\item Provided appropriate test cases, can Landslide find the most common ``deep'' concurrency bugs in student submissions?
-	\item How does Landslide's bug-finding compare to that of Fritz, the class's stress tester?
-	\item How does Landslide's bug-finding compare to that of TAs grading by hand?
-\end{itemize}
-
+%
+%\section{Landslide as a Grading Tool}
+%\label{sec:grading}
+%
+%Beginning in 2017, I will recruit 15-410 Teaching Assistants (TAs) for my user study rather than students.
+%The modified user study will evaluate Landslide as a grading tool,
+%alternative to the ``Fritz'' stress-testing infrastructure which 15-410 currently employs,
+%and possibly substituting as well for the human process of reading student code looking for concurrency bugs by hand.
+%The intent is not for Landslide to ultimately replace 15-410's manual grading process,
+%but rather to demonstrate that comparably-rigorous grading could be deployed automatedly at other schools wishing to adopt the 15-410 curriculum.
+%
+%I will pursue the following evaluation questions:
+%
+%\begin{itemize}
+%	\item Provided appropriate test cases, can Landslide find the most common ``deep'' concurrency bugs in student submissions?
+%	\item How does Landslide's bug-finding compare to that of Fritz, the class's stress tester?
+%	\item How does Landslide's bug-finding compare to that of TAs grading by hand?
+%\end{itemize}
+%
 Finally, I will conclude this chapter in the thesis by discussing how future work could adapt Pebbles's interface to support fully-automatic model checking without compromising the educational value of P3.

### 2017-04-03 proposal/pintos.tex
garth revisions for thesprop

@@
 
-The existing Pintos support includes
-a new set of annotations (e.g., {\tt mutex\_lock} in Pebbles is now called {\tt sema\_down};
-handling several quirks of the basecode's scheduling behaviour; %(ask me if you really care to know);
-extending the heap tracker to handle Pintos's page allocator {\tt palloc} as well as the kernel {\tt malloc},
+So far, just enough Pintos support exists to have run Chapter \ref{chap:quicksand}'s experiments. This includes:
+%The existing Pintos support includes
+\begin{itemize}
+	\item a new set of annotations (e.g., {\tt mutex\_lock} in Pebbles is now called {\tt sema\_down};
+	\item handling several quirks of the basecode's scheduling behaviour; %(ask me if you really care to know);
+	\item extending the heap tracker to handle Pintos's page allocator {\tt palloc} as well as the kernel {\tt malloc},
 		as well as the fact that kernel memory is not direct-mapped like in Pebbles;
-new thread-liveness code to detect when a test case terminates;
-skipping some busy-wait loops in the boot sequence used for device communication,
-an extra fall-back case in Landslide's timer injection algorithm, %the workings of which I don't even remember anymore,
-and extending the lock-set and vector-clock analyses to include disabling interrupts.
+	\item new thread-liveness code to detect when a test case terminates;
+	\item skipping some busy-wait loops in the boot sequence used for device communication;
+	\item an extra fall-back case in Landslide's timer injection algorithm; %the workings of which I don't even remember anymore,
+	\item and extending the lock-set and vector-clock analyses to include disabling interrupts.
+\end{itemize}
 %(which was actually pretty tricky when using vector clocks for pure-HB).
@@
 Other candidate emulation (simulation or virtualization) platforms for Landslide include Bochs, QEMU, VMWare, and Xen.
-I am presently studying which of these will most suit Landslide's needs,
-but so far Bochs seems like the best bet,
+%I am presently studying which of these will most suit Landslide's needs,
+%but so far Bochs seems like the best bet,
+I have begun work porting Landslide to Bochs, which is open-source and provides both instruction and memory tracing, and
 which the Berkeley, Stanford, and U. of Chicago classes already use as their simulator.
@@
 
-Having only two semesters to test Landslide with Pintos students,
-I'll be unable to gather as comprehensive a dataset as I have with 15-410 at CMU.
-This will serve more as a proof-of-concept, demonstrating that MC can realistically replace stress testing in operating systems courses worldwide.
-Nevertheless, I intend also to analyze what little data I can gather, comparing the incidences and types of bugs found between Pebbles and Pintos populations.
-This will perhaps shed light on the advantages and/or shortcomings of either project,
-leading to recommendations for improving either project's educational value (independently of using MC).
+%Having only two semesters to test Landslide with Pintos students,
+%I'll be unable to gather as comprehensive a dataset as I have with 15-410 at CMU.
+%This will serve more as a proof-of-concept, demonstrating that MC can realistically replace stress testing in operating systems courses worldwide.
+%Nevertheless, I intend also to analyze what little data I can gather, comparing the incidences and types of bugs found between Pebbles and Pintos populations.
+%This will perhaps shed light on the advantages and/or shortcomings of either project,
+%leading to recommendations for improving either project's educational value (independently of using MC).
+
+I aim to run a similar user study with Pathos students as I've done so far with Pebbles.
+After finding one or more Pintos-using classes at other schools to collaborate with,
+I will distribute Landslide to the students during the 'threads' and 'userprog' projects,
+have them run a similar suite of tests to those in Chapter \ref{chap:quicksand}'s experiments,
+and collect and analyze the results.
+I plan to analyze the overall incidences and types of bugs found between Pebbles and Pintos populations,
+which will perhaps shed light on the advantages and/or shortcomings of either project,
+and lead to recommendations for improving either project's educational value (whether by incorporating MC or otherwise).

### 2017-04-03 proposal/related.tex
write brief eval plan for tm
garth revisions for thesprop

@@
 To my knowledge, this is the first study of model checking in an educational setting, although teaching concurrency is not itself an unstudied problem.
-\cite{towards-a-framework} presents a promising framework for testing concurrent programs,
+Eytani et al.~\cite{towards-a-framework} present a promising framework for testing concurrent programs,
 which can incorporate model checking as well as static analysis, resource exhaustion, data-race analysis, and coverage analysis.
@@
 %
-\cite{how-studence} presents a survey of how students think about concurrent program behaviour and debugging,
-while \cite{learning-concurrency} investigates how students form cognitive patterns about concurrent programming that could either aid or stunt their reasoning.
+L\"{o}nnberg et al.~\cite{how-studence} present a survey of how students think about concurrent program behaviour and debugging,
+while Kolikant \cite{learning-concurrency} investigates how students form cognitive patterns about concurrent programming that could either aid or stunt their reasoning.
 Both of these studies could help optimize Landslide's bug reports for clarity and student enlightenment.
@@
 Replay analysis extends single-pass data-race analysis to apply model checking (in limited quantities) to classify data-race candidates by their impact on program behaviour.
-It was first introduced in \cite{recordreplaydrs}, which compares the program states immediately after the access pair for differences, preferring still to err on the side of false positives.
+It was first introduced by Narayanasamy et al.~\cite{recordreplaydrs}, which compares the program states immediately after the access pair for differences, preferring still to err on the side of false positives.
 RaceFuzzer \cite{racefuzzer} avoids false positives by requiring an actual failure be exhibited, as we do,
@@
 
-Transactional memory (TM), first introduced in \cite{transactional-memory},
+\section{Transactional Memory}
+\label{sec:related-tm}
+
+Transactional memory (TM), first introduced in 1993 \cite{transactional-memory},
 has received renewed attention in recent years with the announcement of Intel's Haswell architecture \cite{htm-haswell},
-which promises to support hardware transactions using new x86 instructions.
+which supports hardware transactions using new x86 instructions.
 Because hardware transactions can fail for any reason, not just on memory conflicts,
 software TM remains relevant,
-and recent works such as \cite{hybrid-htm-stm} and \cite{stm-relaxed-memory} enhance it to be nested with hardware transactions, or to be used on weak memory architectures, respectively;
+and recent works \cite{hybrid-htm-stm,stm-relaxed-memory} enhance it to be nested with hardware transactions, or to be used on weak memory architectures, respectively;
 which in turn produce ever more complicated concurrency models for people like me to tackle.
@@
 but not that of client programs which use TM.
+SI-TM \cite{si-tm} introduces techniques for reducing HTM's abort rates, but without fully eliminating aborts, MCs must still consider them possible anywhere.
 Getting closer, McRT STM \cite{mc-tm-with-spin} uses SPIN \cite{spin} to model check an STM implementation

### 2017-04-03 proposal/thesprop.tex
garth revisions for thesprop

@@
 Stateless model checking is a powerful technique for testing concurrent programs,
-capable in theory of rooting out any bug or providing total verification on any program,
+capable in theory of rooting out any nondeterministic bug or providing total verification on any program,
 but suffers several problems which relegate that theory to fantasy land \cite{vargomax}.

### 2017-04-03 proposal/tm.tex
garth's minor revisions
garth revisions for thesprop

@@
 The programmer may also explicitly abort the transaction using an abort statement.
-An example transactional program is shown in Figure~\ref{fig:txn-example}.
+An example transactional program is shown in Figure~\ref{fig:txn-example} (slightly extended from the prior example in Section \ref{sec:overview-tm}).
 
@@
 	%The x86 assembly instructions are named {\tt xbegin}, {\tt xend}, and {\tt xabort}; I have named the functions of this imaginary C interface the same.
-	Different behaviours are possible depending whether the transactions are backed by HTM or STM.}
+	%Different behaviours are possible depending whether the transactions are backed by HTM or STM.}
+	These transactions would only abort if backed by HTM, while under STM their disjoint memory accesses guarantee them to succeed.}
 	\label{fig:txn-example}
@@
 STM transactions, on the other hand, will fail only if an actual conflict is observed from another thread.
-
 Consider again the example program: The transactions of the two threads do not conflict, so they may abort only under HTM.
@@
 Hence, some programs which are correct under STM may contain bugs under HTM.
-Supporting TM in Landslide will consist of several steps, extending the concurrency model to incorporate failure injections and extending DPOR to determine when transaction aborts are possible depending on the underlying TM mechanism.
+
+Supporting TM in Landslide will consist of two major steps:
+extending the concurrency model to incorporate failure injections
+and extending DPOR to determine when transaction aborts are possible.
+Failure injections can model the semantics of transaction aborts, as each {\tt \_xbegin()} call can nondeterministically return either success ({\tt \_XBEGIN\_STARTED}) or some abort failure code (such as {\tt \_XABORT\_CONFLICT} or {\tt \_XABORT\_CAPACITY}).
+Moreover, the extent of this nondeterminism depends on the underlying TM mechanism, which I will discuss shortly.
+%in STM, only {\tt \_XABORT\_CONFLICT} may arise, and only when
+(The {\tt \_XABORT\_EXPLICIT} code is never generated nondeterministically but rather by a user invocation of {\tt \_xabort()} from within the success path.)
 
@@
 so Landslide can model the execution of transactional programs using existing primitives if possible.
-In the first stage of the project, I will prove that a transactional program using is equivalent to one with a global mutex swapped for its {\tt xbegin}s and {\tt xend}s (assuming a retry-only policy for handling aborts).
+In the first stage of the project, I will prove that a transactional program using is equivalent
+to one with a global mutex swapped for its {\tt xbegin}s and {\tt xend}s. %(assuming a retry-only policy for handling aborts).
+Here I will assume a retry-only policy for handling aborts;
+in other words, transaction aborts could never direct the program to execute a different logic branch entirely.
 This will allow Landslide to test all observable TM behaviours using its existing infrastructure for mutexes,
@@
 Next, I will extend the concurrency model to support the nondeterminism arising from transaction aborts.
-During execution, Landslide inject a failure to force threads to branch into backup code paths.
+During execution, Landslide will inject a failure to force threads to branch into backup code paths.
 Failure injections add an extra ``dimension'' of non-determinism:
 at each {\tt xbegin} operation which is a preemption point, Landslide may force a normal context switch to re-interleave threads, or it may inject a transaction abort to test the backup code.
-(This also avoids the need to speculatively execute and/or roll-back failing transactions.)
-Because HTM transactions can fail for any reason, outside of the programmer's control, Landslide's DPOR implementation will consider failure injections always ``enabled''.
+(This also avoids the need to speculatively execute and/or roll-back failing transactions.
+Being not testing the TM implementation itself, I'll assume it correctly rolls-back failed transactions non-observably.)
+
+HTM transactions can fail for several reasons outside of a programmer's control,
+including cache evictions,
+false sharing (disjoint simultaneous accesses which nevertheless share the same cache line),
+and page faults or interrupts \cite{si-tm}.
+For example, in Figure \ref{fig:txn-example}, the globals {\tt foo} and {\tt bar} are likely to share the same cache-line and lead to false sharing conflicts.
+Accordingly, Landslide's DPOR implementation will consider failure injections always ``enabled'' under HTM.
 
@@
 \label{sec:tm-reduction-challenge}
-I've identified a case under abort nondeterminism in which the current implementation of DPOR will fail to achieve a possible reduction.
-Let $1A$, $1B$, $2A$, and $2B$ denote the transitions which execute the {\tt if} and {\tt else} branches by each thread, respectively.
-Note that $1A$ conflicts with $2B$, and $1B$ with $2A$, but $1A$ and $2A$ are independent, as are $1B$ and $2B$.
+I've identified at least one case under abort nondeterminism in which the current implementation of DPOR will fail to achieve a possible reduction.
+Let $1A$, $1B$, $2A$, and $2B$ denote the transitions which execute the {\tt if} and {\tt else} branches by each thread in Figure~\ref{fig:txn-example} respectively.
+Note that $1A$ conflicts with $2B$ (thread 1 writes {\tt foo}; thread 2 reads). and $1B$ with $2A$ (same for {\tt bar}).
+Meanwhile, $1A$ and $2A$ are independent (disjint writes to {\tt foo} or {\tt bar} only), as are $1B$ and $2B$ (reads versus reads).
 Figure~\ref{fig:txn-graph} provides a visual aid.
+
 There are two possible reductions: pruning $2A,1A$ after testing $1A,2A$, and pruning $2B,1B$ after testing $1B,2B$;
@@
 
-STM transactions abort only when multiple threads conflict.
+% TODO: Verify doesn't include false sharing of cache lines.
+STM transactions abort only when multiple threads conflict directly on the same memory address(es) -- the same sense of conflict that DPOR already considers.
 Because Landslide already computes memory conflicts among each pair of transitions, it will be natural to extend DPOR to consult the conflict set when deciding whether to exercise a failure injection.
@@
 I plan to develop a theoretical model for how such transactions would affect Landslide's concurrency model,
-although I expect to relegate any implementation thereof to future work.
+and justify its correctness as an extension of the HTM or STM models alone.
+%although I expect to relegate any implementation thereof to future work.
 
@@
 	\item Does Landslide model TM concurrency accurately enough to find all the bugs and verify all the correct programs in a benchmark suite?
-	\item To what degree can DPOR reduce these programs' state spaces?
+	\item To what degree can an improved DPOR reduce these programs' state spaces?
 Assuming I achieve the reduction challenge (Section \ref{sec:tm-reduction-challenge}), how much benefit is conferred, and how commonly across different types of programs?

### 2017-04-04 proposal/thesprop.tex
fix broken section references

@@
 	\item 2017 ???uary: Present thesis proposal (whenever the committee is free).
-	\item 2017 February-March: Develop ``grading'' TA user study and deploy it in 15-410 (Section~\ref{sec:grading}).
+	\item 2017 February-March: Deploy 15-410 student user study for more data (Section~\ref{sec:studence}).
 	\item 2017 March-August: Port Landslide to Bochs for use in Pintos OS classes (Section~\ref{chap:pintos}).
-	\item 2017 March-August: Implement, evaluate, and write paper about transactional memory project (Section~\ref{sec:tm}).
+	\item 2017 March-August: Implement, evaluate, and write paper about transactional memory project (Section~\ref{chap:tm}).
 	\item 2017 August-October: Polish and deploy Pintos user study (Section~\ref{chap:pintos}).
 	\item 2017 ???ember: Resubmit paper about 410 user studies (Section~\ref{sec:studence}) (previously rejected from SIGCSE) (whenever an appropriate conference deadline arises).
-	\item 2017 October-November: Deploy 410 TA grading study again for more data.
+	\item 2017 October-November: Deploy 410 student study again for more data.
 	\item 2017 December: Begin writing thesis.

### 2017-04-24 proposal/talk/proposal.tex
write some thesprop slides

\documentclass[xcolor=dvipsnames]{beamer}
\usepackage{lmodern}
\usepackage[T1]{fontenc}
\usepackage[english]{babel}
\usepackage[utf8]{inputenc}

\usepackage{manfnt}
\usepackage{wasysym}
\usepackage{listings}
\usepackage{graphicx}
\usepackage{url}
\usepackage{ulem}
\usepackage{marvosym}
\usepackage{skull}
\usepackage{proof}
\usepackage{array}
\usepackage{colortbl}
\usepackage{xspace}
\setbeamertemplate{navigation symbols}{}

\title[Thesis Proposal]{{\bf Practical Concurrency Testing}}
\subtitle[]{a thesis proposal}
\author[Ben Blum]{Ben Blum \texttt{(bblum@cs.cmu.edu)}}

\institute[CMU CSD]{Carnegie Mellon University}
\date[]{2017, April 25}

\setbeamertemplate{footline}{\hspace*{.5cm}\scriptsize{\insertauthor\hspace*{50pt} \hfill\insertframenumber\hspace*{.5cm}}}

\usecolortheme{seahorse}
\usecolortheme{rose}
\useoutertheme{infolines}

\usecolortheme[named=ForestGreen]{structure}

\newcommand\noob{\mathsf{noob}}
\newcommand\gibs{\mathsf{gibs}}
\newcommand\dps{\mathsf{dps}}
\newcommand\squig\rightsquigarrow
\newcommand\Coloneqq{\mathrel{\mathop{::}}=}
\newcommand\dmg{\text{\Laserbeam}}
\newcommand\delter\delta
\newcommand\alpher\alpha
\newcommand\defnor{\text{ }|\text{ }}

\newcommand\pimp{\mathop{\supset}}
\newcommand\pand{\mathop{\wedge}}
\newcommand\por{\mathop{\vee}}
\newcommand\ptrue{\top}
\newcommand\pfalse{\bot}

\newcommand\hilight[2]{\color{#1}#2\color{black}}
\definecolor{olivegreen}{RGB}{0,127,0}

\begin{document}
%\renewcommand{\inserttotalframenumber}{39}
\normalem
\begin{frame}
	\titlepage
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\newcommand\linegap{\vspace{0.2in}}
\newcommand\breakslide[1]{\begin{frame}{} \begin{center} #1 \end{center} \end{frame}}

\section{Introduction}
\subsection{Motivation}

\begin{frame}{Thesis Statement}
	Thanks to the new algorithms, heuristics, and concurrency models I have developed,
	stateless model checking is an appropriate and accessible concurrency testing technique
	for programmers in both educational and real-world settings.
\end{frame}

\begin{frame}{Outline}
	guess i should outline my talk huh
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{some kinda introductory}

%\begin{frame}{overview of state of art}
%	\begin{itemize}
%		\item so like, you got all this threads, right?
%		\item n then you just make em go any which way
%		\item see what all bugs come out
%	\end{itemize}
%\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Education}

\breakslide{\Large Educational Use}

\begin{frame}{Pebbles}
	CMU (15-410)
	\linegap

	Kernel (``P3'')
	\begin{itemize}
		%\item UNIX-like educational kernel specification
		\item Students implement scheduler, synchronization, VM %from scratch %% trying not to be too arrogant ._.
		\item UNIX-like process lifecycle ({\tt fork}, {\tt exec}, {\tt vanish}, {\tt wait})
		%\item Teams of 2
		\item Testing with MC requires manual annotation {\em [Blum '12]}
	\end{itemize}
	\pause
	\linegap

	Thread library (``P2'')
	\begin{itemize}
		\item POSIX-like userspace thread library %(also teams of 2)
		\item Students provided reference Pebbles implementation
		\item Fixed APIs $\rightarrow$ instrumentation can be automated
		\item May still need heuristics to handle ad-hoc synchronization
	\end{itemize}
\end{frame}

\begin{frame}{15-410 User Study}
	Method (since spring 2015)
	\begin{itemize}
		\item Guest lecture midway during P2
		\item Landslide (+ Quicksand) provided with 6 test cases and user guide
		\item Student code snapshots and bugs found recorded each use
	\end{itemize}
	\pause
	\linegap

	Results (as of spring 2016)
	\begin{itemize}
		\item 47/90 groups volunteered
		\item 85 concurrency bugs found; 53 which students then fixed
		\item No visible impact on ultimate grades...
	\end{itemize}
\end{frame}

\begin{frame}{15-410 User Study - Grades Results} % lack thereof
	\begin{center}
		\includegraphics[width=\textwidth]{../p2-p3-distribution.pdf}
	\end{center}
\end{frame}

%%%%

\begin{frame}{Pintos}
	U. of Chicago (CMSC 23000), Berkeley (CS162), Stanford (CS140)
	\linegap

	Kernel (``threads'')
	\begin{itemize}
		\item Students implement priority scheduling, sleep
		\item Basecode provides round-robin scheduler, synchronization
	\end{itemize}
	\linegap

	Processes (``userprog'')
	\begin{itemize}
		\item UNIX-like process lifecycle ({\tt exec}, {\tt exit}, {\tt wait})
		\item File descriptor management ({\tt create}, {\tt remove}, {\tt open}, {\tt read}, ...)
	\end{itemize}
	\linegap

	VM, filesys projects not as concurrency-focused as above
\end{frame}

\begin{frame}{Pintos \& Landslide}
	Existing support:
	\begin{itemize}
		\item Lockset/vector-clock analyses extended to interrupt-disabling
		\item Automatically inserting (most) kernel annotations
		%\item (various engineering challenges) % palloc heap, non-direct-mapped kernelspace, skipping driver init loops
		\item Often requires manual intervention
	\end{itemize}
	% "Works enough for OOPSLA but not enough for the studence"
	\linegap

	Still needed:
	\begin{itemize}
		\item Improved automation to support corner-case student designs
		\item More concurrency test cases
		\item Remove proprietary Simics dependence by porting to Bochs
	\end{itemize}
\end{frame}

\begin{frame}{Deliverables}
	Pebbles
	\begin{itemize}
		\item Discuss factors keeping Landslide from improving grades
		%\item Discuss how to make P3 and Landslide compromise to work together
		\item Study accuracy of automation heuristics
	\end{itemize}
	\linegap

	Pintos
	\begin{itemize}
		\item Open-source port of Landslide to Bochs
		\item 2 semesters of user study in Pintos class(es)
		\item Compare incidence/nature of bugs found to Pebbles
	\end{itemize}
	\linegap
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Transactional Memory}
\newcommand\xbegin{\texttt{\_xbegin()}\xspace}
\newcommand\xend{\texttt{\_xend()}\xspace}
\newcommand\xabort{\texttt{\_xabort()}\xspace}

\breakslide{\Large Transactional Memory}

\begin{frame}{Example}
	\begin{center}
		\begin{tabular}{l}
			\texttt{if ((status = \_xbegin()) == \_XBEGIN\_STARTED) \{} \\
			%\texttt{} \\
			%\texttt{} \\
			\texttt{~~~~foo++;} \\
			\texttt{~~~~\_xend();} \\
			\texttt{\} else \{}\\ %if (status != \_XABORT\_EXPLICIT) \{} \\
			\texttt{~~~~mutex\_lock(\&foo\_lock);} \\
			\texttt{~~~~foo++;} \\
			\texttt{~~~~mutex\_unlock(\&foo\_lock);} \\
			\texttt{\}} \\
		\end{tabular}
	\end{center}
\end{frame}

\begin{frame}{HTM Execution Semantics}
	Attempting a transaction
	\begin{itemize}
		\item Code between \xbegin and \xend is provisionally executed
		\item All writes kept in cache until \xend
		\item \xend commits changes atomically (w.r.t. other CPUs)
	\end{itemize}
	\pause
	\linegap

	Transaction failure
	\begin{itemize}
		\item Possible reasons:
			\begin{itemize}
				\item Memory conflict with another thread
				\item False sharing (conflict on same cache-line)
				\item Cache eviction, random interrupt, ...
			\end{itemize}
		\item Changes discarded, local state reverted, \xbegin returns failure
		\item {\tt else} branch runs some user-supplied backup plan
	\end{itemize}
\end{frame}

\begin{frame}{Model Checking Transactional Programs}
	What does it mean to MC a program?
	\begin{itemize}
		\item Capture all sites of nondeterminism as ``branch points'' %in a state space
		\item Control nondeterminism to force each possibility at each point
		\item (Identify equivalences to prune unnecessary branch points)
	\end{itemize}
	\pause
	\linegap

	MCing a transactional program
	\begin{itemize}
		\item Simulate TM, interleaving transactions, rolling-back upon conflicts
			\begin{itemize}
					% TODO: need a bonus slide to justify why this
					% imagine a pair of transactions each with multiple conflicting accesses
					% they should be treated as atomic
				\item slow; invites state space explosion
			\end{itemize}
		\item Enforce atomicity via scheduler, anticipate aborts via analysis
			\begin{itemize}
				\item (under HTM, all transactions might abort)
					% TODO: need a bonus slide on finding when a STM txn may abort.
			\end{itemize}
		\item Suffices to enumerate all {\em observationally-equivalent} schedules.
	\end{itemize}
\end{frame}

\begin{frame}{Concurrency model}
	What are a transactional program's {\em observable} behaviours?
	\begin{itemize}
		\item Transactional writes, if successful, appear atomically %to other threads
		\begin{itemize}
			\item Equivalent to a lock protecting that state
		\end{itemize}
			\pause
		\item Simultaneous transactions succeed only if no conflicts
		\begin{itemize}
			\item Equivalent (modulo performance) to serialize independent transactions
		\end{itemize}
			\pause
		\item Failing transactions' effects rolled-back with no trace
		\begin{itemize}
			\item Treat each \xbegin as a failure injection point
		\end{itemize}
	\end{itemize}
\end{frame}

\begin{frame}{Failure Injection - 2 dimensions of concurrency}
	% TODO - pict of mario man
\end{frame}

\begin{frame}{Deliverables}
	In the thesis, I will:
	\begin{itemize}
		\item Implement the HTM(-equivalent) concurrency model in Landslide
		\item Extend the model to support STM
		\item Prove equivalence to HTM/STM semantics % under MC
		\item Bugfind/verify a suite of unit tests \& real-world TM programs
	\end{itemize}
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{End}

\begin{frame}{Timeline}
	\begin{itemize}
		\item 2017 Mar-Aug: Port Landslide to Bochs for Pintos
		\item 2017 Mar-Aug: Transactional memory
		\item 2017 Aug-Oct: Pintos user study
		\item 2017 Oct-Nov: Pebbles user study (again)
		\item 2017 Dec:     Start writing thesis
		\item 2018 Jan-Apr: More writing \& miscellany
		\item 2018 Feb-Mar: Pintos \& Pebbles (again)
		\item 2018 Apr-May: Defend \& graduate
	\end{itemize}
\end{frame}

\breakslide{\Large Q\&A
%\linegap
%
%\begin{center}
%	\includegraphics[width=0.65\textwidth]{3word-questions.png}
%\end{center}
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Bonus Slides}

\begin{frame}{HTM: Even more dimensions of nondeterminism}
	% The gotcha: When xbegin can fail, it can emit diff error codes, which depend on what happened within.
	\begin{center}
		\begin{tabular}{l}
			\texttt{if ((status = \_xbegin()) == \_XBEGIN\_STARTED) \{} \\
			\texttt{~~~~if (foo < 100)}\\
			\texttt{~~~~~~~~\_xabort();} \\
			\texttt{~~~~foo++;} \\
			\texttt{~~~~\_xend();} \\
			\texttt{\} else if (status != \_XABORT\_EXPLICIT) \{} \\
			\texttt{~~~~mutex\_lock(\&foo\_lock);} \\
			\texttt{~~~~foo++;} \\
			\texttt{~~~~mutex\_unlock(\&foo\_lock);} \\
			\texttt{\}} \\
		\end{tabular}
	\end{center}
	\linegap

	\xbegin can fail with multiple possible error codes
	\begin{itemize}
		\item MC will identify on-the-fly when (e.g.) {\tt \_XABORT\_EXPLICIT} is possible
		\item Treat such transactions as having multiple failure injection possibilities
			% XXX: How to identify when the failure status is never used? Dataflow analysis?
	\end{itemize}
\end{frame}

\begin{frame}{HTM: Possible \xabort error codes}
	From GCC 4.8.2 manual \S 6.56.8, ``X86 transaction memory intrinsics'':
	\begin{itemize}
		\item {\tt \_XABORT\_EXPLICIT} - Transaction explicitely [sic] aborted with \xabort.
			%The parameter passed to \xabort is available with {\tt \_XABORT\_CODE(status)}
		\item {\tt \_XABORT\_RETRY} - Transaction retry is possible.
		\item {\tt \_XABORT\_CONFLICT} - Transaction abort due to a memory conflict with another thread
		\item {\tt \_XABORT\_CAPACITY} - Transaction abort due to the transaction using too much memory
		\item {\tt \_XABORT\_DEBUG} - Transaction abort due to a debug trap
		\item {\tt \_XABORT\_NESTED} - Transaction abort in a inner nested transaction 
	\end{itemize}
\end{frame}

\end{document}

### 2017-04-24 proposal/talk/proposal.tex
finish writing proposal slides barring some bonuses and picts of mario man

@@
 
-\begin{frame}{Thesis Statement}
-	Thanks to the new algorithms, heuristics, and concurrency models I have developed,
-	stateless model checking is an appropriate and accessible concurrency testing technique
-	for programmers in both educational and real-world settings.
+\begin{frame}{Concurrency!}
+	\begin{center}
+	\begin{tabular}{ll}
+		Initially \texttt{x = 0;} \\
+		\\
+		{\bf Thread 1} & {\bf Thread 2} \\
+		\\
+		\texttt{\hilight{orange}{mutex\_lock}(m);}      & \texttt{atomic\_xadd(\&x, 1);} \\
+		\texttt{x++;}				   & \texttt{\hilight{olivegreen}{yield}();} \\
+		\texttt{\hilight{blue}{mutex\_unlock}(m);}      & \texttt{atomic\_xadd(\&x, 1);} \\
+		\texttt{assert(x >= 1);}			& \texttt{assert(x >= 2);} \\
+		\\
+		\\
+		\\
+		\\
+		\\
+	\end{tabular}
+	\end{center}
+\end{frame}
+\begin{frame}{Concurrency!}
+	\begin{center}
+	\begin{tabular}{ll}
+		Initially \texttt{x = 0;} \\
+		\\
+		{\bf Thread 1} & {\bf Thread 2} \\
+		\\
+		\texttt{\hilight{orange}{mutex\_lock}(m);} \\
+		\texttt{int tmp = x;} \\
+								& \texttt{atomic\_xadd(\&x, 1);} \\
+								& \texttt{\hilight{olivegreen}{yield}();} \\
+								& \texttt{atomic\_xadd(\&x, 1);} \\
+		\texttt{x = tmp + 1;} \\
+								& \texttt{\hilight{red}{assert(x >= 2);}} \\
+		\texttt{\hilight{blue}{mutex\_unlock}(m);} \\
+		\texttt{assert(x >= 1);}
+	\end{tabular}
+	\end{center}
 \end{frame}
 
-\begin{frame}{Outline}
-	guess i should outline my talk huh
+\begin{frame}{Stateless Model Checking}
+	Stateless Model Checking {\em [Godefroid '97]} (MC)
+	\begin{itemize}
+		\item Test framework controls thread scheduling
+		\item Each test iteration, test a different interleaving
+		\item Goal: Exhaustively check all possible program behaviours
+	\end{itemize}
+	\pause
+	\linegap
+
+	\begin{center}
+		\includegraphics[width=0.25\textwidth]{../../oopsla/exponential-curve.pdf}
+	\end{center}
 \end{frame}
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\begin{frame}{State Space Example}
+	\begin{center}
+		\includegraphics[width=0.96\textwidth]{../../oopsla/tree-maximal-only.pdf}
+	\end{center}
+	\linegap
 
-\section{some kinda introductory}
+	Possible interleavings represented as a tree
+	\begin{itemize}
+		\item Node: Intermediate execution state
+		\item Edge: State transition from executing a thread
+	\end{itemize}
+	%\linegap
+
+	%State space size depends on which {\bf preemption points} are used.
+\end{frame}
+
+\begin{frame}{Challenges} %\& Existing Coping Techniques}
+	Unpredictable state space explosion
+	\begin{itemize}
+		\item Equivalence reductions: DPOR {\em [Flanagan '05]}, MCR {\em [Huang '15]}
+		\item Heuristic search ordering: ICB {\em [Musuvathi '08]}, PCT {\em [Burckhardt '10]}
+		\item There will always be a {\bf bigger test case}!
+	\end{itemize}
+	\pause
+	\linegap
+
+	Too much manual intervention for non-expert use
+	\begin{itemize}
+		\item Manual kernel annotations: Landslide {\em [Blum '12]} {\tiny that's me!}
+		\item State space size estimation {\em [Simsa '12]}
+		\item Widespread adoption requires {\bf fully automated instrumentation}!
+	\end{itemize}
+	\pause
+	\linegap
+
+	Limited sources of nondeterminism
+	\begin{itemize}
+		\item Timer-driven schedule nondeterminism (all prior work)
+		\item Weak memory nondeterminism {\em [Zhang '15]}
+		\item Event-driven applications {\em [Jensen '15]}
+		\item {\bf Transactional concurrency}?
+	\end{itemize}
+\end{frame}
 
@@
 %\end{frame}
+
+\begin{frame}{Thesis Statement}
+	Thanks to the new algorithms, heuristics, and concurrency models I have developed,
+	stateless model checking is an appropriate and accessible concurrency testing technique
+	for programmers in both educational and real-world settings.
+\end{frame}
+
+\definecolor{gray}{RGB}{128,128,128}
+
+\begin{frame}{Thesis Statement}
+	The Iterative Deepening algorithm makes MC appropriate for both educational and real-world programs.
+	\vspace{4.14em}
+
+	Ad-hoc synchronization heuristics make MC accessible to student programmers.
+	\vspace{4.14em}
+
+	A new transactional concurrency model keeps MC applicable to the latest real-world techniques.
+	\vspace{4.14em}
+\end{frame}
+\begin{frame}{Outline}
+	\hilight{gray}{The Iterative Deepening algorithm makes MC appropriate for both educational and real-world programs.} \\
+	\begin{itemize}
+		\item {\bf Quicksand and Iterative Deepening}
+		\begin{itemize}
+			\item Algorithm overview \& data-race preemption points
+			\item Verification \& experimental results
+		\end{itemize}
+	\end{itemize}
+	\hilight{gray}{Ad-hoc synchronization heuristics make MC accessible to student programmers.} \\
+	\begin{itemize}
+		\item {\bf Educational use}
+		\begin{itemize}
+			\item Pebbles \& 15-410 user study
+			\item Pintos \& upcoming plans
+		\end{itemize}
+	\end{itemize}
+
+	\hilight{gray}{A new transactional concurrency model keeps MC applicable to the latest real-world techniques.} \\
+	\begin{itemize}
+		\item {\bf Transactional memory}
+		\begin{itemize}
+			\item Example \& challenges
+			\item Equivalence \& failure injection
+		\end{itemize}
+	\end{itemize}
+\end{frame}
+
+%\begin{frame}{Outline}
+%	guess i should outline my talk huh
+%\end{frame}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{Quicksand}
+
+\breakslide{\Large Iterative Deepening of State Spaces}
+
+\begin{frame}{Preemption Points}
+	The burning question: ``Which {\bf preemption points} (PPs) are important?''
+	\linegap
+
+	State space of interleavings is {\em parameterized} by PPs.
+	\begin{itemize}
+		\item Too few PPs: Many bugs will go undetected
+		\item Preempt everywhere: Completion is often infeasible % say: "for some programs"
+	\end{itemize}
+	\linegap
+
+	Prior model checkers hard-code a fixed set of PPs, committing to one state space in advance.
+	% say: "...then attempt to cope with the resulting exponentially sized problem, however big it may be"
+\end{frame}
+
+\begin{frame}{Example (Revisited)}
+	\begin{center}
+	\begin{tabular}{ll}
+		Initially \texttt{x = 0;} \\
+		\\
+		{\bf Thread 1} & {\bf Thread 2} \\
+		\\
+		\texttt{\hilight{orange}{mutex\_lock}(m);} \\
+		\texttt{int tmp = x;} \\
+								& \texttt{atomic\_xadd(\&x, 1);} \\
+								& \texttt{\hilight{olivegreen}{yield}();} \\
+								& \texttt{atomic\_xadd(\&x, 1);} \\
+		\texttt{x = tmp + 1;} \\
+								& \texttt{\hilight{red}{assert(x >= 2);}} \\
+		\texttt{\hilight{blue}{mutex\_unlock}(m);} \\
+		\texttt{assert(x >= 1);}
+	\end{tabular}
+	\end{center}
+\end{frame}
+
+\begin{frame}{State Space (Revisited)}
+	\begin{center}
+		\includegraphics[width=0.96\textwidth]{../../oopsla/tree-maximal-only.pdf}
+		\linegap
+
+		None of these branches contain the necessary preemption!
+		\linegap
+
+		A {\bf data-race preemption point} is required to find the bug.
+	\end{center}
+\end{frame}
+
+\begin{frame}{Iterative Deepening (in Pictures)}
+	\begin{center}
+	\vspace{-0.88em}
+	\includegraphics[width=0.86\textwidth]{../../oopsla/dr-jobs-1.pdf}
+	\end{center}
+\end{frame}
+\begin{frame}{Iterative Deepening (in Pictures)}
+	\begin{center}
+	\vspace{-0.88em}
+	\includegraphics[width=0.86\textwidth]{../../oopsla/dr-jobs-2.pdf}
+	\end{center}
+\end{frame}
+\begin{frame}{Iterative Deepening (in Pictures)}
+	\begin{center}
+	\vspace{-0.88em}
+	\includegraphics[width=0.86\textwidth]{../../oopsla/dr-jobs-3.pdf}
+	\end{center}
+\end{frame}
+\begin{frame}{Iterative Deepening (in Pictures)}
+	\begin{center}
+	\vspace{-0.88em}
+	\includegraphics[width=0.86\textwidth]{../../oopsla/dr-jobs-4.pdf}
+	\end{center}
+\end{frame}
+
+\begin{frame}{Total Verification}
+	\textbf{Convergence theorem:} Testing maximal state space, after detecting all data races, $\equiv$ preempting everywhere!
+	\begin{itemize}
+		\item Insight: $\forall$ arbitrary buggy interleaving, $\exists$ equivalent interleaving % say 'under DPOR' i guess?
+			using only sync/data-race PPs.
+	\end{itemize}
+	\linegap
+
+	Hence, Iterative Deepening provides ``best of both worlds'':
+	\begin{itemize}
+		\item Provides total safety guarantee for small tests
+		\item Finds bugs quickly for large tests when completion is infeasible
+	\end{itemize}
+\end{frame}
+
+\begin{frame}{Evaluation}
+	79 ``P2'' thread libraries from CMU, 78 ``Pintos'' kernels from U. Chicago \& Stanford
+	\linegap
+
+	%629 pairs of \{P2 or Pintos\} + test case
+	629 pairs of P2 or Pintos + test case
+	\linegap
+
+	Each tested for 10 CPU-hours with
+	\begin{itemize}
+		\item Quicksand, finding data-race candidates with...
+			\begin{itemize}
+				\item ...``Pure'' Happens-Before analysis {\em [Lamport '78]}
+				\item ...``Limited'' Happens-Before analysis {\em [O'Callahan '03]}
+			\end{itemize}
+		\item Stand-alone Landslide using ICB {\em [Musuvathi '08]}, preempting at...
+			\begin{itemize}
+				\item ...synchronization APIs only {\em [Simsa '10]}
+				\item ...all shared memory accesses {\em [Holzmann '97]}
+			\end{itemize}
+	\end{itemize}
+\end{frame}
+
+\begin{frame}{Results - Bugs found}
+	\begin{center}
+		\includegraphics[width=0.8\textwidth]{../../oopsla/bugs-talk.pdf}
+	\end{center}
+\end{frame}
+\begin{frame}{Results - Full verifications}
+	\begin{center}
+		\includegraphics[width=0.8\textwidth]{../../oopsla/verifs-talk.pdf}
+	\end{center}
+\end{frame}
+
+\begin{frame}{Deliverables}
+	In the thesis, I will:
+	\begin{itemize}
+		\item Restate the paper {\large \smiley}
+		\item Study variance of repeated Quicksand experiment trials
+	\end{itemize}
+\end{frame}
 
@@
 			\begin{itemize}
-					% TODO: need a bonus slide to justify why this
-					% imagine a pair of transactions each with multiple conflicting accesses
-					% they should be treated as atomic
 				\item slow; invites state space explosion
@@
 				\item (under HTM, all transactions might abort)
-					% TODO: need a bonus slide on finding when a STM txn may abort.
 			\end{itemize}
@@
 
+\breakslide{\Large Bonus Slides}
+
+\begin{frame}{15-410: Summary of bugs found}
+	\begin{center}
+	\begin{tabular}{r|cc|cc}
+	%det bugs found det bugs fixed  races fixed     races found
+	& \multicolumn{4}{c}{Number of groups} \\
+	Number & \multicolumn{2}{c|}{Deterministic bugs} & \multicolumn{2}{c}{Concurrency bugs} \\
+	of bugs & found & fixed & found & fixed \\
+	\hline
+	0       & 27    & 32    & 15    & 23    \\
+	1       & 6     & 5     & 7     & 11    \\
+	2       & 9     & 6     & 12    & 8     \\
+	3       & 2     & 1     & 7     & 2     \\
+	4       & 1     & 1     & 3     & 1     \\
+	5       & 2     & 2     & 2     & 1     \\
+	11      &       &       & 1     & 1     \\
+	\hline
+	%Total groups
+	%       & 47    & 47    & 47    & 47    \\
+	Total bugs
+		& 44    & 34    & 85    & 53    \\
+	\end{tabular}
+	\end{center}
+	%\linegap
+
+	%E.g.: 11 groups fixed 1 bug + 8 fixed 2 + 2 $\times$ 3 + 1 $\times$ 4 + 1 $\times$ 5 + 1 overachieving group who fixed 11 = 11 + 16 + 6 + 4 + 5 + 11 = 53.
+\end{frame}
+
 \begin{frame}{HTM: Even more dimensions of nondeterminism}
@@
 
+\begin{frame}{Interleaving t'xns non-atomically $\rightarrow$ state space explosion}
+					% TODO: need a bonus slide to justify why this
+					% imagine a pair of transactions each with multiple conflicting accesses
+					% they should be treated as atomic
+\end{frame}
+
+\begin{frame}{STM: Judging when aborts are possible}
+	% TODO
+\end{frame}
+
+%%%%
+
+\breakslide{\Large Super Extra Bonus Slides Copy-Pasted from OOPSLA}
+
+\begin{frame}{Data-Race Definitions}
+	\textbf{Data race:} 2 threads access the same memory, and...
+	\begin{itemize}
+		\item At least one access is a write
+		\item Threads do not hold the same lock
+		\item No {\em Happens-Before} relation between threads
+	\end{itemize}
+	\pause
+	\linegap
+
+	Variants of Happens-Before (HB)
+	\begin{itemize}
+		\item {\bf Pure HB:} Any synchronization events {\em [Lamport '78]}
+			% Say: "Or another way of putting it I heard from one of yesterday's talks that I'm stealing, one event is guaranteed to be visible to the other"
+			% Or don't, idgaf
+		\item {\bf Limited HB:} Blocking synchronization (e.g. \texttt{cond\_wait()}) enforces one ordering {\em [O'Callahan '03]}
+	\end{itemize}
+\end{frame}
+
+\begin{frame}{Happens-Before Example}
+	\begin{center}
+		\begin{tabular}{ll}
+			{\bf Thread 1}  & {\bf Thread 2} \\
+			\texttt{\hilight{red}{x++;}}    & \\
+			\texttt{mutex\_lock(m);}	& \\
+			\texttt{// unrelated}	   & \\
+			\texttt{mutex\_unlock(m);}      & \\
+				& \texttt{mutex\_lock(m);} \\
+				& \texttt{// unrelated} \\
+				& \texttt{mutex\_unlock(m);} \\
+				& \texttt{\hilight{red}{x++;}} \\
+		\end{tabular}
+		\linegap
+
+		No race under Pure HB; {\em true potential race} under Limited HB.
+	\end{center}
+\end{frame}
+\begin{frame}{Happens-Before Example}
+	\begin{center}
+		\begin{tabular}{ll}
+			{\bf Thread 1}  & {\bf Thread 2} \\
+			\texttt{\hilight{red}{x++;}}    & \\
+			\texttt{mutex\_lock(m);}	& \\
+			\texttt{\hilight{blue}{y = true;}}      & \\
+			\texttt{mutex\_unlock(m);}      & \\
+				& \texttt{mutex\_lock(m);} \\
+				& \texttt{\hilight{blue}{bool tmp = y;}} \\
+				& \texttt{mutex\_unlock(m);} \\
+				& \texttt{\hilight{blue}{if (tmp)}~\hilight{red}{x++;}} \\
+		\end{tabular}
+		\linegap
+
+		No race under Pure HB; {\em false positive} under Limited HB.
+	\end{center}
+\end{frame}
+
+\begin{frame}{Iterative Deepening Reductions}
+	If one job is a subset of another, testing either might let us skip the other.
+	\begin{center}
+		\includegraphics[width=\textwidth]{../../oopsla/reduction-nothing.png}
+	\end{center}
+\end{frame}
+\begin{frame}{Iterative Deepening Reductions}
+	If one job is a subset of another, testing either might let us skip the other.
+	\begin{center}
+		\includegraphics[width=\textwidth]{../../oopsla/reduction-deferred.png}
+	\end{center}
+\end{frame}
+\begin{frame}{Iterative Deepening Reductions}
+	If one job is a subset of another, testing either might let us skip the other.
+	\begin{center}
+		\includegraphics[width=\textwidth]{../../oopsla/reduction-complete.png}
+	\end{center}
+\end{frame}
+
+\begin{frame}{Quicksand results - Bugs found by CPU time}
+	\begin{center}
+		\includegraphics[width=0.8\textwidth]{../../oopsla/bugs-talk.pdf}
+	\end{center}
+\end{frame}
+\begin{frame}{Quicksand results - Bugs found by wall-clock time}
+	\begin{center}
+		\includegraphics[width=0.8\textwidth]{../../oopsla/bugs-wallclock-talk.pdf}
+	\end{center}
+\end{frame}
+\begin{frame}{Quicksand results - Tests fully verified (CPU time)}
+	\begin{center}
+		\includegraphics[width=0.8\textwidth]{../../oopsla/verifs-talk.pdf}
+	\end{center}
+\end{frame}
+\begin{frame}{Quicksand results - MC + DR vs single-pass DR}
+	\begin{center}
+		\includegraphics[width=0.8\textwidth]{../../oopsla/dr-false-negatives-poster.pdf}
+	\end{center}
+\end{frame}
+\begin{frame}{Quicksand results - Representative of real-world programs?}
+	Bugs by required number of preemptions to expose:
+	\begin{center}
+		\footnotesize
+	\begin{tabular}{r|c|c}
+		{\bf Bound} & {\bf SSS (sync PPs only)} & {\bf SSS (preempt everywhere)} \\
+		\hline
+		0       & 2     & 1     \\
+		1       & 82    & 86    \\
+		2       & 16    & 32    \\
+		3       & 2     & 3     \\
+		4+      & 0     & 0     \\
+		\hline
+		Total   & 102   & 122   \\
+	\end{tabular}
+	\end{center}
+	\linegap
+
+	Bugs by type (Pure HB trial):
+	\begin{itemize}
+		\item 56 deadlocks
+		\item 49 heap errors (47 use-after-free)
+		\item 35 assertion failures
+		\item 31 page fault crashes
+		\item 1 infinite loop
+		\item 1 recursive mutex lock
+	\end{itemize}
+\end{frame}
+
+\begin{frame}{References}
+	\small
+	\begin{itemize}
+		\item {\bf [Lamport '78]}:
+			Leslie Lamport. Time, clocks, and the ordering of events in a distributed system.
+			Commun. ACM, 1978.
+		\item {\bf [Godefroid '97]}: Patrice Godefroid.
+			VeriSoft. A tool for the automatic analysis of concurrent reactive software. CAV 1997.
+		\item {\bf [Holzmann '97]}: The model checker SPIN. TOSE 1997.
+		\item {\bf [Engler '03]}: Dawson Engler and Ken Ashcraft.
+			RacerX: effective, static detection of race conditions and deadlocks. SOSP 2003.
+		\item {\bf [O'Callahan '03]}: Robert O'Callahan and Jong-Deok Choi.
+			Hybrid dynamic data race detection. PPoPP '03.
+		\item {\bf [Flanagan '05]}: Cormac Flanagan and Patrice Godefroid. Dynamic partial-order reduction for
+			model checking software. POPL 2005.
+		\item {\bf [Musuvathi '08]}: Madanlal Musuvathi et al. Finding and Reproducing Heisenbugs in Concurrent
+			Programs. OSDI 2008.
+	\end{itemize}
+\end{frame}
+\begin{frame}{References}
+	\small
+	\begin{itemize}
+		\item {\bf [Buckhardt '10]}: A randomized scheduler with probabilistic guarantees of finding bugs. ASPLOS 2010.
+		\item {\bf [Simsa '12]}: Jiri Simsa. Runtime Estimation and Resource Allocation for
+			Concurrency Testing. CMU-PDL-12-113. December 2012.
+		\item {\bf [Blum '12]}: Ben Blum. Landslide: Systematic Dynamic Race Detection in Kernel
+			Space. CMU-CS-12-118. May 2012.
+		\item {\bf [Kasicki '12]}: Baris Kasicki. Data races vs. data race bugs: telling the difference with Portend. ASPLOS '12.
+		\item {\bf [Huang '15]}: Jeff Huang. Stateless model checking concurrent programs with maximal causality reduction. PLDI '15.
+		\item {\bf [Jensen '15]}: Stateless Model Checking of Event-Driven Applications. OOPSLA 2015.
+	\end{itemize}
+\end{frame}
+
 \end{document}

### 2017-04-25 proposal/talk/proposal.tex
finish slides??
finalize slides i think

@@
 \begin{document}
-%\renewcommand{\inserttotalframenumber}{39}
+\renewcommand{\inserttotalframenumber}{39}
 \normalem
@@
 	\end{itemize}
-	%\linegap
+	\linegap
 
-	%State space size depends on which {\bf preemption points} are used.
+	State space size depends on which {\bf preemption points} are used.
 \end{frame}
@@
 		\item Students implement scheduler, synchronization, VM %from scratch %% trying not to be too arrogant ._.
-		\item UNIX-like process lifecycle ({\tt fork}, {\tt exec}, {\tt vanish}, {\tt wait})
+		\item UNIX-like process lifecycle ({\tt fork}, {\tt exec}, {\tt exit}, {\tt wait})
 		%\item Teams of 2
@@
 		\item 85 concurrency bugs found; 53 which students then fixed
+		\item 32/47 found $\ge 1$ concurrency bug, 22/47 fixed $\ge 1$
 		\item No visible impact on ultimate grades...
@@
 	\begin{itemize}
+		\item Basecode provides round-robin scheduler, synchronization
 		\item Students implement priority scheduling, sleep
-		\item Basecode provides round-robin scheduler, synchronization
 	\end{itemize}
@@
 	\begin{itemize}
-		\item Lockset/vector-clock analyses extended to interrupt-disabling
-		\item Automatically inserting (most) kernel annotations
+		\item Data-race analysis
+			%Lockset/vector-clock analyses
+			extended to interrupt-disabling
+		\item Automatically inserts (most) kernel annotations
 		%\item (various engineering challenges) % palloc heap, non-direct-mapped kernelspace, skipping driver init loops
@@
 \begin{frame}{Deliverables}
-	Pebbles
+	In the thesis, I will:
+	\begin{itemize}
+		\item Pebbles
 		\begin{itemize}
@@
 		\end{itemize}
-	\linegap
-
-	Pintos
+		\item Pintos
 		\begin{itemize}
@@
 		\end{itemize}
-	\linegap
+	\end{itemize}
 \end{frame}
@@
 			\end{itemize}
-		\item Enforce atomicity via scheduler, anticipate aborts via analysis
-			\begin{itemize}
-				\item (under HTM, all transactions might abort)
-			\end{itemize}
+		\item Enforce atomicity via scheduler, anticipate aborts on every transaction
+			%via analysis
+			%\begin{itemize}
+			%	\item (under HTM, all transactions might abort)
+			%\end{itemize}
 		\item Suffices to enumerate all {\em observationally-equivalent} schedules.
@@
 		\begin{itemize}
-			\item Equivalent (modulo performance) to serialize independent transactions
+			\item Equivalent
+				%(modulo performance)
+				to serialize independent transactions
 		\end{itemize}
@@
 \begin{frame}{Failure Injection - 2 dimensions of concurrency}
-	% TODO - pict of mario man
+	\begin{center}
+		\includegraphics[width=\textwidth]{failure-injection-0.pdf}
+	\end{center}
+\end{frame}
+\begin{frame}{Failure Injection - 2 dimensions of concurrency}
+	\begin{center}
+		\includegraphics[width=\textwidth]{failure-injection-2.pdf}
+	\end{center}
 \end{frame}
@@
 	\begin{itemize}
-		\item 2017 Mar-Aug: Port Landslide to Bochs for Pintos
-		\item 2017 Mar-Aug: Transactional memory
+		\item 2017 Apr-Aug: Port Landslide to Bochs for Pintos
+		\item 2017 Apr-Aug: Transactional memory
 		\item 2017 Aug-Oct: Pintos user study
@@
 
-\begin{frame}{Interleaving t'xns non-atomically $\rightarrow$ state space explosion}
-					% TODO: need a bonus slide to justify why this
-					% imagine a pair of transactions each with multiple conflicting accesses
-					% they should be treated as atomic
-\end{frame}
+%\begin{frame}{Interleaving t'xns non-atomically $\rightarrow$ state space explosion}
+%	% imagine a pair of transactions each with multiple conflicting accesses
+%	% they should be treated as atomic
+%\end{frame}
 
-\begin{frame}{STM: Judging when aborts are possible}
-	% TODO
+%\begin{frame}{STM: Judging when aborts are possible}
+%\end{frame}
+
+\begin{frame}{HTM: Sources of Benchmarks}
+	\begin{itemize}
+		\item Shamay, https://github.com/cmpxchg16/tsx.me, 2014
+		\item Dehesa-Azuara and Stanley, http://www.contrib.andrew.cmu.edu/\textasciitilde{}mdehesaa/, 2016
+		\item Chapman et al., Hybrid STM/HTM for nested transactions on OpenJDK, OOPSLA 2016
+		\item Wang et al., Eunomia: Scaling Concurrent Search Trees under Contention Using HTM, PPoPP 2017
+	\end{itemize}
 \end{frame}
@@
 %%%%
+
+\section{Warp Zone}
 
@@
 
+\begin{frame}{Don't you need \texttt{if (mutex\_held(\&foo\_lock)) \_xabort();} in that transaction on slide 31?}
+	Yes.
+\end{frame}
+
 \begin{frame}{References}

### 2017-09-27 irb/landslide-guide-p2.tex
add landslide guide p2

--- 2015-02-04 irb/survey.tex
@@
 
-\title{\bf Landslide: Systematic Testing in 15-410 \\ User Guide and Survey}
+\title{\bf Landslide: Systematic Testing in 15-410 \\ User Guide}
 \author{Ben Blum}
@@
 
-%Normally here I would write something researchy about systematic testing and preemption points, but if you're here it probably means you know what this is about.
-If you don't, please refer to the lecture material.
-%\\
-%
-%\noindent Thanks for volunteering to try out Landslide.
-We have worked hard to make it nice enough to give to you, and we hope you're just as excited as we are to try out some systematic testing techniques the world has never seen before.
-%\\
-
-\noindent You should have received a document titled {\em User Study Information Sheet} that describes your rights as a study participant and our confidentiality procedures.
-If you are missing this, please contact me immediately.
-My email address is {\tt bblum@cs.cmu.edu} and my phone number is 412-304-4294; email is preferred.
-% TODO: Host them on maxi or contrib and link to them.
+You should have received a document titled {\em User Study Information Sheet} that describes your rights as a study
+participant and our confidentiality procedures. If you are missing this, please contact me immediately. My email
+address is {\tt bblum@cs.cmu.edu} and my phone number is 412-304-4294; email is preferred.
 \\
 
-\noindent Likewise, please make sure you have read and understand the lecture slides on systematic testing, which are available on the 15-410 course website.
-If you have any questions about the research background I would be happy to answer them.
+\noindent Likewise, please make sure you have read and understand the lecture slides on systematic testing, which are
+available on the 15-410 course website. If you have any questions about the research background I would be
+happy to answer them.
 
@@
 
-The very first thing you should do is {\bf READ THIS DOCUMENT ENTIRELY FIRST!} We need you to print out several survey sheets before you dive in.
-\\
-
-\noindent We encourage you to use Landslide on the GHC cluster or UNIX pool machines ({\tt ghcXX.ghc.andrew.cmu.edu} or {\tt unix.andrew.cmu.edu}).
-Landslide has not been tested with other configurations and will probably fail mysteriously on your personal laptop, and we cannot promise to be of any help.
+We encourage you to use Landslide on the GHC cluster or UNIX pool machines ({\tt ghcXX.ghc.andrew.cmu.edu}
+or {\tt unix.andrew.cmu.edu}). Landslide has not been tested with other configurations and will probably fail
+mysteriously on your personal laptop, and we cannot promise to be of any help.
 
 \begin{enumerate}
-	\item Clone the repository -- {\tt git clone https://github.com/bblum/landslide.git} -- this will be your workspace.
+	\item Clone the repository - {\tt git clone https://github.com/bblum/landslide.git} - this will be your workspace.
 	\item Run {\tt ./p2-setup.sh PATH}, where {\tt PATH} is the absolute path to your p2 implementation.
@@
 			\item This command will import your p2 into {\tt ./pebsim/p2-basecode/} and then attempt to build it.
+				Whenever you update your p2 code, please do so in its original {\tt PATH}
+				and then run {\tt p2-setup.sh} again.
 			\item {\tt PATH} should be the path with {\tt 410kern}, {\tt 410user}, {\tt user/libthread}, etc as subdirectories, not the {\tt user/libthread} directory itself.
 		\end{itemize}
-	\item Next, make sure you have read this document entirely.
-At this point you should have several survey sheets printed out, ready to write on -- see the Survey section below.
-You should also have some means of timing yourself.
-	\item Run {\tt ./landslide OPTIONS} to run tests with Landslide. {\bf Fill out a fresh survey sheet each time you do this.} Repeat until satisfied.
+	\item Run {\tt ./landslide OPTIONS} to run tests with Landslide, for various values of {\tt OPTIONS}:
 		\begin{itemize}
-			\item Use {\tt ./landslide -h} to view command line options. The most important ones are {\tt -t}, max testing time, and {\tt -p}, which test program to run.
-				% TODO: Remove this TODO when you have implemented glue to move the html files to the toplevel directory.
-			\item Landslide's bug reports will show up as HTML files. If you are using a cluster machine in person, you can view them by pointing a browser to {\tt file:///PATH/TO/landslide/FILENAME.html}.
-				\footnote{Another option: move it to {\tt \textasciitilde/www} in your AFS, and view it at {\tt http://www.contrib.andrew.cmu.edu/\textasciitilde{}YOUR\_ANDREWID/FILENAME.html}.
-Note the {\tt www} in the URL is mandatory; note also that this will allow the world to see it, so be careful!}
-				\footnote{If you have your own custom test case, you can put it in {\tt pebsim/p2-basecode/user/progs} and edit \\ {\tt pebsim/p2-basecode/config.mk} to build it, then run {\tt ./p2-setup.sh} again.}
-			\item The available test programs are {\tt thr\_exit\_join} (the default), {\tt broadcast\_test}, {\tt paraguay}, and {\tt rwlock\_downgrade\_read\_test}.
-			\item The suggested sequence of test configurations to run is as follows.
+			\item Use ./landslide -h to view command line options. The most important ones are -t , max testing
+time, and -p , which test program to run.
+			\item Landslide's bug reports will show up as HTML files.
+				\begin{itemize}
+					\item If you are using a cluster machine in person, you can view them by pointing a browser to \\
+						{\tt file:///PATH/TO/landslide/FILENAME.html}
+					\item For remote usage, move the html file to {\tt ~/www} in your AFS, and view it at \\
+						{\tt http://www.contrib.andrew.cmu.edu/~YOUR\_ANDREWID/FILENAME.html}.
+						Note the www in the URL is mandatory;
+						note also that this will allow the world to see it,
+						so be careful to clean up any traces in {\tt ~/www} when you're done!
+				\end{itemize}
+			\item Suggested test configurations to run are as follows.
 				You can, of course, change the time limits or test programs as you see fit.
-For the longer-running configurations, it's recommended to run them inside of a {\tt screen} session.
+				For the longer-running configurations,
+				we recommend running them inside of a {\tt screen} or {\tt tmux} session.
 				\begin{itemize}
-				\item {\tt ./landslide -p thr\_exit\_join -t 30m}
-				\item {\tt ./landslide -p broadcast\_test -t 30m}
-				\item {\tt ./landslide -p paraguay -t 30m}
-				\item {\tt ./landslide -p rwlock\_downgrade\_read\_test -t 30m}
-				\item {\tt ./landslide -p thr\_exit\_join -t 3h}
-				\item {\tt ./landslide -p broadcast\_test -t 3h}
-				\item {\tt ./landslide -p paraguay -t 3h}
-				\item {\tt ./landslide -p rwlock\_downgrade\_read\_test -t 3h}
-				\item {\tt ./landslide -p thr\_exit\_join -t 12h}
-				\item {\tt ./landslide -p broadcast\_test -t 12h}
-				\item {\tt ./landslide -p paraguay -t 12h}
-				\item {\tt ./landslide -p rwlock\_downgrade\_read\_test -t 12h}
+					\item \texttt{./landslide -t 30m -p thr\_exit\_join}
+					\item \texttt{./landslide -t 30m -p mutex\_test}
+					\item \texttt{./landslide -t 30m -p paradise\_lost}
+					\item \texttt{./landslide -t 30m -p broadcast\_test}
+					\item \texttt{./landslide -t 30m -p paraguay}
+					\item \texttt{./landslide -t 30m -p rwlock\_downgrade\_read\_test}
+						\\
+					\item \texttt{./landslide -t 3h -p thr\_exit\_join}
+					\item \texttt{./landslide -t 3h -p mutex\_test}
+					\item \texttt{./landslide -t 3h -p paradise\_lost}
+					\item \texttt{./landslide -t 3h -p broadcast\_test}
+					\item \texttt{./landslide -t 3h -p paraguay}
+					\item \texttt{./landslide -t 3h -p rwlock\_downgrade\_read\_test}
+						\\
+					\item \texttt{./landslide -t 12h -p thr\_exit\_join}
+					\item \texttt{./landslide -t 12h -p mutex\_test}
+					\item \texttt{./landslide -t 12h -p paradise\_lost}
+					\item \texttt{./landslide -t 12h -p broadcast\_test}
+					\item \texttt{./landslide -t 12h -p paraguay}
+					\item \texttt{./landslide -t 12h -p rwlock\_downgrade\_read\_test}
 				\end{itemize}
+			\item To build your own custom test case for use with Landslide,
+				put it in {\tt pebsim/p2-basecode/user/progs}
+				and edit the {\tt STUDENTTESTS} line in {\tt pebsim/p2-basecode/config-incomplete.mk}
+				(note: NOT {\tt config.mk}; the setup script auto-generates that file).
+				Then run {\tt ./p2-setup.sh} again.
 		\end{itemize}
+	\item After fixing any found bugs, re-run Landslide with the same test options to confirm that your fix works.
 \end{enumerate}
 
-\noindent {\bf What to do when something goes wrong:}
-If you find a Landslide bug, such as an assertion/crash, or such as a bug report that you think is actually a bug in Landslide rather than a bug in your P2, or such as it getting stuck and not exiting in the given time limit, please create a tarball of your workspace, just as you would to report a reference kernel bug, but send it to Ben (\texttt{bblum@cs.cmu.edu}) instead of 410 staff.
-For any other technical issues getting it to work, or if you have difficulty understanding a bug report, send email to Ben.
+\noindent {\bf If something goes wrong:} If you find a Landslide bug, such as an assertion/crash, or such as a
+bug report that you think is actually a bug in Landslide rather than a bug in your P2, or such as it getting stuck
+and not exiting in the given time limit, please create a tarball of your workspace, just as you would to report
+a reference kernel bug, but send it to Ben ({\tt bblum@cs.cmu.edu}) instead of 410 staff. For any other technical
+issues getting it to work, or if you have difficulty understanding a bug report, send email to Ben.
 \\
 
-\noindent
-However, we {\em cannot} provide help on how to understand {\em why} your bug happened or how to fix it.
-The point of the study is that Landslide's automation plus your own brain should be enough!
-On the other hand, if you have a design question while deciding between potential ways to fix a bug, send it to 410 course staff as usual.
-
-\newpage
-
-\section{Survey}
-\subsection{Overall Survey Questions}
-
-\begin{center}
-{\bf Print this page out once.}
-\end{center}
-
-\begin{enumerate}
-	\item How much time did you spend directly using Landslide?
-This includes configuring test parameters and interpreting debug output.
-It does not include time spent running Landslide in the background, e.g.
-overnight.
-		\vspace{0.5in}
-
-	\item What test configurations did you use that DIDN'T find bugs? For example, {\tt ./landslide -t 15m -p thr\_exit\_join}
-\end{enumerate}
-
-\newpage
-\subsection{Per-Bug Survey Questions}
-\begin{center}
-{\bf \large PRINT OUT THIS PAGE MULTIPLE TIMES.
-\\
-
-Each time you find a new bug with Landslide, fill this out again.}
-\end{center}
-
-\begin{enumerate}
-		% TODO: find a way to allow ze studence to condense non-bug-finding runs onto 1 page.
-	\item What test configuration did you use? For example, {\tt ./landslide -t 15m -p thr\_exit\_join}
-		\vspace{0.5in}
-	\item Summarize the bug.
-		\vspace{2in}
-	\item How IMPORTANT was this bug? Rate on a scale from 1 to 10, where:
-		\begin{itemize}
-			\item 1 = Trivial; an obvious ``oops'' that does not reflect any key lessons learned during the project. I would expect to not lose any points for this bug.
-			\item 4 = Minor; a logic problem not related to threads/synchronization. I would expect to lose 1 or 2 points at most for this bug.
-			\item 7 = Major; a concurrency-related bug related to subject matter discussed in lecture.
-			\item 10 = Severe; this bug reflects a fundamental lesson about concurrency that we learned during the project.
-I would expect to receive ``see course staff'' in my ink if I turned in code with this bug!
-		\end{itemize}
-		\vspace{0.2in}
-	\item How much EFFORT/TIME did this bug require to diagnose and fix? Rate on a scale from 1 to 10, where:
-		\begin{itemize}
-			\item 1 = Very little; we knew immediately what was wrong and needed to change only 1 or 2 lines of code.
-			\item 5 = Some; the bug took considerable thought to fix, and required nontrivial code changes.
-			\item 10 = Lots; we worked for several hours to fix the bug and/or required a major refactoring of our code.
-		\end{itemize}
-		\vspace{0.2in}
-	\item How likely would a CONVENTIONAL STRESS TEST be able to find the bug (i.e., running the program repeatedly, not using Landslide)? Rate on a scale from 1 to 10, where:
-		\begin{itemize}
-			\item 1 = No hope. The window for a random timer interrupt to expose the bug is so small that a conventional stress test could easily run continuously overnight without encountering it.
-			\item 5 = Moderate chance. Most random timer interrupts would not expose the bug, but the window is more than just a few instructions.
-			\item 10 = Very likely. A random timer interrupt during the test would expose the bug more often than not.
-		\end{itemize}
-\end{enumerate}
+\noindent However, we cannot provide help on how to understand why your bug happened or how to fix it. The point of
+the study is that Landslide's automation plus your own brain should be enough! On the other hand, if you have
+a design question while deciding between potential ways to fix a bug, send it to 410 course staff as usual.
 

### 2017-10-24 resume/cv_4.tex
write a modern resume

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Medium Length Professional CV
% LaTeX Template
% Version 2.0 (8/5/13)
%
% This template has been downloaded from:
% http://www.LaTeXTemplates.com
%
% Original author:
% Trey Hunner (http://www.treyhunner.com/)
%
% Important note:
% This template requires the resume.cls file to be in the same directory as the
% .tex file. The resume.cls file provides the resume style used for structuring the
% document.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%----------------------------------------------------------------------------------------
%	PACKAGES AND OTHER DOCUMENT CONFIGURATIONS
%----------------------------------------------------------------------------------------

\documentclass{resume} % Use the custom resume.cls style

\usepackage[left=0.6in,top=0.6in,right=0.6in,bottom=0.6in]{geometry} % Document margins
\usepackage[charter]{mathdesign}

\name{Ben Blum} % Your name
\address{5000 Forbes Ave \\ Gates-Hillman 9009 \\ Pittsburgh, PA 15213}
\address{\texttt{bblum@cs.cmu.edu} \\ \texttt{https://github.com/bblum}} % Your phone number and email
\address{6704 Thomas Blvd Apt 2L \\ Pittsburgh, PA 15208}

\begin{document}

%----------------------------------------------------------------------------------------
%	EDUCATION SECTION
%----------------------------------------------------------------------------------------

\begin{rSection}{Education}

{\bf Carnegie Mellon University} \hfill {\em June 2011} \\ 
B.S. in Computer Science

{\bf Carnegie Mellon University} \hfill {\em June 2012} \\ 
M.S. in Computer Science

{\bf Carnegie Mellon University} \hfill {\em 2018 (expected)} \\ 
Ph.D. in Computer Science

\end{rSection}

%----------------------------------------------------------------------------------------
%	WORK EXPERIENCE SECTION
%----------------------------------------------------------------------------------------

\begin{rSection}{Experience}

\begin{rSubsection}{Carnegie Mellon University}{Summer 2011 - Present}{Graduate Student}{Pittsburgh, PA}
\item Developed Landslide, a systematic concurrency testing framework for low-level userspace and kernel systems
\item Conducted educational user studies offering Landslide to students during operating systems class projects
\end{rSubsection}

\begin{rSubsection}{Mozilla}{Summers 2012 \& 2013}{Research Intern}{Mountain View, CA}
\item Helped develop the Rust language runtime (green-thread scheduler, stack management, concurrency primitives)
\item Developed an API for type-safe, race-free mutable shared state between threads using Rust's borrow checker
\item Contributed to the design of a sound type system for Rust's closures
\end{rSubsection}

\begin{rSubsection}{Google}{Summers 2009 \& 2010}{Software Engineering Intern}{Mountain View, CA}
\item Implemented atomic per-process migration algorithm for Linux kernel cgroups
\end{rSubsection}

\begin{rSubsection}{Carnegie Mellon University}{Spring 2009 - Fall 2012}{Teaching Assistant and Instructor}{Pittsburgh, PA}
\item TA for 15-213, Introduction to Computer Systems (3 semesters)
\item TA for 15-410, Operating System Design and Implementation (3 semesters)
\item Instructor for 98-172, Great Practical Ideas in Computer Science (1 semester)
\end{rSubsection}

\end{rSection}

%----------------------------------------------------------------------------------------
%	TECHNICAL STRENGTHS SECTION
%----------------------------------------------------------------------------------------

\begin{rSection}{Publications}

\textbf{Stateless Model Checking with Data-Race Preemption Points.} Ben Blum and Garth Gibson. OOPSLA 2016.
%ACM SIGPLAN International Conference on Object-Oriented Programming, Systems, Languages, and Applications (OOPSLA) 2016.

\textbf{Parrot: A Practical Runtime for Deterministic, Stable Threads.} Heming Cui et al. (5th author). SOSP 2013.
%ACM Symposium on Operating Systems Principles (SOSP) 2013.

\textbf{Soundness Proofs for Iterative Deepening.} Ben Blum. Tech report CMU-PDL-16-103.

\textbf{Landslide: Systematic Dynamic Race Detection in Kernel Space.} Ben Blum. MS thesis CMU-CS-12-118.

\end{rSection}

\begin{rSection}{Skills}

\begin{tabular}{ @{} >{\bfseries}l @{\hspace{6ex}} l }
	Languages (proficient) & C, Rust, Haskell, x86 assembly, Bash, LaTeX \\
	Languages (familiar) & C++, SML, Python, Lua \\
	Tools & git, vim, GNU toolchain, simics, bochs
\end{tabular}

\end{rSection}

%----------------------------------------------------------------------------------------
%	EXAMPLE SECTION
%----------------------------------------------------------------------------------------

%\begin{rSection}{Section Name}

%Section content\ldots

%\end{rSection}

%----------------------------------------------------------------------------------------

\end{document}

### 2018-01-03 thesis/graph.py
add page count scripce

@@ -1,0 +1,53 @@
+#!/usr/bin/env python3
+
+import numpy as np
+import matplotlib
+from importlib import import_module
+import os.path
+import sys
+import argparse
+import dateutil.parser
+from matplotlib.dates import DayLocator, HourLocator, DateFormatter, drange
+
+# Work around matplotlib brain damage of binding backend at load
+plt = None
+def init_pyplot(nongraphical=False):
+    global plt
+    if nongraphical:
+        matplotlib.use('Agg')
+    plt = import_module("matplotlib.pyplot")
+def write_pyplot(out):
+    if out:
+        plt.savefig(out)
+    else:
+        plt.show()
+
+HEADER_NAMES = ["commit", "time", "pages", "word"]
+
+def load_file(name):
+    return matplotlib.mlab.csv2rec(name, names=HEADER_NAMES)
+
+def make_graph(name):
+    data = load_file(name)
+    fig, ax = plt.subplots()
+
+    ax.plot_date(data.time, data.pages)
+#    ax.xaxis.set_major_locator(DayLocator())
+ #   ax.fmt_xdata = DateFormatter('%Y-%m-%d %H:%M:%S')
+    ax.set_ylim(0)
+    ax.set_xlim(dateutil.parser.parse("2016-05-01")) # XXX
+    plt.title('pages in thesis.pdf')
+    fig.autofmt_xdate()
+
+def main():
+    parser = argparse.ArgumentParser()
+    parser.add_argument("--out")
+    parser.add_argument("files", nargs='+')
+    args = parser.parse_args()
+
+    init_pyplot(args.out)
+    make_graph(args.files[0])
+    write_pyplot(args.out)
+
+if __name__ == '__main__':
+    sys.exit(main())

### 2018-01-03 thesis/run.sh

@@ -1,0 +1,22 @@
+#!/bin/bash
+
+function die {
+	echo -e "\033[01;31m$1\033[00m"
+	exit 1
+}
+
+#REPO=https://github.com/bblum/talks
+CHECKOUT_DIR=talks
+REPO_SUBDIR=thesis
+LOG=log.csv
+
+LOG_PATH=$PWD/$LOG
+
+#rm -rf $CHECKOUT_DIR
+#git clone $REPO $CHECKOUT_DIR
+
+cd $CHECKOUT_DIR/$REPO_SUBDIR || die "no cd subdir"
+git pull || die "no git pull"
+make >/dev/null || die "no make"
+((git show -s --format=%h,%ci HEAD; pdftk thesis.pdf dump_data | grep NumberOfPages | cut -d' ' -f2; wc -w thesis.tex | cut -d' ' -f1) | tr '\n' ,; echo) >> $LOG_PATH || die "no pdftk"
+make clean >/dev/null

### 2018-01-03 thesis/wrapper.sh

@@ -1,0 +1,15 @@
+#!/bin/bash
+
+while true; do
+	echo -n "updating page count at "
+	date
+	./run.sh
+	if [ "$?" = 0 ]; then
+		./graph.py --out graph.png log.csv
+		if [ "$?" = 0 ]; then
+			mv graph.png /var/www/html/
+		fi
+	fi
+	echo "(done)"
+	sleep 5m
+done

### 2018-01-03 thesis/thesis.tex
write some inspirational quotes

--- 2017-03-01 proposal/thesprop.tex
@@
 %        colohan  = Chris Colohan
+%        comar    = Cyrus Omar
 %        jab      = Justin Boyan
@@
 % some useful packages
-\usepackage{times}
+\usepackage[charter]{mathdesign}
 \usepackage{fullpage}
@@
 \usepackage{amsmath}
-\usepackage{amsthm}
-\usepackage[charter]{mathdesign}
 \usepackage[numbers,sort]{natbib}
@@
 \usepackage{subfigure}
-\usepackage[linesnumbered,ruled]{algorithm2e}
-\usepackage{algpseudocode}
 
@@
 \draftstamp{\today}{DRAFT}
-
-\newcommand\simics{\textsc{Simics}}
-\newcommand{\sect}[1]{\S #1}
-\newcommand\hilight[2]{\color{#1}#2\color{black}}
-\definecolor{orange}{RGB}{192,96,0}
-\definecolor{olivegreen}{RGB}{0,127,0}
-\definecolor{brickred}{RGB}{192,0,0}
-\definecolor{commentblue}{RGB}{0,0,192}
-
-\newtheorem{theorem}{Theorem}
-
-\newcommand\llama[1]{\includegraphics[width=#1]{llama.pdf}}
-\newcommand\llitem{\item[\llama{1.2em}]}
 
@@
 \author{Ben Blum}
-\date{January 2017}
-\Year{2017}
+\date{Decembruary 2018}
+\Year{2018}
 \trnumber{}
@@
 Brandon Lucia \\
-Haryadi Gunawi (University of Chicago)
+Haryadi Gunawi, University of Chicago
 }
@@
 
-% TODO(thesis): Save this one for the thesis itself...
-%\keywords{landslide terminal, baggage claim, ground transportation, ticketing}
+\keywords{landslide terminal, baggage claim, ground transportation, ticketing}
 
@@
 
-%\begin{dedication}
-%	For my dog, Louie.
-%\end{dedication}
+\begin{dedication}
+For Elsie, Margie, Johnny, Mel, Eve, and David, who brought me here today.
+\end{dedication}
 
@@
 
-In this thesis, I propose to make stateless model checking more practical for human use by way of several new techniques.
-I have built Landslide, a stateless model checker specializing in student projects for undergraduate operating systems classes.
-Landslide includes a novel algorithm for automatically managing
+In this thesis, I develop several new techniques to make stateless model checking more practical for human use.
+I have built Landslide, a stateless model checker specializing in undergraduate operating systems class projects.
+Landslide extends the traditional model checking algorithm with a new framework for automatically managing
 %the exploration of
 multiple state spaces according to their estimated completion times,
-which I will show quickly finds bugs should they exist and also quickly verifies correctness otherwise.
-I will evaluate Landslide's suitability for inexpert use by presenting the results of many semesters providing it to students in 15-410, CMU's Operating System Design and Implementation class.
+which I show quickly finds bugs should they exist and also quickly verifies correctness otherwise.
+I evaluate Landslide's suitability for inexpert use by presenting the results of many semesters providing it to students in 15-410, CMU's Operating System Design and Implementation class,
+and more recently, students in the operating systems classes at the University of Chicago and Berkeley.
 %Finally, I will present several new techniques that allow stateless model checking to be practically employed on real-world programs.
-Finally, I will explore broader impact by extending Landslide to test some real-world programs and to be used by students at other universities.
+%Finally, I will explore broader impact by extending Landslide to test some real-world programs and to be used by students at other universities.
+Finally, I extend Landslide with a new concurrency model for hardware transactional memory,
+and evaluate several real-world transactional benchmarks to show that stateless model checking can keep up with the developing concurrency demands of real-world programs.
 \end{abstract}
 
-%\begin{acknowledgments}
-%My advisor is cool.
-%
-%But I am cooler.
-%\end{acknowledgments}
+\begin{acknowledgments}
+I thank the online Android: Netrunner community Stimhack.
+\\
+
+Please hold on. This thesis will be departing shortly for the Landslide terminal, baggage claim, ground transportation, and ticketing.
+\end{acknowledgments}
+
+\newpage
+\begin{figure}[h]
+	\vspace{12em}
+	\begin{center}
+		\includegraphics[width=0.5\textwidth]{how-i-learned.png}
+	\end{center}
+\end{figure}
 
 \tableofcontents
-%\listoffigures
-%\listoftables
+\listoffigures
+\listoftables
 
@@
 
+\newcommand{\inspirationallinebreak}{\vspace{0.25em}}
+\newcommand\inspirationalquote[2]{\begin{flushright}
+	\vspace{-1em}
+	\fontfamily{pzc}\selectfont
+	{\em #1}
+	\inspirationallinebreak
+
+	---{#2}
+	\fontfamily{mdbch}\selectfont
+	\vspace{2em}
+\end{flushright}}
+
 \chapter{Introduction}
 
-Modern computer architectures have turned to increasing CPU core count, rather than clock speed, to improve processing power \cite{mooreslaw}.
-%To take advantage of multiple cores, programs must be written {\em concurrently}
-To take advantage of multiple cores for performance, programmers must write software to execute {\em concurrently} --
-using multiple {\em threads} which execute multiple parts of a program's logic simultaneously.
-However, when threads access the same shared data, they may interleave in unexpected ways which change the outcome of their execution.
-When an unexpected interleaving produces undesirable program behaviour,
-for example, by corrupting shared data structures,
-we call it a {\em concurrency bug}.
-Concurrency bugs are notoriously hard for programmers to find and debug
-because the specific thread interleaving required to trigger them arises at random during normal execution,
-and often with very low probability.
-%Concurrency bugs are notoriously hard to find and reproduce because they only appear in specific thread interleavings, which arise at random during normal program execution.
-% TODO(LAYMAN): give example of trying to open car door at same time as friend turns key to unlock it.
+\chapter{Background}
 
-Most commonly, a programmer searches for concurrency bugs in her code by running it many times (in parallel, in serial, or both),
-hoping that eventually, it will run according to the particular interleaving required to expose a hypothetical bug.
-This technique, known as {\em stress testing}, is unreliable,
-providing no guarantee of finding the failing interleaving in any finite amount of time.
-%, even when a bug does exist.
-It also provides no assurance of correctness:
-when finished, there is no way of knowing how many distinct thread interleavings were actually tested.
-%it may by chance test only a single interleaving over and over again!
-Nevertheless, stress testing remains popular because of how easily a programmer can use it:
-she simply wraps her program in a loop, sets it to run overnight, and kills it if her patience runs out before it finds a bug.
+\chapter{410} % rename
+\inspirationalquote{Knowing the students might one day
+%find a way to
+fix their concurrency bugs...
+it fills you with determination.}{Undertale (paraphrased)}
 
-{\em Stateless model checking} \cite{verisoft} is an alternative way to test for concurrency bugs,
-or to verify their absence,
-which provides more reliable coverage, progress, and verification than stress testing.
-A stateless model checker tests a program by forcing it to execute a new unique thread interleaving on each iteration of the test,
-capturing and controlling the randomness in a finite state space of all possible interleavings.
+\chapter{Related Work}
+\inspirationalquote{I see now that none of us are yet ready. The cycle exists so that we may
+improve ourselves. But \\
+the one who reaches the summit is not our superior, for
+they stand on our shoulders to reach it.}
+{The Shepherd v82.6.0174, The Talos Principle}
 
-Unfortunately, the size of these state spaces is exponentially proportional to the size of the tested program.
-% TODO(LAYMAN): explain exponential explosion by relating the parable of grains of rice on a chessboard.
-For even moderately-sized programs, there may be more possible ways to interleave every thread's every instruction
-than particles in the universe.
-Accordingly, a programmer who wants her test to make reasonable progress through the state space must choose a subset of ways that her threads could interleave,
-focusing on fully testing that subset, while ignoring other possibilities she doesn't care about.
-%However, choosing this subset is a difficult trade-off for humans to make before even knowing whether a bug exists.
-However, it is difficult to choose a subset of thread interleavings that will produce a meaningful, yet feasible test.
-Until computers can automatically navigate this trade-off in some intelligent way,
-programmers will continue to fall back to the random approach of stress testing.
+\chapter{Conclusion}
+\inspirationalquote{
+%	``If you are a god, Zeus, as the stories claim, then why did you create evolution?
+%	Why did you make \\ a world that can only grow through cruelty and pain?''
+%	\inspirationallinebreak
+%
+%For a moment--surely this meant death was near--she thought she heard him answer:
+%``My child, I \\
+%was shaped by the gods that came before me, as you were shaped by me.
+%The choice I had was be- \\
+%tween creation and oblivion, life and death.
+%And I chose life, because any life is better than no life, \\
+%because as long as there is life, there is hope--if not for us, then for some generation to come.''
+%	\inspirationallinebreak
+%
+%``Then how are you a god, if you can offer so little?'' she whispered, feeling death creep closer.
+%	\inspirationallinebreak
+%
+%``I am a god because I take upon myself the burden of creation,'' the statue replied.
+%	\inspirationallinebreak
+%
+%``Then we are all gods,'' Alexandra said, and pushed the button.}
+	\begin{tabular}{p{0.875\textwidth}}
+	``If you are a god, Zeus, as the stories claim, then why did you create evolution?
+	Why did you make a world that can only grow through cruelty and pain?''
+	\\
+	For a moment--surely this meant death was near--she thought she heard him answer:
+	``My child, I was shaped by the gods that came before me, as you were shaped by me.
+	The choice I had was between creation and oblivion, life and death.
+	And I chose life, because any life is better than no life,
+	because as long as there is life, there is hope--if not for us, then for some generation to come.''
+	\\
+	``Then how are you a god, if you can offer so little?'' she whispered, feeling death creep closer.
+	\\
+	``I am a god because I take upon myself the burden of creation,'' the statue replied.
+	\\
+	``Then we are all gods,'' Alexandra said, and pushed the button.
+	\end{tabular}}
+{The Talos Principle (Road to Gehenna)}
 
-Another problem stateless model checking suffers is that certain types of programs cannot be tested without the programmer putting forth some manual instrumentation effort.
-For example, operating system kernels implement their own sources of concurrency and their own synchronization primitives,
-so the checker needs to be told how to identify and control the execution of each thread.
-Some expert concurrency research wizards may be willing to add manual annotations to their code,
-but required manual effort is a serious downside for anyone with a looming deadline,
-and especially so for students who are still learning basic concurrency principles.
-%We should not expect programmers to add effortful manual annotations to their code,
-%or they will abandon our fancy technique to instead simply run stress tests until their deadline tomorrow evening.
-
-This thesis will solve both problems discussed above.
-My thesis statement is as follows:
-
-\vspace{1em}
-
-\begin{center}
-	{\em Thanks to the new algorithms, heuristics, and concurrency models I have developed,
-	stateless model checking is an appropriate and accessible concurrency testing technique
-	for programmers in both educational and real-world settings.}
-\end{center}
-
-\vspace{1em}
-
-I have built Landslide \cite{landslide}, a stateless model checker for thread libraries and kernels,
-and I have developed some techniques for automatically choosing the best thread interleavings to test
-and for automatically instrumenting operating system kernels in an educational setting.
-This thesis will comprise three major contributions:
-
-\begin{itemize}
-	\item {\bf Meaningful state spaces (Chapter \ref{chap:quicksand}).}
-		I will present {\em Iterative Deepening}, a new algorithm for navigating the trade-off in how many preemption points to test at once.
-		Iterative Deepening incorporates state space estimation \cite{estimation} to decide on-the-fly whether each state space is worth pursuing, and uses data race analysis \cite{tsan} to find new preemption point candidates based on a program's dynamic behaviour.
-		This section will include a large evaluation of the technique, comparing its performance to three prior work approaches across 600+ unique tests.
-		I will show that Iterative Deepening of preemption points outperforms prior work in terms both of finding bugs quickly and of completely verifying correctness when no bug exists.
-
-		This work is {\bf completed} and was published at OOPSLA 2016.
-	\item {\bf Educational use (Chapters \ref{chap:410} and \ref{chap:pintos}).}
-		For the past three semesters, I have offered a fully-automated version of Landslide to students in 15-410, CMU's undergraduate Operating System Design and Implementation class \cite{kspec,thrlib}, for use as a debugging aid during the thread library project.
-		I will continue these user studies, and use the data to evaluate the suitability of stateless model checking in an educational setting.
-		%, investigating the following questions in particular:
-		%\begin{itemize}
-		%	\item Does Landslide improve project submission quality when used as a debugging aid?
-		%	\item Does Landslide teach students lessons in concurrent programming
-		%\end{itemize}
-
-		So far, the fully-automatic testing mode is available only for 15-410 thread library projects.
-		To prove these techniques are relevant beyond CMU's walls, I will extend Landslide to handle Pintos kernel projects from other universities \cite{pintos}.
-		I will then collaborate with those schools to deploy Landslide to their operating systems students.
-
-		This is {\bf ongoing work}; I have run the user study for 4 semesters so far and am proposing to continue them
-		and extend the existing evaluation,
-		as well as deploy a Pintos port.
-	\item {\bf Transactional Memory (Chapter \ref{chap:tm}).}
-		Transactional Memory (TM) is a relatively new concurrent programming technique \cite{transactional-memory}
-		which is not yet addressed by modern model checkers.
-		I will extend Landslide's concurrency model to support both hardware (HTM) and software (STM) variants of TM,
-		and test several ``real-world'' TM programs.
-
-		This will be {\bf new work},
-		involving some implementation effort, some proofs, and designing a new evaluation.
-		%, as Landslide presently supports testing Pintoses only with some manual effort,
-		%and does not yet feature instrumentation to run on Linux programs.
-		%and will need its concurrency model extended to incorporate the HTM programming model.
-\end{itemize}
-
-% TODO(thesis): Other chapters of this thesis include...
-
-\input{overview}
-\input{quicksand}
-\input{410}
-\input{pintos}
-\input{tm}
-\input{related}
-
-%Reduction
-%techniques such as Dynamic Partial Order Reduction \cite{dpor} and Maximal Causality Reduction \cite{mcr} expand the limits of feasible test completion,
-%and search ordering strategies such as Iterative Context Bounding \cite{chess-icb} \revision{encourage}~bugs to be found sooner in a given space should they exist.
-
-%Topics of thesis:
-%1. Data race preemption points / iterative deepening
-%2. Education in 410. Evaluation questions are:
-%	2a. Does P2 submission quality improve?
-%	2b. Does P3 submission quality improve correlated with students who used (got the most out of) Landslide during P2?
-%	2c. Does Landslide reach the bottom of the class? (Look at bottom-of-class P0s/P1s, see how they improve)
-%3. Broader impact
-%	3a. Real world programs (and the challenges therein?)
-%	3b. Give to pintos kids (and open-source landslide on bochs in doing so)
-
-\chapter{Conclusion and Timeline}
-
-Stateless model checking is a powerful technique for testing concurrent programs,
-capable in theory of rooting out any bug or providing total verification on any program,
-but suffers several problems which relegate that theory to fantasy land \cite{vargomax}.
-Chief among those problems is exponential explosion of state spaces,
-making it difficult to decide in advance which combinations of preemption points can be productively tested within a given time limit.
-Another major problem is the manual annotation effort required to test certain types of concurrent programs,
-which is especially relevant for operating systems students who implement their own kernels.
-In this document I have proposed a thesis which will solve both problems.
-I leave you now with a proposed timeline for bringing each project to fruition.
-
-\begin{itemize}
-	\item 2017 January: Submit thesis proposal, submit revised IRB proposal for planned user studies.
-	\item 2017 ???uary: Present thesis proposal (whenever the committee is free).
-	\item 2017 February-March: Develop ``grading'' TA user study and deploy it in 15-410 (Section~\ref{sec:grading}).
-	\item 2017 March-August: Port Landslide to Bochs for use in Pintos OS classes (Section~\ref{chap:pintos}).
-	\item 2017 March-August: Implement, evaluate, and write paper about transactional memory project (Section~\ref{sec:tm}).
-	\item 2017 August-October: Polish and deploy Pintos user study (Section~\ref{chap:pintos}).
-	\item 2017 ???ember: Resubmit paper about 410 user studies (Section~\ref{sec:studence}) (previously rejected from SIGCSE) (whenever an appropriate conference deadline arises).
-	\item 2017 October-November: Deploy 410 TA grading study again for more data.
-	\item 2017 December: Begin writing thesis.
-	\item 2018 January-April: Write thesis.
-Also fill in miscellaneous gaps, such as the ``variance'' Quicksand experiments (Section~\ref{chap:quicksand}) and measuring Landslide's heuristics' accuracy (Section~\ref{sec:studence}).
-	\item 2018 February-March: Deploy Pintos student and 410 TA studies one final time.
-	\item 2018 April-May: Defend \& graduate.
-\end{itemize}
-
-%For pretty much any reason (much like hardware transactions),
-%I may need to delay the timeline by a semester, defending instead in 2018 June.
-%For example, the above plan to deploy a Pintos user study within one semester's time of finishing the open-source Landslide port is quite ambitious,
-%assuming among other things smooth negotiations with the OS course staff at another university.
-%Anyway, this is my optimistic proposal. Thanks for reading.
-
+asdfasdasdf
 %\appendix
@@
 % up in the table of contents and pdf bookmarks.
-\renewcommand{\bibsection}{\chapter{\bibname}}
-%\newcommand{\bibpreamble}{This text goes between the ``Bibliography''
+\renewcommand{\bibsection}{\chapter{\bibname}
+\inspirationalquote{Only a doctor of philosophy, Darth.}{Robert Marsh, via Twitter}
+}
+%\renewcommand{\bibpreamble}{This text goes between the ``Bibliography''
 %  header and the actual list of references}
-\bibliographystyle{plainnat}
-\bibliography{citations} %your bib file
+\bibliographystyle{plainnat} % plain n'at
+\bibliography{register} %your bib file
 

### 2018-01-06 thesis/overview.tex
import background chapter from proposal

--- 2017-03-01 proposal/overview.tex
@@
 \subsection{Transactional Memory}
+\label{sec:overview-tm}
 
@@
 My treatment of TM will be largely agnostic to whether a transactional interface is backed by HTM or STM,
-and I will extend it to cover STM as well, as discussed in Section~\ref{sec:tm}.
+and I will extend it to cover STM as well, as discussed in Section~\ref{chap:tm}.
 
@@
 % TODO: fig tree
-I show an example execution tree in Figure~\ref{fig:tree}.
+%I show an example execution tree in Figure~\ref{fig:tree}.
 The number of preemption points in each execution define the depth of this tree, and the number of threads available to run define the branching factor.

### 2018-01-06 thesis/thesis.tex
import background chapter from proposal

@@
 
-\begin {document} 
+\begin{document}
 \frontmatter
@@
 I evaluate Landslide's suitability for inexpert use by presenting the results of many semesters providing it to students in 15-410, CMU's Operating System Design and Implementation class,
-and more recently, students in the operating systems classes at the University of Chicago and Berkeley.
+and more recently, students in similar classes at the University of Chicago and Berkeley.
 %Finally, I will present several new techniques that allow stateless model checking to be practically employed on real-world programs.
@@
 \begin{acknowledgments}
-I thank the online Android: Netrunner community Stimhack.
-\\
+% things to thank
+% stimhack
+% csd bgames
+% undergrad frenz
+% new frenz
+%
+% \\
 
@@
 
-\newpage
-\begin{figure}[h]
-	\vspace{12em}
-	\begin{center}
-		\includegraphics[width=0.5\textwidth]{how-i-learned.png}
-	\end{center}
-\end{figure}
-
 \tableofcontents
@@
 \listoftables
+
+\cleardoublepage
+\pagestyle{empty}
+\begin{center}
+\begin{tabular}{c}
+\vspace{12em} \\
+\includegraphics[width=0.55\textwidth]{how-i-learned.png}
+\end{tabular}
+\end{center}
 
@@
 
+\newcommand\simics{\textsc{Simics}}
+\newcommand{\sect}[1]{\S #1}
+\newcommand\hilight[2]{\color{#1}#2\color{black}}
+\definecolor{orange}{RGB}{192,96,0}
+\definecolor{olivegreen}{RGB}{0,127,0}
+\definecolor{brickred}{RGB}{192,0,0}
+\definecolor{commentblue}{RGB}{0,0,192}
+
+\newtheorem{theorem}{Theorem}
+
+\newcommand\llama[1]{\includegraphics[width=#1]{llama.pdf}}
+\newcommand\llitem{\item[\llama{1.2em}]}
+
 \newcommand{\inspirationallinebreak}{\vspace{0.25em}}
@@
 
-\chapter{Background}
+\input{overview}
 
@@
 \bibliographystyle{plainnat} % plain n'at
-\bibliography{register} %your bib file
+\bibliography{citations}
 

### 2018-01-06 thesis/thesis.tex
work on quotes

@@
 
-\input{overview}
+\input{background}
+
+\chapter{Quicksand}
+\inspirationalquote{Somewhere is the promise of an uncharted trail, with 700 branching limbs and 700 ways to fail.}
+{ThouShaltNot, ``Cardinal Directions''}
 
@@
 \chapter{Related Work}
-\inspirationalquote{I see now that none of us are yet ready. The cycle exists so that we may
-improve ourselves. But \\
-the one who reaches the summit is not our superior, for
-they stand on our shoulders to reach it.}
-{The Shepherd v82.6.0174, The Talos Principle}
+\inspirationalquote{
+\begin{tabular}{p{0.6\textwidth}}
+To test if your paper makes a genuine contribution to its discipline,
+see if you can afford a generous tone in the ``Related Work'' section.
+\end{tabular}}
+{Conor McBride, via Twitter}
 
@@
 	\end{tabular}}
-{The Talos Principle (Road to Gehenna)}
+{The Talos Principle: Road to Gehenna}
 

### 2018-01-06 thesis/background.tex
work on quotes
revise bgnd a bit

--- 2018-01-06 thesis/overview.tex
@@
 \label{chap:background}
+\inspirationalquote{
+\begin{tabular}{p{0.875\textwidth}}
+I see now that none of us are yet ready. The cycle exists so that we may improve ourselves. But
+the one who reaches the summit is not our superior, for they stand on our shoulders to reach it.
+\end{tabular}}
+{The Shepherd v82.6.0174, The Talos Principle}
 
 This chapter will introduce the necessary background material on concurrency, stateless model checking, data-race analysis, and the relevant undergraduate operating systems classes.
-
-Thesis proposal note: I've written this section with the intent of reusing it in the thesis.
-The other chapters, being necessarily more impermanent, are aimed directly at you (my thesis committee) as the audience.
-This section is written with the intention for programmers of any level to read.
+It is intended to be approachable by any intermediate programmer
+already familiar with C programming.
 
@@
 		\texttt{if ((status = \_xbegin()) == \_XBEGIN\_STARTED) \{} \\
-		\texttt{~~~~if (too\_much(foo)) \{} \\
+		\texttt{~~~~if (bad\_value\_of\_foo(foo)) \{} \\
 		\texttt{~~~~~~~~\_xabort();} \\
@@
 		execution will revert to the return of {\tt \_xbegin},
-		and {\tt status} will be assigned an error code indicating the abort reason.
+		{\tt status} will be assigned an error code indicating the abort reason,
+		and control will drop into the {\tt else} branch.
 		The programmer can then use explicit synchronization, such as a mutex, to resolve the conflict.}
@@
 
-In this thesis, I will focus on HTM as my platform for testing transactional programs, to highlight the importance of researching advanced testing techniques in anticipation of upcoming hardware features.
-My treatment of TM will be largely agnostic to whether a transactional interface is backed by HTM or STM,
-and I will extend it to cover STM as well, as discussed in Section~\ref{chap:tm}.
+Concerning possible execution patterns,
+the main difference between STM and HTM is the circumstances under which a transaction may abort.
+A software-backed transaction will abort if and only if a memory conflict occurs therein with another thread.
+HTM, however, is backed by the CPU's cache, and is therefore subject to other circumstances such as cache capacity or interrupt-triggered cache flushes which may force an abort even when no memory conflict occurs.
+I will explore the consequences of this difference further in Chapter~\ref{chap:tm}.
+
+In this thesis, I will focus on HTM as my platform for testing transactional programs,
+to highlight the importance of researching advanced testing techniques in anticipation of upcoming hardware features.
 

### 2018-01-08 thesis/background.tex
add fig:tree

@@
 During each execution, it forces threads to execute serially, thereby confining the program's nondeterminism to controlled thread switches.
+Using a single iteration of the {\tt x++;} loop from Figure~\ref{fig:concurrency-bug} as an example,
+Figure~\ref{fig:tree}(a) shows all possible execution interleavings of the compiled code between 2 threads.
+
+\begin{figure}[p]
+	\begin{tabular}{c}
+		\includegraphics[width=\textwidth]{statespace-list.pdf}
+		\\
+		(a) Interleavings visualized individually, as a list.
+		\\
+		\\
+		\includegraphics[width=\textwidth]{statespace-tree.pdf}
+		\\
+		(b) Interleavings (same order as in (a)), with common prefixes \\
+		combined as ``preemption points'', forming a tree.
+	\end{tabular}
+	\caption{Visualization of interleaving state space for the program in Figure~\ref{fig:concurrency-bug}.
+	Thread 1 is represented by blue ovals, thread 2 by orange squares, and time flows from top to bottom.
+	As the two threads execute the same code, without loss of generality thread 1 is fixed to run first --
+	the full state space is twice the size, and the other half is symmetric to the one shown.}
+	\label{fig:tree}
+\end{figure}
+
 Some model checkers explicitly store the set of visited program states as a means of identifying equivalent interleavings \cite{spin}.
@@
 During execution, a model checker identifies a subset of the program's operations as ``interesting'', i.e.,
-points at which interrupting the current thread to run a different one is likely to produce different behaviour.
-These points, called {\em preemption points}, may be identified by any combination of human intuition and machine analysis.
+where interrupting the current thread to run a different one is likely to produce different behaviour.
+These so-called {\em preemption points} may be identified by any combination of human intuition and machine analysis.
 Typical preemption points include the boundaries of synchronization APIs (e.g., {\tt mutex\_lock}) or accesses to shared variables.
-Considering that at each preemption point multiple threads exist as options to run next, the set of possible ways to execute the program can be viewed as a tree.
-% TODO: fig tree
-%I show an example execution tree in Figure~\ref{fig:tree}.
-The number of preemption points in each execution define the depth of this tree, and the number of threads available to run define the branching factor.
+Considering that at each preemption point multiple threads exist as options to run next,
+the set of possible ways to execute the program can be viewed as a tree.
+Figure~\ref{fig:tree}(b) shows a visualization of the corresponding tree from our example program.
+
+The number of preemption points in each execution defines the depth of this tree,
+and the number of threads available to run defines the branching factor.
 Hence, in a program with $n$ preemption points and $k$ threads available to run at each, the state space size is $O(n^k)$.
+Nevertheless, to fully test all of a program's possible behaviours, we must check the executions corresponding to every branch of the tree.
 Addressing the scaling problem in this exponential relation is the central research problem for all model checkers.
@@
 Dynamic Partial Order Reduction \cite{dpor} (henceforth, DPOR) is the premier algorithm for mitigating the exponential explosion that arises as program size increases.
-It identifies equivalent execution sequences according to Mazurkiewicz trace theory \cite{mazurkiewicz},
+Technically speaking, it identifies equivalent execution sequences according to Mazurkiewicz trace theory \cite{mazurkiewicz},
 and tests at least one execution from each equivalence class.

### 2018-01-09 thesis/background.tex
bgnd

@@
 Consider the program in Figure \ref{fig:concurrency-bug}: Any output between 2 and 2000 is possible\footnote{
-	Exercise for the reader: Show why 2 is a possible output, but 1 is not!
+	Fun exercise for the reader: Show why 2 is a possible output, but 1 is not!
 },
@@
 offering three new instructions: \texttt{xbegin}, \texttt{xend}, and \texttt{xabort}, to begin, commit, and fail a transaction, respectively.
-The example program in Figure~\ref{fig:htm-example} demonstrates how these instructions can be used to synchronize a simple shared access without locking overhead in the common case,
+The example program in Figure~\ref{fig:htm-example} demonstrates how these primitives can be used to synchronize a simple shared access without locking overhead in the common case\footnote{
+	The solution presented here is actually incomplete; stay tuned until Chapter~\ref{chap:tm} for the surprising twist!
+},
 using GCC's compiler intrinsics \cite{htm-gcc}.
@@
 		\begin{tabular}{l}
-		\textbf{(All threads)} \\
-		\hline
-		\texttt{if ((status = \_xbegin()) == \_XBEGIN\_STARTED) \{} \\
-		\texttt{~~~~if (bad\_value\_of\_foo(foo)) \{} \\
-		\texttt{~~~~~~~~\_xabort();} \\
-		\texttt{~~~~\} else \{} \\
-		\texttt{~~~~~~~~foo++;} \\
-		\texttt{~~~~~~~~\_xend();} \\
+		\texttt{void count() \{} \\
+		\texttt{~~~~for (int i = 0; i < 1000; i++) \{} \\
+		\texttt{~~~~~~~~if ((status = \_xbegin()) == \_XBEGIN\_STARTED) \{} \\
+		\texttt{~~~~~~~~~~~~x++;} \\
+		\texttt{~~~~~~~~~~~~\_xend();} \\
+		\texttt{~~~~~~~~\} else \{} \\
+		\texttt{~~~~~~~~~~~~mutex\_lock(\&m);} \\
+		\texttt{~~~~~~~~~~~~x++;} \\
+		\texttt{~~~~~~~~~~~~mutex\_unlock(\&m);} \\
+		\texttt{~~~~~~~~\}} \\
 		\texttt{~~~~\}} \\
-		\texttt{\} else if (status != \_XABORT\_EXPLICIT) \{} \\
-		\texttt{~~~~mutex\_lock(\&m);} \\
-		\texttt{~~~~foo++;} \\
-		\texttt{~~~~mutex\_unlock(\&m);} \\
 		\texttt{\}} \\
@@
 	\end{center}
-	\caption{An example program demonstrating the various HTM primitives.
+	\caption{The example {\tt count} routine from Figure~\ref{fig:mutex}, rewritten to use HTM.
 		If the transaction in the top branch aborts,
-		whether from a conflict or from the programmer's intention,
+		whether from a memory conflict or random system interrupt,
+		%from the programmer's intention,
 		execution will revert to the return of {\tt \_xbegin},
@@
 \section{Stateless Model Checking}
+
+\subsection{Overview}
 
@@
 
+\subsection{On State Space Explosion}
+
 {\bf Reduction techniques.}
@@
 and tests at least one execution from each equivalence class.
-Intuitively, if two thread transitions between preemption points do not conflict on any shared resource access, reordering them produces an equivalent interleaving, i.e., the same program behaviour.
+More intuitively, if two thread transitions between preemption points do not conflict on any shared resource access, reordering them produces an equivalent interleaving, i.e., the same program behaviour.
+% TODO: make a better fiugre based on the tree
 Figure~\ref{fig:dpor} shows a part of an execution tree in which the operations by threads 1 and 2 are independent, and hence only one of the subtrees need be tested.

### 2018-01-09 irb/infosheet-berkeley.tex
old cleanup: irb

--- 2016-04-14 irb/infosheet.tex
@@
 
-\title{\bf Landslide: Systematic Testing in 15-410 \\ User Study Information Sheet}
+% TODO for berkeley: CS162
+\newcommand\classname{CS 162}
+\title{\bf Landslide: Systematic Testing in \classname \\ User Study Information Sheet}
 \author{Ben Blum}
@@
 {\em Systematic testing}, a novel technique for testing concurrent programs, offers a mechanical way to exhaustively explore large state spaces of thread interleavings.
-The purpose of the study is to compare systematic testing against stress testing in the context of 15-410.
+The purpose of the study is to compare systematic testing against stress testing in the context of \classname.
 State-of-the-art systematic testing tools are fundamentally limited by exponential state space explosion: the user must manually configure the set of {\em preemption points} which define the state space in order to achieve meaningful test results within a reasonable CPU budget.
@@
 
+If you agree to participate in this study, your TA/course instructor will use Landslide to debug your submitted projects, and you will receive these debugging reports either before or after the project deadline, depending on the project.
+Landslide's testing results will not be used to determine your project grade.
+After you have received at least one debugging report from Landslide, we will ask you to complete a brief survey that asks about your opinions of the Landslide report.
+As part of this research, we (as researchers at Carnegie Mellon) will collect the debugging reports, snapshots of your code at the time the bug was found, and your survey responses.
+
 \section{Risks and Benefits}
@@
 {\bf Risks.}
-We expect the risks and discomfort associated with participation in this study to be no greater than those ordinarily encountered while working on 15-410 class projects.
+We expect the risks and discomfort associated with participation in this study to be no greater than those ordinarily encountered while working on \classname~class projects.
 As always with concurrent programming, it is possible to spend a lot of time looking for a bug with no meaningful result, while that time could otherwise have been spent on project implementation.
 However, as Landslide is a research tool still under development, it is also possible that a bug in Landslide itself may produce a misleading error report.
-We expect you to protect yourself by responsibly allocating time between using Landslide and working on P2 normally.
+We expect you to protect yourself by responsibly allocating time between Landslide's bug reports and working on the class projects normally.
 It is expected that participation in this study will require from 2 to 8 hours of active work, depending on bug-finding results.
@@
 \noindent {\bf Benefits.}
-Because we are providing you with a more powerful testing tool during P2 than you would normally have access to, we hope you will be able to find and fix bugs faster, resulting in a higher grade.
-%\\
-%
-%\noindent {\bf FALL 2014 NOTE}: As this semester we are conducting a preliminary study on a ``for fun'' basis, at the end of the semester rather than during P2, there will be no impact on your grade.
-15-410 staff will not accept patches based on your findings with Landslide.
+Because we are testing your code with a more powerful testing tool than is normally distributed with the project, we hope you will be able to fix more concurrency bugs, potentially improving the quality of your submitted projects and resulting in a higher grade.
 
@@
 During this study we will collect personally identifiable information from you as follows.
-While you use Landslide, it will automatically record your activity in terms of commands issued, time spent running tests, bug reports, and snapshots of your code.
+When we run Landslide on your code, we will record the same bug reports we distribute to you, as well as snapshots of your code at the time the bug was found.
+We will also collect your answers to the survey, and in this case we will also keep whatever you write confidential from the OS course staff at your university.
 %We will also collect paper surveys which you will fill out while using Landslide.
 Carnegie Mellon may be required to disclose personally identifiable information as required by law, regulation, subpoena or court order.
-Otherwise, we will protect your confidentiality as follows:
-
-\begin{itemize}
-	\item Data collected by Landslide will be stored in AFS accessible only to Ben Blum and Garth Gibson.
-	%\item Paper surveys will be stored in a locked filing cabinet accessible only to Ben Blum and Garth Gibson.
-	\item Within one semester after the study, we will destroy all personally identifying information, keeping only anonymized and aggregate results.
-	\item Professor Eckhardt will have access to study data only after final semester grades are submitted.
-Other members of course staff will never have access to study data.
-Course staff may, however, use Landslide on your final submitted code on their own.
-\end{itemize}
+%Otherwise, we will protect your confidentiality as follows:
+%
+%\begin{itemize}
+%	\item Bug reports and code snapshots will be stored on CMU computers accessible only to Ben Blum and Garth Gibson.
+%	%\item Paper surveys will be stored in a locked filing cabinet accessible only to Ben Blum and Garth Gibson.
+%	\item Within one semester after the study, we will destroy all personally identifying information, keeping only anonymized and aggregate results.
+%	%\item Professor Eckhardt will have access to study data only after final semester grades are submitted. Other members of course staff will never have access to study data.
+%		%Course staff may, however, use Landslide on your final submitted code on their own.
+%		%Course staff will also not have access to Landslide to grade your final submission.
+%\end{itemize}
 %
@@
 You must be 18 years of age or older to participate.
-You are free to stop your participation at any point; likewise, the Principal Investigator may at his discretion remove you from the study.
+You are free to stop your participation at any point; likewise, the Principal Investigator may at their discretion remove you from the study.
 Refusal to participate or discontinued participation in the study will not result in any penalty or loss of benefits or rights to which you might otherwise be entitled.
-Your participation or lack thereof will have no impact in your 15-410 class grade, except as it might (hopefully) help you write better code.
+Your participation or lack thereof will have no impact in your \classname~class grade, except as it might (hopefully) help you write better code.
 \\
@@
 or to report concerns about this study, you should contact the
-Office of Research Integrity and Compliance at Carnegie Mellon University.
-Email: {\tt irb-review@andrew.cmu.edu}. Phone: 412-268-1901 or 412-268-5460.
+Office of Research Integrity and Compliance at Carnegie Mellon University,
+or the IRB at your own institution.
+
+CMU IRB Email: {\tt irb-review@andrew.cmu.edu}. Phone: 412-268-1901 or 412-268-5460.
+
+Berkeley IRB website: {\tt https://cphs.berkeley.edu/}. Email: {\tt ophs@berkeley.edu}.
 
@@
 	\item {\bf Garth Gibson} (Faculty Advisor): Email: {\tt garth@cs.cmu.edu}; Office: GHC 9111
-	\item {\bf David Eckhardt} (Course Instructor): Email: {\tt de0u@andrew.cmu.edu}; Office: GHC 7717
+	%\item {\bf David Eckhardt} (Course Instructor): Email: {\tt de0u@andrew.cmu.edu}; Office: GHC 7717
+	\item {\bf Ion Stoica} (Course Instructor): Email: {\tt istoica@berkeley.edu}; Office: 465 Soda Hall
 \end{itemize}

### 2018-01-09 irb/infosheet-uchicago.tex
old cleanup: irb

--- 2018-01-09 irb/infosheet-berkeley.tex
@@
 % TODO for berkeley: CS162
-\newcommand\classname{CS 162}
+\newcommand\classname{CS 230}
 \title{\bf Landslide: Systematic Testing in \classname \\ User Study Information Sheet}
@@
 
-Berkeley IRB website: {\tt https://cphs.berkeley.edu/}. Email: {\tt ophs@berkeley.edu}.
+University of Chicago IRB website: {\tt https://sbsirb.uchicago.edu/}
 
@@
 	%\item {\bf David Eckhardt} (Course Instructor): Email: {\tt de0u@andrew.cmu.edu}; Office: GHC 7717
-	\item {\bf Ion Stoica} (Course Instructor): Email: {\tt istoica@berkeley.edu}; Office: 465 Soda Hall
+	\item {\bf Haryadi Gunawi} (Course Instructor): Email: {\tt haryadi@cs.uchicago.edu}; Office: Ryerson Hall 259
 \end{itemize}

### 2018-01-09 irb/landslide-guide-p2.tex
old cleanup: irb

@@
 	\item Run {\tt ./landslide OPTIONS} to run tests with Landslide, for various values of {\tt OPTIONS}:
-		\begin{itemize}
+		\begin{enumerate}
 			\item Use ./landslide -h to view command line options. The most important ones are -t , max testing
@@
 				Then run {\tt ./p2-setup.sh} again.
-		\end{itemize}
+		\end{enumerate}
 	\item After fixing any found bugs, re-run Landslide with the same test options to confirm that your fix works.
 \end{enumerate}
+
+\noindent {\bf Note on simultaneous Landslides:} Landslide does not support running multiple tests at once;
+it needs to create some temporary files to annotate your code, and multiple instances of Landslide can step on each other's toes by clobbering those files.
+If you want to try multiple tests at once, simply {\tt git clone} the repository again in a different directory, and run one instance of Landslide in each.
+\\
 

### 2018-01-09 thesis/intro.tex
add incomplete intro

--- 2017-03-01 proposal/thesprop.tex
@@
-%for a more compact document, add the option openany to avoid
-%starting all chapters on odd numbered pages
-\documentclass[12pt]{cmuthesis}
-
-% This is a template for a CMU thesis.  It is 18 pages without any content :-)
-% The source for this is pulled from a variety of sources and people.
-% Here's a partial list of people who may or may have not contributed:
-%
-%        bnoble   = Brian Noble
-%        caruana  = Rich Caruana
-%        colohan  = Chris Colohan
-%        jab      = Justin Boyan
-%        josullvn = Joseph O'Sullivan
-%        jrs      = Jonathan Shewchuk
-%        kosak    = Corey Kosak
-%        mjz      = Matt Zekauskas (mattz@cs)
-%        pdinda   = Peter Dinda
-%        pfr      = Patrick Riley
-%        dkoes = David Koes (me)
-
-% My main contribution is putting everything into a single class files and small
-% template since I prefer this to some complicated sprawling directory tree with
-% makefiles.
-
-% some useful packages
-\usepackage{times}
-\usepackage{fullpage}
-\usepackage{graphicx}
-\usepackage{amsmath}
-\usepackage{amsthm}
-\usepackage[charter]{mathdesign}
-\usepackage[numbers,sort]{natbib}
-\usepackage[backref,pageanchor=true,plainpages=false, pdfpagelabels, bookmarks,bookmarksnumbered,
-%pdfborder=0 0 0,  %removes outlines around hyper links in online display
-]{hyperref}
-\usepackage{subfigure}
-\usepackage[linesnumbered,ruled]{algorithm2e}
-\usepackage{algpseudocode}
-
-% Approximately 1" margins, more space on binding side
-%\usepackage[letterpaper,twoside,vscale=.8,hscale=.75,nomarginpar]{geometry}
-%for general printing (not binding)
-\usepackage[letterpaper,twoside,vscale=.8,hscale=.75,nomarginpar,hmarginratio=1:1]{geometry}
-
-% Provides a draft mark at the top of the document.
-\draftstamp{\today}{DRAFT}
-
-\newcommand\simics{\textsc{Simics}}
-\newcommand{\sect}[1]{\S #1}
-\newcommand\hilight[2]{\color{#1}#2\color{black}}
-\definecolor{orange}{RGB}{192,96,0}
-\definecolor{olivegreen}{RGB}{0,127,0}
-\definecolor{brickred}{RGB}{192,0,0}
-\definecolor{commentblue}{RGB}{0,0,192}
-
-\newtheorem{theorem}{Theorem}
-
-\newcommand\llama[1]{\includegraphics[width=#1]{llama.pdf}}
-\newcommand\llitem{\item[\llama{1.2em}]}
-
-\begin {document}
-\frontmatter
-
-%initialize page style, so contents come out right (see bot) -mjz
-\pagestyle{empty}
-
-\title{ %% {\it \huge Thesis Proposal}\\
-{\bf Practical Concurrency Testing} \\
-\normalsize \vspace{1em}
-{\em or, How I Learned to Stop Worrying and Love the Exponential Explosion}}
-\author{Ben Blum}
-\date{January 2017}
-\Year{2017}
-\trnumber{}
-
-\committee{
-Garth Gibson, Chair \\
-David A. Eckhardt \\
-Brandon Lucia \\
-Haryadi Gunawi (University of Chicago)
-}
-
-\support{}
-\disclaimer{}
-
-% copyright notice generated automatically from Year and author.
-% permission added if \permission{} given.
-
-% TODO(thesis): Save this one for the thesis itself...
-%\keywords{landslide terminal, baggage claim, ground transportation, ticketing}
-
-\maketitle
-
-%\begin{dedication}
-%	For my dog, Louie.
-%\end{dedication}
-
-\pagestyle{plain} % for toc, was empty
-
-%% Obviously, it's probably a good idea to break the various sections of your thesis
-%% into different files and input them into this file...
-
-\begin{abstract}
-Concurrent programming presents a challenge to students and experts alike because of the complexity of multithreaded interactions and the difficulty to reproduce and reason about bugs.
-Stateless model checking is a concurrency testing approach which forces a program to interleave its threads in many different ways, checking for bugs each time.
-This technique is powerful, in principle capable of finding any nondeterministic bug in finite time,
-but suffers from exponential explosion as program size increases.
-Checking an exponential number of thread interleavings is not a practical or predictable approach for programmers to find concurrency bugs before their project deadlines.
-
-In this thesis, I propose to make stateless model checking more practical for human use by way of several new techniques.
-I have built Landslide, a stateless model checker specializing in student projects for undergraduate operating systems classes.
-Landslide includes a novel algorithm for automatically managing
-%the exploration of
-multiple state spaces according to their estimated completion times,
-which I will show quickly finds bugs should they exist and also quickly verifies correctness otherwise.
-I will evaluate Landslide's suitability for inexpert use by presenting the results of many semesters providing it to students in 15-410, CMU's Operating System Design and Implementation class.
-%Finally, I will present several new techniques that allow stateless model checking to be practically employed on real-world programs.
-Finally, I will explore broader impact by extending Landslide to test some real-world programs and to be used by students at other universities.
-\end{abstract}
-
-%\begin{acknowledgments}
-%My advisor is cool.
-%
-%But I am cooler.
-%\end{acknowledgments}
-
-\tableofcontents
-%\listoffigures
-%\listoftables
-
-\mainmatter
-
-%% Double space document for easy review:
-%\renewcommand{\baselinestretch}{1.66}\normalsize
-
-% The other requirements Catherine has:
-%
-%  - avoid large margins.  She wants the thesis to use fewer pages,
-%    especially if it requires colour printing.
-%
-%  - The thesis should be formatted for double-sided printing.  This
-%    means that all chapters, acknowledgements, table of contents, etc.
-%    should start on odd numbered (right facing) pages.
-%
-%  - You need to use the department standard tech report title page.  I
-%    have tried to ensure that the title page here conforms to this
-%    standard.
-%
-%  - Use a nice serif font, such as Times Roman.  Sans serif looks bad.
-%
-% Other than that, just make it look good...
-
 \chapter{Introduction}
+\inspirationalquote{
+\begin{tabular}{p{0.65\textwidth}}
+Please hold on. This thesis will be departing shortly for the Landslide terminal, baggage claim, ground transportation, and ticketing.
+\end{tabular}}
+{Pittsburgh International Airport (paraphrased)}
 
 Modern computer architectures have turned to increasing CPU core count, rather than clock speed, to improve processing power \cite{mooreslaw}.
-%To take advantage of multiple cores, programs must be written {\em concurrently}
 To take advantage of multiple cores for performance, programmers must write software to execute {\em concurrently} --
@@
 providing no guarantee of finding the failing interleaving in any finite amount of time.
-%, even when a bug does exist.
 It also provides no assurance of correctness:
 when finished, there is no way of knowing how many distinct thread interleavings were actually tested.
-%it may by chance test only a single interleaving over and over again!
 Nevertheless, stress testing remains popular because of how easily a programmer can use it:
@@
 focusing on fully testing that subset, while ignoring other possibilities she doesn't care about.
-%However, choosing this subset is a difficult trade-off for humans to make before even knowing whether a bug exists.
 However, it is difficult to choose a subset of thread interleavings that will produce a meaningful, yet feasible test.
@@
 \begin{center}
+	% TODO: this sux, fix it
 	{\em Thanks to the new algorithms, heuristics, and concurrency models I have developed,
@@
 		I will show that Iterative Deepening of preemption points outperforms prior work in terms both of finding bugs quickly and of completely verifying correctness when no bug exists.
+	\item {\bf Educational use (Chapters \ref{chap:410} and \ref{chap:pintos}).}
+		For the past five semesters, I have offered a fully-automated version of Landslide to students in 15-410, CMU's undergraduate Operating System Design and Implementation class \cite{kspec,thrlib}, for use as a debugging aid during the thread library project.
+		Recently I have also extended Landslide to handle Pintos kernel projects from other universities \cite{pintos}.
+		In the two most recent semesters, I collaborated with Operating Systems course staff at two such schools, the University of Chicago and University of California at Berkeley,
+		to provide debugging feedback to their students.
 
-		This work is {\bf completed} and was published at OOPSLA 2016.
-	\item {\bf Educational use (Chapters \ref{chap:410} and \ref{chap:pintos}).}
-		For the past three semesters, I have offered a fully-automated version of Landslide to students in 15-410, CMU's undergraduate Operating System Design and Implementation class \cite{kspec,thrlib}, for use as a debugging aid during the thread library project.
-		I will continue these user studies, and use the data to evaluate the suitability of stateless model checking in an educational setting.
-		%, investigating the following questions in particular:
-		%\begin{itemize}
-		%	\item Does Landslide improve project submission quality when used as a debugging aid?
-		%	\item Does Landslide teach students lessons in concurrent programming
-		%\end{itemize}
-
-		So far, the fully-automatic testing mode is available only for 15-410 thread library projects.
-		To prove these techniques are relevant beyond CMU's walls, I will extend Landslide to handle Pintos kernel projects from other universities \cite{pintos}.
-		I will then collaborate with those schools to deploy Landslide to their operating systems students.
-
-		This is {\bf ongoing work}; I have run the user study for 4 semesters so far and am proposing to continue them
-		and extend the existing evaluation,
-		as well as deploy a Pintos port.
+		At all three universities I then collected statistics on the numbers and types of bugs found,
+		and surveyed students to understand the human experience,
+		This section will present the study's results
+		to evaluate the suitability of stateless model checking in an educational setting.
 	\item {\bf Transactional Memory (Chapter \ref{chap:tm}).}
@@
 		which is not yet addressed by modern model checkers.
-		I will extend Landslide's concurrency model to support both hardware (HTM) and software (STM) variants of TM,
-		and test several ``real-world'' TM programs.
-
-		This will be {\bf new work},
-		involving some implementation effort, some proofs, and designing a new evaluation.
-		%, as Landslide presently supports testing Pintoses only with some manual effort,
-		%and does not yet feature instrumentation to run on Linux programs.
-		%and will need its concurrency model extended to incorporate the HTM programming model.
+		I have extended Landslide's concurrency model to support both hardware (HTM) and software (STM) variants of TM,
+		and tested several ``real-world'' TM programs and benchmarks.
+		This section will discuss the theoretical techniques I used to model the new form of concurrency,
+		present associated correctness proofs of my approach,
+		and show the testing results.
 \end{itemize}
 
-% TODO(thesis): Other chapters of this thesis include...
-
-\input{overview}
-\input{quicksand}
-\input{410}
-\input{pintos}
-\input{tm}
-\input{related}
-
-%Reduction
-%techniques such as Dynamic Partial Order Reduction \cite{dpor} and Maximal Causality Reduction \cite{mcr} expand the limits of feasible test completion,
-%and search ordering strategies such as Iterative Context Bounding \cite{chess-icb} \revision{encourage}~bugs to be found sooner in a given space should they exist.
-
-%Topics of thesis:
-%1. Data race preemption points / iterative deepening
-%2. Education in 410. Evaluation questions are:
-%	2a. Does P2 submission quality improve?
-%	2b. Does P3 submission quality improve correlated with students who used (got the most out of) Landslide during P2?
-%	2c. Does Landslide reach the bottom of the class? (Look at bottom-of-class P0s/P1s, see how they improve)
-%3. Broader impact
-%	3a. Real world programs (and the challenges therein?)
-%	3b. Give to pintos kids (and open-source landslide on bochs in doing so)
-
-\chapter{Conclusion and Timeline}
-
-Stateless model checking is a powerful technique for testing concurrent programs,
-capable in theory of rooting out any bug or providing total verification on any program,
-but suffers several problems which relegate that theory to fantasy land \cite{vargomax}.
-Chief among those problems is exponential explosion of state spaces,
-making it difficult to decide in advance which combinations of preemption points can be productively tested within a given time limit.
-Another major problem is the manual annotation effort required to test certain types of concurrent programs,
-which is especially relevant for operating systems students who implement their own kernels.
-In this document I have proposed a thesis which will solve both problems.
-I leave you now with a proposed timeline for bringing each project to fruition.
-
-\begin{itemize}
-	\item 2017 January: Submit thesis proposal, submit revised IRB proposal for planned user studies.
-	\item 2017 ???uary: Present thesis proposal (whenever the committee is free).
-	\item 2017 February-March: Develop ``grading'' TA user study and deploy it in 15-410 (Section~\ref{sec:grading}).
-	\item 2017 March-August: Port Landslide to Bochs for use in Pintos OS classes (Section~\ref{chap:pintos}).
-	\item 2017 March-August: Implement, evaluate, and write paper about transactional memory project (Section~\ref{sec:tm}).
-	\item 2017 August-October: Polish and deploy Pintos user study (Section~\ref{chap:pintos}).
-	\item 2017 ???ember: Resubmit paper about 410 user studies (Section~\ref{sec:studence}) (previously rejected from SIGCSE) (whenever an appropriate conference deadline arises).
-	\item 2017 October-November: Deploy 410 TA grading study again for more data.
-	\item 2017 December: Begin writing thesis.
-	\item 2018 January-April: Write thesis.
-Also fill in miscellaneous gaps, such as the ``variance'' Quicksand experiments (Section~\ref{chap:quicksand}) and measuring Landslide's heuristics' accuracy (Section~\ref{sec:studence}).
-	\item 2018 February-March: Deploy Pintos student and 410 TA studies one final time.
-	\item 2018 April-May: Defend \& graduate.
-\end{itemize}
-
-%For pretty much any reason (much like hardware transactions),
-%I may need to delay the timeline by a semester, defending instead in 2018 June.
-%For example, the above plan to deploy a Pintos user study within one semester's time of finishing the open-source Landslide port is quite ambitious,
-%assuming among other things smooth negotiations with the OS course staff at another university.
-%Anyway, this is my optimistic proposal. Thanks for reading.
-
-%\appendix
-%\include{appendix}
-
-\backmatter
-
-%\renewcommand{\baselinestretch}{1.0}\normalsize
-
-% By default \bibsection is \chapter*, but we really want this to show
-% up in the table of contents and pdf bookmarks.
-\renewcommand{\bibsection}{\chapter{\bibname}}
-%\newcommand{\bibpreamble}{This text goes between the ``Bibliography''
-%  header and the actual list of references}
-\bibliographystyle{plainnat}
-\bibliography{citations} %your bib file
-
-\end{document}
+The rest of this dissertation is organized as follows.
+% TODO(thesis):

### 2018-01-09 thesis/intro.tex
write intro organization section

@@
 {Pittsburgh International Airport (paraphrased)}
+
+\subsection{Motivation}
 
@@
 
+\subsection{Contribution}
+
 This thesis will solve both problems discussed above.
@@
 
-\begin{itemize}
+\begin{enumerate}
 	\item {\bf Meaningful state spaces (Chapter \ref{chap:quicksand}).}
@@
 		and show the testing results.
-\end{itemize}
+\end{enumerate}
+
+\subsection{Organization}
 
 The rest of this dissertation is organized as follows.
-% TODO(thesis):
+
+\begin{itemize}
+	\item {\bf Background:} Chapter~\ref{chap:background} will present the requisite background material on concurrent programming, stateless model checking, and the various types of programs targeted by Landslide.
+
+		{\bf Special personal note for the curious reader:} I have taken special care to write this chapter to be approachable by any intermediate programmer familiar with basic C programming concepts.
+		My committee members obviously do not require such treatment, but I hope the rare reader who happens upon this work from outside my circle of research collaborators shall also not find the barrier to entry here too high.
+		Concurrency analysis and verification is often very abstract,
+		requiring strong intuition for certain concepts to understand the next algorithm that builds upon them, and so on;
+		and I'd hate for any reader to get left behind off the bat for not knowing what a mutex is.
+		%Abstract as the field is, it's also quite difficult to make good visual aids, unlike for example in graphics research, but I have done my best.
+		Should you find any of my explanations insufficiently illuminating, no matter how far in the future you come from, please do get in touch.
+	\item {\bf Landslide:} Chapter~\ref{chap:landslide} explains the design and implementation of Landslide,
+		the core of my stateless model checking framework,
+		and all the special features it's been equipped with over the years.
+	\item {\bf Quicksand:} Chapter~\ref{chap:quicksand} presents the Iterative Deepening framework I've developed to more intelligently choose which state spaces to test, corresponding to contribution 1 above.
+	\item {\bf Pebbles:} Chapter~\ref{chap:410} discusses my evaluation of Landslide in CMU's 15-410 class environment using the Pebbles kernel.
+	\item {\bf Pintos:} Chapter~\ref{chap:410} discusses my evaluation of Landslide in the University of Chicago's and Berkeley's OS class environments using the Pintos kernel.
+	\item {\bf Transactional Memory:} Chapter~\ref{chap:tm} presents my extension of Landslide's concurrency model to handle transactional concurrency and the evaluation thereof.
+	\item {\bf Related Work:} Chapter~\ref{chap:relatedwork} honors my ancestors and neighbours in research spirit.
+	\item {\bf Conclusion:} Chapter~\ref{chap:conclusion} hopes to leave you with thoughts on the future of this field.
+\end{itemize}

### 2018-01-10 thesis/background.tex
rearrange some stuff bw intro/bgnd

@@
 This chapter will introduce the necessary background material on concurrency, stateless model checking, data-race analysis, and the relevant undergraduate operating systems classes.
-It is intended to be approachable by any intermediate programmer
-already familiar with C programming.
+
+{\bf Special personal note for the curious reader:} I have taken special care to write this chapter to be approachable by any intermediate programmer familiar with basic C programming concepts.
+My committee members obviously do not require such treatment, but I hope the rare reader from outside this relatively narrow field shall also not find the barrier to entry here too high.
+Concurrency analysis and verification is often very abstract,
+requiring strong intuition for certain concepts to understand the next algorithm that builds upon them, and so on;
+and I'd hate for any reader to get left behind off the bat for not knowing what a mutex is.
+%Abstract as the field is, it's also quite difficult to make good visual aids, unlike for example in graphics research, but I have done my best.
+Should you find any of my explanations insufficiently illuminating, please do get in touch.
 

### 2018-01-10 thesis/intro.tex
rearrange some stuff bw intro/bgnd

--- 2018-01-09 thesis/intro.tex
@@
 		and show the testing results.
+\end{enumerate}
+
+\subsection{Organization}
+
+The rest of this dissertation is organized as follows.
+
+\begin{itemize}
+	\item {\bf Background:} Chapter~\ref{chap:background} will present the requisite background material on concurrent programming, stateless model checking, and the various types of programs targeted by Landslide.
+	\item {\bf Landslide:} Chapter~\ref{chap:landslide} explains the design and implementation of Landslide
+		%the core of my stateless model checking framework,
+		and all the special features it's been equipped with over the years.
+	\item {\bf Quicksand:} Chapter~\ref{chap:quicksand} presents the Iterative Deepening framework which more intelligently chooses which state spaces to test, corresponding to contribution 1 above.
+	\item {\bf Pebbles:} Chapter~\ref{chap:410} discusses my evaluation of Landslide in CMU's 15-410 class environment using the Pebbles kernel, corresponding to contribution 2 above.
+	\item {\bf Pintos:} Chapter~\ref{chap:410} discusses my evaluation of Landslide in the University of Chicago's and Berkeley's OS class environments using the Pintos kernel, corresponding to contribution 2 above.
+	\item {\bf Transactional Memory:} Chapter~\ref{chap:tm} presents my extension of Landslide's concurrency model to handle transactional concurrency and the evaluation thereof, corresponding to contribution 3 above.
+	\item {\bf Related Work:} Chapter~\ref{chap:relatedwork} honors my neighbours and ancestors in research spirit.
+	\item {\bf Conclusion:} Chapter~\ref{chap:conclusion} provides some thoughts on the future of the field.
 \end{itemize}
 
-The rest of this dissertation is organized as follows.
-% TODO(thesis):
+\newpage
+\thispagestyle{empty}
+\begin{center}
+\begin{tabular}{c}
+\vspace{12em} \\
+\includegraphics[width=0.55\textwidth]{how-i-learned.png}
+\end{tabular}
+\end{center}

### 2018-01-10 thesis/background.tex
organize bgnd

@@
 
-\section{Concurrency Bugs}
+\section{Concurrency}
 
@@
 In a multithreaded program, multiple execution units (or {\em threads}) execute the same or different sections of code simultaneously.
+This can provide speedups up to a factor of the number of threads running in parallel,
+but may also provide surprising execution results.
 
-{\bf Simultaneity.}
+\subsubsection{Simultaneity}
 This simultaneity of threads is achieved either by executing each one on a separate CPU, or by interleaving them nondeterministically (as controlled by clock interrupts) on the same CPU.
@@
 
-{\bf Shared state.}
+\subsubsection{Shared state}
 When a programming language offers multithreaded parallelism but forbids access to any shared state between threads \cite{rust-language},
@@
 
-{\bf Identifying bugs.}
+\subsection{Identifying bugs}
+
 Even if a program's behaviour is nondeterministic, that does not necessarily mean it has a bug.
@@
 Most often, a concurrency novice who programs with shared state will consider the possible interleavings where one thread's access sequence occurs entirely before the other's, but neglect to consider intermediate outcomes in which the threads' access sequences are interleaved.
-%
+
 Consider the program in Figure \ref{fig:concurrency-bug}: Any output between 2 and 2000 is possible\footnote{
@@
 
-\subsection{Overview}
+\subsection{The state space}
 
@@
 
-{\bf Static versus dynamic analysis.}
+\subsubsection{Static versus dynamic analysis}
+
 Model checking is a {\em dynamic} program analysis, meaning that it observes the operations and accesses performed by the program as its code is executed.
@@
 
-{\bf Preemption points.}
+\subsubsection{Preemption points}
+
 During execution, a model checker identifies a subset of the program's operations as ``interesting'', i.e.,
@@
 
-\subsection{On State Space Explosion}
+\subsection{On the size of state spaces}
 
-{\bf Reduction techniques.}
+At its essence, stateless model checking research is a perpetual struggle to become more and more efficient in order to test and verify bigger and bigger programs.
+But whence this efficiency?
+Techniques for coping with the exponential explosion fall into two categories:
+(1) removing redundant interleavings from the state space when we can prove they are equivalent to some interleaving already tested,
+or {\bf reduction techniques},
+and
+(2) prioritizing interleavings judged as more likely to contain bugs should bugs exist
+in case we are unable to exhaustively test all interleavings after all,
+or {\bf search heuristics}.
+
+\subsubsection{Reduction techniques}
+
 Dynamic Partial Order Reduction \cite{dpor} (henceforth, DPOR) is the premier algorithm for mitigating the exponential explosion that arises as program size increases.
@@
 More intuitively, if two thread transitions between preemption points do not conflict on any shared resource access, reordering them produces an equivalent interleaving, i.e., the same program behaviour.
+
 % TODO: make a better fiugre based on the tree
@@
 
-{\bf Search heuristics.}
+\subsubsection{Search heuristics}
+
 However, even though DPOR can prune an exponential number of redundant interleavings, the state space size is still exponential in the number of {\em dependent} (conflicting) interleavings.
@@
 or missing the chance to complete a full verification.
+
 Iterative Context Bounding \cite{chess-icb} is a popular such technique which heuristically reorders the search to prioritize interleavings with fewer preemptions first.
@@
 although it requires some human intuition to correctly mark those boundaries.
+Iterative Deepening, presented in Chapter~\ref{chap:quicksand}, is another such search heuristic.
 
@@
 
+\subsection{Definition}
+
 Data race analysis \cite{eraser} identifies pairs of unsynchronized memory accesses between threads.
@@
 
-{\bf Variants of Happens-Before.}
-        Most prior work focuses on {\em Happens-Before} \cite{lamport-clocks} as the order relation between accesses.
+\subsection{Happens-Before}
+
+Condition 4 of the above definition expresses the notion that the access pair can be executed concurrently,
+regardless of whether the hardware actually carries out the operations in the same physical instant.
+Several approaches exist to formally representing this condition.
+
+\begin{itemize}
+	\item Most prior work focuses on {\em Happens-Before} \cite{lamport-clocks} as the order relation between accesses.
 \cite{predictive-dr} and \cite{hybriddatarace} identify a problem with this approach:
@@
 We call such unreported access pairs {\em false negatives}.
+
+\item
 \cite{hybriddatarace} introduces the {\em Limited Happens-Before} relation,
@@
 
-%Finally, the {\em Causally-Precedes} relation \cite{predictive-dr} %strikes a middle ground,
-%extends Happens-Before to additionally report a subset of potential races while soundly avoiding false positives.
-%It tracks conflicting accesses in intervening
-%critical sections to determine whether lock events are unrelated to a potential race.
-%Causally-Precedes will identify the potential race in Figure~\ref{fig:hb-example}(a), as the two critical sections do not conflict,
-%although it can still miss true potential races in other cases.
+\item
+In recent work, the {\em Causally-Precedes} relation \cite{predictive-dr} %strikes a middle ground,
+extends Happens-Before to additionally report a subset of potential races while soundly avoiding false positives.
+It tracks conflicting accesses in intervening
+critical sections to determine whether lock events are unrelated to a potential race.
+Causally-Precedes will identify the potential race in Figure~\ref{fig:hb-example}(a), as the two critical sections do not conflict,
+although it can still miss true potential races in other cases.
+\end{itemize}
 
-In Chapter~\ref{chap:quicksand}, we use the Limited Happens-Before relation for our analysis.
-The justification for this is that, while stand-alone data-race analyses must avoid inundating the user with false alarms \cite{racerx},
-my work incorporates data-race analysis in an internal feedback loop, and reports only directly observed failures to the user.
-Hence, I accept some overhead from false positives for the sake of more thorough testing.
+Landslide implements both Happens-Before (henceforth referred to as {\em Pure Happens-Before} for clarity) and Limited Happens-Before.
+Chapter~\ref{chap:quicksand} includes a comparison of the two approaches for the purpose of finding new preemption points for model checking.
+%, we use the Limited Happens-Before relation for our analysis.
+%The justification for this is that, while stand-alone data-race analyses must avoid inundating the user with false alarms \cite{racerx},
+%my work incorporates data-race analysis in an internal feedback loop, and reports only directly observed failures to the user.
+%Hence, I accept some overhead from false positives for the sake of more thorough testing.
 

### 2018-01-10 thesis/landslide.tex
add landslide chapt

--- 2018-01-03 thesis/thesis.tex
@@
-%for a more compact document, add the option openany to avoid
-%starting all chapters on odd numbered pages
-\documentclass[12pt]{cmuthesis}
+\chapter{Landslide}
+\inspirationalquote{Somewhere is the promise of an uncharted trail, with 700 branching limbs and 700 ways to fail.}
+{ThouShaltNot, ``Cardinal Directions''}
 
-% This is a template for a CMU thesis.  It is 18 pages without any content :-)
-% The source for this is pulled from a variety of sources and people.
-% Here's a partial list of people who may or may have not contributed:
-%
-%        bnoble   = Brian Noble
-%        caruana  = Rich Caruana
-%        colohan  = Chris Colohan
-%        comar    = Cyrus Omar
-%        jab      = Justin Boyan
-%        josullvn = Joseph O'Sullivan
-%        jrs      = Jonathan Shewchuk
-%        kosak    = Corey Kosak
-%        mjz      = Matt Zekauskas (mattz@cs)
-%        pdinda   = Peter Dinda
-%        pfr      = Patrick Riley
-%        dkoes = David Koes (me)
-
-% My main contribution is putting everything into a single class files and small
-% template since I prefer this to some complicated sprawling directory tree with
-% makefiles.
-
-% some useful packages
-\usepackage[charter]{mathdesign}
-\usepackage{fullpage}
-\usepackage{graphicx}
-\usepackage{amsmath}
-\usepackage[numbers,sort]{natbib}
-\usepackage[backref,pageanchor=true,plainpages=false, pdfpagelabels, bookmarks,bookmarksnumbered,
-%pdfborder=0 0 0,  %removes outlines around hyper links in online display
-]{hyperref}
-\usepackage{subfigure}
-
-% Approximately 1" margins, more space on binding side
-%\usepackage[letterpaper,twoside,vscale=.8,hscale=.75,nomarginpar]{geometry}
-%for general printing (not binding)
-\usepackage[letterpaper,twoside,vscale=.8,hscale=.75,nomarginpar,hmarginratio=1:1]{geometry}
-
-% Provides a draft mark at the top of the document. 
-\draftstamp{\today}{DRAFT}
-
-\begin {document} 
-\frontmatter
-
-%initialize page style, so contents come out right (see bot) -mjz
-\pagestyle{empty}
-
-\title{ %% {\it \huge Thesis Proposal}\\
-{\bf Practical Concurrency Testing} \\
-\normalsize \vspace{1em}
-{\em or, How I Learned to Stop Worrying and Love the Exponential Explosion}}
-\author{Ben Blum}
-\date{Decembruary 2018}
-\Year{2018}
-\trnumber{}
-
-\committee{
-Garth Gibson, Chair \\
-David A. Eckhardt \\
-Brandon Lucia \\
-Haryadi Gunawi, University of Chicago
-}
-
-\support{}
-\disclaimer{}
-
-% copyright notice generated automatically from Year and author.
-% permission added if \permission{} given.
-
-\keywords{landslide terminal, baggage claim, ground transportation, ticketing}
-
-\maketitle
-
-\begin{dedication}
-For Elsie, Margie, Johnny, Mel, Eve, and David, who brought me here today.
-\end{dedication}
-
-\pagestyle{plain} % for toc, was empty
-
-%% Obviously, it's probably a good idea to break the various sections of your thesis
-%% into different files and input them into this file...
-
-\begin{abstract}
-Concurrent programming presents a challenge to students and experts alike because of the complexity of multithreaded interactions and the difficulty to reproduce and reason about bugs.
-Stateless model checking is a concurrency testing approach which forces a program to interleave its threads in many different ways, checking for bugs each time.
-This technique is powerful, in principle capable of finding any nondeterministic bug in finite time,
-but suffers from exponential explosion as program size increases.
-Checking an exponential number of thread interleavings is not a practical or predictable approach for programmers to find concurrency bugs before their project deadlines.
-
-In this thesis, I develop several new techniques to make stateless model checking more practical for human use.
-I have built Landslide, a stateless model checker specializing in undergraduate operating systems class projects.
-Landslide extends the traditional model checking algorithm with a new framework for automatically managing
-%the exploration of
-multiple state spaces according to their estimated completion times,
-which I show quickly finds bugs should they exist and also quickly verifies correctness otherwise.
-I evaluate Landslide's suitability for inexpert use by presenting the results of many semesters providing it to students in 15-410, CMU's Operating System Design and Implementation class,
-and more recently, students in the operating systems classes at the University of Chicago and Berkeley.
-%Finally, I will present several new techniques that allow stateless model checking to be practically employed on real-world programs.
-%Finally, I will explore broader impact by extending Landslide to test some real-world programs and to be used by students at other universities.
-Finally, I extend Landslide with a new concurrency model for hardware transactional memory,
-and evaluate several real-world transactional benchmarks to show that stateless model checking can keep up with the developing concurrency demands of real-world programs.
-\end{abstract}
-
-\begin{acknowledgments}
-I thank the online Android: Netrunner community Stimhack.
-\\
-
-Please hold on. This thesis will be departing shortly for the Landslide terminal, baggage claim, ground transportation, and ticketing.
-\end{acknowledgments}
-
-\newpage
-\begin{figure}[h]
-	\vspace{12em}
-	\begin{center}
-		\includegraphics[width=0.5\textwidth]{how-i-learned.png}
-	\end{center}
-\end{figure}
-
-\tableofcontents
-\listoffigures
-\listoftables
-
-\mainmatter
-
-%% Double space document for easy review:
-%\renewcommand{\baselinestretch}{1.66}\normalsize
-
-% The other requirements Catherine has:
-%
-%  - avoid large margins.  She wants the thesis to use fewer pages, 
-%    especially if it requires colour printing.
-%
-%  - The thesis should be formatted for double-sided printing.  This
-%    means that all chapters, acknowledgements, table of contents, etc.
-%    should start on odd numbered (right facing) pages.
-%
-%  - You need to use the department standard tech report title page.  I
-%    have tried to ensure that the title page here conforms to this
-%    standard.
-%
-%  - Use a nice serif font, such as Times Roman.  Sans serif looks bad.
-%
-% Other than that, just make it look good...
-
-\newcommand{\inspirationallinebreak}{\vspace{0.25em}}
-\newcommand\inspirationalquote[2]{\begin{flushright}
-	\vspace{-1em}
-	\fontfamily{pzc}\selectfont
-	{\em #1}
-	\inspirationallinebreak
-
-	---{#2}
-	\fontfamily{mdbch}\selectfont
-	\vspace{2em}
-\end{flushright}}
-
-\chapter{Introduction}
-
-\chapter{Background}
-
-\chapter{410} % rename
-\inspirationalquote{Knowing the students might one day
-%find a way to
-fix their concurrency bugs...
-it fills you with determination.}{Undertale (paraphrased)}
-
-\chapter{Related Work}
-\inspirationalquote{I see now that none of us are yet ready. The cycle exists so that we may
-improve ourselves. But \\
-the one who reaches the summit is not our superior, for
-they stand on our shoulders to reach it.}
-{The Shepherd v82.6.0174, The Talos Principle}
-
-\chapter{Conclusion}
-\inspirationalquote{
-%	``If you are a god, Zeus, as the stories claim, then why did you create evolution?
-%	Why did you make \\ a world that can only grow through cruelty and pain?''
-%	\inspirationallinebreak
-%
-%For a moment--surely this meant death was near--she thought she heard him answer:
-%``My child, I \\
-%was shaped by the gods that came before me, as you were shaped by me.
-%The choice I had was be- \\
-%tween creation and oblivion, life and death.
-%And I chose life, because any life is better than no life, \\
-%because as long as there is life, there is hope--if not for us, then for some generation to come.''
-%	\inspirationallinebreak
-%
-%``Then how are you a god, if you can offer so little?'' she whispered, feeling death creep closer.
-%	\inspirationallinebreak
-%
-%``I am a god because I take upon myself the burden of creation,'' the statue replied.
-%	\inspirationallinebreak
-%
-%``Then we are all gods,'' Alexandra said, and pushed the button.}
-	\begin{tabular}{p{0.875\textwidth}}
-	``If you are a god, Zeus, as the stories claim, then why did you create evolution?
-	Why did you make a world that can only grow through cruelty and pain?''
-	\\
-	For a moment--surely this meant death was near--she thought she heard him answer:
-	``My child, I was shaped by the gods that came before me, as you were shaped by me.
-	The choice I had was between creation and oblivion, life and death.
-	And I chose life, because any life is better than no life,
-	because as long as there is life, there is hope--if not for us, then for some generation to come.''
-	\\
-	``Then how are you a god, if you can offer so little?'' she whispered, feeling death creep closer.
-	\\
-	``I am a god because I take upon myself the burden of creation,'' the statue replied.
-	\\
-	``Then we are all gods,'' Alexandra said, and pushed the button.
-	\end{tabular}}
-{The Talos Principle (Road to Gehenna)}
-
-asdfasdasdf
-%\appendix
-%\include{appendix}
-
-\backmatter
-
-%\renewcommand{\baselinestretch}{1.0}\normalsize
-
-% By default \bibsection is \chapter*, but we really want this to show
-% up in the table of contents and pdf bookmarks.
-\renewcommand{\bibsection}{\chapter{\bibname}
-\inspirationalquote{Only a doctor of philosophy, Darth.}{Robert Marsh, via Twitter}
-}
-%\renewcommand{\bibpreamble}{This text goes between the ``Bibliography''
-%  header and the actual list of references}
-\bibliographystyle{plainnat} % plain n'at
-\bibliography{register} %your bib file
-
-\end{document}
+% list of features todo
+% todo.. write this todo list

### 2018-01-10 thesis/landslide.tex
configure fonts

@@
 \inspirationalquote{Somewhere is the promise of an uncharted trail, with 700 branching limbs and 700 ways to fail.}
-{ThouShaltNot, ``Cardinal Directions''}
+{ThouShaltNot, "Cardinal Directions"}
 

### 2018-01-10 thesis/thesis.tex
add fig:tree
add incomplete intro
rearrange some stuff bw intro/bgnd
add landslide chapt
configure fonts

@@
 % some useful packages
-\usepackage[charter]{mathdesign}
 \usepackage{fullpage}
@@
 \usepackage{subfigure}
+\usepackage{xspace}
+
+%% custom fonts
+
+\usepackage{fontspec}
+\usepackage{xeCJK}
+
+\setmainfont{XCharter}
+%\setmainfont{Utopia}[
+%	Path = ./fonts/,
+%	UprightFont = UtopiaStd-Regular.otf,
+%	BoldFont = UtopiaStd-Semibold.otf,
+%	ItalicFont = UtopiaStd-Italic.otf,
+%	BoldItalicFont = UtopiaStd-SemiboldIt.otf
+%	]
+
+\setCJKmainfont{HeiseiMinStd-W5}[Path = ./fonts/]
 
@@
 
-Please hold on. This thesis will be departing shortly for the Landslide terminal, baggage claim, ground transportation, and ticketing.
+I thank my dog, Louie.
 \end{acknowledgments}
@@
 \tableofcontents
+% TODO: figure out how to make latex abbreviate these
 \listoffigures
 \listoftables
-
-\cleardoublepage
-\pagestyle{empty}
-\begin{center}
-\begin{tabular}{c}
-\vspace{12em} \\
-\includegraphics[width=0.55\textwidth]{how-i-learned.png}
-\end{tabular}
-\end{center}
 
@@
 \newcommand{\inspirationallinebreak}{\vspace{0.25em}}
+\newcommand{\inspirationalhyphen}{-\hspace{-0.15em}-\xspace}
 \newcommand\inspirationalquote[2]{\begin{flushright}
 	\vspace{-1em}
-	\fontfamily{pzc}\selectfont
+	{
+	\fontspec[Path=./fonts/]{AlexaStd}
 	{\em #1}
@@
 
-	---{#2}
-	\fontfamily{mdbch}\selectfont
+	-\hspace{-0.15em}-\hspace{-0.15em}-{#2}
+	}
 	\vspace{2em}
@@
 
-\chapter{Introduction}
-
+\input{intro}
 \input{background}
+\input{landslide}
 
 \chapter{Quicksand}
-\inspirationalquote{Somewhere is the promise of an uncharted trail, with 700 branching limbs and 700 ways to fail.}
-{ThouShaltNot, ``Cardinal Directions''}
 
@@
 \inspirationalquote{
-\begin{tabular}{p{0.6\textwidth}}
+\begin{tabular}{p{0.58\textwidth}}
 To test if your paper makes a genuine contribution to its discipline,
-see if you can afford a generous tone in the ``Related Work'' section.
+see if you can afford a generous tone in the "Related Work" section.
 \end{tabular}}
@@
 \inspirationalquote{
-%	``If you are a god, Zeus, as the stories claim, then why did you create evolution?
-%	Why did you make \\ a world that can only grow through cruelty and pain?''
-%	\inspirationallinebreak
-%
-%For a moment--surely this meant death was near--she thought she heard him answer:
-%``My child, I \\
-%was shaped by the gods that came before me, as you were shaped by me.
-%The choice I had was be- \\
-%tween creation and oblivion, life and death.
-%And I chose life, because any life is better than no life, \\
-%because as long as there is life, there is hope--if not for us, then for some generation to come.''
-%	\inspirationallinebreak
-%
-%``Then how are you a god, if you can offer so little?'' she whispered, feeling death creep closer.
-%	\inspirationallinebreak
-%
-%``I am a god because I take upon myself the burden of creation,'' the statue replied.
-%	\inspirationallinebreak
-%
-%``Then we are all gods,'' Alexandra said, and pushed the button.}
-	\begin{tabular}{p{0.875\textwidth}}
-	``If you are a god, Zeus, as the stories claim, then why did you create evolution?
-	Why did you make a world that can only grow through cruelty and pain?''
+	\begin{tabular}{p{0.84\textwidth}}
+	"If you are a god, Zeus, as the stories claim, then why did you create evolution?
+	Why did you make a world that can only grow through cruelty and pain?"
 	\\
-	For a moment--surely this meant death was near--she thought she heard him answer:
-	``My child, I was shaped by the gods that came before me, as you were shaped by me.
+	For a moment\inspirationalhyphen{}surely this meant death was near\inspirationalhyphen{}she thought she heard him answer:
+	"My child, I was shaped by the gods that came before me, as you were shaped by me.
 	The choice I had was between creation and oblivion, life and death.
 	And I chose life, because any life is better than no life,
-	because as long as there is life, there is hope--if not for us, then for some generation to come.''
+	because as long as there is life, there is hope\inspirationalhyphen{}if not for us, then for some generation to come."
 	\\
-	``Then how are you a god, if you can offer so little?'' she whispered, feeling death creep closer.
+	"Then how are you a god, if you can offer so little?" she whispered, feeling death creep closer.
 	\\
-	``I am a god because I take upon myself the burden of creation,'' the statue replied.
+	"I am a god because I take upon myself the burden of creation," the statue replied.
 	\\
-	``Then we are all gods,'' Alexandra said, and pushed the button.
+	"Then we are all gods," Alexandra said, and pushed the button.
 	\end{tabular}}

### 2018-01-11 thesis/landslide.tex
write skeleton for landlside chapter

@@
 
-% list of features todo
-% todo.. write this todo list
+Landslide is a model checker implemented as a plug-in module for x86 full-system simulators.
+The program to be tested runs in a simulated environment,
+and Landslide uses its access to the simulator's internal state to inspect and manipulate the memory and thread scheduling of the program as it executes.
+As of this thesis's writing, Landslide supports the use of two possible simulators:
+
+\begin{itemize}
+	\item {\bf Simics} \cite{simics}, a proprietary simulator licensed commercially by Wind River, used at CMU in 15-410 to run Pebbles thread libraries and kernels, and
+	\item {\bf Bochs} \cite{bochs}, an open-source (LGPL) simulator used at the University of Chicago, Berkeley, and other schools to run Pintos kernels.
+\end{itemize}
+
+% TODO: update this linkerino (repository name)
+% TODO: add some backup linkeroos
+The Bochs port of Landslide is likewise open-source and available at \url{https://github.com/bblum/bochs}.
+% TODO: refresh
+The HEAD commit at the time of writing is 8984021.
+
+This chapter will discuss Landslide's outer and inner workings for both aspiring users and aspiring developers in all gory detail. Strap in.
+
+\section{User interface}
+
+asdfadsf
+
+\section{Architecture}
+
+\subsection{Scheduler}
+\subsection{Memory analysis}
+\subsection{Machine state manipulation}
+\subsection{State space traversal}
+% time travel, explore, arbiter
+\subsection{Debugging output}
+% symtable stuff goes here
+% stack tracing goes here
+% foundabug goes here
+\subsection{Pebbles-specific features}
+\subsection{Pintos-specific features}
+\subsection{Handy scripts}
+
+\section{Algorithms of note}
+
+\subsection{Kernel annotations}
+\subsection{Bug identification}
+\subsection{Data race analysis}
+\subsection{Preemption point identification}
+\subsection{State space estimation}
+\subsection{Dynamic Partial Order Reduction}
+\subsection{Heuristic loop, deadlock, and synchronization detection}

### 2018-01-11 thesis/landslide.tex
write a lil about quicksand options

@@
 The HEAD commit at the time of writing is 8984021.
+The Simics port is available upon request.
 
@@
 
-asdfadsf
+This section describes the features of Landslide the average student user should expect to interact with.
+% TODO: make more specific section reference
+Separate user guides also exist, described in Chapter~\ref{chap:410}.
+
+\subsection{Setup}
+
+Two setup scripts are provided, one for each supported kernel architecture: {\tt p2-setup.sh} and {\tt pintos-setup.sh}.
+The user should supply the directory containing their project implementation.
+The latter script also supports arguments specifying which of the Pintos projects to target.
+For example:
+\begin{itemize}
+	\item {\tt ./p2-setup.sh /path/to/my/p2}
+	\item {\tt ./pintos-setup.sh /path/to/my/threads} (2nd argument defaults to ``{\tt threads}'')
+	\item {\tt ./pintos-setup.sh /path/to/my/userprog userprog}
+\end{itemize}
+
+These scripts accomplish the following setup tasks (among other trivialities):
+\begin{itemize}
+	\item Copy the user's code into {\tt pebsim/p2-basecode/} or {\tt pebsim/pintos/},
+		which contain a pre-annotated Pebbles reference kernel binary or pre-annotated Pintos basecode, respectively.
+	\item Build the code in its new location.
+	\item Run the instrumentation script on the resulting binary to let Landslide know where all the important functions are
+		(see \sect{\ref{sec:landslide-glue}}).
+\end{itemize}
+
+\subsection{Running Landslide through Quicksand}
+
+The preferred method of invoking Landslide is through Quicksand, the Iterative Deepening wrapper program which has all of Chapter~\ref{chap:quicksand} to itself.
+% TODO: add quicksand symlink
+This is done via the {\tt ./landslide} script in the top-level directory, which:
+\begin{itemize}
+	\item Checks if the user needs to run {\tt *-setup.sh} again, in case of their source code being more recently updated than the existing annotated build (a common mistake),
+	\item Passes its arguments through to {\tt id/landslide-id}, the Quicksand binary,
+		and
+	\item (If during the student user study,) compresses the resulting log files,
+		creates a snapshot tarball of them and the current version of the user's code,
+		and sends it to me for nefarious research purposes.
+\end{itemize}
+
+\subsubsection{Command-line argments}
+
+The following command line arguments are recommended for the average user.
+
+\begin{itemize}
+	\item {\tt -p PROGRAM}: the name of the test case to invoke
+	\item {\tt -t TIME}: wall-clock time limit, in seconds; or suffixed with one of {\tt ydhms} for years, days, hours, minutes, or seconds respectively (default 1h)
+	\item {\tt -c CPUS}: maximum number of Landslide instances to run in parallel (defaults to half the number of system CPUs)
+	\item {\tt -i INTERVAL}: interval of time between printing progress reports (default 10s)
+	\item {\tt -d TRACEDIR}: directory for resulting bug traces (default current directory)
+	\item {\tt -v}: verbose mode (issues output for each executed interleaving by each instance of landslide, makes progress reports more detailed, etc)
+	\item {\tt -l}: leave Landslide log files from completed state spaces even when no bug was found (deleted automatically by default)
+	\item {\tt -h}: print help text and exit immediately
+\end{itemize}
+
+The following ``secret'' arguments also exist, primarily for my own use in running experiments or debugging.
+
+\begin{itemize}
+	\item {\tt -C}: enable ``control experiment'' mode, i.e., run only 1 instance of Landslide, with all (non-data-race) preemption points enabled in advance
+		% TODO: put a section reference here
+	\item {\tt -I}: enable Iterative Context Bounding (requires {\tt -C}, although future work may relax this restriction);
+		this generally causes bugs to be found faster should they exist, but degrades completion time
+	\item {\tt -0}: enable Preempt-Everywhere mode (see \sect{\ref{sec:quicksand-eval}}, requires {\tt -C})
+	\item {\tt -H}: use Limited Happens-Before for data-race analysis (see \sect{\ref{sec:background-hb}})
+	\item {\tt -V}: use vector-clock-based Pure Happens-Before for data-race analysis (see \sect{\ref{sec:background-hb}})
+	\item {\tt -X}: support transactional memory (see Chapter~\ref{chap:tm})
+		% TODO: make a more specific section reference
+	\item {\tt -A}: support multiple abort codes during transaction failure (see Chapter~\ref{chap:tm});
+		required for testing programs which behave differently under different abort circumstances,
+		but impacts the state space size
+	\item {\tt -P}: support Pintos architecture (enabled automatically when {\tt pintos-setup.sh} is run)
+	\item {\tt -4}: support Pebbles architecture (enabled automatically when {\tt p2-setup.sh} is run)
+	\item {\tt -e ETAFACTOR}: configure heuristic state space ETA deferring factor (described in detail in {\tt id/option.c}
+	\item {\tt -E ETATHRESH}: configure heuristic threshold of state space progress for judging ETA stability (described in detail in {\tt id/option.c}
+\end{itemize}
+
+Quicksand will automatically generate configuration files and invoke Landslide according to the process described in the next section.
+
+\subsection{Running Landslide directly}
+
+Rather than letting Quicksand juggle multiple instances of Landslide,
+the user may run a single instance directly, optionally configuring the preemption points by hand.
+This is recommended only for enthusiastic users who are annotating their own kernel.
+
+The script {\tt pebsim/landslide} invokes Landslide thus.
+It should be run from within the {\tt pebsim/} directory.
+When supplied no arguments, it reads configuration options from {\tt pebsim/config.landslide}
+(a bash script expected to define certain variables as described in \sect{\ref{sec:landslide-glue}}).
+The user may optionally specify a file containing additional config directives
+% , such as custom preemption points,
+as an argument.
+Such supported options are as follows.
+
+% TODO
+
+\subsection{Bug reports}
 
@@
 \subsection{Handy scripts}
+\label{sec:landslide-glue}
 

### 2018-01-16 thesis/landslide.tex
dyn config opts

@@
 \begin{itemize}
-	\item Checks if the user needs to run {\tt *-setup.sh} again, in case of their source code being more recently updated than the existing annotated build (a common mistake),
+	\item Checks if the user needs to run {\tt *-setup.sh} again, in case their source code was more recently updated than the existing annotated build (a common mistake),
 	\item Passes its arguments through to {\tt id/landslide-id}, the Quicksand binary,
@@
 	\item {\tt -4}: support Pebbles architecture (enabled automatically when {\tt p2-setup.sh} is run)
-	\item {\tt -e ETAFACTOR}: configure heuristic state space ETA deferring factor (described in detail in {\tt id/option.c}
-	\item {\tt -E ETATHRESH}: configure heuristic threshold of state space progress for judging ETA stability (described in detail in {\tt id/option.c}
+	\item {\tt -e ETAFACTOR}: configure heuristic state space ETA deferring factor (described in detail in {\tt id/option.c})
+	\item {\tt -E ETATHRESH}: configure heuristic threshold of state space progress for judging ETA stability (described in detail in {\tt id/option.c})
 \end{itemize}
@@
 % , such as custom preemption points,
-as an argument.
+as an argument\footnote{
+Quicksand actually supplies two such files as arguments: one ``static'' config file and one ''dynamic'' config file.
+The former contains options which require recompiling Landslide (e.g., whether or not to use ICB is controlled by an {\tt \#ifdef} in Landslide's code),
+while the latter contains options which Landslide interprets at runtime (e.g., which preemption points to use).
+The static options do not change between Landslide instances in a single Quicksand run,
+avoiding long Landslide start-up times.
+	}.
 Such supported options are as follows.
 
-% TODO
+\subsubsection{Dynamic configuration options}
+First, the following options may be changed without triggering a recompile of Landslide.
+
+\begin{itemize}
+	\item {\tt within\_function} % TODO
+	\item {\tt within\_user\_function} - as above but finds the function in the userspace test program rather than the kernel code.
+	\item {\tt without\_function} % TODO
+	\item {\tt without\_user\_function} - difference to one above same as stated two above.
+	\item {\tt data\_race ADDR TID LAST\_CALL CURRENT\_SYSCALL} - specifies a data-race preemption point.
+		\begin{itemize}
+			\item {\tt ADDR} shall be the code address (in hex) of the racing address,
+			{\em before} the execution of which a preemption will be issued.
+			\item {\tt TID} indicates a thread ID required to be running for this data race.
+				To specify data race PPs across all threads at once, set {\tt FILTER\_DRS\_BY\_TID=0} (see next section).
+			\item {\tt LAST\_CALL} indicates a code address required to be the site of the last {\tt call} instruction executed
+				(similar to specifying a stack trace, but using a full stack trace here degrades performance too much),
+				or 0 to not use this feature.
+				From personal experience I found this option rather useless and recommend always supplying 0.
+				% TODO: fix this section ref
+				For further discussion see \sect{\ref{sec:quicksand-pps}}.
+			\item {\tt CURRENT\_SYSCALL} indicates the system call number if a user-space data race comes from within a kernel system call which accesses user memory (Pebbles only).
+				Usually 0 (i.e., not in kernel code) but {\tt deschedule}'s system call number is common as well.
+		\end{itemize}
+	\item {\tt input\_pipe FILENAME} - FIFO file used for receiving messages from Quicksand (e.g. to suspend or resume execution).
+		Requires {\tt id\_magic} option to be set (next section below).
+		The odds a human user will find success using this option by hand are infinitesimal.
+	\item {\tt output\_pipe FILENAME} - as above but for sending messages.
+\end{itemize}
+
+\subsubsection{Static configuration options}
+
+Next, configuration options which affect an {\tt \#ifdef} in Landslide and will trigger a recompile upon changing.
+Unless otherwise specified these are boolean flags (1 or 0) and the example value shown indicates the default used if unspecified.
+
+\begin{itemize}
+	\item {\tt FILTER\_DRS\_BY\_TID=1} - configures whether to use the {\tt TID} parameter of {\tt data\_race} described above.
+	\item % TODO moar
+\end{itemize}
 

### 2018-01-18 thesis/landslide.tex
polish and add some chapter quotes

@@
 
-The following command line arguments are recommended for the average user.
+The following command line arguments are recommended for the common user.
 
@@
 	\item {\tt -H}: use Limited Happens-Before for data-race analysis (see \sect{\ref{sec:background-hb}})
+		(default for Pebbles kernelspace mode)
 	\item {\tt -V}: use vector-clock-based Pure Happens-Before for data-race analysis (see \sect{\ref{sec:background-hb}})
+		(default for P2 userspace and Pintos modes)
 	\item {\tt -X}: support transactional memory (see Chapter~\ref{chap:tm})

### 2018-01-18 thesis/landslide.tex
improve dpor bgnd; add some static landslide configs

@@
 \subsection{Running Landslide through Quicksand}
+\label{sec:landslide-quicksand-options}
 
@@
 		this generally causes bugs to be found faster should they exist, but degrades completion time
-	\item {\tt -0}: enable Preempt-Everywhere mode (see \sect{\ref{sec:quicksand-eval}}, requires {\tt -C})
-	\item {\tt -H}: use Limited Happens-Before for data-race analysis (see \sect{\ref{sec:background-hb}})
+		(\sect{\ref{sec:landslide-icb}})
+	\item {\tt -0}: enable Preempt-Everywhere mode (\sect{\ref{sec:quicksand-eval}}, requires {\tt -C})
+	\item {\tt -H}: use Limited Happens-Before for data-race analysis (\sect{\ref{sec:background-hb}})
 		(default for Pebbles kernelspace mode)
-	\item {\tt -V}: use vector-clock-based Pure Happens-Before for data-race analysis (see \sect{\ref{sec:background-hb}})
+	\item {\tt -V}: use vector-clock-based Pure Happens-Before for data-race analysis (\sect{\ref{sec:background-hb}})
 		(default for P2 userspace and Pintos modes)
-	\item {\tt -X}: support transactional memory (see Chapter~\ref{chap:tm})
+	\item {\tt -X}: support transactional memory (Chapter~\ref{chap:tm})
 		% TODO: make a more specific section reference
-	\item {\tt -A}: support multiple abort codes during transaction failure (see Chapter~\ref{chap:tm});
+	\item {\tt -A}: support multiple abort codes during transaction failure (Chapter~\ref{chap:tm});
 		required for testing programs which behave differently under different abort circumstances,
@@
 First, the following options may be changed without triggering a recompile of Landslide.
+They are implemented as bash functions defined in {\tt pebsim/build.sh}.
 
 \begin{itemize}
-	\item {\tt within\_function} % TODO
-	\item {\tt within\_user\_function} - as above but finds the function in the userspace test program rather than the kernel code.
-	\item {\tt without\_function} % TODO
-	\item {\tt without\_user\_function} - difference to one above same as stated two above.
+	\item {\tt within\_function FUNC} - adds {\tt FUNC} to a whitelist of functions required to appear in the current stack trace before identifying a preemption point (see \sect{\ref{sec:landslide-pps}})
+	\item {\tt without\_function FUNC} - as above, but a blacklist instead of a whitelist
+	\item {\tt within\_user\_function FUNC} - as two above but finds the function in the userspace test program rather than the kernel code.
+	\item {\tt without\_user\_function FUNC} - difference to two above same as stated one above.
 	\item {\tt data\_race ADDR TID LAST\_CALL CURRENT\_SYSCALL} - specifies a data-race preemption point.
@@
 		Requires {\tt id\_magic} option to be set (next section below).
-		The odds a human user will find success using this option by hand are infinitesimal.
+		The odds that a human user will find spiritual enlightenment through using this option by hand are infinitesimal.
 	\item {\tt output\_pipe FILENAME} - as above but for sending messages.
@@
 \begin{itemize}
+	\item {\tt TEST\_CASE} - configure the name of the test program to run (mandatory; no default)
+	\item {\tt VERBOSE=0} - enable more verbose output
 	\item {\tt FILTER\_DRS\_BY\_TID=1} - configures whether to use the {\tt TID} parameter of {\tt data\_race} described above.
+	\item {\tt ICB=0} - enable Iterative Context Bounding (\sect{\ref{sec:landslide-icb}});
+		corresponds to {\tt -I} in \sect{\ref{sec:landslide-quicksand-options}}.
+	\item {\tt PREEMPT\_EVERYWHERE=0} - enable Preempt-Everywhere mode (\sect{\ref{sec:quicksand-eval}});
+		corresponds to {\tt -0} in \sect{\ref{sec:landslide-quicksand-options}}.
+	\item {\tt PURE\_HAPPENS\_BEFORE=1} - select Pure Happens-Before (1) or Limited Happens-Before (2) (\sect{\ref{sec:background-hb}});
+		corresponds to {\tt -V}/{\tt -H} in \sect{\ref{sec:landslide-quicksand-options}}.
 	\item % TODO moar
@@
 % time travel, explore, arbiter
-\subsection{Debugging output}
+\subsection{Bug-finding output}
 % symtable stuff goes here
@@
 \subsection{Preemption point identification}
+\label{sec:landslide-pps}
+% TODO: talk about how w/in func and w/out func are processed
 \subsection{State space estimation}
 \subsection{Dynamic Partial Order Reduction}
+\label{sec:landslide-dpor}
+\subsection{Iterative Context Bounding}
+\label{sec:landslide-icb}
 \subsection{Heuristic loop, deadlock, and synchronization detection}

### 2018-01-18 thesis/thesis.tex
add charter font

@@
 \usepackage{amsmath}
+\usepackage{amssymb}
 \usepackage[numbers,sort]{natbib}
@@
 
-\setmainfont{XCharter}
+\setmainfont{Charter}[
+	Path = ./fonts/,
+	UprightFont = XCharter-Roman.otf,
+	BoldFont = XCharter-Bold.otf,
+	ItalicFont = XCharter-Italic.otf,
+	BoldItalicFont = XCharter-BoldItalic.otf
+	]
 %\setmainfont{Utopia}[
@@
 \normalsize \vspace{1em}
-{\em or, How I Learned to Stop Worrying and Love the Exponential Explosion}}
+\begin{tabular}{c}
+{\em or, How I Learned to Stop Worrying and Love the Exponential Explosion}
+\end{tabular}}
 \author{Ben Blum}
@@
 % new frenz
+% 410 ta community
+% emarhavil
 %
@@
 \chapter{Quicksand}
+%\inspirationalquote{
+%\begin{tabular}{p{0.55\textwidth}}
+%There are awful, horrible things in this world; I know that now.
+%But there are a lot of things worth protecting, too.
+%\end{tabular}}
+%{Kaname Madoka, Mahou Shoujo Madoka{\raisebox{0.1em}{$\scriptstyle \bigstar$}}Magica}
+\inspirationalquote{Always, somewhere, someone is fighting for you.
+As long as you remember her, you are not alone.}
+{Mahou Shoujo Madoka{\raisebox{0.1em}{$\scriptstyle \bigstar$}}Magica}
 
-\chapter{410} % rename
+\chapter{Education}
 \inspirationalquote{Knowing the students might one day
@@
 \end{tabular}}
-{Conor McBride, via Twitter}
+{Conor McBride}
+
+\chapter{Future Work}
+% TODO: ask for permission for this one
+% https://twitter.com/Poem4your_sprog/status/878747092788838400
+\inspirationalquote{
+\begin{tabular}{p{0.54\textwidth}}
+A little bit of being kind,
+a tiny open door; \\
+A nicer slice of peace of mind
+begets a little more. \\
+It won't correct the world tonight,
+nor change tomorrow too; \\
+But maybe if you do it right,
+you'll find it changes *you*.
+\end{tabular}}
+{Sam "Poem\_for\_your\_sprog" Garland}
 
@@
 	\end{tabular}}
-{The Talos Principle: Road to Gehenna}
+{Galatea, The Talos Principle}
 
@@
 \renewcommand{\bibsection}{\chapter{\bibname}
-\inspirationalquote{Only a doctor of philosophy, Darth.}{Robert Marsh, via Twitter}
+\inspirationalquote{Only a doctor of philosophy, Darth.}{Robert Marsh}
 }

### 2018-01-23 thesis/intro.tex
configure fonts
polish and add some chapter quotes
finish writing landslide static config opts

@@
 \inspirationalquote{
-\begin{tabular}{p{0.65\textwidth}}
+\begin{tabular}{p{0.7\textwidth}}
 Please hold on. This thesis will be departing shortly for the Landslide terminal, baggage claim, ground transportation, and ticketing.
@@
 	\item {\bf Quicksand:} Chapter~\ref{chap:quicksand} presents the Iterative Deepening framework which more intelligently chooses which state spaces to test, corresponding to contribution 1 above.
-	\item {\bf Pebbles:} Chapter~\ref{chap:410} discusses my evaluation of Landslide in CMU's 15-410 class environment using the Pebbles kernel, corresponding to contribution 2 above.
-	\item {\bf Pintos:} Chapter~\ref{chap:410} discusses my evaluation of Landslide in the University of Chicago's and Berkeley's OS class environments using the Pintos kernel, corresponding to contribution 2 above.
+	\item {\bf Education:} Chapter~\ref{chap:education} discusses my evaluation of Landslide
+		in CMU's 15-410 class environment using the Pebbles kernel,
+		and in the University of Chicago's and Berkeley's OS class environments using the Pintos kernel,
+		corresponding to contribution 2 above.
+	%\item {\bf Pebbles:} Chapter~\ref{chap:410} discusses my evaluation of Landslide in CMU's 15-410 class environment using the Pebbles kernel, corresponding to contribution 2 above.
+	%\item {\bf Pintos:} Chapter~\ref{chap:410} discusses my evaluation of Landslide in the University of Chicago's and Berkeley's OS class environments using the Pintos kernel, corresponding to contribution 2 above.
 	\item {\bf Transactional Memory:} Chapter~\ref{chap:tm} presents my extension of Landslide's concurrency model to handle transactional concurrency and the evaluation thereof, corresponding to contribution 3 above.

### 2018-01-23 thesis/landslide.tex
finish writing landslide static config opts

@@
 	\item {\tt -4}: support Pebbles architecture (enabled automatically when {\tt p2-setup.sh} is run)
+		% TODO: put a section reference
 	\item {\tt -e ETAFACTOR}: configure heuristic state space ETA deferring factor (described in detail in {\tt id/option.c})
@@
 % , such as custom preemption points,
-as an argument\footnote{
+as an argument.\footnote{
 Quicksand actually supplies two such files as arguments: one ``static'' config file and one ''dynamic'' config file.
@@
 avoiding long Landslide start-up times.
-	}.
+}
 Such supported options are as follows.
@@
 Next, configuration options which affect an {\tt \#ifdef} in Landslide and will trigger a recompile upon changing.
+%These span a wide variety of features, sorted below in subsections by roughly how interesting I think they are.
 Unless otherwise specified these are boolean flags (1 or 0) and the example value shown indicates the default used if unspecified.
 
+{\bf Search algorithm options}
 \begin{itemize}
-	\item {\tt TEST\_CASE} - configure the name of the test program to run (mandatory; no default)
-	\item {\tt VERBOSE=0} - enable more verbose output
-	\item {\tt FILTER\_DRS\_BY\_TID=1} - configures whether to use the {\tt TID} parameter of {\tt data\_race} described above.
 	\item {\tt ICB=0} - enable Iterative Context Bounding (\sect{\ref{sec:landslide-icb}});
@@
 		corresponds to {\tt -0} in \sect{\ref{sec:landslide-quicksand-options}}.
+	\item {\tt EXPLORE\_BACKWARDS=0} - configure whether, at each newly encountered preemption point,
+		to allow the current thread to run first then later upon backtracking to preempt (0),
+		or to issue preemptions first and then try continuing the current thread later (1).
+		0 tends to produce shorter preemption traces while 1 tends to find bugs faster (\cite{landslide} \S{}8.7.1).
+		Not compatible with ICB.
+\end{itemize}
+
+{\bf Memory analysis options}
+\begin{itemize}
 	\item {\tt PURE\_HAPPENS\_BEFORE=1} - select Pure Happens-Before (1) or Limited Happens-Before (2) (\sect{\ref{sec:background-hb}});
 		corresponds to {\tt -V}/{\tt -H} in \sect{\ref{sec:landslide-quicksand-options}}.
-	\item % TODO moar
+	\item {\tt FILTER\_DRS\_BY\_TID=1} - configures whether to use the {\tt TID} parameter of {\tt data\_race} described above.
+	\item {\tt FILTER\_DRS\_BY\_LAST\_CALL=0} - configures whether to use the {\tt LAST\_CALL} parameter of {\tt data\_race} described above.
+	\item {\tt ALLOW\_LOCK\_HANDOFF=0} - configures lockset tracking to permit or disallow a lock taken by one thread to be released by another thread.%
+		\footnote{If enabled, accesses performed by the second thread before unlocking will not be considered protected by that lock,
+			as Landslide cannot infer what prior event abstractly represented the lock's ownership changing,
+			leading to spurious data race reports.
+			This could be solved in future work with a new annotation.}
+	\item {\tt ALLOW\_REENTRANT\_MALLOC\_FREE=0} - allow two threads to be in {\tt malloc}, {\tt free}, or so on simultaneously without declaring it a bug.%
+		\footnote{Used in Pintos, where those functions lock/unlock the heap mutex themselves rather than relying on a wrapper function to do so before invoking them.}
 \end{itemize}
+
+{\bf Interface options}
+\begin{itemize}
+	\item {\tt TEST\_CASE} - configure the name of the test program to run (mandatory; no default)
+	\item {\tt VERBOSE=0} - enable more verbose output
+	\item {\tt BREAK\_ON\_BUG=0} - configure whether to exit the simulator or drop into a debug prompt when a bug is found. Simics only and not compatible with Quicksand.
+	\item {\tt DONT\_EXPLORE=0} - if enabled, Landslide will not perform stateless model checking but rather will execute the default thread interleaving then exit (useful for manual inspection of preemption points).
+	\item {\tt PRINT\_DATA\_RACES=0} - as it says on the tin (for stand-alone use; will message them to Quicksand regardless).
+	\item {\tt TABULAR\_TRACE=1} - configure whether to emit bug reports to the console (0) or to an HTML trace file (1)
+\end{itemize}
+
+\subsection{Kernel annotations}
+
+The educational experiments in this thesis focus on projects which students implement on top of provided kernel basecode which Landslide already ``understands''.
+Such understanding is conferred via the annotations described in this section.
+For P2 and Pintos students I supply these annotations behind the scenes,
+but a CMU 15-410 student who wishes to use Landslide on their kernel project shall need to brave forth hereupon.
+
+\subsubsection{config.landslide annotations}
+
+The following annotations are specified in {\tt pebsim/config.landslide} akin to the static configuration options described above.
+These specify the names of kernel functions, global variables, default values, and so on
+which are required to accurately track the kernel's scheduler state:
+{\tt CONTEXT\_SWITCH},
+{\tt EXEC},
+{\tt FIRST\_TID},
+{\tt IDLE\_TID},
+{\tt INIT\_TID},
+{\tt MEMSET},
+{\tt PAGE\_FAULT\_WRAPPER},
+{\tt PREEMPT\_ENABLE\_FLAG},
+{\tt PREEMPT\_ENABLE\_VALUE},
+{\tt READLINE},
+{\tt SFREE},
+{\tt SHELL\_TID},
+{\tt SPURIOUS\_INTERRUPT\_WRAPPER},
+{\tt THREAD\_KILLED\_ARG\_VAL},
+{\tt THREAD\_KILLED\_FUNC},
+{\tt TIMER\_WRAPPER},
+{\tt VM\_USER\_COPY},
+{\tt VM\_USER\_COPY\_TAIL},
+{\tt YIELD}.
+
+The following less self-explanatory secret options also exist:
+\begin{itemize}
+	\item {\tt PINTOS\_KERNEL=0} - configure Landslide for Pebbles (0) or Pintos (1) kernel architecture. Normally set automatically by the setup scripts.
+	\item {\tt TESTING\_USERSPACE=1} - configure Landslide whether to test (i.e., focus preemption points, memory analysis, etc. on) the userspace or kernelspace code.
+	\item {\tt CURRENT\_THREAD\_LIVES\_ON\_RQ=0} - Landslide infers the list of runnable threads from the {\tt on\_rq()} and {\tt off\_rq()} annotations (described below).
+		Some kernels\footnote{most, actually} remove the current thread from their runqueue,
+		such that the abstract set of all runnable threads is actually the runqueue plus the current thread rather than just the runqueue.
+		Other kernels\footnote{the author's own student kernel from long ago}
+		leave the current thread on the runqueue,
+		removing it only when it's descheduling and should actually be considered blocked.
+		Set this option to 0 to support the former kernel type or 1 to support the latter.
+	\item {\tt PATHOS\_SYSCALL\_IRET\_DISTANCE} - indicate how much stack space is used by the reference kernel's system call wrappers (used for cross-kernel-to-userspace stack traces)
+	\item {\tt PDE\_PTE\_POISON} - indicate a poison value used in the page tables to indicate absent VM mappings
+	\item {\tt BUG\_ON\_THREADS\_WEDGED=1} - set to 0 to disable deadlock detection but instead let the kernel keep receiving system interrupts when all threads appear blocked.%
+		\footnote{once used in the bad old days; now recommended for debugging use only}
+	\item {\tt TIMER\_WRAPPER\_DISPATCH} - used to manually indicate a label before the end of the timer interrupt assembly wrapper, in case the {\tt iret} instruction couldn't be found automatically (see {\tt pebsim/definegen.sh}).
+\end{itemize}
+
+\subsubsection{In-kernel code annotations}
+
+% TODO tell_landslide
 

### 2018-01-27 thesis/landslide.tex
finish writing abt kernel annos

@@
 		\footnote{Used in Pintos, where those functions lock/unlock the heap mutex themselves rather than relying on a wrapper function to do so before invoking them.}
+	\item {\tt TESTING\_MUTEXES=0} - configure ``mutex testing'' mode (1),
+		in which the data race analysis will not consider a mutex's implementation to be protected by the mutex itself.
+		In other words, the mutex's internal memory accesses will be flagged as data races,
+		thereby enabling Landslide to verify the mutual exclusion property.
+		Normally (0), Landslide assumes mutual exclusion is provided in order to efficiently find data races in the rest of the code.
+		Quicksand will automatically set this option for P2s when {\tt -t mutex\_test} is specified.
 \end{itemize}
@@
 \begin{itemize}
-	\item {\tt TEST\_CASE} - configure the name of the test program to run (mandatory; no default)
+	\item {\tt TEST\_CASE=NAME} - configure the name of the test program to run (mandatory; no default)
 	\item {\tt VERBOSE=0} - enable more verbose output
@@
 
+% TODO: mb this should be a section, not a subsection, bc there are too many subsubsubsections herein?
 \subsection{Kernel annotations}
@@
 {\tt PAGE\_FAULT\_WRAPPER},
-{\tt PREEMPT\_ENABLE\_FLAG},
-{\tt PREEMPT\_ENABLE\_VALUE},
 {\tt READLINE},
@@
 {\tt SPURIOUS\_INTERRUPT\_WRAPPER},
+\\
 {\tt THREAD\_KILLED\_ARG\_VAL},
@@
 {\tt VM\_USER\_COPY},
+\\
 {\tt VM\_USER\_COPY\_TAIL},
@@
 
-The following less self-explanatory secret options also exist:
+Following are the non-self-explanatory options.
 \begin{itemize}
@@
 	\item {\tt TESTING\_USERSPACE=1} - configure Landslide whether to test (i.e., focus preemption points, memory analysis, etc. on) the userspace or kernelspace code.
-	\item {\tt CURRENT\_THREAD\_LIVES\_ON\_RQ=0} - Landslide infers the list of runnable threads from the {\tt on\_rq()} and {\tt off\_rq()} annotations (described below).
+	\item {\tt CURRENT\_THREAD\_LIVES\_ON\_RQ=0} - Landslide infers the list of runnable threads from the {\tt tell\_landslide\_on\_rq()} and {\tt off\_rq()} annotations (described below).
 		Some kernels\footnote{most, actually} remove the current thread from their runqueue,
@@
 		removing it only when it's descheduling and should actually be considered blocked.
-		Set this option to 0 to support the former kernel type or 1 to support the latter.
-	\item {\tt PATHOS\_SYSCALL\_IRET\_DISTANCE} - indicate how much stack space is used by the reference kernel's system call wrappers (used for cross-kernel-to-userspace stack traces)
-	\item {\tt PDE\_PTE\_POISON} - indicate a poison value used in the page tables to indicate absent VM mappings
+		Set this option to 0 to support the former kernel type or 1 to support the latter.%
+		\footnote{This option replaces the deprecated {\tt kern\_current\_extra\_runnable()} annotation from {\tt student.c} described in \cite{landslide} \S{}6.2.3.}
+	\item {\tt PREEMPT\_ENABLE\_FLAG=NAME} - name of a global variable which the kernel uses to toggle scheduler preemptability, for kernels which may disable preemption without disabling interrupts.
+		For kernels wherein preemptability is corresponds directly by interrupts, leave this option unspecified.
+	\item {\tt PREEMPT\_ENABLE\_VALUE=VAL} - value of the above variable when preemption is enabled
+		(usually 0; note that many kernels use a nesting depth counter where any positive value corresponds to disabled).%
+		\footnote{These two options replace the deprecated {\tt kern\_ready\_for\_timer\_interrupt()} annotation from {\tt student.c} described in \cite{landslide} \${}6.2.3.}
+	\item {\tt PATHOS\_SYSCALL\_IRET\_DISTANCE=VALUE} - indicate how much stack space is used by the reference kernel's system call wrappers.
+		Used for cross-kernel-to-userspace stack traces;
+		if unset, stack traces from kernel space will end at the system call boundary.
+	\item {\tt PDE\_PTE\_POISON=VALUE} - indicate a poison value used in the page tables to indicate absent VM mappings to check for as well as checking the present bit (if unspecified, will check present bit only)
 	\item {\tt BUG\_ON\_THREADS\_WEDGED=1} - set to 0 to disable deadlock detection but instead let the kernel keep receiving system interrupts when all threads appear blocked.%
 		\footnote{once used in the bad old days; now recommended for debugging use only}
-	\item {\tt TIMER\_WRAPPER\_DISPATCH} - used to manually indicate a label before the end of the timer interrupt assembly wrapper, in case the {\tt iret} instruction couldn't be found automatically (see {\tt pebsim/definegen.sh}).
+	\item {\tt TIMER\_WRAPPER\_DISPATCH=NAME} - used to manually indicate a label before the end of the timer interrupt assembly wrapper, in case the {\tt iret} instruction couldn't be found automatically (see {\tt pebsim/definegen.sh}).
+	\item {\tt starting\_threads TID STARTS\_ON\_RQ} - specifies a system thread which already exists at the time {\tt tell\_landslide\_sched\_init\_done()} (see below) is called; {\tt TID} is the thread's ID and {\tt STARTS\_ON\_RQ} is 0 or 1 to indicate whether or not it starts on the system runqueue.
+		Typical threads to use this for are init and idle.
+	\item {\tt ignore\_sym NAME SIZE} - specifies a global variable {\tt NAME} of a given {\tt SIZE} in bytes whose memory accesses should be ignored for the purposes of DPOR and data race analysis.
+		Typical symbols to use this for are the console or heap mutex.
+	\item {\tt sched\_func NAME} - specifies a function whose memory accesses should all be ignored for the purposes of DPOR and data race analysis.
+		Typical functions to use this for are the timer handler and context switcher.
+	\item {\tt disk\_io\_func NAME} - specifies a function which may block a thread waiting for disk I/O (or other external interrupt) rather than blocking on another thread.
+		If any threads are blocked in a disk I/O function during an apparent deadlock,
+		Landslide will allow the kernel to idle until the simulator delivers the appropriate interrupt,
+		rather than declaring a bug.
+	\item {\tt ignore\_dr\_function NAME USERSPACE} - specifies a function whose memory access should not be counted as data races (but still be considered memory conflicts for DPOR).
+		{\tt USERSPACE} should be 0 or 1 to denote a kernel-space or user-space function respectively.
 \end{itemize}
@@
 
-% TODO tell_landslide
+The following annotations are provided as C functions
+which a kernel author shall include in their source code and call at appropriate times.
+The functions' actual implementations are empty;
+rather they serve as labels whose positions the annotation scripts extract
+along with the other various annotations from the previous section.
+Some of these are mandatory for Landslide to function properly,
+while others serve to improve or otherwise manipulate the state space.
+
+{\bf Mandatory annotations}
+
+\begin{itemize}
+\item {\tt tell\_landslide\_thread\_switch(int new\_tid)} - to be called during context switch, indicating the newly-running thread
+	(must be called with interrupts and/or scheduler preemption disabled)
+\item {\tt tell\_landslide\_sched\_init\_done()} - to be called after scheduler initialization,
+	indicating the point after which Landslide should begin analysis.
+	Any threads already initialized before this point (init, idle, etc) should be specified with {\tt starting\_threads} (previous subsection).
+\item {\tt tell\_landslide\_forking()} - to be called whenever a new thread is created,
+	``immediately'' before the next {\tt thread\_switch()} or {\tt on\_rq()} call for that new thread (i.e., this call sets a flag which the next instance of either of the latter will check to see if the indicated thread is new).
+	Most Pebbles kernels will call this twice; once in {\tt fork} and once in {\tt thread\_fork}.
+\item {\tt tell\_landslide\_vanishing()} - to be called whenever a thread ceases to exist,
+	``immediately'' before the next {\tt thread\_switch()} or {\tt off\_rq()} call for the exiting thread
+	(works similarly to above).
+\item {\tt tell\_landslide\_sleeping()} - to be called whenever a thread is about to {\tt sleep()} waiting for timer interrupts,
+	``immediately'' before the next {\tt thread\_switch()} or {\tt off\_rq()} call for the sleeping thread
+	(similar to the above).
+	Landslide considers sleeping threads to be runnable as normal (they will just take more timer interrupts to arrive at),
+	so this call is necessary to distinguish from the case when a thread is descheduled on a non-timer event.
+\item {\tt tell\_landslide\_thread\_on\_rq(int tid)} - to be called when a thread is added to the runqueue
+	(must be called with interrupts and/or scheduler preemption disabled).
+\item {\tt tell\_landslide\_thread\_off\_rq(int tid)} - dual of the above.
+	If {\tt CURRENT\_THREAD\_\allowbreak{}LIVES\_ON\_RQ=0} (described above), this should be invoked (among other times) during context switch with the TID of the thread about to start running.
+	Alternatively (thanks sully), even for a kernel which takes the current thread off its literal runqueue,
+	the annotator may use these two calls to indicate the ``abstract runqueue'' which includes the current thread as well,
+	and set {\tt CURRENT\_THREAD\_LIVES\_ON\_RQ=1}.
+\end{itemize}
+
+{\bf Optional annotations}
+
+\begin{itemize}
+\item {\tt tell\_landslide\_preempt()} - specifies a preemption point.
+	Subject to the constraints of {\tt within\_function}/{\tt without\_function};
+	hence may be ignored if used with Quicksand.
+\item {\tt tell\_landslide\_dump\_stack()} - instructs Landslide to print a stack trace whenever this point is reached (for debugging purposes).
+\end{itemize}
+
+{\bf Optional but strongly recommended annotations}
+
+The following annotations enable Landslide to track locksets for data race analysis.
+If not provided, it will be as if Landslide assumes no guarantees about mutual exclusion or happens-before,
+and hence will identify all memory conflicts as data races.
+
+\begin{itemize}
+\item {\tt tell\_landslide\_mutex\_locking(void *mutex\_addr)} - indicates the beginning of the lock routine for
+	whatever synchronization API Landslide should treat as the primitive for data race detection.
+	In Pintos this is the {\tt sema\_*()} function family; in Pebbles they may be called anything.
+\item {\tt tell\_landslide\_mutex\_blocking(int owner\_tid)} - called ``immediately'' before a thread becomes blocked on the mutex.
+	Definition of ``immediately'' similar to the {\tt forking()} and friends annotations above.
+	{\tt owner\_tid} allows Landslide to efficiently unblock/re-block threads when the mutex holder changes
+	(rather than relying on heuristic yield-loop detection);
+	see {\tt kern\_mutex\_block\_others()} and {\tt deadlocked()} in {\tt schedule.c} for implementation details.
+\item {\tt tell\_landslide\_mutex\_locking\_done(void *mutex\_addr)} - indicates the end of the lock routine.
+\item {\tt tell\_landslide\_mutex\_trylocking(void *mutex\_addr)} - indicates the beginning of the trylock routine (if present).
+\item {\tt tell\_landslide\_mutex\_trylocking\_done(void *mutex\_addr, int succeeded)} -
+	indicates when a thread is finished trylocking, even if it failed to get the lock (indicated by {\tt succeeded}).
+\item {\tt tell\_landslide\_mutex\_unlocking(void *mutex\_addr)} - indicates the beginning of the unlock routine.
+\item {\tt tell\_landslide\_mutex\_unlocking\_done()} - indicates the end of the unlock routine.
+\end{itemize}
 

### 2018-01-27 thesis/landslide.tex
clean up landslide subsection organization

@@
 
-This chapter will discuss Landslide's outer and inner workings for both aspiring users and aspiring developers in all gory detail. Strap in.
+This chapter will discuss Landslide's outer and inner workings in all their gory detail.
+It is intended for the aspiring developer or the ambitious user
+and hence is written in the style of documentation rather than as a report of research results.
 
@@
 \subsection{Running Landslide directly}
+\label{sec:landslide-directly}
 
@@
 
-{\bf Search algorithm options}
+\begin{itemize}
+\item {\bf Search algorithm options}
 \begin{itemize}
@@
 
-{\bf Memory analysis options}
+\item {\bf Memory analysis options}
 \begin{itemize}
@@
 
-{\bf Interface options}
+\item {\bf Interface options}
 \begin{itemize}
@@
 \end{itemize}
+\end{itemize}
 
-% TODO: mb this should be a section, not a subsection, bc there are too many subsubsubsections herein?
-\subsection{Kernel annotations}
+\subsection{Bug reports}
+
+\section{Kernel annotations}
 
@@
 
-\subsubsection{config.landslide annotations}
+\subsection{config.landslide annotations}
 
-The following annotations are specified in {\tt pebsim/config.landslide} akin to the static configuration options described above.
+The following annotations are specified in {\tt pebsim/config.landslide} akin to the static configuration options
+described in \sect{\ref{sec:landslide-directly}}..
 These specify the names of kernel functions, global variables, default values, and so on
@@
 
-Following are the non-self-explanatory options.
+Following are the less self-explanatory options.
 \begin{itemize}
@@
 
-\subsubsection{In-kernel code annotations}
+\subsection{In-kernel code annotations}
 
@@
 
-{\bf Mandatory annotations}
+\subsubsection{Mandatory annotations}
 
@@
 
-{\bf Optional annotations}
+\subsubsection{Optional annotations}
 
@@
 
-{\bf Optional but strongly recommended annotations}
+\subsubsection{Optional but strongly recommended annotations}
 
@@
 and hence will identify all memory conflicts as data races.
+(Note that the corresponding instrumentation for P2s is achieved automatically,
+as the names of the mutex interface are mandated by the project specification.)
 
@@
 \end{itemize}
-
-\subsection{Bug reports}
 

### 2018-01-28 thesis/landslide.tex
put a example bug report

@@
 It is intended for the aspiring developer or the ambitious user
-and hence is written in the style of documentation rather than as a report of research results.
+and hence unlike other chapters is written in the style of documentation rather than as a report of research results.
 
@@
 
+When Landslide finds a bug, it produces an execution trace of the particular interleaving of threads that led to the bug,
+in html format, recommended to be viewed in a web browser.
+This takes the form of a two-dimensional table,
+with a column for each thread,
+and each row representing the continuous execution of one thread between two (not necessarily consecutive) preemption points.
+In each row, the cell in the column corresponding to the executed thread will contain a stack trace,
+indicating the code location of the preemption point {\em at the end} of that thread transition
+(i.e., each stack trace indicates ``this thread ran until it reached the indicated line of code'').
+An example bug report is shown in Figure~\ref{fig:bugreport}.
+
+In addition to the preemption trace, the bug report reports some additional helpful information:
+a stack trace of the current thread at the ultimate point when the bug was executed,
+a message indicating the nature of the bug encountered,
+statistics about the size of the state space,
+and optionally additional information about the bug.%
+\footnote{For certain types of bugs, not pictured here;
+for example, use-after-frees will report separate stack traces
+indicating when the corresponding heap block was last allocated and freed.
+The intrepid source-diver may find all such cases of extra bug details by searching for the macro {\tt FOUND\_A\_BUG\_HTML\_INFO} in Landslide's code.}
+
+% TODO: check figure placement
+\begin{figure}[t]
+	\begin{center}
+		% nb. generated from landslide-trace-1506524837.14.html
+		\includegraphics[width=\textwidth]{bugreport.pdf}
+	\end{center}
+	\caption{Example preemption trace bug report.}
+	\label{fig:bugreport}
+\end{figure}
+
 \section{Kernel annotations}

### 2018-01-30 thesis/landslide.tex
glue scripce

@@
 \subsubsection{Dynamic configuration options}
+\label{sec:landslide-dynamicconfig}
+
 First, the following options may be changed without triggering a recompile of Landslide.
@@
 \subsubsection{Static configuration options}
+\label{sec:landslide-staticconfig}
 
@@
 		corresponds to {\tt -V}/{\tt -H} in \sect{\ref{sec:landslide-quicksand-options}}.
-	\item {\tt FILTER\_DRS\_BY\_TID=1} - configures whether to use the {\tt TID} parameter of {\tt data\_race} described above.
+	\item {\tt FILTER\_DRS\_BY\_TID=1} - configures whether to use the {\tt TID} parameter of {\tt data\_\allowbreak{}race} described above.
 	\item {\tt FILTER\_DRS\_BY\_LAST\_CALL=0} - configures whether to use the {\tt LAST\_CALL} parameter of {\tt data\_race} described above.
@@
 
-When Landslide finds a bug, it produces an execution trace of the particular interleaving of threads that led to the bug,
-in html format, recommended to be viewed in a web browser.
+When Landslide finds a bug, it produces an execution trace of the particular interleaving of threads that led to the bug.
 This takes the form of a two-dimensional table,
@@
 (i.e., each stack trace indicates ``this thread ran until it reached the indicated line of code'').
-An example bug report is shown in Figure~\ref{fig:bugreport}.
+The bug reports are formatted in html, recommended to be viewed in a web browser.
+An example is shown in Figure~\ref{fig:bugreport}.
 
-In addition to the preemption trace, the bug report reports some additional helpful information:
+In addition to the preemption trace, the bug report provides some additional helpful information:
 a stack trace of the current thread at the ultimate point when the bug was executed,
@@
 indicating when the corresponding heap block was last allocated and freed.
-The intrepid source-diver may find all such cases of extra bug details by searching for the macro {\tt FOUND\_A\_BUG\_HTML\_INFO} in Landslide's code.}
+The intrepid source-diver may find all such cases of extra bug details
+by searching for the macro {\tt FOUND\_A\_BUG\_HTML\_INFO} in Landslide's code.}
 
@@
 
+This section documents the organization of code within Landslide.
+Unless otherwise specified, Landslide's code lives in
+%C files live in
+{\tt work/modules/landslide/} (Simics implemenation) or {\tt src/bochs-2.6.8/instrument/landslide/} (Bochs implementation)
+relative to the repository root.
+
 \subsection{Scheduler}
@@
 \label{sec:landslide-glue}
+
+The options specified in \sect{\ref{sec:landslide-directly}}
+are handled by a family of gross shell scripts that live in {\tt pebsim/}.
+
+\begin{itemize}
+	\item {\tt landslide} is the outermost script invoked by Quicksand (or by a \sect{\ref{sec:landslide-directly}} aficionado).
+		It exports several key environment variables used by the other scripts,
+		ensures the instrumentation is up-to-date,
+		and launches the simulator.
+	\item {\tt getfunc.sh} defines several functions commonly used by {\tt build.sh} and {\tt definegen.sh} to extract function or global variable addresses from the program binary.
+	\item {\tt symbols.sh} defines the names of kernel functions that can be instrumented automatically without a corresponding manual annotation (e.g., {\tt malloc} and friends, the names of the {\tt tell\_landslide} family, various library helpers such as {\tt panic}).
+	\item {\tt build.sh} ensures the build of Landslide is up-to-date,
+		and processes any dynamic configuration options which don't require updating the build (\sect{\ref{sec:landslide-dynamicconfig}})
+		It verifies all required {\tt tell\_landslide} annotations are present,
+		verifies all required {\tt config.\allowbreak{}landslide} options,
+		processes the dynamic config options,
+		checks whether or not {\tt definegen.sh} needs to be run again (via hashes stored in {\tt student\_specifics.h} of the program binary and static config options),
+		and does so if necessary.
+	\item {\tt definegen.sh} produces the
+		%automatically-generated
+		content of {\tt student\_specifics.h}.
+		It repeatedly invokes the helpers defined in {\tt getfunc.sh}
+		to find the addresses of both functions specified in the config options
+		and functions whose names are known in advance.%
+		\footnote{You might think it should invoke objdump but once and keep the output in a shell variable,
+		but I tried that and it was mysteriously slower, so I gave up without ever figuring out why.}
+\end{itemize}
+
+The final output of these scripts is an auto-generated header, {\tt student\_specifics.h},
+containing a bunch of {\tt \#define}s of the addresses of important functions in the compiled binary,
+specific features enabled or disabled by the static config options (\sect{\ref{sec:landslide-staticconfig}}),
+and so on.
+The files {\tt kernel\_specifics.c}, {\tt user\_specifics.c}, and {\tt student.c} provide several interface functions
+for interpreting the current program state with respect to these values.
 

### 2018-01-31 thesis/landslide.tex
formatting/indent
write abt kernel specific landslide features

@@
 and hence unlike other chapters is written in the style of documentation rather than as a report of research results.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 \subsection{Bug reports}
+\label{sec:landslide-bugreport}
 
@@
 \end{figure}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \section{Architecture}
@@
 % foundabug goes here
+\sect{\ref{sec:landslide-bugreport}}
+
 \subsection{Pebbles-specific features}
+
+This section lists special cases of instrumentation specific to the Pebbles kernel.
+
+\begin{itemize}
+	\item {\tt mem\_check\_shared\_access()} ({\tt memory.c})
+		will assert that kernel memory is direct-mapped.
+	\item {\tt use\_after\_free()} ({\tt memory.c})
+		will ignore use-after-free reads originating from kernel code during the {\tt swexn} system call
+		(an extremely common and neither harmful nor technically interesting bug among student implementations).
+	\item {\tt cause\_test()} ({\tt test.c})
+		will issue keyboard input to type the test case name and press enter
+		when the initialization sequence completes and the shell is blocked on {\tt readline}.
+	\item {\tt kern\_mutex\_block\_others()} ({\tt schedule.c})
+		will handle the special ``blocked on via mutex'' state changes
+		whenever a mutex is acquired or released,
+		for kernels which use the {\tt tell\_landslide\_kern\_mutex\_blocking()} annotation.
+	\item {\tt sched\_update\_kern\_state\_machine()} ({\tt schedule.c})
+		will handle the reference kernel's invocation of {\tt sched\_unblock()} within {\tt cond\_signal()}
+		as a signal event for happens-before analysis.
+	\item {\tt cause\_timer\_interrupt\_immediately()} ({\tt x86.c})
+		will read the {\tt esp0} value out of the TSS to support user-to-kernel mode switches
+		in timer interrupts injected during userspace execution.
+\end{itemize}
+
 \subsection{Pintos-specific features}
+
+This section lists special cases of instrumentation specific to the Pintos kernel.
+
+\begin{itemize}
+	\item {\tt arbiter\_interested()} ({\tt arbiter.c})
+		will automatically insert preemption points on {\tt intr\_disable()} and {\tt intr\_enable()} calls
+		(immediately before and after the interrupt state is changed, respectively)
+		(as long as they aren't part of the mutex implementation, which has preemption points of its own).
+	\item {\tt lockset\_remove()} ({\tt lockset.c})
+		will warn instead of panic if a lock is unlocked twice,
+		to allow for double {\tt sema\_up()} in cases where the lock is actually a multi-use semaphore rather than a mutex.
+	\item {\tt build.sh} will edit the {\tt bootfd.img} binary to implant the name of the test case to be run in the kernel's boot command.
+	\item {\tt sched\_check\_pintos\_init\_sequence()} ({\tt schedule.c})
+		will force the kernel to skip the {\tt timer\_calibrate()} and {\tt timer\_msleep()} routines
+		used in I/O initialization.
+	\item {\tt keep\_schedule\_inflight()} ({\tt schedule.c})
+		will detect when an attempted thread switch is impossible
+		because the timer handler's try-lock will fail,
+		and will abort the interleaving early as if it never existed (which it shouldn't).
+	\item {\tt sched\_update\_kern\_state\_machine()} ({\tt schedule.c})
+		will:
+		\begin{itemize}
+			\item track invocations of {\tt timer\_sleep()} and {\tt list\_insert\_ordered()}
+				to infer when a thread is sleeping rather than blocked automatically,
+				rather than relying on the {\tt tell\_landslide\_sleeping()} annotation.
+			\item allow {\tt sema\_up()} to reenter itself,
+				which may happen when an IDE interrupt is taken when interrupts are re-enabled at the end of said function.
+			\item handle interrupt disabling/enabling as an abstract global lock for the purposes of happens-before analysis.
+		\end{itemize}
+	\item {\tt sched\_update()} ({\tt schedule.c})
+		will handle ``lock handoff'' of the abstract disable-interrupts lock
+		during a context switch for happens-before analysis.
+	\item {\tt memory.c} (various functions)
+		will handle page allocations from the {\tt palloc} family of functions in a separate memory heap,
+		allowing the usual allocator {\tt malloc} to allocate and free from {\tt palloc}'ed memory,
+		and checking both allocation heaps when checking for use-after-frees.
+	\item {\tt kern\_address\_in\_heap()} ({\tt kernel\_specifics.c})
+		will ignore DMA accesses to the VGA console,
+		which appear to be in Pintos's heap region.
+	\item {\tt test\_update\_state()} ({\tt test.c})
+		will use the boundaries of {\tt run\_test()} to denote the test lifecycle.
+\end{itemize}
+
 \subsection{Handy scripts}
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \section{Algorithms of note}
@@
 \subsection{Data race analysis}
+\subsubsection{Limited Happens-Before}
+\subsubsection{Pure Happens-Before}
 \subsection{Preemption point identification}

### 2018-02-01 thesis/landslide.tex
start talking abt foundabug

@@
 \subsection{Bug-finding output}
-% symtable stuff goes here
-% stack tracing goes here
-% foundabug goes here
-\sect{\ref{sec:landslide-bugreport}}
+
+\subsubsection{Symbol table}
+
+The symbol table logic lives in {\tt symtable.c} and is pretty much a lot of glue code.
+In the Simics version, Landslide relies on the {\tt deflsym} Simics object created by the 15-410 python scripts,
+and queries its attributes using Simics API calls.
+In the Bochs version, function names and line numbers are handled separately:
+Bochs is patched with a new API function named {\tt bx\_dbg\_symbolic\_address\_landslide}%
+\footnote{does the same thing as the existing {\tt bx\_dbg\_symbolic\_address}, but with a better type signature}
+which provides function names and hexadecimal offsets;
+while for line numbers, {\tt pebsim/pintos/build.sh}%
+\footnote{Line numbers in Bochs for Pebbles/P2s are not supported yet; see {\tt p2-setup.sh} for the work-in-progress.}
+generates a header file {\tt line\_numbers.h}
+using {\tt objdump} and {\tt addr2line},
+which the aforementioned hex offset then serves as an index into.
+
+\subsubsection{Stack traces}
+
+% features:
+% noreturn functions, when last instr is a call, will avoid printing the next function in the binary
+% various function eip offsets
+% kernel to user mode swinches
+
+\subsubsection{Preemption traces}
+
+The preemption traces, described and exemplified in \sect{\ref{sec:landslide-bugreport}}, are generated by {\tt found\_a\_bug.c},
+in cooperation with {\tt save.c}.
+Whenever {\tt save.c} creates a preemption point, it captures a stack trace of the current thread at the point it was interrupted, and saves it in the preemption point tree.
+{\tt found\_a\_bug.c} then traverses the current branch of the tree,
+potentially producing both console output and HTML output (controlled by the {\tt HTML\_PRINTF} macro family).
+It should be invoked by the {\tt FOUND\_A\_BUG} macro defined in {\tt found\_a\_bug.h},
+or by {\tt FOUND\_A\_BUG\_HTML\_INFO}, which also allows the caller to specify a callback to print extra details (such as use-after-free stack traces) in the bug report.
 
@@
 	\item {\tt sched\_update\_kern\_state\_machine()} ({\tt schedule.c})
-		will handle the reference kernel's invocation of {\tt sched\_unblock()} within {\tt cond\_signal()}
+		will:
+		\begin{itemize}
+			\item handle the reference kernel's invocation of {\tt sched\_unblock()} within {\tt cond\_signal()}
 				as a signal event for happens-before analysis.
+			\item when a {\tt vanish}ing thread is about to delete its Simics symbol table object,
+				save a ``pre-vanish'' stack trace,
+				which {\tt splice\_pre\_vanish\_trace} ({\tt stack.c}) will then splice into any subsequent stack traces by that thread,
+				so the user can see the userspace execution sequence preceding the {\tt vanish} invocation.
+		\end{itemize}
 	\item {\tt cause\_timer\_interrupt\_immediately()} ({\tt x86.c})

### 2018-02-02 thesis/landslide.tex
write abt stack traces

@@
 \subsection{config.landslide annotations}
+\label{sec:landslide-config-landslide}
 
@@
 
+The stack tracer is implemented in {\tt stack.c}.
+It does the standard approach of following the base pointer chain
+(not supporting code compiled with {\tt -fomit-frame-pointer} by doing anything clever
+like understanding how much stack frame is allocated for each function),
+and printing symbol table information for the pushed return addresses at the top of each frame.
+
+However, it also offers several special-case features
+which even some students have sometimes noticed as being more clever than Simics's stack tracer.
+I document those features here.
+If you prefer to read about these in code than in prose, see {\tt stack\_trace()} in {\tt stack.c}.
+
+\begin{itemize}
+	\item If a function is preempted at its beginning or end
+		such that its corresponding base pointer is missing from the base pointer chain,
+		Landslide will find its ``hidden'' frame and include it in the stack trace in the following cases.%
+		\footnote{Note that in such cases,
+		most other debuggers' stack tracers will be missing not the name of the interrupted function,
+		but the name of the function which called that function,
+		because it's the former's stack frame which should enable the debugger
+		to find the pushed return address for the latter.}
+		\begin{itemize}
+			\item If the last pushed return address is at offset 0 into the body of its containing function,
+				Landslide will find the next pushed return address at {\tt esp+0}.
+			\item If as above but the function is the page fault handler, at {\tt esp+4}.
+			\item If the return address is at offset 1 and the previous instruction was {\tt push ebp},
+				Landslide will find the next pushed return address at {\tt esp+4}.
+			\item If the return address is a multiple of 4 offset
+				and all previous instructions are of the form {\tt mov m32,r32},
+				Landslide will find the next pushed return address at {\tt esp+0}.
+				(This is common in student hand-written assembly functions.)
+		\end{itemize}
+	\item If the instruction at a pushed return address is a {\tt pop} or {\tt popa},
+		Landslide will search for the next non-{\tt pop}({\tt a}) instruction,
+		and if it's {\tt ret} or {\tt iret},
+		treat the function as a system call wrapper
+		(which tend not to preserve the base pointer chain)
+		and find the next return address above where all those registers were pushed.
+	\item If a return address was pushed during a fault or interrupt
+		(determined by checking for the {\tt iret} opcode or the page fault wrapper special case mentioned above),
+		Landslide will read the iret block to determine whether a stack switched happened
+		and if so what {\tt esp} used to be.
+	\item Landslide runs the tortoise/rabbit algorithm to detect cyclic {\tt ebp} chains and terminate after the first time around.
+	\item Two other Pebbles-specifc special cases described in \sect{\ref{sec:landslide-pebblesspecifics}}.
+\end{itemize}
+
+Also implemented in {\tt stack.c} is the backend of the {\tt within\_function}/{\tt without\_function} configuration command,
+which searches a given stack trace for the presence of a function return address within a specified range.
+
 % features:
@@
 \subsection{Pebbles-specific features}
+\label{sec:landslide-pebblesspecifics}
 
@@
 	\item {\tt sched\_update\_kern\_state\_machine()} ({\tt schedule.c})
-		will:
-		\begin{itemize}
-			\item handle the reference kernel's invocation of {\tt sched\_unblock()} within {\tt cond\_signal()}
+		will handle the reference kernel's invocation of {\tt sched\_unblock()} within {\tt cond\_\allowbreak{}signal()}
 		as a signal event for happens-before analysis.
-			\item when a {\tt vanish}ing thread is about to delete its Simics symbol table object,
-				save a ``pre-vanish'' stack trace,
-				which {\tt splice\_pre\_vanish\_trace} ({\tt stack.c}) will then splice into any subsequent stack traces by that thread,
-				so the user can see the userspace execution sequence preceding the {\tt vanish} invocation.
-		\end{itemize}
 	\item {\tt cause\_timer\_interrupt\_immediately()} ({\tt x86.c})
@@
 		in timer interrupts injected during userspace execution.
+	\item {\tt splice\_pre\_vanish\_trace} ({\tt stack.c})
+		will, when a {\tt vanish}ing thread has already deleted its Simics symbol table object,
+		splice in a saved ``pre-vanish'' stack trace
+		(saved previously in {\tt sched\_update\_kern\_state\_machine()})
+		so the user can see the userspace execution sequence preceding the {\tt vanish} invocation.
+	\item {\tt stack\_trace} ({\tt stack.c})
+		will, when {\tt ebp} crosses from kernel- to userspace across a system call boundary
+		(a reference kernel feature to allow Simics stack traces to cross same),
+		use {\tt PATHOS\_SYSCALL\_IRET\_DISTANCE} (\sect{\ref{sec:landslide-config-landslide}})
+		to track {\tt esp}'s value across the stack switch.
 \end{itemize}

### 2018-02-02 thesis/landslide.tex
bit more

@@
 
+The infrastructure for producing the diagnostic output to help users understand their bugs
+can be classified in three parts:
+the symbol table glue, the excessively clever stack tracer, and the preemption table generator.
+
 \subsubsection{Symbol table}
@@
 I document those features here.
-If you prefer to read about these in code than in prose, see {\tt stack\_trace()} in {\tt stack.c}.
+As a common point of implementation among them,
+Landslide traces the stack pointer {\tt esp} in addition to the base pointer {\tt ebp};
+not only updating it whenever dereferencing the base pointer,
+but also when decoding simple assembly routines, finding ``hidden'' stack frames without base pointers,
+identifying system call wrappers, and so on.
+The corresponding code lives in {\tt stack\_trace()} in {\tt stack.c}.
 

### 2018-02-08 thesis/landslide.tex
write abt time travel and friends

@@
 \subsection{State space traversal}
+
+Traversal of the state space is implemented in three parts:
+first, identifying preemption points when first encountered and selecting which thread to run for its first execution, in {\tt arbiter.c},
+second, selecting which preemption point to backtrack to after completing an execution
+and which thread to ``have switched to'' %
+% TODO: post-revisions; uncomment; change " %" above to "%"
+%\footnote{willan on-having switched to \cite{hhgg-reu}}
+instead, in {\tt explore.c},
+and third, rewinding the machine state to implement said backtracking,
+in {\tt timetravel.c} (Bochs version) and {\tt timetravel-simics.c} (Simics version).
+
+\subsubsection{Arbiter}
+
+The arbiter (named after the corresponding component of dBug \cite{dbug-ssv})
+is responsible for checking which code locations during execution should be identified as preemption points
+({\tt arbiter\_interested()}),
+and thereupon for choosing whether to keep running the current thread or to preempt and switch to a new one
+({\tt arbiter\_choose()}).
+Its behaviour in the former case is configured by the options listed in \sect{\ref{sec:landslide-dynamicconfig}},
+and in the latter case by the options listed in \sect{\ref{sec:landslide-staticconfig}}.
+For example, {\tt EXPLORE\_BACKWARDS} is interpreted here;
+if set, it will cause Landslide to always preempt and switch threads the first time it encounters each new preemption point.%
+\footnote{Another secret option, {\tt CHOOSE\_RANDOMLY},
+also exists here to randomize whether to ``explore backwards''
+(choosing independently at each preemption point, resulting in an overall unpredictable exploration order).
+% TODO: decide whether you're gonna put those estimate graphs in here
+%% I
+It's not exposed to {\tt config.landslide} but rather the user must edit it in {\tt arbiter.c} directly,
+whereupon the probability may also be adjusted via {\tt numerator} and {\tt denominator}.}
+
+\subsubsection{Explorer}
+
+Landslide invokes the explorer at the end of each execution of the test case,
+which analyzes the current branch of the interleaving state space tree
+to figure out which alternate branch to try executing next.
+Its contents are largely algorithmic rather than architectural
+and hence further described in \sect{\ref{sec:landslide-dpor}} and \sect{\ref{sec:landslide-icb}}.
+
+\subsubsection{Time Travel}
+
+After the explorer picks a past point of program preemption, % sorry this started out like that and i just had to roll with it
+Landslide collaborates with the simulator to revert the machine state to that point before switching to the desired thread.
+The Simics version is merely a bunch of wrapper glue code around the {\tt set-bookmark} and {\tt skip-to} backtracking commands.
+Bochs however does not support backtracking, so I instead use {\tt fork()} to get Linux to copy the machine state for me.
+
+The big issue to note here is that, while the simulation state should be completely reverted,
+parts of Landslide's state (e.g., scheduler runqueues, thread action flags)
+should likewise be reverted to mirror the change in program state,
+while others (tagged ancestor branches from DPOR, state space estimates) should be preserved from branch to branch.
+In Simics, I simply copy every data structure of the former case ({\tt copy\_sched()} and friends in {\tt save.c}),
+leaving those of the latter undisturbed across backtracks.%
+\footnote{Simics actually wants to save/restore all its modules' internal state on its own,
+offering an attribute set/get API for modules to expose such state (used for other purposes in {\tt simics\_glue.c}),
+but doing deep copies of data structures this way would be more trouble than it's worth.}
+
+In Bochs, {\tt fork()} automatically copies everything, so the reverse holds:
+all data of the latter case must whenever updated be propagaged to all {\tt fork()}ed children processes explicitly.
+I worried while implementing this that I might miss a case, or that future updates to the code could easily forget this step,
+resulting undoubtedly in state corruption bugs which to diagnose would be a thesis in their own right,
+so I enlisted help from my compiler via the oft-ridiculed {\tt const}.
+Every preemption point node in the execution tree ({\tt tree.h}),
+each of whose state is kept generally read-only,
+and all modifications must go through {\tt modify\_hax()} ({\tt timetravel.h}) using a modification callback,
+which internally casts away the {\tt const}, performs the requested modification,
+and also messages all relevant child processes to perform the same ({\tt timetravel\_set()} in {\tt timetravel.c}).
+The {\tt const} is absolutely, inviolably, not to be casted away, at the sacred cost of what little type safety C offers.%
+\footnote{Of course this would be followed by a footnote describing the one place where I cast it away anyway,
+{\tt mem\_check\_shared\_access()} in {\tt memory.c};
+why it's ok is documented in an {\tt XXX} comment.}
+Thence the typechecker enforces that all exploration-related state is properly propagated while scheduler state is automatically reverted.
+
 % time travel, explore, arbiter
@@
 		and if so what {\tt esp} used to be.
+	\item If a return address's offset into its containing function is 0,
+		and the last instruction in the preceding function (binary-wise) is a {\tt call},
+		Landslide will recognize it as a {\tt noreturn} tail-call, and print the correct function name.%
+		\footnote{Normally Landslide reports function/line number information for the return address as-is,
+		which indicates the next line of code after the relevant call rather than the call itself.}
 	\item Landslide runs the tortoise/rabbit algorithm to detect cyclic {\tt ebp} chains and terminate after the first time around.
@@
 
-% features:
-% noreturn functions, when last instr is a call, will avoid printing the next function in the binary
-% various function eip offsets
-% kernel to user mode swinches
-
 \subsubsection{Preemption traces}
 
-The preemption traces, described and exemplified in \sect{\ref{sec:landslide-bugreport}}, are generated by {\tt found\_a\_bug.c},
+The preemption traces, described and exemplified in \sect{\ref{sec:landslide-bugreport}}, are generated by {\tt found\_a\_\allowbreak{}bug.c},
 in cooperation with {\tt save.c}.

### 2018-02-09 thesis/landslide.tex
write abt x86.c

@@
 The HEAD commit at the time of writing is 8984021.
-The Simics port is available upon request.
+The Simics port uses Simics's proprietary API and is hence unlicensed and available upon request.
 
@@
 \subsection{Machine state manipulation}
+
+The interface to inspect and manipulate the simulated machine state lives in {\tt x86.c}.
+
+\subsubsection{Memory}
+
+{\tt read\_memory()} and {\tt write\_memory()} are both provided (with various wrapper macros in {\tt x86.h}).
+The former is used basically everywhere throughout Landslide to query the machine state;
+the latter is used only by the interrupt manipulation below
+and by the scheduler to force Pintos to skip certain parts of its init sequence (\sect{\ref{sec:landslide-pintosspecifics}}).
+Both rely on the helper function {\tt mem\_translate()} for virtual address resolution,
+which at present supports only the normal x86 32-bit addressing mode (no PAE, long mode, etc.).
+
+\subsubsection{Interrupts}
+
+Several routines are provided for manipulating system interrupts.
+Note that the Landslide is called once per fetch-decode-execute loop of the CPU,
+after the CPU processes all already-pending interrupts and decides which instruction to execute,
+but before actually executing the instruction
+(true of both Bochs and Simics).
+I refer to this as the {\em upcoming instruction}.
+Whether or not Landslide wants that instruction to execute before triggering a thread switch is a matter of some concern in the following API.
+
+\begin{itemize}
+	\item {\tt cause\_timer\_interrupt()}
+		triggers a pending timer interrupt, whose handler will be entered as soon as the upcoming instruction is executed.
+	\item {\tt cause\_timer\_interrupt\_immediately()}
+		does as above, but forces the system to enter the interrupt handler before the upcoming instruction is executed.
+		That instruction will be executed upon return from the interrupt.
+	\item {\tt avoid\_timer\_interrupt\_immediately()}
+		suppresses a timer interrupt triggered by the simulator from outside of Landslide's control.
+		It acknowledges the APIC and forces the system to jump to the end of the interrupt handler.
+	\item {\tt delay\_instruction()}
+		forces the system to execute a no-op before the upcoming instruction,
+		effectively converting an invocation of {\tt cause\_timer\_interrupt()} to {\tt cause\_timer\_interrupt\_immediately()}.
+	\item {\tt cause\_keypress()} triggers a keyboard event corresponding to the specified character.
+		The interrupt will be taken after the upcoming instruction is executed
+		(provided no timer interrupt is simultaneously pending).
+		Only {\tt a-z}, {\tt 0-9}, {\tt \_}, space, and newline are supported (enough to name any Pebbles test case).
+	\item {\tt interrupts\_enabled()} queries the CPU's interrupt flag ({\tt eflags:IF}).
+\end{itemize}
+
 \subsection{State space traversal}
@@
 
-After the explorer picks a past point of program preemption, % sorry this started out like that and i just had to roll with it
+After the explorer picks a past point of the program to preempt, % sorry this started out like that and i just had to roll with it
 Landslide collaborates with the simulator to revert the machine state to that point before switching to the desired thread.
@@
 {\tt mem\_check\_shared\_access()} in {\tt memory.c};
-why it's ok is documented in an {\tt XXX} comment.}
+why it's ok is documented in an {\tt XXX} comment in the code.}
 Thence the typechecker enforces that all exploration-related state is properly propagated while scheduler state is automatically reverted.
@@
 \subsection{Pintos-specific features}
+\label{sec:landlside-pintosspecifics}
 

### 2018-02-09 thesis/landslide.tex
describe memory.c

@@
 \subsection{Memory analysis}
+
+{\tt memory.c} is responsible for all manner of memory access analysis.
+It tracks heap allocations, checks reads and writes in the heap region against same;
+tracks reads and writes (in any region) from each thread
+and checks them against each other
+for DPOR (\sect{\ref{sec:landslide-dpor}})
+and data race analysis (\sect{\ref{sect:landslide-datarace}}).
+
+\subsubsection{Heap checking}
+
+{\tt mem\_update()} serves as the main entrypoint for tracking heap allocations.
+It's called every instruction to check for the boundaries of the {\tt malloc} library,
+and behaves in a similar way to the scheduler state machine described above.
+Then, {mem\_check\_shared\_access()} checks
+(after some elaborate manoeuvres to figure out whether to use the kernel- or userspace heap)
+whether, if in the heap region, the memory is contained in a currently-allocated heap block,
+reporting a bug if not.
+
+\subsubsection{Memory conflicts}
+
+{\tt mem\_check\_shared\_access()} also records each such access in a per-thread-transition rbtree,
+which is saved and then cleared at each preemption point.
+This allows {\tt mem\_shm\_intersect()}, called at each preemption point once for each of its ancestors
+($n^2$ total calls per interleaving),
+to perform a set intersection to find any memory conflicts.
+Any such conflicts whicih also fail a lockset (\sect{\ref{sec:landslide-lhb}})
+and/or happens-before (\sect{\ref{sec:landslide-phb}}) check
+are then reported as data races.
+Regardless, all such conflicts are later used by DPOR (\sect{\ref{sec:landslide-dpor}}) to find dependent transition pairs.
+
 \subsection{Machine state manipulation}
@@
 \end{itemize}
+
+Note that {\tt kern\_ready\_for\_timer\_interrupt()} should generally be invoked separately
+from {\tt interrupts\_enabled()} if needed;
+while {\tt interrupts\_enabled()} must be true before invoking {\tt cause\_timer\_interrupt()},
+if the kernel is not ready the interrupt may not be received for a long time.
+Also, use of {\tt cause\_timer\_interrupt\_immediately()} must not be used while the kernel is not ready.
 
@@
 \subsection{Data race analysis}
+\label{sec:landslide-datarace}
 \subsubsection{Limited Happens-Before}
+\label{sec:landslide-lhb}
 \subsubsection{Pure Happens-Before}
+\label{sec:landslide-phb}
 \subsection{Preemption point identification}

### 2018-02-09 thesis/landslide.tex
scheudler

@@
 
+Both simulators invoke Landslide once per instruction and once per memory read or write.
+% just before..
+The entry point is the aptly-named {\tt landslide\_entrypoint()} in {\tt landslide.c},
+which then dispatches to various other modules their respective analyses, described as follows.
+
+\subsection{Execution Tree}
+\label{sec:landslide-save}
+
 \subsection{Scheduler}
+
+The Landslide scheduler, which lives in {\tt schedule.c}, has two main duties:
+to maintain an accurate representation of all the existing threads on the simulated system
+and track what concurrency-relevant actions each is performing at any given time,
+and to orchestrate the sequence of timer interrupts necessary
+to cause the simulated system to context switch to any given thread at any given time.
+System-wide state is stored in a single {\tt struct sched\_state},
+including the thread queues (runqueue, deschedule queue, and sleep queue),
+while per-thread state is stored in {\tt struct agent}s (named after the terminology of \cite{dbug-ssv}) which live on said queues.
+
+It has one main entrypoint, {\tt sched\_update()}, in which both the state machine is updated and scheduling decisions are made.
+The interface also offers helper functions for finding and manipulating {\tt agent}s,
+and {\tt sched\_recover()}, which prepares the scheduler to force a new thread to run
+after a time travel (\sect{\ref{sec:landslide-timetravel}}).
+
+\subsubsection{State machine}
+
+The first part of {\tt sched\_update()} is to update the state machine of thread actions and runnability.
+Much of this functionality is found in
+{\tt sched\_update\_kern\_state\_machine()}
+and
+{\tt sched\_update\_user\_state\_machine()}.
+The current intruction pointer is compared against the known locations of
+the mutex API, system calls, runnable/descheduling {\tt tell\_\allowbreak{}landslide} annotations, and so on,
+and locksets, action flags, and runqueue membership are updated accordingly.
+
+\subsubsection{Interrupt injection}
+
+The second part of {\tt sched\_update()},
+conditional on the arbiter identifying preemption points (\sect{\ref{sec:landslide-arbiter}}),
+manages timer interrupts to switch to a desired thread.
+Whenever a preemption point is reached,
+the scheduler first creates a checkpoint in the execution tree (\sect{\ref{sec:landslide-save}}),
+asks the arbiter which thread to run next,
+and if that thread is different from the current one,
+forces the kernel into its timer interrupt handler (\sect{\ref{sec:landslide-interrupce}}).
+
+Because the kernel is part of the system being tested, Landslide can't necessarily always switch directly to a specific thread,
+but rather must keep triggering context switches until the desired thread is reached;
+any mechanism to tell the kernel which thread we want would necessarily involve modifying the code being testsed
+and hence possibly obscuring bugs or introducing new ones.%
+\footnote{For userspace testing, where I supply a pre-annotated reference kernel,
+such an approach would be more straightforward,
+but the kernel-testing repeated-context-switch approach infrastructure was already in place
+and it was easier to reuse that than to add more code.}
+
+The scheduler marks up to one thread as the ``schedule target'',
+which when set makes Landslide wait until that thread is reached before looking for more preemption points,
+so the kernel may finish its context switches undisturbed.
+Whenever the schedule target is set and the end of the context switcher is reached,
+if the schedule target is not the current thread,
+the scheduler repeats this process until it is.%
+\footnote{Note that this ``loop'' is not structured as an explicit loop in Landslide's code,
+but rather as part of the state machine which updates each time a new instruction is traced.}
+
 \subsection{Memory analysis}
@@
 which is saved and then cleared at each preemption point.
-This allows {\tt mem\_shm\_intersect()}, called at each preemption point once for each of its ancestors
+This allows {\tt mem\_shm\_\allowbreak{}intersect()}, called at each preemption point once for each of its ancestors
 ($n^2$ total calls per interleaving),
 to perform a set intersection to find any memory conflicts.
-Any such conflicts whicih also fail a lockset (\sect{\ref{sec:landslide-lhb}})
-and/or happens-before (\sect{\ref{sec:landslide-phb}}) check
+Any such conflicts which also fail a lockset
+and/or happens-before check (\sect{\ref{sec:landslide-datarace}})
 are then reported as data races.
@@
 \subsubsection{Interrupts}
+\label{sec:landslide-interrupce}
 
@@
 \subsubsection{Arbiter}
+\label{sec:landslide-arbiter}
 
@@
 \subsubsection{Time Travel}
+\label{sec:landslide-timetravel}
 

### 2018-02-10 thesis/landslide.tex
tree.h

@@
 
+The execution tree
+%(or state space, or preemption point log, however calling it suits you)
+is stored as a chain of preemption point nodes named {\tt struct hax} defined in {\tt tree.h}.
+Although the state space of possible interleavings is exponentially-sized,
+Landslide does not actually need to store any nodes for execution sequences outside the current variation
+(see \sect{\ref{sec:landslide-estimate}} and \sect{\ref{sec:landslide-dpor}} for why),
+so the total memory consumption is only $O(n)$ in the number of preemption points in a single program run
+(for the test cases used in this thesis, typically 20-1000).
+Each {\tt hax} stores the following information:
+
+\begin{itemize}
+	\item Basic statistics such as the current instruction pointer, thread ID,
+		stack trace of current thread at the moment of preemption,
+		depth in the tree, parent node pointer, etc.;
+	\item Snapshots of the current state of the scheduler (\sect{\ref{sec:landslide-scheduler}})
+		and memory accesses and heaps (\sect{\ref{sec:landslide-memory}});
+	\item Simulator-dependent data needed to time travel and resume execution
+		from this checkpoint (\sect{\ref{sec:landslide-timetravel}});
+	\item List of parent/ancestor nodes with memory conflicts and/or happens-before edges to this one
+		for DPOR (\sect{\ref{sec:landslide-dpor}});
+	\item Current estimated state space proportion and execution time for the subtree rooted at this node
+		(not necessarily fully explored yet) for estimation \sect{\ref{sec:landslide-estimate}};
+	\item Whether this point is an {\tt xbegin} invocation
+		and if so what {\tt xabort} codes are possible and/or already explored for this transaction
+		(\sect{\ref{chap:txn}}).
+\end{itemize}
+
 \subsection{Scheduler}
+\label{sec:landslide-scheduler}
 
@@
 \subsection{Memory analysis}
+\label{sec:landslide-memory}
 
@@
 and behaves in a similar way to the scheduler state machine described above.
-Then, {mem\_check\_shared\_access()} checks
+Then, {\tt mem\_check\_shared\_access()} checks
 (after some elaborate manoeuvres to figure out whether to use the kernel- or userspace heap)
@@
 \subsection{State space estimation}
+\label{sec:landslide-estimate}
 \subsection{Dynamic Partial Order Reduction}

### 2018-02-10 thesis/overview.tex
overview.tex is much easier to type than background

--- 2017-01-07 proposal/overview.tex
@@
 \label{chap:background}
+\inspirationalquote{
+\begin{tabular}{p{0.83\textwidth}}
+I see now that none of us are yet ready. The cycle exists so that we may improve ourselves. But
+the one who reaches the summit is not our superior, for they stand on our shoulders to reach it.
+\end{tabular}}
+{The Shepherd v82.6.0174, The Talos Principle}
 
@@
 
-Thesis proposal note: I've written this section with the intent of reusing it in the thesis.
-The other chapters, being necessarily more impermanent, are aimed directly at you (my thesis committee) as the audience.
-This section is written with the intention for programmers of any level to read.
+{\bf Special personal note for the curious reader:} I have taken special care to write this chapter to be approachable by any intermediate programmer familiar with basic C programming concepts.
+My committee members obviously do not require such treatment, but I hope the rare reader from outside this relatively narrow field shall also not find the barrier to entry here too high.
+Concurrency analysis and verification is often very abstract,
+requiring strong intuition for certain concepts to understand the next algorithm that builds upon them, and so on;
+and I'd hate for any reader to get left behind off the bat for not knowing what a mutex is.
+%Abstract as the field is, it's also quite difficult to make good visual aids, unlike for example in graphics research, but I have done my best.
+Should you find any of my explanations insufficiently illuminating, please do get in touch.
 
-\section{Concurrency Bugs}
+\section{Concurrency}
 
@@
 Most often, a concurrency novice who programs with shared state will consider the possible interleavings where one thread's access sequence occurs entirely before the other's, but neglect to consider intermediate outcomes in which the threads' access sequences are interleaved.
-%
+
 Consider the program in Figure \ref{fig:concurrency-bug}: Any output between 2 and 2000 is possible\footnote{
-	Exercise for the reader: Show why 2 is a possible output, but 1 is not!
+	Fun exercise for the reader: Show why 2 is a possible output, but 1 is not!
 },
@@
 offering three new instructions: \texttt{xbegin}, \texttt{xend}, and \texttt{xabort}, to begin, commit, and fail a transaction, respectively.
+The example program in Figure~\ref{fig:htm-example} demonstrates how these primitives can be used to synchronize a simple shared access without locking overhead in the common case\footnote{
+	The solution presented here is actually incomplete; stay tuned until Chapter~\ref{chap:tm} for the surprising twist!
+},
+using GCC's compiler intrinsics \cite{htm-gcc}.
 
-In this thesis, I will focus on HTM as my platform for testing transactional programs, to highlight the importance of researching advanced testing techniques in anticipation of upcoming hardware features.
-My treatment of TM will be largely agnostic to whether a transactional interface is backed by HTM or STM,
-and I will extend it to cover STM as well, as discussed in Section~\ref{sec:tm}.
+\begin{figure}[h]
+	\begin{center}
+		\begin{tabular}{l}
+		\texttt{void count() \{} \\
+		\texttt{~~~~for (int i = 0; i < 1000; i++) \{} \\
+		\texttt{~~~~~~~~if ((status = \_xbegin()) == \_XBEGIN\_STARTED) \{} \\
+		\texttt{~~~~~~~~~~~~x++;} \\
+		\texttt{~~~~~~~~~~~~\_xend();} \\
+		\texttt{~~~~~~~~\} else \{} \\
+		\texttt{~~~~~~~~~~~~mutex\_lock(\&m);} \\
+		\texttt{~~~~~~~~~~~~x++;} \\
+		\texttt{~~~~~~~~~~~~mutex\_unlock(\&m);} \\
+		\texttt{~~~~~~~~\}} \\
+		\texttt{~~~~\}} \\
+		\texttt{\}} \\
+		\end{tabular}
+	\end{center}
+	\caption{The example {\tt count} routine from Figure~\ref{fig:mutex}, rewritten to use HTM.
+		If the transaction in the top branch aborts,
+		whether from a memory conflict or random system interrupt,
+		%from the programmer's intention,
+		execution will revert to the return of {\tt \_xbegin},
+		{\tt status} will be assigned an error code indicating the abort reason,
+		and control will drop into the {\tt else} branch.
+		The programmer can then use explicit synchronization, such as a mutex, to resolve the conflict.}
+	\label{fig:htm-example}
+\end{figure}
+
+Concerning possible execution patterns,
+the main difference between STM and HTM is the circumstances under which a transaction may abort.
+A software-backed transaction will abort if and only if a memory conflict occurs therein with another thread.
+HTM, however, is backed by the CPU's cache, and is therefore subject to other circumstances such as cache capacity or interrupt-triggered cache flushes which may force an abort even when no memory conflict occurs.
+I will explore the consequences of this difference further in Chapter~\ref{chap:tm}.
+
+In this thesis, I will focus on HTM as my platform for testing transactional programs,
+to highlight the importance of researching advanced testing techniques in anticipation of upcoming hardware features.
 
@@
 
+\subsection{The state space}
+
 Model checking \cite{verisoft} is a testing technique for systematically exploring the possible thread interleavings of a concurrent program.
@@
 
-{\bf Preemption points.}
+\subsubsection{Preemption points}
+
 During execution, a model checker identifies a subset of the program's operations as ``interesting'', i.e.,
-points at which interrupting the current thread to run a different one is likely to produce different behaviour.
-These points, called {\em preemption points}, may be identified by any combination of human intuition and machine analysis.
+where interrupting the current thread to run a different one is likely to produce different behaviour.
+These so-called {\em preemption points} may be identified by any combination of human intuition and machine analysis.
 Typical preemption points include the boundaries of synchronization APIs (e.g., {\tt mutex\_lock}) or accesses to shared variables.
-Considering that at each preemption point multiple threads exist as options to run next, the set of possible ways to execute the program can be viewed as a tree.
-% TODO: fig tree
-I show an example execution tree in Figure~\ref{fig:tree}.
-The number of preemption points in each execution define the depth of this tree, and the number of threads available to run define the branching factor.
+Considering that at each preemption point multiple threads exist as options to run next,
+the set of possible ways to execute the program can be viewed as a tree.
+Figure~\ref{fig:tree}(b) shows a visualization of the corresponding tree from our example program.
+
+The number of preemption points in each execution defines the depth of this tree,
+and the number of threads available to run defines the branching factor.
 Hence, in a program with $n$ preemption points and $k$ threads available to run at each, the state space size is $O(n^k)$.
+Nevertheless, to fully test all of a program's possible behaviours, we must check the executions corresponding to every branch of the tree.
 Addressing the scaling problem in this exponential relation is the central research problem for all model checkers.
 
-{\bf Reduction techniques.}
+\subsection{On the size of state spaces}
+
+At its essence, stateless model checking research is a perpetual struggle to become more and more efficient in order to test and verify bigger and bigger programs.
+But whence this efficiency?
+Techniques for coping with the exponential explosion fall into two categories:
+(1) removing redundant interleavings from the state space when we can prove they are equivalent to some interleaving already tested,
+or {\bf reduction techniques},
+and
+(2) prioritizing interleavings judged as more likely to contain bugs should bugs exist
+in case we are unable to exhaustively test all interleavings after all,
+or {\bf search heuristics}.
+
+\subsubsection{Reduction techniques}
+
 Dynamic Partial Order Reduction \cite{dpor} (henceforth, DPOR) is the premier algorithm for mitigating the exponential explosion that arises as program size increases.
-It identifies equivalent execution sequences according to Mazurkiewicz trace theory \cite{mazurkiewicz},
+
+{\bf Abstractly speaking:}
+Let {\em independent transitions} denote a pair of executions of two threads, each from one preemption point to the next,
+in which there are no read/write or write/write access pairs to the same memory between threads.
+DPOR reduces a state space, originally exponentially-sized in the number of thread transitions,
+to an equivalent one
+(i.e., testing which suffices to check all program behaviours that could arise in the original state space)
+exponentially-sized in the number of {\em dependent} thread transitions.
+%More intuitively, if two thread transitions between preemption points do not conflict on any shared resource access,
+%reordering them produces an equivalent interleaving, i.e., the same program behaviour.
+More technically, it identifies equivalent execution sequences according to Mazurkiewicz trace theory \cite{mazurkiewicz},
 and tests at least one execution from each equivalence class.
-Intuitively, if two thread transitions between preemption points do not conflict on any shared resource access, reordering them produces an equivalent interleaving, i.e., the same program behaviour.
-Figure~\ref{fig:dpor} shows a part of an execution tree in which the operations by threads 1 and 2 are independent, and hence only one of the subtrees need be tested.
+
+{\bf Concretely speaking:}
+Figure~\ref{fig:dpor} highlights part of an execution tree where the execution ordering of threads 1 and 2 are swapped,
+and each interleaving has a respective ``subtree'' (i.e., possible interleavings given the fixed execution prefix leading up to it).
+The specifics of execution before the thread 1/thread 2 sequence,
+other possible threads to run instead of threads 1 or 2,
+and what logic the program executes in those subtrees
+are all presumably arbitrary.
+In these two highlighted branches,
+if the transitions of threads 1 and 2 are {\em independent},
+%if the operations performed by threads 1 and 2 are independent
+%(i.e., no write/read or write/write access pairs to the same memory),
+DPOR deduces that the subsequent program states (indicated by the red arrow) are equivalent.
+Thence, only one of the two interleavings and its respective subtree needs to be executed
+in order to check all possible program states.
+I explain how DPOR implements such a deduction in more detail in \sect{\ref{sec:landslide-dpor}}.
+
 Over the years, researchers have developed many enhancements to DPOR, such as Optimal DPOR \cite{optimal-dpor}, parallelizable DPOR \cite{parallel-dpor}, SAT-directed model checking \cite{satcheck}, Maximal Causality Reduction \cite{mcr}, and DPOR for relaxed memory architectures \cite{tsopso}.
@@
 
-{\bf Variants of Happens-Before.}
-        Most prior work focuses on {\em Happens-Before} \cite{lamport-clocks} as the order relation between accesses.
+\subsection{Happens-Before}
+\label{sec:background-hb}
+
+Condition 4 of the above definition expresses the notion that the access pair can be executed concurrently,
+regardless of whether the hardware actually carries out the operations in the same physical instant.
+Several approaches exist to formally representing this condition.
+
+\begin{itemize}
+	\item Most prior work focuses on {\em Happens-Before} \cite{lamport-clocks} as the order relation between accesses.
 \cite{predictive-dr} and \cite{hybriddatarace} identify a problem with this approach:

### 2018-02-10 thesis/thesis.tex
Revert "add charter font"
This reverts commit 273453a99843294c4c20125cd14b94b51eaadc2f.
overview.tex is much easier to type than background

@@
 
-\setmainfont{Charter}[
-	Path = ./fonts/,
-	UprightFont = XCharter-Roman.otf,
-	BoldFont = XCharter-Bold.otf,
-	ItalicFont = XCharter-Italic.otf,
-	BoldItalicFont = XCharter-BoldItalic.otf
-	]
+\setmainfont{XCharter}
 %\setmainfont{Utopia}[
@@
 \begin{dedication}
-For Elsie, Margie, Johnny, Mel, Eve, and David, who brought me here today.
+For Elsie, Margie, Johnny, Mel, Eve, and David.
+%, who brought me here today.
 \end{dedication}
@@
 \input{intro}
-\input{background}
+\input{overview}
 \input{landslide}

### 2018-02-11 thesis/landslide.tex
data race

@@
 
-\subsection{Kernel annotations}
-\subsection{Bug identification}
-\subsection{Data race analysis}
-\label{sec:landslide-datarace}
-\subsubsection{Limited Happens-Before}
-\label{sec:landslide-lhb}
-\subsubsection{Pure Happens-Before}
-\label{sec:landslide-phb}
+% not worth discussing?
+%\subsection{Kernel annotations}
+%\subsection{Bug identification}
+
 \subsection{Preemption point identification}
@@
 \label{sec:landslide-dpor}
+
+\subsection{Data race analysis}
+\label{sec:landslide-datarace}
+
+Whenever a memory conflict is identified for DPOR as described above,
+the access pair's corresponding locksets and/or happens-before edges are checked to determine if it's also a data race.
+Note the distinction: DPOR memory conflicts indicate that two thread transitions,
+if reordered, could produce different behaviour, even if all accesses therein are adequately synchronized;
+while a data race indicates furthermore that the two threads can be interleaved precisely at the moment of one or both accesses.
+% nb. i think "only one of the accesses can be preempted at and have the other interleaved inside, but not the other way around,
+% is a property of cli/sti involvement in kernel space only.
+
+The core of the comparison is in {\tt check\_locksets()} in {\tt memory.c}.
+It checks each DPOR memory conflict's locksets, for limited happens-before,
+and happens-before edges, for pure happens-before
+(\sect{\ref{sec:background-hb}}).
+
+\subsubsection{Limited Happens-Before}
+\label{sec:landslide-lhb}
+
+Conditions 1, 2, and 4 defined in \sect{\ref{sec:background-hb}}, provided the Limited Happens-Before definition for \#4,
+coincide exactly with the conditions to be considered a memory conflict under DPOR.
+Hence all that remains to be checked is \#3, the set of locks held by each thread at the time of access.
+
+Routines for recording lockset changes and computing set intersection are found in {\tt lockset.c}.
+Apart from standard data structure manipulation,
+one algorithmic point of note is that locks are distinguished by types in addition to address.
+This allows (e.g.) mutexes stored as part of the implementation of semaphores to protect a different set of accesses than are protected by the semaphore they implement.
+
+% TODO: talk abt free re malloc false positives
+
+\subsubsection{Pure Happens-Before}
+\label{sec:landslide-phb}
+
+In Pure Happens-Before,
+condition 4 is replaced with the traditional distributed systems notion of Happens-Before \cite{lamport-clocks}.
+Landslide implements this via the vector clocks approach described by {\sc FastTrack} \cite{fasttrack}.
+I refer the reader interested in the vector clock algorithm itself to the {\sc FastTrack} paper,
+limiting discussion here to Landslide's corresponding implementation of each inference rule.
+
+Despite being algorithmically worse in the size of the vector clocks,
+I use the {\sc Djit+} rules for reads and writes rather than the {\tt FastTrack} ones;
+because Landslide tests should be limited to few threads to manage the state space size,
+$O(n)$ is usually in the single digits, so I optimize for code simplicity.
+
+\begin{itemize}
+	\item Reads and Writes (writes vs reads comparison already handled by DPOR) ({\tt memory.c})
+	\begin{itemize}
+		\item \textsc{Djit+ read/write same epoch} - {\tt vc\_eq()} case of {\tt add\_lockset\_to\_shm()}
+		\item \textsc{Djit+ read/write} - {\tt vc\_happens\_before()} case of {\tt check\_locksets()}
+	\end{itemize}
+	\item Synchronization ({\tt schedule.c})
+	\begin{itemize}
+		\item \textsc{FT acquire}
+			\begin{itemize}
+				\item {\tt kern\_mutex\_\{,try\}locking\_done()} cases of {kern\_update\_state\_machine()}
+				\item {\tt user\_mutex\_\{,try\}lock\_exiting()} cases of {user\_update\_state\_machine()}
+				\item {\tt cli} case of {\tt kern\_update\_state\_machine()} (Pintos only)
+				\item {\tt cli}/{\tt sti} lock handoff case in {\tt sched\_update()} (Pebbles only)
+			\end{itemize}
+		\item \textsc{FT release}
+			\begin{itemize}
+				\item {\tt kern\_mutex\_unlocking()} case of {kern\_update\_state\_machine()}
+				\item {\tt user\_mutex\_unlock\_entering()} case of {user\_update\_state\_machine()}
+				\item {\tt sti} case of {\tt kern\_update\_state\_machine()} (Pintos only)
+				\item {\tt cli}/{\tt sti} lock handoff case in {\tt sched\_update()} (Pebbles only)
+			\end{itemize}
+		\item \textsc{FT fork} - {\tt agent\_fork()}
+		\item \textsc{FT join} - {\tt sched\_unblock()} case of {\tt kern\_update\_state\_machine()} (Pebbles only)
+	\end{itemize}
+\end{itemize}
+
 \subsection{Iterative Context Bounding}

### 2018-02-11 thesis/landslide.tex
fix some stuff

@@
 
-Despite being algorithmically worse in the size of the vector clocks,
-I use the {\sc Djit+} rules for reads and writes rather than the {\tt FastTrack} ones;
+I use the {\sc Djit+} rules for reads and writes rather than the {\tt FastTrack} ones,
+even though they incur $O(n)$ runtime in the size of the vector clocks more often:
 because Landslide tests should be limited to few threads to manage the state space size,
@@
 \begin{itemize}
-	\item Reads and Writes (writes vs reads comparison already handled by DPOR) ({\tt memory.c})
+	\item Reads and Writes ({\tt memory.c})
 	\begin{itemize}
@@
 			\begin{itemize}
-				\item {\tt kern\_mutex\_\{,try\}locking\_done()} cases of {kern\_update\_state\_machine()}
-				\item {\tt user\_mutex\_\{,try\}lock\_exiting()} cases of {user\_update\_state\_machine()}
+				\item {\tt kern\_mutex\_\{,try\}locking\_done()} cases of {\tt kern\_update\_state\_\allowbreak{}machine()}
+				\item {\tt user\_mutex\_\{,try\}lock\_exiting()} cases of {\tt user\_update\_state\_\allowbreak{}machine()}
 				\item {\tt cli} case of {\tt kern\_update\_state\_machine()} (Pintos only)
@@
 			\begin{itemize}
-				\item {\tt kern\_mutex\_unlocking()} case of {kern\_update\_state\_machine()}
-				\item {\tt user\_mutex\_unlock\_entering()} case of {user\_update\_state\_machine()}
+				\item {\tt kern\_mutex\_unlocking()} case of {\tt kern\_update\_state\_machine()}
+				\item {\tt user\_mutex\_unlock\_entering()} case of {\tt user\_update\_state\_machine()}
 				\item {\tt sti} case of {\tt kern\_update\_state\_machine()} (Pintos only)

### 2018-02-15 landslide-lecture/2018/lecture.tex
landslide lecture - copy 2017 2018

--- 2017-02-28 landslide-lecture/2017/lecture.tex
@@
 \title[Landslide]{{\bf Landslide: A New Race-Finding Tool for 15-410}}
-\subtitle[]{ {\em more clever than } ``\texttt{mandelbrot}'' {\em since 2011.}}
+\subtitle[]{ {\em more clever than } ``\texttt{agility\_drill}'' {\em since 2011.}}
 \author[Ben Blum]{Ben Blum \texttt{(bblum@andrew.cmu.edu)}}
@@
 \institute[CMU 15-410]{Carnegie Mellon University - 15-410}
-\date[]{2017, February 20}
+\date[]{2017, September 27}
 
@@
 \begin{frame}{Outline}
-	\textbf{Theory: Seeing race conditions in a new way}
+	\textbf{Theory: Seeing concurrency bugs in a new way}
 	\begin{itemize}
@@
 \begin{frame}{About The Project}
-	\textbf{About me}: 6th year graduate student, advised by Garth Gibson
+	\textbf{About me}: Final(?) year graduate student, advised by Garth Gibson
 	\begin{itemize}
@@
 
-\begin{frame}{Iterative Deepening of Preemption Points}
-	Goal: Run the best tests for a given CPU budget.
-	%\begin{itemize}
-	%	\item Technique: ``Iterative Deepening''
-	%\end{itemize}
-	\linegap
-
-	Based on experience from past 15-410 student volunteers
-	\begin{itemize}
-		%\item Students worked best with an iterative process
-		\item ``Start small, then add more preemption points as time allows''
-		\item Landslide now automates this process
-	\end{itemize}
-	\linegap
-
-	Named after analogous technique in chess AI.
-	\begin{itemize}
-		\item Chess search is DFS limited by max number of moves (ply).
-		\item Chess AIs repeat DFS, increasing ply, until timeout.
-	\end{itemize}
-\end{frame}
+%\begin{frame}{Iterative Deepening of Preemption Points}
+%	Goal: Run the best tests for a given CPU budget.
+%	%\begin{itemize}
+%	%	\item Technique: ``Iterative Deepening''
+%	%\end{itemize}
+%	\linegap
+%
+%	Based on experience from past 15-410 student volunteers
+%	\begin{itemize}
+%		%\item Students worked best with an iterative process
+%		\item ``Start small, then add more preemption points as time allows''
+%		\item Landslide now automates this process
+%	\end{itemize}
+%	\linegap
+%
+%	Named after analogous technique in chess AI.
+%	\begin{itemize}
+%		\item Chess search is DFS limited by max number of moves (ply).
+%		\item Chess AIs repeat DFS, increasing ply, until timeout.
+%	\end{itemize}
+%\end{frame}
 
@@
 
+\begin{frame}{What makes a Landslide-friendly test}
+	``Why not {\tt largetest}, {\tt juggle}, {\tt cyclone}, {\tt agility\_drill}...?''
+	\linegap
+
+	Sample code from {\tt cyclone}: % abbreviated
+	\begin{center}
+	\begin{tabular}{l}
+		\texttt{for (i = 0; i < MAX\_MISBEHAVE /* 64 */; i++) \{} \\
+		\texttt{~~~~misbehave(i);} \\
+		\texttt{~~~~tid = thr\_create(child\_fn, i);} \\
+		\texttt{~~~~thr\_join(tid, \&status);} \\
+		\texttt{\}}
+	\end{tabular}
+	\end{center}
+	\pause
+	\linegap
+
+	Stress tests expose various interleavings using big loops/many threads;
+	\linegap
+
+	Landslide finds many interleavings itself; even if the test case has no loops.
+	\vspace{0.01in}
+
+	How long do you think Landslide would take to test {\tt cyclone}..?
+\end{frame}
+
 %\begin{frame}{Preliminary Evaluation}
@@
 \begin{frame}{Previous Semesters}
-	%Spring 2015: 7 groups signed up to use Landslide; 5 found bugs
-	%%\begin{itemize}
-	%%	\item (that they didn't find during stress testing)
-	%%\end{itemize}
-	%%%%% hopefully!
-	%\linegap
-
-	%Among all groups, 38 total tests were run
-	%\begin{itemize}
-	%	\item 21 tests ran to completion without finding bugs
-	%	\item 4 {\em deterministic} bugs
-	%	\item 13 distinct {\em non-deterministic} bugs
-	%		\begin{itemize}
-	%			\item 6 in {\tt thr\_exit\_join} \small (use-after-free, deadlock, NULL, infinite loop...)
-	%			\item 4 in {\tt paraguay} (deadlock, infinite loop)
-	%			\item 2 in {\tt broadcast\_test} (deadlock, infinite loop)
-	%			\item 1 in {\tt rwlock\_downgrade\_read\_test} (use-after-free)
-	%		\end{itemize}
-	%		%\begin{itemize}
-	%		%	\item 6 of which required data-race PPs to expose
-	%		%\end{itemize}
-	%\end{itemize}
-	S'15, F'15, S'16: 48 groups signed up to use Landslide; 38 found bugs
-	% 18 groups found races
-	%\begin{itemize}
-	%	\item (that they didn't find during stress testing)
-	%\end{itemize}
-	%%%% hopefully!
-	\linegap
-
-	44 {\em deterministic} bugs (e.g. swexn, initialization)
-	\linegap
-
-	85 distinct {\em non-deterministic} bugs (among 32 groups)
-	\begin{itemize}
-		\item 24 groups (75\%) fixed $\ge 1$ such bug
+	%% S'15, F'15, S'16: 48 groups signed up to use Landslide; 38 found bugs
+	S'15, F'15, S'16, F'16, S'17: 79 groups signed up to use Landslide; 59 found bugs
+	\linegap
+
+	%%44 {\em deterministic} bugs (e.g. swexn, initialization)
+	91 {\em deterministic} bugs (e.g. swexn, initialization)
+	\linegap
+
+	%%85 distinct {\em non-deterministic} bugs (among 32 groups)
+	113 distinct {\em non-deterministic} bugs (among 44 groups)
+	\begin{itemize}
+		\item 33 groups (75\%) fixed $\ge 1$ such bug
+		%%\item 24 groups (75\%) fixed $\ge 1$ such bug
 			\begin{itemize}
@@
 			\end{itemize}
-		\item 15 groups (47\%) fixed {\em all} such bugs
+		\item 24 groups (55\%) fixed {\em all} such bugs
+		%%\item 15 groups (47\%) fixed {\em all} such bugs
 		\item Most ambitious group: 11 distinct races found + fixed!
@@
 	\linegap
-	% 44 dets 85 races
-
-	%Among all groups, 185 distinct tests were run
-	%\begin{itemize}
-	%	\item 137 tests ran to completion without finding bugs
-	%	\item 21 {\em deterministic} bugs
-	%\end{itemize}
-	%\pause % ask: "who can tell me why landslide might find a *deterministic* bug not found during past testing? A: use-after-free
-	%\linegap
-
-	%38 distinct {\em non-deterministic} bugs
-	%\begin{itemize}
-	%		%\begin{itemize}
-	%		%	\item 7 of which required data-race PPs to expose
-	%		%\end{itemize}
-	%	\item 11 groups (50\%) fixed all their bugs
-	%		\begin{itemize}
-	%			\item (as verified by running Landslide again -- not a guarantee!)
-	%		\end{itemize}
-	%	\item 2.7 bugs ``fixed'' (see above) on average per group
-	%	\item Most ambitious group: 11 distinct races found + fixed!
-	%\end{itemize}
 \end{frame}
@@
 			\begin{itemize}
-				\item \url{http://tinyurl.com/landslide-p2-s17}
+				\item (watch your email for the link)
+				%\item \url{http://tinyurl.com/landslide-p2-s17}
 			\end{itemize}
@@
 			\begin{itemize}
-				\item Later this week, room and time TBD
+				\item Next week, room and time TBD
 			\end{itemize}
@@
 	\end{itemize}
-	\linegap
-
-	{\em Really} interested?
-	\begin{itemize}
-		\item I know of an M.S. thesis topic if you want to do post-OS research.
-	\end{itemize}
+	%% im out of time :<<
+	%\linegap
+
+	%{\em Really} interested?
+	%\begin{itemize}
+	%	\item I know of an M.S. thesis topic if you want to do post-OS research.
+	%\end{itemize}
 \end{frame}

### 2018-02-15 thesis/landslide.tex
landslide lecture - copy 2017 2018

@@
 
-I use the {\sc Djit+} rules for reads and writes rather than the {\tt FastTrack} ones,
-even though they incur $O(n)$ runtime in the size of the vector clocks more often:
-because Landslide tests should be limited to few threads to manage the state space size,
-$O(n)$ is usually in the single digits, so I optimize for code simplicity.
+I use the {\sc Djit+} rules for reads and writes rather than the {\sc FastTrack} ones,
+even though they more often incur $O(n)$ runtime in the size of the vector clocks:
+because Landslide tests should be limited to few threads in order to manage the state space size,
+$n$ is always in the single digits, so I optimize for code simplicity.
 
 \begin{itemize}
-	\item Reads and Writes ({\tt memory.c})
+	\item Reads and writes ({\tt memory.c})
 	\begin{itemize}
@@
 		\item \textsc{FT fork} - {\tt agent\_fork()}
-		\item \textsc{FT join} - {\tt sched\_unblock()} case of {\tt kern\_update\_state\_machine()} (Pebbles only)
+		\item \textsc{FT join} - {\tt sched\_unblock()} case of {\tt kern\_update\_state\_machine()}
+			(Pebbles only; Pintos case is handled by above {\tt cli}/{\tt sti} cases in context switch)
 	\end{itemize}
@@
 \label{sec:landslide-icb}
+
+Iterative Context Bounding \cite{chess-icb} is a state space exploration strategy
+that prioritizes interleavings with fewer total preemptions first.
+Let $P(I)$ denote the number of preemptions in an interleaving $I$.
+Then, to summarize in pseudocode a na\"ive exploration of some state space $S$ as:
+
+\begin{algorithm}[h]
+	\ForEach{$I \in S$}{
+		Execute($I$)
+	}
+\end{algorithm}
+
+ICB's approach could likewise be summarized as follows:
+
+\begin{algorithm}[h]
+	\For{$n \in [0..\mathsf{max}_P(S)]$}{
+		\ForEach{$I \in S, P(I) \le n$}{
+			Execute($I$)
+		}
+	}
+\end{algorithm}
+
+%% stupid imperative way to write it
+%\begin{algorithm}[h]
+%	\For{$n \in [0..\infty]$}{
+%		$\mathsf{skipped\_any} := \mathsf{false}$ \\
+%		\ForEach{$I \in S$}{
+%			\uIf{$P(I) \le n$}{
+%				Execute($I$)
+%			} \Else {
+%				$\mathsf{skipped\_any} := \mathsf{true}$
+%			}
+%		}
+%		\If{$\neg \mathsf{skipped\_any}$}{
+%			{\bf Break}
+%		}
+%	}
+%\end{algorithm}
+
+However, when combined with DPOR to prune equivalent interleavings,
+DPOR's reduction might not be sound with respect to the subset of $S$ under $P \le n$.
+Prior work introduced Bounded Partial Order Reduction (BPOR) \cite{bpor},
+a compatibility extension to DPOR for ICB to address this problem.
+I refer the reader interested in the algorithm's details to the citation;
+this section will discuss Landslide's implementation.
+
+% TODO: say how u do it
+
 \subsection{Heuristic loop, deadlock, and synchronization detection}

### 2018-02-16 landslide-lecture/2018/lecture.tex
upd lecture for s18

@@
 \institute[CMU 15-410]{Carnegie Mellon University - 15-410}
-\date[]{2017, September 27}
+\date[]{2018, February 16}
 
@@
 		\end{tabular}
+		\pause
 		\linegap
@@
 		\end{itemize}
-		\item Releasing to paying customers and worrying about correctness later
+		\item \sout{Releasing to paying customers and worrying about correctness later}
 			% Say... "and I'm hoping to prevent you from finding yourself in a future where this happens!"
@@
 \begin{frame}{About The Project}
-	\textbf{About me}: Final(?) year graduate student, advised by Garth Gibson
+	\textbf{About me}: Final year graduate student, advised by Garth Gibson
 	\begin{itemize}
@@
 	%% S'15, F'15, S'16: 48 groups signed up to use Landslide; 38 found bugs
-	S'15, F'15, S'16, F'16, S'17: 79 groups signed up to use Landslide; 59 found bugs
+	%% S'15, F'15, S'16, F'16, S'17: 79 groups signed up to use Landslide; 59 found bugs
+	S'15, F'15, S'16, F'16, S'17, F'17: 94 groups signed up to use Landslide; 71 found bugs
 	\linegap
@@
 	%%44 {\em deterministic} bugs (e.g. swexn, initialization)
-	91 {\em deterministic} bugs (e.g. swexn, initialization)
+	%% F17 lecture - stats up to S17
+	%% 91 {\em deterministic} bugs (e.g. swexn, initialization)
+	% S18 lecture - stats up to F17
+	109 {\em deterministic} bugs (e.g. swexn, initialization)
 	\linegap
@@
 	%%85 distinct {\em non-deterministic} bugs (among 32 groups)
-	113 distinct {\em non-deterministic} bugs (among 44 groups)
-	\begin{itemize}
-		\item 33 groups (75\%) fixed $\ge 1$ such bug
+	%% F17 lecture - stats up to S17
+	%% 113 distinct {\em non-deterministic} bugs (among 44 groups)
+	% S18 lecture - stats up to F17
+	122 distinct {\em non-deterministic} bugs (among 49 groups)
+	\begin{itemize}
+		\item 36 groups (73\%) fixed $\ge 1$ such bug
+		%\item 33 groups (75\%) fixed $\ge 1$ such bug
 		%%\item 24 groups (75\%) fixed $\ge 1$ such bug
@@
 			\end{itemize}
-		\item 24 groups (55\%) fixed {\em all} such bugs
+		\item 26 groups (53\%) fixed {\em all} such bugs
+		%\item 24 groups (55\%) fixed {\em all} such bugs
 		%%\item 15 groups (47\%) fixed {\em all} such bugs

### 2018-03-01 thesis/landslide.tex
define dpor happens-before

@@
 The HEAD commit at the time of writing is 8984021.
-The Simics port uses Simics's proprietary API and is hence unlicensed and available upon request.
+The Simics port uses Simics's proprietary API and is hence unlicensed and available upon request for educational use only.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 Separate user guides also exist, described in Chapter~\ref{chap:410}.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 \end{itemize}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \subsection{Running Landslide directly}
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \subsection{Bug reports}
@@
 but a CMU 15-410 student who wishes to use Landslide on their kernel project shall need to brave forth hereupon.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \subsection{In-kernel code annotations}
@@
 which then dispatches to various other modules their respective analyses, described as follows.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \subsection{Scheduler}
@@
 but rather as part of the state machine which updates each time a new instruction is traced.}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 Regardless, all such conflicts are later used by DPOR (\sect{\ref{sec:landslide-dpor}}) to find dependent transition pairs.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \subsection{State space traversal}
@@
 Thence the typechecker enforces that all exploration-related state is properly propagated while scheduler state is automatically reverted.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 or by {\tt FOUND\_A\_BUG\_HTML\_INFO}, which also allows the caller to specify a callback to print extra details (such as use-after-free stack traces) in the bug report.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \subsection{Pintos-specific features}
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \subsection{Preemption point identification}
@@
 % TODO: talk about how w/in func and w/out func are processed
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \subsection{State space estimation}
 \label{sec:landslide-estimate}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \subsection{Dynamic Partial Order Reduction}
 \label{sec:landslide-dpor}
+
+Landslide implements Dynamic Partial Order Reduction (DPOR) \cite{dpor}
+to identify concurrent yet independent thread transitions
+whose permutations can safely be pruned from the state space
+while still testing all possible program behaviours.
+
+The DPOR implementation consists of 3 parts:
+computing memory conflicts, computing happens-before,
+and tagging alternate branches to explore to drive the state space exploration.
+The former two are computed as each preemption point is reached,
+for the associated thread transition pairwise with all other preceding thread transitions.
+The latter is computed at the end of each full interleaving executed, using the results of the two former,
+and constitutes the bulk of the algorithm.
+
+\subsubsection{Happens-before}
+
+The happens-before relation expresses when two thread transitions can potentially be reordered,
+or in other words, are concurrent (despite the serialized nature of the simulated execution).
+This relation is expressed in the following definitions paraphrased from \cite{dpor}.
+
+\begin{itemize}
+	\item {\bf Enabled}: A transition $t$ is enabled in a state $s$ when a state $s'$ exists such that $s \xrightarrow{t} s'$ exists.
+		In systems terms, the scheduler at state $s$ considers the thread corresponding to $t$ to be runnable.
+	\item {\bf Dependent}: Two transitions $t_i$ and $t_j$ are dependent if
+		\begin{itemize}
+			\item $t_1$ is enabled in $s$ and $s \xrightarrow{t_1} s'$, and
+			\item $t_2$ is enabled in $s$ but not enabled in $s'$, or vice versa.
+		\end{itemize}
+		In systems terms, either $t_1$ and $t_2$ belong to the same thread,
+		or the execution of $t_1$ at $s$ causes $t_2$'s thread
+		to change state from blocked to runnable or vice versa.
+		Landslide computes this relation in {\tt enabled\_by()} in {\tt save.c}.
+		\footnote{The original definition includes a second criterion that, from $s$,
+		the executions $t_1 t_2$ and $t_2 t_1$ result in identical states;
+		this captures the memory independence relation which I defer to the next subsection.}
+	\item {\bf Happens-Before}: The happens-before relation for a transition sequence $S = t_1 \dots t_n$
+		is the smallest relation $\rightarrow_S$ on ${1 \dots n}$ such that
+		\begin{enumerate}
+			\item if $i < j$ and $S_i$ and $S_j$ are dependent then $i \rightarrow_S j$, and
+			\item $\rightarrow_S$ is transitively closed.
+		\end{enumerate}
+		Landslide computes this relation in the aptly-named {\tt compute\_happens\_before()} in {\tt save.c}.
+\end{itemize}
+The happens-before relation is a partial order expressing the scheduling constraints of a given interleaving.
+All pairs of interleavings not included are subject to reordering,
+and hence candidates for new thread interleavings to test.
+
+Note that DPOR's notion of happens-before differs from the traditional distributed systems definition \cite{lamport-clocks}
+or Pure Happens-Before;
+rather, it coincides with condition 3 of Limited Happens-Before as defined in \sect{\ref{sec:background-datarace}}
+(in fact, Landslide's Limited HB implementation simply reuses the same result computed for DPOR's purpose).
+
+\subsubsection{Memory conflicts}
+
+The memory conflict relation expresses when two thread transitions are dependent,
+or in other words, when their behaviour could potentially vary depending which executes first.
+
+When each preemption point is reached,
+Landslide saves the current set of all memory accesses since the previous preemption point,
+compares it to the accesses from all previous transitions by different threads
+which weren't marked as happens-before in the previous section,
+and then empties that set for the next thread transition ({\tt shimsham\_shm()} in {\tt save.c}).
+
+The set intersection is implemented in {\tt mem\_shm\_interseect()} in {\tt memory.c}.
+It checks for read/write and write/write pairs to the same address with an $O(\mathsf{max}(m,n))$ scan of both access sets (pre-sorted).
+Whenever a conflict is identified, it also invokes the data race analysis (\sect{\ref{sec:landslide-datarace}}).
+It checks for {\tt free()}/access conflicts as well as access pairs,
+effectively treating deallocation of a heap block as a ``poisoning'' write to its entire contents,
+which should conflict with accesses to any address therein.
+
+\subsubsection{State space exploration}
+
+\subsubsection{Example}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \subsection{Iterative Context Bounding}
@@
 	}
+	\caption{Straightforward exploration ordering}
+	\label{alg:not-icb}
 \end{algorithm}
 
+% TODO: make sure latex doesn't fuck up the algorithm placement
 ICB's approach could likewise be summarized as follows:
@@
 	}
+	\caption{ICB exploration ordering}
+	\label{alg:icb}
 \end{algorithm}
@@
 
-However, when combined with DPOR to prune equivalent interleavings,
+First of all, note that Algorithm~\ref{alg:icb} is structured in a way that repeats interleavings
+with fewer than $n$ preemptions that have already been checked in previous iterations of the outer loop.
+This is because the number of preemptions in each branch is not known in advance;
+rather, the state space must always be explored in an overall depth-first approach,
+at best skipping too-preemptful interleavings as they are encountered.
+As simple as it would be to state ``{\bf foreach} $I \in {\mathsf{sort}_P(S)}$'' in pseudocode,
+implementing such an ordering would be much less straightforward. %
+% TODO: post committee review
+%\footnote{Translating conference papers' pseudocode into usable implementations is often quite difficult \cite{itg2}.}
+The core ICB loop appears in {\tt time\_travel()} in {\tt landslide.c}.
+%(although not explicitly structured as a C-style loop in the code)
+
+Secondly, prior work \cite{bpor} has shown that
+when combined with DPOR to prune equivalent interleavings,
 DPOR's reduction might not be sound with respect to the subset of $S$ under $P \le n$.
-Prior work introduced Bounded Partial Order Reduction (BPOR) \cite{bpor},
+That work introduced Bounded Partial Order Reduction (BPOR),
 a compatibility extension to DPOR for ICB to address this problem.
-I refer the reader interested in the algorithm's details to the citation;
-this section will discuss Landslide's implementation.
-
+To summarize, % TODO: summarize this once you write the DPOR section so you know what terminology to write in terms of
+%I refer the reader interested in the algorithm's details to the citation;
+%this section will discuss Landslide's implementation.
 % TODO: say how u do it
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \subsection{Heuristic loop, deadlock, and synchronization detection}

### 2018-03-01 thesis/overview.tex
define dpor happens-before

@@
 \section{Data Race Analsyis}
+\label{sec:background-datarace}
 
@@
 		Students typically implement semaphores using mutexes and condition variables, not using atomic instructions or system calls directly.
-	\item Reader-writer locks (rwlocks), with the interface {\tt rwlock\_lock(rwp, mode)} and {\tt rwlock\_unlock(rwp)}. {\tt mode} may be either {\tt RWLOCK\_READ} or {\tt RWLOCK\_}- {\tt WRITE}.
+	\item Reader-writer locks (rwlocks), with the interface {\tt rwlock\_lock(rwp, mode)} and {\tt rwlock\_unlock(rwp)}. {\tt mode} may be either {\tt RWLOCK\_READ} or {\tt RWLOCK\_\allowbreak{}WRITE}.
 		Behaves as mutexes, but multiple readers may access the critical section simultaneously.
@@
 	\item Semaphores (the basic concurrency primitive, implemented using direct scheduler calls): {\tt sema\_up}, {\tt sema\_down}, {\tt sema\_try\_down};
-	\item Locks (which simply wrap a semaphore initialized to 1), {\tt lock\_acquire}, {\tt lock\_}- {\tt release}, {\tt lock\_try\_acquire};
-	\item Condition variables (also implemented using scheduler calls): {\tt cond\_wait}, {\tt cond\_}- {\tt signal}, {\tt cond\_broadcast}, with the same semantics as Pebbles P2 condvars;
+	\item Locks (which wrap a semaphore initialized to 1), {\tt lock\_acquire}, {\tt lock\_\allowbreak{}release}, {\tt lock\_try\_acquire};
+	\item Condition variables (also implemented using scheduler calls): {\tt cond\_wait}, {\tt cond\_\allowbreak{}signal}, {\tt cond\_broadcast}, with the same semantics as Pebbles P2 condvars;
 	\item Basic round-robin scheduling facilities: {\tt thread\_block} (a kernel-level analogue to Pebbles's {\tt deschedule}), {\tt thread\_yield}
@@
 	\item Interrupt and fault handlers;
-	\item A page allocator, {\tt palloc\_get\_page}, {\tt palloc\_get\_multiple}, {\tt palloc\_}- {\tt free\_page}, {\tt palloc\_free\_multiple}
+	\item A page allocator, {\tt palloc\_get\_page}, {\tt palloc\_get\_multiple}, {\tt palloc\_free\_page}, and {\tt palloc\_free\_multiple}
 \end{itemize}

### 2018-03-01 thesis/thesis.tex
landslide lecture - copy 2017 2018
quotes

@@
 \usepackage{xspace}
+\usepackage[linesnumbered,ruled]{algorithm2e}
+\usepackage{algpseudocode}
 
@@
 \chapter{Quicksand}
-%\inspirationalquote{
-%\begin{tabular}{p{0.55\textwidth}}
-%There are awful, horrible things in this world; I know that now.
-%But there are a lot of things worth protecting, too.
-%\end{tabular}}
-%{Kaname Madoka, Mahou Shoujo Madoka{\raisebox{0.1em}{$\scriptstyle \bigstar$}}Magica}
-\inspirationalquote{Always, somewhere, someone is fighting for you.
-As long as you remember her, you are not alone.}
-{Mahou Shoujo Madoka{\raisebox{0.1em}{$\scriptstyle \bigstar$}}Magica}
+\inspirationalquote{
+% original
+% {\footnotesize 嫌な事も悲しい事もあったけど、守りたい物だって沢山この世界にはあったから…} \\
+% trans. slightly paraphrased from crunchyroll subtitle; removed "..."s, added past tense for あった
+There were awful, sad things in this world.
+But there were a lot of things worth protecting, too.
+}
+{Kaname Madoka, Mahou Shoujo Madoka{\raisebox{0.1em}{$\scriptstyle \bigstar$}}Magica}
+%\inspirationalquote{Always, somewhere, someone is fighting for you.
+%As long as you remember her, you are not alone.}
+%{Mahou Shoujo Madoka{\raisebox{0.1em}{$\scriptstyle \bigstar$}}Magica}
 
@@
 it fills you with determination.}{Undertale (paraphrased)}
+
+\chapter{Transactions}
+%\inspirationalquote{
 
@@
 	\end{tabular}}
-{Galatea, The Talos Principle}
+%{Galatea, The Talos Principle}
+{The Talos Principle}
 

### 2018-03-04 thesis/landslide.tex
explain DPOR

@@
 The DPOR implementation consists of 3 parts:
-computing memory conflicts, computing happens-before,
+computing happens-before,
+computing memory conflicts,
 and tagging alternate branches to explore to drive the state space exploration.
@@
 
+In this section $t_i$ will denote a transition between two program states during execution,
+with each state being a preemption point as identified in \sect{\ref{sec:landslide-pps}},
+and $T(t_i)$ will denote the thread which was scheduled to produce that transition.
+
 \subsubsection{Happens-before}
@@
 \begin{itemize}
-	\item {\bf Enabled}: A transition $t$ is enabled in a state $s$ when a state $s'$ exists such that $s \xrightarrow{t} s'$ exists.
-		In systems terms, the scheduler at state $s$ considers the thread corresponding to $t$ to be runnable.
+	\item {\bf Enabled}: A transition $t_i$ is enabled in a state $s$ when a state $s'$ exists such that $s \xrightarrow{t_i} s'$ exists.
+		In systems terms, the scheduler at state $s$ considers $T(t_i)$ to be runnable.
 	\item {\bf Dependent}: Two transitions $t_i$ and $t_j$ are dependent if
-		\begin{itemize}
+		\begin{enumerate}
 			\item $t_1$ is enabled in $s$ and $s \xrightarrow{t_1} s'$, and
 			\item $t_2$ is enabled in $s$ but not enabled in $s'$, or vice versa.
-		\end{itemize}
-		In systems terms, either $t_1$ and $t_2$ belong to the same thread,
-		or the execution of $t_1$ at $s$ causes $t_2$'s thread
+		\end{enumerate}
+		In systems terms, either $T(t_1) = T(t_2)$,
+		or the execution of $t_1$ at $s$ causes $T(t_2)$
 		to change state from blocked to runnable or vice versa.
+		\footnote{The original paper's definition includes a second criterion that, from $s$,
+		%the executions $t_1 t_2$ and $t_2 t_1$ result in identical states;
+		a state $s'$ exists such that both $s \xrightarrow{t_1 t_2} s'$ and $s \xrightarrow{t_2 t_1} s'$.
+		This captures the memory independence relation,
+		but computationally requires direct comparison of program states.
+		{\em Stateless} model checkers compute memory conflicts separately from happens-before,
+		to find and prune such identical states implicitly,
+		as described in the next two subsections.}
 		Landslide computes this relation in {\tt enabled\_by()} in {\tt save.c}.
-		\footnote{The original definition includes a second criterion that, from $s$,
-		the executions $t_1 t_2$ and $t_2 t_1$ result in identical states;
-		this captures the memory independence relation which I defer to the next subsection.}
 	\item {\bf Happens-Before}: The happens-before relation for a transition sequence $S = t_1 \dots t_n$
@@
 		\end{enumerate}
-		Landslide computes this relation in the aptly-named {\tt compute\_happens\_before()} in {\tt save.c}.
+		Landslide computes this relation in
+		%the aptly-named
+		{\tt compute\_happens\_before()} in {\tt save.c}.
 \end{itemize}
@@
 All pairs of interleavings not included are subject to reordering,
-and hence candidates for new thread interleavings to test.
+and hence candidates for new interleavings to test.
 
 Note that DPOR's notion of happens-before differs from the traditional distributed systems definition \cite{lamport-clocks}
-or Pure Happens-Before;
-rather, it coincides with condition 3 of Limited Happens-Before as defined in \sect{\ref{sec:background-datarace}}
+as used in Pure Happens-Before \sect{\ref{sec:background-hb}};
+rather, it coincides with condition 3 of Limited Happens-Before %as defined in \sect{\ref{sec:background-datarace}}
 (in fact, Landslide's Limited HB implementation simply reuses the same result computed for DPOR's purpose).
@@
 
-The memory conflict relation expresses when two thread transitions are dependent,
+The memory conflict relation expresses when two transitions are dependent,
 or in other words, when their behaviour could potentially vary depending which executes first.
 
-When each preemption point is reached,
-Landslide saves the current set of all memory accesses since the previous preemption point,
-compares it to the accesses from all previous transitions by different threads
-which weren't marked as happens-before in the previous section,
-and then empties that set for the next thread transition ({\tt shimsham\_shm()} in {\tt save.c}).
+Upon execution of each $t_j \in S$,
+Landslide saves the current set of all memory accesses since the previous preemption point (call this $M(t_j)$),
+compares it to all $M(t_i)$ with $i < j$ and $i \not\rightarrow_S j$,
+and then begins recording subsequent memory accesses in a new empty set for $t_{j+1}$.
+%When each preemption point is reached,
+%Landslide saves the current set of all memory accesses since the previous preemption point,
+%compares it to the accesses from all previous transitions by different threads
+%which weren't marked as happens-before in the previous section,
+%and then empties that set for the next thread transition
+({\tt shimsham\_shm()} in {\tt save.c}).
+These $M(t)$ sets are mappings from memory addresses to
+instruction pointer value, read-or-write boolean, lockset or vector clock, and various other metadata
+({\tt struct mem\_lockset} in {\tt memory.h}).
 
@@
 It checks for read/write and write/write pairs to the same address with an $O(\mathsf{max}(m,n))$ scan of both access sets (pre-sorted).
+If any address $a$ exists with $a \in M(t_i)$ and $a \in M(t_j)$ and $M(t_i)(a) = \mathsf{write} \vee M(t_j)(a) = \mathsf{write}$
+then $t_i$ and $t_j$ are said to conflict,
+which I will denote $t_i \leftrightsquigarrow_S t_j$.
+
 Whenever a conflict is identified, it also invokes the data race analysis (\sect{\ref{sec:landslide-datarace}}).
@@
 effectively treating deallocation of a heap block as a ``poisoning'' write to its entire contents,
-which should conflict with accesses to any address therein.
+which is considered to conflict with accesses to any address therein
+on the grounds that reordering may expose a use-after-free.
 
@@
 
+The core of the DPOR algorithm is implemented in {\tt explore()} in {\tt explore.c}.
+
+{\bf Definition.}
+Given a transition sequence (execution, interleaving, preemption trace) $S = t_1 \dots t_n$,
+the DPOR algorithm identifies any number of alternate interleavings that must be tested.
+Each such interleaving I will denote in this section as $I_{ij} = (t_1 \dots t_{i-1}, T_j)$,
+where $t_1 \dots t_{i-1}$ is the common execution prefix shared between $S$ and the new interleaving,
+and $T_j$ is the thread ID to be scheduled after $t_{i-1}$, $T_j \ne T(t_i)$.
+%to run {\em instead} of $T(t_i)$.
+\footnote{I describe $T_j$ as a thread ID here, rather than as a thread transition,
+because the nature of the transition (its memory accesses, the subsequent state, etc)
+is unknown until we actually execute it.}
+Landslide's implementation represents $S$ as a list of {\tt struct hax}es, defined in {\tt tree.h},
+each one representing a preemption point, or intermediate state between two transitions.
+
+{\bf Identifying new interleavings.}
+To summarize, for each pair of transitions $t_i$ and $t_j$ ($i<j$),
+DPOR identifies interleavings which will reorder them,
+executing $t_j' t_i'$ to reach a new program state not exposed in the current interleaving.
+If $t_i \rightarrow_S t_j$ then they cannot be reordered,
+and if $t_i \not\leftrightsquigarrow_S t_j$ then reordering them
+will produce a state already encountered in this interleaving;
+hence, DPOR considers only transition pairs
+which are not related by happens-before and are related by memory conflict ({\tt is\_evil\_ancestor()}).
+
+% TODO: change i<4 to i<j.
+% i accidentally typed i<4 here and decided it'd be funny to leave it this way
+For each such pair $t_i$, $t_j$, $i<4$, let $s$ denote the state (preemption point) immediately before $t_i$.
+Then:
+\begin{itemize}
+	\item If $T(t_j)$ is runnable at $s$, return $I_{ij} = (t_1 \dots t_{i-1}, T_j)$ ({\tt tag\_good\_sibling()}).
+	\item Otherwise, there must be some third thread runnable at $s$
+		(from $t_i \not\rightarrow_S t_j$, by contrapositive of the second condition of dependence);
+		then, return all $I_{ik} = (t_1 \dots t_{i-1}, T_k)$ such that $T_k$ is runnable at $s$ and $T_k \ne T(t_i)$
+		({\tt tag\_all\_siblings()}).
+\end{itemize}
+
+{\bf Tracking already-visited interleavings.}
+Let $U(I_{ij})$ denote the sub-state-space (or sub-tree) beginning at the state
+which results from executing $T_j$ after $t_1 \dots t_{i-1}$;
+in other words, all sequences $S' = t_1 \dots t_{i-1}, t'_i, \dots t'_{n'}$ with $T(t'_{i+1}) = T_j$.
+I use a depth-first search ordering,
+so for any such $U$ outside the current interleaving,
+either all or none of its $S'$s will have been tested already.
+Therefore, to avoid repeating interleavings,
+Landslide need only store at each {\tt struct hax} a list of threads
+such that their corresponding subtree $U$ is fully explored,
+and can omit any non-constant-size information about the contents of that $U$
+({\tt struct hax\_child} in the code).
+Hence Landslide's DPOR implementation's memory cost is $O(n*k)$,
+$k$ being the average number of runnable threads at each preemption point.
+
+{\bf Choosing which new interleaving to test next.}
+Among all interleavings chosen by DPOR not already marked as explored,
+Landslide chooses the one with the longest execution prefix matching the current $S$,
+to maintain the depth-first search invariant.
+(In the case of a tie, differing only by which thread to run, it chooses arbitrarily.)
+All other new interleavings are marked to explore later in their corresponding {\tt struct hax},
+and automatically included in the result of any future iterations of DPOR until they are tested.
+Because the one with the longest execution prefix was chosen to test next,
+all others must share their execution prefixes with it,
+so storing markers for the others during the next execution is still $O(1)$ memory.
+
+When DPOR returns no new $I_{ij}$s not already marked in the set of visited subtrees,
+the exploration is complete.
+
 \subsubsection{Example}
+
+Acknowledging that quickly gaining an intuitive understanding for an algorithm
+is basically impossible from just paragraphs of text,
+I provide in this section an example of a single iteration of DPOR
+on the example state space from Figures~\ref{fig:concurrency-bug} and \ref{fig:tree}.
 

### 2018-03-04 thesis/landslide.tex
revise dpor desc

@@
 		(from $t_i \not\rightarrow_S t_j$, by contrapositive of the second condition of dependence);
-		then, return all $I_{ik} = (t_1 \dots t_{i-1}, T_k)$ such that $T_k$ is runnable at $s$ and $T_k \ne T(t_i)$
+		then, return all $I_{ik} = (t_1 \dots t_{i-1}, T_k)$ such that
+		$T_k \ne T(t_i)$ and $T_k$ is runnable at $s$
 		({\tt tag\_all\_siblings()}).
@@
 {\bf Tracking already-visited interleavings.}
-Let $U(I_{ij})$ denote the sub-state-space (or sub-tree) beginning at the state
-which results from executing $T_j$ after $t_1 \dots t_{i-1}$;
-in other words, all sequences $S' = t_1 \dots t_{i-1}, t'_i, \dots t'_{n'}$ with $T(t'_{i+1}) = T_j$.
-I use a depth-first search ordering,
+Let $U(I_{ij})$ denote the sub-state-space (or sub-tree) beginning at the next preemption point reached
+after executing $T_j$ after $t_1 \dots t_{i-1}$;
+in other words, the set of all sequences $S' = t_1 \dots t_{i-1}, u_i, \dots u_{n'}$ with $T(u_i) = T_j$.
+Landslide orders its search depth-first,
 so for any such $U$ outside the current interleaving,
@@
 and can omit any non-constant-size information about the contents of that $U$
-({\tt struct hax\_child} in the code).
-Hence Landslide's DPOR implementation's memory cost is $O(n*k)$,
-$k$ being the average number of runnable threads at each preemption point.
+({\tt struct hax\_child}).
+Hence the memory cost of Landslide's DPOR implementation is $O(nk)$,
+$k$ being the maximum runnable threads at any preemption point
+(which in turn is always single digits for model checking tests).
 
@@
 all others must share their execution prefixes with it,
-so storing markers for the others during the next execution is still $O(1)$ memory.
+preserving the $O(nk)$ memory bound described above.
+%so storing markers for the others during the next execution is still $O(1)$ memory.
 

### 2018-03-04 thesis/landslide.tex
few revisions dpor

@@
 {\bf Identifying new interleavings.}
-To summarize, for each pair of transitions $t_i$ and $t_j$ ($i<j$),
-DPOR identifies interleavings which will reorder them,
-executing $t_j' t_i'$ to reach a new program state not exposed in the current interleaving.
+To find which alternate interleavings need to be tested,
+% TODO: change i<4 to i<j.
+% i accidentally typed i<4 here and decided it'd be funny to leave it this way
+DPOR compares pairwise each pair of transitions $t_i$ and $t_j$, $i<4$, in the current interleaving $S$.
 If $t_i \rightarrow_S t_j$ then they cannot be reordered,
@@
 will produce a state already encountered in this interleaving;
-hence, DPOR considers only transition pairs
-which are not related by happens-before and are related by memory conflict ({\tt is\_evil\_ancestor()}).
-
-% TODO: change i<4 to i<j.
-% i accidentally typed i<4 here and decided it'd be funny to leave it this way
-For each such pair $t_i$, $t_j$, $i<4$, let $s$ denote the state (preemption point) immediately before $t_i$.
+hence, DPOR marks new interleavings only
+when $t_i \not\rightarrow_S t_j$ and $t_i \leftrightsquigarrow_S t_j$
+%for transition pairs not related by happens-before and are related by memory conflict
+({\tt is\_evil\_ancestor()}).
+\footnote{To aid intuition, consider the two extremes:
+if all transitions are related by happens-before,
+the program is not concurrent and no alternate interleavings are possible;
+if all transitions are memory-independent,
+the program exhibits full data isolation between threads and all schedules are observably equivalent.}
+
+For each such pair, let $s$ denote the state (preemption point) before $t_i$.
 Then:
@@
 	\item If $T(t_j)$ is runnable at $s$, return $I_{ij} = (t_1 \dots t_{i-1}, T_j)$ ({\tt tag\_good\_sibling()}).
-	\item Otherwise, there must be some third thread runnable at $s$
-		(from $t_i \not\rightarrow_S t_j$, by contrapositive of the second condition of dependence);
+	\item Otherwise, there must be some third thread runnable at $s$;
+		\footnote{AFSOC $T(t_i)$ is the only runnable thread at $s$,
+		then either $T(t_i)$'s execution at $s$ enabled $T(t_j)$,
+		or it enabled an intermediate transition
+		(whether by $T(t_i)$ or a third thread)
+		which in turn enabled $T(t_j)$.
+		In either case $t_i$ is transitively dependent with $t_j$, contradicting $t_i \not\rightarrow_S t_j$.}
+		%(from $t_i \not\rightarrow_S t_j$, by contrapositive of the second condition of dependence);
 		then, return all $I_{ik} = (t_1 \dots t_{i-1}, T_k)$ such that
@@
 \end{itemize}
+
+To summarize, %for each pair of transitions $t_i$ and $t_j$ ($i<j$) in $S$,
+DPOR identifies $I_{ij}$s which will (eventually) reorder each conflicting, concurrent transition pair in $S$
+%executing some $t_j' t_i'$
+to reach a (possibly) new program state not exposed in the current interleaving.
 
@@
 is basically impossible from just paragraphs of text,
-I provide in this section an example of a single iteration of DPOR
+I conclude this section with an example of a single iteration of DPOR
 on the example state space from Figures~\ref{fig:concurrency-bug} and \ref{fig:tree}.
@@
 if reordered, could produce different behaviour, even if all accesses therein are adequately synchronized;
-while a data race indicates furthermore that the two threads can be interleaved precisely at the moment of one or both accesses.
+while a data race indicates furthermore that the two threads can be interleaved precisely at the moment of one or both accesses,
+supposing that a new preemption point were introduced to split one or both transitions in half.
 % nb. i think "only one of the accesses can be preempted at and have the other interleaved inside, but not the other way around,
@@
 
-Conditions 1, 2, and 4 defined in \sect{\ref{sec:background-hb}}, provided the Limited Happens-Before definition for \#4,
-coincide exactly with the conditions to be considered a memory conflict under DPOR.
+Conditions \#1, \#2, and \#4 defined in \sect{\ref{sec:background-hb}},
+provided the Limited Happens-Before definition for \#4,
+coincide with DPOR's version of happens-before described in the previous section.
+%coincide exactly with the conditions to be considered a memory conflict under DPOR.
 Hence all that remains to be checked is \#3, the set of locks held by each thread at the time of access.

### 2018-03-05 thesis/landslide.tex
summarize need for bpor

@@
 
+% TODO
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 that prioritizes interleavings with fewer total preemptions first.
-Let $P(I)$ denote the number of preemptions in an interleaving $I$.
-Then, to summarize in pseudocode a na\"ive exploration of some state space $S$ as:
+Let $P(S)$ denote the number of preemptions in an execution sequence $S$.
+Then, to summarize in pseudocode a na\"ive exploration of some state space $U$ as:
 
 \begin{algorithm}[h]
-	\ForEach{$I \in S$}{
-		Execute($I$)
+	\ForEach{$S \in U$}{
+		Execute($S$)
 	}
@@
 \begin{algorithm}[h]
-	\For{$n \in [0..\mathsf{max}_P(S)]$}{
-		\ForEach{$I \in S, P(I) \le n$}{
-			Execute($I$)
+	\For{$n \in [0..\mathsf{max}_P(U)]$}{
+		\ForEach{$S \in U, P(S) \le n$}{
+			Execute($S$)
 		}
@@
 %		$\mathsf{skipped\_any} := \mathsf{false}$ \\
-%		\ForEach{$I \in S$}{
-%			\uIf{$P(I) \le n$}{
-%				Execute($I$)
+%		\ForEach{$S \in U$}{
+%			\uIf{$P(S) \le n$}{
+%				Execute($S$)
 %			} \Else {
@@
 at best skipping too-preemptful interleavings as they are encountered.
-As simple as it would be to state ``{\bf foreach} $I \in {\mathsf{sort}_P(S)}$'' in pseudocode,
+As simple as it would be to state ``{\bf foreach} $S \in {\mathsf{sort}_P(U)}$'' in pseudocode,
 implementing such an ordering would be much less straightforward. %
@@
 
+\subsubsection{Bounded Partial-Order Reduction}
+
 Secondly, prior work \cite{bpor} has shown that
 when combined with DPOR to prune equivalent interleavings,
-DPOR's reduction might not be sound with respect to the subset of $S$ under $P \le n$.
+DPOR's reduction might not be sound with respect to the subset of $U$ under $P \le n$.
 That work introduced Bounded Partial Order Reduction (BPOR),
 a compatibility extension to DPOR for ICB to address this problem.
-To summarize, % TODO: summarize this once you write the DPOR section so you know what terminology to write in terms of
+
+To summarize, when DPOR identifies some interleaving $I_{ij}$ to test,
+it may not be possible to execute $T(t_j)$ after $t_i$ without exceeding the current preemption bound.
+However, there may exist another interleaving $J_{i'j}$ within the bound which runs $T(t_j)$ before $t_i$.
+If a DPOR implementation na\"ively configured with ICB simply skipped $I$ on account of the preemption bound,
+there's no guarantee $J$ would get marked for exploration
+from any other iteration and/or any other pair of conflicting transitions.
+Even though restricting the state space to a certain maximum preemption bound is already unsound
+in terms of losing full concurrency coverage,
+failing to test even $J$ would be a failure of DPOR itself
+to soundly prune the already-reduced state space defined by the bound.
+Hence the need for BPOR, to ensure that if such an alternative $J$ to $I$ exists within the bound,
+$J$ gets marked for exploration immediately.
+
 %I refer the reader interested in the algorithm's details to the citation;

### 2018-03-10 thesis/landslide.tex
decribe icb, bpor

@@
 
+\subsubsection{Implementation}
+
 First of all, note that Algorithm~\ref{alg:icb} is structured in a way that repeats interleavings
@@
 %\footnote{Translating conference papers' pseudocode into usable implementations is often quite difficult \cite{itg2}.}
-The core ICB loop appears in {\tt time\_travel()} in {\tt landslide.c}.
-%(although not explicitly structured as a C-style loop in the code)
+
+Therefore, Landslide's ICB implementation combines with DPOR
+when tagging new branches to explore at the end of each branch:
+just as DPOR skips alternate interleavings that are memory-independent,
+ICB further filters interleavings requiring more preemptions than the current bound
+out of the to-explore set.
+The macro {\tt ICB\_BLOCKED},
+defined in {\tt schedule.h},
+decides if a given thread would require a preemption beyond the current bound to switch to.
+The DPOR implementation then checks, for some $I_{ij}$ it wants to mark for exploration,
+whether {\tt ICB\_BLOCKED(}$T_j${\tt )} at the state after $t_i$,
+and skips it if so ({\tt tag\_good\_sibling()}/{\tt tag\_all\_siblings()}).
+\footnote{Since ``voluntary'' context switches (e.g. arising from {\tt yield()})
+are often necessary for correct execution,
+{\tt ICB\_BLOCKED} does not count such switches towards the preemption count.
+Therefore, within a certain preemption bound $b$,
+interleavings with more than $b$ context switches may still be tested.}
+
+Then, the entire state space is repeated with increasing bound until no such are filtered.
+This core ICB loop appears in {\tt time\_travel()} in {\tt landslide.c}.
+Although not explicitly structured as a C-style loop in the code,
+it resets Landslide's progress through the state space,
+allowing exploration to continue until it finally observes all interleavings to have fewer preemptions than the bound.
 
@@
 
-Secondly, prior work \cite{bpor} has shown that
+Prior work \cite{bpor} has shown that
 when combined with DPOR to prune equivalent interleavings,
@@
 If a DPOR implementation na\"ively configured with ICB simply skipped $I$ on account of the preemption bound,
-there's no guarantee $J$ would get marked for exploration
-from any other iteration and/or any other pair of conflicting transitions.
+$J$ may not get marked for exploration
+from any other iteration and/or pair of conflicting transitions.
 Even though restricting the state space to a certain maximum preemption bound is already unsound
-in terms of losing full concurrency coverage,
+in terms of losing full interleaving coverage,
 failing to test even $J$ would be a failure of DPOR itself
-to soundly prune the already-reduced state space defined by the bound.
+to soundly prune the already-reduced state space defined by that bound.
 Hence the need for BPOR, to ensure that if such an alternative $J$ to $I$ exists within the bound,
-$J$ gets marked for exploration immediately.
-
-%I refer the reader interested in the algorithm's details to the citation;
-%this section will discuss Landslide's implementation.
-% TODO: say how u do it
+it gets marked for exploration immediately.
+
+To implement BPOR,
+whenever {\tt ICB\_BLOCKED} causes DPOR to skip an interleaving,
+Landslide searches all transitions $t_k \in S$
+such that $t_k \prec_S t_i$ and $T(t_k) = T(t_i)$ and
+$\neg$\{$\exists t_l \in S$ such that $t_k \preceq_S t_l \prec_S t_i$ and $T(t_l) = T_j$\}
+({\tt stop\_bpor\_backtracking()}).
+All $I_{kj}$s which can be tested within the preemption bound are marked instead of $I_{ij}$
+({\tt tag\_reachable\_aunts()}).
+The reader interested in further algorithm details and the corresponding soundness proof is referred to \cite{bpor}.
 

### 2018-03-10 thesis/education.tex
fix incomplete references/labels/citations, move some chapters to their own files

--- 2018-02-10 thesis/thesis.tex
@@
-%for a more compact document, add the option openany to avoid
-%starting all chapters on odd numbered pages
-\documentclass[12pt]{cmuthesis}
+\chapter{Education}
+\label{chap:education}
 
-% This is a template for a CMU thesis.  It is 18 pages without any content :-)
-% The source for this is pulled from a variety of sources and people.
-% Here's a partial list of people who may or may have not contributed:
-%
-%        bnoble   = Brian Noble
-%        caruana  = Rich Caruana
-%        colohan  = Chris Colohan
-%        comar    = Cyrus Omar
-%        jab      = Justin Boyan
-%        josullvn = Joseph O'Sullivan
-%        jrs      = Jonathan Shewchuk
-%        kosak    = Corey Kosak
-%        mjz      = Matt Zekauskas (mattz@cs)
-%        pdinda   = Peter Dinda
-%        pfr      = Patrick Riley
-%        dkoes = David Koes (me)
-
-% My main contribution is putting everything into a single class files and small
-% template since I prefer this to some complicated sprawling directory tree with
-% makefiles.
-
-% some useful packages
-\usepackage{fullpage}
-\usepackage{graphicx}
-\usepackage{amsmath}
-\usepackage{amssymb}
-\usepackage[numbers,sort]{natbib}
-\usepackage[backref,pageanchor=true,plainpages=false, pdfpagelabels, bookmarks,bookmarksnumbered,
-%pdfborder=0 0 0,  %removes outlines around hyper links in online display
-]{hyperref}
-\usepackage{subfigure}
-\usepackage{xspace}
-
-%% custom fonts
-
-\usepackage{fontspec}
-\usepackage{xeCJK}
-
-\setmainfont{XCharter}
-%\setmainfont{Utopia}[
-%	Path = ./fonts/,
-%	UprightFont = UtopiaStd-Regular.otf,
-%	BoldFont = UtopiaStd-Semibold.otf,
-%	ItalicFont = UtopiaStd-Italic.otf,
-%	BoldItalicFont = UtopiaStd-SemiboldIt.otf
-%	]
-
-\setCJKmainfont{HeiseiMinStd-W5}[Path = ./fonts/]
-
-% Approximately 1" margins, more space on binding side
-%\usepackage[letterpaper,twoside,vscale=.8,hscale=.75,nomarginpar]{geometry}
-%for general printing (not binding)
-\usepackage[letterpaper,twoside,vscale=.8,hscale=.75,nomarginpar,hmarginratio=1:1]{geometry}
-
-% Provides a draft mark at the top of the document.
-\draftstamp{\today}{DRAFT}
-
-\begin{document}
-\frontmatter
-
-%initialize page style, so contents come out right (see bot) -mjz
-\pagestyle{empty}
-
-\title{ %% {\it \huge Thesis Proposal}\\
-{\bf Practical Concurrency Testing} \\
-\normalsize \vspace{1em}
-\begin{tabular}{c}
-{\em or, How I Learned to Stop Worrying and Love the Exponential Explosion}
-\end{tabular}}
-\author{Ben Blum}
-\date{Decembruary 2018}
-\Year{2018}
-\trnumber{}
-
-\committee{
-Garth Gibson, Chair \\
-David A. Eckhardt \\
-Brandon Lucia \\
-Haryadi Gunawi, University of Chicago
-}
-
-\support{}
-\disclaimer{}
-
-% copyright notice generated automatically from Year and author.
-% permission added if \permission{} given.
-
-\keywords{landslide terminal, baggage claim, ground transportation, ticketing}
-
-\maketitle
-
-\begin{dedication}
-For Elsie, Margie, Johnny, Mel, Eve, and David.
-%, who brought me here today.
-\end{dedication}
-
-\pagestyle{plain} % for toc, was empty
-
-%% Obviously, it's probably a good idea to break the various sections of your thesis
-%% into different files and input them into this file...
-
-\begin{abstract}
-Concurrent programming presents a challenge to students and experts alike because of the complexity of multithreaded interactions and the difficulty to reproduce and reason about bugs.
-Stateless model checking is a concurrency testing approach which forces a program to interleave its threads in many different ways, checking for bugs each time.
-This technique is powerful, in principle capable of finding any nondeterministic bug in finite time,
-but suffers from exponential explosion as program size increases.
-Checking an exponential number of thread interleavings is not a practical or predictable approach for programmers to find concurrency bugs before their project deadlines.
-
-In this thesis, I develop several new techniques to make stateless model checking more practical for human use.
-I have built Landslide, a stateless model checker specializing in undergraduate operating systems class projects.
-Landslide extends the traditional model checking algorithm with a new framework for automatically managing
-%the exploration of
-multiple state spaces according to their estimated completion times,
-which I show quickly finds bugs should they exist and also quickly verifies correctness otherwise.
-I evaluate Landslide's suitability for inexpert use by presenting the results of many semesters providing it to students in 15-410, CMU's Operating System Design and Implementation class,
-and more recently, students in similar classes at the University of Chicago and Berkeley.
-%Finally, I will present several new techniques that allow stateless model checking to be practically employed on real-world programs.
-%Finally, I will explore broader impact by extending Landslide to test some real-world programs and to be used by students at other universities.
-Finally, I extend Landslide with a new concurrency model for hardware transactional memory,
-and evaluate several real-world transactional benchmarks to show that stateless model checking can keep up with the developing concurrency demands of real-world programs.
-\end{abstract}
-
-\begin{acknowledgments}
-% things to thank
-% stimhack
-% csd bgames
-% undergrad frenz
-% new frenz
-% 410 ta community
-% emarhavil
-%
-% \\
-
-I thank my dog, Louie.
-\end{acknowledgments}
-
-\tableofcontents
-% TODO: figure out how to make latex abbreviate these
-\listoffigures
-\listoftables
-
-\mainmatter
-
-%% Double space document for easy review:
-%\renewcommand{\baselinestretch}{1.66}\normalsize
-
-% The other requirements Catherine has:
-%
-%  - avoid large margins.  She wants the thesis to use fewer pages,
-%    especially if it requires colour printing.
-%
-%  - The thesis should be formatted for double-sided printing.  This
-%    means that all chapters, acknowledgements, table of contents, etc.
-%    should start on odd numbered (right facing) pages.
-%
-%  - You need to use the department standard tech report title page.  I
-%    have tried to ensure that the title page here conforms to this
-%    standard.
-%
-%  - Use a nice serif font, such as Times Roman.  Sans serif looks bad.
-%
-% Other than that, just make it look good...
-
-\newcommand\simics{\textsc{Simics}}
-\newcommand{\sect}[1]{\S #1}
-\newcommand\hilight[2]{\color{#1}#2\color{black}}
-\definecolor{orange}{RGB}{192,96,0}
-\definecolor{olivegreen}{RGB}{0,127,0}
-\definecolor{brickred}{RGB}{192,0,0}
-\definecolor{commentblue}{RGB}{0,0,192}
-
-\newtheorem{theorem}{Theorem}
-
-\newcommand\llama[1]{\includegraphics[width=#1]{llama.pdf}}
-\newcommand\llitem{\item[\llama{1.2em}]}
-
-\newcommand{\inspirationallinebreak}{\vspace{0.25em}}
-\newcommand{\inspirationalhyphen}{-\hspace{-0.15em}-\xspace}
-\newcommand\inspirationalquote[2]{\begin{flushright}
-	\vspace{-1em}
-	{
-	\fontspec[Path=./fonts/]{AlexaStd}
-	{\em #1}
-	\inspirationallinebreak
-
-	-\hspace{-0.15em}-\hspace{-0.15em}-{#2}
-	}
-	\vspace{2em}
-\end{flushright}}
-
-\input{intro}
-\input{overview}
-\input{landslide}
-
-\chapter{Quicksand}
-%\inspirationalquote{
-%\begin{tabular}{p{0.55\textwidth}}
-%There are awful, horrible things in this world; I know that now.
-%But there are a lot of things worth protecting, too.
-%\end{tabular}}
-%{Kaname Madoka, Mahou Shoujo Madoka{\raisebox{0.1em}{$\scriptstyle \bigstar$}}Magica}
-\inspirationalquote{Always, somewhere, someone is fighting for you.
-As long as you remember her, you are not alone.}
-{Mahou Shoujo Madoka{\raisebox{0.1em}{$\scriptstyle \bigstar$}}Magica}
-
-\chapter{Education}
 \inspirationalquote{Knowing the students might one day
@@
 
-\chapter{Related Work}
-\inspirationalquote{
-\begin{tabular}{p{0.58\textwidth}}
-To test if your paper makes a genuine contribution to its discipline,
-see if you can afford a generous tone in the "Related Work" section.
-\end{tabular}}
-{Conor McBride}
-
-\chapter{Future Work}
-% TODO: ask for permission for this one
-% https://twitter.com/Poem4your_sprog/status/878747092788838400
-\inspirationalquote{
-\begin{tabular}{p{0.54\textwidth}}
-A little bit of being kind,
-a tiny open door; \\
-A nicer slice of peace of mind
-begets a little more. \\
-It won't correct the world tonight,
-nor change tomorrow too; \\
-But maybe if you do it right,
-you'll find it changes *you*.
-\end{tabular}}
-{Sam "Poem\_for\_your\_sprog" Garland}
-
-\chapter{Conclusion}
-\inspirationalquote{
-	\begin{tabular}{p{0.84\textwidth}}
-	"If you are a god, Zeus, as the stories claim, then why did you create evolution?
-	Why did you make a world that can only grow through cruelty and pain?"
-	\\
-	For a moment\inspirationalhyphen{}surely this meant death was near\inspirationalhyphen{}she thought she heard him answer:
-	"My child, I was shaped by the gods that came before me, as you were shaped by me.
-	The choice I had was between creation and oblivion, life and death.
-	And I chose life, because any life is better than no life,
-	because as long as there is life, there is hope\inspirationalhyphen{}if not for us, then for some generation to come."
-	\\
-	"Then how are you a god, if you can offer so little?" she whispered, feeling death creep closer.
-	\\
-	"I am a god because I take upon myself the burden of creation," the statue replied.
-	\\
-	"Then we are all gods," Alexandra said, and pushed the button.
-	\end{tabular}}
-{Galatea, The Talos Principle}
-
-asdfasdasdf
-%\appendix
-%\include{appendix}
-
-\backmatter
-
-%\renewcommand{\baselinestretch}{1.0}\normalsize
-
-% By default \bibsection is \chapter*, but we really want this to show
-% up in the table of contents and pdf bookmarks.
-\renewcommand{\bibsection}{\chapter{\bibname}
-\inspirationalquote{Only a doctor of philosophy, Darth.}{Robert Marsh}
-}
-%\renewcommand{\bibpreamble}{This text goes between the ``Bibliography''
-%  header and the actual list of references}
-\bibliographystyle{plainnat} % plain n'at
-\bibliography{citations}
-
-\end{document}
+% TODO

### 2018-03-10 thesis/intro.tex
fix incomplete references/labels/citations, move some chapters to their own files

@@
 		I will show that Iterative Deepening of preemption points outperforms prior work in terms both of finding bugs quickly and of completely verifying correctness when no bug exists.
-	\item {\bf Educational use (Chapters \ref{chap:410} and \ref{chap:pintos}).}
+	\item {\bf Educational use (Chapter \ref{chap:education}).}
 		For the past five semesters, I have offered a fully-automated version of Landslide to students in 15-410, CMU's undergraduate Operating System Design and Implementation class \cite{kspec,thrlib}, for use as a debugging aid during the thread library project.

### 2018-03-10 thesis/landslide.tex
fix incomplete references/labels/citations, move some chapters to their own files

@@
 \chapter{Landslide}
+\label{chap:landslide}
+
 \inspirationalquote{Somewhere is the promise of an uncharted trail, with 700 branching limbs and 700 ways to fail.}
@@
 % TODO: make more specific section reference
-Separate user guides also exist, described in Chapter~\ref{chap:410}.
+Separate user guides also exist, described in Chapter~\ref{chap:education}.
 
@@
 		and if so what {\tt xabort} codes are possible and/or already explored for this transaction
-		(\sect{\ref{chap:txn}}).
+		(\sect{\ref{chap:tm}}).
 \end{itemize}
@@
 for DPOR (\sect{\ref{sec:landslide-dpor}})
-and data race analysis (\sect{\ref{sect:landslide-datarace}}).
+and data race analysis (\sect{\ref{sec:landslide-datarace}}).
 
@@
 \subsection{Pintos-specific features}
-\label{sec:landlside-pintosspecifics}
+\label{sec:landslide-pintosspecifics}
 

### 2018-03-10 thesis/quicksand.tex
fix incomplete references/labels/citations, move some chapters to their own files

--- 2018-03-01 thesis/thesis.tex
@@
-%for a more compact document, add the option openany to avoid
-%starting all chapters on odd numbered pages
-\documentclass[12pt]{cmuthesis}
-
-% This is a template for a CMU thesis.  It is 18 pages without any content :-)
-% The source for this is pulled from a variety of sources and people.
-% Here's a partial list of people who may or may have not contributed:
-%
-%        bnoble   = Brian Noble
-%        caruana  = Rich Caruana
-%        colohan  = Chris Colohan
-%        comar    = Cyrus Omar
-%        jab      = Justin Boyan
-%        josullvn = Joseph O'Sullivan
-%        jrs      = Jonathan Shewchuk
-%        kosak    = Corey Kosak
-%        mjz      = Matt Zekauskas (mattz@cs)
-%        pdinda   = Peter Dinda
-%        pfr      = Patrick Riley
-%        dkoes = David Koes (me)
-
-% My main contribution is putting everything into a single class files and small
-% template since I prefer this to some complicated sprawling directory tree with
-% makefiles.
-
-% some useful packages
-\usepackage{fullpage}
-\usepackage{graphicx}
-\usepackage{amsmath}
-\usepackage{amssymb}
-\usepackage[numbers,sort]{natbib}
-\usepackage[backref,pageanchor=true,plainpages=false, pdfpagelabels, bookmarks,bookmarksnumbered,
-%pdfborder=0 0 0,  %removes outlines around hyper links in online display
-]{hyperref}
-\usepackage{subfigure}
-\usepackage{xspace}
-\usepackage[linesnumbered,ruled]{algorithm2e}
-\usepackage{algpseudocode}
-
-%% custom fonts
-
-\usepackage{fontspec}
-\usepackage{xeCJK}
-
-\setmainfont{XCharter}
-%\setmainfont{Utopia}[
-%	Path = ./fonts/,
-%	UprightFont = UtopiaStd-Regular.otf,
-%	BoldFont = UtopiaStd-Semibold.otf,
-%	ItalicFont = UtopiaStd-Italic.otf,
-%	BoldItalicFont = UtopiaStd-SemiboldIt.otf
-%	]
-
-\setCJKmainfont{HeiseiMinStd-W5}[Path = ./fonts/]
-
-% Approximately 1" margins, more space on binding side
-%\usepackage[letterpaper,twoside,vscale=.8,hscale=.75,nomarginpar]{geometry}
-%for general printing (not binding)
-\usepackage[letterpaper,twoside,vscale=.8,hscale=.75,nomarginpar,hmarginratio=1:1]{geometry}
-
-% Provides a draft mark at the top of the document.
-\draftstamp{\today}{DRAFT}
-
-\begin{document}
-\frontmatter
-
-%initialize page style, so contents come out right (see bot) -mjz
-\pagestyle{empty}
-
-\title{ %% {\it \huge Thesis Proposal}\\
-{\bf Practical Concurrency Testing} \\
-\normalsize \vspace{1em}
-\begin{tabular}{c}
-{\em or, How I Learned to Stop Worrying and Love the Exponential Explosion}
-\end{tabular}}
-\author{Ben Blum}
-\date{Decembruary 2018}
-\Year{2018}
-\trnumber{}
-
-\committee{
-Garth Gibson, Chair \\
-David A. Eckhardt \\
-Brandon Lucia \\
-Haryadi Gunawi, University of Chicago
-}
-
-\support{}
-\disclaimer{}
-
-% copyright notice generated automatically from Year and author.
-% permission added if \permission{} given.
-
-\keywords{landslide terminal, baggage claim, ground transportation, ticketing}
-
-\maketitle
-
-\begin{dedication}
-For Elsie, Margie, Johnny, Mel, Eve, and David.
-%, who brought me here today.
-\end{dedication}
-
-\pagestyle{plain} % for toc, was empty
-
-%% Obviously, it's probably a good idea to break the various sections of your thesis
-%% into different files and input them into this file...
-
-\begin{abstract}
-Concurrent programming presents a challenge to students and experts alike because of the complexity of multithreaded interactions and the difficulty to reproduce and reason about bugs.
-Stateless model checking is a concurrency testing approach which forces a program to interleave its threads in many different ways, checking for bugs each time.
-This technique is powerful, in principle capable of finding any nondeterministic bug in finite time,
-but suffers from exponential explosion as program size increases.
-Checking an exponential number of thread interleavings is not a practical or predictable approach for programmers to find concurrency bugs before their project deadlines.
-
-In this thesis, I develop several new techniques to make stateless model checking more practical for human use.
-I have built Landslide, a stateless model checker specializing in undergraduate operating systems class projects.
-Landslide extends the traditional model checking algorithm with a new framework for automatically managing
-%the exploration of
-multiple state spaces according to their estimated completion times,
-which I show quickly finds bugs should they exist and also quickly verifies correctness otherwise.
-I evaluate Landslide's suitability for inexpert use by presenting the results of many semesters providing it to students in 15-410, CMU's Operating System Design and Implementation class,
-and more recently, students in similar classes at the University of Chicago and Berkeley.
-%Finally, I will present several new techniques that allow stateless model checking to be practically employed on real-world programs.
-%Finally, I will explore broader impact by extending Landslide to test some real-world programs and to be used by students at other universities.
-Finally, I extend Landslide with a new concurrency model for hardware transactional memory,
-and evaluate several real-world transactional benchmarks to show that stateless model checking can keep up with the developing concurrency demands of real-world programs.
-\end{abstract}
-
-\begin{acknowledgments}
-% things to thank
-% stimhack
-% csd bgames
-% undergrad frenz
-% new frenz
-% 410 ta community
-% emarhavil
-%
-% \\
-
-I thank my dog, Louie.
-\end{acknowledgments}
-
-\tableofcontents
-% TODO: figure out how to make latex abbreviate these
-\listoffigures
-\listoftables
-
-\mainmatter
-
-%% Double space document for easy review:
-%\renewcommand{\baselinestretch}{1.66}\normalsize
-
-% The other requirements Catherine has:
-%
-%  - avoid large margins.  She wants the thesis to use fewer pages,
-%    especially if it requires colour printing.
-%
-%  - The thesis should be formatted for double-sided printing.  This
-%    means that all chapters, acknowledgements, table of contents, etc.
-%    should start on odd numbered (right facing) pages.
-%
-%  - You need to use the department standard tech report title page.  I
-%    have tried to ensure that the title page here conforms to this
-%    standard.
-%
-%  - Use a nice serif font, such as Times Roman.  Sans serif looks bad.
-%
-% Other than that, just make it look good...
-
-\newcommand\simics{\textsc{Simics}}
-\newcommand{\sect}[1]{\S #1}
-\newcommand\hilight[2]{\color{#1}#2\color{black}}
-\definecolor{orange}{RGB}{192,96,0}
-\definecolor{olivegreen}{RGB}{0,127,0}
-\definecolor{brickred}{RGB}{192,0,0}
-\definecolor{commentblue}{RGB}{0,0,192}
-
-\newtheorem{theorem}{Theorem}
-
-\newcommand\llama[1]{\includegraphics[width=#1]{llama.pdf}}
-\newcommand\llitem{\item[\llama{1.2em}]}
-
-\newcommand{\inspirationallinebreak}{\vspace{0.25em}}
-\newcommand{\inspirationalhyphen}{-\hspace{-0.15em}-\xspace}
-\newcommand\inspirationalquote[2]{\begin{flushright}
-	\vspace{-1em}
-	{
-	\fontspec[Path=./fonts/]{AlexaStd}
-	{\em #1}
-	\inspirationallinebreak
-
-	-\hspace{-0.15em}-\hspace{-0.15em}-{#2}
-	}
-	\vspace{2em}
-\end{flushright}}
-
-\input{intro}
-\input{overview}
-\input{landslide}
-
 \chapter{Quicksand}
+\label{chap:quicksand}
 \inspirationalquote{
@@
 
-\chapter{Education}
-\inspirationalquote{Knowing the students might one day
-%find a way to
-fix their concurrency bugs...
-it fills you with determination.}{Undertale (paraphrased)}
+% TODO
 
-\chapter{Transactions}
-%\inspirationalquote{
+\section{Preemption points} % ??? -- label needed from reference to LAST_CALL in landslide.tex
+\label{sec:quicksand-pps}
 
-\chapter{Related Work}
-\inspirationalquote{
-\begin{tabular}{p{0.58\textwidth}}
-To test if your paper makes a genuine contribution to its discipline,
-see if you can afford a generous tone in the "Related Work" section.
-\end{tabular}}
-{Conor McBride}
-
-\chapter{Future Work}
-% TODO: ask for permission for this one
-% https://twitter.com/Poem4your_sprog/status/878747092788838400
-\inspirationalquote{
-\begin{tabular}{p{0.54\textwidth}}
-A little bit of being kind,
-a tiny open door; \\
-A nicer slice of peace of mind
-begets a little more. \\
-It won't correct the world tonight,
-nor change tomorrow too; \\
-But maybe if you do it right,
-you'll find it changes *you*.
-\end{tabular}}
-{Sam "Poem\_for\_your\_sprog" Garland}
-
-\chapter{Conclusion}
-\inspirationalquote{
-	\begin{tabular}{p{0.84\textwidth}}
-	"If you are a god, Zeus, as the stories claim, then why did you create evolution?
-	Why did you make a world that can only grow through cruelty and pain?"
-	\\
-	For a moment\inspirationalhyphen{}surely this meant death was near\inspirationalhyphen{}she thought she heard him answer:
-	"My child, I was shaped by the gods that came before me, as you were shaped by me.
-	The choice I had was between creation and oblivion, life and death.
-	And I chose life, because any life is better than no life,
-	because as long as there is life, there is hope\inspirationalhyphen{}if not for us, then for some generation to come."
-	\\
-	"Then how are you a god, if you can offer so little?" she whispered, feeling death creep closer.
-	\\
-	"I am a god because I take upon myself the burden of creation," the statue replied.
-	\\
-	"Then we are all gods," Alexandra said, and pushed the button.
-	\end{tabular}}
-%{Galatea, The Talos Principle}
-{The Talos Principle}
-
-asdfasdasdf
-%\appendix
-%\include{appendix}
-
-\backmatter
-
-%\renewcommand{\baselinestretch}{1.0}\normalsize
-
-% By default \bibsection is \chapter*, but we really want this to show
-% up in the table of contents and pdf bookmarks.
-\renewcommand{\bibsection}{\chapter{\bibname}
-\inspirationalquote{Only a doctor of philosophy, Darth.}{Robert Marsh}
-}
-%\renewcommand{\bibpreamble}{This text goes between the ``Bibliography''
-%  header and the actual list of references}
-\bibliographystyle{plainnat} % plain n'at
-\bibliography{citations}
-
-\end{document}
+\section{Evaluation}
+\label{sec:quicksand-eval}

### 2018-03-10 thesis/tm.tex
fix incomplete references/labels/citations, move some chapters to their own files

\chapter{Transactions}
\label{chap:tm}

%\inspirationalquote{

% notes to self abour marios tests
% counter: tests k threads on n iterations, benchmark xchg-aways vs txn. removed benchmarking, added assertion that no counts are lost.

% TODO

### 2018-04-03 landslide-lecture/psu/lecture.tex
revise slides for psu suitability

--- 2018-02-16 landslide-lecture/2018/lecture.tex
@@
 \usepackage{colortbl}
+\usepackage{xspace}
 \setbeamertemplate{navigation symbols}{}
 
-\title[Landslide]{{\bf Landslide: A New Race-Finding Tool for 15-410}}
+\newcommand\classname{CMPSC 473\xspace}
+\title[Landslide]{{\bf Landslide: Concurrency Bug-Finding for Pebbles}}
 \subtitle[]{ {\em more clever than } ``\texttt{agility\_drill}'' {\em since 2011.}}
@@
 
-\institute[CMU 15-410]{Carnegie Mellon University - 15-410}
-\date[]{2018, February 16}
+\institute[PSU \classname]{Penn State University - \classname}
+\date[]{2018, April 3}
 
@@
 \begin{document}
-\renewcommand{\inserttotalframenumber}{39}
+\renewcommand{\inserttotalframenumber}{42}
 \normalem
@@
 	\begin{itemize}
-		\item Case study (example)
+		\item Examples
 		\item Tabular execution traces
@@
 	\begin{itemize}
-		\item How it works
 		\item Automatically choosing preemption points
+		\item Demo (the fun part)
 		\item User study (that's you!)
@@
 
-\begin{frame}{Case Study}
+% buggy atomics
+
+\begin{frame}{Case Study 1}
+	\begin{center}
+			\fbox{
+			\begin{tabular}{l}
+				libatomic/atomic.S \\
+				\hline
+				\\
+				\texttt{/* int atomic\_fetch\_add(volatile int* obj, int arg); */} \\
+				\texttt{.global atomic\_fetch\_add} \\
+				\texttt{atomic\_fetch\_add:} \\
+				\texttt{~~mov 4(\%esp), \%ecx~~~~~~// obj} \\
+				\texttt{~~mov 8(\%esp), \%edx~~~~~~// arg} \\
+				\texttt{~~mov~~(\%ecx), \%eax~~~~~~// return old value of obj} \\
+				\texttt{~~lock xadd \%edx, (\%ecx) // atomically increment obj} \\
+				\texttt{~~ret} \\
+			\end{tabular}
+			}
+		\pause
+		\linegap
+
+		\begin{itemize}
+			\item Loading {\tt \%ecx} separately breaks atomicity!
+			\item Two threads can see the same return value.
+		\end{itemize}
+	\end{center}
+\end{frame}
+\begin{frame}{Thread Interleavings (``good'' case)}
+	\begin{center}
+		\begin{tabular}{|l|l|}
+			\hline
+			\cellcolor{thread1} {\bf Thread 1} & \cellcolor{thread2} {\bf Thread 2} \\
+			\hline
+			\texttt{mov 4(\%esp), \%ecx} & \\
+			\texttt{mov 8(\%esp), \%edx} & \\
+			\texttt{mov~~(\%ecx), \%eax} & \\
+			\texttt{lock xadd \%edx, (\%ecx)} & \\
+			\texttt{ret} & \\
+			& \texttt{mov 4(\%esp), \%ecx} \\
+			& \texttt{mov 8(\%esp), \%edx} \\
+			& \texttt{mov~~(\%ecx), \%eax} \\
+			& \texttt{lock xadd \%edx, (\%ecx)} \\
+			& \texttt{ret} \\
+			\hline
+		\end{tabular}
+	\end{center}
+\end{frame}
+\begin{frame}{Thread Interleavings (different ``good'' case)}
+	\begin{center}
+		\begin{tabular}{|l|l|}
+			\hline
+			\cellcolor{thread1} {\bf Thread 1} & \cellcolor{thread2} {\bf Thread 2} \\
+			\hline
+			\texttt{mov 4(\%esp), \%ecx} & \\
+			\texttt{mov 8(\%esp), \%edx} & \\
+			& \texttt{mov 4(\%esp), \%ecx} \\
+			& \texttt{mov 8(\%esp), \%edx} \\
+			& \texttt{mov~~(\%ecx), \%eax} \\
+			& \texttt{lock xadd \%edx, (\%ecx)} \\
+			& \texttt{ret} \\
+			\texttt{mov~~(\%ecx), \%eax} & \\
+			\texttt{lock xadd \%edx, (\%ecx)} & \\
+			\texttt{ret} & \\
+			\hline
+		\end{tabular}
+	\end{center}
+\end{frame}
+\begin{frame}{Thread Interleavings (race condition)}
+	\begin{center}
+		\begin{tabular}{|l|l|}
+			\hline
+			\cellcolor{thread1} {\bf Thread 1} & \cellcolor{thread2} {\bf Thread 2} \\
+			\hline
+			\texttt{mov 4(\%esp), \%ecx} & \\
+			\texttt{mov 8(\%esp), \%edx} & \\
+			\texttt{mov~~(\%ecx), \%eax} & \\
+			& \texttt{mov 4(\%esp), \%ecx} \\
+			& \texttt{mov 8(\%esp), \%edx} \\
+			& \texttt{mov~~(\%ecx), \%eax} \\
+			& \texttt{lock xadd \%edx, (\%ecx)} \\
+			& \texttt{ret} \\
+			\texttt{lock xadd \%edx, (\%ecx)} & \\
+			\texttt{ret} & \\
+			\hline
+		\end{tabular}
+	\end{center}
+\end{frame}
+
+% paradise lost
+
+\begin{frame}{Case Study 2}
 	\begin{center}
@@
 		\begin{itemize}
-			\item See {\bf Paradise Lost} lecture!
+			\item See {\bf Synchronization} lecture!
 			\item {\tt if} vs {\tt while}: Two consumers can race to make one fail.
@@
 		\begin{itemize}
-			\item state of the art in 15-410
+			\item state of the art in \classname
 		\end{itemize}
@@
 
-\begin{frame}{Preemption Point Example (remember this?)}
+\begin{frame}{Preemption Point Example ({\tt atomic\_fetch\_add()})}
 	\fbox{
-	\begin{tabular}{l}
-	\texttt{boolean~want[2]~=~\{~false,~false~\};}\\
-	\\
 	\begin{tabular}{rll}
-		1 & \texttt{want[i]~=~true;} & \\
+		\multicolumn{3}{l}{\texttt{/* int atomic\_fetch\_add(volatile int* obj, int arg); */}} \\
+		1 & \texttt{.global atomic\_fetch\_add} \\
+		2 & \texttt{atomic\_fetch\_add:} \\
+		3 & \texttt{~~mov 4(\%esp), \%ecx} \\
 		  & & \footnotesize (preemption point A) \\
-		2 & \texttt{while~(want[j])} & \\
+		4 & \texttt{~~mov 8(\%esp), \%edx} \\
 		  & & \footnotesize (preemption point B) \\
-		3 & \texttt{~~~~~~~~continue;} & \\
+		5 & \texttt{~~mov~~(\%ecx), \%eax} \\
 		  & & \footnotesize (preemption point C) \\
-		4 & \texttt{//~...critical~section...} & \\
+		6 & \texttt{~~lock xadd \%edx, (\%ecx)} \\
 		  & & \footnotesize (preemption point D) \\
-		5 & \texttt{want[i]~=~false;} & \\
-	\end{tabular}
+		7 & \texttt{~~ret} \\
 	\end{tabular}
@@
 
-\begin{frame}{Preemption Point Example (remember this?)}
+\begin{frame}{Preemption Point Example ({\tt atomic\_fetch\_add()})}
 	\fbox{
-	\begin{tabular}{l}
-	\texttt{boolean~want[2]~=~\{~false,~false~\};}\\
-	\\
 	\begin{tabular}{rll}
-		1 & \texttt{want[i]~=~true;} & \\
+		\multicolumn{3}{l}{\texttt{/* int atomic\_fetch\_add(volatile int* obj, int arg); */}} \\
+		1 & \texttt{.global atomic\_fetch\_add} \\
+		2 & \texttt{atomic\_fetch\_add:} \\
+		3 & \texttt{~~mov 4(\%esp), \%ecx} \\
 		  & & \footnotesize (preemption point A) \\
-		2 & \texttt{while~(want[j])} & \\
-		  & & \footnotesize ~ \\
-		3 & \texttt{~~~~~~~~continue;} & \\
-		  & & \footnotesize ~ \\
-		4 & \texttt{//~...critical~section...} & \\
-		  & & \footnotesize ~ \\
-		5 & \texttt{want[i]~=~false;} & \\
-	\end{tabular}
+		4 & \texttt{~~mov 8(\%esp), \%edx} \\
+		  & & \footnotesize (preemption point B) \\
+		5 & \texttt{~~mov~~(\%ecx), \%eax} \\
+		  & & \footnotesize (preemption point C) \\
+		6 & \texttt{~~lock xadd \%edx, (\%ecx)} \\
+		  & & \footnotesize (preemption point D) \\
+		7 & \texttt{~~ret} \\
 	\end{tabular}
@@
 
-	Here, only preemption point A will trigger a deadlock. {\bf ~}\\
+	Here, only preemption point B will break atomicity. {\bf ~}\\
 	All other interleavings are benign.{\bf ~}
@@
 \begin{frame}{About The Project}
-	\textbf{About me}: Final year graduate student, advised by Garth Gibson
-	\begin{itemize}
-		\item TAed 15-410 for 3 semesters during undergrad
+	\textbf{About me}: Final year graduate student at Carnegie Mellon
+	\begin{itemize}
+		\item TAed CMU OS (same thread project) for 3 semesters %during undergrad
 		%\item 1st graduate year was 5th year M.S., rest in Ph.D. program
@@
 		\begin{itemize}
-			\item Master's thesis
+			\item Master's thesis (CMU)
 			\begin{itemize}
@@
 	\begin{itemize}
-		\item Simics module, which traces:
+		\item Runs your code in a simulator (Bochs) which traces:
 			\begin{itemize}
@@
 		\item Each time test completes, identify a PP to replay differently
-		\item Reset machine state and start over
-		\item Implemented using Simics bookmarks
-			\begin{itemize}
-				\item {\tt set-bookmark} and {\tt skip-to}
-			\end{itemize}
+		\item Reset thread library state and start over
+		% \item Implemented with fork() in Bochs
 		\item Replay test from the beginning, with a different interleaving
@@
 		\item Tool must control all sources of nondeterminism
-		\item In 15-410, just timer and keyboard interrupts
+		\item For thread libraries, just timer interrupts
+		%\item In 15-410, just timer and keyboard interrupts
 			% optional - say: "in real kernels, disk/network io interrupts"
@@
 
-\begin{frame}{Landslide \& You}
-	\begin{center}
-	\includegraphics[width=0.85\textwidth]{landslide-new-blank.pdf}
-	\end{center}
-\end{frame}
+%\begin{frame}{Landslide \& You}
+%	\begin{center}
+%	\includegraphics[width=0.85\textwidth]{landslide-new-blank.pdf}
+%	\end{center}
+%\end{frame}
 
@@
 	\begin{itemize}
-		\item Infinite loop (halting problem)
+		\item Infinite loop (see: halting problem)
 		\item Data race bugs
@@
 	\begin{itemize}
-		\item Bakery alg: Is {\tt number[i]=max(number[0],number[1])+1} bad?
+		%\item Bakery alg: Is {\tt number[i]=max(number[0],number[1])+1} bad?
+		\item Atomic load/store: Is {\tt mov \%ecx, \%edx} bad?
 		%\item Is unprotected {\tt foobar\_statistics\_counter++} bad?
@@
 	\begin{itemize}
-		\item E.g., ``all thread API calls'' or ``all kernel mutex locks/unlocks''
+		\item E.g., ``all thread API calls'' or ``all mutex locks/unlocks''
 		\item Depending on length of test, completion time is unpredictable.
@@
 \begin{frame}{Test Suite}
-	Landslide ships with 6 approved test cases:
-	\linegap
-
-	Standard P2 tests
+	Landslide can check the following thread library tests:
+	%Landslide ships with 6 approved test cases:
+	%\linegap
+
+	Atomic tests
+	\begin{itemize}
+		\item {\tt atomic\_fetch\_add}
+		\item (etc...)
+	\end{itemize}
+
+	Locking tests
+	\begin{itemize}
+		\item mutex: {\tt mutex\_test}
+		\item cvars: {\tt broadcast\_test}, {\tt paraguay}
+		\item semaphores: {\tt paradise\_lost}
+		\item rwlocks: {\tt rwlock\_downgrade\_read\_test}
+	\end{itemize}
+
+	Threading tests
 	\begin{itemize}
 		\item {\tt thr\_exit\_join}
-		\item {\tt paraguay}
-		\item {\tt rwlock\_downgrade\_read\_test}
-	\end{itemize}
-
-	New tests
-	\begin{itemize}
-		\item {\tt broadcast\_test}
-		\item {\tt paradise\_lost} %(new this semester)
-		\item {\tt mutex\_test} %(new this semester)
 	\end{itemize}
@@
 	\begin{tabular}{l}
-		\texttt{for (i = 0; i < MAX\_MISBEHAVE /* 64 */; i++) \{} \\
-		\texttt{~~~~misbehave(i);} \\
+		\texttt{for (i = 0; i < 64; i++) \{} \\
+		%\texttt{for (i = 0; i < MAX\_MISBEHAVE /* 64 */; i++) \{} \\
+		%\texttt{~~~~misbehave(i);} \\
 		\texttt{~~~~tid = thr\_create(child\_fn, i);} \\
@@
 
-\begin{frame}{Previous Semesters}
+\begin{frame}{Previous Semesters in CMU's OS class}
 	%% S'15, F'15, S'16: 48 groups signed up to use Landslide; 38 found bugs
@@
 	% S18 lecture - stats up to F17
-	109 {\em deterministic} bugs (e.g. swexn, initialization)
-	\linegap
+
+	%109 {\em deterministic} bugs (e.g. swexn, initialization)
+	%\linegap
 
@@
 \begin{frame}{User Study}
-	\textbf{Try Landslide on your P2!}
+	\textbf{Try Landslide on your project!}
 	\begin{itemize}
@@
 		\begin{itemize}
-			\item Clone a github URL, run setup script, run tests %, answer survey questions
-			\item Landslide will automatically report test results (as described below)
+			\item Clone a github URL, run setup script, run tests, answer survey
+			%\item Landslide will automatically report test results (as described below)
 		\end{itemize}
@@
 	\begin{itemize}
-		\item You {\em must} pass the P2 hurdle before using Landslide.
-		\begin{itemize}
-			\item {\tt startle}, {\tt agility\_drill}, {\tt cyclone}, {\tt join\_specific\_test}, {\tt thr\_exit\_join}
-		\end{itemize}
-		\item Must have {\em attempted} several stress tests
-		\begin{itemize}
-			\item {\tt juggle 4 3 2 0}, {\tt multitest}, {\tt racer} (15 min), {\tt paraguay}
+		\item Atomic tests: Landslide requires working {\tt thr\_create()}.
+		\begin{itemize}
+			\item You must first pass {\tt startle}
+		\end{itemize}
+		\item Locking and threading tests: also need {\tt thr\_exit()} and {\tt thr\_join()}
+		\begin{itemize}
+			\item Must pass {\tt agility\_drill}, {\tt cyclone}, {\tt join\_specific\_test}, {\tt thr\_exit\_join}
+		\end{itemize}
+		%\item You {\em must} pass the P2 hurdle before using Landslide.
+		%\begin{itemize}
+		%	\item {\tt startle}, {\tt agility\_drill}, {\tt cyclone}, {\tt join\_specific\_test}, {\tt thr\_exit\_join}
+		%\end{itemize}
+		\item Recommended to attempt several stress tests, e.g.:
+		\begin{itemize}
+			\item {\tt juggle 4 3 2 0}, {\tt multitest}, {\tt racer}, {\tt paraguay}
 		\end{itemize}
@@
 		\item CMU IRB has approved this study
-		\item Landslide will collect results while you use it
-		\begin{itemize}
-			\item Record commands issued, take snapshots of your P2 code
-			\item All data will be anonymized before publication
-		\end{itemize}
-		\item No coercion: {\bf \em There is no penalty for not participating.}
+		\item I will survey you when the project is done
+		\begin{itemize}
+			\item How many bugs did you find, was it easy to use...
+			\item All survey responses will be anonymized before publication
+		\end{itemize}
+		%\item Landslide will collect results while you use it
+		%\begin{itemize}
+		%	\item Record commands issued, take snapshots of your P2 code
+		%	\item All data will be anonymized before publication
+		%\end{itemize}
+		\item No coercion: {\em There is no penalty for not participating.}
 		\begin{itemize}
@@
 		\item Risk: Landslide may find no bugs and be a waste of your time.
-		\item Benefit: You might learn something...
+		% sorry dave
+		%\item Benefit: You might learn something...
 	\end{itemize}
@@
 		%\item Review this lecture and study information sheet
-		\item Meet prerequisites of passing P2 tests
+		\item Meet prerequisites of passing thread library tests
 		\item Complete sign-up form online to get further instructions
 			\begin{itemize}
-				\item (watch your email for the link)
+				\item (link will be emailed to you soon)
 				%\item \url{http://tinyurl.com/landslide-p2-s17}
 			\end{itemize}
-		\item Optional ``Landslide clinic'' for in-person tech (or moral) support
-			\begin{itemize}
-				\item Next week, room and time TBD
-			\end{itemize}
+		\item Email me with any Landslide-related questions: {\sf bblum@cs.cmu.edu}
+			\begin{itemize}
+				\item OK: ``How do I interpret the html output as a thread interleaving?''
+				\item OK: ``Landslide said it crashed and to email you for tech support!''
+				\item NOT: ``Landslide found a bug, what's wrong with my code?''
+				\item NOT: ``Landslide found a bug, how do I fix it?''
+			\end{itemize}
+		%\item Optional ``Landslide clinic'' for in-person tech (or moral) support
+		%	\begin{itemize}
+		%		\item Next week, room and time TBD
+		%	\end{itemize}
 		%\item Use Landslide in addition to stress tests until P2 is due!
@@
 \section{Bonus Slides}
+
+\breakslide{\Huge Bonus Slides}
 

### 2018-04-04 landslide-lecture/2018/lecture.tex
some lec revisions
fix some errors in lecture

@@
 	\begin{itemize}
-		\item $n$ PPs and $k$ threads $\Rightarrow$ state space size is $O(n^k)$.
+		\item $n$ PPs and $k$ threads $\Rightarrow$ state space size is $O(k^n)$.
 		\item Need to choose PPs very carefully for test to be effective.
@@
 		\begin{itemize}
-			\item Clone a github URL, run setup script, run tests %, answer survey questions
+			\item Clone a github URL, run setup script, run tests, answer survey
 			\item Landslide will automatically report test results (as described below)
@@
 		\end{itemize}
-		\item Must have {\em attempted} several stress tests
-		\begin{itemize}
-			\item {\tt juggle 4 3 2 0}, {\tt multitest}, {\tt racer} (15 min), {\tt paraguay}
+		\item Recommended to attempt several stress tests, e.g.:
+		\begin{itemize}
+			\item {\tt juggle 4 3 2 0}, {\tt multitest}, {\tt racer}, {\tt paraguay}
 		\end{itemize}

### 2018-04-04 landslide-lecture/psu/lecture.tex
fix some errors in lecture

@@
 
-	Here, only preemption point B will break atomicity. {\bf ~}\\
+	Here, only preemption point C will break atomicity. {\bf ~}\\
 	All other interleavings are benign.{\bf ~}
@@
 		%\item Bakery alg: Is {\tt number[i]=max(number[0],number[1])+1} bad?
-		\item Atomic load/store: Is {\tt mov \%ecx, \%edx} bad?
+		\item Atomic load/store: Is {\tt mov \%ecx, (\%edx)} bad?
 		%\item Is unprotected {\tt foobar\_statistics\_counter++} bad?

### 2018-04-04 irb/landslide-guide-psu.tex
landslide guide psu

--- 2018-01-09 irb/landslide-guide-p2.tex
@@
 \usepackage{wrapfig}
+\usepackage{xspace}
 
@@
 
-\title{\bf Landslide: Systematic Testing in 15-410 \\ User Guide}
+\newcommand\classname{CMPSC 473\xspace}
+\title{\bf Landslide: Systematic Testing in \classname \\ User Guide}
 \author{Ben Blum}
@@
 \noindent Likewise, please make sure you have read and understand the lecture slides on systematic testing, which are
-available on the 15-410 course website. If you have any questions about the research background I would be
+available on the \classname course website. If you have any questions about the research background I would be
 happy to answer them.
@@
 
-We encourage you to use Landslide on the GHC cluster or UNIX pool machines ({\tt ghcXX.ghc.andrew.cmu.edu}
-or {\tt unix.andrew.cmu.edu}). Landslide has not been tested with other configurations and will probably fail
+We encourage you to use Landslide on the Vagrant VM or the W204 cluster machines.
+Landslide has not been tested with other configurations and may fail
 mysteriously on your personal laptop, and we cannot promise to be of any help.
@@
 	\item Clone the repository - {\tt git clone https://github.com/bblum/landslide.git} - this will be your workspace.
-	\item Run {\tt ./p2-setup.sh PATH}, where {\tt PATH} is the absolute path to your p2 implementation.
+		On the Vagrant VM, do this in the home directory, {\tt \textasciitilde{}/},
+		{\em not} in the shared host directory {\tt /vagrant}.
+	\item Run {\tt ./psu-setup.sh /vagrant},
+		or if using on the W204 cluster, replace {\tt /vagrant} with the absolute path to your project directory.
 		\begin{itemize}
-			\item This command will import your p2 into {\tt ./pebsim/p2-basecode/} and then attempt to build it.
-				Whenever you update your p2 code, please do so in its original {\tt PATH}
-				and then run {\tt p2-setup.sh} again.
-			\item {\tt PATH} should be the path with {\tt 410kern}, {\tt 410user}, {\tt user/libthread}, etc as subdirectories, not the {\tt user/libthread} directory itself.
+			\item This command will import your project into the {\tt ./pebsim/p2-basecode/} subdirectory of the landslide repository and then attempt to build it.
+				Whenever you update your thread library code, please do so in its original {\tt /vagrant} location
+				and then run {\tt psu-setup.sh} again.
+			%\item {\tt /vagrant} should be the path with {\tt 410kern}, {\tt 410user}, {\tt user/libthread}, etc as subdirectories, not the {\tt user/libthread} directory itself.
 		\end{itemize}
-	\item Run {\tt ./landslide OPTIONS} to run tests with Landslide, for various values of {\tt OPTIONS}:
+	\item Run {\tt ./landslide OPTIONS} to run tests with Landslide, replacing {\tt OPTIONS} with any of the following:
 		\begin{enumerate}
-			\item Use ./landslide -h to view command line options. The most important ones are -t , max testing
-time, and -p , which test program to run.
-			\item Landslide's bug reports will show up as HTML files.
-				\begin{itemize}
-					\item If you are using a cluster machine in person, you can view them by pointing a browser to \\
-						{\tt file:///PATH/TO/landslide/FILENAME.html}
-					\item For remote usage, move the html file to {\tt ~/www} in your AFS, and view it at \\
-						{\tt http://www.contrib.andrew.cmu.edu/~YOUR\_ANDREWID/FILENAME.html}.
-						Note the www in the URL is mandatory;
-						note also that this will allow the world to see it,
-						so be careful to clean up any traces in {\tt ~/www} when you're done!
-				\end{itemize}
+			\item Use ./landslide -h to view command line options. The most important ones are {\tt -t}, max testing
+				time, and {\tt -p}, which test program to run.
 			\item Suggested test configurations to run are as follows.
 				You can, of course, change the time limits or test programs as you see fit.
-				For the longer-running configurations,
-				we recommend running them inside of a {\tt screen} or {\tt tmux} session.
+				%For the longer-running configurations,
+				%we recommend running them inside of a {\tt screen} or {\tt tmux} session.
+				\begin{enumerate}
+					\item Atomic tests
+					\begin{itemize}
+						\item \texttt{./landslide -t 30m -p atomic\_fetch\_add}
+						\item \texttt{./landslide -t 30m -p atomic\_fetch\_sub}
+						\item \texttt{./landslide -t 30m -p atomic\_exchange}
+						\item \texttt{./landslide -t 30m -p atomic\_compare\_swap}
+					\end{itemize}
+					\item Threading tests
 					\begin{itemize}
@@
 					\end{itemize}
-			\item To build your own custom test case for use with Landslide,
-				put it in {\tt pebsim/p2-basecode/user/progs}
+				\item To build your own custom test case, %for use with Landslide,
+					put it in the {\tt pebsim/p2-basecode/user/progs} subdirectory %of the repository
 					and edit the {\tt STUDENTTESTS} line in {\tt pebsim/p2-basecode/config-incomplete.mk}
 					(note: NOT {\tt config.mk}; the setup script auto-generates that file).
-				Then run {\tt ./p2-setup.sh} again.
+					Then run {\tt ./psu-setup.sh} again.
+					Refer to slide 37 of the Landslide lecture to make sure your test is ``Landslide-friendly''.
+				\end{enumerate}
+			\item Landslide's bug reports will show up in the current directory as HTML files which you can view in a web browser.
+				\begin{itemize}
+					\item On the Vagrant VM, move these files to {\tt /vagrant}, the shared directory, and open them with a web browser on the host.
+					\item On the W204 cluster, you should be able to open the files directly from the landslide repository.
+				\end{itemize}
+				Landslide may also print some {\em data races}, even if it doesn't emit an HTML bug report.
+These are not necessarily bugs; please refer to slide 29 of the Landslide lecture for details about what this means.
 		\end{enumerate}
 	\item After fixing any found bugs, re-run Landslide with the same test options to confirm that your fix works.
+		(Remember to re-run {\tt psu-setup.sh} to make sure Landslide sees your updates.)
 \end{enumerate}
@@
 
-\noindent {\bf If something goes wrong:} If you find a Landslide bug, such as an assertion/crash, or such as a
-bug report that you think is actually a bug in Landslide rather than a bug in your P2, or such as it getting stuck
-and not exiting in the given time limit, please create a tarball of your workspace, just as you would to report
-a reference kernel bug, but send it to Ben ({\tt bblum@cs.cmu.edu}) instead of 410 staff. For any other technical
-issues getting it to work, or if you have difficulty understanding a bug report, send email to Ben.
+% TODO
+% \nointend{\bf Note on disk usage:}
+
+\noindent {\bf Asking questions:} If you have any techical questions about Landslide,
+such as how to interpret its html interleaving traces,
+how to write a custom test case, etc.,
+feel free to email me ({\tt bblum@cs.cmu.edu}).
+I will {\em not} provide advice on the thread library project itself,
+nor will I personally look at your code to help you fix your bugs.
+%Once Landslide shows you a specific thread interleaving that led to a bug, the rest is in your hands.
+%However, we cannot provide help on how to understand why your bug happened or how to fix it.
+The point of
+the study is that Landslide's automation plus your own brain should be enough!
+On the other hand, if you have
+a design question while deciding between potential ways to fix a bug, send it to \classname course staff (not me) as usual.
 \\
 
-\noindent However, we cannot provide help on how to understand why your bug happened or how to fix it. The point of
-the study is that Landslide's automation plus your own brain should be enough! On the other hand, if you have
-a design question while deciding between potential ways to fix a bug, send it to 410 course staff as usual.
+\noindent {\bf If something goes wrong:} If you find a Landslide bug, such as an assertion/crash (it will say ``Landslide crashed. This is not your fault.''), or such as a
+bug report that you think is actually a bug in Landslide rather than a bug in your P2, or such as it getting stuck
+and not exiting in the given time limit, please create a tarball of your workspace
+using the command {\tt tar cjvf landslide-crash.tar.bz2 ./},
+%, just as you would to report a reference kernel bug,
+email it to me ({\tt bblum@cs.cmu.edu}), and I'll try to fix your issue promptly.
+%For any other technical
+%issues getting it to work, or if you have difficulty understanding a bug report, send email to .
+\\
 

### 2018-04-04 irb/landslide-guide-psu.tex
update psu guide per timmys commence:

@@
 participant and our confidentiality procedures. If you are missing this, please contact me immediately. My email
-address is {\tt bblum@cs.cmu.edu} and my phone number is 412-304-4294; email is preferred.
+address is {\tt bblum@cs.cmu.edu}.
 \\
@@
 Landslide has not been tested with other configurations and may fail
-mysteriously on your personal laptop, and we cannot promise to be of any help.
+mysteriously on other development environments, and we cannot promise to be of any help.
 
@@
 		\begin{enumerate}
-			\item Use ./landslide -h to view command line options. The most important ones are {\tt -t}, max testing
+			\item Use {\tt ./landslide -h} to view command line options. The most important ones are {\tt -t}, max testing
 				time, and {\tt -p}, which test program to run.
@@
 						\item \texttt{./landslide -t 30m -p rwlock\_downgrade\_read\_test}
+						%	\\
+						%\item \texttt{./landslide -t 3h -p thr\_exit\_join}
+						%\item \texttt{./landslide -t 3h -p mutex\_test}
+						%\item \texttt{./landslide -t 3h -p paradise\_lost}
+						%\item \texttt{./landslide -t 3h -p broadcast\_test}
+						%\item \texttt{./landslide -t 3h -p paraguay}
+						%\item \texttt{./landslide -t 3h -p rwlock\_downgrade\_read\_test}
 							\\
-						\item \texttt{./landslide -t 3h -p thr\_exit\_join}
-						\item \texttt{./landslide -t 3h -p mutex\_test}
-						\item \texttt{./landslide -t 3h -p paradise\_lost}
-						\item \texttt{./landslide -t 3h -p broadcast\_test}
-						\item \texttt{./landslide -t 3h -p paraguay}
-						\item \texttt{./landslide -t 3h -p rwlock\_downgrade\_read\_test}
-							\\
-						\item \texttt{./landslide -t 12h -p thr\_exit\_join}
-						\item \texttt{./landslide -t 12h -p mutex\_test}
-						\item \texttt{./landslide -t 12h -p paradise\_lost}
-						\item \texttt{./landslide -t 12h -p broadcast\_test}
-						\item \texttt{./landslide -t 12h -p paraguay}
-						\item \texttt{./landslide -t 12h -p rwlock\_downgrade\_read\_test}
+						\item \texttt{./landslide -t 8h -p thr\_exit\_join}
+						\item \texttt{./landslide -t 8h -p mutex\_test}
+						\item \texttt{./landslide -t 8h -p paradise\_lost}
+						\item \texttt{./landslide -t 8h -p broadcast\_test}
+						\item \texttt{./landslide -t 8h -p paraguay}
+						\item \texttt{./landslide -t 8h -p rwlock\_downgrade\_read\_test}
 					\end{itemize}
-				\item To build your own custom test case, %for use with Landslide,
-					put it in the {\tt pebsim/p2-basecode/user/progs} subdirectory %of the repository
-					and edit the {\tt STUDENTTESTS} line in {\tt pebsim/p2-basecode/config-incomplete.mk}
-					(note: NOT {\tt config.mk}; the setup script auto-generates that file).
-					Then run {\tt ./psu-setup.sh} again.
-					Refer to slide 37 of the Landslide lecture to make sure your test is ``Landslide-friendly''.
+				%\item To build your own custom test case, %for use with Landslide,
+				%	put it in the {\tt pebsim/p2-basecode/user/progs} subdirectory %of the repository
+				%	and edit the {\tt STUDENTTESTS} line in {\tt pebsim/p2-basecode/config-incomplete.mk}
+				%	(note: NOT {\tt config.mk}; the setup script auto-generates that file).
+				%	Then run {\tt ./psu-setup.sh} again.
+				%	Refer to slide 37 of the Landslide lecture to make sure your test is ``Landslide-friendly''.
 				\end{enumerate}
@@
 				\end{itemize}
-				Landslide may also print some {\em data races}, even if it doesn't emit an HTML bug report.
+				Landslide may also print some {\em data races},
+				even if it doesn't emit an HTML bug report,
+				for example as follows:
+				\[
+					\texttt{Found a racy access at 0x0100291e in atomic\_compare\_swap (atomic.S:42)}
+				\]
 				These are not necessarily bugs; please refer to slide 29 of the Landslide lecture for details about what this means.
@@
 	\item After fixing any found bugs, re-run Landslide with the same test options to confirm that your fix works.
-		(Remember to re-run {\tt psu-setup.sh} to make sure Landslide sees your updates.)
+		(Remember to re-run {\tt psu-setup.sh} as in step 2 to make sure Landslide sees your updates.)
 \end{enumerate}
@@
 it needs to create some temporary files to annotate your code, and multiple instances of Landslide can step on each other's toes by clobbering those files.
-If you want to try multiple tests at once, simply {\tt git clone} the repository again in a different directory, and run one instance of Landslide in each.
+%If you want to try multiple tests at once, simply {\tt git clone} the repository again in a different directory, and run one instance of Landslide in each.
 \\

### 2018-04-06 irb/infosheet-psu.tex
fix some errors in psu info/guide documence

--- 2018-01-09 irb/infosheet-uchicago.tex
@@
 % TODO for berkeley: CS162
-\newcommand\classname{CS 230}
+\newcommand\classname{CMPSC 473}
 \title{\bf Landslide: Systematic Testing in \classname \\ User Study Information Sheet}
@@
 
-If you agree to participate in this study, your TA/course instructor will use Landslide to debug your submitted projects, and you will receive these debugging reports either before or after the project deadline, depending on the project.
-Landslide's testing results will not be used to determine your project grade.
-After you have received at least one debugging report from Landslide, we will ask you to complete a brief survey that asks about your opinions of the Landslide report.
-As part of this research, we (as researchers at Carnegie Mellon) will collect the debugging reports, snapshots of your code at the time the bug was found, and your survey responses.
+If you agree to participate in this study,
+%, your TA/course instructor will use Landslide to debug your submitted projects, and you will receive these debugging reports either before or after the project deadline, depending on the project.
+you will receive access to Landslide to use freely, just as you would any other debuggging tool, for the extent of the project.
+The course staff will not use Landslide's testing results to determine your project grade.
+%After you have received at least one debugging report from Landslide, we will ask you to complete a brief survey that asks about your opinions of the Landslide report.
+As part of this research, we (as researchers at Carnegie Mellon) will ask you to complete a brief survey that asks about your experience using Landslide.
+%the debugging reports, snapshots of your code at the time the bug was found, and
 
@@
 \noindent {\bf Benefits.}
-Because we are testing your code with a more powerful testing tool than is normally distributed with the project, we hope you will be able to fix more concurrency bugs, potentially improving the quality of your submitted projects and resulting in a higher grade.
+Because we are offering a more powerful testing tool than is normally distributed with the project, we hope you will be able to fix more concurrency bugs, potentially improving the quality of your submitted projects and resulting in a higher grade.
 
@@
 During this study we will collect personally identifiable information from you as follows.
-When we run Landslide on your code, we will record the same bug reports we distribute to you, as well as snapshots of your code at the time the bug was found.
-We will also collect your answers to the survey, and in this case we will also keep whatever you write confidential from the OS course staff at your university.
+%When we run Landslide on your code, we will record the same bug reports we distribute to you, as well as snapshots of your code at the time the bug was found.
+We will collect your answers to the survey, and will keep whatever you write confidential from the OS course staff at your university.
+Any survey responses we publish in our research will be anonymized.
 %We will also collect paper surveys which you will fill out while using Landslide.
@@
 
-University of Chicago IRB website: {\tt https://sbsirb.uchicago.edu/}
+% PSU wants this gone!
+%PSU IRB website: {\tt https://www.research.psu.edu/irb}. Email: {\tt irb-orp@psu.edu}.
 
@@
 	%\item {\bf David Eckhardt} (Course Instructor): Email: {\tt de0u@andrew.cmu.edu}; Office: GHC 7717
-	\item {\bf Haryadi Gunawi} (Course Instructor): Email: {\tt haryadi@cs.uchicago.edu}; Office: Ryerson Hall 259
+	% PSU wants this gone!
+	%\item {\bf Timothy Zhu} (Course Instructor): Email: {\tt timothyz@cse.psu.edu}; Office: W353 Westgate Building
 \end{itemize}

### 2018-04-06 irb/landslide-guide-psu.tex
psu user guide update
fix some errors in psu info/guide documence

@@
 				\end{enumerate}
+				During the course of your debugging you may wish to read the tests' source code to see what they're doing. You can find these tests in {\tt pebsim/p2-basecode/410user/progs/}.
 			\item Landslide's bug reports will show up in the current directory as HTML files which you can view in a web browser.
@@
 \noindent {\bf If something goes wrong:} If you find a Landslide bug, such as an assertion/crash (it will say ``Landslide crashed. This is not your fault.''), or such as a
-bug report that you think is actually a bug in Landslide rather than a bug in your P2, or such as it getting stuck
+bug report that you think is actually a bug in Landslide rather than a bug in your project, or such as it getting stuck
 and not exiting in the given time limit, please create a tarball of your workspace

### 2018-04-25 thesis/overview.tex
talk abt psu and pebwine

@@
 
+The Pebbles kernel architecture
+is used at CMU in 15-410, Operating System Design and Education \cite{kspec,thrlib}.
 In the course of a semester, students work on five programming assignments;
@@
 
+\subsubsection{Use at other universities}
+
+\newcommand\psuos{CMPSC 473\xspace}
+
+In the Spring 2018 semester,
+the Operating Systems class at Penn State University (henceforth \psuos and PSU, respectively)
+offered the P2 thread library project as part of its curriculum.
+Students in this class implemented P2 working alone rather than in pairs,
+skip the {\tt swexn} automatic stack growth portion,
+and rather than running their code with a reference Pebbles kernel binary in a simulator,
+use the Pebwine emulation layer \cite{pebwine}
+to run Pebbles-compatible program binaries in the Linux userspace.
+Otherwise, the project is identical to CMU 15-410's P2.
+
 \subsection{Pintos}

### 2018-04-28 thesis/landslide.tex
work on quotes

@@
 \inspirationalquote{Somewhere is the promise of an uncharted trail, with 700 branching limbs and 700 ways to fail.}
-{ThouShaltNot, "Cardinal Directions"}
+{ThouShaltNot, Cardinal Directions}
 

### 2018-04-28 thesis/thesis.tex
fix incomplete references/labels/citations, move some chapters to their own files
work on quotes

@@
 \input{landslide}
-
-\chapter{Quicksand}
-\inspirationalquote{
-% original
-% {\footnotesize 嫌な事も悲しい事もあったけど、守りたい物だって沢山この世界にはあったから…} \\
-% trans. slightly paraphrased from crunchyroll subtitle; removed "..."s, added past tense for あった
-There were awful, sad things in this world.
-But there were a lot of things worth protecting, too.
-}
-{Kaname Madoka, Mahou Shoujo Madoka{\raisebox{0.1em}{$\scriptstyle \bigstar$}}Magica}
-%\inspirationalquote{Always, somewhere, someone is fighting for you.
-%As long as you remember her, you are not alone.}
-%{Mahou Shoujo Madoka{\raisebox{0.1em}{$\scriptstyle \bigstar$}}Magica}
-
-\chapter{Education}
-\inspirationalquote{Knowing the students might one day
-%find a way to
-fix their concurrency bugs...
-it fills you with determination.}{Undertale (paraphrased)}
-
-\chapter{Transactions}
-%\inspirationalquote{
+\input{quicksand}
+\input{education}
+\input{tm}
 
 \chapter{Related Work}
+\label{chap:relatedwork}
 \inspirationalquote{
@@
 \chapter{Future Work}
+\label{chap:warpzone}
 % TODO: ask for permission for this one
@@
 \inspirationalquote{
-\begin{tabular}{p{0.54\textwidth}}
+\begin{tabular}{p{0.51\textwidth}}
 A little bit of being kind,
@@
 But maybe if you do it right,
-you'll find it changes *you*.
+you'll find it changes you.
 \end{tabular}}
@@
 \chapter{Conclusion}
+\label{chap:conclusion}
 \inspirationalquote{

### 2018-05-02 thesis/landslide.tex
landslide preemption pt identification

@@
 
-% TODO: update this linkerino (repository name)
 % TODO: add some backup linkeroos
-The Bochs port of Landslide is likewise open-source and available at \url{https://github.com/bblum/bochs}.
+The Bochs port of Landslide is likewise open-source and available at \url{https://github.com/bblum/landslide}.
 % TODO: refresh
-The HEAD commit at the time of writing is 8984021.
+The HEAD commit at the time of writing is 5d45e2c.
 The Simics port uses Simics's proprietary API and is hence unlicensed and available upon request for educational use only.
@@
 
-Two setup scripts are provided, one for each supported kernel architecture: {\tt p2-setup.sh} and {\tt pintos-setup.sh}.
+Three setup scripts are provided, one for each supported kernel architecture: {\tt p2-setup.sh}, {\tt psu-setup.sh}, and {\tt pintos-setup.sh}.
 The user should supply the directory containing their project implementation.
-The latter script also supports arguments specifying which of the Pintos projects to target.
+The second of the three is largely the same as the first, with CMU-specific project details replaced by PSU-specific ones.
+The latter of the three also supports arguments specifying which of the Pintos projects to target.
 For example:
@@
 	\item {\tt ./p2-setup.sh /path/to/my/p2}
+	\item {\tt ./psu-setup.sh /path/to/my/thrlib}
 	\item {\tt ./pintos-setup.sh /path/to/my/threads} (2nd argument defaults to ``{\tt threads}'')
@@
 	\item {\tt -0}: enable Preempt-Everywhere mode (\sect{\ref{sec:quicksand-eval}}, requires {\tt -C})
+	\item {\tt -M}: enable Maximal State Space mode,
+		which prioritizes the maximal state space to optimize for fast verification,
+		abandoning all subset jobs even if they might find bugs faster
+		% TODO: more speicfic sectio nreference
+		(\sect{\ref{chap:tm}}, incompatible with {\tt -C})
 	\item {\tt -H}: use Limited Happens-Before for data-race analysis (\sect{\ref{sec:background-hb}})
@@
 	\item {\tt -V}: use vector-clock-based Pure Happens-Before for data-race analysis (\sect{\ref{sec:background-hb}})
-		(default for P2 userspace and Pintos modes)
+		(default for P2/PSU userspace and Pintos modes)
 	\item {\tt -X}: support transactional memory (Chapter~\ref{chap:tm})
@@
 	\item {\tt -P}: support Pintos architecture (enabled automatically when {\tt pintos-setup.sh} is run)
-	\item {\tt -4}: support Pebbles architecture (enabled automatically when {\tt p2-setup.sh} is run)
+	\item {\tt -4}: support Pebbles architecture (enabled automatically when either {\tt p2-setup.sh} or {\tt psu-setup.sh} is run)
 		% TODO: put a section reference
@@
 % TODO: check figure placement
-\begin{figure}[t]
+\begin{figure}[h!]
 	\begin{center}
+		% XXX: at 0.986\textwidth or bigger, latex puts this infinity pages into the future
+		% at 0.985 or lower, it goes on the right page
+		% wtffff latex.
 		% nb. generated from landslide-trace-1506524837.14.html
-		\includegraphics[width=\textwidth]{bugreport.pdf}
+		\includegraphics[width=0.985\textwidth]{bugreport.pdf}
 	\end{center}
@@
 and data race analysis (\sect{\ref{sec:landslide-datarace}}).
+For userspace tests, it also tracks which virtual address space ({\tt cr3}) belongs to the test program
+via a state machine of the test lifecycle,
+which lets it avoid false positive heap errors from other programs which have differently-addressed heaps
+({\tt check\_user\_address\_space()} and {\tt ignore\_user\_access()}.
 
@@
 \subsubsection{Stack traces}
+\label{sec:landslide-stacktrace}
 
@@
 \label{sec:landslide-pps}
-% TODO: talk about how w/in func and w/out func are processed
+
+When should Landslide sunder the universe into alternate realities,
+in which each a different thread executes immediately following the current instruction?
+This singular question determines to what extent the state space of possible interleavings explodes exponentially.
+While other parts of the research decide which lock API calls to consider,
+or which memory accesses constitute a data race,
+interpreting those combinations of preemption point {\em predicates}
+to decide if the current program state constitutes a single preemption {\em point}
+warrants discussion.
+
+Preemption point identification is implemented largely in {\tt pp.c}.
+At startup, {\tt pps\_init()} and {\tt load\_dynamic\_pps()} process the
+statically-configured preemption points (\sect{\ref{sec:landslide-staticconfig}})
+and dynamically-configured preemption points (\sect{\ref{sec:landslide-dynamicconfig}}),
+respectively.
+Each of these configurations may contain any number of
+{\tt within\_function}, {\tt without\_function}, and {\tt data\_race} commands.
+
+\subsubsection{Stack trace inclusion/exclusion}
+
+{\tt check\_withins()} implements the whitelist/blacklist behaviour for the former two of those commands
+(in a similar manner to the Preemption Sealing technique described in prior work \cite{sealing}).
+It invokes the stack tracer (\sect{\ref{sec:landslide-stacktrace}})
+for a list of which functions are on the call stack
+(hence the importance of the stack tracer's complex logic to not miss any frames
+even when interrupts or system calls are involved).
+Then, to determine if the current program state should be considered a valid preemption point,
+or whether it should be rejected,
+it compares each {\tt within} or {\tt without\_function} directive in the following sequence-dependent manner:
+
+\begin{itemize}
+	\item If no {\tt within\_function} commands are given, operate in ``blacklist mode'':
+		the preemption point is by default valid as long as no {\tt without\_function} calls reject it.
+		Otherwise, operate in ``whitelist mode'':
+		the preemption point is by default rejected unless at least one {\tt within\_function} directive matches.
+	\item Subject to the above, find the sequentially-last {\tt *\_function} directive
+		(static preemption point commands considered before dynamic ones)
+		which matches any function in the stack trace.
+		If {\tt within}, accept the preemption point; if {\tt without}, reject it.
+\end{itemize}
+
+The same comparison is done for {\tt within\_user\_function} and {\tt without\_user\_function}.
+
+\subsubsection{Data race predicates}
+
+The {\tt data\_race} command specifies an instruction pointer value to identify as a data race preemption point.
+The point can optionally be qualified by a thread ID, most recent system call number, etc.,
+as described in \sect{\ref{sec:landslide-dynamicconfig}},
+and is queried through {\tt suspected\_data\_race()}.
+
+In Preempt-Everywhere mode, instead Landslide marks all shared memory accesses
+as long as they are not either part of the mutex implementation or part of the running thread's stack frame.
+The {\tt data\_race} command is ignored and {\tt suspected\_data\_race()}
+instead checks whether the instruction pointer is associated with any such shared memory access.
+
+\subsubsection{Preemption point predicates}
+
+{\tt arbiter\_interested()} in {\tt arbiter.c} then
+checks various annotations and hard-coded preemption point predicates
+to decide whether the current program state constitutes a preemption point.
+The following predicates are constrained by {\tt check\_withins()}:
+
+\begin{itemize}
+	\item {\tt suspected\_data\_race()}
+	\item User or kernel {\tt mutex\_lock()} or {\tt mutex\_unlock()} call
+	\item Custom kernel preemption point using {\tt tell\_landslide\_preempt()}
+		(relic of \cite{landslide}, largely obsoleted by data-race preemption pooints)
+\end{itemize}
+
+The following predicates ignore any {\tt within\_function} settings
+(mandatory preemption points needed, for example, to maintain the one-thread-per-transition invariant):
+
+\begin{itemize}
+	\item Voluntary reschedule, e.g. explicit {\tt yield()}
+	\item {\tt hlt} instruction (kernel waiting for interrupt)
+	\item User thread becomes yield- or xchg-blocked (\sect{\ref{sec:landslide-blocking}})
+	\item {\tt \_xbegin()} or {\tt \_xend()}, if testing transactional memory (Chapter~\ref{chap:tm})
+\end{itemize}
+
+Whenever {\tt arbiter\_interested()} returns true,
+Landslide creates a new {\tt struct hax} in the execution tree (\sect{\ref{sec:landslide-save}}),
+creates a checkpoint (\sect{\ref{sec:landslide-timetravel}}),
+and queries {\tt arbiter\_choose()} to decide which thread to run next (\sect{\ref{sec:landslide-arbiter}}).
+
+\subsubsection{Example}
+
+Consider the following examples to illustrate the behaviour of the stack trace directives.
+\begin{enumerate}
+	\item
+		{\tt mutex\_lock()}, in
+		{\tt malloc()}, in
+		{\tt thr\_create()}, in
+		{\tt main()}
+	\item
+		{\tt mutex\_lock()}, in
+		{\tt cond\_wait()}, in
+		{\tt thr\_join()}, in
+		{\tt main()}
+\end{enumerate}
+and the following {\tt within}/{\tt without\_user\_function} combinations:
+\begin{itemize}
+	\item
+		{\tt within\_user\_function mutex\_lock},
+		{\tt without\_user\_function malloc}
+
+		Rejects stack trace 1 (last matching directive is to exclude {\tt malloc}),
+		accepts stack trace 2 (last matching directive is to include {\tt mutex\_lock})
+	\item
+		{\tt within\_user\_function thr\_join}
+
+		No {\tt without}s present, so behaves as a whitelist,
+		rejecting stack trace 1 (not in {\tt thr\_join()}),
+		accepting stack trace 2 (in {\tt thr\_join()}).
+	\item
+		{\tt without\_user\_function cond\_wait}
+
+		No {\tt within}s present, so behaves as a blacklist,
+		accepting stack trace 1 (not in {\tt cond\_\allowbreak{}wait()}),
+		rejecting stack trace 2 (in {\tt cond\_wait()}).
+	\item
+		{\tt without\_user\_function main},
+		{\tt within\_user\_function mutex\_lock}
+
+		Accepts both (last matching directive is to accept {\tt mutex\_lock()},
+		regardless of {\tt main()})
+\end{itemize}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \subsection{Heuristic loop, deadlock, and synchronization detection}
+\label{sec:landslide-blocking}

### 2018-05-03 thesis/landslide.tex
describe bochs simics and which universties more accurately

@@
 	\item {\bf Simics} \cite{simics}, a proprietary simulator licensed commercially by Wind River, used at CMU in 15-410 to run Pebbles thread libraries and kernels, and
-	\item {\bf Bochs} \cite{bochs}, an open-source (LGPL) simulator used at the University of Chicago, Berkeley, and other schools to run Pintos kernels.
+	\item {\bf Bochs} \cite{bochs}, an open-source (LGPL) simulator used at the University of Chicago, Berkeley, Stanford, and other schools to run Pintos kernels.
 \end{itemize}
@@
 The Simics port uses Simics's proprietary API and is hence unlicensed and available upon request for educational use only.
+Development on the Simics port is largely frozen,
+as the Bochs port implements all the same features and more,
+and is also roughly 3x faster.
 
@@
 \label{sec:landslide-estimate}
+
+% TODO
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \label{sec:landslide-blocking}
+
+% TODO

### 2018-05-03 thesis/thesis.tex
describe bochs simics and which universties more accurately

@@
 I evaluate Landslide's suitability for inexpert use by presenting the results of many semesters providing it to students in 15-410, CMU's Operating System Design and Implementation class,
-and more recently, students in similar classes at the University of Chicago and Berkeley.
+and more recently, students in similar classes at the University of Chicago and Penn State University.
 %Finally, I will present several new techniques that allow stateless model checking to be practically employed on real-world programs.

### 2018-05-03 thesis/tm.tex
work on quotes
introduce tm

@@
 
-%\inspirationalquote{
+%\inspirationalquote{This paper presents earth-shaking work that deserves to be read by the entire computer science community.
+%To that end, have you considered putting it in a thesis?}
+%{Anonymous SIGBOVIK reviewer}
+
+\inspirationalquote{
+\begin{tabular}{p{0.47\textwidth}}
+One can only build a sand castle where the sand is wet. \\
+But where the sand is wet, the tide comes. \\
+Yet we still build sand castles.
+\end{tabular}}
+{Yuri, Doki Doki Literature Club}
 
@@
 % TODO
+
+Transactional memory (TM) is a concurrency primitive
+by which programmers may attempt an arbitrary sequence of shared memory accesses,
+which will either be committed (made visible to other processors/threads) atomically,
+such that no intermediate state modification is ever visible,
+or in the case of a conflict which would prevent such,
+discarded with an error code returned to allow the programmer to write a synchronized backup path.
+%
+Transactional memory specifications typically have three API functions, abstractly speaking:
+\begin{itemize}
+	\item  {\sf begin} begins a transaction,
+		staging any subsequent shared memory accesses in some temporary thread-local storage,
+		and checking for conflicts with the accesses of any other threads or CPUs.
+		If the transaction is unsuccessful, as described below,
+		{\sf begin} instead returns an error code
+		indicating the programmer should fall back to some other, possibly slower, synchronization method.
+	\item {\sf end} ends a transaction,
+		attempting to commit all staged accesses to the shared memory atomically with respect to
+		reads or writes from other concurrently-executing code.
+		If any of those accesses conflict (i.e., read/write or write/write)
+		with any other access to the same memory since the transaction started,
+		they are instead discarded and execution state reverts to the {\sf begin} with an error code as described above.
+	\item {\sf abort} explicitly aborts a transaction,
+		regardless of any memory conflicts,
+		discarding changes and reverting execution as described above.
+		Some implementations allow an arbitrary abort code to be specified
+		which will appear in {\sf begin}'s error code.
+\end{itemize}
+
+% do they track changes in pthread-style TLS? or?
+{\bf Implementations.}
+Software TM implementations (STM) typically function as a library,
+tracking staged memory accesses in local memory,
+and aborting whenever a conflict is detected between two transactions' tracked accesses \cite{stm-pldi06}.
+Hardware TM implementations (HTM) use processor-level hardware support,
+which stages changes in per-CPU cache lines,
+and may abort for STM's reason above \cite{htm-experience},
+or additionally whenever a conflict is detected between one transaction's traced access and {\em any} other memory access,
+or whenever a conflict occurs on the same cache line, not necessarily the same address,
+or in case of any system interrupt or cache overflow.
+
+{\bf Terminology.}
+The world of hardware transactional memory is home to several more confusing acronyms in addition to ``HTM''.
+Transactional Synchronization Extensions (TSX)
+refers to Intel's implementation of HTM on the Haswell microarchitecture
+\cite{htm-haswell}.
+Restricted Transactional Memory (RTM)
+refers to the {\tt xbegin}, {\tt xend}, and {\tt xabort} subset of TSX instructions,
+which of course correspond to {\sf begin}, {\tt end}, and {\tt abort} listed above,
+and which GCC exposes as C/C++ intrinsices named {\tt \_xbegin()}, {\tt \_xend()}, and {\tt \_xabort()}
+\cite{htm-gcc}.
+Hardware Lock Elision (HLE)
+refers to the {\tt xacquire} and {\tt xrelease} subset of TSX instructions,
+which extend the traditional interface to offer a slightly higher-level way to access the CPU feature,
+optimized for simplicity for locking-like synchronization patterns
+% other possible things to cite here:
+% https://lwn.net/Articles/534758/
+% https://software.intel.com/en-us/node/683688
+\cite{hardware-lock-elision}.
+In this thesis I focus on RTM, the more general (i.e., expressive (i.e., bug-prone)) interface,
+and among all these acronyms restrict myself to ``HTM''
+(when referring to transactional memory as a concurrency primitive in the abstract)
+and ``TSX''
+(when referring to Intel's implementation and/or GCC's intrinsics interface).
+The non-pedantic reader may treat these as interchangeable.
+
+An example TSX program is shown in \sect{\ref{sec:overview-tm}}.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{Concurrency Model}
+
+% TODO

### 2018-05-03 thesis/tm.tex
add tm proofs

@@
 \label{chap:tm}
+
+% copied from sigbovik - mb move into thesis.tex later?
+% threads
+\definecolor{lavender}{RGB}{102,65,208} % (?? i don't know what happened here anymore but it looks good)
+\definecolor{seafoam}{RGB}{56,158,68} % * 2/3
+\definecolor{salmon}{RGB}{208,89,74} % * 7/8
+% eval data
+\definecolor{darkcyan}{RGB}{0,128,128}
+\definecolor{lime}{RGB}{48,128,0}
+\definecolor{darklavender}{RGB}{68,44,138}
+\definecolor{brownish}{RGB}{128,96,0}
+\definecolor{pinkish}{RGB}{128,34,102} % ff4488
+% syntax hilighting
+\newcommand\flow[1]{\hilight{brown}{#1}\xspace}
+\newcommand\const[1]{\hilight{brickred}{#1}\xspace}
+\newcommand\call[1]{\hilight{pinkish}{#1}}
 
@@
 
-An example TSX program is shown in \sect{\ref{sec:overview-tm}}.
+An example TSX program is shown in Figure~\ref{fig:htm-example} (\sect{\ref{sec:overview-tm}}).
 
@@
 \section{Concurrency Model}
+\label{sec:tm-design}
+
+While up to now Landslide's tested programs' concurrency has been limited to timer-driven thread scheduling,
+HTM presents a fundamentally new dimension of nondeterminism,
+namely the hardware's ability to revert execution sequences
+and the delayed visibility of changes to other threads.
+In order to efficiently test HTM programs in Landslide,
+in this section
+I develop a simpler concurrency model and offer a proof of equivalence to HTM execution semantics.
+I make two major simplifications:
+simulating transaction aborts as immediate failure injections,
+and treating transaction atomicity as a global mutex during data-race analysis;
+and provide corresponding equivalence proofs.
+
+{\bf Notation.} Let $I = TN_1@L_1, TN_2@L_2, ... TN_n@L_n$,
+with $N_i$ a thread ID and $L_i$ a code line number,
+denote the execution sequence of a program as it runs according to the specified thread interleaving.
+This serialization of concurrent execution is told from the perspective of all CPUs at once
+and hence assumes sequential consistency.
+% TODO: fix sect refrance
+For discussion of relaxed memory models refer to Section~\ref{sec:warpzone}.
+
+\subsection{Example}
+
+Consider again the program in Figure~\ref{fig:htm-example}.
+Note that the C-style {\tt x++} operations, when compiled into assembly, %\cite{printy},
+results in multiple memory accesses which can be interleaved with other threads.
+\[
+\begin{tabular}{ll}
+	$2a$ & \texttt{temp <- x;} \\
+	$2b$ & \texttt{temp <- temp + 1;} \\
+	$2c$ & \texttt{x <- temp;} \\
+\end{tabular}
+\]
+
+% T1 T2 T3 but latex doesn't allow #s in cmd names
+\newcommand\ti{\ensuremath{\hilight{lavender}{\mathbf{T1}}}\xspace}
+\newcommand\tj{\ensuremath{\hilight{seafoam}{\mathbf{T2}}}\xspace}
+\newcommand\tk{\ensuremath{\hilight{salmon}{\mathbf{T3}}}\xspace}
+
+\newcommand\tiat[1]{\ensuremath{\hilight{lavender}{\mathbf{T1}@#1}}\xspace}
+\newcommand\tjat[1]{\ensuremath{\hilight{seafoam}{\mathbf{T2}@#1}}\xspace}
+\newcommand\tkat[1]{\ensuremath{\hilight{salmon}{\mathbf{T3}@#1}}\xspace}
+
+If these instructions from the {\tt x++} in the transaction are preempted,
+with another thread's access to {\tt x} interleaved in between,
+the transaction will abort.
+So, the interleaving
+\[
+	\tiat{1}, \tiat{2a}, \tiat{2b}, \tjat{1}, \tjat{2}, \tjat{3}, \tiat{2c}, \tiat{3}
+\]
+or, henceforth abbreviated for clarity:
+\[
+	\tiat{1-2b}, \tjat{1-3}, \tiat{2c-3}
+\]
+is not possible; rather, \ti will fall into the backup path:
+\[
+	\tiat{1-2b}, \tjat{1-3}, \tiat{4-7}
+\]
+However, the {\tt x++} operation from the failure path (correspondingly $6a$, $6b$, $6c$)
+{\em can} be thusly separated with conflicting accesses interleaved in between,
+since the mutex only protects the failure path against other failure paths,
+but not against the transaction itself.
+So (assuming {\tt x} is intended to be a precise counter rather than a sloppy one),
+%losing one of the increments to which constitutes a bug),
+the following interleaving exposes a bug:%
+\footnote{Note also that this bug requires either at least 3 threads or at least 2 iterations between 2 threads to expose;
+this highlights MC's dependence on its test cases to produce meaningful state spaces in the first place.}
+\[
+	\tiat{1-2b}, \tjat{1-3}, \tiat{4-6b}, \tkat{1-3}, \tiat{6c-7}
+\]
+Prior work \cite{htm-mario} proposed the idiom shown in Figure~\ref{fig:htm-fixed}
+to exclude this family of interleavings,
+which shows that correctly synchronizing even the simplest transactions may be surprisingly difficult or complex.
+%further motivating our research.
+
+\begin{figure}[t]
+	\begin{center}
+		\begin{tabular}{ll}
+		%\texttt{void count() \{} \\
+		%\texttt{~~~~for (int i = 0; i < 1000; i++) \{} \\
+		  & \texttt{prevent\_transactions = \const{false};} \\
+		\\
+		0 & \texttt{\flow{while} (prevent\_transactions) \flow{continue};} \\
+		1 & \texttt{\flow{if} ((status = \call{\_xbegin}()) == \const{SUCCESS}) \{} \\
+		2 & \texttt{~~~~\flow{if} (prevent\_transactions)} \\
+		3 & \texttt{~~~~~~~~\call{\_xabort}();} \\
+		4 & \texttt{~~~~x++;} \\
+		5 & \texttt{~~~~\call{\_xend}();} \\
+		6 & \texttt{\} \flow{else} \{} \\
+		7 & \texttt{~~~~\call{mutex\_lock}(\&m);} \\
+		8 & \texttt{~~~~prevent\_transactions = \const{true};} \\
+		9 & \texttt{~~~~x++;} \\
+		A & \texttt{~~~~prevent\_transactions = \const{false};} \\
+		B & \texttt{~~~~\call{mutex\_unlock}(\&m);} \\
+		C & \texttt{\}} \\
+		%\texttt{~~~~\}} \\
+		%\texttt{\}} \\
+		\end{tabular}
+	\end{center}
+	\caption{Variant of the program in Figure~\ref{fig:htm-example},
+		with additional synchronization to protect the failure path from the transactional path.
+		The optional line 0 serves to prevent a cascade of failure paths
+		for the sake of performance
+		by allowing threads to wait until transacting is safe again.}
+	\label{fig:htm-fixed}
+\end{figure}
+
+\subsection{Modeling Transaction Failure}
+
+Left unstated in interleavings such as $\tiat{1-2c}, \tjat{1-3}, \tiat{4-7}$%
+\footnote{For a clearer example here I have reordered \ti's write to {\tt x} before \tj's part, compared to before.}
+are HTM's execution semantics, namely:
+\begin{enumerate}
+	\item any modifications to shared state (such as $2c$) by \ti are not visible to \tj during its execution,
+		despite \tj being executed afterwards, and
+	\item all local and global state changes by \ti between lines $1$ and $2c$ are discarded when jumping to line $4$.
+\end{enumerate}
+% there shouldbe a better word than 'requires' here.. benefits from.. "the point of it is"..
+While use of HTM in production requires the performance advantage
+of temporarily staging such accesses in local CPU cache,
+model checking such programs need be concerned only with the program's {\em observable} behaviours.
+I claim that MCing the simpler interleaving $\tiat{1}, \tjat{1-3}, \tiat{4-7}$
+is an equivalent verification as MCing the one above;
+in fact, this interleaving suffices to check
+all observable behaviours
+of all interleavings
+of all subsets of $\tjat{1-3}$
+with all subsets of $\tiat{2a-2c}$,
+whether they share a memory conflict or not.
+Stated formally, let:
+
+\newcommand\tii{\ensuremath{\hilight{lavender}{\mathbf{Ti}}}\xspace}
+\newcommand\tjj{\ensuremath{\hilight{seafoam}{\mathbf{Tj}}}\xspace}
+\newcommand\tkk{\ensuremath{\hilight{salmon}{\mathbf{Tk}}}\xspace}
+
+\newcommand\tiiat[1]{\ensuremath{\hilight{lavender}{\mathbf{Ti}@#1}}\xspace}
+\newcommand\tjjat[1]{\ensuremath{\hilight{seafoam} {\mathbf{Tj}@#1}}\xspace}
+\newcommand\tkkat[1]{\ensuremath{\hilight{salmon}  {\mathbf{Tk}@#1}}\xspace}
+
+\begin{itemize}
+	\item $\tiiat{\alpha}$ be an HTM begin operation,
+	\item $\tiiat{\beta_1}\dots\tiiat{\beta_n}$ be the transaction body (with $\beta_n$ the HTM end call),
+	\item $\tiiat{\phi_1}\dots\tiiat{\phi_m}$ be the failure path, and
+	\item $\tiiat{\omega_1}\dots\tiiat{\omega_l}$ be the subsequent code executed unconditionally.%
+		\footnote{Arbitrary code may not be structured to distinguish these as nicely as in our examples;
+		e.g., more code may exist in the success branch after {\tt \_xend()};
+		such would be considered part of $\omega$ here.}
+\end{itemize}
+
+Then, without loss of generality (for any number of other threads \tjj/\tkk, and for any number of thread switches away from \tii during the transaction):
+
+\begin{lemma}[Equivalence of Aborts]
+	\label{lem:equiv}
+	For any interleaving prefix
+	\[
+	\begin{tabular}{c}
+		$\tiiat{\alpha},\tiiat{\beta_1}\dots\tiiat{\beta_b},$\\
+		$\tjjat{\gamma_1}\dots\tjjat{\gamma_j},$ \\
+		$\tkkat{\kappa_1}\dots\tkkat{\kappa_k},$ \\
+		$\tiiat{\beta_{b+1}}$ % $\dots\tiiat{\beta_{n-1}}$ -- excluded bc might abort
+	\end{tabular}
+	\]
+	with $b<n$, $j \ne i$, $k \ne i$, etc., either:
+	\begin{enumerate}
+		\item $\tiiat{\alpha},\tjjat{\gamma_1}\dots\tjjat{\gamma_j},\tkkat{\kappa_1}\dots\tkkat{\kappa_k},\tiiat{\phi_1}\dots$
+			(conflicting case), or
+		\item $\tiiat{\alpha},\tiiat{\beta_1}\dots\tiiat{\beta_b}\dots\tiiat{\beta_n},\tjjat{\gamma_1}\dots\tjjat{\gamma_j},$ \\
+			$\tkkat{\kappa_1}\dots\tkkat{\kappa_k}$
+			(independent case)
+	\end{enumerate}
+	exists and is observationally equivalent.
+\end{lemma}
+
+\begin{proof}
+	Case on whether the operations by \tjj and/or \tkk have any memory conflicts (read/write or write/write)
+	with $\tiiat{\beta_1}\dots\tiiat{\beta_n}$.
+	If so, then the hardware will abort \tii's transaction, discarding the effects of $\tiiat{\beta_1}\dots\tiiat{\beta_n}$
+	and jumping to $\tiiat{\phi_1}$,
+	satisfying case 1.
+	% TODO: give a secrion referenace
+	Otherwise, by DPOR's definition of transition dependence \cite{dpor}, %,landslide-phdthesis},
+	$\tiiat{\beta_{b+1}}\dots\tiiat{\beta_n}$ is independent with the transitions of \tjj and \tkk,
+	may be successfully executed until transaction commit,
+	and reordering them produces an equivalent interleaving,
+	satisfying case 2.
+\end{proof}
+
+The claim's second part follows naturally.
+
+\begin{theorem}[Atomicity of Transactions]
+	\label{thm:atom}
+	For any state space $S$ of a transactionally-concurrent program,
+	an equivalent state space exists in which all transactions are either executed atomically or aborted immediately.
+\end{theorem}
+
+\begin{proof}
+	For every $I \in S$ with $\tiiat{\alpha},\tiiat{\beta_1}\dots\tiiat{\beta_b},$ $\tjjat{\dots},\tkkat{\dots},\tiiat{\beta_{b+1}} \in I$,
+	apply Lemma~\ref{lem:equiv} to obtain an equivalent interleaving $I'$ satisfying the theorem condition.
+	The resulting $S'$ can then be MCed without ever simulating HTM rollbacks.
+\end{proof}
+
+\subsection{Memory Access Analysis}
+
+Next comes the issue of memory accesses within transactions with regard to data-race analysis (\sect{\ref{sec:background-datarace}}).
+Theorem~\ref{thm:atom} provides that the body of all transactions may be executed atomically within the MC environment.
+While they may interleave between other non-transactional sequences,
+no other operations (whether transactional or not) will interrupt them.
+I claim this level of atomicity is equivalent to that provided by a global lock,
+and hence abstracting it as such in Landslide's data-race analysis is sound.
+
+Let $\tiiat{\mu},\tjjat{\nu}$ be a pair of memory accesses to the same address, at least one a write,
+in some transactional execution $I$ normalized under Lemma~\ref{lem:equiv}.
+Then let $\mathsf{lockify}_m(\tkkat{L})$ denote a function over instructions in $I$,
+which replaces $\tkkat{L}$ with $\tkkat{\mathsf{lock}(m)}$ if $L$ is a successful HTM begin,
+with a no-op if $L$ is a transaction abort,
+or with $\tkkat{\mathsf{unlock}(m)}$ if $L$ is an HTM end,
+or no replacement otherwise.
+Finally, let $I' = \exists m. \mathsf{lockify}_m(I)$,
+the execution with the boundaries of all successful transactions replaced by an abstract global lock.
+Lemma~\ref{lem:equiv} guarantees mutual exclusion of $m$.
+
+\begin{theorem}[Transactions are a Global Lock]
+	$\tiiat{\mu},\tjjat{\nu}$ is a data race in $I$ iff it is a data race in $I'$.
+\end{theorem}
+
+\begin{proof}
+I prove one case for each variant definiton for data races supported in Landslide \cite{quicksand}.
+For each, I state below what it means to race in an execution with synchronizing HTM instructions.
+
+\begin{itemize}
+	\item {\bf Limited Happens-Before.}
+		%Under this definition,
+		To race in $I$ they must be reorderable at instruction granularity,
+		at least one with a thread switch immediately before or after.
+		\cite{tsan,hybriddatarace}.
+		\begin{itemize}
+			\llitem $I \Rightarrow I'$:
+				If $\tiiat{\mu},\tjjat{\nu}$ race in $I$,
+				then they cannot both be in successful transactions,
+				or else placing \tiiat{\mu} within the boundaries of \tjjat{\nu}'s transaction
+				would cause the latter to abort, invalidating \tjjat{\nu}, or vice versa.
+				Hence they will not both hold $m$ in $I'$.
+				Otherwise their lock-sets and DPOR dependence relation remain unchanged.
+			\llitem $I' \Rightarrow I$:
+				If $\tiiat{\mu},\tjjat{\nu}$ race in $I'$,
+				both corresponding threads cannot hold $m$;
+				WLOG let $\tii$ not hold $m$ during $\tiiat{\mu}$.
+				Then in $I$, $\tiiat{\mu}$ is not in a transaction.
+				With the remainder of their lock-sets still disjoint,
+				and still not DPOR-dependent, $\tjjat{\nu}$ (or its containing transaction)
+				can then be reordered directly before or after $\tiiat{\mu}$.
+		\end{itemize}
+	\item {\bf Pure Happens-Before.}
+		WLOG fix $\tiiat{\mu} \prec \tjjat{\nu} \in I$.
+		Then to race in $I$ there must be no pair of synchronizing instructions
+		$\tiiat{\epsilon}$ (a release edge) and $\tjjat{\chi}$ (an acquire edge) such that
+		\[
+			\tiiat{\mu} \prec \tiiat{\epsilon} \prec \tjjat{\chi} \prec \tjjat{\nu} \in I
+		\]
+		to update the vector clock epoch between \tiiat{\mu} and \tjjat{\nu} \cite{djit,fasttrack}.
+		\begin{itemize}
+			\llitem $I \Rightarrow I'$:
+				If $\tiiat{\mu},\tjjat{\nu}$ race in $I$,
+				then they cannot both be in successful transactions,
+				or else Lemma~\ref{lem:equiv} normalization would provide
+				the corresponding HTM end and begin for $\tiiat{\epsilon}$ and $\tjjat{\chi}$ respectively.
+				Hence there will be no unlock/lock pair on $m$ in $I'$ to satisfy the above sequence.
+			\llitem $I' \Rightarrow I$:
+				If $\tiiat{\mu},\tjjat{\nu}$ race in $I'$,
+				then they cannot both hold $m$,
+				or else $\mathsf{lockify}_m$ would provide the corresponding
+				unlock and lock for $\tiiat{\epsilon}$ and $\tjjat{\chi}$ respectively.
+				Hence there will be no HTM end/begin pair in $I$ to satisfy the above sequence.
+		\end{itemize}
+		%In both cases $I$ and $I'$ otherwise have the same synchronizing instructions.
+\end{itemize}
+Hence, data-race analysis is sound when transaction boundaries are replaced by an abstract global lock.
+\end{proof}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{Implementation}
 
 % TODO
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{Evaluation}
+
+% TODO

### 2018-05-03 thesis/tm.tex
put example figure and syntax hilight it

@@
 \label{chap:tm}
-
-% copied from sigbovik - mb move into thesis.tex later?
-% threads
-\definecolor{lavender}{RGB}{102,65,208} % (?? i don't know what happened here anymore but it looks good)
-\definecolor{seafoam}{RGB}{56,158,68} % * 2/3
-\definecolor{salmon}{RGB}{208,89,74} % * 7/8
-% eval data
-\definecolor{darkcyan}{RGB}{0,128,128}
-\definecolor{lime}{RGB}{48,128,0}
-\definecolor{darklavender}{RGB}{68,44,138}
-\definecolor{brownish}{RGB}{128,96,0}
-\definecolor{pinkish}{RGB}{128,34,102} % ff4488
-% syntax hilighting
-\newcommand\flow[1]{\hilight{brown}{#1}\xspace}
-\newcommand\const[1]{\hilight{brickred}{#1}\xspace}
-\newcommand\call[1]{\hilight{pinkish}{#1}}
 
@@
 
-An example TSX program is shown in Figure~\ref{fig:htm-example} (\sect{\ref{sec:overview-tm}}).
+The example TSX program from Figure~\ref{fig:htm-example} (\sect{\ref{sec:overview-tm}})
+is reproduced here for the reader's convenience.
+
+\begin{figure}[h]
+	\input{fig-htm-example}
+	\caption{Example TSX program.}
+	\label{fig:htm-example-reproduced}
+\end{figure}
 
@@
 
-Consider again the program in Figure~\ref{fig:htm-example}.
+Consider again the program in Figure~\ref{fig:htm-example-reproduced}.
 Note that the C-style {\tt x++} operations, when compiled into assembly, %\cite{printy},
@@
 		%\texttt{~~~~for (int i = 0; i < 1000; i++) \{} \\
-		  & \texttt{prevent\_transactions = \const{false};} \\
+		& \texttt{\ctype{bool} prevent\_transactions = \const{false};} \\
 		\\
 		0 & \texttt{\flow{while} (prevent\_transactions) \flow{continue};} \\
-		1 & \texttt{\flow{if} ((status = \call{\_xbegin}()) == \const{SUCCESS}) \{} \\
+		1 & \texttt{\flow{if} ((status = \call{\_xbegin}()) == \const{\_XBEGIN\_STARTED}) \{} \\
 		2 & \texttt{~~~~\flow{if} (prevent\_transactions)} \\
@@
 	\end{center}
-	\caption{Variant of the program in Figure~\ref{fig:htm-example},
+	\caption{Variant of the program in Figure~\ref{fig:htm-example-reproduced},
 		with additional synchronization to protect the failure path from the transactional path.
@@
 			(conflicting case), or
-		\item $\tiiat{\alpha},\tiiat{\beta_1}\dots\tiiat{\beta_b}\dots\tiiat{\beta_n},\tjjat{\gamma_1}\dots\tjjat{\gamma_j},$ \\
-			$\tkkat{\kappa_1}\dots\tkkat{\kappa_k}$
+		\item $\tiiat{\alpha},\tiiat{\beta_1}\dots\tiiat{\beta_b}\dots\tiiat{\beta_n},\tjjat{\gamma_1}\dots\tjjat{\gamma_j},\tkkat{\kappa_1}\dots\tkkat{\kappa_k}$
 			(independent case)
@@
 The claim's second part follows naturally.
+\vspace{1em}
 
@@
 Lemma~\ref{lem:equiv} guarantees mutual exclusion of $m$.
+\vspace{1em}
 

### 2018-05-14 thesis/tm.tex
improve htm theory discussion

@@
 
-Left unstated in interleavings such as $\tiat{1-2c}, \tjat{1-3}, \tiat{4-7}$%
+In the previous section's examples,
+the way I stated interleavings such as $\tiat{1-2c}, \tjat{1-3}, \tiat{4-7}$%
 \footnote{For a clearer example here I have reordered \ti's write to {\tt x} before \tj's part, compared to before.}
-are HTM's execution semantics, namely:
+glossed over the HTM execution semantics underlying how such a sequence of operations would be carried out.
+For example, thread 1's write during $2c$ is not actually visible to thread 2,
+such as it would be under a thread-scheduling-only concurrency model.
+
+Intel's official TSX documentation summarizes its interface and behaviour in prose \cite{intel-tsx-overview};
+to the best of my knowledge,
+no formal definition exists which could be verified in a mechanized logical framework \cite{x86-semantics}.
+Hence the proofs in this section will likewise be in prose.
+To summarize HTM's execution semantics:
+
+% TODO FIXME: this is not genral semantics, this is jut as epplies to the examineple. rephrase
 \begin{enumerate}
@@
 \end{enumerate}
+
 % there shouldbe a better word than 'requires' here.. benefits from.. "the point of it is"..
@@
 	\item $\tiiat{\phi_1}\dots\tiiat{\phi_m}$ be the failure path, and
-	\item $\tiiat{\omega_1}\dots\tiiat{\omega_l}$ be the subsequent code executed unconditionally.%
-		\footnote{Arbitrary code may not be structured to distinguish these as nicely as in our examples;
+	\item $\tiiat{\omega_1}\dots\tiiat{\omega_l}$ be the subsequent code executed unconditionally.
+\end{itemize}
+Note that arbitrary code may not be structured to distinguish these as nicely as in our examples;
 e.g., more code may exist in the success branch after {\tt \_xend()};
-		such would be considered part of $\omega$ here.}
-\end{itemize}
+such would be considered part of $\omega$ here.
 
 Then, without loss of generality (for any number of other threads \tjj/\tkk, and for any number of thread switches away from \tii during the transaction):
+\vspace{1em}
 
@@
 \begin{proof}
-	Case on whether the operations by \tjj and/or \tkk have any memory conflicts (read/write or write/write)
+	Case on whether the operations by \tjj and/or \tkk have any memory conflicts (read/\allowbreak{}write or write/write)
 	with $\tiiat{\beta_1}\dots\tiiat{\beta_n}$.

### 2018-05-14 thesis/tm.tex
tm implementation (partly)

@@
 results in multiple memory accesses which can be interleaved with other threads.
+% TODO: fix line numbers
+% TODO: put "void count" back in the mario-fixed example program, and fix ALL line numbers thruout proofs
 \[
@@
 
-% TODO
+Support for TSX programs in Landslide is implemented in three parts, broadly speaking,
+the user interface plus one internal part corresponding to each proof above.
+
+\subsection{User Interface}
+
+Landslide provides its own ``implementation'' of the TSX interface,
+which matches GCC's interface exactly,
+located in {\tt 410user/inc/htm.h}
+under {\tt pebsim/p2-basecode/}.
+The interface is implemented in {\tt 410user/libhtm/htm.c},
+perhaps surprisingly, as totally empty functions:
+Landslide hooks these functions' addresses during instrumentation
+and inserts preemption points, failure injections, etcetera as necessary whenever the execution encounters them.
+
+Transactional test programs should be ported to the Pebbles userspace if not already,
+then replace any use of compiler HTM intrinsics with Landslide's interface%
+\footnote{Attempts to execute a real TSX instruction under Landslide, instead of using the custom interface,
+will be reported as ``invalid opcode'' bugs.
+}.
+They should then be put in either {\tt 410user/progs/} or {\tt user/progs/}
+and the {\tt 410TESTS} or {\tt STUDENTTESTS} line (respectively) of {\tt config-incomplete.mk} be edited to add the test name,
+% TODO: better section reference
+before running {\tt p2-setup.sh} (\sect{\ref{chap:education}})
+% TODO(implementation): Provide an obfuscated refp2?
+on a (hopefully) correct P2 as usual.
+Several example HTM tests are provided as {\tt 410user/progs/htm*.c}.
+
+Finally, Quicksand supports the following command-line options to enable various sets of HTM features within Landslide.
+Each option requires each of the previously-listed options to also be present.
+\begin{itemize}
+	\item {\tt -X} (for ``tsX'' or ``Xbegin'') enables the basic features:
+		preemption points on each {\tt \_xbegin()} and {\tt \_xend()} call,
+		failure injection on each of the former
+		(always returning {\tt \_XABORT\_RETRY} as the failure code)
+		(\sect{\ref{sec:txn-failure}}),
+		and treatment of transactional regions during data-race analysis (\sect{\ref{sec:txn-datarace}}).
+	\item {\tt -A} (for ``xAbort codes'')
+		enables multiple xabort failure codes (\sect{\ref{sec:txn-failure}}).
+	% TODO(implementation): implement an option for this as -S for STM.
+	\item {\tt -S} (for ``Stm'')
+		disables the {\tt \_XABORT\_RETRY} failure reason,
+		which causes Landslide to match STM's execution semantics rather than HTM.
+		(This feature was used in \cite{sigbovik-htm}~\sect{5}.)
+\end{itemize}
+
+\subsection{Failure Injection}
+\label{sec:txn-failure}
+
+Each preemption point ({\tt struct hax} as defined in {\tt tree.h})
+includes three new fields.
+The boolean {\tt h->xbegin} is set if the preemption point occurred at an {\tt \_xbegin()} call.
+Then if set (as the tag of an option type),
+the twin lists of integers
+{\tt h->xabort\_codes\_ever}
+and
+{\tt h->xabort\_codes\_todo}
+store possible error codes this {\tt \_xbegin()} call should be tested to possibly return.
+The former list stores all such error codes, whether already tested or yet to be tested,
+while the latter serves as a workqueue that indicates only those not already tested yet
+(serving an analogous purpose as the {\tt all\_explored} flag for thread scheduling).
+
+\subsubsection{Adding possible abort codes}
+
+Up to three abort codes are considered depending on the testing options listed in the previous section.
+All abort codes are simultaneously added to both lists,
+unless already present on {\tt xabort\_codes\_ever},
+in which case not added to {\tt xabort\_codes\_todo} to avoid duplicate work.
+\begin{itemize}
+	\item {\tt \_XABORT\_RETRY}:
+		When a {\tt xbegin} preemption point is created ({\tt save\_setjmp()}),
+		if {\tt -S} is not set,
+		both {\tt xabort\_codes} lists are initialized to contain this code.
+		This represents the possibility for a hardware transaction to fail
+		for reasons outside of the programmer's control such as system interrupts or cache eviction.
+		If {\tt -S} is set they are initialized to empty,
+		representing either STM's policy of failing only when a true memory conflict arises,
+		or a TSX user wrapping all their {\tt \_xbegin()}s in a retry loop such as in \cite{sigbovik-htm}.
+	\item {\tt \_XABORT\_CONFLICT}:
+		Whenever DPOR detects a memory conflict between two preemption points ({\tt shimsham\_shm()}),
+		if {\tt -A} is set,
+		for each of the two which are within a transaction,
+		it finds the nearest ancestor {\tt xbegin} preemption point by the same thread
+		and adds this code to both lists.
+		% FIXME(implementation): make sure that abort_transaction can't cross a xend() boundary?
+		% don't want to cause spurious conflict aborts on post-xend code
+	\item {\tt \_XABORT\_EXPLICIT}:
+		Whenever the program invokes {\tt \_xabort()} ({\tt sched\_update\_user\_\allowbreak{}state\_machine()}),
+		if {\tt -A} is set,
+		this code is added to both lists, bitwise-ored with the user-supplied argument code
+		as specified in \cite{htm-gcc}.
+\end{itemize}
+
+% FIXME(implementation): false sharing should be covered under conflict, not retry!
+
+\subsubsection{Injecting abort codes}
+When traversing the state space (\sect{\ref{sec:landslide-statespace}}),
+in addition to performing DPOR to select non-independent thread interleavings (\sect{\ref{sec:landslide-dpor}}),
+% TODO: reduction challenge?
+the abort codes under each {\tt xbegin} preemption point are also considered ``marked'' paths which must be tested.
+Hence {\tt explore()}, by way of {\tt any\_tagged\_child()},
+will pop off the {\tt xabort\_codes\_todo} queue
+when it's time to explore that preemption point in the usual depth-first manner%
+\footnote{The search order prioritizes abort codes before scheduling other threads
+at such preemption points,
+which is just an implementation detail, not theoretically necessary}.
+The optional abort code is then passed through {\tt arbiter\_append\_choice()}/{\tt arbiter\_pop\_choice()}
+to {\tt cause\_transaction\_failure()},
+which edits the simulation state ({\tt \%eip} and {\tt \%eax})
+to force {\tt \_xbegin()} to return the provided code.
+
+\subsection{Data Race Analysis}
+\label{sec:txn-datarace}
+
+% XXX(implementation): i bet you need to fix your pure hb impl
 

### 2018-05-14 thesis/tm.tex
false sharing

@@
 
-% FIXME(implementation): false sharing should be covered under conflict, not retry!
+Landslide does not yet check for false sharing,
+i.e. read/write or write/write access pairs to different memory addresses that share a cache line.
+On real hardware, these would produce {\tt \_XABORT\_CONFLICT} failures,
+but to find such access pairs would require extending {\tt mem\_shm\_intersect()}'s set intersection algorithm
+to consider a certain degree of ($N$-byte-aligned) fuzziness when comparing addresses,
+as well as adding a command-line option to configure said $N$, the cache-line size to simulate.
+For now, Landslide (wrongly) lumps false sharing in among the ``spurious'' {\tt \_XABORT\_RETRY} failure reasons.
+
+On HTM, even if the user wraps all such spurious failure reasons in a retry loop,
+false sharing should still produce a non-retryable abort,
+begetting a discrepancy with STM,
+which aborts only when the memory addresses match exactly.
+Accordingly, the {\tt -S} option described above provides STM semantics as currently implemented.
+Future work could extend Landslide with an option to configure false sharing conflicts
+to remain faithful to HTM semantics and still abort even under {\tt -S}.
 

### 2018-05-15 thesis/tm.tex
txn data race impl

@@
 		enables multiple xabort failure codes (\sect{\ref{sec:txn-failure}}).
-	% TODO(implementation): implement an option for this as -S for STM.
 	\item {\tt -S} (for ``Stm'')
@@
 		and adds this code to both lists.
-		% FIXME(implementation): make sure that abort_transaction can't cross a xend() boundary?
-		% don't want to cause spurious conflict aborts on post-xend code
 	\item {\tt \_XABORT\_EXPLICIT}:
@@
 
-% XXX(implementation): i bet you need to fix your pure hb impl
+When a thread returns {\tt \_XBEGIN\_STARTED} from {\tt \_xbegin()}
+(analogous to {\tt mutex\_trylock()}),
+Landslide's scheduler sets the ``in transaction'' action flag for that thread (\sect{\ref{sec:landslide-scheduler}}),
+and if using Pure Happens-Before,
+applies \textsc{FT acquire} (\sect{\ref{sec:landslide-phb}}) using a dummy lock address to represent the abstract global lock.
+When a thread reaches {\tt \_xend()},
+the above flag is cleared,
+and under Pure Happens-Before,
+\textsc{FT release} is applied.
+Then when {\tt check\_locksets()} compares an access pair,
+under Limited Happens-Before,
+it is considered not a data race if not both threads' ``in transaction'' flags were set in addition to the usual conditions;
+under Pure Happens-Before,
+the vector clocks are simply checked as usual.
 

### 2018-05-22 thesis/education.tex
introduce educ chapt

@@
 
-% TODO
+Concurrency is taught in as many different ways as there are
+systems programming classes at universities which teach the subject.
+Yet one thing they all have in common is presenting the concurrency bug
+as some elusive menace,
+humanity's best weapon against which is mere random stress testing.
+This chapter will prove stateless model checking's mettle as a better alternative in the educational theatre.
+
+While the previous chapter demonstrated Landslide's bug-finding power
+compared to prior MC techniques in a controlled environment,.
+whether it offers pedagogical merit in the hands of students and/or TAs is a separate question.
+And while my MS thesis \cite{landslide} showed that students
+could annotate P3 Pebbles kernels and thence use Landslide to debug them,
+the annotations alone required 2 hours of effort on average per user,
+meaning the only students who could benefit were the ones already succeeding enough to have such free time.
+% TODO: sect ref
+Since then, I have extended Landslide with a fully-automatic instrumentation process
+for Pebbles thread libraries (P2s) and Pintos kernels
+to improve its accessibility.
+
+I have run several user studies in the Operating Systems classes
+at Carnegie Mellon University (CMU), University of Chicago (U. Chicago), and Penn State University (PSU),
+wherein students get to use Landslide to find and diagnose their own bugs during the semester.
+% TODO: put section refs
+At CMU, I recorded logs and code snapshots as students used Landslide during P2.
+At CMU and PSU, I surveyed students on their experience after submitting their Landslide-debugged P2s.
+At U. Chicago, I collaborated with a TA to check submitted Pintos kernels,
+then return any resulting bug reports to students and likewise survey them on the quality of the diagnostic output.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{Pebbles}
+
+% Possible experiment questions
+% compare e.g. use after free bug reporce from OOPSLA data set, to P2 grade files, see who has thread exit uafs
+% is landslide better at finding thread uafs than TAs
+% same Q for other stuff.. (expect paraguay answer to be "no", explain why)
+
+\subsection{Automatic instrumentation}
+
+\subsection{Evaluation}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{Pintos}
+
+\subsection{Automatic instrumentation}
+
+\subsection{Evaluation}
+
+% survey questions i WISH i had asked
+% - did you have any technical difficulties w landslide that i had to intervene on
+
+% \section{Survey results} % ???

### 2018-05-23 thesis/education.tex
write abt recruiting, etc for eduction

@@
 as some elusive menace,
-humanity's best weapon against which is mere random stress testing.
+against which humanity's best weapon is mere random stress testing.
 This chapter will prove stateless model checking's mettle as a better alternative in the educational theatre.
@@
 While the previous chapter demonstrated Landslide's bug-finding power
-compared to prior MC techniques in a controlled environment,.
+compared to prior MC techniques in a controlled environment,
 whether it offers pedagogical merit in the hands of students and/or TAs is a separate question.
-And while my MS thesis \cite{landslide} showed that students
+And while my MS thesis~\cite{landslide} showed that students
 could annotate P3 Pebbles kernels and thence use Landslide to debug them,
@@
 At U. Chicago, I collaborated with a TA to check submitted Pintos kernels,
-then return any resulting bug reports to students and likewise survey them on the quality of the diagnostic output.
+then returned any resulting bug reports to students and likewise surveyed them on the quality of the diagnostic output.
 
@@
 
-\section{Pebbles}
+\section{P2}
+
+This section presents the user studies done in
+%CMU's 15-410 and PSU's \psuos classes,
+%in semesters Fall 2015 to Spring 2018 and in Spring 2018 alone, respectively,
+%taught by David Eckhardt and Timothy Zhu, respectively.
+CMU's 15-410 in semesters Fall 2015 to Spring 2018,
+taught by David Eckhardt,
+and in PSU's \psuos in Spring 2018,
+taught by Timothy Zhu.
+In both cases the instructors assisted to introduce me during the guest lecture
+%(see below)
+and to distribute the recruiting emails;
+TAs were not involved.
+The in-house user study has CMU IRB approval under study number STUDY2016\_00000425,
+and the external user study under STUDY2017\_00000429.
 
@@
 
+\subsection{Recruiting}
+
+Since the Spring 2015 semester I have given a guest lecture in 15-410
+to recruit students to participate in the user study.
+The 50-minute lecture is given 1 week into the 2.5-week-long P2 project,
+approximately when the students should be getting child threads running in {\tt thr\_create()}
+and experiencing concurrency bugs for the first time.
+It introduces the research subject abstractly
+using an example ``Paradise Lost'' bug from a previous lecture \cite{paradise-lost},
+explains how Landslide works concretely,
+shows a short demo of effortlessly using Landslide to find the example bug,
+and provides the necessary IRB legalese about the risks and benefits of participation.
+The most recent lecture slides are available on the course website at
+\url{http://www.cs.cmu.edu/~410-s18/lectures/L14_Landslide.pdf},
+and all semesters' editions at
+\url{https://github.com/bblum/talks/tree/master/landslide-lecture}.
+
+The PSU version of the lecture
+%was given in Spring 2018, and
+is available at
+\url{http://www.contrib.andrew.cmu.edu/~bblum/psu-lecture.pdf}
+as well as under the github link above.
+Being a 70-minute lecture slot rather than 50, I extended the demo to
+both find and (attempt to) verify a fix for two bugs:
+one a simple data race and the other the more complicated Paradise Lost bug as above.
+After finding each bug, I demonstrated using Landslide on a fixed version of the code
+to show how it proves the test case correct by completing all state spaces,
+or (in the case of Paradise Lost) suffers an exponentially-exploding state space.
+Not that I scientifically measured it or anything,
+but this extended demo seemed to help students more clearly understand Landslide's intended workflow,
+at the cost of about 10-15 extra minutes of lecture time.
+
+At both schools students then signed up using a Google form I emailed them,
+which upon completion linked them to the Landslide user guide,
+which is available at
+\url{http://www.contrib.andrew.cmu.edu/~bblum/landslide-guide-p2.pdf}
+(CMU version)
+and
+\url{http://www.contrib.andrew.cmu.edu/~bblum/landslide-guide-psu.pdf}
+(PSU version)
+and
+\url{https://github.com/bblum/talks/tree/master/irb}
+(both versions).
+
 \subsection{Automatic instrumentation}
+
+As described in \sect{\ref{sec:landslide-setup}},
+all setup from the user's point of view is handled through the {\tt p2-setup.sh} script%
+\footnote{PSU's version is called {\tt psu-setup.sh};
+in this section, unless otherwise noted, {\tt p2-setup.sh} will refer to both.
+}.
+It, its helper scripts (\sect{\ref{sec:landslide-glue}}),
+and the {\tt landslide} script itself contain several checks to prevent
+studence
+from accidentally misusing Landslide in ways that could produce mysterious crashes, false bug reports, and so on
+(the need for each one, as the reader might imagine, discovered through bitter experience).
+These include:
+\begin{itemize}
+	\item {\tt check-need-p2-setup-again.sh} checks if any source files in the original P2 source directory
+		(the argument supplied to {\tt p2-setup.sh}),
+		in case the student hoped to fix some bug and verify their fix but forgot to re-run the setup script.
+	% TODO
+	% pointing p2 setup at the user/libthread directory or whatever other wrong thing
+	% wrong program name like racer or OPTIONS
+	\item
+	\item {\tt landslide} checks if any other instance of itself is simultaneously running in the same directory,
+		and if so, refuses to do so and advises the student
+		to {\tt git clone} the repository afresh for simultaneous use%
+		\footnote{This is ironically implemented with a non-atomic lock file
+		and should really be using {\tt flock} instead.
+		}.
+\end{itemize}
+Landslide also includes several P2-specific instrumentations and features to cope with various student irregularities:
+\begin{itemize}
+		% yield loosps
+		% mutex loosp "suspicious"
+		% HURDLE_VIOLATION
+		% snapshooting
+		% psu magic post-test-case check these flags (rationale why, dont have to thr_exit -- also save state space)
+		% quicksand emit custom without_function directives depending on name of test
+		% TODO
+
+	\item
+\end{itemize}
+
+\subsection{Test cases}
+
+% mx, bcast, paradise, paraguay, rwl dgr, exit join
+
+% for psu additionally - bcast test 2, and all the atomics
 
@@
 
+This section presents the user study done in U. Chicago's \uchos class in the Fall 2017 semester,
+taught by Haryadi Gunawi.
+Kevin Zhao, the TA, assisted to run Landslide on student submissions
+and to distribute recruiting materials and testing results.
+The study has CMU IRB approval under study number STUDY2017\_00000429.
+
+\subsection{Recruiting}
+
+For this study students were recruited remotely via email.
+After each of the Threads and Userprog project deadlines (\sect{\ref{sec:overview-pintos}}),
+% TODO talk about th econtence of the email
+
 \subsection{Automatic instrumentation}
+
+\subsection{Test cases}
+
+% prisema
+% alarm simul
+% donate multiple
+% wait simple
 
@@
 % - did you have any technical difficulties w landslide that i had to intervene on
+% mb anything else from timmys 2nd latest email
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
 % \section{Survey results} % ???
+
+% TODO: addressing bias
+% well, we did the best we could(?)
+% survey email: "please answer honestly rather than flatteringly"
+% anything else?

### 2018-05-23 thesis/quicksand.tex
work on quotes
update quicksand chapter quote

@@
 \inspirationalquote{
-% original
-% {\footnotesize 嫌な事も悲しい事もあったけど、守りたい物だって沢山この世界にはあったから…} \\
-% trans. slightly paraphrased from crunchyroll subtitle; removed "..."s, added past tense for あった
-There were awful, sad things in this world.
-But there were a lot of things worth protecting, too.
+% original - TODO, put in
+%{\footnotesize 嫌な事も悲しい事もあったけど、守りたい物だってたくさんこの世界にはあったから。} \\
+There are awful, sad things in this world.
+But there are a lot of things worth protecting, too.
 }
-{Kaname Madoka, Mahou Shoujo Madoka{\raisebox{0.1em}{$\scriptstyle \bigstar$}}Magica}
+%{Kaname Madoka, Mahou Shoujo Madoka{\raisebox{0.1em}{$\scriptstyle \bigstar$}}Magica}
+{Kaname Madoka, Puella Magi Madoka{\raisebox{0.1em}{$\scriptstyle \bigstar$}}Magica}
 %\inspirationalquote{Always, somewhere, someone is fighting for you.

### 2018-05-23 thesis/overview.tex
work on quotes
put example figure and syntax hilight it
fix some code figures
write abt recruiting, etc for eduction
add some todo commence

@@
 \label{chap:background}
+
+%\inspirationalquote{
+%\begin{tabular}{p{0.83\textwidth}}
+%Our society, our art, everything\inspirationalhyphen{}it's built on thousands of years of human innovation.
+%So as long as you start on that foundation, and take it step by step\dots~you, too, can do amazing things.
+%\end{tabular}}
+%{Monika, Doki Doki Literature Club}
+
 \inspirationalquote{
@@
 				\texttt{void count() \{} \\
-				\texttt{~~~~for (int i = 0; i < 1000;} \\
-				\texttt{~~~~~~~~~i++)} \\
+				\texttt{~~~~for (int i = 0; i < 1000; i++)} \\
 				\texttt{~~~~~~~~x++;} \\
@@
 			\texttt{void count() \{} \\
-			\texttt{~~~~for (int i = 0; i < 1000;} \\
-			\texttt{~~~~~~~~~i++)} \\
+			\texttt{~~~~for (int i = 0; i < 1000; i++) \{} \\
 			\texttt{~~~~~~~~mutex\_lock(\&m);} \\
@@
 			\texttt{~~~~~~~~mutex\_unlock(\&m);} \\
+			\texttt{~~~~\}} \\
 			\texttt{\}} \\
@@
 \begin{figure}[h]
-	\begin{center}
-		\begin{tabular}{l}
-		\texttt{void count() \{} \\
-		\texttt{~~~~for (int i = 0; i < 1000; i++) \{} \\
-		\texttt{~~~~~~~~if ((status = \_xbegin()) == \_XBEGIN\_STARTED) \{} \\
-		\texttt{~~~~~~~~~~~~x++;} \\
-		\texttt{~~~~~~~~~~~~\_xend();} \\
-		\texttt{~~~~~~~~\} else \{} \\
-		\texttt{~~~~~~~~~~~~mutex\_lock(\&m);} \\
-		\texttt{~~~~~~~~~~~~x++;} \\
-		\texttt{~~~~~~~~~~~~mutex\_unlock(\&m);} \\
-		\texttt{~~~~~~~~\}} \\
-		\texttt{~~~~\}} \\
-		\texttt{\}} \\
-		\end{tabular}
-	\end{center}
+	\input{fig-htm-example}
 	\caption{The example {\tt count} routine from Figure~\ref{fig:mutex}, rewritten to use HTM.
@@
 
-\section{Data Race Analsyis}
+\section{Data Race Analysis}
 \label{sec:background-datarace}
@@
 The Pebbles kernel architecture
-is used at CMU in 15-410, Operating System Design and Education \cite{kspec,thrlib}.
+is used at Carnegie Mellon University (CMU) in 15-410, Operating System Design and Education \cite{kspec,thrlib}.
 In the course of a semester, students work on five programming assignments;
@@
 
+\newcommand\uchos{CMSC 23000\xspace}
+
 The Pintos kernel architecture \cite{pintos} is used at several universities, including Berkeley, Stanford, and the University of Chicago.
@@
 and designing correct interactions between the {\tt wait} and {\tt exit} system calls.
+
+\section{Glossary}
+
+This section provides a convenient reference of terminology used throughout the thesis.
+
+% TODO
+
+% concurrency bug
+% race condition: "a confusing term which in some circles means concurrency bug and others data race, which we will avoid"
+% data race
+% atomicity violation (what even is this.)
+% thread
+% transition
+% conflict
+% independent: "see conflict"
+% state space
+% interleaving
+% DPOR
+% estimation (WBE, RE)
+% landslide
+% quicksand
+% iterative deepening
+% MC, stateless MC
+% TA: teaching assistant
+% P2 (note abotu psu - dosent call it p2, but we call it that for both schools here

### 2018-05-23 thesis/tm.tex
add some todo commence

--- 2018-05-14 thesis/tm.tex
@@
 
-% XXX(implementation): i bet you need to fix your pure hb impl
+When a thread returns {\tt \_XBEGIN\_STARTED} from {\tt \_xbegin()}
+(analogous to {\tt mutex\_trylock()}),
+Landslide's scheduler sets the {\tt user\_txn} action flag for that thread (\sect{\ref{sec:landslide-scheduler}}),
+and if using Pure Happens-Before,
+applies \textsc{FT acquire} (\sect{\ref{sec:landslide-phb}}) using a dummy lock address to represent the abstract global lock.
+When a thread reaches {\tt \_xend()},
+the above flag is cleared,
+and under Pure Happens-Before,
+\textsc{FT release} is applied.
+Then when {\tt check\_locksets()} compares an access pair,
+under Limited Happens-Before,
+it is considered not a data race if not both threads' {\tt user\_txn} flags were set in addition to the usual conditions;
+under Pure Happens-Before,
+the vector clocks are simply checked as usual.
 
@@
 % TODO
+
+% TODO: evaluate a reduction in the AVL/etc where you take the xbegin/retry/mutex/retry pattern and encapsulate it in smth like
+% T txn(mutex_t *mp, T body(U env), U env)
+% verify this primitive on some simple mutual exclusion tests up to N,K or w/e
+% then in AVL/etc have landslide, whenever it is reached, force the user into an alternate path which just always takes the mx

### 2018-05-24 thesis/education.tex
test cases

@@
 \subsection{Recruiting}
+\label{sec:education-pebbles-recruiting}
 
@@
 \subsection{Automatic instrumentation}
+\label{sec:education-pebbles-instrumentation}
 
@@
 
-% mx, bcast, paradise, paraguay, rwl dgr, exit join
+Landslide ships with several ``approved'' test cases,
+i.e., programs copied from, derived from, or at least vaguely resembling
+the tests distributed with P2,
+which I curated to produce concurrent behaviour suitable for stateless model checking.
+Some tests are crafted to target specific bugs which,
+from personal experience as a TA, are common in many student submissions;
+others are crafted to exercise generally concurrency-heavy code paths and uncover any number of unforeseen problems.
 
-% for psu additionally - bcast test 2, and all the atomics
+The following tests were released to CMU students:
+\begin{itemize}
+	\item {\tt broadcast\_test}:
+		Tests the {\tt cond\_broadcast()} signalling path with a single waiter.
+	\item {\tt mutex\_test}:
+		Tests student mutexes under 2 threads with 2 iterations
+		(the 2nd iteration serves to expose problems with {\tt mutex\_unlock()} as well as {\tt mutex\_lock()}).
+		This test uses the special {\tt TESTING\_MUTEXES} data-race analysis mode
+		described in \sect{\ref{sec:landslide-staticconfig}}.
+	\item {\tt paradise\_lost}:
+		Written for the sake of the Landslide lecture demo
+		(\sect{\ref{sec:education-pebbles-recruiting}}).
+		Tests for the Paradise Lost bug by attempting to break mutual exclusion.
+	\item {\tt paraguay}:
+		Copied directly from the P2 test suite;
+		tests for proper handling of seemingly ``spurious'' wakeups in {\tt cond\_wait()}.
+		Written by Michael Sullivan.
+	\item {\tt rwlock\_downgrade\_read\_test}:
+		Copied directly from the P2 test suite;
+		tests for mutually-exclusive and deadlock-free {\tt rwlock\_downgrade()}.
+		Written by me (as a TA).
+	\item {\tt thr\_exit\_join}:
+		Copied directly from the P2 test suite;
+		tests for a variety of problems between {\tt thr\_exit()} and {\tt thr\_join()},
+		but especially for memory issues pertaining to stack deallocation.
+\end{itemize}
+The following tests were released to PSU students, in addition to the ones above:
+\begin{itemize}
+	\item {\tt atomic\_compare\_swap}:
+		Tests the {\tt cmpxchg} assembly function for being properly atomic.
+		Uses the {\tt magic\_*} global variables, and invokes {\tt vanish()} directly,
+		to avoid requiring the student to implement {\tt thr\_join()}/{\tt thr\_exit()}
+		before being able to run this test.
+	\item {\tt atomic\_exchange}:
+		As above for {\tt xchg}.
+	\item {\tt atomic\_fetch\_add}:
+		As above for {\tt xadd}.
+	\item {\tt atomic\_fetch\_sub}:
+		As above for {\tt xadd}.
+	\item {\tt broadcast\_two\_waiters}:
+		As {\tt broadcast\_test}, but uses two waiting threads to ensure both get signalled.
+\end{itemize}
 

### 2018-05-28 thesis/education.tex
annotate pintos

@@
 		(the 2nd iteration serves to expose problems with {\tt mutex\_unlock()} as well as {\tt mutex\_lock()}).
-		This test uses the special {\tt TESTING\_MUTEXES} data-race analysis mode
-		described in \sect{\ref{sec:landslide-staticconfig}}.
+		This test uses the {\tt TESTING\_MUTEXES}
+		described in \sect{\ref{sec:landslide-staticconfig}}
+		to enable data-race preemption points within the mutex implementation.
 	\item {\tt paradise\_lost}:
@@
 		Tests the {\tt cmpxchg} assembly function for being properly atomic.
-		Uses the {\tt magic\_*} global variables, and invokes {\tt vanish()} directly,
+		Uses the {\tt magic\_*} global variables (\sect{\ref{sec:education-pebbles-instrumentation}}), and invokes {\tt vanish()} directly,
 		to avoid requiring the student to implement {\tt thr\_join()}/{\tt thr\_exit()}
@@
 \subsection{Automatic instrumentation}
+\label{sec:education-pintos-instrumentation}
+
+As described in \sect{\ref{sec:landslide-setup}},
+all setup from the user's point of view is handled through the {\tt pintos-setup.sh} script.
+It
+and its helper {\tt pebsim/pintos/import-pintos.sh}
+perform most of the same sanity checks as listed in \sect{\ref{sec:education-pebbles-instrumentation}},
+then applies the patch {\tt annotate-\allowbreak{}pintos.patch}
+(plus several more hacks in the script itself)
+to insert the {\tt tell\_landslide\_*()} annotations (\sect{\ref{sec:tell-landslide}})
+into the student's kernel code.
+The following tricks serve to make sure the annotations apply consistently
+to (almost) all variations of code that students commonly submit:
+
+\begin{itemize}
+	\item Finds the declaration of {\tt ready\_list}, the scheduler runqueue declared by the basecode,
+		and detects if the student has modified to be an array of lists rather than a single one.
+		If so, defines the length of that array in a macro to be used by {\tt is\_runqueue()}
+		(part of the patch described below).
+		Either way defines a function {\tt get\_rq\_addr()} to return the address of the (first) list.
+	\item Changes the basecode's definition of {\tt TIME\_SLICE} from 4 to 1 (units of timer ticks)
+		so Landslide's timer injection will properly drive the context switcher.
+	\item Inserts {\tt tell\_landslide\_forking()} into {\tt thread.c}
+		(using {\tt sed} rather than the patch, described below,
+		because it must go in a function which students have to implement,
+		which is likely to disturb the context and make a patch fail).
+	\item Adds the new {\tt priority-donate-multiple} test.
+	\item Applies the {\tt annotate-pintos.patch} patch to the imported student implementation, which:
+	\begin{itemize}
+		\item Adds {\tt tell\_landslide\_thread\_on\_rq()}
+			and {\tt tell\_landslide\_thread\_off\_rq()}
+			annotations
+			to {\tt list\_insert()} and {\tt list\_remove()} respectively
+			(in {\tt lib/kernel/\allowbreak{}list.c}, which the students don't modify),
+			which
+			check whether the argument list
+			is the scheduler runqueue
+			using a helper function {\tt is\_runqueue},
+			which in turn uses {\tt get\_rq\_addr()} and {\tt READY\_LIST\_LENGTH} described above.
+		\item Modifies the existing {\tt priority-sema} and {\tt alarm-simultaneous} tests to be more Landslide-friendly.
+		\item Inserts the {\tt tell\_landslide\_sched\_init\_done()},
+			{\tt tell\_landslide\_vanishing()},
+			and {\tt tell\_landslide\_thread\_switch()}
+			annotations in the appropriate places
+			(which the students generally do not modify).
+	\end{itemize}
+	\item Detects if the student has renamed the {\tt elem} field of the TCB struct,
+		and if so renames its use in {\tt is\_runqueue()} (described above) correspondingly.
+\end{itemize}
 

### 2018-05-28 thesis/landslide.tex
tm implementation (partly)
write abt recruiting, etc for eduction
annotate pintos

@@
 \subsection{Setup}
+\label{sec:landslide-setup}
 
@@
 \subsection{In-kernel code annotations}
+\label{sec:tell-landslide}
 
@@
 \subsection{State space traversal}
+\label{sec:landslide-statespace}
 
@@
 		but I tried that and it was mysteriously slower, so I gave up without ever figuring out why.}
+	\item {\tt p2-basecode/import-p2.sh} and {\tt pintos/import-pintos.sh}
+		are invoked by their respective setup scripts
+		to copy the student implementation into their respective directories.
+		\sect{\ref{sec:education-pebbles-instrumentation}} and \sect{\ref{sec:education-pintos-instrumentation}}
+		describe their office in more detail.
 \end{itemize}

### 2018-05-29 thesis/education.tex
pintos testce

@@
 
+The tests can all be viewed at \url{https://github.com/bblum/landslide/tree/master/pebsim/p2-basecode/410user/progs}.
+
 \subsection{Evaluation}
@@
 For this study students were recruited remotely via email.
-After each of the Threads and Userprog project deadlines (\sect{\ref{sec:overview-pintos}}),
+After each of the {\em threads} and {\em userprog} project deadlines (\sect{\ref{sec:overview-pintos}}),
 % TODO talk about th econtence of the email
@@
 
-% prisema
-% alarm simul
-% donate multiple
-% wait simple
+Like the P2 tests, the Pintos test cases are either hand-picked from the provided unit tests,
+with an eye for which will produce interesting concurrent behaviour,
+or created using a TA's intuition for the most likely student bugs.
+The following tests are approved to be Landslide-friendly:
+
+\begin{itemize}
+	\item {\tt priority-sema}:
+		Modified to be Landslide-friendly from the basecode, for {\em threads}.
+		Creates two child threads to wait on a semaphore and signals them.
+		Replaces threads with different priorities
+		(originally chosen to produce deterministic output which the test checked for)
+		with threads of the same priority.
+	\item {\tt alarm-simultaneous}:
+		Modified to be Landslide-friendly from the basecode, for {\em threads}.
+		Creates two child threads which each invoke {\tt timer\_sleep()} for a different amount of time.
+		Number of (threads,iterations) reduced from (3,5) to (2,1).
+	\item {\tt wait-simple}:
+		Unmodified from the basecode's version, for {\em userprog}.
+		Userspace process {\tt exec()}s a child process, which immediately exits, and {\tt wait()}s on it.
+	\item {\tt priority-donate-multiple}:
+		Written by Kevin Zhao, TA at U. Chicago, for {\em threads}.
+		Tests for a priority donation race during {\tt lock\_release()}
+		in which a thread holding a lock can accidentally keep a contending thread's donated priority
+		after finishing releasing it.
+\end{itemize}
+
+The (unpatched versions of) the first three tests are available at
+\url{https://github.com/Berkeley-CS162/group0/tree/master/pintos/src/tests}.
+The fourth test is available at
+% FIXME: allowbreak makes nasty latex shit show up in the output??
+\url{https://github.com/bblum/landslide/blob/master/pebsim/pintos/priority-donate-multiple.c}.
 

### 2018-05-29 thesis/overview.tex
pintos testce

@@
 Being largely architectural/algorithmic projects rather than concurrency-oriented ones, I am not concerned with these assignments in this thesis.
-The main concurrency challenges in Pintos projects arise from the threads and userprog assignments:
+The main concurrency challenges in Pintos projects arise from the {\em threads} and {\em userprog} assignments:
 implementing a correct {\tt alarm} routine,

### 2018-05-30 thesis/education.tex
write abt pintos recruiting/retrospect

@@
 
+\subsection{Retrospect}
+
+Landslide is great and should be used in 410. Dave wants to use it. That's enough evidence right there.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 After each of the {\em threads} and {\em userprog} project deadlines (\sect{\ref{sec:overview-pintos}}),
-% TODO talk about th econtence of the email
+\uchos staff sent students an email inviting them to volunteer to receive Landslide's bug reports,
+disclaiming that it did not represent part of the official grading process but could help improve their future submissions.
 
@@
 
+\subsection{Retrospect}
+
+I attribute the low participation rate of Pintos students to P2 students to two major factors:
+one, not incentivizing the students to directly improve their grades
+(instead offering only the vague promise of a ``learning experience'' debugging their code only after handin),
+and two, not traveling to the university to introduce the research topic in an in-person lecture
+(leaving the students potentially confused about what advantage, if any, was offered over stress testing).
+
+While part of the point of this experimental design was to evaluate Landslide as a grading tool in the hands of TAs,
+I would be remiss to mention that I also feared the automatic annotation process would not be as robust as the P2 version.
+Indeed, while helping Kevin get oriented with using Landslide,
+I implemented
+several % TODO: how many and what specifically
+fixes/improvements to the setup scripts
+as we found student kernels that failed to automatically annotate
+(for example, those with {\tt ready\_list} changed to an array,
+as described in \sect{\ref{sec:education-pintos-instrumentation}}).
+Had we given Landslide directly to students that semester,
+the students themselves would have had to email me for tech support.
+
+Hypothetically, I could have achieved greater user study participation
+either by offering extra credit to students
+or by offering an autograder-like interface for students to receive bug reports before their deadlines instead of after
+(either way requiring a more rigorous IRB review process).
+Practically, as for non-research use in Pintos classes,
+Landslide can now handle a considerably wider variety of student implementation quirks
+on account of this time's fixes.
+In its current shape I would recommend it for TA use grading,
+but not necessarily directly to students without someone familiar with the codebase on immediate hand for tech support.
+
+% Point two is much more easily addressed, although perhaps not necessary if only being used for grading anyway.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2018-06-05 thesis/education.tex
write more education, everything but eval prety much

@@
 \begin{itemize}
+	\item {\tt p2-setup.sh} checks if the directory argument correctly points at the top-level P2 basecode directory
+		rather than any subdirectories such as {\tt user/libthread/}.
 	\item {\tt check-need-p2-setup-again.sh} checks if any source files in the original P2 source directory
@@
 		in case the student hoped to fix some bug and verify their fix but forgot to re-run the setup script.
-	% TODO
-	% pointing p2 setup at the user/libthread directory or whatever other wrong thing
-	% wrong program name like racer or OPTIONS
-	\item
+	\item {\tt landslide} checks the supplied test name matches one of the endorsed Landslide-friendly tests
+		(students love trying to run Landslide with {\tt racer}, {\tt largetest},
+		or even the string {\tt OPTIONS}).
 	\item {\tt landslide} checks if any other instance of itself is simultaneously running in the same directory,
@@
 \begin{itemize}
-		% yield loosps
-		% mutex loosp "suspicious"
-		% HURDLE_VIOLATION
-		% snapshooting
-		% psu magic post-test-case check these flags (rationale why, dont have to thr_exit -- also save state space)
-		% quicksand emit custom without_function directives depending on name of test
-		% TODO
-
-	\item
+	\item Quicksand emits different combinations of {\tt within\_function}/{\tt without\_function} directives
+		for Landslide depending on the name of the test.
+		For example, for {\tt paradise\_lost} Landslide will not preempt in a function named {\tt critical\_section()},
+		which the test case uses to protect an internal counter used to detect the bug;
+		and it will not preempt in any of the {\tt thr\_*()} thread library API functions
+		for tests intended to target just the concurrency primitives.
+		In future work this could be improved as annotations to be placed inside the test case code itself.
+	\item Landslide finds ad-hoc synchronization patterns which students often open-code,
+		rather than using the prescribed synchronization API,
+		such as {\tt while (!flag) yield();} or {\tt while (xchg(...)) continue;},
+		and treats them as synchronization points as described in \sect{\ref{sec:landslide-blocking}}.
+	\item Landslide finds ``too suspicious'' spinwait-loops in the students' mutex implementations
+		which are neither yield- nor {\tt xchg}-loops (as described above),
+		which would ordinarily be classified as infinite loop bugs,
+		and reports them with a suggestive message ({\tt undesirable\_loop\_html()} in {\tt landslide.c})
+		referring them to the appropriate lecture material
+		\cite{synchronization-2}.
+	\item The {\tt landslide} wrapper script logs the time and command-line options of invocation
+		and captures a snapshot of the student code and results of the test and saves them to AFS after each run.
+		% HURDLE_VIOLATION -- no, this is for p3 (context switcher) features only
 \end{itemize}
@@
 others are crafted to exercise generally concurrency-heavy code paths and uncover any number of unforeseen problems.
+
+\subsubsection{Test case list}
 
@@
 		Tests the {\tt cmpxchg} assembly function for being properly atomic.
-		Uses the {\tt magic\_*} global variables (\sect{\ref{sec:education-pebbles-instrumentation}}), and invokes {\tt vanish()} directly,
+		Uses the {\tt magic\_*} global variables described below, and invokes {\tt vanish()} directly,
 		to avoid requiring the student to implement {\tt thr\_join()}/{\tt thr\_exit()}
@@
 
+\subsubsection{Magic post-test assertions}
+
+Test cases may define global variables of the following names
+to instruct Landslide to assert the following corresponding predicates at the end of each test execution,
+after all threads exit %
+\footnote{Each predicate will be checked iff its first listed variable name is defined;
+if that variable is defined, all others associated must also be;
+any combination of the three first-listed variables
+}.
+(These cannot be implemented as asserts in the test code itself
+without requiring the student to implement {\tt thr\_join()} and {\tt thr\_exit()}.)
+\begin{itemize}
+	\item {\tt magic\_global\_expected\_result == magic\_global\_value}
+	\item {\tt magic\_expected\_sum == magic\_thread\_local\_value\_parent +} \\ {\tt magic\_thread\_local\_value\_child}
+	\item {\tt magic\_expected\_sum\_minus\_result == magic\_thread\_local\_value\_parent +} \\ {\tt magic\_thread\_local\_value\_child - magic\_global\_value}
+\end{itemize}
+
 The tests can all be viewed at \url{https://github.com/bblum/landslide/tree/master/pebsim/p2-basecode/410user/progs}.
@@
 
+% TODO
+
 \subsection{Retrospect}
 
+% TODO FIXME
 Landslide is great and should be used in 410. Dave wants to use it. That's enough evidence right there.
@@
 (plus several more hacks in the script itself)
-to insert the {\tt tell\_landslide\_*()} annotations (\sect{\ref{sec:tell-landslide}})
+to insert the {\tt tell\_landslide()} annotations (\sect{\ref{sec:tell-landslide}})
 into the student's kernel code.
@@
 The fourth test is available at
-% FIXME: allowbreak makes nasty latex shit show up in the output??
 \url{https://github.com/bblum/landslide/blob/master/pebsim/pintos/priority-donate-multiple.c}.
@@
 % mb anything else from timmys 2nd latest email
+
+% TODO
 

### 2018-06-14 thesis/related.tex
related own latex file, add some citations i don't wanna forget abt later

--- 2018-04-28 thesis/thesis.tex
@@
-%for a more compact document, add the option openany to avoid
-%starting all chapters on odd numbered pages
-\documentclass[12pt]{cmuthesis}
-
-% This is a template for a CMU thesis.  It is 18 pages without any content :-)
-% The source for this is pulled from a variety of sources and people.
-% Here's a partial list of people who may or may have not contributed:
-%
-%        bnoble   = Brian Noble
-%        caruana  = Rich Caruana
-%        colohan  = Chris Colohan
-%        comar    = Cyrus Omar
-%        jab      = Justin Boyan
-%        josullvn = Joseph O'Sullivan
-%        jrs      = Jonathan Shewchuk
-%        kosak    = Corey Kosak
-%        mjz      = Matt Zekauskas (mattz@cs)
-%        pdinda   = Peter Dinda
-%        pfr      = Patrick Riley
-%        dkoes = David Koes (me)
-
-% My main contribution is putting everything into a single class files and small
-% template since I prefer this to some complicated sprawling directory tree with
-% makefiles.
-
-% some useful packages
-\usepackage{fullpage}
-\usepackage{graphicx}
-\usepackage{amsmath}
-\usepackage{amssymb}
-\usepackage[numbers,sort]{natbib}
-\usepackage[backref,pageanchor=true,plainpages=false, pdfpagelabels, bookmarks,bookmarksnumbered,
-%pdfborder=0 0 0,  %removes outlines around hyper links in online display
-]{hyperref}
-\usepackage{subfigure}
-\usepackage{xspace}
-\usepackage[linesnumbered,ruled]{algorithm2e}
-\usepackage{algpseudocode}
-
-%% custom fonts
-
-\usepackage{fontspec}
-\usepackage{xeCJK}
-
-\setmainfont{XCharter}
-%\setmainfont{Utopia}[
-%	Path = ./fonts/,
-%	UprightFont = UtopiaStd-Regular.otf,
-%	BoldFont = UtopiaStd-Semibold.otf,
-%	ItalicFont = UtopiaStd-Italic.otf,
-%	BoldItalicFont = UtopiaStd-SemiboldIt.otf
-%	]
-
-\setCJKmainfont{HeiseiMinStd-W5}[Path = ./fonts/]
-
-% Approximately 1" margins, more space on binding side
-%\usepackage[letterpaper,twoside,vscale=.8,hscale=.75,nomarginpar]{geometry}
-%for general printing (not binding)
-\usepackage[letterpaper,twoside,vscale=.8,hscale=.75,nomarginpar,hmarginratio=1:1]{geometry}
-
-% Provides a draft mark at the top of the document.
-\draftstamp{\today}{DRAFT}
-
-\begin{document}
-\frontmatter
-
-%initialize page style, so contents come out right (see bot) -mjz
-\pagestyle{empty}
-
-\title{ %% {\it \huge Thesis Proposal}\\
-{\bf Practical Concurrency Testing} \\
-\normalsize \vspace{1em}
-\begin{tabular}{c}
-{\em or, How I Learned to Stop Worrying and Love the Exponential Explosion}
-\end{tabular}}
-\author{Ben Blum}
-\date{Decembruary 2018}
-\Year{2018}
-\trnumber{}
-
-\committee{
-Garth Gibson, Chair \\
-David A. Eckhardt \\
-Brandon Lucia \\
-Haryadi Gunawi, University of Chicago
-}
-
-\support{}
-\disclaimer{}
-
-% copyright notice generated automatically from Year and author.
-% permission added if \permission{} given.
-
-\keywords{landslide terminal, baggage claim, ground transportation, ticketing}
-
-\maketitle
-
-\begin{dedication}
-For Elsie, Margie, Johnny, Mel, Eve, and David.
-%, who brought me here today.
-\end{dedication}
-
-\pagestyle{plain} % for toc, was empty
-
-%% Obviously, it's probably a good idea to break the various sections of your thesis
-%% into different files and input them into this file...
-
-\begin{abstract}
-Concurrent programming presents a challenge to students and experts alike because of the complexity of multithreaded interactions and the difficulty to reproduce and reason about bugs.
-Stateless model checking is a concurrency testing approach which forces a program to interleave its threads in many different ways, checking for bugs each time.
-This technique is powerful, in principle capable of finding any nondeterministic bug in finite time,
-but suffers from exponential explosion as program size increases.
-Checking an exponential number of thread interleavings is not a practical or predictable approach for programmers to find concurrency bugs before their project deadlines.
-
-In this thesis, I develop several new techniques to make stateless model checking more practical for human use.
-I have built Landslide, a stateless model checker specializing in undergraduate operating systems class projects.
-Landslide extends the traditional model checking algorithm with a new framework for automatically managing
-%the exploration of
-multiple state spaces according to their estimated completion times,
-which I show quickly finds bugs should they exist and also quickly verifies correctness otherwise.
-I evaluate Landslide's suitability for inexpert use by presenting the results of many semesters providing it to students in 15-410, CMU's Operating System Design and Implementation class,
-and more recently, students in similar classes at the University of Chicago and Berkeley.
-%Finally, I will present several new techniques that allow stateless model checking to be practically employed on real-world programs.
-%Finally, I will explore broader impact by extending Landslide to test some real-world programs and to be used by students at other universities.
-Finally, I extend Landslide with a new concurrency model for hardware transactional memory,
-and evaluate several real-world transactional benchmarks to show that stateless model checking can keep up with the developing concurrency demands of real-world programs.
-\end{abstract}
-
-\begin{acknowledgments}
-% things to thank
-% stimhack
-% csd bgames
-% undergrad frenz
-% new frenz
-% 410 ta community
-% emarhavil
-%
-% \\
-
-I thank my dog, Louie.
-\end{acknowledgments}
-
-\tableofcontents
-% TODO: figure out how to make latex abbreviate these
-\listoffigures
-\listoftables
-
-\mainmatter
-
-%% Double space document for easy review:
-%\renewcommand{\baselinestretch}{1.66}\normalsize
-
-% The other requirements Catherine has:
-%
-%  - avoid large margins.  She wants the thesis to use fewer pages,
-%    especially if it requires colour printing.
-%
-%  - The thesis should be formatted for double-sided printing.  This
-%    means that all chapters, acknowledgements, table of contents, etc.
-%    should start on odd numbered (right facing) pages.
-%
-%  - You need to use the department standard tech report title page.  I
-%    have tried to ensure that the title page here conforms to this
-%    standard.
-%
-%  - Use a nice serif font, such as Times Roman.  Sans serif looks bad.
-%
-% Other than that, just make it look good...
-
-\newcommand\simics{\textsc{Simics}}
-\newcommand{\sect}[1]{\S #1}
-\newcommand\hilight[2]{\color{#1}#2\color{black}}
-\definecolor{orange}{RGB}{192,96,0}
-\definecolor{olivegreen}{RGB}{0,127,0}
-\definecolor{brickred}{RGB}{192,0,0}
-\definecolor{commentblue}{RGB}{0,0,192}
-
-\newtheorem{theorem}{Theorem}
-
-\newcommand\llama[1]{\includegraphics[width=#1]{llama.pdf}}
-\newcommand\llitem{\item[\llama{1.2em}]}
-
-\newcommand{\inspirationallinebreak}{\vspace{0.25em}}
-\newcommand{\inspirationalhyphen}{-\hspace{-0.15em}-\xspace}
-\newcommand\inspirationalquote[2]{\begin{flushright}
-	\vspace{-1em}
-	{
-	\fontspec[Path=./fonts/]{AlexaStd}
-	{\em #1}
-	\inspirationallinebreak
-
-	-\hspace{-0.15em}-\hspace{-0.15em}-{#2}
-	}
-	\vspace{2em}
-\end{flushright}}
-
-\input{intro}
-\input{overview}
-\input{landslide}
-\input{quicksand}
-\input{education}
-\input{tm}
-
 \chapter{Related Work}
@@
 
-\chapter{Future Work}
-\label{chap:warpzone}
-% TODO: ask for permission for this one
-% https://twitter.com/Poem4your_sprog/status/878747092788838400
-\inspirationalquote{
-\begin{tabular}{p{0.51\textwidth}}
-A little bit of being kind,
-a tiny open door; \\
-A nicer slice of peace of mind
-begets a little more. \\
-It won't correct the world tonight,
-nor change tomorrow too; \\
-But maybe if you do it right,
-you'll find it changes you.
-\end{tabular}}
-{Sam "Poem\_for\_your\_sprog" Garland}
+% things to cite what you haven't alerady got cited somewhere
 
-\chapter{Conclusion}
-\label{chap:conclusion}
-\inspirationalquote{
-	\begin{tabular}{p{0.84\textwidth}}
-	"If you are a god, Zeus, as the stories claim, then why did you create evolution?
-	Why did you make a world that can only grow through cruelty and pain?"
-	\\
-	For a moment\inspirationalhyphen{}surely this meant death was near\inspirationalhyphen{}she thought she heard him answer:
-	"My child, I was shaped by the gods that came before me, as you were shaped by me.
-	The choice I had was between creation and oblivion, life and death.
-	And I chose life, because any life is better than no life,
-	because as long as there is life, there is hope\inspirationalhyphen{}if not for us, then for some generation to come."
-	\\
-	"Then how are you a god, if you can offer so little?" she whispered, feeling death creep closer.
-	\\
-	"I am a god because I take upon myself the burden of creation," the statue replied.
-	\\
-	"Then we are all gods," Alexandra said, and pushed the button.
-	\end{tabular}}
-%{Galatea, The Talos Principle}
-{The Talos Principle}
+% dejafu - haskell mcing
+% relacy - smth for go?
+% jepsen-io/jepsen
 
-asdfasdasdf
-%\appendix
-%\include{appendix}
+% cchtm - the intel ismm17 paper on nvram htm
+% stamp (mb just cite from txn section)
 
-\backmatter
+\section{Stateless Model Checking}
 
-%\renewcommand{\baselinestretch}{1.0}\normalsize
+D\'{e}j\`{a} Fu \cite{dejafu}
 
-% By default \bibsection is \chapter*, but we really want this to show
-% up in the table of contents and pdf bookmarks.
-\renewcommand{\bibsection}{\chapter{\bibname}
-\inspirationalquote{Only a doctor of philosophy, Darth.}{Robert Marsh}
-}
-%\renewcommand{\bibpreamble}{This text goes between the ``Bibliography''
-%  header and the actual list of references}
-\bibliographystyle{plainnat} % plain n'at
-\bibliography{citations}
+\section{Education}
 
-\end{document}
+Willgrind \cite{willgrind}

### 2018-06-14 thesis/thesis.tex
add tm proofs
put example figure and syntax hilight it
add some todo commence
mysterious latex wizardry from nwf that doesn't seem to work
better dedication
write more education, everything but eval prety much
related own latex file, add some citations i don't wanna forget abt later

@@
 \usepackage{amssymb}
+\usepackage{amsthm}
 \usepackage[numbers,sort]{natbib}
+\PassOptionsToPackage{hyphens}{url}
 \usepackage[backref,pageanchor=true,plainpages=false, pdfpagelabels, bookmarks,bookmarksnumbered,
@@
 \draftstamp{\today}{DRAFT}
+
+% % nwf made this but i don't know how to get it to work?
+% \usepackage{etoolbox}
+% \makeatletter
+% \apptocmd{\ref}{%
+%   \immediate\write\@auxout%
+%     {\expandafter\string\csname nwf@backref\endcsname{#1}{\arabic{page}}}%
+% }{}{}
+%
+% \newcommand{\nwf@backref}[2]{%
+%   \ifcsdef{nwf@backrefs@#1}%
+%    {\csgappto{nwf@backrefs@#1}{, #2}}%
+%    {\csgdef{nwf@backrefs@#1}{#2}}%
+% }
+% \AtEndDocument{\let\nwf@backref\relax}
+%
+% \newcommand{\linksto}[1]{\csname nwf@backrefs@#1\endcsname}
+% \makeatother
 
@@
 \begin{dedication}
-For Elsie, Margie, Johnny, Mel, Eve, and David.
-%, who brought me here today.
+%For Elsie, Margie, Johnny, Mel, Eve, and David.
+%%, who brought me here today.
+For my family, my teachers, and my students.
 \end{dedication}
@@
 \definecolor{commentblue}{RGB}{0,0,192}
+% colors from sigbovik for tsx/etc
+% threads
+\definecolor{lavender}{RGB}{102,65,208} % (?? i don't know what happened here anymore but it looks good)
+\definecolor{seafoam}{RGB}{56,158,68} % * 2/3
+\definecolor{salmon}{RGB}{208,89,74} % * 7/8
+% eval data
+\definecolor{darkcyan}{RGB}{0,128,128}
+\definecolor{lime}{RGB}{48,128,0}
+\definecolor{darklavender}{RGB}{68,44,138}
+\definecolor{brownish}{RGB}{128,96,0}
+\definecolor{pinkish}{RGB}{128,34,102} % ff4488
+% syntax hilighting
+\newcommand\flow[1]{\hilight{brown}{#1}\xspace}
+\newcommand\const[1]{\hilight{brickred}{#1}\xspace}
+\newcommand\call[1]{\hilight{pinkish}{#1}}
+\newcommand\ctype[1]{\hilight{olivegreen}{#1}\xspace}
 
+\newtheorem{lemma}{Lemma}
 \newtheorem{theorem}{Theorem}
+\newtheorem{definition}{Definition}
 
 \newcommand\llama[1]{\includegraphics[width=#1]{llama.pdf}}
-\newcommand\llitem{\item[\llama{1.2em}]}
+\newcommand\llitem{\item[\raisebox{-0.15em}{\llama{1.2em}}]}
 
@@
 \input{tm}
-
-\chapter{Related Work}
-\label{chap:relatedwork}
-\inspirationalquote{
-\begin{tabular}{p{0.58\textwidth}}
-To test if your paper makes a genuine contribution to its discipline,
-see if you can afford a generous tone in the "Related Work" section.
-\end{tabular}}
-{Conor McBride}
+\input{related}
 

### 2018-06-15 thesis/education.tex
describe mysterious test case annos in landslide chapter

@@
 
-\section{P2}
+\section{Pebbles}
 
@@
 \subsection{Test cases}
+\label{sec:education-pebbles-tests}
 
@@
 others are crafted to exercise generally concurrency-heavy code paths and uncover any number of unforeseen problems.
+Many use some of the features/annotations described in \sect{\ref{sec:landslide-testcases}}.
 
@@
 		As {\tt broadcast\_test}, but uses two waiting threads to ensure both get signalled.
-\end{itemize}
-
-\subsubsection{Magic post-test assertions}
-
-Test cases may define global variables of the following names
-to instruct Landslide to assert the following corresponding predicates at the end of each test execution,
-after all threads exit %
-\footnote{Each predicate will be checked iff its first listed variable name is defined;
-if that variable is defined, all others associated must also be;
-any combination of the three first-listed variables
-}.
-(These cannot be implemented as asserts in the test code itself
-without requiring the student to implement {\tt thr\_join()} and {\tt thr\_exit()}.)
-\begin{itemize}
-	\item {\tt magic\_global\_expected\_result == magic\_global\_value}
-	\item {\tt magic\_expected\_sum == magic\_thread\_local\_value\_parent +} \\ {\tt magic\_thread\_local\_value\_child}
-	\item {\tt magic\_expected\_sum\_minus\_result == magic\_thread\_local\_value\_parent +} \\ {\tt magic\_thread\_local\_value\_child - magic\_global\_value}
 \end{itemize}
@@
 \subsection{Test cases}
+\label{sec:education-pintos-tests}
 

### 2018-06-15 thesis/landslide.tex
describe mysterious test case annos in landslide chapter

@@
 \end{itemize}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\subsection{Test cases}
+\label{sec:landslide-testcases}
+
+Landslide depends on human intuition to construct a test case
+that will produce both meaningful in quality and manageable in quantity thread interleavings.
+
+The user may supply custom test cases
+for Pebbles (under {\tt pebsim/p2-basecode})
+by creating a file in {\tt 410user/progs} and adding it to config.mk as usual,
+or for Pintos (under {\tt pebsim/pintos/src/tests/threads})
+by creating a file and adding it to both {\tt tests.c} and {\tt Make.tests}.
+Tests for the most common interactions during the P2 and Pintos projects are of course already supplied,
+as described in \sect{\ref{sec:education-pebbles-tests}} and \sect{\ref{sec:education-pintos-tests}}.
+
+Use of {\tt tell\_landslide()} annotations is not necessary,
+although {\tt tell\_landslide\_\allowbreak{}preempt()} and {\tt tell\_landslide\_dump\_stack()}
+may optionally be used at the user's convenience.
+Additionally, the following ``secret'' annotations
+are occasionally used in the pre-supplied test cases
+to accomplish several mysterious goals described hereupon.
+
+\subsubsection{Magic post-test assertions}
+
+Test cases may define global variables of the following names
+to instruct Landslide to assert the following corresponding predicates at the end of each test execution,
+after all threads exit.
+Each predicate will be checked iff its first listed variable name is defined;
+if that variable is defined, all others associated must also be;
+any combination of the three first-listed variables may be specified at the user's option.
+\begin{itemize}
+	\item {\tt magic\_global\_expected\_result == magic\_global\_value}
+	\item {\tt magic\_expected\_sum == magic\_thread\_local\_value\_parent +} \\ {\tt magic\_thread\_local\_value\_child}
+	\item {\tt magic\_expected\_sum\_minus\_result == magic\_thread\_local\_value\_parent +} \\ {\tt magic\_thread\_local\_value\_child - magic\_global\_value}
+\end{itemize}
+These could not be implemented as asserts in the test code itself
+without requiring the student to implement {\tt thr\_join()} and {\tt thr\_exit()},
+avoiding which is important for tests to be student-accessible earlier in the project implementation timeline.
+
+\subsubsection{Misbehave}
+
+Many of the P2 test cases can be seen to invoke the {\tt misbehave()} system call
+with a mysterious argument (usually {\tt BGND\_BRWN >> FGND\_CYAN})
+before the creation of any child threads.
+The use of terminal color code constants is of course a red herring of obfuscation,
+as the true nature of the Pathos reference kernel's {\tt misbehave} modes
+is a closely-guarded secret among 15-410 course staff.
+The mode in question causes the reference kernel to prioritize scheduling the child thread over the parent
+whenever {\tt thread\_fork} is called,
+and the target thread over the invoking thread
+whenever {\tt make\_runnable} is called,
+which are necessary to allow Landslide to recognize a {\tt yield()} preemption point
+and be able to run the newly-runnable thread as soon as possible.
+
+To illustrate, consider the following program.
+
+% TODO
+% TODO - make a progr that data races immediately after thr create, no mutex es or whatever.
+% TODO - check if it rly happens under pure happens before? or if this is only a concern under limited HB.
+
+Other {\tt misbehave} modes may be used, but are likely to have no effect,
+since Landslide's thread-scheduling algorithm will override
+any Pathos-internal scheduling priorities that may arise therefrom.
+Hypothetically speaking, a reader with access to the top-secret Pebbles source code
+could find further {\tt misbehave} documentation in its {\tt inc/misbehavior.h}.
 

### 2018-07-01 thesis/landslide.tex
minor revisions

@@
 
-Many of the P2 test cases can be seen to invoke the {\tt misbehave()} system call
+Many of the supplied P2 test cases invoke the {\tt misbehave()} system call
 with a mysterious argument (usually {\tt BGND\_BRWN >> FGND\_CYAN})
@@
 any Pathos-internal scheduling priorities that may arise therefrom.
-Hypothetically speaking, a reader with access to the top-secret Pebbles source code
+Hypothetically speaking, a reader with access to the top-secret Pathos source code
 could find further {\tt misbehave} documentation in its {\tt inc/misbehavior.h}.
@@
 Each of these configurations may contain any number of
-{\tt within\_function}, {\tt without\_function}, and {\tt data\_race} commands.
+{\tt within\_function}, {\tt without\_function}, and {\tt data\_\allowbreak{}race} commands.
 

### 2018-07-02 thesis/education.tex
write abt the surbeys

@@
 \end{itemize}
-Landslide also includes several P2-specific instrumentations and features to cope with various student irregularities:
+\vspace{1em}
+
+\noindent Landslide also includes several P2-specific instrumentations and features to cope with various student irregularities:
 \begin{itemize}
@@
 \end{itemize}
-The following tests were released to PSU students, in addition to the ones above:
+\vspace{1em}
+
+\noindent The following tests were released to PSU students, in addition to the ones above:
 \begin{itemize}
@@
 \end{itemize}
+\vspace{1em}
 
-The tests can all be viewed at \url{https://github.com/bblum/landslide/tree/master/pebsim/p2-basecode/410user/progs}.
+\noindent The tests can all be viewed at \url{https://github.com/bblum/landslide/tree/master/pebsim/p2-basecode/410user/progs}.
+
+\subsection{Survey}
+\label{sec:education-survey-pebbles}
+
+Starting in Fall 2017, I sought to gauge the students' personal opinions on their experience with Landslide,
+in addition to simply counting
+from the automatic snapshots
+how many bugs were found.
+Shortly after the P2 submission deadline,
+I asked participants to answer several survey questions, reproduced below.
+%distributed via email as a Google Doc.
+
+\begin{enumerate}
+	\item How many bugs did Landslide help you find in your code? (Please indicate a number.)
+	\item How many of the bugs you found with Landslide do you believe you fixed before submitting your project? (You may answer "all", "none", or a number.)
+	\item How many of the bugs you found with Landslide did you verify you had fixed by running Landslide again to make sure the bug was gone? (You may answer "all", "none", or a number.)
+	\item In addition to the bugs Landslide found, did it report anything that you believe was NOT a bug?
+For example, Landslide printed an execution trace that was actually impossible, or Landlside reported a bug about some behaviour that was actually allowed by the P2 specification.
+(If so, please describe.)
+	\item I found Landslide's debugging output easy to understand.
+		(Multiple choice from strongly disagree to strongly agree.)
+	\item It's easier to diagnose the root cause of a bug with Landslide than with a stress test (e.g. juggle).
+		(Multiple choice from strongly disagree to strongly agree; plus "Not sure" and "Easier for some bugs but harder for others")
+	\item I felt the time I saved by having Landslide to help debug was worth the time it took me to learn how to use Landslide.
+		(Multiple choice from strongly disagree to strongly agree.)
+	\item I feel that by using Landslide I learned to understand concurrency better.
+		(Multiple choice from strongly disagree to strongly agree.)
+	\item Suppose after you submitted your project, % "P2" for cmu, "project" for pintos
+		we gave you 100 CPU-hours on the cloud provider of your choice to test it.
+Then we extended the project deadline by a day for you to use the results to fix bugs and get partial credit.
+How would you divide that CPU time between the staff-provided stress tests and Landslide?
+		(Multiple choice: 0/10/.../100 CPU-hours on Landslide, 100/90/.../0 CPU-hours on stress tests.)
+	\item If I found out next semester that a friend of mine (or a student in my degree program) were taking OS, I would recommend that they should probably invest some time during the project % "P2" for cmu, as above
+		to learn Landslide and try to find bugs with it.
+		(Multiple choice from strongly disagree to strongly agree.)
+	\item Regarding the previous question, why or why not?
+\suspend{enumerate}
+\vspace{1em}
+
+\noindent The following questions were served only on the CMU version of the survey.
+\resume{enumerate}
+	\item Did you answer this survey together with your partner, or on your own while they were busy? (If you both have time for it, please try to submit one survey together.)
+		(Multiple choice: together or alone)
+	\item Your andrew ID
+	\item Your partner's andrew ID (if any)
+\suspend{enumerate}
+\vspace{1em}
+
+\noindent The following questions were served only on the PSU version of the survey.
+\resume{enumerate}
+	\item Any feedback on how Landslide's user interface could be improved / made easier to use or understand?
+(setup process, messages printed while running, or the execution trace / stack traces emitted after a bug is found?)
+	\item Your PSU username
+\end{enumerate}
 
@@
 \end{itemize}
+\vspace{1em}
 
@@
 \url{https://github.com/bblum/landslide/blob/master/pebsim/pintos/priority-donate-multiple.c}.
+
+\subsection{Survey}
+
+Similar to the survey for Pebbles projects
+\sect{\ref{sec:education-survey-pebbles}},
+I surveyed the Pintos user study participants for their opinions.
+Because of the different nature of the user study, of course,
+the questions here focus more on the debugging experience than on using Landslide directly.
+
+\begin{enumerate}
+	\item How many Landslide bug reports did you receive from course staff? (Please indicate a number.)
+	\item Among those bug reports, how many were you able to diagnose and recognize the root cause in your code? (You may answer "all", "none", or a number.)
+	\item Among those bug reports, how many described a behaviour that you believe was NOT a bug?
+For example, Landslide printed an execution trace that was actually impossible, or Landslide reported a bug about some behaviour that was actually allowed by the Pintos specification.
+(You may answer "all", "none", or a number.)
+	\item About how much time did you spend interpreting Landslide's debugging output? (Please indicate a number of minutes, or a range if uncertain, e.g. "30-60 minutes".)
+	\item I found Landslide's debugging output easy to understand.
+		(Multiple choice from strongly disagree to strongly agree.)
+	\item It's easier to diagnose the root cause of a bug with Landslide than with a stress test (for example "exec-multiple").
+		(Multiple choice from strongly disagree to strongly agree; plus "Not sure" and "Easier for some bugs but harder for others")
+	\item I feel that by interpreting Landslide's debugging output I learned to understand concurrency better.
+		(Multiple choice from strongly disagree to strongly agree.)
+	\item These kinds of concurrency bugs are important to fix, even though they don't count against my grade.
+		(Multiple choice from strongly disagree to strongly agree.)
+	\item Suppose after you submitted your pintos, we gave you 100 CPU-hours on the cloud provider of your choice to test it.
+Then we extended the project deadline by a day for you to use the results to fix bugs and get partial credit.
+How would you divide that CPU time between the staff-provided stress tests and Landslide?
+		(Multiple choice: 0/10/.../100 CPU-hours on Landslide, 100/90/.../0 CPU-hours on stress tests.)
+	\item If course staff were to allow students to resubmit updated code after reviewing Landslide bug reports to receive partial credit for each bug that had been fixed, it would be worth my time to try that (even if I could be spending that time working on the next project instead).
+		(Multiple choice from strongly disagree to strongly agree.)
+	\item If a friend of mine took OS next semester, I would recommend that they should sign up to receive Landslide bug reports during projects in the future.
+		(Multiple choice from strongly disagree to strongly agree.)
+	\item Regarding the previous question, why or why not?
+	\item Your name
+	\item Your project partner's name (if applicable)
+\end{enumerate}
 

### 2018-07-02 thesis/education.tex
quotes

@@
 	\item How many bugs did Landslide help you find in your code? (Please indicate a number.)
-	\item How many of the bugs you found with Landslide do you believe you fixed before submitting your project? (You may answer "all", "none", or a number.)
-	\item How many of the bugs you found with Landslide did you verify you had fixed by running Landslide again to make sure the bug was gone? (You may answer "all", "none", or a number.)
+	\item How many of the bugs you found with Landslide do you believe you fixed before submitting your project? (You may answer ``all'', ``none'', or a number.)
+	\item How many of the bugs you found with Landslide did you verify you had fixed by running Landslide again to make sure the bug was gone? (You may answer ``all'', ``none'', or a number.)
 	\item In addition to the bugs Landslide found, did it report anything that you believe was NOT a bug?
@@
 	\item It's easier to diagnose the root cause of a bug with Landslide than with a stress test (e.g. juggle).
-		(Multiple choice from strongly disagree to strongly agree; plus "Not sure" and "Easier for some bugs but harder for others")
+		(Multiple choice from strongly disagree to strongly agree; plus ``Not sure'' and ``Easier for some bugs but harder for others'')
 	\item I felt the time I saved by having Landslide to help debug was worth the time it took me to learn how to use Landslide.
@@
 	\item How many Landslide bug reports did you receive from course staff? (Please indicate a number.)
-	\item Among those bug reports, how many were you able to diagnose and recognize the root cause in your code? (You may answer "all", "none", or a number.)
+	\item Among those bug reports, how many were you able to diagnose and recognize the root cause in your code? (You may answer ``all'', ``none'', or a number.)
 	\item Among those bug reports, how many described a behaviour that you believe was NOT a bug?
 For example, Landslide printed an execution trace that was actually impossible, or Landslide reported a bug about some behaviour that was actually allowed by the Pintos specification.
-(You may answer "all", "none", or a number.)
-	\item About how much time did you spend interpreting Landslide's debugging output? (Please indicate a number of minutes, or a range if uncertain, e.g. "30-60 minutes".)
+(You may answer ``all'', ``none'', or a number.)
+	\item About how much time did you spend interpreting Landslide's debugging output? (Please indicate a number of minutes, or a range if uncertain, e.g. ``30-60 minutes''.)
 	\item I found Landslide's debugging output easy to understand.
 		(Multiple choice from strongly disagree to strongly agree.)
-	\item It's easier to diagnose the root cause of a bug with Landslide than with a stress test (for example "exec-multiple").
-		(Multiple choice from strongly disagree to strongly agree; plus "Not sure" and "Easier for some bugs but harder for others")
+	\item It's easier to diagnose the root cause of a bug with Landslide than with a stress test (for example {\tt exec-multiple}).
+		(Multiple choice from strongly disagree to strongly agree; plus ``Not sure'' and ``Easier for some bugs but harder for others'')
 	\item I feel that by interpreting Landslide's debugging output I learned to understand concurrency better.

### 2018-07-02 thesis/education.tex
reorganize education eval/bias/retro and setup eval Qs

@@
 
-\subsection{Evaluation}
-
-% TODO
-
-\subsection{Retrospect}
-
-% TODO FIXME
-Landslide is great and should be used in 410. Dave wants to use it. That's enough evidence right there.
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-\subsection{Evaluation}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
+\section{Evaluation}
+
+I pose the following evaluation questions.
+
+\begin{itemize}
+	\item How many bugs, and of what severities, does Landslide help students find and fix before submitting their code?
+	\item Does Landslide use result in higher quality submissions, whether directly for P2, or for subsequent projects as well?
+	\item Do students feel the experience is worthwhile, as compared to stress testing?
+	\item How well does Landslide apply to operating systems projects outside of CMU?
+\end{itemize}
+\vspace{1em}
+
+The data set is comprised of Landslide's automatically-generated usage snapshots
+from CMU students from semesters Spring 2015 to Spring 2018 inclusive,
+% and PSU \psuos from Spring 2018, % no snapshots from psu
+CMU's official project grades from same,
+CMU students' survey responses and submitted project code from Fall 2017 and Spring 2018
+PSU students' survey responses and submitted project code from Spring 2018,
+and U. Chicago students' survey responses from Fall 2017.
+
+\subsection{Bug-finding}
+
+\subsection{Student submission quality}
+
+\subsection{Student opinions}
+
+\subsection{Other universities}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{Discussion}
+
+\subsection{Bias}
+
+% TODO: addressing bias
+% well, we did the best we could(?)
+% survey email: "please answer honestly rather than flatteringly"
+% anything else?
+
+%%%% pintos %%%%
 % survey questions i WISH i had asked
@@
 
+\subsection{Retrospect}
+
+Human subjects research is inherently messy.
+More than just drawing firm conclusions from
+the opinions of students who are just learning concurrency to begin with,
+student feedback in turn guided the constant development
+of Landslide, and the experimental design itself,
+as the semesters went by.
+In this section I will fantasize about how I might have run more perfect experiments
+granted the impossible wish of knowing then what I know now.
+
+\subsubsection{Pebbles}
+
 % TODO
 
-\subsection{Retrospect}
+\subsubsection{Pintos}
 
@@
 % Point two is much more easily addressed, although perhaps not necessary if only being used for grading anyway.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-% \section{Survey results} % ???
-
-% TODO: addressing bias
-% well, we did the best we could(?)
-% survey email: "please answer honestly rather than flatteringly"
-% anything else?

### 2018-07-03 thesis/education.tex
describe the buges im investigating

@@
 \subsection{Bug-finding}
+\label{sec:education-eval-bugfinding}
 
@@
 
+I evaluated Landslide's impact on student submission quality in two ways.
+First, I analyzed the overall P2 grades of CMU 15-410 students between Landslide users and non-users,
+to see if Landslide helped them submit overall better implementations,
+as well as their later overall P3 (kernel project) grades,
+to see if Landslide helped them internalize any concurrency debugging skills
+that would help write better kernels even without Landslide's aid.
+% This experiment produced inconclusive or negative results. % say this?
+
+Second, I studied
+several % how many? 3
+individual concurrency bugs that I thought Landslide was likely to detect,
+all already well-known by 15-410 course staff,
+to see if using Landslide correlated with submitting projects absent those bugs.
+% results..?
+
+\subsubsection{Impact on grades}
+
+% TODO
+
+\subsubsection{Common bugs}
+
+Based on the results from \sect{\ref{sec:education-eval-bugfinding}},
+I selected 3 bugs to study in depth to ascertain
+whether Landslide played an instrumental difference
+in helping the students ultimately submit respecitvely correct implementations.
+To avoid bias of picking too obscure and/or trivial bugs
+that Landslide alone might find but even course staff would not expect students to solve,
+I chose only bugs which had substantial penalties in the grading rubric
+(guided, as well, by my own intuition as a former TA).
+% TODO: the results are shown in table...
+
+{\bf Thread exit use-after-free.}
+During {\tt thr\_exit()}, a thread must arrange for its own stack memory allocation to be freed.
+A common student pitfall is to allow a {\tt thr\_join()}ing thread to free said stack space
+while {\tt thr\_exit()} is still executing userspace C code (which inevitably accesses the stack),
+or even for {\tt thr\_exit()} to do it itself.
+In the former case, threads must interleave specifically to exhibit a use-after-free;
+in the latter case, the use-after-free will be deterministic (i.e., present in all interleavings).
+Besides Landslide, however,
+the students have no Valgrind-like heap debugging tool which would report a bug immediately upon any illegal heap access;
+without such, a third thread would have to race to recycle the memory for a new thread stack
+and conflict with the old exiting thread
+before any problem could be detected.
+% TODO: mark which of the tej bugs were deterministic vs race
+%
+The Landslide-friendly test {\tt thr\_exit\_join} is most likely to expose this bug, and
+the official rubric penalties for it are
+{\tt :exit\_stack\_minor -2} and
+{\tt :exit\_stack\_race -4} (for the former case),
+and
+{\tt :uses\_freed\_stack -4} (for the latter).
+
+{\bf Paraguay.}
+Named after a certain Paraguayan 15-410 TA from years gone by,
+this refers to a condition variable bug in which a thread which sleeps on two different condition variables
+can spuriously wake up early from the later {\tt cond\_wait()}.
+The precise reasoning why many common na\"ive student implementations are susceptible to this bug,
+as well as the 3 major ways of fixing it,
+are one of 15-410 staff's closely-guarded secrets.
+Suffice to say that the test invokes a specially-made Pathos {\tt misbehave} mode
+which biases thread scheduling towards the specific interleaving required to exhibit the bug
+(Landslide, of course, replaces this {\tt misbehav}iour with model checking).
+% TODO: i feel tihs sentence should go somewhere else, like after analysis, not in description
+Hence, despite being a subtle concurrency bug,
+the official course test suite is likely to expose it,
+so comparing how many Landslide users and non-users submitted this bug in particular
+speaks more to the impact of Landslide's preemption traces
+in helping students diagnose a bug that a ``stress'' test could already find.
+%
+The Landslide-friendly test {\tt paraguay} is most likely to expose this bug, and
+the official rubric penalty for it is
+{\tt :paraguay -2}.
+
+{\bf R/W-lock downgrade.}
+In addition to the standard R/W-lock interface,
+P2 requires students to implement {\tt rwlock\_downgrade()}:
+called with the lock held in write mode,
+the caller adopts the reader role instead,
+allowing other waiting reader threads to proceed simultaneously,
+all while allowing no waiting writers to access in between.
+The Landslide-friendly test {\tt rwlock\_downgrade\_read\_test}
+checks that readers are allowed simultaneous access after a downgrade, and
+the official rubric penalties related to this bug are
+{\tt :downgrade\_no\_signal -1},
+{\tt :unsafe\_rwlock\_downgrade -2},
+{\tt :downgrade\_lies -3}, and
+{\tt :downgrade\_deadlock -4}.
+
 \subsection{Student opinions}
 
+% TODO
+
 \subsection{Other universities}
+
+% TODO
 
@@
 \section{Discussion}
+
+% TODO intro text
 
@@
 Human subjects research is inherently messy.
-More than just drawing firm conclusions from
+More than just trying to draw firm conclusions from
 the opinions of students who are just learning concurrency to begin with,

### 2018-07-03 thesis/education.tex
table up the p2 bug study, first 3 bugs at least

@@
 % This experiment produced inconclusive or negative results. % say this?
-
+%
 Second, I studied
@@
 (guided, as well, by my own intuition as a former TA).
-% TODO: the results are shown in table...
+Table~\ref{tab:eval-common-bugs} presents the results.
+
+\begin{table}[h]
+	\begin{center}
+	\begin{tabular}{r||cc|cc|c}
+		& \multicolumn{2}{c|}{Landslide users} & \multicolumn{2}{c|}{Non-users} & \\
+		Bug name & \#correct & total & \#correct & total & $\Delta P(\mathsf{correct})$ \\
+		\hline
+		%Exit UAF  & 9	& 21	& 2	& 16	& $+30.4\%$ \\ % not filtering out did not run tej; also using "old" thing..?
+		Exit UAF  & 11	& 21	& 0	& 16	& $+52.4\%$ \\
+		% nb. this is not filter out the did not run paras -- smae for all of the above/below
+		% Paraguay  & 22	& 24	& 12	& 13	& $-0.64\%$ \\ % not filtering out did not run para
+		Paraguay  & 20	& 21	& 14	& 16	& $+07.7\%$ \\
+		% nb this is any downgrade related ding not just dgr tested ones
+		% nb 1 is msising bc didn't impl rwlock at all
+		%Downgrade & 17	& 24	& 8	& 12	& $+4.17\%$ \\ % not filtering out either did not run, or unrel ding
+		Downgrade & 10	& 13	& 19	& 23	& $-05.7\%$ \\
+	\end{tabular}
+	\end{center}
+	\caption{
+		Students who used Landslide in general but whose usage snapshots showed they did not run the test case in question are counted as non-Landslide-users here.
+		$\Delta P$ stands for the increased probability that a student submits a correct P2
+		(with respect to each bug)
+		given that they use Landslide, versus not using Landslide; i.e.,
+		$P(\mathsf{correct}|\mathsf{landslide})-
+		P(\mathsf{correct}|\mathsf{non\text{-}user})$.
+		}
+	\label{tab:eval-common-bugs}
+\end{table}
 
@@
 % TODO: mark which of the tej bugs were deterministic vs race
-%
+
 The Landslide-friendly test {\tt thr\_exit\_join} is most likely to expose this bug, and
@@
 {\tt :uses\_freed\_stack -4} (for the latter).
+Whether each student's submitted P2 solved or suffered from this bug
+was determined by me personally inspecting their code.
 
 {\bf Paraguay.}
-Named after a certain Paraguayan 15-410 TA from years gone by,
+Named after a certain Paraguayan 15-410 TA from years gone by who originally discovered it,
 this refers to a condition variable bug in which a thread which sleeps on two different condition variables
@@
 are one of 15-410 staff's closely-guarded secrets.
-Suffice to say that the test invokes a specially-made Pathos {\tt misbehave} mode
+Suffice to say that the test {\tt paraguay} invokes a specially-made Pathos {\tt misbehave} mode
 which biases thread scheduling towards the specific interleaving required to exhibit the bug
@@
 in helping students diagnose a bug that a ``stress'' test could already find.
-%
-The Landslide-friendly test {\tt paraguay} is most likely to expose this bug, and
+
+{\tt paraguay} is itself the Landslide-friendly test for its eponymous bug, and
 the official rubric penalty for it is
 {\tt :paraguay -2}.
+Whether each student's submitted P2 solved or suffered from this bug
+was determined by me personally inspecting their code.
 
@@
 all while allowing no waiting writers to access in between.
+
 The Landslide-friendly test {\tt rwlock\_downgrade\_read\_test}
@@
 {\tt :downgrade\_no\_signal -1},
-{\tt :unsafe\_rwlock\_downgrade -2},
+%{\tt :unsafe\_rwlock\_downgrade -2}, % don't think this is right; not relevant anyway (overlaps)
+{\tt :downgrade\_partial\_signal -1}, % not sure if this means smth that dgr would expose
 {\tt :downgrade\_lies -3}, and
 {\tt :downgrade\_deadlock -4}.
+Whether each student's submitted P2 solved or suffered from this bug
+was determined by checking their grade files for any of the above-listed penalties
+(which were assessed by the TAs as a result of their manual inspection, rather than mine).
 

### 2018-07-05 thesis/education.tex
survey resulce & false positives

@@
 \subsection{Survey}
+\label{sec:education-survey-pintos}
 
@@
 
-\subsection{Student opinions}
+\subsection{Survey responses}
 
-% TODO
+Analyzing only the raw technical data of how users interacted with Landslide can paint only part of the picture.
+For one, offering students better testing and debugging tools
+may not necessarily find strictly more bugs or help students submit more correct implementations
+than with stress testing;
+it may instead find the same bugs faster,
+affording the students more free time apart from the project,
+a quality-of-life improvement normally invisible to graders.
+For two, Landslide's automatic snapshots, being captured at the time of issuing each preemption trace,
+necessarily miss the student's subsequent experience interpreting them.
+The surveys listed in sections \sect{\ref{sec:education-survey-pebbles}} and \sect{\ref{sec:education-survey-pintos}}
+serve to probe these more qualitative aspects of the Landslide experience.
+
+\begin{figure}[p]
+	\begin{center}
+		\begin{tabular}{cc}
+			\includegraphics[width=0.42\textwidth]{survey1.pdf} &
+			\includegraphics[width=0.42\textwidth]{survey2.pdf} \\
+			\includegraphics[width=0.42\textwidth]{survey3.pdf} &
+			\includegraphics[width=0.42\textwidth]{survey4.pdf} \\
+			\includegraphics[width=0.42\textwidth]{survey5.pdf} &
+			\includegraphics[width=0.42\textwidth]{survey6.pdf} \\
+			\includegraphics[width=0.42\textwidth]{survey7.pdf} &
+			\includegraphics[width=0.42\textwidth]{survey8.pdf} \\
+			\includegraphics[width=0.42\textwidth]{survey9.pdf} &
+			\includegraphics[width=0.42\textwidth]{survey10.pdf} \\
+		\end{tabular}
+	\end{center}
+	\caption{Student survey responses.
+	SD/D/N/A/SA stands for
+	strongly disagree/\allowbreak{}disagree/\allowbreak{}neutral/\allowbreak{}agree/\allowbreak{}strongly agree.}
+	\label{fig:survey}
+\end{figure}
+
+{\bf Response data.}
+The response distributions for each of the surveys' multiple choice questions
+are shown in Figure~\ref{fig:survey}.
+In total, 28 students (or pairs thereof, in the case of CMU and U. Chicago) answered the survey:
+12 pairs from CMU, 15 individuals from PSU, and 1 pair from U. Chicago.
+The first four questions/graphs focus on concrete debugging results,
+and the latter six on the students' subjective opinions.
+%
+Note that two of the questions
+(3 and 7 from \sect{\ref{sec:education-survey-pebbles}})
+were not asked on U. Chicago's version of the survey,
+so their corresponding graphs show only CMU and PSU response data.
+Likewise, U. Chicago's questions 4, 8, and 10
+(see \sect{\ref{sec:education-survey-pintos}}),
+which were not asked on CMU's and PSU's surveys,
+having only one respondent,
+are not pictured;
+the answers thereto were ``15-20 minutes'', ``Agree'', and ``Agree'', respectively.
+
+The survey results were overwhelmingly positive, and largely speak for themselves:
+% TODO: corroborate this with snapshot data?
+students reported being able to both diagnose and verify fixed the vast majority of Landslide's reported bugs,
+compared Landslide favorably to stress testing,
+and found the experience worthwhile and worth recommending.
+Three questions with open-form answers bear further discussion
+are that of what kind of false positives Landslide (regrettably) reported % oopsative mood
+(CMU and PSU only),
+reasons they found it worth recommending to a friend,
+and suggestions for improving the interface
+(PSU only\footnote{I thought to ask this question too late for CMU's surveys.}).
+
+{\bf False positives.}
+Even though 71\% of students reported receiving no false positive bug reports,
+the nature of Landslide's bug-detection algorithms is such
+that it should ideally never report any correct behaviour as wrong,
+so I consider those 29\% that did report such the most negative result among the survey responses.
+They described their false positives, and I either make excuses or own up, as appropriate, as follows.
+Reports from CMU students (Simics version):
+\begin{enumerate}
+	\item Landslide complained of a nonexistent {\tt MAGIC\_BREAK} (Simics debugging function),
+		despite the student's code never invoking it.
+		This arose because of confusion between the test program's and the shell's address spaces,
+		and was subsequently fixed in commit 3f24d67 (Simics repository only).
+	\item Landslide reported ``some weird errors'' and/or mysteriously crashed
+		when multiple instances were run from the same directory.
+		(Multiple students suffering this failure mode contacted me for support,
+		though only one reported it on the survey;
+		in some cases, I recall said weird errors manifesting as false positive invalid heap access reports.)
+		Simultaneous Landslides can clobber certain auto-generated header and/or temporary files,
+		leading to undefined behaviour.
+		I introduced a guard against this in commit 977e8fb (Simics repository) and e49b5df (Bochs repository).
+	\item One student reported ``Data races which I believe they are not''.
+		Looking at their usage snapshots to corroborate this,
+		these appear to be true, yet benign, data races
+		(in several cases corresponding to {\tt mutex\_test}'s verification of the mutex's internal memory accesses
+		(see \sect{\ref{sec:education-pebbles-tests}}));
+		i.e., expected behaviour rather than false positive bug reports.
+		Student confusion about these could be alleviated
+		by improving Landslide's user interface messages when printing data race information. % uuuuuuu its already so
+	\item One student complained of a bug report that, while truly a bug, showed wrong filenames and line numbers in stack traces,
+		and (quite naturally) suggested that accurate stack traces would make debugging easier.
+		Checking against their usage snapshots,
+		this seems to be an issue with Simics's symtable's handling of assembly functions.
+		The Bochs version handles these correctly.
+\suspend{enumerate}
+Reports from PSU students (Bochs version):
+\resume{enumerate}
+	\item Landslide kept crashing for one student while running {\tt paraguay}.
+		This is likely due to exceeding Linux's process limit and/or exhausting the class's official VM's memory,
+		which can arise when many data race candidates cause Quicksand to spawn many Landslide jobs,
+		and (as on {\tt paraguay}) each with very large state spaces that must defer.
+		I had been working on mitigating this problem just before beginning the PSU study;
+		properly addressing it would involve improving Quicksand's memory-exhaustion detection code
+		and/or making Quicksand at all aware of the process limit to begin with.
+		(Note that this is not strictly a false positive bug report, just a Landslide crash.)
+	\item Students who initialized child threads with a base pointer value of {\tt 0xffffffff} observed Landslide crashes,
+		as it attempted to stack trace through that address (accessing memory that wrapped around the address space).
+		(Several students contacted me about this via email, and I issued a prompt fix;
+		one student later reported it on the survey.)
+		Commits 0573e34 (Bochs repository) and 654f459 (Simics repository) fixed this bug.
+		(Like the above, not actually a false positive bug report.)
+	\item Landslide issued false invalid heap access reports to one student
+		who had been using {\tt new\_pages} to allocate thread stacks ``very close'' to the {\tt malloc} heap.
+		They reported this during the study and I fixed it promptly for them in commit 4a26da7 (Bochs repository only),
+		then reminded me of it again in the survey.
+	\item Finally, one student reported simply, ``Race condition''.
+		Without the same usage snapshots to consult as I'd have for a CMU student,
+		or more self-reported detail, I regrettably can offer no comment.
+\end{enumerate}
+
+\noindent
+Overall, most of the issues reported as answers to the ``false positive'' survey question
+were merely Landslide crashes or user interface confusion.
+Those few truly erroneous bug reports (items 1, 2, and 7),
+while certainly guilty of burdening % oopsative mood
+students with worry over whether the problem is their own code or in Landslide itself,
+are at least not too discouraging for two reasons.
+Firstly, in all cases the students were able to recognize the false positive quickly enough to ask for help,
+and I was able to deploy a fix and let them proceed before the project came due.
+Secondly, each such problem, now having been fixed, will befall no future student again --
+not to assert Landslide is completely bug-free now,
+but it at least grows more and more stable with each passing semester.
+
+{\bf Reasons worthwhile.}
+
+{\bf Interface suggestions.}
 

### 2018-07-05 thesis/education.tex
survey other feedbacks

@@
 
-{\bf Response data.}
+\subsubsection{Response data}
+
 The response distributions for each of the surveys' multiple choice questions
 are shown in Figure~\ref{fig:survey}.
-In total, 28 students (or pairs thereof, in the case of CMU and U. Chicago) answered the survey:
+In total, 28 students (or pairs thereof) answered the survey:
 12 pairs from CMU, 15 individuals from PSU, and 1 pair from U. Chicago.
@@
 
-The survey results were overwhelmingly positive, and largely speak for themselves:
+The survey responses were very positive:
 % TODO: corroborate this with snapshot data?
-students reported being able to both diagnose and verify fixed the vast majority of Landslide's reported bugs,
+students reported being able both to diagnose and to verify as fixed the vast majority of Landslide's reported bugs,
 compared Landslide favorably to stress testing,
 and found the experience worthwhile and worth recommending.
-Three questions with open-form answers bear further discussion
-are that of what kind of false positives Landslide (regrettably) reported % oopsative mood
-(CMU and PSU only),
+Three questions with open-form answers bear further discussion:
+what kind of false positives Landslide reported, % oopsative mood
 reasons they found it worth recommending to a friend,
 and suggestions for improving the interface
-(PSU only\footnote{I thought to ask this question too late for CMU's surveys.}).
+(PSU only\footnote{I thought to ask this question too late for CMU's surveys.}). % ososugichatte
 
-{\bf False positives.}
+\subsubsection{False positives}
+
 Even though 71\% of students reported receiving no false positive bug reports,
@@
 		despite the student's code never invoking it.
-		This arose because of confusion between the test program's and the shell's address spaces,
+		This arose because of technical confusion between the test program's and the shell's address spaces,
 		and was subsequently fixed in commit 3f24d67 (Simics repository only).
@@
 	\item Students who initialized child threads with a base pointer value of {\tt 0xffffffff} observed Landslide crashes,
-		as it attempted to stack trace through that address (accessing memory that wrapped around the address space).
+		as it attempted to stack trace through that address and access memory that wrapped around the address space.
 		(Several students contacted me about this via email, and I issued a prompt fix;
@@
 
-{\bf Reasons worthwhile.}
+\subsubsection{Reasons worthwhile}
 
-{\bf Interface suggestions.}
+After the question asking ``Would you recommend a friend taking OS next semester to use Landslide?'',
+I asked the students for open-form reasons why or why not.
+As the former question's answers were exclusively positive (only 1 student even answering ``no opinion''),
+this question's answers turned out to be mostly praise.
+3 students declined to answer this question;
+I reproduce the rest, paraphrasing for clarity and brevity,%
+\footnote{Note that most students used the term ``race condition'' rather than ``concurrency bug'',
+as taught in CMU and PSU lecture material;
+I replaced these while paraphrasing in accordance with \sect{\ref{sec:glossary}}.}
+here.
+Answers from CMU:
+
+\begin{enumerate}
+	\item Easy to use and found nontrivial bugs.
+	\item Useful to find some bugs, but other types of bugs (such as memory corruption) were impossible to find with Landslide.
+	\item Helped verify our atomic primitives were correct.
+	\item Would recommend, but takes too much time (unclear, but seems to be referring to execution time rather than setup/usage).
+	\item Tests are automated and can be left running for a long time.
+		However, faced uptime issues with CMU's linux servers (which reboot every night);
+		wished for a more reliable execution environment.
+	\item It's pretty helpful. % lol
+	\item Seems more reliable than stress tests.
+	\item Found a bug not found by stress tests.
+	\item Finds concurrency bugs with little effort that may be undiscovered otherwise.
+		(Also provided some interface feedback here, since I didn't ask CMU students a separate question for such;
+		see next section.)
+	\item Easy to learn and a simple way to test our P2.
+\suspend{enumerate}
+Answers from PSU:
+\resume{enumerate}
+	\item Although didn't find any concurrency bugs for me, gives me more confidence about my code.
+	\item Helpful, just didn't have a lot of time to use it.
+	\item Does not make concurrency debugging easy, but definitely makes it easier.
+	\item Helpful to find bugs you weren't previously aware of.
+		Makes more sense to use an expressly designed tool rather than (unit/stress) tests.
+	\item Helpful and easy to use.
+	\item Saves time overall, can run long tests overnight.
+	\item Helps to find concurrency bugs and their root causes better than stress tests.
+	\item It finds the concurrency bugs you need to fix for full credit.
+	\item Helpful for finding some uncommon bugs I hadn't found or wasn't looking for.
+	\item Found bugs I kept overlooking, which may have taken many hours to find otherwise.
+	\item Helpful for the difficult step from code being ``finished'' (scare-quotes theirs; presumably meaning ``feature-complete'')
+		to getting rid of all concurrency bugs.
+	\item Easy to use, setup taking no more than 5-10 minutes, and allowing being run overnight.
+	\item Very efficient at concurrency testing.
+		Stress test crash reports do not necessarily point to the root cause, due to memory corruption for example;
+		plus bugs may not show up every time due to nondeterminism.
+	\item Helped find a bug I wouldn't have found otherwise.
+		Did not show an interleaving directly (i.e., did not issue a bug report),
+		but reported a data race that turned out to be a concurrency bug upon inspection.
+\suspend{enumerate}
+Answer from U. Chicago:
+\resume{enumerate}
+	\item Found several subtle, legitimate bugs we wouldn't have easily caught otherwise, but made sense once revealed.
+		Fixing them took little time but allowed us to proceed confidently on the next project.
+		Often wished for Landslide to have been available to use during the next project as well.
+\end{enumerate}
+
+\noindent
+Overall, students most commonly praised Landslide's ease of use,
+its ability to find bugs that elude stress testing,
+and the confidence instilled from verifying bugs had been fixed.
+
+\subsubsection{Interface suggestions}
+
+Lastly, I asked the PSU students for any feedback they might have on making Landslide's interface easier to use or understand.
+
+\begin{enumerate}
+	\item Requested for preemption traces to be more clear about the meaning of each stack trace in each table cell,
+		and complained of inaccurate line numbers
+		(likely referring to how the current behaviour indicates the line of code {\em after} a function call,
+		corresponding to the {\tt call} instruction's pushed return address,
+		rather than the function call itself). (This answer from CMU; see above.)
+	\item Including a manual or tutorial would be helpful (presumably beyond the user guide's instructions, such as recapping the procedure shown in the lecture demo which wasn't written down anywhere).
+	\item Don't print warnings about line length exceeding 80 characters (inconsistency between 15-410 and \psuos compilation options).
+	\item ``It takes too long. But I guess that's impossible to fix.'' (Well, it's an open research problem to fix!)
+	\item Preemption traces should explicitly indicate where in the interleaving the bug occurred.
+		(Root cause identification is its own research area,
+		but Landslide could show, for example, the sets of memory conflicts between threads in the trace as well.)
+	\item Improve explanation of data races (in the user guide, perhaps).
+	\item Happy with it as-is (3 students)
+	\item No response (7 students)
+\end{enumerate}
 

### 2018-07-05 thesis/education.tex
commentate interface feedback, reorg some education, add some todos

@@
 	\begin{center}
+		% TODO: calc p values for this
+		% note, smth like fisher's method which "combines independent studies into a meta p value"
+		% cannot be applied for 2 reasons - 1, these are not independent since the same studence were
+		% running landslide on diff test cases; 2, the hypotheses tested are different for each test
 	\begin{tabular}{r||cc|cc|c}
@@
 		Downgrade & 10	& 13	& 19	& 23	& $-05.7\%$ \\
+		Total & \\ % TODO
 	\end{tabular}
@@
 	\caption{
+		Correlation of student Landslide use with solving certain bugs in their final submission during Fall 2017 and Spring 2018 semesters.
 		Students who used Landslide in general but whose usage snapshots showed they did not run the test case in question are counted as non-Landslide-users here.
@@
 {\tt :downgrade\_partial\_signal -1}, % not sure if this means smth that dgr would expose
-{\tt :downgrade\_lies -3}, and
+{\tt :downgrade\_\allowbreak{}lies -3}, and
 {\tt :downgrade\_deadlock -4}.
@@
 	\item Preemption traces should explicitly indicate where in the interleaving the bug occurred.
-		(Root cause identification is its own research area,
-		but Landslide could show, for example, the sets of memory conflicts between threads in the trace as well.)
+		(Root cause identification is its own research area, but more detail is certainly possible.)
 	\item Improve explanation of data races (in the user guide, perhaps).
@@
 \end{enumerate}
+
+Though I did not ask this question on the CMU survey,
+my experience handling student questions suggests CMU students
+also mostly wish for
+better explanations of data races and
+more detail and clarity in the preemption traces.
+Though I present the formal definition of data races in the lecture
+(\sect{\ref{sec:education-pebbles-recruiting}})
+and refer back to it in Landslide's documentation and output,
+showing a concrete example
+in future iterations of the user guide
+would go a long way to illustrate the abstract concepts.
+The preemption traces could be improved by making it clearer that the stack trace in each cell of the table
+represents executing the thread in question from wherever it previously left off (or its inception)
+all the way until it reaches that stack trace, then preempting it to run another thread.
+They could also easily report more diagnostic information;
+for example, showing the sets of memory conflicts between each thread,
+annotating the type of each preemption point (yield, mutex, or data race),
+and/or indicating the adversarial memory access for each data race preemption point.
 
@@
 % well, we did the best we could(?)
-% survey email: "please answer honestly rather than flatteringly"
+% survey email: "please answer honestly rather than flatteringly" vs. studence trying to be polite
+% selection bias, students who liked landslide more more likely to answer the survey at all
+% TODO - u can verify the bug just by taking survey graph #1 and comparing to landslide snapshots (sep. by groups who answered survey)
 % anything else?
 
-%%%% pintos %%%%
+%%%% pintos %%%% (?? this is psu not pintos)
 % survey questions i WISH i had asked
@@
 
-I attribute the low participation rate of Pintos students to P2 students to two major factors:
-one, not incentivizing the students to directly improve their grades
-(instead offering only the vague promise of a ``learning experience'' debugging their code only after handin),
-and two, not traveling to the university to introduce the research topic in an in-person lecture
-(leaving the students potentially confused about what advantage, if any, was offered over stress testing).
-
 While part of the point of this experimental design was to evaluate Landslide as a grading tool in the hands of TAs,
@@
 I implemented
-several % TODO: how many and what specifically
+several
 fixes/improvements to the setup scripts
@@
 
+I attribute the low participation rate of Pintos students to P2 students to two major factors:
+one, not incentivizing the students to directly improve their grades
+(instead offering only the vague promise of a ``learning experience'' debugging their code only after handin),
+and two, not traveling to the university to introduce the research topic in an in-person lecture
+(leaving the students potentially confused about what advantage, if any, was offered over stress testing).
 Hypothetically, I could have achieved greater user study participation
@@
 (either way requiring a more rigorous IRB review process).
-Practically, as for non-research use in Pintos classes,
+
+\subsection{Future educational use}
+
+% TODO: para for pebbles; discuss with dave
+
+Regarding non-research use in Pintos classes,
 Landslide can now handle a considerably wider variety of student implementation quirks
-on account of this time's fixes.
+on account of the fixes from this time (\sect{\ref{sec:education-pintos-instrumentation}}).
 In its current shape I would recommend it for TA use grading,
 but not necessarily directly to students without someone familiar with the codebase on immediate hand for tech support.
-
-% Point two is much more easily addressed, although perhaps not necessary if only being used for grading anyway.
+However, I also believe Landslide's success in these user studies,
+provided me present to handle technical issues,
+serves as testament for stateless model checking in general in the educational theatre.
+While Pintos's kernel-level environment presents a unique challenge for concurrency testing,
+other, more readily automatic model checkers for user-space programs,
+such as dBug \cite{dbug-ssv},
+could easily be used on other thread-library-like programming projects at any university.

### 2018-07-06 thesis/education.tex
dave copyedit description of bugs to be case studied

@@
 
-{\bf Thread exit use-after-free.}
-During {\tt thr\_exit()}, a thread must arrange for its own stack memory allocation to be freed.
+{\bf Thread-exit use-after-free.}
+%During {\tt thr\_exit()}, a thread must arrange for its own stack memory allocation to be freed.
+After a thread finishes exiting, the memory allocated for its stack must be reclaimed for other uses.
 A common student pitfall is to allow a {\tt thr\_join()}ing thread to free said stack space
-while {\tt thr\_exit()} is still executing userspace C code (which inevitably accesses the stack),
-or even for {\tt thr\_exit()} to do it itself.
+while {\tt thr\_exit()} is still executing userspace C code (which typically accesses the stack),
+or even for {\tt thr\_exit()} to free it itself.
 In the former case, threads must interleave specifically to exhibit a use-after-free;
 in the latter case, the use-after-free will be deterministic (i.e., present in all interleavings).
-Besides Landslide, however,
-the students have no Valgrind-like heap debugging tool which would report a bug immediately upon any illegal heap access;
-without such, a third thread would have to race to recycle the memory for a new thread stack
+Other than Landslide, however,
+the students have no Valgrind-like heap debugging tool which would report a bug immediately upon any illegal heap access.
+This means that a subsequent {\tt thr\_create()} invocation would need to race to recycle the memory for a new thread stack
 and conflict with the old exiting thread
@@
 % TODO: mark which of the tej bugs were deterministic vs race
-
-The Landslide-friendly test {\tt thr\_exit\_join} is most likely to expose this bug, and
-the official rubric penalties for it are
-{\tt :exit\_stack\_minor -2} and
-{\tt :exit\_stack\_race -4} (for the former case),
-and
-{\tt :uses\_freed\_stack -4} (for the latter).
+The Landslide-friendly test {\tt thr\_exit\_join} is most likely to expose this bug.
+%, and
+%the official rubric penalties for it are
+%{\tt :exit\_stack\_minor -2} and
+%{\tt :exit\_stack\_race -4} (for the former case),
+%and
+%{\tt :uses\_freed\_stack -4} (for the latter).
 Whether each student's submitted P2 solved or suffered from this bug
@@
 {\bf Paraguay.}
-Named after a certain Paraguayan 15-410 TA from years gone by who originally discovered it,
-this refers to a condition variable bug in which a thread which sleeps on two different condition variables
-can spuriously wake up early from the later {\tt cond\_wait()}.
+Named after Ivan Jager, the Paraguayan 15-410 TA from 2004-2006 who originally discovered it,
+this refers to a condition-variable bug in which a thread which sequentially sleeps on two different condition variables
+can spuriously wake up early from the second {\tt cond\_wait()}.
 The precise reasoning why many common na\"ive student implementations are susceptible to this bug,
 as well as the 3 major ways of fixing it,
-are one of 15-410 staff's closely-guarded secrets.
-Suffice to say that the test {\tt paraguay} invokes a specially-made Pathos {\tt misbehave} mode
+are one of 15-410 staff's closely-guarded secrets (see \sect{\ref{sec:410-secrecy}}).
+Suffice it to say that the test {\tt paraguay} invokes a specially-made Pathos {\tt misbehave} mode
 which biases thread scheduling towards the specific interleaving required to exhibit the bug
@@
 in helping students diagnose a bug that a ``stress'' test could already find.
-
-{\tt paraguay} is itself the Landslide-friendly test for its eponymous bug, and
-the official rubric penalty for it is
-{\tt :paraguay -2}.
+%
+{\tt paraguay} is itself the Landslide-friendly test for its eponymous bug.
+%, and
+%the official rubric penalty for it is
+%{\tt :paraguay -2}.
 Whether each student's submitted P2 solved or suffered from this bug
@@
 all while allowing no waiting writers to access in between.
-
+%
 The Landslide-friendly test {\tt rwlock\_downgrade\_read\_test}
-checks that readers are allowed simultaneous access after a downgrade, and
-the official rubric penalties related to this bug are
-{\tt :downgrade\_no\_signal -1},
-%{\tt :unsafe\_rwlock\_downgrade -2}, % don't think this is right; not relevant anyway (overlaps)
-{\tt :downgrade\_partial\_signal -1}, % not sure if this means smth that dgr would expose
-{\tt :downgrade\_\allowbreak{}lies -3}, and
-{\tt :downgrade\_deadlock -4}.
-Whether each student's submitted P2 solved or suffered from this bug
+checks that readers are allowed simultaneous access after a downgrade with no possibility for deadlock.
+Rather than one specific bug, this refers to any of several failures that can arise during a downgrade.
+%, and
+%the official rubric penalties related to this bug are
+%{\tt :downgrade\_no\_signal -1},
+%%{\tt :unsafe\_rwlock\_downgrade -2}, % don't think this is right; not relevant anyway (overlaps)
+%{\tt :downgrade\_partial\_signal -1}, % not sure if this means smth that dgr would expose
+%{\tt :downgrade\_\allowbreak{}lies -3}, and
+%{\tt :downgrade\_deadlock -4}.
+Whether each student's submitted P2 solved or suffered from these bugs
 was determined by checking their grade files for any of the above-listed penalties

### 2018-07-06 thesis/landslide.tex
dave copyedit description of bugs to be case studied

@@
 as the true nature of the Pathos reference kernel's {\tt misbehave} modes
-is a closely-guarded secret among 15-410 course staff.
+is a closely-guarded secret among 15-410 course staff (\sect{\ref{sec:410-secrecy}}).
 The mode in question causes the reference kernel to prioritize scheduling the child thread over the parent

### 2018-07-06 thesis/overview.tex
pintos testce
dave copyedit description of bugs to be case studied

@@
 
+\subsubsection{Secrecy}
+\label{sec:410-secrecy}
+
+The 15-410 course staff is notoriously secretive about the nature of many concurrency bugs
+students commonly encounter during P2 and P3.
+This is driven by a desire to cause students to find, diagnose, and fix these bugs on their own during the projects,
+rather than to be surprised by them afterwards during grading
+\cite{de0u-2018}.
+%
+One such example is the {\tt paraguay} unit test distributed with P2 (\sect{\ref{sec:education-pebbles-tests}}),
+which targets a subtle condition-variable bug.
+The test uses the {\tt misbehave} system call to target a particular thread interleaving likely to expose the bug
+which is otherwise very unlikely to arise in normal execution.
+The reference kernel specification \cite{kspec} does not define the {\tt misbehave} modes' behaviours,
+as doing so would deprive students of the learning experience of discovering the interleaving in question on their own.
+%
+In this thesis I will occasionally use intentionally vague phrasing to preserve the mystery of these bugs.
+
 \subsubsection{Use at other universities}
@@
 \section{Glossary}
+\label{sec:glossary}
 
@@
 % P2 (note abotu psu - dosent call it p2, but we call it that for both schools here
+% happens before, all 3 versions - DPOR, LHB, and PHB
+% userspace, kernelspace

### 2018-07-07 thesis/education.tex
very subtle
add mutex test to submissions table

@@
 	\item Regarding the previous question, why or why not?
-	\item Your name
+	\item Your name.
 	\item Your project partner's name (if applicable)
@@
 		%Downgrade & 17	& 24	& 8	& 12	& $+4.17\%$ \\ % not filtering out either did not run, or unrel ding
-		Downgrade & 10	& 13	& 19	& 23	& $-05.7\%$ \\
-		Total & \\ % TODO
+		Downgrade & 10	& 13	& 19	& 23	& $-05.7\%$ \\ % TODO verify these buggers by eye
+		% I spent more hours than I should have looking at this landslide-user-submitted-bug group.
+		% Can't name the group here, but if you're me and have the analysis file,
+		% it's the one from f17 with 3 mutex_test logs in their "u" snapshot.
+		% It seems like a mysterious older version of mutex.c was present between their "a" version
+		% (which gets fully verified very quickly btw) and their "u" version (for which landslide finds
+		% the bug in just a few minutes, no way they could be running the same version; plus the reported
+		% data races and so on in the id-log don't line up to the "u" mutex.c source code).
+		% That version was correct, then they updated it (introducing the bug) and ran a bunch of other tests,
+		% but didn't rerun mutex_test to check their update.
+		Mutex     & 17	& 18	& 19	& 19	& $-05.6\%$ \\
+		\hline
+		% TODO: figure out if calculating a p value (or w/e) for this the same way is valid,
+		% since they summands are arguably not independent? ..or maybe they are?
+		Total     & 58	& 73	& 52	& 74	& $+09.2\%$ \\
 	\end{tabular}
@@
 		P(\mathsf{correct}|\mathsf{non\text{-}user})$.
+		Note that the bottom row's totals double-count students,
+		which only makes sense if you believe the incidences of each bug within a given submission are independent.
 		}
@@
 can spuriously wake up early from the second {\tt cond\_wait()}.
-The precise reasoning why many common na\"ive student implementations are susceptible to this bug,
+The precise reasoning why many na\"ive student implementations are susceptible to this bug,
 as well as the 3 major ways of fixing it,
-are one of 15-410 staff's closely-guarded secrets (see \sect{\ref{sec:410-secrecy}}).
-Suffice it to say that the test {\tt paraguay} invokes a specially-made Pathos {\tt misbehave} mode
-which biases thread scheduling towards the specific interleaving required to exhibit the bug
+are one of 15-410 staff's closely-guarded secrets (\sect{\ref{sec:410-secrecy}}).
+Suffice it to say that the {\tt paraguay} test invokes a custom Pathos {\tt misbehave} mode
+which biases thread scheduling towards the interleaving required to exhibit the bug
 (Landslide, of course, replaces this {\tt misbehav}iour with model checking).

### 2018-07-07 thesis/education.tex
describe mutex bug

@@
 		Exit UAF  & 11	& 21	& 0	& 16	& $+52.4\%$ \\
-		% nb. this is not filter out the did not run paras -- smae for all of the above/below
-		% Paraguay  & 22	& 24	& 12	& 13	& $-0.64\%$ \\ % not filtering out did not run para
-		Paraguay  & 20	& 21	& 14	& 16	& $+07.7\%$ \\
-		% nb this is any downgrade related ding not just dgr tested ones
-		% nb 1 is msising bc didn't impl rwlock at all
-		%Downgrade & 17	& 24	& 8	& 12	& $+4.17\%$ \\ % not filtering out either did not run, or unrel ding
-		Downgrade & 10	& 13	& 19	& 23	& $-05.7\%$ \\ % TODO verify these buggers by eye
 		% I spent more hours than I should have looking at this landslide-user-submitted-bug group.
@@
 		Mutex     & 17	& 18	& 19	& 19	& $-05.6\%$ \\
+		% nb. this is not filter out the did not run paras -- smae for all of the above/below
+		% Paraguay  & 22	& 24	& 12	& 13	& $-0.64\%$ \\ % not filtering out did not run para
+		Paraguay  & 20	& 21	& 14	& 16	& $+07.7\%$ \\
+		% nb this is any downgrade related ding not just dgr tested ones
+		% nb 1 is msising bc didn't impl rwlock at all
+		%Downgrade & 17	& 24	& 8	& 12	& $+4.17\%$ \\ % not filtering out either did not run, or unrel ding
+		Downgrade & 10	& 13	& 19	& 23	& $-05.7\%$ \\ % TODO verify these buggers by eye
 		\hline
@@
 
-{\bf Thread-exit use-after-free.}
+{\bf Thread exit/join.}
 %During {\tt thr\_exit()}, a thread must arrange for its own stack memory allocation to be freed.
@@
 was determined by me personally inspecting their code.
+
+{\bf Mutex.}
+The Landslide-friendly {\tt mutex\_test} checks for the possibility of two contending threads
+accessing a mutex-protected critical section simultaneously.
+It includes one thread repeating once the lock, unlock cycle
+so as to check for unlock/lock races as well as lock/lock interactions,
+and, as mentioned previously (\sect{\ref{sec:education-pebbles-tests}}),
+checks for data race access pairs inside the mutex implementation itself.
+As students have free rein to design their mutex internals,
+this refers to the general class of mutex bugs
+in which any number of things can go wrong depending on the implementation.
+Whether each student's submitted P2 solved or suffered from this bug
+was determined by checking their grade files for any mutex-related penalties (assessed by the TAs),
+then me double-checking their implementations by hand to confirm.
+
+Further investigating the one group who submitted a buggy mutex,
+I found that while they had run {\tt mutex\_test} in Landslide
+(and even found and fixed a separate deterministic bug already),
+Landslide found no bug in what was presumably a correct implementation,
+then they updated % oops
+that implementation, introducing the bug, without testing it again thereafter.
 
@@
 Whether each student's submitted P2 solved or suffered from these bugs
-was determined by checking their grade files for any of the above-listed penalties
+was determined by checking their grade files for any downgrade-related penalties
 (which were assessed by the TAs as a result of their manual inspection, rather than mine).

### 2018-07-12 thesis/education.tex
discuss more abt survey results

@@
 was determined by me personally inspecting their code.
+%
+This bug is especially remarkable in that not a single student who did not use Landslide these semesters
+submitted a correct implementation.
 
@@
 this refers to the general class of mutex bugs
-in which any number of things can go wrong depending on the implementation.
-Whether each student's submitted P2 solved or suffered from this bug
+in which any number of things can go wrong, depending on the implementation,
+leading to mutual exclusion (or otherwise assertion) failure.
+Whether each student's submitted P2 solved or suffered from such bugs
 was determined by checking their grade files for any mutex-related penalties (assessed by the TAs),
@@
 then they updated % oops
-that implementation, introducing the bug, without testing it again thereafter.
+their code, introducing the bug, without testing it again thereafter.
+% TODO: some reference to future work about focused testing on small code diffs? talked with some industry ppl about this sometimes
 
@@
 and found the experience worthwhile and worth recommending.
+Regarding the 100 CPU-hours question in particular,
+students could approach answering it in several different ways:
+do the three students who answered 20/80 think stress testing is four times as good as Landslide at finding bugs,
+or that stress testing requires four times as long to reach the same point of diminishing returns?
+On the other extreme, one student said they would spend all 100 CPU-hours on Landslide,
+which probably speaks more to their enthusiasm than to a careful attempt to maximize expected number of bugs found
+% TODO: put a section reference to future work resource exhaustion / malloc failure injection discussion
+(even the author themself recognizes stress testing's advantages for certain types of bugs, such as resource exhaustion).
+Nevertheless, the responses' overall bias to spending at least half the CPU-time on Landslide
+shows clearly that the students found the experience worthwhile.
+
 Three questions with open-form answers bear further discussion:
@@
 
+Comparing the survey response distributions between CMU versus PSU and U. Chicago (Figure~\ref{fig:survey}),
+CMU students tended to verify their bugfixes more often by re-running Landslide
+and found the preemption traces easier to understand,
+while PSU and U. Chicago students generally compared Landslide more favourably to stress testing
+(while CMU students comparatively preferred the more nuanced answer that it depends on the type of bug),
+and reported more often that helped them understand concurrency better.
+%Considering that CMU's 15-410 demands substantial prerequisite concurrency experience
+%(15-213 \cite{sigcse01:CSaPP} or equivalent)
+%whereas PSU's \psuos is more towards the average between CMU's 15-213 and 15-410
+%(tempering the difficulty of P2 to make it more accessible (\sect{\ref{sec:overview-psu}})),
+Considering the higher demand CMU's 15-410 makes for prerequisite concurrency experience
+compared to PSU's relative tempering of P2's difficulty to make it more accessible (\sect{\ref{sec:pebbles}}),
+% TODO: represent u c in this sentents?
+these trends seem to correlate with the different levels of preparation each course's students had,
+showing that students of various skill levels
+can each benefit from the experience in different ways.
+
 % TODO

### 2018-07-12 thesis/overview.tex
discuss more abt survey results

@@
 The other three (a stack-crawling backtrace utility, a bare-metal game with device drivers, and a small extension to the P3 kernel) are not of concern in this thesis.
-The course's prerequisite is Introduction to Computer Systems \cite{sigcse01:CSaPP}.
+The course's prerequisite is 15-213, Introduction to Computer Systems \cite{sigcse01:CSaPP}.
 Both P2 and P3 are built using the {\em Pebbles} system call specification, outlined in Table~\ref{tab:syscalls}
@@
 \subsubsection{Use at other universities}
+\label{sec:overview-psu}
 
@@
 offered the P2 thread library project as part of its curriculum.
-Students in this class implemented P2 working alone rather than in pairs,
+Students in this class implemented P2
+on a 6 week project timeline (compared to 2 weeks at CMU),
+work alone rather than in pairs,
 skip the {\tt swexn} automatic stack growth portion,

### 2018-07-14 thesis/landslide.tex
describe sleep sets in landlside

@@
 to reach a (possibly) new program state not exposed in the current interleaving.
+Prior work \cite{dpor,optimal-dpor} refers to the set of these $I_{ij}$s,
+for a given $i$,
+as the {\em persistent set} at the preemption point after $t_{i-1}$.
 
@@
 
+{\bf Further pruning.}
+In the presence of non-conflicting transitions as well as conflicting ones,
+DPOR's approach as described so far can still end up testing equivalent interleavings.
+For example, consider the execution sequence
+in Figure~\ref{fig:sleepsets}(a).
+After executing this interleaving, DPOR will detect the conflict between $t_a$ and $t_c$,
+and tag $I_{a2}$ (i.e., $T_2$ to run before $t_a$), causing us to next execute
+Figure~\ref{fig:sleepsets}(b).
+However, DPOR does not ``remember'' why it wanted to reorder $T_2$ with $T_1$
+(i.e., whether $t_b$ or $t_c$ was the important one to run before $t_a$),
+so from this branch,
+it next tags $I_{c'1}$, causing
+Figure~\ref{fig:sleepsets}(c).
+to be tested next, which is regrettably equivalent to the original interleaving
+($t_a$ and $t_b$ having been reordered all told, which are independent).
+To recover our forgotten original purpose,
+which was already fulfilled by testing the second interleaving,
+DPOR can check just before tagging a new interleaving (here, $I_{c'1}$),
+among all preceding transitions equivalent to the conflicting one (here, $t_{b'}$ equivalent to $t_{a'}$),
+for an already-explored interleaving beginning with the target thread.
+If such exists,
+%then because of the depth-first search ordering,
+the new interleaving is guaranteed to be equivalent to one already checked,
+and can be skipped.
+
+\begin{figure}[h]
+	\begin{center}
+		\begin{tabular}{p{0.3\textwidth}p{0.25\textwidth}p{0.35\textwidth}}
+	\begin{tabular}{l}
+	$T_1$@$t_a$: \texttt{write~x} \\
+	$T_2$@$t_b$: \texttt{read~y} \\
+	$T_2$@$t_c$: \texttt{read~x} \\
+	\end{tabular}
+	&
+	\begin{tabular}{l}
+	$T_2$@$t_{b'}$: \texttt{read~y} \\
+	$T_2$@$t_{c'}$: \texttt{read~x} \\
+	$T_1$@$t_{a'}$: \texttt{write~x} \\
+	\end{tabular}
+	&
+	\begin{tabular}{l}
+	$T_2$@$t_{b'}$: \texttt{read~y} \\
+	$T_1$@$t_{a''}$: \texttt{write~x} \\
+	$T_2$@$t_{c''}$: \texttt{read~x} \\
+	\end{tabular}
+	\\
+	\\
+	(a) Original interleaving.
+	&
+	(b) Goal interleaving.
+	&
+	(c) Redundant interleaving.
+	\end{tabular}
+	\end{center}
+	\caption{Motivating example for the sleep sets optimization.
+	For clarity, threads are labeled here with numbers ($T_n$) and the transitions thereby with letters ($t_i$).}
+	\label{fig:sleepsets}
+\end{figure}
+Landslide implements this check in {\tt equiv\_already\_explored()}.
+This corresponds to the {\em sleep sets} optimization described by prior work \cite{dpor,optimal-dpor},
+so named because it effectively puts $T_1$ ``to sleep'' until after the true conflicting access in $t_c$.
+Further optimizations such as {\em source sets} and {\em wakeup trees},
+which achieve optimality (i.e., executing exactly one interleaving per equivalence class) \cite{optimal-dpor}
+are not yet implemented,
+although to the best of my knowledge,
+they provide further reduction only in cases of 3 or more threads;
+I suspect (without proof) that sleep-set DPOR is optimal for 2 threads.
+
+% TODO: make sure it's clear that the quicksand chapter expts were run before this optmz was added,
+% and that the HTM expts fro msigbovik paper were re-run for its chapater herein
+
 {\bf Choosing which new interleaving to test next.}
@@
 
+{\bf Termination.}
 When DPOR returns no new $I_{ij}$s not already marked in the set of visited subtrees,

### 2018-07-15 thesis/tm.tex
import tsx eval from sigbovik

@@
 {Yuri, Doki Doki Literature Club}
-
-% notes to self abour marios tests
-% counter: tests k threads on n iterations, benchmark xchg-aways vs txn. removed benchmarking, added assertion that no counts are lost.
-
-% TODO
 
@@
 which of course correspond to {\sf begin}, {\tt end}, and {\tt abort} listed above,
-and which GCC exposes as C/C++ intrinsices named {\tt \_xbegin()}, {\tt \_xend()}, and {\tt \_xabort()}
+as well as {\tt xtest}, an instruction which returns whether or not the CPU is executing transactionally.
+GCC and Clang expose these as C/C++ intrinsics named {\tt \_xbegin()}, {\tt \_xend()}, {\tt \_xabort()}, and {\tt \_xtest()}
 \cite{htm-gcc}.
@@
 \]
-Prior work \cite{htm-mario} proposed the idiom shown in Figure~\ref{fig:htm-fixed}
+Prior work \cite{tm-benchmark-cmu} proposed the idiom shown in Figure~\ref{fig:htm-fixed}
 to exclude this family of interleavings,
@@
 Landslide hooks these functions' addresses during instrumentation
-and inserts preemption points, failure injections, etcetera as necessary whenever the execution encounters them.
+and inserts preemption points, failure injections,
+et cetera as necessary whenever the execution encounters them.
 
 Transactional test programs should be ported to the Pebbles userspace if not already,
-then replace any use of compiler HTM intrinsics with Landslide's interface%
+then any use of compiler HTM intrinsics replaced with Landslide's interface%
 \footnote{Attempts to execute a real TSX instruction under Landslide, instead of using the custom interface,
-will be reported as ``invalid opcode'' bugs.
+will be reported as ``invalid opcode'' bugs,
+as neither of its supported simulation platforms support the feature.
 }.
@@
 and the {\tt 410TESTS} or {\tt STUDENTTESTS} line (respectively) of {\tt config-incomplete.mk} be edited to add the test name,
-% TODO: better section reference
-before running {\tt p2-setup.sh} (\sect{\ref{chap:education}})
+before running {\tt p2-setup.sh} (\sect{\ref{sec:landslide-setup}})
 % TODO(implementation): Provide an obfuscated refp2?
@@
 
+While prior work has focused on verifying implementations of transactional memory themselves
+\cite{specifying-verifying-tm,tm-correctness,tm-completeness,mc-tm-with-spin},
+Landslide is to the best of my knowledge the first model checker to support transactional client code.
+% mumble smth new concurrency model
+Accordingly, there is no baseline against which to compare Landslide's performance.
+Likewise, since Landslide's HTM semantics emulation relies on the equivalences proved in \sect{\ref{sec:tm-design}},
+we did not actually implement an HTM-style speculative-execution-and-rollback simulation mode.
+% so have no basis to measure how much it saves, or smth like that
+On this count, at least,
+we hope the reader finds it self-evident that
+our equivalence proofs provide exponential state space reduction
+compared to actually testing (and thereafter aborting) every combination of instructions within transactions.
+Beyond those, this chapter's evaluation will take a relatively green-field and exploratory approach.
+I pose the following evaluation questions.
+
+% on one hand,
+% comparing with \cite{mc-tm-with-spin}, succeeding for (2,3) then burning out with (2,4) is consistent with prior work
+% on the other hand, they were verifying STM implementation, which is no doubt more operations per iteration?
+% whatever, don't need to make this argument
+
+\begin{enumerate}
+	\item How quickly does Landslide find bugs in incorrect transactional programs? %of varying sizes?
+	\item Does Landslide find any previously-unknown bugs in real-world transactional code?%
+		\footnote{The savvy reader will realize that whether or not the author poses this evaluation question to begin with spoils its answer.}
+	%\item How quickly does Landslide verify correct transactional programs of varying sizes?
+	\item How does Iterative Deepening's performance compare to Maximal State Space mode?%
+		\footnote{Not necessarily related to HTM,
+		but the latter was implemented well after Chapter~\ref{chap:quicksand}'s conference paper was published,
+		so this was the most convenient test suite to evaluate it on.}
+	\item How well does Landslide's verification scale
+		with increasing thread/iteration count for correct transactional programs?
+	\item What further reduction can be achieved beyond that provided by
+		the global-lock/failure-injection equivalences?
+	%\item By the way, should MC research papers quantify variance in their CPU-time performance experiments?
+\end{enumerate}
+
+The evaluation suite comprises several hand-written unit tests,
+microbenchmarks and transactional data structures from \cite{tm-benchmark-cmu},
+a transactional spinlock from \cite{spinlock-rtm-github},
+and various combinations thereof,
+as follows.
+
+% notes to self abour marios tests
+% counter: tests k threads on n iterations, benchmark xchg-aways vs txn. removed benchmarking, added assertion that no counts are lost.
+
+\begin{itemize}
+	\item Microbenchmarks
+	\begin{itemize}
+		\item {\tt htm1}: The bug from Figure~\ref{fig:htm-example}. %(assertion failure on {\tt count}).
+		\item {\tt htm2}: The fixed version as in Figure~\ref{fig:htm-fixed}.
+		\item {\tt counter}: Microbenchmark version of {\tt htm2} which replaces the complex failure path with a simple {\tt xadd}.
+		\item {\tt swap}: Microbenchmark that swaps values in an array.
+		\item {\tt swapbug}: {\tt swap} modified to introduce circular locking in the failure path. %(deadlock).
+	\end{itemize}
+	\item Data structure tests
+	\begin{itemize}
+		\item {\tt avl\_insert}: AVL tree concurrent insertion test \cite{tm-benchmark-cmu}.
+		\item {\tt avl\_fixed}: {\tt avl\_insert} with the AVL bug fixed (spoilers!!).
+		\item {\tt map\_basic}: Separate-chaining hashmap concurrent insertion test \cite{tm-benchmark-cmu}.
+		\item {\tt map\_basicer}: {\tt map\_basic} modified with a larger initial size to skip the resizing step.
+		\item {\tt avl\_mutex}: {\tt avl\_fixed} with transactional sections simplified by abstraction into a mutex.
+		\item {\tt map\_mutex}: {\tt map\_basic} simplified likewise.
+	\end{itemize}
+	\item Lock abstraction tests.
+		These are each parameterized over implementations {\tt spinlock} (from \cite{spinlock-rtm-github}),
+		{\tt spinlock\_fixed} (spoilers!!),
+		and {\tt mutex} (replaces the spinlock with a Landslide-annotated mutex to reduce state space size).
+	\begin{itemize}
+		\item {\tt lock}(): Checks that multiple threads using a transactional lock
+			cannot access the critical section simultaneously.
+		\item {\tt lock\_fast}(): Checks that a transactional lock's fast path will not suffer conflict aborts
+			if its client threads' critical sections are independent.
+	\end{itemize}
+\end{itemize}
+
+The notation {\tt testname}$(K,N)$ will denote a test configuration of $K$ threads, each running $N$ iterations of the test logic.
+All tests were run on an 8-core 2.7GHz Core i7 with 32 GB RAM.
+To minimize variance in CPU-time measurements,
+I ensured the test machine was not loaded beyond normal web browser use,
+and ran only one instance of Landslide at a time
+\cite{sigbovik-htm}.
+
+\newcommand\ETA[1]{\hilight{brownish}{{\em #1}}\xspace}
+\newcommand\cpu[1]{\hilight{darkcyan}{{#1}}\xspace}
+\newcommand\wtm[1]{\hilight{lime}{{#1}}\xspace}
+\newcommand\ints[1]{\hilight{pinkish}{{#1}}\xspace}
+\newcommand\ETAdag[1]{\ETA{\ensuremath{\dagger}#1}}
+
+\subsection{Bugs}
+
+\begin{table*}[t]
+	\begin{center}
+		\footnotesize
+	\begin{tabular}{cc||r|r|r||r|r|r|r}
+			&	&\multicolumn{3}{c||}{\bf Quicksand mode}&\multicolumn{4}{c}{{\bf Maximal state space mode} ({\tt -M})} \\
+		\bf buggy test	& \bf params&\cpu{\bf cpu (s)}&\wtm{\bf wall (s)}&\ints{\bf int's}&\cpu{\bf cpu (s)}&\wtm{\bf wall (s)}&\ints{\bf int's}& \ETA{\bf SS size (est.)} \\
+		\hline
+		\hline
+		{\tt htm1}
+			& 2,1	&\cpu{  45.78}&\wtm{9.70}&\ints{21}& \cpu{*9.47}& \wtm{*6.40}& \ints{21	}& \ETA{213} \\
+		(assertion)
+			& 2,2	&\cpu{  84.14}&\wtm{13.59}&\ints{*33}& \cpu{*10.39}& \wtm{*7.70}& \ints{49	}& \ETA{1536} \\
+			& 2,3	&\cpu{ 131.91}&\wtm{20.44}&\ints{*73}& \cpu{*12.83}& \wtm{*9.67}& \ints{113	}& \ETA{10752} \\
+			& 2,4	&\cpu{ 255.75}&\wtm{37.56}&\ints{257}& \cpu{*18.63}& \wtm{*15.86}& \ints{257	}& \ETA{73728} \\
+		\cline{2-9}
+			& 3,1	&\cpu{ 114.06}&\wtm{17.45}&\ints{*15}& \cpu{*9.50}& \wtm{*6.79}& \ints{21	}& \ETA{13653} \\
+			& 3,2	&\cpu{ 109.60}&\wtm{26.16}&\ints{49}& \cpu{*10.72}& \wtm{*7.97}& \ints{49	}& \ETA{393216} \\
+			& 3,3	&\cpu{ 124.80}&\wtm{20.40}&\ints{*73}& \cpu{*13.84}& \wtm{*11.01}& \ints{113	}& \ETA{11010048} \\
+			& 3,4	&\cpu{ 227.49}&\wtm{35.15}&\ints{*161}& \cpu{*31.37}& \wtm{*28.53}& \ints{257	}& \ETA{301989888} \\
+		\cline{2-9}
+			& 4,1	&\cpu{ 53.08}&\wtm{9.79}&\ints{*15}& \cpu{*9.82}& \wtm{*7.00}& \ints{21	}& \ETA{873813} \\
+			& 4,2	&\cpu{117.07}&\wtm{19.09}&\ints{*33}& \cpu{*11.54}& \wtm{*8.55}& \ints{49	}& \ETA{100663296} \\
+		\hline
+		{\tt swapbug}
+			& 2,1	&\cpu{ 70.95}&\wtm{13.45}&\ints{*16}& \cpu{*38.96}& \wtm{*13.15}& \ints{109	}& \ETA{194} \\
+		(deadlock)
+			& 2,2	&\cpu{ 107.28}&\wtm{*17.45}&\ints{*146}& \cpu{*44.73}& \wtm{19.47}& \ints{281	}& \ETA{1620} \\
+			& 2,3	&\cpu{ 280.05}&\wtm{38.70}&\ints{*352}& \cpu{*60.30}& \wtm{*35.55}& \ints{718	}& \ETA{12748} \\
+			& 2,4	&\cpu{ 617.94}&\wtm{*81.50}&\ints{*834}& \cpu{*108.58}& \wtm{82.60}& \ints{1820	}& \ETA{97823} \\
+		\cline{2-9}
+			& 3,1	&\cpu{*1275.04}&\wtm{*163.42}&\ints{*771}&\ETA{--}&\ETA{>30m}&\ETA{--}& \ETA{184984} \\
+			& 3,2	&\ETA{--}&\ETA{>30m}&\ETA{--}&\ETA{--}&\ETA{>30m}&\ETA{--}& \ETA{3099225} \\
+		\hline
+		{\tt avl\_insert}
+			& 2,2	&\cpu{488.07}&\wtm{64.77}&\ints{*83}	&\cpu{*81.00}&\wtm{*40.30}&\ints{336}&\ETA{379982} \\
+		(segfault+)
+			& 2,3	&\cpu{2670.87}&\wtm{*330.45}&\ints{*3066}&\cpu{*1331.79}&\wtm{1274.36}&\ints{13926}&\ETA{96248131} \\
+			& 2,4	&\cpu{*3259.37}&\wtm{*436.50}&\ints{*1639}&\ETA{--}&\ETA{>30m}&\ETA{--}& \ETA{36019973} \\
+		\cline{2-9}
+			& 3,1	&\cpu{222.02}&\wtm{40.04}&\ints{*28}	&\cpu{*69.99}&\wtm{*24.25}&\ints{78}&\ETA{1572107} \\
+			& 3,2	&\cpu{*1569.09}&\wtm{*216.85}&\ints{*209}&\ETA{--}&\ETA{>30m}&\ETA{--}& \ETA{1402363529} \\
+	\end{tabular}
+	\end{center}
+	\caption{Landslide's bug-finding performance on various test configurations.
+		Iterative Deepening, optimized for fast bug-finding,
+		is compared against Maximal State Space mode,
+		optimized for fast verification.
+		For each, I list the CPU-time and wall-clock time elapsed,
+		plus the number of interleavings of the ultimately buggy state space tested,
+		before the bug was found.
+		* marks the winning measurements between each series.
+		Lastly, state space estimation, % \sect{\ref{sec....}}
+		though approximate at best,
+		confers a sense of the exponential explosion.
+		{\bf THESIS DRAFT NOTE:} These numbers all need to be re-run on account of some state space reductions and/or soundness fixes that have been added since these numbers were calculated in \cite{sigbovik-htm}.
+	}
+	\label{tab:buges}
+\end{table*}
+
+Table~\ref{tab:buges}
+presents the bug-finding results.
+% TODO: better sec'n reference?
+I configured Landslide to run Quicksand's Iterative Deepening algorithm (\sect{\ref{chap:quicksand}}),
+shown left,
+as well as to prioritize the maximal state space (\sect{\ref{sec:landslide-quicksand-options}}),
+shown right,
+% TODO: make 1 hour
+each with a time limit of 30 minutes.
+
+{\bf Finding bugs quickly.}
+As the test parameters increase,
+the multiplicative factor in bug-finding speed (2-4x, eyeballing) is generally smaller
+than that of the total number of interleavings (10-100x).
+In other words,
+should they exist,
+Landslide find bugs reasonably quickly in these transactional programs
+despite prohibitive exponential explosion in total state space size.
+This corroborates the results from Chapter~\ref{chap:quicksand},
+extending its good news to the world of HTM.
+
+{\bf Finding new bugs.}
+In addition to the bugs I intentionally wrote in {\tt htm1} and {\tt swapbug},
+%to my pleasant surprise % only the former of these is a surprise now
+Landslide also found two bugs in the ``real-world'' transactional algorithms I tested.
+\begin{itemize}
+	\item
+	{\bf Atomicity violation.}
+	{\tt avl\_insert} with any parameters higher than (2,1)
+	exposed a previously-unknown bug in the transactional AVL tree.
+	Figure~\ref{fig:avlbug} shows the root cause, essentially the {\tt htm1} bug in disguise.
+	% landslide-trace-1521093835.18927.html
+	This manifested
+	alternately
+	as a segfault (for test parameters (2,2) and (3,1)) % TODO double check
+	% landslide-trace-1521092404.23772.html
+	and
+	as a consistency-check assertion failure (for test parameters (2,3)).
+	The presence of {\tt while (\_retry);} makes the necessary preemption window extremely small
+	(between it and {\tt \_xbegin()}),
+	making the bug extremely unlikely to manifest under stress testing,
+	but Landslide is blind to such matters of chance.
+
+	As a matter of full disclosure,
+	while editing this test to make it Landslide-friendly,
+	I noted that the loop does not affect the test's possible behaviours,
+	only its {\em likely} ones,
+	and so removed it to keep the state space size manageable.
+	One could then argue that manual inspection is at least partly to credit for this result.
+	% TODO justify this
+	However, adding the loop back in to dispel any doubt,
+	% converting it to a {\tt yield()} loop % TODO: necessary?
+	Landslide still finds the bug,
+	albeit slightly slower,
+	in % TODO how much time
+	\item
+	% TODO
+	{\bf Spurious spinlock abort.}
+	{\tt lock\_fast}({\tt spinlock})
+\end{itemize}
+Note that in the AVL tree bug,
+the code's author was the very same person who proposed the protocol in Figure~\ref{fig:htm-fixed},
+yet still got it wrong once, having to write it out by hand throughout both data structures.
+This motivates the need for model checking such programs,
+no matter how much of a concurrency expert the author may be.
+It also suggests HTM primitives should be encapsulated behind higher-level abstractions,
+such as lock elision \cite{lock-elision}
+or a simple spinlock \cite{spinlock-rtm-github},
+which can be verified in isolation with smaller state spaces
+then trusted in turn when checking their client programs \cite{dbug-phdthesis}.
+\sect{\ref{sec:tm-verif}} explores this further.
+
+\begin{figure}[t]
+	\begin{center}
+		\begin{tabular}{l}
+		\texttt{\flow{while} (\_retry);} \\
+		\texttt{\flow{if} (\call{\_xbegin}() == \const{SUCCESS}) \{} \\
+		\texttt{~~~~\hilight{brickred}{tie}(\_root,inserted) = \call{\_insert}(\_root,n); } \\
+		\texttt{~~~~\call{\_xend}();} \\
+		\texttt{\} \flow{else} \{} \\
+		\texttt{~~~~\call{pthread\_mutex\_lock}(\&\_tree\_lock);} \\
+		\texttt{~~~~\_retry = \const{true};} \\
+		\texttt{~~~~\hilight{brickred}{tie}(\_root,inserted) = \call{\_insert}(\_root,n); } \\
+		\texttt{~~~~\_retry = \const{false};} \\
+		\texttt{~~~~\call{pthread\_mutex\_unlock}(\&\_tree\_lock);} \\
+		\texttt{\}} \\
+		\end{tabular}
+	\end{center}
+	%\caption{mario man had a very large house. or a castle.}
+	\caption{Unmodified code from {\tt htmavl.hpp} showing the previously-unknown bug found by Landslide.
+		The transaction path fails to check {\tt \_retry},
+		leading to data races and corruption just as in {\tt htm1}.
+		}
+	\label{fig:avlbug}
+\end{figure}
+
+\begin{figure}[t]
+	\begin{center}
+		% TODO synhi
+		\begin{tabular}{l}
+		\texttt{bool hle\_spinlock\_isfree(spinlock\_t* lock) \{} \\
+		\texttt{~~~~return (\_\_sync\_bool\_compare\_and\_swap(\&lock->v, 0, 0) ? true : false);} \\
+		\texttt{\}} \\
+		\texttt{void rtm\_spinlock\_acquire(spinlock\_t* lock) \{} \\
+		\texttt{~~~~if ((tm\_status = \_xbegin()) == \_XBEGIN\_STARTED) \{} \\
+		\texttt{~~~~~~~~if (hle\_spinlock\_isfree(lock)) return;} \\
+		\texttt{~~~~~~~~\_xabort(0xff);} \\
+		\texttt{~~~~\} else \{} \\
+		\texttt{~~~~~~~~// ... retrying, etc. abbreviated for brevity ...} \\
+		\texttt{~~~~~~~~hle\_spinlock\_acquire(lock);} \\
+		\texttt{~~~~\}} \\
+		\texttt{\}} \\
+		\texttt{void rtm\_spinlock\_release(spinlock\_t* lock)\{} \\
+		\texttt{~~~~if (hle\_spinlock\_isfree(lock)) \{} \\
+		\texttt{~~~~~~~~\_xend();} \\
+		\texttt{~~~~\} else \{} \\
+		\texttt{~~~~~~~~hle\_spinlock\_release(lock);} \\
+		\texttt{~~~~\}} \\
+		\texttt{\}} \\
+		\end{tabular}
+	\end{center}
+	% TODO describe
+	\caption{ff}
+	\label{fig:spinlockbug}
+\end{figure}
+
+\subsection{Performance}
+
+Quicksand's ability to find bugs in fewer distinct interleavings (i.e., overall smaller state spaces)
+does not necessarily correlate with better performance in terms of CPU-time.
+Comparing Table~\ref{tab:buges}'s trends
+to the break-even point in Quicksand's evaluation (\sect{\ref{sec:quicksand-eval}}),
+most of these tests are too small for its approach to pay off,
+% TODO: re check this after re running
+with {\tt swapbug}(3,1) and {\tt avl\_insert} as its notable wins.
+While plenty more wins were observed in \sect{\ref{sec:quicksand-eval}},
+%an upper bound also exists:
+%in {\tt swapbug}(3,2),
+%not even the {\em minimal} state space was completed in time.
+this suggests
+future MCs could prioritize state spaces using not just size estimation
+but a hybrid approach also conisdering state space maximality and preemption bounds \cite{chess-icb}
+to soften the trade-off
+both for smaller tests and for verification.
+
+\subsection{Verification}
+\label{sec:tm-verif}
+
+Landslide proved the following tests correct.
+With no bugs to find, comparing the different testing modes against each other is meaningless;
+I simply present their state space sizes and runtime
+(measured under {\tt -M})
+in Table~\ref{tab:verifs}.
+
+\begin{table}[h]
+	\begin{center}
+		\begin{tabular}{cc|r|r}
+			& & \cpu{\bf cpu (s)} & \ints{\bf SS size} \\
+			\bf test & \bf params & \em (or \ETAdag{\bf \em ETA}) & \em (or \ETAdag{\bf \em est.}) \\
+			\hline
+			\hline
+			{\tt htm2}
+			& 2,1 & \cpu{18.57}	& \ints{294}	\\
+			& 2,2 & \cpu{133.78}	& \ints{4902}	\\
+			& 2,3 &\cpu{1986.98}	& \ints{79017}	\\
+			& 3,1 &\cpu{11672.15}	& \ints{467730}	\\
+			& 3,2 &	\ETAdag{10d 14h}& \ETAdag{13763999}	\\
+			%[JOB 4] progress: 127032/13763999 brs (0.922929%), ETA 10d 13h 52m 44s (elapsed 59m 56s)
+			%[JOB 4] TIMED OUT (0.922929%; ETA 10d 13h 52m 44s)
+			\hline
+			{\tt counter}
+			& 2,1 & \cpu{5.57}	& \ints{30}	\\
+			& 2,2 & \cpu{15.53}	& \ints{384}	\\
+			& 2,3 &\cpu{155.00}	& \ints{5280}	\\
+			& 2,4 &\cpu{2211.10}	& \ints{75264}	\\
+			& 3,1 & \cpu{57.90}	& \ints{1960}	\\
+			& 3,2 & \cpu{10028.93}	& \ints{329888}	\\ % TODO running atm
+			\hline
+			{\tt swap}
+			%& 2,1 & \cpu{38.77}	& \ints{228}	\\ % pre-ignore_thr_func fix
+			& 2,1 & \cpu{32.98}	& \ints{193}	\\
+			& 2,2 & \cpu{3652.91}	& \ints{101150}	\\
+			%[JOB 8] progress: 97913/161897 brs (60.478362%), ETA 1h 29m 30s (elapsed 59m 52s)
+			% [JOB 8] TIMED OUT (60.478362%; ETA 1h 29m 30s)
+
+			& 3,1 & \ETAdag{8d 12h} & \ETAdag{411312}	\\
+			%[JOB 8] progress: 103873/411312 brs (25.254021%), ETA 8d 11h 36m 0s (elapsed 59m 42s)
+			%[JOB 8] TIMED OUT (25.254021%; ETA 8d 11h 36m 0s)
+			\hline
+			{\tt avl\_insert}
+			& 2,1 & \cpu{1083.35}	& \ints{40062}	\\
+			%& 2,2 & \multicolumn{2}{c}{\em (bug!)} \\
+			\hline
+			{\tt avl\_fixed}
+			& 2,1 & \cpu{1079.03} 	& \ints{45078} 	\\
+			& 2,2 & \ETAdag{2762y}	& \ETAdag{8714863} 	\\
+			%[JOB 19] progress: 45863/8714863 brs (0.526262%), ETA 2762y 28d 9h 57m 14s (elapsed 59m 25s)
+			%[JOB 19] TIMED OUT (0.526262%; ETA 2762y 28d 9h 57m 14s)
+			% the below is from avlfixed2,2 after running it overnight
+			% eta is ever worse but i'm using the 1hr one as promised
+			%[JOB 19] progress: 315672/29992026 brs (1.052520%), ETA 321008y 136d 18h 22m 12s (elapsed 6h 58m 15s)
+			& 3,1 & \ETAdag{12d 2h}	& \ETAdag{11498545}	\\
+
+			\hline
+			{\tt map\_basic}
+			& 2,1 & \ETAdag{10d 17h} & \ETAdag{16388977} \\
+			%[JOB 15] progress: 70330/16388977 brs (0.429130%), ETA 10d 16h 34m 15s (elapsed 43m 21s)
+			%[JOB 15] TIMED OUT (0.429130%; ETA 10d 16h 34m 15s)
+			\hline
+			{\tt map\_basicer}
+			& 2,1 & \cpu{877.44}& \ints{28635}	\\
+			& 2,2 & \ETAdag{2d 7h} & \ETAdag{5925634} \\
+			& 3,1 & \ETAdag{468d 13h} & \ETAdag{35893653} \\
+			% [JOB 5] progress: 100181/35893653 brs (0.279105%), ETA 1y 103d 12h 50m 47s (elapsed 59m 24s)
+			%[JOB 5] TIMED OUT (0.279105%; ETA 1y 103d 12h 50m 47s)
+			% i don't know what this is but it was hanging out under 3,1 here :<
+			%[JOB 6] progress: 35042/5925634 brs (0.591363%), ETA 2d 6h 40m 30s (elapsed 21m 12s)
+			%[JOB 6] TIMED OUT (0.591363%; ETA 2d 6h 40m 30s)
+		\end{tabular}
+	\end{center}
+	\caption{Transactional tests verified (or not) by Landslide.
+		{\bf THESIS DRAFT NOTE:} These numbers all need to be re-run on account of some state space reductions and/or soundness fixes that have been added since these numbers were calculated in \cite{sigbovik-htm}.
+		}
+	\label{tab:verifs}
+\end{table}
+
+% TODO rephrase the outdated para from sigbovik here,
+% probably after rerunning expts
+
+{\bf State space reduction.}
+
+% as for the sleep-set-alike optimization,
+% we see it would provide 25% reduction in the example (from thesprop)
+% however, evaluating its potential impact in general is tantamount to implementing it.
+
 % TODO
@@
 % then in AVL/etc have landslide, whenever it is reached, force the user into an alternate path which just always takes the mx
+
+% new spinlock test resulce - htm_spinlock
+% (2,1) without isfree fix
+%%%% total time elapsed: 9s
+%%%% [JOB 4] COMPLETE (41 interleavings tested; 3s elapsed)
+%%%% total CPU time consumed: 14s (14231661 usecs) (core saturation: 37%)
+
+% (2,2) without isfree fix
+%%%% total time elapsed: 1h 22m 20s
+%%%% [JOB 4] COMPLETE (38544 interleavings tested; 1h 22m 13s elapsed)
+%%%% total CPU time consumed: 1h 22m 24s (4944373889 usecs) (core saturation: 25%)
+
+% (3,1) without isfree fix
+
+% (2,1) with isfree fix
+%%%% total time elapsed: 10s
+%%%% [JOB 4] COMPLETE (31 interleavings tested; 2s elapsed)
+%%%% total CPU time consumed: 13s (13256415 usecs) (core saturation: 30%)
+
+% (2,2) with isfree fix
+%%%% total time elapsed: 25m 1s
+%%%% [JOB 4] COMPLETE (15076 interleavings tested; 24m 41s elapsed)
+%%%% total CPU time consumed: 25m 4s (1504809436 usecs) (core saturation: 25%)
+% (nb. it's not known whether this was run from a "hot" start (no recompiling) -- but probably wasnt!
+
+% (2,3)
+%% this will be an overnight 10hr job
+% [DBG] [JOB 4] progress: 30353/444731 brs (6.825013%),
+% [JOB 4] Running (6.825013%; ETA 5d 9h 4m 41s)
+
+% (3,1) with isfree fix
+% hot start confirmed
+%%%% total time elapsed: 41m 26s
+%%%% [JOB 4] COMPLETE (24790 interleavings tested; 41m 18s elapsed)
+%%%% total CPU time consumed: 41m 28s (2488953483 usecs) (core saturation: 25%)
+
+%%%% avl_mutex
+% % 2,1
+% [JOB 0] COMPLETE (3 interleavings tested; 2s elapsed)
+% total CPU time consumed: 2s (2489515 usecs) (core saturation: 24%)
+% % 2,2
+% [JOB 0] COMPLETE (3 interleavings tested; 1s elapsed)
+% total CPU time consumed: 2s (2443451 usecs) (core saturation: 24%)
+% % 2,3
+% [JOB 0] COMPLETE (3 interleavings tested; 1s elapsed)
+% total CPU time consumed: 2s (2448127 usecs) (core saturation: 24%)
+% % 2,4
+% [JOB 0] COMPLETE (3 interleavings tested; 1s elapsed)
+% total CPU time consumed: 2s (2407903 usecs) (core saturation: 24%)
+% % 3,1
+% [JOB 0] COMPLETE (12 interleavings tested; 2s elapsed)
+% total CPU time consumed: 2s (2685913 usecs) (core saturation: 24%)
+% %3,2
+% [JOB 0] COMPLETE (12 interleavings tested; 2s elapsed)
+% total CPU time consumed: 2s (2666826 usecs) (core saturation: 24%)
+% % 3,3
+% [JOB 0] COMPLETE (13 interleavings tested; 2s elapsed)
+% total CPU time consumed: 2s (2761157 usecs) (core saturation: 24%)
+% % 3,4
+% [JOB 0] COMPLETE (13 interleavings tested; 2s elapsed)
+% total CPU time consumed: 2s (2738169 usecs) (core saturation: 24%)
+% % 4,1
+% [JOB 0] COMPLETE (60 interleavings tested; 3s elapsed)
+% total CPU time consumed: 4s (4034411 usecs) (core saturation: 24%)
+% % 4,2
+% [JOB 0] COMPLETE (66 interleavings tested; 3s elapsed)
+% total CPU time consumed: 4s (4338442 usecs) (core saturation: 24%)
+% % 4,3
+% [JOB 0] COMPLETE (71 interleavings tested; 4s elapsed)
+% total CPU time consumed: 4s (4526641 usecs) (core saturation: 24%)
+% % 4,4
+% [JOB 0] COMPLETE (73 interleavings tested; 4s elapsed)
+% total CPU time consumed: 4s (4752864 usecs) (core saturation: 24%)
+
+%%%% map mutex
+% % TODO - bc of this weird nonlinear behaviour around 2,1+, yo ushoud prob try 2,2 for map_basic as well
+% % 2,1
+% total time elapsed: 19s
+% [JOB 7] COMPLETE (57 interleavings tested; 5s elapsed)
+%        PPs: { 'mutex_lock' 'mutex_unlock' 'DR @ 0x100053b' [NONDET] 'DR @ 0x1000552' [NONDET] 'DR @ 0x1000521' [NONDET]
+%        'DR @ 0x100055d' [NONDET] 'DR @ 0x100055a' [NONDET] 'DR @ 0x1000916' [NONDET] 'DR @ 0x1000919' [NONDET] }
+% total CPU time consumed: 31s (31198379 usecs) (core saturation: 39%)
+% % 2,2
+% total time elapsed: 3s
+% [JOB 0] COMPLETE (6 interleavings tested; 2s elapsed)
+%        PPs: { 'mutex_lock' 'mutex_unlock' }
+% =========================
+% progress report thr exiting
+% total CPU time consumed: 3s (3192831 usecs) (core saturation: 24%)
+% % 2,3
+% ==== PROGRESS REPORT ====
+% total time elapsed: 3s
+% [JOB 0] COMPLETE (6 interleavings tested; 2s elapsed)
+%        PPs: { 'mutex_lock' 'mutex_unlock' }
+% =========================
+% progress report thr exiting
+% total CPU time consumed: 3s (3240367 usecs) (core saturation: 24%)
+% 2,4
+% 3,1
+% 3,2
+
+%%%%%%%%%%% htm_mutex
+%%%% -- nb. these numbers were obtained with a oops-exploded state space from a misplaced assert on hm->held
+% % 2,1
+% total time elapsed: 6s
+% [JOB 2] COMPLETE (24 interleavings tested; 3s elapsed)
+% total CPU time consumed: 9s (9642348 usecs) (core saturation: 35%)
+% % 2,2
+% total time elapsed: 1m 20s
+% [JOB 2] COMPLETE (2672 interleavings tested; 1m 16s elapsed)
+% total CPU time consumed: 1m 23s (83001932 usecs) (core saturation: 25%)
+% % 2,3
+% total time elapsed: 2h 3m 45s
+% [JOB 2] COMPLETE (220338 interleavings tested; 2h 3m 40s elapsed)
+% total CPU time consumed: 2h 3m 47s (7427873599 usecs) (core saturation: 25%)
+% % 3,1
+% total time elapsed: 51s
+% [JOB 2] COMPLETE (1594 interleavings tested; 47s elapsed)
+% total CPU time consumed: 54s (54572810 usecs) (core saturation: 26%)
+
+%%%% -- nb. better version
+% % 2,1
+% total time elapsed: 6s
+% [JOB 2] COMPLETE (22 interleavings tested; 2s elapsed)
+% total CPU time consumed: 9s (9415448 usecs) (core saturation: 34%)
+% % 2,2
+% total time elapsed: 40s
+% [JOB 2] COMPLETE (1208 interleavings tested; 37s elapsed)
+% total CPU time consumed: 44s (44068381 usecs) (core saturation: 27%)
+% 2,3
+% 2,4
+% % 3,1
+% total time elapsed: 38s
+% [JOB 2] COMPLETE (1148 interleavings tested; 35s elapsed)
+% total CPU time consumed: 42s (42108005 usecs) (core saturation: 27%)
+% % 3,2
+% [JOB 2] progress: 920684/4793666 brs (19.206260%), ETA 11d 4h 17m 50s (elapsed 9h 59m 56s)
+% total time elapsed: 10h 0s
+% [JOB 2] TIMED OUT (19.206260%; ETA 11d 4h 17m 50s)
+% total CPU time consumed: 10h 3s (36003199876 usecs) (core saturation: 25%)
+
+% commentary -- writing this test case well requires expert knowledge of noobs (of writing landslide friendly tests)
+% - placing the asserts well, using sync_test_and_set and asserting on the result rather than having a separate read,
+%   has an impact on the state space, as seen above
+% - design of critical_section and how it uses xtest to yield only in failure path, and why that's justified
+
+% "Landslide cannot in general test for performance degradation bugs,
+% such as in a program which requires a transaction abort rate below a certain threshhold
+% owing to the nature of its workload,
+% because all interleavings are equal in Landslide's eyes, and probability is no object.
+% However, the lock_noconflict test illustrates that model checking can still test for certain performance properties
+% as long as the element of probability can be removed."
+
+% other one to test -
+% https://github.com/vamsikc/leveldb-tsx/blob/master/ext/xsync/include/scope.hpp
+
+% "Although we skipped implementing it to begin with, the sheer size of this reduction also
+% reflects on the numbers one might see betwee the original reduction in theorem so-and-so that lets us
+% treat transactional success paths as mutually-exclusive (and, once started, abort-proof) to begin with"
+% this sentence is a complete mess....

### 2018-07-15 thesis/tm.tex
fill in lots of eval stuff, start filling in verif table, discuss abstraction reduction

@@
 in addition to performing DPOR to select non-independent thread interleavings (\sect{\ref{sec:landslide-dpor}}),
-% TODO: reduction challenge?
 the abort codes under each {\tt xbegin} preemption point are also considered ``marked'' paths which must be tested.
@@
 
+\subsubsection{Tracking abort sets}
+\label{sec:tm-abortsets}
+
+% TODO
+
 \subsection{Data Race Analysis}
@@
 When a thread reaches {\tt \_xend()},
-the above flag is cleared,
+the flag is cleared,
 and under Pure Happens-Before,
@@
 under Limited Happens-Before,
-it is considered not a data race if not both threads' {\tt user\_txn} flags were set in addition to the usual conditions;
+it is considered a data race only if at least one thread's {\tt user\_txn} was not set
+in addition to the usual conditions; % in multiplication to..? && c.c
 under Pure Happens-Before,
@@
 	%\item How quickly does Landslide verify correct transactional programs of varying sizes?
-	\item How does Iterative Deepening's performance compare to Maximal State Space mode?%
+	\item How does Iterative Deepening's (\sect{\ref{sec:quicksand-id}})
+		performance compare to Maximal State Space mode (\sect{\ref{sec:landslide-quicksand-options}})?%
 		\footnote{Not necessarily related to HTM,
@@
 			cannot access the critical section simultaneously.
+			% htm_spinlock and htm_mutex
 		\item {\tt lock\_fast}(): Checks that a transactional lock's fast path will not suffer conflict aborts
 			if its client threads' critical sections are independent.
+			% htm_spinlock_fastpath and htm_mutex_fastpath
 	\end{itemize}
@@
 \subsection{Bugs}
-
-\begin{table*}[t]
+\label{sec:tm-eval-bugs}
+
+\begin{table*}[p]
 	\begin{center}
@@
 			& 3,2	&\cpu{*1569.09}&\wtm{*216.85}&\ints{*209}&\ETA{--}&\ETA{>30m}&\ETA{--}& \ETA{1402363529} \\
+		\hline
+		{\tt lock\_fast}
+			& 2,1	& \\
+		%({\tt spinlock})
+		(perf)
+			& 9,9	& \\
 	\end{tabular}
@@
 	\caption{Landslide's bug-finding performance on various test configurations.
-		Iterative Deepening, optimized for fast bug-finding,
-		is compared against Maximal State Space mode,
+		Iterative Deepening (\sect{\ref{sec:quicksand-id}}), optimized for fast bug-finding,
+		is compared against Maximal State Space mode (\sect{\ref{sec:landslide-quicksand-options}}),
 		optimized for fast verification.
@@
 		* marks the winning measurements between each series.
-		Lastly, state space estimation, % \sect{\ref{sec....}}
+		Lastly, state space estimation (\sect{\ref{sec:landslide-estimate}}),
 		though approximate at best,
@@
 presents the bug-finding results.
-% TODO: better sec'n reference?
-I configured Landslide to run Quicksand's Iterative Deepening algorithm (\sect{\ref{chap:quicksand}}),
+I configured Landslide to run Quicksand's Iterative Deepening algorithm, %(\sect{\ref{sec:quicksand-id}}),
 shown left,
-as well as to prioritize the maximal state space (\sect{\ref{sec:landslide-quicksand-options}}),
+as well as to prioritize the maximal state space, %(\sect{\ref{sec:landslide-quicksand-options}}),
 shown right,
@@
 
-{\bf Finding bugs quickly.}
+\subsubsection{Finding bugs quickly}
+
 As the test parameters increase,
@@
 
-{\bf Finding new bugs.}
+\subsubsection{Finding new bugs}
+
 In addition to the bugs I intentionally wrote in {\tt htm1} and {\tt swapbug},
@@
 	\item
+	{\bf Spurious spinlock abort.}
 	% TODO
-	{\bf Spurious spinlock abort.}
 	{\tt lock\_fast}({\tt spinlock})
+	\footnote{{\tt lock\_fast}'s unusual (9,9) parameter shows that this state space size is constant: %in $(K,N)$:
+	DPOR will always either deem all thread transitions independent and end exploration immediately,
+	or the test's assertion will trip as soon as the first conflict is found.}
+
+	As another matter of full disclosure,
+	I noticed this bug through manual inspection,
+	while adapting the spinlock's client code to be Landslide-friendly,
+	then wrote {\tt lock\_fast} specifically to target this behaviour,
+	so unlike {\tt avl\_insert}, it does not count as Landslide finding a previously-unknown bug.
+	However, I feel in retrospect that how and when
+	an HTM-backed concurrency abstraction
+	will fall into its slow path
+	is a reasonable performance property for a user to want to verify,
+	so I consider Landslide confirming the bug (and later verifying its absence, in \sect{\ref{sec:tm-verif}})
+	a positive result anyway.
 \end{itemize}
@@
 \sect{\ref{sec:tm-verif}} explores this further.
+
+As HTM is fundamentally a performance-minded concurrency extension,
+users are likely to care about more probabilistic properties of their code,
+such as requiring a transaction abort rate below a certain threshhold
+owing to the nature of its workload.
+Landslide cannot in general test for performance degradation bugs,
+because all interleavings are equal in Landslide's eyes, and probability is no object.
+However, {\tt lock\_fast} illustrates that model checking can still check some interesting performance properties
+as long as the element of probability can be removed.
+Future work may attempt to verify a wider range of performance properties,
+but with a hybrid approach between model checking and what other technique is not yet known.
 
@@
 	%\caption{mario man had a very large house. or a castle.}
-	\caption{Unmodified code from {\tt htmavl.hpp} showing the previously-unknown bug found by Landslide.
+	\caption{Unmodified code from {\tt htmavl.hpp} showing the previously-unknown bug
+		Landslide found in {\tt avl\_insert}.
 		The transaction path fails to check {\tt \_retry},
@@
 		\texttt{bool hle\_spinlock\_isfree(spinlock\_t* lock) \{} \\
-		\texttt{~~~~return (\_\_sync\_bool\_compare\_and\_swap(\&lock->v, 0, 0) ? true : false);} \\
+		\texttt{~~~~// XXX: should be "return lock->v == 0;"} \\
+		\texttt{~~~~return \_\_sync\_bool\_compare\_and\_swap(\&lock->v, 0, 0);} \\
 		\texttt{\}} \\
@@
 		\texttt{~~~~\} else \{} \\
-		\texttt{~~~~~~~~// ... retrying, etc. abbreviated for brevity ...} \\
+			\texttt{~~~~~~~~// ... retrying \&c abbreviated for brevity ...} \\
 		\texttt{~~~~~~~~hle\_spinlock\_acquire(lock);} \\
@@
 	% TODO describe
-	\caption{ff}
+	\caption{Code from {\tt spinlock-rtm.c}, modified only to remove unrelated logic for brevity,
+		showing the performance bug Landslide found in {\tt lock\_fast}({\tt spinlock}).
+		The {\tt isfree()} routine uses an atomic read-and-write operation where just a read would suffice,
+		which leads to superfluous memory conflicts in the transactional path
+		(seen at both of its callsites below).
+	}
 	\label{fig:spinlockbug}
@@
 	\begin{center}
+		\footnotesize
 		\begin{tabular}{cc|r|r}
@@
 			%[JOB 6] TIMED OUT (0.591363%; ETA 2d 6h 40m 30s)
+
+			\hline
+			\multicolumn{4}{c}{\em everything above this line needs rerun} \\
 		\end{tabular}
@@
 	\caption{Transactional tests verified (or not) by Landslide.
+		Run with {\tt landslide -M -X}, i.e., retry aborts enabled, different xabort codes not distinguished.
+		State space estimates measured after a timeout of 10 hours.
 		{\bf THESIS DRAFT NOTE:} These numbers all need to be re-run on account of some state space reductions and/or soundness fixes that have been added since these numbers were calculated in \cite{sigbovik-htm}.
@@
 
-{\bf State space reduction.}
-
-% as for the sleep-set-alike optimization,
-% we see it would provide 25% reduction in the example (from thesprop)
-% however, evaluating its potential impact in general is tantamount to implementing it.
-
-% TODO
-
-% TODO: evaluate a reduction in the AVL/etc where you take the xbegin/retry/mutex/retry pattern and encapsulate it in smth like
-% T txn(mutex_t *mp, T body(U env), U env)
-% verify this primitive on some simple mutual exclusion tests up to N,K or w/e
-% then in AVL/etc have landslide, whenever it is reached, force the user into an alternate path which just always takes the mx
-
+\subsection{Reduction}
+
+While verifying the soundness of Figure~\ref{fig:htm-fixed}'s atomicity protocol is a positive result,
+the previous section's attempts at several larger real-world programs quickly suffered exponential explosion
+for even small thread/iteration counts,
+in one case failing to make any guarantee about even the minimum possible test configuration.
+I explore two possible mitigation approaches.
+
+\subsubsection{Abstraction reduction}
+
+Visual inspection of the AVL tree and separate-chaining map implementations \cite{tm-benchmark-cmu},
+after correcting the former's atomicity bug (Figure~\ref{fig:avlbug}),
+reveals that every use of HTM followed exactly the same pattern,
+running identical data structure logic in both the transactional and abort paths,
+as though HTM were merely a mutual exclusion lock with fancy performance characteristics.
+% this citation is a lil problematic
+Prior work \cite{dbug-phdthesis} proposed {\em abstraction reduction},
+in which the user identifies program components that can be separated by a well-understood API,
+then tests each one against the API individually,
+effectively turning multiplicative state space size factors into additive ones.
+
+In this case, I split the lock-like HTM use
+and the mutually-exclusive data structure code
+into separate tests,
+{\tt lock}, which checks the use of HTM guarantees mutual exclusion,
+and {\tt avl\_mutex}/{\tt map\_mutex},
+which replace the open-coded HTM use with an already-trusted P2 mutex.
+{\tt lock\_fast},
+a bonus test,
+checks the transactional lock's performance by asserting that
+its internal logic won't trigger conflict aborts
+even when the client's accesses are independent.
+Figure~\ref{fig:htm-lock} shows the core logic of these tests.
+Finally, I parameterized them over how the lock was implemented:
+{\tt spinlock} ``real-world'' code from \cite{spinlock-rtm-github},
+{\tt spin\_fixed} the same with the performance bug from \sect{\ref{sec:tm-eval-bugs}} fixed,
+and {\tt mutex} using Landslide-annotated P2 mutexes (as the AVL and map implementations do).
+
+\begin{figure}[h]
+	% TODO: syn hi
+	\begin{center}
+	\begin{tabular}{p{0.47\textwidth}p{0.5\textwidth}}
+		\begin{tabular}{l}
+			\texttt{static int num\_in\_section = 0;} \\
+			\texttt{for (int i = 0; i < NITERS; i++) \{} \\
+			\texttt{~~~~rtm\_spinlock\_acquire(\&lock);} \\
+			\texttt{~~~~num\_in\_section++;} \\
+			\texttt{~~~~if (!\_xtest())} \\
+			\texttt{~~~~~~~~thr\_yield(-1);} \\
+			\texttt{~~~~assert(num\_in\_section == 1); } \\
+			\texttt{~~~~num\_in\_section--;} \\
+			\texttt{~~~~rtm\_spinlock\_release(\&lock);} \\
+			\texttt{\}} \\
+		\end{tabular}
+		&
+		\begin{tabular}{ll}
+			\texttt{for (int i = 0; i < NITERS; i++) \{} \\
+			\texttt{~~~~rtm\_spinlock\_acquire(\&lock);} \\
+			\texttt{~~~~assert(\_xtest());} \\
+			\texttt{~~~~rtm\_spinlock\_release(\&lock);} \\
+			\texttt{\}} \\
+		\end{tabular}
+		\\
+		\\
+		(a) {\tt lock}(), tests mutual exclusion.
+		&
+		(b) {\tt lock\_fast}(), tests for no spurious aborts.
+	\end{tabular}
+	\end{center}
+	\caption{Abstraction reduction test cases.}
+	\label{fig:htm-lock}
+\end{figure}
+
+Table~\ref{tab:verifs2} shows the new resulting levels of verification
+Landslide reached before the same 10-hour timeout.
+Provided that one trusts the {\tt lock} tests correctly check the desired properties,
+and that open-coding hadn't introduced any new bugs (such as Figure~\ref{fig:avlbug}'s),
+the benefit is clear:
+for example,
+summing the testing times of {\tt lock}({\tt mutex})$(K,N)$ and {\tt avl\_mutex}$(K,N)$
+produces the same verification as {\tt avl\_fixed}$(K,N)$ far more cheaply.
+Furthermore, {\tt lock}'s verification can be reused,
+whereas {\tt avl\_fixed} and {\tt map\_basic} effectively duplicated the mutex verification between them.
+Note also the impact that fixing Figure~\ref{fig:spinlockbug}'s performance bug
+(changing a read+write to a read only)
+had on even the correctness tests:
+{\tt lock}({\tt spin\_fixed})'s state spaces were reduced by more than half compared to {\tt lock}({\tt spinlock}).
+
+\begin{table}[h]
+	\begin{center}
+		\footnotesize
+		%\begin{tabular}{cc}
+		%\begin{tabular}{cc||r|r||r|r}
+		%	& & \multicolumn{2}{c||}{\bf DPOR} & \multicolumn{2}{c}{\bf DPOR + abort sets} \\
+		%	& & \cpu{\bf cpu (s)} & \ints{\bf SS size}
+		%	  & \cpu{\bf cpu (s)} & \ints{\bf SS size} \\
+		%	\bf test & \bf params & \em (or \ETAdag{\bf \em ETA}) & \em (or \ETAdag{\bf \em est.})
+		%	  & \em (or \ETAdag{\bf \em ETA}) & \em (or \ETAdag{\bf \em est.}) \\
+		\begin{tabular}{cc||r|r}
+			& & \cpu{\bf cpu (s)} & \ints{\bf SS size} \\
+			\bf test & \bf params & \em (or \ETAdag{\bf \em ETA}) & \em (or \ETAdag{\bf \em est.}) \\
+			\hline
+			\hline
+			{\tt lock}
+			& 2,1 & \cpu{14.23} & \ints{41} \\
+			({\tt spinlock})
+			& 2,2 & \cpu{4944.37} & \ints{38544} \\
+			& 2,3 & \\ % \cpu{0} & \ints{0} \\
+			& 3,1 & \\ % \cpu{0} & \ints{0} \\ % TODO currently running, i think,
 			% new spinlock test resulce - htm_spinlock
@@
 			%%%% total CPU time consumed: 14s (14231661 usecs) (core saturation: 37%)
-
 			% (2,2) without isfree fix
@@
 			%%%% total CPU time consumed: 1h 22m 24s (4944373889 usecs) (core saturation: 25%)
-
 			% (3,1) without isfree fix
-
+			\hline
+			{\tt lock}
+			& 2,1 & \cpu{12.26} & \ints{31} \\
+			({\tt spin\_fixed})
+			& 2,2 & \cpu{1662.21} & \ints{15076} \\
+			& 2,3 & \\ %\cpu{0} & \ints{0} \\
+			& 3,1 & \cpu{2488.95} & \ints{24970} \\
 			% (2,1) with isfree fix
@@
 			%%%% total CPU time consumed: 13s (13256415 usecs) (core saturation: 30%)
-
 			% (2,2) with isfree fix
-%%%% total time elapsed: 25m 1s
-%%%% [JOB 4] COMPLETE (15076 interleavings tested; 24m 41s elapsed)
-%%%% total CPU time consumed: 25m 4s (1504809436 usecs) (core saturation: 25%)
-% (nb. it's not known whether this was run from a "hot" start (no recompiling) -- but probably wasnt!
-
+			% total time elapsed: 27m 39s
+			% [JOB 4] COMPLETE (15076 interleavings tested; 27m 28s elapsed)
+			% total CPU time consumed: 27m 42s (1662213886 usecs) (core saturation: 25%)
 			% (2,3)
@@
 			% [JOB 4] Running (6.825013%; ETA 5d 9h 4m 41s)
-
 			% (3,1) with isfree fix
@@
 			%%%% total CPU time consumed: 41m 28s (2488953483 usecs) (core saturation: 25%)
-
+			\hline
+			{\tt lock}
+			& 2,1 & \cpu{9.42} & \ints{22}	\\
+			({\tt mutex})
+			& 2,2 & \cpu{44.07} & \ints{1208}	\\
+			& 2,3 & \cpu{2647.38} & \ints{71136}	\\
+			& 2,4 & \ETAdag{2d 1h} & \ETAdag{3223055}	\\
+			& 3,1 & \cpu{42.11} & \ints{1148}	\\
+			& 3,2 & \ETAdag{11d 4h} & \ETAdag{4793666}	\\
+			& 4,1 & \cpu{5586.37} & \ints{159664}	\\
+			%%%%%%%%%%% htm_mutex
+			%%%% -- nb. these numbers were obtained with a oops-exploded state space
+			%%%% from a misplaced assert on hm->held
+			% % 2,1
+			% total time elapsed: 6s
+			% [JOB 2] COMPLETE (24 interleavings tested; 3s elapsed)
+			% total CPU time consumed: 9s (9642348 usecs) (core saturation: 35%)
+			% % 2,2
+			% total time elapsed: 1m 20s
+			% [JOB 2] COMPLETE (2672 interleavings tested; 1m 16s elapsed)
+			% total CPU time consumed: 1m 23s (83001932 usecs) (core saturation: 25%)
+			% % 2,3
+			% total time elapsed: 2h 3m 45s
+			% [JOB 2] COMPLETE (220338 interleavings tested; 2h 3m 40s elapsed)
+			% total CPU time consumed: 2h 3m 47s (7427873599 usecs) (core saturation: 25%)
+			% % 3,1
+			% total time elapsed: 51s
+			% [JOB 2] COMPLETE (1594 interleavings tested; 47s elapsed)
+			% total CPU time consumed: 54s (54572810 usecs) (core saturation: 26%)
+			%%%% -- nb. better version
+			% % 2,1
+			% total time elapsed: 6s
+			% [JOB 2] COMPLETE (22 interleavings tested; 2s elapsed)
+			% total CPU time consumed: 9s (9415448 usecs) (core saturation: 34%)
+			% % 2,2
+			% total time elapsed: 40s
+			% [JOB 2] COMPLETE (1208 interleavings tested; 37s elapsed)
+			% total CPU time consumed: 44s (44068381 usecs) (core saturation: 27%)
+			% % 2,3
+			% total time elapsed: 44m 4s
+			% [JOB 2] COMPLETE (71136 interleavings tested; 44m 0s elapsed)
+			% total CPU time consumed: 44m 7s (2647380230 usecs) (core saturation: 25%)
+			% % 2,4
+			% [JOB 2] progress: 957248/3223055 brs (29.700013%), ETA 2d 1h 24m 44s (elapsed 9h 59m 56s)
+			% total time elapsed: 10h 0s
+			% [JOB 2] TIMED OUT (29.700013%; ETA 2d 1h 24m 44s)
+			% total CPU time consumed: 10h 3s (36003560403 usecs) (core saturation: 25%)
+			% % 3,1
+			% total time elapsed: 38s
+			% [JOB 2] COMPLETE (1148 interleavings tested; 35s elapsed)
+			% total CPU time consumed: 42s (42108005 usecs) (core saturation: 27%)
+			% % 3,2
+			% [JOB 2] progress: 920684/4793666 brs (19.206260%), ETA 11d 4h 17m 50s (elapsed 9h 59m 56s)
+			% total time elapsed: 10h 0s
+			% [JOB 2] TIMED OUT (19.206260%; ETA 11d 4h 17m 50s)
+			% total CPU time consumed: 10h 3s (36003199876 usecs) (core saturation: 25%)
+			% % 4,1
+			% total time elapsed: 1h 33m 3s
+			% [JOB 2] COMPLETE (159664 interleavings tested; 1h 32m 59s elapsed)
+			% total CPU time consumed: 1h 33m 6s (5586369677 usecs) (core saturation: 25%)
+			\hline
+			{\tt lock\_fast}
+			& 2,1 & \cpu{0} & \ints{0}	\\
+			({\tt spin\_fixed})
+			& 9,9 & \cpu{0} & \ints{0}	\\
+			\hline
+			{\tt lock\_fast}
+			& 2,1 & \cpu{0} & \ints{0}	\\
+			({\tt mutex})
+			& 9,9 & \cpu{0} & \ints{0}	\\
+			%\end{tabular}
+			%&
+			%\begin{tabular}{cc||r|r}
+			%& & \cpu{\bf cpu (s)} & \ints{\bf SS size} \\
+			%\bf test & \bf params & \em (or \ETAdag{\bf \em ETA}) & \em (or \ETAdag{\bf \em est.}) \\
+			\hline
+			\hline
+			{\tt avl\_mutex}
+			& 2,1 & \cpu{2.49} & \ints{3} \\
+			& 2,2 & \cpu{2.44} & \ints{3} \\
+			& 2,3 & \cpu{2.45} & \ints{3} \\
+			& 2,4 & \cpu{2.41} & \ints{3} \\
+			& 3,1 & \cpu{2.69} & \ints{12} \\
+			& 3,2 & \cpu{2.67} & \ints{12} \\
+			& 3,3 & \cpu{2.76} & \ints{13} \\
+			& 3,4 & \cpu{2.74} & \ints{13} \\
+			& 4,1 & \cpu{4.03} & \ints{60} \\
+			& 4,2 & \cpu{4.34} & \ints{66} \\
+			& 4,3 & \cpu{4.53} & \ints{71} \\
+			& 4,4 & \cpu{4.75} & \ints{73} \\
 			%%%% avl_mutex
@@
 			% total CPU time consumed: 4s (4752864 usecs) (core saturation: 24%)
+			\cline{1-4}
+			{\tt map\_mutex}
+			& 2,1 & \\ %\cpu{0} & \ints{0} \\
+			& 2,2 & \\ %\cpu{0} & \ints{0} \\
+			& 2,3 & \\ %\cpu{0} & \ints{0} \\
+			& 3,1 & \\ %\cpu{0} & \ints{0} \\
+			& 3,2 & \\ %\cpu{0} & \ints{0} \\
+		\end{tabular}
+		%\end{tabular}
+	\end{center}
+	\caption{Continuation of Table~\ref{tab:verifs}.
+		Demonstrates abstraction reduction \cite{dbug-phdthesis}
+		on the {\tt avl\_fixed} and {\tt map\_basic} tests
+		by verifying HTM mutex implementations separately.
+		Run with {\tt landslide -M -X -A -S}, i.e.,
+		conflict and explicit aborts tested separately,
+		and retry aborts suppressed
+		(both lock implementations include a retry loop).
+		The reduced {\tt avl\_mutex} and {\tt map\_mutex} are, of course, no longer transactional.
+		%% i mean ok xabort sets doesn't really help for ANY test tho
+		%so the abort sets optimization does not apply to them.
+		}
+	\label{tab:verifs2}
+\end{table}
+
+\subsubsection{Abort set reduction}
+
+% TODO
+
+(\sect{\ref{sec:tm-abortsets}}).
+
+% as for the sleep-set-alike optimization,
+% we see it would provide 25% reduction in the example (from thesprop)
+% however, evaluating its potential impact in general is tantamount to implementing it.
 
@@
 
-%%%%%%%%%%% htm_mutex
-%%%% -- nb. these numbers were obtained with a oops-exploded state space from a misplaced assert on hm->held
-% % 2,1
-% total time elapsed: 6s
-% [JOB 2] COMPLETE (24 interleavings tested; 3s elapsed)
-% total CPU time consumed: 9s (9642348 usecs) (core saturation: 35%)
-% % 2,2
-% total time elapsed: 1m 20s
-% [JOB 2] COMPLETE (2672 interleavings tested; 1m 16s elapsed)
-% total CPU time consumed: 1m 23s (83001932 usecs) (core saturation: 25%)
-% % 2,3
-% total time elapsed: 2h 3m 45s
-% [JOB 2] COMPLETE (220338 interleavings tested; 2h 3m 40s elapsed)
-% total CPU time consumed: 2h 3m 47s (7427873599 usecs) (core saturation: 25%)
-% % 3,1
-% total time elapsed: 51s
-% [JOB 2] COMPLETE (1594 interleavings tested; 47s elapsed)
-% total CPU time consumed: 54s (54572810 usecs) (core saturation: 26%)
-
-%%%% -- nb. better version
-% % 2,1
-% total time elapsed: 6s
-% [JOB 2] COMPLETE (22 interleavings tested; 2s elapsed)
-% total CPU time consumed: 9s (9415448 usecs) (core saturation: 34%)
-% % 2,2
-% total time elapsed: 40s
-% [JOB 2] COMPLETE (1208 interleavings tested; 37s elapsed)
-% total CPU time consumed: 44s (44068381 usecs) (core saturation: 27%)
-% 2,3
-% 2,4
-% % 3,1
-% total time elapsed: 38s
-% [JOB 2] COMPLETE (1148 interleavings tested; 35s elapsed)
-% total CPU time consumed: 42s (42108005 usecs) (core saturation: 27%)
-% % 3,2
-% [JOB 2] progress: 920684/4793666 brs (19.206260%), ETA 11d 4h 17m 50s (elapsed 9h 59m 56s)
-% total time elapsed: 10h 0s
-% [JOB 2] TIMED OUT (19.206260%; ETA 11d 4h 17m 50s)
-% total CPU time consumed: 10h 3s (36003199876 usecs) (core saturation: 25%)
-
 % commentary -- writing this test case well requires expert knowledge of noobs (of writing landslide friendly tests)
@@
 
-% "Landslide cannot in general test for performance degradation bugs,
-% such as in a program which requires a transaction abort rate below a certain threshhold
-% owing to the nature of its workload,
-% because all interleavings are equal in Landslide's eyes, and probability is no object.
-% However, the lock_noconflict test illustrates that model checking can still test for certain performance properties
-% as long as the element of probability can be removed."
-
 % other one to test -

### 2018-07-15 thesis/tm.tex
lock_fast

@@
 Likewise, since Landslide's HTM semantics emulation relies on the equivalences proved in \sect{\ref{sec:tm-design}},
-we did not actually implement an HTM-style speculative-execution-and-rollback simulation mode.
+I did not actually implement an HTM-style speculative-execution-and-rollback simulation mode.
 % so have no basis to measure how much it saves, or smth like that
 On this count, at least,
-we hope the reader finds it self-evident that
-our equivalence proofs provide exponential state space reduction
+I hope the reader finds it self-evident that
+the equivalence proofs provide exponential state space reduction
 compared to actually testing (and thereafter aborting) every combination of instructions within transactions.
@@
 	\end{itemize}
-	\item Lock abstraction tests.
-		These are each parameterized over implementations {\tt spinlock} (from \cite{spinlock-rtm-github}),
-		{\tt spinlock\_fixed} (spoilers!!),
-		and {\tt mutex} (replaces the spinlock with a Landslide-annotated mutex to reduce state space size).
+	\item Lock abstraction tests
 	\begin{itemize}
@@
 	\end{itemize}
+		These are each parameterized over implementations {\tt spinlock} (from \cite{spinlock-rtm-github}),
+		{\tt spin\_\allowbreak{}fixed} (spoilers!!),
+		and {\tt mutex} (replaces the spinlock with a Landslide-annotated P2 mutex to reduce state space size).
 \end{itemize}
@@
 		\hline
+			\multicolumn{9}{c}{\em everything above this line needs rerun} \\ \hline
 		{\tt lock\_fast}
-			& 2,1	& \\
 		%({\tt spinlock})
+			& 2,1	&\cpu{10.48}&\wtm{3.22}&\ints{2}&\cpu{*2.98}&\wtm{*2.98}&\ints{2}& \ETA{8} \\
 		(perf)
-			& 9,9	& \\
+			& 9,9	&\cpu{37.41}&\wtm{9.96}&\ints{2}&\cpu{*4.93}&\wtm{*4.93}&\ints{2}& \ETA{inf} \\
 	\end{tabular}
@@
 	{\bf Spurious spinlock abort.}
-	% TODO
-	{\tt lock\_fast}({\tt spinlock})
+	{\tt lock\_fast} discovered a spurious transactional-path write conflict
+	in the {\tt spinlock} HTM-lock implementation.%
 	\footnote{{\tt lock\_fast}'s unusual (9,9) parameter shows that this state space size is constant: %in $(K,N)$:
@@
 	or the test's assertion will trip as soon as the first conflict is found.}
+	This ``performance bug'' causes the lock to suffer slow-path spin-locking
+	even in cases where the user's thread transitions are completely independent
+	(for example, locking the root of an AVL tree,
+	then traversing in different directions to make disjoint modifications),
+	which the test case detects with {\tt \_xtest()}.
+	Figure~\ref{fig:spinlockbug} shows the root cause:
+	the {\tt isfree()} routine (corresponding to the AVL's {\tt \_retry})
+	used an atomic compare-and-swap that would always write to memory even without modifying it.
+	I corrected this in the {\tt spin\_fixed} implementation by replacing it
+	with a normal read
+	(being used only in the transactional path, no barriers are required to protect it).
+	A cursory search on Github found one user of this code,
+	a transactional LevelDB implementation \cite{htm-leveldb-github},
+	whose author had also noticed and corrected this problem in the same way.
 
@@
 			& 3,1 & \cpu{57.90}	& \ints{1960}	\\
-			& 3,2 & \cpu{10028.93}	& \ints{329888}	\\ % TODO running atm
+			& 3,2 & \cpu{10028.93}	& \ints{329888}	\\
 			\hline
@@
 			% (3,1) without isfree fix
+			% in progress but smth like
+			% [JOB 4] progress: 39726/83689 brs (47.468314%), ETA 3h 27m 10s (elapsed 1h 30m 21s)
 			\hline
@@
 			{\tt lock\_fast}
-			& 2,1 & \cpu{0} & \ints{0}	\\
+			& 2,1 & \cpu{2.96} & \ints{1}	\\
 			({\tt spin\_fixed})
-			& 9,9 & \cpu{0} & \ints{0}	\\
+			& 9,9 & \cpu{4.33} & \ints{1}	\\
 			\hline
 			{\tt lock\_fast}
-			& 2,1 & \cpu{0} & \ints{0}	\\
+			& 2,1 & \cpu{3.05} & \ints{1}	\\
 			({\tt mutex})
-			& 9,9 & \cpu{0} & \ints{0}	\\
+			& 9,9 & \cpu{4.38} & \ints{1}	\\
 			%\end{tabular}

### 2018-07-15 thesis/tm.tex
synhi

@@
 		\begin{tabular}{l}
-		\texttt{bool hle\_spinlock\_isfree(spinlock\_t* lock) \{} \\
-		\texttt{~~~~// XXX: should be "return lock->v == 0;"} \\
-		\texttt{~~~~return \_\_sync\_bool\_compare\_and\_swap(\&lock->v, 0, 0);} \\
+		\texttt{\ctype{bool} \call{hle\_spinlock\_isfree}(\ctype{spinlock\_t} *lock) \{} \\
+		\texttt{~~~~\ccomment{// XXX: should be "return lock->v == 0;"}} \\
+		\texttt{~~~~\flow{return} \call{\_\_sync\_bool\_compare\_and\_swap}(\&lock->v, \const{0}, \const{0});} \\
 		\texttt{\}} \\
-		\texttt{void rtm\_spinlock\_acquire(spinlock\_t* lock) \{} \\
-		\texttt{~~~~if ((tm\_status = \_xbegin()) == \_XBEGIN\_STARTED) \{} \\
-		\texttt{~~~~~~~~if (hle\_spinlock\_isfree(lock)) return;} \\
-		\texttt{~~~~~~~~\_xabort(0xff);} \\
-		\texttt{~~~~\} else \{} \\
-			\texttt{~~~~~~~~// ... retrying \&c abbreviated for brevity ...} \\
-		\texttt{~~~~~~~~hle\_spinlock\_acquire(lock);} \\
+		\texttt{\ctype{void} \call{rtm\_spinlock\_acquire}(\ctype{spinlock\_t} *lock) \{} \\
+		\texttt{~~~~\flow{if} ((tm\_status = \call{\_xbegin}()) == \const{\_XBEGIN\_STARTED}) \{} \\
+		\texttt{~~~~~~~~\flow{if} (\call{hle\_spinlock\_isfree}(lock)) \flow{return};} \\
+		\texttt{~~~~~~~~\call{\_xabort}(\const{0xff});} \\
+		\texttt{~~~~\} \flow{else} \{} \\
+		\texttt{~~~~~~~~\ccomment{// ... retrying \&c abbreviated for brevity ...}} \\
+		\texttt{~~~~~~~~\call{hle\_spinlock\_acquire}(lock);} \\
 		\texttt{~~~~\}} \\
 		\texttt{\}} \\
-		\texttt{void rtm\_spinlock\_release(spinlock\_t* lock)\{} \\
-		\texttt{~~~~if (hle\_spinlock\_isfree(lock)) \{} \\
-		\texttt{~~~~~~~~\_xend();} \\
-		\texttt{~~~~\} else \{} \\
-		\texttt{~~~~~~~~hle\_spinlock\_release(lock);} \\
+		\texttt{\ctype{void} \call{rtm\_spinlock\_release}(\ctype{spinlock\_t} *lock)\{} \\
+		\texttt{~~~~\flow{if} (\call{hle\_spinlock\_isfree}(lock)) \{} \\
+		\texttt{~~~~~~~~\call{\_xend}();} \\
+		\texttt{~~~~\} \flow{else} \{} \\
+		\texttt{~~~~~~~~\call{hle\_spinlock\_release}(lock);} \\
 		\texttt{~~~~\}} \\
@@
 		\begin{tabular}{l}
-			\texttt{static int num\_in\_section = 0;} \\
-			\texttt{for (int i = 0; i < NITERS; i++) \{} \\
-			\texttt{~~~~rtm\_spinlock\_acquire(\&lock);} \\
+			\texttt{\ctype{static int} num\_in\_section = 0;} \\
+			\texttt{\flow{for} (\ctype{int} i = \const{0}; i < \const{NITERS}; i++) \{} \\
+			\texttt{~~~~\call{rtm\_spinlock\_acquire}(\&lock);} \\
 			\texttt{~~~~num\_in\_section++;} \\
-			\texttt{~~~~if (!\_xtest())} \\
-			\texttt{~~~~~~~~thr\_yield(-1);} \\
-			\texttt{~~~~assert(num\_in\_section == 1); } \\
+			\texttt{~~~~\flow{if} (!\call{\_xtest()})} \\
+			\texttt{~~~~~~~~\call{thr\_yield}(\const{-1});} \\
+			\texttt{~~~~\flow{assert}(num\_in\_section == \const{1}); } \\
 			\texttt{~~~~num\_in\_section--;} \\
-			\texttt{~~~~rtm\_spinlock\_release(\&lock);} \\
+			\texttt{~~~~\call{rtm\_spinlock\_release}(\&lock);} \\
 			\texttt{\}} \\
@@
 		\begin{tabular}{ll}
-			\texttt{for (int i = 0; i < NITERS; i++) \{} \\
-			\texttt{~~~~rtm\_spinlock\_acquire(\&lock);} \\
-			\texttt{~~~~assert(\_xtest());} \\
-			\texttt{~~~~rtm\_spinlock\_release(\&lock);} \\
+			\texttt{\flow{for} (\ctype{int} i = \const{0}; i < \const{NITERS}; i++) \{} \\
+			\texttt{~~~~\call{rtm\_spinlock\_acquire}(\&lock);} \\
+			\texttt{~~~~\flow{assert}(\_xtest());} \\
+			\texttt{~~~~\call{rtm\_spinlock\_release}(\&lock);} \\
 			\texttt{\}} \\
@@
 the benefit is clear:
-for example,
+the data structure tests suddenly reach unprecedented thread/iteration counts with ease,
+their state space growth now defined only by their internal conflicts from tree rebalancing, map resizing, and so on.
+In total,
 summing the testing times of {\tt lock}({\tt mutex})$(K,N)$ and {\tt avl\_mutex}$(K,N)$

### 2018-07-17 thesis/landslide.tex
do more tm experimence

@@
 \subsubsection{Misbehave}
+\label{sec:landslide-friendly-misbehave}
 
@@
 		{\tt USERSPACE} should be 0 or 1 to denote a kernel-space or user-space function respectively.
+	\item {\tt thrlib\_function NAME} - specifies a function whose memory accesses should be ignored both by the data race analysis and by DPOR.
+		This is recommended for marking trusted-correct thread library code when testing multithreaded client code thereof,
+		in order to avoid unnecessarily checking, for example,
+		all the different ways {\tt thr\_exit()} and {\tt thr\_join()} could interleave.
+		Users of this option should be careful to enable the proper {\tt thr\_create()} misbehave mode in their test case (\sect{\ref{sec:landslide-friendly-misbehave}}).
 \end{itemize}
@@
 \subsubsection{State space exploration}
+\label{sec:landslide-explore}
 
@@
 to reach a (possibly) new program state not exposed in the current interleaving.
-Prior work \cite{dpor,optimal-dpor} refers to the set of these $I_{ij}$s,
+Prior work \cite{partial-order-methods,dpor,optimal-dpor} refers to the set of these $I_{ij}$s,
 for a given $i$,
@@
 Landslide implements this check in {\tt equiv\_already\_explored()}.
-This corresponds to the {\em sleep sets} optimization described by prior work \cite{dpor,optimal-dpor},
+This corresponds to the {\em sleep sets} optimization described by prior work \cite{partial-order-methods,dpor,optimal-dpor},
 so named because it effectively puts $T_1$ ``to sleep'' until after the true conflicting access in $t_c$.

### 2018-07-17 thesis/tm.tex
do more tm experimence

@@
 
+% In contrast to Landslide's implementation of the sleep set reduction (\sect{\ref{sec:landslide-explore}}),
+% blah blah which is implicit and works in retrospect by pruning at tag-sibling time, which figures out what would have been in a sleep set by analyzing the past independences and explored siblings,
+% this cannot work that way,
+% because in the first place, the former works with the "keep running current thread" ordering heuristic,
+% (which automatically takes care of the first half of the sleep set problem)
+% whereas no such similar heuristic exists (can exist??) for xbegin PPs, we always just run the success path first
+
 % TODO
+
+% ok so heres an example of why this is rly hard
+
+% ==== thread 1 ====
+% if (xbegin()) {
+% 	xabort(0) // whatever, executed in a past branch to be aborting in the present
+% } else {
+% 	x++;
+% }
+%
+% ==== thread 3 ====
+% conflict_with_thread_2 = true;
+%
+% ==== thread 2 ====
+% if (xbegin()) {
+% 	conflict_with_thread_2 ++ ; // whatever, just conflict and fail w thr 3
+% } else {
+% 	x++; // conf with thr 1
+% }
+%
+% so we execute this interleaving (T1 abort, T3, T2 abort)
+% dpor sees the conflict between T2 and T1 and wants to tag T2 to run as transition(T1-abort)'s sibling to run
+% it would use the "abort set" of (T2 [CONFLICT], T1 [EXPLICIT])
+% but like, how do we know that thread 3 is necessary to reach this?
+% and even, how do we know that it's valid to inject CONFLCIT in T2?
+% you pretty much hafta reorder T3 (dpor doesn't see why you need to)
+% but if you inject conflict in T2 without also making sure T3 executes "near" it, that might not necessarily be sound
 
@@
 
+\subsubsection{Experimental setup}
+
 The evaluation suite comprises several hand-written unit tests,
@@
 		\item {\tt htm2}: The fixed version as in Figure~\ref{fig:htm-fixed}.
-		\item {\tt counter}: Microbenchmark version of {\tt htm2} which replaces the complex failure path with a simple {\tt xadd}.
-		\item {\tt swap}: Microbenchmark that swaps values in an array.
+		\item {\tt counter}: Microbenchmark version of {\tt htm2} which replaces the complex locking failure path with an atomic {\tt xadd}, from \cite{tm-benchmark-cmu}.
+		\item {\tt swap}: Microbenchmark that swaps values in an array, from \cite{tm-benchmark-cmu}.
 		\item {\tt swapbug}: {\tt swap} modified to introduce circular locking in the failure path. %(deadlock).
@@
 
+Finally, the keen-eyed reader will notice the state space sizes reported here
+differ from those reported in \cite{sigbovik-htm}.
+All experiments have been re-run
+on account of three updates to Landslide's exploration algorithm
+implemented since then:
+the sleep sets optimization for DPOR (\sect{\ref{sec:landslide-explore}}, commit 0447666),
+the {\tt thrlib\_function} directive
+to mark internal thread library logic as trusted (\sect{\ref{sec:landslide-config-landslide}}, commit 64a02e4),
+and fixing a soundness bug in which Landslide could neglect to inject transaction failures
+immediately after a thread switch (commit dcae85b).
+On account of the former two updates, some state spaces may be smaller than before;
+on account of the third, some may be larger.
+These updates do not discredit the bug-finding results (a bug is a bug),
+but the previously-published verification results should be considered outdated.
+
 \newcommand\ETA[1]{\hilight{brownish}{{\em #1}}\xspace}
@@
 I configured Landslide to run Quicksand's Iterative Deepening algorithm, %(\sect{\ref{sec:quicksand-id}}),
+% TODO how many cores?
 shown left,
@@
 should they exist,
-Landslide find bugs reasonably quickly in these transactional programs
+Landslide is likely to find bugs reasonably quickly in these transactional programs
 despite prohibitive exponential explosion in total state space size.
@@
 \end{itemize}
-Note that in the AVL tree bug,
-the code's author was the very same person who proposed the protocol in Figure~\ref{fig:htm-fixed},
-yet still got it wrong once, having to write it out by hand throughout both data structures.
-This motivates the need for model checking such programs,
-no matter how much of a concurrency expert the author may be.
-It also suggests HTM primitives should be encapsulated behind higher-level abstractions,
-such as lock elision \cite{lock-elision}
-or a simple spinlock \cite{spinlock-rtm-github},
-which can be verified in isolation with smaller state spaces
-then trusted in turn when checking their client programs \cite{dbug-phdthesis}.
-\sect{\ref{sec:tm-verif}} explores this further.
-
-As HTM is fundamentally a performance-minded concurrency extension,
-users are likely to care about more probabilistic properties of their code,
-such as requiring a transaction abort rate below a certain threshhold
-owing to the nature of its workload.
-Landslide cannot in general test for performance degradation bugs,
-because all interleavings are equal in Landslide's eyes, and probability is no object.
-However, {\tt lock\_fast} illustrates that model checking can still check some interesting performance properties
-as long as the element of probability can be removed.
-Future work may attempt to verify a wider range of performance properties,
-but with a hybrid approach between model checking and what other technique is not yet known.
 
@@
 	\begin{center}
-		% TODO synhi
 		\begin{tabular}{l}
@@
 
+Note that in the AVL tree bug,
+the code's author was the very same person who proposed the protocol in Figure~\ref{fig:htm-fixed},
+yet still got it wrong once, having to write it out by hand throughout both data structures.
+This motivates the need for model checking such programs,
+no matter how much of a concurrency expert the author may be.
+It also suggests HTM primitives should be encapsulated behind higher-level abstractions,
+such as lock elision \cite{lock-elision}
+or a simple spinlock \cite{spinlock-rtm-github},
+which can be verified in isolation with smaller state spaces
+then trusted in turn when checking their client programs \cite{dbug-phdthesis}.
+\sect{\ref{sec:tm-verif}} explores this further.
+
+Regarding the spinlock bug,
+as HTM is fundamentally a performance-minded concurrency extension,
+users are likely to care about more probabilistic properties of their code,
+such as requiring a transaction abort rate below a certain threshhold
+owing to the nature of its workload.
+Landslide cannot in general test for performance degradation bugs,
+because all interleavings are equal in Landslide's eyes, and probability is no object.
+However, {\tt lock\_fast} illustrates that model checking can still check some interesting performance properties
+as long as the element of probability can be removed.
+Future work may attempt to verify a wider range of performance properties,
+but with a hybrid approach between model checking and what other technique is not yet known.
+
 \subsection{Performance}
@@
 			{\tt htm2}
-			& 2,1 & \cpu{18.57}	& \ints{294}	\\
-			& 2,2 & \cpu{133.78}	& \ints{4902}	\\
-			& 2,3 &\cpu{1986.98}	& \ints{79017}	\\
-			& 3,1 &\cpu{11672.15}	& \ints{467730}	\\
-			& 3,2 &	\ETAdag{10d 14h}& \ETAdag{13763999}	\\
-			%[JOB 4] progress: 127032/13763999 brs (0.922929%), ETA 10d 13h 52m 44s (elapsed 59m 56s)
-			%[JOB 4] TIMED OUT (0.922929%; ETA 10d 13h 52m 44s)
+			& 2,1 & \cpu{16.30}	& \ints{22}	\\
+			& 2,2 & \cpu{69.80}	& \ints{1970}	\\
+			& 2,3 &\cpu{3510.00}	& \ints{104914}	\\
+			%& 2,4 &	\ETAdag{...}	& \ETAdag{...}	\\
+			& 3,1 &\cpu{41.07}	& \ints{941}	\\
+			& 3,2 &	\ETAdag{todo}	& \ETAdag{todo}	\\
+			& 4,1 & \cpu{12699.65}	& \ints{344240}	\\
 			\hline
 			{\tt counter}
-			& 2,1 & \cpu{5.57}	& \ints{30}	\\
-			& 2,2 & \cpu{15.53}	& \ints{384}	\\
-			& 2,3 &\cpu{155.00}	& \ints{5280}	\\
-			& 2,4 &\cpu{2211.10}	& \ints{75264}	\\
-			& 3,1 & \cpu{57.90}	& \ints{1960}	\\
-			& 3,2 & \cpu{10028.93}	& \ints{329888}	\\
+			& 2,1 & \cpu{5.40}	& \ints{10}	\\
+			& 2,2 & \cpu{10.91}	& \ints{190}	\\
+			& 2,3 & \cpu{134.21}	& \ints{3970}	\\
+			& 2,4 & \cpu{3009.40}	& \ints{86950}	\\
+			& 3,1 & \cpu{8.30}	& \ints{120}	\\
+			& 3,2 & \cpu{2038.90}	& \ints{60606}	\\
+			& 4,1 & \cpu{101.95}	& \ints{3006}	\\
 			\hline
 			{\tt swap}
-			%& 2,1 & \cpu{38.77}	& \ints{228}	\\ % pre-ignore_thr_func fix
-			& 2,1 & \cpu{32.98}	& \ints{193}	\\
-			& 2,2 & \cpu{3652.91}	& \ints{101150}	\\
-			%[JOB 8] progress: 97913/161897 brs (60.478362%), ETA 1h 29m 30s (elapsed 59m 52s)
-			% [JOB 8] TIMED OUT (60.478362%; ETA 1h 29m 30s)
-
-			& 3,1 & \ETAdag{8d 12h} & \ETAdag{411312}	\\
-			%[JOB 8] progress: 103873/411312 brs (25.254021%), ETA 8d 11h 36m 0s (elapsed 59m 42s)
-			%[JOB 8] TIMED OUT (25.254021%; ETA 8d 11h 36m 0s)
+			& 2,1 & \cpu{38.79}	& \ints{99}	\\
+			& 2,2 & \cpu{10299.27}	& \ints{233396}	\\
+			& 3,1 & \ETAdag{todo} & \ETAdag{todo}	\\
 			\hline
 			{\tt avl\_insert}
-			& 2,1 & \cpu{1083.35}	& \ints{40062}	\\
-			%& 2,2 & \multicolumn{2}{c}{\em (bug!)} \\
+			& 2,1 & \cpu{491.98}	& \ints{15056}	\\
 			\hline
 			{\tt avl\_fixed}
-			& 2,1 & \cpu{1079.03} 	& \ints{45078} 	\\
-			& 2,2 & \ETAdag{2762y}	& \ETAdag{8714863} 	\\
-			%[JOB 19] progress: 45863/8714863 brs (0.526262%), ETA 2762y 28d 9h 57m 14s (elapsed 59m 25s)
-			%[JOB 19] TIMED OUT (0.526262%; ETA 2762y 28d 9h 57m 14s)
-			% the below is from avlfixed2,2 after running it overnight
-			% eta is ever worse but i'm using the 1hr one as promised
-			%[JOB 19] progress: 315672/29992026 brs (1.052520%), ETA 321008y 136d 18h 22m 12s (elapsed 6h 58m 15s)
-			& 3,1 & \ETAdag{12d 2h}	& \ETAdag{11498545}	\\
+			& 2,1 & \cpu{572.49} 	& \ints{20420} 	\\
+			& 2,2 & \ETAdag{todo}	& \ETAdag{todo} 	\\
+			& 3,1 & \ETAdag{todo}	& \ETAdag{todo}	\\
 
@@
 			{\tt map\_basic}
-			& 2,1 & \ETAdag{10d 17h} & \ETAdag{16388977} \\
-			%[JOB 15] progress: 70330/16388977 brs (0.429130%), ETA 10d 16h 34m 15s (elapsed 43m 21s)
-			%[JOB 15] TIMED OUT (0.429130%; ETA 10d 16h 34m 15s)
+			& 2,1 & \ETAdag{todo} & \ETAdag{todo} \\
 			\hline
 			{\tt map\_basicer}
-			& 2,1 & \cpu{877.44}& \ints{28635}	\\
-			& 2,2 & \ETAdag{2d 7h} & \ETAdag{5925634} \\
-			& 3,1 & \ETAdag{468d 13h} & \ETAdag{35893653} \\
-			% [JOB 5] progress: 100181/35893653 brs (0.279105%), ETA 1y 103d 12h 50m 47s (elapsed 59m 24s)
-			%[JOB 5] TIMED OUT (0.279105%; ETA 1y 103d 12h 50m 47s)
-			% i don't know what this is but it was hanging out under 3,1 here :<
-			%[JOB 6] progress: 35042/5925634 brs (0.591363%), ETA 2d 6h 40m 30s (elapsed 21m 12s)
-			%[JOB 6] TIMED OUT (0.591363%; ETA 2d 6h 40m 30s)
-
-			\hline
-			\multicolumn{4}{c}{\em everything above this line needs rerun} \\
+			& 2,1 & \cpu{36.17}& \ints{426}	\\
+			& 2,2 & \ETAdag{todo} & \ETAdag{todo} \\ % currently running
+			& 3,1 & \ETAdag{todo} & \ETAdag{todo} \\
 		\end{tabular}
@@
 		State space estimates measured after a timeout of 10 hours.
-		{\bf THESIS DRAFT NOTE:} These numbers all need to be re-run on account of some state space reductions and/or soundness fixes that have been added since these numbers were calculated in \cite{sigbovik-htm}.
 		}
@@
 			\texttt{~~~~~~~~\call{thr\_yield}(\const{-1});} \\
-			\texttt{~~~~\flow{assert}(num\_in\_section == \const{1}); } \\
+			\texttt{~~~~\flow{assert}{}(num\_in\_section == \const{1}); } \\
 			\texttt{~~~~num\_in\_section--;} \\
@@
 			\texttt{~~~~\call{rtm\_spinlock\_acquire}(\&lock);} \\
-			\texttt{~~~~\flow{assert}(\_xtest());} \\
+			\texttt{~~~~\flow{assert}{}(\call{\_xtest}());} \\
 			\texttt{~~~~\call{rtm\_spinlock\_release}(\&lock);} \\
@@
 			& 3,1 & \\ % \cpu{0} & \ints{0} \\ % TODO currently running, i think,
+			& 3,2 & \\ % \cpu{0} & \ints{0} \\
 			% new spinlock test resulce - htm_spinlock
@@
 			& 2,2 & \cpu{1662.21} & \ints{15076} \\
-			& 2,3 & \\ %\cpu{0} & \ints{0} \\
+			& 2,3 & \ETAdag{3d 20h} & \ETAdag{2003906} \\
 			& 3,1 & \cpu{2488.95} & \ints{24970} \\
+			& 3,2 & \\ % \cpu{0} & \ints{0} \\
 			% (2,1) with isfree fix
@@
 			% (2,3)
-			%% this will be an overnight 10hr job
-			% [DBG] [JOB 4] progress: 30353/444731 brs (6.825013%),
-			% [JOB 4] Running (6.825013%; ETA 5d 9h 4m 41s)
+			% [JOB 4] progress: 242266/2003906 brs (12.089688%), ETA 3d 20h 2m 12s (elapsed 9h 59m 49s)
+			% [JOB 4] TIMED OUT (12.089688%; ETA 3d 20h 2m 12s)
 			% (3,1) with isfree fix

### 2018-07-18 thesis/tm.tex
rerun all the maximals in the tm bugs found table

@@
 All tests were run on an 8-core 2.7GHz Core i7 with 32 GB RAM.
+Reported CPU-times include time spent on all state spaces
+Quicksand saw fit to run,
+not just the maximal or the buggy state space;
+for verification tests (run with {\tt -M}),
+this still includes abandoned smaller jobs that were run to saturate the set of data-race preemption points
+(Chapter~\ref{chap:quicksand}).
 To minimize variance in CPU-time measurements,
@@
 		{\tt htm1}
-			& 2,1	&\cpu{  45.78}&\wtm{9.70}&\ints{21}& \cpu{*9.47}& \wtm{*6.40}& \ints{21	}& \ETA{213} \\
+			& 2,1	&\cpu{  45.78}&\wtm{9.70}&\ints{21}& \cpu{*8.21}& \wtm{*5.78}& \ints{5	}& \ETA{12} \\
+			% 5 interleavings tested; 1 preemptions; job time 2s; pldi time 5781000; new-fixed pldi cputime 8214902
 		(assertion)
-			& 2,2	&\cpu{  84.14}&\wtm{13.59}&\ints{*33}& \cpu{*10.39}& \wtm{*7.70}& \ints{49	}& \ETA{1536} \\
-			& 2,3	&\cpu{ 131.91}&\wtm{20.44}&\ints{*73}& \cpu{*12.83}& \wtm{*9.67}& \ints{113	}& \ETA{10752} \\
-			& 2,4	&\cpu{ 255.75}&\wtm{37.56}&\ints{257}& \cpu{*18.63}& \wtm{*15.86}& \ints{257	}& \ETA{73728} \\
+			& 2,2	&\cpu{  84.14}&\wtm{13.59}&\ints{*33}& \cpu{*8.42}& \wtm{*5.97}& \ints{9	}& \ETA{102} \\
+			% 9 interleavings tested; 1 preemptions; job time 2s; pldi time 5972944; new-fixed pldi cputime 8418505
+			& 2,3	&\cpu{ 131.91}&\wtm{20.44}&\ints{*73}& \cpu{*8.80}& \wtm{*6.31}& \ints{17	}& \ETA{819} \\
+			% 17 interleavings tested; 1 preemptions; job time 2s; pldi time 6314386; new-fixed pldi cputime 8795639
+			& 2,4	&\cpu{ 255.75}&\wtm{37.56}&\ints{257}& \cpu{*9.72}& \wtm{*7.19}& \ints{33	}& \ETA{6553} \\
+			% 33 interleavings tested; 1 preemptions; job time 3s; pldi time 7193093; new-fixed pldi cputime 9717902
 		\cline{2-9}
-			& 3,1	&\cpu{ 114.06}&\wtm{17.45}&\ints{*15}& \cpu{*9.50}& \wtm{*6.79}& \ints{21	}& \ETA{13653} \\
-			& 3,2	&\cpu{ 109.60}&\wtm{26.16}&\ints{49}& \cpu{*10.72}& \wtm{*7.97}& \ints{49	}& \ETA{393216} \\
-			& 3,3	&\cpu{ 124.80}&\wtm{20.40}&\ints{*73}& \cpu{*13.84}& \wtm{*11.01}& \ints{113	}& \ETA{11010048} \\
-			& 3,4	&\cpu{ 227.49}&\wtm{35.15}&\ints{*161}& \cpu{*31.37}& \wtm{*28.53}& \ints{257	}& \ETA{301989888} \\
+			& 3,1	&\cpu{ 114.06}&\wtm{17.45}&\ints{*15}& \cpu{*8.15}& \wtm{*5.71}& \ints{5	}& \ETA{76} \\
+			% 5 interleavings tested; 1 preemptions; job time 2s; pldi time 5710637; new-fixed pldi cputime 8153919
+			& 3,2	&\cpu{ 109.60}&\wtm{26.16}&\ints{49}& \cpu{*8.37}& \wtm{*5.91}& \ints{9	}& \ETA{3686} \\
+			% 9 interleavings tested; 1 preemptions; job time 2s; pldi time 5907009; new-fixed pldi cputime 8367804
+			& 3,3	&\cpu{ 124.80}&\wtm{20.40}&\ints{*73}& \cpu{*8.71}& \wtm{*6.26}& \ints{17	}& \ETA{176947} \\
+			% 17 interleavings tested; 1 preemptions; job time 2s; pldi time 6263778; new-fixed pldi cputime 8711332
+			& 3,4	&\cpu{ 227.49}&\wtm{35.15}&\ints{*161}& \cpu{*10.36}& \wtm{*7.15}& \ints{33	}& \ETA{8493465} \\
+			% 33 interleavings tested; 1 preemptions; job time 3s; pldi time 7153968; new-fixed pldi cputime 10355961
 		\cline{2-9}
-			& 4,1	&\cpu{ 53.08}&\wtm{9.79}&\ints{*15}& \cpu{*9.82}& \wtm{*7.00}& \ints{21	}& \ETA{873813} \\
-			& 4,2	&\cpu{117.07}&\wtm{19.09}&\ints{*33}& \cpu{*11.54}& \wtm{*8.55}& \ints{49	}& \ETA{100663296} \\
+			& 4,1	&\cpu{ 53.08}&\wtm{9.79}&\ints{*15}& \cpu{*7.77}& \wtm{*5.30}& \ints{5	}& \ETA{460} \\
+			% 5 interleavings tested; 1 preemptions; job time 2s; pldi time 5295360; new-fixed pldi cputime 7770389
+			& 4,2	&\cpu{117.07}&\wtm{19.09}&\ints{*33}& \cpu{*8.32}& \wtm{*5.59}& \ints{9	}& \ETA{132710} \\
+			% 132710 9 interleavings tested; 1 preemptions; job time 2s; pldi time 5588002; new-fixed pldi cputime 8317173
 		\hline
 		{\tt swapbug}
-			& 2,1	&\cpu{ 70.95}&\wtm{13.45}&\ints{*16}& \cpu{*38.96}& \wtm{*13.15}& \ints{109	}& \ETA{194} \\
+			& 2,1	&\cpu{ 70.95}&\wtm{13.45}&\ints{*16}& \cpu{*28.32}& \wtm{*11.29}& \ints{33	}& \ETA{73} \\
+			% 33 interleavings tested; 1 preemptions; job time 3s; pldi time 11286074; new-fixed pldi cputime 28324164
 		(deadlock)
-			& 2,2	&\cpu{ 107.28}&\wtm{*17.45}&\ints{*146}& \cpu{*44.73}& \wtm{19.47}& \ints{281	}& \ETA{1620} \\
-			& 2,3	&\cpu{ 280.05}&\wtm{38.70}&\ints{*352}& \cpu{*60.30}& \wtm{*35.55}& \ints{718	}& \ETA{12748} \\
-			& 2,4	&\cpu{ 617.94}&\wtm{*81.50}&\ints{*834}& \cpu{*108.58}& \wtm{82.60}& \ints{1820	}& \ETA{97823} \\
+			& 2,2	&\cpu{ 107.28}&\wtm{*17.45}&\ints{*146}& \cpu{*30.55}& \wtm{13.50}& \ints{85	}& \ETA{860} \\
+			% 85 interleavings tested; 1 preemptions; job time 5s; pldi time 13499606; new-fixed pldi cputime 30550037
+			& 2,3	&\cpu{ 280.05}&\wtm{38.70}&\ints{*352}& \cpu{*32.27}& \wtm{*19.42}& \ints{217	}& \ETA{9120} \\
+			% 217 interleavings tested; 1 preemptions; job time 11s; pldi time 19419243; new-fixed pldi cputime 37272562
+			& 2,4	&\cpu{ 617.94}&\wtm{*81.50}&\ints{*834}& \cpu{*53.44}& \wtm{35.80}& \ints{537	}& \ETA{91239} \\
+			% 537 interleavings tested; 1 preemptions; job time 27s; pldi time 35796053; new-fixed pldi cputime 53441873
 		\cline{2-9}
-			& 3,1	&\cpu{*1275.04}&\wtm{*163.42}&\ints{*771}&\ETA{--}&\ETA{>30m}&\ETA{--}& \ETA{184984} \\
-			& 3,2	&\ETA{--}&\ETA{>30m}&\ETA{--}&\ETA{--}&\ETA{>30m}&\ETA{--}& \ETA{3099225} \\
+			& 3,1	&\cpu{*1275.04}&\wtm{*163.42}&\ints{*771}&\cpu{*65.13}& \wtm{52.23}& \ints{980   }& \ETA{3417} \\
+			% 980 interleavings tested; 2 preemptions; job time 40s; pldi time 52225013; new-fixed pldi cputime 65125824
+			& 3,2	&\ETA{--}&\ETA{>30m}&\ETA{--}&\ETA{--}&\ETA{>1h}&\ETA{--}& \ETA{2256553} \\
+			% [JOB 8] progress: 63137/2256553 brs (2.797940%), ETA 130d 19h 28m 29s (elapsed 59m 45s)
 		\hline
 		{\tt avl\_insert}
-			& 2,2	&\cpu{488.07}&\wtm{64.77}&\ints{*83}	&\cpu{*81.00}&\wtm{*40.30}&\ints{336}&\ETA{379982} \\
+			& 2,2	&\cpu{488.07}&\wtm{64.77}&\ints{*83}	&\cpu{*66.32}&\wtm{*31.81}&\ints{127}&\ETA{254078} \\
 		(segfault+)
-			& 2,3	&\cpu{2670.87}&\wtm{*330.45}&\ints{*3066}&\cpu{*1331.79}&\wtm{1274.36}&\ints{13926}&\ETA{96248131} \\
-			& 2,4	&\cpu{*3259.37}&\wtm{*436.50}&\ints{*1639}&\ETA{--}&\ETA{>30m}&\ETA{--}& \ETA{36019973} \\
+			& 2,3	&\cpu{2670.87}&\wtm{*330.45}&\ints{*3066}&\cpu{*256.94}&\wtm{193.71}&\ints{1391}&\ETA{22773586} \\
+			& 2,4	&\cpu{*3259.37}&\wtm{*436.50}&\ints{*1639}&\ETA{--}&\ETA{>1h}&\ETA{--}& \ETA{91704203} \\
+			% [JOB 12] progress: 28140/91704203 brs (0.030686%), ETA 18y 54d 18h 42m 58s (elapsed 51m 12s)
 		\cline{2-9}
-			& 3,1	&\cpu{222.02}&\wtm{40.04}&\ints{*28}	&\cpu{*69.99}&\wtm{*24.25}&\ints{78}&\ETA{1572107} \\
-			& 3,2	&\cpu{*1569.09}&\wtm{*216.85}&\ints{*209}&\ETA{--}&\ETA{>30m}&\ETA{--}& \ETA{1402363529} \\
-		\hline
-			\multicolumn{9}{c}{\em everything above this line needs rerun} \\ \hline
+			& 3,1	&\cpu{222.02}&\wtm{40.04}&\ints{*28}	&\cpu{*64.70}&\wtm{*25.59}&\ints{53}&\ETA{343250} \\
+			% 53 interleavings tested; 3 preemptions; job time 6s; pldi time 25585758; new-fixed pldi cputime 64695226
+			& 3,2	&\cpu{*1569.09}&\wtm{*216.85}&\ints{*209}&\ETA{--}&\ETA{>1h}&\ETA{--}& \ETA{466977272} \\
+			%[JOB 38] progress: 3812/466977272 brs (0.000816%), ETA INF (elapsed 13m 50s)
+		\hline
+			\multicolumn{5}{c}{\em every quicksand above this line needs rerun} \\ \hline
 		{\tt lock\_fast}
@@
 each with a time limit of 30 minutes.
+Tests {\tt htm1}, {\tt swapbug}, and {\tt avl\_insert} were run with {\tt landslide -X}
+(i.e., retry aborts enabled and different abort codes not distinguished);
+{\tt lock\_fast} was run with {\tt landslide -X -A -S}
+(i.e., suppressing retry aborts, due to the spinlock's use of a retry loop confirmed with manual inspection).
 
@@
 			& 2,2 & \cpu{4944.37} & \ints{38544} \\
-			& 2,3 & \\ % \cpu{0} & \ints{0} \\
-			& 3,1 & \\ % \cpu{0} & \ints{0} \\ % TODO currently running, i think,
-			& 3,2 & \\ % \cpu{0} & \ints{0} \\
+			% if lock(spin_fixed) will time out on 2,3 and 3,2, this is guaranteed to aswell
+			% & 2,3 & \\ % \cpu{0} & \ints{0} \\
+			& 3,1 & \cpu{13649.32} & \ints{97462} \\
+			% & 3,2 & \\ % \cpu{0} & \ints{0} \\
 			% new spinlock test resulce - htm_spinlock
@@
 			& 3,1 & \cpu{2488.95} & \ints{24970} \\
-			& 3,2 & \\ % \cpu{0} & \ints{0} \\
+			& 3,2 & \ETAdag{207d 20h} & \ETAdag{6514618} \\
+			& 4,1 & \\ % TODO
 			% (2,1) with isfree fix
@@
 			%%%% total CPU time consumed: 41m 28s (2488953483 usecs) (core saturation: 25%)
+			% (3,2)
+			% [JOB 4] progress: 187522/6514618 brs (2.878480%), ETA 207d 20h 23m 10s (elapsed 9h 59m 51s)
+			% [JOB 4] TIMED OUT (2.878480%; ETA 207d 20h 23m 10s)
 			\hline

### 2018-07-18 thesis/tm.tex
rerun quicksand tm bugs

@@
 I ensured the test machine was not loaded beyond normal web browser use,
-and ran only one instance of Landslide at a time
-\cite{sigbovik-htm}.
+and ran only one instance of Landslide at a time;
+for further discussion of variance see \cite{sigbovik-htm}.
 
@@
 		{\tt htm1}
-			& 2,1	&\cpu{  45.78}&\wtm{9.70}&\ints{21}& \cpu{*8.21}& \wtm{*5.78}& \ints{5	}& \ETA{12} \\
+			& 2,1	&\cpu{  26.48}&\wtm{6.48}&\ints{5}& \cpu{*8.21}& \wtm{*5.78}& \ints{5	}& \ETA{12} \\
+			% 5 interleavings tested; 1 preemptions; job time 3s; pldi time 6482137; new-fixed pldi cputime 26480574
 			% 5 interleavings tested; 1 preemptions; job time 2s; pldi time 5781000; new-fixed pldi cputime 8214902
 		(assertion)
-			& 2,2	&\cpu{  84.14}&\wtm{13.59}&\ints{*33}& \cpu{*8.42}& \wtm{*5.97}& \ints{9	}& \ETA{102} \\
+			& 2,2	&\cpu{  45.23}&\wtm{8.28}&\ints{9}& \cpu{*8.42}& \wtm{*5.97}& \ints{9	}& \ETA{102} \\
+			% 9 interleavings tested; 1 preemptions; job time 4s; pldi time 8281422; new-fixed pldi cputime 45289375
 			% 9 interleavings tested; 1 preemptions; job time 2s; pldi time 5972944; new-fixed pldi cputime 8418505
-			& 2,3	&\cpu{ 131.91}&\wtm{20.44}&\ints{*73}& \cpu{*8.80}& \wtm{*6.31}& \ints{17	}& \ETA{819} \\
+			& 2,3	&\cpu{ 53.88}&\wtm{9.92}&\ints{17}& \cpu{*8.80}& \wtm{*6.31}& \ints{17	}& \ETA{819} \\
+			% 17 interleavings tested; 1 preemptions; job time 5s; pldi time 9921305; new-fixed pldi cputime 53884499
 			% 17 interleavings tested; 1 preemptions; job time 2s; pldi time 6314386; new-fixed pldi cputime 8795639
-			& 2,4	&\cpu{ 255.75}&\wtm{37.56}&\ints{257}& \cpu{*9.72}& \wtm{*7.19}& \ints{33	}& \ETA{6553} \\
+			& 2,4	&\cpu{ 90.24}&\wtm{14.63}&\ints{33}& \cpu{*9.72}& \wtm{*7.19}& \ints{33	}& \ETA{6553} \\
+			% 33 interleavings tested; 1 preemptions; job time 7s; pldi time 14629409; new-fixed pldi cputime 90244293
 			% 33 interleavings tested; 1 preemptions; job time 3s; pldi time 7193093; new-fixed pldi cputime 9717902
 		\cline{2-9}
-			& 3,1	&\cpu{ 114.06}&\wtm{17.45}&\ints{*15}& \cpu{*8.15}& \wtm{*5.71}& \ints{5	}& \ETA{76} \\
+			& 3,1	&\cpu{ 39.75}&\wtm{8.21}&\ints{5}& \cpu{*8.15}& \wtm{*5.71}& \ints{5	}& \ETA{76} \\
+			% 5 interleavings tested; 1 preemptions; job time 4s; pldi time 8213084; new-fixed pldi cputime 39752062
 			% 5 interleavings tested; 1 preemptions; job time 2s; pldi time 5710637; new-fixed pldi cputime 8153919
-			& 3,2	&\cpu{ 109.60}&\wtm{26.16}&\ints{49}& \cpu{*8.37}& \wtm{*5.91}& \ints{9	}& \ETA{3686} \\
+			& 3,2	&\cpu{ 46.91}&\wtm{8.40}&\ints{9}& \cpu{*8.37}& \wtm{*5.91}& \ints{9	}& \ETA{3686} \\
+			% 9 interleavings tested; 1 preemptions; job time 4s; pldi time 8397105; new-fixed pldi cputime 46909424
 			% 9 interleavings tested; 1 preemptions; job time 2s; pldi time 5907009; new-fixed pldi cputime 8367804
-			& 3,3	&\cpu{ 124.80}&\wtm{20.40}&\ints{*73}& \cpu{*8.71}& \wtm{*6.26}& \ints{17	}& \ETA{176947} \\
+			& 3,3	&\cpu{ 41.39}&\wtm{8.50}&\ints{17}& \cpu{*8.71}& \wtm{*6.26}& \ints{17	}& \ETA{176947} \\
+			% 17 interleavings tested; 1 preemptions; job time 4s; pldi time 8503682; new-fixed pldi cputime 41386987
 			% 17 interleavings tested; 1 preemptions; job time 2s; pldi time 6263778; new-fixed pldi cputime 8711332
-			& 3,4	&\cpu{ 227.49}&\wtm{35.15}&\ints{*161}& \cpu{*10.36}& \wtm{*7.15}& \ints{33	}& \ETA{8493465} \\
+			& 3,4	&\cpu{ 92.98}&\wtm{15.19}&\ints{33}& \cpu{*10.36}& \wtm{*7.15}& \ints{33	}& \ETA{8493465} \\
+			% 33 interleavings tested; 1 preemptions; job time 8s; pldi time 15185585; new-fixed pldi cputime 92982811
 			% 33 interleavings tested; 1 preemptions; job time 3s; pldi time 7153968; new-fixed pldi cputime 10355961
 		\cline{2-9}
-			& 4,1	&\cpu{ 53.08}&\wtm{9.79}&\ints{*15}& \cpu{*7.77}& \wtm{*5.30}& \ints{5	}& \ETA{460} \\
+			& 4,1	&\cpu{ 45.13}&\wtm{8.56}&\ints{5}& \cpu{*7.77}& \wtm{*5.30}& \ints{5	}& \ETA{460} \\
+			% 5 interleavings tested; 1 preemptions; job time 4s; pldi time 8556561; new-fixed pldi cputime 45128135
 			% 5 interleavings tested; 1 preemptions; job time 2s; pldi time 5295360; new-fixed pldi cputime 7770389
-			& 4,2	&\cpu{117.07}&\wtm{19.09}&\ints{*33}& \cpu{*8.32}& \wtm{*5.59}& \ints{9	}& \ETA{132710} \\
+			& 4,2	&\cpu{64.99}&\wtm{11.33}&\ints{9}& \cpu{*8.32}& \wtm{*5.59}& \ints{9	}& \ETA{132710} \\
+			% 9 interleavings tested; 1 preemptions; job time 5s; pldi time 11333162; new-fixed pldi cputime 64991248
 			% 132710 9 interleavings tested; 1 preemptions; job time 2s; pldi time 5588002; new-fixed pldi cputime 8317173
@@
 		{\tt swapbug}
-			& 2,1	&\cpu{ 70.95}&\wtm{13.45}&\ints{*16}& \cpu{*28.32}& \wtm{*11.29}& \ints{33	}& \ETA{73} \\
+			& 2,1	&\cpu{ 84.26}&\wtm{15.49}&\ints{*6}& \cpu{*28.32}& \wtm{*11.29}& \ints{33	}& \ETA{73} \\
+			% 6 interleavings tested; 1 preemptions; job time 5s; pldi time 15485317; new-fixed pldi cputime 84260417
 			% 33 interleavings tested; 1 preemptions; job time 3s; pldi time 11286074; new-fixed pldi cputime 28324164
 		(deadlock)
-			& 2,2	&\cpu{ 107.28}&\wtm{*17.45}&\ints{*146}& \cpu{*30.55}& \wtm{13.50}& \ints{85	}& \ETA{860} \\
+			& 2,2	&\cpu{ 64.65}&\wtm{*11.90}&\ints{*38}& \cpu{*30.55}& \wtm{13.50}& \ints{85	}& \ETA{860} \\
+			% 38 interleavings tested; 2 preemptions; job time 6s; pldi time 11899011; new-fixed pldi cputime 64653180
 			% 85 interleavings tested; 1 preemptions; job time 5s; pldi time 13499606; new-fixed pldi cputime 30550037
-			& 2,3	&\cpu{ 280.05}&\wtm{38.70}&\ints{*352}& \cpu{*32.27}& \wtm{*19.42}& \ints{217	}& \ETA{9120} \\
+			& 2,3	&\cpu{ 94.59}&\wtm{*14.68}&\ints{*48}& \cpu{*32.27}& \wtm{19.42}& \ints{217	}& \ETA{9120} \\
+			% 48 interleavings tested; 2 preemptions; job time 9s; pldi time 14681610; new-fixed pldi cputime 94594891
 			% 217 interleavings tested; 1 preemptions; job time 11s; pldi time 19419243; new-fixed pldi cputime 37272562
-			& 2,4	&\cpu{ 617.94}&\wtm{*81.50}&\ints{*834}& \cpu{*53.44}& \wtm{35.80}& \ints{537	}& \ETA{91239} \\
+			& 2,4	&\cpu{ 109.37}&\wtm{*17.05}&\ints{*68}& \cpu{*53.44}& \wtm{35.80}& \ints{537	}& \ETA{91239} \\
+			% 68 interleavings tested; 2 preemptions; job time 10s; pldi time 17049141; new-fixed pldi cputime 109374628
 			% 537 interleavings tested; 1 preemptions; job time 27s; pldi time 35796053; new-fixed pldi cputime 53441873
 		\cline{2-9}
-			& 3,1	&\cpu{*1275.04}&\wtm{*163.42}&\ints{*771}&\cpu{*65.13}& \wtm{52.23}& \ints{980   }& \ETA{3417} \\
+			& 3,1	&\cpu{221.58}&\wtm{*33.44}&\ints{*32}&\cpu{*65.13}& \wtm{52.23}& \ints{980   }& \ETA{3417} \\
+			% 32 interleavings tested; 4 preemptions; job time 6s; pldi time 33442378; new-fixed pldi cputime 221580261
 			% 980 interleavings tested; 2 preemptions; job time 40s; pldi time 52225013; new-fixed pldi cputime 65125824
-			& 3,2	&\ETA{--}&\ETA{>30m}&\ETA{--}&\ETA{--}&\ETA{>1h}&\ETA{--}& \ETA{2256553} \\
+			& 3,2	&\cpu{*3283.16}&\wtm{*413.59}&\ints{*260}&\ETA{--}&\ETA{>1h}&\ETA{--}& \ETA{2256553} \\
+			% 260 interleavings tested; 4 preemptions; job time 32s; pldi time 413587352; new-fixed pldi cputime 3283157217
 			% [JOB 8] progress: 63137/2256553 brs (2.797940%), ETA 130d 19h 28m 29s (elapsed 59m 45s)
@@
 		{\tt avl\_insert}
-			& 2,2	&\cpu{488.07}&\wtm{64.77}&\ints{*83}	&\cpu{*66.32}&\wtm{*31.81}&\ints{127}&\ETA{254078} \\
+			& 2,2	&\cpu{862.54}&\wtm{111.64}&\ints{*57}	&\cpu{*66.32}&\wtm{*31.81}&\ints{127}&\ETA{254078} \\
+			% 57 interleavings tested; 1 preemptions; job time 13s; pldi time 111641721; new-fixed pldi cputime 862539202
 		(segfault+)
-			& 2,3	&\cpu{2670.87}&\wtm{*330.45}&\ints{*3066}&\cpu{*256.94}&\wtm{193.71}&\ints{1391}&\ETA{22773586} \\
-			& 2,4	&\cpu{*3259.37}&\wtm{*436.50}&\ints{*1639}&\ETA{--}&\ETA{>1h}&\ETA{--}& \ETA{91704203} \\
+			& 2,3	&\cpu{1490.20}&\wtm{*191.56}&\ints{*221}&\cpu{*256.94}&\wtm{193.71}&\ints{1391}&\ETA{22773586} \\
+			% 221 interleavings tested; 2 preemptions; job time 27s; pldi time 191557235; new-fixed pldi cputime 1490204572
+			& 2,4	&\cpu{*7598.19}&\wtm{*970.67}&\ints{*2633}&\ETA{--}&\ETA{>1h}&\ETA{--}& \ETA{91704203} \\
+			% 2633 interleavings tested; 3 preemptions; job time 5m 31s; pldi time 970674411; new-fixed pldi cputime 7598188798
 			% [JOB 12] progress: 28140/91704203 brs (0.030686%), ETA 18y 54d 18h 42m 58s (elapsed 51m 12s)
 		\cline{2-9}
-			& 3,1	&\cpu{222.02}&\wtm{40.04}&\ints{*28}	&\cpu{*64.70}&\wtm{*25.59}&\ints{53}&\ETA{343250} \\
+			& 3,1	&\cpu{62.69}&\wtm{*10.36}&\ints{*23}	&\cpu{*64.70}&\wtm{25.59}&\ints{53}&\ETA{343250} \\
+			% 23 interleavings tested; 3 preemptions; job time 6s; pldi time 10359882; new-fixed pldi cputime 62690171
 			% 53 interleavings tested; 3 preemptions; job time 6s; pldi time 25585758; new-fixed pldi cputime 64695226
-			& 3,2	&\cpu{*1569.09}&\wtm{*216.85}&\ints{*209}&\ETA{--}&\ETA{>1h}&\ETA{--}& \ETA{466977272} \\
+			& 3,2	&\cpu{*2445.90}&\wtm{*325.81}&\ints{*233}&\ETA{--}&\ETA{>1h}&\ETA{--}& \ETA{466977272} \\
+			% 233 interleavings tested; 3 preemptions; job time 30s; pldi time 325806017; new-fixed pldi cputime 2445895897)
+			% was also found with 169 interleavings but later on
 			%[JOB 38] progress: 3812/466977272 brs (0.000816%), ETA INF (elapsed 13m 50s)
 		\hline
-			\multicolumn{5}{c}{\em every quicksand above this line needs rerun} \\ \hline
 		{\tt lock\_fast}
@@
 		For each, I list the CPU-time and wall-clock time elapsed,
-		plus the number of interleavings of the ultimately buggy state space tested,
+		plus the number of interleavings tested in the ultimately buggy state space,
 		before the bug was found.
@@
 		confers a sense of the exponential explosion.
-		{\bf THESIS DRAFT NOTE:} These numbers all need to be re-run on account of some state space reductions and/or soundness fixes that have been added since these numbers were calculated in \cite{sigbovik-htm}.
 	}
@@
 presents the bug-finding results.
-I configured Landslide to run Quicksand's Iterative Deepening algorithm, %(\sect{\ref{sec:quicksand-id}}),
-% TODO how many cores?
+I configured Landslide to run Quicksand's Iterative Deepening algorithm on 8 cores, %(\sect{\ref{sec:quicksand-id}}),
 shown left,
@@
 shown right,
-% TODO: make 1 hour
-each with a time limit of 30 minutes.
+each with a time limit of 1 hour.
 Tests {\tt htm1}, {\tt swapbug}, and {\tt avl\_insert} were run with {\tt landslide -X}
@@
 			{\tt map\_basicer}
-			& 2,1 & \cpu{36.17}& \ints{426}	\\
-			& 2,2 & \ETAdag{todo} & \ETAdag{todo} \\ % currently running
+			& 2,1 & \cpu{36.17}& \ints{426}	\\ % TODO make sure this gets updated if you update the test
+			& 2,2 & \ETAdag{todo} & \ETAdag{todo} \\
 			& 3,1 & \ETAdag{todo} & \ETAdag{todo} \\

### 2018-07-19 thesis/tm.tex
fix conflict abort discussion to account for causality

@@
 	\item {\tt \_XABORT\_CONFLICT}:
-		Whenever DPOR detects a memory conflict between two preemption points ({\tt shimsham\_shm()}),
+		Whenever DPOR detects a memory conflict between two transitions ({\tt shimsham\_shm()}),
 		if {\tt -A} is set,
-		for each of the two which are within a transaction,
-		it finds the nearest ancestor {\tt xbegin} preemption point by the same thread
-		and adds this code to both lists.
+		% for each of the two which are within a transaction, % no!!!!!!!
+		%it finds the nearest ancestor {\tt xbegin} preemption point by the same thread
+		if the later-executed transition is a transactional success path,
+		it adds this code to both lists of the immediately preceding preemption point.
+		Note that a transaction should suffer a conflict abort
+		only when executed {\em after} a conflicting memory access
+		to avoid circular causality
+		(see {\tt 410user/progs/htm\_causality.c} for rationale).
+		Should the transaction happen to be executed first,
+		DPOR will first try to reorder it as normal,
+		and then abort it as described if the conflict persists.
 	\item {\tt \_XABORT\_EXPLICIT}:
@@
 
+% don't forget to note that this only applies to xabort retry;
+
+The proof of soundness for this optimization is left to future work;
+while the basic idea is motivated by straightforward analogy to the known-sound sleep sets,
+its intersection with Landslide's other exploration features may cause unforeseen problems.
+Accordingly, I implemented the optimization as conservatively as possible,
+falling back to no pruning in cases where how to prune soundly is not clear
+(for example, when a second pair of conflicting transactions involving a third thread
+interleaves while a first abort set is already being tracked).
+
 % TODO
 
+% TODO - rewrite the following example so that it only uses retry aborts
 % ok so heres an example of why this is rly hard

### 2018-07-19 thesis/tm.tex
rerun all -X -A -S tests bc of causality violation

@@
 		%({\tt spinlock})
-			& 2,1	&\cpu{10.48}&\wtm{3.22}&\ints{2}&\cpu{*2.98}&\wtm{*2.98}&\ints{2}& \ETA{8} \\
+			& 2,1	&\cpu{8.17}&\wtm{2.65}&\ints{2}&\cpu{*2.39}&\wtm{*2.39}&\ints{2}& \ETA{4} \\
 		(perf)
-			& 9,9	&\cpu{37.41}&\wtm{9.96}&\ints{2}&\cpu{*4.93}&\wtm{*4.93}&\ints{2}& \ETA{inf} \\
+			& 9,9	&\cpu{36.64}&\wtm{9.78}&\ints{2}&\cpu{*4.75}&\wtm{*4.75}&\ints{2}& \ETA{inf} \\
 	\end{tabular}
@@
 whereas {\tt avl\_fixed} and {\tt map\_basic} effectively duplicated the mutex verification between them.
-Note also the impact that fixing Figure~\ref{fig:spinlockbug}'s performance bug
+
+Note two curiosities:
+firstly,
+the impact that fixing Figure~\ref{fig:spinlockbug}'s performance bug
 (changing a read+write to a read only)
@@
 {\tt lock}({\tt spin\_fixed})'s state spaces were reduced by more than half compared to {\tt lock}({\tt spinlock}).
+Secondly, {\tt spinlock} and {\tt spin\_fixed} take longer to test per interleaving
+(CPU time divided by state space size)
+than {\tt mutex},
+because while {\tt mutex} abstracts away threads needing to wait their turn for the critical section
+behind an API Landslide understands,
+the spinlock's wait loop is open-coded, and Landslide must fall back on its costlier heuristic synchronization detection
+(\sect{\ref{sec:landslide-blocking}}).
+In this way (and also, of course, because {\tt mutex}'s state spaces are smaller overall),
+{\tt mutex} can be seen in turn as a further abstraction reduction of {\tt spinlock}.
 
@@
 			{\tt lock}
-			& 2,1 & \cpu{14.23} & \ints{41} \\
+			& 2,1 & \cpu{2.49} & \ints{4} \\ % certified causality ok
 			({\tt spinlock})
-			& 2,2 & \cpu{4944.37} & \ints{38544} \\
+			& 2,2 & \cpu{174.26} & \ints{1702} \\ % certified causality ok
 			% if lock(spin_fixed) will time out on 2,3 and 3,2, this is guaranteed to aswell
 			% & 2,3 & \\ % \cpu{0} & \ints{0} \\
-			& 3,1 & \cpu{13649.32} & \ints{97462} \\
+			& 3,1 & \cpu{26.97} & \ints{246} \\ % certified causality ok
 			% & 3,2 & \\ % \cpu{0} & \ints{0} \\
@@
 			{\tt lock}
-			& 2,1 & \cpu{12.26} & \ints{31} \\
+			& 2,1 & \cpu{2.47} & \ints{4} \\ % certified causality ok
 			({\tt spin\_fixed})
-			& 2,2 & \cpu{1662.21} & \ints{15076} \\
-			& 2,3 & \ETAdag{3d 20h} & \ETAdag{2003906} \\
-			& 3,1 & \cpu{2488.95} & \ints{24970} \\
-			& 3,2 & \ETAdag{207d 20h} & \ETAdag{6514618} \\
+			& 2,2 & \cpu{94.64} & \ints{998} \\ % certified causality ok
+			& 2,3 & \ETAdag{todo} & \ETAdag{todo} \\
+			& 3,1 & \cpu{218.02} & \ints{185} \\ % certified causality ok
+			& 3,2 & \ETAdag{todo} & \ETAdag{todo} \\
 			& 4,1 & \\ % TODO
@@
 			{\tt lock}
-			& 2,1 & \cpu{9.42} & \ints{22}	\\
+			& 2,1 & \cpu{2.49} & \ints{4}		\\ % certified causality ok
 			({\tt mutex})
-			& 2,2 & \cpu{44.07} & \ints{1208}	\\
-			& 2,3 & \cpu{2647.38} & \ints{71136}	\\
-			& 2,4 & \ETAdag{2d 1h} & \ETAdag{3223055}	\\
-			& 3,1 & \cpu{42.11} & \ints{1148}	\\
-			& 3,2 & \ETAdag{11d 4h} & \ETAdag{4793666}	\\
-			& 4,1 & \cpu{5586.37} & \ints{159664}	\\
+			& 2,2 & \cpu{12.72} & \ints{180}	\\ % certified causality ok
+			& 2,3 & \cpu{273.45} & \ints{8356}	\\ % certified causality ok
+			& 2,4 & \ETAdag{todo} & \ETAdag{todo}	\\
+			& 3,1 & \cpu{11.21} & \ints{132}	\\ % certified causality ok
+			& 3,2 & \ETAdag{todo} & \ETAdag{todo}	\\
+			& 4,1 & \cpu{493.38} & \ints{15064}	\\ % certified causality ok
+			& 4,2 & \\ % TODO maybe?????? well its 10hrs
 			%%%%%%%%%%% htm_mutex
@@
 			{\tt lock\_fast}
-			& 2,1 & \cpu{2.96} & \ints{1}	\\
+			& 2,1 & \cpu{2.42} & \ints{1}	\\
 			({\tt spin\_fixed})
-			& 9,9 & \cpu{4.33} & \ints{1}	\\
+			& 9,9 & \cpu{3.69} & \ints{1}	\\
 			\hline
 			{\tt lock\_fast}
-			& 2,1 & \cpu{3.05} & \ints{1}	\\
+			& 2,1 & \cpu{2.35} & \ints{1}	\\
 			({\tt mutex})
-			& 9,9 & \cpu{4.38} & \ints{1}	\\
+			& 9,9 & \cpu{3.79} & \ints{1}	\\
 			%\end{tabular}

### 2018-07-28 thesis/landslide.tex
ok i think retry set impl is final, so write about it

@@
 
-{\bf Further pruning.}
+{\bf Choosing which new interleaving to test next.}
+Among all interleavings chosen by DPOR not already marked as explored,
+Landslide chooses the one with the longest execution prefix matching the current $S$,
+to maintain the depth-first search invariant.
+(In the case of a tie, differing only by which thread to run, it chooses arbitrarily.)
+All other new interleavings are marked to explore later in their corresponding {\tt struct hax},
+and automatically included in the result of any future iterations of DPOR until they are tested.
+Because the one with the longest execution prefix was chosen to test next,
+all others must share their execution prefixes with it,
+preserving the $O(nk)$ memory bound described above.
+%so storing markers for the others during the next execution is still $O(1)$ memory.
+
+{\bf Termination.}
+When DPOR returns no new $I_{ij}$s not already marked in the set of visited subtrees,
+the exploration is complete.
+
+\subsubsection{Sleep set reduction}
+\label{sec:landslide-sleepsets}
+
 In the presence of non-conflicting transitions as well as conflicting ones,
@@
 \end{figure}
+
 Landslide implements this check in {\tt equiv\_already\_explored()}.
@@
 so named because it effectively puts $T_1$ ``to sleep'' until after the true conflicting access in $t_c$.
+Landslide's implementation differs from prior work by not explicitly tracking
+sets of reordered threads and memory accesses they should expect to conflict with,
+and instead identifies where the reduction should take place in subsequent iterations of DPOR.
 Further optimizations such as {\em source sets} and {\em wakeup trees},
@@
 % and that the HTM expts fro msigbovik paper were re-run for its chapater herein
-
-{\bf Choosing which new interleaving to test next.}
-Among all interleavings chosen by DPOR not already marked as explored,
-Landslide chooses the one with the longest execution prefix matching the current $S$,
-to maintain the depth-first search invariant.
-(In the case of a tie, differing only by which thread to run, it chooses arbitrarily.)
-All other new interleavings are marked to explore later in their corresponding {\tt struct hax},
-and automatically included in the result of any future iterations of DPOR until they are tested.
-Because the one with the longest execution prefix was chosen to test next,
-all others must share their execution prefixes with it,
-preserving the $O(nk)$ memory bound described above.
-%so storing markers for the others during the next execution is still $O(1)$ memory.
-
-{\bf Termination.}
-When DPOR returns no new $I_{ij}$s not already marked in the set of visited subtrees,
-the exploration is complete.
 

### 2018-07-28 thesis/tm.tex
some mx verif results
more 10h expts
ok i think retry set impl is final, so write about it

@@
 \subsubsection{Adding possible abort codes}
+\label{sec:txn-abort-codes}
 
@@
 
-\subsubsection{Tracking abort sets}
-\label{sec:tm-abortsets}
-
-% In contrast to Landslide's implementation of the sleep set reduction (\sect{\ref{sec:landslide-explore}}),
-% blah blah which is implicit and works in retrospect by pruning at tag-sibling time, which figures out what would have been in a sleep set by analyzing the past independences and explored siblings,
-% this cannot work that way,
-% because in the first place, the former works with the "keep running current thread" ordering heuristic,
-% (which automatically takes care of the first half of the sleep set problem)
-% whereas no such similar heuristic exists (can exist??) for xbegin PPs, we always just run the success path first
-
-% don't forget to note that this only applies to xabort retry;
-
-The proof of soundness for this optimization is left to future work;
-while the basic idea is motivated by straightforward analogy to the known-sound sleep sets,
-its intersection with Landslide's other exploration features may cause unforeseen problems.
-Accordingly, I implemented the optimization as conservatively as possible,
-falling back to no pruning in cases where how to prune soundly is not clear
-(for example, when a second pair of conflicting transactions involving a third thread
-interleaves while a first abort set is already being tracked).
-
-% TODO
-
-% TODO - rewrite the following example so that it only uses retry aborts
-% ok so heres an example of why this is rly hard
-
-% ==== thread 1 ====
-% if (xbegin()) {
-% 	xabort(0) // whatever, executed in a past branch to be aborting in the present
-% } else {
-% 	x++;
-% }
-%
-% ==== thread 3 ====
-% conflict_with_thread_2 = true;
-%
-% ==== thread 2 ====
-% if (xbegin()) {
-% 	conflict_with_thread_2 ++ ; // whatever, just conflict and fail w thr 3
-% } else {
-% 	x++; // conf with thr 1
-% }
-%
-% so we execute this interleaving (T1 abort, T3, T2 abort)
-% dpor sees the conflict between T2 and T1 and wants to tag T2 to run as transition(T1-abort)'s sibling to run
-% it would use the "abort set" of (T2 [CONFLICT], T1 [EXPLICIT])
-% but like, how do we know that thread 3 is necessary to reach this?
-% and even, how do we know that it's valid to inject CONFLCIT in T2?
-% you pretty much hafta reorder T3 (dpor doesn't see why you need to)
-% but if you inject conflict in T2 without also making sure T3 executes "near" it, that might not necessarily be sound
-
 \subsection{Data Race Analysis}
@@
 the vector clocks are simply checked as usual.
+
+\subsection{Retry independence}
+\label{sec:tm-retrysets}
+
+Finally, I identified a specific pattern of transactional code where existing state-space reduction algorithms
+will fail to identify and prune equivalent thread interleavings.
+Figure~\ref{fig:retry-set-example} shows a minimal example program which exhibits this problem.
+In this program, each thread's transactional path conflicts with the other thread's abort path,
+while the two transactional paths (disjoint memory accesses) and the two abort paths (reads only) are pairwise independent.
+
+\begin{figure}[h]
+        \begin{center}
+	% TODO syntax hilight
+        \begin{tabular}{ll}
+		\multicolumn{2}{c}{Initially {\tt \ctype{int} foo, bar = \const{0}; \ctype{mutex\_t} m;}} \\
+                \\
+                \begin{tabular}{l}
+                {\bf Thread 1} \\
+                \hline
+			\texttt{\flow{if} (\call{\_xbegin}() == \const{STARTED}) \{} \\
+			\texttt{~~~~foo++;} \\
+			\texttt{~~~~\call{\_xend}();} \\
+			\texttt{\} \flow{else} \{} \\
+			\texttt{~~~~\call{mutex\_lock}(\&m);} \\
+			\texttt{~~~~\flow{assert}(foo + bar < \const{2});} \\
+			\texttt{~~~~\call{mutex\_unlock}(\&m);} \\
+			\texttt{\}} \\
+                \end{tabular}
+                &
+                \begin{tabular}{l}
+                {\bf Thread 2} \\
+                \hline
+			\texttt{\flow{if} (\call{\_xbegin}() == \const{STARTED}) \{} \\
+			\texttt{~~~~bar++;} \\
+			\texttt{~~~~\call{\_xend}();} \\
+			\texttt{\} \flow{else} \{} \\
+			\texttt{~~~~\call{mutex\_lock}(\&m);} \\
+			\texttt{~~~~\flow{assert}(foo + bar < \const{2});} \\
+			\texttt{~~~~\call{mutex\_unlock}(\&m);} \\
+			\texttt{\}} \\
+                \end{tabular}
+        \end{tabular}
+        \end{center}
+	\caption{Motivating example for retry independence reduction.}
+	\label{fig:retry-set-example}
+\end{figure}
+
+Ordinarily, in the state space subset which schedules thread 1 before thread 2,
+there would be 4 combinations of each thread succeeding or aborting their respective transactions;
+among those, at least one would show a memory conflict,
+causing DPOR to explore into the other half of the state space which schedules thread 2 first,
+where the same 4 combinations of transaction reuslts would be tested in the other order,
+even though 2 are redundant under schedule reordering.
+Consequently, na\"ively testing both success and retry aborts in both threads regardless of reordering
+will unnecessarily execute % oops
+both equivalent interleavings from each such pair;
+to identify such equivalences,
+DPOR must somehow remember which combination of {\tt xbegin} results led to the memory conflict in the first place.
+
+While Figure~\ref{fig:retry-set-example}'s example may seem contrived
+(what program's transactions would just give up and do no work if aborted?),
+it is easy to imagine a larger transactional data structure,
+two insertions into which might operate on disjoint nodes or array indices,
+allowing simultaneous transactions to usually succeed,
+while the more uncommon abort paths might take the opportunity to assert a full consistency check of all elements,
+ultimately resulting in a similar conflict pattern.
+Also, the evaluation will later show
+that even programs with fully conflicting success/abort paths
+may still exhibit some equivalent thread interleavings of this pattern
+after their failure paths are split apart into smaller transitions by data-race preemption points
+(\sect{\ref{sec:tm-eval-retrysets}}).
+
+To solve this,
+I extended Landslide
+with an experimental feature called {\em retry set reduction},
+so named after the analogous {\em sleep sets} from prior work
+(\cite{partial-order-methods,dpor,optimal-dpor}, \sect{\ref{sec:landslide-sleepsets}}).
+Whenever DPOR tags a new branch for exploration (\sect{\ref{sec:landslide-explore}}),
+if either or both sides of a memory conflict were part of a transaction,
+it records a {\em retry set},
+i.e., a pair of {\tt xbegin} results executed by the conflicting threads.
+% ah-- not compatible after all!
+%in addition to tracking the set of abort codes possible at each preemption point.
+Later, when traversing that new branch,
+Landslide remembers those threads' expected {\tt xbegin} results
+and refuses to test any others
+(unless DPOR separately found them, too, in conflict),
+thereby skipping over (i.e., pruning) reorderings of any other {\tt xbegin} results that would be independent.
+The {\tt landslide} repository's {\tt retry-sets} branch implements this feature%
+\footnote{The upcoming evaluation showed it too unstable, let alone not yet proved sound, to include in {\tt master}.}.
+
+% In contrast to Landslide's implementation of the sleep set reduction (\sect{\ref{sec:landslide-explore}}),
+% blah blah which is implicit and works in retrospect by pruning at tag-sibling time, which figures out what would have been in a sleep set by analyzing the past independences and explored siblings,
+% this cannot work that way,
+% because in the first place, the former works with the "keep running current thread" ordering heuristic,
+% (which automatically takes care of the first half of the sleep set problem)
+% whereas no such similar heuristic exists (can exist??) for xbegin PPs, we always just run the success path first
+
+% don't forget to note that this only applies to xabort retry;
+
+The proof of soundness for this optimization is left to future work;
+while the basic idea is motivated by straightforward analogy to the known-sound sleep sets,
+its intersection with Landslide's other exploration features may cause unforeseen problems.
+Accordingly, I implemented the optimization as conservatively as possible,
+falling back to no pruning in cases where how to prune soundly is not clear
+(for example, when a second pair of conflicting transactions involving a third thread
+interleaves while a first retry set is already being tracked).
+Note also that {\tt xbegin} results other than {\tt \_XABORT\_RETRY}
+may depend on the execution logic
+(explicit aborts may be conditional on some change by a conflicting thread,
+and conflict aborts cannot occur before their conflicting transaction to begin with%
+\footnote{See {\tt 410user/progs/htm\_causality.c}; that was a fun realization to have
+already halfway into implementing this feature the wrong way at first.}%
+),
+so how to soundly prune under multiple abort codes
+and/or when retry aborts are suppressed is left to future work.
+For now, retry sets, true to their name,
+are supported only under {\tt landslide -X},
+and incompatible with both the {\tt -A} and {\tt -S} options.
 
@@
 
-\subsection{Bugs}
+\subsection{Bug-finding}
 \label{sec:tm-eval-bugs}
@@
 			& 3,1 &\cpu{41.07}	& \ints{941}	\\
-			& 3,2 &	\ETAdag{todo}	& \ETAdag{todo}	\\
+			& 3,2 &	\ETAdag{9d 19h}	& \ETAdag{5058149}	\\ % add 10h ok
+			% [JOB 4] progress: 923822/5058149 brs (18.264030%), ETA 9d 8h 30m 38s (elapsed 9h 59m 55s)
 			& 4,1 & \cpu{12699.65}	& \ints{344240}	\\
@@
 for even small thread/iteration counts,
-in one case failing to make any guarantee about even the minimum possible test configuration.
-I explore two possible mitigation approaches.
+in one case failing to make any guarantee about even the minimum possible test configuration,
+and in another failing to verify that a previously-found bug had been fixed.
+I explore three possible mitigation approaches.
 
@@
 \begin{figure}[h]
-	% TODO: syn hi
 	\begin{center}
@@
 		%\begin{tabular}{cc||r|r||r|r}
-		%	& & \multicolumn{2}{c||}{\bf DPOR} & \multicolumn{2}{c}{\bf DPOR + abort sets} \\
+		%	& & \multicolumn{2}{c||}{\bf DPOR} & \multicolumn{2}{c}{\bf DPOR + retry sets} \\
 		%	& & \cpu{\bf cpu (s)} & \ints{\bf SS size}
@@
 			& 2,2 & \cpu{94.64} & \ints{998} \\ % certified causality ok
-			& 2,3 & \ETAdag{todo} & \ETAdag{todo} \\
+			& 2,3 & \ETAdag{28h 6m} & \ETAdag{486966} \\ % certified causality ok, add 10h ok
+			% [JOB 4] progress: 234307/486966 brs (48.115630%), ETA 18h 5m 37s (elapsed 9h 59m 50s)
 			& 3,1 & \cpu{218.02} & \ints{185} \\ % certified causality ok
@@
 			& 2,3 & \cpu{273.45} & \ints{8356}	\\ % certified causality ok
-			& 2,4 & \ETAdag{todo} & \ETAdag{todo}	\\
+			& 2,4 & \cpu{13545.35} & \ints{384512}	\\ % certified causality ok
 			& 3,1 & \cpu{11.21} & \ints{132}	\\ % certified causality ok
-			& 3,2 & \ETAdag{todo} & \ETAdag{todo}	\\
+			& 3,2 & \cpu{25434.85} & \ints{718690}	\\ % certified causality ok
 			& 4,1 & \cpu{493.38} & \ints{15064}	\\ % certified causality ok
@@
 		The reduced {\tt avl\_mutex} and {\tt map\_mutex} are, of course, no longer transactional.
-		%% i mean ok xabort sets doesn't really help for ANY test tho
-		%so the abort sets optimization does not apply to them.
 		}
@@
 
-\subsubsection{Abort set reduction}
+\subsubsection{Retry set reduction}
+\label{sec:tm-eval-retrysets}
 
@@
 
-(\sect{\ref{sec:tm-abortsets}}).
+(\sect{\ref{sec:tm-retrysets}}).
 

### 2018-07-28 thesis/tm.tex
more tm expts

@@
 		with increasing thread/iteration count for correct transactional programs?
-	\item What further reduction can be achieved beyond that provided by
+	\item What further reduction can be achieved beyond the baseline provided by
 		the global-lock/failure-injection equivalences?
@@
 		\item {\tt swapbug}: {\tt swap} modified to introduce circular locking in the failure path. %(deadlock).
+			% TODO: make sure figure number reference is right
+		\item {\tt fig63}: Generalized version of Figure~\ref{fig:retry-set-example},
+			contrived to induce as much reduction as possible from the retry sets optimization.
 	\end{itemize}
@@
 			\hline
+			{\tt fig63}
+			& 2,1 & \cpu{todo}	& \ints{todo}	\\
+			& 2,2 & \cpu{todo}	& \ints{todo}	\\
+			& 3,2 & \ETAdag{todo} & \ETAdag{todo}	\\
+			\hline
 			{\tt avl\_insert}
@@
 			& 2,1 & \cpu{572.49} 	& \ints{20420} 	\\
-			& 2,2 & \ETAdag{todo}	& \ETAdag{todo} 	\\
+			& 2,2 & \ETAdag{133y}	& \ETAdag{51484570} 	\\ % added 10hr but dosent matter
+			% [JOB 19] progress: 408805/51484570 brs (0.794034%), ETA 132y 356d 8h 55m 16s (elapsed 9h 59m 19s)
 			& 3,1 & \ETAdag{todo}	& \ETAdag{todo}	\\
@@
 had on even the correctness tests:
-{\tt lock}({\tt spin\_fixed})'s state spaces were reduced by more than half compared to {\tt lock}({\tt spinlock}).
+{\tt lock}({\tt spin\_fixed})'s state spaces were reduced by more than half compared to {\tt lock}({\tt spinlock}),
+on account of DPOR no longer needing to reorder the (now) read-read access pairs.
 Secondly, {\tt spinlock} and {\tt spin\_fixed} take longer to test per interleaving
-(CPU time divided by state space size)
-than {\tt mutex},
+%(CPU time divided by state space size)
+than {\tt mutex}
+(roughly 9 interleavings per second for the former, 28 for the latter),
 because while {\tt mutex} abstracts away threads needing to wait their turn for the critical section
@@
 			% [JOB 4] progress: 234307/486966 brs (48.115630%), ETA 18h 5m 37s (elapsed 9h 59m 50s)
-			& 3,1 & \cpu{218.02} & \ints{185} \\ % certified causality ok
-			& 3,2 & \ETAdag{todo} & \ETAdag{todo} \\
-			& 4,1 & \\ % TODO
+			& 3,1 & \cpu{21.68} & \ints{186} \\ % certified causality ok
+			& 3,2 & \ETAdag{8y 153d} & \ETAdag{1590416} \\ % caus ok
+			% [JOB 4] progress: 187162/1590416 brs (11.768113%), ETA 8y 152d 16h 25m 38s (elapsed 9h 59m 49s)
+			& 4,1 & \cpu{30408.58} & \ints {282084} \\ % causality ok
 			% (2,1) with isfree fix
@@
 			& 4,1 & \cpu{493.38} & \ints{15064}	\\ % certified causality ok
-			& 4,2 & \\ % TODO maybe?????? well its 10hrs
 			%%%%%%%%%%% htm_mutex

### 2018-07-28 thesis/tm.tex
extend verifs table and run buncha small retry-set and abort-code verif tests

@@
 		this code is added to both lists, bitwise-ored with the user-supplied argument code
-		as specified in \cite{htm-gcc}.
+		as specified in \cite{htm-gcc},
+		and execution of the transactional path immediately stops.
+	\item {\tt \_XABORT\_CAPACITY}:
+		Whenever the program invokes a system call during a transaction,
+		if {\tt -A} is set,
+		this code is added to both lists,
+		and execution of the transactional path immediately stops.
 \end{itemize}
@@
 I extended Landslide
-with an experimental feature called {\em retry set reduction},
-so named after the analogous {\em sleep sets} from prior work
+with {\em retry set reduction},
+an experimental feature named after the analogous {\em sleep sets} from prior work
 (\cite{partial-order-methods,dpor,optimal-dpor}, \sect{\ref{sec:landslide-sleepsets}}).
@@
 it records a {\em retry set},
-i.e., a pair of {\tt xbegin} results executed by the conflicting threads.
+i.e., the pair of {\tt xbegin} results executed by the conflicting threads.
 % ah-- not compatible after all!
@@
 The {\tt landslide} repository's {\tt retry-sets} branch implements this feature%
-\footnote{The upcoming evaluation showed it too unstable, let alone not yet proved sound, to include in {\tt master}.}.
+\footnote{The upcoming evaluation showed it too unstable, let alone not yet proved sound, to include in {\tt master}.},
+available with the {\tt -R} command-line option.
 
@@
 
+% TODO
+% discuss the diff reducion approaches
+
+% Please note that retry sets are a theoretically-founded reduction that should apply to all programs,
+% while distinguishing abort codes to suppress retries is potentially unsound,
+% depending on the user's ability to manually verify the presence (and correctness) of retry loops in the code.
+
 \begin{table}[h]
@@
 		\footnotesize
-		\begin{tabular}{cc|r|r}
-			& & \cpu{\bf cpu (s)} & \ints{\bf SS size} \\
-			\bf test & \bf params & \em (or \ETAdag{\bf \em ETA}) & \em (or \ETAdag{\bf \em est.}) \\
+		\begin{tabular}{cc||r|r||r|r||r|r}
+			& & \multicolumn{2}{c||}{\bf Retry aborts only}
+			  & \multicolumn{2}{c||}{\bf Retry sets ({\tt -R})}
+			  & \multicolumn{2}{c}{\bf Abort codes ({\tt -A -S})} \\
+			& & \cpu{\bf cpu (s)} & \ints{\bf SS size}
+			  & \cpu{\bf cpu (s)} & \ints{\bf SS size}
+			  & \cpu{\bf cpu (s)} & \ints{\bf SS size} \\
+			\bf test & \bf params
+			  & \em (or \ETAdag{\bf \em ETA}) & \em (or \ETAdag{\bf \em est.})
+			  & \em (or \ETAdag{\bf \em ETA}) & \em (or \ETAdag{\bf \em est.})
+			  & \em (or \ETAdag{\bf \em ETA}) & \em (or \ETAdag{\bf \em est.}) \\
 			\hline
@@
 			{\tt htm2}
-			& 2,1 & \cpu{16.30}	& \ints{22}	\\
-			& 2,2 & \cpu{69.80}	& \ints{1970}	\\
-			& 2,3 &\cpu{3510.00}	& \ints{104914}	\\
+			& 2,1 & \cpu{16.30}	& \ints{22}	& \cpu{19.82}	& \ints{17}	& \cpu{3.32}	& \ints{4} \\
+			& 2,2 & \cpu{69.80}	& \ints{1970}	& \cpu{38.71}	& \ints{906}	& \cpu{26.59}	& \ints{288} \\
+			& 2,3 &\cpu{3510.00}	& \ints{104914}	& \cpu{}	& \ints{}	& \cpu{}	& \ints{} \\
 			%& 2,4 &	\ETAdag{...}	& \ETAdag{...}	\\
-			& 3,1 &\cpu{41.07}	& \ints{941}	\\
+			& 3,1 &\cpu{41.07}	& \ints{941}	& \cpu{29.38}	& \ints{553}	& \cpu{23.03}	& \ints{162} \\
 			& 3,2 &	\ETAdag{9d 19h}	& \ETAdag{5058149}	\\ % add 10h ok
 			% [JOB 4] progress: 923822/5058149 brs (18.264030%), ETA 9d 8h 30m 38s (elapsed 9h 59m 55s)
-			& 4,1 & \cpu{12699.65}	& \ints{344240}	\\
+			& 4,1 & \cpu{12699.65}	& \ints{344240}	& \cpu{}	& \ints{}	& \cpu{}	& \ints{} \\
 			\hline
 			{\tt counter}
-			& 2,1 & \cpu{5.40}	& \ints{10}	\\
-			& 2,2 & \cpu{10.91}	& \ints{190}	\\
-			& 2,3 & \cpu{134.21}	& \ints{3970}	\\
-			& 2,4 & \cpu{3009.40}	& \ints{86950}	\\
-			& 3,1 & \cpu{8.30}	& \ints{120}	\\
-			& 3,2 & \cpu{2038.90}	& \ints{60606}	\\
-			& 4,1 & \cpu{101.95}	& \ints{3006}	\\
+			& 2,1 & \cpu{5.40}	& \ints{10}	& \cpu{6.70}	& \ints{8}	& \cpu{3.22}	& \ints{4} \\
+			& 2,2 & \cpu{10.91}	& \ints{190}	& \cpu{14.91}	& \ints{288}	& \cpu{7.90}	& \ints{48} \\
+			& 2,3 & \cpu{134.21}	& \ints{3970}	& \cpu{379.04}	& \ints{11512}	& \cpu{35.00}	& \ints{904} \\
+			& 2,4 & \cpu{3009.40}	& \ints{86950}	& \cpu{}	& \ints{}	& \cpu{}	& \ints{} \\
+			& 3,1 & \cpu{8.30}	& \ints{120}	& \cpu{8.90}	& \ints{74}	& \cpu{7.55}	& \ints{40} \\
+			& 3,2 & \cpu{2038.90}	& \ints{60606}	& \cpu{}	& \ints{}	& \cpu{}	& \ints{} \\
+			& 4,1 & \cpu{101.95}	& \ints{3006}	& \cpu{99.22}	& \ints{2854}	& \cpu{32.95}	& \ints{848} \\
 			\hline
 			{\tt swap}
-			& 2,1 & \cpu{38.79}	& \ints{99}	\\
-			& 2,2 & \cpu{10299.27}	& \ints{233396}	\\
-			& 3,1 & \ETAdag{todo} & \ETAdag{todo}	\\
+			& 2,1 & \cpu{38.79}	& \ints{99}	& \cpu{35.80}	& \ints{84}	& \cpu{3.25}	& \ints{4} \\
+			& 2,2 & \cpu{10299.27}	& \ints{233396}	& \cpu{}	& \ints{}	& \cpu{}	& \ints{} \\
+			& 3,1 & \ETAdag{todo}	& \ETAdag{todo}	& \cpu{}	& \ints{}	& \cpu{}	& \ints{} \\
 			\hline
 			{\tt fig63}
-			& 2,1 & \cpu{todo}	& \ints{todo}	\\
-			& 2,2 & \cpu{todo}	& \ints{todo}	\\
-			& 3,2 & \ETAdag{todo} & \ETAdag{todo}	\\
+			& 2,1 & \cpu{6.66}	& \ints{18}	& \cpu{6.50}	& \ints{6}	& \cpu{3.10}	& \ints{1} \\
+			& 2,2 & \cpu{11.52}	& \ints{244}	& \cpu{10.74}	& \ints{192}	& \cpu{3.12}	& \ints{1} \\
+			& 2,3 & \cpu{107.04}	& \ints{4428}	& \cpu{146.96}	& \ints{6151}	& \cpu{3.13}	& \ints{1} \\
+			& 3,1 & \cpu{13.68}	& \ints{296}	& \cpu{7.29}	& \ints{21}	& \cpu{3.26}	& \ints{1} \\
+			& 3,2 & \ETAdag{todo}	& \ETAdag{todo}	& \cpu{}	& \ints{}	& \cpu{3.35}	& \ints{1} \\
+			& 4,1 & \cpu{220.27}	& \ints{7382}	& \cpu{11.40}	& \ints{157}	& \cpu{3.27}	& \ints{1} \\
 			\hline
 			{\tt avl\_insert}
-			& 2,1 & \cpu{491.98}	& \ints{15056}	\\
+			& 2,1 & \cpu{491.98}	& \ints{15056}	& \cpu{220.53}	& \ints{6332}	& \cpu{98.77}	& \ints{2776} \\
+			% (2,2) - bug found on regular w/ 127 ints, on retry sets w/ 201 ints, and codes w/ 45 ints
 			\hline
 			{\tt avl\_fixed}
-			& 2,1 & \cpu{572.49} 	& \ints{20420} 	\\
+			& 2,1 & \cpu{572.49} 	& \ints{20420} 	& \cpu{271.11}	& \ints{9739}	& \cpu{99.16}	& \ints{2776} \\
 			& 2,2 & \ETAdag{133y}	& \ETAdag{51484570} 	\\ % added 10hr but dosent matter
@@
 	\caption{Transactional tests verified (or not) by Landslide.
-		Run with {\tt landslide -M -X}, i.e., retry aborts enabled, different xabort codes not distinguished.
+		Run with {\tt -M -X}, plus any additional reduction options listed.
+		``Retry aborts only'' always tests every abort path without distinguishing among failure reasons;
+		``Retry sets'' skips equivalent retry abort reorderings (\sect{\ref{sec:tm-retrysets}});
+		``Abort codes'' suppresses retries and dynamically detects when to inject conflict aborts and so on
+		(\sect{\ref{sec:txn-abort-codes}}).
 		State space estimates measured after a timeout of 10 hours.

### 2018-07-28 thesis/education.tex
small copyedits education

@@
 		P(\mathsf{correct}|\mathsf{non\text{-}user})$.
-		Note that the bottom row's totals double-count students,
+		Note that the totals in the bottom row double-count students,
 		which only makes sense if you believe the incidences of each bug within a given submission are independent.
@@
 Though I did not ask this question on the CMU survey,
-my experience handling student questions suggests CMU students
+my experience handling student questions in person
+suggests CMU students
 also mostly wish for

### 2018-07-28 thesis/overview.tex
small copyedits education

@@
 offered the P2 thread library project as part of its curriculum.
-Students in this class implemented P2
+Students in this class implement P2
 on a 6 week project timeline (compared to 2 weeks at CMU),

### 2018-07-28 thesis/tm.tex
say why not using stamp

@@
 for further discussion of variance see \cite{sigbovik-htm}.
+
+I investigated the popular transactional benchmark suite STAMP \cite{stamp}
+to include in this test suite,
+but found that all transactional code therein was written without failure paths,
+so would likely not contribute any theoretical depth to the evaluation.
+STAMP uses the OpenTM interface \cite{opentm},
+which allows the programmer to specify transactional code regions to be implemented atomically,
+however the underlying architecture may require
+(whether HTM or STM; presumably with retry loops on HTM).
+On one hand,
+this lends credence to my upcoming conclusion
+that HTM programs should be written at a higher abstraction level than calling {\tt \_xbegin()} directly;
+on the other,
+OpenTM's requirement of {\em virtualized transaction} semantics
+(i.e., being unconstrained by memory footprint, able to make system calls, and able to nest arbitrarily)
+is more suited to STM and glosses over many pitfalls of HTM programming.
+%% #overlyhonestmethods
+% I also wanted to test the Eunomia algorithm for transactional search trees \cite{eunomia},
+% but its implementation is not open source,
+% and the authors ignored my email requesting to use it.
 

### 2018-07-28 thesis/tm.tex
xabort capacity

@@
 
+{\bf Limitations.}
 Landslide does not yet check for false sharing,
-i.e. read/write or write/write access pairs to different memory addresses that share a cache line.
+i.e. read/write or write/\allowbreak{}write access pairs to different memory addresses that share a cache line.
 On real hardware, these would produce {\tt \_XABORT\_CONFLICT} failures,
@@
 For now, Landslide (wrongly) lumps false sharing in among the ``spurious'' {\tt \_XABORT\_RETRY} failure reasons.
-
+%
 On HTM, even if the user wraps all such spurious failure reasons in a retry loop,
@@
 to remain faithful to HTM semantics and still abort even under {\tt -S}.
+
+Likewise, Landslide does not check for a transaction's memory footprint exceeding the CPU's cache capacity,
+which on real hardware would trigger a {\tt \_XABORT\_CAPACITY} abort.
+Landslide's memory tracking could simulate this check as well,
+perhaps with a configurable cache size,
+but the implementation would likely be theoretically uninteresting
+(all programs in the upcoming evaluation have trivial memory usage),
+and so is left to future work.
 

### 2018-07-29 thesis/tm.tex
fix all the busted retry-set tests since i fixed the last(?) bug in it

@@
 			{\tt htm2}
-			& 2,1 & \cpu{16.30}	& \ints{22}	& \cpu{19.82}	& \ints{17}	& \cpu{3.32}	& \ints{4} \\
-			& 2,2 & \cpu{69.80}	& \ints{1970}	& \cpu{38.71}	& \ints{906}	& \cpu{26.59}	& \ints{288} \\
+			& 2,1 & \cpu{16.30}	& \ints{22}	& \cpu{17.93}	& \ints{15}	& \cpu{3.32}	& \ints{4} \\
+			& 2,2 & \cpu{69.80}	& \ints{1970}	& \cpu{27.87}	& \ints{408}	& \cpu{26.59}	& \ints{288} \\
 			& 2,3 &\cpu{3510.00}	& \ints{104914}	& \cpu{}	& \ints{}	& \cpu{}	& \ints{} \\
 			%& 2,4 &	\ETAdag{...}	& \ETAdag{...}	\\
-			& 3,1 &\cpu{41.07}	& \ints{941}	& \cpu{29.38}	& \ints{553}	& \cpu{23.03}	& \ints{162} \\
+			& 3,1 &\cpu{41.07}	& \ints{941}	& \cpu{23.86}	& \ints{273}	& \cpu{23.03}	& \ints{162} \\
 			& 3,2 &	\ETAdag{9d 19h}	& \ETAdag{5058149}	\\ % add 10h ok
@@
 			{\tt counter}
-			& 2,1 & \cpu{5.40}	& \ints{10}	& \cpu{6.70}	& \ints{8}	& \cpu{3.22}	& \ints{4} \\
-			& 2,2 & \cpu{10.91}	& \ints{190}	& \cpu{14.91}	& \ints{288}	& \cpu{7.90}	& \ints{48} \\
-			& 2,3 & \cpu{134.21}	& \ints{3970}	& \cpu{379.04}	& \ints{11512}	& \cpu{35.00}	& \ints{904} \\
+			& 2,1 & \cpu{5.40}	& \ints{10}	& \cpu{7.00}	& \ints{8}	& \cpu{3.22}	& \ints{4} \\
+			& 2,2 & \cpu{10.91}	& \ints{190}	& \cpu{10.57}	& \ints{122}	& \cpu{7.90}	& \ints{48} \\
+			& 2,3 & \cpu{134.21}	& \ints{3970}	& \cpu{85.29}	& \ints{2376}	& \cpu{35.00}	& \ints{904} \\
 			& 2,4 & \cpu{3009.40}	& \ints{86950}	& \cpu{}	& \ints{}	& \cpu{}	& \ints{} \\
-			& 3,1 & \cpu{8.30}	& \ints{120}	& \cpu{8.90}	& \ints{74}	& \cpu{7.55}	& \ints{40} \\
+			& 3,1 & \cpu{8.30}	& \ints{120}	& \cpu{9.40}	& \ints{74}	& \cpu{7.55}	& \ints{40} \\
 			& 3,2 & \cpu{2038.90}	& \ints{60606}	& \cpu{}	& \ints{}	& \cpu{}	& \ints{} \\
-			& 4,1 & \cpu{101.95}	& \ints{3006}	& \cpu{99.22}	& \ints{2854}	& \cpu{32.95}	& \ints{848} \\
+			& 4,1 & \cpu{101.95}	& \ints{3006}	& \cpu{82.16}	& \ints{2148}	& \cpu{32.95}	& \ints{848} \\
 			\hline
 			{\tt swap}
-			& 2,1 & \cpu{38.79}	& \ints{99}	& \cpu{35.80}	& \ints{84}	& \cpu{3.25}	& \ints{4} \\
+			& 2,1 & \cpu{38.79}	& \ints{99}	& \cpu{37.33}	& \ints{59}	& \cpu{3.25}	& \ints{4} \\
 			& 2,2 & \cpu{10299.27}	& \ints{233396}	& \cpu{}	& \ints{}	& \cpu{}	& \ints{} \\
-			& 3,1 & \ETAdag{todo}	& \ETAdag{todo}	& \cpu{}	& \ints{}	& \cpu{}	& \ints{} \\
+			& 3,1 & \cpu{2359.85}	& \ints{59850}	& \cpu{}	& \ints{}	& \cpu{}	& \ints{} \\
 			\hline
 			{\tt fig63}
-			& 2,1 & \cpu{6.66}	& \ints{18}	& \cpu{6.50}	& \ints{6}	& \cpu{3.10}	& \ints{1} \\
-			& 2,2 & \cpu{11.52}	& \ints{244}	& \cpu{10.74}	& \ints{192}	& \cpu{3.12}	& \ints{1} \\
-			& 2,3 & \cpu{107.04}	& \ints{4428}	& \cpu{146.96}	& \ints{6151}	& \cpu{3.13}	& \ints{1} \\
-			& 3,1 & \cpu{13.68}	& \ints{296}	& \cpu{7.29}	& \ints{21}	& \cpu{3.26}	& \ints{1} \\
+			& 2,1 & \cpu{6.66}	& \ints{18}	& \cpu{6.98}	& \ints{6}	& \cpu{3.10}	& \ints{1} \\
+			& 2,2 & \cpu{11.52}	& \ints{244}	& \cpu{8.75}	& \ints{76}	& \cpu{3.12}	& \ints{1} \\
+			& 2,3 & \cpu{107.04}	& \ints{4428}	& \cpu{32.26}	& \ints{1060}	& \cpu{3.13}	& \ints{1} \\
+			& 3,1 & \cpu{13.68}	& \ints{296}	& \cpu{7.55}	& \ints{21}	& \cpu{3.26}	& \ints{1} \\
 			& 3,2 & \ETAdag{todo}	& \ETAdag{todo}	& \cpu{}	& \ints{}	& \cpu{3.35}	& \ints{1} \\
-			& 4,1 & \cpu{220.27}	& \ints{7382}	& \cpu{11.40}	& \ints{157}	& \cpu{3.27}	& \ints{1} \\
+			& 4,1 & \cpu{220.27}	& \ints{7382}	& \cpu{11.13}	& \ints{130}	& \cpu{3.27}	& \ints{1} \\
 			\hline
 			{\tt avl\_insert}
-			& 2,1 & \cpu{491.98}	& \ints{15056}	& \cpu{220.53}	& \ints{6332}	& \cpu{98.77}	& \ints{2776} \\
+			& 2,1 & \cpu{491.98}	& \ints{15056}	& \cpu{228.23}	& \ints{6296}	& \cpu{98.77}	& \ints{2776} \\
 			% (2,2) - bug found on regular w/ 127 ints, on retry sets w/ 201 ints, and codes w/ 45 ints
@@
 			{\tt avl\_fixed}
-			& 2,1 & \cpu{572.49} 	& \ints{20420} 	& \cpu{271.11}	& \ints{9739}	& \cpu{99.16}	& \ints{2776} \\
-			& 2,2 & \ETAdag{133y}	& \ETAdag{51484570} 	\\ % added 10hr but dosent matter
+			& 2,1 & \cpu{572.49} 	& \ints{20420} 	& \cpu{284.37}	& \ints{9685}	& \cpu{99.16}	& \ints{2776} \\
+			& 2,2 & \ETAdag{133y 2d}	& \ETAdag{51484570} 	\\ % added 10hr but dosent matter
 			% [JOB 19] progress: 408805/51484570 brs (0.794034%), ETA 132y 356d 8h 55m 16s (elapsed 9h 59m 19s)
-			& 3,1 & \ETAdag{todo}	& \ETAdag{todo}	\\
-
+			& 3,1 & \ETAdag{30y 320d}	& \ETAdag{93493027}	\\
+			% [JOB 18] progress: 492976/93493027 brs (0.011233%), ETA 30y 319d 14h 48m 32s (elapsed 9h 59m 35s)
 			\hline
@@
 		Run with {\tt -M -X}, plus any additional reduction options listed.
-		``Retry aborts only'' always tests every abort path without distinguishing among failure reasons;
-		``Retry sets'' skips equivalent retry abort reorderings (\sect{\ref{sec:tm-retrysets}});
+		``Retry aborts only'' always tests every abort path and does not distinguish among failure reasons,
+		as the baseline;
+		``Retry sets'' skips equivalent success path and/or retry abort reorderings (\sect{\ref{sec:tm-retrysets}});
 		``Abort codes'' suppresses retries and dynamically detects when to inject conflict aborts and so on

### 2018-08-01 thesis/tm.tex
rerun buncha tm verifs bc of dpor-preferred-tid fix; start writing verifs section, interpreting verif guarantee

@@
 implemented since then:
-the sleep sets optimization for DPOR (\sect{\ref{sec:landslide-explore}}, commit 0447666),
+the sleep sets optimization for DPOR (\sect{\ref{sec:landslide-sleepsets}}, commit 0447666),
 the {\tt thrlib\_function} directive
@@
 			& 9,9	&\cpu{36.64}&\wtm{9.78}&\ints{2}&\cpu{*4.75}&\wtm{*4.75}&\ints{2}& \ETA{inf} \\
+			\hline \multicolumn{4}{c}{everything above this line needs rerun} \\
 	\end{tabular}
@@
 		confers a sense of the exponential explosion.
+		{\bf RERUN NOTE: this entire table needs to be rerun on account of dpor-preferred-tid}
 	}
-	\label{tab:buges}
+	\label{tab:tm-buges}
 \end{table*}
 
-Table~\ref{tab:buges}
+Table~\ref{tab:tm-buges}
 presents the bug-finding results.
@@
 
-\begin{figure}[t]
+% TODO check figure placement
+\begin{figure}[h]
 	\begin{center}
@@
 	\end{center}
-	% TODO describe
 	\caption{Code from {\tt spinlock-rtm.c}, modified only to remove unrelated logic for brevity,
@@
 does not necessarily correlate with better performance in terms of CPU-time.
-Comparing Table~\ref{tab:buges}'s trends
+Comparing Table~\ref{tab:tm-buges}'s trends
 to the break-even point in Quicksand's evaluation (\sect{\ref{sec:quicksand-eval}}),
@@
 
-Landslide proved the following tests correct.
-With no bugs to find, comparing the different testing modes against each other is meaningless;
-I simply present their state space sizes and runtime
-(measured under {\tt -M})
-in Table~\ref{tab:verifs}.
-
-% TODO
-% discuss the diff reducion approaches
-
-% Please note that retry sets are a theoretically-founded reduction that should apply to all programs,
-% while distinguishing abort codes to suppress retries is potentially unsound,
-% depending on the user's ability to manually verify the presence (and correctness) of retry loops in the code.
-
-\begin{table}[h]
+For the test cases with no bugs found, I sought to provide Landslide's verification guarantee
+% TODO better section reference
+(\sect{\ref{sec:quicksand-id}})
+for up to as many threads and test iterations as possible under a reasonable time limit.
+The results, obtained using the same {\tt -X -M} configuration options as
+in the previous section,
+%were used in Table~\ref{tab:tm-buges},
+are shown in Table~\ref{tab:tm-verifs} in the ``Baseline DPOR'' column.
+%Of course, because stateless model checking is a perpetual existence
+%of staring up the sheer cliff face that is the exponential curve,
+%many tests were only able to achieve verification for only modest values of $(K, N)$
+%before becoming intractable.
+%After a timeout of 10 wall-clock hours,
+%I measured these tests'
+%estimated remaining time and state space size and list those instead,
+%\ETAdag{emphasized thusly}.
+For test configurations which could not be verified
+within 10 wall-clock hours,
+I report their estimated state space size and runtime measured after that timeout instead,
+\ETAdag{typeset thusly}.
+
+\begin{table}[p]
 	\begin{center}
@@
 		\begin{tabular}{cc||r|r||r|r||r|r}
-			& & \multicolumn{2}{c||}{\bf Retry aborts only}
+			& & \multicolumn{2}{c||}{\bf Baseline DPOR}
 			  & \multicolumn{2}{c||}{\bf Retry sets ({\tt -R})}
-			  & \multicolumn{2}{c}{\bf Abort codes ({\tt -A -S})} \\
+			  & \multicolumn{2}{c}{\bf STM ({\tt -A -S})} \\
 			& & \cpu{\bf cpu (s)} & \ints{\bf SS size}
@@
 			{\tt htm2}
-			& 2,1 & \cpu{16.30}	& \ints{22}	& \cpu{17.93}	& \ints{15}	& \cpu{3.32}	& \ints{4} \\
-			& 2,2 & \cpu{69.80}	& \ints{1970}	& \cpu{27.87}	& \ints{408}	& \cpu{26.59}	& \ints{288} \\
-			& 2,3 &\cpu{3510.00}	& \ints{104914}	& \cpu{}	& \ints{}	& \cpu{}	& \ints{} \\
-			%& 2,4 &	\ETAdag{...}	& \ETAdag{...}	\\
-			& 3,1 &\cpu{41.07}	& \ints{941}	& \cpu{23.86}	& \ints{273}	& \cpu{23.03}	& \ints{162} \\
-			& 3,2 &	\ETAdag{9d 19h}	& \ETAdag{5058149}	\\ % add 10h ok
-			% [JOB 4] progress: 923822/5058149 brs (18.264030%), ETA 9d 8h 30m 38s (elapsed 9h 59m 55s)
-			& 4,1 & \cpu{12699.65}	& \ints{344240}	& \cpu{}	& \ints{}	& \cpu{}	& \ints{} \\
+			& 2,1 & \cpu{19.82}	& \ints{22}	& \cpu{19.01}	& \ints{15}	& \cpu{3.45}	& \ints{4} \\
+			& 2,2 &\cpu{73.02}	& \ints{1970}	& \cpu{26.96}	& \ints{356}	& \cpu{27.38}	& \ints{288} \\
+			& 2,3 &\cpu{qqq}	& \ints{qqq}	& \cpu{qqq}	& \ints{qqq}	& \cpu{910.80}	& \ints{22409} \\
+			& 2,4 & \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq} \\
+			& 3,1 & \cpu{40.74}	& \ints{774}	& \cpu{24.27}	& \ints{224}	& \cpu{23.26}	& \ints{140} \\
+			& 3,2 & \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq} \\
+			& 4,1 & \cpu{8202.71}	& \ints{212146}	& \cpu{333.26}	& \ints{11973}	& \cpu{1794.06}	& \ints{44995} \\
+			% measured with !CONSIDER_ONLY_MOST_RECENT_DPOR_PREFERRED_TID
+			%& 2,2 & \cpu{72.40}	& \ints{1970}	& \cpu{27.28}	& \ints{356}	& \cpu{26.84}	& \ints{288} \\
+			%& 3,1 &\cpu{38.76}	& \ints{708}	& \cpu{23.84}	& \ints{224}	& \cpu{23.38}	& \ints{140} \\
+			%& 3,2 &\ETAdag{vvv}	& \ETAdag{vvv}	& \ETAdag{vvv}	& \ETAdag{vvv}	&\ETAdag{4d 21h}&\ETAdag{3456001}\\
+			%[JOB 4] progress: 709997/3456001 brs (20.543887%), ETA 4d 10h 36m 25s (elapsed 9h 59m 54s)
 			\hline
 			{\tt counter}
-			& 2,1 & \cpu{5.40}	& \ints{10}	& \cpu{7.00}	& \ints{8}	& \cpu{3.22}	& \ints{4} \\
-			& 2,2 & \cpu{10.91}	& \ints{190}	& \cpu{10.57}	& \ints{122}	& \cpu{7.90}	& \ints{48} \\
-			& 2,3 & \cpu{134.21}	& \ints{3970}	& \cpu{85.29}	& \ints{2376}	& \cpu{35.00}	& \ints{904} \\
-			& 2,4 & \cpu{3009.40}	& \ints{86950}	& \cpu{}	& \ints{}	& \cpu{}	& \ints{} \\
-			& 3,1 & \cpu{8.30}	& \ints{120}	& \cpu{9.40}	& \ints{74}	& \cpu{7.55}	& \ints{40} \\
-			& 3,2 & \cpu{2038.90}	& \ints{60606}	& \cpu{}	& \ints{}	& \cpu{}	& \ints{} \\
-			& 4,1 & \cpu{101.95}	& \ints{3006}	& \cpu{82.16}	& \ints{2148}	& \cpu{32.95}	& \ints{848} \\
+			& 2,1 & \cpu{qqq}	& \ints{qqq}	& \cpu{qqq}	& \ints{qqq}	& \cpu{qqq}	& \ints{qqq} \\
+			& 2,2 & \cpu{12.81}	& \ints{190}	& \cpu{10.58}	& \ints{102}	& \cpu{8.19}	& \ints{48} \\
+			& 2,3 & \cpu{144.83}	& \ints{3970}	& \cpu{65.66}	& \ints{1558}	& \cpu{38.18}	& \ints{904} \\
+			& 2,4 & \cpu{qqq}	& \ints{qqq}	& \cpu{qqq}	& \ints{qqq}	& \cpu{qqq}	& \ints{qqq} \\
+			& 3,1 & \cpu{10.57}	& \ints{120}	& \cpu{92.30}	& \ints{64}	& \cpu{7.88}	& \ints{40} \\
+			& 3,2 & \cpu{qqq}	& \ints{qqq}	& \cpu{qqq}	& \ints{qqq}	& \cpu{qqq}	& \ints{qqq} \\
+			& 4,1 & \cpu{121.96}	& \ints{3006}	& \cpu{61.85}	& \ints{1296}	& \cpu{39.23}	& \ints{848} \\
 			\hline
 			{\tt swap}
-			& 2,1 & \cpu{38.79}	& \ints{99}	& \cpu{37.33}	& \ints{59}	& \cpu{3.25}	& \ints{4} \\
-			& 2,2 & \cpu{10299.27}	& \ints{233396}	& \cpu{}	& \ints{}	& \cpu{}	& \ints{} \\
-			& 3,1 & \cpu{2359.85}	& \ints{59850}	& \cpu{}	& \ints{}	& \cpu{}	& \ints{} \\
+			& 2,1 & \cpu{38.76}	& \ints{99}	& \cpu{36.91}	& \ints{59}	& \cpu{3.29}	& \ints{4} \\
+			& 2,2 & \cpu{qqq}	& \ints{qqq}	& \cpu{qqq}	& \ints{qqq}	& \cpu{qqq}	& \ints{qqq} \\
+			& 3,1 & \cpu{2975.30}	& \ints{60912}	& \cpu{522.07}	& \ints{10706}	& \cpu{86.21}	& \ints{1014} \\
+			% measured with !CONSIDER_ONLY_MOST_RECENT_DPOR_PREFERRED_TID
+			%& 2,1 & \cpu{38.43}	& \ints{99}	& \cpu{36.12}	& \ints{59}	& \cpu{3.34}	& \ints{4} \\
+			%& 3,1 & \cpu{vvv}	& \ints{vvv}	& \cpu{730.94}	& \ints{17112}	& \cpu{74.53}	& \ints{1064} \\
 			\hline
 			{\tt fig63}
-			& 2,1 & \cpu{6.66}	& \ints{18}	& \cpu{6.98}	& \ints{6}	& \cpu{3.10}	& \ints{1} \\
-			& 2,2 & \cpu{11.52}	& \ints{244}	& \cpu{8.75}	& \ints{76}	& \cpu{3.12}	& \ints{1} \\
-			& 2,3 & \cpu{107.04}	& \ints{4428}	& \cpu{32.26}	& \ints{1060}	& \cpu{3.13}	& \ints{1} \\
-			& 3,1 & \cpu{13.68}	& \ints{296}	& \cpu{7.55}	& \ints{21}	& \cpu{3.26}	& \ints{1} \\
-			& 3,2 & \ETAdag{todo}	& \ETAdag{todo}	& \cpu{}	& \ints{}	& \cpu{3.35}	& \ints{1} \\
-			& 4,1 & \cpu{220.27}	& \ints{7382}	& \cpu{11.13}	& \ints{130}	& \cpu{3.27}	& \ints{1} \\
+			& 2,1 & \cpu{7.03}	& \ints{10}	& \cpu{6.96}	& \ints{6}	& \cpu{3.42}	& \ints{1} \\
+			& 2,2 & \cpu{9.32}	& \ints{108}	& \cpu{8.78}	& \ints{76}	& \cpu{3.43}	& \ints{1} \\
+			& 2,3 & \cpu{54.41}	& \ints{1934}	& \cpu{31.65}	& \ints{977}	& \cpu{3.27}	& \ints{1} \\
+			& 2,4 & \cpu{997.09} 	& \ints{36600} 	& \cpu{414.27}	& \ints{14512}	& \cpu{3.34}	& \ints{1} \\
+			% nb this 22 would be a 21 with !CONSIDER_ONLY... ..but, if it's the biggest loss, really n b d
+			& 3,1 & \cpu{10.40}	& \ints{148}	& \cpu{7.72}	& \ints{22}	& \cpu{3.33}	& \ints{1} \\
+			& 3,2 & \cpu{708.86}	& \ints{21642}	& \cpu{196.45}	& \ints{6467}	& \cpu{3.23}	& \ints{1} \\
+			& 4,1 & \cpu{106.21}	& \ints{3064}	& \cpu{11.29}	& \ints{130}	& \cpu{3.43}	& \ints{1} \\
+			& 4,2 & \cpu{qqq} 	& \ints{qqq} 	& \cpu{qqq}	& \ints{todo atm}	& \cpu{3.28}	& \ints{1} \\
+			% measured with !CONSIDER_ONLY_MOST_RECENT_DPOR_PREFERRED_TID
+			%& 2,2 & \cpu{9.40}	& \ints{108}	& \cpu{8.92}	& \ints{76}	& \cpu{3.37}	& \ints{1} \\
+			%& 3,1 & \cpu{10.30}	& \ints{132}	& \cpu{7.50}	& \ints{21}	& \cpu{3.64}	& \ints{1} \\
 			\hline
 			{\tt avl\_insert}
-			& 2,1 & \cpu{491.98}	& \ints{15056}	& \cpu{228.23}	& \ints{6296}	& \cpu{98.77}	& \ints{2776} \\
-			% (2,2) - bug found on regular w/ 127 ints, on retry sets w/ 201 ints, and codes w/ 45 ints
+			& 2,1 & \cpu{545.45}	& \ints{15052}	& \cpu{244.20}	& \ints{6288}	& \cpu{109.84}	& \ints{2774} \\
+			% (2,2) - bug found on regular w/ ?? ints, on retry sets w/ ?? ints, and codes w/ ?? ints
 			\hline
 			{\tt avl\_fixed}
-			& 2,1 & \cpu{572.49} 	& \ints{20420} 	& \cpu{284.37}	& \ints{9685}	& \cpu{99.16}	& \ints{2776} \\
-			& 2,2 & \ETAdag{133y 2d}	& \ETAdag{51484570} 	\\ % added 10hr but dosent matter
-			% [JOB 19] progress: 408805/51484570 brs (0.794034%), ETA 132y 356d 8h 55m 16s (elapsed 9h 59m 19s)
-			& 3,1 & \ETAdag{30y 320d}	& \ETAdag{93493027}	\\
-			% [JOB 18] progress: 492976/93493027 brs (0.011233%), ETA 30y 319d 14h 48m 32s (elapsed 9h 59m 35s)
+			& 2,1 & \cpu{642.73} 	& \ints{20416} 	& \cpu{293.70}	& \ints{9677}	& \cpu{108.69}	& \ints{2774} \\
+			& 2,2 & \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq} \\
+			& 3,1 & \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq} \\
 			\hline
 			{\tt map\_basic}
-			& 2,1 & \ETAdag{todo} & \ETAdag{todo} \\
+			& 2,1 & \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq} \\
 			\hline
 			{\tt map\_basicer}
-			& 2,1 & \cpu{36.17}& \ints{426}	\\ % TODO make sure this gets updated if you update the test
-			& 2,2 & \ETAdag{todo} & \ETAdag{todo} \\
-			& 3,1 & \ETAdag{todo} & \ETAdag{todo} \\
+			& 2,1 & \cpu{qqq} 	& \ints{qqq} 	& \cpu{qqq}	& \ints{qqq}	& \cpu{qqq}	& \ints{qqq} \\
+			& 2,2 & \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq} \\
+			& 3,1 & \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq} \\
+
+			%%%% nb. below is a fake version of the results, basically measured with abort set reduction,
+			%%%% after implementing abort set coalescing/merging, but not with either "dpor preferred tid"
+			%%%% or "trust thr join" optimizations.
+			%%%% this causes both (1) baseline to look worse than it should, and (2) abort sets to be
+			%%%% pretty inconsistent about whether they actually provide reduction, or accidentally inflate
+			%%%% the state space instead.
+			%%%% nevertheless, i'm keeping it around for posterity so you can compare, e.g., this fake baseline
+			%%%% against real baseline, to evaluate the impact of the non-abort-set dpor optimizations alone.
+			% {\tt htm2}
+			% & 2,1 & \cpu{16.30}	& \ints{22}	& \cpu{17.93}	& \ints{15}	& \cpu{3.32}	& \ints{4} \\
+			% & 2,2 & \cpu{69.80}	& \ints{1970}	& \cpu{27.87}	& \ints{408}	& \cpu{26.59}	& \ints{288} \\
+			% & 2,3 &\cpu{3510.00}	& \ints{104914}	& \cpu{}	& \ints{}	& \cpu{}	& \ints{} \\
+			% %& 2,4 &	\ETAdag{...}	& \ETAdag{...}	\\
+			% & 3,1 &\cpu{41.07}	& \ints{941}	& \cpu{23.86}	& \ints{273}	& \cpu{23.03}	& \ints{162} \\
+			% & 3,2 &	\ETAdag{9d 19h}	& \ETAdag{5058149}	\\ % add 10h ok
+			% % [JOB 4] progress: 923822/5058149 brs (18.264030%), ETA 9d 8h 30m 38s (elapsed 9h 59m 55s)
+			% & 4,1 & \cpu{12699.65}	& \ints{344240}	& \cpu{}	& \ints{}	& \cpu{}	& \ints{} \\
+			% \hline
+			% {\tt counter}
+			% & 2,1 & \cpu{5.40}	& \ints{10}	& \cpu{7.00}	& \ints{8}	& \cpu{3.22}	& \ints{4} \\
+			% & 2,2 & \cpu{10.91}	& \ints{190}	& \cpu{10.57}	& \ints{122}	& \cpu{7.90}	& \ints{48} \\
+			% & 2,3 & \cpu{134.21}	& \ints{3970}	& \cpu{85.29}	& \ints{2376}	& \cpu{35.00}	& \ints{904} \\
+			% & 2,4 & \cpu{3009.40}	& \ints{86950}	& \cpu{}	& \ints{}	& \cpu{}	& \ints{} \\
+			% & 3,1 & \cpu{8.30}	& \ints{120}	& \cpu{9.40}	& \ints{74}	& \cpu{7.55}	& \ints{40} \\
+			% & 3,2 & \cpu{2038.90}	& \ints{60606}	& \cpu{}	& \ints{60140}	& \cpu{486.56}	& \ints{14304} \\
+			% & 4,1 & \cpu{101.95}	& \ints{3006}	& \cpu{82.16}	& \ints{2148}	& \cpu{32.95}	& \ints{848} \\
+			% \hline
+			% {\tt swap}
+			% % these definitely have a different retry sets data point, run with some coalescing updates,
+			% % but i think before the dpor-preferred-tid search ordering fix
+			% % i think fig63 might have been updated this way as well? but not sure
+			% %& 2,1 & \cpu{38.79}	& \ints{99}	& \cpu{37.33}	& \ints{59}	& \cpu{3.25}	& \ints{4} \\
+			% %& 2,2 & \cpu{10299.27}	& \ints{233396}	& \cpu{433.01}	& \ints{11588}	& \cpu{382.28}	& \ints{9046} \\
+			% %& 3,1 & \cpu{2359.85}	& \ints{59850}	& \cpu{382.27}	& \ints{9712}	& \cpu{71.41}	& \ints{1014} \\
+			% & 2,1 & \cpu{38.79}	& \ints{99}	& \cpu{}	& \ints{}	& \cpu{3.25}	& \ints{4} \\
+			% & 2,2 & \cpu{10299.27}	& \ints{233396}	& \cpu{}	& \ints{}	& \cpu{382.28}	& \ints{9046} \\
+			% & 3,1 & \cpu{2359.85}	& \ints{59850}	& \cpu{}	& \ints{}	& \cpu{71.41}	& \ints{1014} \\
+			% \hline
+			% {\tt fig63}
+			% & 2,1 & \cpu{6.66}	& \ints{18}	& \cpu{6.97}	& \ints{6}	& \cpu{3.10}	& \ints{1} \\
+			% & 2,2 & \cpu{11.52}	& \ints{244}	& \cpu{8.75}	& \ints{76}	& \cpu{3.12}	& \ints{1} \\
+			% & 2,3 & \cpu{107.04}	& \ints{4428}	& \cpu{32.26}	& \ints{1060}	& \cpu{3.13}	& \ints{1} \\
+			% & 3,1 & \cpu{13.68}	& \ints{296}	& \cpu{7.55}	& \ints{21}	& \cpu{3.26}	& \ints{1} \\
+			% & 3,2 & \ETAdag{}	& \ETAdag{}	& \cpu{126.65}	& \ints{4042}	& \cpu{3.35}	& \ints{1} \\
+			% & 4,1 & \cpu{220.27}	& \ints{7382}	& \cpu{11.13}	& \ints{130}	& \cpu{3.27}	& \ints{1} \\
+			% \hline
+			% {\tt avl\_insert}
+			% & 2,1 & \cpu{491.98}	& \ints{15056}	& \cpu{228.23}	& \ints{6296}	& \cpu{98.77}	& \ints{2776} \\
+			% % (2,2) - bug found on regular w/ 127 ints, on retry sets w/ 201 ints, and codes w/ 45 ints
+			% \hline
+			% {\tt avl\_fixed}
+			% & 2,1 & \cpu{572.49} 	& \ints{20420} 	& \cpu{284.37}	& \ints{9685}	& \cpu{99.16}	& \ints{2776} \\
+			% & 2,2 & \ETAdag{133y}	& \ETAdag{51484570} % added 10hr but dosent matter
+			% % the order of magnitude of these time estimates is basically completely random
+			% % and fluctuates wildly from branch to branch
+			% % [JOB 19] progress: 408805/51484570 brs (0.794034%), ETA 132y 356d 8h 55m 16s (elapsed 9h 59m 19s)
+			% 	& \ETAdag{}	& \ETAdag{}
+			% 	& \ETAdag{1060y}	& \ETAdag{31302769} \\
+			% % for example if this one had timed out just 7 branches earlier, it woulda been 6y 246d instead
+			% % [JOB 19] progress: 321936/31302769 brs (1.028459%), ETA 1059y 260d 9h 16m 8s (elapsed 9h 59m 0s)
+			% & 3,1 & \ETAdag{30y 320d}	& \ETAdag{93493027}	\\
+			% % [JOB 18] progress: 492976/93493027 brs (0.011233%), ETA 30y 319d 14h 48m 32s (elapsed 9h 59m 35s)
 		\end{tabular}
@@
 		Run with {\tt -M -X}, plus any additional reduction options listed.
-		``Retry aborts only'' always tests every abort path and does not distinguish among failure reasons,
-		as the baseline;
+		``Baseline DPOR'' always tests every abort path, without distinguish among failure reasons
+		(i.e., injecting only {\tt \_XABORT\_RETRY}).
 		``Retry sets'' skips equivalent success path and/or retry abort reorderings (\sect{\ref{sec:tm-retrysets}});
-		``Abort codes'' suppresses retries and dynamically detects when to inject conflict aborts and so on
-		(\sect{\ref{sec:txn-abort-codes}}).
+		``STM'' suppresses retry aborts and dynamically detects when to inject conflict aborts and so on
+		(\sect{\ref{sec:txn-abort-codes}}),
+		ultimately matching STM's execution semantics.
 		State space estimates measured after a timeout of 10 hours.
 		}
-	\label{tab:verifs}
+	\label{tab:tm-verifs}
 \end{table}
 
-% TODO rephrase the outdated para from sigbovik here,
-% probably after rerunning expts
-
-\subsection{Reduction}
-
-While verifying the soundness of Figure~\ref{fig:htm-fixed}'s atomicity protocol is a positive result,
-the previous section's attempts at several larger real-world programs quickly suffered exponential explosion
+\subsubsection{Interpreting the verification guarantee}
+
+Landslide was able to verify most of these tests for a fair range of thread and iteration counts,
+often reaching up to 2 threads with 3 iterations each, 3 with 2 each, or 4 with 1.
+In the case of {\tt htm2}, for example,
+verifying up to $(K, N)$ represents a guarantee that,
+even repeating Figure~\ref{fig:htm-fixed}'s atomicity protocol $N$ times
+in any scheduling sequence or combination of transaction aborts,
+it is impossible for $K$ threads to violate the intended atomicity property
+(i.e., get 2 threads in the critical section simultaneously).
+
+It is difficult to discern from prior work a concrete standard
+for what values of $(K, N)$ constitute a ``good'' degree of verification.
+One recent paper \cite{tsopso},
+which likewise extended DPOR with a new dimension of concurrency (weak memory orderings)
+reported verifying programs with up to 10 concurrent events,
+presumably shared memory accesses.
+Another \cite{optimal-dpor}
+reported test cases with as many as 19 threads,
+although with what must be very little synchronization or memory conflicts,
+%although with what nature of synchronization or memory conflicts is not clear,
+as even their baseline DPOR checked only 4096 interleavings on that test.
+%
+Table~\ref{tab:tm-test-case-evence} confers a sense of the complexity of this evaluation's test suite,
+with the final column showing the approximate maximum number of preemption points
+reached among Landslide's verifications
+(i.e., $({\sf txn}+{\sf sync}+{\sf race})*K*N$ for the highest $(K,N)$ completed.
+%%%% on second thought, i'm not gonna do this -- any branch that xend shows up in, is gonna lose any of the pps coming
+%%%% from the failure path's mxes or data races, so xend will never be an *additional* pp
+%%%% and of course it will scare the committee
+% \footnote{The 1.5 multiplier on {\sf txn} represents the fact that for each transaction,
+% the {\tt xend} preemption point marking the end of its success path shows up about half the time,
+% the other half being failure injections.
+% % in reality it's usually less than half, because (1) drpps can show up in failure paths, which will get interleaved around,
+% % thereby giving the abort path more interleavings, and (2) under -A -S, you might get multiple codes for duplicate abort paths
+% % abort sets could theoretically push them above half but probably not in any test case here
+% }).
+
+\begin{table}[h]
+	\begin{center}
+		\begin{tabular}{c||c|c|c||c}
+			{\bf test} & {\bf \#txn} & {\bf \#sync} & {\bf \#race} & {\bf max events verified} \\
+			\hline
+			{\tt htm2}		& 1	& 2	& 4	& 42	\\ % 2,3
+			{\tt counter}		& 1	& 0	& 1	& 16	\\ % 2,4
+			{\tt swap}		& 1	& 4	& 8	& 52	\\ % 2,2
+			{\tt fig63}		& 1	& 0	& $K$	& 24	\\ % 3,2 k=3 -> 24; 4,2 -> 40 (not yet); 2,4 -> 24
+			{\tt avl\_insert}	& 1	& 2	& 7	& 20	\\ % 2,1 nb. only 5 drpps under -A -S but w/e
+			{\tt avl\_fixed}	& 1	& 2	& 9	& 24	\\ % 2,1 nb. again, only 5 under -A -S
+			% the extra 2 over basicer is from resize, which only 1 thread does, but K*N is 0 here anyway so vov
+			{\tt map\_basic}	& 1	& 4	& ?	& 0	\\ % lol
+			{\tt map\_basicer}	& 1	& 2	& ?	& ?	\\
+		\end{tabular}
+	\end{center}
+	\caption{Number of concurrency events per iteration of each test case.
+	Note that no test used any synchronization besides mutexes % ;
+	%while they do use the P2 thread API, %in the first place,
+	(the P2 thread API was annotated as trusted (\sect{\ref{sec:landslide-config-landslide}})
+	and so does not contribute to state space size).
+	Also note that ``\#race'' means the number of unique accesses identified as racy,
+	rather than racing pairs
+	(the other half of a pair might well be within a transaction, which cannot be preempted on).
+	}
+	\label{tab:tm-test-case-evence}
+\end{table}
+
+In the case of {\tt htm2},
+it is easy to look at the program with human intuition and judge that,
+because the protocol's only state is stored in a single boolean,
+with no use of unbounded-capacity data structures or 
+contention-dependent exponential backoff loops,
+it would be unimaginable that adding a 5th thread to the system
+could make any difference in correctness where 4 threads could not,
+or that a 4th repetition between 2 threads could make a difference where 3 could not,
+and ultimately that it must safely generalize to all $(K, N)$.
+% under certain invisible assumptions like the system runs with SC memory! (spoilers!!)
+For more complicated algorithms,
+which may involve complex conflict patterns such as tree rebalancing or map resizing,
+generalizing the verification is not so straightforward;
+other formal verification approaches aside,
+the user must ultimately be content with the probabilistic assurance that
+as $K$ and $N$ increase,
+the likelihood that a bug exists which requires more threads or iterations to expose grows ever lower
+% (in the author's fancy, it may as well drop exponentially in the same way as the state spaces grow)
+(none of the bugs in \sect{\ref{sec:tm-eval-bugs}} required any higher parameters than $(2,2)$ to expose).
+% and by analogy with ICB, none of the bugs in quicksand evaluation took more than a preemption bound of 3 to expose
+% but it's not clear how to state this "analogy" concisely in text
+
+Nevertheless, pushing $K$ and $N$ higher is obviously desirable,
+even if it means applying reductions that require human intuition to trust are sound.
+Moreover, much as verifying {\tt htm2}'s soundness is a positive result,
+the attempts at larger data structures quickly suffered exponential explosion
 for even small thread/iteration counts,
@@
 and in another failing to verify that a previously-found bug had been fixed.
-I explore three possible mitigation approaches.
+In the following subsections I explore three possible mitigation approaches.
+
+\subsubsection{Retry set reduction}
+\label{sec:tm-eval-retrysets}
+
+% TODO
+% discuss the diff reducion approaches
+
+% Please note that retry sets are a theoretically-founded reduction that should apply to all programs,
+% while distinguishing abort codes to suppress retries is potentially unsound,
+% depending on the user's ability to manually verify the presence (and correctness) of retry loops in the code.
+
+% TODO
+
+% discussion items
+
+% note that even programs which have fully conflicting txn/abort paths, under SSS-MC abort sets have nothing to reduce,
+% but once you split those abort paths apart by dr pps, abort sets become useful again
+
+(\sect{\ref{sec:tm-retrysets}}).
+
+% as for the sleep-set-alike optimization,
+% we see it would provide 25% reduction in the example (from thesprop)
+% however, evaluating its potential impact in general is tantamount to implementing it.
 
@@
 
-Table~\ref{tab:verifs2} shows the new resulting levels of verification
+Table~\ref{tab:tm-verifs2} shows the new resulting levels of verification
 Landslide reached before the same 10-hour timeout.
@@
 			& 3,2 & \\ %\cpu{0} & \ints{0} \\
+			\hline \multicolumn{4}{c}{everything above this line needs rerun} \\
 		\end{tabular}
@@
 	\end{center}
-	\caption{Continuation of Table~\ref{tab:verifs}.
+	\caption{Continuation of Table~\ref{tab:tm-verifs}.
 		Demonstrates abstraction reduction \cite{dbug-phdthesis}
@@
 		The reduced {\tt avl\_mutex} and {\tt map\_mutex} are, of course, no longer transactional.
+		{\bf RERUN NOTE: this entire table needs to be rerun on account of dpor-preferred-tid}
 		}
-	\label{tab:verifs2}
+	\label{tab:tm-verifs2}
 \end{table}
-
-\subsubsection{Retry set reduction}
-\label{sec:tm-eval-retrysets}
-
-% TODO
-
-(\sect{\ref{sec:tm-retrysets}}).
-
-% as for the sleep-set-alike optimization,
-% we see it would provide 25% reduction in the example (from thesprop)
-% however, evaluating its potential impact in general is tantamount to implementing it.
 

### 2018-08-01 thesis/tm.tex
buncha minor revisions tm

@@
 
-Support for TSX programs in Landslide is implemented in three parts, broadly speaking,
-the user interface plus one internal part corresponding to each proof above.
+Support for TSX programs in Landslide is implemented in four parts, broadly speaking:
+the user interface,
+plus one internal part corresponding to each proof above,
+plus a bonus optimization for pruning equivalent interleavings under the new concurrency model.
 
@@
 The interface is implemented in {\tt 410user/libhtm/htm.c},
-perhaps surprisingly, as totally empty functions:
+perhaps surprisingly, as totally empty functions;
 Landslide hooks these functions' addresses during instrumentation
@@
 		enables multiple xabort failure codes (\sect{\ref{sec:txn-failure}}).
-	\item {\tt -S} (for ``Stm'')
+	\item {\tt -S} (for ``Stm'' or ``Suppress retries'')
 		disables the {\tt \_XABORT\_RETRY} failure reason,
@@
 Also, the evaluation will later show
+(\sect{\ref{sec:tm-eval-retrysets}})
 that even programs with fully conflicting success/abort paths
 may still exhibit some equivalent thread interleavings of this pattern
-after their failure paths are split apart into smaller transitions by data-race preemption points
-(\sect{\ref{sec:tm-eval-retrysets}}).
+after their failure paths are split apart into smaller transitions by data-race preemption points.
 
@@
 thereby skipping over (i.e., pruning) reorderings of any other {\tt xbegin} results that would be independent.
+% TODO: make a real commit hash instead; put on master
 The {\tt landslide} repository's {\tt retry-sets} branch implements this feature%
@@
 the sleep sets optimization for DPOR (\sect{\ref{sec:landslide-sleepsets}}, commit 0447666),
-the {\tt thrlib\_function} directive
+the {\tt thrlib\_function} and {\tt TRUSTED\_THR\_JOIN} directives
 to mark internal thread library logic as trusted (\sect{\ref{sec:landslide-config-landslide}}, commit 64a02e4),
@@
 			& 4,1 & \cpu{106.21}	& \ints{3064}	& \cpu{11.29}	& \ints{130}	& \cpu{3.43}	& \ints{1} \\
-			& 4,2 & \cpu{qqq} 	& \ints{qqq} 	& \cpu{qqq}	& \ints{todo atm}	& \cpu{3.28}	& \ints{1} \\
+			& 4,2 & \ETAdag{qqq} 	& \ETAdag{qqq} 	& \ETAdag{qqq}	& \ETAdag{todo atm}	& \cpu{3.28}	& \ints{1} \\
 			% measured with !CONSIDER_ONLY_MOST_RECENT_DPOR_PREFERRED_TID
@@
 because the protocol's only state is stored in a single boolean,
-with no use of unbounded-capacity data structures or 
+with no unbounded-capacity data structures or
 contention-dependent exponential backoff loops,
@@
 % under certain invisible assumptions like the system runs with SC memory! (spoilers!!)
-For more complicated algorithms,
-which may involve complex conflict patterns such as tree rebalancing or map resizing,
-generalizing the verification is not so straightforward;
+Generalizing the verification is not so straightforward
+for more complicated algorithms,
+which may involve complex conflict patterns such as tree rebalancing or map resizing;
 other formal verification approaches aside,
 the user must ultimately be content with the probabilistic assurance that
-as $K$ and $N$ increase,
+as verified $K$ and $N$ increase,
 the likelihood that a bug exists which requires more threads or iterations to expose grows ever lower
 % (in the author's fancy, it may as well drop exponentially in the same way as the state spaces grow)
-(none of the bugs in \sect{\ref{sec:tm-eval-bugs}} required any higher parameters than $(2,2)$ to expose).
+(for example, none of the bugs in \sect{\ref{sec:tm-eval-bugs}} required any higher parameters than $(2,2)$ to expose).
 % and by analogy with ICB, none of the bugs in quicksand evaluation took more than a preemption bound of 3 to expose
@@
 
+\subsubsection{Abort code (STM) reduction}
+\label{sec:tm-eval-stm}
+
 \subsubsection{Abstraction reduction}
@@
 after correcting the former's atomicity bug (Figure~\ref{fig:avlbug}),
-reveals that every use of HTM followed exactly the same pattern,
+reveals that every use of HTM followed exactly the same pattern:
 running identical data structure logic in both the transactional and abort paths,
@@
 into separate tests,
-{\tt lock}, which checks the use of HTM guarantees mutual exclusion,
+{\tt lock}, which checks that the use of HTM guarantees mutual exclusion,
 and {\tt avl\_mutex}/{\tt map\_mutex},
@@
 even when the client's accesses are independent.
-Figure~\ref{fig:htm-lock} shows the core logic of these tests.
+Figure~\ref{fig:htm-lock} shows their core logic.
 Finally, I parameterized them over how the lock was implemented:
-{\tt spinlock} ``real-world'' code from \cite{spinlock-rtm-github},
-{\tt spin\_fixed} the same with the performance bug from \sect{\ref{sec:tm-eval-bugs}} fixed,
-and {\tt mutex} using Landslide-annotated P2 mutexes (as the AVL and map implementations do).
+a real-world {\tt spinlock} implementation from \cite{spinlock-rtm-github},
+{\tt spin\_fixed}, the same with the performance bug from \sect{\ref{sec:tm-eval-bugs}} fixed,
+and {\tt mutex}, using Landslide-annotated P2 mutexes (as the AVL and map implementations do).
 
@@
 
-Note two curiosities:
-firstly,
-the impact that fixing Figure~\ref{fig:spinlockbug}'s performance bug
-(changing a read+write to a read only)
-had on even the correctness tests:
-{\tt lock}({\tt spin\_fixed})'s state spaces were reduced by more than half compared to {\tt lock}({\tt spinlock}),
-on account of DPOR no longer needing to reorder the (now) read-read access pairs.
-Secondly, {\tt spinlock} and {\tt spin\_fixed} take longer to test per interleaving
-%(CPU time divided by state space size)
-than {\tt mutex}
-(roughly 9 interleavings per second for the former, 28 for the latter),
-because while {\tt mutex} abstracts away threads needing to wait their turn for the critical section
-behind an API Landslide understands,
-the spinlock's wait loop is open-coded, and Landslide must fall back on its costlier heuristic synchronization detection
-(\sect{\ref{sec:landslide-blocking}}).
-In this way (and also, of course, because {\tt mutex}'s state spaces are smaller overall),
-{\tt mutex} can be seen in turn as a further abstraction reduction of {\tt spinlock}.
-
-\begin{table}[h]
+\begin{table}[t]
 	\begin{center}
 		\footnotesize
-		%\begin{tabular}{cc}
+		\begin{tabular}{cc}
 		%\begin{tabular}{cc||r|r||r|r}
@@
 			& 9,9 & \cpu{3.79} & \ints{1}	\\
-			%\end{tabular}
-			%&
-			%\begin{tabular}{cc||r|r}
-			%& & \cpu{\bf cpu (s)} & \ints{\bf SS size} \\
-			%\bf test & \bf params & \em (or \ETAdag{\bf \em ETA}) & \em (or \ETAdag{\bf \em est.}) \\
+			\end{tabular}
+			&
+			\begin{tabular}{cc||r|r}
+			& & \cpu{\bf cpu (s)} & \ints{\bf SS size} \\
+			\bf test & \bf params & \em (or \ETAdag{\bf \em ETA}) & \em (or \ETAdag{\bf \em est.}) \\
 			\hline
@@
 			\hline \multicolumn{4}{c}{everything above this line needs rerun} \\
+			\\
+			\\
 		\end{tabular}
-		%\end{tabular}
+		\\
+		\\
+		(a) Verifying HTM locks alone.
+		&
+		(b) Verifying the lock's client code.
+		\end{tabular}
 	\end{center}
-	\caption{Continuation of Table~\ref{tab:tm-verifs}.
-		Demonstrates abstraction reduction \cite{dbug-phdthesis}
+	\caption{Continuation of Table~\ref{tab:tm-verifs},
+		demonstrating abstraction reduction \cite{dbug-phdthesis}
 		on the {\tt avl\_fixed} and {\tt map\_basic} tests
@@
 		(both lock implementations include a retry loop).
-		The reduced {\tt avl\_mutex} and {\tt map\_mutex} are, of course, no longer transactional.
+		The reduced {\tt avl\_mutex} and {\tt map\_mutex} are, of course, no longer transactional at all.
 		{\bf RERUN NOTE: this entire table needs to be rerun on account of dpor-preferred-tid}
@@
 \end{table}
+
+Note two curiosities:
+firstly,
+the impact that fixing Figure~\ref{fig:spinlockbug}'s performance bug
+(changing a read+write to a read only)
+had on even the correctness tests:
+{\tt lock}({\tt spin\_fixed})'s state spaces were reduced by more than half compared to {\tt lock}({\tt spinlock}),
+on account of DPOR no longer needing to reorder the (now) read-read access pairs.
+Secondly, {\tt spinlock} and {\tt spin\_fixed} take longer to test per interleaving
+%(CPU time divided by state space size)
+than {\tt mutex}
+(roughly 9 interleavings per second for the former, 28 for the latter),
+because while {\tt mutex} abstracts away threads needing to wait their turn for the critical section
+behind an API Landslide understands,
+the spinlock's wait loop is open-coded, and Landslide must fall back on its costlier heuristic synchronization detection
+(\sect{\ref{sec:landslide-blocking}}).
+In this way (and also, of course, because {\tt mutex}'s state spaces are smaller overall),
+{\tt mutex} can be seen in turn as a further abstraction reduction of {\tt spinlock}.
 

### 2018-08-01 thesis/landslide.tex
TRUSTED_THR_JOIN

@@
 		Users of this option should be careful to enable the proper {\tt thr\_create()} misbehave mode in their test case (\sect{\ref{sec:landslide-friendly-misbehave}}).
+	\item {\tt TRUSTED\_THR\_JOIN=0} - if set to 1, forces Landslide to add a happens-before edge (\sect{\ref{sec:landslide-dpor-hb}}) between the exiting of some thread $N$ and the end of any subsequent {\tt thr\_join}$(N)$ call,
+		even if that {\tt join} would not ordinarily block.
+		This is useful for state space reduction when testing threaded client code;
+		for example,
+		in the interleaving
+		TID1: {\tt x++; thr\_exit();},
+		TID2: {\tt thr\_join(1); print(x);},
+		%this {\tt join} will wait on no condition variables, just take a few uncontended mutexes, reap the thread,
+		%and go along on its way, so
+		DPOR, not automatically trusting {\tt join}'s behaviour,
+		will attempt to test the TID2, TID1 interleaving to reorder the accesses on {\tt x},
+		whereupon {\tt join} will block, forcing these interleavings to be equivalent.
+		% if you had tested them in the other order to begin with, dpor would detect this dependency,
+		% and would skip the other equivalent interleaving
+		This option allows DPOR to skip checking that {\tt join} behaves properly
+		and prune the second interleaving
+		by teaching it the expected blocking semantics. %to begin with.
+		Obviously, not for use when actually testing {\tt thr\_join} itself!
 \end{itemize}
@@
 \subsubsection{Happens-before}
+\label{sec:landslide-dpor-hb}
 

### 2018-08-01 thesis/tm.tex
few more tm runs

@@
 for further discussion of variance see \cite{sigbovik-htm}.
+The number of interleavings in each state space are, of course, deterministic and do not vary across runs.
 
@@
 			& 2,2 &\cpu{73.02}	& \ints{1970}	& \cpu{26.96}	& \ints{356}	& \cpu{27.38}	& \ints{288} \\
-			& 2,3 &\cpu{qqq}	& \ints{qqq}	& \cpu{qqq}	& \ints{qqq}	& \cpu{910.80}	& \ints{22409} \\
+			& 2,3 &\cpu{3923.32}	& \ints{104914}	& \cpu{230.71}	& \ints{7528}	& \cpu{910.80}	& \ints{22409} \\
 			& 2,4 & \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq} \\
@@
 			{\tt counter}
-			& 2,1 & \cpu{qqq}	& \ints{qqq}	& \cpu{qqq}	& \ints{qqq}	& \cpu{qqq}	& \ints{qqq} \\
+			& 2,1 & \cpu{6.90}	& \ints{10}	& \cpu{7.04}	& \ints{8}	& \cpu{3.35}	& \ints{4} \\
 			& 2,2 & \cpu{12.81}	& \ints{190}	& \cpu{10.58}	& \ints{102}	& \cpu{8.19}	& \ints{48} \\
 			& 2,3 & \cpu{144.83}	& \ints{3970}	& \cpu{65.66}	& \ints{1558}	& \cpu{38.18}	& \ints{904} \\
-			& 2,4 & \cpu{qqq}	& \ints{qqq}	& \cpu{qqq}	& \ints{qqq}	& \cpu{qqq}	& \ints{qqq} \\
+			& 2,4 & \cpu{3433.08}	& \ints{86950}	& \cpu{1134.62}	& \ints{25398}	& \cpu{752.51}	& \ints{19128} \\
 			& 3,1 & \cpu{10.57}	& \ints{120}	& \cpu{92.30}	& \ints{64}	& \cpu{7.88}	& \ints{40} \\
-			& 3,2 & \cpu{qqq}	& \ints{qqq}	& \cpu{qqq}	& \ints{qqq}	& \cpu{qqq}	& \ints{qqq} \\
+			& 3,2 & \cpu{2492.17}	& \ints{60606}	& \cpu{2040.98}	& \ints{44862}	& \cpu{565.97}	& \ints{14304} \\
 			& 4,1 & \cpu{121.96}	& \ints{3006}	& \cpu{61.85}	& \ints{1296}	& \cpu{39.23}	& \ints{848} \\
@@
 			& 2,1 & \cpu{38.76}	& \ints{99}	& \cpu{36.91}	& \ints{59}	& \cpu{3.29}	& \ints{4} \\
-			& 2,2 & \cpu{qqq}	& \ints{qqq}	& \cpu{qqq}	& \ints{qqq}	& \cpu{qqq}	& \ints{qqq} \\
+			& 2,2 & \cpu{qqq}	& \ints{qqq}	& \cpu{497.11}	& \ints{11427}	& \cpu{461.54}	& \ints{9046} \\
 			& 3,1 & \cpu{2975.30}	& \ints{60912}	& \cpu{522.07}	& \ints{10706}	& \cpu{86.21}	& \ints{1014} \\
@@
 			& 4,1 & \cpu{106.21}	& \ints{3064}	& \cpu{11.29}	& \ints{130}	& \cpu{3.43}	& \ints{1} \\
-			& 4,2 & \ETAdag{qqq} 	& \ETAdag{qqq} 	& \ETAdag{qqq}	& \ETAdag{todo atm}	& \cpu{3.28}	& \ints{1} \\
+			& 4,2 & \ETAdag{qqq} 	& \ETAdag{qqq} 	& \ETAdag{22d 17h}	& \ETAdag{3129034}	& \cpu{3.28}	& \ints{1} \\
+			%[JOB 1] progress: 827215/3129034 brs (26.436748%), ETA 22d 7h 18m 34s (elapsed 9h 59m 56s)
+
 			% measured with !CONSIDER_ONLY_MOST_RECENT_DPOR_PREFERRED_TID
@@
 reached among Landslide's verifications
-(i.e., $({\sf txn}+{\sf sync}+{\sf race})*K*N$ for the highest $(K,N)$ completed.
+(i.e., $({\sf txn}+{\sf sync}+{\sf race}) \times K \times N$ for the highest $K \times N$ completed.
 %%%% on second thought, i'm not gonna do this -- any branch that xend shows up in, is gonna lose any of the pps coming
@@
 	\begin{center}
+		\small
 		\begin{tabular}{c||c|c|c||c}
@@
 % (in the author's fancy, it may as well drop exponentially in the same way as the state spaces grow)
-(for example, none of the bugs in \sect{\ref{sec:tm-eval-bugs}} required any higher parameters than $(2,2)$ to expose).
+(for example, none of the bugs in \sect{\ref{sec:tm-eval-bugs}} required any higher parameters than (2,2) to expose).
 % and by analogy with ICB, none of the bugs in quicksand evaluation took more than a preemption bound of 3 to expose
@@
 \label{sec:tm-eval-stm}
+
+As an example, note how STM semantics reduced {\tt fig63}'s state space on all $(K,N)$ configurations to 1.
+Because its transactional paths are all mutually independent,
+DPOR identifies no need either to inject conflict aborts or to reorder threads.
+However, this skips the slow-path consistency assertion completely,
+which the programmer may have intended to run ``every so often''
 

### 2018-08-01 thesis/landslide.tex
and to prune
KEEP_RUNNING_DPORS_CHOSEN_TID

@@
 		This option allows DPOR to skip checking that {\tt join} behaves properly
-		and prune the second interleaving
+		and to prune the second interleaving
 		by teaching it the expected blocking semantics. %to begin with.
@@
 and instead identifies where the reduction should take place in subsequent iterations of DPOR.
+% KEEP_RUNNING_DPORS_CHOSEN_TID
+This approach also relies on the search ordering strategy ({\tt arbiter\_choose()})
+to prefer scheduling the thread previously chosen for reordering by DPOR,
+to force whatever its conflicting memory access was to happen before the other, preempted thread gets a chance to run.
 Further optimizations such as {\em source sets} and {\em wakeup trees},

### 2018-08-03 thesis/tm.tex
prose for abort set reduction; run smore tests

@@
 
-Up to three abort codes are considered depending on the testing options listed in the previous section.
+Up to four abort codes are considered depending on the testing options listed in the previous section.
 All abort codes are simultaneously added to both lists,
@@
 On HTM, even if the user wraps all such spurious failure reasons in a retry loop,
-false sharing should still produce a non-retryable abort,
+false sharing (i.e., disjoint memory accesses that share a cache line)
+should still produce a non-retryable abort,
 begetting a discrepancy with STM,
@@
 			%[JOB 38] progress: 3812/466977272 brs (0.000816%), ETA INF (elapsed 13m 50s)
+			\hline \multicolumn{9}{c}{everything above this line needs rerun for dpor-preferred tid} \\
 		\hline
@@
 		%({\tt spinlock})
-			& 2,1	&\cpu{8.17}&\wtm{2.65}&\ints{2}&\cpu{*2.39}&\wtm{*2.39}&\ints{2}& \ETA{4} \\
+			& 2,1	&\cpu{12.33}&\wtm{3.69}&\ints{2}&\cpu{*3.32}&\wtm{*3.32}&\ints{2}& \ETA{4} \\
 		(perf)
-			& 9,9	&\cpu{36.64}&\wtm{9.78}&\ints{2}&\cpu{*4.75}&\wtm{*4.75}&\ints{2}& \ETA{inf} \\
-			\hline \multicolumn{4}{c}{everything above this line needs rerun} \\
+			& 9,9	&\cpu{20.49}&\wtm{5.74}&\ints{2}&\cpu{*4.78}&\wtm{*4.78}&\ints{2}& \ETA{inf} \\
 	\end{tabular}
@@
 	alternately
-	as a segfault (for test parameters (2,2) and (3,1)) % TODO double check
-	% landslide-trace-1521092404.23772.html
+	as a segfault (for test parameters (2,2) and (3,1))
 	and
+	% landslide-trace-1533274266.417257.html
 	as a consistency-check assertion failure (for test parameters (2,3)).
@@
 	As a matter of full disclosure,
-	while editing this test to make it Landslide-friendly,
 	I noted that the loop does not affect the test's possible behaviours,
 	only its {\em likely} ones,
-	and so removed it to keep the state space size manageable.
-	One could then argue that manual inspection is at least partly to credit for this result.
-	% TODO justify this
-	However, adding the loop back in to dispel any doubt,
-	% converting it to a {\tt yield()} loop % TODO: necessary?
-	Landslide still finds the bug,
-	albeit slightly slower,
-	in % TODO how much time
+	and so removed it to make the test more Landslide-friendly.
+	%keep the state space size manageable.
+	%Lest any reader think I hacked the test after already knowing about the bug,
+	To dispel any doubt about bias or test hacking,
+	I confirmed that Landslide still finds the bug with the spin loop unmodified,
+	on (3,1) in the same 53 interleavings,
+	although it suffers resource exhaustion
+	% from dealing with the loop
+	on parameters of (2,2) or greater.
+	% specifically, it runs out of file descriptors, since on bochs you need 1 fd per PP,
+	% and the branch just gets 1000+ PPs deep
 	\item
@@
 	As another matter of full disclosure,
-	I noticed this bug through manual inspection,
+	I noticed this bug through manual inspection
 	while adapting the spinlock's client code to be Landslide-friendly,
@@
 			\bf test & \bf params
-			  & \em (or \ETAdag{\bf \em ETA}) & \em (or \ETAdag{\bf \em est.})
-			  & \em (or \ETAdag{\bf \em ETA}) & \em (or \ETAdag{\bf \em est.})
-			  & \em (or \ETAdag{\bf \em ETA}) & \em (or \ETAdag{\bf \em est.}) \\
+			  & \em (or \ETAdag{\bf \em est.}) & \em (or \ETAdag{\bf \em est.})
+			  & \em (or \ETAdag{\bf \em est.}) & \em (or \ETAdag{\bf \em est.})
+			  & \em (or \ETAdag{\bf \em est.}) & \em (or \ETAdag{\bf \em est.}) \\
 			\hline
@@
 			& 2,3 &\cpu{3923.32}	& \ints{104914}	& \cpu{230.71}	& \ints{7528}	& \cpu{910.80}	& \ints{22409} \\
-			& 2,4 & \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq} \\
+			& 2,4 & \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{20h 50m}	& \ETAdag{1814844} \\
+			% [JOB 4] progress: 805070/1814844 brs (44.360267%), ETA 10h 49m 33s (elapsed 9h 59m 54s)
 			& 3,1 & \cpu{40.74}	& \ints{774}	& \cpu{24.27}	& \ints{224}	& \cpu{23.26}	& \ints{140} \\
@@
 			& 2,1 & \cpu{38.76}	& \ints{99}	& \cpu{36.91}	& \ints{59}	& \cpu{3.29}	& \ints{4} \\
-			& 2,2 & \cpu{qqq}	& \ints{qqq}	& \cpu{497.11}	& \ints{11427}	& \cpu{461.54}	& \ints{9046} \\
+			& 2,2 & \cpu{12009.77}	& \ints{233396}	& \cpu{497.11}	& \ints{11427}	& \cpu{461.54}	& \ints{9046} \\
 			& 3,1 & \cpu{2975.30}	& \ints{60912}	& \cpu{522.07}	& \ints{10706}	& \cpu{86.21}	& \ints{1014} \\
@@
 			& 3,2 & \cpu{708.86}	& \ints{21642}	& \cpu{196.45}	& \ints{6467}	& \cpu{3.23}	& \ints{1} \\
+			% 3,3 with -R was tried for about 90 min, and stopped with ETA of 2367715 / 2d22h
+			% without -R would of course be bigger still
 			& 4,1 & \cpu{106.21}	& \ints{3064}	& \cpu{11.29}	& \ints{130}	& \cpu{3.43}	& \ints{1} \\
@@
 			{\tt map\_basicer}	& 1	& 2	& ?	& ?	\\
+
+			{\tt lock}({\tt spinlock})	& 1	& 2	& 4	& 28	\\ % 2,2
+			{\tt lock}({\tt spin\_fixed})	& 1	& 2	& 4	& 28	\\ % 2,2 or 4,1
+			{\tt lock}({\tt mutex})		& 1	& 2	& 2	& 40	\\ % 2,4
+			% not putting lock_fast because it like, doesn't have a "state space" per se
 		\end{tabular}
@@
 
+Secondly, the state spaces could be reduced
+simply by restricting the concurrency model
+to only a subset of nondeterministic {\tt xbegin} outcomes possible under HTM.
+Concretely speaking,
+the {\tt -X -A -S} combination of Landslide options
+suppresses retry aborts
+(\sect{\ref{sec:txn-abort-codes}}),
+which must be checked at every transactional preemption point,
+replacing them with explicit and conflict aborts,
+which Landslide injects only after identifying memory conflicts through DPOR
+or encountering an {\tt xabort},
+respectively,
+ultimately simulating the semantics of STM rather than HTM.
+This can allow for state space reduction %result in smaller state spaces compared to the baseline
+when transactions happen to be non-conflicting,
+but more impactfully,
+conflict aborts can occur only {\em after} the other thread's conflicting access,
+so between a pair of transactions
+% only the first two during execution, not true for subsequent pairs, if they are in conflict with previous pairs at all
+only the success, success and success, abort sequences need be tested;
+abort, success and abort, abort may (in fact, must) be skipped.
+The final column in Table~\ref{tab:tm-verifs} shows the result of STM semantics verification,
+which always results in at least 2x reduction compared to the baseline,
+although retry sets can make up some of the lost ground in some cases.
+
+However, suppressing retry aborts is not guaranteed to faithfully test all possible behaviours
+observable under HTM%
+\footnote{A user who wishes to distinguish conflict aborts, retry aborts, and so on during testing
+without glossing over any of HTM's peculiarities
+could supply {\tt -X -A} without {\tt -S};
+however, this could produce only state spaces at least as large as the baseline.}.
 As an example, note how STM semantics reduced {\tt fig63}'s state space on all $(K,N)$ configurations to 1.
@@
 DPOR identifies no need either to inject conflict aborts or to reorder threads.
-However, this skips the slow-path consistency assertion completely,
-which the programmer may have intended to run ``every so often''
+However, this skips the slow-path consistency assertion completely.
+If the programmer had intended it to run ``every so often'' at the whim of the timer interrupt,
+applying this reduction would be unsound.
+Also note the state space size of 4 for many $(2,1)$ test configurations,
+corresponding exactly to the aforementioned success, success and success, abort sequences
+(times two ways to interleave the two threads).
+Because of the scheduling dependency for conflict aborts,
+Landslide cannot recognize the failure path's data races
+without a third freely-reorderable iteration;
+STM mode must be run with $K \times N \ge 3$ to meaningfully test conflicts between failure and success paths at all.
+
+On the other hand, some programs may clearly annotate their intention
+for abort paths to be executed only in case of actual memory conflicts.
+{\tt swap}, {\tt avl\_insert}, and {\tt map\_basic} abstract their {\tt \_xbegin()} calls
+behind an interface which can be implemented either with or without retry loops,
+while {\tt lock}({\tt spinlock}) and {\tt lock}({\tt mutex}) implement the retry loop directly.
+In these cases, the user can check this reduction's soundness by visual inspection.
+Future work could automatically identify a program's retry loops
+and give up on {\tt \_XABORT\_RETRY} by switching to STM mode on-the-fly,
+much like Landslide's heuristic synchronization detection does for yield loops (\sect{\ref{sec:landslide-blocking}})%
+\footnote{Landslide's current implementation gets stuck in infinitely deep interleavings
+when it encounters a retry loop,
+so for now the user must inspect the test case to determine which testing mode to use.
+}.
 
@@
 		\begin{tabular}{cc||r|r}
+			& & \multicolumn{2}{c}{\bf STM ({\tt -A -S})} \\
 			& & \cpu{\bf cpu (s)} & \ints{\bf SS size} \\
-			\bf test & \bf params & \em (or \ETAdag{\bf \em ETA}) & \em (or \ETAdag{\bf \em est.}) \\
+			\bf test & \bf params & \em (or \ETAdag{\bf \em est.}) & \em (or \ETAdag{\bf \em est.}) \\
 			\hline
@@
 			{\tt lock}
-			& 2,1 & \cpu{2.49} & \ints{4} \\ % certified causality ok
+			& 2,1 & \cpu{3.19} & \ints{4} \\
 			({\tt spinlock})
-			& 2,2 & \cpu{174.26} & \ints{1702} \\ % certified causality ok
+			& 2,2 & \cpu{184.51} & \ints{1702} \\
 			% if lock(spin_fixed) will time out on 2,3 and 3,2, this is guaranteed to aswell
 			% & 2,3 & \\ % \cpu{0} & \ints{0} \\
-			& 3,1 & \cpu{26.97} & \ints{246} \\ % certified causality ok
+			& 3,1 & \cpu{33.29} & \ints{246} \\
 			% & 3,2 & \\ % \cpu{0} & \ints{0} \\
-			% new spinlock test resulce - htm_spinlock
-			% (2,1) without isfree fix
-			%%%% total time elapsed: 9s
-			%%%% [JOB 4] COMPLETE (41 interleavings tested; 3s elapsed)
-			%%%% total CPU time consumed: 14s (14231661 usecs) (core saturation: 37%)
-			% (2,2) without isfree fix
-			%%%% total time elapsed: 1h 22m 20s
-			%%%% [JOB 4] COMPLETE (38544 interleavings tested; 1h 22m 13s elapsed)
-			%%%% total CPU time consumed: 1h 22m 24s (4944373889 usecs) (core saturation: 25%)
-			% (3,1) without isfree fix
-			% in progress but smth like
-			% [JOB 4] progress: 39726/83689 brs (47.468314%), ETA 3h 27m 10s (elapsed 1h 30m 21s)
 			\hline
 			{\tt lock}
-			& 2,1 & \cpu{2.47} & \ints{4} \\ % certified causality ok
+			& 2,1 & \cpu{qqq} & \ints{4} \\
 			({\tt spin\_fixed})
-			& 2,2 & \cpu{94.64} & \ints{998} \\ % certified causality ok
-			& 2,3 & \ETAdag{28h 6m} & \ETAdag{486966} \\ % certified causality ok, add 10h ok
-			% [JOB 4] progress: 234307/486966 brs (48.115630%), ETA 18h 5m 37s (elapsed 9h 59m 50s)
-			& 3,1 & \cpu{21.68} & \ints{186} \\ % certified causality ok
+			& 2,2 & \cpu{104.77} & \ints{998} \\
+			& 2,3 & \ETAdag{qqq} & \ETAdag{qqq} \\
+			& 3,1 & \cpu{26.96} & \ints{186} \\
+			\hline \multicolumn{4}{c}{everything below this line needs rerun} \\
 			& 3,2 & \ETAdag{8y 153d} & \ETAdag{1590416} \\ % caus ok
@@
 			& 4,1 & \cpu{493.38} & \ints{15064}	\\ % certified causality ok
+			\hline \multicolumn{4}{c}{everything above this line needs rerun} \\
 			%%%%%%%%%%% htm_mutex
@@
 			%%%% from a misplaced assert on hm->held
+			%%%% (nb. also before dpor-preferred-tid, although that doesn't seem to affect ss size)
 			% % 2,1
@@
 			{\tt lock\_fast}
-			& 2,1 & \cpu{2.42} & \ints{1}	\\
+			& 2,1 & \cpu{3.36} & \ints{1}	\\
 			({\tt spin\_fixed})
-			& 9,9 & \cpu{3.69} & \ints{1}	\\
+			& 9,9 & \cpu{4.91} & \ints{1}	\\
 			\hline
 			{\tt lock\_fast}
-			& 2,1 & \cpu{2.35} & \ints{1}	\\
+			& 2,1 & \cpu{3.40} & \ints{1}	\\
 			({\tt mutex})
-			& 9,9 & \cpu{3.79} & \ints{1}	\\
+			& 9,9 & \cpu{4.71} & \ints{1}	\\
 			\end{tabular}
@@
 			\begin{tabular}{cc||r|r}
+			& & \multicolumn{2}{c}{\bf Non-transactional} \\
 			& & \cpu{\bf cpu (s)} & \ints{\bf SS size} \\
@@
 		by verifying HTM mutex implementations separately.
-		Run with {\tt landslide -M -X -A -S}, i.e.,
-		conflict and explicit aborts tested separately,
-		and retry aborts suppressed
-		(both lock implementations include a retry loop).
-		The reduced {\tt avl\_mutex} and {\tt map\_mutex} are, of course, no longer transactional at all.
+		Tested with STM semantics, as both lock implementations include a retry loop.
+		%Run with {\tt landslide -M -X -A -S}, i.e.,
+		%conflict and explicit aborts tested separately,
+		%and retry aborts suppressed
+		%(both lock implementations include a retry loop).
+		%The reduced {\tt avl\_mutex} and {\tt map\_mutex} are, of course, no longer transactional at all.
 		{\bf RERUN NOTE: this entire table needs to be rerun on account of dpor-preferred-tid}

### 2018-08-03 thesis/tm.tex
describe retry sets with a nice figure

@@
 			& 3,1 & \cpu{40.74}	& \ints{774}	& \cpu{24.27}	& \ints{224}	& \cpu{23.26}	& \ints{140} \\
-			& 3,2 & \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq} \\
+			& 3,2 & \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{8d 8h}& \ETAdag{2562278} \\
+			% [JOB 4] progress: 713090/2562278 brs (27.830313%), ETA 7d 22h 27m 42s (elapsed 9h 59m 54s)
 			& 4,1 & \cpu{8202.71}	& \ints{212146}	& \cpu{333.26}	& \ints{11973}	& \cpu{1794.06}	& \ints{44995} \\
@@
 
-% TODO
-% discuss the diff reducion approaches
-
-% Please note that retry sets are a theoretically-founded reduction that should apply to all programs,
-% while distinguishing abort codes to suppress retries is potentially unsound,
-% depending on the user's ability to manually verify the presence (and correctness) of retry loops in the code.
-
-% TODO
-
-% discussion items
-
-% note that even programs which have fully conflicting txn/abort paths, under SSS-MC abort sets have nothing to reduce,
-% but once you split those abort paths apart by dr pps, abort sets become useful again
-
-(\sect{\ref{sec:tm-retrysets}}).
-
-% as for the sleep-set-alike optimization,
-% we see it would provide 25% reduction in the example (from thesprop)
-% however, evaluating its potential impact in general is tantamount to implementing it.
+Firstly, the middle column of Table~\ref{tab:tm-verifs}
+shows the impact of Landslide's experimental retry set reduction (\sect{\ref{sec:tm-retrysets}}).
+The biggest win is apparent in {\tt fig63},
+the test contrived to induce as much retry set reduction as possible
+using transactional paths that conflict only with abort paths and not with each other, and vice versa.
+Retry set reduction provides
+In {\tt fig63}(2,1), corresponding to Figure~\ref{fig:retry-set-example},
+retry-set-enabled DPOR correctly prunes down to the optimal 6 interleavings,
+while the baseline treats it identically to the fully-conflicting {\tt counter}(2,1).
+Figure~\ref{fig:retry-reduction} visualizes the difference between the state spaces explored by the two approaches.
+
+\begin{figure}[h]
+	\begin{center}
+		\includegraphics[width=\textwidth]{retry-sets.pdf}
+	\end{center}
+	\caption{Visualization of retry set state space reduction.
+	On both {\tt counter}(2,1) and {\tt fig63}(2,1), baseline DPOR tests % oops
+	all 10 interleavings pictured,
+	with the middle 2 arising from the data-race preemption point within the abort path.
+	On {\tt counter}, all transitions marked as memory conflicts conflict with each other;
+	on {\tt fig63}, only pairs of one {\tt xend} and one {\tt data race} each conflict.
+	}
+	\label{fig:retry-reduction}
+\end{figure}
+
+Perhaps surprisingly, retry sets also provide reduction even on tests with fully conflicting transactions
+(i.e., all tests besides {\tt fig63}).
+With just synchronization preemption points (including {\tt \_xbegin()} and {\tt \_xend()}),
+both the baseline and retry sets would explore exactly all 8 permutations of success and abort between the two threads.
+However, in the presence of data-race preemption points,
+even for example on {\tt counter}(2,1)
+(whose abort path is just 1 {\tt xadd},
+and whose optimal state space size should be 8 regardless of data-race preemption points),
+baseline DPOR tests reorderings of one thread's transaction both with the other's failure path,
+and with the other's data race therein.
+Retry sets on the other hand identify and skip that equivalence
+(technically speaking, retry set DPOR reorders with the data race first during depth-first search,
+then skips generating a retry set for reordering the full abort path).
+
+% --- counter-retry       2018-08-03 13:50:37.324021777 -0400
+% +++ counter-baseline    2018-08-03 13:50:53.059021559 -0400
+%  0--1@105c41 1-4@105c41 2-5@xbeg-delay 3-5@xbegin 4-5@xend 5-5@105c41 6-4@1001042 7-4@105c41 8-6@xbeg-delay 9-6@xbegin 10-6@xend 11-6@105c41 12-4@1001042 13-4@105c41 14-3@1028e9
+% #[EXPLORE]       dbg #2/tid5 retry set: {TID6 (1,0), TID5 (1,0)}
+% #[EXPLORE]       from #14/tid3, chose tid 6 (xabort injection), child of #9/tid6
+%  0--1@105c41 1-4@105c41 2-5@xbeg-delay 3-5@xbegin 4-5@xend 5-5@105c41 6-4@1001042 7-4@105c41 8-6@xbeg-delay 9-6@xbegin 10-6@datarace-a 11-6@105c41 12-4@1001042 13-4@105c41 14-3@1028e9
+% #[EXPLORE]       dbg #2/tid5 retry set: {TID6 (1,0), TID5 (1,0)}
+% #[EXPLORE]       dbg #2/tid5 retry set: {TID6 (0,1), TID5 (1,0)}
+% #[EXPLORE]       from #14/tid3, chose tid 5 (xabort injection), child of #3/tid5
+%  0--1@105c41 1-4@105c41 2-5@xbeg-delay 3-5@xbegin 4-5@datarace-a 5-5@105c41 6-4@1001042 7-4@105c41 8-6@xbeg-delay 9-6@xbegin 10-6@xend 11-6@105c41 12-4@1001042 13-4@105c41 14-3@1028e9
+% #[EXPLORE]       from #14/tid3, chose tid 6 (xabort injection), child of #9/tid6
+%  0--1@105c41 1-4@105c41 2-5@xbeg-delay 3-5@xbegin 4-5@datarace-a 5-5@105c41 6-4@1001042 7-4@105c41 8-6@xbeg-delay 9-6@xbegin 10-6@datarace-a 11-6@105c41 12-4@1001042 13-4@105c41 14-3@1028e9
+% #[EXPLORE]       from #14/tid3, chose tid 6, child of #4/tid5
+%  0--1@105c41 1-4@105c41 2-5@xbeg-delay 3-5@xbegin 4-5@datarace-a 5-6@xbeg-delay 6-6@xbegin 7-6@xend 8-6@105c41 9-5@105c41 10-4@1001042 11-4@1001042 12-4@105c41 13-3@1028e9
+% #[EXPLORE]       from #13/tid3, chose tid 6 (xabort injection), child of #6/tid6
+%  0--1@105c41 1-4@105c41 2-5@xbeg-delay 3-5@xbegin 4-5@datarace-a 5-6@xbeg-delay 6-6@xbegin 7-6@datarace-a 8-6@105c41 9-5@105c41 10-4@1001042 11-4@1001042 12-4@105c41 13-3@1028e9
+% #[EXPLORE]       from #13/tid3, chose tid 6, child of #2/tid5
+% #[D-MAIL]        tt'ing abort set: {TID6 (1,0), TID5 (1,0)}
+%  0--1@105c41 1-4@105c41 2-5@xbeg-delay 3-6@xbeg-delay 4-6@xbegin 5-6@xend 6-6@105c41 7-5@xbegin 8-5@xend 9-5@105c41 10-4@1001042 11-4@1001042 12-4@105c41 13-3@1028e9
+% #[EXPLORE]       from #13/tid3, chose tid 6, child of #2/tid5
+% #[D-MAIL]        tt'ing abort set: {TID6 (0,1), TID5 (1,0)}
+% +0--1@105c41 1-4@105c41 2-5@xbeg-delay 3-6@xbeg-delay 4-6@xbegin 5-6@xend 6-6@105c41 7-5@xbegin 8-5@datarace-a 9-5@105c41 10-4@1001042 11-4@1001042 12-4@105c41 13-3@1028e9
+%  0--1@105c41 1-4@105c41 2-5@xbeg-delay 3-6@xbeg-delay 4-6@xbegin 5-6@datarace-a 6-6@105c41 7-5@xbegin 8-5@xend 9-5@105c41 10-4@1001042 11-4@1001042 12-4@105c41 13-3@1028e9
+% +0--1@105c41 1-4@105c41 2-5@xbeg-delay 3-6@xbeg-delay 4-6@xbegin 5-6@datarace-a 6-6@105c41 7-5@xbegin 8-5@datarace-a 9-5@105c41 10-4@1001042 11-4@1001042 12-4@105c41 13-3@1028e9
+
+% --- fig-retry   2018-08-03 16:41:45.910879780 -0400
+% +++ fig-baseline        2018-08-03 16:42:02.822879546 -0400
+% @@ -16 +110 @@
+%  7-5@xbegin 8-5@xend 9-5@105c41 10-6@105c41 11-4@105c41 12-6@105c41 13-4@105c41 14-6@1008178 15-6@xbegin 16-6@xend 17-6@105c41 18-3@1028e9
+%  7-5@xbegin 8-5@xend 9-5@105c41 10-6@105c41 11-4@105c41 12-6@105c41 13-4@105c41 14-6@1008178 15-6@xbegin 16-6@datarace-a 17-6@datarace 18-6@105c41 19-3@1028e9
+% #[EXPLORE]       dbg #6/tid5 retry set: {TID6 (0,1), TID5 (1,0)}
+% #[EXPLORE]       from #19/tid3, chose tid 5 (xbegin failure injection), child of #7/tid5
+%  7-5@xbegin 8-5@datarace-a 9-5@datarace 10-5@105c41 11-6@105c41 12-4@105c41 13-6@105c41 14-4@105c41 15-6@1008178 16-6@xbegin 17-6@xend 18-6@105c41 19-3@1028e9
+% #[EXPLORE]       dbg #9/tid5 retry set: {TID6 (1,0), TID5 (1,1)}
+% #[EXPLORE]       dbg #6/tid5 retry set: {TID6 (0,1), TID5 (1,0)}
+% #[EXPLORE]       from #19/tid3, chose tid 6 (xbegin failure injection), child of #16/tid6
+%  7-5@xbegin 8-5@datarace-a 9-5@datarace 10-5@105c41 11-6@105c41 12-4@105c41 13-6@105c41 14-4@105c41 15-6@1008178 16-6@xbegin 17-6@datarace-a 18-6@datarace 19-6@105c41 20-3@1028e9
+% #[EXPLORE]       dbg #9/tid5 retry set: {TID6 (1,0), TID5 (1,1)}
+% #[EXPLORE]       dbg #6/tid5 retry set: {TID6 (0,1), TID5 (1,0)}
+% #[EXPLORE]       from #20/tid3, chose tid 6, child of #9/tid5
+% #[D-MAIL]        tt'ing abort set: {TID6 (1,0), TID5 (1,1)}
+%  7-5@xbegin 8-5@datarace-a 9-5@datarace 10-6@105c41 11-4@105c41 12-6@105c41 13-4@105c41 14-6@1008178 15-6@xbegin 16-6@xend 17-6@105c41 18-5@105c41 19-3@1028e9
+% #[EXPLORE]       dbg #6/tid5 retry set: {TID6 (0,1), TID5 (1,0)}
+% #[EXPLORE]       from #19/tid3, chose tid 6, child of #6/tid5
+% #[D-MAIL]        tt'ing abort set: {TID6 (0,1), TID5 (1,0)}
+% +7-5@xbegin 8-5@datarace-a 9-5@datarace 10-6@105c41 11-4@105c41 12-6@105c41 13-4@105c41 14-6@1008178 15-6@xbegin 16-6@datarace-a 17-6@datarace 18-6@105c41 19-5@105c41 20-3@1028e9
+% +7-6@105c41 8-4@105c41 9-6@105c41 10-4@105c41 11-6@1008178 12-6@xbegin 13-6@xend 14-6@105c41 15-5@xbegin 16-5@xend 17-5@105c41 18-3@1028e9
+% +7-6@105c41 8-4@105c41 9-6@105c41 10-4@105c41 11-6@1008178 12-6@xbegin 13-6@xend 14-6@105c41 15-5@xbegin 16-5@datarace-a 17-5@datarace 18-5@105c41 19-3@1028e9
+%  7-6@105c41 8-4@105c41 9-6@105c41 10-4@105c41 11-6@1008178 12-6@xbegin 13-6@datarace-a 14-6@datarace 15-6@105c41 16-5@xbegin 17-5@xend 18-5@105c41 19-3@1028e9
+% +7-6@105c41 8-4@105c41 9-6@105c41 10-4@105c41 11-6@1008178 12-6@xbegin 13-6@datarace-a 14-6@datarace 15-6@105c41 16-5@xbegin 17-5@datarace-a 18-5@datarace 19-5@105c41 20-3@1028e9
 
@@
 			& 3,2 & \cpu{25434.85} & \ints{718690}	\\ % certified causality ok
-			& 4,1 & \cpu{493.38} & \ints{15064}	\\ % certified causality ok
 			\hline \multicolumn{4}{c}{everything above this line needs rerun} \\
+			& 4,1 & \cpu{547.96} & \ints{15064}	\\
 			%%%%%%%%%%% htm_mutex

### 2018-08-03 thesis/tm.tex
revise retry set fig
update same again

@@
 while the baseline treats it identically to the fully-conflicting {\tt counter}(2,1).
-Figure~\ref{fig:retry-reduction} visualizes the difference between the state spaces explored by the two approaches.
+Figure~\ref{fig:retry-reduction} depicts the difference between the state spaces explored by the two approaches.
 
@@
 	with the middle 2 arising from the data-race preemption point within the abort path.
-	On {\tt counter}, all transitions marked as memory conflicts conflict with each other;
-	on {\tt fig63}, only pairs of one {\tt xend} and one {\tt data race} each conflict.
+	%On {\tt counter}, all transitions marked as memory conflicts conflict with each other;
+	%on {\tt fig63}, only pairs of one {\tt xend} and one {\tt data race} each conflict.
+	When {\tt -R} is supplied, however,
+	after the 4th and 6th branches Landslide activates the retry set
+	indicated at the top of the next upcoming subtree
+	to remember which parts of it to skip.
 	}

### 2018-08-04 thesis/tm.tex
tm discussion/limitations/warpzone section; also write some more about retry sets

@@
 to remain faithful to HTM semantics and still abort even under {\tt -S}.
-
+%
 Likewise, Landslide does not check for a transaction's memory footprint exceeding the CPU's cache capacity,
@@
 perhaps with a configurable cache size,
-but the implementation would likely be theoretically uninteresting
+but would likely be theoretically uninteresting
 (all programs in the upcoming evaluation have trivial memory usage),
-and so is left to future work.
+and so is left unimplemented for now.
+%
+Finally,
+{\tt \_XABORT\_NESTED},
+the last abort code specified by \cite{htm-gcc},
+depends on currently-unsupported {\tt xbegin} nesting, which I discuss further in
+\sect{\ref{sec:tm-warpzone-nested}}.
 
@@
 % whereas no such similar heuristic exists (can exist??) for xbegin PPs, we always just run the success path first
-
-% don't forget to note that this only applies to xabort retry;
-
-The proof of soundness for this optimization is left to future work;
-while the basic idea is motivated by straightforward analogy to the known-sound sleep sets,
-its intersection with Landslide's other exploration features may cause unforeseen problems.
-Accordingly, I implemented the optimization as conservatively as possible,
-falling back to no pruning in cases where how to prune soundly is not clear
-(for example, when a second pair of conflicting transactions involving a third thread
-interleaves while a first retry set is already being tracked).
-Note also that {\tt xbegin} results other than {\tt \_XABORT\_RETRY}
-may depend on the execution logic
-(explicit aborts may be conditional on some change by a conflicting thread,
-and conflict aborts cannot occur before their conflicting transaction to begin with%
-\footnote{See {\tt 410user/progs/htm\_causality.c}; that was a fun realization to have
-already halfway into implementing this feature the wrong way at first.}%
-),
-so how to soundly prune under multiple abort codes
-and/or when retry aborts are suppressed is left to future work.
-For now, retry sets, true to their name,
-are supported only under {\tt landslide -X},
-and incompatible with both the {\tt -A} and {\tt -S} options.
 
@@
 shows the impact of Landslide's experimental retry set reduction (\sect{\ref{sec:tm-retrysets}}).
+Despite its conservative implementation,
+it provides roughly 2-6x reduction in most tests, with up to 20x in extreme cases.
 The biggest win is apparent in {\tt fig63},
-the test contrived to induce as much retry set reduction as possible
-using transactional paths that conflict only with abort paths and not with each other, and vice versa.
-Retry set reduction provides
+the test contrived to induce as much reduction as possible
+using transactional paths that conflict only with aborts and not with each other, and vice versa.
 In {\tt fig63}(2,1), corresponding to Figure~\ref{fig:retry-set-example},
@@
 	%on {\tt fig63}, only pairs of one {\tt xend} and one {\tt data race} each conflict.
-	When {\tt -R} is supplied, however,
+	With the reduction enabled,
 	after the 4th and 6th branches Landslide activates the retry set
@@
 
-Perhaps surprisingly, retry sets also provide reduction even on tests with fully conflicting transactions
+Perhaps surprisingly, retry sets also provide reduction
+even when transactional success and abort paths are fully conflicting,
 (i.e., all tests besides {\tt fig63}).
 With just synchronization preemption points (including {\tt \_xbegin()} and {\tt \_xend()}),
-both the baseline and retry sets would explore exactly all 8 permutations of success and abort between the two threads.
+both the baseline and retry sets would explore exactly all 8 permutations of success and abort between two threads.
 However, in the presence of data-race preemption points,
 even for example on {\tt counter}(2,1)
-(whose abort path is just 1 {\tt xadd},
+(whose abort path is just 1 {\tt xadd} operation),
 and whose optimal state space size should be 8 regardless of data-race preemption points),
 baseline DPOR tests reorderings of one thread's transaction both with the other's failure path,
-and with the other's data race therein.
+and with the other's data race therein%
+\footnote{
+Not pictured in Figure~\ref{fig:retry-reduction} is the symmetric subtree of branches 5-6 in the right half of the state space,
+which would occur after branch 10, reordering the blue thread before the pink's data race.
+Such would be equivalent to branches 2 and 4,
+and is pruned by the normal sleep set algorithm (\sect{\ref{sec:landslide-sleepsets}}, {\tt equiv\_already\_explored()})
+even in baseline DPOR, with no need for retry sets.
+}.
 Retry sets on the other hand identify and skip that equivalence
@@
 
-However, suppressing retry aborts is not guaranteed to faithfully test all possible behaviours
-observable under HTM%
-\footnote{A user who wishes to distinguish conflict aborts, retry aborts, and so on during testing
-without glossing over any of HTM's peculiarities
-could supply {\tt -X -A} without {\tt -S};
-however, this could produce only state spaces at least as large as the baseline.}.
-As an example, note how STM semantics reduced {\tt fig63}'s state space on all $(K,N)$ configurations to 1.
-Because its transactional paths are all mutually independent,
-DPOR identifies no need either to inject conflict aborts or to reorder threads.
-However, this skips the slow-path consistency assertion completely.
-If the programmer had intended it to run ``every so often'' at the whim of the timer interrupt,
-applying this reduction would be unsound.
-Also note the state space size of 4 for many $(2,1)$ test configurations,
-corresponding exactly to the aforementioned success, success and success, abort sequences
-(times two ways to interleave the two threads).
-Because of the scheduling dependency for conflict aborts,
-Landslide cannot recognize the failure path's data races
-without a third freely-reorderable iteration;
-STM mode must be run with $K \times N \ge 3$ to meaningfully test conflicts between failure and success paths at all.
-
-On the other hand, some programs may clearly annotate their intention
-for abort paths to be executed only in case of actual memory conflicts.
-{\tt swap}, {\tt avl\_insert}, and {\tt map\_basic} abstract their {\tt \_xbegin()} calls
-behind an interface which can be implemented either with or without retry loops,
-while {\tt lock}({\tt spinlock}) and {\tt lock}({\tt mutex}) implement the retry loop directly.
-In these cases, the user can check this reduction's soundness by visual inspection.
-Future work could automatically identify a program's retry loops
-and give up on {\tt \_XABORT\_RETRY} by switching to STM mode on-the-fly,
-much like Landslide's heuristic synchronization detection does for yield loops (\sect{\ref{sec:landslide-blocking}})%
-\footnote{Landslide's current implementation gets stuck in infinitely deep interleavings
-when it encounters a retry loop,
-so for now the user must inspect the test case to determine which testing mode to use.
-}.
-
 \subsubsection{Abstraction reduction}
+\label{sec:tm-abstraction}
 
@@
 % this sentence is a complete mess....
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{Discussion}
+
+In this section I review some of the evaluation's results in a broader context,
+list the current limitations of Landslide's implementation,
+and discuss open problems for future work.
+
+\subsubsection{Retry set optimality}
+
+For all the reduction retry sets demonstrated in Table~\ref{tab:tm-verifs},
+some inefficiencies remain in its strategy.
+For example, it is not clear how to prune soundly
+when three or more threads must be reorderd around one transactional preemption point,
+or when a second pair of partially-independent transactions interleaves while an existing retry set is already active.
+Accordingly, I implemented the optimization as conservatively as possible in these cases,
+``saturating'' the retry sets to fall back to no pruning
+({\tt update\_hax\_abort\_set()} and
+{\tt update\_hax\_abandon\_abort\_set()}, respectively).
+Similarly, {\tt xbegin} results other than {\tt \_XABORT\_RETRY}
+may depend on the execution logic
+(explicit aborts may be conditional on some change by a conflicting thread,
+and conflict aborts cannot occur before their conflicting transaction to begin with%
+\footnote{See {\tt 410user/progs/htm\_causality.c}; that was a fun realization to have
+already halfway into implementing this feature the wrong way at first.}%
+),
+so how to soundly prune either success paths or retry aborts
+while other abort codes are in play
+remains an open problem.
+For now, retry sets, true to their name,
+%are supported only under {\tt landslide -X},
+are incompatible with both the {\tt -A} and {\tt -S} options.
+
+Even so, I must also leave a soundness proof for this optimization to future work.
+While
+%the basic idea is
+motivated by straightforward analogy to the known-sound sleep sets,
+its intersection with Landslide's other exploration features may cause unforeseen problems.
+For now, its use is prohibited in conjunction with other state-space-affecting features
+such as ICB (\sect{\ref{sec:landslide-icb}})
+as well as multiple abort codes.
+I personally believe retry set reduction to be sound under these restrictions,
+having carefully scrutinized its behaviour
+while constructing Figure~\ref{fig:retry-set-example}
+and from inspecting larger test parameters' state spaces while implementing it;
+nevertheless, this is no formal proof.
+
+\subsubsection{STM reduction soundness}
+
+In \sect{\ref{sec:tm-eval-stm}} I showed that state spaces could be reduced even further than with retry sets
+by assuming an HTM interface which abstracts away {\tt \_XABORT\_RETRY} behind a loop.
+However, suppressing retry aborts is not guaranteed to faithfully test all possible behaviours
+observable under HTM.
+As an example, note in Table~\ref{tab:tm-verifs}
+how STM semantics reduced % oops
+{\tt fig63}'s state space on all $(K,N)$ configurations to 1.
+Because its transactional paths are all mutually independent,
+DPOR identifies no need either to inject conflict aborts or to reorder threads.
+However, this skips the slow-path consistency assertion completely.
+If the programmer had intended it to run ``every so often'' at the whim of the timer interrupt,
+applying this reduction would be unsound.
+Also note the state space size of 4 for many $(2,1)$ test configurations,
+corresponding exactly to the aforementioned success, success and success, abort sequences
+(times two ways to interleave the two threads).
+Because of the scheduling dependency for conflict aborts,
+Landslide cannot recognize the failure path's data races
+without a third freely-reorderable iteration;
+STM mode must be run with $K \times N \ge 3$ to meaningfully test conflicts between failure and success paths at all.
+A user wishing to distinguish conflict aborts, retry aborts, and so on during testing
+without glossing over any of HTM's peculiarities
+could supply the {\tt -X -A} options without {\tt -S};
+however, this will inevitably result in state spaces at least as large as the baseline.
+
+On the other hand, some programs may clearly annotate their intention
+for abort paths to be executed only in case of actual memory conflicts.
+{\tt swap}, {\tt avl\_insert}, and {\tt map\_basic} abstract their {\tt \_xbegin()} calls
+behind an interface which can be implemented either with or without retry loops,
+while {\tt lock}({\tt spinlock}) and {\tt lock}({\tt mutex}) implement the retry loop directly.
+In these cases, the user can assure themself of STM reduction's soundness by visual inspection.
+In fact, Landslide's current implementation gets stuck in infinitely deep interleavings
+whenever it encounters a retry loop
+% not actually sure why, as it should be around-a-pp and not on the 1st branch...
+(bypassing even its heuristic infinite loop detection),
+so for now the user must inspect the test case to determine which testing mode to use.
+Future work could automatically identify a program's retry loops
+and give up on {\tt \_XABORT\_RETRY} by switching to STM mode on-the-fly,
+much like Landslide's heuristic synchronization detection does for yield loops (\sect{\ref{sec:landslide-blocking}}).
+
+\subsubsection{Nested transactions}
+\label{sec:tm-warpzone-nested}
+
+Whenever {\tt \_xbegin()} is called with a transaction already active, or {\tt \_xend()} while not,
+Landslide's current implementation immediately stops and reports a bug.
+However, just as concurrent programs often hold multiple locks simultaneously,
+one may wish to conduct multiple transactional routines simultaneously,
+especially as they may be abstracted across different code modules when projects grow in scale.
+Recent work \cite{hybrid-htm-stm,relaxed-transactions-popl} has developed
+both implementations and formal semantics for executing nested transactions,
+so future work should extend the verification concurrency model to permit such programs.
+For now, the best Landslide can offer is to check the transactional components and their client code
+separately against their APIs
+with abstraction reduction (\sect{\ref{sec:tm-abstraction}}),
+then check the rest of the program with (for example) traditional mutexes that can nest safely.
+
+\subsubsection{Relaxed memory orderings}
+\label{sec:tm-warpzone-relaxed}
+
+Section~\ref{sec:tm-design}'s formalization of thread interleavings does not account for read/write reorderings
+possible on relaxed consistency architectures \cite{memory-consistency-models}.
+In fact,
+even after \cite{tm-benchmark-cmu}'s proposed fix to the atomicity protocol in Figure~\ref{fig:htm-fixed},
+it is still incorrect on Total Store Order (TSO) architectures such as x86,
+let alone on weaker memory models.
+Despite stores being totally-ordered, x86 may still reorder stores after subsequent loads
+\cite{sully-thesis}.
+Accordingly, an execution of lines $8,9a,9b,9c$
+may be locally visible to another thread as $9a,8,9b,9c$,
+and hence an apparent interleaving of
+\[
+        \tiat{1},\tjat{1-5},\tiat{7},\underline{\tiat{9a}},\tkat{1-5},\underline{\tiat{8}},\tiat{9b-B}
+\]
+is possible
+(reordered accesses underlined for emphasis).
+An acquire barrier is needed between lines 8 and 9 to solve this problem on TSO \cite{tsx-need-barrier}.
+(on x86, either {\tt mfence} or {\tt xchg}/{\tt xadd}).
+% gcc generates the latter for its __sync_lock_test_and_set intrinsic
+On Partial Store Order (PSO) architectures,
+%even more barriers may be necessary.
+a release barrier is necessary before line A, likewise.
+Recent work \cite{relaxed-transactions-pldi} also discovered similar unsoundness in lock elision on ARMv8 (PSO).
+
+Because Landslide's concurrency model includes only instruction-level thread nondeterminism,
+not per-CPU memory buffer reorderings,
+its current HTM implementation cannot find this bug.
+In fact, it erroneously verifies the corresponding test {\tt htm2}(3,1)
+%% sigbovik version, probably pre-thrlib-function and definitely pre-dpor-preferred-tid and pre-trusted-join
+% in 3 CPU-hours,
+% with 467730 distinct interleavings in total,
+in 40 CPU-seconds,
+with 774 interleavings in total,
+none of which include the above-listed sequence.
+Recent work has extended DPOR to support TSO and PSO memory nondeterminism \cite{tsopso},
+as well as proposed formal execution semantics for HTM on these architectures
+\cite{relaxed-transactions-popl,relaxed-transactions-pldi};
+if both of these advances were incorporated into Landslide's concurrency model,
+it could find or verify the absence of such bugs.
+Visual inspection of \cite{tm-benchmark-cmu}'s HTM data structures found no barriers used in this implementation pattern;
+I would urge any reader interested in using those to add them in by hand first.
+The test case {\tt lock}({\tt mutex}) ({\tt 410user/progs/htm\_mutex.c} in the repository)
+provides an example of how to use compiler intrinsics to emit the necessary barriers.

### 2018-08-04 thesis/tm.tex
first revision pass on tm.tex, improve section references, commit hashes, clean up todos

@@
 and hence assumes sequential consistency.
-% TODO: fix sect refrance
-For discussion of relaxed memory models refer to Section~\ref{sec:warpzone}.
+For discussion of relaxed memory models refer to Section~\ref{sec:tm-warpzone-relaxed}.
 
@@
 	satisfying case 1.
-	% TODO: give a secrion referenace
-	Otherwise, by DPOR's definition of transition dependence \cite{dpor}, %,landslide-phdthesis},
+	Otherwise, by DPOR's definition of transition dependence (\cite{dpor}, \sect{\ref{sec:landslide-dpor-hb}}),
 	$\tiiat{\beta_{b+1}}\dots\tiiat{\beta_n}$ is independent with the transitions of \tjj and \tkk,
@@
 at such preemption points,
-which is just an implementation detail, not theoretically necessary}.
+which is just an implementation detail, not theoretically necessary.}.
 The optional abort code is then passed through {\tt arbiter\_append\_choice()}/{\tt arbiter\_pop\_choice()}
@@
         \begin{center}
-	% TODO syntax hilight
         \begin{tabular}{ll}
@@
 thereby skipping over (i.e., pruning) reorderings of any other {\tt xbegin} results that would be independent.
-% TODO: make a real commit hash instead; put on master
-The {\tt landslide} repository's {\tt retry-sets} branch implements this feature%
-\footnote{The upcoming evaluation showed it too unstable, let alone not yet proved sound, to include in {\tt master}.},
+%The {\tt landslide} repository's {\tt retry-sets} branch implements this feature,
+This feature is implemented in commit 86657c7,
 available with the {\tt -R} command-line option.
@@
 
-The evaluation suite comprises several hand-written unit tests,
+The evaluation suite comprises several unit tests hand-written by yours truly,
 microbenchmarks and transactional data structures from \cite{tm-benchmark-cmu},
@@
 \begin{itemize}
-	\item Microbenchmarks
+	\item Unit tests and microbenchmarks
 	\begin{itemize}
@@
 implemented since then:
-the sleep sets optimization for DPOR (\sect{\ref{sec:landslide-sleepsets}}, commit 0447666),
+the sleep sets optimization for DPOR (\sect{\ref{sec:landslide-sleepsets}}, commits 0447666 and 588687c),
 the {\tt thrlib\_function} and {\tt TRUSTED\_THR\_JOIN} directives
-to mark internal thread library logic as trusted (\sect{\ref{sec:landslide-config-landslide}}, commit 64a02e4),
+to mark internal thread library logic as trusted
+(\sect{\ref{sec:landslide-config-landslide}}, commits 64a02e4 and a50d4ea),
 and fixing a soundness bug in which Landslide could neglect to inject transaction failures
@@
 
-\begin{table*}[p]
+\begin{table*}[t]
 	\begin{center}
@@
 
-% TODO check figure placement
 \begin{figure}[h]
@@
 			& 2,4 & \cpu{3433.08}	& \ints{86950}	& \cpu{1134.62}	& \ints{25398}	& \cpu{752.51}	& \ints{19128} \\
-			& 3,1 & \cpu{10.57}	& \ints{120}	& \cpu{92.30}	& \ints{64}	& \cpu{7.88}	& \ints{40} \\
+			& 3,1 & \cpu{10.57}	& \ints{120}	& \cpu{9.23}	& \ints{64}	& \cpu{7.88}	& \ints{40} \\
 			& 3,2 & \cpu{2492.17}	& \ints{60606}	& \cpu{2040.98}	& \ints{44862}	& \cpu{565.97}	& \ints{14304} \\
@@
 
-\subsubsection{Abort code (STM) reduction}
+\subsubsection{STM (abort code) reduction}
 \label{sec:tm-eval-stm}
@@
 			& 2,3 & \\ %\cpu{0} & \ints{0} \\
+			& 2,4 & \\ %\cpu{0} & \ints{0} \\
 			& 3,1 & \\ %\cpu{0} & \ints{0} \\
 			& 3,2 & \\ %\cpu{0} & \ints{0} \\
+			& 3,3 & \\ %\cpu{0} & \ints{0} \\
+			& 3,4 & \\ %\cpu{0} & \ints{0} \\
+			& 4,1 & \\ %\cpu{0} & \ints{0} \\
+			%& 4,2 & \\ %\cpu{0} & \ints{0} \\
 			\hline \multicolumn{4}{c}{everything above this line needs rerun} \\
-			\\
-			\\
 		\end{tabular}

### 2018-08-05 thesis/tm.tex
lotta revisions to tm.tex prose, esp from related work

@@
 or in case of any system interrupt or cache overflow.
+Intel's commercial implementation of HTM has a rocky history of hardware bugs \cite{intel-disable-tsx,intel-tsx-update},
+attesting to the feature's complexity and the need for formal verification on both sides of its API.
 
@@
 Transactional Synchronization Extensions (TSX)
-refers to Intel's implementation of HTM on the Haswell microarchitecture
+refers to Intel's implementation of HTM on Haswell and more recent microarchitectures
 \cite{htm-haswell}.
@@
 % https://software.intel.com/en-us/node/683688
-\cite{hardware-lock-elision}.
+\cite{speculative-lock-elision,lock-elision}
 In this thesis I focus on RTM, the more general (i.e., expressive (i.e., bug-prone)) interface,
@@
 the way I stated interleavings such as $\tiat{1-2c}, \tjat{1-3}, \tiat{4-7}$%
-\footnote{For a clearer example here I have reordered \ti's write to {\tt x} before \tj's part, compared to before.}
-glossed over the HTM execution semantics underlying how such a sequence of operations would be carried out.
-For example, thread 1's write during $2c$ is not actually visible to thread 2,
-such as it would be under a thread-scheduling-only concurrency model.
-
-Intel's official TSX documentation summarizes its interface and behaviour in prose \cite{intel-tsx-overview};
-to the best of my knowledge,
-no formal definition exists which could be verified in a mechanized logical framework \cite{x86-semantics}.
-Hence the proofs in this section will likewise be in prose.
+\footnote{For a clearer example to follow, I have reordered \ti's write to {\tt x} before \tj's part, compared to before.}
+glossed over
+%the HTM execution semantics underlying how such a sequence of operations would be carried out.
+how such a sequence of operations would be carried out under HTM.
+For example, \ti's write during $2c$ is not actually visible to \tj,
+although it would be under a thread-scheduling-only concurrency model.
+
+Intel's official TSX documentation summarizes its interface and behaviour in prose \cite{intel-tsx-overview}.
+Recent work has used proof assistants to formalize some of the execution semantics
+of x86 in general \cite{x86-semantics}
+and of transactions in particular (at both hardware- and language-level) \cite{relaxed-transactions-pldi}.
+However, state-of-the-art advances in model checking algorithms
+still state their theorems and proofs in prose \cite{bpor,optimal-dpor,tsopso,satcheck,mcr,quicksand},
+so this section's proofs will regrettably do likewise,
+leaving the rigor of mechanization to future work.
+The reader may at least draw assurance from the existing formal verifications of
+the transactional semantics required herein.
+
 To summarize HTM's execution semantics:
 
-% TODO FIXME: this is not genral semantics, this is jut as epplies to the examineple. rephrase
 \begin{enumerate}
-	\item any modifications to shared state (such as $2c$) by \ti are not visible to \tj during its execution,
-		despite \tj being executed afterwards, and
-	\item all local and global state changes by \ti between lines $1$ and $2c$ are discarded when jumping to line $4$.
+	\item Any modifications to shared state (such as $2c$) by a transacting thread
+		are not visible to any other during the transaction
+		(such as $2c$ in this example, despite \tj executing afterwards).
+	\item All local and global state changes during a transaction
+		(such as \ti's lines $1-2c$ in this example)
+		are discarded when returning an abort code from {\tt xbegin}
+		(jumping to line $4$, in this example).
 \end{enumerate}
@@
 Finally, Quicksand supports the following command-line options to enable various sets of HTM features within Landslide.
-Each option requires each of the previously-listed options to also be present.
+%Each option requires each of the previously-listed options to also be present.
 \begin{itemize}
@@
 		enables multiple xabort failure codes (\sect{\ref{sec:txn-failure}}).
+		Requires {\tt -X}.
 	\item {\tt -S} (for ``Stm'' or ``Suppress retries'')
 		disables the {\tt \_XABORT\_RETRY} failure reason,
-		which causes Landslide to match STM's execution semantics rather than HTM.
-		(This feature was used in \cite{sigbovik-htm}~\sect{5}.)
+		causing Landslide to emulate the semantics of STM rather than HTM.
+		%(This feature was used in \cite{sigbovik-htm}~\sect{5}.)
+		Requires {\tt -X -A}.
+	\item {\tt -R} (for ``Retry sets'')
+		enables state space reduction based on independences between individual {\tt xbegin} results
+		(\sect{\ref{sec:tm-retrysets}}).
+		Requires {\tt -X} and {\em not} {\tt -A}.
 \end{itemize}
@@
 
-To solve this,
+To address this,
 I extended Landslide
-with {\em retry set reduction},
-an experimental feature named after the analogous {\em sleep sets} from prior work
+with {\em retry set reduction} (commit 86657c7),
+an experimental feature named after prior work's analogous {\em sleep sets}
 (\cite{partial-order-methods,dpor,optimal-dpor}, \sect{\ref{sec:landslide-sleepsets}}).
@@
 and refuses to test any others
-(unless DPOR separately found them, too, in conflict),
+(unless DPOR separately found them, too, to conflict),
 thereby skipping over (i.e., pruning) reorderings of any other {\tt xbegin} results that would be independent.
 %The {\tt landslide} repository's {\tt retry-sets} branch implements this feature,
-This feature is implemented in commit 86657c7,
-available with the {\tt -R} command-line option.
+Upcoming in the evaluation, Figure~\ref{fig:retry-reduction} will visualize the reduction achieved in two test cases.
 
@@
 	\begin{itemize}
-		\item {\tt htm1}: The bug from Figure~\ref{fig:htm-example}. %(assertion failure on {\tt count}).
+		\item {\tt htm1}: The bug from Figure~\ref{fig:htm-example}/\ref{fig:htm-example-reproduced}.
+			%(assertion failure on {\tt count}).
 		\item {\tt htm2}: The fixed version as in Figure~\ref{fig:htm-fixed}.
@@
 		For each, I list the CPU-time and wall-clock time elapsed,
-		plus the number of interleavings tested in the ultimately buggy state space,
-		before the bug was found.
+		plus the number of interleavings tested in the ultimately buggy state space
+		until the bug was found.
 		* marks the winning measurements between each series.
-		Lastly, state space estimation (\sect{\ref{sec:landslide-estimate}}),
-		though approximate at best,
+		Lastly, state space estimation (\sect{\ref{sec:landslide-estimate}})
+		%though approximate at best,
 		confers a sense of the exponential explosion.
-		{\bf RERUN NOTE: this entire table needs to be rerun on account of dpor-preferred-tid}
 	}
@@
 the multiplicative factor in bug-finding speed (2-4x, eyeballing) is generally smaller
-than that of the total number of interleavings (10-100x).
+than that of the total number of interleavings (10-1000x).
 In other words,
-should they exist,
-Landslide is likely to find bugs reasonably quickly in these transactional programs
+should transactional bugs exist,
+Landslide is likely to find them reasonably quickly %in these transactional programs
 despite prohibitive exponential explosion in total state space size.
@@
 
-\subsection{Performance}
+\subsubsection{Performance}
 
@@
 Landslide was able to verify most of these tests for a fair range of thread and iteration counts,
-often reaching up to 2 threads with 3 iterations each, 3 with 2 each, or 4 with 1.
+often reaching up to 2 threads with 3-4 iterations each, 3 with 2 each, or 4 with 1.
 In the case of {\tt htm2}, for example,
@@
 One recent paper \cite{tsopso},
-which likewise extended DPOR with a new dimension of concurrency (weak memory orderings)
+which likewise extended DPOR with a new dimension of concurrency (weak memory orderings),
 reported verifying programs with up to 10 concurrent events,
@@
 			{\tt swap}		& 1	& 4	& 8	& 52	\\ % 2,2
-			{\tt fig63}		& 1	& 0	& $K$	& 24	\\ % 3,2 k=3 -> 24; 4,2 -> 40 (not yet); 2,4 -> 24
+			{\tt fig63}		& 1	& 0	& $K$-1	& 18	\\ % 3,2 k=3 -> 18; 2,4 -> 16
 			{\tt avl\_insert}	& 1	& 2	& 7	& 20	\\ % 2,1 nb. only 5 drpps under -A -S but w/e
@@
 for more complicated algorithms,
-which may involve complex conflict patterns such as tree rebalancing or map resizing;
-other formal verification approaches aside,
+which may involve complex conflict patterns such as tree rebalancing or map resizing.
+Other formal verification approaches aside,
 the user must ultimately be content with the probabilistic assurance that
@@
 	With the reduction enabled,
-	after the 4th and 6th branches Landslide activates the retry set
-	indicated at the top of the next upcoming subtree
-	to remember which parts of it to skip.
+	after the 4th and 6th branches
+	(i.e., when preempting to reorder threads),
+	Landslide activates the retry set
+	indicated at the top of the next upcoming subtree,
+	allowing it to identify and skip 2 redundant branches in {\tt counter}
+	and 4 in {\tt fig63}.
+	%to remember which parts of it to skip.
 	}
@@
 \footnote{See {\tt 410user/progs/htm\_causality.c}; that was a fun realization to have
-already halfway into implementing this feature the wrong way at first.}%
+already halfway into implementing retry sets the wrong way at first.}%
 ),
@@
 and from inspecting larger test parameters' state spaces while implementing it;
-nevertheless, this is no formal proof.
+nevertheless, this falls well short of formal proof.
 
@@
 applying this reduction would be unsound.
-Also note the state space size of 4 for many $(2,1)$ test configurations,
+Also note the state space size of 4 for many (2,1) test configurations,
 corresponding exactly to the aforementioned success, success and success, abort sequences
@@
 one may wish to conduct multiple transactional routines simultaneously,
-especially as they may be abstracted across different code modules when projects grow in scale.
+especially as they may be abstracted across different code modules as a project grows in scale.
 Recent work \cite{hybrid-htm-stm,relaxed-transactions-popl} has developed
@@
 (reordered accesses underlined for emphasis).
-An acquire barrier is needed between lines 8 and 9 to solve this problem on TSO \cite{tsx-need-barrier}.
+An acquire barrier is needed between lines 8 and 9 to solve this problem on TSO \cite{tsx-need-barrier}
 (on x86, either {\tt mfence} or {\tt xchg}/{\tt xadd}).
 % gcc generates the latter for its __sync_lock_test_and_set intrinsic
-On Partial Store Order (PSO) architectures,
-%even more barriers may be necessary.
-a release barrier is necessary before line A, likewise.
-Recent work \cite{relaxed-transactions-pldi} also discovered similar unsoundness in lock elision on ARMv8 (PSO).
+Recent work \cite{relaxed-transactions-pldi} also demonstrated unsoundness
+in a similar lock elision implementation on ARMv8 (PSO),
+in which the transactional path reads the lock's internal state directly
+rather than using a separate flag.
+in Figure~\ref{fig:htm-fixed}, a release barrier before line A is also necessary under PSO.
+% On Partial Store Order (PSO) architectures,
+% %even more barriers may be necessary.
+% a release barrier is necessary before line A as well.
+% Recent work \cite{relaxed-transactions-pldi} also discovered similar unsoundness
+% in lock elision \cite{speculative-lock-elision}
+% on ARMv8 (PSO).
+% Recent work \cite{relaxed-transactions-pldi} also discovered similar unsoundness
 

### 2018-08-06 thesis/tm.tex
smore tm revision

@@
 However, state-of-the-art advances in model checking algorithms
-still state their theorems and proofs in prose \cite{bpor,optimal-dpor,tsopso,satcheck,mcr,quicksand},
+still state their theorems and proofs in prose \cite{bpor,optimal-dpor,tsopso,satcheck,mcr,quicksand,cpp-model-checking},
 so this section's proofs will regrettably do likewise,
 leaving the rigor of mechanization to future work.
-The reader may at least draw assurance from the existing formal verifications of
-the transactional semantics required herein.
+The reader may at least rest assured that the proofs herein
+rely on transactional semantics that have themselves been formally verified.
 
@@
 {\tt update\_hax\_abandon\_abort\_set()}, respectively).
-Similarly, {\tt xbegin} results other than {\tt \_XABORT\_RETRY}
+
+Likewise, the cases of {\tt htm2}(2,3) and (4,1),
+in which retry sets achieved better reduction than STM mode,
+show that the latter does not necessarily subsume the former,
+and that combining the two could in theory achieve further reduction still.
+However, {\tt xbegin} results other than {\tt \_XABORT\_RETRY}
 may depend on the execution logic
@@
 remains an open problem.
-For now, retry sets, true to their name,
-%are supported only under {\tt landslide -X},
-are incompatible with both the {\tt -A} and {\tt -S} options.
-
-Even so, I must also leave a soundness proof for this optimization to future work.
+
 While
@@
 motivated by straightforward analogy to the known-sound sleep sets,
-its intersection with Landslide's other exploration features may cause unforeseen problems.
+the intersection of retry sets with Landslide's other exploration features may cause unforeseen problems.
 For now, its use is prohibited in conjunction with other state-space-affecting features
@@
 while constructing Figure~\ref{fig:retry-set-example}
-and from inspecting larger test parameters' state spaces while implementing it;
-nevertheless, this falls well short of formal proof.
+and from inspecting state spaces arising from larger test parameters as well; %while implementing it;
+nevertheless, this falls well short of formal proof, which I must leave to future work.
 
@@
 rather than using a separate flag.
-in Figure~\ref{fig:htm-fixed}, a release barrier before line A is also necessary under PSO.
+In Figure~\ref{fig:htm-fixed}, a release barrier before line A is also necessary under PSO.
 % On Partial Store Order (PSO) architectures,

### 2018-08-07 thesis/tm.tex
if this isn't the last rerun of tm tests i'll have to do i'm gonna cry

@@
 			& 2,1	&\cpu{  26.48}&\wtm{6.48}&\ints{5}& \cpu{*8.21}& \wtm{*5.78}& \ints{5	}& \ETA{12} \\
-			% 5 interleavings tested; 1 preemptions; job time 3s; pldi time 6482137; new-fixed pldi cputime 26480574
-			% 5 interleavings tested; 1 preemptions; job time 2s; pldi time 5781000; new-fixed pldi cputime 8214902
 		(assertion)
 			& 2,2	&\cpu{  45.23}&\wtm{8.28}&\ints{9}& \cpu{*8.42}& \wtm{*5.97}& \ints{9	}& \ETA{102} \\
-			% 9 interleavings tested; 1 preemptions; job time 4s; pldi time 8281422; new-fixed pldi cputime 45289375
-			% 9 interleavings tested; 1 preemptions; job time 2s; pldi time 5972944; new-fixed pldi cputime 8418505
 			& 2,3	&\cpu{ 53.88}&\wtm{9.92}&\ints{17}& \cpu{*8.80}& \wtm{*6.31}& \ints{17	}& \ETA{819} \\
-			% 17 interleavings tested; 1 preemptions; job time 5s; pldi time 9921305; new-fixed pldi cputime 53884499
-			% 17 interleavings tested; 1 preemptions; job time 2s; pldi time 6314386; new-fixed pldi cputime 8795639
 			& 2,4	&\cpu{ 90.24}&\wtm{14.63}&\ints{33}& \cpu{*9.72}& \wtm{*7.19}& \ints{33	}& \ETA{6553} \\
-			% 33 interleavings tested; 1 preemptions; job time 7s; pldi time 14629409; new-fixed pldi cputime 90244293
-			% 33 interleavings tested; 1 preemptions; job time 3s; pldi time 7193093; new-fixed pldi cputime 9717902
 		\cline{2-9}
 			& 3,1	&\cpu{ 39.75}&\wtm{8.21}&\ints{5}& \cpu{*8.15}& \wtm{*5.71}& \ints{5	}& \ETA{76} \\
-			% 5 interleavings tested; 1 preemptions; job time 4s; pldi time 8213084; new-fixed pldi cputime 39752062
-			% 5 interleavings tested; 1 preemptions; job time 2s; pldi time 5710637; new-fixed pldi cputime 8153919
 			& 3,2	&\cpu{ 46.91}&\wtm{8.40}&\ints{9}& \cpu{*8.37}& \wtm{*5.91}& \ints{9	}& \ETA{3686} \\
-			% 9 interleavings tested; 1 preemptions; job time 4s; pldi time 8397105; new-fixed pldi cputime 46909424
-			% 9 interleavings tested; 1 preemptions; job time 2s; pldi time 5907009; new-fixed pldi cputime 8367804
 			& 3,3	&\cpu{ 41.39}&\wtm{8.50}&\ints{17}& \cpu{*8.71}& \wtm{*6.26}& \ints{17	}& \ETA{176947} \\
-			% 17 interleavings tested; 1 preemptions; job time 4s; pldi time 8503682; new-fixed pldi cputime 41386987
-			% 17 interleavings tested; 1 preemptions; job time 2s; pldi time 6263778; new-fixed pldi cputime 8711332
 			& 3,4	&\cpu{ 92.98}&\wtm{15.19}&\ints{33}& \cpu{*10.36}& \wtm{*7.15}& \ints{33	}& \ETA{8493465} \\
-			% 33 interleavings tested; 1 preemptions; job time 8s; pldi time 15185585; new-fixed pldi cputime 92982811
-			% 33 interleavings tested; 1 preemptions; job time 3s; pldi time 7153968; new-fixed pldi cputime 10355961
 		\cline{2-9}
 			& 4,1	&\cpu{ 45.13}&\wtm{8.56}&\ints{5}& \cpu{*7.77}& \wtm{*5.30}& \ints{5	}& \ETA{460} \\
-			% 5 interleavings tested; 1 preemptions; job time 4s; pldi time 8556561; new-fixed pldi cputime 45128135
-			% 5 interleavings tested; 1 preemptions; job time 2s; pldi time 5295360; new-fixed pldi cputime 7770389
 			& 4,2	&\cpu{64.99}&\wtm{11.33}&\ints{9}& \cpu{*8.32}& \wtm{*5.59}& \ints{9	}& \ETA{132710} \\
-			% 9 interleavings tested; 1 preemptions; job time 5s; pldi time 11333162; new-fixed pldi cputime 64991248
-			% 132710 9 interleavings tested; 1 preemptions; job time 2s; pldi time 5588002; new-fixed pldi cputime 8317173
 		\hline
 		{\tt swapbug}
-			& 2,1	&\cpu{ 84.26}&\wtm{15.49}&\ints{*6}& \cpu{*28.32}& \wtm{*11.29}& \ints{33	}& \ETA{73} \\
-			% 6 interleavings tested; 1 preemptions; job time 5s; pldi time 15485317; new-fixed pldi cputime 84260417
-			% 33 interleavings tested; 1 preemptions; job time 3s; pldi time 11286074; new-fixed pldi cputime 28324164
+			& 2,1	&\cpu{104.37}&\wtm{17.00}&\ints{*6}& \cpu{*43.76}& \wtm{*12.51}& \ints{33	}& \ETA{73} \\
 		(deadlock)
-			& 2,2	&\cpu{ 64.65}&\wtm{*11.90}&\ints{*38}& \cpu{*30.55}& \wtm{13.50}& \ints{85	}& \ETA{860} \\
-			% 38 interleavings tested; 2 preemptions; job time 6s; pldi time 11899011; new-fixed pldi cputime 64653180
-			% 85 interleavings tested; 1 preemptions; job time 5s; pldi time 13499606; new-fixed pldi cputime 30550037
-			& 2,3	&\cpu{ 94.59}&\wtm{*14.68}&\ints{*48}& \cpu{*32.27}& \wtm{19.42}& \ints{217	}& \ETA{9120} \\
-			% 48 interleavings tested; 2 preemptions; job time 9s; pldi time 14681610; new-fixed pldi cputime 94594891
-			% 217 interleavings tested; 1 preemptions; job time 11s; pldi time 19419243; new-fixed pldi cputime 37272562
-			& 2,4	&\cpu{ 109.37}&\wtm{*17.05}&\ints{*68}& \cpu{*53.44}& \wtm{35.80}& \ints{537	}& \ETA{91239} \\
-			% 68 interleavings tested; 2 preemptions; job time 10s; pldi time 17049141; new-fixed pldi cputime 109374628
-			% 537 interleavings tested; 1 preemptions; job time 27s; pldi time 35796053; new-fixed pldi cputime 53441873
+			& 2,2	&\cpu{ 66.47}&\wtm{*12.42}&\ints{*38}& \cpu{*45.22}& \wtm{15.30}& \ints{85	}& \ETA{860} \\
+			& 2,3	&\cpu{ 99.41}&\wtm{*15.92}&\ints{*48}& \cpu{*53.57}& \wtm{21.34}& \ints{217	}& \ETA{9120} \\
+			& 2,4	&\cpu{125.23}&\wtm{*19.46}&\ints{*68}& \cpu{*71.01}& \wtm{38.86}& \ints{537	}& \ETA{91239} \\
 		\cline{2-9}
-			& 3,1	&\cpu{221.58}&\wtm{*33.44}&\ints{*32}&\cpu{*65.13}& \wtm{52.23}& \ints{980   }& \ETA{3417} \\
-			% 32 interleavings tested; 4 preemptions; job time 6s; pldi time 33442378; new-fixed pldi cputime 221580261
-			% 980 interleavings tested; 2 preemptions; job time 40s; pldi time 52225013; new-fixed pldi cputime 65125824
-			& 3,2	&\cpu{*3283.16}&\wtm{*413.59}&\ints{*260}&\ETA{--}&\ETA{>1h}&\ETA{--}& \ETA{2256553} \\
-			% 260 interleavings tested; 4 preemptions; job time 32s; pldi time 413587352; new-fixed pldi cputime 3283157217
-			% [JOB 8] progress: 63137/2256553 brs (2.797940%), ETA 130d 19h 28m 29s (elapsed 59m 45s)
+			& 3,1	&\cpu{313.36}&\wtm{*44.50}&\ints{*34}&\cpu{*78.23}& \wtm{60.49}& \ints{1016	}& \ETA{3543} \\
+			& 3,2	&\cpu{*6500.36}&\wtm{*816.27}&\ints{*262}&\ETA{--}&\ETA{>1h}&\ETA{--}& \ETA{2161715} \\
 		\hline
 		{\tt avl\_insert}
-			& 2,2	&\cpu{862.54}&\wtm{111.64}&\ints{*57}	&\cpu{*66.32}&\wtm{*31.81}&\ints{127}&\ETA{254078} \\
-			% 57 interleavings tested; 1 preemptions; job time 13s; pldi time 111641721; new-fixed pldi cputime 862539202
+			& 2,2	&\cpu{1355.05}&\wtm{174.09}&\ints{*29}	&\cpu{*93.28}&\wtm{*30.37}&\ints{79}&\ETA{157286} \\
 		(segfault+)
-			& 2,3	&\cpu{1490.20}&\wtm{*191.56}&\ints{*221}&\cpu{*256.94}&\wtm{193.71}&\ints{1391}&\ETA{22773586} \\
-			% 221 interleavings tested; 2 preemptions; job time 27s; pldi time 191557235; new-fixed pldi cputime 1490204572
-			& 2,4	&\cpu{*7598.19}&\wtm{*970.67}&\ints{*2633}&\ETA{--}&\ETA{>1h}&\ETA{--}& \ETA{91704203} \\
-			% 2633 interleavings tested; 3 preemptions; job time 5m 31s; pldi time 970674411; new-fixed pldi cputime 7598188798
-			% [JOB 12] progress: 28140/91704203 brs (0.030686%), ETA 18y 54d 18h 42m 58s (elapsed 51m 12s)
+			& 2,3	&\cpu{402.91}&\wtm{*54.60}&\ints{*33}&\cpu{*228.16}&\wtm{125.89}&\ints{835}&\ETA{13664151} \\
+			% TODO: these 2,3 ones alone are new + ok
+			% 33 interleavings; 1 preemptions; job time 12s; pldi time 179368977; new-fixed pldi cputime 1403360474
+			% 835 interleavings; 1 preemptions; job time 1m 42s; pldi time 135292827; new-fixed pldi cputime 235664920
+			& 2,4	&\ETA{--}&\ETA{>1h}&\ETA{--}&\ETA{--}&\ETA{>1h}&\ETA{--}& \ETA{79020639} \\
+			% this quicksand jerk required ulimit -n higher than 1024
+			%
+			% the jerk didn't even find the bug anymore, after dpor-preferred tid
+			% heres the stats from how it did before
+			% & 2,4	&\cpu{*7598.19}&\wtm{*970.67}&\ints{*2633}&\ETA{--}&\ETA{>1h}&\ETA{--}& \ETA{79020639} \\
+			% anyway this was the maximal
+			%
 		\cline{2-9}
-			& 3,1	&\cpu{62.69}&\wtm{*10.36}&\ints{*23}	&\cpu{*64.70}&\wtm{25.59}&\ints{53}&\ETA{343250} \\
-			% 23 interleavings tested; 3 preemptions; job time 6s; pldi time 10359882; new-fixed pldi cputime 62690171
-			% 53 interleavings tested; 3 preemptions; job time 6s; pldi time 25585758; new-fixed pldi cputime 64695226
-			& 3,2	&\cpu{*2445.90}&\wtm{*325.81}&\ints{*233}&\ETA{--}&\ETA{>1h}&\ETA{--}& \ETA{466977272} \\
-			% 233 interleavings tested; 3 preemptions; job time 30s; pldi time 325806017; new-fixed pldi cputime 2445895897)
-			% was also found with 169 interleavings but later on
-			%[JOB 38] progress: 3812/466977272 brs (0.000816%), ETA INF (elapsed 13m 50s)
-			\hline \multicolumn{9}{c}{everything above this line needs rerun for dpor-preferred tid} \\
+			& 3,1	&\cpu{*86.43}&\wtm{*14.10}&\ints{*23}	&\cpu{91.97}&\wtm{24.80}&\ints{53}&\ETA{343250} \\
+			& 3,2	&\cpu{*2177.86}&\wtm{*287.66}&\ints{*169}&\ETA{--}&\ETA{>1h}&\ETA{--}& \ETA{425939133} \\
+			% TODO
+			%\multicolumn{9}{c}{everything above this line needs rerun for dpor-preferred tid + thr_bottom1} \\
+
 		\hline
@@
 		%({\tt spinlock})
-			& 2,1	&\cpu{12.33}&\wtm{3.69}&\ints{2}&\cpu{*3.32}&\wtm{*3.32}&\ints{2}& \ETA{4} \\
+			& 2,1	&\cpu{18.33}&\wtm{5.19}&\ints{2}&\cpu{*3.12}&\wtm{*3.12}&\ints{2}& \ETA{4} \\
+			% 2 interleavings; 0 preemptions; job time 4s; pldi time 5191170; new-fixed pldi cputime 18328224
+			% 2 interleavings; 0 preemptions; job time 2s; pldi time 3119081; new-fixed pldi cputime 3118912
 		(perf)
-			& 9,9	&\cpu{20.49}&\wtm{5.74}&\ints{2}&\cpu{*4.78}&\wtm{*4.78}&\ints{2}& \ETA{inf} \\
+			& 9,9	&\cpu{22.43}&\wtm{6.24}&\ints{2}&\cpu{*4.71}&\wtm{*4.71}&\ints{2}& \ETA{inf} \\
+			% 2 interleavings; 0 preemptions; job time 3s; pldi time 6238515; new-fixed pldi cputime 22431755
+			% 2 interleavings; 0 preemptions; job time 3s; pldi time 4712012; new-fixed pldi cputime 4711884
 	\end{tabular}
@@
 			\hline
+			% TODO: everything above this line needs rerun
+			% TODO: everything below is fine up to dpor-preferred & thr_bottom1 updates
 			{\tt fig63}
-			& 2,1 & \cpu{7.03}	& \ints{10}	& \cpu{6.96}	& \ints{6}	& \cpu{3.42}	& \ints{1} \\
-			& 2,2 & \cpu{9.32}	& \ints{108}	& \cpu{8.78}	& \ints{76}	& \cpu{3.43}	& \ints{1} \\
-			& 2,3 & \cpu{54.41}	& \ints{1934}	& \cpu{31.65}	& \ints{977}	& \cpu{3.27}	& \ints{1} \\
-			& 2,4 & \cpu{997.09} 	& \ints{36600} 	& \cpu{414.27}	& \ints{14512}	& \cpu{3.34}	& \ints{1} \\
+			& 2,1 & \cpu{7.12}	& \ints{10}	& \cpu{6.92}	& \ints{6}	& \cpu{3.40}	& \ints{1} \\
+			& 2,2 & \cpu{9.24}	& \ints{108}	& \cpu{8.61}	& \ints{76}	& \cpu{3.46}	& \ints{1} \\
+			& 2,3 & \cpu{54.68}	& \ints{1934}	& \cpu{31.76}	& \ints{977}	& \cpu{3.39}	& \ints{1} \\
+			& 2,4 & \cpu{1054.80} 	& \ints{36600} 	& \cpu{417.99}	& \ints{14512}	& \cpu{3.58}	& \ints{1} \\
 			% nb this 22 would be a 21 with !CONSIDER_ONLY... ..but, if it's the biggest loss, really n b d
-			& 3,1 & \cpu{10.40}	& \ints{148}	& \cpu{7.72}	& \ints{22}	& \cpu{3.33}	& \ints{1} \\
-			& 3,2 & \cpu{708.86}	& \ints{21642}	& \cpu{196.45}	& \ints{6467}	& \cpu{3.23}	& \ints{1} \\
-			% 3,3 with -R was tried for about 90 min, and stopped with ETA of 2367715 / 2d22h
-			% without -R would of course be bigger still
-			& 4,1 & \cpu{106.21}	& \ints{3064}	& \cpu{11.29}	& \ints{130}	& \cpu{3.43}	& \ints{1} \\
-			& 4,2 & \ETAdag{qqq} 	& \ETAdag{qqq} 	& \ETAdag{22d 17h}	& \ETAdag{3129034}	& \cpu{3.28}	& \ints{1} \\
-			%[JOB 1] progress: 827215/3129034 brs (26.436748%), ETA 22d 7h 18m 34s (elapsed 9h 59m 56s)
-
-			% measured with !CONSIDER_ONLY_MOST_RECENT_DPOR_PREFERRED_TID
-			%& 2,2 & \cpu{9.40}	& \ints{108}	& \cpu{8.92}	& \ints{76}	& \cpu{3.37}	& \ints{1} \\
-			%& 3,1 & \cpu{10.30}	& \ints{132}	& \cpu{7.50}	& \ints{21}	& \cpu{3.64}	& \ints{1} \\
+			& 3,1 & \cpu{11.15}	& \ints{148}	& \cpu{7.76}	& \ints{22}	& \cpu{3.53}	& \ints{1} \\
+			& 3,2 & \cpu{717.53}	& \ints{21642}	& \cpu{217.30}	& \ints{6467}	& \cpu{3.52}	& \ints{1} \\
+			& 4,1 & \cpu{111.83}	& \ints{3064}	& \cpu{11.32}	& \ints{130}	& \cpu{3.49}	& \ints{1} \\
 			\hline
 			{\tt avl\_insert}
-			& 2,1 & \cpu{545.45}	& \ints{15052}	& \cpu{244.20}	& \ints{6288}	& \cpu{109.84}	& \ints{2774} \\
-			% (2,2) - bug found on regular w/ ?? ints, on retry sets w/ ?? ints, and codes w/ ?? ints
+			& 2,1 & \cpu{672.68}	& \ints{15125}	& \cpu{307.94}	& \ints{6287}	& \cpu{136.60}	& \ints{2774} \\
 			\hline
 			{\tt avl\_fixed}
-			& 2,1 & \cpu{642.73} 	& \ints{20416} 	& \cpu{293.70}	& \ints{9677}	& \cpu{108.69}	& \ints{2774} \\
+			& 2,1 & \cpu{739.60} 	& \ints{20459} 	& \cpu{332.32}	& \ints{9675}	& \cpu{122.40}	& \ints{2774} \\
 			& 2,2 & \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq} \\
@@
 			{\tt map\_basic}
-			& 2,1 & \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq} \\
+			& 2,1 & \cpu{1950.48}	& \ints{30719}	& \cpu{867.68}	& \ints{13237}	& \cpu{367.91}	& \ints{5446} \\
+			& 2,2 & \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq} \\
+			& 3,1 & \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq} \\
 			\hline
 			{\tt map\_basicer}
-			& 2,1 & \cpu{qqq} 	& \ints{qqq} 	& \cpu{qqq}	& \ints{qqq}	& \cpu{qqq}	& \ints{qqq} \\
-			& 2,2 & \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq} \\
+			& 2,1 & \cpu{28.40} 	& \ints{150} 	& \cpu{94}	& \ints{26.55}	& \cpu{14.88}	& \ints{9} \\
+			& 2,2 & \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \cpu{1285.06}	& \ints{21684} \\
 			& 3,1 & \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq} \\
@@
 		State space estimates measured after a timeout of 10 hours.
+		{\em Thesis draft note: I need to rerun these tests to ensure consistent cpu times with the next table,
+		but I expect the state space sizes not to change substantially.}
 		}
@@
 			{\tt avl\_insert}	& 1	& 2	& 7	& 20	\\ % 2,1 nb. only 5 drpps under -A -S but w/e
+			% nb less than 9 on stm
 			{\tt avl\_fixed}	& 1	& 2	& 9	& 24	\\ % 2,1 nb. again, only 5 under -A -S
-			% the extra 2 over basicer is from resize, which only 1 thread does, but K*N is 0 here anyway so vov
-			{\tt map\_basic}	& 1	& 4	& ?	& 0	\\ % lol
-			{\tt map\_basicer}	& 1	& 2	& ?	& ?	\\
-
+			% nb stm takes it down from 13 to 10 races
+			{\tt map\_basic}	& 1	& 4	& 13	& 36	\\ % 2,1
+			% mysteriously had this at 6 at one point but ive no idea where that came from
+			{\tt map\_basicer}	& 1	& 2	& 5	& 32	\\ % 2,2
+
+			% TODO: check these again, will 2,4 even complete on mutex
 			{\tt lock}({\tt spinlock})	& 1	& 2	& 4	& 28	\\ % 2,2
 			{\tt lock}({\tt spin\_fixed})	& 1	& 2	& 4	& 28	\\ % 2,2 or 4,1
-			{\tt lock}({\tt mutex})		& 1	& 2	& 2	& 40	\\ % 2,4
+			{\tt lock}({\tt mutex})		& 1	& 2	& 2	& 30	\\ % 2,3 gives 30; 2,4 would give 40
 			% not putting lock_fast because it like, doesn't have a "state space" per se
@@
 for even small thread/iteration counts,
-in one case failing to make any guarantee about even the minimum possible test configuration,
-and in another failing to verify that a previously-found bug had been fixed.
+in one case
+%failing to make any guarantee about even the minimum possible test configuration,
+%and in another
+failing to verify whether or not a previously-found bug had actually been fixed.
 In the following subsections I explore three possible mitigation approaches.
@@
 the data structure tests suddenly reach unprecedented thread/iteration counts with ease,
+% 'so on' here includes free/re-malloc mem conflicts; see comment in avl_mutex.cpp
 their state space growth now defined only by their internal conflicts from tree rebalancing, map resizing, and so on.
@@
 		\begin{tabular}{cc}
-		%\begin{tabular}{cc||r|r||r|r}
-		%	& & \multicolumn{2}{c||}{\bf DPOR} & \multicolumn{2}{c}{\bf DPOR + retry sets} \\
-		%	& & \cpu{\bf cpu (s)} & \ints{\bf SS size}
-		%	  & \cpu{\bf cpu (s)} & \ints{\bf SS size} \\
-		%	\bf test & \bf params & \em (or \ETAdag{\bf \em ETA}) & \em (or \ETAdag{\bf \em est.})
-		%	  & \em (or \ETAdag{\bf \em ETA}) & \em (or \ETAdag{\bf \em est.}) \\
 		\begin{tabular}{cc||r|r}
@@
 			{\tt lock}
-			& 2,1 & \cpu{3.19} & \ints{4} \\
+			& 2,1 & \cpu{3.41} & \ints{4} \\
 			({\tt spinlock})
-			& 2,2 & \cpu{184.51} & \ints{1702} \\
+			& 2,2 & \cpu{198.77} & \ints{1702} \\
 			% if lock(spin_fixed) will time out on 2,3 and 3,2, this is guaranteed to aswell
-			% & 2,3 & \\ % \cpu{0} & \ints{0} \\
-			& 3,1 & \cpu{33.29} & \ints{246} \\
-			% & 3,2 & \\ % \cpu{0} & \ints{0} \\
+			& 3,1 & \cpu{35.28} & \ints{246} \\
 			\hline
 			{\tt lock}
-			& 2,1 & \cpu{qqq} & \ints{4} \\
+			& 2,1 & \cpu{3.55} & \ints{4} \\
 			({\tt spin\_fixed})
-			& 2,2 & \cpu{104.77} & \ints{998} \\
+			& 2,2 & \cpu{105.57} & \ints{998} \\
 			& 2,3 & \ETAdag{qqq} & \ETAdag{qqq} \\
-			& 3,1 & \cpu{26.96} & \ints{186} \\
-			\hline \multicolumn{4}{c}{everything below this line needs rerun} \\
-			& 3,2 & \ETAdag{8y 153d} & \ETAdag{1590416} \\ % caus ok
-			% [JOB 4] progress: 187162/1590416 brs (11.768113%), ETA 8y 152d 16h 25m 38s (elapsed 9h 59m 49s)
-			& 4,1 & \cpu{30408.58} & \ints {282084} \\ % causality ok
-			% (2,1) with isfree fix
-			%%%% total time elapsed: 10s
-			%%%% [JOB 4] COMPLETE (31 interleavings tested; 2s elapsed)
-			%%%% total CPU time consumed: 13s (13256415 usecs) (core saturation: 30%)
-			% (2,2) with isfree fix
-			% total time elapsed: 27m 39s
-			% [JOB 4] COMPLETE (15076 interleavings tested; 27m 28s elapsed)
-			% total CPU time consumed: 27m 42s (1662213886 usecs) (core saturation: 25%)
-			% (2,3)
-			% [JOB 4] progress: 242266/2003906 brs (12.089688%), ETA 3d 20h 2m 12s (elapsed 9h 59m 49s)
-			% [JOB 4] TIMED OUT (12.089688%; ETA 3d 20h 2m 12s)
-			% (3,1) with isfree fix
-			% hot start confirmed
-			%%%% total time elapsed: 41m 26s
-			%%%% [JOB 4] COMPLETE (24790 interleavings tested; 41m 18s elapsed)
-			%%%% total CPU time consumed: 41m 28s (2488953483 usecs) (core saturation: 25%)
-			% (3,2)
-			% [JOB 4] progress: 187522/6514618 brs (2.878480%), ETA 207d 20h 23m 10s (elapsed 9h 59m 51s)
-			% [JOB 4] TIMED OUT (2.878480%; ETA 207d 20h 23m 10s)
+			& 3,1 & \cpu{28.27} & \ints{186} \\
+			& 3,2 & \ETAdag{qqq} & \ETAdag{qqq} \\
+			& 4,1 & \cpu{qqq} & \ints {qqq} \\ % was: 30408, 282084 pre mkrun
 			\hline
 			{\tt lock}
-			& 2,1 & \cpu{2.49} & \ints{4}		\\ % certified causality ok
+			& 2,1 & \cpu{3.44} & \ints{4}		\\
 			({\tt mutex})
-			& 2,2 & \cpu{12.72} & \ints{180}	\\ % certified causality ok
-			& 2,3 & \cpu{273.45} & \ints{8356}	\\ % certified causality ok
-			& 2,4 & \cpu{13545.35} & \ints{384512}	\\ % certified causality ok
-			& 3,1 & \cpu{11.21} & \ints{132}	\\ % certified causality ok
-			& 3,2 & \cpu{25434.85} & \ints{718690}	\\ % certified causality ok
-			\hline \multicolumn{4}{c}{everything above this line needs rerun} \\
-			& 4,1 & \cpu{547.96} & \ints{15064}	\\
-			%%%%%%%%%%% htm_mutex
-			%%%% -- nb. these numbers were obtained with a oops-exploded state space
-			%%%% from a misplaced assert on hm->held
-			%%%% (nb. also before dpor-preferred-tid, although that doesn't seem to affect ss size)
-			% % 2,1
-			% total time elapsed: 6s
-			% [JOB 2] COMPLETE (24 interleavings tested; 3s elapsed)
-			% total CPU time consumed: 9s (9642348 usecs) (core saturation: 35%)
-			% % 2,2
-			% total time elapsed: 1m 20s
-			% [JOB 2] COMPLETE (2672 interleavings tested; 1m 16s elapsed)
-			% total CPU time consumed: 1m 23s (83001932 usecs) (core saturation: 25%)
-			% % 2,3
-			% total time elapsed: 2h 3m 45s
-			% [JOB 2] COMPLETE (220338 interleavings tested; 2h 3m 40s elapsed)
-			% total CPU time consumed: 2h 3m 47s (7427873599 usecs) (core saturation: 25%)
-			% % 3,1
-			% total time elapsed: 51s
-			% [JOB 2] COMPLETE (1594 interleavings tested; 47s elapsed)
-			% total CPU time consumed: 54s (54572810 usecs) (core saturation: 26%)
-			%%%% -- nb. better version
-			% % 2,1
-			% total time elapsed: 6s
-			% [JOB 2] COMPLETE (22 interleavings tested; 2s elapsed)
-			% total CPU time consumed: 9s (9415448 usecs) (core saturation: 34%)
-			% % 2,2
-			% total time elapsed: 40s
-			% [JOB 2] COMPLETE (1208 interleavings tested; 37s elapsed)
-			% total CPU time consumed: 44s (44068381 usecs) (core saturation: 27%)
-			% % 2,3
-			% total time elapsed: 44m 4s
-			% [JOB 2] COMPLETE (71136 interleavings tested; 44m 0s elapsed)
-			% total CPU time consumed: 44m 7s (2647380230 usecs) (core saturation: 25%)
-			% % 2,4
-			% [JOB 2] progress: 957248/3223055 brs (29.700013%), ETA 2d 1h 24m 44s (elapsed 9h 59m 56s)
-			% total time elapsed: 10h 0s
-			% [JOB 2] TIMED OUT (29.700013%; ETA 2d 1h 24m 44s)
-			% total CPU time consumed: 10h 3s (36003560403 usecs) (core saturation: 25%)
-			% % 3,1
-			% total time elapsed: 38s
-			% [JOB 2] COMPLETE (1148 interleavings tested; 35s elapsed)
-			% total CPU time consumed: 42s (42108005 usecs) (core saturation: 27%)
-			% % 3,2
-			% [JOB 2] progress: 920684/4793666 brs (19.206260%), ETA 11d 4h 17m 50s (elapsed 9h 59m 56s)
-			% total time elapsed: 10h 0s
-			% [JOB 2] TIMED OUT (19.206260%; ETA 11d 4h 17m 50s)
-			% total CPU time consumed: 10h 3s (36003199876 usecs) (core saturation: 25%)
-			% % 4,1
-			% total time elapsed: 1h 33m 3s
-			% [JOB 2] COMPLETE (159664 interleavings tested; 1h 32m 59s elapsed)
-			% total CPU time consumed: 1h 33m 6s (5586369677 usecs) (core saturation: 25%)
+			& 2,2 & \cpu{16.83} & \ints{180}	\\
+			& 2,3 & \cpu{405.21} & \ints{9372}	\\
+			& 2,4 & \cpu{qqq} & \ints{qqq}	\\ % was: 15266 sec 384512 pre mkrun
+			& 3,1 & \cpu{15.00} & \ints{132}	\\
+			%& 3,2 & \cpu{29432.87} & \ints{707086}	\\ % pre mkrun fix (let alone bottom1)
+			& 3,2 & \ETAdag{qqq} & \ETAdag{qqq} \\
+			& 4,1 & \cpu{665.89} & \ints{15064}	\\
 			\hline
 			{\tt lock\_fast}
-			& 2,1 & \cpu{3.36} & \ints{1}	\\
+			& 2,1 & \cpu{3.25} & \ints{1}	\\
 			({\tt spin\_fixed})
-			& 9,9 & \cpu{4.91} & \ints{1}	\\
+			& 9,9 & \cpu{4.61} & \ints{1}	\\
 			\hline
 			{\tt lock\_fast}
-			& 2,1 & \cpu{3.40} & \ints{1}	\\
+			& 2,1 & \cpu{3.19} & \ints{1}	\\
 			({\tt mutex})
-			& 9,9 & \cpu{4.71} & \ints{1}	\\
+			& 9,9 & \cpu{4.62} & \ints{1}	\\
 			\end{tabular}
@@
 			{\tt avl\_mutex}
-			& 2,1 & \cpu{2.49} & \ints{3} \\
-			& 2,2 & \cpu{2.44} & \ints{3} \\
-			& 2,3 & \cpu{2.45} & \ints{3} \\
-			& 2,4 & \cpu{2.41} & \ints{3} \\
-			& 3,1 & \cpu{2.69} & \ints{12} \\
-			& 3,2 & \cpu{2.67} & \ints{12} \\
-			& 3,3 & \cpu{2.76} & \ints{13} \\
-			& 3,4 & \cpu{2.74} & \ints{13} \\
-			& 4,1 & \cpu{4.03} & \ints{60} \\
-			& 4,2 & \cpu{4.34} & \ints{66} \\
-			& 4,3 & \cpu{4.53} & \ints{71} \\
-			& 4,4 & \cpu{4.75} & \ints{73} \\
-			%%%% avl_mutex
-			% % 2,1
-			% [JOB 0] COMPLETE (3 interleavings tested; 2s elapsed)
-			% total CPU time consumed: 2s (2489515 usecs) (core saturation: 24%)
-			% % 2,2
-			% [JOB 0] COMPLETE (3 interleavings tested; 1s elapsed)
-			% total CPU time consumed: 2s (2443451 usecs) (core saturation: 24%)
-			% % 2,3
-			% [JOB 0] COMPLETE (3 interleavings tested; 1s elapsed)
-			% total CPU time consumed: 2s (2448127 usecs) (core saturation: 24%)
-			% % 2,4
-			% [JOB 0] COMPLETE (3 interleavings tested; 1s elapsed)
-			% total CPU time consumed: 2s (2407903 usecs) (core saturation: 24%)
-			% % 3,1
-			% [JOB 0] COMPLETE (12 interleavings tested; 2s elapsed)
-			% total CPU time consumed: 2s (2685913 usecs) (core saturation: 24%)
-			% %3,2
-			% [JOB 0] COMPLETE (12 interleavings tested; 2s elapsed)
-			% total CPU time consumed: 2s (2666826 usecs) (core saturation: 24%)
-			% % 3,3
-			% [JOB 0] COMPLETE (13 interleavings tested; 2s elapsed)
-			% total CPU time consumed: 2s (2761157 usecs) (core saturation: 24%)
-			% % 3,4
-			% [JOB 0] COMPLETE (13 interleavings tested; 2s elapsed)
-			% total CPU time consumed: 2s (2738169 usecs) (core saturation: 24%)
-			% % 4,1
-			% [JOB 0] COMPLETE (60 interleavings tested; 3s elapsed)
-			% total CPU time consumed: 4s (4034411 usecs) (core saturation: 24%)
-			% % 4,2
-			% [JOB 0] COMPLETE (66 interleavings tested; 3s elapsed)
-			% total CPU time consumed: 4s (4338442 usecs) (core saturation: 24%)
-			% % 4,3
-			% [JOB 0] COMPLETE (71 interleavings tested; 4s elapsed)
-			% total CPU time consumed: 4s (4526641 usecs) (core saturation: 24%)
-			% % 4,4
-			% [JOB 0] COMPLETE (73 interleavings tested; 4s elapsed)
-			% total CPU time consumed: 4s (4752864 usecs) (core saturation: 24%)
+			& 2,1 & \cpu{3.48} & \ints{7} \\
+			& 2,2 & \cpu{6.25} & \ints{85} \\
+			& 2,3 & \cpu{24.69} & \ints{561} \\
+			& 2,4 & \cpu{217.98} & \ints{4984} \\
+			& 3,1 & \cpu{8.26} & \ints{129} \\
+			& 3,2 & \cpu{1403.46} & \ints{30653} \\
+			& 3,3 & \ETAdag{qqq} & \ETAdag{qqq} \\
+			& 4,1 & \cpu{199.96} & \ints{4488} \\
+			& 4,2 & \ETAdag{qqq} & \ETAdag{qqq} \\
 			\cline{1-4}
 			{\tt map\_mutex}
-			& 2,1 & \\ %\cpu{0} & \ints{0} \\
-			& 2,2 & \\ %\cpu{0} & \ints{0} \\
-			& 2,3 & \\ %\cpu{0} & \ints{0} \\
-			& 2,4 & \\ %\cpu{0} & \ints{0} \\
-			& 3,1 & \\ %\cpu{0} & \ints{0} \\
-			& 3,2 & \\ %\cpu{0} & \ints{0} \\
-			& 3,3 & \\ %\cpu{0} & \ints{0} \\
-			& 3,4 & \\ %\cpu{0} & \ints{0} \\
-			& 4,1 & \\ %\cpu{0} & \ints{0} \\
-			%& 4,2 & \\ %\cpu{0} & \ints{0} \\
-			\hline \multicolumn{4}{c}{everything above this line needs rerun} \\
+			& 2,1 & \cpu{39.81} & \ints{83} \\
+			& 2,2 & \cpu{qqq} & \ints{qqq} \\ % 30k and rising, looks like mb 5 hr? or 10+hr. 50/50 odds
+			& 2,3 & \ETAdag{qqq} & \ETAdag{qqq} \\
+			& 3,1 & \cpu{qqq} & \ints{qqq} \\ % 200k and rising, same tbh
+			& 3,2 & \ETAdag{qqq} & \ETAdag{qqq} \\
+			& 4,1 & \ETAdag{qqq} & \ETAdag{qqq} \\
+			\\
+			\\
+			\\
+			\\
+			\\
 		\end{tabular}
@@
 		by verifying HTM mutex implementations separately.
-		Tested with STM semantics, as both lock implementations include a retry loop.
-		%Run with {\tt landslide -M -X -A -S}, i.e.,
-		%conflict and explicit aborts tested separately,
-		%and retry aborts suppressed
-		%(both lock implementations include a retry loop).
-		%The reduced {\tt avl\_mutex} and {\tt map\_mutex} are, of course, no longer transactional at all.
-		{\bf RERUN NOTE: this entire table needs to be rerun on account of dpor-preferred-tid}
-		}
+		Tested with STM semantics, as both lock implementations include a retry loop.}
 	\label{tab:tm-verifs2}
@@
 than {\tt mutex}
-(roughly 9 interleavings per second for the former, 28 for the latter),
+% TODO recalc
+(roughly 9 interleavings per second for the former, 14 for the latter),
 because while {\tt mutex} abstracts away threads needing to wait their turn for the critical section
@@
 {\tt mutex} can be seen in turn as a further abstraction reduction of {\tt spinlock}.
-
-%%%% map mutex
-% % TODO - bc of this weird nonlinear behaviour around 2,1+, yo ushoud prob try 2,2 for map_basic as well
-% % 2,1
-% total time elapsed: 19s
-% [JOB 7] COMPLETE (57 interleavings tested; 5s elapsed)
-%        PPs: { 'mutex_lock' 'mutex_unlock' 'DR @ 0x100053b' [NONDET] 'DR @ 0x1000552' [NONDET] 'DR @ 0x1000521' [NONDET]
-%        'DR @ 0x100055d' [NONDET] 'DR @ 0x100055a' [NONDET] 'DR @ 0x1000916' [NONDET] 'DR @ 0x1000919' [NONDET] }
-% total CPU time consumed: 31s (31198379 usecs) (core saturation: 39%)
-% % 2,2
-% total time elapsed: 3s
-% [JOB 0] COMPLETE (6 interleavings tested; 2s elapsed)
-%        PPs: { 'mutex_lock' 'mutex_unlock' }
-% =========================
-% progress report thr exiting
-% total CPU time consumed: 3s (3192831 usecs) (core saturation: 24%)
-% % 2,3
-% ==== PROGRESS REPORT ====
-% total time elapsed: 3s
-% [JOB 0] COMPLETE (6 interleavings tested; 2s elapsed)
-%        PPs: { 'mutex_lock' 'mutex_unlock' }
-% =========================
-% progress report thr exiting
-% total CPU time consumed: 3s (3240367 usecs) (core saturation: 24%)
-% 2,4
-% 3,1
-% 3,2
 

### 2018-08-07 thesis/landslide.tex
revise tm, cleanup todos in bgnd/landslide, fix cross-chapter section references, etc -- sending tm chapter off to committee here

@@
 		abandoning all subset jobs even if they might find bugs faster
-		% TODO: more speicfic sectio nreference
-		(\sect{\ref{chap:tm}}, incompatible with {\tt -C})
+		(\sect{\ref{sec:tm-eval}}, incompatible with {\tt -C})
 	\item {\tt -H}: use Limited Happens-Before for data-race analysis (\sect{\ref{sec:background-hb}})
@@
 	\item {\tt -X}: support transactional memory (Chapter~\ref{chap:tm})
-		% TODO: make a more specific section reference
-	\item {\tt -A}: support multiple abort codes during transaction failure (Chapter~\ref{chap:tm});
+	\item {\tt -A}: support multiple abort codes during transaction failure (\sect{\ref{sec:tm-implementation}});
 		required for testing programs which behave differently under different abort circumstances,
 		but impacts the state space size
+	\item {\tt -S}: suppress retry aborts during transaction failure (\sect{\ref{sec:tm-implementation}})
+	\item {\tt -R}: enable retry-set state space reduction for transactional tests (\sect{\ref{sec:tm-implementation}})
 	\item {\tt -P}: support Pintos architecture (enabled automatically when {\tt pintos-setup.sh} is run)
@@
 	\item {\tt interrupts\_enabled()} queries the CPU's interrupt flag ({\tt eflags:IF}).
+	\item {\tt cause\_transaction\_failure()} forces {\tt \_xbegin()} to return a specified abort code.
 \end{itemize}
@@
 if the kernel is not ready the interrupt may not be received for a long time.
-Also, use of {\tt cause\_timer\_interrupt\_immediately()} must not be used while the kernel is not ready.
+Also, {\tt cause\_timer\_interrupt\_immediately()} must not be used while the kernel is not ready.
 

### 2018-08-07 thesis/tm.tex
revise tm, cleanup todos in bgnd/landslide, fix cross-chapter section references, etc -- sending tm chapter off to committee here

@@
 \begin{itemize}
-	\item  {\sf begin} begins a transaction,
-		staging any subsequent shared memory accesses in some temporary thread-local storage,
+	\item  {\sf xbegin} begins a transaction,
+		staging any subsequent shared memory accesses in some temporary thread-/processor-local storage,
 		and checking for conflicts with the accesses of any other threads or CPUs.
@@
 		indicating the programmer should fall back to some other, possibly slower, synchronization method.
-	\item {\sf end} ends a transaction,
+	\item {\sf xend} ends a transaction,
 		attempting to commit all staged accesses to the shared memory atomically with respect to
@@
 		they are instead discarded and execution state reverts to the {\sf begin} with an error code as described above.
-	\item {\sf abort} explicitly aborts a transaction,
+	\item {\sf xabort} explicitly aborts a transaction,
 		regardless of any memory conflicts,
@@
 \begin{figure}[h]
-	\input{fig-htm-example}
+	\begin{center}
+		\begin{tabular}{cl}
+			1 & \texttt{\flow{if} ((status = \call{\_xbegin}()) == \const{\_XBEGIN\_STARTED}) \{} \\
+			2 & \texttt{~~~~x++;} \\
+			3 & \texttt{~~~~\call{\_xend}();} \\
+			4 & \texttt{\} \flow{else} \{} \\
+			5 & \texttt{~~~~\call{mutex\_lock}(\&m);} \\
+			6 & \texttt{~~~~x++;} \\
+			7 & \texttt{~~~~\call{mutex\_unlock}(\&m);} \\
+			8 & \texttt{\}} \\
+		\end{tabular}
+	\end{center}
 	\caption{Example TSX program.}
@@
 results in multiple memory accesses which can be interleaved with other threads.
-% TODO: fix line numbers
-% TODO: put "void count" back in the mario-fixed example program, and fix ALL line numbers thruout proofs
 \[
@@
 \section{Implementation}
+\label{sec:tm-implementation}
 
@@
 			\texttt{\} \flow{else} \{} \\
-			\texttt{~~~~\call{mutex\_lock}(\&m);} \\
-			\texttt{~~~~\flow{assert}(foo + bar < \const{2});} \\
-			\texttt{~~~~\call{mutex\_unlock}(\&m);} \\
+			%\texttt{~~~~\call{mutex\_lock}(\&m);} \\
+			\texttt{~~~~\flow{assert}{}(foo + bar < \const{2});} \\
+			%\texttt{~~~~\call{mutex\_unlock}(\&m);} \\
 			\texttt{\}} \\
@@
 \section{Evaluation}
+\label{sec:tm-eval}
 
@@
 most of these tests are too small for its approach to pay off,
-% TODO: re check this after re running
-with {\tt swapbug}(3,1) and {\tt avl\_insert} as its notable wins.
+with {\tt swapbug} and {\tt avl\_insert} as its notable wins.
 While plenty more wins were observed in \sect{\ref{sec:quicksand-eval}},
@@
 For the test cases with no bugs found, I sought to provide Landslide's verification guarantee
-% TODO better section reference
-(\sect{\ref{sec:quicksand-id}})
+(\sect{\ref{sec:quicksand-soundness}})
 for up to as many threads and test iterations as possible under a reasonable time limit.
@@
 		Run with {\tt -M -X}, plus any additional reduction options listed.
-		``Baseline DPOR'' always tests every abort path, without distinguish among failure reasons
+		``Baseline DPOR'' always tests every abort path, without distinguishing among failure reasons
 		(i.e., injecting only {\tt \_XABORT\_RETRY}).
@@
 		``STM'' suppresses retry aborts and dynamically detects when to inject conflict aborts and so on
-		(\sect{\ref{sec:txn-abort-codes}}),
-		ultimately matching STM's execution semantics.
-		State space estimates measured after a timeout of 10 hours.
-		{\em Thesis draft note: I need to rerun these tests to ensure consistent cpu times with the next table,
-		but I expect the state space sizes not to change substantially.}
+		(\sect{\ref{sec:txn-abort-codes}}).
+		State space estimates measured after a timeout of 10 hours (and include those 10 hours in the predicted total).
 		}
@@
 reached among Landslide's verifications
-(i.e., $({\sf txn}+{\sf sync}+{\sf race}) \times K \times N$ for the highest $K \times N$ completed.
+(i.e., $({\sf txn}+{\sf sync}+{\sf race}) \times K \times N$ for the highest $K \times N$ completed).
 %%%% on second thought, i'm not gonna do this -- any branch that xend shows up in, is gonna lose any of the pps coming
@@
 the benefit is clear:
-the data structure tests suddenly reach unprecedented thread/iteration counts with ease,
+the data structure tests' state spaces become much more tractable,
+%the data structure tests suddenly reach unprecedented thread/iteration counts with ease,
 % 'so on' here includes free/re-malloc mem conflicts; see comment in avl_mutex.cpp
@@
 % TODO recalc
-(roughly 9 interleavings per second for the former, 14 for the latter),
+(roughly 9 interleavings per second for the former, 22 for the latter),
 because while {\tt mutex} abstracts away threads needing to wait their turn for the critical section

### 2018-08-08 thesis/tm.tex
more tm reruns (same rerun pass as last commit, don't worry not crying yet)

@@
 			% TODO
-			%\multicolumn{9}{c}{everything above this line needs rerun for dpor-preferred tid + thr_bottom1} \\
+			\multicolumn{9}{c}{everything above this line needs rerun for dpor-preferred tid + thr\_bottom1} \\
 
@@
 			& 3,1 & \cpu{10.57}	& \ints{120}	& \cpu{9.23}	& \ints{64}	& \cpu{7.88}	& \ints{40} \\
-			& 3,2 & \cpu{2492.17}	& \ints{60606}	& \cpu{2040.98}	& \ints{44862}	& \cpu{565.97}	& \ints{14304} \\
-			& 4,1 & \cpu{121.96}	& \ints{3006}	& \cpu{61.85}	& \ints{1296}	& \cpu{39.23}	& \ints{848} \\
-			\hline
-			{\tt swap}
-			& 2,1 & \cpu{38.76}	& \ints{99}	& \cpu{36.91}	& \ints{59}	& \cpu{3.29}	& \ints{4} \\
-			& 2,2 & \cpu{12009.77}	& \ints{233396}	& \cpu{497.11}	& \ints{11427}	& \cpu{461.54}	& \ints{9046} \\
-			& 3,1 & \cpu{2975.30}	& \ints{60912}	& \cpu{522.07}	& \ints{10706}	& \cpu{86.21}	& \ints{1014} \\
-			% measured with !CONSIDER_ONLY_MOST_RECENT_DPOR_PREFERRED_TID
-			%& 2,1 & \cpu{38.43}	& \ints{99}	& \cpu{36.12}	& \ints{59}	& \cpu{3.34}	& \ints{4} \\
-			%& 3,1 & \cpu{vvv}	& \ints{vvv}	& \cpu{730.94}	& \ints{17112}	& \cpu{74.53}	& \ints{1064} \\
-			\hline
+			% stm ok, rts ok, baseline alone needs redo
+			& 3,2 & \cpu{2492.17}	& \ints{60606}	& \cpu{2363.41}	& \ints{44862}	& \cpu{639.62}	& \ints{14304} \\
 			% TODO: everything above this line needs rerun
 			% TODO: everything below is fine up to dpor-preferred & thr_bottom1 updates
+			\multicolumn{8}{c}{everything above this line needs rerun for dpor-preferred tid + thr\_bottom1} \\
+			& 4,1 & \cpu{129.25}	& \ints{3006}	& \cpu{64.91}	& \ints{1296}	& \cpu{40.78}	& \ints{848} \\
+			\hline
+			{\tt swap}
+			& 2,1 & \cpu{65.72}	& \ints{99}	& \cpu{66.05}	& \ints{59}	& \cpu{3.40}	& \ints{4} \\
+			& 2,2 & \cpu{18124.06}	& \ints{277824}	& \cpu{1030.23}	& \ints{19542}	& \cpu{703.82}	& \ints{11600} \\
+			& 3,1 & \cpu{3820.64}	& \ints{60912}	& \cpu{608.48}	& \ints{10706}	& \cpu{89.69}	& \ints{1014} \\
+			\hline
 			{\tt fig63}
@@
 		State space estimates measured after a timeout of 10 hours (and include those 10 hours in the predicted total).
+		{\em Thesis draft note: Some of these tests still need to be rerun on account of landslide implementation updates
+		to ensure consistent cpu times with the next table,
+		but I expect the state space sizes not to change substantially.}
 		}
@@
 			& 2,1 & \cpu{39.81} & \ints{83} \\
-			& 2,2 & \cpu{qqq} & \ints{qqq} \\ % 30k and rising, looks like mb 5 hr? or 10+hr. 50/50 odds
-			& 2,3 & \ETAdag{qqq} & \ETAdag{qqq} \\
-			& 3,1 & \cpu{qqq} & \ints{qqq} \\ % 200k and rising, same tbh
-			& 3,2 & \ETAdag{qqq} & \ETAdag{qqq} \\
-			& 4,1 & \ETAdag{qqq} & \ETAdag{qqq} \\
+			& 2,2 & \ETAdag{26h 21m} & \ETAdag{1085126} \\
+			% [JOB 9] progress: 496530/1085126 brs (45.757812%), ETA 16h 21m 0s (elapsed 9h 59m 38s)
+			% total CPU time consumed: 10h 19s (36019507965 usecs) (core saturation: 12%)
+			& 3,1 & \ETAdag{qqq} & \ETAdag{qqq} \\
 			\\

### 2018-08-09 thesis/landslide.tex
friendly dpor description (and a bit about estimation)

@@
 		abandoning all subset jobs even if they might find bugs faster
-		(\sect{\ref{sec:tm-eval}}, incompatible with {\tt -C})
+		(\sect{\ref{sec:tm-eval}}, incompatible with {\tt -C}).
+		According to \sect{\ref{sec:quicksand-soundness}}'s soundness proofs,
+		this is equivalent to {\tt -0}
+		(and according to my experience, way faster as well).
 	\item {\tt -H}: use Limited Happens-Before for data-race analysis (\sect{\ref{sec:background-hb}})
@@
 
-\subsection{State space estimation}
-\label{sec:landslide-estimate}
-
-% TODO
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
 \subsection{Dynamic Partial Order Reduction}
@@
 
+\subsubsection{Example}
+
+Although a superhuman reader may quickly reach intuitive understanding
+of complex algorithms from dense prose and mathematical notation alone,
+I also present here a friendly example of using DPOR on Figure~\ref{fig:concurrency-bug}'s example program,
+whose original state space is shown in Figure~\ref{fig:tree}.
+%Acknowledging that quickly gaining an intuitive understanding for an algorithm
+%is basically impossible from just paragraphs of text,
+%I conclude this section with an example of a single iteration of DPOR
+%on the example state space from Figures~\ref{fig:concurrency-bug} and \ref{fig:tree}.
+First, let us consider a single iteration of DPOR, applied after executing the first branch.
+The result is shown in Figure~\ref{fig:dpor-example-0}.
+
+\begin{figure}[h]
+	\begin{center}
+		\includegraphics[width=\textwidth]{dpor-example-0.pdf}
+	\end{center}
+	\caption{Result of a single iteration of DPOR.}
+	\label{fig:dpor-example-0}
+\end{figure}
+
+In this interleaving, DPOR identifies 3 memory conflicts, two read/write and one write/write,
+among the threads' 4 accesses to {\tt x}.
+For each such pair, it ``marks'' an alternate interleaving,
+which shall begin by preempting the thread of the first half of the conflict just before its execution thereof.
+The ultimate goal is to execute an interleaving which reorders the conflict, which may expose new behaviour.
+These marked interleavings form a work-queue which defines the exploration.
+DPOR consumes from this set in depth-first order
+(note the reversed order of $\bigstar$1 and $\bigstar$2)
+to avoid storing in memory any representation of exponentially-sized subtrees outside of the current branch.
+Note also that in $\bigstar$2,
+the reordered {\tt tmp2 <- x} is not directly part of the memory conflict which marked it,
+but it must be executed first to reach the conflicting {\tt x <- tmp2}.
+
+Now, let us run multiple iterations of DPOR
+to advance through the first half of the full state space shown in Figure~\ref{fig:tree}(b).
+Figure~\ref{fig:dpor-example-1} shows the outcome
+(with the previously-marked $\bigstar$2, now re-labeled as $\bigstar$3, having its subtree abbreviated).
+
+\begin{figure}[h]
+	\begin{center}
+		\includegraphics[width=\textwidth]{dpor-example-1.pdf}
+	\end{center}
+	\caption{Result of 3 DPOR iterations, pruning 7 redundant branches.}
+	\label{fig:dpor-example-1}
+\end{figure}
+
+After marking $\bigstar$1 and $\bigstar$3 from Figure~\ref{fig:dpor-example-0}'s interleaving, now labeled A,
+DPOR advances to interleaving B, preferring to schedule the second thread before switching back to the first
+to ensure the memory access is properly reordered.
+From there, it identifies a new memory conflict, marks $\bigstar$2, and advances to C,
+where it finds no memory conflicts that would mark anything not already marked and/or explored
+(memory conflicts that were already reordered in old branches are not highlighted with arrows).
+From C, $\bigstar$3 alone remains in the work-queue,
+so DPOR advances to the second (symmetric) half of the state space,
+skipping (thereby pruning) branches D through J.
+
+To see why branches D through J need not be tested, consider that each thread's {\tt tmpN++} is a thread-local event,
+participating in no memory conflicts,
+and hence any two interleavings differing only by reordering {\tt tmpN++}s must be equivalent.
+The dashed blue arrows denote such equivalences;
+note the two disjoint equivalence classes \{B,E,G,I\} and \{C,D,F,H,J\},
+distinguished by the order of the two final {\tt x <- tmpN}s.
+Note also that
+%(as shown in Figure~\ref{fig:tree}(b)),
+although B and C also have the same outcome ({\tt x==1}),
+this depends on the {\em values} written to memory rather than {\em addresses}
+(and would change if one thread's {\tt tmpN++} were a {\tt tmpN+=2}, for example),
+which DPOR does not consider.
+Recent work \cite{mcr} has extended DPOR to find such value-based equivalences,
+although is beyond this explanation's scope.
+
+Finally, let us consider the final result after DPOR runs out of remaining unexplored marked branches,
+shown in Figure~\ref{fig:dpor-example-2}.
+
+\begin{figure}[h]
+	\begin{center}
+		\includegraphics[width=\textwidth]{dpor-example-2.pdf}
+	\end{center}
+	\caption{DPOR's termination state, having reduced 20 interleavings to 6.}
+	\label{fig:dpor-example-2}
+\end{figure}
+
+Ultimately, the second half of the state space is pruned symmetrically.
+In general, the number of ways to interleave two threads executing $N$ and $M$ events each is given by ${N+M \choose N}$;
+in this case, the original state space's size was ${3+3 \choose 3} = 20$.
+DPOR's reduction is characterized by replacing $N$ and $M$ with the number of {\em conflicting} events only;
+in this case, ignoring all {\tt tmpN++} reorderings and testing only ${2+2 \choose 2} = 6$ branches.
+
+However, as the legend spoils, this subset of the state space still contains redundant interleavings,
+arising from the fact that one pair of those 4 events is two reads, and hence not actually in conflict.
+Visual inspection shows that $\bigstar$5, while locally justified in trying to reorder {\tt tmp1 <- x} before {\tt x <- tmp2},
+effectively serves % oops
+only to reorder it with {\tt tmp2 <- x}
+relative to the first symmetric subtree.
+Landslide identifies this equivalence by checking,
+%in this case during the iteration of DPOR after interleaving K,
+in this case after K,
+that if the first event to be reordered (here, {\tt tmp <- x})
+has already been tested in an equivalent reordering around any number of preceding events (here, {\tt tmp2 <- x} and {\tt tmp2++}),
+then the newly marked subtree is safe to prune.
+The next section describes the implementation of this check in general.
+
 \subsubsection{Sleep set reduction}
@@
 
+% TODO: condense this by removing the 'new' example and just continuing with the previous section's
 In the presence of non-conflicting transitions as well as conflicting ones,
@@
 
-\subsubsection{Example}
-
-Acknowledging that quickly gaining an intuitive understanding for an algorithm
-is basically impossible from just paragraphs of text,
-I conclude this section with an example of a single iteration of DPOR
-on the example state space from Figures~\ref{fig:concurrency-bug} and \ref{fig:tree}.
-
-% TODO
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\subsection{State space estimation}
+\label{sec:landslide-estimate}
+
+For both the user's convenience and for Quicksand's prioritization algorithm (\sect{\ref{sec:quicksand-id}}),
+Landslide attempts to guess how big partially-explored state spaces will ultimately end up being upon completion.
+Because the backtracking implementation uses checkpointing
+rather than replaying similar interleavings' shared execution prefixes from the beginning \sect{\ref{sec:landslide-timetravel}},
+the total number of interleavings (i.e., leaf nodes in the execution tree)
+must be estimated separately from the total runtime (i.e., sum of all edge weights in the tree).
+
+% TODO: once dpor example tree is made
+% switch this to that
+As a concrete example, consider the state space of Figure~\ref{fig:tree}(b) (untouched by DPOR),
+and suppose each transition to the next preemption point takes 1 second to execute.
+While the first branch executes in 5 seconds,
+the second branch, sharing the first transition as a common prefix, takes 4 seconds,
+and the one after that only 2;
+the state space being ultimately completed in 33 seconds.
+Even with perfect hindsight,
+na\"ively multiplying the total interleavings (10)
+by the total execution time per branch (5) would double-count common prefixes
+and grossly overestimate (50) the total runtime.
+
+Hence, Landslide uses two differently suitable algorithms for each of size and runtime estimation:
+the Weighted Backtrack Estimator (WBE) and the Recursive Estimator (RE), respectively,
+first introduced in \cite{estimating-search-tree-size} and later adapted to DPOR by \cite{estimation}.
+In principle, both calculate the current progress as a proportion of the expected total
+by counting how many branches DPOR has marked for future exploration (\sect{\ref{sec:landslide-explore}})
+and assuming the sizes of their resulting subtrees are predicted by the known sizes of similar already-explored subtrees.
+In practice, the calculation strategy differs between the two approaches,
+which can occasionally result in drastically differing outputs (\sect{\ref{sec:tm-verif}}).
+I summarize their respective implementations here.
+% TODO more specific section reference
+Chapter~\ref{chap:warpzone} will discuss their limitations and some opportunities for future improvement.
+
+\subsubsection{Size (Weighted Backtrack) estimation}
+
+\subsubsection{Runtime (Recursive) estimation}
 

### 2018-08-09 thesis/overview.tex
revise tm, cleanup todos in bgnd/landslide, fix cross-chapter section references, etc -- sending tm chapter off to committee here
friendly dpor description (and a bit about estimation)

@@
 \begin{figure}[h]
-	\input{fig-htm-example}
+	\begin{center}
+		\begin{tabular}{l}
+		\texttt{\ctype{void} \call{count}() \{} \\
+		\texttt{~~~~\flow{for} (\ctype{int} i = \const{0}; i < \const{1000}; i++) \{} \\
+		\texttt{~~~~~~~~\flow{if} ((status = \call{\_xbegin}()) == \const{\_XBEGIN\_STARTED}) \{} \\
+		\texttt{~~~~~~~~~~~~x++;} \\
+		\texttt{~~~~~~~~~~~~\call{\_xend}();} \\
+		\texttt{~~~~~~~~\} \flow{else} \{} \\
+		\texttt{~~~~~~~~~~~~\call{mutex\_lock}(\&m);} \\
+		\texttt{~~~~~~~~~~~~x++;} \\
+		\texttt{~~~~~~~~~~~~\call{mutex\_unlock}(\&m);} \\
+		\texttt{~~~~~~~~\}} \\
+		\texttt{~~~~\}} \\
+		\texttt{\}} \\
+		\end{tabular}
+	\end{center}
 	\caption{The example {\tt count} routine from Figure~\ref{fig:mutex}, rewritten to use HTM.
@@
 	\caption{Visualization of interleaving state space for the program in Figure~\ref{fig:concurrency-bug}.
-	Thread 1 is represented by blue ovals, thread 2 by orange squares, and time flows from top to bottom.
+	Thread 1 is represented by purple ovals, thread 2 by yellow squares, and time flows from top to bottom.
 	As the two threads execute the same code, without loss of generality thread 1 is fixed to run first --
@@
 
-Dynamic Partial Order Reduction \cite{dpor} (henceforth, DPOR) is the premier algorithm for mitigating the exponential explosion that arises as program size increases.
+Dynamic Partial Order Reduction \cite{dpor} (henceforth, DPOR) is the most popular algorithm for mitigating the exponential explosion that arises as program size increases.
 

### 2018-08-09 thesis/tm.tex
finish rerunning easy tests from tm verif (still not crying)

@@
 			{\tt htm2}
-			& 2,1 & \cpu{19.82}	& \ints{22}	& \cpu{19.01}	& \ints{15}	& \cpu{3.45}	& \ints{4} \\
-			& 2,2 &\cpu{73.02}	& \ints{1970}	& \cpu{26.96}	& \ints{356}	& \cpu{27.38}	& \ints{288} \\
-			& 2,3 &\cpu{3923.32}	& \ints{104914}	& \cpu{230.71}	& \ints{7528}	& \cpu{910.80}	& \ints{22409} \\
-			& 2,4 & \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{20h 50m}	& \ETAdag{1814844} \\
-			% [JOB 4] progress: 805070/1814844 brs (44.360267%), ETA 10h 49m 33s (elapsed 9h 59m 54s)
-			& 3,1 & \cpu{40.74}	& \ints{774}	& \cpu{24.27}	& \ints{224}	& \cpu{23.26}	& \ints{140} \\
-			& 3,2 & \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{8d 8h}& \ETAdag{2562278} \\
-			% [JOB 4] progress: 713090/2562278 brs (27.830313%), ETA 7d 22h 27m 42s (elapsed 9h 59m 54s)
-			& 4,1 & \cpu{8202.71}	& \ints{212146}	& \cpu{333.26}	& \ints{11973}	& \cpu{1794.06}	& \ints{44995} \\
+			& 2,1 & \cpu{18.26}	& \ints{22}	& \cpu{17.99}	& \ints{15}	& \cpu{3.25}	& \ints{4} \\
+			& 2,2 & \cpu{63.61}	& \ints{1446}	& \cpu{26.40}	& \ints{334}	& \cpu{28.04}	& \ints{286} \\
+			& 2,3 & \cpu{3429.04}	& \ints{86536}	& \cpu{196.18}	& \ints{6366}	& \cpu{997.91}	& \ints{24740} \\
+			& 2,4 & \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq} \\
+			& 3,1 & \cpu{43.24}	& \ints{774}	& \cpu{24.02}	& \ints{224}	& \cpu{22.79}	& \ints{140} \\
+			& 3,2 & \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq} \\
+			& 4,1 & \cpu{9225.52}	& \ints{212146}	& \cpu{376.70}	& \ints{11973}	& \cpu{2158.93}	& \ints{44995} \\
+			%% i don't even remember which test case this commented stuff was for anymore
 			% measured with !CONSIDER_ONLY_MOST_RECENT_DPOR_PREFERRED_TID
@@
 			{\tt counter}
-			& 2,1 & \cpu{6.90}	& \ints{10}	& \cpu{7.04}	& \ints{8}	& \cpu{3.35}	& \ints{4} \\
-			& 2,2 & \cpu{12.81}	& \ints{190}	& \cpu{10.58}	& \ints{102}	& \cpu{8.19}	& \ints{48} \\
-			& 2,3 & \cpu{144.83}	& \ints{3970}	& \cpu{65.66}	& \ints{1558}	& \cpu{38.18}	& \ints{904} \\
-			& 2,4 & \cpu{3433.08}	& \ints{86950}	& \cpu{1134.62}	& \ints{25398}	& \cpu{752.51}	& \ints{19128} \\
-			& 3,1 & \cpu{10.57}	& \ints{120}	& \cpu{9.23}	& \ints{64}	& \cpu{7.88}	& \ints{40} \\
-			% stm ok, rts ok, baseline alone needs redo
-			& 3,2 & \cpu{2492.17}	& \ints{60606}	& \cpu{2363.41}	& \ints{44862}	& \cpu{639.62}	& \ints{14304} \\
-			% TODO: everything above this line needs rerun
-			% TODO: everything below is fine up to dpor-preferred & thr_bottom1 updates
-			\multicolumn{8}{c}{everything above this line needs rerun for dpor-preferred tid + thr\_bottom1} \\
+			& 2,1 & \cpu{6.92}	& \ints{10}	& \cpu{6.81}	& \ints{8}	& \cpu{3.28}	& \ints{4} \\
+			& 2,2 & \cpu{13.27}	& \ints{190}	& \cpu{10.59}	& \ints{102}	& \cpu{8.73}	& \ints{48} \\
+			& 2,3 & \cpu{155.09}	& \ints{3970}	& \cpu{67.82}	& \ints{1558}	& \cpu{40.08}	& \ints{904} \\
+			& 2,4 & \cpu{3664.66}	& \ints{86950}	& \cpu{1150.34}	& \ints{25398}	& \cpu{805.85}	& \ints{19128} \\
+			& 3,1 & \cpu{11.26}	& \ints{120}	& \cpu{9.30}	& \ints{64}	& \cpu{8.26}	& \ints{40} \\
+			& 3,2 & \cpu{2572.13}	& \ints{60606}	& \cpu{2363.41}	& \ints{44862}	& \cpu{639.62}	& \ints{14304} \\
 			& 4,1 & \cpu{129.25}	& \ints{3006}	& \cpu{64.91}	& \ints{1296}	& \cpu{40.78}	& \ints{848} \\
@@
 		State space estimates measured after a timeout of 10 hours (and include those 10 hours in the predicted total).
-		{\em Thesis draft note: Some of these tests still need to be rerun on account of landslide implementation updates
-		to ensure consistent cpu times with the next table,
-		but I expect the state space sizes not to change substantially.}
 		}

### 2018-08-10 thesis/tm.tex
more tm tests

@@
 		\hline
+			\multicolumn{9}{c}{everything above this line needs rerun for dpor-preferred tid + thr\_bottom1} \\
 		{\tt avl\_insert}
-			& 2,2	&\cpu{1355.05}&\wtm{174.09}&\ints{*29}	&\cpu{*93.28}&\wtm{*30.37}&\ints{79}&\ETA{157286} \\
+			& 2,2	&\cpu{2494.77}&\wtm{315.46}&\ints{*29}	&\cpu{*95.53}&\wtm{*30.56}&\ints{79}&\ETA{158505} \\
+			% 79 interleavings; 1 preemptions; job time 9s; pldi time 30558236; new-fixed pldi cputime 95530430
+			% 29 interleavings; 1 preemptions; job time 14s; pldi time 315461192; new-fixed pldi cputime 2494772471
 		(segfault+)
-			& 2,3	&\cpu{402.91}&\wtm{*54.60}&\ints{*33}&\cpu{*228.16}&\wtm{125.89}&\ints{835}&\ETA{13664151} \\
-			% TODO: these 2,3 ones alone are new + ok
-			% 33 interleavings; 1 preemptions; job time 12s; pldi time 179368977; new-fixed pldi cputime 1403360474
-			% 835 interleavings; 1 preemptions; job time 1m 42s; pldi time 135292827; new-fixed pldi cputime 235664920
-			& 2,4	&\ETA{--}&\ETA{>1h}&\ETA{--}&\ETA{--}&\ETA{>1h}&\ETA{--}& \ETA{79020639} \\
-			% this quicksand jerk required ulimit -n higher than 1024
+			& 2,3	&\cpu{308.10}&\wtm{*43.95}&\ints{*33}&\cpu{*249.42}&\wtm{144.33}&\ints{835}&\ETA{13664203} \\
+			% 835 interleavings; 1 preemptions; job time 1m 50s; pldi time 144332259; new-fixed pldi cputime 249418666
+			% 33 interleavings; 1 preemptions; job time 10s; pldi time 43954391; new-fixed pldi cputime 308100075
+			& 2,4	&\cpu{*2979.29}&\wtm{*390.34}&\ints{*1457}&\ETA{--}&\ETA{>1h}&\ETA{--}& \ETA{61882736} \\
+			% 1457 interleavings; 4 preemptions; job time 3m 11s; pldi time 390342652; new-fixed pldi cputime 2979291067
+			% maximal: progress: 22854/61882736 brs (0.036931%), ETA 9y 307d 18h 4m 41s (elapsed 51m 54s)
+			% update for below, post thr-bottom1 fix - ok it FABed after all;
+			% not sure if higher ulimit still necessary but i still used it to be safe
+			% (old comment:) this quicksand jerk required ulimit -n higher than 1024
 			%
 			% the jerk didn't even find the bug anymore, after dpor-preferred tid
-			% heres the stats from how it did before
+			% heres the stats from how it did before (older comment:)
 			% & 2,4	&\cpu{*7598.19}&\wtm{*970.67}&\ints{*2633}&\ETA{--}&\ETA{>1h}&\ETA{--}& \ETA{79020639} \\
-			% anyway this was the maximal
-			%
 		\cline{2-9}
-			& 3,1	&\cpu{*86.43}&\wtm{*14.10}&\ints{*23}	&\cpu{91.97}&\wtm{24.80}&\ints{53}&\ETA{343250} \\
-			& 3,2	&\cpu{*2177.86}&\wtm{*287.66}&\ints{*169}&\ETA{--}&\ETA{>1h}&\ETA{--}& \ETA{425939133} \\
-			% TODO
-			\multicolumn{9}{c}{everything above this line needs rerun for dpor-preferred tid + thr\_bottom1} \\
-
+			& 3,1	&\cpu{*87.10}&\wtm{*14.81}&\ints{*14}	&\cpu{94.08}&\wtm{23.60}&\ints{24}&\ETA{207575} \\
+			% 14 interleavings tested; 3 preemptions; job time 8s; pldi time 14807247; new-fixed pldi cputime 87102600
+			% 24 interleavings tested; 3 preemptions; job time 5s; pldi time 23600679; new-fixed pldi cputime 94082790
+			& 3,2	&\cpu{*3672.84}&\wtm{*475.03}&\ints{*145}&\ETA{--}&\ETA{>1h}&\ETA{--}& \ETA{1635075071} \\
+			% maximal: progress: 6571/1635075071 brs (0.000402%), ETA 489418866284y 68d 4h 41m 44s (elapsed 29m 54s)
+			% nb. this quicksand one is actually the slower of 2 FABs, but so close in time to each other
+			% that the one with substantially fewer interleavings. they were:
+			% 145 interleavings; 1 preemptions; job time 23s; pldi time 475030016; new-fixed pldi cputime 3672836411
+			% 225 interleavings; 2 preemptions; job time 32s; pldi time 465874053; new-fixed pldi cputime 3601564200
+			% not like maximal found the bug at all so it's not like there's a comparison to make fair/unfair anyway
 		\hline
@@
 	Figure~\ref{fig:avlbug} shows the root cause, essentially the {\tt htm1} bug in disguise.
-	% landslide-trace-1521093835.18927.html
 	This manifested
 	alternately
+	% landslide-trace-1533850116.527842.html
+	% landslide-trace-1533850236.873032.html
 	as a segfault (for test parameters (2,2) and (3,1))
 	and
-	% landslide-trace-1533274266.417257.html
+	% landslide-trace-1533854371.201589.html
 	as a consistency-check assertion failure (for test parameters (2,3)).
@@
 			& 2,1 & \cpu{739.60} 	& \ints{20459} 	& \cpu{332.32}	& \ints{9675}	& \cpu{122.40}	& \ints{2774} \\
-			& 2,2 & \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq} \\
+			& 2,2 & \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	&\ETAdag{36393y}&\ETAdag{35094477}\\
+			% progress: 302022/35094477 brs (0.860597%), ETA 36392y 304d 6h 53m 45s (elapsed 9h 58m 49s)
 			& 3,1 & \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq} \\
@@
 			& 2,1 & \cpu{28.40} 	& \ints{150} 	& \cpu{94}	& \ints{26.55}	& \cpu{14.88}	& \ints{9} \\
-			& 2,2 & \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \cpu{1285.06}	& \ints{21684} \\
-			& 3,1 & \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq} \\
+			& 2,2 & \ETAdag{qqq}	& \ETAdag{qqq}	&\cpu{16455.24}	& \ints{283756}	& \cpu{1285.06}	& \ints{21684} \\
+			& 3,1 & \ETAdag{qqq}	& \ETAdag{qqq}	&\cpu{21153.64}	& \ints{366030}	& \cpu{705.04}	& \ints{12707} \\
 
@@
 Despite its conservative implementation,
-it provides roughly 2-6x reduction in most tests, with up to 20x in extreme cases.
+it provides roughly 2-6x reduction in most tests, with up to 17x in extreme cases.
 The biggest win is apparent in {\tt fig63},

### 2018-08-11 thesis/landslide.tex
finish DPOR friendly example and rewrite sleep sets description to follow

@@
 and $T(t_i)$ will denote the thread which was scheduled to produce that transition.
+A visual example will be given at the end to help reinforce the intuition behind the formalism.
 
@@
 
+% trying to approximate the pastel background colors in something dark enough to render text in
+\newcommand\dporTA[1]{\hilight{darklavender}{#1}\xspace}
+\newcommand\dporTB[1]{\hilight{goldish}     {#1}\xspace}
+\newcommand\dporTAcode[1]{\dporTA{\ensuremath{\mathbf{T_1}}: {\tt #1}}\xspace}
+\newcommand\dporTBcode[1]{\dporTB{\ensuremath{\mathbf{T_2}}: {\tt #1}}\xspace}
+
 In this interleaving, DPOR identifies 3 memory conflicts, two read/write and one write/write,
@@
 Note also that in $\bigstar$2,
-the reordered {\tt tmp2 <- x} is not directly part of the memory conflict which marked it,
-but it must be executed first to reach the conflicting {\tt x <- tmp2}.
+the reordered \dporTB{{\tt tmp2 <- x}} is not directly part of the memory conflict which marked it,
+but it must be executed first to reach the conflicting \dporTB{{\tt x <- tmp2}}.
 
@@
 
-However, as the legend spoils, this subset of the state space still contains redundant interleavings,
-arising from the fact that one pair of those 4 events is two reads, and hence not actually in conflict.
-Visual inspection shows that $\bigstar$5, while locally justified in trying to reorder {\tt tmp1 <- x} before {\tt x <- tmp2},
-effectively serves % oops
-only to reorder it with {\tt tmp2 <- x}
-relative to the first symmetric subtree.
-Landslide identifies this equivalence by checking,
-%in this case during the iteration of DPOR after interleaving K,
-in this case after K,
-that if the first event to be reordered (here, {\tt tmp <- x})
-has already been tested in an equivalent reordering around any number of preceding events (here, {\tt tmp2 <- x} and {\tt tmp2++}),
-then the newly marked subtree is safe to prune.
-The next section describes the implementation of this check in general.
-
 \subsubsection{Sleep set reduction}
@@
 
-% TODO: condense this by removing the 'new' example and just continuing with the previous section's
 In the presence of non-conflicting transitions as well as conflicting ones,
 DPOR's approach as described so far can still end up testing equivalent interleavings.
-For example, consider the execution sequence
-in Figure~\ref{fig:sleepsets}(a).
-After executing this interleaving, DPOR will detect the conflict between $t_a$ and $t_c$,
-and tag $I_{a2}$ (i.e., $T_2$ to run before $t_a$), causing us to next execute
-Figure~\ref{fig:sleepsets}(b).
-However, DPOR does not ``remember'' why it wanted to reorder $T_2$ with $T_1$
-(i.e., whether $t_b$ or $t_c$ was the important one to run before $t_a$),
-so from this branch,
-it next tags $I_{c'1}$, causing
-Figure~\ref{fig:sleepsets}(c).
-to be tested next, which is regrettably equivalent to the original interleaving
-($t_a$ and $t_b$ having been reordered all told, which are independent).
-To recover our forgotten original purpose,
-which was already fulfilled by testing the second interleaving,
-DPOR can check just before tagging a new interleaving (here, $I_{c'1}$),
-among all preceding transitions equivalent to the conflicting one (here, $t_{b'}$ equivalent to $t_{a'}$),
-for an already-explored interleaving beginning with the target thread.
-If such exists,
-%then because of the depth-first search ordering,
-the new interleaving is guaranteed to be equivalent to one already checked,
-and can be skipped.
+As the presence of more equivalence arrows in Figure~\ref{fig:dpor-example-2} hints,
+its reduced subset state space still contains redundancy,
+arising from the fact that one pair of those 4 events is two reads, and hence not actually in conflict.
+Visual inspection shows that $\bigstar$4, while locally justified
+in trying to reorder \dporTA{{\tt tmp1 <- x}} before \dporTB{{\tt x <- tmp2}},
+effectively serves % oops
+only to reorder it with \dporTB{{\tt tmp2 <- x}}
+relative to the first symmetric subtree ($\bigstar$1).
+In other words,
+even though DPOR marked each new branch with the intent only to reorder conflicting accesses,
+$\bigstar$1 and $\bigstar$4 contained interleavings equivalent up to independent reorderings anyway.
+Figure~\ref{fig:sleepsets} summarizes the relevant interleavings to highlight one such equivalence.
 
@@
 	\begin{center}
-		\begin{tabular}{p{0.3\textwidth}p{0.25\textwidth}p{0.35\textwidth}}
+		\begin{tabular}{p{0.30\textwidth}p{0.30\textwidth}p{0.30\textwidth}}
+	\begin{center}
 	\begin{tabular}{l}
-	$T_1$@$t_a$: \texttt{write~x} \\
-	$T_2$@$t_b$: \texttt{read~y} \\
-	$T_2$@$t_c$: \texttt{read~x} \\
-	\end{tabular}
-	&
-	\begin{tabular}{l}
-	$T_2$@$t_{b'}$: \texttt{read~y} \\
-	$T_2$@$t_{c'}$: \texttt{read~x} \\
-	$T_1$@$t_{a'}$: \texttt{write~x} \\
-	\end{tabular}
-	&
-	\begin{tabular}{l}
-	$T_2$@$t_{b'}$: \texttt{read~y} \\
-	$T_1$@$t_{a''}$: \texttt{write~x} \\
-	$T_2$@$t_{c''}$: \texttt{read~x} \\
-	\end{tabular}
-	\\
-	\\
-	(a) Original interleaving.
-	&
-	(b) Goal interleaving.
-	&
-	(c) Redundant interleaving.
+		\dporTAcode{tmp1 <- x} \\
+		\dporTBcode{tmp2 <- x} \\
+		\dporTAcode{x <- tmp1} \\
+		\dporTBcode{x <- tmp2} \\
+		% old example, pre dpor-example figure
+		%$T_1$@$t_a$: \texttt{write~x} \\
+		%$T_2$@$t_b$: \texttt{read~y} \\
+		%$T_2$@$t_c$: \texttt{read~x} \\
 	\end{tabular}
 	\end{center}
-	\caption{Motivating example for the sleep sets optimization.
-	For clarity, threads are labeled here with numbers ($T_n$) and the transitions thereby with letters ($t_i$).}
+	&
+	\begin{center}
+	\begin{tabular}{l}
+		\dporTBcode{tmp2 <- x} \\
+		\dporTBcode{x <- tmp2} \\
+		\dporTAcode{tmp1 <- x} \\
+		\dporTAcode{x <- tmp1} \\
+		%$T_2$@$t_{b'}$: \texttt{read~y} \\
+		%$T_2$@$t_{c'}$: \texttt{read~x} \\
+		%$T_1$@$t_{a'}$: \texttt{write~x} \\
+	\end{tabular}
+	\end{center}
+	&
+	\begin{center}
+	\begin{tabular}{l}
+		\dporTBcode{tmp2 <- x} \\
+		\dporTAcode{tmp1 <- x} \\
+		\dporTAcode{x <- tmp1} \\
+		\dporTBcode{x <- tmp2} \\
+		%$T_2$@$t_{b'}$: \texttt{read~y} \\
+		%$T_1$@$t_{a''}$: \texttt{write~x} \\
+		%$T_2$@$t_{c''}$: \texttt{read~x} \\
+	\end{tabular}
+	\end{center}
+	\\
+	\begin{center}
+	(a) Original branch (C).
+	\end{center}
+	&
+	\begin{center}
+	(b) Goal branch (K).
+	\end{center}
+	&
+	\begin{center}
+	(c) Redundant branch (L).
+	\end{center}
+	\end{tabular}
+	\end{center}
+	\caption{Motivating example for the sleep sets optimization,
+	highlighting three interleavings from Figure~\ref{fig:dpor-example-2}.
+	The always-independent {\tt tmpN++}s are omitted for brevity.
+	}
 	\label{fig:sleepsets}
@@
 
-Landslide implements this check in {\tt equiv\_already\_explored()}.
+Intuitively speaking,
+when DPOR entered the $\bigstar$3 subtree,
+it did not ``remember'' which memory conflict it wanted to reorder \dporTA{$\mathbf{T_1}$} around
+(i.e., that \dporTB{{\tt x <- tmp2}} should come before \dporTA{{\tt tmp1 <- x}}).
+Upon witnessing the conflict in the new (intended) order,
+it then tried to reorder it again,
+producing interleavings regrettably equivalent to ones already tested
+(all told, only \dporTB{{\tt tmp2 <- x}} and \dporTB{{\tt tmp2++}} having been reordered around \dporTA{{\tt tmp1 <- x}}).
+%
+To ``remember'' the original purpose of testing subtree $\bigstar$3,
+which was already fulfilled by testing K,
+DPOR can check just before tagging a new subtree (here, $\bigstar$4)
+among all preceding transitions independent with the conflicting one
+(here, \dporTB{{\tt tmp2 <- x}} and \dporTB{{\tt tmp2++}} independent with \dporTA{{\tt tmp1 <- x}})
+for an already-explored interleaving beginning with the target thread.
+%(here, the original sibling of $\bigstar 3$).
+If such exists,
+the new subtree is guaranteed to be equivalent to one already checked,
+and can safely be skipped.
+
+Landslide implements this check in {\tt equiv\_already\_explored()},
+which checks (in this case after executing K),
+that if the first event to be reordered (here, \dporTA{{\tt tmp1 <- x}})
+has already been tested in an equivalent reordering around any number of preceding events
+(here, \dporTB{{\tt tmp2 <- x}} and \dporTB{{\tt tmp2++}}),
+then the newly marked subtree is safe to prune.
 This corresponds to the {\em sleep sets} optimization described by prior work \cite{partial-order-methods,dpor,optimal-dpor},
-so named because it effectively puts $T_1$ ``to sleep'' until after the true conflicting access in $t_c$.
-Landslide's implementation differs from prior work by not explicitly tracking
-sets of reordered threads and memory accesses they should expect to conflict with,
-and instead identifies where the reduction should take place in subsequent iterations of DPOR.
+so named because it effectively puts \dporTA{$\mathbf{T_1}$}
+``to sleep'' until after the true conflicting access of \dporTB{{\tt x <- tmp2}}.
+
+Landslide's implementation differs from prior work,
+which explicitly tracks sets of reordered threads and expected conflicting accesses,
+by instead identifying
+where the reduction should occur
+during subsequent DPOR iterations.
 % KEEP_RUNNING_DPORS_CHOSEN_TID
@@
 to prefer scheduling the thread previously chosen for reordering by DPOR,
-to force whatever its conflicting memory access was to happen before the other, preempted thread gets a chance to run.
+to ensure the conflicting access happens before the preempted thread gets a chance to run again.
+%to force whatever its conflicting memory access was to happen before the other, preempted thread gets a chance to run.
 Further optimizations such as {\em source sets} and {\em wakeup trees},
-which achieve optimality (i.e., executing exactly one interleaving per equivalence class) \cite{optimal-dpor}
-are not yet implemented,
-although to the best of my knowledge,
+which prior work has shown achieve optimality
+(i.e., executing exactly one interleaving per equivalence class) \cite{optimal-dpor}
+are not yet implemented.
+To the best of my knowledge,
 they provide further reduction only in cases of 3 or more threads;
@@
 
-% TODO: make sure it's clear that the quicksand chapter expts were run before this optmz was added,
-% and that the HTM expts fro msigbovik paper were re-run for its chapater herein
+Chapter~\ref{chap:quicksand}'s experiments and Chapter~\ref{chap:education}'s user studies
+were conducted before this optimization was implemented.
+Note that its absence has no bearing on DPOR's soundness, only its efficiency,
+and that Landslide showed good bug-finding performance even without it.
+Chapter~\ref{chap:tm}'s experiments include this optimization,
+because its presence was required to fairly compare the other reduction strategies presented therein.
 

### 2018-08-11 thesis/landslide.tex
describe estimation algs and make an example with RE (still tryna remember how WBE works tbh)

@@
 
-\subsection{Execution Tree}
+\subsection{Execution tree}
 \label{sec:landslide-save}
@@
 
-\subsubsection{Time Travel}
+\subsubsection{Time travel}
 \label{sec:landslide-timetravel}
@@
 This singular question determines to what extent the state space of possible interleavings explodes exponentially.
-While other parts of the research decide which lock API calls to consider,
+While other parts of
+%the research
+the great work % :relaxed:
+decide which lock API calls to consider,
 or which memory accesses constitute a data race,
@@
 whose original state space is shown in Figure~\ref{fig:tree}.
+In this program both threads are always enabled, imposing no scheduling constraints,
+so memory conflicts alone will drive exploration.
 %Acknowledging that quickly gaining an intuitive understanding for an algorithm
@@
 	\begin{tabular}{l}
-		\dporTAcode{tmp1 <- x} \\
-		\dporTBcode{tmp2 <- x} \\
-		\dporTAcode{x <- tmp1} \\
-		\dporTBcode{x <- tmp2} \\
+		\dporTAcode{tmp1 <- x (read)} \\
+		\dporTBcode{tmp2 <- x (read)} \\
+		\dporTAcode{x <- tmp1 (write)} \\
+		\dporTBcode{x <- tmp2 (write)} \\
 		% old example, pre dpor-example figure
@@
 	\begin{tabular}{l}
-		\dporTBcode{tmp2 <- x} \\
-		\dporTBcode{x <- tmp2} \\
-		\dporTAcode{tmp1 <- x} \\
-		\dporTAcode{x <- tmp1} \\
+		\dporTBcode{tmp2 <- x (read)} \\
+		\dporTBcode{x <- tmp2 (write)} \\
+		\dporTAcode{tmp1 <- x (read)} \\
+		\dporTAcode{x <- tmp1 (write)} \\
 		%$T_2$@$t_{b'}$: \texttt{read~y} \\
@@
 	\begin{tabular}{l}
-		\dporTBcode{tmp2 <- x} \\
-		\dporTAcode{tmp1 <- x} \\
-		\dporTAcode{x <- tmp1} \\
-		\dporTBcode{x <- tmp2} \\
+		\dporTBcode{tmp2 <- x (read)} \\
+		\dporTAcode{tmp1 <- x (read)} \\
+		\dporTAcode{x <- tmp1 (write)} \\
+		\dporTBcode{x <- tmp2 (write)} \\
 		%$T_2$@$t_{b'}$: \texttt{read~y} \\
@@
 	\end{center}
-	\caption{Motivating example for the sleep sets optimization,
-	highlighting three interleavings from Figure~\ref{fig:dpor-example-2}.
-	The always-independent {\tt tmpN++}s are omitted for brevity.
+	\caption{Motivating example for the sleep sets optimization.
+	Three of Figure~\ref{fig:dpor-example-2}'s interleavings are highlighted,
+	with the always-independent {\tt tmpN++}s omitted for brevity.
 	}
@@
 then the newly marked subtree is safe to prune.
+Note that this does not require storing any full subtrees outside of the current branch;
+only the subtree's root node need be saved to
+prove that an equivalent interleaving beginning with \dporTA{$\mathbf{T_1}$} therein was already checked,
+preserving DPOR's $O(n)$ memory footprint.
+
 This corresponds to the {\em sleep sets} optimization described by prior work \cite{partial-order-methods,dpor,optimal-dpor},
@@
 ``to sleep'' until after the true conflicting access of \dporTB{{\tt x <- tmp2}}.
-
 Landslide's implementation differs from prior work,
@@
 
-% TODO: once dpor example tree is made
-% switch this to that
-As a concrete example, consider the state space of Figure~\ref{fig:tree}(b) (untouched by DPOR),
+As a concrete example, consider the state space of Figure~\ref{fig:dpor-example-2},
 and suppose each transition to the next preemption point takes 1 second to execute.
-While the first branch executes in 5 seconds,
+While the first branch executes in 6 seconds,
 the second branch, sharing the first transition as a common prefix, takes 4 seconds,
 and the one after that only 2;
-the state space being ultimately completed in 33 seconds.
+the state space being ultimately completed in 24 seconds.
 Even with perfect hindsight,
-na\"ively multiplying the total interleavings (10)
-by the total execution time per branch (5) would double-count common prefixes
-and grossly overestimate (50) the total runtime.
+na\"ively multiplying the total interleavings (6)
+by the total execution time per branch (6) would double-count common prefixes
+and grossly overestimate (36) the total runtime.
 
@@
 which can occasionally result in drastically differing outputs (\sect{\ref{sec:tm-verif}}).
-I summarize their respective implementations here.
+
+Implementation-wise, Landslide reports size estimates as both the percentage and as a total number of branches,
+and time estimates as an ETA.
+Quicksand's {\tt -v} option (\sect{\ref{sec:landslide-quicksand-options}})
+will cause it to print them each time a new interleaving is tested; for example:
+\begin{center}
+	{\tt \small [JOB 1] progress: 66101/94825 brs (69.708252\%), ETA 13m 37s (elapsed 46m 10s)}
+\end{center}
+Both estimates are computed simultaneously in {\tt \_estimate()} in {\tt estimate.c}
+(which, I might add, is well-commented in case the following prose is insufficient).
+%I summarize their respective implementations here.
 % TODO more specific section reference
@@
 
+The WBE, used to estimate total number of interleavings,
+computes the {\em proportion} of the total size
+that the already-explored branches are expected to comprise,
+using DPOR's workqueue to anticipate how many unexplored marked branches remain.
+This serves as a ``progress bar'' that represents the estimated overall progress,
+approaching 1 (not necessarily monotonically) as exploration continues.
+
+Summarizing prior work's formal definition \cite{estimation},
+the proportion at a terminal node $v_n$%
+\footnote{Prior work \cite{estimating-search-tree-size,estimation} refers to this instead as {\em probability},
+i.e., the probability that the node will appear in a branch chosen uniformly at random from the completed tree.
+I find ``proportion'' to be more illuminating on how the algorithm works.
+},
+preceded by an execution sequence $(v_1 \dots v_{n-1})$,
+is computed as:%
+\footnote{
+Simplified from \cite{estimation}: the missing $F(v_i)$ is 0, using the empty fit strategy.
+}
+
+\[
+	\mathsf{proportion}(v_1 \dots v_n) = \displaystyle\prod_{i=1}^{n-1} \frac{1}{|\mathsf{marked~children}(v_i)|}
+\]
+
+where $\mathsf{marked~children}(v_i)$ is the number of enabled thread transitions at $v_i$
+which have either already been explored or been marked by DPOR.
+Then, the total estimate is given as the sum over all branches $b = (v_1 \dots v_n)$ explored so far:%
+\footnote{
+Simplified from \cite{estimation}: $t(b)$, the time for each branch, is 1, because we are counting them.
+}
+
+\[
+	\mathsf{estimate} = \frac{1}{\sum_{b \in B} \mathsf{proportion}(b)}
+\]
+
+It is easy to see how these formulas might fit into DPOR's incremental search procedure:
+at the end of each branch compute its $\mathsf{proportion}$ and add it in to a global $\mathsf{estimate}$ value.
+However, DPOR may tag new branches to explore that would affect past branches' proportions,
+in which case they would have to be recomputed,
+which is not feasible without storing the entire exponentially-sized tree in memory.
+Instead, Landslide also stores per-subtree estimates at intermediate $v_i$ nodes, $1 < i < n$,
+along the current branch.
+Whenever DPOR marks a new $k$th branch for exploration
+at some $v_i$
+%with $\mathsf{marked~children}(v_i) = k$,
+its estimate
+is multiplied by $(k-1)/k$ to retroactively adjust all past branches' proportions contributing to that estimate.
+The change is also propagated to its sub-subtrees,
+whose estimates must also incorporate the new $\mathsf{marked~children}$ value.
+This allows Landslide to update the global estimate after each branch in $O(n)$ time and memory,
+without recomputing past branch proportions individually.
+
 \subsubsection{Runtime (Recursive) estimation}
+
+The RE, used to estimate total execution time,
+computes at each node the expected time to execute all subtrees rooted at children of that node,
+assuming unvisited subtrees' times will be an average of their visited siblings.
+This estimate at the root node, minus the current time elapsed so far,
+serves as a guess at how long until completion.
+Let $\mathsf{usecs}(v_i)$ denote the time elapsed during execution of the transition $v_{i-1} \rightarrow v_i$.
+Then a node's estimate is given by:
+
+\[
+	\mathsf{estimate}(v_i) = \mathsf{usecs}(v_i) +
+	\frac{|\mathsf{marked~children}(v_i)|}{|\mathsf{explored~children}(v_i)|}
+	\sum_{v_j \in \mathsf{explored~children}(v_i)} \mathsf{estimate}(v_j)
+\]
+
+Like the WBE, whenever DPOR tags a new $k$th child at some $v_i$,
+its estimate is multiplied by $k/(k-1)$ (note the reciprocal of before)
+to retroactively re-weight previously explored subtrees' estimates.
+Unlike the WBE, this change does not need to be propagated to descendant subtrees' estimates.
+This estimate also takes $O(n)$ time and memory.
+
+\subsubsection{Accuracy example}
+
+To illustrate how the two estimators can under-estimate the total tree size and/or diverge from each other,
+consider the state space from Figure~\ref{fig:dpor-example-2}, of size 6.
+Suppose for RE that each transition takes 1 second to execute.
+
+\begin{enumerate}
+	\item After branch A, two tags exist, $\bigstar$1 and $\bigstar$3.
+
+		Under RE, the estimate at \dporTA{{\tt tmp1++}} will be 9 seconds,
+		and the root estimate will be 20 seconds, an underestimate.
+	\item After branch B, $\bigstar$2 is now marked.
+
+		Under RE, the estimate at \dporTB{{\tt tmp2++}} is 5 seconds,
+		the estimate at \dporTA{{\tt tmp1++}} is updated to 11 seconds,
+		and the root estimate to 24 seconds, accurate.
+	\item After branch C, nothing new was marked.
+
+		Under RE, no estimates change from after B.
+	\item After branch K, $\bigstar$4 now exists.
+
+		Under RE, the estimate at \dporTB{{\tt tmp2++}} is 9 seconds,
+		and the root estimate is 22 seconds, an underestimate.
+	\item After branch L, $\bigstar$5 joins the party.
+
+		Under RE, the estimate at \dporTB{{\tt tmp2++}} is 11 seconds,
+		and the root estimate is 24 seconds, accurate.
+	\item After branch M, both estimators have perfect information and converge to accuracy.
+\end{enumerate}
+
+To illustrate how the estimators can over-estimate the total tree size,
+consider the same state space,
+except with the $\bigstar$4 subtree also pruned by DPOR's sleep sets extension (\sect{\ref{sec:landslide-sleepsets}});
+i.e., only branches A, B, C, and K remain,
+with an 18 second execution time.
+Both estimators' behaviour is identical through branch C,
+only now WBE's prediction is accurate at A but overestimates at B and C,
+while RE's predictions are all overestimates.
+As before, both reach perfect accuracy upon completion, now occurring at K.
+Intuitively speaking, the estimators underpredict when DPOR keeps finding new branches to tag as it makes progress,
+and overpredict when sleep set reduction achieves extra pruning on right subtrees.
+Not shown in this example, interleaving-dependent control flow can, of course,
+beget unexpected state space structure in essentially arbitrary other ways.
 

### 2018-08-11 thesis/tm.tex
tm: abort codes / retry sets <--> estimation; also some more 10hrs

@@
 (serving an analogous purpose as the {\tt all\_explored} flag for thread scheduling).
+The state space estimators (\sect{\ref{sec:landslide-estimate}})
+check the length of these lists, in addition to DPOR's tagged threads,
+when computing the number of marked children.
 
@@
 it records a {\em retry set},
-i.e., the pair of {\tt xbegin} results executed by the conflicting threads.
+i.e., the pair of {\tt xbegin} results executed by the conflicting threads,
+to accompany that branch.
+Like the {\tt xabort\_codes} lists, the state space estimators (\sect{\ref{sec:landslide-estimate}})
+check the number of retry sets when counting the number of marked children.
 % ah-- not compatible after all!
 %in addition to tracking the set of abort codes possible at each preemption point.
-Later, when traversing that new branch,
+Later, when traversing the new branch,
 Landslide remembers those threads' expected {\tt xbegin} results
@@
 		{\tt avl\_insert}
+			% note the somewhat chaotic behaviour between quicksand 2,2 and 2,3; the latter surprisingly FABs faster
+			% this is likely because non-drpp SSes were not big enough to defer for longer in 2,2,
+			% whereas in 2,3 it went straight to the buggy SS asap
 			& 2,2	&\cpu{2494.77}&\wtm{315.46}&\ints{*29}	&\cpu{*95.53}&\wtm{*30.56}&\ints{79}&\ETA{158505} \\
@@
 			& 3,2 & \ETAdag{qqq} & \ETAdag{qqq} \\
-			& 4,1 & \cpu{qqq} & \ints {qqq} \\ % was: 30408, 282084 pre mkrun
+			& 4,1 & \ETAdag{16h 26m} & \ETAdag{432628} \\ % was: 30408, 282084 pre mkrun
+			% [JOB 4] progress: 286910/432628 brs (66.317896%), ETA 6h 26m 25s (elapsed 9h 59m 48s)
+			% total CPU time consumed: 10h 4s (36004288982 usecs) (core saturation: 12%)
 			\hline
@@
 			% total CPU time consumed: 10h 19s (36019507965 usecs) (core saturation: 12%)
-			& 3,1 & \ETAdag{qqq} & \ETAdag{qqq} \\
+			& 3,1 & \ETAdag{173d 17h} & \ETAdag{12572187} \\
+			% [JOB 9] progress: 496281/12572187 brs (3.947451%), ETA 173d 6h 38m 55s (elapsed 9h 59m 43s)
+			% total CPU time consumed: 10h 20s (36020148608 usecs) (core saturation: 12%)
 			\\
@@
 had on even the correctness tests:
-{\tt lock}({\tt spin\_fixed})'s state spaces were reduced by more than half compared to {\tt lock}({\tt spinlock}),
+{\tt lock}({\tt spin\_fixed})'s state spaces were reduced by nearly half compared to {\tt lock}({\tt spinlock}),
 on account of DPOR no longer needing to reorder the (now) read-read access pairs.

### 2018-08-11 thesis/overview.tex
rearrange some overview mc prose

@@
 
+Model checking \cite{verisoft} is a testing technique for systematically exploring the possible thread interleavings of a concurrent program.
+A model checker executes the program repeatedly, each time according to a new thread interleaving, until all interleavings have been tested or the CPU budget is exhausted.
+During each execution, it forces threads to execute serially, thereby confining the program's nondeterminism to scheduler thread switches.
+It then controls the scheduling decisions
+%, choosing a different thread from one chosen before at a certain execution states
+to guarantee a unique interleaving is tested each iteration.
+
 \subsection{The state space}
 
-Model checking \cite{verisoft} is a testing technique for systematically exploring the possible thread interleavings of a concurrent program.
-A model checker executes the program repeatedly, each time according to a new thread interleaving, until the state space (or the CPU budget) is exhausted.
-During each execution, it forces threads to execute serially, thereby confining the program's nondeterminism to controlled thread switches.
-Using a single iteration of the {\tt x++;} loop from Figure~\ref{fig:concurrency-bug} as an example,
-Figure~\ref{fig:tree}(a) shows all possible execution interleavings of the compiled code between 2 threads.
+To understand what it means to exhaustively test all possible thread interleavings,
+one must define the possible execution sequences as a finite {\em state space}.
+To visualize this,
+using a single iteration of the {\tt x++;} loop from Figure~\ref{fig:concurrency-bug} as an example,
+with {\tt x++;} expanded into its three corresponding pseudo-assembly instructions,
+Figure~\ref{fig:tree}(a) shows all possible execution interleavings
+%of the compiled code
+between 2 threads.
 
@@
 \end{figure}
-
-Some model checkers explicitly store the set of visited program states as a means of identifying equivalent interleavings \cite{spin}.
-This approach is called {\em stateful} model checking.
-In this thesis, I focus on {\em stateless} model checking,
-which instead analyzes the sequence of execution events to avoid a prohibitive memory footprint.
-Henceforth I will abbreviate ``stateless model checking'' simply as ``model checking'' for brevity.
 
@@
 the set of possible ways to execute the program can be viewed as a tree.
-Figure~\ref{fig:tree}(b) shows a visualization of the corresponding tree from our example program.
+Figure~\ref{fig:tree}(b) shows a visualization of the corresponding tree from our example program,
+using each pseudo-assembly instruction as a preemption point.
 
@@
 Addressing the scaling problem in this exponential relation is the central research problem for all model checkers.
+
+Some model checkers explicitly store the set of visited program states as a means of identifying equivalent interleavings \cite{spin}.
+From the perspective of such tools, state spaces such as these wherein equivalent states may be reached by multiple paths
+are represented as a directed acyclic graph (DAG) instead of as a tree.
+This approach is called {\em stateful} model checking.
+In this thesis, I focus on {\em stateless} model checking (and execution trees, not DAGs),
+which instead analyzes the sequence of execution events to avoid a prohibitive memory footprint.
+Henceforth I will abbreviate ``stateless model checking'' simply as ``model checking'' for brevity.
+Also, the term ``state space'' was originally coined to refer to the stateful approach's emphasis on the DAG's nodes
+(i.e., program states);
+while stateless checkers emphasize the tree's branches (i.e., execution sequences) instead,
+I will continue to use ``state space'' for consistency with prior work.
 
@@
 % estimation (WBE, RE)
+% ETA
 % landslide

### 2018-08-11 thesis/landslide.tex
the busy bee moves randomly around the screen to show that the system is processing

@@
 using DPOR's workqueue to anticipate how many unexplored marked branches remain.
-This serves as a ``progress bar'' that represents the estimated overall progress,
+This serves as a ``progress bar'' that represents the estimated overall progress \cite{progress-bar},
 approaching 1 (not necessarily monotonically) as exploration continues.

### 2018-08-11 thesis/landslide.tex
finish estimation example, wbe really kinda sucks the more i think about it

--- 2018-08-11 thesis/landslide.tex
@@
 using DPOR's workqueue to anticipate how many unexplored marked branches remain.
-This serves as a ``progress bar'' that represents the estimated overall progress,
-approaching 1 (not necessarily monotonically) as exploration continues.
+This serves as a progress bar \cite{progress-bar} that represents the estimated percentage towards completion,
+approaching 100\% (not necessarily monotonically) as exploration continues.
 
@@
 However, DPOR may tag new branches to explore that would affect past branches' proportions,
-in which case they would have to be recomputed,
+requiring them to be recomputed,
 which is not feasible without storing the entire exponentially-sized tree in memory.
@@
 	\item After branch A, two tags exist, $\bigstar$1 and $\bigstar$3.
-
-		Under RE, the estimate at \dporTA{{\tt tmp1++}} will be 9 seconds,
+		Under WBE, the subtree estimate at \dporTA{{\tt tmp1++}} will first be 1/2
+		(half its children being fully explored),
+		and the root estimate will be 1/4,
+		half that,
+		which is propagated back down to \dporTA{{\tt tmp1++}}, becoming also 1/4.
+		Dividing the current progress (1) by that yields 4 total branches, an underestimate.
+
+		Under RE, the estimate at \dporTA{{\tt tmp1++}} will be 9 seconds
+		(incorrectly assuming $\bigstar$1's subtree will be 1 branch),
 		and the root estimate will be 20 seconds, an underestimate.
 	\item After branch B, $\bigstar$2 is now marked.
+		Under WBE, the subtree estimate at \dporTB{{\tt tmp2++}} is 1/2,
+		which at \dporTA{{\tt tmp1++}} is then divided by its $\mathsf{marked~children}$
+		and added to its estimate, yielding 3/4.
+		Note that it has ``forgotten'' that only branch A, alone, contributed to its original 1/2,
+		rather than two branches as in this subtree.
+		The root
+		the subtree estimate at \dporTA{{\tt tmp1++}} is updated (and propagated down) to 3/8.
+		Dividing the current progress (2) by that yields 5.33 branches, an underestimate.
 
@@
 	\item After branch C, nothing new was marked.
+		The subtree estimate at \dporTB{{\tt tmp2++}} is 1
+		(having been completely explored)
+		and the root estimate is 1/2.
+		Dividing the current progress (3) by that yields 6, accurate.
 
@@
 	\item After branch K, $\bigstar$4 now exists.
+		\dporTB{{\tt tmp2++}}'s subtree estimate is at first 1/2,
+		then the root estimate and it get updated to 3/4.
+		Dividing into the current progress (4), 5.33, an underestimate.
 
@@
 	\item After branch L, $\bigstar$5 joins the party.
+		\dporTB{{\tt tmp2++}}'s estimate is updated to 3/4,
+		and the root estimate ultimately becomes 7/8.
+		Dividing into the current progress (5), 5.7, an underestimate.
 
@@
 Both estimators' behaviour is identical through branch C,
-only now WBE's prediction is accurate at A but overestimates at B and C,
+only now WBE's prediction happens to be accurate at A (although for the wrong reasons),
+but overestimates at B and C,
 while RE's predictions are all overestimates.

### 2018-08-13 thesis/intro.tex
how to read the thesis

@@
 		and all the special features it's been equipped with over the years.
+		It is the foundation upon which all three contributions above build.
 	\item {\bf Quicksand:} Chapter~\ref{chap:quicksand} presents the Iterative Deepening framework which more intelligently chooses which state spaces to test, corresponding to contribution 1 above.
@@
 	\item {\bf Related Work:} Chapter~\ref{chap:relatedwork} honors my neighbours and ancestors in research spirit.
+	\item {\bf Future Work:} Chapter~\ref{chap:warpzone} % TODO
 	\item {\bf Conclusion:} Chapter~\ref{chap:conclusion} provides some thoughts on the future of the field.
 \end{itemize}
+
+\subsubsection{How to read this thesis}
+
+I have made an effort to make this document accessible to readers of all programming experience levels,
+although some of the research being theoretical and several layers of abstractions deep,
+I cannot promise all easy reading.
+The Background chapter~(\ref{chap:background}) provides what I hope are friendly concrete examples
+to help an reader feel comfortable with each level of intuition that upcoming algorithms will build upon.
+These should suffice for the experiments and overall contributions,
+if not necessarily the details of each algorithm or soundness proof.
+The Education chapter~(\ref{chap:education}) in particular
+may be approached with no knowledge of concurrency or model checking,
+taking it merely as a study of a magic new debugging tool in the classroom setting.
+The more ambitious reader may proceed to the Landslide chapter's further examples~(\sect{\ref{sec:landslide-algs}}),
+which should equip them to understand every detail herein.
+Readers who did not come to read this end-to-end should at least consult the glossary~(\sect{\ref{sec:glossary}})
+to help clarify any terminology confusion.
+
+Color will be used in figures and graphs to add visual contrast and make the data easier to navigate at a glance,
+but only in redundant ways also signaled by symbols.
+I have made some effort to choose palettes friendly to color-blindness,
+but should the reader find contrasts too low anyway,
+whether being color-blind or reading a physical copy printed in greyscale,
+they may be assured all important distinctions still render in monochrome;
+for example, ovals and rectangles to depict different threads,
+or $\dagger$ to distinguish state space estimates from completed verifications.
+
+Pronoun use will vary between specific singular and ambiguously-singular-or-plural to convey additional nuance.
+The singular ``I'' will present my own research contributions,
+while the royal ``we'' should be taken to include the reader,
+such as when surveying background material or related work,
+to which the author's and reader's relationships are more similar.
+The abstract ``the programmer'' (or ``the user'') will be referred to as ``she''
+to highlight her role as the intended audience for concurrency testing tools
+rather than the underlying research,
+although real students who participated in the user studies will be given the more inclusive ``they''.
+Gender-neutral pronouns are to be used on the author themself.
+
+% TODO:
+% Finally, this thesis also contains a small puzzle-hunt to reward dedicated readers who make it all the way through.
 

### 2018-08-13 thesis/overview.tex
how to read the thesis

@@
 This chapter will introduce the necessary background material on concurrency, stateless model checking, data-race analysis, and the relevant undergraduate operating systems classes.
-
-{\bf Special personal note for the curious reader:} I have taken special care to write this chapter to be approachable by any intermediate programmer familiar with basic C programming concepts.
-My committee members obviously do not require such treatment, but I hope the rare reader from outside this relatively narrow field shall also not find the barrier to entry here too high.
-Concurrency analysis and verification is often very abstract,
-requiring strong intuition for certain concepts to understand the next algorithm that builds upon them, and so on;
-and I'd hate for any reader to get left behind off the bat for not knowing what a mutex is.
-%Abstract as the field is, it's also quite difficult to make good visual aids, unlike for example in graphics research, but I have done my best.
-Should you find any of my explanations insufficiently illuminating, please do get in touch.
 

### 2018-08-14 thesis/landslide.tex
minor fixups landslide.tex

@@
 and hence unlike other chapters is written in the style of documentation rather than as a report of research results.
+The reader interested only in a theoretical introduction to model checking's foundational algorithms,
+with detailed and friendly examples to help establish intuitions the later chapters may require,
+may skip to \sect{\ref{sec:landslide-algs}}.
 
@@
 
-\section{Algorithms of note}
-
-% not worth discussing?
-%\subsection{Kernel annotations}
-%\subsection{Bug identification}
+\section{Algorithms}
+\label{sec:landslide-algs}
+
+This section describes Landslide's model-checking algorithms from a theoretical perspective.
+The more complex ones are accompanied by concrete examples to hopefully help the reader build a solid intuition,
+which upcoming chapters will require in their soundness proofs.
 
@@
 
-The set intersection is implemented in {\tt mem\_shm\_interseect()} in {\tt memory.c}.
+The set intersection is implemented in {\tt mem\_shm\_intersect()} in {\tt memory.c}.
 It checks for read/write and write/write pairs to the same address with an $O(\mathsf{max}(m,n))$ scan of both access sets (pre-sorted).
@@
 
-\subsubsection{Accuracy example}
+\subsubsection{Example}
 
@@
 		rather than two branches as in this subtree.
-		The root
-		the subtree estimate at \dporTA{{\tt tmp1++}} is updated (and propagated down) to 3/8.
+		The root and \dporTA{{\tt tmp1++}}'s subtree estimates are updated (and propagated down) to 3/8.
 		Dividing the current progress (2) by that yields 5.33 branches, an underestimate.

### 2018-08-14 thesis/education.tex
challenge unconscious bias

@@
 		and treats them as synchronization points as described in \sect{\ref{sec:landslide-blocking}}.
-	\item Landslide finds ``too suspicious'' spinwait-loops in the students' mutex implementations
+	\item Landslide finds ``too suspicious'' spinwait-loops in mutex implementations
 		which are neither yield- nor {\tt xchg}-loops (as described above),
@@
 		and reports them with a suggestive message ({\tt undesirable\_loop\_html()} in {\tt landslide.c})
-		referring them to the appropriate lecture material
+		referring the student to the appropriate lecture material
 		\cite{synchronization-2}.
@@
 % TODO: addressing bias
-% well, we did the best we could(?)
+% well, I did the best I could(?)
 % survey email: "please answer honestly rather than flatteringly" vs. studence trying to be polite
@@
 fixes/improvements to the setup scripts
-as we found student kernels that failed to automatically annotate
+as I found student kernels that failed to automatically annotate
 (for example, those with {\tt ready\_list} changed to an array,
 as described in \sect{\ref{sec:education-pintos-instrumentation}}).
-Had we given Landslide directly to students that semester,
+Had I given Landslide directly to students that semester,
 the students themselves would have had to email me for tech support.

### 2018-08-14 thesis/intro.tex
challenge unconscious bias

--- 2018-03-10 thesis/intro.tex
@@
 	\item {\bf Related Work:} Chapter~\ref{chap:relatedwork} honors my neighbours and ancestors in research spirit.
+	\item {\bf Future Work:} Chapter~\ref{chap:warpzone} % TODO
 	\item {\bf Conclusion:} Chapter~\ref{chap:conclusion} provides some thoughts on the future of the field.
 \end{itemize}
+
+\subsubsection{Notes on reading this thesis}
+
+I have tried to make this document accessible to readers of all programming experience levels,
+although some of the research being theoretical and several layers of abstractions deep,
+I cannot promise all easy reading.
+Chapter~\ref{chap:background}, Background, provides what I hope are friendly concrete examples
+to help the reader feel comfortable with each level of intuition that upcoming algorithms will build upon.
+These should suffice for the experiments and overall contributions,
+if not necessarily the details of each algorithm or soundness proof.
+In particular, Chapter~\ref{chap:education}, Education,
+may be approached with no knowledge of concurrency or model checking,
+taking it merely as a study of a magic new debugging tool in the classroom setting.
+The more ambitious reader may proceed to the Landslide chapter's algorithm walkthroughs~(\sect{\ref{sec:landslide-algs}}),
+which should equip them to understand every detail herein.
+Readers who are here only to skim and skip around should at least be aware of the glossary~(\sect{\ref{sec:glossary}})
+to help clarify any terminology confusion.
+
+Color will be used in figures and graphs to add visual contrast and make the data easier to navigate at a glance,
+but only in redundant ways also signaled by symbols.
+I have made some effort to choose palettes friendly to color-blindness,
+but should the reader find contrasts too low anyway,
+whether being color-blind or reading a physical copy printed in greyscale,
+they may be assured all important distinctions still render in monochrome;
+for example, ovals and rectangles to depict different threads,
+or $\dagger$ to distinguish state space estimates from completed verifications.
+
+Pronoun use will vary between more specific and more ambiguous to convey additional nuance.
+The singular ``I'' is associated with my own research contributions,
+while the royal ``we'' should be taken to include the reader,
+such as when surveying background material or related work,
+to which the author and reader share more similar relationships.
+The impersonal ``the programmer'' will be referred to as she/her
+to distinguish her from the thesis audience concerned with the underlying research,
+and also to challenge readers' unconscious bias about gender in computer science.
+Individual students who participated in the user studies will be given the more inclusive they/them.
+Gender-neutral pronouns are also to be used on the author themself.
+
+This document is, in a way, only half the work of the thesis, the other half being Landslide's implementation.
+While some readers may prefer to be taught in prose and/or mathematical notation how an algorithm works,
+others may find that disorienting and wish to see things in a way a compiler would understand.
+The beginning of Chapter~\ref{chap:landslide} provides source code links,
+and the rest of it serves as a guide to browsing the repository.
+Later chapters will often make parenthetical references to specific files and/or functions
+therein which implement a feature under discussion.
+
+% TODO:
+% Finally, this thesis also contains a small puzzle-hunt to reward dedicated readers who make it all the way through.
 

### 2018-08-14 thesis/landslide.tex
challenge unconscious bias

@@
 Three setup scripts are provided, one for each supported kernel architecture: {\tt p2-setup.sh}, {\tt psu-setup.sh}, and {\tt pintos-setup.sh}.
-The user should supply the directory containing their project implementation.
+The user should supply the directory containing her project implementation.
 The second of the three is largely the same as the first, with CMU-specific project details replaced by PSU-specific ones.
@@
 \begin{itemize}
-	\item Checks if the user needs to run {\tt *-setup.sh} again, in case their source code was more recently updated than the existing annotated build (a common mistake),
+	\item Checks if the user needs to run {\tt *-setup.sh} again, in case her source code was more recently updated than the existing annotated build (a common mistake),
 	\item Passes its arguments through to {\tt id/landslide-id}, the Quicksand binary,
@@
 the user may run a single instance directly, optionally configuring the preemption points by hand.
-This is recommended only for enthusiastic users who are annotating their own kernel.
+This is recommended only for the enthusiastic user annotating her own kernel.
 
@@
 For P2 and Pintos students I supply these annotations behind the scenes,
-but a CMU 15-410 student who wishes to use Landslide on their kernel project shall need to brave forth hereupon.
+but a CMU 15-410 student who wishes to use Landslide on her kernel project shall need to brave forth hereupon.
 
@@
 		all the different ways {\tt thr\_exit()} and {\tt thr\_join()} could interleave.
-		Users of this option should be careful to enable the proper {\tt thr\_create()} misbehave mode in their test case (\sect{\ref{sec:landslide-friendly-misbehave}}).
+		The user should be careful with this option to also enable the proper {\tt thr\_create()} misbehave mode
+		in her test case (\sect{\ref{sec:landslide-friendly-misbehave}}).
 	\item {\tt TRUSTED\_THR\_JOIN=0} - if set to 1, forces Landslide to add a happens-before edge (\sect{\ref{sec:landslide-dpor-hb}}) between the exiting of some thread $N$ and the end of any subsequent {\tt thr\_join}$(N)$ call,
@@
 The following annotations are provided as C functions
-which a kernel author shall include in their source code and call at appropriate times.
+which a kernel author shall include in her source code and call at appropriate times.
 The functions' actual implementations are empty;
@@
 The entry point is the aptly-named {\tt landslide\_entrypoint()} in {\tt landslide.c},
-which then dispatches to various other modules their respective analyses, described as follows.
+which then dispatches to various other modules' respective analyses, described as follows.
 
@@
 but rather must keep triggering context switches until the desired thread is reached;
-any mechanism to tell the kernel which thread we want would necessarily involve modifying the code being testsed
+any mechanism to tell the kernel which thread it wants would necessarily involve modifying the code being testsed
 and hence possibly obscuring bugs or introducing new ones.%
@@
 because the nature of the transition (its memory accesses, the subsequent state, etc)
-is unknown until we actually execute it.}
+is unknown until actually executed.}
 Landslide's implementation represents $S$ as a list of {\tt struct hax}es, defined in {\tt tree.h},

### 2018-08-14 thesis/tm.tex
nother tm result
more txn bugs
challenge unconscious bias

@@
 which shows that correctly synchronizing even the simplest transactions may be surprisingly difficult or complex.
-%further motivating our research.
+%further motivating the research.
 
@@
 \end{itemize}
-Note that arbitrary code may not be structured to distinguish these as nicely as in our examples;
+Note that arbitrary code may not be structured to distinguish these as nicely as in the examples;
 e.g., more code may exist in the success branch after {\tt \_xend()};
@@
 		representing either STM's policy of failing only when a true memory conflict arises,
-		or a TSX user wrapping all their {\tt \_xbegin()}s in a retry loop such as in \cite{sigbovik-htm}.
+		or a TSX user wrapping all her {\tt \_xbegin()}s in a retry loop such as in \cite{sigbovik-htm}.
 	\item {\tt \_XABORT\_CONFLICT}:
@@
 		{\tt swapbug}
-			& 2,1	&\cpu{104.37}&\wtm{17.00}&\ints{*6}& \cpu{*43.76}& \wtm{*12.51}& \ints{33	}& \ETA{73} \\
+			& 2,1	&\cpu{104.37}&\wtm{17.00}&\ints{*6}& \cpu{*43.76}& \wtm{*12.51}& \ints{33}	& \ETA{73} \\
+			% TODO: for maximal:
+			% 33 interleavings tested; 1 preemptions; job time 4s; pldi time 12326606; new-fixed pldi cputime 43294461
+			% 73 brs eta
 		(deadlock)
-			& 2,2	&\cpu{ 66.47}&\wtm{*12.42}&\ints{*38}& \cpu{*45.22}& \wtm{15.30}& \ints{85	}& \ETA{860} \\
-			& 2,3	&\cpu{ 99.41}&\wtm{*15.92}&\ints{*48}& \cpu{*53.57}& \wtm{21.34}& \ints{217	}& \ETA{9120} \\
-			& 2,4	&\cpu{125.23}&\wtm{*19.46}&\ints{*68}& \cpu{*71.01}& \wtm{38.86}& \ints{537	}& \ETA{91239} \\
+			& 2,2	&\cpu{ 66.47}&\wtm{*12.42}&\ints{*38}& \cpu{*45.22}& \wtm{15.30}& \ints{85}	& \ETA{860} \\
+			& 2,3	&\cpu{ 99.41}&\wtm{*15.92}&\ints{*48}& \cpu{*53.57}& \wtm{21.34}& \ints{217}	& \ETA{9120} \\
+			\multicolumn{9}{c}{everything above this line needs rerun for dpor-preferred tid + thr\_bottom1} \\
+			& 2,4	&\cpu{*38.92}&\wtm{*8.61}&\ints{*34}& \cpu{82.95}& \wtm{48.31}& \ints{537}	& \ETA{91239} \\
+			% 34 interleavings tested; 1 preemptions; job time 6s; pldi time 8614371; new-fixed pldi cputime 38917975
+			% 537 interleavings tested; 1 preemptions; job time 38s; pldi time 48314592; new-fixed pldi cputime 82952352
 		\cline{2-9}
-			& 3,1	&\cpu{313.36}&\wtm{*44.50}&\ints{*34}&\cpu{*78.23}& \wtm{60.49}& \ints{1016	}& \ETA{3543} \\
-			& 3,2	&\cpu{*6500.36}&\wtm{*816.27}&\ints{*262}&\ETA{--}&\ETA{>1h}&\ETA{--}& \ETA{2161715} \\
-		\hline
-			\multicolumn{9}{c}{everything above this line needs rerun for dpor-preferred tid + thr\_bottom1} \\
+			& 3,1	&\cpu{*38.59}&\wtm{*8.50}&\ints{*32}&\cpu{88.28}& \wtm{72.41}& \ints{1016}	& \ETA{3543} \\
+			% 32 interleavings; 258 preemptions; job time 5s; pldi time 8506344; new-fixed pldi cputime 38594745
+			% 1016 interleavings; 258 preemptions; job time 58s; pldi time 72406995; new-fixed pldi cputime 88281009
+			& 3,2	&\cpu{*1572.83}&\wtm{*199.98}&\ints{*262}&\ETA{--}&\ETA{>1h}&\ETA{--}& \ETA{1683509} \\
+			% 262 interleavings; 260 preemptions; job time 47s; pldi time 199982231; new-fixed pldi cputime 1572825801
+			% [JOB 8] progress: 47047/1683509 brs (2.794578%), ETA 1y 237d 20h 41m 57s (elapsed 59m 44s)
+		\hline
 		{\tt avl\_insert}
@@
 as HTM is fundamentally a performance-minded concurrency extension,
-users are likely to care about more probabilistic properties of their code,
+the user may also care about more probabilistic properties of her code,
 such as requiring a transaction abort rate below a certain threshhold
@@
 % 'so on' here includes free/re-malloc mem conflicts; see comment in avl_mutex.cpp
-their state space growth now defined only by their internal conflicts from tree rebalancing, map resizing, and so on.
+their state space growth now defined only by internal conflicts from tree rebalancing, map resizing, and so on.
 In total,
@@
 			& 3,1 & \cpu{28.27} & \ints{186} \\
-			& 3,2 & \ETAdag{qqq} & \ETAdag{qqq} \\
+			& 3,2 & \ETAdag{13y 281d} & \ETAdag{1443676} \\
+			% [JOB 4] progress: 169893/1443676 brs (11.768080%), ETA 13y 280d 17h 30m 54s (elapsed 9h 59m 48s)
 			& 4,1 & \ETAdag{16h 26m} & \ETAdag{432628} \\ % was: 30408, 282084 pre mkrun
@@
 
-% "Although we skipped implementing it to begin with, the sheer size of this reduction also
+% "Although I skipped implementing it to begin with, the sheer size of this reduction also
 % reflects on the numbers one might see betwee the original reduction in theorem so-and-so that lets us
@@
 while {\tt lock}({\tt spinlock}) and {\tt lock}({\tt mutex}) implement the retry loop directly.
-In these cases, the user can assure themself of STM reduction's soundness by visual inspection.
+In these cases, the user can assure herself of STM reduction's soundness by visual inspection.
 In fact, Landslide's current implementation gets stuck in infinitely deep interleavings

### 2018-08-14 thesis/intro.tex
#NewThesisStatement

@@
 \begin{center}
-	% TODO: this sux, fix it
-	{\em Thanks to the new algorithms, heuristics, and concurrency models I have developed,
-	stateless model checking is an appropriate and accessible concurrency testing technique
-	for programmers in both educational and real-world settings.}
+	{\em Combining both theoretically-founded automatic reduction techniques
+	and user-informed heuristic ones,
+	stateless model checking can sufficiently mitigate exponential explosion
+	to be a practical testing technique even for inexperienced users.}
 \end{center}

### 2018-08-14 thesis/landslide.tex
test case misbehave (is dumb)

@@
 
-Many of the supplied P2 test cases invoke the {\tt misbehave()} system call
+Many of the supplied P2 test cases invoke the {\tt misbehave} system call
 with a mysterious argument (usually {\tt BGND\_BRWN >> FGND\_CYAN})
@@
 
-To illustrate, consider the following program.
-
-% TODO
-% TODO - make a progr that data races immediately after thr create, no mutex es or whatever.
-% TODO - check if it rly happens under pure happens before? or if this is only a concern under limited HB.
+To illustrate, consider the following program in Figure~\ref{fig:misbehave-example},
+and suppose Landslide is configured to preempt only on mutex API calls
+(such as in the first step of Iterative Deepening (\sect{\ref{sec:quicksand-id}})).
+Because Landslide ignores all kernel-level synchronization short of context switches when testing user-level code,
+if the kernel created the child thread and returned from {\tt thread\_fork}
+(the system call underlying {\tt thr\_create()})
+without yielding first,
+the next preemption point will not occur until {\tt thr\_join()} waits for the child to exit.
+Hence, DPOR will erroneously think everything before that {\tt thr\_join()}
+happens-before (\sect{\ref{sec:landslide-dpor-hb}}) anything the child does,
+and will fail to identify the racing accesses on {\tt x}.
+
+\begin{figure}[h]
+	\begin{center}
+	\begin{tabular}{l}
+		\texttt{\ctype{void} \call{child}(\ctype{int} *xp) \{} \\
+		\texttt{~~~~*xp++;} \\
+		\texttt{\}} \\
+		\\
+		\texttt{\ctype{void} \call{parent}() \{} \\
+		\texttt{~~~~\ctype{int} x = \const{0};} \\
+		\texttt{~~~~\ctype{int} tid = \call{thr\_create}(child, \&x);} \\
+		\texttt{~~~~x++;} \\
+		\texttt{~~~~\call{thr\_join}(tid, \const{NULL});} \\
+		\texttt{~~~~\flow{assert}{}(x == \const{2});} \\
+		\texttt{\}} \\
+	\end{tabular}
+	\end{center}
+	\caption{Example demonstrating the need for {\tt misbehave} to force the kernel
+	to yield during {\tt thread\_fork}.}
+	\label{fig:misbehave-example}
+\end{figure}
+
+Though Iterative Deepening's soundness (\sect{\ref{sec:quicksand-soundness}})
+guarantees all data races will eventually be detected starting from just synchronization preemption points,
+it assumes threads becoming runnable counts among those.
+In that sense, {\tt misbehave} serves to restore the last synchronization preemptions where they belong.
+If at this point the reader wonders why Landslide doesn't just identify
+the {\tt thread\_create} and {\tt make\_runnable} system calls in the arbiter itself (\sect{\ref{sec:landslide-arbiter}})
+and skip this mysterious user-visible complexity,
+they would be right to ask:
+I have left it this way for no better reason than to maintain consistency
+with the upcoming chapters' experimental environments,
+and intend on fixing it in a future update.
 

### 2018-08-14 thesis/intro.tex
gayer colors

@@
 but only in redundant ways also signaled by symbols.
-I have made some effort to choose palettes friendly to color-blindness,
-but should the reader find contrasts too low anyway,
+I have made some effort to choose palettes friendly to color-blindness;
+should the reader find contrasts too low anyway,
 whether being color-blind or reading a physical copy printed in greyscale,
-they may be assured all important distinctions still render in monochrome;
-for example, ovals and rectangles to depict different threads,
-or $\dagger$ to distinguish state space estimates from completed verifications.
+they may be assured all important distinctions still render in monochrome.
+For example, ovals and rectangles typically depict different threads,
+and $\dagger$ distinguishes state space estimates from completed verifications.
 

### 2018-08-14 thesis/tm.tex
more tm tests

@@
         \begin{tabular}{ll}
-		\multicolumn{2}{c}{Initially {\tt \ctype{int} foo, bar = \const{0}; \ctype{mutex\_t} m;}} \\
+		\multicolumn{2}{c}{Initially {\tt \ctype{int} foo, bar = \const{0};}} \\
                 \\
@@
 			{\tt map\_basicer}
-			& 2,1 & \cpu{28.40} 	& \ints{150} 	& \cpu{94}	& \ints{26.55}	& \cpu{14.88}	& \ints{9} \\
+			& 2,1 & \cpu{28.40} 	& \ints{150} 	& \cpu{26.55}	& \ints{94}	& \cpu{14.88}	& \ints{9} \\
 			& 2,2 & \ETAdag{qqq}	& \ETAdag{qqq}	&\cpu{16455.24}	& \ints{283756}	& \cpu{1285.06}	& \ints{21684} \\
@@
 			& 2,2 & \cpu{105.57} & \ints{998} \\
-			& 2,3 & \ETAdag{qqq} & \ETAdag{qqq} \\
+			& 2,3 & \ETAdag{33h 13m} & \ETAdag{321553} \\
+			% [JOB 4] progress: 138911/321553 brs (43.200015%), ETA 23h 12m 50s (elapsed 9h 59m 48s)
+			% total CPU time consumed: 10h 4s (36004577784 usecs) (core saturation: 12%)
 			& 3,1 & \cpu{28.27} & \ints{186} \\
@@
 			& 2,3 & \cpu{405.21} & \ints{9372}	\\
-			& 2,4 & \cpu{qqq} & \ints{qqq}	\\ % was: 15266 sec 384512 pre mkrun
+			& 2,4 & \cpu{24999.68} & \ints{489480}	\\ % was: 15266 sec 384512 pre mkrun
+			% total CPU time consumed: 6h 56m 39s (24999680839 usecs) (core saturation: 12%)
 			& 3,1 & \cpu{15.00} & \ints{132}	\\
 			%& 3,2 & \cpu{29432.87} & \ints{707086}	\\ % pre mkrun fix (let alone bottom1)
-			& 3,2 & \ETAdag{qqq} & \ETAdag{qqq} \\
+			& 3,2 & \ETAdag{26h 42m} & \ETAdag{1223955} \\
+			% [JOB 2] progress: 714882/1223955 brs (58.407517%), ETA 16h 41m 58s (elapsed 9h 59m 55s)
+			% total CPU time consumed: 10h 3s (36003621109 usecs) (core saturation: 12%)
 			& 4,1 & \cpu{665.89} & \ints{15064}	\\

### 2018-08-14 thesis/overview.tex
some new inspirational quotes

@@
 
-\inspirationalquote{
-\begin{tabular}{p{0.83\textwidth}}
-I see now that none of us are yet ready. The cycle exists so that we may improve ourselves. But
-the one who reaches the summit is not our superior, for they stand on our shoulders to reach it.
-\end{tabular}}
-{The Shepherd v82.6.0174, The Talos Principle}
+%\inspirationalquote{
+%\begin{tabular}{p{0.83\textwidth}}
+%I see now that none of us are yet ready. The cycle exists so that we may improve ourselves. But
+%the one who reaches the summit is not our superior, for they stand on our shoulders to reach it.
+%\end{tabular}}
+%{The Shepherd v82.6.0174, The Talos Principle}
+
+\inspirationalquote{We pray to our ancestors. We've met all of them. None of them were better than anyone else.}
+{Mina, Iconoclasts}
 

### 2018-08-14 thesis/related.tex
some new inspirational quotes

@@
 \label{chap:relatedwork}
+%\inspirationalquote{
+%\begin{tabular}{p{0.58\textwidth}}
+%To test if your paper makes a genuine contribution to its discipline,
+%see if you can afford a generous tone in the "Related Work" section.
+%\end{tabular}}
+%{Conor McBride}
+
 \inspirationalquote{
-\begin{tabular}{p{0.58\textwidth}}
-To test if your paper makes a genuine contribution to its discipline,
-see if you can afford a generous tone in the "Related Work" section.
-\end{tabular}}
-{Conor McBride}
+\begin{tabular}{p{0.51\textwidth}}
+It is important to draw wisdom from many different places.
+If you take it from only one place, it becomes rigid and stale.
+\end{tabular}
+}
+{Iroh, Avatar: The Last Airbender}
 

### 2018-08-14 thesis/overview.tex
overview 1st p pronouns

@@
 Because clock interrupts can occur at any instruction\footnote{
-	With some exceptions in kernel-level programming, which I discuss later.
+	With some exceptions in kernel-level programming, discussed later.
 },
@@
 at the cost of increasing the implementation and theoretical complexity of the analysis.
-In this thesis, I will consider locks and descheduling to be the only concurrency primitives,
+This thesis will consider locks, descheduling, and transaction begin/end as the only concurrency primitives,
 and assume the others listed above are implemented using those as building blocks (an exercise for the reader \cite{thrlib}).
@@
 HTM, however, is backed by the CPU's cache, and is therefore subject to other circumstances such as cache capacity or interrupt-triggered cache flushes which may force an abort even when no memory conflict occurs.
-I will explore the consequences of this difference further in Chapter~\ref{chap:tm}.
-
-In this thesis, I will focus on HTM as my platform for testing transactional programs,
+Chapter~\ref{chap:tm} will explore the consequences of this difference further.
+%
+This thesis will focus on HTM as my platform for testing transactional programs,
 to highlight the importance of researching advanced testing techniques in anticipation of upcoming hardware features.
@@
 This approach is called {\em stateful} model checking.
-In this thesis, I focus on {\em stateless} model checking (and execution trees, not DAGs),
+This thesis focuses on {\em stateless} model checking (and execution trees, not DAGs),
 which instead analyzes the sequence of execution events to avoid a prohibitive memory footprint.
-Henceforth I will abbreviate ``stateless model checking'' simply as ``model checking'' for brevity.
+Henceforth ``stateless model checking'' will be abbreviated simply as ``model checking'' for brevity.
 Also, the term ``state space'' was originally coined to refer to the stateful approach's emphasis on the DAG's nodes
@@
 in order to check all possible program states.
-I explain how DPOR implements such a deduction in more detail in \sect{\ref{sec:landslide-dpor}}.
+\sect{\ref{sec:landslide-dpor}} explains how DPOR implements such a deduction in more detail.
 
@@
 
-In this thesis I will tackle Pebbles and Pintos, two different system architectures used in educational operating systems courses.
+This thesis will tackle Pebbles and Pintos, two different system architectures used in educational operating systems courses.
 This section describes the projects which students implement and which Landslide tests.
@@
 %
-In this thesis I will occasionally use intentionally vague phrasing to preserve the mystery of these bugs.
+I will occasionally use intentionally vague phrasing to preserve the mystery of these bugs.
 
@@
 Some schools further offer a virtual memory project, extending the provided VM with a frame table and supplemental page table and fault handler \cite{standford-cs140,uchicago-cs230}, or supplemental HTTP server and {\tt malloc} assignments \cite{berkeley-cs162}.
-Being largely architectural/algorithmic projects rather than concurrency-oriented ones, I am not concerned with these assignments in this thesis.
+Being largely architectural/algorithmic projects rather than concurrency-oriented ones, this thesis is not concerned with these assignments.
 The main concurrency challenges in Pintos projects arise from the {\em threads} and {\em userprog} assignments:

### 2018-08-14 thesis/intro.tex
revise some out of date stuff

@@
 		to evaluate the suitability of stateless model checking in an educational setting.
-	\item {\bf Transactional Memory (Chapter \ref{chap:tm}).}
-		Transactional Memory (TM) is a relatively new concurrent programming technique \cite{transactional-memory}
+	\item {\bf Hardware Transactional Memory (HTM) (Chapter \ref{chap:tm}).}
+		HTM is a relatively new concurrent programming technique \cite{transactional-memory}
 		which is not yet addressed by modern model checkers.
-		I have extended Landslide's concurrency model to support both hardware (HTM) and software (STM) variants of TM,
-		and tested several ``real-world'' TM programs and benchmarks.
+		I have extended Landslide's concurrency model to support
+		HTM's execution semantics and several advanced features, % abort codes p much
+		%both hardware (HTM) and software (STM) variants of TM,
+		and tested several ``real-world'' HTM programs and benchmarks.
 		This section will discuss the theoretical techniques I used to model the new form of concurrency,
 		present associated correctness proofs of my approach,
-		and show the testing results.
+		and show the verification results.
 \end{enumerate}

### 2018-08-14 thesis/landslide.tex
revise some out of date stuff

@@
 defined in {\tt schedule.h},
-decides if a given thread would require a preemption beyond the current bound to switch to.
-The DPOR implementation then checks, for some $I_{ij}$ it wants to mark for exploration,
-whether {\tt ICB\_BLOCKED(}$T_j${\tt )} at the state after $t_i$,
-and skips it if so ({\tt tag\_good\_sibling()}/{\tt tag\_all\_siblings()}).
+decides if a given thread would require a preemption beyond the current bound to switch to%
 \footnote{Since ``voluntary'' context switches (e.g. arising from {\tt yield()})
@@
 Therefore, within a certain preemption bound $b$,
-interleavings with more than $b$ context switches may still be tested.}
+interleavings with more than $b$ context switches may still be tested.
+}.
+The DPOR implementation then checks, for some $I_{ij}$ it wants to mark for exploration,
+whether {\tt ICB\_BLOCKED(}$T_j${\tt )} at the state after $t_i$,
+and skips it if so ({\tt tag\_good\_sibling()}/{\tt tag\_all\_siblings()}).
 

### 2018-08-16 thesis/education.tex
bias and pebbles retro

@@
 (both versions).
+
+During the last week of P2 at CMU, I held several ``Landslide clinic'' sessions
+(basically office hours, but given a different name to remind students
+to limit themselves to questions a normal TA couldn't answer),
+where students could receive in-person technical and/or moral support.
+Collecting study information during these sessions was not included in the IRB protocol.
+In the PSU study, I had returned to Pittsburgh shortly after giving the lecture,
+so technical support for PSU students was limited to email correspondence.
 
@@
 
-%%%% pintos %%%% (?? this is psu not pintos)
-% survey questions i WISH i had asked
-% - did you have any technical difficulties w landslide that i had to intervene on
-% mb anything else from timmys 2nd latest email
+As long as an educational user study is run on a volunteer basis,
+one cannot completely avoid selection bias:
+those with enough free time to participate are more likely to be the most capable students already,
+who are least in need of the extra debugging help.
+This was especially evident in the annotation-required P3 study from \cite{landslide},
+in which only 5 groups volunteered (15\% of the 34 total who submitted P3 that semester),
+among which 2 had enough after just the annotation phase and did not continue to do any in-depth testing.
+In contrast, since switching to the automatically-annotatable P2,
+the participation rate rose to NN\% % TODO
+(XX out of YY P2-submitting groups across ZZ semesters (TODO actually count them but obviously it will be higher)). % TODO
+This alone could be seen as a major step in mitigating said selection bias.
+
+Beyond participation, I anticipated several forms of bias affecting the survey responses.
+First of all, I suspected that students might feel compelled to be overly polite in their answers
+(whether consciously or subconsciously).
+I attempted to counteract this by concluding my survey link emails with,
+{\em Please try to answer honestly rather than flatteringly--if any part
+of the experience was bad for you, I want to hear about it to make
+Landslide better in the future!}
+It's also possible that survey respondents % respondence
+were more likely to be those who enjoyed Landslide the most,
+meaning I might not hear as much negative feedback as I should.
+Table~\ref{tab:TODO} % TODO -- make a table or graph showing correlation between bugfinding and who responded on the survey
+is my attempt to dispel this possibility.
+% TODO
+% possible flaws in table: students who found bugs might have had either a good (bug!) or bad (false positive) experience
+% could studence who didn't find bugs also have had a good experience? (think of how...)
+
+% TODO: is this ok to report without IRB approval? ask dave
+I cannot think of any reason gender, racial, or other demographic-based bias might affect this study.
+In fact, according to my memory alone, the proportion of women among students
+attending the Landslide clinics (\sect{\ref{sec:education-pebbles-recruiting}})
+was considerably higher than the class's overall gender ratio,
+although I did not record any such statistics scientifically.
 
@@
 
-% TODO
+The survey was introduced into the study only in time for two semesters' worth of student responses,
+after several iterations of collecting only usage and bug report snapshots.
+Apart from the obvious improvement of having been surveying students from the beginning,
+the following questions could have improved the survey.
+\begin{enumerate}
+	\item {\em Did you have any technical difficulties with Landslide that I had to intervene on,
+		whether in person or over email?}
+		(Some students reported this in the ``false positives'' question,
+		although fewer than I helped overall, so others must have not mentioned it.)
+		Comparing answers to this question across subsequent semesters would give a sense
+		of how much Landslide's stability was improving over time
+		and whether it was mature enough for unsupervised use in the future.
+	\item {\em How well do you feel you understood the research challenges explained in the lecture?}
+		and,
+		{\em How well do you feel a user should need to understand same in order to benefit from Landslide's bug reports?}
+		(Answers on a scale from ``Not at all; Landslide is a magic black box to me''
+		to ``I'm ready to work on research in this field myself''.)
+		These questions would help fine-tune the lecture material and user guide to maximize student comfort,
+		and potentially also corroborate the claim that Landslide is accessible even to novice users.
+	\item {\em What additional debugging information would you want displayed on the preemption traces?}
+		Knowing now that interpreting preemption traces was a sticking point for many users,
+		I would hope to identify the most wished-for features to know what to prioritize improving.
+		This could also assess students' understanding of what kinds of information
+		would or would not be reasonable for Landslide to record and report.
+\end{enumerate}
+
+Some students (between 0 and 2 per semester)
+emailed me during P3
+to ask if they could test their kernels with Landslide just like their thread libraries.
+I answered each by explaining that it would take more effort on their part,
+and if they were still interested,
+guided them through the annotation process on a case-by-case basis.
+This process was not included in the IRB-approved study protocol,
+so I collected no results from them.
+If I had planned in advance, I could have supported this ``bonus stage'' officially,
+and further surveyed the brave volunteers about how P3 Landslide could be made more generally accessible.
+
+Finally, to evaluate whether the experience of using Landslide
+left the students with any lasting lessons learned,
+a follow-up survey could have been given one or two years later.
+Such a survey would ask, for example,
+{\em Have you encountered any debugging problems since finishing OS that made you wish for a tool like Landslide?}
+and
+{\em Do you feel the way you think about testing, debugging, and program correctness
+has been influenced in any way by using Landslide?}.
 

### 2018-08-18 thesis/education.tex
education eval overall conclusions, table of participance, improve discussion of bias

@@
 and U. Chicago students' survey responses from Fall 2017.
+Table~\ref{tab:photo-of-ze-studence} shows the student participation rate across semesters
+% TODO update for psu
+for the thread library study.
+% and PSU
+
+\begin{table}[h]
+	\begin{center}
+		% ugh this looks so stupid
+		% but better than latex's output looking stupid
+		\begin{tabular}{c|crcrc|r}
+			& \multicolumn{5}{c|}{\bf used Landslide?} & \\
+			\bf semester & & \bf yes & & \bf no & & \bf total \\
+			\hline
+			CMU S'15	& &   8	& &  18	& &  26 \\
+			CMU F'15	& &  22	& &   8	& &  30 \\
+			CMU S'16	& &  18	& &  16	& &  34 \\
+			CMU F'16	& &  12	& &   6	& &  18 \\
+			CMU S'17	& &  19	& &  13	& &  32 \\
+			CMU F'17	& &  14	& &   5	& &  19 \\
+			CMU S'18	& &  10	& &   8	& &  18 \\
+			% NB. CMU's participance are counted by how many ppl i got at least one usage tarball for,
+			% while PSU's are by how many sign-up forms i got.
+			% the latter may be overcounting, because a student may sign up and then never use it;
+			% but i don't have tarballs for PSU so this is the best you're getting.
+			% TODO
+			% PSU s18	& &  38
+			\hline
+			\bf total	& & 103	& &  74	& & 177 \\
+		\end{tabular}
+	\end{center}
+	\caption{Student participation rates across semesters,
+		among students who submitted thread library implementations.}
+	\label{tab:photo-of-ze-studence}
+\end{table}
 
@@
 \subsection{Student submission quality}
+\label{sec:education-eval-grades}
 
@@
 (guided, as well, by my own intuition as a former TA).
+I also counted students who used Landslide in general but whose usage snapshots showed they did not run the test case in question as non-users.
 Table~\ref{tab:eval-common-bugs} presents the results.
-
-\begin{table}[h]
+% two-tailed btw, the better (more conservative) way,
+% tbh one tailed might even make sense here, but it doesn't change anything wrt the cutoffs anyway, so
+The $p$ values are calculated in R \cite{r-lang} using Fisher's exact test \cite{fishers-exact-test},
+treating each row as an independent 2x2 contingency table,
+and dividing the significance cutoff by the number of bugs to conservatively account for multiplicity \cite{xkcd-jellybeans}.
+
+% TODO check figure placement
+\begin{table}[t]
 	\begin{center}
-		% TODO: calc p values for this
+		\small
 		% note, smth like fisher's method which "combines independent studies into a meta p value"
@@
 		% running landslide on diff test cases; 2, the hypotheses tested are different for each test
-	\begin{tabular}{r||cc|cc|c}
-		& \multicolumn{2}{c|}{Landslide users} & \multicolumn{2}{c|}{Non-users} & \\
-		Bug name & \#correct & total & \#correct & total & $\Delta P(\mathsf{correct})$ \\
+	\begin{tabular}{r||cc|cc|c|cc}
+		& \multicolumn{2}{c|}{Landslide users} & \multicolumn{2}{c|}{Non-users} & & \multicolumn{2}{c}{Significance} \\
+		Bug name & \#correct & \#buggy & \#correct & \#buggy & $\Delta$\%correct & $p$ & cutoff \\
 		\hline
-		%Exit UAF  & 9	& 21	& 2	& 16	& $+30.4\%$ \\ % not filtering out did not run tej; also using "old" thing..?
-		Exit UAF  & 11	& 21	& 0	& 16	& $+52.4\%$ \\
+		% calculated in R with (rest done similarly)
+		% fisher.test(rbind(c(11,10),c(0,16)))$p.value --> 0.0006143355
+		Exit UAF  & 11	& 10	& 0	& 16	& $+52.4\%$	& 0.00061	& 0.0125 \\
 		% I spent more hours than I should have looking at this landslide-user-submitted-bug group.
@@
 		% but didn't rerun mutex_test to check their update.
-		Mutex     & 17	& 18	& 19	& 19	& $-05.6\%$ \\
-		% nb. this is not filter out the did not run paras -- smae for all of the above/below
-		% Paraguay  & 22	& 24	& 12	& 13	& $-0.64\%$ \\ % not filtering out did not run para
-		Paraguay  & 20	& 21	& 14	& 16	& $+07.7\%$ \\
+		Mutex     & 17	& 1	& 19	& 0	& $-05.6\%$	& 0.49 & 0.0125 \\
+		Paraguay  & 20	& 1	& 14	& 2	& $+07.7\%$	& 0.57	& 0.0125 \\
 		% nb this is any downgrade related ding not just dgr tested ones
 		% nb 1 is msising bc didn't impl rwlock at all
-		%Downgrade & 17	& 24	& 8	& 12	& $+4.17\%$ \\ % not filtering out either did not run, or unrel ding
-		Downgrade & 10	& 13	& 19	& 23	& $-05.7\%$ \\ % TODO verify these buggers by eye
+		% nice p value
+		Downgrade & 10	& 3	& 19	& 4	& $-05.7\%$	& 0.69	& 0.0125 \\ % TODO verify these buggers by eye
 		\hline
-		% TODO: figure out if calculating a p value (or w/e) for this the same way is valid,
-		% since they summands are arguably not independent? ..or maybe they are?
-		Total     & 58	& 73	& 52	& 74	& $+09.2\%$ \\
+		Total     & 58	& 15	& 52	& 22	& $+09.2\%$	& 0.26 & 0.05 \\
 	\end{tabular}
@@
 		Correlation of student Landslide use with solving certain bugs in their final submission during Fall 2017 and Spring 2018 semesters.
-		Students who used Landslide in general but whose usage snapshots showed they did not run the test case in question are counted as non-Landslide-users here.
-		$\Delta P$ stands for the increased probability that a student submits a correct P2
-		(with respect to each bug)
-		given that they use Landslide, versus not using Landslide; i.e.,
-		$P(\mathsf{correct}|\mathsf{landslide})-
-		P(\mathsf{correct}|\mathsf{non\text{-}user})$.
+		%$\Delta P$ stands for the increased probability that a student submits a correct P2
+		%(with respect to each bug)
+		%given that they use Landslide, versus not using Landslide; i.e.,
+		%$P(\mathsf{correct}|\mathsf{landslide})-
+		%P(\mathsf{correct}|\mathsf{non\text{-}user})$.
 		Note that the totals in the bottom row double-count students,
-		which only makes sense if you believe the incidences of each bug within a given submission are independent.
+		which makes sense only if you believe the incidences of each bug in a given submission are independent.
 		}
@@
 \subsection{Survey responses}
+\label{sec:education-eval-survey}
 
@@
 
-% TODO intro text
+Human subjects research is inherently messy.
+Each of the previous section's approaches to evaluating Landslide's educational value
+was accompanied by some drawback which prevented it from being perfectly objective science,
+but many of them presented tentatively positive results nonetheless.
+Landslide helped many students find and fix many bugs (\sect{\ref{sec:education-eval-bugfinding}}),
+but making a direct comparison to stress testing, the prior state of the art, is not straightforward.
+Little, if any,
+%and not statistically significant,
+immediate improvement in students' project grades was measured
+(\sect{\ref{sec:education-eval-grades}}),
+but impact on grades alone is a very narrow measure of pedagogical value.
+Landslide's debugging power was also found to be statistically significant
+for the {\tt thr\_exit\_join} bug in particular, even after adjusting for multiplicity.
+Students responded overwhelmingly positively in the survey (\sect{\ref{sec:education-eval-survey}}),
+although it is easy to imagine students being equally happy with a ``debugging tool'' that just tells them all the answers.
+
+Nevertheless, I believe each of these partial results taken together
+paint an overall picture of success:
+students fixed their own bugs {\em and} were happy about it;
+students were able to ask for help rather than be deterred by inevitable technical difficulties;
+students provided intelligent feedback suggesting they truly understood the debugging process;
+% well not the student who said "traces should point to where the bug is" but eh can't win em all c.c
+et cetera.
+%
+The rest of this section will discuss the study's limitations and provide some perspective for the future.
 
@@
 
-% TODO: addressing bias
-% well, I did the best I could(?)
-% survey email: "please answer honestly rather than flatteringly" vs. studence trying to be polite
-% selection bias, students who liked landslide more more likely to answer the survey at all
 % TODO - u can verify the bug just by taking survey graph #1 and comparing to landslide snapshots (sep. by groups who answered survey)
@@
 In contrast, since switching to the automatically-annotatable P2,
-the participation rate rose to NN\% % TODO
-(XX out of YY P2-submitting groups across ZZ semesters (TODO actually count them but obviously it will be higher)). % TODO
-This alone could be seen as a major step in mitigating said selection bias.
+the participation rate rose to 58\% (Table~\ref{tab:photo-of-ze-studence})
+among all P2-submitting groups.
+Reaching over half the class could alone be seen as a major step in mitigating said selection bias.
 
@@
 % TODO: is this ok to report without IRB approval? ask dave
-I cannot think of any reason gender, racial, or other demographic-based bias might affect this study.
-In fact, according to my memory alone, the proportion of women among students
-attending the Landslide clinics (\sect{\ref{sec:education-pebbles-recruiting}})
-was considerably higher than the class's overall gender ratio,
-although I did not record any such statistics scientifically.
+I took no special measures to compensate for bias in gender, race, or being non-natively English-speaking
+during recruitment or the survey.
+Surveying students to measure any differences in these demographics between study participants and the overall class population
+would have required mandatory survey participation,
+and in turn, a more rigorous IRB approval process.
+According only to my memory of students who attended the Landslide clinics
+(\sect{\ref{sec:education-pebbles-recruiting}}),
+the racial diversity was roughly representative of the class at large,
+and the proportion of women was in fact somewhat higher than the overall gender ratio.
+Any more scientific analysis of such statistics was deemed beyond the study's scope for now.
 
@@
 
-Human subjects research is inherently messy.
 More than just trying to draw firm conclusions from

### 2018-08-19 thesis/tm.tex
more tm tests

@@
 Note that the C-style {\tt x++} operations, when compiled into assembly, %\cite{printy},
-results in multiple memory accesses which can be interleaved with other threads.
+become multiple memory accesses which can be interleaved with other threads.
 \[
@@
 			& 4,2	&\cpu{64.99}&\wtm{11.33}&\ints{9}& \cpu{*8.32}& \wtm{*5.59}& \ints{9	}& \ETA{132710} \\
+			\multicolumn{9}{c}{everything above this line needs rerun for dpor-preferred tid + thr\_bottom1} \\
 		\hline
 		{\tt swapbug}
-			& 2,1	&\cpu{104.37}&\wtm{17.00}&\ints{*6}& \cpu{*43.76}& \wtm{*12.51}& \ints{33}	& \ETA{73} \\
-			% TODO: for maximal:
-			% 33 interleavings tested; 1 preemptions; job time 4s; pldi time 12326606; new-fixed pldi cputime 43294461
-			% 73 brs eta
+			& 2,1	&\cpu{*26.25}&\wtm{*6.42}&\ints{*6}& \cpu{47.80}& \wtm{13.43}& \ints{33}	& \ETA{73} \\
+			% using fewer interleavings one for quicksand out of 2 close in cputime bugs
+			% 6 interleavings tested; 1 preemptions; job time 4s; pldi time 6415943; new-fixed pldi cputime 26251146
+			% 13 interleavings tested; 1 preemptions; job time 3s; pldi time 4654607; new-fixed pldi cputime 16966236
+			% 33 interleavings tested; 1 preemptions; job time 4s; pldi time 13430843; new-fixed pldi cputime 47803093
 		(deadlock)
-			& 2,2	&\cpu{ 66.47}&\wtm{*12.42}&\ints{*38}& \cpu{*45.22}& \wtm{15.30}& \ints{85}	& \ETA{860} \\
-			& 2,3	&\cpu{ 99.41}&\wtm{*15.92}&\ints{*48}& \cpu{*53.57}& \wtm{21.34}& \ints{217}	& \ETA{9120} \\
-			\multicolumn{9}{c}{everything above this line needs rerun for dpor-preferred tid + thr\_bottom1} \\
+			& 2,2	&\cpu{*18.08}&\wtm{*4.98}&\ints{*10}& \cpu{51.37}& \wtm{16.78}& \ints{85}	& \ETA{860} \\
+			% 10 interleavings tested; 1 preemptions; job time 3s; pldi time 4980992; new-fixed pldi cputime 18078134
+			% 85 interleavings tested; 1 preemptions; job time 7s; pldi time 16777253; new-fixed pldi cputime 51372309
+			& 2,3	&\cpu{*20.93}&\wtm{*5.57}&\ints{*18}& \cpu{57.87}& \wtm{23.99}& \ints{217}	& \ETA{9120} \\
+			% 18 interleavings tested; 1 preemptions; job time 4s; pldi time 5573337; new-fixed pldi cputime 20932814
+			% 217 interleavings tested; 1 preemptions; job time 15s; pldi time 23987230; new-fixed pldi cputime 57868535
 			& 2,4	&\cpu{*38.92}&\wtm{*8.61}&\ints{*34}& \cpu{82.95}& \wtm{48.31}& \ints{537}	& \ETA{91239} \\
@@
 			% progress: 302022/35094477 brs (0.860597%), ETA 36392y 304d 6h 53m 45s (elapsed 9h 58m 49s)
-			& 3,1 & \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq} \\
+			& 3,1 & \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	&\ETAdag{97y 100d}&\ETAdag{1698185036} \\
+			% [JOB 18] progress: 357586/1698185036 brs (0.021057%), ETA 97y 99d 8h 44m 33s (elapsed 9h 59m 41s)
 			\hline
@@
 			& 2,1 & \cpu{28.40} 	& \ints{150} 	& \cpu{26.55}	& \ints{94}	& \cpu{14.88}	& \ints{9} \\
-			& 2,2 & \ETAdag{qqq}	& \ETAdag{qqq}	&\cpu{16455.24}	& \ints{283756}	& \cpu{1285.06}	& \ints{21684} \\
-			& 3,1 & \ETAdag{qqq}	& \ETAdag{qqq}	&\cpu{21153.64}	& \ints{366030}	& \cpu{705.04}	& \ints{12707} \\
+			& 2,2 &\ETAdag{11h 17m}	&\ETAdag{727759}&\cpu{16455.24}	& \ints{283756}	& \cpu{1285.06}	& \ints{21684} \\
+			% [JOB 7] progress: 634229/727759 brs (87.148190%), ETA 1h 16m 33s (elapsed 9h 36m 22s)
+			% total CPU time consumed: 10h 11s (36011250596 usecs) (core saturation: 12%)
+			& 3,1 &\ETAdag{26h 15m}	&\ETAdag{1451708}&\cpu{21153.64}& \ints{366030}	& \cpu{705.04}	& \ints{12707} \\
+			% [JOB 5] progress: 655856/1451708 brs (45.178208%), ETA 16h 14m 59s (elapsed 9h 59m 50s)
+			% total CPU time consumed: 10h 10s (36010204482 usecs) (core saturation: 12%)
 

### 2018-08-19 thesis/tm.tex
params -> K,N in tm tables

@@
 			&	&\multicolumn{3}{c||}{\bf Quicksand mode}&\multicolumn{4}{c}{{\bf Maximal state space mode} ({\tt -M})} \\
-		\bf buggy test	& \bf params&\cpu{\bf cpu (s)}&\wtm{\bf wall (s)}&\ints{\bf int's}&\cpu{\bf cpu (s)}&\wtm{\bf wall (s)}&\ints{\bf int's}& \ETA{\bf SS size (est.)} \\
+		\bf buggy test	& \bf K,N &\cpu{\bf cpu (s)}&\wtm{\bf wall (s)}&\ints{\bf int's}&\cpu{\bf cpu (s)}&\wtm{\bf wall (s)}&\ints{\bf int's}& \ETA{\bf SS size (est.)} \\
 		\hline
@@
 			  & \cpu{\bf cpu (s)} & \ints{\bf SS size} \\
-			\bf test & \bf params
+			\bf test & \bf K,N
 			  & \em (or \ETAdag{\bf \em est.}) & \em (or \ETAdag{\bf \em est.})
@@
 			& & \cpu{\bf cpu (s)} & \ints{\bf SS size} \\
-			\bf test & \bf params & \em (or \ETAdag{\bf \em est.}) & \em (or \ETAdag{\bf \em est.}) \\
+			\bf test & \bf K,N  & \em (or \ETAdag{\bf \em est.})
+			                      & \em (or \ETAdag{\bf \em est.}) \\
 			\hline
@@
 			& & \cpu{\bf cpu (s)} & \ints{\bf SS size} \\
-			\bf test & \bf params & \em (or \ETAdag{\bf \em ETA}) & \em (or \ETAdag{\bf \em est.}) \\
+			\bf test & \bf K,N & \em (or \ETAdag{\bf \em ETA}) & \em (or \ETAdag{\bf \em est.}) \\
 			\hline

### 2018-08-19 thesis/education.tex
add photo of ze studence box plot; rough draft for now; colors need work etc
education prose revise

@@
 \footnote{PSU's version is called {\tt psu-setup.sh};
-in this section, unless otherwise noted, {\tt p2-setup.sh} will refer to both.
+in this section {\tt p2-setup.sh} refers to both unless otherwise noted.
 }.
@@
 		for Landslide depending on the name of the test.
-		For example, for {\tt paradise\_lost} Landslide will not preempt in a function named {\tt critical\_section()},
+		For example, for {\tt paradise\_\allowbreak{}lost} Landslide will not preempt in a function named {\tt critical\_section()},
 		which the test case uses to protect an internal counter used to detect the bug;
@@
 		In future work this could be improved as annotations to be placed inside the test case code itself.
-	\item Landslide finds ad-hoc synchronization patterns which students often open-code,
-		rather than using the prescribed synchronization API,
-		such as {\tt while (!flag) yield();} or {\tt while (xchg(...)) continue;},
+	\item Landslide finds ad-hoc synchronization patterns,
+		such as {\tt while (!flag) yield()} or {\tt while (xchg(...)) continue},
+		which students often open-code rather than using the prescribed synchronization API,
 		and treats them as synchronization points as described in \sect{\ref{sec:landslide-blocking}}.
@@
 		which would ordinarily be classified as infinite loop bugs,
-		and reports them with a suggestive message ({\tt undesirable\_loop\_html()} in {\tt landslide.c})
+		and reports them with a suggestive message ({\tt undesirable\_\allowbreak{}loop\_html()} in {\tt landslide.c})
 		referring the student to the appropriate lecture material
@@
 
-\subsubsection{Test case list}
+%\subsubsection{Test case list}
 
@@
 % TODO
+
+Figure~\ref{Fig:photo-of-ze-studence} shows some resulce
+
+\begin{figure}[t]
+	\includegraphics[width=\textwidth]{photo-of-ze-studence.pdf}
+	\caption{sadfasdfa}
+	\label{fig:photo-of-ze-studence}
+\end{figure}
 
@@
 \subsubsection{Response data}
+
+% probably just better to state this in one line of prose
+%\begin{table}[h]
+%	\begin{center}
+%		\begin{tabular}{r|crcrc|r}
+%			& \multicolumn{5}{c|}{\bf Did survey?} & \\
+%			\bf Semester & & \bf yes & & \bf no & & \bf Total \\
+%			\hline
+%			CMU F'17	& &  7 & &  7 & & 14 \\
+%			U. Chicago F'17	& &  1 & &  3 & &  4 \\
+%			CMU S'18	& &  5 & &  5 & & 10 \\
+%			PSU S'18	& & 11 & & 27 & & 38 \\
+%		\end{tabular}
+%	\end{center}
+%	\caption{Survey response rate among students who used Landslide. (The survey was first introduced in F'17.)}
+%	\label{tab:photo-of-ze-surveys}
+%\end{table}
 
@@
 among all P2-submitting groups.
-Reaching over half the class could alone be seen as a major step in mitigating said selection bias.
+Reaching over half the class could already be seen as a major step in mitigating said selection bias.
 
@@
 the racial diversity was roughly representative of the class at large,
-and the proportion of women was in fact somewhat higher than the overall gender ratio.
-Any more scientific analysis of such statistics was deemed beyond the study's scope for now.
+and the proportion of women I perceived was in fact somewhat higher than the overall gender ratio.
+More scientific analysis of such statistics was deemed beyond the study's scope for now.
 

### 2018-08-19 thesis/intro.tex
education prose revise

@@
 
+% TODO: god.........
 Modern computer architectures have turned to increasing CPU core count, rather than clock speed, to improve processing power \cite{mooreslaw}.
@@
 The impersonal ``the programmer'' will be referred to as she/her
-to distinguish her from the thesis audience concerned with the underlying research,
+to highlight her role as the intended user, separate from the underlying research,
+%to distinguish her from the thesis audience concerned with the underlying research,
 and also to challenge readers' unconscious bias about gender in computer science.

### 2018-08-20 thesis/education.tex
impact on grades (photo of ze studence)

@@
 I evaluated Landslide's impact on student submission quality in two ways.
-First, I analyzed the overall P2 grades of CMU 15-410 students between Landslide users and non-users,
-to see if Landslide helped them submit overall better implementations,
-as well as their later overall P3 (kernel project) grades,
-to see if Landslide helped them internalize any concurrency debugging skills
-that would help write better kernels even without Landslide's aid.
-% This experiment produced inconclusive or negative results. % say this?
-%
+First, I analyzed CMU 15-410 students' overall project grades between Landslide users and non-users
+to see if Landslide helped them submit overall better implementations.
 Second, I studied
@@
 
-% TODO
-
-Figure~\ref{Fig:photo-of-ze-studence} shows some resulce
+Figure~\ref{fig:photo-of-ze-studence} shows the distribution of project grades in CMU 15-410
+between Fall 2013 and Spring 2018,
+grouped by whether or not students used Landslide during P2.
+%Landslide users are, of course, the experimental group.
+The control group is further distinguished by whether Landslide was offered and the student delined,
+or whether the study had not yet begun that semester.
+Intuition suggests students in the former group would be more likely
+to be struggling too much to have free time to volunteer in the first place,
+so any comparison between them and Landslide users is vulnerable to selection bias.
+The latter control group, in which students did not have a choice, mitigates such bias,
+although itself may be vulnerable to other confounding factors such as grading criteria varying across semesters.
+%
+In addition to P2, I also show the subsequent P3 (kernel project) grades,
+likewise broken down by who used Landslide previously during P2%
+\footnote{The P3 distributions are slightly smaller than those from P2,
+as some students dropped the class in between.
+A few new project groups also tend to appear during P3 as partnerships break up or re-form,
+but as these cannot be meaningfully classified as Landslide users or non-users,
+I omit them here.
+}.
+Should Landslide use be correlated with higher grades on the later, more difficult concurrency project,
+one explanation could be that Landslide helped students
+internalize new concurrency programming and/or debugging skills
+that would help them write better kernels even without Landslide's aid.
+%
+
+% TODO: contingency table for drop rate? but.... need to get studence names to make sure they didn't just solbo, or w/e
+
+% colors for these graphs in case i need them again later
+% #720000
+% #FF8080
+% #804700
+% #FFC680
+% #807700
+% #FFF780
+% #004111
+% #ABFFC0
+% #002680
+% #ABC3FF
+% #3B0144
+% #F4ABFF
+\begin{figure}[p]
+	\begin{center}
+	\begin{tabular}{c}
+		\includegraphics[width=0.96\textwidth]{photo-of-ze-studence.pdf}
+		\\
+		(a) Visualized as a box plot.
+		\\
+		\\
+		\includegraphics[width=0.96\textwidth]{photo-of-ze-studence-cdf.pdf}
+		\\
+		(b) Visualized as a CDF.
+	\end{tabular}
+	\end{center}
+	\caption{Distribution of project grades among Landslide users and non-users.
+	Study participants (S'15 to S'18) are the experimental group;
+	non-participants (S'15 to S'18)
+	and students from semesters predating the study (F'13-F14) are the control groups.
+	Compare Landslide users to non-users for within-semester differences,
+	or users to pre-study students to mitigate selection bias.}
+	\label{fig:photo-of-ze-studence}
+\end{figure}
+
+%We see in Figure~\ref{fig:photo-of-ze-studence} that, indeed,
+{\bf Results.}
+Overall,
+Landslide users did indeed receive
+slightly better grades overall on their P2 submissions than non-users from the same semesters,
+although comparing to the second control group lessens that difference somewhat.
+The difference is also less pronounced in both comparisons among P3 grades
+(unsurprisingly, as learning lasting lessons about concurrency should be harder than fixing bugs case-by-case),
+although the experimental group still maintains a tiny lead.
+
+{\bf Statistical significance.}
+Table~\ref{tab:table-of-ze-studence} presents more detailed statistics from these six distributions.
+To assess statistical significance,
+% discrete == I used "version 2" of the AD metric that R outputs.
+I use the $k$-sample discrete Anderson-Darling test \cite{anderson-darling}
+provided by the R statistical programming environment \cite{r-lang,r-ksamples},
+comparing each control group to its respective Landslide users group.
+Anderson-Darling uses a weighted sum-of-squares distance metric to compare two CDFs,
+testing if their samples are likely to arise from the same underlying, though unspecified, distribution%
+\footnote{
+I choose Anderson-Darling over the simpler Kolmogorov-Smirnov,
+which computes only the maximum instantaneous distance between CDFs,
+for its better sensitivity both to repeated deviations and to tail differences \cite{beware-kolmogorov-smirnov}.
+}.
+I deem the difference in grade distributions significant when $p < 0.05$%
+% FIXME: is this legit?
+% should you be using the combined/grouped AD test instead?
+\footnote{I choose not to correct for multiplicity among these 4 comparsons (see next section)
+because each one is testing a different hypothesis.
+},
+although as in all $p$-value calculations,
+this rejects only the null hypothesis,
+not alternative hypotheses (such as ``more skilled students were more likely to sign up to begin with'').
+
+Ultimately, only the same-semester comparisons were significant
+However, the distribution difference being smaller in P3 than in P2
+% TODO: is there a good way to treat this more formally?
+further suggests, albeit informally,
+that the larger impact in P2 is not attributable {\em only} to selection bias.
+% let me have this exclamation mark. please
+That aside, ``Landslide helps students submit better P2s'' is the next most likely hypothesis I can think of!
+
+% P2 vs opt out
+% ad.test(c(42,50,51,55,55,55,56,57,57,58,62,63,65,65,65,66,66,67,67,67,67,68,68,68,69,69,69,70,70,70,70,71,71,71,71,71,72,73,73,73,73,73,74,74,74,74,74,75,75,75,75,75,76,76,76,76,77,77,77,77,78,78,78,78,79,79,79,79,79,79,80,80,80,80,81,81,81,82,82,82,82,83,84,84,84,84,85,85,86,86,86,87,87,87,88,89,90,91,92,93,93,93,95),c(38,42,47,50,52,53,55,55,56,56,57,57,58,58,58,59,60,60,61,62,62,63,65,66,67,68,68,68,69,70,70,70,71,71,71,71,72,72,72,73,73,73,74,74,76,76,76,77,77,78,78,78,78,78,79,79,79,80,82,82,83,83,83,84,84,84,85,86,86,88,90,95,96,96))
+%               AD  T.AD  asympt. P-value
+% version 1: 2.611 2.139          0.04263
+% version 2: 2.760 2.334          0.03576
+
+% P2 vs pre study
+% ad.test(c(42,50,51,55,55,55,56,57,57,58,62,63,65,65,65,66,66,67,67,67,67,68,68,68,69,69,69,70,70,70,70,71,71,71,71,71,72,73,73,73,73,73,74,74,74,74,74,75,75,75,75,75,76,76,76,76,77,77,77,77,78,78,78,78,79,79,79,79,79,79,80,80,80,80,81,81,81,82,82,82,82,83,84,84,84,84,85,85,86,86,86,87,87,87,88,89,90,91,92,93,93,93,95),c(43,51,56,58,59,59,60,60,60,60,61,61,61,61,61,62,63,65,65,67,68,69,69,69,69,69,70,70,71,71,72,72,72,73,73,73,73,73,74,74,74,74,75,75,75,75,76,76,77,77,77,77,78,78,78,78,79,79,79,80,80,80,80,80,80,81,81,82,82,82,82,83,83,83,83,85,85,85,86,86,86,88,88,90,92,93))
+%                AD    T.AD  asympt. P-value
+% version 1: 0.6248 -0.4977           0.6275
+% version 2: 0.5790 -0.5581           0.6706
+
+% P3 vs opt out
+% ad.test(c(65,68,72,72,73,75,75,75,75,75,76,76,76,76,77,77,77,78,78,78,78,78,78,79,79,80,80,80,80,80,80,80,81,82,82,82,82,82,82,83,83,83,83,83,83,84,84,84,84,85,85,85,85,85,85,85,85,85,85,86,86,86,86,86,86,86,86,86,87,87,87,88,88,88,88,88,89,89,89,89,90,90,90,90,90,91,91,91,91,92,92,93,93,93,94,95,95,95),c(45,58,66,70,72,72,72,72,74,74,74,75,75,75,77,77,77,78,78,78,79,79,79,79,79,79,80,80,81,81,81,81,81,82,82,83,83,83,83,83,84,84,85,85,86,86,86,86,86,86,87,87,89,89,89,89,90,91,91,91,93,93,93))
+%               AD  T.AD  asympt. P-value
+% version 1: 2.645 2.186          0.04085
+% version 2: 2.800 2.398          0.03376
+
+% P3 vs pre-study
+% ad.test(c(65,68,72,72,73,75,75,75,75,75,76,76,76,76,77,77,77,78,78,78,78,78,78,79,79,80,80,80,80,80,80,80,81,82,82,82,82,82,82,83,83,83,83,83,83,84,84,84,84,85,85,85,85,85,85,85,85,85,85,86,86,86,86,86,86,86,86,86,87,87,87,88,88,88,88,88,89,89,89,89,90,90,90,90,90,91,91,91,91,92,92,93,93,93,94,95,95,95),c(60,62,63,64,65,68,69,72,72,74,75,75,75,75,76,76,77,77,77,77,78,78,78,78,79,80,80,80,81,81,81,81,81,81,81,81,81,82,82,82,82,82,83,83,83,84,84,84,84,84,84,84,85,86,86,87,87,88,88,88,88,88,88,88,89,89,89,89,89,90,90,90,91,92,92,92,93,94,95))
+%               AD   T.AD  asympt. P-value
+% version 1: 1.552 0.7322           0.1633
+% version 2: 1.640 0.8473           0.1454
 
 \begin{figure}[t]
-	\includegraphics[width=\textwidth]{photo-of-ze-studence.pdf}
-	\caption{sadfasdfa}
-	\label{fig:photo-of-ze-studence}
+	\begin{center}
+		\small
+	\begin{tabular}{cr|r||r|r|r|r||cc}
+					&	&	&	&	& \multicolumn{2}{c}{Significance}
+		\multicolumn{2}{c|}{Population}
+					& $N$	& min	& mean	& max	& stddev & $p$ vs users	& cutoff\\
+		\hline
+			& users		& 103	& 42	& 74.8	& 95	& 10.3	& --	& --	\\
+		P2	& non-users	& 74	& 38	& 70.9	& 96	& 12.5	& 0.036	& 0.05	\\
+			& pre-study	& 86	& 43	& 73.7	& 93	& 9.7	& 0.671	& 0.05	\\
+			\hline
+			& users		& 98	& 65	& 83.6	& 95	& 6.2	& --	& --	\\
+		P3	& non-users	& 63	& 45	& 80.7	& 93	& 8.2	& 0.034	& 0.05	\\
+			& pre-study	& 79	& 60	& 81.6	& 95	& 7.6	& 0.145	& 0.05	\\
+	\end{tabular}
+	\end{center}
+	\caption{Detailed statistics from student grade distributions.}
+	\label{tab:table-of-ze-studence}
 \end{figure}
@@
 Table~\ref{tab:eval-common-bugs} presents the results.
-% two-tailed btw, the better (more conservative) way,
-% tbh one tailed might even make sense here, but it doesn't change anything wrt the cutoffs anyway, so
-The $p$ values are calculated in R \cite{r-lang} using Fisher's exact test \cite{fishers-exact-test},
-treating each row as an independent 2x2 contingency table,
-and dividing the significance cutoff by the number of bugs to conservatively account for multiplicity \cite{xkcd-jellybeans}.
 
@@
 
+Each of the four bugs studied is described in detail as follows.
+
+\begin{enumerate}
+\item
 {\bf Thread exit/join.}
@@
 
+\item
 {\bf Mutex.}
@@
 
+\item
 {\bf Paraguay.}
@@
 
+\item
 {\bf R/W-lock downgrade.}
@@
 (which were assessed by the TAs as a result of their manual inspection, rather than mine).
+\end{enumerate}
+
+{\bf Statistical significance.}
+% two-tailed btw, the better (more conservative) way,
+% tbh one tailed might even make sense here, but it doesn't change anything wrt the cutoffs anyway, so
+The $p$ values
+in Table~\ref{tab:eval-common-bugs}
+are calculated in R \cite{r-lang} using Fisher's exact test \cite{fishers-exact-test},
+treating each row as an independent 2x2 contingency table.
+I divide 0.05, the standard significance cutoff,
+by the number of bugs to conservatively account for multiplicity \cite{xkcd-jellybeans}
+(not that it matters).
+% TODO write more
 

### 2018-08-20 thesis/education.tex
oops
some impact-on-grades revisions

@@
 Table~\ref{tab:table-of-ze-studence} presents more detailed statistics from these six distributions.
-To assess statistical significance,
+To assess whether the differences are significant,
 % discrete == I used "version 2" of the AD metric that R outputs.
@@
 not alternative hypotheses (such as ``more skilled students were more likely to sign up to begin with'').
-
-Ultimately, only the same-semester comparisons were significant
+%
+Ultimately, only the same-semester comparisons were significant.
 However, the distribution difference being smaller in P3 than in P2
@@
 
+% TODO: should i report the T.AD test statistic? i don't even know what it means lol but it's what the p value like, applies to..
 \begin{figure}[t]
@@
 	\begin{tabular}{cr|r||r|r|r|r||cc}
-					&	&	&	&	& \multicolumn{2}{c}{Significance}
-		\multicolumn{2}{c|}{Population}
-					& $N$	& min	& mean	& max	& stddev & $p$ vs users	& cutoff\\
+		\multicolumn{3}{c||}{Population} & \multicolumn{4}{c||}{Distribution} & \multicolumn{2}{c}{Significance} \\
+		project	& group		& $N$	& min	& mean	& max	& stddev & $p$ vs users	& cutoff\\
 		\hline
@@
 but making a direct comparison to stress testing, the prior state of the art, is not straightforward.
-Little, if any,
 %and not statistically significant,
-immediate improvement in students' project grades was measured
+Modest immediate improvement in students' project grades was observed,
 (\sect{\ref{sec:education-eval-grades}}),
-but impact on grades alone is a very narrow measure of pedagogical value.
+although statistical significance was lost when attempting to account for selection bias;
+and impact on grades alone is a very narrow measure of pedagogical value anyway.
 Landslide's debugging power was also found to be statistically significant

### 2018-08-20 thesis/education.tex
discuss significance for submitted bug counts

@@
 % should you be using the combined/grouped AD test instead?
-\footnote{I choose not to correct for multiplicity among these 4 comparsons (see next section)
-because each one is testing a different hypothesis.
+\footnote{I choose not to correct for multiplicity among these 4 comparsons like I do in the next section
+because each tests a different hypothesis.
 },
@@
 that the larger impact in P2 is not attributable {\em only} to selection bias.
+% idk about this
+%Though ``P3 grading criteria results in a tighter bell curve than P2'' is another potential hypothesis,
 % let me have this exclamation mark. please
@@
 before any problem could be detected.
-% TODO: mark which of the tej bugs were deterministic vs race
 The Landslide-friendly test {\tt thr\_exit\_join} is most likely to expose this bug.
@@
 %
-This bug is especially remarkable in that not a single student who did not use Landslide these semesters
-submitted a correct implementation.
 
@@
 by the number of bugs to conservatively account for multiplicity \cite{xkcd-jellybeans}
-(not that it matters).
-% TODO write more
+(not that it matters with these $p$ values).
+
+The latter three bugs' occurences are thoroughly uncorrelated with Landslide use,
+the {\tt thr\_exit()} bug standing alone with extremely high significance:
+not a single student who did not use Landslide these semesters
+submitted a correct implementation.
+This difference is easily explainable:
+the class-provided unit tests already do a good enough job catching the other three bugs
+that students are able to find and fix them before submission regardless of what testing tool they used.
+(The high correct-to-buggy ratio corroborates this.)
+Nevertheless, that does not mean Landslide is pointless for these bugs;
+it may well have helped the students reproduce them more reliably and/or diagnose them faster.
+This is merely a null result for submission quality, not necessarily a negative one,
+and the next section will attempt to evaluate such quality-of-life improvements instead.
+
+On the other hand, because the {\tt thr\_exit()} bug typically manifests as a use-after-free,
+the official tests must stress the thread library until the memory in question gets re-allocated
+in just the right interleaving pattern to cause corruption that leads to a visible crash.
+Landslide, on the other hand, detects this error immediately upon access (\sect{\ref{sec:landslide-valgrind-mode}}),
+with no need for such complex corruption conditions.
+Among the 11 Landslide users who submitted correct {\tt thr\_exit()}s in this regard,
+it is difficult to say whether the heap checking alone, together with unit or stress tests,
+would have been sufficient,
+or whether heap checking and model checking were both necessary in concert to help them.
+To address this doubt,
+I simulated who among the buggy {\tt thr\_exit()} implementations
+would have found and fixed the bug
+given a heap-checking unit test
+by checking whether Landslide finds their bug ``deterministically''
+% TODO - more specific section reference - quicksand det/nondet drpps eval
+(i.e., on the first interleaving, similarly to \sect{\ref{sec:quicksand-eval}}).
+So doing,
+the 11-10-0-16 distribution becomes
+NN-MM-PP-QQ, % TODO
+with a new $p$ value of
+0.ppp, % TODO
+%% which one, will it be??? :-O
+% still a significant difference.
+% still positively correlated, but no longer significant.
+whose significance is a MYSTERYYYY.
 

### 2018-08-22 thesis/education.tex
study the exit uaf bug further, put good section refs in educ intro

@@
 meaning the only students who could benefit were the ones already succeeding enough to have such free time.
-% TODO: sect ref
 Since then, I have extended Landslide with a fully-automatic instrumentation process
-for Pebbles thread libraries (P2s) and Pintos kernels
+for Pebbles thread libraries (P2s) (\sect{\ref{sec:education-pebbles-instrumentation}})
+and Pintos kernels (\sect{\ref{sec:education-pintos-instrumentation}})
 to improve its accessibility.
@@
 wherein students get to use Landslide to find and diagnose their own bugs during the semester.
-% TODO: put section refs
-At CMU, I recorded logs and code snapshots as students used Landslide during P2.
-At CMU and PSU, I surveyed students on their experience after submitting their Landslide-debugged P2s.
+At CMU, I analyzed logs and code snapshots taken as students used Landslide during P2
+(\sect{\ref{sec:education-eval-bugfinding}}),
+as well as the grades ultimately assigned after students who either did or didn't use Landslide submitted their projects
+(\sect{\ref{sec:education-eval-grades}}).
+At CMU and PSU, I surveyed students on their experience after submitting their Landslide-debugged P2s
+(\sect{\ref{sec:education-eval-survey}}).
 At U. Chicago, I collaborated with a TA to check submitted Pintos kernels,
-then returned any resulting bug reports to students and likewise surveyed them on the quality of the diagnostic output.
+then returned any resulting bug reports to students
+(\sect{\ref{sec:education-eval-bugfinding}})
+and likewise surveyed them on the quality of the diagnostic output
+(\sect{\ref{sec:education-eval-survey}}).
 
@@
 that would help them write better kernels even without Landslide's aid.
-%
-
-% TODO: contingency table for drop rate? but.... need to get studence names to make sure they didn't just solbo, or w/e
 
@@
 		\\
-		(b) Visualized as a CDF.
+		(b) Visualized as an EDF.
 	\end{tabular}
@@
 comparing each control group to its respective Landslide users group.
-Anderson-Darling uses a weighted sum-of-squares distance metric to compare two CDFs,
+Anderson-Darling uses a weighted sum-of-squares distance metric to compare two empirical distribution functions (EDFs),
 testing if their samples are likely to arise from the same underlying, though unspecified, distribution%
@@
 I choose Anderson-Darling over the simpler Kolmogorov-Smirnov,
-which computes only the maximum instantaneous distance between CDFs,
+which computes only the maximum instantaneous distance between EDFs,
 for its better sensitivity both to repeated deviations and to tail differences \cite{beware-kolmogorov-smirnov}.
@@
 further suggests, albeit informally,
-that the larger impact in P2 is not attributable {\em only} to selection bias.
-% idk about this
-%Though ``P3 grading criteria results in a tighter bell curve than P2'' is another potential hypothesis,
-% let me have this exclamation mark. please
-That aside, ``Landslide helps students submit better P2s'' is the next most likely hypothesis I can think of!
+that the larger impact in P2 is not attributable {\em only} to selection bias,
+or else a similar difference should have been observed in the P3 distributions.
+Another possible explanation could simply be that the P3 grading critera result in less grade variance overall,
+but this pattern is at least consistent with the ``Landslide helps students submit better P2s'' hypothesis.
 
@@
 (guided, as well, by my own intuition as a former TA).
-I also counted students who used Landslide in general but whose usage snapshots showed they did not run the test case in question as non-users.
+While checking for a given bug's presence by manual inspection,
+I blinded myself to whether each group had been a Landslide user or not.
+After unblinding, I then also classified students who used Landslide in general,
+but whose usage snapshots showed they did not run the test case in question,
+as non-users.
 Table~\ref{tab:eval-common-bugs} presents the results.
@@
 or whether heap checking and model checking were both necessary in concert to help them.
-To address this doubt,
-I simulated who among the buggy {\tt thr\_exit()} implementations
-would have found and fixed the bug
-given a heap-checking unit test
-by checking whether Landslide finds their bug ``deterministically''
+To isolate the impact of model checking,
+I simulated the students having access to a stand-alone heap checker
+by checking only the first thread interleaving with Landslide
 % TODO - more specific section reference - quicksand det/nondet drpps eval
-(i.e., on the first interleaving, similarly to \sect{\ref{sec:quicksand-eval}}).
-So doing,
+(similarly to \sect{\ref{sec:quicksand-eval}})
+and assuming the students would find and fix any such ``deterministic'' use-after-free bugs before the deadline.
+Re-classifying these into the ``correct'' group,
 the 11-10-0-16 distribution becomes
-NN-MM-PP-QQ, % TODO
-with a new $p$ value of
-0.ppp, % TODO
-%% which one, will it be??? :-O
-% still a significant difference.
-% still positively correlated, but no longer significant.
-whose significance is a MYSTERYYYY.
+% fisher.test(rbind(c(11,10),c(3,13)))$p.value -->  0.0475343
+11-10-3-13,
+with a new $p$ value of 0.048:
+still positively correlated, but no longer significant under the multiplicity-corrected cutoff.
 
@@
 %and not statistically significant,
-Modest immediate improvement in students' project grades was observed,
+Immediate improvement in students' project grades was observed,
 (\sect{\ref{sec:education-eval-grades}}),
@@
 Landslide's debugging power was also found to be statistically significant
-for the {\tt thr\_exit\_join} bug in particular, even after adjusting for multiplicity.
+for the {\tt thr\_exit\_join} bug in particular.
 Students responded overwhelmingly positively in the survey (\sect{\ref{sec:education-eval-survey}}),
@@
 meaning I might not hear as much negative feedback as I should.
-Table~\ref{tab:TODO} % TODO -- make a table or graph showing correlation between bugfinding and who responded on the survey
-is my attempt to dispel this possibility.
-% TODO
+% trying to isolate this bias would be pretty fraught...
 % possible flaws in table: students who found bugs might have had either a good (bug!) or bad (false positive) experience
-% could studence who didn't find bugs also have had a good experience? (think of how...)
+% could studence who didn't find bugs also have had a good experience, like enjoying the verification guarantee
+% think of how to say it presentably...
 

### 2018-08-22 thesis/education.tex
bugs found at u chicago, and associated discussion

@@
 	\item The {\tt landslide} wrapper script logs the time and command-line options of invocation
-		and captures a snapshot of the student code and results of the test and saves them to AFS after each run.
+		and captures a snapshot of the student code and results of the test and saves them to AFS
+		(CMU's network file system)
+		after each run.
 		% HURDLE_VIOLATION -- no, this is for p3 (context switcher) features only
@@
 into the student's kernel code.
-The following tricks serve to make sure the annotations apply consistently
-to (almost) all variations of code that students commonly submit:
+The following tricks were developed after trial-and-error on student submissions from the same semester,
+and serve to make sure the annotations apply consistently
+to (almost) all variations of commonly-submitted code.
 
@@
 		and if so renames its use in {\tt is\_runqueue()} (described above) correspondingly.
+	\item Detects if the student has renamed the {\tt cur} (currently running thread) variable
+		in the context switcher, and if so renames it back.
 \end{itemize}
@@
 		Userspace process {\tt exec()}s a child process, which immediately exits, and {\tt wait()}s on it.
+	\item {\tt wait-twice}:
+		Unmodified from the basecode's version, for {\em userprog}.
+		Slightly more complicated version of {\tt wait-simple},
+		intended to expose failure-path bugs if the former finds no common-case ones.
 	\item {\tt priority-donate-multiple}:
@@
 
-The (unpatched versions of) the first three tests are available at
+The (unpatched versions of) the first four tests are available at
 \url{https://github.com/Berkeley-CS162/group0/tree/master/pintos/src/tests}.
-The fourth test is available at
+The fifth test is available at
 \url{https://github.com/bblum/landslide/blob/master/pebsim/pintos/priority-donate-multiple.c}.
@@
 % TODO update for psu
-for the thread library study.
+for the thread library study%
+\footnote{
+Participation at CMU was determined by receiving any automatic usage snapshots;
+participation at PSU was determined, for lack of snapshots, by sign-up form submissions, which might over-count slightly.
+}.
 % and PSU
@@
 \label{sec:education-eval-bugfinding}
+
+Firstly, I sought to prove Landslide does as advertised:
+finds concurrency bugs of severity, subtlety, and difficulty
+consistent with the lessons an advanced operating systems class should hope to teach its students,
+and provides diagnostic output that helps the student to understand and solve them.
+% without necessarily spoiling the answer to the puzzle ... not scientifically evaluatable,
+% but at least the trace format is de0u-approved.
+
+At CMU, I configured Landslide to save a usage snapshot every time a student ran Landslide,
+including command-line options issued, the current version of their project code,
+a log of Landslide's output,
+and the preemption traces for any bugs found.
+These I analyzed by hand to determine how many distinct bugs were reported
+(as multiple preemption traces may refer to the same bug; inspecting their code if necessary),
+how many were deterministic
+(i.e., Landslide reported them on the very first interleaving tested, with no need for artificial preemptions)
+and how many were concurrency bugs,
+whether any were false positives,
+and whether the student was able to fix them thereafter
+(determined if a subsequent snapshot showed them re-running the test and verifying the bug's absence;
+an approximate measure at best, but far easier to implement than checking all the bugs by hand).
+
+At U. Chicago,
+course staff ran Landslide on student submissions behind-the-scenes
+after each project deadline had passed,
+and returned its preemption traces to any study volunteers.
+I collaborated with the course staff
+to confirm whether each trace represented a real bug and not a false positive
+before distributing them to the students.
+We decided on this procedure to optimize for most benefit to the students rather than for scientific rigor,
+due to the small sample size (4 volunteering student groups among 21 submitted projects).
+
+As PSU's version of the study was planned on relatively short notice,
+and without immediate access to a shared, yet confidential, academic file system (such as CMU's AFS),
+I was unable to examine its students' objective usage data or bug reports for this section.
+Results from PSU are presented instead in \sect{\ref{sec:education-eval-survey}}.
+
+\subsubsection{CMU}
+\label{sec:education-eval-bugs-cmu}
+
+% cmu - fps marked in the reuslts files? is this gonna be an accurate count?
+\subsubsection{University of Chicago}
+\label{sec:education-eval-bugs-uc}
+
+Among the 4 groups volunteering to receive Landslide bug reports,
+Landslide found 7 distinct bugs across the two projects tested,
+which I confirmed by manual inspection of the students' code.
+3 among these were deterministic; 4 were concurrency bugs.
+
+\begin{enumerate}
+	\item {\tt priority-donate-multiple} found 2 deterministic bugs and 1 concurrency bug in {\tt threads} projects.
+		It found the targeted priority-leak bug
+		described in \sect{\ref{sec:education-pintos-tests}} in one group,
+		and exposed a {\tt NULL} pointer dereference in another
+		when attempting to donate priority to an already-exited thread.
+		The latter bug was observed deterministically.
+		A third group neglected to implement priority donation during {\tt lock\_release()} at all,
+		which was found deterministically.
+	\item {\tt wait-simple} found 1 deterministic bug and 3 concurrency bugs in {\tt userprog} projects.
+		It found a deadlock in two groups' implementations,
+		% for one of these groups 2 traces were emitted, same stack trace (i think) just w diff pps
+		wherein {\tt process\_wait()}'s synchronization assumed
+		it would always be called before {\tt process\_exit()}.
+		In both cases the former's {\tt cond\_wait()} call would block forever if reordered after the latter.
+		This test also found several use-after-frees for one of those two groups,
+		whose {\tt process\_exit()} didn't guard against a parent process exiting and deallocating its memory first;
+		this root cause manifested in heap errors in 3 separate locations.
+		Lastly, the heap checking alone found a deterministic use-after-free in a third group's {\tt exec()}.
+\end{enumerate}
+
+In cases where Landslide emitted multiple preemption traces for the same bug,
+whether manifested the same way through different combinations of preemption points
+or with different stack traces entirely arising from the same underlying cause,
+course staff sent them all to the student,
+including a disclaimer along the lines of ``some of these may indicate the same bug''.
+%
+The other 3 tests ({\tt priority-sema}, {\tt alarm-simultaneous}, and {\tt wait-twice}) found no bugs among the 4 groups.
+One of the 4 groups had no bugs found among any of the 5 tests.
+
+%%%% i think it'd be an irb violation even to publish this at all
+%%%% and the potential false positive issue makes it not scientific enough anyway
+% Course staff also ran Landslide with wait-simple on submissions from even non-volunteers;
+% in total finding bugs in 11 of 18 groups' implementations.
+% However, ... couldn't check for false positives because of non-volunteers ...
+
+% XXX: did I actually fix this in the annotation scripts, or did I just hackhackhack to suppress it?
+% my email to kevin just says, "I suppressed the false positive bug from wait-simple in GROUPNAME and got a real bug next".
+Finally, one false positive was found while testing {\tt wait-simple}.
+Landslide mistakenly reported that {\tt free()} had reentered {\tt malloc()},
+which would be a potential heap corruption bug,
+due to a technical discrepancy between when Pebbles and Pintos kernels take and release their heap allocation lock
+with regard to the rest of the {\tt malloc()} API.
+After discarding the bug report and suppressing this false positive,
+Landslide moved on to find the true deterministic use-after-free described above.
+
+Overall, I consider all of the 7 bugs to be ``severe'':
+they are all either correctness violations (priority mis-donation)
+or stability issues (crash, deadlock, or data corruption).
+In the upcoming survey section, although only one of these groups did the survey,
+they enthusiastically reported finding the bug reports very helpful (\sect{\ref{sec:education-survey-response-data}}).
+Despite the small sample size, I consider this
+a positive result that Landslide can be a useful concurrency programming aid
+even at universities with different class projects than CMU's 15-410.
+As for 1 false positive among 8 total reports,
+it is difficult to decide what rate of mistakes is acceptable \cite{racerx},
+but the upcoming survey results (\sect{\ref{sec:education-eval-survey-falsepositives}})
+suggest that students are generally at least patient enough
+to ask for help and make progress when technical support is available.
 
@@
 \subsubsection{Response data}
+\label{sec:education-survey-response-data}
 
@@
 the answers thereto were ``15-20 minutes'', ``Agree'', and ``Agree'', respectively.
+Also, this respondent's answer of 6 on the ``How many bugs'' question
+appears to indicate the total number of preemption trace files course staff sent them;
+%despite the attached ``some of these may indicate the same bug'' disclaimer;
+upon further investigation,
+the 6 traces seem to represent 2 distinct bugs among them (\sect{\ref{sec:education-eval-bugs-uc}}).
 
@@
 \subsubsection{False positives}
+\label{sec:education-eval-survey-falsepositives}
 
@@
 		Simultaneous Landslides can clobber certain auto-generated header and/or temporary files,
-		leading to undefined behaviour.
+		scrambling its instrumentation and leading to chaotic behaviour.
 		I introduced a guard against this in commit 977e8fb (Simics repository) and e49b5df (Bochs repository).
@@
 		would or would not be reasonable for Landslide to record and report.
+	\item {\em For each bug Landslide found in your code, how trivial or severe do you feel it was?}
+		This would help get a sense of how the students regarded Landslide
+		on a spectrum between annoying style check or as a life-saver,
+		and potentially suggest options to make Landslide suppress certain types of bug reports.
+		For example, it currently reports spin-wait loops in {\tt mutex\_lock()} as bugs
+		with a special message referring students to relevant lecture slides,
+		but it's possible refusing to test any code beyond until that bug was fixed
+		might have made Landslide less useful overall.
 \end{enumerate}
@@
 on account of the fixes from this time (\sect{\ref{sec:education-pintos-instrumentation}}).
+In its original shape
+(before the F'17 semester,
+having only enough instrumentation necessary for the Pintoses used in \sect{\ref{sec:quicksand-eval}}'s experiments),
+Landslide was already able to automatically instrument 18 out of the 21 {\tt threads} project submissions at U. Chicago.
+I was able to quickly deploy a fix to make the annotation scripts handle the other 3 cases,
+although such technical support is not something any TA would be able to do.
+
 In its current shape I would recommend it for TA use grading,

### 2018-08-22 thesis/education.tex
small revisions to bugfinding

@@
 consistent with the lessons an advanced operating systems class should hope to teach its students,
-and provides diagnostic output that helps the student to understand and solve them.
+and provides diagnostic output that helps students understand and solve them.
 % without necessarily spoiling the answer to the puzzle ... not scientifically evaluatable,
@@
 after each project deadline had passed,
-and returned its preemption traces to any study volunteers.
+and returned its preemption traces to the study volunteers.
 I collaborated with the course staff
@@
 before distributing them to the students.
-We decided on this procedure to optimize for most benefit to the students rather than for scientific rigor,
-due to the small sample size (4 volunteering student groups among 21 submitted projects).
+On account of the small sample size (4 volunteering groups),
+we decided on this approach,
+rather than to study how students would cope with false positives themselves,
+to optimize for student happiness over scientific rigor.
 
@@
 Finally, one false positive was found while testing {\tt wait-simple}.
-Landslide mistakenly reported that {\tt free()} had reentered {\tt malloc()},
-which would be a potential heap corruption bug,
+Landslide mistakenly reported that {\tt free()} had reentered {\tt malloc()}
+(normally a heap corruption bug)
 due to a technical discrepancy between when Pebbles and Pintos kernels take and release their heap allocation lock

### 2018-08-23 thesis/education.tex
future edu pebbles

@@
 
-% TODO: para for pebbles; discuss with dave
-
+%{\bf Pebbles.}
+Now that I'm done collecting student usage data to publish as research results,
+and no longer bound by the IRB's requirement that Landslide not be used to assign grades
+lest it be seen as coercion to participate,
+Professor Eckhardt and I have been discussing options to deploy it as an official part of 15-410's curriculum.
+This section has already clearly shown students are capable of debugging with it on their own time,
+and I believe it well-automated enough to supplement Fritz (the existing stress testing infrastructure)
+in the class's grading process as well.
+TAs could also, at their option, use Landslide by hand to confirm any bugs encountered during manual inspection
+and/or write new Landslide-friendly tests to expose bugs not yet targeted by the 6 tests offered here.
+Over the study's seven semesters,
+I believe the stability of Landslide's instrumentation process has improved enough
+to require little to no ongoing technical support anymore,
+although Landslide-specific office hours may still prove helpful.
+I am also willing to continue giving the guest lecture as long as proximity and the curriculum allow.
+Future problems to address include grading bias
+(i.e., students submitting blindly-hacked code that just barely passes Landslide,
+even if not necessarily correct,
+thereby gaming the autograder),
+and improving usability to reach even the most struggling students
+(i.e., those past 42\% who submitted P2s without participating in the study).
+
+%{\bf Pintos.}
 Regarding non-research use in Pintos classes,
@@
 although such technical support is not something any TA would be able to do.
-
+%
 In its current shape I would recommend it for TA use grading,

### 2018-08-24 thesis/education.tex
cmu bug finding tallies (at last)

@@
 
-% cmu - fps marked in the reuslts files? is this gonna be an accurate count?
+In the recruiting lectures for every semester after the first,
+I included tallies of Landslide's bug-finding achievements to date
+as an extra advertisement to tempt students.
+These included each of the measures I counted by hand as described above,
+as well as aggregate totals of how many groups participated,
+how many groups found bugs,
+how many found at least one concurrency bug,
+and how many were able to fix and subsequently verify any or all thereof.
+Among the concurrency bugs, a wide variety of types were found:
+deadlocks, use-after-frees, segfaults, infinite loops, assertion failure, and unit test failure.
+Table~\ref{tab:cmubugs} shows these statistics for each semester.
+% may also be an overestimate if they only hacked around the bug, but that's less likely
+Considering the method for deciding whether students fixed each bug, described above,
+note that the tallies thereof may be undercounting
+on account of the possibility some students may have skipped the verification step despite truly fixing them.
+
+\begin{table}[t]
+	% TODO
+	\begin{center}
+		\small
+	\begin{tabular}{r||r|r|r|r|r|r|r||r}
+			& \bf S'15 & \bf F'15 & \bf S'16 & \bf F'16 & \bf S'17 & \bf F'17 & \bf S'18 & \bf Total \\
+			\hline
+			%			S15	F15	S16	F16	S17	F17	S18
+			\bf Participating groups& 8	& 22	& 18	& 12	& 19	& 14	& 10	& 103	\\
+			% nb not counting FPs
+			Groups w/any bugs found	& 5	& 21	& 12	& 7	& 14	& 10	& 6	& 75	\\
+			Groups w/any bugs fixed	& 4	& 18	& 9	& 6	& 9	& 7	& 5	& 58	\\
+			Groups w/all (>0) bugs fixed
+						& 2	& 11	& 3	& 3	& 7	& 5	& 4	& 35	\\
+			\hline
+			% the 2 for f15 were pretty messy; i had to go back and get em by hand by grepping for NOTE
+			% but this semester also had a LOT of seemingly true race NOTE mx lock inf loop,
+			% so i counted only the det ones, of which there were 2 (both also in "fixed" category)
+			% otherwise 1 race recursive mx and 6 race mutex loops and 1 race cvar loop were left as races.
+
+			% this 1 for f17 was salvaged from results-raw; the grupo who had it it was their only bug
+			% and it was an invl heap that i think came from multiple landslides
+			% (and i hadn't listed them as "fixing" it; they were listed as det 1 race 0 fixed 0)
+			% so i'm -1ing for "det bugs found" and "groups w any bugs found" to compensate
+			Known false positives	& 1	& 2	& 6	& 5	& 3	& 1	& 0	& 18 	\\
+			\hline
+			Deteterministic bugs found ($\le$)
+						& 5	& 20	& 18	& 18	& 30	& 15	& 9	& 115	\\
+			Concurrency bugs found	& 5	& 56	& 19	& 15	& 13	& 9	& 5	& 122	\\
+			\bf Total bugs found	& 10	& 76	& 37	& 33	& 43	& 24	& 14	& 237	\\
+			\hline
+			Deteterministic bugs fixed ($\ge$/$\le$)
+						& 3	& 19	& 9	& 15	& 22	& 12	& 7	& 87	\\
+			Concurrency bugs fixed ($\ge$)
+						& 1	& 38	& 11	& 10	& 12	& 4	& 4	& 80	\\
+			\bf Total bugs fixed	& 4	& 57	& 20	& 25	& 34	& 16	& 11	& 167	\\
+	\end{tabular}
+	\end{center}
+	\caption{Landslide's bug-finding performance across all semesters of the CMU 15-410 study.
+	$\le$ marks possible overcounts on account of unidentified false positives;
+	$\ge$ marks possible undercounts on account of students not necessarily re-running to verify bugfixes.}
+	\label{tab:cmubugs}
+\end{table}
+
+This table also shows the number of false positives, as self-reported by Landslide
+as described in \sect{\ref{sec:education-pebbles-instrumentation}}.
+% nah where actually doing this, where making it hapen.
+%which I started counting since S'16.
+Please note that this approach %to counting false positives
+is limited by Landslide's ability to classify them as ``suspicious'', even if not definitely bugs.
+These represent technical obstacles that either
+prevented Landslide from being able to analyze the synchronization in play
+(e.g., recursive {\tt mutex\_lock()} invocations)
+or were deemed too pedagogically important to allow the student to proceed without fixing
+(e.g., busy spin-wait synchronization loops)%
+\footnote{
+Landslide was also configured to issue a ``bug report''
+if {\tt MAGIC\_BREAK}, a Simics debugging trap, was ever invoked,
+with specific instructions to remove it;
+I consider these closer to compilation errors than bug reports and so did not count them
+even among false positives.
+}.
+Other false positives arising from bugs in Landslide itself
+required more individual effort to confirm;
+to classify these I relied on students to help report them
+during the Landslide clinics and survey,
+and I report on them in \sect{\ref{sec:education-eval-survey-falsepositives}}.
+Nevertheless,
+I am aware of no case in which Landslide would output a false positive {\em nondeterministic} bug%
+% this is really shaky
+% not too shaky for me to feel uncomfortable publishing it
+% but too shaky for me to feel comfortable leaving it unsaid too...
+\footnote{I did observe Landslide's ``suspicious loop'' message
+for mutex or condition variable internal synchronization
+on subsequent interleavings,
+but at this point the line between style warning and true bug blurs,
+so I count these as true concurrency bugs for simplicity's sake.}
+% i *think*.......
+all unexpected false positives encountered to date were output on the first interleaving tested.
+Though unscientific,
+this provides some reassurance that Table~\ref{tab:cmubugs}'s count of concurrency bugs is accurate,
+even if the deterministic tally may overcount slightly.
+
+Overall, Landslide helped students find and fix a lot of bugs.
+Among the 103 participating groups across all semesters,
+roughly three-quarters received bug reports,
+slightly more than half were able to verify their fix for at least one,
+and one-third overall were able to fix and verify {\em all} such bugs.
+Even avoiding the deterministic bug series for their possible overcounting,
+it's fair to conclude that Landslide caused at least two-thirds
+of all concurrency bugs it found to get fixed before project handin.
+Although lacking a control experiment to make a more scientific comparison of these tallies,
+I tentatively conclude the students were well rewarded for their time.
+The questions of whether it was a {\em better} use of time than conventional stress testing,
+or whether the bug-finding resulted in submissions that course staff also, independently, judged better than before,
+are beyond the scope of just tallying bug reports and will instead be addressed in the upcoming sections.
+
 \subsubsection{University of Chicago}

### 2018-08-24 thesis/education.tex
assorted revisions across education; prose and table headings

@@
 		Slightly more complicated version of {\tt wait-simple},
-		intended to expose failure-path bugs if the former finds no common-case ones.
+		intended to expose failure-path bugs if the former finds no easier ones.
 	\item {\tt priority-donate-multiple}:
@@
 		% but better than latex's output looking stupid
-		\begin{tabular}{c|crcrc|r}
-			& \multicolumn{5}{c|}{\bf used Landslide?} & \\
-			\bf semester & & \bf yes & & \bf no & & \bf total \\
+		\begin{tabular}{r|crcrc|r}
+			& \multicolumn{5}{c|}{\bf Used Landslide?} & \\
+			\bf Semester & & \bf yes & & \bf no & & \bf Total \\
 			\hline
@@
 			CMU S'18	& &  10	& &   8	& &  18 \\
-			% NB. CMU's participance are counted by how many ppl i got at least one usage tarball for,
-			% while PSU's are by how many sign-up forms i got.
-			% the latter may be overcounting, because a student may sign up and then never use it;
-			% but i don't have tarballs for PSU so this is the best you're getting.
-			% TODO
-			% PSU s18	& &  38
 			\hline
-			\bf total	& & 103	& &  74	& & 177 \\
+			U. Chicago F'17	& &   4	& &  17	& &  21	\\ % 21 taken from number of submitted 'threads', idk abt userprog
+			% TODO when you adjust this, adjust the % of participance down in the discussion para
+			PSU S'18	& &  38	& &   ?	& &   ?	\\
+			\hline
+			\bf CMU total	& & 103	& &  74	& & 177 \\
+			\bf Non-CMU total
+					& &  42	& & 17+?& &  21+? \\
+			\bf Total	& & 145 & & 91+?& & 191+? \\
 		\end{tabular}
 	\end{center}
-	\caption{Student participation rates across semesters,
-		among students who submitted thread library implementations.}
+	\caption{Participation rates across semesters,
+	among students who submitted thread libraries (CMU, PSU) or kernels (U. Chicago).}
 	\label{tab:photo-of-ze-studence}
@@
 (i.e., Landslide reported them on the very first interleaving tested, with no need for artificial preemptions)
-and how many were concurrency bugs,
+how many were concurrency bugs,
 whether any were false positives,
@@
 	\begin{tabular}{cr|r||r|r|r|r||cc}
-		\multicolumn{3}{c||}{Population} & \multicolumn{4}{c||}{Distribution} & \multicolumn{2}{c}{Significance} \\
-		project	& group		& $N$	& min	& mean	& max	& stddev & $p$ vs users	& cutoff\\
+		\multicolumn{3}{c||}{\bf Population} & \multicolumn{4}{c||}{\bf Distribution} & \multicolumn{2}{c}{\bf Significance} \\
+		\bf project & \bf group & $N$ & \bf min & \bf mean & \bf max & \bf stddev & \bf $p$ vs users & \bf cutoff\\
 		\hline
@@
 	\begin{tabular}{r||cc|cc|c|cc}
-		& \multicolumn{2}{c|}{Landslide users} & \multicolumn{2}{c|}{Non-users} & & \multicolumn{2}{c}{Significance} \\
-		Bug name & \#correct & \#buggy & \#correct & \#buggy & $\Delta$\%correct & $p$ & cutoff \\
+		& \multicolumn{2}{c|}{\bf Landslide users} & \multicolumn{2}{c|}{\bf Non-users} & & \multicolumn{2}{c}{\bf Significance} \\
+		\bf Bug name & \bf \#correct & \bf \#buggy & \bf \#correct & \bf \#buggy & \bf $\Delta$\%correct & $p$ & \bf cutoff \\
 		\hline
@@
 		\hline
-		Total     & 58	& 15	& 52	& 22	& $+09.2\%$	& 0.26 & 0.05 \\
+		\bf Total & 58	& 15	& 52	& 22	& $+09.2\%$	& 0.26 & 0.05 \\
 	\end{tabular}
@@
 students fixed their own bugs {\em and} were happy about it;
-students were able to ask for help rather than be deterred by inevitable technical difficulties;
+students were able to ask for help rather than be deterred by inevitable technical difficulties as far as I know;
 students provided intelligent feedback suggesting they truly understood the debugging process;
@@
 
-Some students (between 0 and 2 per semester)
+Some students (around 0 to 2 per semester if memory serves)
 emailed me during P3
@@
 and if they were still interested,
-guided them through the annotation process on a case-by-case basis.
+guided them through the annotations on a case-by-case basis.
 This process was not included in the IRB-approved study protocol,
@@
 although Landslide-specific office hours may still prove helpful.
-I am also willing to continue giving the guest lecture as long as proximity and the curriculum allow.
+I am also willing to continue giving the guest lecture as long as proximity and curriculum allow.
 Future problems to address include grading bias

### 2018-08-30 thesis/thesis.tex
write abt the surbeys
do more tm experimence
finish DPOR friendly example and rewrite sleep sets description to follow
start writing acks

@@
 \usepackage{amsthm}
+\usepackage{mdwlist}
 \usepackage[numbers,sort]{natbib}
@@
 \begin{acknowledgments}
-% things to thank
-% stimhack
-% csd bgames
-% undergrad frenz
-% new frenz
-% 410 ta community
-% emarhavil
-%
-% \\
+Even as I built Landslide by myself,
+implementation is of course barely the first step of scientific contribution.
+I would honor here the many people who supported me during these 7 years and made this thesis possible.
 
-I thank my dog, Louie.
+\subsubsection{Thesis}
+
+% garth,
+
+Professor David Eckhardt
+provided unending guidance on education, writing, life, you name it,
+helped revise each semester's iteration of the recruiting lecture with a keen eye for the student mind,
+and supplied constant enthusiasm and reassurance that Landslide was a boon to 15-410 that kept me going.
+
+% jiri,
+% .. orient me in the world of research ..
+
+% these 3 names ar ethe ones without which the thesis flat out could not have happened
+
+Many others have collaborated with me directly on the research itself,
+whom I thank here in approximate chapter order.
+Michael J. Sullivan and Nathaniel Filardo were Landslide's first users besides myself back in its prototype days.
+Joshua Wise, Josh Zimmerman, and Glenn Willen helped revise my explanation of DPOR in Chapter~\ref{chap:background}.
+Professor Brandon Lucia provided valuable comments and revisions to Chapter~\ref{chap:quicksand}'s content
+that ultimately led to its publication at OOPSLA 2016.
+Michael J. Sullivan double-checked the proofs therein.
+Professor Timothy Zhu from Penn State University
+and Professor Haryadi Gunawi and TA Kevin Zhao from the University of Chicago
+graciously allowed me to use their students as guinea pigs
+and helped greatly to improve Landslide's stability and robustness for use beyond CMU's walls,
+greatly enriching Chapter~\ref{chap:education}'s evaluation.
+David Blum and Skye Toor provided invaluable advice on statistical analysis
+that helped me evaluate the student grade distributions therein.
+Mario Dehesa-Azuara generously contributed the transactional memory data structures
+that make up about half of Chapter~\ref{chap:tm}'s evaluation suite.
+
+\subsubsection{Academia}
+
+Without SIGBOVIK's unique brand of humorous, self-aware, and {\em definitely legitimate} research,
+I probably would not have been interested in grad school in the first place,
+or perhaps would have grown too frustrated for lack of a cathartic outlet for conference paper woes.
+I thank all who've represented Harry Bovik over the years,
+from the conference's original founders,
+to every paper author,
+to all future organizers who'll keep those ``mainstream'' conferences on their toes well into the future.
+
+% [ ] 410 ta community
+% [ ] emarhavil
+% [ ] pdl consortium industry ppl who offered good advices
+% [ ] anyone who ever said how cool it was that i helped students for my research
+% [ ] adam and co. for GPI, if you feel like it
+
+% blurry line
+
+I am grateful to the ThursDz Council,
+my group of now-former grad student, Shadyside-bound (in spirit if not in person) restaurant-hoppers,
+for both friendship and mentorship over the years.
+In particular,
+%in alphabetical for lack of a better order,
+Chris Martens,
+Jason Reed,
+Jim McCann,
+Rob Simmons,
+Tom Murphy 7,
+and
+William Lovas:
+you all have been role models to me each in your own way perhaps more than you know.
+
+\subsubsection{Life}
+
+% [ ] csd bgames
+% [ ] undergrad frenz
+% [ ] new frenz
+% [ ] family, incl louie
+
+Lastly,
+the Android: Netrunner community has been such an important part of my life
+for nearly as long as I've been a grad student.
+I thank in particular the other Stimhack moderators,
+the members of {\sf \#rainbow-coalition},
+and the local Pittsburgh crew
+for their support and encouragement
+and friendships very real and deep, even when forged only in text,
+%\footnote{and Slack emojis},
+that I hope will well outlive the silly card game itself.
+%It'd be an injustice to list only a few names here and omit others, but you know who you are.
+
 \end{acknowledgments}
@@
 \definecolor{salmon}{RGB}{208,89,74} % * 7/8
+\definecolor{goldish}{RGB}{128,96,32}
 % eval data
@@
 % syntax hilighting
-\newcommand\flow[1]{\hilight{brown}{#1}\xspace}
+\newcommand\flow[1]{\hilight{brownish}{#1}\xspace}
 \newcommand\const[1]{\hilight{brickred}{#1}\xspace}
 \newcommand\call[1]{\hilight{pinkish}{#1}}
-\newcommand\ctype[1]{\hilight{olivegreen}{#1}\xspace}
+\newcommand\ctype[1]{\hilight{lime}{#1}\xspace}
+\newcommand\ccomment[1]{\hilight{darkcyan}{#1}\xspace}
 

### 2018-08-30 thesis/related.tex
start relwork - sct tools

@@
 
-% things to cite what you haven't alerady got cited somewhere
+Had I a dollar for every programmer before me who thought to ``solve'' concurrency with a perfect debugging tool,
+well,
+it probably would not be quite enough to retire on.
+Even so,
+this field is built of the contributions of many a brilliant mind
+trying to carve out a presentable space in an overall impossible problem,
+each making their own tradeoffs along the way.
+While the previous chapters cited prior work as necessary in background discussions, algorithm descriptions, and so on,
+this chapter aims to comprehensively tour the field,
+orienting the reader's understanding of Landslide in the space of said tradeoffs.
 
-% dejafu - haskell mcing
-% relacy - smth for go?
+\section{Systematic Concurrency Testing}
+
+Equal partners in concurrency testing are the practical and the theoretical:
+the former meaning tool implementations that target specific problem domains and help users as best one can,
+and the latter meaning algorithmic advances to bring ever-larger state spaces within the realm of computational feasibility.
+I discuss my most closely related works split in two sections accordingly.
+
+\subsubsection{Tools}
+
+Systematic concurrency testing dates back to Verisoft \cite{verisoft},
+the 1997 tool which first attempted to exhaustively explore the possibile ways to interleave threads.
+Since then, researchers have built many tools along the same lines to test many kinds of programs.
+One of the best-known SCTs is Microsoft Research's CHESS \cite{chess},
+a checker for userspace C++ programs which preempts on synchronization APIs by default,
+supporting compiler instrumentation to preempt on memory accesses as well,
+and which pioneered the ICB search strategy discussed below.
+
+Many checkers exist which target programs written for various different types of concurrent environments.
+MaceMC \cite{macemc}, MoDist \cite{modist}, SAMC \cite{samc}, ETA \cite{dbug-retreat}, and Concuerror \cite{concuerror},
+focus on distributed systems, where concurrent events are limited to message-passing and may span across multiple machines.
+R4 \cite{r4} and EventRacer \cite{eventracer} check event-driven concurrent programs typical in mobile applications.
+dBug \cite{dbug-ssv}, another CMU original similar to CHESS,
+integrates with Parrot \cite{parrot}, a determinizing runtime scheduler,
+to approach state space reduction from a different angle by limiting which thread interleavings are possible to begin with.
+
+Other checkers target specific programming languages' concurrency models and/or thread communication APIs.
+Relacy \cite{relacy} is a header-only C++ SCT library for verifying synchronization primitives.
+It instruments the existing C++ atomics API,
+although requires custom annotations for other types of memory accesses,
+and notably includes relaxed memory nondeterminism in its concurrency model.
+Inspect \cite{inspect} uses a static alias analysis to instrument and preempt all memory accesses to potentially-shared data.
+SPIN \cite{spin} defines the PROMELA domain-specific language
+to specialize in verifying synchronization primitives such as RCU \cite{rcu}.
+%
+D\'{e}j\`{a} Fu \cite{dejafu} is a SCT
+for the Haskell language,
+whose strong type system guarantees that thread communication be confined to trusted, type-safe APIs.
+%to a trusted API that implements internal synchronization to preserve type safety.
+%Supporting both abstraction reduction and STM,
+D\'{e}j\`{a} Fu instruments these interfaces (STM among them)
+to check for deadlocks or nondeterministic behaviour in general,
+which may arise despite the static guarantee of no data races.
+The Rust language \cite{rust-language}
+provides a more C++-like type system that also
+%, which I contributed to personally,
+achieves the same guarantee, although I know of no existing SCT for it yet.
+
+If I might indulge by listing Landslide in its own related work section \cite{this-thesis},
+I would distinguish it by its ability to find data-race preemption points via dynamic memory tracing,
+rather than relying on user annotations or imprecise compiler instrumentation
+as other tools do.
+Compared to all other tools I know of,
+it implements a wider range of state space explosion coping techniques,
+some theoretical and some heuristic,
+to help the user receive meaningful results as promptly as possible.
+Its choice of a familiar pthread-like C threading API makes it suitable for inexpert users,
+and its recent extension to HTM shows that it can support more modern concurrency patterns as well.
+
+\subsubsection{Algorithms}
+
+% transdpor - older than optimal dpor but still part of history?
+% http://mir.cs.illinois.edu/marinov/publications/TasharofiETAL12TransDPOR.pdf
+
+% effective stateless model checking for c/c++ concurrency - popl 18
+
+% Surveying concurrency bug detectors based on types of detected bugs - idk what this is, from an email
+
+%%%% non-smc testing tools -- even worth talking abt?
+
 % jepsen-io/jepsen
+
+\section{Data race analysis}
+
+\section{Educational debugging}
+
+Willgrind \cite{willgrind}
+
+\section{Transactional memory}
 
@@
 
-\section{Stateless Model Checking}
+% transactions in relaxed mem architecutres - popl 18
 
-D\'{e}j\`{a} Fu \cite{dejafu}
-
-\section{Education}
-
-Willgrind \cite{willgrind}
+% pldi 2018
+% The Semantics of Transactions and Weak Memory in x86, Power, ARM, and C++

### 2018-08-31 thesis/related.tex
relwork dpor algs

@@
 it probably would not be quite enough to retire on.
-Even so,
+At any rate,
 this field is built of the contributions of many a brilliant mind
@@
 Inspect \cite{inspect} uses a static alias analysis to instrument and preempt all memory accesses to potentially-shared data.
-SPIN \cite{spin} defines the PROMELA domain-specific language
-to specialize in verifying synchronization primitives such as RCU \cite{rcu}.
+SPIN \cite{spin} tests algorithms defined in the PROMELA domain-specific language,
+instruments every memory access, and
+specializes in verifying synchronization primitives such as RCU \cite{rcu}.
 %
@@
 %Supporting both abstraction reduction and STM,
-D\'{e}j\`{a} Fu instruments these interfaces (STM among them)
+It instruments these interfaces (STM among them)
 to check for deadlocks or nondeterministic behaviour in general,
-which may arise despite the static guarantee of no data races.
+which either may arise despite the static no-data-race guarantee.
 The Rust language \cite{rust-language}
@@
 If I might indulge by listing Landslide in its own related work section \cite{this-thesis},
-I would distinguish it by its ability to find data-race preemption points via dynamic memory tracing,
+I would distinguish it by its ability to find shared memory preemption points via dynamic tracing,
 rather than relying on user annotations or imprecise compiler instrumentation
@@
 Compared to all other tools I know of,
-it implements a wider range of state space explosion coping techniques,
+it implements a wider range of exponential explosion coping techniques,
 some theoretical and some heuristic,
+some inherited and some novel,
 to help the user receive meaningful results as promptly as possible.
-Its choice of a familiar pthread-like C threading API makes it suitable for inexpert users,
-and its recent extension to HTM shows that it can support more modern concurrency patterns as well.
+Its choice of a familiar pthread-like synchronization API makes it suitable for inexpert users,
+and its recent extension to HTM adds support for more modern concurrency patterns as well.
+
+% TODO
+% Surveying concurrency bug detectors based on types of detected bugs - idk what this is, from an email
 
@@
 
-% transdpor - older than optimal dpor but still part of history?
-% http://mir.cs.illinois.edu/marinov/publications/TasharofiETAL12TransDPOR.pdf
+To date a number of techniques have been proposed to mitigate exponential explosion,
+the Sisyphean rock of SCT.
+The notion that some interleavings of concurrent threads could lead to indistinguishable program states and be therefore redundant,
+known as {\em partial order reduction} (POR),
+was first proposed in \cite{partial-model-checking}
+and explored in detail in \cite{partial-order-methods}.
+{\em Dynamic POR} (DPOR) was later developed in \cite{dpor},
+proposing to track communication events between threads on-the-fly (i.e., dynamically)
+rather than to rely on imprecise static alias analyses,
+and is now considered the baseline for all subsequent state space reduction approaches in SCT.
+That paper includes the {\em sleep sets} extension,
+which Landslide includes in its implementation.
+It is a {\em sound} reduction algorithm, meaning it will never fail to test a possible program behavior, despite skipping many execution sequences.
+\sect{\ref{sec:landslide-dpor}} provides a detailed walk-through of how DPOR works,
+as many of this thesis's contributions build directly upon it.
 
+DPOR has since been extended in several ways to achieve further reduction
+and to incorporate new concurrency models.
+Optimal DPOR \cite{optimal-dpor} extends sleep sets into the more expressive {\em wakeup trees},
+which provably tests exactly one interleaving from each equivalence class,
+i.e., the optimal possible reduction,
+at least under the memory independence definition of equivalence.
+Extending the equivalence relation itself to capture not just memory {\em address} conflicts
+but also the {\em values} read and written,
+SATCheck \cite{satcheck} and Maximal Causality Reduction (MCR) \cite{mcr}
+use an SMT solver \cite{z3} to identify additional pruning opportunities.
+Implementing wakeup trees or SMT-driven exploration in Landslide is left to future work.
+
+Several other recent advances extend DPOR to new concurrency models,
+beyond the shared-memory-threading model outlined in \sect{\ref{sec:landslide-dpor}}.
+TransDPOR \cite{transdpor} provides extra domain-specific reduction for message-passing actor programs
+by exploiting the fact that the dependency relation is transitive in the absence of shared state.
+The $R^4$ algorithm \cite{r4} (corresponding to the R4 checker mentioned above)
+extends DPOR to event-driven programs by separating the notion of enabled events from that of multiple threads.
+DPOR for TSO and PSO \cite{tsopso}
+extends the concurrency model to include store buffer reordering,
+which can expose bugs not even possible in the strong consistency model
+such as discussed in \sect{\ref{sec:tm-warpzone-relaxed}}.
+It also introduced a heuristic {\em buffer bounding} technique, analogous to ICB,
+to mitigate the corresponding increase in state space size.
+MCR was recently extended to support relaxed memory models likewise \cite{mcr-tsopso}.
+Along the same lines, this work's Chapter~\ref{chap:tm}
+extended DPOR's concurrency model for HTM,
+and proposed two new reduction strategies, one sound and one heuristic,
+to find new types of equivalences.
+% TODO
 % effective stateless model checking for c/c++ concurrency - popl 18
 
-% Surveying concurrency bug detectors based on types of detected bugs - idk what this is, from an email
+Of course, no matter how optimal a sound reduction, there will always be programs too large to test.
+To provide even partial results for state spaces that exceed the testing budget
+(whether as predicted by automatic estimation \cite{estimation} or by a human's wild guess),
+various heuristic exploration strategies have been proposed.
+Preemption Sealing \cite{sealing} allows programmers to manually exclude preemption points
+arising from trusted source code modules;
+Landslide implements this as the {\tt without\_function} command (\sect{\ref{sec:landslide-pps}}).
+Iterative Context Bounding (ICB) \cite{chess-icb} (\sect{\ref{sec:landslide-icb}})
+orders the search space by increasing number of preemptions in each branch,
+which is more likely to expose bugs sooner should they exist;
+BPOR \cite{bpor} extends DPOR to preserve soundness
+thereunder. % what a word
+Landslide implements ICB and BPOR for Chapter~\ref{chap:quicksand}'s control experiments,
+although does not yet incorporate it into this thesis's own contributions.
+Chapter~\ref{chap:quicksand}'s Quicksand algorithm is, effectively, another such heuristic search strategy,
+focusing on preemption point subsets rather than context switch bounding.
+Each of these approaches is compatible (and indeed, throughout this thesis used often in concert)
+with the sound reduction analyses listed above.
+% mentioning PCT would disrupt the flow here and it's a dumb technique anyway so i won't bother
 

### 2018-09-01 thesis/related.tex
relwork relaxed mem checkers

@@
 Other checkers target specific programming languages' concurrency models and/or thread communication APIs.
-Relacy \cite{relacy} is a header-only C++ SCT library for verifying synchronization primitives.
-It instruments the existing C++ atomics API,
-although requires custom annotations for other types of memory accesses,
-and notably includes relaxed memory nondeterminism in its concurrency model.
 Inspect \cite{inspect} uses a static alias analysis to instrument and preempt all memory accesses to potentially-shared data.
@@
 achieves the same guarantee, although I know of no existing SCT for it yet.
+
+The problem of relaxed memory nondeterminism alone has inspired the creation of several new SCTs in the past few years.
+Relacy \cite{relacy}, a header-only C++ SCT library for checking synchronization primitives,
+was the first to broach this field,
+although requires custom annotations for non-atomic memory accesses
+and
+%(according to later citing papers) % cdschecker
+does not fully model all possible relaxed memory behaviours.
+%designed for verifying synchronization primitives,
+CDSChecker \cite{cdschecker} extends DPOR with a {\em reads-from} relation
+to capture most of the C++11 memory model's new behaviours.
+Nidhugg \cite{nidhugg} is a SCT for TSO and PSO which instruments LLVM abstract assembly,
+% i don't actually know what the difference is #overlyhonestmethods
+although does not yet support the C++11 memory model.
+%% actually it doesn't -- they say they take the Source DPOR worse version. im confus?
+% and extends the Optimal DPOR algorithm (discussed below) to include store buffer nondeterminism.
+rInspect \cite{tsopso}
+%models TSO and PSO differently, using shadow threads, and
+offers further heuristic state space reduction using buffer bounding (described below).
+RCMC \cite{rcmc} models a ``repaired'' version of the C++11 memory model known as RC11 \cite{rc11},
+and professes to achieve the best state space reduction to date.
+These tools each use various heuristics to account for spin-wait loops,
+ranging from delay bounding \cite{bpor} to a rigid rewrite rule,
+and provide only limited support so far for read-modify-write atomics
+(at best, supporting them by introducing % oops
+some redundant exploration).
+%
+No relaxed-memory SCT has yet proposed a satisfactory model for the ``thin-air'' problem \cite{sully-thesis},
+which can cause state space cycles in a way not yet well-understood and remains future work.
+%and notably includes relaxed memory nondeterminism in its concurrency model.
+They also identify all data races
+% TODO check if ther'es not a better section reference
+(by the C++ definition rather than \sect{\ref{sec:quicksand-soundness}}'s)
+as bugs immediately, rather than checking them for benign or buggy outcomes.
+All of these tools are notably open-source -- an encouraging recent trend in the field.
 
@@
 DPOR for TSO and PSO \cite{tsopso}
-extends the concurrency model to include store buffer reordering,
+extends the concurrency model
+using {\em shadow threads}, which interleave with traditional threads to represent store buffer nondeterminism,
 which can expose bugs not even possible in the strong consistency model
@@
 to mitigate the corresponding increase in state space size.
+The same year, Nidhugg \cite{nidhugg} proposed a DPOR extension to account for TSO and PSO
+using {\em chronological traces}.
 MCR was recently extended to support relaxed memory models likewise \cite{mcr-tsopso}.
-Along the same lines, this work's Chapter~\ref{chap:tm}
-extended DPOR's concurrency model for HTM,
-and proposed two new reduction strategies, one sound and one heuristic,
-to find new types of equivalences.
-% TODO
-% effective stateless model checking for c/c++ concurrency - popl 18
+Just this year, RCMC \cite{rcmc} proposed to replace the interleaving model entirely with {\em execution graphs},
+offering further reduction still.
+Somewhat analogously for HTM, this work's Chapter~\ref{chap:tm}
+extended DPOR's concurrency model to include failure injection,
+% not "new" -- abstraction reduction is the same as always
+and proposed three reduction strategies, one sound and two heuristic,
+%to make some progress up the exponential mountain.
+to keep state spaces manageable.
 

### 2018-09-01 thesis/landslide.tex
relwork relaxed mem checkers
icb complexity discussion

@@
 \subsubsection{Example}
+\label{sec:landslide-dpor-example}
 
@@
 Ultimately, the second half of the state space is pruned symmetrically.
-In general, the number of ways to interleave two threads executing $N$ and $M$ events each is given by ${N+M \choose N}$;
+In general, the number of ways to interleave two threads executing $N$ and $M$ events each is given by ${N+M \choose N}$%
+\footnote{Generalizing to $K$ threads, and simplifying to the same number $N$ of events each,
+this formula becomes $\frac{n!k!}{n!^k}$.};
 in this case, the original state space's size was ${3+3 \choose 3} = 20$.
@@
 \begin{algorithm}[h]
-	\For{$n \in [0..\mathsf{max}_P(U)]$}{
-		\ForEach{$S \in U, P(S) \le n$}{
+	\For{$B \in [0..\mathsf{max}_P(U)]$}{
+		\ForEach{$S \in U, P(S) \le B$}{
 			Execute($S$)
@@
 %\begin{algorithm}[h]
-%	\For{$n \in [0..\infty]$}{
+%	\For{$B \in [0..\infty]$}{
 %		$\mathsf{skipped\_any} := \mathsf{false}$ \\
 %		\ForEach{$S \in U$}{
-%			\uIf{$P(S) \le n$}{
+%			\uIf{$P(S) \le B$}{
 %				Execute($S$)
@@
 {\tt ICB\_BLOCKED} does not count such switches towards the preemption count.
-Therefore, within a certain preemption bound $b$,
-interleavings with more than $b$ context switches may still be tested.
+Therefore, within a certain preemption bound $B$,
+interleavings with more than $B$ context switches may still be tested.
 }.
@@
 allowing exploration to continue until it finally observes all interleavings to have fewer preemptions than the bound.
+
+\subsubsection{Complexity}
+
+If the search is terminated early after reaching a predetermined fixed bound for $B$,
+ICB in principle reduces the state space from exponentially-sized%
+\footnote{Combinatorial, to be precise; see \sect{\ref{sec:landslide-dpor-example}}.}
+in both $K$, the number of threads,
+and $N$, the number of events,
+to still exponential in $K$ (typically small) but only polynomial in $N$ (typically large).
+Under a preemption bound of $B$, there are only $B+K$ opportunities for context switching%
+\footnote{This $K$ appears from the ``mandatory'' context switches at thread exit;
+more of which could also be introduced from blocking synchronization.},
+% "roughly" - i.e., this formulation allows choosing all B preemptions from the same thread,
+% which is only possible when K >= B i guess
+% it's an overapproximation but getting it more precise seems like it'd be way more complex,
+% and the overapproximation is good enough to make this point anyway
+% also an underapprox because this doesn't count interleavings with b<B preemptions but that's a constant factor
+so the corresponding state space size is at most ${KN \choose B}(B+K)!$.
+All $N$-related factors therein are bounded above by $N^B$.
+
+Prior work often recommends 2 for such a cutoff \cite{chess-icb,smc-empirical-study,dejafu},
+although \sect{\ref{sec:quicksand-eval}}'s larger dataset suggests 3 would be considerably more thorough.
+On the other hand, any finite such bound can provide only a heuristic verification guarantee anymore.
+Preserving the full formal verification, i.e., continuing iteration until $B = \mathsf{max}_P(U)$
+not only remains exponential in $N$,
+% again overapproximating because each previous iteration is combinatorially smaller than the next
+but also introduces a factor of $\mathsf{max}_P(U)$ repeated work.
+Future work could memoize already-tested interleavings so that each iteration of $B$ could test only
+those schedules with exactly $B$ preemptions,
+restoring the original $B$-independent (but still exponential) complexity.
 

### 2018-09-01 thesis/education.tex
final pre-send-to-committee revisions to education chapater

@@
 The (unpatched versions of) the first four tests are available at
-\url{https://github.com/Berkeley-CS162/group0/tree/master/pintos/src/tests}.
+\url{https://github.com/bblum/pintos} (a fork of the main Pintos basecode repository).
+%\url{https://github.com/Berkeley-CS162/group0/tree/master/pintos/src/tests}.
 The fifth test is available at
@@
 			% TODO when you adjust this, adjust the % of participance down in the discussion para
-			PSU S'18	& &  38	& &   ?	& &   ?	\\
+			PSU S'18	& &  38	& &  98	& & 136 \\
 			\hline
@@
 			\bf Non-CMU total
-					& &  42	& & 17+?& &  21+? \\
-			\bf Total	& & 145 & & 91+?& & 191+? \\
+					& &  42	& & 115	& & 157 \\
+			\bf Total	& & 145	& & 189	& & 327 \\
 		\end{tabular}
@@
 I included tallies of Landslide's bug-finding achievements to date
-as an extra advertisement to tempt students.
+as an extra advertisement to entice students to participate.
 These included each of the measures I counted by hand as described above,
@@
 % may also be an overestimate if they only hacked around the bug, but that's less likely
-Considering the method for deciding whether students fixed each bug, described above,
-note that the tallies thereof may be undercounting
-on account of the possibility some students may have skipped the verification step despite truly fixing them.
+Note that the tallies of bugs fixed may be undercounting
+on account of the possibility some students may have truly fixed them but skipped the verification step thereafter.
 
@@
 			Groups w/any bugs fixed	& 4	& 18	& 9	& 6	& 9	& 7	& 5	& 58	\\
-			Groups w/all (>0) bugs fixed
+			Groups w/all ($\ne$0) bugs fixed
 						& 2	& 11	& 3	& 3	& 7	& 5	& 4	& 35	\\
@@
 with specific instructions to remove it;
-I consider these closer to compilation errors than bug reports and so did not count them
+I consider these closer in spirit to compilation errors than bug reports and so did not count them
 even among false positives.
@@
 Nevertheless,
-I am aware of no case in which Landslide would output a false positive {\em nondeterministic} bug%
-% this is really shaky
-% not too shaky for me to feel uncomfortable publishing it
-% but too shaky for me to feel comfortable leaving it unsaid too...
-\footnote{I did observe Landslide's ``suspicious loop'' message
-for mutex or condition variable internal synchronization
-on subsequent interleavings,
-but at this point the line between style warning and true bug blurs,
-so I count these as true concurrency bugs for simplicity's sake.}
-% i *think*.......
+apart from the spin-wait synchronization loops
+(which can arise nondeterministically due to lock contention, but which Landslide already self-identifies),
+I am aware of no cases of false positive {\em nondeterministic} bugs;
 all unexpected false positives encountered to date were output on the first interleaving tested.
 Though unscientific,
-this provides some reassurance that Table~\ref{tab:cmubugs}'s count of concurrency bugs is accurate,
-even if the deterministic tally may overcount slightly.
+this provides some assurance that Table~\ref{tab:cmubugs}'s count of concurrency bugs is accurate,
+even if the deterministic tally may overcount.
 
@@
 %
-The other 3 tests ({\tt priority-sema}, {\tt alarm-simultaneous}, and {\tt wait-twice}) found no bugs among the 4 groups.
+The other 3 tests, {\tt priority-sema}, {\tt alarm-simultaneous}, and {\tt wait-twice},
+found no bugs among the 4 groups
+({\tt wait-twice} being run only on the group that passed {\tt wait-simple}).
 One of the 4 groups had no bugs found among any of the 5 tests.
@@
 Landslide mistakenly reported that {\tt free()} had reentered {\tt malloc()}
-(normally a heap corruption bug)
 due to a technical discrepancy between when Pebbles and Pintos kernels take and release their heap allocation lock
 with regard to the rest of the {\tt malloc()} API.
+This typically indicates a heap corruption bug, but in this case,
+{\tt malloc()} was preempted after it released the heap lock, which was perfectly safe.
 After discarding the bug report and suppressing this false positive,
@@
 As for 1 false positive among 8 total reports,
-it is difficult to decide what rate of mistakes is acceptable \cite{racerx},
+it is difficult to decide what rate of mistakes is acceptable when the user's patience is at stake \cite{racerx},
 but the upcoming survey results (\sect{\ref{sec:education-eval-survey-falsepositives}})
-suggest that students are generally at least patient enough
+suggest that students are generally at least willing
 to ask for help and make progress when technical support is available.
@@
 %Landslide users are, of course, the experimental group.
-The control group is further distinguished by whether Landslide was offered and the student delined,
+The control group is further distinguished by whether Landslide was offered and the student declined,
 or whether the study had not yet begun that semester.
@@
 as some students dropped the class in between.
-A few new project groups also tend to appear during P3 as partnerships break up or re-form,
+Some new project groups also tend to appear during P3 as students either solo or find new partners,
 but as these cannot be meaningfully classified as Landslide users or non-users,
@@
 this rejects only the null hypothesis,
-not alternative hypotheses (such as ``more skilled students were more likely to sign up to begin with'').
+not alternative hypotheses,
+such as ``more skilled students were more likely to sign up to begin with''.
 %
@@
 % TODO: is there a good way to treat this more formally?
+% TODO: run a combined A-D test for each of the two control pairs
 further suggests, albeit informally,
@@
 
+% combined test
+% > ad.test.combined(list(c(42,50,51,55,55,55,56,57,57,58,62,63,65,65,65,66,66,67,67,67,67,68,68,68,69,69,69,70,70,70,70,71,71,71,71,71,72,73,73,73,73,73,74,74,74,74,74,75,75,75,75,75,76,76,76,76,77,77,77,77,78,78,78,78,79,79,79,79,79,79,80,80,80,80,81,81,81,82,82,82,82,83,84,84,84,84,85,85,86,86,86,87,87,87,88,89,90,91,92,93,93,93,95),c(43,51,56,58,59,59,60,60,60,60,61,61,61,61,61,62,63,65,65,67,68,69,69,69,69,69,70,70,71,71,72,72,72,73,73,73,73,73,74,74,74,74,75,75,75,75,76,76,77,77,77,77,78,78,78,78,79,79,79,80,80,80,80,80,80,81,81,82,82,82,82,83,83,83,83,85,85,85,86,86,86,88,88,90,92,93)), list(c(65,68,72,72,73,75,75,75,75,75,76,76,76,76,77,77,77,78,78,78,78,78,78,79,79,80,80,80,80,80,80,80,81,82,82,82,82,82,82,83,83,83,83,83,83,84,84,84,84,85,85,85,85,85,85,85,85,85,85,86,86,86,86,86,86,86,86,86,87,87,87,88,88,88,88,88,89,89,89,89,90,90,90,90,90,91,91,91,91,92,92,93,93,93,94,95,95,95),c(60,62,63,64,65,68,69,72,72,74,75,75,75,75,76,76,77,77,77,77,78,78,78,78,79,80,80,80,81,81,81,81,81,81,81,81,81,82,82,82,82,82,83,83,83,84,84,84,84,84,84,84,85,86,86,87,87,88,88,88,88,88,88,88,89,89,89,89,89,90,90,90,91,92,92,92,93,94,95)))
+%            AD.comb T.comb  asympt. P-value
+% version 1:   2.176 0.1655           0.3324
+% version 2:   2.219 0.2055           0.3191
+%
+% > ad.test.combined(list(c(42,50,51,55,55,55,56,57,57,58,62,63,65,65,65,66,66,67,67,67,67,68,68,68,69,69,69,70,70,70,70,71,71,71,71,71,72,73,73,73,73,73,74,74,74,74,74,75,75,75,75,75,76,76,76,76,77,77,77,77,78,78,78,78,79,79,79,79,79,79,80,80,80,80,81,81,81,82,82,82,82,83,84,84,84,84,85,85,86,86,86,87,87,87,88,89,90,91,92,93,93,93,95),c(38,42,47,50,52,53,55,55,56,56,57,57,58,58,58,59,60,60,61,62,62,63,65,66,67,68,68,68,69,70,70,70,71,71,71,71,72,72,72,73,73,73,74,74,76,76,76,77,77,78,78,78,78,78,79,79,79,80,82,82,83,83,83,84,84,84,85,86,86,88,90,95,96,96)),list(c(65,68,72,72,73,75,75,75,75,75,76,76,76,76,77,77,77,78,78,78,78,78,78,79,79,80,80,80,80,80,80,80,81,82,82,82,82,82,82,83,83,83,83,83,83,84,84,84,84,85,85,85,85,85,85,85,85,85,85,86,86,86,86,86,86,86,86,86,87,87,87,88,88,88,88,88,89,89,89,89,90,90,90,90,90,91,91,91,91,92,92,93,93,93,94,95,95,95),c(45,58,66,70,72,72,72,72,74,74,74,75,75,75,77,77,77,78,78,78,79,79,79,79,79,79,80,80,81,81,81,81,81,82,82,83,83,83,83,83,84,84,85,85,86,86,86,86,86,86,87,87,89,89,89,89,90,91,91,91,93,93,93)))
+%            AD.comb T.comb  asympt. P-value
+% version 1:   5.256  3.058          0.01503
+% version 2:   5.560  3.343          0.01109
+
 \subsubsection{Common bugs}
@@
 Based on the results from \sect{\ref{sec:education-eval-bugfinding}},
-I selected 3 bugs to study in depth to ascertain
+I selected 4 bugs to study in depth to ascertain
 whether Landslide played an instrumental difference
-in helping the students ultimately submit respecitvely correct implementations.
+in helping the students ultimately submit respectively correct implementations.
 To avoid bias of picking too obscure and/or trivial bugs
@@
 I blinded myself to whether each group had been a Landslide user or not.
-After unblinding, I then also classified students who used Landslide in general,
+After unblinding, I then re-classified students who used Landslide in general,
 but whose usage snapshots showed they did not run the test case in question,
@@
 so comparing how many Landslide users and non-users submitted this bug in particular
-speaks more to the impact of Landslide's preemption traces
-in helping students diagnose a bug that a ``stress'' test could already find.
+would speak more to the impact of Landslide's preemption traces
+in helping to diagnose a bug that a ``stress'' test could already find.
 %
@@
 in just the right interleaving pattern to cause corruption that leads to a visible crash.
-Landslide, on the other hand, detects this error immediately upon access (\sect{\ref{sec:landslide-valgrind-mode}}),
-with no need for such complex corruption conditions.
+Landslide, meanwhile, detects this error immediately upon access (\sect{\ref{sec:landslide-valgrind-mode}}),
+with no need for complex corruption conditions.
 Among the 11 Landslide users who submitted correct {\tt thr\_exit()}s in this regard,
@@
 not to assert Landslide is completely bug-free now,
-but it at least grows more and more stable with each passing semester.
+but at least that it grows more and more stable with each passing semester.
 
@@
 3 students declined to answer this question;
-I reproduce the rest, paraphrasing for clarity and brevity,%
+I reproduce the rest here, paraphrasing for clarity and brevity%
 \footnote{Note that most students used the term ``race condition'' rather than ``concurrency bug'',
 as taught in CMU and PSU lecture material;
-I replaced these while paraphrasing in accordance with \sect{\ref{sec:glossary}}.}
-here.
+I replaced these while paraphrasing in accordance with \sect{\ref{sec:glossary}}.}.
 Answers from CMU:
@@
 
-\subsection{Other universities}
+\subsubsection{Other universities}
 
@@
 %and not statistically significant,
-Immediate improvement in students' project grades was observed,
+Immediate improvement in students' project grades was observed
 (\sect{\ref{sec:education-eval-grades}}),
@@
 
-Beyond participation, I anticipated several forms of bias affecting the survey responses.
-First of all, I suspected that students might feel compelled to be overly polite in their answers
+The survey may also be susceptible to several sources of bias beyond participation itself.
+I suspected students might feel inclined to be overly polite in their answers
 (whether consciously or subconsciously).
@@
 		This would help get a sense of how the students regarded Landslide
-		on a spectrum between annoying style check or as a life-saver,
+		on a spectrum between annoying style checker or a life-saver,
 		and potentially suggest options to make Landslide suppress certain types of bug reports.
@@
 I answered each by explaining that it would take more effort on their part,
-and if they were still interested,
+and then if they were still interested,
 guided them through the annotations on a case-by-case basis.
@@
 {\em Do you feel the way you think about testing, debugging, and program correctness
-has been influenced in any way by using Landslide?}.
+has been influenced in any way by using Landslide?}
+to evaluate its lasting impact on their understanding of concurrency.
 
@@
 
-I attribute the low participation rate of Pintos students to P2 students to two major factors:
+I attribute the comparatively low participation rate of Pintos students to two major factors:
 one, not incentivizing the students to directly improve their grades
@@
 %{\bf Pebbles.}
-Now that I'm done collecting student usage data to publish as research results,
-and no longer bound by the IRB's requirement that Landslide not be used to assign grades
+Now being done collecting student usage data to publish as research results,
+and no longer bound by the IRB's requirement that Landslide be isolated from the grading process
 lest it be seen as coercion to participate,
-Professor Eckhardt and I have been discussing options to deploy it as an official part of 15-410's curriculum.
+Professor Eckhardt and I have discussed options to deploy it as an official part of 15-410's curriculum.
 This section has already clearly shown students are capable of debugging with it on their own time,
@@
 and/or write new Landslide-friendly tests to expose bugs not yet targeted by the 6 tests offered here.
+
 Over the study's seven semesters,
@@
 although Landslide-specific office hours may still prove helpful.
-I am also willing to continue giving the guest lecture as long as proximity and curriculum allow.
+I am willing to continue giving the guest lecture as long as proximity and curriculum allow,
+although I also hope the documentation herein be enough to pass the mantle like any other piece of the course infrastructure.
 Future problems to address include grading bias
@@
 and improving usability to reach even the most struggling students
-(i.e., those past 42\% who submitted P2s without participating in the study).
+(i.e., that last 42\% who submitted P2s without participating in the study).
 
@@
 other, more readily automatic model checkers for user-space programs,
-such as dBug \cite{dbug-ssv},
+such as dBug \cite{dbug-ssv} or CHESS \cite{chess},
 could easily be used on other thread-library-like programming projects at any university.

### 2018-09-01 thesis/thesis.tex
minor intro updates
acknowledgemence updates

@@
 \usepackage{algpseudocode}
+\usepackage{microtype}
 
@@
 \begin{tabular}{c}
-{\em or, How I Learned to Stop Worrying and Love the Exponential Explosion}
+{\em or: How I Learned to Stop Worrying and Love the Exponential Explosion}
 \end{tabular}}
@@
 helped revise each semester's iteration of the recruiting lecture with a keen eye for the student mind,
+contributed substantially to the design of the survey,
 and supplied constant enthusiasm and reassurance that Landslide was a boon to 15-410 that kept me going.
+I can't express the depth of my gratitude to Dave strongly enough,
+even though three of the Landslide-friendly tests never made it upstream into the official P2 test suite.
 
@@
 Michael J. Sullivan and Nathaniel Filardo were Landslide's first users besides myself back in its prototype days.
-Joshua Wise, Josh Zimmerman, and Glenn Willen helped revise my explanation of DPOR in Chapter~\ref{chap:background}.
+Joshua Wise, Miriam Zimmerman, and Glenn Willen helped revise my explanation of DPOR in Chapter~\ref{chap:background}.
 Professor Brandon Lucia provided valuable comments and revisions to Chapter~\ref{chap:quicksand}'s content
@@
 or perhaps would have grown too frustrated for lack of a cathartic outlet for conference paper woes.
-I thank all who've represented Harry Bovik over the years,
+I thank all who've worn the mantle of Harry Bovik over the years,
 from the conference's original founders,
@@
 William Lovas:
-you all have been role models to me each in your own way perhaps more than you know.
+you each have been role models to me
+%each in your own way
+perhaps more than you know.
 
@@
 %\footnote{and Slack emojis},
-that I hope will well outlive the silly card game itself.
+that I hope will well outlive the card game that brought us together.
 %It'd be an injustice to list only a few names here and omit others, but you know who you are.

### 2018-09-01 thesis/tm.tex
more tm revisions and 10hr resulce

@@
 However, state-of-the-art advances in model checking algorithms
-still state their theorems and proofs in prose \cite{bpor,optimal-dpor,tsopso,satcheck,mcr,quicksand,cpp-model-checking},
+still state their theorems and proofs in prose \cite{bpor,optimal-dpor,tsopso,satcheck,mcr,quicksand,rcmc},
 so this section's proofs will regrettably do likewise,
@@
 thereby skipping over (i.e., pruning) reorderings of any other {\tt xbegin} results that would be independent.
+Like sleep sets, it also considers the preempted thread ``retry-set blocked''
+(like sleep-set blocking \cite{optimal-dpor}),
+and refuses to run it until the conflicting thread runs its transaction first,
+unless such would result in deadlock%
+\footnote{
+	Before the search ordering update to Landslide's normal sleep sets implementation
+	(\sect{\ref{sec:landslide-sleepsets}}),
+	I observed false-positive retry-set-blocked deadlocks fairly often;
+	after the update,
+	it took until 981773 interleavings (>9 hours) into {\tt htm2}(3,2) to find one
+	and confirm the need to still explicitly avoid them by abandoning the retry set in such cases.
+	% TODO put a commit # here
+}.
 %The {\tt landslide} repository's {\tt retry-sets} branch implements this feature,
@@
 			& 2,3 & \cpu{3429.04}	& \ints{86536}	& \cpu{196.18}	& \ints{6366}	& \cpu{997.91}	& \ints{24740} \\
-			& 2,4 & \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq} \\
+			& 2,4 & \ETAdag{29d 8h}	&\ETAdag{2710056}&\cpu{4771.17}	& \ints{123140}	&\ETAdag{5d 1h}	&\ETAdag{1792330} \\
+			% [JOB 4] progress: 683895/2710056 brs (25.235453%), ETA 28d 21h 54m 32s (elapsed 9h 59m 54s)
+			% [JOB 4] progress: 714585/1792330 brs (39.869037%), ETA 4d 14h 46m 38s (elapsed 9h 59m 54s)
 			& 3,1 & \cpu{43.24}	& \ints{774}	& \cpu{24.02}	& \ints{224}	& \cpu{22.79}	& \ints{140} \\
-			& 3,2 & \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq} \\
+			& 3,2 &\ETAdag{1y 299d}	&\ETAdag{4510472}& \ETAdag{qqq}	& \ETAdag{qqq}	&\ETAdag{10d 16h}&\ETAdag{2322150}\\
+			% [JOB 4] progress: 759740/4510472 brs (16.843911%), ETA 1y 298d 22h 14m 13s (elapsed 9h 59m 55s)
+			% [JOB 4] progress: 646606/2322150 brs (27.845142%), ETA 10d 5h 50m 15s (elapsed 9h 59m 55s)
 			& 4,1 & \cpu{9225.52}	& \ints{212146}	& \cpu{376.70}	& \ints{11973}	& \cpu{2158.93}	& \ints{44995} \\
@@
 			% progress: 302022/35094477 brs (0.860597%), ETA 36392y 304d 6h 53m 45s (elapsed 9h 58m 49s)
-			& 3,1 & \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	&\ETAdag{97y 100d}&\ETAdag{1698185036} \\
+			& 3,1 & \ETAdag{5y 292d}&\ETAdag{2873664642}&\ETAdag{qqq}& \ETAdag{qqq}	&\ETAdag{97y 100d}&\ETAdag{1698185036} \\
+			% [JOB 18] progress: 447687/2873664642 brs (0.015579%), ETA 5y 291d 11h 13m 30s (elapsed 9h 59m 41s)
 			% [JOB 18] progress: 357586/1698185036 brs (0.021057%), ETA 97y 99d 8h 44m 33s (elapsed 9h 59m 41s)
@@
 			& 2,1 & \cpu{1950.48}	& \ints{30719}	& \cpu{867.68}	& \ints{13237}	& \cpu{367.91}	& \ints{5446} \\
-			& 2,2 & \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq} \\
-			& 3,1 & \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq} \\
+			& 2,2 & \ETAdag{1y 176d}&\ETAdag{75516602}&\ETAdag{qqq}	& \ETAdag{qqq}	&\ETAdag{13d 11h}&\ETAdag{565334} \\
+			% [JOB 22] progress: 87777/75516602 brs (0.116235%), ETA 1y 176d 16h 38m 50s (elapsed 3h 1m 48s)
+			% [JOB 22] progress: 18380/565334 brs (3.251171%), ETA 13d 1h 5m 3s (elapsed 38m 33s)
+			& 3,1 &\ETAdag{6y 133d}&\ETAdag{1847957714}&\ETAdag{qqq}& \ETAdag{qqq}	&\ETAdag{4y 156d}&\ETAdag{888242178} \\
+			% [JOB 22] progress: 284530/1847957714 brs (0.015397%), ETA 6y 132d 6h 20m 26s (elapsed 9h 49m 5s)
+			% [JOB 22] progress: 274532/888242178 brs (0.030907%), ETA 4y 155d 23h 50m 51s (elapsed 9h 47m 5s)
 			\hline

### 2018-09-01 thesis/thesis.tex
add lotsa comment proto-prose thoughts for future work speculation and conclusion

@@
 
+% test case design, writing code that actually checks the properties you're interested in
+% for example the mutex test, with the if (!xtest) yield check
+% then even after you got the right properties, optimizing the state space
+% with without_function, ignore_sym, etc, while preserving the guarantee
+
+% envision a framework that uses some code coverage tester to guide choice of pps and/or the user's without_function calls 
+% to get that coverage -- although relating flow control coverage directly to concurrency coverage is not straightforward
+% framework can also use 'git diff's, like comparing the most recent commit against an already-trusted codebase,
+% to figure out how to isolate just the stuff the change affects (memory intersection & data flow resulting from those conflix)
+% and then automatically assign some without_functions/etc to test just that commit
+
+% failure injection, testing malloc/etc failures
+% eg student condvars which loop around malloc failures can deadlock
+% can use retry set reduction in principle to prune the same kind of equivalences
+
+% also here we've got discussion of the exit-UAF buggers where landslide oops verifs the test anyway
+% because they're using a free pool and not free()
+% need a 3rd thread at least, to expose this (by actually getting corruption)
+
+% things not to forget about
+% in the maximal state space, the "tag all evil ancestors" thing justified in my MS thesis figure 31337
+% is not necessary, bc if it's the maximal state space, you know that data race PPs would separate the accesses in C1/C2/C3.
+% but, actually figuring out if you're in the maximal state space on-the-fly to justify not tagging?
+% (but but, does this even matter, is this an optimization at all?)
+% i'm actually not even sure this would be sound even on maximal, because of RMW instructions, their 2 evence are inseparable
+
+In the author's experience,
+the WBE algorithm (used in Landslide for estimating number of interleavings) consistently underestimates,
+being often seen to count (nearly) monotonically up toward the true value as exploration progresses,
+while the RE algorithm (used in Landslide for estimating total execution time) can be unstable and erratic,
+bouncing wildly among orders of magnitude even in the space of ten or fewer adjacent interleavings.
+These flaws are especially visible in the {\tt avl\_fixed} transactional verification results (Table~\ref{tab:tm-verifs}).
+% TODO: simpler approach would be to account for like, "T1->T1" prefix subtree likely to be smaller than "T1->T2" prefix subtree
+% because T1->T1->... means T1 has 2 fewer events left to schedule and (M+N-2 choose M-2) tends to be smaller than (M+N-1 choose M-1)
+% for roughly similar M/N;
+% estimator could account for this somehow... not sure how but
+% TODO: cite smth about machine learning
+% TODO: publish the htm evaluation logs
+% state space estimation could probably be drastically improved with machine learning
+% accordingly, all the state space estimate logs (ie, quicksand logs) from the tsx chapter are made open-source on so-and-so
+
+% arugment why ML is NOT appropriate for smc in general -- if you try to train it to choose thread interleavings / search ordering,
+% even if you're already in not-enough-cpu-time heuristics land, you're biasing it to find only the types of bugs you've already seen
+% and destroying its ability to find novelties
+% if your ML provides a partial verification, it is impossible for a human to understand what that verification means
+
 \chapter{Conclusion}
@@
 
+Concurrency testing is, in a way, a microcosm for research in general:
+an exponentially large problem
+% (literally and figuratively for each side of the analogy, respectively)
+that invariably thwarts they whoever try to solve it ``perfectly'';
+it bears fruit only to one who can find their own human-oriented meaning in partial results and compromises.
+
 asdfasdasdf
+
+% TODO: Get a little bit preachy about how computer programs should be created to help humans,
+% not to look as much like scifi as possible, by knowing their limits, asking for help when they get confused, etc.
+
+% ways humans cooperate to make landslide better:
+% - test case design incl. preemption sealing
+% - abstraction reduction
+% - manual inspection of data races
+
 %\appendix

### 2018-09-03 thesis/related.tex
data race and tm rela work; just existing citations, still needta read new papeers

@@
 
-Had I a dollar for every programmer before me who thought to ``solve'' concurrency with a perfect debugging tool,
-well,
-it probably would not be quite enough to retire on.
-At any rate,
-this field is built of the contributions of many a brilliant mind
+%Had I a dollar for every programmer before me who thought to ``solve'' concurrency with a perfect debugging tool,
+%well,
+%it probably would not be quite enough to retire on.
+%At any rate,
+This field is built of the contributions of many a brilliant mind
 trying to carve out a presentable space in an overall impossible problem,
 each making their own tradeoffs along the way.
-While the previous chapters cited prior work as necessary in background discussions, algorithm descriptions, and so on,
+While previous chapters cited prior work as necessary in background discussions, algorithm descriptions, and so on,
 this chapter aims to comprehensively tour the field,
@@
 as bugs immediately, rather than checking them for benign or buggy outcomes.
-All of these tools are notably open-source -- an encouraging recent trend in the field.
+All the tools in this paragraph are notably open-source -- an encouraging recent trend in the field.
 
@@
 Just this year, RCMC \cite{rcmc} proposed to replace the interleaving model entirely with {\em execution graphs},
+which precisely model the executions legal under the RC11 memory model,
 offering further reduction still.
@@
 
+Data race analysis, originating with the lockset-only analysis of Eraser \cite{eraser},
+has since grown into a mature field in its own right,
+which Landslide more borrows as building blocks for its own methods rather than contributing new techniques to.
+% TODO fix terminology not to conflict with dpor
+Race detectors are largely distinguished by their particular flavour of the Happens-Before (HB) relation,
+as discussed in \sect{\ref{sec:background-hb}}.
+Djit+ \cite{djit} and FastTrack \cite{fasttrack} are among those
+which soundly avoid false positives using ``Pure'' HB,
+tracking Lamport-style vector clocks \cite{lamport-clocks}
+for each lock and each thread to compute a global partial order on shared state accesses,
+and flag any access pair not related thereby.
+FastTrack optimizes Djit+'s analysis rules to remove $O(K)$ runtime factors (i.e., linear in the number of threads)
+from several common read and write tracing events;
+however, because $K$ is relatively small in model checking's use cases,
+Landslide uses the Djit+ rules for the sake of implementation simplicity.
+Meanwhile, the ``hybrid'' approach which combines DPOR-style happens-before with locksets \cite{hybriddatarace},
+used in tools such as ThreadSanitizer \cite{tsan},
+compute a more relaxed partial order to find more potential races in a single pass at the cost of false positives.
+I called this ``Limited'' HB on account of how it excludes only those access pairs separated by blocking synchronization,
+not those separated by just locks or barriers,
+as compared to Pure HB.
+Landslide's Limited HB implementation piggy-backs on DPOR's computed happens-before relation,
+supplemented with straightforward lock-sets and heuristic treatment of lock hand-off
+(often common in kernels).
+
+Since these foundational algorithms,
+many more recent works have contributed to make data-race analysis more precise, more performant,
+and/or more domain-specific.
+The Causally-Precedes relation \cite{predictive-dr} is a refinement of Limited HB which avoids the most common cases of false positives,
+% TODO section refrance
+including \sect{\ref{sec:quicksand-soundness}}'s reallocation false positives.
+It could strike a middle ground in the bug-finding/verification tradeoff
+between Pure and Limited HB (\sect{\ref{sec:quicksand-eval}})
+that would be a welcome enhancement in Quicksand.
+IFRit \cite{ifrit}
+improves the performance of Pure HB using an interference analysis,
+which could allow future work to avoid tracing every memory access in a simulator such as Bochs \cite{bochs} or Simics \cite{simics}.
+%
+DroidRacer \cite{droidracer} and CAFA \cite{cafa} find data races in Android applications,
+using domain-specific heuristics (orthogonal to Quicksand's method) to reduce false positives.
+DataCollider \cite{datacollider} finds data races in kernel code
+by using hardware breakpoints and random sampling to achieve high performance.
+% TODO: read some more recent confs for data race shit since thesprop
+% probably like weak memory c.c
+
+Although many SCTs listed in the previous section are content to report any data races as outright bugs,
+RacerX \cite{racerx} showed that tools must be careful not to overwhelm the user with benign warnings
+that she doesn't care about fixing.
+This has motivated replay analysis
+to classify data-race candidates by their impact on program behaviour
+by extending single-pass data-race analysis to many thread interleavings.
+It was first introduced in \cite{recordreplaydrs},
+which compares the program states immediately after the access pair for differences,
+preferring still to err on the side of false positives (as different program states might not necessarily lead to a failure).
+RaceFuzzer \cite{racefuzzer} avoids false positives by requiring an actual failure be exhibited, as Quicksand does,
+although it uses random schedule fuzzing rather than systematic testing for its concurrency coverage.
+%
+Portend \cite{portend} is closest in spirit to Quicksand:
+it tests alternate executions based on single-pass data-race candidates to classify them in a taxonomy of likely severity,
+including non-failing races which nevertheless cause nondeterministic output
+in addition to obvious failures.
+However, it does not
+test alternate interleavings in advance of knowing any specific data races,
+% TODO more speicifc section refrance
+which \sect{\ref{sec:quicksand-eval}} showed is necessary to find certain bugs.
+Quicksand builds on Portend's approach by introducing a feedback loop between the data-race analysis and SCT,
+which results in a stronger verification property
+when the test can be fully completed (\sect{\ref{sec:quicksand-soundness}}).
+Portend also uses symbolic execution to test input nondeterminism as well as schedule nondeterminism,
+while Quicksand remains at the mercy of manual test case design.
+Future work could incorporate Portend's taxonomy to better help the user understand any non-failing data races
+when the test is too large to complete,
+as well as its symbolic execution to help user-provided tests achieve better coverage automatically.
+
 \section{Educational debugging}
@@
 \section{Transactional memory}
+
+Transactional memory (TM), first introduced in 1993 \cite{transactional-memory},
+% TODO expand this to talk about some stm implementations?
+has received renewed attention in recent years with the announcement of Intel's Haswell architecture \cite{htm-haswell},
+which supports hardware transactions (HTM) using new x86 instructions.
+Since then, many studies have evaluated the increased performance it offers over traditional locking and/or STM
+\cite{tm-benchmark-cmu}. % TODO more
+%
+HTM's performance comes at an increased cost in complexity to the programmer,
+who must avoid system calls or transaction nesting, respect the CPU cache capacity,
+and consider retry loops for spurious failure.
+SI-TM \cite{si-tm} introduces techniques for reducing HTM's abort rates for performance's sake,
+but without eliminating them altogether, any full verification must still consider them possible anywhere.
+For programmers who wish to avoid such concerns,
+the simpler STM programming model remains relevant.
+One recent work \cite{hybrid-htm-stm} enhances STM transactions to nest with HTM ones,
+while another \cite{stm-relaxed-memory} adds support for relaxed memory models.
+Such extensions come with the challenge of even more complicated behavioural semantics
+for SCT to accurately model and verify in future work.
+
+Testing approaches for transactional programs are sparsely represented in the literature so far.
+Although several related works \cite{tm-correctness,tm-completeness,specifying-verifying-tm}
+are building up to formal proofs of the correctness of underlying TM {\em implementations},
+Landslide is the first I know of to verify client programs thereof.
+McRT STM \cite{mc-tm-with-spin} uses SPIN \cite{spin} to model check an STM implementation
+up to 2 threads running 1 transaction each with up to 3 memory accesses.
+This kind of verification,
+analogous to \sect{\ref{sec:education-pebbles-tests}}'s {\tt mutex\_test},
+is an important stepping stone for trusting the results Landslide will provide.
+% TODO more tm impl verifs
+
+% TODO: find a home for this sentence
+{\em Learning from Mistakes} \cite{learning-from-mistakes},
+a survey of the characteristics of many types of concurrency bugs,
+found that TM could potentially fix some, but not all, of the studied bugs,
+while in other cases it must be combined with other concurrency primitives to be fully correct,
+motivating the use of SCT to verify such transactional code.
 

### 2018-09-04 thesis/related.tex
related educ

@@
 Although many SCTs listed in the previous section are content to report any data races as outright bugs,
-RacerX \cite{racerx} showed that tools must be careful not to overwhelm the user with benign warnings
-that she doesn't care about fixing.
+RacerX \cite{racerx} showed that tools must be careful not to overwhelm users with benign warnings
+they don't care about fixing.
 This has motivated replay analysis
@@
 
-\section{Educational debugging}
+\section{Concurrency in Education}
 
-Willgrind \cite{willgrind}
+The operating systems curriculum at CMU has used the Pebbles project infrastructure
+and assigned the thread library \cite{thrlib} and kernel \cite{kspec} projects
+in something recognizably close to their modern forms since the Fall 2003 semester.
+% the author's first semester of high school
+I chose Pebbles to target with Landslide because it is closest to home, naturally.
+To indulge my bias as a former member of 15-410 course staff,
+I also believe that Pebbles's open-ended, design-oriented project structure is best suited
+to train students to design robust concurrent code and debug it efficiently,
+as it forces them to consider interactions between many different parts of their design simultaneously.
+However, the difficulty of its concurrency problems (mostly having to do with thread lifecycle)
+leaves little time left in the semester to cover more modern topics
+such as multicore scheduling let alone transactions or relaxed memory
+(all relegated to lecture material not reinforced by the assignments).
+
+Pintos \cite{pintos} has recently emerged as the most popular educational kernel
+(by count of top CS schools in the USA who teach by it);
+it trades off the prevalence of its concurrency challenges to cover various OS topics more broadly,
+especially advanced scheduling algorithms and filesystems.
+Pintos is the stand-alone evolution of its predecessor, Nachos \cite{nachos},
+which originally ran as a UNIX process with simulated device drivers.
+Its popularity motivated me to extend Landslide to support it as an additional kernel architecture
+(an unfortunately arduous task)
+to prove Landslide's mettle beyond CMU's walls.
+Xv6 \cite{xv6}, from MIT, is another major educational kernel, which is also UNIX-like and runs in QEMU,
+and a natural target for model checking in future work.
+Recently, Columbia introduced a new Android-focused OS course \cite{teaching-android},
+which perhaps highlights the importance of related work on model-checking event-driven applications \cite{r4}.
+
+To my knowledge, this is the first study of model checking in an educational setting,
+although teaching concurrency is not itself an unstudied problem.
+%% tf??
+% Eytani et al.~\cite{towards-a-framework} present a promising framework for testing concurrent programs,
+% which can incorporate model checking as well as static analysis, resource exhaustion, data-race analysis, and coverage analysis.
+% However, it lacks an evaluation, and makes mention of its educational value only in its future work remarks.
+%
+%One recent study~
+\cite{how-studence} surveyed how students think about
+testing and debugging during a concurrent programming project,
+finding that unguided, students often approach testing haphazardly,
+not understanding the goal of good concurrency coverage,
+and also had difficulty understanding single failing executions.
+In fact, the study explicitly recommended tool support for testing many thread interleavings automatically (SCT)
+and for execution traces to communicate sequences of important events (preemption traces),
+which I dare say I have achieved in this thesis.
+For future work, I would also recommend studying in more detail the students' thought process
+between receiving a buggy preemption trace and fixing their bug,
+which could perhaps inspire a ``warm up'' Landslide assignment to teach debugging skills explicitly
+and ultimately lead to a higher solve rate on Landslide's bug reports during P2 (\sect{\ref{sec:education-eval-bugs-cmu}}).
+%% this 2nd paper... not freely available.......
+%while Kolikant \cite{learning-concurrency} investigates how students form cognitive patterns about concurrent programming that could either aid or stunt their reasoning.
+%Both of these studies could help optimize Landslide's bug reports for clarity and student enlightenment.
+Willgrind \cite{willgrind} is a tool recently developed at Virginia Tech
+that targets a fork-join parallelism project
+and checks for memory errors (using the Valgrind \cite{valgrind} framework)
+as well as deadlocks, assertion failures, and data races, similarly to Landslide,
+although unlike Landslide, its thread interleaving coverage is as yet limited to stress testing.
+Its GUI-based debugging output is perhaps more friendly than Landslide's HTML preemption traces,
+and its user survey found that students appreciated detailed debugging info especially for deadlocks
+(future work for Landslide),
+but also that students had little patience for even a 5-minute stress test
+when no assurance against false negatives could be provided.
+This suggests motivating students with Landslide's verification guarantee,
+although it is tricky to avoid accientally encouraging % oops
+students to limit possible interleavings by just using one global lock for everything,
+which is counter to 15-410's educational goals.
 

### 2018-09-18 thesis/intro.tex
minor intro updates
write a ton of relwork

@@
 	\item {\bf Hardware Transactional Memory (HTM) (Chapter \ref{chap:tm}).}
-		HTM is a relatively new concurrent programming technique \cite{transactional-memory}
+		HTM is a relatively new concurrent programming technique \cite{htm-haswell,htm-experience}
 		which is not yet addressed by modern model checkers.
@@
 Individual students who participated in the user studies will be given the more inclusive they/them.
-Gender-neutral pronouns are also to be used on the author themself.
+Gender-neutral pronouns will also be used on the author themself.
 

### 2018-09-18 thesis/overview.tex
comments in glossary
write a ton of relwork

@@
 Software transactions (STM) \cite{stm-pldi06} can be used on any commodity processor, but must impose runtime overhead associated with logging.
-Hardware transactions (HTM) \cite{htm-experience, htm-performance} achieve better performance by reusing existing cache coherence logic to detect conflicts, but require explicit support from the CPU, which is not yet widespread.
+Hardware transactions (HTM) \cite{intel-tsx-overview}
+%\cite{htm-experience, htm-performance}
+achieve better performance by reusing existing cache coherence logic to detect conflicts, but require explicit support from the CPU, which is not yet widespread.
 Haswell \cite{htm-haswell} is the first x86 architecture to support HTM,
@@
 \section{Stateless Model Checking}
+\label{sec:overview-stateless} % ugh this label name
 
@@
 
-% TODO
+\begin{enumerate}
+	\item
+\end{enumerate}
 
@@
 % atomicity violation (what even is this.)
+% false positive false negative
 % thread
@@
 % conflict
-% independent: "see conflict"
+% independent: "see conflict"; and note diff b/w this (applies to transitions) equivalent (applies to interleavings)
 % state space
@@
 % iterative deepening
-% MC, stateless MC
+% MC, stateless MC, systematic concurrency testing (TODO: stop calling it SMC start saying SCT)
 % TA: teaching assistant
 % P2 (note abotu psu - dosent call it p2, but we call it that for both schools here
-% happens before, all 3 versions - DPOR, LHB, and PHB
+% happens before, all 3 versions - DPOR, LHB, and PHB (TODO: replace PHB uses of 'happens before' with vector clock
 % userspace, kernelspace
+% "landslide friendly"

### 2018-09-18 thesis/related.tex
write a ton of relwork

@@
 
-Many checkers exist which target programs written for various different types of concurrent environments.
+Many checkers exist which target programs written for various different types of
+concurrent execution and/or programming environments.
 MaceMC \cite{macemc}, MoDist \cite{modist}, SAMC \cite{samc}, ETA \cite{dbug-retreat}, and Concuerror \cite{concuerror},
@@
 R4 \cite{r4} and EventRacer \cite{eventracer} check event-driven concurrent programs typical in mobile applications.
+Like Landslide, SimTester \cite{simtester} is a Simics~\cite{simics}-based tool for kernel-level code,
+although it focuses on interrupt nondeterminism for testing device drivers,
+and is limited to injecting at most one interrupt per test run (as if under ICB with a bound of 1).
+%
+%Other checkers target specific programming languages' concurrency models and/or thread communication APIs.
 dBug \cite{dbug-ssv}, another CMU original similar to CHESS,
-integrates with Parrot \cite{parrot}, a determinizing runtime scheduler,
-to approach state space reduction from a different angle by limiting which thread interleavings are possible to begin with.
-
-Other checkers target specific programming languages' concurrency models and/or thread communication APIs.
-Inspect \cite{inspect} uses a static alias analysis to instrument and preempt all memory accesses to potentially-shared data.
+tests natively-executing programs
+using a dynamic library preload to insert preemption points at pthread and MPI interface boundaries.
+Inspect \cite{inspect} uses a static alias analysis to instrument and preempt all memory accesses to potentially-shared data
+at compile time, in addition to common synchronization APIs.
+RacePRO~\cite{racepro} targets multi-process programs using system calls such as the filesystem API as preemption points
+to find bugs which can corrupt persistent system resources.
 SPIN \cite{spin} tests algorithms defined in the PROMELA domain-specific language,
-instruments every memory access, and
-specializes in verifying synchronization primitives such as RCU \cite{rcu}.
+instruments every memory access,
+uses explicit state tracking rather than the stateless approach (\sect{\ref{sec:overview-stateless}}),
+and specializes in verifying synchronization primitives such as RCU \cite{rcu}.
+TLC \cite{tlc} checks formal models of concurrent program behaviour
+written in the specification language TLA+ \cite{tlaplus},
+and is arguably one of the only true concurrency {\em model checkers}
+as it checks specifications separate from the programs themselves
+%using explicit state tracking,
+rather than attempting to exhaustively exercise every thread interleaving directly.
 %
@@
 which either may arise despite the static no-data-race guarantee.
-The Rust language \cite{rust-language}
-provides a more C++-like type system that also
-%, which I contributed to personally,
-achieves the same guarantee, although I know of no existing SCT for it yet.
 
@@
 % TODO check if ther'es not a better section reference
-(by the C++ definition rather than \sect{\ref{sec:quicksand-soundness}}'s)
+(under the C++ definition rather than \sect{\ref{sec:quicksand-soundness}}'s)
 as bugs immediately, rather than checking them for benign or buggy outcomes.
@@
 and its recent extension to HTM adds support for more modern concurrency patterns as well.
-
-% TODO
-% Surveying concurrency bug detectors based on types of detected bugs - idk what this is, from an email
 
@@
 orders the search space by increasing number of preemptions in each branch,
-which is more likely to expose bugs sooner should they exist;
+which is empirically more likely to expose bugs sooner should they exist;
 BPOR \cite{bpor} extends DPOR to preserve soundness
@@
 Landslide implements ICB and BPOR for Chapter~\ref{chap:quicksand}'s control experiments,
-although does not yet incorporate it into this thesis's own contributions.
+although does not yet incorporate it into this thesis's own contributions
+(as discussed in Chapter~\ref{chap:warpzone}).
 Chapter~\ref{chap:quicksand}'s Quicksand algorithm is, effectively, another such heuristic search strategy,
 focusing on preemption point subsets rather than context switch bounding.
+DeMeter \cite{demeter} adapted abstraction reduction to distributed systems verification under the name Dynamic Interface Reduction,
+while dBug \cite{dbug-phdthesis} applied abstraction reduction to synchronization primitives,
+and I showed how it could be applied similarly to transactional memory in \sect{\ref{sec:tm-abstraction}}.
 Each of these approaches is compatible (and indeed, throughout this thesis used often in concert)
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \section{Data race analysis}
+\label{sec:related-data-race}
 
@@
 
-\section{Concurrency in Education}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{Concurrency in education}
 
@@
 Pintos \cite{pintos} has recently emerged as the most popular educational kernel
-(by count of top CS schools in the USA who teach by it);
+(by count of top CS schools in the United States who teach by it);
 it trades off the prevalence of its concurrency challenges to cover various OS topics more broadly,
@@
 which I dare say I have achieved in this thesis.
-For future work, I would also recommend studying in more detail the students' thought process
-between receiving a buggy preemption trace and fixing their bug,
-which could perhaps inspire a ``warm up'' Landslide assignment to teach debugging skills explicitly
-and ultimately lead to a higher solve rate on Landslide's bug reports during P2 (\sect{\ref{sec:education-eval-bugs-cmu}}).
+A more recent study \cite{novices-programmers}
+examined in detail the students' thought process
+during the diagnosis and fixing phases,
+although its participants were drawn from novice-level programming classes,
+and the experiment was set up with more elementary bugs like syntax and logic errors correspondingly.
+Nevertheless, the authors recommended teaching debugging skills explicitly
+via systematic exposure to different kinds of bugs,
+which suggests future work for even advanced operating systems curricula
+to offer a ``warm up'' Landslide assignment
+(for example, the {\tt atomic\_*} tests from \sect{\ref{sec:education-pebbles-tests}})
+that could ultimately lead to a higher solve rate on Landslide's bug reports during P2 (\sect{\ref{sec:education-eval-bugs-cmu}}).
+
 %% this 2nd paper... not freely available.......
@@
 This suggests motivating students with Landslide's verification guarantee,
+% TODO fix this explanation its too roundabout
 although it is tricky to avoid accientally encouraging % oops
-students to limit possible interleavings by just using one global lock for everything,
+them to limit possible interleavings by just using one global lock for everything,
 which is counter to 15-410's educational goals.
@@
 Since then, many studies have evaluated the increased performance it offers over traditional locking and/or STM
-\cite{tm-benchmark-cmu}. % TODO more
+\cite{htm-experience,htm-performance,tm-benchmark-cmu}.
 %
@@
 while another \cite{stm-relaxed-memory} adds support for relaxed memory models.
+Meanwhile, two recent papers \cite{relaxed-transactions-pldi,relaxed-transactions-popl}
+have proposed formal models of HTM's execution semantics under relaxed memory likewise.
 Such extensions come with the challenge of even more complicated behavioural semantics
@@
 is an important stepping stone for trusting the results Landslide will provide.
-% TODO more tm impl verifs
+STAMP \cite{stamp} is a benchmark suite transactional programs,
+implemented using the OpenTM interface \cite{opentm},
+used by many papers in the field to evaluate the performance of both STM and HTM implementations alike.
+Even so, the more recent Stampede suite \cite{scalable-tm} argues that STAMP's benchmarks
+were constructed under a programming model poorly-suited to fully take advantage of HTM's performance,
+and that scalable HTM programs should
+seek to minimize incidental conflicts and to handle aborts more flexibly than with blind retry loops.
+The programming complexity needed to achieve these goals calls,
+of course,
+for correspondingly advanced verification approaches such as Landslide.
+Finally, TxRace \cite{txrace}
+tests non-transactional programs for data races
+by inserting HTM calls via compiler instrumentation,
+% they don't actually seem to check the failure reason lmao
+% i'll phrase this in a slightly weasel way to point out that's what they *should* be doing
+% and that in turn motivates the stuff i did
+% even tho they use a diff way to filter them post hoc (well, they'd still need to, bc of false sharing)
+relying on conflict aborts to point out access pairs that would be unsafe in the original program.
+This citation perhaps half belongs in \sect{\ref{sec:related-data-race}} as well;
+I include it here to highlight the importance of Landslide's
+ability to distinguish different abort reasons (\sect{\ref{sec:txn-abort-codes}}).
 
-% TODO: find a home for this sentence
+An article from relatively early in the timeline of TSX \cite{htm-subtleties}
+warns of several false equivalence pitfalls when converting conventionally-locking code to use transactions,
+although these pitfalls depend on multiple existing locks used locally and disjointly,
+so this does not invalidate the equivalence proved in \sect{\ref{sec:tm-design-locks}}.
+Rather, Landslide could be used to ensure that freshly-converted transactional code avoids the warned-of pitfalls.
 {\em Learning from Mistakes} \cite{learning-from-mistakes},
@@
 found that TM could potentially fix some, but not all, of the studied bugs,
-while in other cases it must be combined with other concurrency primitives to be fully correct,
-motivating the use of SCT to verify such transactional code.
+while in other cases it must be combined with other concurrency primitives to be fully correct.
+A subsequent paper \cite{applying-tm-bugs}
+found a majority of bugs in their study to be easily fixable with hand-written transactions,
+while others remained out of scope due to blocking {\tt cond\_wait()} operations and the like;
+more recently, the tool BugTM \cite{bugtm} aims to deploy such repairs in production code fully automatically.
+However, these studies all optimize for empirical correctness at best,
+as well as maintaining good performance,
+which motivates the use of tools like Landslide to ensure these rewrites
+are actually correct,
+% oops
+rather than merely shrinking the necessary preemption window required to expose them.
 
-% cchtm - the intel ismm17 paper on nvram htm
-% stamp (mb just cite from txn section)
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-% transactions in relaxed mem architecutres - popl 18
+\section{Other concurrency verification approaches}
 
-% pldi 2018
-% The Semantics of Transactions and Weak Memory in x86, Power, ARM, and C++
+% .. rather than just executing it a bunch of times and seeing what behaviours fall out ..
+
+\subsubsection{Programming language design}
+
+% erlang
+% lvish
+% sully rmc
+
+The Rust language \cite{rust-language}
+provides a more C++-like type system that also
+%, which I contributed to personally,
+achieves the same guarantee, although I know of no existing SCT for it yet.
+
+% TODO: will there be enough content for this?
+% \subsubsection{Other dynamic analyses}
+% symbolic exec, incl concolic (w e that is)
+
+\subsubsection{Deterministic multithreading}
+
+Coming at nondeterminsm from the opposite angle as SCT,
+which aims to push the frontier of testing coverage to expand as many interleavings as possible,
+is deterministic multithreading,
+which reduces the number of interleavings possible to begin with enough that said frontier can reach it more easily.
+Kendo~\cite{kendo} and CoreDet \cite{coredet} were among the first systems to implement this,
+but were limited in which sources of nondeterminism they could control
+and suffered high performance overhead.
+DThreads \cite{dthreads} then extended the scope of determinization to include data races,
+while Peregrine \cite{peregrine}
+improved performance by using record-and-replay to compute a set of possible safe schedules.
+%
+Parrot \cite{parrot} later integrated with the aforementioned dBug \cite{dbug-ssv}
+to offer a partially-determinizing runtime scheduler
+that offered near-baseline performance by allowing the programmer to
+manually annotate speed-critical nondeterministic sections
+and then check the resulting state spaces using dBug's SCT as normal.
+%
+Most recently, Sofritas \cite{sofritas}
+proposed the Ordering-Free Region execution model
+which restricts nondeterminism to only order-enforcing operations such as blocking,
+and automatically suggests refinement annotations to the programmer
+when that would be too aggressive for the intended behaviour.
+These determinizing runtimes serve a different purpose than SCT:
+they seek to preserve the stability of existing code already running in production,
+whether or not concurrency bugs may exist,
+while this thesis aims to eradicate as many such bugs as possible beforehand.
+As Parrot demonstrated, the two approaches are compatible in cases where either extreme be infeasible.
+
+\subsubsection{Kernel verification}
+
+seL4 \cite{sel4} is a microkernel fully designed and specified in Haskell and translated into C.
+Its proofs guarantee not only standard security properties such as process isolation and bounded interrupt latency,
+but also that the C code faithfully implements the specification.
+It addresses concurrency by enabling system interrupts only at carefully-chosen code points,
+and proving bounded runtime besides to ensure good preemptibility.
+%and that the executable binary is correctly compiled from the C.
+This degree of verification must however come at a cost:
+seL4's authors reported over 2 person-years of development effort, with the majority spent on the Haskell specification.
+%
+CertiKOS \cite{certikos} extends this approach to include full concurrency and fine-grained locking
+in the scope of verification,
+using a proof in Coq that also took 2 person-years to complete.
+Its safety properties hold under all possible interleavings,
+and include data-race freedom as well as standard sequential properties
+such as no null dereference and no buffer or integer overflow,
+although it stops short of reasoning about relaxed memory orderings or the TLB cache.
+Many programmers would find a verification cost measured in person-years far too prohibitive,
+while others might argue that for safety-critical kernel code you can't afford {\em not} to verify so thoroughly.
+%
+More recently, Hyperkernel \cite{hyperkernel} extended the xv6 educational kernel \cite{xv6}
+to allow for partial, case-by-case verification of system call behaviour using state-machine specification in Python
+checked by an SMT solver.
+To limit verification complexity, it assumes not only uniprocessor execution
+but also that interrupts be perpetually disabled,
+taking concurrency entirely out of the equation to allow for greater extensibility
+and lessen the programmer's verification burden.
+%
+Heroic as such end-to-end formal verification projects are,
+this thesis finds that trading off thoroughness for accessibility is also acceptable
+if it means helping more users overall.
+Future work could check preemptive and/or multiprocessor kernels
+%by combining Hyperkernel's approach,
+by first checking safety properties in the absence of concurrency,
+then checking with Landslide that concurrency introduces no new program behaviour not already verified,
+to provide a less formal, but still hopefully useful, verification guarantee.

### 2018-09-18 thesis/tm.tex
write a ton of relwork

@@
 which stages changes in per-CPU cache lines,
-and may abort for STM's reason above \cite{htm-experience},
+and may abort for STM's reason above \cite{intel-tsx-overview},
+%\cite{htm-experience},
 or additionally whenever a conflict is detected between one transaction's traced access and {\em any} other memory access,
@@
 \subsection{Modeling Transaction Failure}
+\label{sec:tm-design-failure}
 
@@
 \subsection{Memory Access Analysis}
+\label{sec:tm-design-locks}
 

### 2018-09-20 thesis/related.tex
more rela

@@
 implemented using the OpenTM interface \cite{opentm},
-used by many papers in the field to evaluate the performance of both STM and HTM implementations alike.
+used by many papers in the field to evaluate the performance of both STM and HTM implementations alike,
+although as discussed in \sect{\ref{sec:tm-eval-exp-setup}},
+focuses more on performance than on interesting concurrency properties.
+% stamp bugfixes - https://github.com/mfs409/transmem/blob/master/benchmarks/stamp_c/VERSIONS.
+% stamp need not be considered harmful put here
 Even so, the more recent Stampede suite \cite{scalable-tm} argues that STAMP's benchmarks
@@
 
-% .. rather than just executing it a bunch of times and seeing what behaviours fall out ..
+It should come as no surprise that many avenues of research towards writing correct programs
+have been explored
+apart from just executing them a bunch of times to check all the thread interleavings.
+Though not as directly related to Landslide as the works referenced above,
+this section explores such other approaches,
+ranging from building safety guarantees into a language's type system
+to checking, proving, and/or enforcing execution properties post-hoc.
 
@@
 
+\subsubsection{Symbolic execution}
+
+Analogous to SCT,
+which seeks good coverage of possible thread execution paths under schedule nondeterminism,
+another popular testing approach is symbolic execution \cite{symbolic-execution},
+which seeks good coverage of possible flow control paths under input nondeterminism.
+Symbolic executors
+abstract a program's variables and use constraint solving engines such as Z3 \cite{z3}
+to work backwards and, effectively,
+synthesize combinations of test inputs which can lead to a failure.
+KLEE \cite{klee}, one well-known and open-source implementation of this technique,
+offers over 90\% code coverage on average across many tests,
+often outdoing that achieved by programmers' own hand-written tests.
+Later, Contessa \cite{contessa} extended symbolic execution to include concurrency nondeterminism as well,
+by using a DPOR-like analysis on individual execution traces
+then incorporating reordering possibilities into its SMT constraints.
+%rather than using explicit scheduling as SCT does.
+This simultaneously exercises both input and schedule nondeterminism,
+and achieves good bug-finding performance in practice,
+but does not provide the same verification guarantees as repeated DPOR iterations with explicit scheduling.
+Exploring both kinds of state space at once thoroughly enough to provide strong verification
+is undoubtedly subject to further state space explosion,
+and remains future work.
+
 \subsubsection{Kernel verification}

### 2018-09-20 thesis/related.tex
related - pls

@@
 
-% erlang
-% lvish
-% sully rmc
-
-The Rust language \cite{rust-language}
-provides a more C++-like type system that also
-%, which I contributed to personally,
-achieves the same guarantee, although I know of no existing SCT for it yet.
-
-% TODO: will there be enough content for this?
-% \subsubsection{Other dynamic analyses}
-% symbolic exec, incl concolic (w e that is)
+While C's extremely rudimentary type system allows the compiler to statically check programs
+for properties such as not accidentally dereferencing raw integer values as if they were pointers,
+more advanced programming languages may make guarantees about concurrent execution.
+%
+Erlang \cite{erlang}, an early concurrent functional language,
+introduced the actor model for concurrency,
+in which threads share no state and must communicate only by message-passing.
+%One might say this statically guarantees freedom from data races;
+%from another point of view, data races do not even exist as a concept in this programming model.
+While this statically guarantees the absence of data races,
+programs may still execute nondeterministically,
+so concurrency bugs, especially deadlocks, are not ruled out.
+Concuerror \cite{concuerror}, discussed above, is a SCT tool for Erlang programs.
+%
+Haskell \cite{haskell} offers a more sophisticated interface to concurrency:
+threads may reference the same objects and even update shared references
+using monads that encapsulate mutation,
+but at the (garbage-collected) execution level all data is immutable once created,
+which preserves type soundness and data-race freedom.
+The aforementioned D\'{e}j\`{a} Fu \cite{dejafu} checks concurrent Haskell programs.
+%
+Rust \cite{rust-language}
+presents a type system with more explicit memory management,
+in-place mutation,
+and mutable references
+to appear familiar and approachable to those already versed in C++.
+It proposes a borrow-check analysis to ensure memory and type safety
+despite mutable references,
+and a trait system to ensure no shared state between threads by default.
+Its concurrency libraries then offer interfaces which relax this restriction,
+allowing threads even to simultaneously reference shared mutable state,
+using the type system to enforce sound locking discipline across such accesses,
+again preserving type soundness and data-race freedom%
+\footnote{The author contributed the original design for this latter feature.}.
+I know of no existing SCT for Rust as of yet.
+%
+The Relaxed Memory Calculus \cite{sully-thesis}
+proposes to extend C++ with annotations for weak memory atomics,
+which allows for static formal analysis of memory access reorderings.
+Although not ruling out data races,
+this approach is an important step towards compilers
+which can statically reason about program execution under more advanced concurrency models.
+%
+Finally, LVish \cite{lvish}
+features a type system that enforces deterministic behaviour by construction,
+using shared state called LVars which allow writes only in ways that update order is not observable.
+This renders thread interleavings entirely irrelevant,
+obviating any need for runtime verification,
+but at the cost of a more restrictive programming model.
 
@@
 which reduces the number of interleavings possible to begin with enough that said frontier can reach it more easily.
+Unlike LVish, described above, these systems provide deterministic execution
+even for the familiar, C-like, shared-state multithreading programming model.
 Kendo~\cite{kendo} and CoreDet \cite{coredet} were among the first systems to implement this,

### 2018-09-20 thesis/related.tex
ffinish related

@@
 extends DPOR to event-driven programs by separating the notion of enabled events from that of multiple threads.
+TaxDC \cite{taxdc}, a taxonomy study of distributed systems concurrency bugs,
+showed that for completeness distributed model checkers must incorporate many forms of nondeterminism,
+including message reordering, timeouts, network disconnections, and crashes and reboots, in addition to local threads.
 DPOR for TSO and PSO \cite{tsopso}
@@
 
-It should come as no surprise that many avenues of research towards writing correct programs
+%It should come as no surprise that
+Naturally, many avenues of research towards writing correct programs
 have been explored
-apart from just executing them a bunch of times to check all the thread interleavings.
-Though not as directly related to Landslide as the works referenced above,
-this section explores such other approaches,
-ranging from building safety guarantees into a language's type system
+apart from just executing them a bunch of times to check all the interleavings.
+Though not as directly related as the works referenced above,
+this section explores such approaches,
+ranging from expressing safety guarantees in a language's type system
 to checking, proving, and/or enforcing execution properties post-hoc.
@@
 again preserving type soundness and data-race freedom%
-\footnote{The author contributed the original design for this latter feature.}.
+\footnote{The author themself contributed the original design for this latter feature.}.
 I know of no existing SCT for Rust as of yet.
@@
 Symbolic executors
-abstract a program's variables and use constraint solving engines such as Z3 \cite{z3}
-to work backwards and, effectively,
-synthesize combinations of test inputs which can lead to a failure.
-KLEE \cite{klee}, one well-known and open-source implementation of this technique,
+abstract a program's variables and use constraint solvers such as Z3 \cite{z3}
+to work backwards and synthesize combinations of test inputs which can lead to a failure.
+%
+KLEE \cite{klee}, one well-known and open-source implementation,
 offers over 90\% code coverage on average across many tests,
 often outdoing that achieved by programmers' own hand-written tests.
+%
 Later, Contessa \cite{contessa} extended symbolic execution to include concurrency nondeterminism as well,
 by using a DPOR-like analysis on individual execution traces
-then incorporating reordering possibilities into its SMT constraints.
+then including reordering possibilities in its SMT constraints.
 %rather than using explicit scheduling as SCT does.
 This simultaneously exercises both input and schedule nondeterminism,
-and achieves good bug-finding performance in practice,
+%and achieves good bug-finding performance in practice,
 but does not provide the same verification guarantees as repeated DPOR iterations with explicit scheduling.
@@
 and remains future work.
+%
+Symbiosis \cite{symbiosis} starts from the known root cause of an existing failure
+and uses symbolic execution to synthesize a schedule to reproduce it,
+then further searches for a non-failing schedule and compares them to produce
+a minimum sequence of events necessary for the failure.
+This approach skips the initial verification step entirely,
+but greatly reduces the diagnosis effort required of the user,
+which was a common complaint about Landslide's preemption traces.
 

### 2018-09-21 thesis/landslide.tex
final pre-send-to-committee revisions to education chapater
i don't know if blacklist/whitelist is truly racist but it cost literally 0 dollars and 0 cents to change ur language just to be safe

@@
 \begin{itemize}
-	\item {\tt within\_function FUNC} - adds {\tt FUNC} to a whitelist of functions required to appear in the current stack trace before identifying a preemption point (see \sect{\ref{sec:landslide-pps}})
-	\item {\tt without\_function FUNC} - as above, but a blacklist instead of a whitelist
+	\item {\tt within\_function FUNC} - adds {\tt FUNC} to an allowlist of functions required to appear in the current stack trace before identifying a preemption point (see \sect{\ref{sec:landslide-pps}})
+	\item {\tt without\_function FUNC} - as above, but a denylist instead of an allowlist
 	\item {\tt within\_user\_function FUNC} - as two above but finds the function in the userspace test program rather than the kernel code.
@@
 \subsubsection{Heap checking}
+\label{sec:landslide-valgrind-mode}
 
@@
 
-{\tt check\_withins()} implements the whitelist/blacklist behaviour for the former two of those commands
+{\tt check\_withins()} implements the allow/denylist behaviour for the former two of those commands
 (in a similar manner to the Preemption Sealing technique described in prior work \cite{sealing}).
@@
 \begin{itemize}
-	\item If no {\tt within\_function} commands are given, operate in ``blacklist mode'':
+	\item If no {\tt within\_function} commands are given, operate in ``denylist mode'':
 		the preemption point is by default valid as long as no {\tt without\_function} calls reject it.
-		Otherwise, operate in ``whitelist mode'':
+		Otherwise, operate in ``allowlist mode'':
 		the preemption point is by default rejected unless at least one {\tt within\_function} directive matches.
@@
 
-		No {\tt without}s present, so behaves as a whitelist,
+		No {\tt without}s present, so behaves as an allowlist,
 		rejecting stack trace 1 (not in {\tt thr\_join()}),
@@
 
-		No {\tt within}s present, so behaves as a blacklist,
+		No {\tt within}s present, so behaves as a denylist,
 		accepting stack trace 1 (not in {\tt cond\_\allowbreak{}wait()}),

### 2018-09-25 thesis/overview.tex
start wirting glossary

@@
 This section provides a convenient reference of terminology used throughout the thesis.
+Use of {\em italics} within definitions will indicate terms also defined in this list.
 
 \begin{enumerate}
-	\item
+	\item {\bf API.}
+		Short for {\em application programming interface} (which nobody ever says).
+		% TODO
+	\item {\bf Atomic.}
+		Said of a sequence of operations within a single thread,
+		whose intermediate state cannot be observed by or interfered with
+		by certain conflicting operations from another thread.
+		Must always be qualified as ``atomic {\em with respect to} X'',
+		where X is the set of possibly-conflicting operations to be protected against.
+		Apart from hardware-provided atomic assembly instructions (e.g., {\tt xchg}),
+		i.e., for any multi-instruction sequence,
+		setting X as ``everything'' can be surprisingly difficult to achieve safely in practice;
+		for example, disabling interrupts in kernel code does not prevent other CPUs' simultaneous execution.
+	\item {\bf Atomicity violation.}
+		A term introduced by AVIO \cite{avio} to refer to a particular class of {\em concurrency bug}
+		in which the programmer's expectation of {\em atomicity} around a certain sequence of operations
+		is violated by another thread interleaving in between.
+		This is distinct from language-level {\em data races},
+		which may result in bugs by being reordered by the compiler or hardware;
+		% TODO: put a section reference for weak memory
+		% mb have to write a section on it here
+		however, this thesis checks programs at the executable level and assumes sequential consistency,
+		so all data races that still result in bugs (e.g., unprotected {\tt x++})
+		must technically also be atomicity violations.
+		Also, this thesis is focused on identifying failures, not root causes,
+		so will simply avoid the term entirely.
+	\item {\bf Branch.}
+		Another term for {\em interleaving}, with an emphasis on the {\em execution tree}
+		interpretation of {\em state spaces}.
+		May also refer to an entire subtree outside of the current interleaving,
+		characterized by the choice to run a certain {\em thread} at a certain past {\em preemption point},
+		comprising all possible interleavings which may arise with that as a prefix.
+	\item {\bf Concurrency bug.}
+		A program behaviour which manifests nondeterministically depending on the thread scheduling pattern,
+		which is not desired or intended by the programmer.
+		This thesis further assumes these to be identifiable by computable predicates on the execution state,
+		excluding some real bugs which may result in data corruption that never actually leads to crashes or assertion
+		failures under the test case in question.
+	\item {\bf Conflict.}
+		% TODO
+	\item {\bf Data race.}
+		An unsynchronized pair of memory accesses between two threads that can, may, or did execute concurrently.
+		{\em Race} for short.
+		Distinguished further in this thesis as {\em limited happens-before} races and {\em pure happens-before} races;
+		see \sect{\ref{sec:background-datarace}}.
+		Distinct from the more nebulously-defined {\em concurrency bug},
+		in that they are identified precisely by analysis on execution traces of one (or all) thread interleavings.
+		This thesis considers any memory accesses not protected by a known trusted {\em mutex API}
+		to be potential races;
+		this may include atomic instructions such as {\tt xchg},
+		which most all-data-races-are-bugs research does not include,
+		but are necessary for {\em Iterative Deepening}'s {\em verification} guarantee
+		(\sect{\ref{sec:quicksand-soundness}}).
+	\item {\bf Deterministic bug.}
+		As {\em concurrency bug}, but manifesting in all (or at least most) thread interleavings,
+		making it easy for programmers to find and debug without advanced tooling.
+		This thesis will occasionally use ``manifests on the 1st interleaving Landslide tested''
+		as an approximation, without necessarily checking for its presence or absence in other branches.
+	\item {\bf DPOR.} See {\em dynamic partial order reduction}.
+	\item {\bf Dynamic partial order reduction}.
+		Exploration algorithm for {\em stateless model checkers} which
+		identifies alternate interleavings to explore likely to result
+		in reorderings of {\em memory conflicts} observed in the current branch.
+		Runs iteratively upon completion of each branch, tagging one or more sibling branches
+		for future exploration, skipping over ones guaranteed to be {\em independent}.
+		See \sect{\ref{sec:landslide-dpor}}.
+	\item {\bf Equivalent.}
+		Said of two {\em interleavings} where one can be transformed into the other using
+		permutations of only {\em independent interleavings}.
+		Note the difference in ``type signature'' with {\em independent}, which applies to {\em transitions} instead.
+	\item {\bf Estimation.}
+		An algorithm whose input is the known structure of a partially-explored {\em state space}
+		and whose output is a guess at the ultimate size of the state space when exploration finish. % subjunctive woo
+		Output may be in units of either time or branches.
+		See \sect{\ref{sec:landslide-estimate}}.
+	\item {\bf ETA.}
+		Short for {\em estimation} (or ``state space estimate''),
+		typically with nuance for units of time rather than branches.
+	\item {\bf Execution tree.}
+		% TODO
+	\item {\bf False negative.}
+		Said of a tool's failure to produce a bug report when a bug does exist after all.
+		In theory this may mean that the {\em test case} might have been insufficient to expose the bug
+		(i.e., the code under test is wrong in any way whatsoever);
+		however, for pragmatic, tool-oriented discussion,
+		this thesis will use the term to mean bugs observable under the given test case which the tool itself fails to find.
+	\item {\bf False positive.}
+		Said of a bug report which,
+		upon subjective manual inspection by the programmer,
+		does not indicate a bug after all.
+		In principle, Landslide's bug-detection predicates always indicate something truly wrong,
+		so it cannot produce false positive bug reports;
+		in practice, however, infinite loop detection and deadlock detection in the presence of ad-hoc synchronization
+		are both heuristic (\sect{\ref{sec:landslide-blocking}}),
+		so may produce false positives after all, although I have never witnessed one to date.
+	\item {\bf Happens-before.}
+		% TODO
+	\item {\bf Heuristic.}
+		% TODO
+	\item {\bf Independent.}
+		% TODO
+		Note the difference in ``type signature'' with {\em equivalent}, which applies to {\em interleavings} instead.
+	\item {\bf Interleaving.}
+		% TODO
+	\item {\bf Iterative Deepening.}
+		The heuristic search ordering strategy used by {\em Quicksand}.
+		One of this thesis's major contributions.
+		See \sect{\ref{sec:quicksand-id}}.
+	\item {\bf Kernelspace.}
+		Opposite of {\em userspace}.
+		Execution mode in which code has the ability to modify critical system state.
+		In addition to having access to privileged CPU instructions,
+		kernelspace code is also responsible for implementing safe
+		concurrent scheduling, safe virtual memory protection and isolation guarantees, et cetera.
+	\item {\bf Landslide.}
+		Your friendly local {\em stateless model checking} implementation, written by yours truly.
+		The framework upon which all this thesis's contributions are built.
+		Documented in Chapter \ref{chap:landslide}.
+	\item {\bf Landslide-friendly.}
+		Said of a {\em test case} that is well-suited for {\em model checking},
+		i.e.,
+		contains few or no loops which,
+		under normal circumstances, would increase the probability of a bug manifesting,
+		but in principle do not make any new behaviours possible that would be impossible without the loop.
+	\item {\bf Limited happens-before.}
+		% TODO
+	\item {\bf Lock.}
+		% TODO
+	\item {\bf Memory conflict.} See {\em conflict}.
+	\item {\bf mutex.}
+		Short for {\em mutual exclusion lock}. The name {\em P2} and {\em pthreads} use for their {\em lock API}.
+	\item {\bf Model checking.}
+		In general, refers to the whole research field of checking program implementations against
+		formal specifications in a theoretically rigorous way.
+		In this thesis, used as shorthand for {\em stateless model checking}.
+	\item {\bf P2.}
+		% TODO
+		{\em API} is very similar to that of {\em pthreads}.
+		Often used here to refer to the same thread library project at both CMU and PSU,
+		even though the latter did not call it by that name.
+	\item {\bf pthreads.}
+		Thread library for Linux programs provided by POSIX.
+	\item {\bf Pure happens-before.}
+		% TODO
+	\item {\bf Quicksand.}
+		% TODO
+	\item {\bf Race condition.}
+		A confusing term which, while CMU 15-410 teaches it to mean {\em concurrency bug},
+		in research circles is generally taken to mean {\em data race}.
+		This thesis will avoid it entirely, using those two more precise terms instead.
+	\item {\bf Schedule.}
+		Another term for {\em interleaving} with a more theoretical nuance.
+	\item {\bf Stateless model checking.}
+		% TODO
+	\item {\bf State space.}
+		% TODO
+	\item {\bf Systematic testing.}
+		Another name for {\em stateless model checking}.
+	\item {\bf TA.} See {\em teaching assistant}.
+	\item {\bf Teaching assistant.}
+		% TODO
+	\item {\bf Test case.}
+		% TODO
+	\item {\bf Thread.}
+		% TODO
+	\item {\bf Transition.}
+		% not the kind of transition that's for people
+		A sequence of execution steps by a single thread,
+		characterized by the two different program states at beginning and end.
+		In this thesis, will be used exclusively when the terminal program states are {\em preemption points},
+		and no intermediate state is a preemption point as well.
+	\item {\bf Userspace.}
+		Opposite of {\em kernelspace}.
+		Execution mode in which programs, such as {\em P2} or client code thereof,
+		run with limited privileges,
+		such as no access to the CPU's control registers, interrupt descriptor table, ability to perform device I/O,
+		et cetera.
+		Assumed to run with the virtual memory protections standard in any modern kernel
+		(such as crashing when dereferencing a null pointer).
+	\item {\bf Vector clock.}
+		% TODO
+	\item {\bf Verification.}
+		Formally certifying the correctness of a program under certain properties.
+		In this thesis, will always be constrained by the set of behaviours
+		the {\em test case} is capable of exposing in the first place.
 \end{enumerate}
-
-% concurrency bug
-% race condition: "a confusing term which in some circles means concurrency bug and others data race, which we will avoid"
-% data race
-% atomicity violation (what even is this.)
-% false positive false negative
-% thread
-% transition
-% conflict
-% independent: "see conflict"; and note diff b/w this (applies to transitions) equivalent (applies to interleavings)
-% state space
-% interleaving
-% DPOR
-% estimation (WBE, RE)
-% ETA
-% landslide
-% quicksand
-% iterative deepening
-% MC, stateless MC, systematic concurrency testing (TODO: stop calling it SMC start saying SCT)
-% TA: teaching assistant
-% P2 (note abotu psu - dosent call it p2, but we call it that for both schools here
-% happens before, all 3 versions - DPOR, LHB, and PHB (TODO: replace PHB uses of 'happens before' with vector clock
-% userspace, kernelspace
-% "landslide friendly"

### 2018-09-25 thesis/overview.tex
write more glosses

@@
 This section provides a convenient reference of terminology used throughout the thesis.
-Use of {\em italics} within definitions will indicate terms also defined in this list.
+Use of {\em italics} within definitions will indicate terms also defined elsewhere in the list.
 
 \begin{enumerate}
+	\item {\bf Actor.}
+		(Also ``actor concurrency model''.)
+		An independent execution component of a concurrent system.
+		In {\em pthread}-like multithreaded programs, each {\em thread} is an actor.
+		In distributed systems, each machine is an actor.
 	\item {\bf API.}
-		Short for {\em application programming interface} (which nobody ever says).
-		% TODO
+		The set of function signatures which comprise the interface to a body of code.
+		For example, the standard {\em mutex} API consists of
+		{\tt mutex\_init()}, {\tt mutex\_lock()}, {\tt mutex\_unlock()},
+		and optionally {\tt mutex\_trylock()} and {\tt mutex\_destroy()}.
+		From ``application programming interface'' (which nobody ever says).
 	\item {\bf Atomic.}
@@
 		comprising all possible interleavings which may arise with that as a prefix.
+	\item {\bf Completeness.}
+		%Dual of {\em soundness}.
+		In logic, said of an inference system in which any proposition consistent with the axioms can be proved.
+		When speaking of {\em state space reduction},
+		could mean opposite things depending on perspective:
+		complete from an interleaving coverage perspective would mean not skipping any buggy interleavings;
+		complete from a state space reduction perspective would mean not testing any two equivalent interleavings
+		without pruning one.
+		I define {\em soundness} as the dual from the reduction perspective (i.e., not pruning anything unsafe to prune),
+		which is actually identical to the coverage perspective definition of completeness.
+		To avoid confusion, this thesis avoids this term altogether,
+		using {\em optimality} to refer to the reduction perspective definition instead.
 	\item {\bf Concurrency bug.}
@@
 		failures under the test case in question.
-	\item {\bf Conflict.}
-		% TODO
+	\item {\bf Conflict.} See {\em memory conflict}.
 	\item {\bf Data race.}
@@
 		(\sect{\ref{sec:quicksand-soundness}}).
+	\item {\bf Deadlock.}
+		An execution state among a set of N {\em threads},
+		where each thread is blocked waiting for an event that can be caused only by another thread in the set
+		\cite{deadlock}.
+		May arise either deterministically or concurrently; {\em Landslide} considers it a bug in any case.
 	\item {\bf Deterministic bug.}
@@
 		as an approximation, without necessarily checking for its presence or absence in other branches.
+	\item {\bf Disjoint.}
+		Said of a pair or of two sets of memory accesses
+		when there are no two elements that share an address and also are not both reads.
 	\item {\bf DPOR.} See {\em dynamic partial order reduction}.
+	\item {\bf Dynamic analysis.}
+		Program analysis technique which involves analyzing concrete execution(s) of a program,
+		either through tracing (via compiler-inserted instrumentation, simulation, or other means)
+		or through post-hoc analysis of event logs.
+		Searches for concrete evidence of undesirable program behaviour,
+		and is blind to what the source code looks like (or even what language it be written in).
+		Opposite of {\em static analysis}.
 	\item {\bf Dynamic partial order reduction}.
@@
 		typically with nuance for units of time rather than branches.
+		From ``estimated time of arrival''.
 	\item {\bf Execution tree.}
@@
 		in practice, however, infinite loop detection and deadlock detection in the presence of ad-hoc synchronization
-		are both heuristic (\sect{\ref{sec:landslide-blocking}}),
+		are both {\em heuristic} (\sect{\ref{sec:landslide-blocking}}),
 		so may produce false positives after all, although I have never witnessed one to date.
 	\item {\bf Happens-before.}
-		% TODO
+		Relation between two events executed by different threads in a single execution trace
+		which captures the notion of concurrent execution.
+		Comes in two flavours, {\em pure happens-before} and {\em limited happens-before}.
 	\item {\bf Heuristic.}
-		% TODO
+		A testing strategy informed by empirical evidence rather than theoretically-founded algorithms.
+		For example, {\em ICB} is a heuristic way of prioritizing {\em interleavings} with fewer {\em preemptions},
+		based on the observation that most concurrency bugs in practice can be exposed with as few as 2-3 preemptions.
+		For example, {\em Landslide} heuristically judges the program to be stuck in an infinite loop
+		when it has executed at least 4000 times longer than previously-observed ``safe'' executions,
+		based on the observation that no student has ever written correct thread library or kernel code
+		that can exhibit such extreme interleaving-dependent variance.
+	\item {\bf ICB.} Short for ``Iterative Context Bounding''. See \sect{\ref{sec:landslide-icb}}.
 	\item {\bf Independent.}
-		% TODO
+		Said of two {\em transitions} by different threads,
+		when they share no {\em memory conflicts};
+		i.e., they can have no effect on each other's behaviour even if reordered.
 		Note the difference in ``type signature'' with {\em equivalent}, which applies to {\em interleavings} instead.
 	\item {\bf Interleaving.}
-		% TODO
+		A sequence of {\em transitions} of many threads representing the complete or partial execution
+		of a {\em test case}.
 	\item {\bf Iterative Deepening.}
-		The heuristic search ordering strategy used by {\em Quicksand}.
+		The {\em heuristic} search ordering strategy and framework for incorporating {\em data race preemption points}
+		used by {\em Quicksand}.
 		One of this thesis's major contributions.
@@
 	\item {\bf Limited happens-before.}
-		% TODO
+		Variant of {\em happens-before} that relates two events only when a blocking/signalling synchronization affecting
+		the second thread's runnability, such as {\tt cond\_wait()}/{\tt cond\_signal()} or {\tt thr\_fork()}
+		occurs in between.
+		In contrast with {\em pure happens-before},
+		indicates not just that the events were observed
+		to happen non-concurrently in the execution in question,
+		but also that they could not be reordered even in alternate interleavings.
 	\item {\bf Lock.}
 		% TODO
-	\item {\bf Memory conflict.} See {\em conflict}.
+	\item {\bf Memory conflict.}
+		A pair of memory accesses between two {\em threads}, where at least one is a write,
+		meaning that executing them in a different order from the one observed may produce different program behaviour.
+		If both are writes, reordering them may cause a different value to be globally visible in that location
+		to subsequent parts of the program.
+		If one is a read and one is a write, reordering them may cause the read access to return a different value,
+		possibly affecting that thread's subsequent logic.
 	\item {\bf mutex.}
-		Short for {\em mutual exclusion lock}. The name {\em P2} and {\em pthreads} use for their {\em lock API}.
+		Short for ``mutual exclusion lock''. The name {\em P2} and {\em pthreads} use for their {\em lock API}.
 	\item {\bf Model checking.}
@@
 		In this thesis, used as shorthand for {\em stateless model checking}.
+	\item {\bf Optimality.} See {\em completeness}.
 	\item {\bf P2.}
-		% TODO
+		Operating systems class project in which students must implement synchronization primitives
+		and thread lifecycle functions.
 		{\em API} is very similar to that of {\em pthreads}.
@@
 		even though the latter did not call it by that name.
+		See \sect{\ref{sec:pebbles}}.
 	\item {\bf pthreads.}
 		Thread library for Linux programs provided by POSIX.
+	\item {\bf Preemption.}
+		The act of interrupting one thread's execution to force a context switch to another.
+		Can be performed either deliberately by the {\em model checker}
+		or randomly by the {\em scheduler} as driven by system interrupts.
+	\item {\bf Preemption point.}
+		A single instance of a program state or code location where a {\em preemption point predicate} returns true.
+		This thesis will also use this as shorthand for preemption point predicates,
+		using the longer term only when the precise difference is important.
+	\item {\bf Preemption point predicate.}
+		% TODO
 	\item {\bf Pure happens-before.}
-		% TODO
+		Variant of {\em happens-before} that relates two events when any synchronization operation introducing a memory
+		barrier, which would prevent compiler or hardware access reorderings, occurs in between.
+		In contrast with {\em limited happens-before},
+		indicates that the events were observed to execute non-concurrently in this execution,
+		but could possibly be reordered to be concurrent in alternate interleavings.
 	\item {\bf Quicksand.}
-		% TODO
+		Testing framework which manages the execution of multiple {\em Landslide} instances,
+		each configured with a different set of {\em preemption points},
+		using the {\em Iterative Deepening} algorithm.
+		Also supports various other execution modes in which only one Landslide instance is run at a time.
+		See Chapter~\ref{chap:quicksand}.
+		In later chapters, ``Landslide'' will refer to the project as a whole, including Quicksand,
+		while the term ``Quicksand'' will be used only when specific reference to Iterative Deepening is required.
 	\item {\bf Race condition.}
-		A confusing term which, while CMU 15-410 teaches it to mean {\em concurrency bug},
-		in research circles is generally taken to mean {\em data race}.
+		A confusing term which CMU 15-410 teaches to mean {\em concurrency bug},
+		but in research circles is generally taken to mean {\em data race}.
 		This thesis will avoid it entirely, using those two more precise terms instead.
+	\item {\bf Reduction.}
+		The act of making a {\em state space} smaller by finding multiple execution sequences that can be proved equivalent,
+		such that all but one may be skipped over.
 	\item {\bf Schedule.}
 		Another term for {\em interleaving} with a more theoretical nuance.
+	\item {\bf Scheduler.}
+		The component of an operating system kernel responsible for tracking the set of runnable threads,
+		handling system interrupts,
+		and allocating execution time among them (in round-robin, priority-driven, or other manner).
+		May also to refer to the analogous component of {\em Landslide}
+		which overrides the decision of the kernel scheduler.
+	\item {\bf Soundness.}
+		Said of a search algorithm
+		which provides the property that if a bug exists under the given test case,
+		given arbitrary CPU time,
+		it will eventually be found;
+		in other words, admits no {\em false negatives}.
+		When said of a {\em reduction} strategy as applies to an existing search algorithm,
+		means that any bugs found by the original algorithm will still be found under the reduction,
+		even if there were false negatives
+		(e.g., applying {\em DPOR} without using {\em data race preemption points}).
+		Used by analogy with logic, in which said of an inference system,
+		means contradiction cannot be derived from the axioms.
+		See also {\em completeness}.
 	\item {\bf Stateless model checking.}
-		% TODO
+		Concurrency testing {\em dynamic analysis} technique involving controlling a system's nondeterminism
+		to exhaustively check all possible thread interleavings.
+		See \sect{\ref{sec:overview-stateless}}.
+		Often ``MC'' for short in this thesis (to mean either ``model checking'' or ''model checker'' depending on context).
+		This name is potentially confusing for tools which explore {\em state spaces}:
+		the term ``stateless'' refers to the implicit way they use {\em DPOR} to ensure {\em sound} checking
+		of all possible program states, without actually explicitly storing those states in memory.
+		The second half of the name is another historical accident,
+		as most modern MCs do not check ``models'' in the sense that
+		programs be verified against external formal specifications,
+		but rather the program's internal assertions serve as informal specifications,
+		alongside the MC's own bug-detection predicates.
 	\item {\bf State space.}
 		% TODO
+	\item {\bf Static analysis.}
+		Class of program analysis techniques which process the source code,
+		usually at syntax-tree or LLVM IR level,
+		looking for suspicious implementation patterns,
+		rather than actually observing behaviours that arise during execution.
+		Opposite of {\em dynamic analysis}.
 	\item {\bf Systematic testing.}
@@
 	\item {\bf Teaching assistant.}
-		% TODO
+		A student who volunteers to help run a class, holding office hours, grading projects, and so on,
+		compensated either with course credit or pay from the university.
+		Probably has taken the class herself in the past.
 	\item {\bf Test case.}
-		% TODO
+		A piece of code, separate from the code under test,
+		to drive execution of the code under test according to a certain pattern designed to expose interesting behaviour.
+		In the context of {\em P2}, the thread library is the code under test,
+		and client code thereof, such as {\em paradise\_lost.c}, is the test case.
 	\item {\bf Thread.}
-		% TODO
+		Unit of concurrent execution in a typical modern program,
+		concretely represented as a set of private CPU registers, including program counter.
+		Probably has its own stack as well, allocated by the userspace thread library.
+		A program may have as many parts of its code executing simultaneously as threads exist.
 	\item {\bf Transition.}
@@
 	\item {\bf Vector clock.}
-		% TODO
+		Data structure for time-keeping in distributed systems which uses a list of epoch numbers
+		to track when the execution of any given pair of {\em actors} should be considered simultaneous/concurrent.
+		Used in {\em Landslide} to implement {\em pure happens-before}.
 	\item {\bf Verification.}

### 2018-09-26 thesis/overview.tex
finish glossing

@@
 
-\subsection{Concurrency Primitives}
+\subsection{Concurrency primitives}
+\label{sec:overview-mutex}
 
@@
 \subsection{The state space}
+\label{sec:overview-statespace}
 
@@
 	\item {\bf Concurrency bug.}
-		A program behaviour which manifests nondeterministically depending on the thread scheduling pattern,
+		A program behaviour which manifests {\em nondeterministically} depending on the thread scheduling pattern,
 		which is not desired or intended by the programmer.
@@
 		when there are no two elements that share an address and also are not both reads.
-	\item {\bf DPOR.} See {\em dynamic partial order reduction}.
+	\item {\bf DPOR.} {\em Dynamic partial order reduction}.
 	\item {\bf Dynamic analysis.}
@@
 	\item {\bf Execution tree.}
-		% TODO
+		Representation of the {\em state space} by joining interleavings with common prefixes
+		into a subtree with a branch at each point they diverge, forming a tree overall.
+		See \sect{\ref{sec:overview-statespace}}; Figure~\ref{fig:tree}.
 	\item {\bf False negative.}
@@
 	\item {\bf Lock.}
-		% TODO
+		A data structure used to ensure {\em atomic} access to shared state
+		with respect to other {\em threads} using the same lock.
+		See \sect{\ref{sec:overview-mutex}}.
 	\item {\bf Memory conflict.}
@@
 		In this thesis, used as shorthand for {\em stateless model checking}.
+	\item {\bf Nondeterminism.}
+		Property of a program which can exhibit different behaviours across multiple runs,
+		despite no change in the input data.
 	\item {\bf Optimality.} See {\em completeness}.
@@
 	\item {\bf Preemption point predicate.}
-		% TODO
+		A test of an intermediate execution state returning whether the {\em model checker}
+		should inject {\em preemptions} to force another thread to interleave.
+		Often simply ``is it the beginning of {\tt mutex\_lock()} or {\tt mutex\_unlock()}?'',
+		but may be extended to include the program counter values of known {\em data races},
+		or to further restrict by predicates on the current stack trace.
+		See \sect{\ref{sec:landslide-pps}}.
+		Often called {\em preemption point} for short, with context hopefully making the distinction clear.
 	\item {\bf Pure happens-before.}
@@
 	\item {\bf Reduction.}
-		The act of making a {\em state space} smaller by finding multiple execution sequences that can be proved equivalent,
+		The act of making a {\em state space} smaller by finding multiple execution sequences
+		that can be proved {\em equivalent},
 		such that all but one may be skipped over.
@@
 	\item {\bf Stateless model checking.}
-		Concurrency testing {\em dynamic analysis} technique involving controlling a system's nondeterminism
-		to exhaustively check all possible thread interleavings.
+		Concurrency testing {\em dynamic analysis} technique involving controlling a system's {\em nondeterminism}
+		to exhaustively check all possible {\em thread interleavings}.
 		See \sect{\ref{sec:overview-stateless}}.
@@
 	\item {\bf State space.}
-		% TODO
+		The set of all possible behaviours of a program under any interleaving of threads.
+		As opposed to modern {\em stateless model checkers}, older MCs stored the set of states explicitly,
+		attempting to cover them by traversing as few edges (executing as few {\em transitions}) as possible;
+		hence the name.
+		Modern MCs instead cover these states implicitly by instead exhaustively executing all {\em interleavings},
+		using {\em DPOR} to prune redundant {\em equivalent} ones which could only revisit old states.
+		The slightly misleading name is kept for consistency with prior work (sorry).
 	\item {\bf Static analysis.}
@@
 		Another name for {\em stateless model checking}.
-	\item {\bf TA.} See {\em teaching assistant}.
+	\item {\bf TA.} {\em Teaching assistant}.
 	\item {\bf Teaching assistant.}
@@
 		A sequence of execution steps by a single thread,
-		characterized by the two different program states at beginning and end.
+		bounded by the two different program states at beginning and end.
 		In this thesis, will be used exclusively when the terminal program states are {\em preemption points},

### 2018-09-26 thesis/related.tex
trim related

@@
 implemented using the OpenTM interface \cite{opentm},
-used by many papers in the field to evaluate the performance of both STM and HTM implementations alike,
-although as discussed in \sect{\ref{sec:tm-eval-exp-setup}},
-focuses more on performance than on interesting concurrency properties.
+used by many papers in the field to evaluate the performance of both STM and HTM implementations alike.
+As discussed in \sect{\ref{sec:tm-eval-exp-setup}},
+it focuses more on performance than on interesting concurrency properties.
 % stamp bugfixes - https://github.com/mfs409/transmem/blob/master/benchmarks/stamp_c/VERSIONS.
@@
 were constructed under a programming model poorly-suited to fully take advantage of HTM's performance,
-and that scalable HTM programs should
-seek to minimize incidental conflicts and to handle aborts more flexibly than with blind retry loops.
+and that scalable HTM programs must
+minimize incidental conflicts and handle aborts more flexibly than with blind retry loops.
 The programming complexity needed to achieve these goals calls,
@@
 relying on conflict aborts to point out access pairs that would be unsafe in the original program.
-This citation perhaps half belongs in \sect{\ref{sec:related-data-race}} as well;
+This citation arguably belongs in \sect{\ref{sec:related-data-race}} as well;
 I include it here to highlight the importance of Landslide's

### 2018-09-27 irb/landslide-guide-f18.tex
copy lecture for f18; make it non-research; add non-research user guide

--- 2018-01-09 irb/landslide-guide-p2.tex
@@
 
-\section{Introduction}
+\section{Preface}
 
-You should have received a document titled {\em User Study Information Sheet} that describes your rights as a study
-participant and our confidentiality procedures. If you are missing this, please contact me immediately. My email
-address is {\tt bblum@cs.cmu.edu} and my phone number is 412-304-4294; email is preferred.
-\\
+%You should have received a document titled {\em User Study Information Sheet} that describes your rights as a study
+%participant and our confidentiality procedures. If you are missing this, please contact me immediately. My email
+%address is {\tt bblum@cs.cmu.edu} and my phone number is 412-304-4294; email is preferred.
+%\\
 
-\noindent Likewise, please make sure you have read and understand the lecture slides on systematic testing, which are
+Please make sure you have read and understand the lecture slides on systematic testing, which are
 available on the 15-410 course website. If you have any questions about the research background I would be
-happy to answer them.
+happy to answer them; you can get in touch at {\tt bblum@cs.cmu.edu}.
 
@@
 We encourage you to use Landslide on the GHC cluster or UNIX pool machines ({\tt ghcXX.ghc.andrew.cmu.edu}
-or {\tt unix.andrew.cmu.edu}). Landslide has not been tested with other configurations and will probably fail
-mysteriously on your personal laptop, and we cannot promise to be of any help.
+or {\tt unix.andrew.cmu.edu}).
+%Landslide has not been tested with other configurations and will probably fail
+%mysteriously on your personal laptop, and we cannot promise to be of any help.
+Since Landslide uses the open-source simulator Bochs rather than the proprietary Simics,
+you can also try it on any personal machine which runs Linux.
+(The Landslide repository includes the Bochs source; you do not need to install it separately.)
 
@@
 			\item {\tt PATH} should be the path with {\tt 410kern}, {\tt 410user}, {\tt user/libthread}, etc as subdirectories, not the {\tt user/libthread} directory itself.
+			\item This step may take a few minutes, as it needs to compile the Bochs simulator from source.
 		\end{itemize}
@@
 time, and -p , which test program to run.
-			\item Landslide's bug reports will show up as HTML files.
-				\begin{itemize}
-					\item If you are using a cluster machine in person, you can view them by pointing a browser to \\
-						{\tt file:///PATH/TO/landslide/FILENAME.html}
-					\item For remote usage, move the html file to {\tt ~/www} in your AFS, and view it at \\
-						{\tt http://www.contrib.andrew.cmu.edu/~YOUR\_ANDREWID/FILENAME.html}.
-						Note the www in the URL is mandatory;
-						note also that this will allow the world to see it,
-						so be careful to clean up any traces in {\tt ~/www} when you're done!
-				\end{itemize}
 			\item Suggested test configurations to run are as follows.
 				You can, of course, change the time limits or test programs as you see fit.
-				For the longer-running configurations,
-				we recommend running them inside of a {\tt screen} or {\tt tmux} session.
 				\begin{itemize}
@@
 						\\
-					\item \texttt{./landslide -t 3h -p thr\_exit\_join}
-					\item \texttt{./landslide -t 3h -p mutex\_test}
-					\item \texttt{./landslide -t 3h -p paradise\_lost}
-					\item \texttt{./landslide -t 3h -p broadcast\_test}
-					\item \texttt{./landslide -t 3h -p paraguay}
-					\item \texttt{./landslide -t 3h -p rwlock\_downgrade\_read\_test}
-						\\
+					%\item \texttt{./landslide -t 3h -p thr\_exit\_join}
+					%\item \texttt{./landslide -t 3h -p mutex\_test}
+					%\item \texttt{./landslide -t 3h -p paradise\_lost}
+					%\item \texttt{./landslide -t 3h -p broadcast\_test}
+					%\item \texttt{./landslide -t 3h -p paraguay}
+					%\item \texttt{./landslide -t 3h -p rwlock\_downgrade\_read\_test}
+					%	\\
 					\item \texttt{./landslide -t 12h -p thr\_exit\_join}
@@
 					\item \texttt{./landslide -t 12h -p rwlock\_downgrade\_read\_test}
+				\end{itemize}
+				Please check slide 34 of the lecture before attempting to run {\tt cyclone}, {\tt racer}, {\tt largetest}, etc.
+
+				For the longer-running configurations,
+				we recommend running them inside of a {\tt screen} or {\tt tmux} session.
+				The GHC and UNIX cluster machines reboot nightly, so using a personal machine is recommended
+				for running tests overnight.
+			\item Landslide's bug reports will show up as HTML files.
+				\begin{itemize}
+					\item If you are using a cluster machine in person, or a personal machine,
+						you can view them by pointing a browser to \\
+						{\tt file:///PATH/TO/landslide/FILENAME.html}
+					\item For remote usage, move the html file to {\tt ~/www} in your AFS, and view it at \\
+						{\tt http://www.contrib.andrew.cmu.edu/~YOUR\_ANDREWID/FILENAME.html}.
+						Note the www in the URL is mandatory;
+						note also that this will allow the world to see it,
+						so be careful to clean up any traces in {\tt ~/www} when you're done!
 				\end{itemize}
@@
 
-\noindent {\bf If something goes wrong:} If you find a Landslide bug, such as an assertion/crash, or such as a
+\section{Troubleshooting}
+
+If you find a Landslide bug, such as an assertion/crash, or such as a
 bug report that you think is actually a bug in Landslide rather than a bug in your P2, or such as it getting stuck
@@
 
-\noindent However, we cannot provide help on how to understand why your bug happened or how to fix it. The point of
-the study is that Landslide's automation plus your own brain should be enough! On the other hand, if you have
+\noindent
+This is the first semester we have released Landslide with Bochs instead of Simics,
+so there may be any manner of unexpected bugs (it's fairly well tested on my own laptop, though!).
+If you are trying Landslide on a personal machine and have problems with the setup step,
+please try again on one of the official CMU servers before sending me a bug report.
+If all else fails, you may try the known-good Simics version:
+\url{https://github.com/bblum/landslide-simics}
+(which requires a GHC or UNIX cluster machine for the software license).
+\\
+
+\noindent Finally, we cannot provide help on how to understand why your bug happened or how to fix it. The point of
+the research is that Landslide's automation plus your own brain should be enough! On the other hand, if you have
 a design question while deciding between potential ways to fix a bug, send it to 410 course staff as usual.

### 2018-09-27 landslide-lecture/2018f/lecture.tex
copy lecture for f18; make it non-research; add non-research user guide

--- 2018-04-04 landslide-lecture/2018/lecture.tex
@@
 \begin{document}
-\renewcommand{\inserttotalframenumber}{39}
+\renewcommand{\inserttotalframenumber}{38}
 \normalem
@@
 	\begin{itemize}
-		\item Simics module, which traces:
+		%\item Simics module, which traces:
+		\item Simulator* module, which traces:
 			\begin{itemize}
@@
 				\item Every memory access read/written
+				\item *Used to be Simics, now Bochs (open-source, 3x faster)
 			\end{itemize}
@@
 		\item Reset machine state and start over
-		\item Implemented using Simics bookmarks
-			\begin{itemize}
-				\item {\tt set-bookmark} and {\tt skip-to}
-			\end{itemize}
+		% ehh.. don't feel like explaining
+		%\item Implemented using {\tt fork()}
+		%\item Implemented using Simics bookmarks
+		%	\begin{itemize}
+		%		\item {\tt set-bookmark} and {\tt skip-to}
+		%	\end{itemize}
 		\item Replay test from the beginning, with a different interleaving
@@
 \begin{frame}{Test Suite}
-	Landslide ships with 6 approved test cases:
-	\linegap
-
-	Standard P2 tests
+	6 of the official P2 tests are ``Landslide-friendly'':
 	\begin{itemize}
@@
 		\item {\tt rwlock\_downgrade\_read\_test}
-	\end{itemize}
-
-	New tests
-	\begin{itemize}
 		\item {\tt broadcast\_test}
@@
 
-\begin{frame}{User Study}
-	\textbf{Try Landslide on your P2!}
-	\begin{itemize}
-		\item Bare minimum effort: No more than 1 hour
+\begin{frame}{Try Landslide on your P2!}
+	Bare minimum effort: No more than 1 hour
 	\begin{itemize}
@@
 	\end{itemize}
-		\item Full study plan: 4-8 hours of active attention
+	\linegap
+
+	Full study plan: 4-8 hours of active attention
 	\begin{itemize}
@@
 	\end{itemize}
-	\end{itemize}
 	\pause
@@
 
-\begin{frame}{User Study - Additional Information}
-	Human Subjects Research
-	\begin{itemize}
-		\item CMU IRB has approved this study
-		\item Landslide will collect results while you use it
-		\begin{itemize}
-			\item Record commands issued, take snapshots of your P2 code
-			\item All data will be anonymized before publication
-		\end{itemize}
-		\item No coercion: {\bf \em There is no penalty for not participating.}
-		\begin{itemize}
-			\item I am not on course staff, cannot influence your grade
-			\item Course staff will not have access to study data during semester
-		\end{itemize}
-	\end{itemize}
-	\pause
-	\linegap
-
-	Risks \& Benefits
-	\begin{itemize}
-		\item Benefit: Landslide may help you find/fix bugs, improving your grade!
-		\item Risk: Landslide may find no bugs and be a waste of your time.
-		\item Benefit: You might learn something...
-	\end{itemize}
-	% Say: Accordingly, although I hope you will use landslide enough to get good results for me, you should also responsibly allocate time between testing with landslide versus actually implementing your P2.
-\end{frame}
-
-\begin{frame}{User Study - How to Participate}
+%\begin{frame}{User Study - Additional Information}
+%	\sout{Human Subjects Research} not anymore!
+%	\begin{itemize}
+%		\item CMU IRB has approved this study
+%		\item Landslide will collect results while you use it
+%		\begin{itemize}
+%			\item Record commands issued, take snapshots of your P2 code
+%			\item All data will be anonymized before publication
+%		\end{itemize}
+%		\item No coercion: {\bf \em There is no penalty for not participating.}
+%		\begin{itemize}
+%			\item I am not on course staff, cannot influence your grade
+%			\item Course staff will not have access to study data during semester
+%		\end{itemize}
+%	\end{itemize}
+%	\pause
+%	\linegap
+%
+%	Risks \& Benefits
+%	\begin{itemize}
+%		\item Benefit: Landslide may help you find/fix bugs, improving your grade!
+%		\item Risk: Landslide may find no bugs and be a waste of your time.
+%		\item Benefit: You might learn something...
+%	\end{itemize}
+%	% Say: Accordingly, although I hope you will use landslide enough to get good results for me, you should also responsibly allocate time between testing with landslide versus actually implementing your P2.
+%\end{frame}
+
+\begin{frame}{Try Landslide on your P2!}
 	Interested?
@@
 		\item Meet prerequisites of passing P2 tests
-		\item Complete sign-up form online to get further instructions
-			\begin{itemize}
-				\item (watch your email for the link)
+		\item Read \& follow user guide instructions
+			\begin{itemize}
+				\item \url{http://www.contrib.andrew.cmu.edu/~bblum/landslide-guide-f18.pdf}
+				\item (this will also be emailed to you)
 				%\item \url{http://tinyurl.com/landslide-p2-s17}
@@
 	\end{itemize}
-	%% im out of time :<<
-	%\linegap
-
-	%{\em Really} interested?
-	%\begin{itemize}
-	%	\item I know of an M.S. thesis topic if you want to do post-OS research.
-	%\end{itemize}
 \end{frame}

### 2018-09-27 landslide-lecture/2018f/lecture.tex
study and survey resulce

@@
 \begin{document}
-\renewcommand{\inserttotalframenumber}{38}
+\renewcommand{\inserttotalframenumber}{39}
 \normalem
@@
 	%% S'15, F'15, S'16, F'16, S'17: 79 groups signed up to use Landslide; 59 found bugs
-	S'15, F'15, S'16, F'16, S'17, F'17: 94 groups signed up to use Landslide; 71 found bugs
-	\linegap
-
-	%%44 {\em deterministic} bugs (e.g. swexn, initialization)
-	%% F17 lecture - stats up to S17
-	%% 91 {\em deterministic} bugs (e.g. swexn, initialization)
-	% S18 lecture - stats up to F17
-	109 {\em deterministic} bugs (e.g. swexn, initialization)
-	\linegap
-
-	%%85 distinct {\em non-deterministic} bugs (among 32 groups)
-	%% F17 lecture - stats up to S17
-	%% 113 distinct {\em non-deterministic} bugs (among 44 groups)
-	% S18 lecture - stats up to F17
-	122 distinct {\em non-deterministic} bugs (among 49 groups)
-	\begin{itemize}
-		\item 36 groups (73\%) fixed $\ge 1$ such bug
-		%\item 33 groups (75\%) fixed $\ge 1$ such bug
-		%%\item 24 groups (75\%) fixed $\ge 1$ such bug
+	%% S'15, F'15, S'16, F'16, S'17, F'17: 94 groups signed up to use Landslide; 75 found bugs % not sure why 75 -> 75
+	%%S'15, F'15, S'16, F'16, S'17, F'17, S'18: 103 groups signed up to use Landslide; 75 found bugs
+	S'15--S'18: 103 groups signed up to use Landslide; 75 found bugs
+	\linegap
+
+	% all this now from table 5.2 in thesis
+	115 {\em deterministic} bugs (e.g. swexn, initialization)
+	\begin{itemize}
+		\item 87 (76\%) of these were fixed before submitting P2
 			\begin{itemize}
@@
 			\end{itemize}
-		\item 26 groups (53\%) fixed {\em all} such bugs
-		%\item 24 groups (55\%) fixed {\em all} such bugs
-		%%\item 15 groups (47\%) fixed {\em all} such bugs
-		\item Most ambitious group: 11 distinct races found + fixed!
-	\end{itemize}
-	\linegap
+	\end{itemize}
+	\linegap
+
+	122 distinct {\em non-deterministic} bugs
+	\begin{itemize}
+		\item 80 (66\%) of these were fixed before submitting P2
+			\begin{itemize}
+				\item (verified by running Landslide again, same)
+			\end{itemize}
+	\end{itemize}
+	\linegap
+
+	75 groups in total found any bugs
+	\begin{itemize}
+		\item 58 (77\%) groups fixed at least 1 bug
+		\item 35 (47\%) groups fixed all found bugs
+		\item Most ambitious group: 11 bugs found \& fixed!
+	\end{itemize}
+\end{frame}
+
+\begin{frame}{Survey Results}
+	F'17, S'18: Asked students what they thought of it all
+	\begin{itemize}
+		\item (also ran similar study at Penn State U. \& U. of Chicago)
+	\end{itemize}
+
+	\begin{center}
+	\includegraphics[width=0.8\textwidth]{survey6.pdf}
+	\end{center}
+\end{frame}
+
+\begin{frame}{Survey Results}
+	\addtocounter{framenumber}{-1}
+	F'17, S'18: Asked students what they thought of it all
+	\begin{itemize}
+		\item (also ran similar study at Penn State U. \& U. of Chicago)
+	\end{itemize}
+
+	\begin{center}
+	\includegraphics[width=0.8\textwidth]{survey9.pdf}
+	\end{center}
+\end{frame}
+
+\begin{frame}{Survey Results}
+	\addtocounter{framenumber}{-1}
+	F'17, S'18: Asked students what they thought of it all
+	\begin{itemize}
+		\item (also ran similar study at Penn State U. \& U. of Chicago)
+	\end{itemize}
+
+	\begin{center}
+	\includegraphics[width=0.8\textwidth]{survey10.pdf}
+	\end{center}
 \end{frame}
@@
 
+\begin{frame}{More Survey Results}
+	(To dispel any accusations that I might have omitted survey questions with less good results)
+	\linegap
+
+	\begin{center}
+	\includegraphics[width=0.8\textwidth]{survey1.pdf}
+	\end{center}
+\end{frame}
+
+\begin{frame}{More Survey Results}
+	(To dispel any accusations that I might have omitted survey questions with less good results)
+	\linegap
+
+	\begin{center}
+	\includegraphics[width=0.8\textwidth]{survey2.pdf}
+	\end{center}
+\end{frame}
+
+\begin{frame}{More Survey Results}
+	(To dispel any accusations that I might have omitted survey questions with less good results)
+	\linegap
+
+	\begin{center}
+	\includegraphics[width=0.8\textwidth]{survey3.pdf}
+	\end{center}
+\end{frame}
+
+\begin{frame}{More Survey Results}
+	(To dispel any accusations that I might have omitted survey questions with less good results)
+	\linegap
+
+	\begin{center}
+	\includegraphics[width=0.8\textwidth]{survey4.pdf}
+	\end{center}
+\end{frame}
+
+\begin{frame}{More Survey Results}
+	(To dispel any accusations that I might have omitted survey questions with less good results)
+	\linegap
+
+	\begin{center}
+	\includegraphics[width=0.8\textwidth]{survey5.pdf}
+	\end{center}
+\end{frame}
+
+\begin{frame}{More Survey Results}
+	(To dispel any accusations that I might have omitted survey questions with less good results)
+	\linegap
+
+	\begin{center}
+	\includegraphics[width=0.8\textwidth]{survey7.pdf}
+	\end{center}
+\end{frame}
+
+\begin{frame}{More Survey Results}
+	(To dispel any accusations that I might have omitted survey questions with less good results)
+	\linegap
+
+	\begin{center}
+	\includegraphics[width=0.8\textwidth]{survey8.pdf}
+	\end{center}
+\end{frame}
+
 \end{document}

### 2018-09-29 thesis/thesis.tex
more future work musings
write more acks

@@
 \begin{dedication}
-%For Elsie, Margie, Johnny, Mel, Eve, and David.
-%%, who brought me here today.
 For my family, my teachers, and my students.
@@
 \pagestyle{plain} % for toc, was empty
-
-%% Obviously, it's probably a good idea to break the various sections of your thesis
-%% into different files and input them into this file...
 
@@
 \begin{acknowledgments}
-Even as I built Landslide by myself,
-implementation is of course barely the first step of scientific contribution.
-I would honor here the many people who supported me during these 7 years and made this thesis possible.
+I honor here the many people who supported me during these 7 years and made this journey possible.
 
@@
 
-% garth,
+Professor Garth Gibson is largely responsible for shaping me into the researcher I am today.
+%Thanks to Garth's guidance,
+%this thesis has grown into something that past-me,
+%first starting research with the simple vision ``make a better autograder for 410'',
+%could not have then imagined.
+Despite having primary research interests in other fields,
+Garth enthusiastically took on the project,
+and always pushed me to explore new problem domains,
+to design scientifically thorough experiments,
+to write prose explaining difficult concepts approachably,
+to appreciate related work as charitably as possible,
+and to seek guidance from industry people with relevant experience,
+despite my often stubborn refusal to leave my little comfort zone.
+So, even when it took me sending 10+ email reminders to get your attention,
+truly,
+thanks for everything, Garth.
 
@@
 
-% jiri,
-% .. orient me in the world of research ..
+Ji\v{r}\'{i} \v{S}im\v{s}a, another student of Garth's,
+first sparked my interest in the field,
+and then gave invaluable support during my first few years.
+Ji\v{r}\'{i}'s mentorship
+oriented me in the world of research and helped me find my bearings as a grad student,
+from teaching me how DPOR works
+to helping revise my paper drafts
+to offering the kind of candid career advice you can't get from a professor.
+Had I not met Ji\v{r}\'{i} and learned about dBug during grad school applications way back,
+none of this would have happened.
 
-% these 3 names ar ethe ones without which the thesis flat out could not have happened
+% FIXME - cmon latex pls??
+\pagebreak
 
@@
 
-Without SIGBOVIK's unique brand of humorous, self-aware, and {\em definitely legitimate} research,
-I probably would not have been interested in grad school in the first place,
-or perhaps would have grown too frustrated for lack of a cathartic outlet for conference paper woes.
-I thank all who've worn the mantle of Harry Bovik over the years,
-from the conference's original founders,
-to every paper author,
-to all future organizers who'll keep those ``mainstream'' conferences on their toes well into the future.
+The community of 15-410 TAs,
+having surrounded me pretty much since starting undergrad,
+nurtured my love of teaching
+well before taking the class myself and well after graduation.
+As skilled as the instructor is,
+15-410 could not be the world's best operating systems class
+without TAs who value thorough grading, the Socratic method,
+and empathy to struggling students as much
+from year to year as the crew I have known.
+It was a blessing to serve, and later do research, among you all.
+I am also grateful to Adam Blank and Miriam Zimmerman
+for founding and organizing the class Great Practical Ideas,
+which teaches oft-neglected programming skills such as version control and debugging,
+which I had the opportunity to teach as head instructor one semester.
 
-% [ ] 410 ta community
-% [ ] emarhavil
-% [ ] pdl consortium industry ppl who offered good advices
-% [ ] anyone who ever said how cool it was that i helped students for my research
-% [ ] adam and co. for GPI, if you feel like it
-
-% blurry line
+I thank the Parallel Data Lab, my research group at CMU,
+for supporting me at the PDL retreat every fall,
+pushing me to refine my presentation skills and to network with strangers (now familiar faces) from industry.
+I also thank the retreat's industry attendees from each of the PDL Consortium member companies
+for their interest even in more theoretical work
+and the enlightening conversations resulting therefrom.
+I thank everyone outside my immediate research circle who,
+whether at the retreat or otherwise,
+told me how cool they thought my research was for directly helping students --
+every such reminder helped me keep my head up during darker times.
 
@@
 you each have been role models to me
-%each in your own way
 perhaps more than you know.
+
+Without SIGBOVIK's unique brand of humorous, self-aware, and {\em definitely legitimate} research,
+I probably would not have been interested in grad school in the first place,
+or perhaps would have grown too frustrated along the way for lack of a cathartic outlet for conference paper woes.
+I thank all who've worn the mantle of Harry Bovik over the years,
+from the conference's original founders,
+to every paper author,
+to all future organizers who'll keep those ``mainstream'' conferences on their toes well into the future.
 
@@
 for nearly as long as I've been a grad student.
+Netrunner is a tabletop strategy deckbuilding card game
+that depicts cyberspace hacking battles (completely different from real-life computer security)
+alongside dystopian social issues (some of which seem oddly more familiar with each passing year).
+Its characters have a diversity of ethnicity, gender identity, and sexual orientation
+unparalleled in any other game I know of,
+and the community that has grown around it reflects those values.
 I thank in particular the other Stimhack moderators,
@@
 %\footnote{and Slack emojis},
-that I hope will well outlive the card game that brought us together.
+that I hope will long outlive the silly card game that brought us together.
 %It'd be an injustice to list only a few names here and omit others, but you know who you are.
@@
 % estimator could account for this somehow... not sure how but
+% likewise, estimator could account for retry sets by checking the size of the retry set and multiplying by 1/4 or 1/2 or w/e
 % TODO: cite smth about machine learning
@@
 % if your ML provides a partial verification, it is impossible for a human to understand what that verification means
+
+% ICB incorporations
+% into quicksand - when SS estimates are too high, fall back on a preemption bounded version to find bugs - improve the partial verif
+% into education - provide more partial verifs to students (but is it clear they'll understand the subtleties of it?)
+% into txn - incorporate failure injection into the model
+
+% soundness concerns - the xbegin/delay-xbegin pp soundness bug popped up in the implementation, despite the soundness thm
+% current SoA for this is to manually inspect the state spaces that arize for small K,N and just hope...
+% as concurrency model becomes more complex, more opportunities for the implementation to deviate from what
+% the soundness theorem actually proves
+% how to prove landslide is actually doing what you think? analogous to sel4 spec/impl verif...
 

### 2018-09-30 thesis/future.tex
welcome to warp zone

--- 2018-09-01 thesis/thesis.tex
@@
-%for a more compact document, add the option openany to avoid
-%starting all chapters on odd numbered pages
-\documentclass[12pt]{cmuthesis}
-
-% This is a template for a CMU thesis.  It is 18 pages without any content :-)
-% The source for this is pulled from a variety of sources and people.
-% Here's a partial list of people who may or may have not contributed:
-%
-%        bnoble   = Brian Noble
-%        caruana  = Rich Caruana
-%        colohan  = Chris Colohan
-%        comar    = Cyrus Omar
-%        jab      = Justin Boyan
-%        josullvn = Joseph O'Sullivan
-%        jrs      = Jonathan Shewchuk
-%        kosak    = Corey Kosak
-%        mjz      = Matt Zekauskas (mattz@cs)
-%        pdinda   = Peter Dinda
-%        pfr      = Patrick Riley
-%        dkoes = David Koes (me)
-
-% My main contribution is putting everything into a single class files and small
-% template since I prefer this to some complicated sprawling directory tree with
-% makefiles.
-
-% some useful packages
-\usepackage{fullpage}
-\usepackage{graphicx}
-\usepackage{amsmath}
-\usepackage{amssymb}
-\usepackage{amsthm}
-\usepackage{mdwlist}
-\usepackage[numbers,sort]{natbib}
-\PassOptionsToPackage{hyphens}{url}
-\usepackage[backref,pageanchor=true,plainpages=false, pdfpagelabels, bookmarks,bookmarksnumbered,
-%pdfborder=0 0 0,  %removes outlines around hyper links in online display
-]{hyperref}
-\usepackage{subfigure}
-\usepackage{xspace}
-\usepackage[linesnumbered,ruled]{algorithm2e}
-\usepackage{algpseudocode}
-\usepackage{microtype}
-
-%% custom fonts
-
-\usepackage{fontspec}
-\usepackage{xeCJK}
-
-\setmainfont{XCharter}
-%\setmainfont{Utopia}[
-%	Path = ./fonts/,
-%	UprightFont = UtopiaStd-Regular.otf,
-%	BoldFont = UtopiaStd-Semibold.otf,
-%	ItalicFont = UtopiaStd-Italic.otf,
-%	BoldItalicFont = UtopiaStd-SemiboldIt.otf
-%	]
-
-\setCJKmainfont{HeiseiMinStd-W5}[Path = ./fonts/]
-
-% Approximately 1" margins, more space on binding side
-%\usepackage[letterpaper,twoside,vscale=.8,hscale=.75,nomarginpar]{geometry}
-%for general printing (not binding)
-\usepackage[letterpaper,twoside,vscale=.8,hscale=.75,nomarginpar,hmarginratio=1:1]{geometry}
-
-% Provides a draft mark at the top of the document.
-\draftstamp{\today}{DRAFT}
-
-% % nwf made this but i don't know how to get it to work?
-% \usepackage{etoolbox}
-% \makeatletter
-% \apptocmd{\ref}{%
-%   \immediate\write\@auxout%
-%     {\expandafter\string\csname nwf@backref\endcsname{#1}{\arabic{page}}}%
-% }{}{}
-%
-% \newcommand{\nwf@backref}[2]{%
-%   \ifcsdef{nwf@backrefs@#1}%
-%    {\csgappto{nwf@backrefs@#1}{, #2}}%
-%    {\csgdef{nwf@backrefs@#1}{#2}}%
-% }
-% \AtEndDocument{\let\nwf@backref\relax}
-%
-% \newcommand{\linksto}[1]{\csname nwf@backrefs@#1\endcsname}
-% \makeatother
-
-\begin{document}
-\frontmatter
-
-%initialize page style, so contents come out right (see bot) -mjz
-\pagestyle{empty}
-
-\title{ %% {\it \huge Thesis Proposal}\\
-{\bf Practical Concurrency Testing} \\
-\normalsize \vspace{1em}
-\begin{tabular}{c}
-{\em or: How I Learned to Stop Worrying and Love the Exponential Explosion}
-\end{tabular}}
-\author{Ben Blum}
-\date{Decembruary 2018}
-\Year{2018}
-\trnumber{}
-
-\committee{
-Garth Gibson, Chair \\
-David A. Eckhardt \\
-Brandon Lucia \\
-Haryadi Gunawi, University of Chicago
-}
-
-\support{}
-\disclaimer{}
-
-% copyright notice generated automatically from Year and author.
-% permission added if \permission{} given.
-
-\keywords{landslide terminal, baggage claim, ground transportation, ticketing}
-
-\maketitle
-
-\begin{dedication}
-%For Elsie, Margie, Johnny, Mel, Eve, and David.
-%%, who brought me here today.
-For my family, my teachers, and my students.
-\end{dedication}
-
-\pagestyle{plain} % for toc, was empty
-
-%% Obviously, it's probably a good idea to break the various sections of your thesis
-%% into different files and input them into this file...
-
-\begin{abstract}
-Concurrent programming presents a challenge to students and experts alike because of the complexity of multithreaded interactions and the difficulty to reproduce and reason about bugs.
-Stateless model checking is a concurrency testing approach which forces a program to interleave its threads in many different ways, checking for bugs each time.
-This technique is powerful, in principle capable of finding any nondeterministic bug in finite time,
-but suffers from exponential explosion as program size increases.
-Checking an exponential number of thread interleavings is not a practical or predictable approach for programmers to find concurrency bugs before their project deadlines.
-
-In this thesis, I develop several new techniques to make stateless model checking more practical for human use.
-I have built Landslide, a stateless model checker specializing in undergraduate operating systems class projects.
-Landslide extends the traditional model checking algorithm with a new framework for automatically managing
-%the exploration of
-multiple state spaces according to their estimated completion times,
-which I show quickly finds bugs should they exist and also quickly verifies correctness otherwise.
-I evaluate Landslide's suitability for inexpert use by presenting the results of many semesters providing it to students in 15-410, CMU's Operating System Design and Implementation class,
-and more recently, students in similar classes at the University of Chicago and Penn State University.
-%Finally, I will present several new techniques that allow stateless model checking to be practically employed on real-world programs.
-%Finally, I will explore broader impact by extending Landslide to test some real-world programs and to be used by students at other universities.
-Finally, I extend Landslide with a new concurrency model for hardware transactional memory,
-and evaluate several real-world transactional benchmarks to show that stateless model checking can keep up with the developing concurrency demands of real-world programs.
-\end{abstract}
-
-\begin{acknowledgments}
-Even as I built Landslide by myself,
-implementation is of course barely the first step of scientific contribution.
-I would honor here the many people who supported me during these 7 years and made this thesis possible.
-
-\subsubsection{Thesis}
-
-% garth,
-
-Professor David Eckhardt
-provided unending guidance on education, writing, life, you name it,
-helped revise each semester's iteration of the recruiting lecture with a keen eye for the student mind,
-contributed substantially to the design of the survey,
-and supplied constant enthusiasm and reassurance that Landslide was a boon to 15-410 that kept me going.
-I can't express the depth of my gratitude to Dave strongly enough,
-even though three of the Landslide-friendly tests never made it upstream into the official P2 test suite.
-
-% jiri,
-% .. orient me in the world of research ..
-
-% these 3 names ar ethe ones without which the thesis flat out could not have happened
-
-Many others have collaborated with me directly on the research itself,
-whom I thank here in approximate chapter order.
-Michael J. Sullivan and Nathaniel Filardo were Landslide's first users besides myself back in its prototype days.
-Joshua Wise, Miriam Zimmerman, and Glenn Willen helped revise my explanation of DPOR in Chapter~\ref{chap:background}.
-Professor Brandon Lucia provided valuable comments and revisions to Chapter~\ref{chap:quicksand}'s content
-that ultimately led to its publication at OOPSLA 2016.
-Michael J. Sullivan double-checked the proofs therein.
-Professor Timothy Zhu from Penn State University
-and Professor Haryadi Gunawi and TA Kevin Zhao from the University of Chicago
-graciously allowed me to use their students as guinea pigs
-and helped greatly to improve Landslide's stability and robustness for use beyond CMU's walls,
-greatly enriching Chapter~\ref{chap:education}'s evaluation.
-David Blum and Skye Toor provided invaluable advice on statistical analysis
-that helped me evaluate the student grade distributions therein.
-Mario Dehesa-Azuara generously contributed the transactional memory data structures
-that make up about half of Chapter~\ref{chap:tm}'s evaluation suite.
-
-\subsubsection{Academia}
-
-Without SIGBOVIK's unique brand of humorous, self-aware, and {\em definitely legitimate} research,
-I probably would not have been interested in grad school in the first place,
-or perhaps would have grown too frustrated for lack of a cathartic outlet for conference paper woes.
-I thank all who've worn the mantle of Harry Bovik over the years,
-from the conference's original founders,
-to every paper author,
-to all future organizers who'll keep those ``mainstream'' conferences on their toes well into the future.
-
-% [ ] 410 ta community
-% [ ] emarhavil
-% [ ] pdl consortium industry ppl who offered good advices
-% [ ] anyone who ever said how cool it was that i helped students for my research
-% [ ] adam and co. for GPI, if you feel like it
-
-% blurry line
-
-I am grateful to the ThursDz Council,
-my group of now-former grad student, Shadyside-bound (in spirit if not in person) restaurant-hoppers,
-for both friendship and mentorship over the years.
-In particular,
-%in alphabetical for lack of a better order,
-Chris Martens,
-Jason Reed,
-Jim McCann,
-Rob Simmons,
-Tom Murphy 7,
-and
-William Lovas:
-you each have been role models to me
-%each in your own way
-perhaps more than you know.
-
-\subsubsection{Life}
-
-% [ ] csd bgames
-% [ ] undergrad frenz
-% [ ] new frenz
-% [ ] family, incl louie
-
-Lastly,
-the Android: Netrunner community has been such an important part of my life
-for nearly as long as I've been a grad student.
-I thank in particular the other Stimhack moderators,
-the members of {\sf \#rainbow-coalition},
-and the local Pittsburgh crew
-for their support and encouragement
-and friendships very real and deep, even when forged only in text,
-%\footnote{and Slack emojis},
-that I hope will well outlive the card game that brought us together.
-%It'd be an injustice to list only a few names here and omit others, but you know who you are.
-
-\end{acknowledgments}
-
-\tableofcontents
-% TODO: figure out how to make latex abbreviate these
-\listoffigures
-\listoftables
-
-\mainmatter
-
-%% Double space document for easy review:
-%\renewcommand{\baselinestretch}{1.66}\normalsize
-
-% The other requirements Catherine has:
-%
-%  - avoid large margins.  She wants the thesis to use fewer pages,
-%    especially if it requires colour printing.
-%
-%  - The thesis should be formatted for double-sided printing.  This
-%    means that all chapters, acknowledgements, table of contents, etc.
-%    should start on odd numbered (right facing) pages.
-%
-%  - You need to use the department standard tech report title page.  I
-%    have tried to ensure that the title page here conforms to this
-%    standard.
-%
-%  - Use a nice serif font, such as Times Roman.  Sans serif looks bad.
-%
-% Other than that, just make it look good...
-
-\newcommand\simics{\textsc{Simics}}
-\newcommand{\sect}[1]{\S #1}
-\newcommand\hilight[2]{\color{#1}#2\color{black}}
-\definecolor{orange}{RGB}{192,96,0}
-\definecolor{olivegreen}{RGB}{0,127,0}
-\definecolor{brickred}{RGB}{192,0,0}
-\definecolor{commentblue}{RGB}{0,0,192}
-% colors from sigbovik for tsx/etc
-% threads
-\definecolor{lavender}{RGB}{102,65,208} % (?? i don't know what happened here anymore but it looks good)
-\definecolor{seafoam}{RGB}{56,158,68} % * 2/3
-\definecolor{salmon}{RGB}{208,89,74} % * 7/8
-\definecolor{goldish}{RGB}{128,96,32}
-% eval data
-\definecolor{darkcyan}{RGB}{0,128,128}
-\definecolor{lime}{RGB}{48,128,0}
-\definecolor{darklavender}{RGB}{68,44,138}
-\definecolor{brownish}{RGB}{128,96,0}
-\definecolor{pinkish}{RGB}{128,34,102} % ff4488
-% syntax hilighting
-\newcommand\flow[1]{\hilight{brownish}{#1}\xspace}
-\newcommand\const[1]{\hilight{brickred}{#1}\xspace}
-\newcommand\call[1]{\hilight{pinkish}{#1}}
-\newcommand\ctype[1]{\hilight{lime}{#1}\xspace}
-\newcommand\ccomment[1]{\hilight{darkcyan}{#1}\xspace}
-
-\newtheorem{lemma}{Lemma}
-\newtheorem{theorem}{Theorem}
-\newtheorem{definition}{Definition}
-
-\newcommand\llama[1]{\includegraphics[width=#1]{llama.pdf}}
-\newcommand\llitem{\item[\raisebox{-0.15em}{\llama{1.2em}}]}
-
-\newcommand{\inspirationallinebreak}{\vspace{0.25em}}
-\newcommand{\inspirationalhyphen}{-\hspace{-0.15em}-\xspace}
-\newcommand\inspirationalquote[2]{\begin{flushright}
-	\vspace{-1em}
-	{
-	\fontspec[Path=./fonts/]{AlexaStd}
-	{\em #1}
-	\inspirationallinebreak
-
-	-\hspace{-0.15em}-\hspace{-0.15em}-{#2}
-	}
-	\vspace{2em}
-\end{flushright}}
-
-\input{intro}
-\input{overview}
-\input{landslide}
-\input{quicksand}
-\input{education}
-\input{tm}
-\input{related}
-
 \chapter{Future Work}
@@
 {Sam "Poem\_for\_your\_sprog" Garland}
+
+Each of the previous three chapters concluded by
+discussing specific limitations,
+listing concrete and immediate ways to address them
+with existing techniques
+(\sect{\ref{sec:quicksand-discussion}},
+\sect{\ref{sec:education-discussion}},
+\sect{\ref{sec:tm-discussion}}).
+Meanwhile, this chapter takes a broader interpretation of ``future work'',
+namely,
+how might future Landslides solve research problems I didn't even pose to begin with.
+%The reader looking for their own thesis topic would most likely find it here.
+
+\subsubsection{User-friendliness}
+
+In the user study survey (\sect{\ref{sec:education-eval-survey}}),
+students most often complained
+that interpreting Landslide's preemption traces to diagnose and understand their bugs was too difficult.
+While the understanding step of course requires human intuition,
+there is certainly room to improve the diagnosis step beyond just showing the user one static HTML table.
+Related works like Symbiosis \cite{symbiosis} can find a minimal difference between the failing trace and a non-failing one,
+which would allow the user to effortlessly pinpoint which preemptions are closest to the true root cause.
+Further, using ICB \cite{chess-icb} to show the user a minimum set of preemptions necessary to expose the bug
+could help her narrow down possible diagnoses more quickly.
+Finally, the preemption trace itself could be interactive,
+allowing the user to click and drag to reorder thread transitions and see how the interleaving would change,
+or to click and drag preemption points across the source code to figure out how much need to be atomic.
+
+State space size management remains an issue, as ever.
+While Quicksand's professed selling point is that the user need input only a CPU budget,
+at the same time,
+pruning out the overall state space would allow Quicksand to achieve more meaningful results in that same budget.
+Theoretical advances in state space reduction come out every year (\sect{\ref{sec:related-algs}}),
+but the user's human intuition can also contribute if properly encouraged.
+For the P2 tests (\sect{\ref{sec:education-pebbles-tests}}),
+I configured Quicksand by hand to issue appropriate {\tt without\_function} commands to Landslide
+(\sect{\ref{sec:landslide-pps}}),
+and even more still for the HTM tests (\sect{\ref{sec:tm-eval-exp-setup}}),
+but a user writing her own tests would have to configure DPOR by hand.
+A more mature version would coach the user to decide which functions focus the test on,
+using state space estimation to give an idea of expected testing time,
+and listing DPOR's memory conflicts to help her identify more candidate functions to potentially ignore.
+% there are a LOT of "hmm"s here for future research to address
+% most basically like, a modification could have a side effect that would lead to changed behaviour
+% arising from a preemption in a *subsequent* function
+% so really this can't be limited to just PPs in the changed functions, but dataflow arising therefrom as well
+Finally, Landslide could integrate with a version control system to do incremental testing,
+automatically analyzing the functions touched by each patch,
+and heuristically prioritizing preemption points therein
+to quickly check small updates on top of an already-verified codebase.
+
+\subsubsection{Test cases}
 
@@
 
-% things not to forget about
-% in the maximal state space, the "tag all evil ancestors" thing justified in my MS thesis figure 31337
-% is not necessary, bc if it's the maximal state space, you know that data race PPs would separate the accesses in C1/C2/C3.
-% but, actually figuring out if you're in the maximal state space on-the-fly to justify not tagging?
-% (but but, does this even matter, is this an optimization at all?)
-% i'm actually not even sure this would be sound even on maximal, because of RMW instructions, their 2 evence are inseparable
+\subsubsection{Verification}
+
+% soundness concerns - the xbegin/delay-xbegin pp soundness bug popped up in the implementation, despite the soundness thm
+% current SoA for this is to manually inspect the state spaces that arize for small K,N and just hope...
+% as concurrency model becomes more complex, more opportunities for the implementation to deviate from what
+% the soundness theorem actually proves
+% how to prove landslide is actually doing what you think? analogous to sel4 spec/impl verif...
+
+% extending landslide's verification for a limited nr of K N to infinite with induction/proof assistance
+
+% - this should prob go in quicksand chapater
+% identifying & skipping redundancies in smaller ID state spaces via memoization (same applies to ICB!)
+
+% arugment why ML is NOT appropriate for smc in general -- if you try to train it to choose thread interleavings / search ordering,
+% even if you're already in not-enough-cpu-time heuristics land, you're biasing it to find only the types of bugs you've already seen
+% and destroying its ability to find novelties
+% if your ML provides a partial verification, it is impossible for a human to understand what that verification means
+
+\subsubsection{Heuristics}
 
@@
 
-% arugment why ML is NOT appropriate for smc in general -- if you try to train it to choose thread interleavings / search ordering,
-% even if you're already in not-enough-cpu-time heuristics land, you're biasing it to find only the types of bugs you've already seen
-% and destroying its ability to find novelties
-% if your ML provides a partial verification, it is impossible for a human to understand what that verification means
+% ICB incorporations
+% into quicksand - when SS estimates are too high, fall back on a preemption bounded version to find bugs - improve the partial verif
+% into education - provide more partial verifs to students (but is it clear they'll understand the subtleties of it?)
+% into txn - incorporate failure injection into the model
 
-\chapter{Conclusion}
-\label{chap:conclusion}
-\inspirationalquote{
-	\begin{tabular}{p{0.84\textwidth}}
-	"If you are a god, Zeus, as the stories claim, then why did you create evolution?
-	Why did you make a world that can only grow through cruelty and pain?"
-	\\
-	For a moment\inspirationalhyphen{}surely this meant death was near\inspirationalhyphen{}she thought she heard him answer:
-	"My child, I was shaped by the gods that came before me, as you were shaped by me.
-	The choice I had was between creation and oblivion, life and death.
-	And I chose life, because any life is better than no life,
-	because as long as there is life, there is hope\inspirationalhyphen{}if not for us, then for some generation to come."
-	\\
-	"Then how are you a god, if you can offer so little?" she whispered, feeling death creep closer.
-	\\
-	"I am a god because I take upon myself the burden of creation," the statue replied.
-	\\
-	"Then we are all gods," Alexandra said, and pushed the button.
-	\end{tabular}}
-%{Galatea, The Talos Principle}
-{The Talos Principle}
+%%%%
 
-Concurrency testing is, in a way, a microcosm for research in general:
-an exponentially large problem
-% (literally and figuratively for each side of the analogy, respectively)
-that invariably thwarts they whoever try to solve it ``perfectly'';
-it bears fruit only to one who can find their own human-oriented meaning in partial results and compromises.
+% not gonna bother with this
 
-asdfasdasdf
-
-% TODO: Get a little bit preachy about how computer programs should be created to help humans,
-% not to look as much like scifi as possible, by knowing their limits, asking for help when they get confused, etc.
-
-% ways humans cooperate to make landslide better:
-% - test case design incl. preemption sealing
-% - abstraction reduction
-% - manual inspection of data races
-
-%\appendix
-%\include{appendix}
-
-\backmatter
-
-%\renewcommand{\baselinestretch}{1.0}\normalsize
-
-% By default \bibsection is \chapter*, but we really want this to show
-% up in the table of contents and pdf bookmarks.
-\renewcommand{\bibsection}{\chapter{\bibname}
-\inspirationalquote{Only a doctor of philosophy, Darth.}{Robert Marsh}
-}
-%\renewcommand{\bibpreamble}{This text goes between the ``Bibliography''
-%  header and the actual list of references}
-\bibliographystyle{plainnat} % plain n'at
-\bibliography{citations}
-
-\end{document}
+% in the maximal state space, the "tag all evil ancestors" thing justified in my MS thesis figure 31337
+% is not necessary, bc if it's the maximal state space, you know that data race PPs would separate the accesses in C1/C2/C3.
+% but, actually figuring out if you're in the maximal state space on-the-fly to justify not tagging?
+% (but but, does this even matter, is this an optimization at all?)
+% i'm actually not even sure this would be sound even on maximal, because of RMW instructions, their 2 evence are inseparable

### 2018-09-30 thesis/thesis.tex
welcome to warp zone

@@
 \input{related}
-
-\chapter{Future Work}
-\label{chap:warpzone}
-% TODO: ask for permission for this one
-% https://twitter.com/Poem4your_sprog/status/878747092788838400
-\inspirationalquote{
-\begin{tabular}{p{0.51\textwidth}}
-A little bit of being kind,
-a tiny open door; \\
-A nicer slice of peace of mind
-begets a little more. \\
-It won't correct the world tonight,
-nor change tomorrow too; \\
-But maybe if you do it right,
-you'll find it changes you.
-\end{tabular}}
-{Sam "Poem\_for\_your\_sprog" Garland}
-
-% test case design, writing code that actually checks the properties you're interested in
-% for example the mutex test, with the if (!xtest) yield check
-% then even after you got the right properties, optimizing the state space
-% with without_function, ignore_sym, etc, while preserving the guarantee
-
-% envision a framework that uses some code coverage tester to guide choice of pps and/or the user's without_function calls 
-% to get that coverage -- although relating flow control coverage directly to concurrency coverage is not straightforward
-% framework can also use 'git diff's, like comparing the most recent commit against an already-trusted codebase,
-% to figure out how to isolate just the stuff the change affects (memory intersection & data flow resulting from those conflix)
-% and then automatically assign some without_functions/etc to test just that commit
-
-% failure injection, testing malloc/etc failures
-% eg student condvars which loop around malloc failures can deadlock
-% can use retry set reduction in principle to prune the same kind of equivalences
-
-% also here we've got discussion of the exit-UAF buggers where landslide oops verifs the test anyway
-% because they're using a free pool and not free()
-% need a 3rd thread at least, to expose this (by actually getting corruption)
-
-% things not to forget about
-% in the maximal state space, the "tag all evil ancestors" thing justified in my MS thesis figure 31337
-% is not necessary, bc if it's the maximal state space, you know that data race PPs would separate the accesses in C1/C2/C3.
-% but, actually figuring out if you're in the maximal state space on-the-fly to justify not tagging?
-% (but but, does this even matter, is this an optimization at all?)
-% i'm actually not even sure this would be sound even on maximal, because of RMW instructions, their 2 evence are inseparable
-
-In the author's experience,
-the WBE algorithm (used in Landslide for estimating number of interleavings) consistently underestimates,
-being often seen to count (nearly) monotonically up toward the true value as exploration progresses,
-while the RE algorithm (used in Landslide for estimating total execution time) can be unstable and erratic,
-bouncing wildly among orders of magnitude even in the space of ten or fewer adjacent interleavings.
-These flaws are especially visible in the {\tt avl\_fixed} transactional verification results (Table~\ref{tab:tm-verifs}).
-% TODO: simpler approach would be to account for like, "T1->T1" prefix subtree likely to be smaller than "T1->T2" prefix subtree
-% because T1->T1->... means T1 has 2 fewer events left to schedule and (M+N-2 choose M-2) tends to be smaller than (M+N-1 choose M-1)
-% for roughly similar M/N;
-% estimator could account for this somehow... not sure how but
-% likewise, estimator could account for retry sets by checking the size of the retry set and multiplying by 1/4 or 1/2 or w/e
-% TODO: cite smth about machine learning
-% TODO: publish the htm evaluation logs
-% state space estimation could probably be drastically improved with machine learning
-% accordingly, all the state space estimate logs (ie, quicksand logs) from the tsx chapter are made open-source on so-and-so
-
-% arugment why ML is NOT appropriate for smc in general -- if you try to train it to choose thread interleavings / search ordering,
-% even if you're already in not-enough-cpu-time heuristics land, you're biasing it to find only the types of bugs you've already seen
-% and destroying its ability to find novelties
-% if your ML provides a partial verification, it is impossible for a human to understand what that verification means
-
-% ICB incorporations
-% into quicksand - when SS estimates are too high, fall back on a preemption bounded version to find bugs - improve the partial verif
-% into education - provide more partial verifs to students (but is it clear they'll understand the subtleties of it?)
-% into txn - incorporate failure injection into the model
-
-% soundness concerns - the xbegin/delay-xbegin pp soundness bug popped up in the implementation, despite the soundness thm
-% current SoA for this is to manually inspect the state spaces that arize for small K,N and just hope...
-% as concurrency model becomes more complex, more opportunities for the implementation to deviate from what
-% the soundness theorem actually proves
-% how to prove landslide is actually doing what you think? analogous to sel4 spec/impl verif...
+\input{future}
 

### 2018-10-02 thesis/future.tex
future test case design

@@
 %The reader looking for their own thesis topic would most likely find it here.
+%The reader should beware of wild speculation ahead
+%which may assume solutions to extremely difficult problems with a wave of the hand.
 
@@
 
+In its current form, Landslide is limited to testing only those behaviours that the test case it's hooked up to can generate.
+The most obvious limitation of this is resource exhaustion scenarios:
+stateless model checking simply cannot handle tests long-running enough to exhaust system memory
+(succumbing to exponential explosion long before),
+% "involve" here, for subjunctive?
+so cannot exercise any flow control that involves {\tt malloc()} failing, for example.
+This specific issue could be solved using by-hand annotations to block all preemption points until just before exhaustion,
+%(requires model-checking expertise on the programmer's part)
+or by extending Landslide to inject allocation failures at {\tt malloc()} callsites (akin to {\tt \_xbegin()}).
+%(assumes certain behavioural properties of the allocation library).
+However,
+these require the user to realize in advance that she should worry about allocation-failure bugs,
+and to configure Landslide specifically to target them.
+In general, a mature testing tool should not require the user to
+know in advance where her bugs might be
+before being able to conduct an effective test.
+
+In future work, a stateless model checking framework could collaborate with its user to semi-automatically design better tests.
+% i hate the way maple actually does it, in principle, but i need to cite smth here
+Concurrency coverage metrics such as proposed by Maple \cite{maple}
+% wild speculation
+could be extended to capture possible program behaviours under any test input,
+not just within the fixed state space under one given test,
+and symbolic execution frameworks such as Contessa \cite{contessa}
+could search for possible test inputs to suggest changes to the test case which might expose them.
+In cases where not-yet-covered conditional branches require
+certain function return values,
+such as {\tt malloc()} failing,
+the tool could offer to add failure injection with the user's approval,
+or at least prompt her to write a new test with that as a subgoal.
+
+In the case study of submitted P2 bugs (\sect{\ref{sec:education-eval-bug-case-study}}),
+I noticed several submissions of a common pattern
+in which Landslide overlooked the {\tt thr\_exit()} use-after-free bug:
+they all manually recycled the exited thread's stack using an internal free-list,
+rather than calling {\tt free()} or {\tt remove\_pages},
+so Landslide's heap-checker failed to see anything out of the ordinary about the subsequent writes to it.
+Explicit annotations about the custom free-list's invariants could make Landslide treat it like {\tt free()}
+and catch the errant write,
+but that begs the question of knowing about the bug in advance.
+Catching this bug would require extending the test case to {\tt thr\_fork()} a new thread
+to reuse the old stack and suffer data corruption from the stray write.
+In this case, the user might happen to find this just by increasing the number of threads/iterations,
+which is just good verification principles,
+but in general, automatically inferring such data structure invariants
+to suggest better assertions
+% is this true??
+remains an open problem.
+
 % test case design, writing code that actually checks the properties you're interested in
@@
 % with without_function, ignore_sym, etc, while preserving the guarantee
-
-% envision a framework that uses some code coverage tester to guide choice of pps and/or the user's without_function calls 
-% to get that coverage -- although relating flow control coverage directly to concurrency coverage is not straightforward
-% framework can also use 'git diff's, like comparing the most recent commit against an already-trusted codebase,
-% to figure out how to isolate just the stuff the change affects (memory intersection & data flow resulting from those conflix)
-% and then automatically assign some without_functions/etc to test just that commit
-
-% failure injection, testing malloc/etc failures
-% eg student condvars which loop around malloc failures can deadlock
-% can use retry set reduction in principle to prune the same kind of equivalences
-
-% also here we've got discussion of the exit-UAF buggers where landslide oops verifs the test anyway
-% because they're using a free pool and not free()
-% need a 3rd thread at least, to expose this (by actually getting corruption)
 

### 2018-10-05 thesis/future.tex
more future - verification etc, some more related works too

@@
 
-\subsubsection{User-friendliness}
+\section{User-friendliness}
 
@@
 at the same time,
-pruning out the overall state space would allow Quicksand to achieve more meaningful results in that same budget.
+pruning uninteresting branches out of the overall state space
+would allow Quicksand to achieve more meaningful results in that same budget.
 Theoretical advances in state space reduction come out every year (\sect{\ref{sec:related-algs}}),
@@
 using state space estimation to give an idea of expected testing time,
-and listing DPOR's memory conflicts to help her identify more candidate functions to potentially ignore.
+and listing the variables/locations of DPOR's memory conflicts
+to help her identify more candidate functions to potentially ignore.
 % there are a LOT of "hmm"s here for future research to address
@@
 
-\subsubsection{Test cases}
+\section{Test cases}
 
@@
 In this case, the user might happen to find this just by increasing the number of threads/iterations,
-which is just good verification principles,
+which is just good testing practice,
 but in general, automatically inferring such data structure invariants
@@
 remains an open problem.
+Likewise, \sect{\ref{sec:tm-abstraction}}'s abstraction reduction requires the user
+to believe that {\tt lock}({\tt spinlock}) and friends correctly express the mutual exclusion property in informal code%
+\footnote{In Figure~\ref{fig:htm-lock}(a), the
+{\tt \flow{if} (!\call{\_xtest()}) \call{thr\_yield}(\const{-1});}
+part required expert knowledge of HTM semantics to write correctly:
+an unconditional yield there
+would cause % oops
+the transaction to abort every time,
+verifying mutual exclusion in the failure path only.}.
+Future work could further check such tests against external formal specifications
+% better citation? that silly planning assignment of ours even?
+%analogously to seL4 \cite{sel4},
+to make the abstraction reduction more trustworthy.
 
-% test case design, writing code that actually checks the properties you're interested in
-% for example the mutex test, with the if (!xtest) yield check
-% then even after you got the right properties, optimizing the state space
-% with without_function, ignore_sym, etc, while preserving the guarantee
+\section{Verification}
 
-\subsubsection{Verification}
-
-% soundness concerns - the xbegin/delay-xbegin pp soundness bug popped up in the implementation, despite the soundness thm
-% current SoA for this is to manually inspect the state spaces that arize for small K,N and just hope...
-% as concurrency model becomes more complex, more opportunities for the implementation to deviate from what
-% the soundness theorem actually proves
-% how to prove landslide is actually doing what you think? analogous to sel4 spec/impl verif...
+As important as Quicksand's convergence theorem (\sect{\ref{sec:quicksand-soundness}}),
+the HTM equivalence proofs (\sect{\ref{sec:tm-design}}),
+and other soundness results from prior work are,
+they fall short of fully end-to-end formal verification
+in terms of trusting the output of stateless model checkers in practice.
+In addition to the issue of test cases,
+discussed above,
+there is also currently no guarantee that Landslide's concrete implementation
+matches the theoretical properties.
+My personal faith in Landslide's implementation comes from years of empirical observation:
+inspecting small state spaces
+(such as Figure~\ref{fig:retry-reduction})
+by hand
+to check that preemption points show up in the right places
+and that DPOR prunes the expected equivalences;
+as well as checking that the relationships between larger state spaces satisfy expected invariants,
+such as when testing two sets of preemption points, one a subset of the other,
+or testing the same but one with an additional reduction applied,
+% this first "the" realllly helps disambiguate the sentence
+the one's resulting state space must be smaller than the other's.
+Nevertheless, discrepancies may still lurk undiscovered:
+the failure injection bug described in \sect{\ref{sec:tm-eval-exp-setup}}
+took until implementing retry set reduction to discover,
+well after publishing the first experimental results under it in \cite{sigbovik-htm}.
+As the concurrency model becomes more complex,
+more opportunities arise for the implementation to deviate from what the soundness theorem actually proves.
+Kernels like seL4 \cite{sel4} and CertiKOS \cite{certikos}
+and concurrency-aware compilers like CompCertTSO \cite{compcerttso},
+all formally verified,
+provide important pioneering steps in this direction
+\footnote{Of course, being verified themselves, these projects have no need for a verified Landslide,
+but checking unverified (e.g. student) code with such would still be an improvement.}.
 
@@
 
-\subsubsection{Heuristics}
+\section{Heuristics}
 

### 2018-10-05 thesis/related.tex
more future - verification etc, some more related works too

@@
 \subsubsection{Algorithms}
+\label{sec:related-algs}
 
@@
 focusing on preemption point subsets rather than context switch bounding.
+\revision{Maple \cite{maple} proposed a heuristic metric for measuring the amount of ``concurrency coverage'',
+analogous to line-by-line coverage for sequential programs,
+and prioritizes testing interleavings which increase the coverage metric,
+although it makes several limiting assumptions such as
+% "vulnerability window", in its words
+preemption locality
+% this is like, TOTAL bs, completely results-oriented empiricism
+% and at worst, intentionally blinding yourself to the actually difficult bugs
+and independence of values read/written.}
 DeMeter \cite{demeter} adapted abstraction reduction to distributed systems verification under the name Dynamic Interface Reduction,
@@
 
-\subsubsection{Kernel verification}
+\subsubsection{Formal verification}
 
@@
 in the scope of verification,
-using a proof in Coq that also took 2 person-years to complete.
+using a proof in the Coq proof assistant
+\cite{coq}
+that also took 2 person-years to complete.
 Its safety properties hold under all possible interleavings,
@@
 to provide a less formal, but still hopefully useful, verification guarantee.
+%
+\revision{Lastly, CompCertTSO \cite{compcerttso} extends the CompCert verified compiler \cite{compcert}
+to capture x86's relaxed memory semantics,
+guaranteeing that a program written in the ClightTSO subset of C
+is translated accurately to assembly with the same behaviour.
+Like CertiKOS, its implementation is verified in Coq.
+Although Landslide checks programs directly at the executable level,
+blind to the code the programmer personally wrote,
+extending that pipeline with a certifying compiler would
+improve the overall verification,
+ensuring that Landslide was checking the behaviour the programmer actually intended to write.}

### 2018-10-05 thesis/future.tex
future - generalize from k,n; combine test case and verif sects

@@
 
-\section{Test cases}
+%\section{Test cases}
+\section{Verification}
 
@@
 before being able to conduct an effective test.
-
+%
 In future work, a stateless model checking framework could collaborate with its user to semi-automatically design better tests.
@@
 % wild speculation
-could be extended to capture possible program behaviours under any test input,
+could be extended to capture possible behaviours under any test input,
 not just within the fixed state space under one given test,
 and symbolic execution frameworks such as Contessa \cite{contessa}
-could search for possible test inputs to suggest changes to the test case which might expose them.
+could search for possible inputs to suggest changes which might expose them.
 In cases where not-yet-covered conditional branches require
@@
 remains an open problem.
+Future work might combine Landslide's ability to prove correctness for finite $K,N$ parameters
+with a formal induction proof that generalizes the result,
+finding a minimum number of threads necessary to expose all behaviours
+(at least 3 in this example),
+% 2 papers rec'd from https://twitter.com/bblum0/status/1018233122750640130
+% although i checked these & couldn't figure out how if at all they were doing generalization
+although I know of no related work which addresses the second half of this problem.
 Likewise, \sect{\ref{sec:tm-abstraction}}'s abstraction reduction requires the user
@@
 
-\section{Verification}
+%\section{Verification}
 
@@
 In addition to the issue of test cases,
-discussed above,
+%discussed above,
 there is also currently no guarantee that Landslide's concrete implementation
@@
 and concurrency-aware compilers like CompCertTSO \cite{compcerttso},
-all formally verified,
+all formally verified against external specifications,
 provide important pioneering steps in this direction
@@
 but checking unverified (e.g. student) code with such would still be an improvement.}.
-
-% extending landslide's verification for a limited nr of K N to infinite with induction/proof assistance
-
-% - this should prob go in quicksand chapater
-% identifying & skipping redundancies in smaller ID state spaces via memoization (same applies to ICB!)
-
-% arugment why ML is NOT appropriate for smc in general -- if you try to train it to choose thread interleavings / search ordering,
-% even if you're already in not-enough-cpu-time heuristics land, you're biasing it to find only the types of bugs you've already seen
-% and destroying its ability to find novelties
-% if your ML provides a partial verification, it is impossible for a human to understand what that verification means
 
@@
 
+% arugment why ML is NOT appropriate for smc in general -- if you try to train it to choose thread interleavings / search ordering,
+% even if you're already in not-enough-cpu-time heuristics land, you're biasing it to find only the types of bugs you've already seen
+% and destroying its ability to find novelties
+% if your ML provides a partial verification, it is impossible for a human to understand what that verification means
+
 % ICB incorporations

### 2018-10-06 thesis/future.tex
future talk abt etas and ml

@@
 
-In the author's experience,
+In my experience,
 the WBE algorithm (used in Landslide for estimating number of interleavings) consistently underestimates,
@@
 These flaws are especially visible in the {\tt avl\_fixed} transactional verification results (Table~\ref{tab:tm-verifs}).
-% TODO: simpler approach would be to account for like, "T1->T1" prefix subtree likely to be smaller than "T1->T2" prefix subtree
-% because T1->T1->... means T1 has 2 fewer events left to schedule and (M+N-2 choose M-2) tends to be smaller than (M+N-1 choose M-1)
-% for roughly similar M/N;
-% estimator could account for this somehow... not sure how but
-% likewise, estimator could account for retry sets by checking the size of the retry set and multiplying by 1/4 or 1/2 or w/e
-% TODO: cite smth about machine learning
-% TODO: publish the htm evaluation logs
-% state space estimation could probably be drastically improved with machine learning
-% accordingly, all the state space estimate logs (ie, quicksand logs) from the tsx chapter are made open-source on so-and-so
+Both estimation algorithms
+use the known structure of the tree as a model for the unknown
+(see \sect{\ref{sec:landslide-estimate}}),
+but make no use of domain-specific knowledge such as which threads were chosen to run at each preemption point.
+For example, consider the root of the left subtree in Figure~\ref{fig:tree},
+thread 1's \dporTA{{\tt tmp1++}}.
+At that moment thread 1 has 1 more instruction left to execute, while thread 2 has 3.
+If thread 1's last instruction is chosen to run next,
+there can only be one way to run thread 2's 3 steps thereafter (${3+0 \choose 0} = 1$);
+if thread 2's first instruction is chosen to run next,
+there will be 3 ways to interleave thread 1's last among thread 2's remaining 2 (${2+1 \choose 1} = 3$).
+In general, the largest child subtree at any preemption point will be the one resulting from running the thread with the most transitions left,
+or more formally, fixing $N+M=C$, ${N+M \choose N}$ is maximized when $N \approx M$.
+WBE and RE could both be relatively easily extended to incorporate this observation,
+using knowledge from branches already tested to guess (still heuristically) appropriate values for $N$ and $M$.
+Predicting more advanced aspects of state space structure,
+such as subtrees pruned by sleep set reduction
+(\sect{\ref{sec:landslide-sleepsets}}; $\bigstar$4 in Figure~\ref{fig:dpor-example-2}),
+would likely require analyzing DPOR's memory conflicts as well.
+
+There is much room to expand Landslide's use of heuristics to find bugs as soon as possible should they exist.
+% TODO
+incorporate ICB \cite{chess-icb}
+% ICB incorporations
+% into quicksand - when SS estimates are too high, fall back on a preemption bounded version to find bugs - improve the partial verif
+% into education - provide more partial verifs to students (but is it clear they'll understand the subtleties of it?)
+% into txn - incorporate failure injection into the model
 
@@
 % if your ML provides a partial verification, it is impossible for a human to understand what that verification means
-
-% ICB incorporations
-% into quicksand - when SS estimates are too high, fall back on a preemption bounded version to find bugs - improve the partial verif
-% into education - provide more partial verifs to students (but is it clear they'll understand the subtleties of it?)
-% into txn - incorporate failure injection into the model
+As a matter of personal philosophy,
+I would strongly caution against using machine learning to try to expose bugs faster.
+Machine learning has become popular recently as a magic cure-all for many programming problems,
+but is also notorious for amplifying any biases in its training set in often unpredictable ways
+\cite{conceptnet-bias}.
+In the domain of concurrency testing, this would translate to being able to find only types of bugs
+already seen in the training data, abandoning any novel bug patterns to further obscurity.
+Granted, Iterative Deepening and ICB also deprioritize ``deeper'' bugs,
+as measured by number of unique or total preemptions, respectively,
+but this strategy is at least backed up by empirical evidence
+% TODO: upgrade citation to also have a section reference to that icb bound table from quicksand eval
+\cite{chess-icb},
+and when a bug is not found,
+the strategy is transparent enough that the user can understand the nature of the partial verification.
+%
+Machine learning would be more appropriate for improving state space estimation,
+whose output is already a black-box oracle from the user's perspective,
+and where the worst case of being wrong
+is that Quicksand prioritize its jobs suboptimally. % subjunctive
+In-progress estimates fluctuate depending on many factors,
+such as interleaving-dependent flow control,
+and I suspect deeper patterns may exist among multiple different state spaces,
+for example,
+from the same test program with different $K,N$ parameters.
+For any future researchers who wish to study such patterns,
+% with machine learning... don't feel like i can do justice to discussing its flaws with real citations
+I have made the estimation logs from the HTM experiments (\sect{\ref{sec:tm-eval}})
+freely available at \url{https://github.com/bblum/thesis-htm-logs}.
+%
+Nevertheless,
+any overall search heuristic for concurrency testing
+must be designed to support the human user's agency first and foremost,
+that she may readily understand partial verification results
+and participate effectively in the test case refinement strategies discussed
+upchapter. % this is a word now.
 

### 2018-10-06 thesis/thesis.tex
future - generalize from k,n; combine test case and verif sects
update fonts so charter in mathmode too
update acks

@@
 
-\usepackage{fontspec}
+%\usepackage{fontspec}
+\usepackage{mathspec}
 \usepackage{xeCJK}
 
-\setmainfont{XCharter}
+%\setmainfont{XCharter}
+\setallmainfonts(Digits,Latin){XCharter}
 %\setmainfont{Utopia}[
@@
 to design scientifically thorough experiments,
-to write prose explaining difficult concepts approachably,
+to explain difficult concepts approachably,
 to appreciate related work as charitably as possible,
@@
 Professor David Eckhardt
-provided unending guidance on education, writing, life, you name it,
+provided unending guidance on education, writing, career, life, you name it,
 helped revise each semester's iteration of the recruiting lecture with a keen eye for the student mind,
 contributed substantially to the design of the survey,
-and supplied constant enthusiasm and reassurance that Landslide was a boon to 15-410 that kept me going.
-I can't express the depth of my gratitude to Dave strongly enough,
-even though three of the Landslide-friendly tests never made it upstream into the official P2 test suite.
+and supplied constant enthusiasm and encouragement that Landslide was a boon to 15-410.
+I can't express my gratitude to Dave strongly enough,
+even if it took until just this week for half of the Landslide-friendly tests to be included in the official P2 test suite.
+%even though three of the Landslide-friendly tests never made it upstream into the official P2 test suite.
 
@@
 
-% FIXME - cmon latex pls??
-\pagebreak
-
-Many others have collaborated with me directly on the research itself,
+{Many others have collaborated with me directly on the research itself,
 whom I thank here in approximate chapter order.
@@
 greatly enriching Chapter~\ref{chap:education}'s evaluation.
-David Blum and Skye Toor provided invaluable advice on statistical analysis
-that helped me evaluate the student grade distributions therein.
-Mario Dehesa-Azuara generously contributed the transactional memory data structures
+David Blum and Skye Toor provided
+invaluable advice on statistical analysis
+that helped me evaluate the student
+% justify last line bc of manual page break
+\unskip\parfillskip 0pt \par}
+
+% FIXME - cmon latex pls??
+\pagebreak
+
+\noindent
+grade distributions therein.
+Mario Dehesa-Azuara generously contributed the
+transactional memory data structures
 that make up about half of Chapter~\ref{chap:tm}'s evaluation suite.
+
+Most importantly of all, I thank every student who ever used Landslide.
+Though the introduction section may say differently,
+their smiles and gratitude were the true motivation of this thesis.
 
@@
 It was a blessing to serve, and later do research, among you all.
+Thanks also to Wind River for generously providing CMU with Simics educational licenses.
 I am also grateful to Adam Blank and Miriam Zimmerman
@@
 pushing me to refine my presentation skills and to network with strangers (now familiar faces) from industry.
-I also thank the retreat's industry attendees from each of the PDL Consortium member companies
-for their interest even in more theoretical work
+I also thank those industry attendees from each of the PDL Consortium member companies
+for their interest, even in more theoretical work,
 and the enlightening conversations resulting therefrom.
@@
 whether at the retreat or otherwise,
-told me how cool they thought my research was for directly helping students --
+insisted how cool my research was for directly helping students --
 every such reminder helped me keep my head up during darker times.
@@
 %\footnote{and Slack emojis},
-that I hope will long outlive the silly card game that brought us together.
+that will long outlive the silly card game that brought us together.
 %It'd be an injustice to list only a few names here and omit others, but you know who you are.
@@
 
+\newcommand\revision[1]{#1\xspace}
+
 \input{intro}

### 2018-10-06 thesis/intro.tex
intro subsubsections
assorted minor prose revisions

@@
 
-\subsection{Motivation}
+\subsubsection{Motivation}
 
@@
 
-\subsection{Contribution}
+\subsubsection{Contribution}
 
@@
 
-\subsection{Organization}
+\subsubsection{Organization}
 
@@
 	\item {\bf Related Work:} Chapter~\ref{chap:relatedwork} honors my neighbours and ancestors in research spirit.
-	\item {\bf Future Work:} Chapter~\ref{chap:warpzone} % TODO
+	\item {\bf Future Work:} Chapter~\ref{chap:warpzone} lists stateless model checking's remaining shortcomings and how new research might address them.
 	\item {\bf Conclusion:} Chapter~\ref{chap:conclusion} provides some thoughts on the future of the field.
@@
 The impersonal ``the programmer'' will be referred to as she/her
-to highlight her role as the intended user, separate from the underlying research,
+to challenge readers' unconscious bias about gender in computer science,
+as well as to highlight her role as the intended user, separate from the underlying research.
 %to distinguish her from the thesis audience concerned with the underlying research,
-and also to challenge readers' unconscious bias about gender in computer science.
 Individual students who participated in the user studies will be given the more inclusive they/them.
@@
 others may find that disorienting and wish to see things in a way a compiler would understand.
-The beginning of Chapter~\ref{chap:landslide} provides source code links,
-and the rest of it serves as a guide to browsing the repository.
+Chapter~\ref{chap:landslide} provides source code links and serves as a guide to browsing the repository.
 Later chapters will often make parenthetical references to specific files and/or functions

### 2018-10-06 thesis/landslide.tex
assorted minor prose revisions

@@
 of complex algorithms from dense prose and mathematical notation alone,
-I also present here a friendly example of using DPOR on Figure~\ref{fig:concurrency-bug}'s example program,
+mortal readers may prefer the following example of using DPOR on the program from Figure~\ref{fig:concurrency-bug},
 whose original state space is shown in Figure~\ref{fig:tree}.
@@
 In general, the number of ways to interleave two threads executing $N$ and $M$ events each is given by ${N+M \choose N}$%
-\footnote{Generalizing to $K$ threads, and simplifying to the same number $N$ of events each,
+\footnote{Generalizing to $K$ threads, and simplifying to $N$ events each,
 this formula becomes $\frac{n!k!}{n!^k}$.};
@@
 
-It is easy to see how these formulas might fit into DPOR's incremental search procedure:
+It is easy to see how these might fit into DPOR's incremental search procedure:
 at the end of each branch compute its $\mathsf{proportion}$ and add it in to a global $\mathsf{estimate}$ value.
@@
 requiring them to be recomputed,
-which is not feasible without storing the entire exponentially-sized tree in memory.
+%which is not feasible without
+which would require
+storing the entire exponentially-sized tree in memory.
 Instead, Landslide also stores per-subtree estimates at intermediate $v_i$ nodes, $1 < i < n$,
 along the current branch.
-Whenever DPOR marks a new $k$th branch for exploration
+Whenever DPOR marks a new $k$th branch
+%for exploration
 at some $v_i$
@@
 
-\subsubsection{Runtime (Recursive) estimation}
+\subsubsection{Run-time (Recursive) estimation}
 

### 2018-10-06 thesis/overview.tex
assorted minor prose revisions

@@
 
-This chapter will introduce the necessary background material on concurrency, stateless model checking, data-race analysis, and the relevant undergraduate operating systems classes.
+This chapter introduces the necessary background material on concurrency, stateless model checking, data-race analysis, and the relevant undergraduate operating systems classes.
 
@@
 		based on the observation that most concurrency bugs in practice can be exposed with as few as 2-3 preemptions.
-		For example, {\em Landslide} heuristically judges the program to be stuck in an infinite loop
+		For another example, {\em Landslide} heuristically judges the program to be stuck in an infinite loop
 		when it has executed at least 4000 times longer than previously-observed ``safe'' executions,

### 2018-10-06 thesis/related.tex
assorted minor prose revisions

@@
 Equal partners in concurrency testing are the practical and the theoretical:
-the former meaning tool implementations that target specific problem domains and help users as best one can,
-and the latter meaning algorithmic advances to bring ever-larger state spaces within the realm of computational feasibility.
-I discuss my most closely related works split in two sections accordingly.
+%the former meaning
+tool implementations that target specific problem domains
+%and manage practical tradeoffs,
+%and help users as best one can,
+%and the latter meaning
+and algorithmic advances to make ever-larger state spaces computationally feasible.
+%within the realm of computational feasibility.
+The following two subsections discuss the most closely related prior work accordingly.
+%I discuss my most closely related works split in two sections accordingly.
 
@@
 as many of this thesis's contributions build directly upon it.
-
+%
 DPOR has since been extended in several ways to achieve further reduction
@@
 Although Landslide checks programs directly at the executable level,
-blind to the code the programmer personally wrote,
+blind to the source code the programmer personally wrote,
 extending that pipeline with a certifying compiler would
 improve the overall verification,
-ensuring that Landslide was checking the behaviour the programmer actually intended to write.}
+ensuring that Landslide was actually checking the behaviour the programmer intended.}

### 2018-10-06 thesis/tm.tex
add missing section label in tm
assorted minor prose revisions

@@
 \subsubsection{Experimental setup}
+\label{sec:tm-eval-exp-setup}
 
@@
 than {\tt mutex}
-% TODO recalc
-(roughly 9 interleavings per second for the former, 22 for the latter),
+% old ver.
+%(roughly 9 interleavings per second for the former, 22 for the latter),
+% lock spin fixed:
+% (4+998+138911+186+169893+286910) / (355+10557+3600000+2827+3600000+3600000) = 5.51
+% lock mutex:
+% (4+180+9372+489480+132+714882+15064) / (344+1683+40521+2499968+1500+3600000+66589) = 19.79
+(roughly 6 interleavings per second for the former, 20 for the latter),
 because while {\tt mutex} abstracts away threads needing to wait their turn for the critical section
@@
 \section{Discussion}
+\label{sec:tm-discussion}
 
@@
 
-Likewise, the cases of {\tt htm2}(2,3) and (4,1),
+% TODO: is this true? at all? htm2 is bigger in STM only because it's got a separate expliit abort
+Likewise, the cases of {\tt htm2}(2,3), (2,4), and (4,1),
 in which retry sets achieved better reduction than STM mode,

### 2018-10-06 thesis/future.tex
future heuristics qs/icb

@@
 
-There is much room to expand Landslide's use of heuristics to find bugs as soon as possible should they exist.
-% TODO
-incorporate ICB \cite{chess-icb}
+There is much room to expand Landslide's use of heuristics to balance verification with fast bug-finding.
+Landslide's maximal state space mode (\sect{\ref{sec:tm-verif}}) currently requires the user to decide in advance
+whether she thinks a full verification is possible within the time limit,
+and supply {\tt -M} if so,
+sacrificing some bug-finding power up front
+(or just running the test twice).
+Quicksand could make this decision automatically,
+always dedicating one of its available CPUs to the maximal state space
+while the rest run Iterative Deepening as normal.
+Currently, Landslide explores each state space sequentially,
+but if extended with Distributed DPOR \cite{parallel-dpor},
+Quicksand could also dynamically allocate more or fewer CPUs to the maximal state space according to its ETA.
 % ICB incorporations
@@
 % into txn - incorporate failure injection into the model
+Likewise, Quicksand could
+incorporate ICB \cite{chess-icb}
+to get the best of both worlds:
+when testing smaller data-race jobs,
+start them in ICB mode to begin with,
+and when the ETAs of larger jobs (including the maximal one) grow too large,
+downgrade them to ICB to try to at least get a partial verification for that preemption point set
+rather than discarding it entirely.
+ICB could also be extended to include HTM failure injections as part of its model,
+counting them either as part of the overall preemption bound
+or under a separate abort bound.
+State space estimation under ICB currently
+counts only DPOR's tagged branches that don't get skipped under the preemption bound,
+estimating that bound's state space size
+and resetting whenever the bound increases;
+it could also be extended to be ICB-aware,
+counting branches skipped due to the bound separately
+and computing two estimates at once,
+which would in turn allow Quicksand to make more informed decisions about when to use ICB.
 
@@
 already seen in the training data, abandoning any novel bug patterns to further obscurity.
-Granted, Iterative Deepening and ICB also deprioritize ``deeper'' bugs,
+Granted,
+Iterative Deepening and
+ICB also deprioritize ``deeper'' bugs,
 as measured by number of unique or total preemptions, respectively,
-but this strategy is at least backed up by empirical evidence
-% TODO: upgrade citation to also have a section reference to that icb bound table from quicksand eval
-\cite{chess-icb},
-and when a bug is not found,
-the strategy is transparent enough that the user can understand the nature of the partial verification.
+%in terms of minimum necessary preemption count,
+%but this strategy is informed by empirical evidence
+%% TODO: upgrade citation to also have a section reference to that icb bound table from quicksand eval
+%\cite{chess-icb},
+%and
+but when a bug is not found
+these strategies are also transparent enough that the user can understand the nature of the partial verification.
 %

### 2018-10-06 thesis/tm.tex
fix retry set fp deadlock issue and finish retry set column
add map basicer mutex tm test

@@
 		\item {\tt map\_mutex}: {\tt map\_basic} simplified likewise.
+		\item {\tt map\_basicer\_mutex}: {\tt map\_basicer} simplified likewise.
 	\end{itemize}
@@
 			& 3,1 & \cpu{43.24}	& \ints{774}	& \cpu{24.02}	& \ints{224}	& \cpu{22.79}	& \ints{140} \\
-			& 3,2 &\ETAdag{1y 299d}	&\ETAdag{4510472}& \ETAdag{qqq}	& \ETAdag{qqq}	&\ETAdag{10d 16h}&\ETAdag{2322150}\\
+			& 3,2 &\ETAdag{1y 299d}&\ETAdag{4510472}&\ETAdag{294d 5h}&\ETAdag{8185793}&\ETAdag{10d 16h}&\ETAdag{2322150}\\
+			% argh.. since re running the retry set i feel like i have to re run the others...
+			% the small state spaces though all end up similar perf numbers though;
+			% i wonder why this one alone broke 1mil?
 			% [JOB 4] progress: 759740/4510472 brs (16.843911%), ETA 1y 298d 22h 14m 13s (elapsed 9h 59m 55s)
+			% [JOB 4] progress: 1133447/8185793 brs (13.846513%), ETA 293d 19h 13m 36s (elapsed 9h 59m 55s) % good retry
 			% [JOB 4] progress: 646606/2322150 brs (27.845142%), ETA 10d 5h 50m 15s (elapsed 9h 59m 55s)
+			% old, owned retry set --
+			% ls-id-log-20180817141549.log.ByMMKx.gz
+			% [JOB 4] BUG FOUND: landslide-trace-1534529755.156051.html (981774 interleavings; 132 preemptions; ...)
+			% it's almost certainly getting retry set blocked
+			% so you just have to detect a retry set block deadlock, and un deadlock it
+			% this should not affect other tests (not even in cpu time (beyond icache considerations))
+			% when you fix it and re run it, remember to use leave-logs, and put a ALWAYS prince in landslide
+			% to see when it happens
 			& 4,1 & \cpu{9225.52}	& \ints{212146}	& \cpu{376.70}	& \ints{11973}	& \cpu{2158.93}	& \ints{44995} \\
@@
 			& 2,1 & \cpu{739.60} 	& \ints{20459} 	& \cpu{332.32}	& \ints{9675}	& \cpu{122.40}	& \ints{2774} \\
-			& 2,2 & \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	& \ETAdag{qqq}	&\ETAdag{36393y}&\ETAdag{35094477}\\
+			& 2,2 &\ETAdag{58880y}&\ETAdag{4.62$\times\mathit{10^7}$}&\ETAdag{51191y}&\ETAdag{3.33$\times\mathit{10^7}$}&\ETAdag{36393y}&\ETAdag{3.51$\times\mathit{10^7}$}\\
+			% [JOB 19] progress: 366191/46150466 brs (0.793472%), ETA 58879y 219d 6h 4m 16s (elapsed 9h 59m 28s)
+			% [JOB 19] progress: 351391/33318009 brs (1.054658%), ETA 51191y 111d 17h 37m 55s (elapsed 9h 59m 28s)
 			% progress: 302022/35094477 brs (0.860597%), ETA 36392y 304d 6h 53m 45s (elapsed 9h 58m 49s)
-			& 3,1 & \ETAdag{5y 292d}&\ETAdag{2873664642}&\ETAdag{qqq}& \ETAdag{qqq}	&\ETAdag{97y 100d}&\ETAdag{1698185036} \\
+			& 3,1 & \ETAdag{5y 292d}&\ETAdag{2.87$\times\mathit{10^9}$}&\ETAdag{735y 212d}&\ETAdag{3.35$\times\mathit{10^8}$}&\ETAdag{97y 100d}&\ETAdag{1.70$\times\mathit{10^9}$} \\
 			% [JOB 18] progress: 447687/2873664642 brs (0.015579%), ETA 5y 291d 11h 13m 30s (elapsed 9h 59m 41s)
+			% [JOB 25] progress: 18151/335288683 brs (0.005414%), ETA 735y 211d 9h 1m 8s (elapsed 22m 25s)
 			% [JOB 18] progress: 357586/1698185036 brs (0.021057%), ETA 97y 99d 8h 44m 33s (elapsed 9h 59m 41s)
@@
 			& 2,1 & \cpu{1950.48}	& \ints{30719}	& \cpu{867.68}	& \ints{13237}	& \cpu{367.91}	& \ints{5446} \\
-			& 2,2 & \ETAdag{1y 176d}&\ETAdag{75516602}&\ETAdag{qqq}	& \ETAdag{qqq}	&\ETAdag{13d 11h}&\ETAdag{565334} \\
+			& 2,2 & \ETAdag{1y 176d}&\ETAdag{7.55$\times\mathit{10^7}$}&\ETAdag{11y 94d}&\ETAdag{9.57$\times\mathit{10^7}$}&\ETAdag{13d 11h}&\ETAdag{565334} \\
 			% [JOB 22] progress: 87777/75516602 brs (0.116235%), ETA 1y 176d 16h 38m 50s (elapsed 3h 1m 48s)
+			% [JOB 22] progress: 128715/95718464 brs (0.134472%), ETA 11y 93d 8h 25m 39s (elapsed 4h 52m 39s)
 			% [JOB 22] progress: 18380/565334 brs (3.251171%), ETA 13d 1h 5m 3s (elapsed 38m 33s)
-			& 3,1 &\ETAdag{6y 133d}&\ETAdag{1847957714}&\ETAdag{qqq}& \ETAdag{qqq}	&\ETAdag{4y 156d}&\ETAdag{888242178} \\
+			& 3,1 &\ETAdag{6y 133d}&\ETAdag{1.85$\times\mathit{10^9}$}&\ETAdag{8y 282d}&\ETAdag{7.90$\times\mathit{10^8}$}&\ETAdag{4y 156d}&\ETAdag{8.88$\times\mathit{10^8}$} \\
 			% [JOB 22] progress: 284530/1847957714 brs (0.015397%), ETA 6y 132d 6h 20m 26s (elapsed 9h 49m 5s)
+			% [JOB 22] progress: 282358/789973638 brs (0.035743%), ETA 8y 281d 14h 50m 3s (elapsed 9h 49m 54s)
 			% [JOB 22] progress: 274532/888242178 brs (0.030907%), ETA 4y 155d 23h 50m 51s (elapsed 9h 47m 5s)
@@
 			% total CPU time consumed: 10h 20s (36020148608 usecs) (core saturation: 12%)
-			\\
-			\\
-			\\
-			\\
+			\cline{1-4}
+			{\tt map\_basicer}
+			& 2,1 & \cpu{14.34} & \ints{9} \\
+			{\tt \_mutex}
+			& 2,2 & \cpu{147.27} & \ints{2953} \\
+			& 2,3 & \cpu{12946.21} & \ints{244691} \\
+			& 3,1 & \cpu{637.81} & \ints{12707} \\
+			& 3,2 & \cpu{qqq} & \ints{qqq} \\
+			& 4,1 & \ETAdag{33h 33m} & \ETAdag{1778661} \\
+			% [JOB 3] progress: 601402/1778661 brs (33.812059%), ETA 23h 33m 16s (elapsed 9h 59m 55s)
 			\\

### 2018-10-06 thesis/thesis.tex
conclusion.

@@
 an exponentially large problem
-% (literally and figuratively for each side of the analogy, respectively)
 that invariably thwarts they whoever try to solve it ``perfectly'';
 it bears fruit only to one who can find their own human-oriented meaning in partial results and compromises.
+I speak of course literally on one side of this analogy, figuratively on the other.
+This thesis has demonstrated such meaning for stateless model checking,
+choosing perhaps one of the most challenging demographics of programmers in need of concurrency testing,
+the undergraduate operating systems student,
+as its target audience.
+Along the way, it justified its techniques' applicability to problems beyond CMU's walls as well
+by showing positive results at other universities
+and by addressing new concurrency models developed by industry in just the last decade.
+% also by comparing the preemption bounds of quicksand eval's bugs to prior work in ICB
+% and finding them to actually be skewed to DEEPER bugs than in the "real" "world", so take that
+As I write this I am also collaborating with 15-410 staff
+to integrate Landslide as an official part of the course grading infrastructure,
+so in some sense the na\"{i}ve wish to make an ``automatic TA'' that first made me interested in grad school
+has also been fulfilled.
 
-asdfasdasdf
-
-% TODO: Get a little bit preachy about how computer programs should be created to help humans,
-% not to look as much like scifi as possible, by knowing their limits, asking for help when they get confused, etc.
-
+For as long as people have written science fiction books,
+we have fantasized about robots that can make any complex intuitive decision a human can
+in addition to the superhuman arithmetic ability that comes standard with silicon.
+Recent trends in machine learning have pursued this aesthetic,
+using statistics and pattern recognition to provide convincing output even on never-before-seen inputs,
+but few such systems are designed to also alert the user when their output confidence is low.
+% appropriate for serving ads (penalty for being wrong -- no click)
+% appropriate for drug synthesis (clinical trial vets the output)
+% not appropriate for... govt surveillance o.o;
+% not appropriate for program verification
+As tempting as it is to fantasize an oracle capable of telling for certain whether your program is free of concurrency bugs,
+that would be a disservice to both user and machine:
+the human skill that artificial intelligence should really try to emulate is
+knowing its limits and asking for help when it gets confused.
 % ways humans cooperate to make landslide better:
@@
 % - manual inspection of data races
+
+It's become fashionable among my peers lately,
+whenever a new exploit is announced in the news,
+to declare that security is doomed,
+that computers were a mistake,
+that we should all retire and become llama farmers,
+and so on.
+There is no doubt that we are burdened with massive technical debt from the mistakes made
+during computer science's infancy.
+Nevertheless,
+safety-oriented programming languages and type systems are growing ever more popular,
+and meanwhile formal verification techniques to check existing code,
+written in the old unsafe ways, grow ever more powerful.
+Landslide's ability to reach impressionable student minds is my little contribution in this long battle.
+I have hope that we may one day live in peace with our silicon creations,
+understanding and respecting their limitations just as they complement ours.
+This thesis has been a profoundly meaningful journey for me and I hope you enjoyed reading it.
 

### 2018-10-08 thesis/related.tex
parallel dpor in related

@@
 and to incorporate new concurrency models.
+\revision{Distributed DPOR \cite{parallel-dpor}
+allows the exploration to be parallelized,
+with a minimum of overhead from redundant interleavings that would ordinarily be pruned in sequential DPOR.}
 Optimal DPOR \cite{optimal-dpor} extends sleep sets into the more expressive {\em wakeup trees},
@@
 use an SMT solver \cite{z3} to identify additional pruning opportunities.
-Implementing wakeup trees or SMT-driven exploration in Landslide is left to future work.
+Implementing parallelization, wakeup trees, or SMT-driven exploration in Landslide is left to future work.
 

### 2018-10-08 thesis/landslide.tex
landslide - heuristic loop/sync/deadlock

@@
 The DPOR implementation then checks, for some $I_{ij}$ it wants to mark for exploration,
-whether {\tt ICB\_BLOCKED(}$T_j${\tt )} at the state after $t_i$,
+whether {\tt ICB\_BLOCKED}($T_j$) at the state after $t_i$,
 and skips it if so ({\tt tag\_good\_sibling()}/{\tt tag\_all\_siblings()}).
@@
 
-\subsection{Heuristic loop, deadlock, and synchronization detection}
+\subsection{Heuristic loop, synchronization, and deadlock detection}
 \label{sec:landslide-blocking}
 
-% TODO
+Despite the ease of automatically instrumenting a fixed concurrency API such as P2's,
+the variety of student implementations inevitably results in many behaviours outside Landslide's model.
+This section documents the heuristics Landslide uses to approximate a program's formal behaviour in such situations.
+
+\subsubsection{Infinite loop detection}
+
+All modern presentations of stateless model checking assume finite program length.
+Even though all test cases used in this thesis's experiments are hand-written to ensure runtime
+not just finite but also short (on account of exponential state space sizes),
+bugs may still cause a program to get stuck in an infinite loop unexpectedly.
+Detecting such loops in general is of course uncomputable \cite{entscheidungsproblem},
+but Landslide has the benefit of knowledge from past iterations to inform its sense of how long the program ``should'' run.
+
+\begin{figure}[h]
+	\begin{center}
+		\begin{tabular}{cc}
+			\includegraphics[width=0.28\textwidth]{heuristic-livelock.pdf}
+			&
+			\includegraphics[width=0.28\textwidth]{heuristic-tightloop.pdf}
+			\\
+			(a) Infinite loop around preemption points.
+			&
+			(b) Stuck between preemption points.
+		\end{tabular}
+	\end{center}
+	\caption{Detecting infinite loops heuristically by comparison to past interleavings.}
+	\label{fig:heuristic-loop}
+\end{figure}
+
+Landslide checks for two distinct categories of infinite loops,
+visualized in Figure~\ref{fig:heuristic-loop}.
+Firstly, it keeps a running average of how many preemption points deep each completed interleaving has been in the past.
+Whenever a new preemption point is reached, Landslide checks
+if its depth is greater than the heuristic constant factor of 20
+({\tt PROGRESS\_DEPTH\_\allowbreak{}FACTOR})
+times the previous average.
+If the depth exceeds that cutoff, it reports an infinite loop bug.
+Whether this represents a livelock or just a mundane sequential logic bug is for the user to decide.
+However, as
+different interleavings do
+often execute different program logic and vary in length accordingly,
+Landslide waits to test 10 interleavings ({\tt PROGRESS\_CONFIDENT\_\allowbreak{}BRANCHES})
+before applying this heuristic;
+in the case of fewer,
+it scales the depth factor by a heuristic exponential factor of 1.1
+to represent its lower confidence
+({\tt PROGRESS\_BRANCH\_UNCERTAINTY\_EXPONENT}).
+In the special case of the first branch ever tested,
+Landslide will abort after a fixed preemption point depth limit of 4000 ({\tt TOO\_DEEP\_0TH\_BRANCH})
+-- a program with such $N$ would probably have an impossibly large state space anyway.
+
+Secondly, a program may get stuck in a loop where no preemption points are encountered each new iteration.
+For example, race-induced data corruption may cause a list to end up circularly linked, %oops
+leading a search or append operation to fail to terminate.
+Landslide also maintains a running average of instructions per transition,
+and each instruction, compares if more instructions have elapsed since the last preemption point
+than a constant times that average.
+Because transitions may themselves vary greatly in number of instructions as each represents completely different program logic,
+this heuristic is much more lenient than the preemption-point-counting one,
+using a multiplicative factor of 4000 ({\tt PROGRESS\_TRIGGER\_FACTOR}).
+If such a loop is encountered within the P2 synchronization primitives,
+which should generally be free of $O(n)$ operations,
+Landslide uses a more aggressive cutoff of 2000
+({\tt PROGRESS\_AGGRESSIVE\_TRIGGER\_FACTOR}).
+Also, as transitions between data-race preemption points may have as few as 1 instruction each,
+Landslide caps the average from below at a minimum of 1000 ({\tt PROGRESS\_MIN\_TRIGGER\_AVERAGE})
+to keep the average relatively stable.
+%(100 was found, empirically, to be not enough).
+
+These checks are both implemented in {\tt check\_infinite\_loop()} in {\tt landslide.c}.
+
+\subsubsection{Yield-loop detection}
+
+One very common student implementation pattern in P2s and kernels is to open-code
+synchronization between threads using a loop that spins around a condition that the loop itself cannot fulfill,
+waiting for another thread to allow it to proceed,
+rather than using the established synchronization API%
+\footnote{Even when the student doesn't open-code any such synchronization and always uses {\tt deschedule}
+or primitives built thereupon,
+{\tt mutex\_test} (\sect{\ref{sec:education-pebbles-tests}})
+often relies on this functionality, as data-race preemption points within {\tt mutex\_lock()}
+would otherwise disrupt Landslide's ability to recognize blocking on a contended lock.}.
+If Landslide failed to recognize that the thread was in principle blocked,
+just as if it had called {\tt deschedule} or {\tt cond\_wait()},
+this would result in an infinitely deep branch
+as it keeps trying to schedule the waiting thread,
+blocking all of its attempted context switches to the thread that could make progress.
+These loops need not necessarily {\tt yield} each iteration:
+they may spin blindly,
+assuming progress is being made on another CPU,
+which is not possible under Landslide as it serializes execution.
+
+Landslide detects such ad-hoc {\em yield-loop blocking}
+by keeping a counter for each thread to track how many {\tt yield}s it has invoked since the last ``interesting'' activity.
+Interesting here is defined heuristically as any other known P2 API call,
+with the exception of {\tt mutex\_lock()} and {\tt mutex\_unlock()}%
+\footnote{Landslide must recognize yield-blocking loops that contain mutex operations
+to allow for open-coded reimplementations of {\tt cond\_wait()},
+% see comment above "USER_MUTEX_YIELD_ACTIVITY" in user sync.c
+which for example the {\tt paraguay} test uses intentionally,
+because it is testing the correctness of the student's {\tt cond\_wait()}.}.
+Whenever this counter reaches the heuristic cutoff of 10 ({\tt TOO\_MANY\_YIELDS}),
+Landslide declares the thread blocked,
+and treats it just as if it had invoked {\tt deschedule()} for purposes of DPOR
+({\tt check\_user\_yield\_activity()} in {\tt user\_sync.c}).
+% this also involves propagating yield-blockedness back up the branch, along those 10 pps where it was still counting up to 10
+% and also refusing to run any other threads while yield is going on - KEEP_RUNNING_YIELDING_THREADS
+%
+In cases where {\tt yield} itself is not involved,
+Landslide also counts the number of atomic instructions ({\tt xchg}, {\tt xadd}, {\tt cmpxchg}, etc),
+and marks the thread blocked if it exceeds 100 such ({\tt TOO\_MANY\_XCHGS\_TIGHT\_LOOP})%
+\footnote{If preemption points exist in between, instead only 20 such ({\tt TOO\_MANY\_XCHGS\_WITH\_PPS}),
+to avoid stressing DPOR's $O(n)$ independence computation.}.
+When such a loop contains neither {\tt yield} nor atomics,
+Landslide falls back on the standard infinite loop detector and reports a bug directly, as described above.
+Future work could extend this to include more modern ways of establishing memory safety such as
+acquire/release barriers
+%\cite{sully-thesis}
+and hardware transactions.
+
+In order to detect when a thread should be unblocked from its yield loop,
+Landslide simply leverages DPOR's existing computation of memory conflicts:
+whatever condition the blocked thread was waiting for
+will show up in the conflicts between it and whichever thread fulfills it.
+Hence, every memory conflict detected during DPOR is also checked against any
+currently yield-loop-blocked threads, unblocking them in the case of a match
+({\tt check\_unblock\_yield\_loop()} in {\tt user\_sync.c}).
+If that memory access is not sufficient to let the blocked thread start making progress again,
+it will simply trigger the yield-blocking heuristic again.
+In theory, this could result in a livelock between two threads,
+each in principle blocked in a yield loop,
+but where the conditions of the loop happen to conflict with each other,
+causing the threads to keep waking each other up;
+however,
+% hmm... is this true? like if they use mutexes on each other or smth, or use HTM to check a condition
+% where the htm failure path uses a stop the world thing...
+I have never observed this in practice, as the conditions of such blocking loops are usually read-only.
+Future work could heuristically address this by deprioritizing such threads,
+as measured by the number of times they've yield-blocked,
+so that a third thread which could actually make progress may run if it exists.
+
+% mutex learning
+% to unblock from open coded accesses in eg unlock_and_vanish
+% this is more for blocked_on_addr stuff, not yield blocked stuff
+% kind of clunky and uninteresting to talk abt tbh
+
+\subsubsection{False-positive deadlock avoidance}
+
+In cases where the yield-loop heuristic described above produces false positives,
+i.e., blocking a thread which could make progress on its own after all,
+Landslide must avoid reporting a deadlock bug if no other thread ends up waking it up through memory conflicts.
+After all other threads quiesce,
+which ordinarily would trigger deadlock detection,
+Landslide checks the system for any yield-looping threads that were blocked heuristically%
+\footnote{This includes threads blocked on specific mutexes using the {\tt blocked\_on\_addr} field,
+which is set when {\tt yield} is invoked within {\tt mutex\_lock()} without waiting for 10 loops.}.
+If any exist, it forces them awake and allows them to proceed;
+if they are truly blocked in principle,
+they will merely trip the yield-loop limit and go back to sleep again.
+
+Two other forms of heuristic blocking exist in Landslide which are also subject to this retry procedure.
+Firstly, when an interleaving has already exhausted the number of preemptions allowed under ICB's current bound,
+other threads which are runnable but which require preemptions to switch to are considered ``ICB-blocked''.
+When yield-looping mixes with ICB-blocking, the yielding thread will require a preemption
+for the other thread to fulfill its blocking condition.
+In such a case, simply spending all 128 deadlock-avoidance retries on the yield-looping thread will not solve anything,
+so the ICB-blocked thread must be forced awake with higher priority,
+disregarding the preemption bound,
+to allow the system to progress.
+Secondly, in programs which use HTM,
+threads blocked by retry sets (\sect{\ref{sec:tm-retrysets}})
+must be forced awake with priority before yield-blocked threads,
+along similar reasoning.
+
+Landslide will retry this process up to a heuristic limit of 128 times ({\tt DEADLOCK\_FP\_MAX\_\allowbreak()ATTEMPTS}
+before issuing a deadlock bug report.
+The overhead of this check is linear time in the number of retries permitted,
+although this is typically negligible compared to the exponential size of state spaces overall.
+If the heuristic limit is too small, very ``loopy'' programs
+(which invoke xchg or yield therein)
+could falsely exhaust this limit while not being truly deadlocked,
+so a good limit should be well higher than the total number of concurrency events expected for any Landslide-friendly test.
+The cost of a high limit manifests in the length of preemption traces when deadlock is declared,
+which will display a proportional number of meaningless preemption points,
+although future work could easily truncate them retrospectively.
+This check is implemented in {\tt try\_avoid\_fp\_deadlock()} in {\tt arbiter.c}.

### 2018-10-08 thesis/thesis.tex
add page count figure after conclusion

@@
 to integrate Landslide as an official part of the course grading infrastructure,
-so in some sense the na\"{i}ve wish to make an ``automatic TA'' that first made me interested in grad school
+so in some sense the na\"{i}ve wish to make an ``automatic TA'' that drew my attention to grad school to begin with
 has also been fulfilled.
@@
 
+% TODO TODO TODO
+% update this before you turn in duh
+% log in to office machine and be like, ./graph.py --out graph.svg logshort.csv
+% where logshort is log.csv but you did 'uniq' on it so the output datapoint set is not humongous
+\newpage
+\thispagestyle{empty}
+\begin{center}
+\begin{tabular}{c}
+\vspace{12em} \\
+\includegraphics[width=\textwidth]{pages.pdf}
+\end{tabular}
+\end{center}
+
 %\appendix

### 2018-10-08 thesis/overview.tex
new quote for bgnd
syntax hilight code figures in bgnd

@@
 
-\inspirationalquote{We pray to our ancestors. We've met all of them. None of them were better than anyone else.}
-{Mina, Iconoclasts}
+%\inspirationalquote{We pray to our ancestors. We've met all of them. None of them were better than anyone else.}
+%{Mina, Iconoclasts}
+
+\begin{flushright}
+	{
+	\fontspec[Path=./fonts/]{AlexaStd}
+	\begin{tabular}{p{0.49\textwidth}}
+	{\em "Well I don't *feel* brave. I feel scared, and insignificant."}
+	\\
+	{\em "That's what heroes always say, at first."}
+	\end{tabular}
+	}
+	\vspace{2em}
+\end{flushright}
 
@@
 		\begin{tabular}{p{0.45\textwidth}p{0.5\textwidth}}
-			{\footnotesize
+			{\small
 			\begin{tabular}{l}
-				\texttt{int x;} \\
-				\texttt{void count() \{} \\
-				\texttt{~~~~for (int i = 0; i < 1000; i++)} \\
+				\texttt{\ctype{int} x;} \\
+				\texttt{\ctype{void} \call{count}() \{} \\
+				\texttt{~~~~\call{for} (\ctype{int} i = \const{0}; i < \const{1000}; i++)} \\
 				\texttt{~~~~~~~~x++;} \\
 				\texttt{\}} \\
-				\texttt{void main() \{} \\
-				\texttt{~~~~tid1 = thr\_create(count);} \\
-				\texttt{~~~~tid2 = thr\_create(count);} \\
-				\texttt{~~~~thr\_join(tid1);} \\
-				\texttt{~~~~thr\_join(tid2);} \\
-				\texttt{~~~~printf("\%d\textbackslash{}n", x);} \\
+				\texttt{\ctype{void} \call{main}() \{} \\
+				\texttt{~~~~tid1 = \call{thr\_create}(count);} \\
+				\texttt{~~~~tid2 = \call{thr\_create}(count);} \\
+				\texttt{~~~~\call{thr\_join}(tid1);} \\
+				\texttt{~~~~\call{thr\_join}(tid2);} \\
+				\texttt{~~~~\call{printf}(\const{"\%d\textbackslash{}n"}, x);} \\
 				\texttt{\}} \\
@@
 			&
-			{\footnotesize \begin{tabular}{ll}
-				{\bf \normalsize Thread 1} & {\bf \normalsize Thread 2} \\
+			{\small \begin{tabular}{ll}
+				{\bf Thread 1} & {\bf Thread 2} \\
 				\hline
@@
 	\begin{tabular}{p{0.45\textwidth}p{0.5\textwidth}}
-		{\footnotesize
+		{\small
 		\begin{tabular}{l}
-			\texttt{typedef struct mutex \{} \\
-			\texttt{~~~~volatile int held;} \\
-			\texttt{~~~~int owner;} \\
-			\texttt{\} mutex\_t;} \\
-			\texttt{void mutex\_lock(mutex\_t *mp) \{}\\
-			\texttt{~~~~while (xchg(mp->held, 1))} \\
-			\texttt{~~~~~~~~yield(mp->owner);} \\
-			\texttt{~~~~mp->owner = gettid();} \\
+			\texttt{\ctype{typedef struct mutex} \{} \\
+			\texttt{~~~~\ctype{volatile int} held;} \\
+			\texttt{~~~~\ctype{int} owner;} \\
+			\texttt{\} \ctype{mutex\_t};} \\
+			\texttt{\ctype{void} \call{mutex\_lock}(\ctype{mutex\_t} *mp) \{}\\
+			\texttt{~~~~\flow{while} (\call{xchg}(mp->held, \const{1}))} \\
+			\texttt{~~~~~~~~\call{yield}(mp->owner);} \\
+			\texttt{~~~~mp->owner = \call{gettid}();} \\
 			\texttt{\}} \\
-			\texttt{void mutex\_unlock(mutex\_t *mp) \{}\\
-			\texttt{~~~~mp->owner = -1;} \\
-			\texttt{~~~~mp->held = 0;} \\
+			\texttt{\ctype{void} \call{mutex\_unlock}(\ctype{mutex\_t} *mp) \{}\\
+			\texttt{~~~~mp->owner = \ctype{-1};} \\
+			\texttt{~~~~mp->held = \ctype{0};} \\
 			\texttt{\}} \\
@@
 		&
-		{\footnotesize
+		{\small
 		\begin{tabular}{l}
-			\texttt{int x;} \\
-			\texttt{mutex\_t m;} \\
-			\texttt{void count() \{} \\
-			\texttt{~~~~for (int i = 0; i < 1000; i++) \{} \\
-			\texttt{~~~~~~~~mutex\_lock(\&m);} \\
+			\texttt{\ctype{int} x;} \\
+			\texttt{\ctype{mutex\_t} m;} \\
+			\texttt{\ctype{void} \call{count}() \{} \\
+			\texttt{~~~~\flow{for} (\ctype{int} i = \const{0}; i < \const{1000}; i++) \{} \\
+			\texttt{~~~~~~~~\call{mutex\_lock}(\&m);} \\
 			\texttt{~~~~~~~~x++;} \\
-			\texttt{~~~~~~~~mutex\_unlock(\&m);} \\
+			\texttt{~~~~~~~~\call{mutex\_unlock}(\&m);} \\
 			\texttt{~~~~\}} \\
@@
 
-\begin{figure}[h]
+\begin{figure}[t]
 	\begin{center}
+		\small
 		\begin{tabular}{l}
@@
 \begin{tabular}{rll}
-        & \multicolumn{2}{c}{\texttt{int x = 0; bool y = false; mutex\_t mx;}} \\
+	& \multicolumn{2}{c}{\texttt{\ctype{int} x = \const{0}; \ctype{bool} y = \const{false}; \ctype{mutex\_t} mx;}} \\
         \\
         & {\bf Thread 1} & {\bf Thread 2} \\
-        1 & \texttt{\hilight{brickred}{x++;}~// A1} & \\
-        2 & \texttt{mutex\_lock(\&mx);} & \\
-        3 & \texttt{mutex\_unlock(\&mx);} & \\
-        4 & & \texttt{mutex\_lock(\&mx);} \\
-        5 & & \texttt{mutex\_unlock(\&mx);} \\
-        6 & & \texttt{\hilight{brickred}{x++;}~// A2} \\
+	1 & \texttt{\hilight{brickred}{x++;}~\ccomment{// A1}} & \\
+	2 & \texttt{\call{mutex\_lock}(\&mx);} & \\
+	3 & \texttt{\call{mutex\_unlock}(\&mx);} & \\
+	4 & & \texttt{\call{mutex\_lock}(\&mx);} \\
+	5 & & \texttt{\call{mutex\_unlock}(\&mx);} \\
+	6 & & \texttt{\hilight{brickred}{x++;}~\ccomment{// A2}} \\
 \end{tabular}
@@
         & {\bf Thread 1} & {\bf Thread 2} \\
-        1 & \texttt{\hilight{brickred}{x++;}~// B1} & \\
-        2 & \texttt{mutex\_lock(\&mx);} & \\
-        3 & \texttt{y = true;} & \\
-        4 & \texttt{mutex\_unlock(\&mx);} & \\
-        5 & & \texttt{mutex\_lock(\&mx);} \\
-        6 & & \texttt{bool tmp = y;} \\
-        7 & & \texttt{mutex\_unlock(\&mx);} \\
-        8 & & \texttt{if (tmp) \hilight{brickred}{x++;}~// B2} \\
+	1 & \texttt{\hilight{brickred}{x++;}~\ccomment{// B1}} & \\
+	2 & \texttt{\call{mutex\_lock}(\&mx);} & \\
+	3 & \texttt{y = \const{true};} & \\
+	4 & \texttt{\call{mutex\_unlock}(\&mx);} & \\
+	5 & & \texttt{\call{mutex\_lock}(\&mx);} \\
+	6 & & \texttt{\ctype{bool} tmp = y;} \\
+	7 & & \texttt{\call{mutex\_unlock}(\&mx);} \\
+	8 & & \texttt{\flow{if} (tmp) \hilight{brickred}{x++;}~\ccomment{// B2}} \\
         %8 & & \texttt{if (tmp)} \\

### 2018-10-08 thesis/future.tex
try not to aggro garth so much abt ml
change future quote

@@
 \label{chap:warpzone}
-% TODO: ask for permission for this one
-% https://twitter.com/Poem4your_sprog/status/878747092788838400
 \inspirationalquote{
-\begin{tabular}{p{0.51\textwidth}}
-A little bit of being kind,
-a tiny open door; \\
-A nicer slice of peace of mind
-begets a little more. \\
-It won't correct the world tonight,
-nor change tomorrow too; \\
-But maybe if you do it right,
-you'll find it changes you.
+\begin{tabular}{p{0.75\textwidth}}
+We can't be free until we learn to laugh at ourselves.
+Once you look in the mirror and see just how foolish we can be,
+laughter is inevitable.
+And from laughter comes wisdom.
 \end{tabular}}
-{Sam "Poem\_for\_your\_sprog" Garland}
+{G'Kar, Babylon 5}
 
@@
 
+% TODO: does this need some estimate progress graphs
+% some RE ones would even need a log scale y axis maybe
 In my experience,
@@
 % if your ML provides a partial verification, it is impossible for a human to understand what that verification means
-As a matter of personal philosophy,
-I would strongly caution against using machine learning to try to expose bugs faster.
+
+%% commenting this first part out bc it will probably aggro garth...
+%As a matter of personal philosophy,
+%I would strongly caution against using machine learning to try to expose bugs faster.
 Machine learning has become popular recently as a magic cure-all for many programming problems,

### 2018-10-09 thesis/future.tex
rearrange quotes and add one from 17776

@@
 \inspirationalquote{
-\begin{tabular}{p{0.75\textwidth}}
-We can't be free until we learn to laugh at ourselves.
-Once you look in the mirror and see just how foolish we can be,
-laughter is inevitable.
-And from laughter comes wisdom.
+	\begin{tabular}{p{0.84\textwidth}}
+	"If you are a god, Zeus, as the stories claim, then why did you create evolution?
+	Why did you make a world that can only grow through cruelty and pain?"
+	\\
+	For a moment\inspirationalhyphen{}surely this meant death was near\inspirationalhyphen{}she thought she heard him answer:
+	"My child, I was shaped by the gods that came before me, as you were shaped by me.
+	The choice I had was between creation and oblivion, life and death.
+	And I chose life, because any life is better than no life,
+	because as long as there is life, there is hope\inspirationalhyphen{}if not for us, then for some generation to come."
+	\\
+	"Then how are you a god, if you can offer so little?" she whispered, feeling death creep closer.
+	\\
+	"I am a god because I take upon myself the burden of creation," the statue replied.
+	\\
+	"Then we are all gods," Alexandra said, and pushed the button.
 	\end{tabular}}
-{G'Kar, Babylon 5}
+%{Galatea, The Talos Principle}
+{The Talos Principle}
 

### 2018-10-09 thesis/thesis.tex
i have a date now
real keywords
rearrange quotes and add one from 17776

@@
 \author{Ben Blum}
-\date{Decembruary 2018}
+\date{October 2018}
 \Year{2018}
@@
 
-\keywords{landslide terminal, baggage claim, ground transportation, ticketing}
+%\keywords{landslide terminal, baggage claim, ground transportation, ticketing}
+\keywords{concurrency, debugging, verification, model checking, data races, education, transactional memory}
 
@@
 \inspirationalquote{
-	\begin{tabular}{p{0.84\textwidth}}
-	"If you are a god, Zeus, as the stories claim, then why did you create evolution?
-	Why did you make a world that can only grow through cruelty and pain?"
-	\\
-	For a moment\inspirationalhyphen{}surely this meant death was near\inspirationalhyphen{}she thought she heard him answer:
-	"My child, I was shaped by the gods that came before me, as you were shaped by me.
-	The choice I had was between creation and oblivion, life and death.
-	And I chose life, because any life is better than no life,
-	because as long as there is life, there is hope\inspirationalhyphen{}if not for us, then for some generation to come."
-	\\
-	"Then how are you a god, if you can offer so little?" she whispered, feeling death creep closer.
-	\\
-	"I am a god because I take upon myself the burden of creation," the statue replied.
-	\\
-	"Then we are all gods," Alexandra said, and pushed the button.
+\begin{tabular}{p{0.6\textwidth}}
+"This is the end, right? The end of this story?"
+
+"We were always at the end. It's a free play, buddy. Clock's all zeroes."
 \end{tabular}}
-%{Galatea, The Talos Principle}
-{The Talos Principle}
+%{Nine \& Ten, 17776}
+{17776}
 

### 2018-10-11 thesis/defense/snakefight.tex
start writing snake fight talk

--- 2017-04-25 proposal/talk/proposal.tex
@@
 
-\title[Thesis Proposal]{{\bf Practical Concurrency Testing}}
-\subtitle[]{a thesis proposal}
-\author[Ben Blum]{Ben Blum \texttt{(bblum@cs.cmu.edu)}}
+\title[]{{\bf Practical Concurrency Testing}}
+\subtitle[]{or, \\ How I Learned to Stop Worrying and Love the Exponential Explosion}
+\author[]{Ben Blum \texttt{(bblum@cs.cmu.edu)}}
 
-\institute[CMU CSD]{Carnegie Mellon University}
-\date[]{2017, April 25}
+\institute[]{Carnegie Mellon University}
+\date[]{2018, October 17}
 
@@
 
-\section{Introduction}
+\section{Part 0}
 \subsection{Motivation}
@@
 		\item Manual kernel annotations: Landslide {\em [Blum '12]} {\tiny that's me!}
+			% TODO: smth about choosing pps?
 		\item State space size estimation {\em [Simsa '12]}
+			% TODO change this bullet pt to be more explicitly about 
 		\item Widespread adoption requires {\bf fully automated instrumentation}!
@@
 		\item Event-driven applications {\em [Jensen '15]}
-		\item {\bf Transactional concurrency}?
+		\item {\bf Hardware transactional memory}?
 	\end{itemize}
@@
 \begin{frame}{Thesis Statement}
-	Thanks to the new algorithms, heuristics, and concurrency models I have developed,
-	stateless model checking is an appropriate and accessible concurrency testing technique
-	for programmers in both educational and real-world settings.
+	% data race pps (as well as txn equivalence)
+	Combining both theoretically-founded automatic reduction techniques
+	% iterative deepening
+	% (say smth like: "landslide also has a wealth of other user-tunable heuristics
+	% for expert users, which you can read about in the document; for this talk i will focus just on the CPU budget thing")
+	% XXX: need to prepare for the question -- "what evidence do you have that users actually want this cpu budget stuff",
+	% ",or, that it helps vs just running something unlimited vs a normal icb test"
+	% (well, for the 2nd half of the Q, i can talk about how ID outperforms ICB on verficiations)
+	and user-informed heuristic ones,
+	stateless model checking can sufficiently mitigate exponential explosion to be
+	% education
+	a practical testing technique for inexperienced users
+	% htm
+	% TODO: this last part you just added for this talk; does the thesis version of the statmeent need to include too?
+	and real-world programs alike.
 \end{frame}
@@
 
-\section{Quicksand}
+\section{Part 1}
+\subsection{Quicksand}
 
-\breakslide{\Large Iterative Deepening of State Spaces}
+\breakslide{\Large Quicksand}
 
@@
 
+\begin{frame}{Data Race Analysis}
+	A {\bf data race} is a pair of memory accesses between two threads, where:
+	\begin{itemize}
+		\item At least one of the accesses is a write
+		\item The threads are not holding the same mutex
+		\item One access {\em happens-before} the other
+			(e.g., no \texttt{cond\_signal()} in between)
+	\end{itemize}
+	\pause
+	\linegap
+
+	Two flavors
+	\begin{itemize}
+		\item {\bf Pure HB}: Whether accesses were {\em observed simultaneously}, false-negative prone
+			{\em [Lamport '78]}
+		\item {\bf Limited HB}: Whether threads {\em could be reordered}, false-positive prone
+			{\em [O'Callahan '03]}
+	\end{itemize}
+\end{frame}
+
 \begin{frame}{Iterative Deepening (in Pictures)}
@@
 
+% TODO put a pict of mario-man, i.e maximal state space
+
 \begin{frame}{Total Verification}
@@
 		\item Insight: $\forall$ arbitrary buggy interleaving, $\exists$ equivalent interleaving % say 'under DPOR' i guess?
-			using only sync/data-race PPs.
+			using only sync/data-race PPs, \& Quicksand will eventually reach it.
 	\end{itemize}
@@
 				\item ...synchronization APIs only {\em [Simsa '10]}
-				\item ...all shared memory accesses {\em [Holzmann '97]}
+				\item ...all shared memory accesses
+					%{\em [Holzmann '97]} % not spin
+					{\em [Yang '08]} % inspect
 			\end{itemize}
@@
 
-\begin{frame}{Deliverables}
-	In the thesis, I will:
-	\begin{itemize}
-		\item Restate the paper {\large \smiley}
-		\item Study variance of repeated Quicksand experiment trials
-	\end{itemize}
+\begin{frame}{Summary}
+	Incorporate data-race analysis to add new PPs in a feedback loop
+	\linegap
+
+	Enough time to saturate \& test every data-race PP? Full verification
+	\linegap
+
+	Otherwise? Quicksand's heuristic search ordering finds bugs faster anyway
+	\pause
+	\linegap
+
+	%Future work
+	%\begin{itemize}
+	%	\item Extend proof to weak memory
+	%	\item Incorporate ICB in Quicksand
+	%	\item Improve estimation accuracy
+	%\end{itemize}
 \end{frame}
@@
 
-\section{Education}
+\section{Part 2}
+\subsection{Education}
 
-\breakslide{\Large Educational Use}
+\breakslide{\Large Education}
 
-\begin{frame}{Pebbles}
-	CMU (15-410)
-	\linegap
-
-	Kernel (``P3'')
+\begin{frame}{Setting}
+	CMU (15-410, S'15-S'18), PSU (CMPSC 473, S'18)
 	\begin{itemize}
-		%\item UNIX-like educational kernel specification
-		\item Students implement scheduler, synchronization, VM %from scratch %% trying not to be too arrogant ._.
-		\item UNIX-like process lifecycle ({\tt fork}, {\tt exec}, {\tt exit}, {\tt wait})
-		%\item Teams of 2
-		\item Testing with MC requires manual annotation {\em [Blum '12]}
+		\item Thread library (``P2'')
+			\begin{itemize}
+				\item POSIX-like userspace thread library %(also teams of 2)
+				\item Reference ``Pebbles'' kernel provided
+					% say: later they implement this in p3
+				\item Fixed APIs $\rightarrow$ instrumentation can be automated
+				\item Still needs heuristics to handle ad-hoc synchronization
+			\end{itemize}
+			% FIXME: need introduce p3 here?
 	\end{itemize}
@@
 
-	Thread library (``P2'')
+	U. Chicago (CMSC 23000, F'17)
 	\begin{itemize}
-		\item POSIX-like userspace thread library %(also teams of 2)
-		\item Students provided reference Pebbles implementation
-		\item Fixed APIs $\rightarrow$ instrumentation can be automated
-		\item May still need heuristics to handle ad-hoc synchronization
+		\item Pintos %(``threads'' and ``userprog'')
+		\begin{itemize}
+			\item Basecode provides round-robin scheduler, synchronization, VM, loader
+			\item ``threads'': Students implement priority scheduling, sleep
+		%\end{itemize}
+		%\linegap
+
+		%\item Processes (``userprog'')
+		%\begin{itemize}
+				% fork is part of exec and the loader is already provided
+			\item ``userprog'': UNIX-like process lifecycle
+				%({\tt exec},
+				({\tt exit}, {\tt wait})
+			%\item File descriptor management ({\tt create}, {\tt remove}, {\tt open}, {\tt read}, ...)
+				% filesys project is beyond scope of landslide
+				% DONT say: the u chicago professor chose to discard the hard concurrency parts
+				% DO say, if pressured: device driver concurrency would require more engineering,
+				% so we chose to limit it to these two projex
+		\end{itemize}
 	\end{itemize}
@@
 
-\begin{frame}{15-410 User Study}
-	Method (since spring 2015)
+\begin{frame}{Instrumentation}
+	% Unlike kernels [Blum '12], P2 can be......
+	% TODO talk abt yield loops, 
+	% compare to prior technique - delay abounding - which sucks bc (verify this by reading BPOR) it doesn't actually
+	% wake up the other thread at the right time (using memory conflix)
+
+	% pintos - bunch of gross patch stuff lmao
+\end{frame}
+
+% do even need this slide?
+%\begin{frame}{Evaluation Questions}
+%	asfda % TODO
+%	% submit better projects?
+%	% get better grades?
+%	% like the experience?
+%\end{frame}
+
+\begin{frame}{Method}
+	P2 (CMU + PSU)
 	\begin{itemize}
-		\item Guest lecture midway during P2
-		\item Landslide (+ Quicksand) provided with 6 test cases and user guide
-		\item Student code snapshots and bugs found recorded each use
+		\item Guest lecture midway during project
+		\item Landslide (+ Quicksand) provided to students with 6 test cases %and user guide
+		\item Recorded code snapshots and bugs found each use (CMU only)
+		\item Surveyed students on usability, false positives, etc.
 	\end{itemize}
@@
 
-	Results (as of spring 2016)
+	Pintos (U. Chicago)
 	\begin{itemize}
-		\item 47/90 groups volunteered
-		\item 85 concurrency bugs found; 53 which students then fixed
-		\item 32/47 found $\ge 1$ concurrency bug, 22/47 fixed $\ge 1$
-		\item No visible impact on ultimate grades...
+		\item TA %(Kevin Zhao)
+			ran Landslide on {\em threads} and {\em userprog} submissions
+		\item Returned preemption traces to students after deadline
+		\item Surveyed students on ease of debugging, etc.
+	\end{itemize}
+	% TODO: irb stuff in a bonus slide
+\end{frame}
+
+\begin{frame}{Results}
+	Participation
+	\begin{itemize}
+		\item CMU: 103/177 students (7 semesters)
+		\item PSU: 38/136 students (1 semester)
+		\item U. Chicago: 4/21 students (1 semester)
+	\end{itemize}
+	\linegap
+
+	Bug-finding
+	\begin{itemize}
+		\item CMU: 115 deterministic, 122 concurrency bugs in 75/103 P2s
+		\item U. Chicago: 3 deterministic, 4 concurrency bugs in 3/4 kernels
+		% \item Case study tej
 	\end{itemize}
@@
 
-\begin{frame}{15-410 User Study - Grades Results} % lack thereof
+\begin{frame}{Effect on Grades}
+	Compared Landslide users' vs. non-users' submitted P2 grades...
+	\begin{itemize}
+		\item ``Does Landslide help students submit better projects?''
+	\end{itemize}
+	\linegap
+
+	...as well as subsequent P3 (kernel) grades
+	\begin{itemize}
+		\item ``Does Landslide teach students lasting lessons about concurrency?''
+	\end{itemize}
+\end{frame}
+
+\begin{frame}{Effect on Grades}
 	\begin{center}
-		\includegraphics[width=\textwidth]{../p2-p3-distribution.pdf}
+		\includegraphics[width=\textwidth]{../photo-of-ze-studence.pdf}
 	\end{center}
@@
 
-%%%%
+\begin{frame}{Effect on Grades}
+	\begin{center}
+		\includegraphics[width=\textwidth]{../photo-of-ze-studence-cdf.pdf}
+	\end{center}
+\end{frame}
 
-\begin{frame}{Pintos}
-	U. of Chicago (CMSC 23000), Berkeley (CS162), Stanford (CS140)
+\begin{frame}{Statistical Significance}
+	Compared users' vs. non-users' grade distributions with k-sample Anderson-Darling {\em [Anderson, Darling '52]}
 	\linegap
 
-	Kernel (``threads'')
+	Control 1: non-users from {\bf same semester}
 	\begin{itemize}
-		\item Basecode provides round-robin scheduler, synchronization
-		\item Students implement priority scheduling, sleep
+		\item {\bf P2} grade distributions differ with $p = 0.036 < 0.05$
+		\item {\bf P3} grade distributions differ with $p = 0.034 < 0.05$
+			\pause
+		\item Possible bias: stronger students volunteer for Landslide in first place
+	\end{itemize}
+	\pause
+	\linegap
+
+	Control 2: non-users from {\bf past semesters} (when Landslide not offered)
+	\begin{itemize}
+		\item {\bf P2} grade distributions differ with $p = 0.671$...
+		\item {\bf P3} grade distributions differ with $p = 0.145$...
+			\pause
+		\item Possible bias: grading criteria may drift across semesters
+			% FIXME does this bullet belong in this sub list
+		\item Possible bias: maybe Landslide finds bugs the official tests don't
+	\end{itemize}
+\end{frame}
+
+\begin{frame}{Survey}
+	14(ish) questions, investigate both education \& happiness
+	\begin{itemize}
+		\item {\em Education:} Did Landslide help students learn more from P2?
+		\item {\em Happiness:} Did Landslide help students save time/effort?
 	\end{itemize}
@@
 
-	Processes (``userprog'')
+	Self-report any false positives
 	\begin{itemize}
-		\item UNIX-like process lifecycle ({\tt exec}, {\tt exit}, {\tt wait})
-		\item File descriptor management ({\tt create}, {\tt remove}, {\tt open}, {\tt read}, ...)
+		\item Landslide crashes aside, 3 FP bug reports, all simple bugs (now fixed)
+			% nonexistent MAGIC_BREAK bc of different address space in shell
+			% simultaneous landslides corrupt each other
+			% invalid heap access when new_pages "too close"
+		\item No FPs related to yield-loop/etc heuristics
+	\end{itemize}
+\end{frame}
+
+\begin{frame}{Survey Results (educational value)}
+	\begin{center}
+		\includegraphics[width=\textwidth]{../survey6.pdf} % easier to diagnose?
+	\end{center}
+\end{frame}
+
+\begin{frame}{Survey Results (happiness value)}
+	\begin{center}
+		\includegraphics[width=\textwidth]{../survey5.pdf} % easy to understand?
+	\end{center}
+\end{frame}
+
+\begin{frame}{Survey Results (educational value)}
+	\begin{center}
+		\includegraphics[width=\textwidth]{../survey9.pdf} % 100 cpu hours
+	\end{center}
+\end{frame}
+
+\begin{frame}{Survey Results (educational value)}
+	\begin{center}
+		\includegraphics[width=\textwidth]{../survey8.pdf} % learned to understand concurrency
+	\end{center}
+\end{frame}
+
+\begin{frame}{Survey Results (happiness value)}
+	\begin{center}
+		\includegraphics[width=\textwidth]{../survey10.pdf} % recommend to friend
+	\end{center}
+\end{frame}
+
+\begin{frame}{Summary}
+	Landslide helps students find \& fix bugs during projects
+	\begin{itemize}
+		\item 73\% of CMU users found $\ge 1$ bug
+		\item 56\% of CMU users fixed $\ge 1$ bug
+		\item 34\% of CMU users fixed all found bugs
 	\end{itemize}
@@
 
-	VM, filesys projects not as concurrency-focused as above
-\end{frame}
-
-\begin{frame}{Pintos \& Landslide}
-	Existing support:
+	Effect on overall project grades: ???
 	\begin{itemize}
-		\item Data-race analysis
-			%Lockset/vector-clock analyses
-			extended to interrupt-disabling
-		\item Automatically inserts (most) kernel annotations
-		%\item (various engineering challenges) % palloc heap, non-direct-mapped kernelspace, skipping driver init loops
-		\item Often requires manual intervention
+		\item Compensating for experimental bias difficult w/human subjects...
 	\end{itemize}
-	% "Works enough for OOPSLA but not enough for the studence"
 	\linegap
 
-	Still needed:
-	\begin{itemize}
-		\item Improved automation to support corner-case student designs
-		\item More concurrency test cases
-		\item Remove proprietary Simics dependence by porting to Bochs
-	\end{itemize}
-\end{frame}
+	Survey results showed students enlightenedly happy
+	\linegap
 
-\begin{frame}{Deliverables}
-	In the thesis, I will:
-	\begin{itemize}
-		\item Pebbles
-		\begin{itemize}
-			\item Discuss factors keeping Landslide from improving grades
-			%\item Discuss how to make P3 and Landslide compromise to work together
-			\item Study accuracy of automation heuristics
-		\end{itemize}
-		\item Pintos
-		\begin{itemize}
-			\item Open-source port of Landslide to Bochs
-			\item 2 semesters of user study in Pintos class(es)
-			\item Compare incidence/nature of bugs found to Pebbles
-		\end{itemize}
-	\end{itemize}
+	15-410 course staff now including Landslide in official grading :)
 \end{frame}
@@
 
-\section{Transactional Memory}
+\section{Part 3}
+\subsection{Transactional Memory}
 \newcommand\xbegin{\texttt{\_xbegin()}\xspace}
@@
 	\begin{itemize}
-		\item Capture all sites of nondeterminism as ``branch points'' %in a state space
+		\item Capture all sites of nondeterminism as preemption points
 		\item Control nondeterminism to force each possibility at each point
-		\item (Identify equivalences to prune unnecessary branch points)
+		\item (Identify equivalences to prune unnecessary preemption points)
 	\end{itemize}
@@
 			\item Equivalent to a lock protecting that state
+				% note that also inserting DR PPs within a transaction is not allowed
 		\end{itemize}
@@
 
-\begin{frame}{Timeline}
-	\begin{itemize}
-		\item 2017 Apr-Aug: Port Landslide to Bochs for Pintos
-		\item 2017 Apr-Aug: Transactional memory
-		\item 2017 Aug-Oct: Pintos user study
-		\item 2017 Oct-Nov: Pebbles user study (again)
-		\item 2017 Dec:     Start writing thesis
-		\item 2018 Jan-Apr: More writing \& miscellany
-		\item 2018 Feb-Mar: Pintos \& Pebbles (again)
-		\item 2018 Apr-May: Defend \& graduate
-	\end{itemize}
-\end{frame}
-
 \breakslide{\Large Q\&A
@@
 
-\begin{frame}{15-410: Summary of bugs found}
-	\begin{center}
-	\begin{tabular}{r|cc|cc}
-	%det bugs found det bugs fixed  races fixed     races found
-	& \multicolumn{4}{c}{Number of groups} \\
-	Number & \multicolumn{2}{c|}{Deterministic bugs} & \multicolumn{2}{c}{Concurrency bugs} \\
-	of bugs & found & fixed & found & fixed \\
-	\hline
-	0       & 27    & 32    & 15    & 23    \\
-	1       & 6     & 5     & 7     & 11    \\
-	2       & 9     & 6     & 12    & 8     \\
-	3       & 2     & 1     & 7     & 2     \\
-	4       & 1     & 1     & 3     & 1     \\
-	5       & 2     & 2     & 2     & 1     \\
-	11      &       &       & 1     & 1     \\
-	\hline
-	%Total groups
-	%       & 47    & 47    & 47    & 47    \\
-	Total bugs
-		& 44    & 34    & 85    & 53    \\
-	\end{tabular}
-	\end{center}
-	%\linegap
-
-	%E.g.: 11 groups fixed 1 bug + 8 fixed 2 + 2 $\times$ 3 + 1 $\times$ 4 + 1 $\times$ 5 + 1 overachieving group who fixed 11 = 11 + 16 + 6 + 4 + 5 + 11 = 53.
+\begin{frame}{15-410...}
+	% TODO:
+	% study of tej/etc specific bugs
+	% list tests cases provided to studence & PSU-specific ones too
 \end{frame}
@@
 			\texttt{if ((status = \_xbegin()) == \_XBEGIN\_STARTED) \{} \\
-			\texttt{~~~~if (foo < 100)}\\
+			\texttt{~~~~if (count < 100)}\\
 			\texttt{~~~~~~~~\_xabort();} \\
-			\texttt{~~~~foo++;} \\
+			\texttt{~~~~count++;} \\
 			\texttt{~~~~\_xend();} \\
 			\texttt{\} else if (status != \_XABORT\_EXPLICIT) \{} \\
-			\texttt{~~~~mutex\_lock(\&foo\_lock);} \\
-			\texttt{~~~~foo++;} \\
-			\texttt{~~~~mutex\_unlock(\&foo\_lock);} \\
+			\texttt{~~~~mutex\_lock(\&count\_lock);} \\
+			\texttt{~~~~count++;} \\
+			\texttt{~~~~mutex\_unlock(\&count\_lock);} \\
 			\texttt{\}} \\
@@
 		\item Treat such transactions as having multiple failure injection possibilities
-			% XXX: How to identify when the failure status is never used? Dataflow analysis?
+		\item Future work: dataflow analysis to identify when failure reason never used
 	\end{itemize}
@@
 	\begin{itemize}
-		\item Shamay, https://github.com/cmpxchg16/tsx.me, 2014
+		%\item Shamay, https://github.com/cmpxchg16/tsx.me, 2014
 		\item Dehesa-Azuara and Stanley, http://www.contrib.andrew.cmu.edu/\textasciitilde{}mdehesaa/, 2016
-		\item Chapman et al., Hybrid STM/HTM for nested transactions on OpenJDK, OOPSLA 2016
-		\item Wang et al., Eunomia: Scaling Concurrent Search Trees under Contention Using HTM, PPoPP 2017
+		%\item Chapman et al., Hybrid STM/HTM for nested transactions on OpenJDK, OOPSLA 2016
+		%\item Wang et al., Eunomia: Scaling Concurrent Search Trees under Contention Using HTM, PPoPP 2017
+			% TODO: add, that one lock thingy
 	\end{itemize}
@@
 
+% TODO: slide number c.c
 \begin{frame}{Don't you need \texttt{if (mutex\_held(\&foo\_lock)) \_xabort();} in that transaction on slide 31?}
@@
 			VeriSoft. A tool for the automatic analysis of concurrent reactive software. CAV 1997.
-		\item {\bf [Holzmann '97]}: The model checker SPIN. TOSE 1997.
+		%\item {\bf [Holzmann '97]}: The model checker SPIN. TOSE 1997.
 		\item {\bf [Engler '03]}: Dawson Engler and Ken Ashcraft.
@@
 			Programs. OSDI 2008.
+		\item {\bf [Yang '08]}: Yu Yang et al. Efficient stateful dynamic partial order reduction. SPIN 2008.
 	\end{itemize}

### 2018-10-12 thesis/defense/snakefight.tex
snakefight finish educ section and write some tsx, mostly bonus slides

@@
 			% FIXME: need introduce p3 here?
+			% Or cut?
+		\item Kernel (``P3'') - reimplement said reference kernel (CMU only)
 	\end{itemize}
@@
 			\item ``userprog'': UNIX-like process lifecycle
-				%({\tt exec},
-				({\tt exit}, {\tt wait})
+				({\tt exec}, % students don't really implement any complciated part of this though
+				% i think just like. the argument copying??
+				{\tt exit}, {\tt wait})
 			%\item File descriptor management ({\tt create}, {\tt remove}, {\tt open}, {\tt read}, ...)
@@
 \begin{frame}{Instrumentation}
+	Understanding concurrency primitives/APIs
+	\begin{itemize}
+		\item Kernel-level testing requires manual annotation {\em [Blum '12]}
+		\item P2 (fixed APIs), Pintos (basecode provided) annotated automatically
+	\end{itemize}
+	\pause
+	\linegap
+
+	Ad-hoc synchronization (e.g. {\tt while (!ready) yield();})
+	\begin{itemize}
+		\item Not an infinite loop bug!
+		\item Heuristically ``block'' threads when 10+ {\tt yield()}s %({\tt xchg}, etc.)
+		\item Existing memory analysis %(DPOR)
+			on loop condition to unblock
+			\begin{itemize}
+					% TODO: cut this and just say it out loud?
+				\item (e.g., what other thread writes to {\tt ready}?)
+				% it's kinda... idk how to critique its unsoundness.
+				% they self-admit to the unsoundness but like, leaving the cycle in,
+				% is the wrong approach to begin with
+				\item Prior work {\em [Coons '13]} likely to leave threads blocked too long
+			\end{itemize}
+		\item Check for false-positive blocking to avoid false-positive deadlocks
+	\end{itemize}
+
 	% Unlike kernels [Blum '12], P2 can be......
@@
 	\begin{center}
+		\includegraphics[width=\textwidth]{photo-of-ze-studence-cdf-no-pre-study.pdf}
+	\end{center}
+\end{frame}
+
+\begin{frame}{Effect on Grades}
+	\begin{center}
 		\includegraphics[width=\textwidth]{../photo-of-ze-studence-cdf.pdf}
@@
 \begin{frame}{Statistical Significance}
-	Compared users' vs. non-users' grade distributions with k-sample Anderson-Darling {\em [Anderson, Darling '52]}
+	Compared %users' vs. non-users'
+	grade distributions with k-sample Anderson-Darling {\em [A., D. '52]}
 	\linegap
@@
 		\begin{tabular}{l}
-			\texttt{if ((status = \_xbegin()) == \_XBEGIN\_STARTED) \{} \\
+			\texttt{if ((status = \_xbegin()) == SUCCESS) \{} \\ %\_XBEGIN\_STARTED) \{} \\
 			%\texttt{} \\
 			%\texttt{} \\
-			\texttt{~~~~foo++;} \\
+			\texttt{~~~~count++;} \\
 			\texttt{~~~~\_xend();} \\
 			\texttt{\} else \{}\\ %if (status != \_XABORT\_EXPLICIT) \{} \\
-			\texttt{~~~~mutex\_lock(\&foo\_lock);} \\
-			\texttt{~~~~foo++;} \\
-			\texttt{~~~~mutex\_unlock(\&foo\_lock);} \\
+			\texttt{~~~~mutex\_lock(\&count\_lock);} \\
+			\texttt{~~~~count++;} \\
+			\texttt{~~~~mutex\_unlock(\&count\_lock);} \\
 			\texttt{\}} \\
@@
 		\item Code between \xbegin and \xend is provisionally executed
-		\item All writes kept in cache until \xend
+		\item All writes kept in CPU cache until \xend
 		\item \xend commits changes atomically (w.r.t. other CPUs)
@@
 		\item Control nondeterminism to force each possibility at each point
-		\item (Identify equivalences to prune unnecessary preemption points)
+			\begin{itemize}
+				\item (identifying equivalences to prune redundant branches)
+			\end{itemize}
 	\end{itemize}
@@
 
+\begin{frame}{Example}
+	\begin{center}
+		\begin{tabular}{l|l}
+			{\bf Thread 1} & {\bf Thread 2} \\
+			\hline
+			{\tt \_xbegin() == SUCCESS} \\
+			{\tt count++;} \\
+				& {\tt \_xbegin() == SUCCESS} \\
+				& {\tt count++;} \\
+				& {\tt xend();} \\
+			{\tt xend()}...? \\
+			{\em <conflict detected>} \\
+			\\
+			{\tt \_xbegin() == FAIL} \\
+			{\tt mutex\_lock(...);} \\
+			{\tt count++;} \\
+			{\tt mutex\_unlock(...);} \\
+		\end{tabular}
+	\end{center}
+\end{frame}
+
 \begin{frame}{Concurrency model}
@@
 	\end{itemize}
+	\pause
+	\linegap
+
+	Equivalence theorem
+	\begin{itemize}
+		\item Safe to automatically skip all interleavings within transactions
+			\begin{itemize}
+				\item Exponential state space reduction
+				\item Traditional reduction (DPOR) applied on top
+			\end{itemize}
+		\item More in bonus slides...
+	\end{itemize}
 \end{frame}
@@
 	\begin{center}
+		\includegraphics[width=\textwidth]{failure-injection-1.pdf}
+	\end{center}
+\end{frame}
+\begin{frame}{Failure Injection - 2 dimensions of concurrency}
+	\begin{center}
 		\includegraphics[width=\textwidth]{failure-injection-2.pdf}
@@
 
-\begin{frame}{Deliverables}
-	In the thesis, I will:
-	\begin{itemize}
-		\item Implement the HTM(-equivalent) concurrency model in Landslide
-		\item Extend the model to support STM
-		\item Prove equivalence to HTM/STM semantics % under MC
-		\item Bugfind/verify a suite of unit tests \& real-world TM programs
-	\end{itemize}
-\end{frame}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{End}
-
-\breakslide{\Large Q\&A
+\begin{frame}{Evaluation}
+\end{frame}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{Conclusion}
+
+\breakslide{
+\begin{center}
+	% TODO: replace with HIL for honban
+	\Large Q\&A
+	%\includegraphics[width=0.4\textwidth]{../how-i-learned.png}
+\end{center}
 %\linegap
@@
 
-\section{Bonus Slides}
-
-\breakslide{\Large Bonus Slides}
-
-\begin{frame}{15-410...}
-	% TODO:
-	% study of tej/etc specific bugs
-	% list tests cases provided to studence & PSU-specific ones too
-\end{frame}
-
-\begin{frame}{HTM: Even more dimensions of nondeterminism}
-	% The gotcha: When xbegin can fail, it can emit diff error codes, which depend on what happened within.
-	\begin{center}
-		\begin{tabular}{l}
-			\texttt{if ((status = \_xbegin()) == \_XBEGIN\_STARTED) \{} \\
-			\texttt{~~~~if (count < 100)}\\
-			\texttt{~~~~~~~~\_xabort();} \\
-			\texttt{~~~~count++;} \\
-			\texttt{~~~~\_xend();} \\
-			\texttt{\} else if (status != \_XABORT\_EXPLICIT) \{} \\
-			\texttt{~~~~mutex\_lock(\&count\_lock);} \\
-			\texttt{~~~~count++;} \\
-			\texttt{~~~~mutex\_unlock(\&count\_lock);} \\
-			\texttt{\}} \\
-		\end{tabular}
-	\end{center}
-	\linegap
-
-	\xbegin can fail with multiple possible error codes
-	\begin{itemize}
-		\item MC will identify on-the-fly when (e.g.) {\tt \_XABORT\_EXPLICIT} is possible
-		\item Treat such transactions as having multiple failure injection possibilities
-		\item Future work: dataflow analysis to identify when failure reason never used
-	\end{itemize}
-\end{frame}
-
-\begin{frame}{HTM: Possible \xabort error codes}
-	From GCC 4.8.2 manual \S 6.56.8, ``X86 transaction memory intrinsics'':
-	\begin{itemize}
-		\item {\tt \_XABORT\_EXPLICIT} - Transaction explicitely [sic] aborted with \xabort.
-			%The parameter passed to \xabort is available with {\tt \_XABORT\_CODE(status)}
-		\item {\tt \_XABORT\_RETRY} - Transaction retry is possible.
-		\item {\tt \_XABORT\_CONFLICT} - Transaction abort due to a memory conflict with another thread
-		\item {\tt \_XABORT\_CAPACITY} - Transaction abort due to the transaction using too much memory
-		\item {\tt \_XABORT\_DEBUG} - Transaction abort due to a debug trap
-		\item {\tt \_XABORT\_NESTED} - Transaction abort in a inner nested transaction 
-	\end{itemize}
-\end{frame}
-
-%\begin{frame}{Interleaving t'xns non-atomically $\rightarrow$ state space explosion}
-%	% imagine a pair of transactions each with multiple conflicting accesses
-%	% they should be treated as atomic
-%\end{frame}
-
-%\begin{frame}{STM: Judging when aborts are possible}
-%\end{frame}
-
-\begin{frame}{HTM: Sources of Benchmarks}
-	\begin{itemize}
-		%\item Shamay, https://github.com/cmpxchg16/tsx.me, 2014
-		\item Dehesa-Azuara and Stanley, http://www.contrib.andrew.cmu.edu/\textasciitilde{}mdehesaa/, 2016
-		%\item Chapman et al., Hybrid STM/HTM for nested transactions on OpenJDK, OOPSLA 2016
-		%\item Wang et al., Eunomia: Scaling Concurrent Search Trees under Contention Using HTM, PPoPP 2017
-			% TODO: add, that one lock thingy
-	\end{itemize}
-\end{frame}
-
-%%%%
-
 \section{Warp Zone}
 
-\breakslide{\Large Super Extra Bonus Slides Copy-Pasted from OOPSLA}
+\subsection{Quicksand}
+\breakslide{\Large Bonus Slides - Quicksand}
 
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Education}
+\breakslide{\Large Bonus Slides - Education}
+
+\begin{frame}{15-410...}
+	% TODO:
+	% study of tej/etc specific bugs
+	% list tests cases provided to studence & PSU-specific ones too
+\end{frame}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{HTM}
+\breakslide{\Large Bonus Slides - HTM}
+
+\begin{frame}{HTM: Even more dimensions of nondeterminism}
+	% The gotcha: When xbegin can fail, it can emit diff error codes, which depend on what happened within.
+	\begin{center}
+		\begin{tabular}{l}
+			\texttt{if ((status = \_xbegin()) == \_XBEGIN\_STARTED) \{} \\
+			\texttt{~~~~if (count >= 100)}\\
+			\texttt{~~~~~~~~\_xabort();} \\
+			\texttt{~~~~count++;} \\
+			\texttt{~~~~\_xend();} \\
+			\texttt{\} else if (status != \_XABORT\_EXPLICIT) \{} \\
+			\texttt{~~~~mutex\_lock(\&count\_lock);} \\
+			\texttt{~~~~count++;} \\
+			\texttt{~~~~mutex\_unlock(\&count\_lock);} \\
+			\texttt{\}} \\
+		\end{tabular}
+	\end{center}
+	\linegap
+
+	\xbegin can fail with multiple possible error codes
+	\begin{itemize}
+		\item MC will identify on-the-fly when (e.g.) {\tt \_XABORT\_EXPLICIT} is possible
+		\item Treat such transactions as having multiple failure injection possibilities
+		\item Future work: dataflow analysis to know when failure reason never used
+	\end{itemize}
+\end{frame}
+
+\begin{frame}{HTM: Possible \xabort error codes}
+	From GCC 4.8.2 manual \S 6.56.8, ``X86 transaction memory intrinsics'':
+	\begin{itemize}
+		\item {\tt \_XABORT\_EXPLICIT} - Transaction explicitely [sic] aborted with \xabort.
+			%The parameter passed to \xabort is available with {\tt \_XABORT\_CODE(status)}
+		\item {\tt \_XABORT\_RETRY} - Transaction retry is possible.
+		\item {\tt \_XABORT\_CONFLICT} - Transaction abort due to a memory conflict with another thread
+		\item {\tt \_XABORT\_CAPACITY} - Transaction abort due to the transaction using too much memory
+		\item {\tt \_XABORT\_DEBUG} - Transaction abort due to a debug trap
+		\item {\tt \_XABORT\_NESTED} - Transaction abort in a inner nested transaction
+	\end{itemize}
+\end{frame}
+
+%\begin{frame}{Interleaving t'xns non-atomically $\rightarrow$ state space explosion}
+%	% imagine a pair of transactions each with multiple conflicting accesses
+%	% they should be treated as atomic
+%\end{frame}
+
+%\begin{frame}{STM: Judging when aborts are possible}
+%\end{frame}
+
+\begin{frame}{HTM: Sources of Benchmarks}
+	\begin{itemize}
+		%\item Shamay, https://github.com/cmpxchg16/tsx.me, 2014
+		\item Dehesa-Azuara and Stanley, http://www.contrib.andrew.cmu.edu/\textasciitilde{}mdehesaa/, 2016
+		%\item Chapman et al., Hybrid STM/HTM for nested transactions on OpenJDK, OOPSLA 2016
+		%\item Wang et al., Eunomia: Scaling Concurrent Search Trees under Contention Using HTM, PPoPP 2017
+			% TODO: add, that one lock thingy
+		\item vamsikc, TSX LevelDB implementation, https://github.com/vamsikc/leveldb-tsx/, 2014
+	\end{itemize}
+\end{frame}
+
+\definecolor{lavender}{RGB}{102,65,208} % (?? i don't know what happened here anymore but it looks good)
+\definecolor{seafoam}{RGB}{56,158,68} % * 2/3
+\definecolor{salmon}{RGB}{208,89,74} % * 7/8
+\definecolor{goldish}{RGB}{128,96,32}
+\newcommand\tii{\ensuremath{\hilight{lavender}{\mathbf{Ti}}}\xspace}
+\newcommand\tjj{\ensuremath{\hilight{seafoam}{\mathbf{Tj}}}\xspace}
+\newcommand\tkk{\ensuremath{\hilight{salmon}{\mathbf{Tk}}}\xspace}
+
+\newcommand\tiiat[1]{\ensuremath{\hilight{lavender}{\mathbf{Ti}@#1}}\xspace}
+\newcommand\tjjat[1]{\ensuremath{\hilight{seafoam} {\mathbf{Tj}@#1}}\xspace}
+\newcommand\tkkat[1]{\ensuremath{\hilight{salmon}  {\mathbf{Tk}@#1}}\xspace}
+
+\begin{frame}{HTM failure injection equivalence}
+		For any interleaving %prefix
+	\[
+	\begin{tabular}{c}
+		$\tiiat{\alpha},\tiiat{\beta_1}\dots\tiiat{\beta_b},$\\
+		$\tjjat{\gamma_1}\dots\tjjat{\gamma_j},$ \\
+		$\tkkat{\kappa_1}\dots\tkkat{\kappa_k},$ \\
+		$\tiiat{\beta_{b+1}}$ % $\dots\tiiat{\beta_{n-1}}$ -- excluded bc might abort
+	\end{tabular}
+	\]
+	with $b<n$, $j \ne i$, $k \ne i$, etc., either:
+	\begin{enumerate}
+		\item $\tiiat{\alpha},\tjjat{\gamma_1}\dots\tjjat{\gamma_j},\tkkat{\kappa_1}\dots\tkkat{\kappa_k},\tiiat{\phi_1}\dots$
+			\begin{itemize}
+				\item (conflicting case), or
+			\end{itemize}
+		\item $\tiiat{\alpha},\tiiat{\beta_1}\dots\tiiat{\beta_b}\dots\tiiat{\beta_n},\tjjat{\gamma_1}\dots\tjjat{\gamma_j},\tkkat{\kappa_1}\dots\tkkat{\kappa_k}$
+			\begin{itemize}
+				\item (independent case)
+			\end{itemize}
+	\end{enumerate}
+	exists and is observationally equivalent.
+\end{frame}
+\begin{frame}{HTM data-race soundness}
+	Let:
+	\begin{itemize}
+		\item
+			$\tiiat{\mu},\tjjat{\nu}$ be a memory conflict in an interleaving $\mathcal{I}$
+		\begin{itemize}
+			\item (normalized under failure injection equivalence)
+		\end{itemize}
+		\item
+		$\mathsf{lockify}_m(\mathcal{I})$ be a function over interleavings
+			\begin{itemize}
+				\item replaces {\tt xbegin} with {\tt mutex\_lock}($m$) (if successful)
+				\item replaces {\tt xend} with {\tt mutex\_unlock}($m$)
+			\end{itemize}
+		%$\tkkat{L}$ with $\tkkat{\mathsf{lock}(m)}$ if $L$ is a successful HTM begin,
+		%with a no-op if $L$ is a transaction abort,
+		%or with $\tkkat{\mathsf{unlock}(m)}$ if $L$ is an HTM end,
+		%or no replacement otherwise.
+		\item
+			$\mathcal{I}' = \exists m. \mathsf{lockify}_m(\mathcal{I})$,
+			\begin{itemize}
+				\item $m$ represents a unique abstract global lock
+				\item (failure injection equivalence guarantees mutual exclusion of $m$)
+			\end{itemize}
+	\end{itemize}
+	\linegap
+
+	Then $\tiiat{\mu},\tjjat{\nu}$ is a data race in $\mathcal{I}$ iff it is a data race in $\mathcal{I}'$.
+\end{frame}
+
+\begin{frame}{Retry Reduction}
+	% TODO: split this into a step by step
+	\begin{center}
+		\includegraphics[width=\textwidth]{../retry-sets.pdf}
+	\end{center}
+\end{frame}
+
 % TODO: slide number c.c
-\begin{frame}{Don't you need \texttt{if (mutex\_held(\&foo\_lock)) \_xabort();} in that transaction on slide 31?}
+\begin{frame}{Don't you need \texttt{if (mutex\_held(\&count\_lock)) \_xabort();} in that transaction on slide 42?}
+	\begin{center}
 	Yes.
-\end{frame}
+	\end{center}
+\end{frame}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Bibliography}
 
 \begin{frame}{References}
-	\small
-	\begin{itemize}
+	\footnotesize
+	\begin{itemize}
+		\item {\bf [A., D. '52]}: T. Anderson and D. Darling.
+			Asymptotic theory of certain ``goodness-of-fit'' criteria based on stochastic processes.
+			Annals of Statistics, 1952.
 		\item {\bf [Lamport '78]}:
@@
 \begin{frame}{References}
-	\small
+	\footnotesize
 	\begin{itemize}
@@
 		\item {\bf [Kasicki '12]}: Baris Kasicki. Data races vs. data race bugs: telling the difference with Portend. ASPLOS '12.
+		\item {\bf [Coons '13]}: Katherine Coons et al. Bounded Partial Order Reduction. OOPSLA 2013.
 		\item {\bf [Huang '15]}: Jeff Huang. Stateless model checking concurrent programs with maximal causality reduction. PLDI '15.

### 2018-10-12 thesis/defense/snakefight.tex
htms

@@
 \newcommand\hilight[2]{\color{#1}#2\color{black}}
-\definecolor{olivegreen}{RGB}{0,127,0}
+\definecolor{darkorange}{RGB}{192,128,0}
+\definecolor{darkblue}{RGB}{0,97,192}
+\definecolor{olivegreen}{RGB}{0,128,0}
 
@@
 		\\
-		\texttt{\hilight{orange}{mutex\_lock}(m);}      & \texttt{atomic\_xadd(\&x, 1);} \\
+		\texttt{\hilight{darkorange}{mutex\_lock}(m);}      & \texttt{atomic\_xadd(\&x, 1);} \\
 		\texttt{x++;}				   & \texttt{\hilight{olivegreen}{yield}();} \\
-		\texttt{\hilight{blue}{mutex\_unlock}(m);}      & \texttt{atomic\_xadd(\&x, 1);} \\
+		\texttt{\hilight{darkblue}{mutex\_unlock}(m);}      & \texttt{atomic\_xadd(\&x, 1);} \\
 		\texttt{assert(x >= 1);}			& \texttt{assert(x >= 2);} \\
@@
 		\\
-		\texttt{\hilight{orange}{mutex\_lock}(m);} \\
+		\texttt{\hilight{darkorange}{mutex\_lock}(m);} \\
 		\texttt{int tmp = x;} \\
@@
 								& \texttt{\hilight{red}{assert(x >= 2);}} \\
-		\texttt{\hilight{blue}{mutex\_unlock}(m);} \\
+		\texttt{\hilight{darkblue}{mutex\_unlock}(m);} \\
 		\texttt{assert(x >= 1);}
@@
 		\begin{tabular}{l}
-			\texttt{if ((status = \_xbegin()) == SUCCESS) \{} \\ %\_XBEGIN\_STARTED) \{} \\
+			\texttt{if ((status = \hilight{darkorange}{\_xbegin}()) == SUCCESS) \{} \\ %\_XBEGIN\_STARTED) \{} \\
 			%\texttt{} \\
@@
 			\texttt{~~~~count++;} \\
-			\texttt{~~~~\_xend();} \\
+			\texttt{~~~~\hilight{darkblue}{\_xend}();} \\
 			\texttt{\} else \{}\\ %if (status != \_XABORT\_EXPLICIT) \{} \\
-			\texttt{~~~~mutex\_lock(\&count\_lock);} \\
+			\texttt{~~~~\hilight{darkorange}{mutex\_lock}(\&count\_lock);} \\
 			\texttt{~~~~count++;} \\
-			\texttt{~~~~mutex\_unlock(\&count\_lock);} \\
+			\texttt{~~~~\hilight{darkblue}{mutex\_unlock}(\&count\_lock);} \\
 			\texttt{\}} \\
@@
 	\end{center}
+	% TODO: cite orig software TM, intel HTM
 \end{frame}
@@
 			\hline
-			{\tt \_xbegin() == SUCCESS} \\
+			{\tt \hilight{darkorange}{\_xbegin}() == SUCCESS} \\
 			{\tt count++;} \\
-				& {\tt \_xbegin() == SUCCESS} \\
+				& {\tt \hilight{darkorange}{\_xbegin}() == SUCCESS} \\
 				& {\tt count++;} \\
-				& {\tt xend();} \\
-			{\tt xend()}...? \\
+				& {\tt \hilight{darkblue}{\_xend}();} \\
+			{\tt \hilight{darkblue}{\_xend}()}...? \\
 			{\em <conflict detected>} \\
 			\\
-			{\tt \_xbegin() == FAIL} \\
-			{\tt mutex\_lock(...);} \\
+			{\tt \hilight{darkorange}{\_xbegin}() == FAIL} \\
+			{\tt \hilight{darkorange}{mutex\_lock}(...);} \\
 			{\tt count++;} \\
-			{\tt mutex\_unlock(...);} \\
+			{\tt \hilight{darkblue}{mutex\_unlock}(...);} \\
 		\end{tabular}
@@
 	\begin{itemize}
-		\item Safe to automatically skip all interleavings within transactions
+		\item Safe to automatically skip all interleavings within transactions %$\qed$
 			\begin{itemize}
@@
 
+\begin{frame}{Example bug}
+	\begin{center}
+		\begin{tabular}{l|l}
+			{\bf Thread 1} & {\bf Thread 2} \\
+			\hline
+			{\tt \hilight{darkorange}{\_xbegin}() == FAIL} \\
+			{\tt \hilight{darkorange}{mutex\_lock}(...);} \\
+			{\tt count++;} \\
+			{\tt \hilight{darkblue}{mutex\_unlock}(...);} \\
+				& {\tt \hilight{darkorange}{\_xbegin}() == SUCCESS} \\
+				& {\tt count++;} \\
+				& {\tt \hilight{darkblue}{\_xend}();} \\
+			\\
+			\\
+		\end{tabular}
+	\end{center}
+\end{frame}
+
+\begin{frame}{Example bug}
+	\begin{center}
+		\begin{tabular}{l|l}
+			{\bf Thread 1} & {\bf Thread 2} \\
+			\hline
+			{\tt \hilight{darkorange}{\_xbegin}() == FAIL} \\
+			{\tt \hilight{darkorange}{mutex\_lock}(...);} \\
+			{\tt int tmp = count;} \\
+			{\em \hilight{red}{<data-race PP>}} \\
+				& {\tt \hilight{darkorange}{\_xbegin}() == SUCCESS} \\
+				& {\tt count++;} \\
+				& {\tt \hilight{darkblue}{\_xend}();} \\
+			{\tt count = tmp + 1;} \\
+			{\tt \hilight{darkblue}{mutex\_unlock}(...);} \\
+		\end{tabular}
+	\end{center}
+\end{frame}
+
+\begin{frame}{Example bug - Solution {\em [Dehesa-Azuara '16]}}
+	\begin{center}
+		\begin{tabular}{l}
+			\texttt{if ((status = \hilight{darkorange}{\_xbegin}()) == SUCCESS) \{} \\ %\_XBEGIN\_STARTED) \{} \\
+			%\texttt{} \\
+			%\texttt{} \\
+			\texttt{~~~~\hilight{red}{if (mutex\_held(\&count\_lock))}} \\
+			\texttt{~~~~~~~~\hilight{red}{\_xabort();}} \\
+			\texttt{~~~~count++;} \\
+			\texttt{~~~~\hilight{darkblue}{\_xend}();} \\
+			\texttt{\} else \{}\\ %if (status != \_XABORT\_EXPLICIT) \{} \\
+			\texttt{~~~~\hilight{darkorange}{mutex\_lock}(\&count\_lock);} \\
+			\texttt{~~~~count++;} \\
+			\texttt{~~~~\hilight{darkblue}{mutex\_unlock}(\&count\_lock);} \\
+			\texttt{\}} \\
+		\end{tabular}
+	\end{center}
+\end{frame}
+
+\begin{frame}{Example bug - Solution {\em [Dehesa-Azuara '16]}}
+	\begin{center}
+		\begin{tabular}{l|l}
+			{\bf Thread 1} & {\bf Thread 2} \\
+			\hline
+			{\tt \hilight{darkorange}{\_xbegin}() == FAIL} \\
+			{\tt \hilight{darkorange}{mutex\_lock}(...);} \\
+			{\tt int tmp = count;} \\
+			{\em \hilight{red}{<data-race PP>}} \\
+				& {\tt \hilight{darkorange}{\_xbegin}() == SUCCESS} \\
+				& {\tt \hilight{red}{if (mutex\_held(...))}} \\
+				& {\tt \hilight{red}{~~~~\_xabort();}} \\
+			{\tt count = tmp + 1;} \\
+			{\tt \hilight{darkblue}{mutex\_unlock}(...);} \\
+				& {\tt \hilight{darkorange}{mutex\_lock}(...);} \\
+				& {\tt count++;} \\
+				& {\tt \hilight{darkblue}{mutex\_unlock}(...);} \\
+		\end{tabular}
+	\end{center}
+\end{frame}
+
 \begin{frame}{Evaluation}
+	Questions
+	\begin{itemize}
+		\item Does Landslide find previously-unknown bugs in HTM programs?
+		\item Does Landslide verify HTM code in reasonable time?
+	\end{itemize}
+	\linegap
+
+	Test suite
+	\begin{itemize}
+		\item Hand-written unit tests
+		\item Transactional data structures {\em [Dehesa-Azuara '16]}
+		\item Spinlock from TSX LevelDB implementation {\em [vamsikc '14]}
+	\end{itemize}
+\end{frame}
+
+\begin{frame}{Evaluation - Bug-finding}
+	Example - {\tt count++} data race
+	\begin{itemize}
+		\item Landslide finds assertion failure in 7 seconds % 2,1 with -M, 5 interleavings, wall clock time
+	\end{itemize}
+	\linegap
+
+	AVL tree - atomicity violation
+	\begin{itemize}
+		\item Transactions omitted the {\tt mutex\_held()} check
+		\item Landslide finds segfault in 15 seconds % 3,1 with quicksand; 14 interleavings, wall clock time
+		%\item Same author who proposed the fix... % this bullet exists in summary slide
+	\end{itemize}
+	\linegap
+
+	Spinlock - performance degradation
+	\begin{itemize}
+		\item Implemented {\tt mutex\_held()} using {\tt cmpxchg} (= atomic read + write)
+		\item Unconditional write during {\tt xbegin} path $\rightarrow$ always aborts!
+		\item Landslide finds assertion failure in 3 seconds % 2,2 with -M, 2 interleavings, wall clock time
+	\end{itemize}
+\end{frame}
+
+\begin{frame}{Evaluation - Verification}
+	$K,N$ = $K$ threads, each $N$ iterations of test logic
+	\linegap
+
+	Microbenchmarks
+	\begin{itemize}
+			% numbers taken from retry sets column
+		\item {\tt mutex\_held()} pattern: $K,N$ = 2,4 (79min); 4,1 (6min)
+		\item Using {\tt xadd} instead: $K,N$ = 2,4 (19min); 3,2 (39min), 4,1 (1min)
+		\item More, see thesis...
+	\end{itemize}
+	\pause
+	\linegap
+
+	HTM data structures
+	\begin{itemize}
+			% numbers taken from retry sets column
+		\item Hashmap insert: $K,N$ = 2,1 (27sec) % map_basicer. not dealing with resizing here
+		\item AVL tree insert: $K,N$ = 2,1 (5min)
+			\pause
+		\begin{itemize}
+			\item but... requires $K \times N \ge 3$ to expose the segfault...
+			\begin{itemize}
+				\item Estimated time for $K,N$ = 3,1... 736 years...
+					% 2,2 -> 51000 years lmao
+			\end{itemize}
+		\end{itemize}
+	\end{itemize}
+
+\end{frame}
+
+\begin{frame}{Evaluation - Verification}
+	Abstraction reduction {\em [Simsa '13]}
+	\begin{itemize}
+		\item Verify locking primitives \& client code separately against common API
+	\end{itemize}
+	\linegap
+
+	HTM locks
+	\begin{itemize}
+			% numbers are now using STM semantics instead
+			% bc of retry loop around the xbegin
+			% can't really mention this in the talk
+		\item Spinlock, $K,N$ = 2,2 (2min), 3,1 (28sec)
+			\begin{itemize}
+				\item Depends on ad-hoc yield-loop heuristics; slow
+			\end{itemize}
+		\item Spin replaced w/P2 mutex: $K,N$ = 2,4 (7hrs), 4,1 (11min)
+	\end{itemize}
+	\pause
+	\linegap
+
+	HTM data structures w/verified locks
+	\begin{itemize}
+		\item Hashmap insert: $K,N$ = 2,3 (4hrs), 3,1 (11min)
+		\item AVL tree insert: $K,N$ = 2,4 (4min), 3,2 (23min), 4,1 (3min)
+	\end{itemize}
+\end{frame}
+
+\begin{frame}{Summary}
+	HTM adds new ``dimension'' of concurrency
+	\begin{itemize}
+		\item Equivalence theorem lets Landslide emulate execution semantics
+		\item Allows exponential reduction of non-observable behaviours
+	\end{itemize}
+	\linegap
+
+	Even simple programs tricky to get right
+	\begin{itemize}
+		\item Author of {\tt mutex\_held()} pattern got it wrong once...
+		\item Landslide can check even unpredictable performance characteristics
+	\end{itemize}
+	\linegap
+
+	Abstraction reduction eases both implementation \& verification burden
 \end{frame}
@@
 \section{Conclusion}
+
+\begin{frame}{Conclusion}
+\end{frame}
 
@@
 			\texttt{mutex\_lock(m);}	& \\
-			\texttt{\hilight{blue}{y = true;}}      & \\
+			\texttt{\hilight{darkblue}{y = true;}}      & \\
 			\texttt{mutex\_unlock(m);}      & \\
 				& \texttt{mutex\_lock(m);} \\
-				& \texttt{\hilight{blue}{bool tmp = y;}} \\
+				& \texttt{\hilight{darkblue}{bool tmp = y;}} \\
 				& \texttt{mutex\_unlock(m);} \\
-				& \texttt{\hilight{blue}{if (tmp)}~\hilight{red}{x++;}} \\
+				& \texttt{\hilight{darkblue}{if (tmp)}~\hilight{red}{x++;}} \\
 		\end{tabular}
@@
 
-% TODO: slide number c.c
-\begin{frame}{Don't you need \texttt{if (mutex\_held(\&count\_lock)) \_xabort();} in that transaction on slide 42?}
-	\begin{center}
-	Yes.
-	\end{center}
-\end{frame}
+%% including this stuff in the talk now so this bonus slide is frivolous
+%% TODO: slide number c.c
+%\begin{frame}{Don't you need \texttt{if (mutex\_held(\&count\_lock)) \_xabort();} in that transaction on slide 42?}
+%	\begin{center}
+%	Yes.
+%	\end{center}
+%\end{frame}
 
@@
 		\item {\bf [Kasicki '12]}: Baris Kasicki. Data races vs. data race bugs: telling the difference with Portend. ASPLOS '12.
+		\item {\bf [Simsa '13]}: Systematic and Scalable Testing of Concurrent Programs.
+			CMU-CS-13-133. December 2013.
 		\item {\bf [Coons '13]}: Katherine Coons et al. Bounded Partial Order Reduction. OOPSLA 2013.
+		\item {\bf [vamsikc '14]}: leveldb-tsx. https://github.com/vamsikc/leveldb-tsx/
 		\item {\bf [Huang '15]}: Jeff Huang. Stateless model checking concurrent programs with maximal causality reduction. PLDI '15.
 		\item {\bf [Jensen '15]}: Stateless Model Checking of Event-Driven Applications. OOPSLA 2015.
+		\item {\bf [Dehesa-Azuara '16]}: Hardware transactional memory with intel's TSX.
+			http://www.contrib.andrew.cmu.edu/\textasciitilde{}mdehesaa/, 2016.
 	\end{itemize}

### 2018-10-12 thesis/defense/snakefight.tex
write intro and concl

@@
 \definecolor{olivegreen}{RGB}{0,128,0}
+\definecolor{lavender}{RGB}{102,65,208} % (?? i don't know what happened here anymore but it looks good)
+\definecolor{seafoam}{RGB}{56,158,68} % * 2/3
+\definecolor{salmon}{RGB}{208,89,74} % * 7/8
+\definecolor{goldish}{RGB}{128,96,32}
+\definecolor{gray}{RGB}{128,128,128}
 
@@
 		\item Heuristic search ordering: ICB {\em [Musuvathi '08]}, PCT {\em [Burckhardt '10]}
-		\item There will always be a {\bf bigger test case}!
-	\end{itemize}
-	\pause
-	\linegap
-
-	Too much manual intervention for non-expert use
+		\item There will always be a {\bf bigger test case}...
+	\end{itemize}
+	\pause
+	\linegap
+
+	Too much manual intervention %for non-expert use
 	\begin{itemize}
@@
 			% TODO change this bullet pt to be more explicitly about 
-		\item Widespread adoption requires {\bf fully automated instrumentation}!
-	\end{itemize}
-	\pause
-	\linegap
-
-	Limited sources of nondeterminism
+		\item Widespread adoption requires {\bf fully automated instrumentation}...
+	\end{itemize}
+	\pause
+	\linegap
+
+	Sources of nondeterminism
 	\begin{itemize}
@@
 		\item Event-driven applications {\em [Jensen '15]}
-		\item {\bf Hardware transactional memory}?
+		\item {\bf Hardware transactional memory}...
 	\end{itemize}
@@
 
-\definecolor{gray}{RGB}{128,128,128}
-
 \begin{frame}{Thesis Statement}
-	The Iterative Deepening algorithm makes MC appropriate for both educational and real-world programs.
-	\vspace{4.14em}
-
-	Ad-hoc synchronization heuristics make MC accessible to student programmers.
-	\vspace{4.14em}
-
-	A new transactional concurrency model keeps MC applicable to the latest real-world techniques.
-	\vspace{4.14em}
+	% data race pps (as well as txn equivalence)
+	\hilight{gray}{Combining both theoretically-founded automatic reduction techniques
+	% iterative deepening
+	% (say smth like: "landslide also has a wealth of other user-tunable heuristics
+	% for expert users, which you can read about in the document; for this talk i will focus just on the CPU budget thing")
+	% XXX: need to prepare for the question -- "what evidence do you have that users actually want this cpu budget stuff",
+	% ",or, that it helps vs just running something unlimited vs a normal icb test"
+	% (well, for the 2nd half of the Q, i can talk about how ID outperforms ICB on verficiations)
+	and user-informed heuristic ones, stateless model checking...}
+	\begin{itemize}
+		\item ...can sufficiently mitigate exponential explosion...
+		% education
+		\item ...to be a practical testing technique for inexperienced users...
+		% htm
+		% TODO: this last part you just added for this talk; does the thesis version of the statmeent need to include too?
+		\item ...and real-world programs alike.
+	\end{itemize}
+\end{frame}
+
+\begin{frame}{Thesis Statement}
+	The Iterative Deepening algorithm %combines theory and heuristics to
+	practically debugs and verifies many exponential state spaces.
+	\vspace{4.54em}
+
+	Landslide's automatic instrumentation makes MC accessible to students.
+	\vspace{4.54em}
+
+	New reduction techniques for transactional memory keep MC up-to-date
+	with the latest real-world concurrency techniques.
+	\vspace{4.54em}
 \end{frame}
 \begin{frame}{Outline}
-	\hilight{gray}{The Iterative Deepening algorithm makes MC appropriate for both educational and real-world programs.} \\
+	\hilight{gray}{The Iterative Deepening algorithm %combines theory and heuristics to
+	practically debugs and verifies many exponential state spaces.} \\
+	%\hilight{gray}{The Iterative Deepening algorithm makes MC appropriate for both educational and real-world programs.} \\
 	\begin{itemize}
@@
 		\begin{itemize}
-			\item Algorithm overview \& data-race preemption points
+			\item Data-race preemption points \& algorithm overview
 			\item Verification \& experimental results
@@
 	\end{itemize}
-	\hilight{gray}{Ad-hoc synchronization heuristics make MC accessible to student programmers.} \\
+	\hilight{gray}{Landslide's automatic instrumentation makes MC accessible to students.} \\
+	%\hilight{gray}{Ad-hoc synchronization heuristics make MC accessible to student programmers.} \\
 	\begin{itemize}
@@
 		\begin{itemize}
-			\item Pebbles \& 15-410 user study
-			\item Pintos \& upcoming plans
-		\end{itemize}
-	\end{itemize}
-
-	\hilight{gray}{A new transactional concurrency model keeps MC applicable to the latest real-world techniques.} \\
+			\item Pebbles \& Pintos OS projects
+			\item Bug-finding, grades, \& surveys
+		\end{itemize}
+	\end{itemize}
+
+	\hilight{gray}{New reduction techniques for transactional memory keep MC up-to-date
+	with the latest real-world concurrency techniques.} \\
+	%\hilight{gray}{A new transactional concurrency model keeps MC applicable to the latest real-world techniques.} \\
 	\begin{itemize}
@@
 		\begin{itemize}
-			\item Example \& challenges
-			\item Equivalence \& failure injection
+			\item Concurrency model \& failure injection
+			\item Bug-finding \& verification
 		\end{itemize}
@@
 
-\begin{frame}{Results - Bugs found}
+\begin{frame}{Results - Bugs found (CPU time)}
 	\begin{center}
 		\includegraphics[width=0.8\textwidth]{../../oopsla/bugs-talk.pdf}
+	\end{center}
+\end{frame}
+\begin{frame}{Results - Bugs found (wall-clock time)}
+	\begin{center}
+		\includegraphics[width=0.8\textwidth]{../../oopsla/bugs-wallclock-talk.pdf}
 	\end{center}
@@
 	Otherwise? Quicksand's heuristic search ordering finds bugs faster anyway
-	\pause
-	\linegap
+	%\pause
+	%\linegap
 
@@
 		\item CMU: 103/177 students (7 semesters)
+		\item U. Chicago: 4/21 students (1 semester)
 		\item PSU: 38/136 students (1 semester)
-		\item U. Chicago: 4/21 students (1 semester)
 	\end{itemize}
@@
 		\item U. Chicago: 3 deterministic, 4 concurrency bugs in 3/4 kernels
+			% TODO: what to do abt this bullet pts
+			%\pause
+		\item PSU: (no log collection due to IRB issues)
 		% \item Case study tej
@@
 
-\begin{frame}{Effect on Grades}
+\begin{frame}{Effect on Grades (CMU)}
 	Compared Landslide users' vs. non-users' submitted P2 grades...
@@
 
-\begin{frame}{Effect on Grades}
+\begin{frame}{Effect on Grades (CMU)}
 	\begin{center}
@@
 	\begin{itemize}
-		\item {\em Education:} Did Landslide help students learn more from P2?
-		\item {\em Happiness:} Did Landslide help students save time/effort?
-	\end{itemize}
-	\linegap
-
-	Self-report any false positives
+		\item {\bf Education:} Did Landslide help students learn more from P2?
+		\item {\bf Happiness:} Did Landslide help students save time/effort?
+	\end{itemize}
+	\pause
+	\linegap
+
+	Self-report any {\bf false positives}
 	\begin{itemize}
@@
 	\end{itemize}
+	\linegap
+
+	Self-report opinions {\bf whether it was worthwhile}
+	\begin{itemize}
+		\item Mostly ``worked as advertised'' praise
+		\item One PSU student found a bug from inspecting a data-race by hand,
+			after Landslide timed out testing with no bug report
+	\end{itemize}
 \end{frame}
@@
 	Landslide helps students find \& fix bugs during projects
+	% TODO: should this be moved ot the earlier bug finding slide
 	\begin{itemize}
@@
 
-	Survey results showed students enlightenedly happy
+	Survey results showed students ``enlightenedly happy'' % dave eckhardt scare quotes if i ever saw some
 	\linegap
@@
 			\\
-			{\tt \hilight{darkorange}{\_xbegin}() == FAIL} \\
+			{\tt \hilight{darkorange}{\_xbegin}() == ABORTED} \\
 			{\tt \hilight{darkorange}{mutex\_lock}(...);} \\
@@
 			\end{itemize}
+			% TODO prob just comment this out
 		\item More in bonus slides...
@@
 			\hline
-			{\tt \hilight{darkorange}{\_xbegin}() == FAIL} \\
+			{\tt \hilight{darkorange}{\_xbegin}() == ABORTED} \\
 			{\tt \hilight{darkorange}{mutex\_lock}(...);} \\
@@
 \begin{frame}{Evaluation - Verification}
-	$K,N$ = $K$ threads, each $N$ iterations of test logic
+	$K,N$ = $K$ threads, each $N$ iterations of test logic; 10 hour timeout
 	\linegap
@@
 		\item Using {\tt xadd} instead: $K,N$ = 2,4 (19min); 3,2 (39min), 4,1 (1min)
-		\item More, see thesis...
+		\item Is that ``enough''? Generalize with manual inspection?
+		%\item More, see thesis...
 	\end{itemize}
@@
 		\item Verify locking primitives \& client code separately against common API
+			% say: again relying on human to write the right test case
+		\item {\em Add} verification costs instead of {\em multiplying}
 	\end{itemize}
@@
 			% can't really mention this in the talk
-		\item Spinlock, $K,N$ = 2,2 (2min), 3,1 (28sec)
+		\item Spinlock: $K,N$ = 2,2 (2min), 3,1 (28sec)
 			\begin{itemize}
@@
 \begin{frame}{Conclusion}
+	% data race pps (as well as txn equivalence)
+	Combining %both
+	{\bf theoretically-founded automatic reduction techniques}
+	\vspace{4.82em}
+
+	and {\bf user-informed heuristic ones},
+	\vspace{4.82em}
+
+	stateless model checking...
+	\begin{itemize}
+		\item ...can sufficiently mitigate exponential explosion...
+		\item ...to be a practical testing technique for inexperienced users...
+		\item ...and real-world programs alike.
+	\end{itemize}
+\end{frame}
+\begin{frame}{Conclusion}
+	Combining %both
+	{\bf theoretically-founded automatic reduction techniques}
+	\begin{itemize}
+		\item Verification with data-race preemption points
+		\item Yield-loop blocking
+		\item Equivalent emulation of HTM semantics
+	\end{itemize}
+	and {\bf user-informed heuristic ones},
+	\begin{itemize}
+		\item Quicksand's CPU budget argument
+		\item Exposing data-race reports to students
+		\item Abstraction reduction for HTM primitives
+	\end{itemize}
+	\hilight{gray}{stateless model checking...}
+	\begin{itemize}
+		\item \hilight{gray}{...can sufficiently mitigate exponential explosion...}
+		\item \hilight{gray}{...to be a practical testing technique for inexperienced users...}
+		\item \hilight{gray}{...and real-world programs alike.}
+	\end{itemize}
+\end{frame}
+\begin{frame}{Closing remarks}
+	``Verifying C programs is doomed, why can't PL design handle this?''
+	\begin{itemize}
+		\item seL4 {\em [Klein '09]}: secure kernel verified end-to-end in Haskell
+		\item Rust {\em [Matsakis '14]}: concurrency-aware type system, no data races
+	\end{itemize}
+	\pause
+	\linegap
+
+	But... we still have to try.
+	\begin{itemize}
+		\item Computers can't handle $O(N^K)$
+			\begin{itemize}
+				\item ``ok, $N^K$, that's, wow, 700 years...''
+			\end{itemize}
+		\item Humans can help?
+			\begin{itemize}
+				\item ``can we get same results with smaller $N$, $K$?''
+			\end{itemize}
+	\end{itemize}
+	\pause
+	\linegap
+
+	One day live in peace with our silicon children?
 \end{frame}
@@
 			\texttt{mutex\_lock(m);}	& \\
-			\texttt{// unrelated}	   & \\
+			\texttt{\hilight{lavender}{// unrelated}}	   & \\
 			\texttt{mutex\_unlock(m);}      & \\
 				& \texttt{mutex\_lock(m);} \\
-				& \texttt{// unrelated} \\
+				& \texttt{\hilight{lavender}{// unrelated}} \\
 				& \texttt{mutex\_unlock(m);} \\
@@
 			\texttt{mutex\_lock(m);}	& \\
-			\texttt{\hilight{darkblue}{y = true;}}      & \\
+			\texttt{\hilight{lavender}{y = true;}}      & \\
 			\texttt{mutex\_unlock(m);}      & \\
 				& \texttt{mutex\_lock(m);} \\
-				& \texttt{\hilight{darkblue}{bool tmp = y;}} \\
+				& \texttt{\hilight{lavender}{bool tmp = y;}} \\
 				& \texttt{mutex\_unlock(m);} \\
-				& \texttt{\hilight{darkblue}{if (tmp)}~\hilight{red}{x++;}} \\
+				& \texttt{\hilight{lavender}{if (tmp)}~\hilight{red}{x++;}} \\
 		\end{tabular}
@@
 
-\begin{frame}{Quicksand results - Bugs found by CPU time}
-	\begin{center}
-		\includegraphics[width=0.8\textwidth]{../../oopsla/bugs-talk.pdf}
-	\end{center}
-\end{frame}
-\begin{frame}{Quicksand results - Bugs found by wall-clock time}
-	\begin{center}
-		\includegraphics[width=0.8\textwidth]{../../oopsla/bugs-wallclock-talk.pdf}
-	\end{center}
-\end{frame}
-\begin{frame}{Quicksand results - Tests fully verified (CPU time)}
-	\begin{center}
-		\includegraphics[width=0.8\textwidth]{../../oopsla/verifs-talk.pdf}
-	\end{center}
-\end{frame}
+%\begin{frame}{Quicksand results - Bugs found by CPU time}
+%	\begin{center}
+%		\includegraphics[width=0.8\textwidth]{../../oopsla/bugs-talk.pdf}
+%	\end{center}
+%\end{frame}
+%\begin{frame}{Quicksand results - Bugs found by wall-clock time}
+%	\begin{center}
+%		\includegraphics[width=0.8\textwidth]{../../oopsla/bugs-wallclock-talk.pdf}
+%	\end{center}
+%\end{frame}
+%\begin{frame}{Quicksand results - Tests fully verified (CPU time)}
+%	\begin{center}
+%		\includegraphics[width=0.8\textwidth]{../../oopsla/verifs-talk.pdf}
+%	\end{center}
+%\end{frame}
 \begin{frame}{Quicksand results - MC + DR vs single-pass DR}
@@
 
-\definecolor{lavender}{RGB}{102,65,208} % (?? i don't know what happened here anymore but it looks good)
-\definecolor{seafoam}{RGB}{56,158,68} % * 2/3
-\definecolor{salmon}{RGB}{208,89,74} % * 7/8
-\definecolor{goldish}{RGB}{128,96,32}
 \newcommand\tii{\ensuremath{\hilight{lavender}{\mathbf{Ti}}}\xspace}
@@
 
+\begin{frame}{How good is $K,N$?}
+	\begin{center}
+		\small
+		\begin{tabular}{c||c|c|c||c}
+			{\bf test} & {\bf \#txn} & {\bf \#sync} & {\bf \#race} & {\bf max events verified} \\
+			\hline
+			{\tt htm2}	      & 1     & 2     & 4     & 42    \\ % 2,3
+			{\tt counter}	   & 1     & 0     & 1     & 16    \\ % 2,4
+			{\tt swap}	      & 1     & 4     & 8     & 52    \\ % 2,2
+			{\tt fig63}	     & 1     & 0     & $K$-1 & 18    \\ % 3,2 k=3 -> 18; 2,4 -> 16
+			{\tt avl\_insert}       & 1     & 2     & 7     & 20    \\ % 2,1 nb. only 5 drpps under -A -S but w/e
+			% nb less than 9 on stm
+			{\tt avl\_fixed}	& 1     & 2     & 9     & 24    \\ % 2,1 nb. again, only 5 under -A -S
+			% nb stm takes it down from 13 to 10 races
+			{\tt map\_basic}	& 1     & 4     & 13    & 36    \\ % 2,1
+			% mysteriously had this at 6 at one point but ive no idea where that came from
+			{\tt map\_basicer}      & 1     & 2     & 5     & 32    \\ % 2,2
+
+			% TODO: check these again, will 2,4 even complete on mutex
+			{\tt lock}({\tt spinlock})      & 1     & 2     & 4     & 28    \\ % 2,2
+			{\tt lock}({\tt spin\_fixed})   & 1     & 2     & 4     & 28    \\ % 2,2 or 4,1
+			{\tt lock}({\tt mutex})	 & 1     & 2     & 2     & 30    \\ % 2,3 gives 30; 2,4 would give 40
+			% not putting lock_fast because it like, doesn't have a "state space" per se
+		\end{tabular}
+	\end{center}
+	\linegap
+
+	\begin{itemize}
+		\item {\em [Abdulla '14]} reports 19 threads (??)
+			\begin{itemize}
+				\item {\sf filesystem(19)}: classic DPOR: 4096; (their) optimal DPOR: 64
+			\end{itemize}
+			% not really a quote, but their x axis from figure 9 graph, whose y axis goes up to 10k
+		\item {\em [Zhang '15]} verifs up to 10 ``concurrent events'' (weak mem r/w?)
+	\end{itemize}
+\end{frame}
+
 %% including this stuff in the talk now so this bonus slide is frivolous
@@
 			Leslie Lamport. Time, clocks, and the ordering of events in a distributed system.
-			Commun. ACM, 1978.
+			Communications of the ACM, 1978.
 		\item {\bf [Godefroid '97]}: Patrice Godefroid.
@@
 			Programs. OSDI 2008.
-		\item {\bf [Yang '08]}: Yu Yang et al. Efficient stateful dynamic partial order reduction. SPIN 2008.
+		\item {\bf [Yang '08]}: Yu Yang et al. Efficient stateful DPOR. SPIN 2008.
+		\item {\bf [Buckhardt '10]}: A randomized scheduler with probabilistic guarantees of finding bugs. ASPLOS 2010.
 	\end{itemize}
@@
 	\begin{itemize}
-		\item {\bf [Buckhardt '10]}: A randomized scheduler with probabilistic guarantees of finding bugs. ASPLOS 2010.
 		\item {\bf [Simsa '12]}: Jiri Simsa. Runtime Estimation and Resource Allocation for
-			Concurrency Testing. CMU-PDL-12-113. December 2012.
-		\item {\bf [Blum '12]}: Ben Blum. Landslide: Systematic Dynamic Race Detection in Kernel
-			Space. CMU-CS-12-118. May 2012.
+			Concurrency Testing. Tech report (CMU-PDL-12-113). December 2012.
+		\item {\bf [Blum '12]}: Ben Blum. Landslide: Systematic Dynamic Race Detection in Kernel Space.
+			MS thesis (CMU-CS-12-118). May 2012.
 		\item {\bf [Kasicki '12]}: Baris Kasicki. Data races vs. data race bugs: telling the difference with Portend. ASPLOS '12.
 		\item {\bf [Simsa '13]}: Systematic and Scalable Testing of Concurrent Programs.
-			CMU-CS-13-133. December 2013.
-		\item {\bf [Coons '13]}: Katherine Coons et al. Bounded Partial Order Reduction. OOPSLA 2013.
-		\item {\bf [vamsikc '14]}: leveldb-tsx. https://github.com/vamsikc/leveldb-tsx/
-		\item {\bf [Huang '15]}: Jeff Huang. Stateless model checking concurrent programs with maximal causality reduction. PLDI '15.
-		\item {\bf [Jensen '15]}: Stateless Model Checking of Event-Driven Applications. OOPSLA 2015.
+			Ph.D. thesis (CMU-CS-13-133). December 2013.
+		\item {\bf [Coons '13]}: Katherine Coons et al. Bounded POR. OOPSLA 2013.
+		\item {\bf [Abdulla '14]}: Parosh Abdulla et al. Optimal DPOR. POPL 2014.
+		\item {\bf [vamsikc '14]}: leveldb-tsx. https://github.com/vamsikc/leveldb-tsx/, 2014.
+		\item {\bf [Huang '15]}: Jeff Huang. Stateless MCing concurrent programs with maximal causality reduction. PLDI '15.
+		\item {\bf [Jensen '15]}: Stateless MCing of Event-Driven Applications. OOPSLA 2015.
 		\item {\bf [Dehesa-Azuara '16]}: Hardware transactional memory with intel's TSX.

### 2018-10-14 thesis/future.tex
turns out oathbringer had a good quote after all

--- 2018-10-08 thesis/future.tex
@@
 \inspirationalquote{
-\begin{tabular}{p{0.75\textwidth}}
-We can't be free until we learn to laugh at ourselves.
-Once you look in the mirror and see just how foolish we can be,
-laughter is inevitable.
-And from laughter comes wisdom.
+	\begin{tabular}{p{0.62\textwidth}}
+		"Journey before destination."
+		\\
+		Some may call it a simple platitude, but it is far more.
+		A journey will have pain and failure.
+		It is not only the steps forward that we must accept.
+		It is the stumbles. The trials. The knowledge that we will fail.
+		% That we will hurt those around us.
+		\\
+		% But
+		If we stop, if we accept the person we are when we fall, the journey ends.
+		% That failure /becomes/ our destination.
+		To love the journey is to accept no such end.
+		% I have found, through painful experience,
+		% that the most important step a person can take is always the /next/ one.
 	\end{tabular}}
-{G'Kar, Babylon 5}
+%{Galatea, The Talos Principle}
+{Dalinar Kholin, Oathbringer}
 

### 2018-10-14 thesis/defense/snakefight.tex
update snaekfight after first practice

@@
 \definecolor{salmon}{RGB}{208,89,74} % * 7/8
+\definecolor{pinkish}{RGB}{224,0,56} % * 7/8
 \definecolor{goldish}{RGB}{128,96,32}
@@
 \begin{document}
-\renewcommand{\inserttotalframenumber}{39}
+\renewcommand{\inserttotalframenumber}{67}
 \normalem
@@
 	\begin{tabular}{ll}
-		Initially \texttt{x = 0;} \\
+		Initially \texttt{count = 0;} \\
 		\\
@@
 		\\
-		\texttt{\hilight{darkorange}{mutex\_lock}(m);}      & \texttt{atomic\_xadd(\&x, 1);} \\
-		\texttt{x++;}				   & \texttt{\hilight{olivegreen}{yield}();} \\
-		\texttt{\hilight{darkblue}{mutex\_unlock}(m);}      & \texttt{atomic\_xadd(\&x, 1);} \\
-		\texttt{assert(x >= 1);}			& \texttt{assert(x >= 2);} \\
+		\texttt{\hilight{darkorange}{mutex\_lock}(m);}      & \texttt{atomic\_xadd(\&count, 1);} \\
+		\texttt{count++;}				   & \texttt{\hilight{olivegreen}{yield}();} \\
+		\texttt{\hilight{darkblue}{mutex\_unlock}(m);}      & \texttt{atomic\_xadd(\&count, 1);} \\
+		\texttt{assert(count >= 1);}			& \texttt{assert(count >= 2);} \\
 		\\
@@
 		\texttt{\hilight{darkorange}{mutex\_lock}(m);} \\
-		\texttt{int tmp = x;} \\
-								& \texttt{atomic\_xadd(\&x, 1);} \\
+		\texttt{int tmp = count;} \\
+								& \texttt{atomic\_xadd(\&count, 1);} \\
 								& \texttt{\hilight{olivegreen}{yield}();} \\
-								& \texttt{atomic\_xadd(\&x, 1);} \\
-		\texttt{x = tmp + 1;} \\
-								& \texttt{\hilight{red}{assert(x >= 2);}} \\
+								& \texttt{atomic\_xadd(\&count, 1);} \\
+		\texttt{count = tmp + 1;} \\
+								& \texttt{\hilight{pinkish}{assert(count >= 2);}} \\
 		\texttt{\hilight{darkblue}{mutex\_unlock}(m);} \\
-		\texttt{assert(x >= 1);}
+		\texttt{assert(count >= 1);}
 	\end{tabular}
@@
 	\begin{itemize}
-		\item Equivalence reductions: DPOR {\em [Flanagan '05]}, MCR {\em [Huang '15]}
-		\item Heuristic search ordering: ICB {\em [Musuvathi '08]}, PCT {\em [Burckhardt '10]}
+		%\item Equivalence reductions: DPOR {\em [Flanagan '05]}, MCR {\em [Huang '15]}
+		\item Dynamic Partial Order Reduction (DPOR) {\em [Flanagan '05]}
+			\begin{itemize}
+				\item identifies \& skips equivalent interleavings
+			\end{itemize}
+		%\item Heuristic search ordering: ICB {\em [Musuvathi '08]}, PCT {\em [Burckhardt '10]}
+		\item Iterative Context Bounding (ICB) {\em [Musuvathi '08]}
+			\begin{itemize}
+				\item finds bugs faster by prioritizing interleavings with fewer preemptions
+			\end{itemize}
+			\pause
 		\item There will always be a {\bf bigger test case}...
@@
 	\begin{itemize}
+		\item State space size estimation {\em [Simsa '12]}
+			% TODO: i just switched the order of these; see how that feels
 		\item Manual kernel annotations: Landslide {\em [Blum '12]} {\tiny that's me!}
-			% TODO: smth about choosing pps?
-		\item State space size estimation {\em [Simsa '12]}
-			% TODO change this bullet pt to be more explicitly about 
+			\pause
 		\item Widespread adoption requires {\bf fully automated instrumentation}...
@@
 		\item Timer-driven schedule nondeterminism (all prior work)
-		\item Weak memory nondeterminism {\em [Zhang '15]}
-		\item Event-driven applications {\em [Jensen '15]}
+		\item Weak memory {\em [Zhang '15]},
+			event-driven {\em [Jensen '15]}
+			\pause
 		\item {\bf Hardware transactional memory}...
@@
 	% data race pps (as well as txn equivalence)
-	Combining both theoretically-founded automatic reduction techniques
+	Combining %both
+	theoretically-founded automatic reduction techniques
 	% iterative deepening
@@
 	% data race pps (as well as txn equivalence)
-	\hilight{gray}{Combining both theoretically-founded automatic reduction techniques
+	\hilight{gray}{Combining %both
+	theoretically-founded automatic reduction techniques
 	% iterative deepening
@@
 \begin{frame}{Thesis Statement}
-	The Iterative Deepening algorithm %combines theory and heuristics to
+	The Quicksand algorithm %combines theory and heuristics to
 	practically debugs and verifies many exponential state spaces.
@@
 \begin{frame}{Outline}
-	\hilight{gray}{The Iterative Deepening algorithm %combines theory and heuristics to
+	\hilight{gray}{The Quicksand algorithm %combines theory and heuristics to
 	practically debugs and verifies many exponential state spaces.} \\
-	%\hilight{gray}{The Iterative Deepening algorithm makes MC appropriate for both educational and real-world programs.} \\
-	\begin{itemize}
-		\item {\bf Quicksand and Iterative Deepening}
+	%\hilight{gray}{The Quicksandalgorithm makes MC appropriate for both educational and real-world programs.} \\
+	\begin{itemize}
+		\item {\bf Quicksand}
 		\begin{itemize}
@@
 	\end{itemize}
-	\linegap
-
-	Prior model checkers hard-code a fixed set of PPs, committing to one state space in advance.
-	% say: "...then attempt to cope with the resulting exponentially sized problem, however big it may be"
 \end{frame}
@@
 		\includegraphics[width=0.96\textwidth]{../../oopsla/tree-maximal-only.pdf}
+	\end{center}
 	\linegap
@@
 	A {\bf data-race preemption point} is required to find the bug.
-	\end{center}
 \end{frame}
@@
 		\item One access {\em happens-before} the other
-			(e.g., no \texttt{cond\_signal()} in between)
-	\end{itemize}
-	\pause
-	\linegap
-
-	Two flavors
-	\begin{itemize}
-		\item {\bf Pure HB}: Whether accesses were {\em observed simultaneously}, false-negative prone
-			{\em [Lamport '78]}
-		\item {\bf Limited HB}: Whether threads {\em could be reordered}, false-positive prone
-			{\em [O'Callahan '03]}
-	\end{itemize}
-\end{frame}
-
-\begin{frame}{Iterative Deepening (in Pictures)}
+			%(e.g., no \texttt{cond\_signal()} in between)
+			(e.g., no \texttt{cond\_wait()})
+	\end{itemize}
+	\pause
+	\linegap
+
+	%Prior model checkers hard-code a fixed set of PPs, committing to one state space in advance.
+	Prior model checkers commit to a fixed set of PPs in advance.
+	\begin{itemize}
+		\item ``Preempt on {\tt mutex\_lock()}, {\tt mutex\_unlock()} only''
+		\item ``Preempt on every single memory access''
+	\end{itemize}
+	\linegap
+
+	Insight: Decide where to preempt on-the-fly
+	% say: "...then attempt to cope with the resulting exponentially sized problem, however big it may be"
+
+	%Two flavors
+	%\begin{itemize}
+	%	\item {\bf Pure HB}: Whether accesses were {\em observed simultaneously}, false-negative prone
+	%		{\em [Lamport '78]}
+	%	\item {\bf Limited HB}: Whether threads {\em could be reordered}, false-positive prone
+	%		{\em [O'Callahan '03]}
+	%\end{itemize}
+\end{frame}
+
+\begin{frame}{Quicksand (in Pictures)}
 	\begin{center}
@@
 \end{frame}
-\begin{frame}{Iterative Deepening (in Pictures)}
+\begin{frame}{Quicksand (in Pictures)}
 	\begin{center}
@@
 \end{frame}
+\begin{frame}{Quicksand (in Pictures)}
+	\begin{center}
+	\vspace{-0.88em}
+	\includegraphics[width=0.86\textwidth]{dr-jobs-maximal.pdf}
+	\end{center}
+\end{frame}
 
@@
 	\textbf{Convergence theorem:} Testing maximal state space, after detecting all data races, $\equiv$ preempting everywhere!
-	\begin{itemize}
-		\item Insight: $\forall$ arbitrary buggy interleaving, $\exists$ equivalent interleaving % say 'under DPOR' i guess?
-			using only sync/data-race PPs, \& Quicksand will eventually reach it.
-	\end{itemize}
-	\linegap
-
-	Hence, Iterative Deepening provides ``best of both worlds'':
+	%\begin{itemize}
+	%	\item Insight: $\forall$ arbitrary buggy interleaving, $\exists$ equivalent interleaving % say 'under DPOR' i guess?
+	%		using only sync/data-race PPs, \& Quicksand will eventually reach it.
+	%\end{itemize}
+	\linegap
+
+	Hence, Quicksand provides ``best of both worlds'':
 	\begin{itemize}
@@
 	\begin{itemize}
-		\item Quicksand, finding data-race candidates with...
-			\begin{itemize}
-				\item ...``Pure'' Happens-Before analysis {\em [Lamport '78]}
-				\item ...``Limited'' Happens-Before analysis {\em [O'Callahan '03]}
-			\end{itemize}
+		\item Quicksand % parallelized 10-fold
+			%, finding data-race candidates with...
+			%\begin{itemize}
+			%	\item ...``Pure'' Happens-Before analysis {\em [Lamport '78]}
+			%	\item ...``Limited'' Happens-Before analysis {\em [O'Callahan '03]}
+			%\end{itemize}
 		\item Stand-alone Landslide using ICB {\em [Musuvathi '08]}, preempting at...
@@
 	\begin{center}
-		\includegraphics[width=0.8\textwidth]{../../oopsla/bugs-talk.pdf}
+		\includegraphics[width=0.8\textwidth]{../quicksand-bugs-talkver.pdf}
 	\end{center}
@@
 	\begin{center}
-		\includegraphics[width=0.8\textwidth]{../../oopsla/bugs-wallclock-talk.pdf}
+		\includegraphics[width=0.8\textwidth]{../quicksand-bugs-wallclock-talkver.pdf}
 	\end{center}
@@
 	\begin{center}
-		\includegraphics[width=0.8\textwidth]{../../oopsla/verifs-talk.pdf}
+		\includegraphics[width=0.8\textwidth]{../quicksand-verifs-talkver.pdf}
 	\end{center}
@@
 \begin{frame}{Setting}
-	CMU (15-410, S'15-S'18), PSU (CMPSC 473, S'18)
+	%CMU (15-410, S'15-S'18), PSU (CMPSC 473, S'18)
+	CMU (15-410), PSU (CMPSC 473)
 	\begin{itemize}
@@
 
-	U. Chicago (CMSC 23000, F'17)
+	%U. Chicago (CMSC 23000, F'17)
+	U. Chicago (CMSC 23000)
 	\begin{itemize}
@@
 				% is the wrong approach to begin with
-				\item Prior work {\em [Coons '13]} likely to leave threads blocked too long
-			\end{itemize}
-		\item Check for false-positive blocking to avoid false-positive deadlocks
+					% TODO: move this to bonus slides, save a min of expl
+				%\item Prior work {\em [Coons '13]} likely to leave threads blocked too long
+			\end{itemize}
+			\pause
+		\item ``Ok, what if I just {\tt yield()} 10 times because I feel like it?''
+			\begin{itemize}
+				\item If deadlock suspected, try waking up to avoid false positives
+			\end{itemize}
+		%\item Check for false-positive blocking to avoid false-positive deadlocks
 	\end{itemize}
@@
 
-	...as well as subsequent P3 (kernel) grades
+	%...as well as subsequent P3 (kernel) grades
+	...as well as subsequent kernel grades
 	\begin{itemize}
@@
 		\begin{tabular}{l}
-			\texttt{if ((status = \hilight{darkorange}{\_xbegin}()) == SUCCESS) \{} \\ %\_XBEGIN\_STARTED) \{} \\
+			\texttt{if (\hilight{darkorange}{\_xbegin}() == SUCCESS) \{} \\ %\_XBEGIN\_STARTED) \{} \\
 			%\texttt{} \\
@@
 			{\tt int tmp = count;} \\
-			{\em \hilight{red}{<data-race PP>}} \\
+			{\em \hilight{pinkish}{<data-race PP>}} \\
 				& {\tt \hilight{darkorange}{\_xbegin}() == SUCCESS} \\
@@
 		\begin{tabular}{l}
-			\texttt{if ((status = \hilight{darkorange}{\_xbegin}()) == SUCCESS) \{} \\ %\_XBEGIN\_STARTED) \{} \\
+			\texttt{if (\hilight{darkorange}{\_xbegin}() == SUCCESS) \{} \\ %\_XBEGIN\_STARTED) \{} \\
 			%\texttt{} \\
 			%\texttt{} \\
-			\texttt{~~~~\hilight{red}{if (mutex\_held(\&count\_lock))}} \\
-			\texttt{~~~~~~~~\hilight{red}{\_xabort();}} \\
+			\texttt{~~~~\hilight{pinkish}{if (mutex\_is\_locked(\&count\_lock))}} \\
+			\texttt{~~~~~~~~\hilight{pinkish}{\_xabort();}} \\
 			\texttt{~~~~count++;} \\
@@
 			{\tt int tmp = count;} \\
-			{\em \hilight{red}{<data-race PP>}} \\
+			{\em \hilight{pinkish}{<data-race PP>}} \\
 				& {\tt \hilight{darkorange}{\_xbegin}() == SUCCESS} \\
-				& {\tt \hilight{red}{if (mutex\_held(...))}} \\
-				& {\tt \hilight{red}{~~~~\_xabort();}} \\
+				& {\tt \hilight{pinkish}{if (mutex\_is\_locked(...))}} \\
+				& {\tt \hilight{pinkish}{~~~~\_xabort();}} \\
 			{\tt count = tmp + 1;} \\
@@
 	\begin{itemize}
-		\item Transactions omitted the {\tt mutex\_held()} check
+		\item Forgot the {\tt mutex\_is\_locked()} check $\rightarrow$ data corruption
 		\item Landslide finds segfault in 15 seconds % 3,1 with quicksand; 14 interleavings, wall clock time
@@
 	\begin{itemize}
-		\item Implemented {\tt mutex\_held()} using {\tt cmpxchg} (= atomic read + write)
+		\item {\tt mutex\_is\_locked()} used {\tt cmpxchg} (= atomic read + write)
 		\item Unconditional write during {\tt xbegin} path $\rightarrow$ always aborts!
@@
 			% numbers taken from retry sets column
-		\item {\tt mutex\_held()} pattern: $K,N$ = 2,4 (79min); 4,1 (6min)
-		\item Using {\tt xadd} instead: $K,N$ = 2,4 (19min); 3,2 (39min), 4,1 (1min)
+		\item {\tt mutex\_is\_locked()} pattern: $K,N$ = 2,4 (79min); 4,1 (6min)
+		%\item Using {\tt xadd} instead: $K,N$ = 2,4 (19min); 3,2 (39min), 4,1 (1min)
+		\item (more tests in thesis)
 		\item Is that ``enough''? Generalize with manual inspection?
-		%\item More, see thesis...
 	\end{itemize}
@@
 	\begin{itemize}
-		\item Author of {\tt mutex\_held()} pattern got it wrong once...
+		\item Author of {\tt mutex\_is\_locked()} pattern got it wrong once...
 		\item Landslide can check even unpredictable performance characteristics
@@
 
-\section{Conclusion}
-
+\section{Part 4}
+\subsection{End}
+
+\breakslide{\Large Conclusion}
+
+\begin{frame}{Conclusion}
+	% data race pps (as well as txn equivalence)
+	Combining %both
+	theoretically-founded automatic reduction techniques
+	% iterative deepening
+	% (say smth like: "landslide also has a wealth of other user-tunable heuristics
+	% for expert users, which you can read about in the document; for this talk i will focus just on the CPU budget thing")
+	% XXX: need to prepare for the question -- "what evidence do you have that users actually want this cpu budget stuff",
+	% ",or, that it helps vs just running something unlimited vs a normal icb test"
+	% (well, for the 2nd half of the Q, i can talk about how ID outperforms ICB on verficiations)
+	and user-informed heuristic ones, stateless model checking...
+	\begin{itemize}
+		\item ...can sufficiently mitigate exponential explosion...
+		% education
+		\item ...to be a practical testing technique for inexperienced users...
+		% htm
+		% TODO: this last part you just added for this talk; does the thesis version of the statmeent need to include too?
+		\item ...and real-world programs alike.
+	\end{itemize}
+\end{frame}
 \begin{frame}{Conclusion}
@@
 \begin{frame}{Conclusion}
-	Combining %both
-	{\bf theoretically-founded automatic reduction techniques}
+	\hilight{gray}{Combining %both
+	{\bf theoretically-founded automatic reduction techniques}}
 	\begin{itemize}
 		\item Verification with data-race preemption points
-		\item Yield-loop blocking
+		\item Student yield-loop blocking % student (adj). from "studence"
 		\item Equivalent emulation of HTM semantics
 	\end{itemize}
-	and {\bf user-informed heuristic ones},
+	\hilight{gray}{and {\bf user-informed heuristic ones},}
 	\begin{itemize}
@@
 			{\bf Thread 1}  & {\bf Thread 2} \\
-			\texttt{\hilight{red}{x++;}}    & \\
+			\texttt{\hilight{pinkish}{x++;}}    & \\
 			\texttt{mutex\_lock(m);}	& \\
@@
 				& \texttt{mutex\_unlock(m);} \\
-				& \texttt{\hilight{red}{x++;}} \\
+				& \texttt{\hilight{pinkish}{x++;}} \\
 		\end{tabular}
@@
 				& \texttt{mutex\_unlock(m);} \\
-				& \texttt{\hilight{lavender}{if (tmp)}~\hilight{red}{x++;}} \\
+				& \texttt{\hilight{lavender}{if (tmp)}~\hilight{pinkish}{x++;}} \\
 		\end{tabular}
@@
 
-\begin{frame}{Iterative Deepening Reductions}
+\begin{frame}{Quicksand (Iterative Deepening) Reductions}
 	If one job is a subset of another, testing either might let us skip the other.
@@
 \end{frame}
-\begin{frame}{Iterative Deepening Reductions}
+\begin{frame}{Quicksand (Iterative Deepening) Reductions}
 	If one job is a subset of another, testing either might let us skip the other.
@@
 
-%\begin{frame}{Quicksand results - Bugs found by CPU time}
-%	\begin{center}
-%		\includegraphics[width=0.8\textwidth]{../../oopsla/bugs-talk.pdf}
-%	\end{center}
-%\end{frame}
-%\begin{frame}{Quicksand results - Bugs found by wall-clock time}
-%	\begin{center}
-%		\includegraphics[width=0.8\textwidth]{../../oopsla/bugs-wallclock-talk.pdf}
-%	\end{center}
-%\end{frame}
-%\begin{frame}{Quicksand results - Tests fully verified (CPU time)}
-%	\begin{center}
-%		\includegraphics[width=0.8\textwidth]{../../oopsla/verifs-talk.pdf}
-%	\end{center}
-%\end{frame}
+\begin{frame}{Quicksand results - Bugs found by CPU time}
+	\begin{center}
+		\includegraphics[width=0.8\textwidth]{../quicksand-bugs-hb-talkver.pdf}
+	\end{center}
+\end{frame}
+\begin{frame}{Quicksand results - Bugs found by wall-clock time}
+	\begin{center}
+		\includegraphics[width=0.8\textwidth]{../quicksand-bugs-hb-wallclock-talkver.pdf}
+	\end{center}
+\end{frame}
+\begin{frame}{Quicksand results - Tests fully verified (CPU time)}
+	\begin{center}
+		\includegraphics[width=0.8\textwidth]{../quicksand-verifs-hb-talkver.pdf}
+	\end{center}
+\end{frame}
 \begin{frame}{Quicksand results - MC + DR vs single-pass DR}
@@
 	\begin{tabular}{r|c|c}
-		{\bf Bound} & {\bf SSS (sync PPs only)} & {\bf SSS (preempt everywhere)} \\
+		{\bf Bound} & {\bf ICB (sync PPs only)} & {\bf ICB (preempt everywhere)} \\
 		\hline
@@
 
-	Bugs by type (Pure HB trial):
+	Bugs by type (Pure HB Quicksand):
 	\begin{itemize}
@@
 		\item 35 assertion failures
-		\item 31 page fault crashes
+		\item 31 page/segmentation fault crashes
 		\item 1 infinite loop
@@
 		\begin{tabular}{l}
-			\texttt{if ((status = \_xbegin()) == \_XBEGIN\_STARTED) \{} \\
+			\texttt{if ((status = \hilight{darkorange}{\_xbegin}()) == \_XBEGIN\_STARTED) \{} \\
 			\texttt{~~~~if (count >= 100)}\\
-			\texttt{~~~~~~~~\_xabort();} \\
+			\texttt{~~~~~~~~\hilight{pinkish}{\_xabort}();} \\
 			\texttt{~~~~count++;} \\
-			\texttt{~~~~\_xend();} \\
+			\texttt{~~~~\hilight{darkblue}{\_xend}();} \\
 			\texttt{\} else if (status != \_XABORT\_EXPLICIT) \{} \\
-			\texttt{~~~~mutex\_lock(\&count\_lock);} \\
+			\texttt{~~~~\hilight{darkorange}{mutex\_lock}(\&count\_lock);} \\
 			\texttt{~~~~count++;} \\
-			\texttt{~~~~mutex\_unlock(\&count\_lock);} \\
+			\texttt{~~~~\hilight{darkblue}{mutex\_unlock}(\&count\_lock);} \\
 			\texttt{\}} \\
@@
 
-\begin{frame}{HTM: Sources of Benchmarks}
-	\begin{itemize}
-		%\item Shamay, https://github.com/cmpxchg16/tsx.me, 2014
-		\item Dehesa-Azuara and Stanley, http://www.contrib.andrew.cmu.edu/\textasciitilde{}mdehesaa/, 2016
-		%\item Chapman et al., Hybrid STM/HTM for nested transactions on OpenJDK, OOPSLA 2016
-		%\item Wang et al., Eunomia: Scaling Concurrent Search Trees under Contention Using HTM, PPoPP 2017
-			% TODO: add, that one lock thingy
-		\item vamsikc, TSX LevelDB implementation, https://github.com/vamsikc/leveldb-tsx/, 2014
-	\end{itemize}
-\end{frame}
+%\begin{frame}{HTM: Sources of Benchmarks}
+%	\begin{itemize}
+%		%\item Shamay, https://github.com/cmpxchg16/tsx.me, 2014
+%		\item Dehesa-Azuara and Stanley, http://www.contrib.andrew.cmu.edu/\textasciitilde{}mdehesaa/, 2016
+%		%\item Chapman et al., Hybrid STM/HTM for nested transactions on OpenJDK, OOPSLA 2016
+%		%\item Wang et al., Eunomia: Scaling Concurrent Search Trees under Contention Using HTM, PPoPP 2017
+%			% TODO: add, that one lock thingy
+%		\item vamsikc, TSX LevelDB implementation, https://github.com/vamsikc/leveldb-tsx/, 2014
+%	\end{itemize}
+%\end{frame}
 
@@
 %% TODO: slide number c.c
-%\begin{frame}{Don't you need \texttt{if (mutex\_held(\&count\_lock)) \_xabort();} in that transaction on slide 42?}
+%\begin{frame}{Don't you need \texttt{if (mutex\_is\_locked(\&count\_lock)) \_xabort();} in that transaction on slide 42?}
 %	\begin{center}
@@
 		\item {\bf [Yang '08]}: Yu Yang et al. Efficient stateful DPOR. SPIN 2008.
-		\item {\bf [Buckhardt '10]}: A randomized scheduler with probabilistic guarantees of finding bugs. ASPLOS 2010.
+		\item {\bf [Klein '09]}: Gerwin Klein et al. seL4: formal verif. of an OS kernel. SOSP 2009.
 	\end{itemize}
@@
 	\begin{itemize}
+		%\item {\bf [Burckhardt '10]}: A randomized scheduler with probabilistic guarantees of finding bugs. ASPLOS 2010.
 		\item {\bf [Simsa '12]}: Jiri Simsa. Runtime Estimation and Resource Allocation for
@@
 			Ph.D. thesis (CMU-CS-13-133). December 2013.
-		\item {\bf [Coons '13]}: Katherine Coons et al. Bounded POR. OOPSLA 2013.
+		%\item {\bf [Coons '13]}: Katherine Coons et al. Bounded POR. OOPSLA 2013.
 		\item {\bf [Abdulla '14]}: Parosh Abdulla et al. Optimal DPOR. POPL 2014.
+		\item {\bf [Matsakis '14]}: Nicholas Matsakis et al. The Rust Language. HILT 2014.
 		\item {\bf [vamsikc '14]}: leveldb-tsx. https://github.com/vamsikc/leveldb-tsx/, 2014.
-		\item {\bf [Huang '15]}: Jeff Huang. Stateless MCing concurrent programs with maximal causality reduction. PLDI '15.
+		%\item {\bf [Huang '15]}: Jeff Huang. Stateless MCing concurrent programs with maximal causality reduction. PLDI '15.
 		\item {\bf [Jensen '15]}: Stateless MCing of Event-Driven Applications. OOPSLA 2015.

### 2018-10-16 thesis/defense/snakefight.tex
buncha updates from practice 2, color coding etc

@@
 
+\definecolor{sect-quicksand}{RGB}{0,96,128}
+\definecolor{sect-410}{RGB}{0,128,64}
+\definecolor{sect-htm}{RGB}{128,0,64}
+\definecolor{sect-pastel-quicksand}{RGB}{144,180,192}
+\definecolor{sect-pastel-410}{RGB}{144,192,168}
+\definecolor{sect-pastel-htm}{RGB}{192,144,168}
+
 \begin{document}
-\renewcommand{\inserttotalframenumber}{67}
+\renewcommand{\inserttotalframenumber}{69}
 \normalem
@@
 
+% TODO: on the title slide, shouldn't there be a "PResented in partial requirement..."?
+
 \section{Part 0}
@@
 
-\begin{frame}{Concurrency!}
+\begin{frame}{Concurrency}
 	\begin{center}
@@
 \end{frame}
-\begin{frame}{Concurrency!}
+\begin{frame}{Concurrency}
 	\begin{center}
@@
 
+\begin{frame}{Concurrency Testing} % "..is hard"?
+	% what's so bad abt concurrecny bugs
+	% - manifest at random depending on interrupts/etc
+	%   - (fine in testing but in production, loaded server, more interrupts, more interleavings, more crashes)
+	% - difficult to reproduce
+	% - difficult to know yr testing "coverage" - did you *really* fix that bug, or just make it more rare?
+	% - difficult to use, as evidenced by most systems classes still teaching kids to use stress (or worse, unit) tests
+	Concurrency bugs...
+	\begin{itemize}
+		\item Manifest at random
+		\item Difficult to reproduce
+		\item Difficult to know when truly fixed % what is ``coverage''?
+	\end{itemize}
+	\linegap
+
+	Testing approaches
+	\begin{itemize}
+		\item Unit tests: only 1 ``most likely'' interleaving
+		\item Stress tests: exposes ``less likely'' interleavings (if lucky), % to expose "less likely" ones
+			cannot verify
+		\item Verification: steep learning curve, unpredictable runtime
+			%still no widespread adoption % -- why?
+	\end{itemize}
+\end{frame}
+\begin{frame}{Practical Concurrency Testing}
+	%Goals
+	\textbf{Landslide}, this thesis's solution
+	\begin{itemize}
+		\item No randomness, despite nondeterminism
+		\item Confirm when you fixed bugs, not just find them % not 'verify'
+			% TODO: make alnguage match, "Automatic so and so"
+		\item User-friendly enough for students
+		% \item Test wide variety of types of concurrency % txn
+			% TODO : figure out how to ref 
+		% ????
+		\item Combines with human intuition to verify tests neither could do alone
+	\end{itemize}
+	%
+	% enter landslide
+	% - test yr nondeterministic program without randomness/unpredictability
+	% - provide formal verification guarantees so you know when you actually fixed that bug
+	%   - (well, subject to assumptions about the test case you're running under, etc etc)
+	% - user friendly enough even for ze studence
+	% 
+	% say: "What if there was a program that... ...I'm here to tell you that I've built such a program,
+	% it's called Landslide, and why you should let me graduate for it" (figure out smth more diplomatic here...)
+\end{frame}
+
 \begin{frame}{Stateless Model Checking}
-	Stateless Model Checking {\em [Godefroid '97]} (MC)
+	Stateless Model Checking (MC) {\em [Godefroid '97]}
 	\begin{itemize}
@@
 	\begin{center}
-		\includegraphics[width=0.25\textwidth]{../../oopsla/exponential-curve.pdf}
-	\end{center}
-\end{frame}
-
-\begin{frame}{State Space Example}
+		\includegraphics[width=0.4\textwidth]{exponence.pdf}
+	\end{center}
+\end{frame}
+
+% better: remember to say, "If you're wondering what the deal is with "stateless"...
+% the name is a historical accident & we can talk abt that in the Q+As"
+\begin{frame}{Stateless Model Checking - The State Space}
 	\begin{center}
@@
 
-	Possible interleavings represented as a tree
+	Possible interleavings (under given test case) represented as a tree
 	\begin{itemize}
@@
 			% TODO: i just switched the order of these; see how that feels
-		\item Manual kernel annotations: Landslide {\em [Blum '12]} {\tiny that's me!}
+		\item Manual kernel annotations {\em [Blum '12]} %{\tiny that's me!}
 			\pause
@@
 	% education
-	a practical testing technique for inexperienced users
+	practical for inexperienced users
+	%a practical testing technique for inexperienced users
 	% htm
@@
 	\begin{itemize}
-		\item ...can sufficiently mitigate exponential explosion...
+		\item \hilight{sect-quicksand}{...can sufficiently mitigate exponential explosion...}\xspace
 		% education
-		\item ...to be a practical testing technique for inexperienced users...
+		%\item ...to be a practical testing technique for inexperienced users...
+		\item \hilight{sect-410}{...to be practical for inexperienced users...}\xspace
 		% htm
 		% TODO: this last part you just added for this talk; does the thesis version of the statmeent need to include too?
-		\item ...and real-world programs alike.
-	\end{itemize}
-\end{frame}
-
-\begin{frame}{Thesis Statement}
-	The Quicksand algorithm %combines theory and heuristics to
-	practically debugs and verifies many exponential state spaces.
+		\item \hilight{sect-htm}{...and real-world programs alike.}\xspace
+	\end{itemize}
+\end{frame}
+
+\begin{frame}{Contributions}
+	\hilight{sect-quicksand}{The Quicksand algorithm %combines theory and heuristics to
+	practically debugs and verifies many exponential state spaces.}
 	\vspace{4.54em}
 
-	Landslide's automatic instrumentation makes MC accessible to students.
+	\hilight{sect-410}{Landslide's automatic instrumentation makes MC accessible to students.}
 	\vspace{4.54em}
 
-	New reduction techniques for transactional memory keep MC up-to-date
-	with the latest real-world concurrency techniques.
+	\hilight{sect-htm}{New reduction techniques for transactional memory keep MC up-to-date
+	with the latest real-world concurrency techniques.}
 	\vspace{4.54em}
@@
 \begin{frame}{Outline}
-	\hilight{gray}{The Quicksand algorithm %combines theory and heuristics to
+	\hilight{sect-pastel-quicksand}{The Quicksand algorithm %combines theory and heuristics to
 	practically debugs and verifies many exponential state spaces.} \\
@@
 	\end{itemize}
-	\hilight{gray}{Landslide's automatic instrumentation makes MC accessible to students.} \\
+	\hilight{sect-pastel-410}{Landslide's automatic instrumentation makes MC accessible to students.} \\
 	%\hilight{gray}{Ad-hoc synchronization heuristics make MC accessible to student programmers.} \\
@@
 
-	\hilight{gray}{New reduction techniques for transactional memory keep MC up-to-date
+	\hilight{sect-pastel-htm}{New reduction techniques for transactional memory keep MC up-to-date
 	with the latest real-world concurrency techniques.} \\
@@
 
-\breakslide{\Large Quicksand}
+\breakslide{\Large \bf \hilight{sect-quicksand}{Quicksand}}
 
 \begin{frame}{Preemption Points}
-	The burning question: ``Which {\bf preemption points} (PPs) are important?''
-	\linegap
-
-	State space of interleavings is {\em parameterized} by PPs.
-	\begin{itemize}
-		\item Too few PPs: Many bugs will go undetected
+	The burning question: ``Which {\bf preemption points} are important?''
+	\linegap
+
+	State space %of interleavings
+	is {\em parameterized} by preemption points.
+	\begin{itemize}
+		\item Too few: Many bugs will go undetected
 		\item Preempt everywhere: Completion is often infeasible % say: "for some programs"
@@
 		\item The threads are not holding the same mutex
-		\item One access {\em happens-before} the other
+		\item Neither access {\em happens-before} the other
 			%(e.g., no \texttt{cond\_signal()} in between)
@@
 	%Prior model checkers hard-code a fixed set of PPs, committing to one state space in advance.
-	Prior model checkers commit to a fixed set of PPs in advance.
+	%Prior model checkers commit to a fixed set of PPs in advance.
+	Prior model checkers commit to fixed preemption points in advance.
 	\begin{itemize}
@@
 	\textbf{Convergence theorem:} Testing maximal state space, after detecting all data races, $\equiv$ preempting everywhere!
-	%\begin{itemize}
-	%	\item Insight: $\forall$ arbitrary buggy interleaving, $\exists$ equivalent interleaving % say 'under DPOR' i guess?
-	%		using only sync/data-race PPs, \& Quicksand will eventually reach it.
-	%\end{itemize}
+	\begin{itemize}
+		\item $\forall$ arbitrary buggy interleaving, $\exists$ equivalent interleaving % say 'under DPOR' i guess?
+			using only sync/data-race PPs, \& Quicksand will eventually reach it.
+			\begin{itemize}
+				\item ({\em equivalence} defined under DPOR {\em [Flanagan '05]})
+			\end{itemize}
+	\end{itemize}
 	\linegap
@@
 \begin{frame}{Summary}
-	Incorporate data-race analysis to add new PPs in a feedback loop
-	\linegap
-
-	Enough time to saturate \& test every data-race PP? Full verification
+	Use data-race analysis to add new preemption points in a feedback loop
+	\linegap
+
+	Enough time to saturate \& test every data-race? Full verification
 	\linegap
@@
 
-\breakslide{\Large Education}
+\breakslide{\Large \bf \hilight{sect-410}{Education}}
 
@@
 			%\pause
-		\item PSU: (no log collection due to IRB issues)
+		\item PSU: (no log collection) %due to IRB issues)
 		% \item Case study tej
@@
 
-\begin{frame}{Effect on Grades}
+% TODO: put p values
+\begin{frame}{Effect on Grades (CMU)} % - statistically significant?}
 	\begin{center}
@@
 
-\begin{frame}{Effect on Grades}
+\begin{frame}{Effect on Grades (CMU)} % - statistically significant?}
 	\begin{center}
@@
 
-\begin{frame}{Statistical Significance}
-	Compared %users' vs. non-users'
-	grade distributions with k-sample Anderson-Darling {\em [A., D. '52]}
-	\linegap
-
-	Control 1: non-users from {\bf same semester}
-	\begin{itemize}
-		\item {\bf P2} grade distributions differ with $p = 0.036 < 0.05$
-		\item {\bf P3} grade distributions differ with $p = 0.034 < 0.05$
-			\pause
-		\item Possible bias: stronger students volunteer for Landslide in first place
-	\end{itemize}
-	\pause
-	\linegap
-
-	Control 2: non-users from {\bf past semesters} (when Landslide not offered)
-	\begin{itemize}
-		\item {\bf P2} grade distributions differ with $p = 0.671$...
-		\item {\bf P3} grade distributions differ with $p = 0.145$...
-			\pause
-		\item Possible bias: grading criteria may drift across semesters
-			% FIXME does this bullet belong in this sub list
-		\item Possible bias: maybe Landslide finds bugs the official tests don't
-	\end{itemize}
-\end{frame}
+%\begin{frame}{Statistical Significance}
+%	Compared %users' vs. non-users'
+%	grade distributions with k-sample Anderson-Darling {\em [A., D. '52]}
+%	\linegap
+%
+%	Control 1: non-users from {\bf same semester}
+%	\begin{itemize}
+%		\item {\bf P2} grade distributions differ with $p = 0.036 < 0.05$
+%		\item {\bf P3} grade distributions differ with $p = 0.034 < 0.05$
+%			\pause
+%		\item Possible bias: stronger students volunteer for Landslide in first place
+%	\end{itemize}
+%	\pause
+%	\linegap
+%
+%	Control 2: non-users from {\bf past semesters} (when Landslide not offered)
+%	\begin{itemize}
+%		\item {\bf P2} grade distributions differ with $p = 0.671$...
+%		\item {\bf P3} grade distributions differ with $p = 0.145$...
+%			\pause
+%		\item Possible bias: grading criteria may drift across semesters
+%			% FIXME does this bullet belong in this sub list
+%		\item Possible bias: maybe Landslide finds bugs the official tests don't
+%	\end{itemize}
+%\end{frame}
 
@@
 
-\begin{frame}{Survey Results (educational value)}
+\begin{frame}{Survey Results} % (educational value)}
 	\begin{center}
@@
 
-\begin{frame}{Survey Results (happiness value)}
-	\begin{center}
-		\includegraphics[width=\textwidth]{../survey5.pdf} % easy to understand?
-	\end{center}
-\end{frame}
-
-\begin{frame}{Survey Results (educational value)}
+%\begin{frame}{Survey Results} % (happiness value)}
+%	\begin{center}
+%		\includegraphics[width=\textwidth]{../survey5.pdf} % easy to understand?
+%	\end{center}
+%\end{frame}
+
+\begin{frame}{Survey Results} % (educational value)}
 	\begin{center}
@@
 
-\begin{frame}{Survey Results (happiness value)}
+\begin{frame}{Survey Results} % (happiness value)}
 	\begin{center}
@@
 
-\breakslide{\Large Transactional Memory}
+\breakslide{\Large \bf \hilight{sect-htm}{Transactional Memory}}
 
@@
 			\end{itemize}
-			% TODO prob just comment this out
-		\item More in bonus slides...
+		%\item More in bonus slides...
 	\end{itemize}
@@
 
-\breakslide{\Large Conclusion}
+\breakslide{\Large \bf Conclusion}
 
@@
 	% (well, for the 2nd half of the Q, i can talk about how ID outperforms ICB on verficiations)
-	and user-informed heuristic ones, stateless model checking...
-	\begin{itemize}
-		\item ...can sufficiently mitigate exponential explosion...
-		% education
-		\item ...to be a practical testing technique for inexperienced users...
-		% htm
-		% TODO: this last part you just added for this talk; does the thesis version of the statmeent need to include too?
-		\item ...and real-world programs alike.
-	\end{itemize}
+	and user-informed heuristic ones,
+	\linegap
+
+	stateless model checking
+	\hilight{sect-quicksand}{can sufficiently mitigate exponential explosion}\xspace
+	\hilight{sect-410}{to be practical for inexperienced users}\xspace
+	\hilight{sect-htm}{and real-world programs alike.}\xspace
 \end{frame}
@@
 	\vspace{4.82em}
-
-	stateless model checking...
-	\begin{itemize}
-		\item ...can sufficiently mitigate exponential explosion...
-		\item ...to be a practical testing technique for inexperienced users...
-		\item ...and real-world programs alike.
-	\end{itemize}
+	\linegap
+
+	stateless model checking
+	\hilight{sect-quicksand}{can sufficiently mitigate exponential explosion}\xspace
+	\hilight{sect-410}{to be practical for inexperienced users}\xspace
+	\hilight{sect-htm}{and real-world programs alike.}\xspace
 \end{frame}
 \begin{frame}{Conclusion}
-	\hilight{gray}{Combining %both
-	{\bf theoretically-founded automatic reduction techniques}}
-	\begin{itemize}
-		\item Verification with data-race preemption points
-		\item Student yield-loop blocking % student (adj). from "studence"
-		\item Equivalent emulation of HTM semantics
-	\end{itemize}
-	\hilight{gray}{and {\bf user-informed heuristic ones},}
-	\begin{itemize}
-		\item Quicksand's CPU budget argument
-		\item Exposing data-race reports to students
-		\item Abstraction reduction for HTM primitives
-	\end{itemize}
-	\hilight{gray}{stateless model checking...}
-	\begin{itemize}
-		\item \hilight{gray}{...can sufficiently mitigate exponential explosion...}
-		\item \hilight{gray}{...to be a practical testing technique for inexperienced users...}
-		\item \hilight{gray}{...and real-world programs alike.}
-	\end{itemize}
+	%\hilight{gray}{Combining %both
+	%{\bf theoretically-founded automatic reduction techniques}}
+	Combining %both
+	{\bf theoretically-founded automatic reduction techniques}
+	\begin{itemize}
+		\item \hilight{sect-quicksand}{Verification with data-race preemption points}
+		\item \hilight{sect-410}{Student yield-loop blocking} % student (adj). from "studence"
+		\item \hilight{sect-htm}{Equivalent emulation of HTM semantics}
+	\end{itemize}
+	% TODO: incremental reveal
+	%\hilight{gray}{and {\bf user-informed heuristic ones},}
+	and {\bf user-informed heuristic ones},
+	\begin{itemize}
+		\item \hilight{sect-quicksand}{Quicksand's CPU budget argument}
+		\item \hilight{sect-410}{Exposing data-race reports to students}
+		%\item Encouraging students to exercise their intuition % ??
+		\item \hilight{sect-htm}{Abstraction reduction for HTM primitives}
+	\end{itemize}
+	\linegap
+
+	\hilight{gray}{stateless model checking}\xspace
+	\hilight{sect-pastel-quicksand}{can sufficiently mitigate exponential explosion}\xspace
+	\hilight{sect-pastel-410}{to be practical for inexperienced users}\xspace
+	\hilight{sect-pastel-htm}{and real-world programs alike.}\xspace
 \end{frame}
@@
 \subsection{Quicksand}
-\breakslide{\Large Bonus Slides - Quicksand}
+\breakslide{\Large \bf \hilight{sect-quicksand}{Bonus Slides - Quicksand}}
 
@@
 \subsection{Education}
-\breakslide{\Large Bonus Slides - Education}
+\breakslide{\Large \bf \hilight{sect-410}{Bonus Slides - Education}}
 
@@
 \subsection{HTM}
-\breakslide{\Large Bonus Slides - HTM}
+\breakslide{\Large \bf \hilight{sect-htm}{Bonus Slides - HTM}}
 
@@
 	\begin{itemize}
-		\item {\bf [A., D. '52]}: T. Anderson and D. Darling.
-			Asymptotic theory of certain ``goodness-of-fit'' criteria based on stochastic processes.
-			Annals of Statistics, 1952.
+		%\item {\bf [A., D. '52]}: T. Anderson and D. Darling.
+		%	Asymptotic theory of certain ``goodness-of-fit'' criteria based on stochastic processes.
+		%	Annals of Statistics, 1952.
 		\item {\bf [Lamport '78]}:

### 2018-10-17 thesis/defense/snakefight.tex
snakefight updates after practice 3 upto bonus slides

@@
 			cannot verify
-		\item Verification: steep learning curve, unpredictable runtime
+		\item Verification: steep learning curve, prohibitive runtime
 			%still no widespread adoption % -- why?
@@
 		\item Edge: State transition from executing a thread
-	\end{itemize}
-	\linegap
-
-	State space size depends on which {\bf preemption points} are used.
+		\item Nondeterminism captured as different branches
+		\item State space size depends on which {\bf preemption points} used
+	\end{itemize}
 \end{frame}
@@
 \begin{frame}{Challenges} %\& Existing Coping Techniques}
-	Unpredictable state space explosion
+	State space explosion
 	\begin{itemize}
 		%\item Equivalence reductions: DPOR {\em [Flanagan '05]}, MCR {\em [Huang '15]}
-		\item Dynamic Partial Order Reduction (DPOR) {\em [Flanagan '05]}
-			\begin{itemize}
-				\item identifies \& skips equivalent interleavings
-			\end{itemize}
+		%\item Dynamic Partial Order Reduction (DPOR) {\em [Flanagan '05]}
+		\item Reduction algorithms to skip equivalent interleavings {\em [Flanagan '05]}
+			%\begin{itemize}
+			%	\item identifies \& skips equivalent interleavings
+			%\end{itemize}
+		\item Heuristic search ordering to expose bugs faster {\em [Musuvathi '08]}
 		%\item Heuristic search ordering: ICB {\em [Musuvathi '08]}, PCT {\em [Burckhardt '10]}
-		\item Iterative Context Bounding (ICB) {\em [Musuvathi '08]}
-			\begin{itemize}
-				\item finds bugs faster by prioritizing interleavings with fewer preemptions
-			\end{itemize}
+		%\item Iterative Context Bounding (ICB) {\em [Musuvathi '08]}
+		%	\begin{itemize}
+		%		\item finds bugs faster by prioritizing interleavings with fewer preemptions
+		%	\end{itemize}
 			\pause
@@
 	\begin{itemize}
+			% TODO: remember to say, "that lets you choose more intelligently, but still by hand, yr PPs
 		\item State space size estimation {\em [Simsa '12]}
@@
 \begin{frame}{Contributions}
+	% this vspace used to be 4.54 for all 3 before i rephrased the quicksand bullet
+	% even tho it was 2 lines before and 2 lines now
+	% i had to do... this??? latex makes no friggin sense
 	\hilight{sect-quicksand}{The Quicksand algorithm %combines theory and heuristics to
-	practically debugs and verifies many exponential state spaces.}
-	\vspace{4.54em}
+	practically debugs and verifies many tests despite exponential explosion.}
+	\vspace{4.00em}
 
 	\hilight{sect-410}{Landslide's automatic instrumentation makes MC accessible to students.}
-	\vspace{4.54em}
+	\vspace{4.58em}
 
@@
 	\vspace{4.54em}
+
 \end{frame}
@@
 	\hilight{sect-pastel-quicksand}{The Quicksand algorithm %combines theory and heuristics to
-	practically debugs and verifies many exponential state spaces.} \\
+	practically debugs and verifies many tests despite exponential explosion.}
 	%\hilight{gray}{The Quicksandalgorithm makes MC appropriate for both educational and real-world programs.} \\
@@
 
-\begin{frame}{Data Race Analysis}
+\begin{frame}{Data Race Analysis {\em [Savage '97]}}
 	A {\bf data race} is a pair of memory accesses between two threads, where:
@@
 		\item Neither access {\em happens-before} the other
+		%\item Neither access happens-before the other
 			%(e.g., no \texttt{cond\_signal()} in between)
@@
 	\end{itemize}
-	\pause
-	\linegap
-
+\end{frame}
+\begin{frame}{Quicksand}
 	%Prior model checkers hard-code a fixed set of PPs, committing to one state space in advance.
@@
 
-	Insight: Decide where to preempt on-the-fly
+	Insight: decide where to preempt on-the-fly
+	\begin{itemize}
+		\item Data-race analysis identifies new preemption points
+		\item User supplies CPU budget
+		\item Prioritize based on state space estimation {\em [Simsa '12]}
+	\end{itemize}
 	% say: "...then attempt to cope with the resulting exponentially sized problem, however big it may be"
@@
 	\vspace{-0.88em}
-	\includegraphics[width=0.86\textwidth]{../../oopsla/dr-jobs-3.pdf}
+	%\includegraphics[width=0.86\textwidth]{../../oopsla/dr-jobs-3.pdf}
+	\includegraphics[width=0.86\textwidth]{dr-jobs-3-alt.pdf}
 	\end{center}
@@
 			using only sync/data-race PPs, \& Quicksand will eventually reach it.
-			\begin{itemize}
-				\item ({\em equivalence} defined under DPOR {\em [Flanagan '05]})
-			\end{itemize}
+			%\begin{itemize}
+			%	\item ({\em equivalence} defined under DPOR {\em [Flanagan '05]})
+			%\end{itemize}
 	\end{itemize}
@@
 				\item POSIX-like userspace thread library %(also teams of 2)
+				\item Implement {\tt thr\_create()}/{\tt exit()}/{\tt join()}, {\tt mutex\_lock()}, etc.
 				\item Reference ``Pebbles'' kernel provided
 					% say: later they implement this in p3
-				\item Fixed APIs $\rightarrow$ instrumentation can be automated
-				\item Still needs heuristics to handle ad-hoc synchronization
-			\end{itemize}
-			% FIXME: need introduce p3 here?
-			% Or cut?
+				%\item Fixed APIs $\rightarrow$ instrumentation can be automated
+				%\item Still needs heuristics to handle ad-hoc synchronization
+			\end{itemize}
 		\item Kernel (``P3'') - reimplement said reference kernel (CMU only)
@@
 	\begin{itemize}
-		\item Kernel-level testing requires manual annotation {\em [Blum '12]}
+		\item Testing Pebbles kernels requires manual annotation {\em [Blum '12]}
 		\item P2 (fixed APIs), Pintos (basecode provided) annotated automatically
@@
 
-\begin{frame}{Method}
+\begin{frame}{User Study}
 	P2 (CMU + PSU)
@@
 		\item CMU: 115 deterministic, 122 concurrency bugs in 75/103 P2s
+			%\begin{itemize}
+			%	\item 3 false positives (self-reported)
+			%\end{itemize}
 		\item U. Chicago: 3 deterministic, 4 concurrency bugs in 3/4 kernels
@@
 \begin{frame}{Effect on Grades (CMU)}
-	Compared Landslide users' vs. non-users' submitted P2 grades...
+	Compared Landslide users' vs. non-users' submitted P2 grades
 	\begin{itemize}
@@
 \end{frame}
-
-% TODO: put p values
 \begin{frame}{Effect on Grades (CMU)} % - statistically significant?}
@@
 \end{frame}
-
 \begin{frame}{Effect on Grades (CMU)} % - statistically significant?}
 	\begin{center}
-		\includegraphics[width=\textwidth]{../photo-of-ze-studence-cdf.pdf}
-	\end{center}
-\end{frame}
-
-%\begin{frame}{Statistical Significance}
-%	Compared %users' vs. non-users'
-%	grade distributions with k-sample Anderson-Darling {\em [A., D. '52]}
-%	\linegap
-%
-%	Control 1: non-users from {\bf same semester}
-%	\begin{itemize}
-%		\item {\bf P2} grade distributions differ with $p = 0.036 < 0.05$
-%		\item {\bf P3} grade distributions differ with $p = 0.034 < 0.05$
-%			\pause
-%		\item Possible bias: stronger students volunteer for Landslide in first place
-%	\end{itemize}
-%	\pause
-%	\linegap
-%
-%	Control 2: non-users from {\bf past semesters} (when Landslide not offered)
-%	\begin{itemize}
-%		\item {\bf P2} grade distributions differ with $p = 0.671$...
-%		\item {\bf P3} grade distributions differ with $p = 0.145$...
-%			\pause
-%		\item Possible bias: grading criteria may drift across semesters
-%			% FIXME does this bullet belong in this sub list
-%		\item Possible bias: maybe Landslide finds bugs the official tests don't
-%	\end{itemize}
-%\end{frame}
+		\includegraphics[width=\textwidth]{photo-of-ze-studence-cdf.pdf}
+	\end{center}
+\end{frame}
 
@@
 			\begin{itemize}
-				\item slow; invites state space explosion
+				\item Slow; invites state space explosion
 			\end{itemize}
 		\item Enforce atomicity via scheduler, anticipate aborts on every transaction
-			%via analysis
-			%\begin{itemize}
-			%	\item (under HTM, all transactions might abort)
-			%\end{itemize}
-		\item Suffices to enumerate all {\em observationally-equivalent} schedules.
-	\end{itemize}
-\end{frame}
-
-\begin{frame}{Example}
+			\vspace{-1em} % latex... WTF??
+			\begin{itemize}
+				\item Suffices to enumerate all observationally-equivalent schedules.
+			\end{itemize}
+	\end{itemize}
+\end{frame}
+
+\begin{frame}{Example (Hardware POV)}
 	\begin{center}
@@
 		\item Transactional writes, if successful, appear atomically %to other threads
-		\begin{itemize}
-			\item Equivalent to a lock protecting that state
-				% note that also inserting DR PPs within a transaction is not allowed
-		\end{itemize}
-			\pause
+		%\begin{itemize}
+		%	\item Equivalent to a lock protecting that state
+		%		% note that also inserting DR PPs within a transaction is not allowed
+		%\end{itemize}
+		%	\pause
 		\item Simultaneous transactions succeed only if no conflicts
-		\begin{itemize}
-			\item Equivalent
-				%(modulo performance)
-				to serialize independent transactions
-		\end{itemize}
-			\pause
+		%\begin{itemize}
+		%	\item Equivalent
+		%		%(modulo performance)
+		%		to serialize independent transactions
+		%\end{itemize}
+		%	\pause
 		\item Failing transactions' effects rolled-back with no trace
-		\begin{itemize}
-			\item Treat each \xbegin as a failure injection point
-		\end{itemize}
+		%\begin{itemize}
+		%	\item Treat each \xbegin as a failure injection point
+		%\end{itemize}
 	\end{itemize}
@@
 	\begin{itemize}
+		\item $\forall$ interleaving within transactions, $\exists$ equivalent interleaving
+			treating them atomically, which either succeeds or aborts immediately
 		\item Safe to automatically skip all interleavings within transactions %$\qed$
@@
 				\item Exponential state space reduction
-				\item Traditional reduction (DPOR) applied on top
+				\item Traditional reduction %(DPOR)
+					applied on top
 			\end{itemize}
@@
 \begin{center}
-	% TODO: replace with HIL for honban
-	\Large Q\&A
-	%\includegraphics[width=0.4\textwidth]{../how-i-learned.png}
-\end{center}
-%\linegap
-%
-%\begin{center}
-%	\includegraphics[width=0.65\textwidth]{3word-questions.png}
-%\end{center}
+	\includegraphics[width=0.4\textwidth]{../how-i-learned.png}
+\end{center}
 }
@@
 
-\subsection{Quicksand}
-\breakslide{\Large \bf \hilight{sect-quicksand}{Bonus Slides - Quicksand}}
+\subsection{Prior Work}
+\breakslide{\Large \bf Bonus Slides - Prior Work}
+
+\begin{frame}{Dynamic Partial Order Reduction (DPOR) {\em [Flanagan '05]}}
+	State space for 2 threads doing {\tt tmp = count; tmp++; count = tmp;} (fixing one thread to run first)
+	\begin{center}
+		\includegraphics[width=\textwidth]{../statespace-tree.pdf}
+	\end{center}
+	\linegap
+
+	Number of ways to interleave $M$ and $N$ events: ${M+N \choose M}$
+\end{frame}
+%\begin{frame}{DPOR example}
+\begin{frame}{Dynamic Partial Order Reduction (DPOR) {\em [Flanagan '05]}}
+	Result of a single iteration of DPOR: alternate branches
+	which would reorder memory conflicts are ``tagged'' for exploration
+	\begin{center}
+		\includegraphics[width=\textwidth]{../dpor-example-0.pdf}
+	\end{center}
+\end{frame}
+\begin{frame}{Dynamic Partial Order Reduction (DPOR) {\em [Flanagan '05]}}
+	Result after 3 iterations; D-J pruned
+	\begin{center}
+		\includegraphics[width=\textwidth]{../dpor-example-1.pdf}
+	\end{center}
+\end{frame}
+\begin{frame}{Dynamic Partial Order Reduction (DPOR) {\em [Flanagan '05]}}
+	Right subtree symmetric; pruning L+M possible with {\em sleep set reduction}
+	\begin{center}
+		\includegraphics[width=0.9\textwidth]{../dpor-example-2.pdf}
+	\end{center}
+\end{frame}
 
@@
 
-\begin{frame}{Quicksand (Iterative Deepening) Reductions}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Quicksand}
+\breakslide{\Large \bf \hilight{sect-quicksand}{Bonus Slides - Quicksand}}
+
+\begin{frame}{Quicksand Reductions}
 	If one job is a subset of another, testing either might let us skip the other.
@@
 \end{frame}
-\begin{frame}{Quicksand (Iterative Deepening) Reductions}
+\begin{frame}{Quicksand Reductions}
 	If one job is a subset of another, testing either might let us skip the other.
@@
 
-\begin{frame}{Quicksand results - Bugs found by CPU time}
+\begin{frame}{Pure HB vs Limited HB - Bugs found by CPU time}
 	\begin{center}
@@
 \end{frame}
-\begin{frame}{Quicksand results - Bugs found by wall-clock time}
+\begin{frame}{Pure HB vs Limited HB - Bugs found by wall-clock time}
 	\begin{center}
@@
 \end{frame}
-\begin{frame}{Quicksand results - Tests fully verified (CPU time)}
+\begin{frame}{Pure HB vs Limited HB - Tests fully verified (CPU time)}
 	\begin{center}
@@
 \end{frame}
-\begin{frame}{Quicksand results - MC + DR vs single-pass DR}
+\begin{frame}{Pure HB vs Limited HB - MC + DR vs single-pass DR}
 	\begin{center}
@@
 
-\begin{frame}{15-410...}
-	% TODO:
-	% study of tej/etc specific bugs
-	% list tests cases provided to studence & PSU-specific ones too
+\begin{frame}{Statistical Significance}
+	Compared %users' vs. non-users'
+	grade distributions with k-sample Anderson-Darling {\em [A., D. '52]}
+	\linegap
+
+	Control 1: non-users from {\bf same semester}
+	\begin{itemize}
+		\item {\bf P2} grade distributions differ with $p = 0.036 < 0.05$
+		\item {\bf P3} grade distributions differ with $p = 0.034 < 0.05$
+			\pause
+		\item Possible bias: stronger students volunteer for Landslide in first place
+	\end{itemize}
+	\pause
+	\linegap
+
+	Control 2: non-users from {\bf past semesters} (when Landslide not offered)
+	\begin{itemize}
+		\item {\bf P2} grade distributions differ with $p = 0.671$...
+		\item {\bf P3} grade distributions differ with $p = 0.145$...
+			\pause
+		\item Possible bias: grading criteria may drift across semesters
+			% FIXME does this bullet belong in this sub list
+		\item Possible bias: maybe Landslide finds bugs the official tests don't
+	\end{itemize}
 \end{frame}
@@
 	\begin{itemize}
-		%\item {\bf [A., D. '52]}: T. Anderson and D. Darling.
-		%	Asymptotic theory of certain ``goodness-of-fit'' criteria based on stochastic processes.
-		%	Annals of Statistics, 1952.
+		\item {\bf [A., D. '52]}: T. Anderson and D. Darling.
+			Asymptotic theory of certain ``goodness-of-fit'' criteria based on stochastic processes.
+			Annals of Statistics, 1952.
 		\item {\bf [Lamport '78]}:
@@
 			VeriSoft. A tool for the automatic analysis of concurrent reactive software. CAV 1997.
+		\item {\bf [Savage '97]}: Stefan Savage.
+			Eraser: a dynamic data race detector for multi-threaded programs. SOSP 1997.
 		%\item {\bf [Holzmann '97]}: The model checker SPIN. TOSE 1997.

### 2018-10-17 thesis/defense/snakefight.tex
last maybe

@@
 		\item Confirm when you fixed bugs, not just find them % not 'verify'
-			% TODO: make alnguage match, "Automatic so and so"
 		\item User-friendly enough for students
 		% \item Test wide variety of types of concurrency % txn
-			% TODO : figure out how to ref 
 		% ????
@@
 		\item State space size estimation {\em [Simsa '12]}
-			% TODO: i just switched the order of these; see how that feels
 		\item Manual kernel annotations {\em [Blum '12]} %{\tiny that's me!}
@@
 	% htm
-	% TODO: this last part you just added for this talk; does the thesis version of the statmeent need to include too?
 	and real-world programs alike.
@@
 		% htm
-		% TODO: this last part you just added for this talk; does the thesis version of the statmeent need to include too?
 		\item \hilight{sect-htm}{...and real-world programs alike.}\xspace
@@
 
-% TODO put a pict of mario-man, i.e maximal state space
-
 \begin{frame}{Total Verification}
-	\textbf{Convergence theorem:} Testing maximal state space, after detecting all data races, $\equiv$ preempting everywhere!
+	Soundness theorem: Testing maximal state space, after detecting all data races, $\equiv$ preempting everywhere!
 	\begin{itemize}
 		\item $\forall$ arbitrary buggy interleaving, $\exists$ equivalent interleaving % say 'under DPOR' i guess?
-			using only sync/data-race PPs, \& Quicksand will eventually reach it.
+			using only sync/data-race preemption points %, \& Quicksand will eventually reach it.
 			%\begin{itemize}
@@
 			%\end{itemize}
-		\item Stand-alone Landslide using ICB {\em [Musuvathi '08]}, preempting at...
-			\begin{itemize}
-				\item ...synchronization APIs only {\em [Simsa '10]}
+		\item Stand-alone Landslide %using ICB {\em [Musuvathi '08]},
+			preempting at...
+			\begin{itemize}
+				\item ...synchronization APIs only {\em [Simsa '13]}
 				\item ...all shared memory accesses
@@
 			\begin{itemize}
-					% TODO: cut this and just say it out loud?
 				\item (e.g., what other thread writes to {\tt ready}?)
@@
 				% is the wrong approach to begin with
-					% TODO: move this to bonus slides, save a min of expl
 				%\item Prior work {\em [Coons '13]} likely to leave threads blocked too long
@@
 
-	% Unlike kernels [Blum '12], P2 can be......
-	% TODO talk abt yield loops, 
-	% compare to prior technique - delay abounding - which sucks bc (verify this by reading BPOR) it doesn't actually
-	% wake up the other thread at the right time (using memory conflix)
-
 	% pintos - bunch of gross patch stuff lmao
 \end{frame}
-
-% do even need this slide?
-%\begin{frame}{Evaluation Questions}
-%	asfda % TODO
-%	% submit better projects?
-%	% get better grades?
-%	% like the experience?
-%\end{frame}
 
@@
 	\end{itemize}
-	% TODO: irb stuff in a bonus slide
 \end{frame}
@@
 		\item CMU: 115 deterministic, 122 concurrency bugs in 75/103 P2s
-			%\begin{itemize}
-			%	\item 3 false positives (self-reported)
-			%\end{itemize}
 		\item U. Chicago: 3 deterministic, 4 concurrency bugs in 3/4 kernels
-			% TODO: what to do abt this bullet pts
-			%\pause
 		\item PSU: (no log collection) %due to IRB issues)
@@
 
-%\begin{frame}{Survey Results} % (happiness value)}
-%	\begin{center}
-%		\includegraphics[width=\textwidth]{../survey5.pdf} % easy to understand?
-%	\end{center}
-%\end{frame}
-
 \begin{frame}{Survey Results} % (educational value)}
 	\begin{center}
+		\includegraphics[width=\textwidth]{../survey8.pdf} % learned to understand concurrency
+	\end{center}
+\end{frame}
+
+\begin{frame}{Survey Results} % (educational value)}
+	\begin{center}
 		\includegraphics[width=\textwidth]{../survey9.pdf} % 100 cpu hours
-	\end{center}
-\end{frame}
-
-\begin{frame}{Survey Results} % (educational value)}
-	\begin{center}
-		\includegraphics[width=\textwidth]{../survey8.pdf} % learned to understand concurrency
 	\end{center}
@@
 	Landslide helps students find \& fix bugs during projects
-	% TODO: should this be moved ot the earlier bug finding slide
 	\begin{itemize}
@@
 
-	stateless model checking
-	\hilight{sect-quicksand}{can sufficiently mitigate exponential explosion}\xspace
-	\hilight{sect-410}{to be practical for inexperienced users}\xspace
-	\hilight{sect-htm}{and real-world programs alike.}\xspace
+	%stateless model checking
+	%\hilight{sect-quicksand}{can sufficiently mitigate exponential explosion}\xspace
+	%\hilight{sect-410}{to be practical for inexperienced users}\xspace
+	%\hilight{sect-htm}{and real-world programs alike.}\xspace
+	\hilight{gray}{stateless model checking}\xspace
+	\hilight{sect-pastel-quicksand}{can sufficiently mitigate exponential explosion}\xspace
+	\hilight{sect-pastel-410}{to be practical for inexperienced users}\xspace
+	\hilight{sect-pastel-htm}{and real-world programs alike.}\xspace
 \end{frame}
@@
 	\end{itemize}
-	% TODO: incremental reveal
+	%\hilight{gray}{and {\bf user-informed heuristic ones},}
+	and {\bf user-informed heuristic ones},
+	\vspace{4.82em}
+	\linegap
+
+	\hilight{gray}{stateless model checking}\xspace
+	\hilight{sect-pastel-quicksand}{can sufficiently mitigate exponential explosion}\xspace
+	\hilight{sect-pastel-410}{to be practical for inexperienced users}\xspace
+	\hilight{sect-pastel-htm}{and real-world programs alike.}\xspace
+\end{frame}
+\begin{frame}{Conclusion}
+	%\hilight{gray}{Combining %both
+	%{\bf theoretically-founded automatic reduction techniques}}
+	Combining %both
+	{\bf theoretically-founded automatic reduction techniques}
+	\begin{itemize}
+		\item \hilight{sect-quicksand}{Verification with data-race preemption points}
+		\item \hilight{sect-410}{Student yield-loop blocking} % student (adj). from "studence"
+		\item \hilight{sect-htm}{Equivalent emulation of HTM semantics}
+	\end{itemize}
 	%\hilight{gray}{and {\bf user-informed heuristic ones},}
@@
 
-\subsection{Prior Work}
-\breakslide{\Large \bf Bonus Slides - Prior Work}
+\subsection{Background}
+\breakslide{\Large \bf Bonus Slides - Background}
 
@@
 \breakslide{\Large \bf \hilight{sect-quicksand}{Bonus Slides - Quicksand}}
+
+\begin{frame}{Verification Soundness}
+	\textbf{Reachability} of preemption points
+	\begin{itemize}
+		\item {\tt mutex\_lock()}, {\tt mutex\_unlock()}, {\tt yield()}, etc, are reachable.
+		\item A data race PP is reachable if it will be identified while preempting on other reachable PPs.
+	\end{itemize}
+	\linegap
+
+	{\bf Convergence}: Quicksand will eventually detect all reachable data races.
+	\linegap
+
+	{\bf Verification}: $\forall$ buggy interleavings, $\exists$ an equivalent interleaving exposing the same bug
+	that switches threads only on reachable PPs.
+	\linegap
+
+	No bugs found testing reachable PPs $\rightarrow$ no buggy interleaving can exist
+\end{frame}
 
@@
 
-\begin{frame}{Statistical Significance}
+\begin{frame}{Grades - Statistical Significance}
 	Compared %users' vs. non-users'
@@
 		\item {\bf P3} grade distributions differ with $p = 0.034 < 0.05$
-			\pause
+			%\pause
 		\item Possible bias: stronger students volunteer for Landslide in first place
 	\end{itemize}
-	\pause
+	%\pause
 	\linegap
@@
 		\item {\bf P3} grade distributions differ with $p = 0.145$...
-			\pause
+			%\pause
 		\item Possible bias: grading criteria may drift across semesters
@@
 \end{frame}
+
+\begin{frame}{Survey Results} \begin{center} \includegraphics[width=\textwidth]{../survey1.pdf} \end{center} \end{frame}
+\begin{frame}{Survey Results} \begin{center} \includegraphics[width=\textwidth]{../survey2.pdf} \end{center} \end{frame}
+\begin{frame}{Survey Results} \begin{center} \includegraphics[width=\textwidth]{../survey3.pdf} \end{center} \end{frame}
+\begin{frame}{Survey Results} \begin{center} \includegraphics[width=\textwidth]{../survey4.pdf} \end{center} \end{frame}
+\begin{frame}{Survey Results} \begin{center} \includegraphics[width=\textwidth]{../survey5.pdf} \end{center} \end{frame}
+\begin{frame}{Survey Results} \begin{center} \includegraphics[width=\textwidth]{../survey7.pdf} \end{center} \end{frame}
 
@@
 %		%\item Wang et al., Eunomia: Scaling Concurrent Search Trees under Contention Using HTM, PPoPP 2017
-%			% TODO: add, that one lock thingy
 %		\item vamsikc, TSX LevelDB implementation, https://github.com/vamsikc/leveldb-tsx/, 2014
@@
 
-\begin{frame}{Retry Reduction}
-	% TODO: split this into a step by step
+\begin{frame}{Retry Set Reduction}
 	\begin{center}
@@
 
-			% TODO: check these again, will 2,4 even complete on mutex
 			{\tt lock}({\tt spinlock})      & 1     & 2     & 4     & 28    \\ % 2,2
@@
 
-%% including this stuff in the talk now so this bonus slide is frivolous
-%% TODO: slide number c.c
-%\begin{frame}{Don't you need \texttt{if (mutex\_is\_locked(\&count\_lock)) \_xabort();} in that transaction on slide 42?}
-%	\begin{center}
-%	Yes.
-%	\end{center}
-%\end{frame}
+\begin{frame}{HTM \& Weak Memory Reorderings}
+	TSO (x86): Stores may be reordered after subsequent loads
+	\linegap
+
+	\begin{center}
+	\begin{tabular}{l}
+		\texttt{if (\hilight{darkorange}{\_xbegin}() == \_XBEGIN\_STARTED) \{} \\
+		\texttt{~~~~if (prevent\_txn)} \\
+		\texttt{~~~~~~~~\_xabort();} \\
+		\texttt{~~~~count++;} \\
+		\texttt{~~~~\hilight{darkblue}{\_xend}();} \\
+		\texttt{\} else \{} \\
+		\texttt{~~~~\hilight{darkorange}{mutex\_lock}(\&m);} \\
+		\texttt{~~~~prevent\_txn = 1;} \\
+		\texttt{~~~~\hilight{red}{// XXX barrier needed here}} \\
+		%\texttt{~~~~// \_\_sync\_synchronize(); // uncomment to solve the bug} \\
+		\texttt{~~~~count++;} \\
+		\texttt{~~~~prevent\_txn = 0;} \\
+		\texttt{~~~~\hilight{darkblue}{mutex\_unlock}(\&m);} \\
+		\texttt{\}} \\
+	\end{tabular}
+	\end{center}
+\end{frame}
 
@@
 		%\item {\bf [Holzmann '97]}: The model checker SPIN. TOSE 1997.
-		\item {\bf [Engler '03]}: Dawson Engler and Ken Ashcraft.
+		%\item {\bf [Engler '03]}: Dawson Engler and Ken Ashcraft.
 			RacerX: effective, static detection of race conditions and deadlocks. SOSP 2003.
@@
 		\item {\bf [Yang '08]}: Yu Yang et al. Efficient stateful DPOR. SPIN 2008.
-		\item {\bf [Klein '09]}: Gerwin Klein et al. seL4: formal verif. of an OS kernel. SOSP 2009.
 	\end{itemize}
@@
 	\begin{itemize}
+		\item {\bf [Klein '09]}: Gerwin Klein et al. seL4: formal verif. of an OS kernel. SOSP 2009.
 		%\item {\bf [Burckhardt '10]}: A randomized scheduler with probabilistic guarantees of finding bugs. ASPLOS 2010.
@@
 			MS thesis (CMU-CS-12-118). May 2012.
-		\item {\bf [Kasicki '12]}: Baris Kasicki. Data races vs. data race bugs: telling the difference with Portend. ASPLOS '12.
+		%\item {\bf [Kasicki '12]}: Baris Kasicki. Data races vs. data race bugs: telling the difference with Portend. ASPLOS '12.
 		\item {\bf [Simsa '13]}: Systematic and Scalable Testing of Concurrent Programs.

### 2018-10-17 thesis/defense/snakefight.tex
revisions aftre 3am practice to self ;_;

@@
 \begin{frame}{Evaluation}
-	79 ``P2'' thread libraries from CMU, 78 ``Pintos'' kernels from U. Chicago \& Stanford
-	\linegap
-
-	%629 pairs of \{P2 or Pintos\} + test case
-	629 pairs of P2 or Pintos + test case
-	\linegap
+	629 tests of submitted ``P2'' thread libraries (CMU) and ``Pintos'' kernels (U. Chicago \& Stanford)
+	%79 ``P2'' thread libraries from CMU, 78 ``Pintos'' kernels from U. Chicago \& Stanford
+	\linegap
+
+	%%629 pairs of \{P2 or Pintos\} + test case
+	%629 pairs of P2 or Pintos + test case
+	%\linegap
 
@@
 	\end{itemize}
-	\pause
-	\linegap
-
+\end{frame}
+
+\begin{frame}{Survey Results} % (educational value)}
+	\begin{center}
+		\includegraphics[width=\textwidth]{../survey8.pdf} % learned to understand concurrency
+	\end{center}
+\end{frame}
+
+\begin{frame}{Survey Results} % (educational value)}
+	\begin{center}
+		\includegraphics[width=\textwidth]{../survey9.pdf} % 100 cpu hours
+	\end{center}
+\end{frame}
+
+\begin{frame}{Survey Results} % (happiness value)}
+	\begin{center}
+		\includegraphics[width=\textwidth]{../survey10.pdf} % recommend to friend
+	\end{center}
+\end{frame}
+
+\begin{frame}{Survey Results}
 	Self-report any {\bf false positives}
@@
 	\end{itemize}
+	\pause
 	\linegap
@@
 	\end{itemize}
-\end{frame}
-
-\begin{frame}{Survey Results} % (educational value)}
-	\begin{center}
-		\includegraphics[width=\textwidth]{../survey6.pdf} % easier to diagnose?
-	\end{center}
-\end{frame}
-
-\begin{frame}{Survey Results} % (educational value)}
-	\begin{center}
-		\includegraphics[width=\textwidth]{../survey8.pdf} % learned to understand concurrency
-	\end{center}
-\end{frame}
-
-\begin{frame}{Survey Results} % (educational value)}
-	\begin{center}
-		\includegraphics[width=\textwidth]{../survey9.pdf} % 100 cpu hours
-	\end{center}
-\end{frame}
-
-\begin{frame}{Survey Results} % (happiness value)}
-	\begin{center}
-		\includegraphics[width=\textwidth]{../survey10.pdf} % recommend to friend
-	\end{center}
 \end{frame}
@@
 
-\breakslide{\Large \bf \hilight{sect-htm}{Transactional Memory}}
+\breakslide{\Large \bf \hilight{sect-htm}{Hardware Transactional Memory}}
 
@@
 				\item Memory conflict with another thread
-				\item False sharing (conflict on same cache-line)
-				\item Cache eviction, random interrupt, ...
+				%\item False sharing (conflict on same cache-line)
+				\item Cache conflicts, cache eviction, random interrupt, ...
 			\end{itemize}
@@
 \begin{frame}{Concurrency model}
-	What are a transactional program's {\em observable} behaviours?
-	\begin{itemize}
-		\item Transactional writes, if successful, appear atomically %to other threads
+	That first {\tt count++} never happened...!
+	\linegap
+
+	%What are a transactional program's {\em observable} behaviours?
 	%\begin{itemize}
-		%	\item Equivalent to a lock protecting that state
-		%		% note that also inserting DR PPs within a transaction is not allowed
+	%	\item Transactional writes, if successful, appear atomically %to other threads
+	%	%\begin{itemize}
+	%	%	\item Equivalent to a lock protecting that state
+	%	%		% note that also inserting DR PPs within a transaction is not allowed
+	%	%\end{itemize}
+	%	%	\pause
+	%	\item Simultaneous transactions succeed only if no conflicts
+	%	%\begin{itemize}
+	%	%	\item Equivalent
+	%	%		%(modulo performance)
+	%	%		to serialize independent transactions
+	%	%\end{itemize}
+	%	%	\pause
+	%	\item Failing transactions' effects rolled-back with no trace
+	%	%\begin{itemize}
+	%	%	\item Treat each \xbegin as a failure injection point
+	%	%\end{itemize}
 	%\end{itemize}
-		%	\pause
-		\item Simultaneous transactions succeed only if no conflicts
-		%\begin{itemize}
-		%	\item Equivalent
-		%		%(modulo performance)
-		%		to serialize independent transactions
-		%\end{itemize}
-		%	\pause
-		\item Failing transactions' effects rolled-back with no trace
-		%\begin{itemize}
-		%	\item Treat each \xbegin as a failure injection point
-		%\end{itemize}
-	\end{itemize}
-	\pause
-	\linegap
+	%\pause
+	%\linegap
 
@@
 
-	Even simple programs tricky to get right
-	\begin{itemize}
-		\item Author of {\tt mutex\_is\_locked()} pattern got it wrong once...
-		\item Landslide can check even unpredictable performance characteristics
-	\end{itemize}
-	\linegap
-
-	Abstraction reduction eases both implementation \& verification burden
+	Landslide found surprising bugs in AVL tree, spinlock
+	\linegap
+
+	%Even simple programs tricky to get right
+	%\begin{itemize}
+	%	\item Author of {\tt mutex\_is\_locked()} pattern got it wrong once...
+	%	\item Landslide can check even unpredictable performance characteristics
+	%\end{itemize}
+	%\linegap
+
+	Abstraction reduction eases %both implementation \&
+	verification burden
 \end{frame}
@@
 \begin{frame}{Survey Results} \begin{center} \includegraphics[width=\textwidth]{../survey5.pdf} \end{center} \end{frame}
+\begin{frame}{Survey Results} \begin{center} \includegraphics[width=\textwidth]{../survey6.pdf} \end{center} \end{frame}
 \begin{frame}{Survey Results} \begin{center} \includegraphics[width=\textwidth]{../survey7.pdf} \end{center} \end{frame}

### 2018-10-17 thesis/defense/snakefight.tex
snakefight last draft

@@
 				\item Implement {\tt thr\_create()}/{\tt exit()}/{\tt join()}, {\tt mutex\_lock()}, etc.
-				\item Reference ``Pebbles'' kernel provided
+				\item Reference ``Pebbles'' kernel provided, later reimplemented
 					% say: later they implement this in p3
@@
 			\end{itemize}
-		\item Kernel (``P3'') - reimplement said reference kernel (CMU only)
+		%\item Kernel (``P3'') - reimplement said reference kernel (CMU only)
 	\end{itemize}
@@
 	\begin{itemize}
-		\item CMU: 115 deterministic, 122 concurrency bugs in 75/103 P2s
-		\item U. Chicago: 3 deterministic, 4 concurrency bugs in 3/4 kernels
+		\item CMU: %115 deterministic,
+			122 concurrency bugs in 75/103 P2s
+		\item U. Chicago: %3 deterministic,
+			4 concurrency bugs in 3/4 kernels
 		\item PSU: (no log collection) %due to IRB issues)
@@
 
-	%...as well as subsequent P3 (kernel) grades
-	...as well as subsequent kernel grades
-	\begin{itemize}
-		\item ``Does Landslide teach students lasting lessons about concurrency?''
-	\end{itemize}
+	%%...as well as subsequent P3 (kernel) grades
+	%...as well as subsequent kernel grades
+	%\begin{itemize}
+	%	\item ``Does Landslide teach students lasting lessons about concurrency?''
+	%\end{itemize}
 \end{frame}
@@
 	\begin{center}
-		\includegraphics[width=\textwidth]{../photo-of-ze-studence.pdf}
+		\includegraphics[width=\textwidth]{photo-of-ze-studence-box-p2-only.pdf}
 	\end{center}
@@
 	\begin{center}
-		\includegraphics[width=\textwidth]{photo-of-ze-studence-cdf-no-pre-study.pdf}
+		\includegraphics[width=\textwidth]{photo-of-ze-studence-cdf-p2-only-no-pre-study.pdf}
 	\end{center}
@@
 	\begin{center}
-		\includegraphics[width=\textwidth]{photo-of-ze-studence-cdf.pdf}
+		\includegraphics[width=\textwidth]{photo-of-ze-studence-cdf-p2-only.pdf}
 	\end{center}
@@
 	\begin{itemize}
-		\item \hilight{sect-quicksand}{Quicksand's CPU budget argument}
+		\item \hilight{sect-quicksand}{Prioritizing jobs based on fixed CPU budget}
 		\item \hilight{sect-410}{Exposing data-race reports to students}
@@
 \breakslide{\Large \bf \hilight{sect-410}{Bonus Slides - Education}}
+
+\begin{frame}{Effect on Grades (CMU)}
+	\begin{center}
+		\includegraphics[width=\textwidth]{../photo-of-ze-studence.pdf}
+	\end{center}
+\end{frame}
+\begin{frame}{Effect on Grades (CMU)} % - statistically significant?}
+	\begin{center}
+		\includegraphics[width=\textwidth]{photo-of-ze-studence-cdf-no-pre-study.pdf}
+	\end{center}
+\end{frame}
+\begin{frame}{Effect on Grades (CMU)} % - statistically significant?}
+	\begin{center}
+		\includegraphics[width=\textwidth]{photo-of-ze-studence-cdf.pdf}
+	\end{center}
+\end{frame}
 

### 2018-10-26 thesis/thesis.tex
turns out oathbringer had a good quote after all
snakefight and parent acks

--- 2018-10-08 thesis/thesis.tex
@@
 that make up about half of Chapter~\ref{chap:tm}'s evaluation suite.
+%
+Carlo Angiuli, Evan Cavallo, Ziv Scully, Jim McCann,
+Matthew Maurer, Stefan Muller, Guillaume Didier,
+Sol Boucher, and Michael J. Sullivan
+helped me rehearse and polish the snake fight talk.
 
@@
 \subsubsection{Life}
+
+My parents, Eve and David Blum, and my grandparents, Margie Granach and Elsie Blum,
+provided bottomless emotional support throughout my education.
+I am grateful beyond words for the privileged life they provided
+that allowed me the opportunity to pursue higher education.
 
@@
 \label{chap:conclusion}
+%\inspirationalquote{
+%\begin{tabular}{p{0.6\textwidth}}
+%"This is the end, right? The end of this story?"
+%
+%"We were always at the end. It's a free play, buddy. Clock's all zeroes."
+%\end{tabular}}
+%%{Nine \& Ten, 17776}
+%{17776}
 \inspirationalquote{

### 2018-10-26 thesis/education.tex
future - generalize from k,n; combine test case and verif sects
some more citations

@@
 % discrete == I used "version 2" of the AD metric that R outputs.
-I use the $k$-sample discrete Anderson-Darling test \cite{anderson-darling}
+I use the $k$-sample discrete Anderson-Darling test \cite{anderson1952,anderson-darling}
 provided by the R statistical programming environment \cite{r-lang,r-ksamples},
@@
 \subsubsection{Common bugs}
+\label{sec:education-eval-bug-case-study}
 
@@
 \section{Discussion}
+\label{sec:education-discussion}
 

### 2018-10-26 thesis/related.tex
some more citations

@@
 %
+\revision{CSPEC \cite{cspec} is another recent system for formally specifying and verifying concurrent systems using Coq,
+which is then extracted into an executable version in Haskell.
+Its type system incorporates the notion of reordering independence,
+much akin to DPOR,
+although the complexity overhead is very high:
+a locked counter on TSO weak memory requires 10 abstraction layers
+to formally specify,
+requiring among other things reasoning about the lock's previous owner as well as the current one.}
+%
 Heroic as such end-to-end formal verification projects are,
@@
 if it means helping more users overall.
-Future work could check preemptive and/or multiprocessor kernels
-%by combining Hyperkernel's approach,
-by first checking safety properties in the absence of concurrency,
-then checking with Landslide that concurrency introduces no new program behaviour not already verified,
-to provide a less formal, but still hopefully useful, verification guarantee.
+\revision{
+Stateless model checking offers a middle ground for users to achieve
+slightly less formal,
+but still theoretically-founded,
+verification guarantees
+even for unsafe programming languages.}
+%Future work could check preemptive and/or multiprocessor kernels
+%%by combining Hyperkernel's approach,
+%by first checking safety properties in the absence of concurrency,
+%then checking with Landslide that concurrency introduces no new program behaviour not already verified,
+%to provide a less formal, but still hopefully useful, verification guarantee.
 %

### 2018-10-27 thesis/education.tex
final pre-defense draft revisions

@@
 Table~\ref{tab:photo-of-ze-studence} shows the student participation rate across semesters
-% TODO update for psu
 for the thread library study%
@@
 			\hline
+			% group-[redacted] had a p2/ dir, but not in grade files
 			CMU S'15	& &   8	& &  18	& &  26 \\
@@
 			U. Chicago F'17	& &   4	& &  17	& &  21 \\ % 21 taken from number of submitted 'threads', idk abt userprog
-			% TODO when you adjust this, adjust the % of participance down in the discussion para
 			PSU S'18	& &  38	& &  98	& & 136 \\

### 2018-10-27 thesis/intro.tex
final pre-defense draft revisions

@@
 \begin{center}
-	{\em Combining both theoretically-founded automatic reduction techniques
+	{\em Combining %both
+	theoretically-founded automatic reduction techniques
 	and user-informed heuristic ones,
 	stateless model checking can sufficiently mitigate exponential explosion
-	to be a practical testing technique even for inexperienced users.}
+	to be a practical testing technique for inexperienced users and real-world programs alike.}
 \end{center}

### 2018-10-27 thesis/landslide.tex
final pre-defense draft revisions

@@
 by keeping a counter for each thread to track how many {\tt yield}s it has invoked since the last ``interesting'' activity.
+% nb. not system calls, not even d/mr, as some student mutexes may use those and you don't want to count them
 Interesting here is defined heuristically as any other known P2 API call,
@@
 
-Landslide will retry this process up to a heuristic limit of 128 times ({\tt DEADLOCK\_FP\_MAX\_\allowbreak()ATTEMPTS}
+Landslide will retry this process up to a heuristic limit of 128 times ({\tt DEADLOCK\_FP\_MAX\_\allowbreak{}ATTEMPTS}
 before issuing a deadlock bug report.
-The overhead of this check is linear time in the number of retries permitted,
-although this is typically negligible compared to the exponential size of state spaces overall.
+The overhead of this check is quadratic time in the number of retries permitted
+(as DPOR is quadratic in the overall branch depth),
+although this is negligible compared to the exponential size of state spaces overall.
 If the heuristic limit is too small, very ``loopy'' programs

### 2018-11-03 thesis/intro.tex
dave revisions intro

--- 2018-10-06 thesis/intro.tex
@@
 \subsubsection{Motivation}
+
+% "Concurrency is notoriously difficult to get right." [....]
+% dejafu A Concurrency Testing Library for Haskell (2016 ver.)
+% parrot?
+% TODO more
 
@@
 Accordingly, a programmer who wants her test to make reasonable progress through the state space must choose a subset of ways that her threads could interleave,
-focusing on fully testing that subset, while ignoring other possibilities she doesn't care about.
+focusing on fully testing that subset, while ignoring other possibilities she doesn't \revisionminor{think she cares} about.
 However, it is difficult to choose a subset of thread interleavings that will produce a meaningful, yet feasible test.
@@
 For example, operating system kernels implement their own sources of concurrency and their own synchronization primitives,
-so the checker needs to be told how to identify and control the execution of each thread.
+so \revisionminor{a checker must} be told how to identify and control the execution of each thread.
 Some expert concurrency research wizards may be willing to add manual annotations to their code,
@@
 
-\subsubsection{Contribution}
+\subsubsection{Contributions}
 
@@
 
-\vspace{1em}
-
 \begin{center}
-	{\em Combining both theoretically-founded automatic reduction techniques
+\begin{tabular}{p{0.8\textwidth}}
+	\revisionminor{{\em Combining %both
+	theoretically-founded automatic reduction techniques
 	and user-informed heuristic ones,
 	stateless model checking can sufficiently mitigate exponential explosion
-	to be a practical testing technique even for inexperienced users.}
+	to be a practical testing technique for inexperienced users and real-world programs alike.}}
+\end{tabular}
 \end{center}
 
-\vspace{1em}
+\vspace{0.75em}
 
@@
 		Iterative Deepening incorporates state space estimation \cite{estimation} to decide on-the-fly whether each state space is worth pursuing, and uses data race analysis \cite{tsan} to find new preemption point candidates based on a program's dynamic behaviour.
-		This section will include a large evaluation of the technique, comparing its performance to three prior work approaches across 600+ unique tests.
-		I will show that Iterative Deepening of preemption points outperforms prior work in terms both of finding bugs quickly and of completely verifying correctness when no bug exists.
+		This section will include a large evaluation of the technique, comparing its performance to three \revisionminor{prior-work} approaches across 600+ unique tests.
+		I will show that Iterative Deepening of preemption points outperforms prior work in terms \revisionminor{of both} finding bugs quickly and completely verifying correctness when no bug exists.
 	\item {\bf Educational use (Chapter \ref{chap:education}).}
@@
 		At all three universities I then collected statistics on the numbers and types of bugs found,
-		and surveyed students to understand the human experience,
+		and surveyed students to understand the human experience.
 		This section will present the study's results

### 2018-11-03 thesis/education.tex
add retrospective survey question did you fully verify

@@
 		might have made Landslide less useful overall.
+	\item \revision{{\em In addition to finding bugs, did you manage to fully verify your code under
+		any of the tests by letting Landslide complete all state spaces before reaching the specified time out?}
+		This would provide a sense of how thoroughly students understood the underlying research technique,
+		and serve as a follow-up to the ``how many bug fixes verified'' question
+		(where students often stopped Landslide midway through after a little while).}
 \end{enumerate}

### 2018-11-13 thesis/overview.tex
relocate bgnd data race figure
rearrange quotes and add one from 17776
turns out oathbringer had a good quote after all
fix colors / generally modernize fig 2.5

@@
 
-\begin{flushright}
-	{
-	\fontspec[Path=./fonts/]{AlexaStd}
-	\begin{tabular}{p{0.49\textwidth}}
-	{\em "Well I don't *feel* brave. I feel scared, and insignificant."}
-	\\
-	{\em "That's what heroes always say, at first."}
-	\end{tabular}
-	}
-	\vspace{2em}
-\end{flushright}
+%% something i wrote myself??
+%\begin{flushright}
+%	{
+%	\fontspec[Path=./fonts/]{AlexaStd}
+%	\begin{tabular}{p{0.49\textwidth}}
+%	{\em "Well I don't *feel* brave. I feel scared, and insignificant."}
+%	\\
+%	{\em "That's what heroes always say, at first."}
+%	\end{tabular}
+%	}
+%	\vspace{2em}
+%\end{flushright}
+\inspirationalquote{
+\begin{tabular}{p{0.75\textwidth}}
+We can't be free until we learn to laugh at ourselves.
+Once you look in the mirror and see just how foolish we can be,
+laughter is inevitable.
+And from laughter, comes wisdom.
+\end{tabular}}
+{G'Kar, Babylon 5}
 
@@
 	\begin{center}
-	\includegraphics[width=0.4\textwidth]{dpor.pdf}
+	\includegraphics[width=0.6\textwidth]{dpor-new-colors.pdf}
 	\end{center}
-	\caption{DPOR identifies independent transitions by different threads which can commute without affecting program behaviour.
-Here, if the transitions marked 1 and 2 have no shared memory conflicts, the states marked with the red arrow are guaranteed identical.
-Hence, only one of the subtrees need be explored.}
+	\caption{DPOR identifies independent transitions by different threads which can commute
+	without affecting program behaviour.
+	\revisionminor{Here, because each thread's local operations on {\tt tmp1}/{\tt tmp2} are not shared memory conflicts,}
+	the states marked with the red arrow are guaranteed identical. Hence, only one of the subtrees need be explored.}
 	\label{fig:dpor}
@@
 
-\begin{figure}[t]
+Data race analysis \cite{eraser} identifies pairs of unsynchronized memory accesses between threads.
+
+\subsection{Definition}
+
+Two instructions are said to race if:
+\begin{enumerate}
+	\item they both access the same memory address,
+	\item at least one is a write,
+	\item the threads do not hold the same lock,
+	\item and no synchronization enforces an order on the thread transitions (the {\em Happens-Before} relation, described below).
+\end{enumerate}
+%In Figure~\ref{fig:example}, lines 3 and 5 each race with 2 and 6, and line 6 races with 8.
+In Figure~\ref{fig:hb-example}, the pairs of lines marked with comments (A1 and A2, B1 and B2) race.
+
+\begin{figure}[h]
         \small
@@
 \end{figure}
-
-\subsection{Definition}
-
-Data race analysis \cite{eraser} identifies pairs of unsynchronized memory accesses between threads.
-Two instructions are said to race if:
-\begin{enumerate}
-	\item they both access the same memory address,
-	\item at least one is a write,
-	\item the threads do not hold the same lock,
-	\item and no synchronization enforces an order on the thread transitions (the {\em Happens-Before} relation, described below).
-\end{enumerate}
-%In Figure~\ref{fig:example}, lines 3 and 5 each race with 2 and 6, and line 6 races with 8.
-In Figure~\ref{fig:hb-example}, the pairs of lines marked with comments (A1 and A2, B1 and B2) race.
 

### 2018-11-15 thesis/education.tex
dave's revisions to background chapter

@@
 		and captures a snapshot of the student code and results of the test and saves them to AFS
-		(CMU's network file system)
+		\revisionminor{(CMU's network file system)~\cite{afs}}
 		after each run.

### 2018-11-15 thesis/overview.tex
dave's revisions to background chapter

@@
 so we further restrict {\em concurrency bug} to denote a program behaviour which is mechanically identifiable,
-according to commonly-accepted notions of what programs behaviours are always bad.
+according to commonly-accepted notions of \revisionminor{which} programs behaviours are always bad.
 %
@@
 			&
-			(b) Example interleaving of the compiled assembly for (a),
-			in which 2 concurrent iterations of the loop yield 1 net increment of {\tt x}.
+			% revisionminor can't go here i guess
+			(b) Example interleaving of (a)'s compiled assembly,
+			in which two concurrent iterations of the loop yield one net increment of {\tt x}.
 		\end{tabular}
@@
 			\texttt{~~~~\ctype{volatile int} held;} \\
-			\texttt{~~~~\ctype{int} owner;} \\
+			%\texttt{~~~~\ctype{int} owner;} \\
 			\texttt{\} \ctype{mutex\_t};} \\
@@
 			\texttt{~~~~\flow{while} (\call{xchg}(mp->held, \const{1}))} \\
-			\texttt{~~~~~~~~\call{yield}(mp->owner);} \\
-			\texttt{~~~~mp->owner = \call{gettid}();} \\
+			%\texttt{~~~~~~~~\call{yield}(mp->owner);} \\
+			\texttt{~~~~~~~~\call{yield}(-1);} \\
+			%\texttt{~~~~mp->owner = \call{gettid}();} \\
 			\texttt{\}} \\
 			\texttt{\ctype{void} \call{mutex\_unlock}(\ctype{mutex\_t} *mp) \{}\\
-			\texttt{~~~~mp->owner = \ctype{-1};} \\
+			%\texttt{~~~~mp->owner = \ctype{-1};} \\
 			\texttt{~~~~mp->held = \ctype{0};} \\
@@
 
-Each such abstraction provides certain semantics about what thread interleavings can arise surrounding their use.
+Each such abstraction provides certain semantics about \revisionminor{which} thread interleavings can arise surrounding their use.
 When building a tool for testing concurrent programs,
@@
 Annotating a certain abstraction's semantics treats it as a trusted concurrency primitive in its own right,
-and allows the testing tool to reduce the possible space of interleavings (or the set of false positive data-race candidates reported, etc.),
-at the cost of increasing the implementation and theoretical complexity of the analysis.
+and allows the testing tool to reduce the possible space of interleavings
+(or the set of false positive data-race candidates reported, etc.;
+\revisionminor{see \sect{\ref{sec:landslide-config-landslide}}})
+at the cost of increasing the implementation and theoretical complexity of the analysis
+\revisionminor{(\sect{\ref{sec:tm-abstraction}})}.
 This thesis will consider locks, descheduling, and transaction begin/end as the only concurrency primitives,
@@
 Locks (or {\em mutexes}, short for ``mutual exclusion locks'') are objects, shared by multiple threads, which allow the programmer to mark certain {\em critical sections} of code that must not interleave with each other.
-When one thread completes a call to {\tt mutex\_lock(mp)}, all invocations by other threads on the same {\tt mp} will wait (or ``block'') until the corresponding {\tt mutex\_unlock(mp)}.
+When one thread completes a call to {\tt mutex\_lock(mp)}, all invocations by other threads on the same {\tt mp}
+will \revisionminor{block (i.e., wait)} until the corresponding {\tt mutex\_unlock(mp)}.
 Figure~\ref{fig:mutex}(a) shows how a yielding mutex (not the best implementation, but the simplest) may be implemented using {\tt xchg},
@@
 
-Transactional memory (TM) may be implemented either in hardware, using special instructions and existing cache coherence algorithms,
+Transactional memory (TM) may be implemented either in hardware,
+using special instructions and existing cache logic, %coherence algorithms,
 or in software, via library calls and a log-based commit approach.
@@
 
-Model checking \cite{verisoft} is a testing technique for systematically exploring the possible thread interleavings of a concurrent program.
+\revisionminor{Model checking \cite{partial-model-checking,verisoft}} is a testing technique for systematically exploring the possible thread interleavings of a concurrent program.
 A model checker executes the program repeatedly, each time according to a new thread interleaving, until all interleavings have been tested or the CPU budget is exhausted.
@@
 	\caption{Visualization of interleaving state space for the program in Figure~\ref{fig:concurrency-bug}.
-	Thread 1 is represented by purple ovals, thread 2 by yellow squares, and time flows from top to bottom.
+	Thread~1 is represented by purple ovals, thread~2 by yellow squares, and time flows from top to bottom.
 	As the two threads execute the same code, without loss of generality thread 1 is fixed to run first --
@@
 so each bug report is accompanied by concrete evidence of the violation.
-Assertions, segfaults, use-after-free of heap memory, and deadlock are examples of such failures we check for,
+Assertions, \revisionminor{access violations}, use-after-free of heap memory,
+and deadlock are examples of \revisionminor{commonly-checked failures},
+%such failures we check for,
 although a checker may also include arbitrary program-specific predicates.
@@
 Nevertheless, to fully test all of a program's possible behaviours, we must check the executions corresponding to every branch of the tree.
-Addressing the scaling problem in this exponential relation is the central research problem for all model checkers.
+Addressing the scaling problem in this exponential relation is the central research problem for all model checkers,
+\revisionminor{which this thesis will address in Chapter~\ref{chap:quicksand}}.
 
@@
 are represented as a directed acyclic graph (DAG) instead of as a tree.
-This approach is called {\em stateful} model checking.
+This approach is called \revisionminor{{\em stateful} model checking \cite{spin}}.
 This thesis focuses on {\em stateless} model checking (and execution trees, not DAGs),
@@
 Techniques for coping with the exponential explosion fall into two categories:
-(1) removing redundant interleavings from the state space when we can prove they are equivalent to some interleaving already tested,
-or {\bf reduction techniques},
+%(1)
+{\em reduction techniques},
+i.e.,
+removing redundant interleavings from the state space when we can prove they are equivalent to some interleaving already tested,
 and
-(2) prioritizing interleavings judged as more likely to contain bugs should bugs exist
-in case we are unable to exhaustively test all interleavings after all,
-or {\bf search heuristics}.
+%(2)
+{\em search heuristics},
+i.e.,
+prioritizing interleavings judged as more likely to contain bugs (should bugs exist)
+in case we are unable to exhaustively test all interleavings after all.
 
@@
 	without affecting program behaviour.
-	\revisionminor{Here, because each thread's local operations on {\tt tmp1}/{\tt tmp2} are not shared memory conflicts,}
+	Here, because each thread's local operations on {\tt tmp1}/{\tt tmp2} are not shared memory conflicts,
 	the states marked with the red arrow are guaranteed identical. Hence, only one of the subtrees need be explored.}
@@
 	\item the threads do not hold the same lock,
-	\item and no synchronization enforces an order on the thread transitions (the {\em Happens-Before} relation, described below).
+	\item and no synchronization enforces an order on the thread transitions
+		(the {\em Happens-Before} \revisionminor{relation~\cite{lamport-clocks,hybriddatarace}}, described below).
 \end{enumerate}
@@
 The Pebbles kernel architecture
-is used at Carnegie Mellon University (CMU) in 15-410, Operating System Design and Education \cite{kspec,thrlib}.
+is used at Carnegie Mellon University (CMU) in 15-410, Operating System Design and \revisionminor{Implementation} \cite{kspec,thrlib}.
 In the course of a semester, students work on five programming assignments;
@@
 The course's prerequisite is 15-213, Introduction to Computer Systems \cite{sigcse01:CSaPP}.
-Both P2 and P3 are built using the {\em Pebbles} system call specification, outlined in Table~\ref{tab:syscalls}
+Both P2 and P3 are built using the {\em Pebbles} system call specification, outlined in Table~\ref{tab:syscalls}.
 
@@
 The thread library project \cite{thrlib} has two main components: implementing concurrency primitives, and implementing thread lifecycle and management routines.
-The required concurrency primitives are as follows:
+\revision{Students implement this project on top of a reference Pebbles kernel implementation (named Pathos~\cite{pathos}),
+which is distributed as an obfuscated binary to protect the subsequent challenges of P3.}
+The required concurrency primitives \revisionminor{for P2} are as follows:
 \begin{itemize}
@@
 		The primary challenge of this exercise is ensuring the aforementioned atomicity between {\tt cond\_wait}'s unlock and deschedule, with respect to the rest of the interface.
-	\item Semaphores, with the interface {\tt sem\_wait(sp)} and {\tt sem\_signal(sp)} (sometimes called {\em proberen} and {\em verhogen} in other literature).
+	\item Semaphores, with the interface {\tt sem\_wait(sp)} and {\tt sem\_signal(sp)}
+		(sometimes called \revisionminor{$P$ and $V$ in other literature,
+		from Dutch,
+		originally {\em passering} and {\em vrijgave} \cite{ewd35},
+		later often explained as {\em proberen} and {\em verhogen}}).
 		The semaphore can be initialized to any integer value; if initialized to 1, it behaves like a mutex.
@@
 \begin{itemize}
-	\item {\tt thr\_init(stack\_size)} initializes the thread library, setting a default stack size to be allocated to new threads.
+	\item {\tt thr\_init(stack\_size)} initializes the thread library, setting a \revisionminor{custom, fixed} stack size to be allocated to new threads.
 	\item {\tt thr\_create(child\_func, child\_arg)} spawns a new thread to run the specified function with the specified argument.
@@
 		Returns an integer thread ID of the newly created thread.
-	\item {\tt thr\_exit(status)} aborts execution of the calling thread, recording an exit status value.
+	\item {\tt thr\_exit(status)} \revisionminor{completes} execution of the calling thread, recording an exit status value.
+		\revisionminor{This function does not return to its callsite.}
 		The main challenge of this function is to allow another thread to free the memory used for the exiting thread's stack,
@@
 Pebbles adopts the Mach \cite{DBLP:conf/usenix/AccettaBBGRTY86} distinction between {\em tasks}, which are resource containers, and {\em threads}, each of which executes within a single task.
-This requires less implementation complexity than the more featureful Plan 9's {\tt rfork} \cite{Pike90plan9} or Linux's {\tt clone} models.
-
-Although the internal interfaces are not mandated like they were in P2, all Pebbles kernels must necessarily contain the same abstract components. These include:
+This requires less implementation complexity than the more featureful
+\revisionminor{{\tt rfork} model, of Plan 9 \cite{Pike90plan9}, or the {\tt clone} model, of Linux}.
+
+Although the interfaces are not mandated \revisionminor{as} they were in P2, all Pebbles kernels must necessarily contain the same abstract components. These include:
 \begin{itemize}
@@
 	\item Some approach to locking, often analogous to P2's concurrency primitives (henceforth referred to as ``kernel mutexes''),
-	 ll       and some approach to blocking threads indefinitely;
-	\item A virtual memory implementation, including a program loader;
-	\item Lifecycle management code for creation and destruction of kernel threads and processes;
+		and some approach to blocking threads indefinitely;
+	\item A virtual memory implementation \revisionminor{(page directories and tables, frame allocator, et cetera)};
+	\item A program loader;
+	\item Lifecycle management code for creation and destruction of \revisionminor{threads and tasks};
 	\item Other miscellany such as a suite of fault handlers to ensure no user program can cause the kernel itself to crash.
@@
 which targets a subtle condition-variable bug.
-The test uses the {\tt misbehave} system call to target a particular thread interleaving likely to expose the bug
+The test uses the {\tt misbehave} system call to \revisionminor{amplify the probability of} a particular thread interleaving,
+likely to expose the bug,
 which is otherwise very unlikely to arise in normal execution.
@@
 The Pintos kernel architecture \cite{pintos} is used at several universities, including Berkeley, Stanford, and the University of Chicago.
-The Pintos basecode implements a rudimentary kernel, consisting of a context switcher, round-robin scheduler, locking primitives, and program loader.
+The Pintos basecode implements a rudimentary kernel, consisting of a context switcher, round-robin scheduler, locking primitives, and program loader,
 upon which students add more features in several projects.
@@
 Some schools further offer a virtual memory project, extending the provided VM with a frame table and supplemental page table and fault handler \cite{standford-cs140,uchicago-cs230}, or supplemental HTTP server and {\tt malloc} assignments \cite{berkeley-cs162}.
-Being largely architectural/algorithmic projects rather than concurrency-oriented ones, this thesis is not concerned with these assignments.
+\revisionminor{This thesis is not concerned with these assignments.
+as they are are largely architectural/algorithmic projects rather than concurrency-oriented ones.}
 The main concurrency challenges in Pintos projects arise from the {\em threads} and {\em userprog} assignments:
@@
 	\item {\bf API.}
+		\revisionminor{Short for ``application programming interface'' (which nobody ever says).}
 		The set of function signatures which comprise the interface to a body of code.
@@
 		and optionally {\tt mutex\_trylock()} and {\tt mutex\_destroy()}.
-		From ``application programming interface'' (which nobody ever says).
 	\item {\bf Atomic.}
@@
 		so will simply avoid the term entirely.
+	\item {\bf Benign.}
+		Said of a {\em data race}, or of {\em nondeterministic} behaviour in general,
+		which does not lead to a {\em concurrency bug} in any {\em interleaving}.
 	\item {\bf Branch.}
@@
 		comprising all possible interleavings which may arise with that as a prefix.
+	\item \revisionminor{{\bf Bug.}
+		See {\em concurrency bug}.}
 	\item {\bf Completeness.}
@@
 		An unsynchronized pair of memory accesses between two threads that can, may, or did execute concurrently.
-		{\em Race} for short.
+		{\em Race} for short \revisionminor{(see also {\em race condition})}.
 		Distinguished further in this thesis as {\em limited happens-before} races and {\em pure happens-before} races;
@@
 		Opposite of {\em static analysis}.
-	\item {\bf Dynamic partial order reduction}.
+	\item {\bf Dynamic partial order reduction} \revisionminor{(DPOR)}.
 		Exploration algorithm for {\em stateless model checkers} which
@@
 		In theory this may mean that the {\em test case} might have been insufficient to expose the bug
-		(i.e., the code under test is wrong in any way whatsoever);
+		(\revisionminor{which the programmer may discover later via manual inspection,
+		or when her code fails in production,
+		or which may remain unnoticed}),
 		however, for pragmatic, tool-oriented discussion,
@@
 		possibly affecting that thread's subsequent logic.
+	\item \revision{{\bf misbehave.}
+		An optional system call offered by {\em Pebbles} kernels, especially the official reference kernel,
+		that changes the scheduler's behaviour, according to a numeric mode argument,
+		to amplify the likelihood of specific illustrative low-probability {\em thread interleavings}.
+		Used in several {\em P2} test cases,
+		although exactly how thread scheduling changes according to each specified mode
+		is kept secret to preserve the intended challenge of debugging (see \sect{\ref{sec:410-secrecy}}).}
 	\item {\bf mutex.}
@@
 		See \sect{\ref{sec:pebbles}}.
+	\item \revision{{\bf Pebbles.}
+		The educational kernel architecture used in 15-410 at CMU. See \sect{\ref{sec:pebbles}}.}
+	\item \revision{{\bf Pintos.}
+		The educational kernel architecture used in \uchos at the University of Chicago,
+		as well as several other universities. See \sect{\ref{sec:overview-pintos}}.}
 	\item {\bf pthreads.}
 		Thread library for Linux programs provided by POSIX.
+		\revision{The modern pthreads model traces its lineage back to
+		Mach's C threads \cite{mach-c-threads}
+		and DEC's Systems Research Center threads \cite{src-threads},
+		and is now specified by ISO/IEC/IEEE \cite{iso-posix}.}
 	\item {\bf Preemption.}
@@
 	\item {\bf Race condition.}
-		A confusing term which CMU 15-410 teaches to mean {\em concurrency bug},
-		but in research circles is generally taken to mean {\em data race}.
-		This thesis will avoid it entirely, using those two more precise terms instead.
+		\revision{Another term for {\em nondeterminism} in general,
+		%or {\em concurrency bug}.
+		which may also carry more specific nuance depending on context.
+		The term dates back (at least) to 1954, describing electrical signals' nondeterministic order of arrival
+		in circuit design \cite{switching-circuits}.
+		That paper further distinguished race conditions as either ``critical'' or ``non-critical'',
+		corresponding to {\em buggy} or {\em benign} in our terminology.
+		%
+		CMU 15-410 teaches the term consistently with this definition,
+		although may also refer to the buggy subset only (i.e., {\em concurrency bugs}) for brevity.
+		%
+		Recently, many systems research papers have used the term to refer to {\em data races} in particular.%
+		\footnote{In the author's personal experience, readers of their M.S. thesis \cite{landslide}
+		have misinterpreted its title to believe % oops
+		that Landslide be only a data-race detector,
+		rather than a model checker.}
+		To avoid confusion, this thesis will avoid the term entirely,
+		using the above two more precise terms instead.}
 	\item {\bf Reduction.}

### 2018-11-22 thesis/education.tex
edit education.tex with some of dave's comments

@@
 whether it offers pedagogical merit in the hands of students and/or TAs is a separate question.
-And while my MS thesis~\cite{landslide} showed that students
+And while my M.S. thesis~\cite{landslide} showed that students
 could annotate P3 Pebbles kernels and thence use Landslide to debug them,
@@
 I have run several user studies in the Operating Systems classes
-at Carnegie Mellon University (CMU), University of Chicago (U. Chicago), and Penn State University (PSU),
+at Carnegie Mellon University (CMU), University of Chicago (U. Chicago),
+and \revisionminor{The Pennsylvania} State University (PSU),
 wherein students get to use Landslide to find and diagnose their own bugs during the semester.
@@
 At U. Chicago, I collaborated with a TA to check submitted Pintos kernels,
-then returned any resulting bug reports to students
+then \revisionminor{they} returned any resulting bug reports to students
 (\sect{\ref{sec:education-eval-bugfinding}})
@@
 or (in the case of Paradise Lost) suffers an exponentially-exploding state space.
-Not that I scientifically measured it or anything,
-but this extended demo seemed to help students more clearly understand Landslide's intended workflow,
-at the cost of about 10-15 extra minutes of lecture time.
+%Not that I scientifically measured it or anything,
+This extended demo seemed to help students more clearly understand Landslide's intended workflow,
+at the cost of about 10-15 extra minutes of lecture time
+\revisionminor{(of course, this is an anecdotal opinion, not a scientific conclusion).}
 
@@
 As described in \sect{\ref{sec:landslide-setup}},
-all setup from the user's point of view is handled through the {\tt p2-setup.sh} script%
+all setup from the user's point of view is handled through the {\tt p2-setup.sh} script.%
 \footnote{PSU's version is called {\tt psu-setup.sh};
 in this section {\tt p2-setup.sh} refers to both unless otherwise noted.
-}.
+}
 It, its helper scripts (\sect{\ref{sec:landslide-glue}}),
@@
 from accidentally misusing Landslide in ways that could produce mysterious crashes, false bug reports, and so on
-(the need for each one, as the reader might imagine, discovered through bitter experience).
+(the need for each \revisionminor{check}, as the reader might imagine, discovered through bitter experience).
 These include:
@@
 	\item {\tt check-need-p2-setup-again.sh} checks if any source files in the original P2 source directory
-		(the argument supplied to {\tt p2-setup.sh}),
+		(the argument supplied to {\tt p2-setup.sh}) \revisionminor{have been updated},
 		in case the student hoped to fix some bug and verify their fix but forgot to re-run the setup script.
@@
 		and if so, refuses to do so and advises the student
-		to {\tt git clone} the repository afresh for simultaneous use%
+		to {\tt git clone} the repository afresh for simultaneous use.%
 		\footnote{This is ironically implemented with a non-atomic lock file
 		and should really be using {\tt flock} instead.
-		}.
+		}
 \end{itemize}
@@
 		for Landslide depending on the name of the test.
-		For example, for {\tt paradise\_\allowbreak{}lost} Landslide will not preempt in a function named {\tt critical\_section()},
+		For example, for {\tt paradise\_\allowbreak{}lost},
+		\revisionminor{designed to test student semaphore code,}
+		Landslide will not preempt in a function named {\tt critical\_section()},
 		which the test case uses to protect an internal counter used to detect the bug;
@@
 		Tests student mutexes under 2 threads with 2 iterations
-		(the 2nd iteration serves to expose problems with {\tt mutex\_unlock()} as well as {\tt mutex\_lock()}).
+		(the \revisionminor{second} iteration serves to expose problems with {\tt mutex\_unlock()} as well as {\tt mutex\_lock()}).
 		This test uses the {\tt TESTING\_MUTEXES}
@@
 		tests for proper handling of seemingly ``spurious'' wakeups in {\tt cond\_wait()}.
-		Written by Michael Sullivan.
+		Written by Michael J. Sullivan.
 	\item {\tt rwlock\_downgrade\_read\_test}:
 		Copied directly from the P2 test suite;
-		tests for mutually-exclusive and deadlock-free {\tt rwlock\_downgrade()}.
+		tests for mutually-exclusive and deadlock-free {\tt rwlock\_downgrade()}
+		\revisionminor{(a function which atomically converts a lock held in exclusive write mode to shared read mode)}.
 		Written by me (as a TA).
@@
 	\item Finds the declaration of {\tt ready\_list}, the scheduler runqueue declared by the basecode,
-		and detects if the student has modified to be an array of lists rather than a single one.
-		If so, defines the length of that array in a macro to be used by {\tt is\_runqueue()}
+		%and detects if the student has modified to be an array of lists rather than a single one.
+		\revisionminor{and detects if the student has replaced the default definition
+		with a different name and/or data structure.}
+		If so,
+		%defines the length of that array in a macro to be used by {\tt is\_runqueue()}
+		\revisionminor{emits macros to configure {\tt is\_runqueue()}'s behaviour to handle
+		certain common alternate approaches}
 		(part of the patch described below).
-		Either way defines a function {\tt get\_rq\_addr()} to return the address of the (first) list.
+		Either way defines a function {\tt get\_rq\_addr()} to return the address of the %(first)
+		list.
 	\item Changes the basecode's definition of {\tt TIME\_SLICE} from 4 to 1 (units of timer ticks)
@@
 
-The data set is comprised of Landslide's automatically-generated usage snapshots
+The data set \revisionminor{comprises} Landslide's automatically-generated usage snapshots
 from CMU students from semesters Spring 2015 to Spring 2018 inclusive,
@@
 CMU's official project grades from same,
-CMU students' survey responses and submitted project code from Fall 2017 and Spring 2018
+CMU students' survey responses and submitted project code from Fall 2017 and Spring 2018,
 PSU students' survey responses and submitted project code from Spring 2018,
@@
 Table~\ref{tab:photo-of-ze-studence} shows the student participation rate across semesters
-for the thread library study%
+for the thread library study.%
 \footnote{
@@
 participation at PSU was determined, for lack of snapshots, by sign-up form submissions, which might over-count slightly.
-}.
+}
 % and PSU
@@
 how many were deterministic
-(i.e., Landslide reported them on the very first interleaving tested, with no need for artificial preemptions)
+(i.e., Landslide reported them on the very first interleaving tested, with no need for artificial preemptions),
 how many were concurrency bugs,
@@
 I collaborated with the course staff
-to confirm whether each trace represented a real bug and not a false positive
+%to confirm whether each trace represented a real bug and not a false positive
+\revisionminor{to filter out false positives}
 before distributing them to the students.
@@
 Note that the tallies of bugs fixed may be undercounting
-on account of the possibility some students may have truly fixed them but skipped the verification step thereafter.
+% lmao
+%on account of the possibility
+\revisionminor{because} some students may have truly fixed them but skipped the verification step thereafter.
 
@@
 or were deemed too pedagogically important to allow the student to proceed without fixing
-(e.g., busy spin-wait synchronization loops)%
+(e.g., busy spin-wait synchronization loops).%
 \footnote{
@@
 even among false positives.
-}.
+}
 Other false positives arising from bugs in Landslide itself
@@
 
-Among the 4 groups volunteering to receive Landslide bug reports,
-Landslide found 7 distinct bugs across the two projects tested,
+\revisionminor{Among the four groups volunteering to receive Landslide bug reports,
+Landslide found seven distinct bugs across the two projects tested,
 which I confirmed by manual inspection of the students' code.
-3 among these were deterministic; 4 were concurrency bugs.
+three among these were deterministic; four were concurrency bugs.}
 
 \begin{enumerate}
-	\item {\tt priority-donate-multiple} found 2 deterministic bugs and 1 concurrency bug in {\tt threads} projects.
+	\item {\tt priority-donate-multiple} found \revisionminor{two}
+		deterministic bugs and \revisionminor{one} concurrency bug in {\tt threads} projects.
 		It found the targeted priority-leak bug
@@
 		which was found deterministically.
-	\item {\tt wait-simple} found 1 deterministic bug and 3 concurrency bugs in {\tt userprog} projects.
+	\item {\tt wait-simple} found \revisionminor{one} deterministic bug
+		and \revisionminor{three} concurrency bugs in {\tt userprog} projects.
 		It found a deadlock in two groups' implementations,
@@
 		whose {\tt process\_exit()} didn't guard against a parent process exiting and deallocating its memory first;
-		this root cause manifested in heap errors in 3 separate locations.
+		this root cause manifested in heap errors in \revisionminor{three} separate locations.
 		Lastly, the heap checking alone found a deterministic use-after-free in a third group's {\tt exec()}.
@@
 In the upcoming survey section, although only one of these groups did the survey,
-they enthusiastically reported finding the bug reports very helpful (\sect{\ref{sec:education-survey-response-data}}).
+they enthusiastically reported \revisionminor{having found}
+the bug reports very helpful (\sect{\ref{sec:education-survey-response-data}}).
 Despite the small sample size, I consider this
@@
 First, I analyzed CMU 15-410 students' overall project grades between Landslide users and non-users
-to see if Landslide helped them submit overall better implementations.
+to see if Landslide \revisionminor{might have} helped them submit overall better implementations.
 Second, I studied
@@
 or whether the study had not yet begun that semester.
-Intuition suggests students in the former group would be more likely
+Intuition suggests students in the former \revisionminor{(``declined'')} group would be more likely
 to be struggling too much to have free time to volunteer in the first place,
@@
 In addition to P2, I also show the subsequent P3 (kernel project) grades,
-likewise broken down by who used Landslide previously during P2%
+likewise broken down by who used Landslide previously during P2.%
 \footnote{The P3 distributions are slightly smaller than those from P2,
@@
 I omit them here.
-}.
+}
 Should Landslide use be correlated with higher grades on the later, more difficult concurrency project,
@@
 Anderson-Darling uses a weighted sum-of-squares distance metric to compare two empirical distribution functions (EDFs),
-testing if their samples are likely to arise from the same underlying, though unspecified, distribution%
+testing if their samples are likely to arise from the same underlying, though unspecified, distribution.%
 \footnote{
@@
 for its better sensitivity both to repeated deviations and to tail differences \cite{beware-kolmogorov-smirnov}.
-}.
+}
 I deem the difference in grade distributions significant when $p < 0.05$%
@@
 % should you be using the combined/grouped AD test instead?
-\footnote{I choose not to correct for multiplicity among these 4 comparsons like I do in the next section
+\footnote{I choose not to correct for multiplicity among these 4 comparsons \revisionminor{as} I do in the next section
 because each tests a different hypothesis.
@@
 It includes one thread repeating once the lock, unlock cycle
+\revisionminor{(i.e., one contender calls {\tt critical\_section()} twice)}
 so as to check for unlock/lock races as well as lock/lock interactions,
@@
 As students have free rein to design their mutex internals,
-this refers to the general class of mutex bugs
+this
+%refers to
+\revisionminor{test probes}
+the general class of mutex bugs
 in which any number of things can go wrong, depending on the implementation,
@@
 a quality-of-life improvement normally invisible to graders.
-For two, Landslide's automatic snapshots, being captured at the time of issuing each preemption trace,
+For \revisionminor{another}, Landslide's automatic snapshots, being captured at the time of issuing each preemption trace,
 necessarily miss the student's subsequent experience interpreting them.
@@
 \begin{enumerate}
-	\item Landslide complained of a nonexistent {\tt MAGIC\_BREAK} (Simics debugging function),
+	\item Landslide complained of a nonexistent {\tt MAGIC\_BREAK} (\revisionminor{a} Simics debugging function),
 		despite the student's code never invoking it.
@@
 		Commits 0573e34 (Bochs repository) and 654f459 (Simics repository) fixed this bug.
-		(Like the above, not actually a false positive bug report.)
+		(\revisionminor{As} above, not actually a false positive bug report.)
 	\item Landslide issued false invalid heap access reports to one student
@@
 
-After the question asking ``Would you recommend a friend taking OS next semester to use Landslide?'',
+After
+%the question
+asking ``Would you recommend a friend taking OS next semester to use Landslide?'',
 I asked the students for open-form reasons why or why not.
@@
 this question's answers turned out to be mostly praise.
-3 students declined to answer this question;
-I reproduce the rest here, paraphrasing for clarity and brevity%
+\revisionminor{Three} students declined to answer this question;
+I reproduce the rest here, paraphrasing for clarity and brevity,
+\revisionminor{denoting my own clarifications and commentary with [square brackets]}.%
 \footnote{Note that most students used the term ``race condition'' rather than ``concurrency bug'',
 as taught in CMU and PSU lecture material;
-I replaced these while paraphrasing in accordance with \sect{\ref{sec:glossary}}.}.
+I replaced these while paraphrasing in accordance with \sect{\ref{sec:glossary}}.}
 Answers from CMU:
@@
 	\item Helped verify our atomic primitives were correct.
-	\item Would recommend, but takes too much time (unclear, but seems to be referring to execution time rather than setup/usage).
+	\item Would recommend, but takes too much time [unclear, but seems to be referring to execution time rather than setup/usage].
 	\item Tests are automated and can be left running for a long time.
-		However, faced uptime issues with CMU's linux servers (which reboot every night);
+		However, faced uptime issues with CMU's Linux servers (which reboot every night);
 		wished for a more reliable execution environment.
@@
 	\item Finds concurrency bugs with little effort that may be undiscovered otherwise.
-		(Also provided some interface feedback here, since I didn't ask CMU students a separate question for such;
-		see next section.)
+		[This respondent also provided some interface feedback here, since I didn't ask CMU students a separate question for such;
+		see next section.]
 	\item Easy to learn and a simple way to test our P2.
@@
 	\item Helpful to find bugs you weren't previously aware of.
-		Makes more sense to use an expressly designed tool rather than (unit/stress) tests.
+		Makes more sense to use an expressly designed tool rather than [unit/stress] tests
+		\revisionminor{which may or may not find bugs}.
 	\item Helpful and easy to use.
@@
 	\item Found bugs I kept overlooking, which may have taken many hours to find otherwise.
-	\item Helpful for the difficult step from code being ``finished'' (scare-quotes theirs; presumably meaning ``feature-complete'')
+	\item Helpful for the difficult step from code being ``finished''
+		[scare-quotes theirs; presumably meaning ``feature-complete'']
 		to getting rid of all concurrency bugs.
@@
 	\item Helped find a bug I wouldn't have found otherwise.
-		Did not show an interleaving directly (i.e., did not issue a bug report),
+		Did not show an interleaving directly [i.e., did not issue a bug report],
 		but reported a data race that turned out to be a concurrency bug upon inspection.
@@
 		and complained of inaccurate line numbers
-		(likely referring to how the current behaviour indicates the line of code {\em after} a function call,
+		[likely referring to how the current behaviour indicates the line of code {\em after} a function call,
 		corresponding to the {\tt call} instruction's pushed return address,
-		rather than the function call itself). (This answer from CMU; see above.)
-	\item Including a manual or tutorial would be helpful (presumably beyond the user guide's instructions, such as recapping the procedure shown in the lecture demo which wasn't written down anywhere).
-	\item Don't print warnings about line length exceeding 80 characters (inconsistency between 15-410 and \psuos compilation options).
-	\item ``It takes too long. But I guess that's impossible to fix.'' (Well, it's an open research problem to fix!)
+		rather than the function call itself]. [This answer from CMU; see above.]
+	\item Including a manual or tutorial would be helpful [presumably beyond the user guide's instructions, such as recapping the procedure shown in the lecture demo which wasn't written down anywhere].
+	\item Don't print warnings about line length exceeding 80 characters [inconsistency between 15-410 and \psuos compilation options].
+	\item ``It takes too long. But I guess that's impossible to fix.'' [Well, it's an open research problem to fix!]
 	\item Preemption traces should explicitly indicate where in the interleaving the bug occurred.
-		(Root cause identification is its own research area, but more detail is certainly possible.)
-	\item Improve explanation of data races (in the user guide, perhaps).
-	\item Happy with it as-is (3 students)
-	\item No response (7 students)
+		[Root cause identification is its own research area, but more detail is certainly possible.]
+	\item Improve explanation of data races [in the user guide, perhaps].
+	\item Happy with it as-is [\revisionminor{reported by three} students]
+	\item [\revisionminor{Seven non-respondents}]
 \end{enumerate}
@@
 suggests CMU students
+\revisionminor{(as well as Professor Eckhardt)}
 also mostly wish for
@@
 while PSU and U. Chicago students generally compared Landslide more favourably to stress testing
-(while CMU students comparatively preferred the more nuanced answer that it depends on the type of bug),
-and reported more often that helped them understand concurrency better.
+(CMU students comparatively preferred the more nuanced answer that it depends on the type of bug),
+and reported more often that \revisionminor{it} helped them understand concurrency better.
 %Considering that CMU's 15-410 demands substantial prerequisite concurrency experience
@@
 was accompanied by some drawback which prevented it from being perfectly objective science,
-but many of them presented tentatively positive results nonetheless.
+but many of them presented tentatively positive results nonetheless,
+\revisionminor{and relatively very little negative feedback such as false positives}.
 Landslide helped many students find and fix many bugs (\sect{\ref{sec:education-eval-bugfinding}}),
@@
 who are least in need of the extra debugging help.
-This was especially evident in the annotation-required P3 study from \cite{landslide},
+This was especially evident in the annotation-required P3 study from \revisionminor{2012} \cite{landslide},
 in which only 5 groups volunteered (15\% of the 34 total who submitted P3 that semester),
@@
 
-More than just trying to draw firm conclusions from
+\revisionminor{Aside from} just trying to draw firm conclusions from
 the opinions of students who are just learning concurrency to begin with,
@@
 		This would help get a sense of how the students regarded Landslide
-		on a spectrum between annoying style checker or a life-saver,
+		on a spectrum between annoying style checker \revisionminor{and} life-saver,
 		and potentially suggest options to make Landslide suppress certain types of bug reports.
@@
 I answered each by explaining that it would take more effort on their part,
-and then if they were still interested,
+and then, if they were still interested,
 guided them through the annotations on a case-by-case basis.
@@
 While part of the point of this experimental design was to evaluate Landslide as a grading tool in the hands of TAs,
-I would be remiss to mention that I also feared the automatic annotation process would not be as robust as the P2 version.
+I would be remiss \revisionminor{not}
+to mention that I also feared the automatic annotation process would not be as robust as the P2 version.
 Indeed, while helping Kevin get oriented with using Landslide,
@@
 one, not incentivizing the students to directly improve their grades
-(instead offering only the vague promise of a ``learning experience'' debugging their code only after handin),
+(instead offering only the vague promise of a ``learning experience'' debugging their code after handin),
 and two, not traveling to the university to introduce the research topic in an in-person lecture
@@
 Hypothetically, I could have achieved greater user study participation
-either by offering extra credit to students
+either by \revision{giving a lecture remotely via videoconference},
+by offering extra credit to students,
 or by offering an autograder-like interface for students to receive bug reports before their deadlines instead of after

### 2018-11-23 thesis/thesis.tex
configure revision changebars
David A. Eckhardt

@@
 \usepackage{xeCJK}
+\usepackage[xetex]{xechangebar}
 
@@
 
+%\newcommand\revision[1]{#1\xspace}
+\newcommand\revision[1]{\cbstart{}#1\cbend\xspace}
+%\newcommand\revisionminor[1]{#1\xspace}
+\newcommand\revisionminor[1]{\cbstart{}#1\cbend\xspace}
+
 \begin{abstract}
@@
 
-Professor David Eckhardt
+Professor David A. Eckhardt
 provided unending guidance on education, writing, career, life, you name it,
@@
 
-\newcommand\revision[1]{#1\xspace}
-
 \input{intro}

### 2018-11-23 thesis/landslide.tex
dave revisions on landslide chapater

@@
 		creates a snapshot tarball of them and the current version of the user's code,
-		and sends it to me for nefarious research purposes.
+		and sends it to me for nefarious research purposes
+		\revisionminor{(see \sect{\ref{sec:education-pebbles-instrumentation}})}.
 \end{itemize}
@@
 		this is equivalent to {\tt -0}
-		(and according to my experience, way faster as well).
+		(and according to my experience, \revisionminor{much} faster as well).
 	\item {\tt -H}: use Limited Happens-Before for data-race analysis (\sect{\ref{sec:background-hb}})
@@
 
-Use of {\tt tell\_landslide()} annotations is not necessary,
+Use of {\tt tell\_landslide()} annotations
+\revisionminor{(\sect{\ref{sec:tell-landslide}})}
+is not necessary,
 although {\tt tell\_landslide\_\allowbreak{}preempt()} and {\tt tell\_landslide\_dump\_stack()}
@@
 \end{itemize}
-These could not be implemented as asserts in the test code itself
+\revision{Because Landslide tests the ultimate value of these variables after all threads have completed execution,}
+these could not be implemented as asserts in the test code itself
 without requiring the student to implement {\tt thr\_join()} and {\tt thr\_exit()},
@@
 (i.e., each stack trace indicates ``this thread ran until it reached the indicated line of code'').
-The bug reports are formatted in html, recommended to be viewed in a web browser.
+The bug reports are formatted in \revisionminor{HTML}, recommended to be viewed in a web browser.
 An example is shown in Figure~\ref{fig:bugreport}.
@@
 In addition to the preemption trace, the bug report provides some additional helpful information:
-a stack trace of the current thread at the ultimate point when the bug was executed,
+a stack trace of the current thread at the ultimate point when the bug was executed
+\revision{(the same as the stack trace in the cell corresponding to that thread in the bottom row of the table)},
 a message indicating the nature of the bug encountered,
@@
 by searching for the macro {\tt FOUND\_A\_BUG\_HTML\_INFO} in Landslide's code.}
+\revision{\sect{\ref{sec:future-friendly}} discusses how these bug reports might be further improved
+by adding more information still,
+such as identifying data-race preemption points or listing memory conflicts that occurred during each transition.}
 
@@
 
-The educational experiments in this thesis focus on projects which students implement on top of provided kernel basecode which Landslide already ``understands''.
+\revision{My M.S. thesis \cite{landslide}
+investigated the annotation overhead required for a user to instrument a Pebbles kernel for use with Landslide.
+On average, students who volunteered for the study (conducted during P3) required 2 hours for this process.
+While many of these students then went on to find and diagnose bugs,
+this was deemed an unacceptable burden for an educational tool to impose on struggling students.
+%(as opposed to those who had enough free time and ambition to volunteer for a study
+%that was advertised in advance as requiring manual annotations)
+
+Hence,}
+the educational experiments in this thesis focus on projects which students implement
+on top of provided kernel basecode which Landslide already ``understands''.
 Such understanding is conferred via the annotations described in this section.
@@
 Following are the less self-explanatory options.
+\revision{Those defined with an equals sign between name and value
+are implemented as bash variables,
+which the annotation scripts check after processing the configuration to emit a corresponding {\tt \#define}
+within Landslide itself;
+those defined with no equals sign
+are implemented as bash functions,
+which the annotation scripts define in advance of processing the configuration to do something more complicated.}
 \begin{itemize}
@@
 	\item {\tt CURRENT\_THREAD\_LIVES\_ON\_RQ=0} - Landslide infers the list of runnable threads from the {\tt tell\_landslide\_on\_rq()} and {\tt off\_rq()} annotations (described below).
-		Some kernels\footnote{most, actually} remove the current thread from their runqueue,
+		Some kernels\footnote{Most, actually.} remove the current thread from their runqueue,
 		such that the abstract set of all runnable threads is actually the runqueue plus the current thread rather than just the runqueue.
-		Other kernels\footnote{the author's own student kernel from long ago}
+		Other kernels\footnote{The author's own student kernel from long ago.}
 		leave the current thread on the runqueue,
@@
 		\footnote{This option replaces the deprecated {\tt kern\_current\_extra\_runnable()} annotation from {\tt student.c} described in \cite{landslide} \S{}6.2.3.}
+		\revision{Whether or not a kernel requires this annotation could be auto-detected in future work.}
 	\item {\tt PREEMPT\_ENABLE\_FLAG=NAME} - name of a global variable which the kernel uses to toggle scheduler preemptability, for kernels which may disable preemption without disabling interrupts.
@@
 		(usually 0; note that many kernels use a nesting depth counter where any positive value corresponds to disabled).%
-		\footnote{These two options replace the deprecated {\tt kern\_ready\_for\_timer\_interrupt()} annotation from {\tt student.c} described in \cite{landslide} \${}6.2.3.}
+		\footnote{These two options replace the deprecated {\tt kern\_ready\_for\_timer\_interrupt()} annotation from {\tt student.c} described in \cite{landslide} \sect{6.2.3}.}
 	\item {\tt PATHOS\_SYSCALL\_IRET\_DISTANCE=VALUE} - indicate how much stack space is used by the reference kernel's system call wrappers.
@@
 	\item {\tt BUG\_ON\_THREADS\_WEDGED=1} - set to 0 to disable deadlock detection but instead let the kernel keep receiving system interrupts when all threads appear blocked.%
-		\footnote{once used in the bad old days; now recommended for debugging use only}
+		\footnote{Once used in the bad old days; now recommended for debugging use only.}
 	\item {\tt TIMER\_WRAPPER\_DISPATCH=NAME} - used to manually indicate a label before the end of the timer interrupt assembly wrapper, in case the {\tt iret} instruction couldn't be found automatically (see {\tt pebsim/definegen.sh}).
@@
 
-The following annotations enable Landslide to track locksets for data race analysis.
+The following annotations enable Landslide to track locksets~\cite{eraser} for data race analysis.
 If not provided, it will be as if Landslide assumes no guarantees about mutual exclusion or happens-before,
@@
 to maintain an accurate representation of all the existing threads on the simulated system
-and track what concurrency-relevant actions each is performing at any given time,
+and track \revisionminor{which} concurrency-relevant actions each is performing at any given time,
 and to orchestrate the sequence of timer interrupts necessary
@@
 System-wide state is stored in a single {\tt struct sched\_state},
-including the thread queues (runqueue, deschedule queue, and sleep queue),
+including \revisionminor{several queues to track threads in various states of runnability}
+(runnable queue, deschedule queue, and sleep queue),
 while per-thread state is stored in {\tt struct agent}s (named after the terminology of \cite{dbug-ssv}) which live on said queues.
@@
 	\item {\tt cause\_timer\_interrupt()}
-		triggers a pending timer interrupt, whose handler will be entered as soon as the upcoming instruction is executed.
+		triggers a pending timer interrupt, whose handler will be entered as soon as
+		\revisionminor{the execution of the upcoming instruction is completed}.
 	\item {\tt cause\_timer\_interrupt\_immediately()}
@@
 The Simics version is merely a bunch of wrapper glue code around the {\tt set-bookmark} and {\tt skip-to} backtracking commands.
-Bochs however does not support backtracking, so I instead use {\tt fork()} to get Linux to copy the machine state for me.
+Bochs however does not support backtracking, so I instead use {\tt fork()} to get Linux
+to copy the \revisionminor{Bochs process and thus the simulation} state for me.
 
@@
 The options specified in \sect{\ref{sec:landslide-directly}}
-are handled by a family of gross shell scripts that live in {\tt pebsim/}.
+are handled by a family of \revisionminor{distasteful} shell scripts that live in {\tt pebsim/}.
 
@@
 with each state being a preemption point as identified in \sect{\ref{sec:landslide-pps}},
-and $T(t_i)$ will denote the thread which was scheduled to produce that transition.
+and $T(t_i)$ will denote the thread which was scheduled \revisionminor{(switched to)} to produce that transition.
 A visual example will be given at the end to help reinforce the intuition behind the formalism.
@@
 The happens-before relation expresses when two thread transitions can potentially be reordered,
-or in other words, are concurrent (despite the serialized nature of the simulated execution).
+or in other words, are \revisionminor{logically} concurrent (despite the serialized nature of the simulated execution).
 This relation is expressed in the following definitions paraphrased from \cite{dpor}.
@@
 	\item {\bf Enabled}: A transition $t_i$ is enabled in a state $s$ when a state $s'$ exists such that $s \xrightarrow{t_i} s'$ exists.
-		In systems terms, the scheduler at state $s$ considers $T(t_i)$ to be runnable.
+		In systems \revisionminor{research} terms, the scheduler at state $s$ considers $T(t_i)$ to be runnable.
 	\item {\bf Dependent}: Two transitions $t_i$ and $t_j$ are dependent if
@@
 		or the execution of $t_1$ at $s$ causes $T(t_2)$
-		to change state from blocked to runnable or vice versa.
+		to change state from blocked to runnable or vice versa.%
 		\footnote{The original paper's definition includes a second criterion that, from $s$,
@@
 
-Whenever a conflict is identified, it also invokes the data race analysis (\sect{\ref{sec:landslide-datarace}}).
+Whenever a conflict is identified,
+\revisionminor{Landslide} also invokes the data race analysis (\sect{\ref{sec:landslide-datarace}}).
 It checks for {\tt free()}/access conflicts as well as access pairs,
@@
 To find which alternate interleavings need to be tested,
-% TODO: change i<4 to i<j.
-% i accidentally typed i<4 here and decided it'd be funny to leave it this way
-DPOR compares pairwise each pair of transitions $t_i$ and $t_j$, $i<4$, in the current interleaving $S$.
+DPOR compares pairwise each pair of transitions $t_i$ and $t_j$, $i<j$, in the current interleaving $S$.
 If $t_i \rightarrow_S t_j$ then they cannot be reordered,
@@
 	\item If $T(t_j)$ is runnable at $s$, return $I_{ij} = (t_1 \dots t_{i-1}, T_j)$ ({\tt tag\_good\_sibling()}).
-	\item Otherwise, there must be some third thread runnable at $s$;
+	\item Otherwise, there must be some third thread runnable at $s$;%
 		\footnote{AFSOC $T(t_i)$ is the only runnable thread at $s$,
@@
 
+% TODO: check figure placement (should be exactly in line with text) - same for subsequent 2
+% TODO: maybe it's ok t ohave them not in line?
 \begin{figure}[h]
@@
 Note also that in $\bigstar$2,
-the reordered \dporTB{{\tt tmp2 <- x}} is not directly part of the memory conflict which marked it,
-but it must be executed first to reach the conflicting \dporTB{{\tt x <- tmp2}}.
+the reordered \dporTB{{\tt tmp2~<-~x}} is not directly part of the memory conflict which marked it,
+but it must be executed first to reach the conflicting \dporTB{{\tt x~<-~tmp2}}.
 
@@
 note the two disjoint equivalence classes \{B,E,G,I\} and \{C,D,F,H,J\},
-distinguished by the order of the two final {\tt x <- tmpN}s.
+distinguished by the order of the two final {\tt x~<-~tmpN}s.
 Note also that
@@
 its reduced subset state space still contains redundancy,
-arising from the fact that one pair of those 4 events is two reads, and hence not actually in conflict.
+arising from the fact that one pair of those \revisionminor{four} events is two reads, and hence not actually in conflict.
 Visual inspection shows that $\bigstar$4, while locally justified
-in trying to reorder \dporTA{{\tt tmp1 <- x}} before \dporTB{{\tt x <- tmp2}},
+in trying to reorder \dporTA{{\tt tmp1~<-~x}} before \dporTB{{\tt x~<-~tmp2}},
 effectively serves % oops
-only to reorder it with \dporTB{{\tt tmp2 <- x}}
+only to reorder it with \dporTB{{\tt tmp2~<-~x}}
 relative to the first symmetric subtree ($\bigstar$1).
@@
 it did not ``remember'' which memory conflict it wanted to reorder \dporTA{$\mathbf{T_1}$} around
-(i.e., that \dporTB{{\tt x <- tmp2}} should come before \dporTA{{\tt tmp1 <- x}}).
+(i.e., that \dporTB{{\tt x~<-~tmp2}} should come before \dporTA{{\tt tmp1~<-~x}}).
 Upon witnessing the conflict in the new (intended) order,
@@
 producing interleavings regrettably equivalent to ones already tested
-(all told, only \dporTB{{\tt tmp2 <- x}} and \dporTB{{\tt tmp2++}} having been reordered around \dporTA{{\tt tmp1 <- x}}).
+(all told, only \dporTB{{\tt tmp2~<-~x}} and \dporTB{{\tt tmp2++}} having been reordered around \dporTA{{\tt tmp1~<-~x}}).
 %
@@
 among all preceding transitions independent with the conflicting one
-(here, \dporTB{{\tt tmp2 <- x}} and \dporTB{{\tt tmp2++}} independent with \dporTA{{\tt tmp1 <- x}})
+(here, \dporTB{{\tt tmp2~<-~x}} and \dporTB{{\tt tmp2++}} independent with \dporTA{{\tt tmp1~<-~x}})
 for an already-explored interleaving beginning with the target thread.
@@
 which checks (in this case after executing K),
-that if the first event to be reordered (here, \dporTA{{\tt tmp1 <- x}})
+that if the first event to be reordered (here, \dporTA{{\tt tmp1~<-~x}})
 has already been tested in an equivalent reordering around any number of preceding events
-(here, \dporTB{{\tt tmp2 <- x}} and \dporTB{{\tt tmp2++}}),
+(here, \dporTB{{\tt tmp2~<-~x}} and \dporTB{{\tt tmp2++}}),
 then the newly marked subtree is safe to prune.
@@
 so named because it effectively puts \dporTA{$\mathbf{T_1}$}
-``to sleep'' until after the true conflicting access of \dporTB{{\tt x <- tmp2}}.
+``to sleep'' until after the true conflicting access of \dporTB{{\tt x~<-~tmp2}}.
 Landslide's implementation differs from prior work,
@@
 In Pure Happens-Before,
-condition 4 is replaced with the traditional distributed systems notion of Happens-Before \cite{lamport-clocks}.
+condition \#4 is replaced with the traditional distributed systems notion of Happens-Before \cite{lamport-clocks}.
 Landslide implements this via the vector clocks approach described by {\sc FastTrack} \cite{fasttrack}.
@@
 
-I use the {\sc Djit+} rules for reads and writes rather than the {\sc FastTrack} ones,
+I use the \revisionminor{{\sc Djit+} rules \cite{djit} (as presented in \cite{fasttrack})}
+for reads and writes rather than the {\sc FastTrack} ones,
 even though they more often incur $O(n)$ runtime in the size of the vector clocks:
@@
 defined in {\tt schedule.h},
-decides if a given thread would require a preemption beyond the current bound to switch to%
+decides if a given thread would require a preemption beyond the current bound to switch to.%
 \footnote{Since ``voluntary'' context switches (e.g. arising from {\tt yield()})
@@
 interleavings with more than $B$ context switches may still be tested.
-}.
+}
 The DPOR implementation then checks, for some $I_{ij}$ it wants to mark for exploration,
@@
 although \sect{\ref{sec:quicksand-eval}}'s larger dataset suggests 3 would be considerably more thorough.
-On the other hand, any finite such bound can provide only a heuristic verification guarantee anymore.
+On the other hand, any finite such bound can provide only a heuristic verification guarantee. %anymore.
 Preserving the full formal verification, i.e., continuing iteration until $B = \mathsf{max}_P(U)$
@@
 but also introduces a factor of $\mathsf{max}_P(U)$ repeated work.
+\revisionminor{Landslide takes this approach for now (rather than stopping at any finite cutoff).}
 Future work could memoize already-tested interleavings so that each iteration of $B$ could test only
@@
 Even though all test cases used in this thesis's experiments are hand-written to ensure runtime
+\revisionminor{which is}
 not just finite but also short (on account of exponential state space sizes),
@@
 
-Landslide checks for two distinct categories of infinite loops,
+Landslide checks for two distinct categories of \revisionminor{potentially-infinite} loops,
 visualized in Figure~\ref{fig:heuristic-loop}.
@@
 waiting for another thread to allow it to proceed,
-rather than using the established synchronization API%
+rather than using the established synchronization API.%
 \footnote{Even when the student doesn't open-code any such synchronization and always uses {\tt deschedule}
@@
 often relies on this functionality, as data-race preemption points within {\tt mutex\_lock()}
-would otherwise disrupt Landslide's ability to recognize blocking on a contended lock.}.
+would otherwise disrupt Landslide's ability to recognize blocking on a contended lock.}
 If Landslide failed to recognize that the thread was in principle blocked,
@@
 % nb. not system calls, not even d/mr, as some student mutexes may use those and you don't want to count them
-Interesting here is defined heuristically as any other known P2 API call,
-with the exception of {\tt mutex\_lock()} and {\tt mutex\_unlock()}%
+``Interesting'' here is defined heuristically as any other known P2 API call,
+with the exception of {\tt mutex\_lock()} and {\tt mutex\_unlock()}.%
 \footnote{Landslide must recognize yield-blocking loops that contain mutex operations
@@
 which for example the {\tt paraguay} test uses intentionally,
-because it is testing the correctness of the student's {\tt cond\_wait()}.}.
+because it is testing the correctness of the student's {\tt cond\_wait()}.}
 Whenever this counter reaches the heuristic cutoff of 10 ({\tt TOO\_MANY\_YIELDS}),
@@
 Landslide also counts the number of atomic instructions ({\tt xchg}, {\tt xadd}, {\tt cmpxchg}, etc),
-and marks the thread blocked if it exceeds 100 such ({\tt TOO\_MANY\_XCHGS\_TIGHT\_LOOP})%
+and marks the thread blocked if it exceeds 100 such ({\tt TOO\_MANY\_XCHGS\_TIGHT\_LOOP}).%
 \footnote{If preemption points exist in between, instead only 20 such ({\tt TOO\_MANY\_XCHGS\_WITH\_PPS}),
-to avoid stressing DPOR's $O(n)$ independence computation.}.
+to avoid stressing DPOR's $O(n)$ independence computation.}
 When such a loop contains neither {\tt yield} nor atomics,
@@
 % where the htm failure path uses a stop the world thing...
-I have never observed this in practice, as the conditions of such blocking loops are usually read-only.
+I have never observed this in practice, as the conditions \revisionminor{checked by} such blocking loops are usually read-only.
 Future work could heuristically address this by deprioritizing such threads,
@@
 which ordinarily would trigger deadlock detection,
-Landslide checks the system for any yield-looping threads that were blocked heuristically%
+Landslide checks the system for any yield-looping threads that were blocked heuristically.%
 \footnote{This includes threads blocked on specific mutexes using the {\tt blocked\_on\_addr} field,
-which is set when {\tt yield} is invoked within {\tt mutex\_lock()} without waiting for 10 loops.}.
+which is set when {\tt yield} is invoked within {\tt mutex\_lock()} without waiting for 10 loops.}
 If any exist, it forces them awake and allows them to proceed;
 if they are truly blocked in principle,
-they will merely trip the yield-loop limit and go back to sleep again.
+they will merely trip the yield-loop limit and go back to sleep again,
+\revisionminor{whereupon Landslide will issue a deadlock bug report after all.}
 

### 2018-11-28 thesis/future.tex
some more citations
dave revisions on landslide chapater
dave's (easy) revisions to related and future

@@
 \section{User-friendliness}
+\label{sec:future-friendly}
 
@@
 there is certainly room to improve the diagnosis step beyond just showing the user one static HTML table.
-Related works like Symbiosis \cite{symbiosis} can find a minimal difference between the failing trace and a non-failing one,
+Related works \revisionminor{such as} Symbiosis \cite{symbiosis}
+can find a minimal difference between the failing trace and a non-failing one,
 which would allow the user to effortlessly pinpoint which preemptions are closest to the true root cause.
@@
 not just within the fixed state space under one given test,
-and symbolic execution frameworks such as Contessa \cite{contessa}
+and symbolic execution frameworks such as Contessa~\cite{contessa}
 could search for possible inputs to suggest changes which might expose them.
@@
 finding a minimum number of threads necessary to expose all behaviours
-(at least 3 in this example),
+(at least \revisionminor{three} in this example),
 % 2 papers rec'd from https://twitter.com/bblum0/status/1018233122750640130
@@
 all formally verified against external specifications,
-provide important pioneering steps in this direction
+provide important pioneering steps in this direction.%
 \footnote{Of course, being verified themselves, these projects have no need for a verified Landslide,
-but checking unverified (e.g. student) code with such would still be an improvement.}.
+but checking unverified (e.g., student) code with such would still be an improvement.}
 
@@
 but is also notorious for amplifying any biases in its training set in often unpredictable ways
-\cite{conceptnet-bias}.
+\cite{conceptnet-bias,amazon-ai}.
 In the domain of concurrency testing, this would translate to being able to find only types of bugs

### 2018-11-28 thesis/related.tex
dave's (easy) revisions to related and future

@@
 
-\section{Systematic Concurrency Testing}
+\section{Stateless Model Checking}
 
@@
 
-Systematic concurrency testing dates back to Verisoft \cite{verisoft},
-the 1997 tool which first attempted to exhaustively explore the possibile ways to interleave threads.
+Stateless model checking dates back to Verisoft \cite{verisoft},
+the 1997 tool which first attempted to exhaustively explore the \revisionminor{possible} ways to interleave threads.
 Since then, researchers have built many tools along the same lines to test many kinds of programs.
-One of the best-known SCTs is Microsoft Research's CHESS \cite{chess},
+One of the best-known MCs is Microsoft Research's CHESS \cite{chess},
 a checker for userspace C++ programs which preempts on synchronization APIs by default,
@@
 although it focuses on interrupt nondeterminism for testing device drivers,
-and is limited to injecting at most one interrupt per test run (as if under ICB with a bound of 1).
+and is limited to injecting at most one interrupt per test run (as if under ICB with a bound of~1).
 %
@@
 %
-D\'{e}j\`{a} Fu \cite{dejafu} is a SCT
+D\'{e}j\`{a} Fu \cite{dejafu} is a model checker
 for the Haskell language,
@@
 
-The problem of relaxed memory nondeterminism alone has inspired the creation of several new SCTs in the past few years.
-Relacy \cite{relacy}, a header-only C++ SCT library for checking synchronization primitives,
+The problem of relaxed memory nondeterminism alone has inspired the creation of several new tools in the past few years.
+Relacy \cite{relacy}, a header-only C++ model checking library for checking synchronization primitives,
 was the first to broach this field,
-although requires custom annotations for non-atomic memory accesses
+although \revisionminor{it} requires custom annotations for non-atomic memory accesses
 and
@@
 to capture most of the C++11 memory model's new behaviours.
-Nidhugg \cite{nidhugg} is a SCT for TSO and PSO which instruments LLVM abstract assembly,
+Nidhugg \cite{nidhugg} is a checker for TSO and PSO which instruments LLVM abstract assembly,
 % i don't actually know what the difference is #overlyhonestmethods
@@
 %
-No relaxed-memory SCT has yet proposed a satisfactory model for the ``thin-air'' problem \cite{sully-thesis},
+No relaxed-memory MC has yet proposed a satisfactory model for the ``thin-air'' problem \cite{sully-thesis},
 which can cause state space cycles in a way not yet well-understood and remains future work.
@@
 To date a number of techniques have been proposed to mitigate exponential explosion,
-the Sisyphean rock of SCT.
+the Sisyphean rock of stateless MC.
 The notion that some interleavings of concurrent threads could lead to indistinguishable program states and be therefore redundant,
@@
 rather than to rely on imprecise static alias analyses,
-and is now considered the baseline for all subsequent state space reduction approaches in SCT.
+and is now considered the baseline for all subsequent state space reduction approaches in stateless MC.
 That paper includes the {\em sleep sets} extension,
@@
 
-Of course, no matter how optimal a sound reduction, there will always be programs too large to test.
+Of course, no matter how \revisionminor{powerful} a sound reduction, there will always be programs too large to test.
 To provide even partial results for state spaces that exceed the testing budget
@@
 
-Although many SCTs listed in the previous section are content to report any data races as outright bugs,
+Although many MCs listed in the previous section are content to report any data races as outright bugs,
 RacerX \cite{racerx} showed that tools must be careful not to overwhelm users with benign warnings
@@
 RaceFuzzer \cite{racefuzzer} avoids false positives by requiring an actual failure be exhibited, as Quicksand does,
-although it uses random schedule fuzzing rather than systematic testing for its concurrency coverage.
+although it uses random schedule fuzzing rather than model checking for its concurrency coverage.
 %
@@
 which \sect{\ref{sec:quicksand-eval}} showed is necessary to find certain bugs.
-Quicksand builds on Portend's approach by introducing a feedback loop between the data-race analysis and SCT,
+Quicksand builds on Portend's approach by introducing a feedback loop between the data-race analysis and model checking,
 which results in a stronger verification property
@@
 and also had difficulty understanding single failing executions.
-In fact, the study explicitly recommended tool support for testing many thread interleavings automatically (SCT)
+In fact, the study explicitly recommended tool support for testing many thread interleavings automatically (model checking)
 and for execution traces to communicate sequences of important events (preemption traces),
@@
 
+\revision{The Deadlock Empire~\cite{deadlockempire} is an educational introduction to concurrent programming and debugging,
+presented as an online puzzle-hunt of sorts, backed up by a fantasy-themed narrative.
+Each puzzle requires the user to manually step through the execution of multiple threads
+to find buggy interleavings.
+%which in some cases requires the use of an ``expand'' button that
+Though itself lacking in scientific analysis of its educational power,
+it demonstrates the kind of user-friendly features and overall approach
+that many students expressed desire for in their survey responses (\sect{\ref{sec:education-eval-survey}}).
+Future work could extend Landslide's preemption traces to include similar interactivity.}
+
+% TODO: more papers to look up:
+% Tool Design and Student Testing Behavior in an Introductory Java Course
+% Testing Strategies for the Automated Grading of Student Programs
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \section{Transactional memory}
@@
 Such extensions come with the challenge of even more complicated behavioural semantics
-for SCT to accurately model and verify in future work.
+for MCs to accurately model and verify in future work.
 
@@
 Landslide is the first I know of to verify client programs thereof.
-McRT STM \cite{mc-tm-with-spin} uses SPIN \cite{spin} to model check an STM implementation
+% TODO: citation as noun - global scan
+\revision{\cite{mc-tm-with-spin} used SPIN \cite{spin} to model check McRT STM \cite{mcrt-stm},
+an STM implementation built by Intel,
+verifying}
 up to 2 threads running 1 transaction each with up to 3 memory accesses.
@@
 %
+\revision{Language-level concurrency abstractions were first introduced by the Communicating Sequential Processes (CSP)
+model~\cite{communicating-sequential-processes},
+which defines a concurrent program as a set of isolated processes
+allowed to communicate only via blocking input/output commands.
 Erlang \cite{erlang}, an early concurrent functional language,
-introduced the actor model for concurrency,
-in which threads share no state and must communicate only by message-passing.
+introduced the actor model for concurrency based on the CSP model,
+in which threads likewise} share no state and must communicate only by message-passing.
 %One might say this statically guarantees freedom from data races;
@@
 so concurrency bugs, especially deadlocks, are not ruled out.
-Concuerror \cite{concuerror}, discussed above, is a SCT tool for Erlang programs.
+Concuerror \cite{concuerror}, discussed above, is a MC tool for Erlang programs.
 %
@@
 using the type system to enforce sound locking discipline across such accesses,
-again preserving type soundness and data-race freedom%
-\footnote{The author themself contributed the original design for this latter feature.}.
-I know of no existing SCT for Rust as of yet.
+again preserving type soundness and data-race freedom.%
+\footnote{The author themself contributed the original design for this latter feature.}
+I know of no existing model checker for Rust as of yet.
 %
@@
 
-Coming at nondeterminsm from the opposite angle as SCT,
+Coming at nondeterminsm from the opposite angle as model checking,
 which aims to push the frontier of testing coverage to expand as many interleavings as possible,
@@
 manually annotate speed-critical nondeterministic sections
-and then check the resulting state spaces using dBug's SCT as normal.
+and then check the resulting state spaces using dBug's model checking as normal.
 %
@@
 when that would be too aggressive for the intended behaviour.
-These determinizing runtimes serve a different purpose than SCT:
+These determinizing runtimes serve a different purpose than MC:
 they seek to preserve the stability of existing code already running in production,
@@
 
-Analogous to SCT,
+Analogous to stateless MC,
 which seeks good coverage of possible thread execution paths under schedule nondeterminism,
@@
 Symbolic executors
-abstract a program's variables and use constraint solvers such as Z3 \cite{z3}
+abstract a program's variables and use constraint solvers such as Z3~\cite{z3}
 to work backwards and synthesize combinations of test inputs which can lead to a failure.
@@
 then including reordering possibilities in its SMT constraints.
-%rather than using explicit scheduling as SCT does.
+%rather than using explicit scheduling as MC does.
 This simultaneously exercises both input and schedule nondeterminism,
@@
 %
-CertiKOS \cite{certikos} extends this approach to include full concurrency and fine-grained locking
+\revision{Verve \cite{verve}
+is another fully-verified kernel which uses type safety, rather than virtual memory isolation,
+to provide strong stability and isolation guarantees.
+It uses a small core written in typed assembly language (TAL)~\cite{tal}
+to provide runtime services such as stack management, interrupt handling, and garbage collection.
+The remaining higher-level kernel services are written in C\# and compiled to TAL.
+%
+Unlike seL4, whose proofs relied on human-driven interactive theorem proving,
+Verve's verification comes from automated type- and invariant-checking,
+relying on pre/postcondition and loop invariant annotations for its TAL core only,
+thereby imposing much less burden on its programmers.
+seL4's microkernel nature allows it to run untrusted code, such as drivers, in the safety of virtual memory isolation,
+while Verve's type safety restricts what programs can be run at all.
+Both approaches accept the limitation of non-parallel, uniprocessor execution.
+}
+%
+CertiKOS \cite{certikos} extends seL4's approach to include full concurrency and fine-grained locking
 in the scope of verification,
@@
 while others might argue that for safety-critical kernel code you can't afford {\em not} to verify so thoroughly.
-%
+
 More recently, Hyperkernel \cite{hyperkernel} extended the xv6 educational kernel \cite{xv6}

### 2018-11-28 thesis/tm.tex
final pre-defense draft revisions
finish rerun htm1 bugfinding expt
dave's revisions to tm

@@
 
-Transactional memory (TM) is a concurrency primitive
+Transactional memory \revisionminor{(TM)~\cite{transactional-memory}} is a concurrency primitive
 by which programmers may attempt an arbitrary sequence of shared memory accesses,
@@
 such that no intermediate state modification is ever visible,
-or in the case of a conflict which would prevent such,
-discarded with an error code returned to allow the programmer to write a synchronized backup path.
+or, in the case of a conflict which would prevent such,
+discarded with an error code returned to allow the programmer to \revisionminor{invoke} a synchronized backup path.
 %
@@
 \begin{itemize}
-	\item  {\sf xbegin} begins a transaction,
+	\item  {\sf \revisionminor{begin}} begins a transaction,
 		staging any subsequent shared memory accesses in some temporary thread-/processor-local storage,
 		and checking for conflicts with the accesses of any other threads or CPUs.
+		\revisionminor{When the transaction is started, {\sf begin} returns a success code.}
 		If the transaction is unsuccessful, as described below,
@@
 		indicating the programmer should fall back to some other, possibly slower, synchronization method.
-	\item {\sf xend} ends a transaction,
+	\item {\sf \revisionminor{end}} ends a transaction,
 		attempting to commit all staged accesses to the shared memory atomically with respect to
@@
 		they are instead discarded and execution state reverts to the {\sf begin} with an error code as described above.
-	\item {\sf xabort} explicitly aborts a transaction,
+	\item {\sf \revisionminor{abort}} explicitly aborts a transaction,
 		regardless of any memory conflicts,
@@
 		Some implementations allow an arbitrary abort code to be specified
-		which will appear in {\sf begin}'s error code.
+		which will appear \revisionminor{as} {\sf begin}'s error code.
 \end{itemize}
@@
 refers to the {\tt xbegin}, {\tt xend}, and {\tt xabort} subset of TSX instructions,
-which of course correspond to {\sf begin}, {\tt end}, and {\tt abort} listed above,
+which of course correspond to {\sf begin}, {\sf end}, and {\sf abort} listed above,
 as well as {\tt xtest}, an instruction which returns whether or not the CPU is executing transactionally.
@@
 % https://software.intel.com/en-us/node/683688
-\cite{speculative-lock-elision,lock-elision}
+\cite{speculative-lock-elision,lock-elision}.
 In this thesis I focus on RTM, the more general (i.e., expressive (i.e., bug-prone)) interface,
@@
 I claim that MCing the simpler interleaving $\tiat{1}, \tjat{1-3}, \tiat{4-7}$
-is an equivalent verification as MCing the one above;
+is an equivalent verification \revisionminor{to} MCing the one above;
 in fact, this interleaving suffices to check
@@
 	and reordering them produces an equivalent interleaving,
-	satisfying case 2.
+	satisfying case~2.
 \end{proof}
@@
 		To race in $I$ they must be reorderable at instruction granularity,
-		at least one with a thread switch immediately before or after.
+		at least one with a thread switch immediately before or after
 		\cite{tsan,hybriddatarace}.
@@
 Transactional test programs should be ported to the Pebbles userspace if not already,
-then any use of compiler HTM intrinsics replaced with Landslide's interface%
+then any use of compiler HTM intrinsics replaced with Landslide's interface.%
 \footnote{Attempts to execute a real TSX instruction under Landslide, instead of using the custom interface,
@@
 as neither of its supported simulation platforms support the feature.
-}.
+}
 They should then be put in either {\tt 410user/progs/} or {\tt user/progs/}
@@
 		(always returning {\tt \_XABORT\_RETRY} as the failure code)
-		(\sect{\ref{sec:txn-failure}}),
+		(\revisionminor{\sect{\ref{sec:tm-design-failure}}}, \sect{\ref{sec:txn-failure}}),
 		and treatment of transactional regions during data-race analysis (\sect{\ref{sec:txn-datarace}}).
@@
 		This represents the possibility for a hardware transaction to fail
-		for reasons outside of the programmer's control such as system interrupts or cache eviction.
+		for reasons outside of the programmer's control, such as system interrupts or cache eviction.
 		If {\tt -S} is set they are initialized to empty,
@@
 		and execution of the transactional path immediately stops.
+		\revision{So called because doing so would trigger a mode switch,
+		which the specification allows to abort for any reason.%
+		\footnote{As far as I know, the reason for syscall aborts is both hardware- and kernel-dependent,
+		and in fact nondeterministic;
+		on my Haswell Core i7 running Linux 4.19, I observed a majority of {\tt \_XABORT\_CONFLICT}s,
+		a scant few {\tt \_XABORT\_CAPACITY}s,
+		and surprisingly many with no reason specified whatever;
+		however, not once did a transaction succeed with even a single {\tt gettid()} call
+		\cite{tsx-syscall}.
+		For simplicity of both implementation and client programs,
+		Landslide always distinguishes system call aborts with {\tt \_XABORT\_CAPACITY}.}
+		See below for discussion of other cache capacity concerns.}
 \end{itemize}
@@
 will pop off the {\tt xabort\_codes\_todo} queue
-when it's time to explore that preemption point in the usual depth-first manner%
+when it's time to explore that preemption point in the usual depth-first manner.%
 \footnote{The search order prioritizes abort codes before scheduling other threads
 at such preemption points,
-which is just an implementation detail, not theoretically necessary.}.
+which is just an implementation detail, not theoretically necessary.}
 The optional abort code is then passed through {\tt arbiter\_append\_choice()}/{\tt arbiter\_pop\_choice()}
@@
         \begin{tabular}{ll}
-		\multicolumn{2}{c}{Initially {\tt \ctype{int} foo, bar = \const{0};}} \\
+		\multicolumn{2}{c}{Initially {\tt \ctype{int} foo = \const{0}, bar = \const{0};}} \\
                 \\
@@
 and refuses to run it until the conflicting thread runs its transaction first,
-unless such would result in deadlock%
+unless such would result in deadlock.%
 \footnote{
@@
 	and confirm the need to still explicitly avoid them by abandoning the retry set in such cases.
-	% TODO put a commit # here
-}.
+	\revisionminor{Commit 02b70c8 implements this fix.}
+}
 %The {\tt landslide} repository's {\tt retry-sets} branch implements this feature,
@@
 for further discussion of variance see \cite{sigbovik-htm}.
-The number of interleavings in each state space are, of course, deterministic and do not vary across runs.
+The number of interleavings in each state space \revisionminor{is}, of course, deterministic and do not vary across runs.
 
@@
 		{\tt htm1}
-			& 2,1	&\cpu{  26.48}&\wtm{6.48}&\ints{5}& \cpu{*8.21}& \wtm{*5.78}& \ints{5	}& \ETA{12} \\
+			& 2,1	&\cpu{  32.73}&\wtm{8.02}&\ints{5}& \cpu{*9.24}& \wtm{*6.49}& \ints{5}& \ETA{12} \\
+			% 5 interleavings tested; 1 preemptions; job time 3s; pldi time 8015847; new-fixed pldi cputime 32726660
+			% 5 interleavings tested; 1 preemptions; job time 2s; pldi time 6492339; new-fixed pldi cputime 9242577 -M
 		(assertion)
-			& 2,2	&\cpu{  45.23}&\wtm{8.28}&\ints{9}& \cpu{*8.42}& \wtm{*5.97}& \ints{9	}& \ETA{102} \\
-			& 2,3	&\cpu{ 53.88}&\wtm{9.92}&\ints{17}& \cpu{*8.80}& \wtm{*6.31}& \ints{17	}& \ETA{819} \\
-			& 2,4	&\cpu{ 90.24}&\wtm{14.63}&\ints{33}& \cpu{*9.72}& \wtm{*7.19}& \ints{33	}& \ETA{6553} \\
+			& 2,2	&\cpu{  50.72}&\wtm{9.22}&\ints{9}& \cpu{*9.93}& \wtm{*7.03}& \ints{9}& \ETA{102} \\
+			% 9 interleavings tested; 1 preemptions; job time 5s; pldi time 9220760; new-fixed pldi cputime 50719368
+			% 9 interleavings tested; 1 preemptions; job time 3s; pldi time 7028111; new-fixed pldi cputime 9933516 -M
+			& 2,3	&\cpu{ 88.35}&\wtm{14.63}&\ints{17}& \cpu{*10.03}& \wtm{*7.12}& \ints{17}& \ETA{819} \\
+			% 17 interleavings tested; 1 preemptions; job time 7s; pldi time 14632911; new-fixed pldi cputime 88346856
+			% 17 interleavings tested; 1 preemptions; job time 3s; pldi time 7119866; new-fixed pldi cputime 10025347 -M
+			& 2,4	&\cpu{108.31}&\wtm{17.44}&\ints{33}& \cpu{*11.24}& \wtm{*8.29}& \ints{33}& \ETA{6553} \\
+			% 33 interleavings tested; 1 preemptions; job time 9s; pldi time 17442317; new-fixed pldi cputime 108309778
+			% 33 interleavings tested; 1 preemptions; job time 3s; pldi time 8294161; new-fixed pldi cputime 11235910 -M
 		\cline{2-9}
-			& 3,1	&\cpu{ 39.75}&\wtm{8.21}&\ints{5}& \cpu{*8.15}& \wtm{*5.71}& \ints{5	}& \ETA{76} \\
-			& 3,2	&\cpu{ 46.91}&\wtm{8.40}&\ints{9}& \cpu{*8.37}& \wtm{*5.91}& \ints{9	}& \ETA{3686} \\
-			& 3,3	&\cpu{ 41.39}&\wtm{8.50}&\ints{17}& \cpu{*8.71}& \wtm{*6.26}& \ints{17	}& \ETA{176947} \\
-			& 3,4	&\cpu{ 92.98}&\wtm{15.19}&\ints{33}& \cpu{*10.36}& \wtm{*7.15}& \ints{33	}& \ETA{8493465} \\
+			& 3,1	&\cpu{ 60.63}&\wtm{11.63}&\ints{5}& \cpu{*9.63}& \wtm{*6.70}& \ints{5}& \ETA{76} \\
+			% 5 interleavings tested; 1 preemptions; job time 5s; pldi time 11631147; new-fixed pldi cputime 60628270
+			% 5 interleavings tested; 1 preemptions; job time 2s; pldi time 6695336; new-fixed pldi cputime 9627076 -M
+			& 3,2	&\cpu{ 78.69}&\wtm{13.17}&\ints{9}& \cpu{*9.43}& \wtm{*6.37}& \ints{9}& \ETA{3686} \\
+			% 9 interleavings tested; 1 preemptions; job time 6s; pldi time 13173912; new-fixed pldi cputime 78690603
+			% 9 interleavings tested; 1 preemptions; job time 2s; pldi time 6370046; new-fixed pldi cputime 9427268 -M
+			& 3,3	&\cpu{ 70.50}&\wtm{13.05}&\ints{17}& \cpu{*10.43}& \wtm{*6.96}& \ints{17}& \ETA{176947} \\
+			% 17 interleavings tested; 1 preemptions; job time 6s; pldi time 13045016; new-fixed pldi cputime 70504465
+			% 17 interleavings tested; 1 preemptions; job time 3s; pldi time 6962930; new-fixed pldi cputime 10427482 -M
+			& 3,4	&\cpu{ 70.69}&\wtm{12.96}&\ints{33}& \cpu{*11.74}& \wtm{*8.71}& \ints{33}& \ETA{8493465} \\
+			% 33 interleavings tested; 1 preemptions; job time 7s; pldi time 12964843; new-fixed pldi cputime 70692866
+			% 33 interleavings tested; 1 preemptions; job time 4s; pldi time 8710265; new-fixed pldi cputime 11743862 -M
 		\cline{2-9}
-			& 4,1	&\cpu{ 45.13}&\wtm{8.56}&\ints{5}& \cpu{*7.77}& \wtm{*5.30}& \ints{5	}& \ETA{460} \\
-			& 4,2	&\cpu{64.99}&\wtm{11.33}&\ints{9}& \cpu{*8.32}& \wtm{*5.59}& \ints{9	}& \ETA{132710} \\
-			\multicolumn{9}{c}{everything above this line needs rerun for dpor-preferred tid + thr\_bottom1} \\
+			& 4,1	&\cpu{ 51.83}&\wtm{9.85}&\ints{5}& \cpu{*9.56}& \wtm{*6.63}& \ints{5}& \ETA{460} \\
+			% 5 interleavings tested; 1 preemptions; job time 5s; pldi time 9851604; new-fixed pldi cputime 51833724
+			% 5 interleavings tested; 1 preemptions; job time 2s; pldi time 6627275; new-fixed pldi cputime 9560320 -M
+			& 4,2	&\cpu{ 44.83}&\wtm{8.94}&\ints{9}& \cpu{*9.83}& \wtm{*6.84}& \ints{9}& \ETA{132710} \\
+			% 9 interleavings tested; 1 preemptions; job time 4s; pldi time 8936964; new-fixed pldi cputime 44826654
+			% 9 interleavings tested; 1 preemptions; job time 2s; pldi time 6835025; new-fixed pldi cputime 9830626 -M
 		\hline
@@
 Nevertheless, pushing $K$ and $N$ higher is obviously desirable,
-even if it means applying reductions that require human intuition to trust are sound.
+even if it means applying reductions that require human intuition to trust \revisionminor{their soundness}. %are sound.
 Moreover, much as verifying {\tt htm2}'s soundness is a positive result,
@@
 
-\begin{figure}[h]
+\begin{figure}[t]
 	\begin{center}
@@
 baseline DPOR tests reorderings of one thread's transaction both with the other's failure path,
-and with the other's data race therein%
+and with the other's data race therein.%
 \footnote{
@@
 even in baseline DPOR, with no need for retry sets.
-}.
+}
 Retry sets on the other hand identify and skip that equivalence
@@
 			& 2,3 & \cpu{12946.21} & \ints{244691} \\
+			& 2,4 & \ETAdag{qqq} & \ETAdag{qqq} \\
 			& 3,1 & \cpu{637.81} & \ints{12707} \\
-			& 3,2 & \cpu{qqq} & \ints{qqq} \\
+			& 3,2 & \ETAdag{102d 13h} & \ETAdag{13220616} \\
+			% [JOB 3] progress: 560711/13220616 brs (4.241187%), ETA 102d 2h 56m 44s (elapsed 9h 59m 55s)
 			& 4,1 & \ETAdag{33h 33m} & \ETAdag{1778661} \\

### 2018-11-28 thesis/overview.tex
dave revisions on landslide chapater
fix a bunch of citations; convert bib to alpha

@@
 The other three (a stack-crawling backtrace utility, a bare-metal game with device drivers, and a small extension to the P3 kernel) are not of concern in this thesis.
-The course's prerequisite is 15-213, Introduction to Computer Systems \cite{sigcse01:CSaPP}.
+The course's prerequisite is 15-213, Introduction to Computer Systems \cite{csapp,sigcse01:CSaPP}.
 Both P2 and P3 are built using the {\em Pebbles} system call specification, outlined in Table~\ref{tab:syscalls}.
@@
 The thread library project \cite{thrlib} has two main components: implementing concurrency primitives, and implementing thread lifecycle and management routines.
-\revision{Students implement this project on top of a reference Pebbles kernel implementation (named Pathos~\cite{pathos}),
+\revision{Students implement it on top of a reference Pebbles kernel implementation (named Pathos~\cite{pathos}),
 which is distributed as an obfuscated binary to protect the subsequent challenges of P3.}
@@
 		See \sect{\ref{sec:pebbles}}.
+	\item \revision{{\bf P3.}
+		Operating systems class project, occuring after {\em P2},
+		in which students implement their own {\em Pebbles} kernel from the ground up.}
 	\item \revision{{\bf Pebbles.}

### 2018-11-29 thesis/thesis.tex
fix a bunch of citations; convert bib to alpha
revise thesis front matter

@@
 %\keywords{landslide terminal, baggage claim, ground transportation, ticketing}
-\keywords{concurrency, debugging, verification, model checking, data races, education, transactional memory}
+\keywords{concurrency, testing, debugging, verification, model checking, data races, education, transactional memory}
 
@@
 \begin{abstract}
-Concurrent programming presents a challenge to students and experts alike because of the complexity of multithreaded interactions and the difficulty to reproduce and reason about bugs.
-Stateless model checking is a concurrency testing approach which forces a program to interleave its threads in many different ways, checking for bugs each time.
+Concurrent programming presents a challenge to students and experts alike because of the complexity of multithreaded interactions and the difficulty \revisionminor{of reproducing and reasoning} about bugs.
+Stateless model checking is a
+%concurrency
+testing approach which forces a program to interleave its threads in many different ways, checking for bugs each time.
 This technique is powerful, in principle capable of finding any nondeterministic bug in finite time,
@@
 %  header and the actual list of references}
-\bibliographystyle{plainnat} % plain n'at
+\bibliographystyle{alpha}
 \bibliography{citations}

### 2018-11-30 thesis/landslide.tex
いちおう acknowledge friends

@@
 		If {\tt CURRENT\_THREAD\_\allowbreak{}LIVES\_ON\_RQ=0} (described above), this should be invoked (among other times) during context switch with the TID of the thread about to start running.
-		Alternatively (thanks sully), even for a kernel which takes the current thread off its literal runqueue,
+		Alternatively, %(thanks sully),
+		even for a kernel which takes the current thread off its literal runqueue,
 		the annotator may use these two calls to indicate the ``abstract runqueue'' which includes the current thread as well,

### 2018-11-30 thesis/thesis.tex
いちおう acknowledge friends

@@
 \begin{acknowledgments}
-I honor here the many people who supported me during these 7 years and made this journey possible.
+\revision{I honor here the many people who supported me during these 7 years and made this journey possible.
 
@@
 and to seek guidance from industry people with relevant experience,
-despite my often stubborn refusal to leave my little comfort zone.
+despite my often stubborn refusal to leave my own little comfort zone.
 So, even when it took me sending 10+ email reminders to get your attention,
@@
 Professor David A. Eckhardt
-provided unending guidance on education, writing, career, life, you name it,
-helped revise each semester's iteration of the recruiting lecture with a keen eye for the student mind,
-contributed substantially to the design of the survey,
-and supplied constant enthusiasm and encouragement that Landslide was a boon to 15-410.
-I can't express my gratitude to Dave strongly enough,
-even if it took until just this week for half of the Landslide-friendly tests to be included in the official P2 test suite.
+provided unending guidance on education, writing, career, and life in general, %life, you name it,
+helped revise each semester's %iteration of the
+recruiting lecture with a keen eye for the student mind,
+contributed several clever survey questions, %substantially to the design of the survey,
+and supplied constant %enthusiasm and
+encouragement about Landslide's value to 15-410, %that Landslide was a boon to 15-410,
+not to mention the red-ink-encrusted printed draft of this document I received immediately after my defense.
+My debt of thanks to Dave is of such magnitude that I may never hope to pay it back, but perhaps forward instead.
+%I can't express my gratitude to Dave strongly enough,
+%even if it took until just this week for half of the Landslide-friendly tests to be included in the official P2 test suite.
 %even though three of the Landslide-friendly tests never made it upstream into the official P2 test suite.
 
-Ji\v{r}\'{i} \v{S}im\v{s}a, another student of Garth's,
+Ji\v{r}\'{i} \v{S}im\v{s}a, another former student of Garth's,
 first sparked my interest in the field,
@@
 whom I thank here in approximate chapter order.
-Michael J. Sullivan and Nathaniel Filardo were Landslide's first users besides myself back in its prototype days.
-Joshua Wise, Miriam Zimmerman, and Glenn Willen helped revise my explanation of DPOR in Chapter~\ref{chap:background}.
+Michael J. Sullivan and Nathaniel Wesley Filardo were Landslide's first users besides myself back in its prototype days.
+Joshua Wise, Miriam Zimmerman, and Glenn Willen helped revise my explanation of DPOR in Chapter~\ref{chap:landslide}.
 Professor Brandon Lucia provided valuable comments and revisions to Chapter~\ref{chap:quicksand}'s content
@@
 
-The community of 15-410 TAs,
-having surrounded me pretty much since starting undergrad,
+The community of 15-410 TAs
+%having surrounded me pretty much since starting undergrad,
 nurtured my love of teaching
 well before taking the class myself and well after graduation.
-As skilled as the instructor is,
+%As skilled as the instructor is,
 15-410 could not be the world's best operating systems class
-without TAs who value thorough grading, the Socratic method,
+without %TAs
+course staff %(staves)
+who value thorough grading, the Socratic method,
 and empathy to struggling students as much
 from year to year as the crew I have known.
-It was a blessing to serve, and later do research, among you all.
-Thanks also to Wind River for generously providing CMU with Simics educational licenses.
-I am also grateful to Adam Blank and Miriam Zimmerman
+It was a blessing to serve, and later to do research, among you all.
+Thanks to Wind River for generously
+providing CMU with Simics educational licenses.
+%I am also grateful
+Thanks also to Adam Blank and Miriam Zimmerman
 for founding and organizing the class Great Practical Ideas,
@@
 pushing me to refine my presentation skills and to network with strangers (now familiar faces) from industry.
-I also thank those industry attendees from each of the PDL Consortium member companies
-for their interest, even in more theoretical work,
-and the enlightening conversations resulting therefrom.
-I thank everyone outside my immediate research circle who,
-whether at the retreat or otherwise,
+I also thank those %industry
+attendees from each of the PDL Consortium member companies
+for their interest %, even in more theoretical work,
+and for the enlightening conversations resulting therefrom.
+%Thanks to the anonymous OOPSLA reviewer who went the extra mile to read my supplementary proofs
+%and urged me to improve the formalism to be more foundational
+%(the lack of any stronger criticism a silent vote of confidence for their overall correctness).
+Thanks to Niko Matsakis for the mentorship and collaboration during my two summers as a Rust research intern at Mozilla,
+and to the rest of the Rust community as well for their extremely important work
+bringing better concurrent programs to the world from a different angle.
+I thank everyone outside my %immediate
+research circle who
+%whether at the retreat or otherwise,
 insisted how cool my research was for directly helping students --
@@
 
-I am grateful to the ThursDz Council,
+%I am grateful to
+The ThursDz Council,
 my group of now-former grad student, Shadyside-bound (in spirit if not in person) restaurant-hoppers,
-for both friendship and mentorship over the years.
+surrounded me with both friendship and mentorship over the years.
 In particular,
@@
 provided bottomless emotional support throughout my education.
-I am grateful beyond words for the privileged life they provided
+I appreciate beyond words for the privileged life they provided
 that allowed me the opportunity to pursue higher education.
 
-% [ ] csd bgames
-% [ ] undergrad frenz
-% [ ] new frenz
-% [ ] family, incl louie
+Looking back on my 11 years at CMU,
+I am especially grateful for the friendships made early on %in my undergrad years
+that continued well past first graduation.
+From those who stayed in Pittsburgh %with me
+to those who %moved away but
+kept in touch,
+from my regular tabletop, board, and/or dance gaming partners
+to those who %inexhaustibly
+welcomed me with %company and
+a couch to sleep on during travel,
+I thank
+%-- and in any order other than alphabetical here would be far too cruel --
+Adam Blank,
+Alan Vangpat,
+Alex Yuschik,
+Anand Subramanian,
+Andrew Krieger,
+Car Bauer,
+Carolyn Sawyer,
+Elizabeth Kemp,
+Elly Fong-Jones,
+Emma Cating-Subramanian,
+Eric Faust,
+Gabe Routh,
+Greg Hanneman,
+Hazel Vird\'{o},
+Jack Ferris,
+Jake Lengyel,
+Jason Deng,
+Josh Keller,
+Joshua Wise,
+Josiah Boning,
+Julia Tuttle,
+Kellie Medlin,
+Laura Abbott,
+Margaret Meyerhofer,
+Michael ``rntz'' Arntzenius,
+Michael J. ``Sully'' Sullivan,
+Miriam Zimmerman,
+Naomi Saphra,
+Nathaniel Wesley Filardo,
+Philip ``8'' Gianfortoni,
+Rhett Lauffenburger,
+Richard Ha,
+Rory Glenn,
+Ryan Pearl,
+and
+Todd Eisenberger
+for their wonderful friendship.
+I am equally grateful to the new friends I made during my graduate years,
+whether through board/card/dance gaming, tea drinking, %and/or
+movie watching,
+and/or as more permanent Pittsburgh residents that I am extremely torn up about moving away from:
+%leaving behind (for now at least):
+Alexandra Lee Falk,
+Alexis Dyer,
+Barbara Jensen,
+Brendan McShane,
+Brian Saghy,
+Carlo Angiuli,
+Cassie Orr,
+Dan Guzek,
+Danny Gratzer,
+David Renshaw,
+Emily Forney,
+Evan Cavallo,
+Gabriele Farina,
+Grant Wu,
+Jenny Lin,
+Karl Schulz,
+Kristy Gardner,
+Lea Albaugh,
+Maija Mednieks,
+Marlena Abraham,
+Matt Stanec,
+Matthew Maurer,
+Sarah R. Allen,
+Sol Boucher,
+Stefan Muller,
+Zachary Waldman,
+Ziv Scully,
+and the rest of ThursDz already named previously.
+And to anyone who may read this list
+and feel even slightly disappointed that their name was left off:
+I owe you a nice dinner. Get in touch.
+
+日本語を
+一緒に
+勉強したり、
+会話したり、
+色々教えてくださったりした
+学習者さんたちとネイティブスピーカーさんたちへ、
+心から感謝を申し上げます。
+外国語を学ぶのは一生の二番目の達成感かもしれません。
 
@@
 for nearly as long as I've been a grad student.
-Netrunner is a tabletop strategy deckbuilding card game
+Netrunner is a deckbuilding strategy card game
 that depicts cyberspace hacking battles (completely different from real-life computer security)
-alongside dystopian social issues (some of which seem oddly more familiar with each passing year).
-Its characters have a diversity of ethnicity, gender identity, and sexual orientation
+alongside dystopian social issues (becoming strangely more familiar with each passing year).
+Its characters have a diversity of ethnicity and gender identity %and sexual orientation
 unparalleled in any other game I know of,
 and the community that has grown around it reflects those values.
-I thank in particular the other Stimhack moderators,
+I thank especially the other Stimhack moderators,
 the members of {\sf \#rainbow-coalition},
 and the local Pittsburgh crew
-for their support and encouragement
-and friendships very real and deep, even when forged only in text,
+for their support and encouragement and
+friendships very real and deep, even when forged only in text,
 %\footnote{and Slack emojis},
@@
 %It'd be an injustice to list only a few names here and omit others, but you know who you are.
+}
 

### 2018-11-30 thesis/education.tex
David A. Eckhardt
dodge latex changebar bug
fix a bunch of citations; convert bib to alpha
add short versions of fig/tab capts for frontmatter list

@@
 %in semesters Fall 2015 to Spring 2018 and in Spring 2018 alone, respectively,
-%taught by David Eckhardt and Timothy Zhu, respectively.
+%taught by David A. Eckhardt and Timothy Zhu, respectively.
 CMU's 15-410 in semesters Fall 2015 to Spring 2018,
-taught by David Eckhardt,
+taught by David A. Eckhardt,
 and in PSU's \psuos in Spring 2018,
@@
 	\end{center}
-	\caption{Participation rates across semesters,
+	\caption[Participation rates across semesters.]
+		{Participation rates across semesters,
 	among students who submitted thread libraries (CMU, PSU) or kernels (U. Chicago).}
@@
 	\end{center}
-	\caption{Landslide's bug-finding performance across all semesters of the CMU 15-410 study.
+	\caption[Landslide's bug-finding performance across all semesters.]
+		{Landslide's bug-finding performance across all semesters of the CMU 15-410 study.
 	$\le$ marks possible overcounts on account of unidentified false positives;
@@
 	\end{center}
-	\caption{Distribution of project grades among Landslide users and non-users.
+	\caption[Distribution of project grades among Landslide users and non-users.]
+		{Distribution of project grades among Landslide users and non-users.
 	Study participants (S'15 to S'18) are the experimental group;
@@
 	\end{center}
-	\caption{
+	\caption[Correlation of student Landslide use with solving certain bugs.]
+		{
 		Correlation of student Landslide use with solving certain bugs in their final submission during Fall 2017 and Spring 2018 semesters.
@@
 	\end{center}
-	\caption{Student survey responses.
+	\caption[Student survey responses.]
+		{Student survey responses.
 	SD/D/N/A/SA stands for
@@
 	\item Happy with it as-is [\revisionminor{reported by three} students]
-	\item [\revisionminor{Seven non-respondents}]
+		% wow, latex -- if you put [] outside of the revision, everything is awful forever
+	\item \revisionminor{[Seven non-respondents]}
 \end{enumerate}
@@
 %Considering that CMU's 15-410 demands substantial prerequisite concurrency experience
-%(15-213 \cite{sigcse01:CSaPP} or equivalent)
+%(15-213 \cite{csapp} or equivalent)
 %whereas PSU's \psuos is more towards the average between CMU's 15-213 and 15-410

### 2018-11-30 thesis/landslide.tex
add short versions of fig/tab capts for frontmatter list

@@
 	\end{center}
-	\caption{Example demonstrating the need for {\tt misbehave} to force the kernel
+	\caption[Example demonstrating the need for {\tt misbehave}.]
+		{Example demonstrating the need for {\tt misbehave} to force the kernel
 	to yield during {\tt thread\_fork}.}
@@
 	\end{center}
-	\caption{Motivating example for the sleep sets optimization.
+	\caption[Motivating example for the sleep sets optimization.]
+		{Motivating example for the sleep sets optimization.
 	Three of Figure~\ref{fig:dpor-example-2}'s interleavings are highlighted,
@@
 	}
-	\caption{Straightforward exploration ordering}
+	\caption{Straightforward exploration ordering.}
 	\label{alg:not-icb}
@@
 	}
-	\caption{ICB exploration ordering}
+	\caption{ICB exploration ordering.}
 	\label{alg:icb}

### 2018-11-30 thesis/overview.tex
add short versions of fig/tab capts for frontmatter list

@@
 	\end{tabular}
-	\caption{Example concurrent program in which simultaneous accesses to shared state may interleave to produce unexpected results.}
+	\caption[Example concurrent program.]{Example concurrent program in which simultaneous accesses to shared state may interleave to produce unexpected results.}
 	\label{fig:concurrency-bug}
@@
 	\end{tabular}
-	\caption{Using a locking primitive to protect accesses to shared state.}
+	\caption[Using a locking primitive to protect accesses to shared state.]
+		{Using a locking primitive to protect accesses to shared state.}
 	\label{fig:mutex}
@@
 	\end{center}
-	\caption{The example {\tt count} routine from Figure~\ref{fig:mutex}, rewritten to use HTM.
+	\caption[The example {\tt count} routine from Figure~\ref{fig:mutex}, rewritten to use HTM.]
+		{The example {\tt count} routine from Figure~\ref{fig:mutex}, rewritten to use HTM.
 		If the transaction in the top branch aborts,
@@
 	\end{tabular}
-	\caption{Visualization of interleaving state space for the program in Figure~\ref{fig:concurrency-bug}.
+	\caption[Visualization of interleaving state space for the program in Figure~\ref{fig:concurrency-bug}.]
+		{Visualization of interleaving state space for the program in Figure~\ref{fig:concurrency-bug}.
 	Thread~1 is represented by purple ovals, thread~2 by yellow squares, and time flows from top to bottom.
@@
 	\end{center}
-	\caption{DPOR identifies independent transitions by different threads which can commute
+	\caption[DPOR identifies independent transitions by different threads.]
+		{DPOR identifies independent transitions by different threads which can commute
 	without affecting program behaviour.
@@
 	\end{center}
-\caption{{Data-race analyses may be prone to either {\em false negatives} or {\em false positives}.
+	\caption[Data-race analyses may be prone to either {\em false negatives} or {\em false positives}.]
+		{{Data-race analyses may be prone to either {\em false negatives} or {\em false positives}.
 	Applying Happens-Before to program (a) will miss the potential race possible between A1/A2 in an alternate interleaving,
@@
         \end{tabular}
-        \caption{The Pebbles specifcation defines 25 system calls. Students are not required to implement ones marked with an asterisk (*), though the reference kernel provides them. }
+	\caption[The Pebbles specifcation defines 25 system calls.]
+		{The Pebbles specifcation defines 25 system calls. Students are not required to implement ones marked with an asterisk (*), though the reference kernel provides them.}
         \label{tab:syscalls}

### 2018-11-30 thesis/tm.tex
fix a bunch of citations; convert bib to alpha
add short versions of fig/tab capts for frontmatter list

@@
 	\end{center}
-	\caption{Variant of the program in Figure~\ref{fig:htm-example-reproduced},
+	\caption[Variant of the program in Figure~\ref{fig:htm-example-reproduced}.]
+		{Variant of the program in Figure~\ref{fig:htm-example-reproduced},
 		with additional synchronization to protect the failure path from the transactional path.
@@
 	\end{center}
-	\caption{Landslide's bug-finding performance on various test configurations.
+	\caption[Landslide's bug-finding performance on various test configurations.]
+		{Landslide's bug-finding performance on various test configurations.
 		Iterative Deepening (\sect{\ref{sec:quicksand-id}}), optimized for fast bug-finding,
@@
 	%\caption{mario man had a very large house. or a castle.}
-	\caption{Unmodified code from {\tt htmavl.hpp} showing the previously-unknown bug
+	\caption[Code from {\tt htmavl.hpp} showing the segfault bug Landslide found.]
+		{Unmodified code from {\tt htmavl.hpp} showing the previously-unknown segfault bug
 		Landslide found in {\tt avl\_insert}.
@@
 	\end{center}
-	\caption{Code from {\tt spinlock-rtm.c}, modified only to remove unrelated logic for brevity,
+	\caption[Code from {\tt spinlock-rtm.c} %, modified only to remove unrelated logic for brevity,
+		showing the performance bug Landslide found.]
+		{Code from {\tt spinlock-rtm.c}, modified only to remove unrelated logic for brevity,
 		showing the performance bug Landslide found in {\tt lock\_fast}({\tt spinlock}).
@@
 	\end{center}
-	\caption{Transactional tests verified (or not) by Landslide.
+	\caption[Transactional tests verified (or not) by Landslide.]
+		{Transactional tests verified (or not) by Landslide.
 		Run with {\tt -M -X}, plus any additional reduction options listed.
@@
 	\end{center}
-	\caption{Number of concurrency events per iteration of each test case.
+	\caption[Number of concurrency events per iteration of each test case.]
+		{Number of concurrency events per iteration of each test case.
 	Note that no test used any synchronization besides mutexes % ;
@@
 	\end{center}
-	\caption{Visualization of retry set state space reduction.
+	\caption[Number of concurrency events per iteration of each test case.]
+		{Visualization of retry set state space reduction.
 	On both {\tt counter}(2,1) and {\tt fig63}(2,1), baseline DPOR tests % oops
@@
 	\end{center}
-	\caption{Continuation of Table~\ref{tab:tm-verifs},
-		demonstrating abstraction reduction \cite{dbug-phdthesis}
+	\caption[Continuation of Table~\ref{tab:tm-verifs}, demonstrating abstraction reduction.]
+		{Continuation of Table~\ref{tab:tm-verifs},
+		% XXX: if you put this citation, it will cause "(document)" to appear in the list of back-references
+		% in the corresponding bibliography entry, rather than an appropriate section number
+		demonstrating abstraction reduction %\cite{dbug-phdthesis}
 		on the {\tt avl\_fixed} and {\tt map\_basic} tests

### 2018-11-30 thesis/overview.tex
relaxed memory in glossary and other overview todos

@@
 
-% TODO: Talk about the hurdle (both for p2 and p3).
+% haha-never: Talk about the hurdle (both for p2 and p3).
 
@@
 		a priority scheduling algorithm, and a multi-level feedback queue scheduler.
-		% TODO: Later. Talk about how concurrency testing can only test certain parts of this crap.
+		% haha-never: Later. Talk about how concurrency testing can only test certain parts of this crap.
 	\item {\em Userprog}: Provided with rudimentary virtual memory and ELF loader implementations, students must implement argument passing and several system calls associated with userspace programs, including {\tt exec}, {\tt exit}, {\tt wait}, and file descriptor management.
@@
 		This is distinct from language-level {\em data races},
-		which may result in bugs by being reordered by the compiler or hardware;
-		% TODO: put a section reference for weak memory
-		% mb have to write a section on it here
-		however, this thesis checks programs at the executable level and assumes sequential consistency,
+		which may result in bugs by being reordered by the compiler or hardware (see {\em relaxed memory});
+		however, this thesis checks programs at the executable level and assumes {\em sequential consistency},
 		so all data races that still result in bugs (e.g., unprotected {\tt x++})
@@
 		in which students implement their own {\em Pebbles} kernel from the ground up.}
+	\item \revision{{\bf Partial store order.} See {\em relaxed memory.}}
 	\item \revision{{\bf Pebbles.}
@@
 		Often called {\em preemption point} for short, with context hopefully making the distinction clear.
+	\item \revision{{\bf PSO.} {\em Partial store order}. See {\em relaxed memory}.}
 	\item {\bf Pure happens-before.}
@@
 		Recently, many systems research papers have used the term to refer to {\em data races} in particular.%
-		\footnote{In the author's personal experience, readers of their M.S. thesis \cite{landslide}
+		\footnote{In the author's personal experience, many readers of their M.S. thesis \cite{landslide}
 		have misinterpreted its title to believe % oops
@@
 		such that all but one may be skipped over.
+	\item \revision{{\bf Relaxed memory.}
+		Also known as {\em weak memory}.
+		A family of multiprocessor execution semantics
+		that allows reads and/or writes of shared memory to become visible to different CPUs in different sequences.
+		Under {\em total store order} (TSO),
+		implemented for example on x86,
+		one CPU's writes may appear reordered after its subsequent loads, from the perspective of another CPU.
+		Under {\em partial store order} (PSO),
+		either writes and/or reads may appear reordered after other subsequent writes and/or reads.
+		On ARM's implementation of PSO, all such combinations are possible.
+		Safe concurrent programming on such architectures requires careful use
+		of acquire and/or release memory barriers
+		\cite{sully-thesis}.
+		By contrast, {\em sequential consistency} (SC) permits no such reorderings.
+		% TODO: check that quicksand discussion shouldn't perhaps get a more subsection'ed sect reference
+		For further discussion see \sect{\ref{sec:quicksand-discussion}} and \sect{\ref{sec:tm-warpzone-relaxed}}.}
 	\item {\bf Schedule.}
@@
 		which overrides the decision of the kernel scheduler.
+	\item {\bf Sequential consistency.} See {\em relaxed memory}.
 	\item {\bf Soundness.}
@@
 		and no intermediate state is a preemption point as well.
+	\item \revision{{\bf Total store order.} See {\em relaxed memory.}}
+	\item \revision{{\bf TSO.} {\em Total store order}. See {\em relaxed memory.}}
 	\item {\bf Userspace.}
@@
 		the {\em test case} is capable of exposing in the first place.
+	\item \revision{{\bf Weak memory.}
+		See {\em relaxed memory.}}
 \end{enumerate}

### 2018-11-30 thesis/thesis.tex
update acks

@@
 From those who stayed in Pittsburgh %with me
-to those who %moved away but
+to those who moved away but
 kept in touch,
 from my regular tabletop, board, and/or dance gaming partners
-to those who %inexhaustibly
-welcomed me with %company and
+to those who inexhaustibly
+welcomed me with company and
 a couch to sleep on during travel,
@@
 Julia Tuttle,
+Kartik Subramanian,
 Kellie Medlin,
@@
 Ryan Pearl,
+Todd Eisenberger,
 and
-Todd Eisenberger
+Zachary McCord
 for their wonderful friendship.
@@
 色々教えてくださったりした
-学習者さんたちとネイティブスピーカーさんたちへ、
+仲間の学習者さんたちと
+ネイティブスピーカーさんたちへ、
 心から感謝を申し上げます。
-外国語を学ぶのは一生の二番目の達成感かもしれません。
+外国語を学ぶのは
+一生の二番
+%充実感
+達成感
+だと考えます。
 
 Lastly,
-the Android: Netrunner community has been such an important part of my life
-for nearly as long as I've been a grad student.
+the Android: Netrunner community has become such an important part of my life
+over the last half decade or so.
+%for the better part of my time as a grad student.
 Netrunner is a deckbuilding strategy card game
@@
 and the local Pittsburgh crew
-for their support and encouragement and
-friendships very real and deep, even when forged only in text,
-%\footnote{and Slack emojis},
-that will long outlive the silly card game that brought us together.
-%It'd be an injustice to list only a few names here and omit others, but you know who you are.
+for their support, encouragement,
+and mutual exchange of inclusive values in these turbulent yet hopeful times.
+%and
+%friendships very real and deep, even when forged only in text,
+%%\footnote{and Slack emojis},
+%that will long outlive the silly card game that brought us together.
+%%It'd be an injustice to list only a few names here and omit others, but you know who you are.
 }
@@
 \tableofcontents
-% TODO: figure out how to make latex abbreviate these
 \listoffigures

### 2018-12-01 thesis/landslide.tex
fix citations to be colorful insted of having ugly borders

@@
 		or to issue preemptions first and then try continuing the current thread later (1).
-		0 tends to produce shorter preemption traces while 1 tends to find bugs faster (\cite{landslide} \S{}8.7.1).
+		0 tends to produce shorter preemption traces while 1 tends to find bugs faster (\cite[\S{}8.7.1]{landslide}).
 		Not compatible with ICB.
@@
 		Set this option to 0 to support the former kernel type or 1 to support the latter.%
-		\footnote{This option replaces the deprecated {\tt kern\_current\_extra\_runnable()} annotation from {\tt student.c} described in \cite{landslide} \S{}6.2.3.}
+		\footnote{This option replaces the deprecated {\tt kern\_current\_extra\_runnable()} annotation from {\tt student.c} described in \cite[\S{}6.2.3]{landslide}.}
 		\revision{Whether or not a kernel requires this annotation could be auto-detected in future work.}
@@
 {\tt check\_withins()} implements the allow/denylist behaviour for the former two of those commands
-(in a similar manner to the Preemption Sealing technique described in prior work \cite{sealing}).
+(in a similar manner to \revisionminor{prior work's Preemption Sealing} \cite{sealing}).
 It invokes the stack tracer (\sect{\ref{sec:landslide-stacktrace}})

### 2018-12-01 thesis/education.tex
more assorted revisions

@@
 %
-Ultimately, only the same-semester comparisons were significant.
+Ultimately, only the same-semester comparisons,
+\revisionminor{which do not address that alternative hypothesis,
+were significant,
+while the cross-semester comparisons,
+which attempted to do so,
+failed to confidently reject even the null hypothesis.}
 However, the distribution difference being smaller in P3 than in P2

### 2018-12-01 thesis/tm.tex
more assorted revisions

@@
 Note that the C-style {\tt x++} operations, when compiled into assembly, %\cite{printy},
-become multiple memory accesses which can be interleaved with other threads.
-\[
+become multiple memory accesses which can be interleaved with other threads,
+\revisionminor{as shown below in Figure~\ref{fig:htm-example-assemblified}.}
+
+\begin{figure}[h]
+	\begin{center}
 		\begin{tabular}{ll}
-	$2a$ & \texttt{temp <- x;} \\
-	$2b$ & \texttt{temp <- temp + 1;} \\
-	$2c$ & \texttt{x <- temp;} \\
+			1 & \texttt{\flow{if} ((status = \call{\_xbegin}()) == \const{\_XBEGIN\_STARTED}) \{} \\
+			%2 & \texttt{~~~~x++;} \\
+			$2a$ & \texttt{~~~~temp <- x;} \\
+			$2b$ & \texttt{~~~~temp <- temp + 1;} \\
+			$2c$ & \texttt{~~~~x <- temp;} \\
+			3 & \texttt{~~~~\call{\_xend}();} \\
+			4 & \texttt{\} \flow{else} \{} \\
+			5 & \texttt{~~~~\call{mutex\_lock}(\&m);} \\
+			$6a$ & \texttt{~~~~temp <- x;} \\
+			$6b$ & \texttt{~~~~temp <- temp + 1;} \\
+			$6c$ & \texttt{~~~~x <- temp;} \\
+			7 & \texttt{~~~~\call{mutex\_unlock}(\&m);} \\
+			8 & \texttt{\}} \\
 		\end{tabular}
-\]
+	\end{center}
+	\caption{Figure~\ref{fig:htm-example-reproduced} with non-atomic operations shown as pseudo-assembly.}
+	\label{fig:htm-example-assemblified}
+\end{figure}
 
@@
 
+\begin{figure}[t]
+	\begin{center}
+		\begin{tabular}{ll}
+			\hilight{lavender}{\bf Thread 1} &
+			\hilight{seafoam}{\bf Thread 2} \\
+			\hline
+			\texttt{status = \call{\_xbegin}()} & \\
+			\texttt{temp <- x;} & \\
+			\texttt{temp <- temp + 1;} & \\
+			& \texttt{status = \call{\_xbegin}()} \\
+			& \texttt{x++;} \\
+			& \texttt{\call{\_xend}();} \\
+			\texttt{x <- temp;} & \\
+			\texttt{\call{\_xend}();} & \\
+		\end{tabular}
+	\end{center}
+	\caption[Impossible hypothetical interleaving within transactions of Figure~\ref{fig:htm-example-reproduced}.]
+		{Impossible hypothetical interleaving within transactions of Figure~\ref{fig:htm-example-reproduced}.
+		The effects of $\tiat{2a-2b}$ are never globally visible, as its transasction must abort,
+		and moreover $\tiat{2c-3}$ will not execute at all.}
+	\label{fig:htm-example-interleaving}
+\end{figure}
+
 If these instructions from the {\tt x++} in the transaction are preempted,
@@
 the transaction will abort.
-So, the interleaving
+So, a hypothetical interleaving such as shown in Figure~\ref{fig:htm-example-interleaving},
+or equivalently written sequentially as:
 \[
@@
 \]
-or, henceforth abbreviated for clarity:
+or, henceforth further \revisionminor{simplified} for clarity:
 \[
@@
 
-\begin{figure}[t]
+\begin{figure}[h]
 	\begin{center}

### 2018-12-01 thesis/thesis.tex
fix citations to be colorful insted of having ugly borders
more acks

@@
 \usepackage[backref,pageanchor=true,plainpages=false, pdfpagelabels, bookmarks,bookmarksnumbered,
+colorlinks=true,
+citecolor=citec,
+linkcolor=linkc,
+urlcolor=urlc,
 %pdfborder=0 0 0,  %removes outlines around hyper links in online display
@@
 \usepackage{microtype}
+\usepackage{cleveref}
 
@@
 % \makeatother
+
+% NB. needs to be defined before mainmatter so they work in hyperref
+\definecolor{citec}{RGB}{168,0,84}
+\definecolor{linkc}{RGB}{0,84,168}
+\definecolor{urlc} {RGB}{168,84,0}
 
@@
 %(the lack of any stronger criticism a silent vote of confidence for their overall correctness).
-Thanks to Niko Matsakis for the mentorship and collaboration during my two summers as a Rust research intern at Mozilla,
-and to the rest of the Rust community as well for their extremely important work
+Thanks to Nicholas D. Matsakis for the mentorship and collaboration during my two summers as a Rust research intern at Mozilla,
+and to the rest of the Rust community as well for their %extremely important
+work
 bringing better concurrent programs to the world from a different angle.
@@
 From those who stayed in Pittsburgh %with me
-to those who moved away but
+to those who %moved away but
 kept in touch,
 from my regular tabletop, board, and/or dance gaming partners
-to those who inexhaustibly
+to those who %inexhaustibly
 welcomed me with company and
@@
 %-- and in any order other than alphabetical here would be far too cruel --
+8 Gianfortoni,
 Adam Blank,
@@
 Elly Fong-Jones,
+Emily Leathers,
 Emma Cating-Subramanian,
@@
 Nathaniel Wesley Filardo,
-Philip ``8'' Gianfortoni,
 Rhett Lauffenburger,
@@
 And to anyone who may read this list
-and feel even slightly disappointed that their name was left off:
+and feel even slightly disappointed that their name was left out:
 I owe you a nice dinner. Get in touch.

### 2018-12-03 thesis/intro.tex
assorted minor revs, intro, notes on reading this, etc

@@
 
-\subsubsection{Motivation}
+\section{Motivation}
 
@@
 
-\subsubsection{Contributions}
+\section{Contributions}
 
@@
 
-\subsubsection{Organization}
+\section{Organization}
 
@@
 
-\subsubsection{Notes on reading this thesis}
+\section{Notes on reading this thesis}
+
+\subsubsection{Experience}
 
@@
 
+\subsubsection{Vision}
+
 Color will be used in figures and graphs to add visual contrast and make the data easier to navigate at a glance,
@@
 and $\dagger$ distinguishes state space estimates from completed verifications.
+\revision{Should any vision-impaired reader require fully-textual figure descriptions,
+I would be happy to provide them upon request.}
+
+\subsubsection{Language}
 
@@
 %to distinguish her from the thesis audience concerned with the underlying research,
-Individual students who participated in the user studies will be given the more inclusive they/them.
-Gender-neutral pronouns will also be used on the author themself.
+Gender-neutral pronouns will be used to refer to
+individual students who participated in the user studies,
+%will be given the more inclusive they/them.
+%Gender-neutral pronouns will also be used
+as well as
+on the author themself.
+
+\subsubsection{Code}
+
+\newcommand\landslidelicense{BSD\xspace}
 
@@
 others may find that disorienting and wish to see things in a way a compiler would understand.
-Chapter~\ref{chap:landslide} provides source code links and serves as a guide to browsing the repository.
+\revisionminor{
+The Landslide codebase is open-source under the \landslidelicense license,
+available at \url{https://github.com/bblum/landslide}.
+Chapter~\ref{chap:landslide} provides more detail and}
+serves as a guide to browsing the repository.
 Later chapters will often make parenthetical references to specific files and/or functions

### 2018-12-03 thesis/landslide.tex
update landslide head commit pointer in case i forget later
assorted minor revs, intro, notes on reading this, etc

@@
 % TODO: add some backup linkeroos
-The Bochs port of Landslide is likewise open-source and available at \url{https://github.com/bblum/landslide}.
+The Bochs port of Landslide is likewise open-source
+\revisionminor{under the \landslidelicense license}
+and available at \url{https://github.com/bblum/landslide}.
 % TODO: refresh
-The HEAD commit at the time of writing is 5d45e2c.
+The HEAD commit at the time of writing is 8127151.
 The Simics port uses Simics's proprietary API and is hence unlicensed and available upon request for educational use only.
@@
 	\item {\tt -h}: print help text and exit immediately
+	\item \revision{{\tt -s}: include ``secret'' options when printing help text}
 \end{itemize}

### 2018-12-03 thesis/thesis.tex
more acks

@@
 Elly Fong-Jones,
+Emily Forney,
 Emily Leathers,
@@
 Laura Abbott,
+Maija Mednieks,
 Margaret Meyerhofer,
@@
 Ryan Pearl,
-Todd Eisenberger,
 and
-Zachary McCord
+Todd Eisenberger
 for their wonderful friendship.
@@
 David Renshaw,
-Emily Forney,
 Evan Cavallo,
@@
 Karl Schulz,
+Kevin Nguyen,
 Kristy Gardner,
 Lea Albaugh,
-Maija Mednieks,
 Marlena Abraham,
@@
 And to anyone who may read this list
-and feel even slightly disappointed that their name was left out:
+and feel even slightly dejected that their name was left out:
 I owe you a nice dinner. Get in touch.

### 2018-12-03 thesis/education.tex
assorted minor revs, intro, notes on reading this, etc
use cleveref

@@
 Since then, I have extended Landslide with a fully-automatic instrumentation process
-for Pebbles thread libraries (P2s) (\sect{\ref{sec:education-pebbles-instrumentation}})
-and Pintos kernels (\sect{\ref{sec:education-pintos-instrumentation}})
+for Pebbles thread libraries (P2s) (\cref{sec:education-pebbles-instrumentation})
+and Pintos kernels (\cref{sec:education-pintos-instrumentation})
 to improve its accessibility.
@@
 At CMU, I analyzed logs and code snapshots taken as students used Landslide during P2
-(\sect{\ref{sec:education-eval-bugfinding}}),
+(\cref{sec:education-eval-bugfinding}),
 as well as the grades ultimately assigned after students who either did or didn't use Landslide submitted their projects
-(\sect{\ref{sec:education-eval-grades}}).
+(\cref{sec:education-eval-grades}).
 At CMU and PSU, I surveyed students on their experience after submitting their Landslide-debugged P2s
-(\sect{\ref{sec:education-eval-survey}}).
+(\cref{sec:education-eval-survey}).
 At U. Chicago, I collaborated with a TA to check submitted Pintos kernels,
 then \revisionminor{they} returned any resulting bug reports to students
-(\sect{\ref{sec:education-eval-bugfinding}})
+(\cref{sec:education-eval-bugfinding})
 and likewise surveyed them on the quality of the diagnostic output
-(\sect{\ref{sec:education-eval-survey}}).
+(\cref{sec:education-eval-survey}).
 
@@
 
-As described in \sect{\ref{sec:landslide-setup}},
+As described in \cref{sec:landslide-setup},
 all setup from the user's point of view is handled through the {\tt p2-setup.sh} script.%
@@
 }
-It, its helper scripts (\sect{\ref{sec:landslide-glue}}),
+It, its helper scripts (\cref{sec:landslide-glue}),
 and the {\tt landslide} script itself contain several checks to prevent
@@
 		which students often open-code rather than using the prescribed synchronization API,
-		and treats them as synchronization points as described in \sect{\ref{sec:landslide-blocking}}.
+		and treats them as synchronization points as described in \cref{sec:landslide-blocking}.
 	\item Landslide finds ``too suspicious'' spinwait-loops in mutex implementations
@@
 others are crafted to exercise generally concurrency-heavy code paths and uncover any number of unforeseen problems.
-Many use some of the features/annotations described in \sect{\ref{sec:landslide-testcases}}.
+Many use some of the features/annotations described in \cref{sec:landslide-testcases}.
 
@@
 		This test uses the {\tt TESTING\_MUTEXES}
-		described in \sect{\ref{sec:landslide-staticconfig}}
+		described in \cref{sec:landslide-staticconfig}
 		to enable data-race preemption points within the mutex implementation.
@@
 		Written for the sake of the Landslide lecture demo
-		(\sect{\ref{sec:education-pebbles-recruiting}}).
+		(\cref{sec:education-pebbles-recruiting}).
 		Tests for the Paradise Lost bug by attempting to break mutual exclusion.
@@
 For this study students were recruited remotely via email.
-After each of the {\em threads} and {\em userprog} project deadlines (\sect{\ref{sec:overview-pintos}}),
+After each of the {\em threads} and {\em userprog} project deadlines (\cref{sec:overview-pintos}),
 \uchos staff sent students an email inviting them to volunteer to receive Landslide's bug reports,
@@
 
-As described in \sect{\ref{sec:landslide-setup}},
+As described in \cref{sec:landslide-setup},
 all setup from the user's point of view is handled through the {\tt pintos-setup.sh} script.
@@
 and its helper {\tt pebsim/pintos/import-pintos.sh}
-perform most of the same sanity checks as listed in \sect{\ref{sec:education-pebbles-instrumentation}},
+perform most of the same sanity checks as listed in \cref{sec:education-pebbles-instrumentation},
 then applies the patch {\tt annotate-\allowbreak{}pintos.patch}
 (plus several more hacks in the script itself)
-to insert the {\tt tell\_landslide()} annotations (\sect{\ref{sec:tell-landslide}})
+to insert the {\tt tell\_landslide()} annotations (\cref{sec:tell-landslide})
 into the student's kernel code.
@@
 Similar to the survey for Pebbles projects
-\sect{\ref{sec:education-survey-pebbles}},
+\cref{sec:education-survey-pebbles},
 I surveyed the Pintos user study participants for their opinions.
@@
 and U. Chicago students' survey responses from Fall 2017.
-Table~\ref{tab:photo-of-ze-studence} shows the student participation rate across semesters
+\Cref{tab:photo-of-ze-studence} shows the student participation rate across semesters
 for the thread library study.%
@@
 I was unable to examine its students' objective usage data or bug reports for this section.
-Results from PSU are presented instead in \sect{\ref{sec:education-eval-survey}}.
+Results from PSU are presented instead in \cref{sec:education-eval-survey}.
 
@@
 deadlocks, use-after-frees, segfaults, infinite loops, assertion failure, and unit test failure.
-Table~\ref{tab:cmubugs} shows these statistics for each semester.
+\Cref{tab:cmubugs} shows these statistics for each semester.
 % may also be an overestimate if they only hacked around the bug, but that's less likely
@@
 This table also shows the number of false positives, as self-reported by Landslide
-as described in \sect{\ref{sec:education-pebbles-instrumentation}}.
+as described in \cref{sec:education-pebbles-instrumentation}.
 % nah where actually doing this, where making it hapen.
@@
 during the Landslide clinics and survey,
-and I report on them in \sect{\ref{sec:education-eval-survey-falsepositives}}.
+and I report on them in \cref{sec:education-eval-survey-falsepositives}.
 Nevertheless,
@@
 Though unscientific,
-this provides some assurance that Table~\ref{tab:cmubugs}'s count of concurrency bugs is accurate,
+this provides some assurance that \Cref{tab:cmubugs}'s count of concurrency bugs is accurate,
 even if the deterministic tally may overcount.
@@
 		It found the targeted priority-leak bug
-		described in \sect{\ref{sec:education-pintos-tests}} in one group,
+		described in \cref{sec:education-pintos-tests} in one group,
 		and exposed a {\tt NULL} pointer dereference in another
@@
 		% for one of these groups 2 traces were emitted, same stack trace (i think) just w diff pps
-		wherein {\tt process\_wait()}'s synchronization assumed
+		wherein {\tt process\_\allowbreak{}wait()}'s synchronization assumed
 		it would always be called before {\tt process\_exit()}.
@@
 they enthusiastically reported \revisionminor{having found}
-the bug reports very helpful (\sect{\ref{sec:education-survey-response-data}}).
+the bug reports very helpful (\cref{sec:education-survey-response-data}).
 Despite the small sample size, I consider this
@@
 it is difficult to decide what rate of mistakes is acceptable when the user's patience is at stake \cite{racerx},
-but the upcoming survey results (\sect{\ref{sec:education-eval-survey-falsepositives}})
+but the upcoming survey results (\cref{sec:education-eval-survey-falsepositives})
 suggest that students are generally at least willing
@@
 
-Figure~\ref{fig:photo-of-ze-studence} shows the distribution of project grades in CMU 15-410
+\Cref{fig:photo-of-ze-studence} shows the distribution of project grades in CMU 15-410
 between Fall 2013 and Spring 2018,
@@
 
-%We see in Figure~\ref{fig:photo-of-ze-studence} that, indeed,
+%We see in \Cref{fig:photo-of-ze-studence} that, indeed,
 {\bf Results.}
@@
 {\bf Statistical significance.}
-Table~\ref{tab:table-of-ze-studence} presents more detailed statistics from these six distributions.
+\Cref{tab:table-of-ze-studence} presents more detailed statistics from these six distributions.
 To assess whether the differences are significant,
@@
 
-Based on the results from \sect{\ref{sec:education-eval-bugfinding}},
+Based on the results from \cref{sec:education-eval-bugfinding},
 I selected 4 bugs to study in depth to ascertain
@@
 as non-users.
-Table~\ref{tab:eval-common-bugs} presents the results.
+\Cref{tab:eval-common-bugs} presents the results.
 
@@
 so as to check for unlock/lock races as well as lock/lock interactions,
-and, as mentioned previously (\sect{\ref{sec:education-pebbles-tests}}),
+and, as mentioned previously (\cref{sec:education-pebbles-tests}),
 checks for data race access pairs inside the mutex implementation itself.
@@
 as well as the 3 major ways of fixing it,
-are one of 15-410 staff's closely-guarded secrets (\sect{\ref{sec:410-secrecy}}).
+are one of 15-410 staff's closely-guarded secrets (\cref{sec:410-secrecy}).
 Suffice it to say that the {\tt paraguay} test invokes a custom Pathos {\tt misbehave} mode
@@
 The $p$ values
-in Table~\ref{tab:eval-common-bugs}
+in \Cref{tab:eval-common-bugs}
 are calculated in R \cite{r-lang} using Fisher's exact test \cite{fishers-exact-test},
@@
 in just the right interleaving pattern to cause corruption that leads to a visible crash.
-Landslide, meanwhile, detects this error immediately upon access (\sect{\ref{sec:landslide-valgrind-mode}}),
+Landslide, meanwhile, detects this error immediately upon access (\cref{sec:landslide-valgrind-mode}),
 with no need for complex corruption conditions.
@@
 % TODO - more specific section reference - quicksand det/nondet drpps eval
-(similarly to \sect{\ref{sec:quicksand-eval}})
+(similarly to \cref{sec:quicksand-eval})
 and assuming the students would find and fix any such ``deterministic'' use-after-free bugs before the deadline.
@@
 necessarily miss the student's subsequent experience interpreting them.
-The surveys listed in sections \sect{\ref{sec:education-survey-pebbles}} and \sect{\ref{sec:education-survey-pintos}}
+The surveys listed in sections \cref{sec:education-survey-pebbles} and \cref{sec:education-survey-pintos}
 serve to probe these more qualitative aspects of the Landslide experience.
@@
 The response distributions for each of the surveys' multiple choice questions
-are shown in Figure~\ref{fig:survey}.
+are shown in \Cref{fig:survey}.
 In total, 28 students (or pairs thereof) answered the survey:
@@
 Note that two of the questions
-(3 and 7 from \sect{\ref{sec:education-survey-pebbles}})
+(3 and 7 from \cref{sec:education-survey-pebbles})
 were not asked on U. Chicago's version of the survey,
@@
 Likewise, U. Chicago's questions 4, 8, and 10
-(see \sect{\ref{sec:education-survey-pintos}}),
+(see \cref{sec:education-survey-pintos}),
 which were not asked on CMU's and PSU's surveys,
@@
 upon further investigation,
-the 6 traces seem to represent 2 distinct bugs among them (\sect{\ref{sec:education-eval-bugs-uc}}).
+the 6 traces seem to represent 2 distinct bugs among them (\cref{sec:education-eval-bugs-uc}).
 
@@
 		(in several cases corresponding to {\tt mutex\_test}'s verification of the mutex's internal memory accesses
-		(see \sect{\ref{sec:education-pebbles-tests}}));
+		(see \cref{sec:education-pebbles-tests}));
 		i.e., expected behaviour rather than false positive bug reports.
@@
 as taught in CMU and PSU lecture material;
-I replaced these while paraphrasing in accordance with \sect{\ref{sec:glossary}}.}
+I replaced these while paraphrasing in accordance with \cref{sec:glossary}.}
 Answers from CMU:
@@
 Though I present the formal definition of data races in the lecture
-(\sect{\ref{sec:education-pebbles-recruiting}})
+(\cref{sec:education-pebbles-recruiting})
 and refer back to it in Landslide's documentation and output,
@@
 
-Comparing the survey response distributions between CMU versus PSU and U. Chicago (Figure~\ref{fig:survey}),
+Comparing the survey response distributions between CMU versus PSU and U. Chicago (\Cref{fig:survey}),
 CMU students tended to verify their bugfixes more often by re-running Landslide
@@
 %whereas PSU's \psuos is more towards the average between CMU's 15-213 and 15-410
-%(tempering the difficulty of P2 to make it more accessible (\sect{\ref{sec:overview-psu}})),
+%(tempering the difficulty of P2 to make it more accessible (\cref{sec:overview-psu})),
 Considering the higher demand CMU's 15-410 makes for prerequisite concurrency experience
-compared to PSU's relative tempering of P2's difficulty to make it more accessible (\sect{\ref{sec:pebbles}}),
+compared to PSU's relative tempering of P2's difficulty to make it more accessible (\cref{sec:pebbles}),
 % TODO: represent u c in this sentents?
@@
 \revisionminor{and relatively very little negative feedback such as false positives}.
-Landslide helped many students find and fix many bugs (\sect{\ref{sec:education-eval-bugfinding}}),
+Landslide helped many students find and fix many bugs (\cref{sec:education-eval-bugfinding}),
 but making a direct comparison to stress testing, the prior state of the art, is not straightforward.
@@
 Immediate improvement in students' project grades was observed
-(\sect{\ref{sec:education-eval-grades}}),
+(\cref{sec:education-eval-grades}),
 although statistical significance was lost when attempting to account for selection bias;
@@
 for the {\tt thr\_exit\_join} bug in particular.
-Students responded overwhelmingly positively in the survey (\sect{\ref{sec:education-eval-survey}}),
+Students responded overwhelmingly positively in the survey (\cref{sec:education-eval-survey}),
 although it is easy to imagine students being equally happy with a ``debugging tool'' that just tells them all the answers.
@@
 In contrast, since switching to the automatically-annotatable P2,
-the participation rate rose to 58\% (Table~\ref{tab:photo-of-ze-studence})
+the participation rate rose to 58\% (\Cref{tab:photo-of-ze-studence})
 among all P2-submitting groups.
@@
 According only to my memory of students who attended the Landslide clinics
-(\sect{\ref{sec:education-pebbles-recruiting}}),
+(\cref{sec:education-pebbles-recruiting}),
 the racial diversity was roughly representative of the class at large,
@@
 (for example, those with {\tt ready\_list} changed to an array,
-as described in \sect{\ref{sec:education-pintos-instrumentation}}).
+as described in \cref{sec:education-pintos-instrumentation}).
 Had I given Landslide directly to students that semester,
@@
 Landslide can now handle a considerably wider variety of student implementation quirks
-on account of the fixes from this time (\sect{\ref{sec:education-pintos-instrumentation}}).
+on account of the fixes from this time (\cref{sec:education-pintos-instrumentation}).
 In its original shape
 (before the F'17 semester,
-having only enough instrumentation necessary for the Pintoses used in \sect{\ref{sec:quicksand-eval}}'s experiments),
+having only enough instrumentation necessary for the Pintoses used in \cref{sec:quicksand-eval}'s experiments),
 Landslide was already able to automatically instrument 18 out of the 21 {\tt threads} project submissions at U. Chicago.

### 2018-12-03 thesis/future.tex
more assorted revisions
use cleveref

@@
 \label{chap:warpzone}
+\revision{
 \inspirationalquote{
@@
 	\end{tabular}}
-%{Galatea, The Talos Principle}
 {Dalinar Kholin, Oathbringer}
+}
 
@@
 with existing techniques
-(\sect{\ref{sec:quicksand-discussion}},
-\sect{\ref{sec:education-discussion}},
-\sect{\ref{sec:tm-discussion}}).
+(\cref{sec:quicksand-discussion},
+\cref{sec:education-discussion},
+\cref{sec:tm-discussion}).
 Meanwhile, this chapter takes a broader interpretation of ``future work'',
@@
 
-In the user study survey (\sect{\ref{sec:education-eval-survey}}),
+In the user study survey (\cref{sec:education-eval-survey}),
 students most often complained
@@
 would allow Quicksand to achieve more meaningful results in that same budget.
-Theoretical advances in state space reduction come out every year (\sect{\ref{sec:related-algs}}),
+Theoretical advances in state space reduction come out every year (\cref{sec:related-algs}),
 but the user's human intuition can also contribute if properly encouraged.
-For the P2 tests (\sect{\ref{sec:education-pebbles-tests}}),
+For the P2 tests (\cref{sec:education-pebbles-tests}),
 I configured Quicksand by hand to issue appropriate {\tt without\_function} commands to Landslide
-(\sect{\ref{sec:landslide-pps}}),
-and even more still for the HTM tests (\sect{\ref{sec:tm-eval-exp-setup}}),
+(\cref{sec:landslide-pps}),
+and even more still for the HTM tests (\cref{sec:tm-eval-exp-setup}),
 but a user writing her own tests would have to configure DPOR by hand.
@@
 
-In the case study of submitted P2 bugs (\sect{\ref{sec:education-eval-bug-case-study}}),
+In the case study of submitted P2 bugs (\cref{sec:education-eval-bug-case-study}),
 I noticed several submissions of a common pattern
@@
 although I know of no related work which addresses the second half of this problem.
-Likewise, \sect{\ref{sec:tm-abstraction}}'s abstraction reduction requires the user
+Likewise, \cref{sec:tm-abstraction}'s abstraction reduction requires the user
 to believe that {\tt lock}({\tt spinlock}) and friends correctly express the mutual exclusion property in informal code%
-\footnote{In Figure~\ref{fig:htm-lock}(a), the
+\footnote{In \Cref{fig:htm-lock}(a), the
 {\tt \flow{if} (!\call{\_xtest()}) \call{thr\_yield}(\const{-1});}
@@
 
-As important as Quicksand's convergence theorem (\sect{\ref{sec:quicksand-soundness}}),
-the HTM equivalence proofs (\sect{\ref{sec:tm-design}}),
+As important as Quicksand's convergence theorem (\cref{sec:quicksand-soundness}),
+the HTM equivalence proofs (\cref{sec:tm-design}),
 and other soundness results from prior work are,
@@
 inspecting small state spaces
-(such as Figure~\ref{fig:retry-reduction})
+(such as \Cref{fig:retry-reduction})
 by hand
@@
 Nevertheless, discrepancies may still lurk undiscovered:
-the failure injection bug described in \sect{\ref{sec:tm-eval-exp-setup}}
+the failure injection bug described in \cref{sec:tm-eval-exp-setup}
 took until implementing retry set reduction to discover,
@@
 \section{Heuristics}
+\label{sec:warpzone-heuristics}
 
@@
 bouncing wildly among orders of magnitude even in the space of ten or fewer adjacent interleavings.
-These flaws are especially visible in the {\tt avl\_fixed} transactional verification results (Table~\ref{tab:tm-verifs}).
+These flaws are especially visible in the {\tt avl\_fixed} transactional verification results (\Cref{tab:tm-verifs}).
 Both estimation algorithms
 use the known structure of the tree as a model for the unknown
-(see \sect{\ref{sec:landslide-estimate}}),
+(see \cref{sec:landslide-estimate}),
 but make no use of domain-specific knowledge such as which threads were chosen to run at each preemption point.
-For example, consider the root of the left subtree in Figure~\ref{fig:tree},
+For example, consider the root of the left subtree in \Cref{fig:tree},
 thread 1's \dporTA{{\tt tmp1++}}.
@@
 such as subtrees pruned by sleep set reduction
-(\sect{\ref{sec:landslide-sleepsets}}; $\bigstar$4 in Figure~\ref{fig:dpor-example-2}),
+(\cref{sec:landslide-sleepsets}; $\bigstar$4 in \Cref{fig:dpor-example-2}),
 would likely require analyzing DPOR's memory conflicts as well.
@@
 There is much room to expand Landslide's use of heuristics to balance verification with fast bug-finding.
-Landslide's maximal state space mode (\sect{\ref{sec:tm-verif}}) currently requires the user to decide in advance
+Landslide's maximal state space mode (\cref{sec:tm-verif}) currently requires the user to decide in advance
 whether she thinks a full verification is possible within the time limit,
@@
 % with machine learning... don't feel like i can do justice to discussing its flaws with real citations
-I have made the estimation logs from the HTM experiments (\sect{\ref{sec:tm-eval}})
+I have made the estimation logs from the HTM experiments (\cref{sec:tm-eval})
 freely available at \url{https://github.com/bblum/thesis-htm-logs}.

### 2018-12-03 thesis/intro.tex
use cleveref

@@
 \begin{enumerate}
-	\item {\bf Meaningful state spaces (Chapter \ref{chap:quicksand}).}
+	\item {\bf Meaningful state spaces (\Cref{chap:quicksand}).}
 		I will present {\em Iterative Deepening}, a new algorithm for navigating the trade-off in how many preemption points to test at once.
@@
 		I will show that Iterative Deepening of preemption points outperforms prior work in terms \revisionminor{of both} finding bugs quickly and completely verifying correctness when no bug exists.
-	\item {\bf Educational use (Chapter \ref{chap:education}).}
+	\item {\bf Educational use (\Cref{chap:education}).}
 		For the past five semesters, I have offered a fully-automated version of Landslide to students in 15-410, CMU's undergraduate Operating System Design and Implementation class \cite{kspec,thrlib}, for use as a debugging aid during the thread library project.
@@
 		to evaluate the suitability of stateless model checking in an educational setting.
-	\item {\bf Hardware Transactional Memory (HTM) (Chapter \ref{chap:tm}).}
+	\item {\bf Hardware Transactional Memory (HTM) (\Cref{chap:tm}).}
 		HTM is a relatively new concurrent programming technique \cite{htm-haswell,htm-experience}
@@
 \begin{itemize}
-	\item {\bf Background:} Chapter~\ref{chap:background} will present the requisite background material on concurrent programming, stateless model checking, and the various types of programs targeted by Landslide.
-	\item {\bf Landslide:} Chapter~\ref{chap:landslide} explains the design and implementation of Landslide
+	\item {\bf Background:} \Cref{chap:background} will present the requisite background material on concurrent programming, stateless model checking, and the various types of programs targeted by Landslide.
+	\item {\bf Landslide:} \Cref{chap:landslide} explains the design and implementation of Landslide
 		%the core of my stateless model checking framework,
@@
 		It is the foundation upon which all three contributions above build.
-	\item {\bf Quicksand:} Chapter~\ref{chap:quicksand} presents the Iterative Deepening framework which more intelligently chooses which state spaces to test, corresponding to contribution 1 above.
-	\item {\bf Education:} Chapter~\ref{chap:education} discusses my evaluation of Landslide
+	\item {\bf Quicksand:} \Cref{chap:quicksand} presents the Iterative Deepening framework which more intelligently chooses which state spaces to test, corresponding to contribution 1 above.
+	\item {\bf Education:} \Cref{chap:education} discusses my evaluation of Landslide
 		in CMU's 15-410 class environment using the Pebbles kernel,
@@
 		corresponding to contribution 2 above.
-	%\item {\bf Pebbles:} Chapter~\ref{chap:410} discusses my evaluation of Landslide in CMU's 15-410 class environment using the Pebbles kernel, corresponding to contribution 2 above.
-	%\item {\bf Pintos:} Chapter~\ref{chap:410} discusses my evaluation of Landslide in the University of Chicago's and Berkeley's OS class environments using the Pintos kernel, corresponding to contribution 2 above.
-	\item {\bf Transactional Memory:} Chapter~\ref{chap:tm} presents my extension of Landslide's concurrency model to handle transactional concurrency and the evaluation thereof, corresponding to contribution 3 above.
-	\item {\bf Related Work:} Chapter~\ref{chap:relatedwork} honors my neighbours and ancestors in research spirit.
-	\item {\bf Future Work:} Chapter~\ref{chap:warpzone} lists stateless model checking's remaining shortcomings and how new research might address them.
-	\item {\bf Conclusion:} Chapter~\ref{chap:conclusion} provides some thoughts on the future of the field.
+	%\item {\bf Pebbles:} \Cref{chap:410} discusses my evaluation of Landslide in CMU's 15-410 class environment using the Pebbles kernel, corresponding to contribution 2 above.
+	%\item {\bf Pintos:} \Cref{chap:410} discusses my evaluation of Landslide in the University of Chicago's and Berkeley's OS class environments using the Pintos kernel, corresponding to contribution 2 above.
+	\item {\bf Transactional Memory:} \Cref{chap:tm} presents my extension of Landslide's concurrency model to handle transactional concurrency and the evaluation thereof, corresponding to contribution 3 above.
+	\item {\bf Related Work:} \Cref{chap:relatedwork} honors my neighbours and ancestors in research spirit.
+	\item {\bf Future Work:} \Cref{chap:warpzone} lists stateless model checking's remaining shortcomings and how new research might address them.
+	\item {\bf Conclusion:} \Cref{chap:conclusion} provides some thoughts on the future of the field.
 \end{itemize}
@@
 I cannot promise all easy reading.
-Chapter~\ref{chap:background}, Background, provides what I hope are friendly concrete examples
+\Cref{chap:background}, Background, provides what I hope are friendly concrete examples
 to help the reader feel comfortable with each level of intuition that upcoming algorithms will build upon.
@@
 if not necessarily the details of each algorithm or soundness proof.
-In particular, Chapter~\ref{chap:education}, Education,
+In particular, \Cref{chap:education}, Education,
 may be approached with no knowledge of concurrency or model checking,
 taking it merely as a study of a magic new debugging tool in the classroom setting.
-The more ambitious reader may proceed to the Landslide chapter's algorithm walkthroughs~(\sect{\ref{sec:landslide-algs}}),
+The more ambitious reader may proceed to the Landslide chapter's algorithm walkthroughs~(\cref{sec:landslide-algs}),
 which should equip them to understand every detail herein.
-Readers who are here only to skim and skip around should at least be aware of the glossary~(\sect{\ref{sec:glossary}})
+Readers who are here only to skim and skip around should at least be aware of the glossary~(\cref{sec:glossary})
 to help clarify any terminology confusion.
@@
 available at \url{https://github.com/bblum/landslide}.
-Chapter~\ref{chap:landslide} provides more detail and}
+\Cref{chap:landslide} provides more detail and}
 serves as a guide to browsing the repository.

### 2018-12-03 thesis/landslide.tex
use cleveref

@@
 with detailed and friendly examples to help establish intuitions the later chapters may require,
-may skip to \sect{\ref{sec:landslide-algs}}.
+may skip to \cref{sec:landslide-algs}.
 
@@
 % TODO: make more specific section reference
-Separate user guides also exist, described in Chapter~\ref{chap:education}.
+Separate user guides also exist, described in \Cref{chap:education}.
 
@@
 	\item Run the instrumentation script on the resulting binary to let Landslide know where all the important functions are
-		(see \sect{\ref{sec:landslide-glue}}).
+		(see \cref{sec:landslide-glue}).
 \end{itemize}
@@
 
-The preferred method of invoking Landslide is through Quicksand, the Iterative Deepening wrapper program which has all of Chapter~\ref{chap:quicksand} to itself.
+The preferred method of invoking Landslide is through Quicksand, the Iterative Deepening wrapper program which has all of \Cref{chap:quicksand} to itself.
 % TODO: add quicksand symlink
@@
 		and sends it to me for nefarious research purposes
-		\revisionminor{(see \sect{\ref{sec:education-pebbles-instrumentation}})}.
+		\revisionminor{(see \cref{sec:education-pebbles-instrumentation})}.
 \end{itemize}
@@
 		this generally causes bugs to be found faster should they exist, but degrades completion time
-		(\sect{\ref{sec:landslide-icb}})
-	\item {\tt -0}: enable Preempt-Everywhere mode (\sect{\ref{sec:quicksand-eval}}, requires {\tt -C})
+		(\cref{sec:landslide-icb})
+	\item {\tt -0}: enable Preempt-Everywhere mode (\cref{sec:quicksand-eval}, requires {\tt -C})
 	\item {\tt -M}: enable Maximal State Space mode,
@@
 		abandoning all subset jobs even if they might find bugs faster
-		(\sect{\ref{sec:tm-eval}}, incompatible with {\tt -C}).
-		According to \sect{\ref{sec:quicksand-soundness}}'s soundness proofs,
+		(\cref{sec:tm-eval}, incompatible with {\tt -C}).
+		According to \cref{sec:quicksand-soundness}'s soundness proofs,
 		this is equivalent to {\tt -0}
 		(and according to my experience, \revisionminor{much} faster as well).
-	\item {\tt -H}: use Limited Happens-Before for data-race analysis (\sect{\ref{sec:background-hb}})
+	\item {\tt -H}: use Limited Happens-Before for data-race analysis (\cref{sec:background-hb})
 		(default for Pebbles kernelspace mode)
-	\item {\tt -V}: use vector-clock-based Pure Happens-Before for data-race analysis (\sect{\ref{sec:background-hb}})
+	\item {\tt -V}: use vector-clock-based Pure Happens-Before for data-race analysis (\cref{sec:background-hb})
 		(default for P2/PSU userspace and Pintos modes)
-	\item {\tt -X}: support transactional memory (Chapter~\ref{chap:tm})
-	\item {\tt -A}: support multiple abort codes during transaction failure (\sect{\ref{sec:tm-implementation}});
+	\item {\tt -X}: support transactional memory (\Cref{chap:tm})
+	\item {\tt -A}: support multiple abort codes during transaction failure (\cref{sec:tm-implementation});
 		required for testing programs which behave differently under different abort circumstances,
 		but impacts the state space size
-	\item {\tt -S}: suppress retry aborts during transaction failure (\sect{\ref{sec:tm-implementation}})
-	\item {\tt -R}: enable retry-set state space reduction for transactional tests (\sect{\ref{sec:tm-implementation}})
+	\item {\tt -S}: suppress retry aborts during transaction failure (\cref{sec:tm-implementation})
+	\item {\tt -R}: enable retry-set state space reduction for transactional tests (\cref{sec:tm-implementation})
 	\item {\tt -P}: support Pintos architecture (enabled automatically when {\tt pintos-setup.sh} is run)
@@
 When supplied no arguments, it reads configuration options from {\tt pebsim/config.landslide}
-(a bash script expected to define certain variables as described in \sect{\ref{sec:landslide-glue}}).
+(a bash script expected to define certain variables as described in \cref{sec:landslide-glue}).
 The user may optionally specify a file containing additional config directives
@@
 \begin{itemize}
-	\item {\tt within\_function FUNC} - adds {\tt FUNC} to an allowlist of functions required to appear in the current stack trace before identifying a preemption point (see \sect{\ref{sec:landslide-pps}})
+	\item {\tt within\_function FUNC} - adds {\tt FUNC} to an allowlist of functions required to appear in the current stack trace before identifying a preemption point (see \cref{sec:landslide-pps})
 	\item {\tt without\_function FUNC} - as above, but a denylist instead of an allowlist
@@
 				% TODO: fix this section ref
-				For further discussion see \sect{\ref{sec:quicksand-pps}}.
+				For further discussion see \cref{sec:quicksand-pps}.
 			\item {\tt CURRENT\_SYSCALL} indicates the system call number if a user-space data race comes from within a kernel system call which accesses user memory (Pebbles only).
@@
 \begin{itemize}
-	\item {\tt ICB=0} - enable Iterative Context Bounding (\sect{\ref{sec:landslide-icb}});
-		corresponds to {\tt -I} in \sect{\ref{sec:landslide-quicksand-options}}.
-	\item {\tt PREEMPT\_EVERYWHERE=0} - enable Preempt-Everywhere mode (\sect{\ref{sec:quicksand-eval}});
-		corresponds to {\tt -0} in \sect{\ref{sec:landslide-quicksand-options}}.
+	\item {\tt ICB=0} - enable Iterative Context Bounding (\cref{sec:landslide-icb});
+		corresponds to {\tt -I} in \cref{sec:landslide-quicksand-options}.
+	\item {\tt PREEMPT\_EVERYWHERE=0} - enable Preempt-Everywhere mode (\cref{sec:quicksand-eval});
+		corresponds to {\tt -0} in \cref{sec:landslide-quicksand-options}.
 	\item {\tt EXPLORE\_BACKWARDS=0} - configure whether, at each newly encountered preemption point,
@@
 \begin{itemize}
-	\item {\tt PURE\_HAPPENS\_BEFORE=1} - select Pure Happens-Before (1) or Limited Happens-Before (2) (\sect{\ref{sec:background-hb}});
-		corresponds to {\tt -V}/{\tt -H} in \sect{\ref{sec:landslide-quicksand-options}}.
+	\item {\tt PURE\_HAPPENS\_BEFORE=1} - select Pure Happens-Before (1) or Limited Happens-Before (2) (\cref{sec:background-hb});
+		corresponds to {\tt -V}/{\tt -H} in \cref{sec:landslide-quicksand-options}.
 	\item {\tt FILTER\_DRS\_BY\_TID=1} - configures whether to use the {\tt TID} parameter of {\tt data\_\allowbreak{}race} described above.
@@
 Tests for the most common interactions during the P2 and Pintos projects are of course already supplied,
-as described in \sect{\ref{sec:education-pebbles-tests}} and \sect{\ref{sec:education-pintos-tests}}.
+as described in \cref{sec:education-pebbles-tests} and \cref{sec:education-pintos-tests}.
 
 Use of {\tt tell\_landslide()} annotations
-\revisionminor{(\sect{\ref{sec:tell-landslide}})}
+\revisionminor{(\cref{sec:tell-landslide})}
 is not necessary,
@@
 as the true nature of the Pathos reference kernel's {\tt misbehave} modes
-is a closely-guarded secret among 15-410 course staff (\sect{\ref{sec:410-secrecy}}).
+is a closely-guarded secret among 15-410 course staff (\cref{sec:410-secrecy}).
 The mode in question causes the reference kernel to prioritize scheduling the child thread over the parent
@@
 
-To illustrate, consider the following program in Figure~\ref{fig:misbehave-example},
+To illustrate, consider the following program in \Cref{fig:misbehave-example},
 and suppose Landslide is configured to preempt only on mutex API calls
-(such as in the first step of Iterative Deepening (\sect{\ref{sec:quicksand-id}})).
+(such as in the first step of Iterative Deepening (\cref{sec:quicksand-id})).
 Because Landslide ignores all kernel-level synchronization short of context switches when testing user-level code,
@@
 Hence, DPOR will erroneously think everything before that {\tt thr\_join()}
-happens-before (\sect{\ref{sec:landslide-dpor-hb}}) anything the child does,
+happens-before (\cref{sec:landslide-dpor-hb}) anything the child does,
 and will fail to identify the racing accesses on {\tt x}.
@@
 
-Though Iterative Deepening's soundness (\sect{\ref{sec:quicksand-soundness}})
+Though Iterative Deepening's soundness (\cref{sec:quicksand-soundness})
 guarantees all data races will eventually be detected starting from just synchronization preemption points,
@@
 If at this point the reader wonders why Landslide doesn't just identify
-the {\tt thread\_create} and {\tt make\_runnable} system calls in the arbiter itself (\sect{\ref{sec:landslide-arbiter}})
+the {\tt thread\_create} and {\tt make\_runnable} system calls in the arbiter itself (\cref{sec:landslide-arbiter})
 and skip this mysterious user-visible complexity,
@@
 The bug reports are formatted in \revisionminor{HTML}, recommended to be viewed in a web browser.
-An example is shown in Figure~\ref{fig:bugreport}.
+An example is shown in \Cref{fig:bugreport}.
 
@@
 by searching for the macro {\tt FOUND\_A\_BUG\_HTML\_INFO} in Landslide's code.}
-\revision{\sect{\ref{sec:future-friendly}} discusses how these bug reports might be further improved
+\revision{\cref{sec:future-friendly} discusses how these bug reports might be further improved
 by adding more information still,
@@
 The following annotations are specified in {\tt pebsim/config.landslide} akin to the static configuration options
-described in \sect{\ref{sec:landslide-directly}}..
+described in \cref{sec:landslide-directly}..
 These specify the names of kernel functions, global variables, default values, and so on
@@
 		The user should be careful with this option to also enable the proper {\tt thr\_create()} misbehave mode
-		in her test case (\sect{\ref{sec:landslide-friendly-misbehave}}).
-	\item {\tt TRUSTED\_THR\_JOIN=0} - if set to 1, forces Landslide to add a happens-before edge (\sect{\ref{sec:landslide-dpor-hb}}) between the exiting of some thread $N$ and the end of any subsequent {\tt thr\_join}$(N)$ call,
+		in her test case (\cref{sec:landslide-friendly-misbehave}).
+	\item {\tt TRUSTED\_THR\_JOIN=0} - if set to 1, forces Landslide to add a happens-before edge (\cref{sec:landslide-dpor-hb}) between the exiting of some thread $N$ and the end of any subsequent {\tt thr\_join}$(N)$ call,
 		even if that {\tt join} would not ordinarily block.
@@
 Landslide does not actually need to store any nodes for execution sequences outside the current variation
-(see \sect{\ref{sec:landslide-estimate}} and \sect{\ref{sec:landslide-dpor}} for why),
+(see \cref{sec:landslide-estimate} and \cref{sec:landslide-dpor} for why),
 so the total memory consumption is only $O(n)$ in the number of preemption points in a single program run
@@
 		depth in the tree, parent node pointer, etc.;
-	\item Snapshots of the current state of the scheduler (\sect{\ref{sec:landslide-scheduler}})
-		and memory accesses and heaps (\sect{\ref{sec:landslide-memory}});
+	\item Snapshots of the current state of the scheduler (\cref{sec:landslide-scheduler})
+		and memory accesses and heaps (\cref{sec:landslide-memory});
 	\item Simulator-dependent data needed to time travel and resume execution
-		from this checkpoint (\sect{\ref{sec:landslide-timetravel}});
+		from this checkpoint (\cref{sec:landslide-timetravel});
 	\item List of parent/ancestor nodes with memory conflicts and/or happens-before edges to this one
-		for DPOR (\sect{\ref{sec:landslide-dpor}});
+		for DPOR (\cref{sec:landslide-dpor});
 	\item Current estimated state space proportion and execution time for the subtree rooted at this node
-		(not necessarily fully explored yet) for estimation \sect{\ref{sec:landslide-estimate}};
+		(not necessarily fully explored yet) for estimation \cref{sec:landslide-estimate};
 	\item Whether this point is an {\tt xbegin} invocation
 		and if so what {\tt xabort} codes are possible and/or already explored for this transaction
-		(\sect{\ref{chap:tm}}).
+		(\cref{chap:tm}).
 \end{itemize}
@@
 and {\tt sched\_recover()}, which prepares the scheduler to force a new thread to run
-after a time travel (\sect{\ref{sec:landslide-timetravel}}).
+after a time travel (\cref{sec:landslide-timetravel}).
 
@@
 The second part of {\tt sched\_update()},
-conditional on the arbiter identifying preemption points (\sect{\ref{sec:landslide-arbiter}}),
+conditional on the arbiter identifying preemption points (\cref{sec:landslide-arbiter}),
 manages timer interrupts to switch to a desired thread.
 Whenever a preemption point is reached,
-the scheduler first creates a checkpoint in the execution tree (\sect{\ref{sec:landslide-save}}),
+the scheduler first creates a checkpoint in the execution tree (\cref{sec:landslide-save}),
 asks the arbiter which thread to run next,
 and if that thread is different from the current one,
-forces the kernel into its timer interrupt handler (\sect{\ref{sec:landslide-interrupce}}).
+forces the kernel into its timer interrupt handler (\cref{sec:landslide-interrupce}).
 
@@
 and checks them against each other
-for DPOR (\sect{\ref{sec:landslide-dpor}})
-and data race analysis (\sect{\ref{sec:landslide-datarace}}).
+for DPOR (\cref{sec:landslide-dpor})
+and data race analysis (\cref{sec:landslide-datarace}).
 For userspace tests, it also tracks which virtual address space ({\tt cr3}) belongs to the test program
@@
 Any such conflicts which also fail a lockset
-and/or happens-before check (\sect{\ref{sec:landslide-datarace}})
+and/or happens-before check (\cref{sec:landslide-datarace})
 are then reported as data races.
-Regardless, all such conflicts are later used by DPOR (\sect{\ref{sec:landslide-dpor}}) to find dependent transition pairs.
+Regardless, all such conflicts are later used by DPOR (\cref{sec:landslide-dpor}) to find dependent transition pairs.
 
@@
 the latter is used only by the interrupt manipulation below
-and by the scheduler to force Pintos to skip certain parts of its init sequence (\sect{\ref{sec:landslide-pintosspecifics}}).
+and by the scheduler to force Pintos to skip certain parts of its init sequence (\cref{sec:landslide-pintosspecifics}).
 Both rely on the helper function {\tt mem\_translate()} for virtual address resolution,
@@
 ({\tt arbiter\_choose()}).
-Its behaviour in the former case is configured by the options listed in \sect{\ref{sec:landslide-dynamicconfig}},
-and in the latter case by the options listed in \sect{\ref{sec:landslide-staticconfig}}.
+Its behaviour in the former case is configured by the options listed in \cref{sec:landslide-dynamicconfig},
+and in the latter case by the options listed in \cref{sec:landslide-staticconfig}.
 For example, {\tt EXPLORE\_BACKWARDS} is interpreted here;
@@
 Its contents are largely algorithmic rather than architectural
-and hence further described in \sect{\ref{sec:landslide-dpor}} and \sect{\ref{sec:landslide-icb}}.
+and hence further described in \cref{sec:landslide-dpor} and \cref{sec:landslide-icb}.
 
@@
 	\item Landslide runs the tortoise/rabbit algorithm to detect cyclic {\tt ebp} chains and terminate after the first time around.
-	\item Two other Pebbles-specifc special cases described in \sect{\ref{sec:landslide-pebblesspecifics}}.
+	\item Two other Pebbles-specifc special cases described in \cref{sec:landslide-pebblesspecifics}.
 \end{itemize}
@@
 
-The preemption traces, described and exemplified in \sect{\ref{sec:landslide-bugreport}}, are generated by {\tt found\_a\_\allowbreak{}bug.c},
+The preemption traces, described and exemplified in \cref{sec:landslide-bugreport}, are generated by {\tt found\_a\_\allowbreak{}bug.c},
 in cooperation with {\tt save.c}.
@@
 		(a reference kernel feature to allow Simics stack traces to cross same),
-		use {\tt PATHOS\_SYSCALL\_IRET\_DISTANCE} (\sect{\ref{sec:landslide-config-landslide}})
+		use {\tt PATHOS\_SYSCALL\_IRET\_DISTANCE} (\cref{sec:landslide-config-landslide})
 		to track {\tt esp}'s value across the stack switch.
@@
 
-The options specified in \sect{\ref{sec:landslide-directly}}
+The options specified in \cref{sec:landslide-directly}
 are handled by a family of \revisionminor{distasteful} shell scripts that live in {\tt pebsim/}.
@@
 \begin{itemize}
-	\item {\tt landslide} is the outermost script invoked by Quicksand (or by a \sect{\ref{sec:landslide-directly}} aficionado).
+	\item {\tt landslide} is the outermost script invoked by Quicksand (or by a \cref{sec:landslide-directly} aficionado).
 		It exports several key environment variables used by the other scripts,
@@
 	\item {\tt build.sh} ensures the build of Landslide is up-to-date,
-		and processes any dynamic configuration options which don't require updating the build (\sect{\ref{sec:landslide-dynamicconfig}})
+		and processes any dynamic configuration options which don't require updating the build (\cref{sec:landslide-dynamicconfig})
 		It verifies all required {\tt tell\_landslide} annotations are present,
@@
 		to copy the student implementation into their respective directories.
-		\sect{\ref{sec:education-pebbles-instrumentation}} and \sect{\ref{sec:education-pintos-instrumentation}}
+		\cref{sec:education-pebbles-instrumentation} and \cref{sec:education-pintos-instrumentation}
 		describe their office in more detail.
@@
 containing a bunch of {\tt \#define}s of the addresses of important functions in the compiled binary,
-specific features enabled or disabled by the static config options (\sect{\ref{sec:landslide-staticconfig}}),
+specific features enabled or disabled by the static config options (\cref{sec:landslide-staticconfig}),
 and so on.
@@
 At startup, {\tt pps\_init()} and {\tt load\_dynamic\_pps()} process the
-statically-configured preemption points (\sect{\ref{sec:landslide-staticconfig}})
-and dynamically-configured preemption points (\sect{\ref{sec:landslide-dynamicconfig}}),
+statically-configured preemption points (\cref{sec:landslide-staticconfig})
+and dynamically-configured preemption points (\cref{sec:landslide-dynamicconfig}),
 respectively.
@@
 (in a similar manner to \revisionminor{prior work's Preemption Sealing} \cite{sealing}).
-It invokes the stack tracer (\sect{\ref{sec:landslide-stacktrace}})
+It invokes the stack tracer (\cref{sec:landslide-stacktrace})
 for a list of which functions are on the call stack
@@
 The point can optionally be qualified by a thread ID, most recent system call number, etc.,
-as described in \sect{\ref{sec:landslide-dynamicconfig}},
+as described in \cref{sec:landslide-dynamicconfig},
 and is queried through {\tt suspected\_data\_race()}.
@@
 	\item {\tt hlt} instruction (kernel waiting for interrupt)
-	\item User thread becomes yield- or xchg-blocked (\sect{\ref{sec:landslide-blocking}})
-	\item {\tt \_xbegin()} or {\tt \_xend()}, if testing transactional memory (Chapter~\ref{chap:tm})
+	\item User thread becomes yield- or xchg-blocked (\cref{sec:landslide-blocking})
+	\item {\tt \_xbegin()} or {\tt \_xend()}, if testing transactional memory (\Cref{chap:tm})
 \end{itemize}
@@
 Whenever {\tt arbiter\_interested()} returns true,
-Landslide creates a new {\tt struct hax} in the execution tree (\sect{\ref{sec:landslide-save}}),
-creates a checkpoint (\sect{\ref{sec:landslide-timetravel}}),
-and queries {\tt arbiter\_choose()} to decide which thread to run next (\sect{\ref{sec:landslide-arbiter}}).
+Landslide creates a new {\tt struct hax} in the execution tree (\cref{sec:landslide-save}),
+creates a checkpoint (\cref{sec:landslide-timetravel}),
+and queries {\tt arbiter\_choose()} to decide which thread to run next (\cref{sec:landslide-arbiter}).
 
@@
 In this section $t_i$ will denote a transition between two program states during execution,
-with each state being a preemption point as identified in \sect{\ref{sec:landslide-pps}},
+with each state being a preemption point as identified in \cref{sec:landslide-pps},
 and $T(t_i)$ will denote the thread which was scheduled \revisionminor{(switched to)} to produce that transition.
@@
 Note that DPOR's notion of happens-before differs from the traditional distributed systems definition \cite{lamport-clocks}
-as used in Pure Happens-Before \sect{\ref{sec:background-hb}};
-rather, it coincides with condition 3 of Limited Happens-Before %as defined in \sect{\ref{sec:background-datarace}}
+as used in Pure Happens-Before \cref{sec:background-hb};
+rather, it coincides with condition 3 of Limited Happens-Before %as defined in \cref{sec:background-datarace}
 (in fact, Landslide's Limited HB implementation simply reuses the same result computed for DPOR's purpose).
@@
 Whenever a conflict is identified,
-\revisionminor{Landslide} also invokes the data race analysis (\sect{\ref{sec:landslide-datarace}}).
+\revisionminor{Landslide} also invokes the data race analysis (\cref{sec:landslide-datarace}).
 It checks for {\tt free()}/access conflicts as well as access pairs,
@@
 of complex algorithms from dense prose and mathematical notation alone,
-mortal readers may prefer the following example of using DPOR on the program from Figure~\ref{fig:concurrency-bug},
-whose original state space is shown in Figure~\ref{fig:tree}.
+mortal readers may prefer the following example of using DPOR on the program from \Cref{fig:concurrency-bug},
+whose original state space is shown in \Cref{fig:tree}.
 In this program both threads are always enabled, imposing no scheduling constraints,
@@
 %I conclude this section with an example of a single iteration of DPOR
-%on the example state space from Figures~\ref{fig:concurrency-bug} and \ref{fig:tree}.
+%on the example state space from \Cref{fig:concurrency-bug,fig:tree}.
 First, let us consider a single iteration of DPOR, applied after executing the first branch.
-The result is shown in Figure~\ref{fig:dpor-example-0}.
+The result is shown in \Cref{fig:dpor-example-0}.
 
@@
 Now, let us run multiple iterations of DPOR
-to advance through the first half of the full state space shown in Figure~\ref{fig:tree}(b).
-Figure~\ref{fig:dpor-example-1} shows the outcome
+to advance through the first half of the full state space shown in \Cref{fig:tree}(b).
+\Cref{fig:dpor-example-1} shows the outcome
 (with the previously-marked $\bigstar$2, now re-labeled as $\bigstar$3, having its subtree abbreviated).
@@
 
-After marking $\bigstar$1 and $\bigstar$3 from Figure~\ref{fig:dpor-example-0}'s interleaving, now labeled A,
+After marking $\bigstar$1 and $\bigstar$3 from \Cref{fig:dpor-example-0}'s interleaving, now labeled A,
 DPOR advances to interleaving B, preferring to schedule the second thread before switching back to the first
@@
 Note also that
-%(as shown in Figure~\ref{fig:tree}(b)),
+%(as shown in \Cref{fig:tree}(b)),
 although B and C also have the same outcome ({\tt x==1}),
@@
 Finally, let us consider the final result after DPOR runs out of remaining unexplored marked branches,
-shown in Figure~\ref{fig:dpor-example-2}.
+shown in \Cref{fig:dpor-example-2}.
 
@@
 DPOR's approach as described so far can still end up testing equivalent interleavings.
-As the presence of more equivalence arrows in Figure~\ref{fig:dpor-example-2} hints,
+As the presence of more equivalence arrows in \Cref{fig:dpor-example-2} hints,
 its reduced subset state space still contains redundancy,
@@
 $\bigstar$1 and $\bigstar$4 contained interleavings equivalent up to independent reorderings anyway.
-Figure~\ref{fig:sleepsets} summarizes the relevant interleavings to highlight one such equivalence.
+\Cref{fig:sleepsets} summarizes the relevant interleavings to highlight one such equivalence.
 
@@
 		{Motivating example for the sleep sets optimization.
-	Three of Figure~\ref{fig:dpor-example-2}'s interleavings are highlighted,
+	Three of \Cref{fig:dpor-example-2}'s interleavings are highlighted,
 	with the always-independent {\tt tmpN++}s omitted for brevity.
@@
 
-Chapter~\ref{chap:quicksand}'s experiments and Chapter~\ref{chap:education}'s user studies
+\Cref{chap:quicksand}'s experiments and \Cref{chap:education}'s user studies
 were conducted before this optimization was implemented.
@@
 and that Landslide showed good bug-finding performance even without it.
-Chapter~\ref{chap:tm}'s experiments include this optimization,
+\Cref{chap:tm}'s experiments include this optimization,
 because its presence was required to fairly compare the other reduction strategies presented therein.
@@
 
-For both the user's convenience and for Quicksand's prioritization algorithm (\sect{\ref{sec:quicksand-id}}),
+For both the user's convenience and for Quicksand's prioritization algorithm (\cref{sec:quicksand-id}),
 Landslide attempts to guess how big partially-explored state spaces will ultimately end up being upon completion.
 Because the backtracking implementation uses checkpointing
-rather than replaying similar interleavings' shared execution prefixes from the beginning \sect{\ref{sec:landslide-timetravel}},
+rather than replaying similar interleavings' shared execution prefixes from the beginning \cref{sec:landslide-timetravel},
 the total number of interleavings (i.e., leaf nodes in the execution tree)
@@
 
-As a concrete example, consider the state space of Figure~\ref{fig:dpor-example-2},
+As a concrete example, consider the state space of \Cref{fig:dpor-example-2},
 and suppose each transition to the next preemption point takes 1 second to execute.
@@
 In principle, both calculate the current progress as a proportion of the expected total
-by counting how many branches DPOR has marked for future exploration (\sect{\ref{sec:landslide-explore}})
+by counting how many branches DPOR has marked for future exploration (\cref{sec:landslide-explore})
 and assuming the sizes of their resulting subtrees are predicted by the known sizes of similar already-explored subtrees.
 In practice, the calculation strategy differs between the two approaches,
-which can occasionally result in drastically differing outputs (\sect{\ref{sec:tm-verif}}).
+which can occasionally result in drastically differing outputs (\cref{sec:tm-verif}).
 
@@
 and time estimates as an ETA.
-Quicksand's {\tt -v} option (\sect{\ref{sec:landslide-quicksand-options}})
+Quicksand's {\tt -v} option (\cref{sec:landslide-quicksand-options})
 will cause it to print them each time a new interleaving is tested; for example:
@@
 %I summarize their respective implementations here.
-% TODO more specific section reference
-Chapter~\ref{chap:warpzone} will discuss their limitations and some opportunities for future improvement.
+\Cref{chap:warpzone-heuristics} will discuss their limitations and some opportunities for future improvement.
 
@@
 To illustrate how the two estimators can under-estimate the total tree size and/or diverge from each other,
-consider the state space from Figure~\ref{fig:dpor-example-2}, of size 6.
+consider the state space from \Cref{fig:dpor-example-2}, of size 6.
 Suppose for RE that each transition takes 1 second to execute.
@@
 consider the same state space,
-except with the $\bigstar$4 subtree also pruned by DPOR's sleep sets extension (\sect{\ref{sec:landslide-sleepsets}});
+except with the $\bigstar$4 subtree also pruned by DPOR's sleep sets extension (\cref{sec:landslide-sleepsets});
 i.e., only branches A, B, C, and K remain,
@@
 and happens-before edges, for pure happens-before
-(\sect{\ref{sec:background-hb}}).
+(\cref{sec:background-hb}).
 
@@
 
-Conditions \#1, \#2, and \#4 defined in \sect{\ref{sec:background-hb}},
+Conditions \#1, \#2, and \#4 defined in \cref{sec:background-hb},
 provided the Limited Happens-Before definition for \#4,
@@
 
-First of all, note that Algorithm~\ref{alg:icb} is structured in a way that repeats interleavings
+First of all, note that \Cref{alg:icb} is structured in a way that repeats interleavings
 with fewer than $n$ preemptions that have already been checked in previous iterations of the outer loop.
@@
 ICB in principle reduces the state space from exponentially-sized%
-\footnote{Combinatorial, to be precise; see \sect{\ref{sec:landslide-dpor-example}}.}
+\footnote{Combinatorial, to be precise; see \cref{sec:landslide-dpor-example}.}
 in both $K$, the number of threads,
@@
 Prior work often recommends 2 for such a cutoff \cite{chess-icb,smc-empirical-study,dejafu},
-although \sect{\ref{sec:quicksand-eval}}'s larger dataset suggests 3 would be considerably more thorough.
+although \cref{sec:quicksand-eval}'s larger dataset suggests 3 would be considerably more thorough.
 On the other hand, any finite such bound can provide only a heuristic verification guarantee. %anymore.
@@
 Landslide checks for two distinct categories of \revisionminor{potentially-infinite} loops,
-visualized in Figure~\ref{fig:heuristic-loop}.
+visualized in \Cref{fig:heuristic-loop}.
 Firstly, it keeps a running average of how many preemption points deep each completed interleaving has been in the past.
@@
 or primitives built thereupon,
-{\tt mutex\_test} (\sect{\ref{sec:education-pebbles-tests}})
+{\tt mutex\_test} (\cref{sec:education-pebbles-tests})
 often relies on this functionality, as data-race preemption points within {\tt mutex\_lock()}
@@
 Secondly, in programs which use HTM,
-threads blocked by retry sets (\sect{\ref{sec:tm-retrysets}})
+threads blocked by retry sets (\cref{sec:tm-retrysets})
 must be forced awake with priority before yield-blocked threads,

### 2018-12-03 thesis/overview.tex
use cleveref

@@
 
-Consider the program in Figure \ref{fig:concurrency-bug}: Any output between 2 and 2000 is possible\footnote{
+Consider the program in \Cref{fig:concurrency-bug}: Any output between 2 and 2000 is possible\footnote{
 	Fun exercise for the reader: Show why 2 is a possible output, but 1 is not!
@@
 
-To prevent unexpected interleavings such as the example in Figure~\ref{fig:concurrency-bug}(b),
+To prevent unexpected interleavings such as the example in \Cref{fig:concurrency-bug}(b),
 most concurrent programs use {\em concurrency primitives} to control which interleavings are possible.
@@
 		&
-		(b) The {\tt count} function from Figure~\ref{fig:concurrency-bug}, adjusted to use a mutex to ensure each increment of {\tt x} is uninterruptible.
+		(b) The {\tt count} function from \Cref{fig:concurrency-bug}, adjusted to use a mutex to ensure each increment of {\tt x} is uninterruptible.
 	\end{tabular}
@@
 (or the set of false positive data-race candidates reported, etc.;
-\revisionminor{see \sect{\ref{sec:landslide-config-landslide}}})
+\revisionminor{see \cref{sec:landslide-config-landslide}})
 at the cost of increasing the implementation and theoretical complexity of the analysis
-\revisionminor{(\sect{\ref{sec:tm-abstraction}})}.
+\revisionminor{(\cref{sec:tm-abstraction})}.
 This thesis will consider locks, descheduling, and transaction begin/end as the only concurrency primitives,
@@
 will \revisionminor{block (i.e., wait)} until the corresponding {\tt mutex\_unlock(mp)}.
-Figure~\ref{fig:mutex}(a) shows how a yielding mutex (not the best implementation, but the simplest) may be implemented using {\tt xchg},
-and (b) shows how a mutex may be used to fix the example from Figure~\ref{fig:concurrency-bug}.
+\Cref{fig:mutex}(a) shows how a yielding mutex (not the best implementation, but the simplest) may be implemented using {\tt xchg},
+and (b) shows how a mutex may be used to fix the example from \Cref{fig:concurrency-bug}.
 
@@
 offering three new instructions: \texttt{xbegin}, \texttt{xend}, and \texttt{xabort}, to begin, commit, and fail a transaction, respectively.
-The example program in Figure~\ref{fig:htm-example} demonstrates how these primitives can be used to synchronize a simple shared access without locking overhead in the common case\footnote{
-	The solution presented here is actually incomplete; stay tuned until Chapter~\ref{chap:tm} for the surprising twist!
+The example program in \Cref{fig:htm-example} demonstrates how these primitives can be used to synchronize a simple shared access without locking overhead in the common case\footnote{
+	The solution presented here is actually incomplete; stay tuned until \Cref{chap:tm} for the surprising twist!
 },
@@
 	\end{center}
-	\caption[The example {\tt count} routine from Figure~\ref{fig:mutex}, rewritten to use HTM.]
-		{The example {\tt count} routine from Figure~\ref{fig:mutex}, rewritten to use HTM.
+	\caption[The example {\tt count} routine from \Cref{fig:mutex}, rewritten to use HTM.]
+		{The example {\tt count} routine from \Cref{fig:mutex}, rewritten to use HTM.
 		If the transaction in the top branch aborts,
@@
 HTM, however, is backed by the CPU's cache, and is therefore subject to other circumstances such as cache capacity or interrupt-triggered cache flushes which may force an abort even when no memory conflict occurs.
-Chapter~\ref{chap:tm} will explore the consequences of this difference further.
+\Cref{chap:tm} will explore the consequences of this difference further.
 %
@@
 To visualize this,
-using a single iteration of the {\tt x++;} loop from Figure~\ref{fig:concurrency-bug} as an example,
+using a single iteration of the {\tt x++;} loop from \Cref{fig:concurrency-bug} as an example,
 with {\tt x++;} expanded into its three corresponding pseudo-assembly instructions,
-Figure~\ref{fig:tree}(a) shows all possible execution interleavings
+\Cref{fig:tree}(a) shows all possible execution interleavings
 %of the compiled code
@@
 	\end{tabular}
-	\caption[Visualization of interleaving state space for the program in Figure~\ref{fig:concurrency-bug}.]
-		{Visualization of interleaving state space for the program in Figure~\ref{fig:concurrency-bug}.
+	\caption[Visualization of interleaving state space for the program in \Cref{fig:concurrency-bug}.]
+		{Visualization of interleaving state space for the program in \Cref{fig:concurrency-bug}.
 	Thread~1 is represented by purple ovals, thread~2 by yellow squares, and time flows from top to bottom.
@@
 the set of possible ways to execute the program can be viewed as a tree.
-Figure~\ref{fig:tree}(b) shows a visualization of the corresponding tree from our example program,
+\Cref{fig:tree}(b) shows a visualization of the corresponding tree from our example program,
 using each pseudo-assembly instruction as a preemption point.
@@
 Addressing the scaling problem in this exponential relation is the central research problem for all model checkers,
-\revisionminor{which this thesis will address in Chapter~\ref{chap:quicksand}}.
+\revisionminor{which this thesis will address in \Cref{chap:quicksand}}.
 
@@
 {\bf Concretely speaking:}
-Figure~\ref{fig:dpor} highlights part of an execution tree where the execution ordering of threads 1 and 2 are swapped,
+\Cref{fig:dpor} highlights part of an execution tree where the execution ordering of threads 1 and 2 are swapped,
 and each interleaving has a respective ``subtree'' (i.e., possible interleavings given the fixed execution prefix leading up to it).
@@
 in order to check all possible program states.
-\sect{\ref{sec:landslide-dpor}} explains how DPOR implements such a deduction in more detail.
+\cref{sec:landslide-dpor} explains how DPOR implements such a deduction in more detail.
 
@@
 although it requires some human intuition to correctly mark those boundaries.
-Iterative Deepening, presented in Chapter~\ref{chap:quicksand}, is another such search heuristic.
+Iterative Deepening, presented in \Cref{chap:quicksand}, is another such search heuristic.
 
@@
 \end{enumerate}
-%In Figure~\ref{fig:example}, lines 3 and 5 each race with 2 and 6, and line 6 races with 8.
-In Figure~\ref{fig:hb-example}, the pairs of lines marked with comments (A1 and A2, B1 and B2) race.
+%In \Cref{fig:example}, lines 3 and 5 each race with 2 and 6, and line 6 races with 8.
+In \Cref{fig:hb-example}, the pairs of lines marked with comments (A1 and A2, B1 and B2) race.
 
@@
 it cannot identify access pairs separated by an unrelated lock operation which could race in an alternate interleaving,
-as shown in the example program in Figure~\ref{fig:hb-example}(a).
+as shown in the example program in \Cref{fig:hb-example}(a).
 We call such unreported access pairs {\em false negatives}.
@@
 by considering only blocking operations like {\tt cond\_wait} to enforce the order.
-However, consider the similar program in Figure~\ref{fig:hb-example}(b),
+However, consider the similar program in \Cref{fig:hb-example}(b),
 in which the access pair ceases to exist in the alternate interleaving.
@@
 critical sections to determine whether lock events are unrelated to a potential race.
-Causally-Precedes will identify the potential race in Figure~\ref{fig:hb-example}(a), as the two critical sections do not conflict,
+Causally-Precedes will identify the potential race in \Cref{fig:hb-example}(a), as the two critical sections do not conflict,
 although it can still miss true potential races in other cases.
@@
 Landslide implements both Happens-Before (henceforth referred to as {\em Pure Happens-Before} for clarity) and Limited Happens-Before.
-Chapter~\ref{chap:quicksand} includes a comparison of the two approaches for the purpose of finding new preemption points for model checking.
+\Cref{chap:quicksand} includes a comparison of the two approaches for the purpose of finding new preemption points for model checking.
 %, we use the Limited Happens-Before relation for our analysis.
@@
 The course's prerequisite is 15-213, Introduction to Computer Systems \cite{csapp,sigcse01:CSaPP}.
-Both P2 and P3 are built using the {\em Pebbles} system call specification, outlined in Table~\ref{tab:syscalls}.
+Both P2 and P3 are built using the {\em Pebbles} system call specification, outlined in \Cref{tab:syscalls}.
 
@@
 \subsubsection{P3}
-In P3, students implement a kernel which provides the same system calls shown in Table~\ref{tab:syscalls}, previously provided by the reference kernel.
+In P3, students implement a kernel which provides the same system calls shown in \Cref{tab:syscalls}, previously provided by the reference kernel.
 Pebbles adopts the Mach \cite{DBLP:conf/usenix/AccettaBBGRTY86} distinction between {\em tasks}, which are resource containers, and {\em threads}, each of which executes within a single task.
@@
 %
-One such example is the {\tt paraguay} unit test distributed with P2 (\sect{\ref{sec:education-pebbles-tests}}),
+One such example is the {\tt paraguay} unit test distributed with P2 (\cref{sec:education-pebbles-tests}),
 which targets a subtle condition-variable bug.
@@
 		Distinguished further in this thesis as {\em limited happens-before} races and {\em pure happens-before} races;
-		see \sect{\ref{sec:background-datarace}}.
+		see \cref{sec:background-datarace}.
 		Distinct from the more nebulously-defined {\em concurrency bug},
@@
 		but are necessary for {\em Iterative Deepening}'s {\em verification} guarantee
-		(\sect{\ref{sec:quicksand-soundness}}).
+		(\cref{sec:quicksand-soundness}).
 	\item {\bf Deadlock.}
@@
 		for future exploration, skipping over ones guaranteed to be {\em independent}.
-		See \sect{\ref{sec:landslide-dpor}}.
+		See \cref{sec:landslide-dpor}.
 	\item {\bf Equivalent.}
@@
 		Output may be in units of either time or branches.
-		See \sect{\ref{sec:landslide-estimate}}.
+		See \cref{sec:landslide-estimate}.
 	\item {\bf ETA.}
@@
 		into a subtree with a branch at each point they diverge, forming a tree overall.
-		See \sect{\ref{sec:overview-statespace}}; Figure~\ref{fig:tree}.
+		See \cref{sec:overview-statespace}; \Cref{fig:tree}.
 	\item {\bf False negative.}
@@
 		in practice, however, infinite loop detection and deadlock detection in the presence of ad-hoc synchronization
-		are both {\em heuristic} (\sect{\ref{sec:landslide-blocking}}),
+		are both {\em heuristic} (\cref{sec:landslide-blocking}),
 		so may produce false positives after all, although I have never witnessed one to date.
@@
 		that can exhibit such extreme interleaving-dependent variance.
-	\item {\bf ICB.} Short for ``Iterative Context Bounding''. See \sect{\ref{sec:landslide-icb}}.
+	\item {\bf ICB.} Short for ``Iterative Context Bounding''. See \cref{sec:landslide-icb}.
 	\item {\bf Independent.}
@@
 		One of this thesis's major contributions.
-		See \sect{\ref{sec:quicksand-id}}.
+		See \cref{sec:quicksand-id}.
 	\item {\bf Kernelspace.}
@@
 		The framework upon which all this thesis's contributions are built.
-		Documented in Chapter \ref{chap:landslide}.
+		Documented in \Cref{chap:landslide}.
 	\item {\bf Landslide-friendly.}
@@
 		with respect to other {\em threads} using the same lock.
-		See \sect{\ref{sec:overview-mutex}}.
+		See \cref{sec:overview-mutex}.
 	\item {\bf Memory conflict.}
@@
 		although exactly how thread scheduling changes according to each specified mode
-		is kept secret to preserve the intended challenge of debugging (see \sect{\ref{sec:410-secrecy}}).}
+		is kept secret to preserve the intended challenge of debugging (see \cref{sec:410-secrecy}).}
 	\item {\bf mutex.}
@@
 		even though the latter did not call it by that name.
-		See \sect{\ref{sec:pebbles}}.
+		See \cref{sec:pebbles}.
 	\item \revision{{\bf P3.}
@@
 	\item \revision{{\bf Pebbles.}
-		The educational kernel architecture used in 15-410 at CMU. See \sect{\ref{sec:pebbles}}.}
+		The educational kernel architecture used in 15-410 at CMU. See \cref{sec:pebbles}.}
 	\item \revision{{\bf Pintos.}
 		The educational kernel architecture used in \uchos at the University of Chicago,
-		as well as several other universities. See \sect{\ref{sec:overview-pintos}}.}
+		as well as several other universities. See \cref{sec:overview-pintos}.}
 	\item {\bf pthreads.}
@@
 		or to further restrict by predicates on the current stack trace.
-		See \sect{\ref{sec:landslide-pps}}.
+		See \cref{sec:landslide-pps}.
 		Often called {\em preemption point} for short, with context hopefully making the distinction clear.
@@
 		Also supports various other execution modes in which only one Landslide instance is run at a time.
-		See Chapter~\ref{chap:quicksand}.
+		See \Cref{chap:quicksand}.
 		In later chapters, ``Landslide'' will refer to the project as a whole, including Quicksand,
@@
 		% TODO: check that quicksand discussion shouldn't perhaps get a more subsection'ed sect reference
-		For further discussion see \sect{\ref{sec:quicksand-discussion}} and \sect{\ref{sec:tm-warpzone-relaxed}}.}
+		For further discussion see \cref{sec:quicksand-discussion} and \cref{sec:tm-warpzone-relaxed}.}
 	\item {\bf Schedule.}
@@
 		to exhaustively check all possible {\em thread interleavings}.
-		See \sect{\ref{sec:overview-stateless}}.
+		See \cref{sec:overview-stateless}.
 		Often ``MC'' for short in this thesis (to mean either ``model checking'' or ''model checker'' depending on context).

### 2018-12-03 thesis/related.tex
use cleveref

@@
 instruments every memory access,
-uses explicit state tracking rather than the stateless approach (\sect{\ref{sec:overview-stateless}}),
+uses explicit state tracking rather than the stateless approach (\cref{sec:overview-stateless}),
 and specializes in verifying synchronization primitives such as RCU \cite{rcu}.
@@
 They also identify all data races
-% TODO check if ther'es not a better section reference
-(under the C++ definition rather than \sect{\ref{sec:quicksand-soundness}}'s)
+(under the C++ definition rather than \cref{sec:quicksand-soundness}'s; see \cref{sec:background-hb})
 as bugs immediately, rather than checking them for benign or buggy outcomes.
@@
 It is a {\em sound} reduction algorithm, meaning it will never fail to test a possible program behavior, despite skipping many execution sequences.
-\sect{\ref{sec:landslide-dpor}} provides a detailed walk-through of how DPOR works,
+\cref{sec:landslide-dpor} provides a detailed walk-through of how DPOR works,
 as many of this thesis's contributions build directly upon it.
@@
 Several other recent advances extend DPOR to new concurrency models,
-beyond the shared-memory-threading model outlined in \sect{\ref{sec:landslide-dpor}}.
+beyond the shared-memory-threading model outlined in \cref{sec:landslide-dpor}.
 TransDPOR \cite{transdpor} provides extra domain-specific reduction for message-passing actor programs
@@
 which can expose bugs not even possible in the strong consistency model
-such as discussed in \sect{\ref{sec:tm-warpzone-relaxed}}.
+such as discussed in \cref{sec:tm-warpzone-relaxed}.
 It also introduced a heuristic {\em buffer bounding} technique, analogous to ICB,
@@
 offering further reduction still.
-Somewhat analogously for HTM, this work's Chapter~\ref{chap:tm}
+Somewhat analogously for HTM, this work's \Cref{chap:tm}
 extended DPOR's concurrency model to include failure injection,
@@
 arising from trusted source code modules;
-Landslide implements this as the {\tt without\_function} command (\sect{\ref{sec:landslide-pps}}).
-Iterative Context Bounding (ICB) \cite{chess-icb} (\sect{\ref{sec:landslide-icb}})
+Landslide implements this as the {\tt without\_function} command (\cref{sec:landslide-pps}).
+Iterative Context Bounding (ICB) \cite{chess-icb} (\cref{sec:landslide-icb})
 orders the search space by increasing number of preemptions in each branch,
@@
 thereunder. % what a word
-Landslide implements ICB and BPOR for Chapter~\ref{chap:quicksand}'s control experiments,
+Landslide implements ICB and BPOR for \Cref{chap:quicksand}'s control experiments,
 although does not yet incorporate it into this thesis's own contributions
-(as discussed in Chapter~\ref{chap:warpzone}).
-Chapter~\ref{chap:quicksand}'s Quicksand algorithm is, effectively, another such heuristic search strategy,
+(as discussed in \Cref{chap:warpzone}).
+\Cref{chap:quicksand}'s Quicksand algorithm is, effectively, another such heuristic search strategy,
 focusing on preemption point subsets rather than context switch bounding.
@@
 while dBug \cite{dbug-phdthesis} applied abstraction reduction to synchronization primitives,
-and I showed how it could be applied similarly to transactional memory in \sect{\ref{sec:tm-abstraction}}.
+and I showed how it could be applied similarly to transactional memory in \cref{sec:tm-abstraction}.
 Each of these approaches is compatible (and indeed, throughout this thesis used often in concert)
@@
 which Landslide more borrows as building blocks for its own methods rather than contributing new techniques to.
-% TODO fix terminology not to conflict with dpor
-Race detectors are largely distinguished by their particular flavour of the Happens-Before (HB) relation,
-as discussed in \sect{\ref{sec:background-hb}}.
+Data race detectors are largely distinguished by their particular flavour of the Happens-Before (HB) relation,
+as discussed in \cref{sec:background-hb}.
 Djit+ \cite{djit} and FastTrack \cite{fasttrack} are among those
@@
 The Causally-Precedes relation \cite{predictive-dr} is a refinement of Limited HB which avoids the most common cases of false positives,
-% TODO section refrance
-including \sect{\ref{sec:quicksand-soundness}}'s reallocation false positives.
+including \cref{sec:quicksand-realloc}'s reallocation false positives.
 It could strike a middle ground in the bug-finding/verification tradeoff
-between Pure and Limited HB (\sect{\ref{sec:quicksand-eval}})
+between Pure and Limited HB (\cref{sec:quicksand-eval})
 that would be a welcome enhancement in Quicksand.
@@
 test alternate interleavings in advance of knowing any specific data races,
-% TODO more speicifc section refrance
-which \sect{\ref{sec:quicksand-eval}} showed is necessary to find certain bugs.
+which \cref{sec:quicksand-eval-nondets} showed is necessary to find certain bugs.
 Quicksand builds on Portend's approach by introducing a feedback loop between the data-race analysis and model checking,
 which results in a stronger verification property
-when the test can be fully completed (\sect{\ref{sec:quicksand-soundness}}).
+when the test can be fully completed (\cref{sec:quicksand-soundness}).
 Portend also uses symbolic execution to test input nondeterminism as well as schedule nondeterminism,
@@
 to offer a ``warm up'' Landslide assignment
-(for example, the {\tt atomic\_*} tests from \sect{\ref{sec:education-pebbles-tests}})
-that could ultimately lead to a higher solve rate on Landslide's bug reports during P2 (\sect{\ref{sec:education-eval-bugs-cmu}}).
+(for example, the {\tt atomic\_*} tests from \cref{sec:education-pebbles-tests})
+that could ultimately lead to a higher solve rate on Landslide's bug reports during P2 (\cref{sec:education-eval-bugs-cmu}).
 
@@
 it demonstrates the kind of user-friendly features and overall approach
-that many students expressed desire for in their survey responses (\sect{\ref{sec:education-eval-survey}}).
+that many students expressed desire for in their survey responses (\cref{sec:education-eval-survey}).
 Future work could extend Landslide's preemption traces to include similar interactivity.}
@@
 This kind of verification,
-analogous to \sect{\ref{sec:education-pebbles-tests}}'s {\tt mutex\_test},
+analogous to \cref{sec:education-pebbles-tests}'s {\tt mutex\_test},
 is an important stepping stone for trusting the results Landslide will provide.
@@
 used by many papers in the field to evaluate the performance of both STM and HTM implementations alike.
-As discussed in \sect{\ref{sec:tm-eval-exp-setup}},
+As discussed in \cref{sec:tm-eval-exp-setup},
 it focuses more on performance than on interesting concurrency properties.
@@
 relying on conflict aborts to point out access pairs that would be unsafe in the original program.
-This citation arguably belongs in \sect{\ref{sec:related-data-race}} as well;
+This citation arguably belongs in \cref{sec:related-data-race} as well;
 I include it here to highlight the importance of Landslide's
-ability to distinguish different abort reasons (\sect{\ref{sec:txn-abort-codes}}).
+ability to distinguish different abort reasons (\cref{sec:txn-abort-codes}).
 
@@
 although these pitfalls depend on multiple existing locks used locally and disjointly,
-so this does not invalidate the equivalence proved in \sect{\ref{sec:tm-design-locks}}.
+so this does not invalidate the equivalence proved in \cref{sec:tm-design-locks}.
 Rather, Landslide could be used to ensure that freshly-converted transactional code avoids the warned-of pitfalls.

### 2018-12-03 thesis/thesis.tex
use cleveref

@@
 \usepackage{microtype}
-\usepackage{cleveref}
+\usepackage[nameinlink]{cleveref}
 
@@
 
-% % nwf made this but i don't know how to get it to work?
-% \usepackage{etoolbox}
-% \makeatletter
-% \apptocmd{\ref}{%
-%   \immediate\write\@auxout%
-%     {\expandafter\string\csname nwf@backref\endcsname{#1}{\arabic{page}}}%
-% }{}{}
-%
-% \newcommand{\nwf@backref}[2]{%
-%   \ifcsdef{nwf@backrefs@#1}%
-%    {\csgappto{nwf@backrefs@#1}{, #2}}%
-%    {\csgdef{nwf@backrefs@#1}{#2}}%
-% }
-% \AtEndDocument{\let\nwf@backref\relax}
-%
-% \newcommand{\linksto}[1]{\csname nwf@backrefs@#1\endcsname}
-% \makeatother
+% NB. needs to be defined before mainmatter so they work in hyperref
+\definecolor{citec}{RGB}{128,0,64}
+\definecolor{linkc}{RGB}{0,64,128}
+\definecolor{urlc} {RGB}{128,64,0}
 
-% NB. needs to be defined before mainmatter so they work in hyperref
-\definecolor{citec}{RGB}{168,0,84}
-\definecolor{linkc}{RGB}{0,84,168}
-\definecolor{urlc} {RGB}{168,84,0}
+% format section references with no space after the \S
+% #1 is the ref number; the #2 and #3 here are the hyperlink boundaries
+\crefformat{section}{#2\S{}#1#3}
+\Crefformat{section}{#2\S{}#1#3}
 
@@
 Michael J. Sullivan and Nathaniel Wesley Filardo were Landslide's first users besides myself back in its prototype days.
-Joshua Wise, Miriam Zimmerman, and Glenn Willen helped revise my explanation of DPOR in Chapter~\ref{chap:landslide}.
-Professor Brandon Lucia provided valuable comments and revisions to Chapter~\ref{chap:quicksand}'s content
+Joshua Wise, Miriam Zimmerman, and Glenn Willen helped revise my explanation of DPOR in \Cref{chap:landslide}.
+Professor Brandon Lucia provided valuable comments and revisions to \Cref{chap:quicksand}'s content
 that ultimately led to its publication at OOPSLA 2016.
@@
 and helped greatly to improve Landslide's stability and robustness for use beyond CMU's walls,
-greatly enriching Chapter~\ref{chap:education}'s evaluation.
+greatly enriching \Cref{chap:education}'s evaluation.
 David Blum and Skye Toor provided
@@
 transactional memory data structures
-that make up about half of Chapter~\ref{chap:tm}'s evaluation suite.
+that make up about half of \Cref{chap:tm}'s evaluation suite.
 %

### 2018-12-03 thesis/tm.tex
use cleveref

@@
 
-The example TSX program from Figure~\ref{fig:htm-example} (\sect{\ref{sec:overview-tm}})
+The example TSX program from \Cref{fig:htm-example} (\cref{sec:overview-tm})
 is reproduced here for the reader's convenience.
@@
 and hence assumes sequential consistency.
-For discussion of relaxed memory models refer to Section~\ref{sec:tm-warpzone-relaxed}.
+For discussion of relaxed memory models refer to \cref{sec:tm-warpzone-relaxed}.
 
@@
 
-Consider again the program in Figure~\ref{fig:htm-example-reproduced}.
+Consider again the program in \Cref{fig:htm-example-reproduced}.
 Note that the C-style {\tt x++} operations, when compiled into assembly, %\cite{printy},
 become multiple memory accesses which can be interleaved with other threads,
-\revisionminor{as shown below in Figure~\ref{fig:htm-example-assemblified}.}
+\revisionminor{as shown below in \Cref{fig:htm-example-assemblified}.}
 
@@
 	\end{center}
-	\caption{Figure~\ref{fig:htm-example-reproduced} with non-atomic operations shown as pseudo-assembly.}
+	\caption{\Cref{fig:htm-example-reproduced} with non-atomic operations shown as pseudo-assembly.}
 	\label{fig:htm-example-assemblified}
@@
 	\end{center}
-	\caption[Impossible hypothetical interleaving within transactions of Figure~\ref{fig:htm-example-reproduced}.]
-		{Impossible hypothetical interleaving within transactions of Figure~\ref{fig:htm-example-reproduced}.
+	\caption[Impossible hypothetical interleaving within transactions of \Cref{fig:htm-example-reproduced}.]
+		{Impossible hypothetical interleaving within transactions of \Cref{fig:htm-example-reproduced}.
 		The effects of $\tiat{2a-2b}$ are never globally visible, as its transasction must abort,
@@
 the transaction will abort.
-So, a hypothetical interleaving such as shown in Figure~\ref{fig:htm-example-interleaving},
+So, a hypothetical interleaving such as shown in \Cref{fig:htm-example-interleaving},
 or equivalently written sequentially as:
@@
 \]
-Prior work \cite{tm-benchmark-cmu} proposed the idiom shown in Figure~\ref{fig:htm-fixed}
+Prior work \cite{tm-benchmark-cmu} proposed the idiom shown in \Cref{fig:htm-fixed}
 to exclude this family of interleavings,
@@
 	\end{center}
-	\caption[Variant of the program in Figure~\ref{fig:htm-example-reproduced}.]
-		{Variant of the program in Figure~\ref{fig:htm-example-reproduced},
+	\caption[Variant of the program in \Cref{fig:htm-example-reproduced}.]
+		{Variant of the program in \Cref{fig:htm-example-reproduced},
 		with additional synchronization to protect the failure path from the transactional path.
@@
 	satisfying case 1.
-	Otherwise, by DPOR's definition of transition dependence (\cite{dpor}, \sect{\ref{sec:landslide-dpor-hb}}),
+	Otherwise, by DPOR's definition of transition dependence (\cite{dpor}, \cref{sec:landslide-dpor-hb}),
 	$\tiiat{\beta_{b+1}}\dots\tiiat{\beta_n}$ is independent with the transitions of \tjj and \tkk,
@@
 	For every $I \in S$ with $\tiiat{\alpha},\tiiat{\beta_1}\dots\tiiat{\beta_b},$ $\tjjat{\dots},\tkkat{\dots},\tiiat{\beta_{b+1}} \in I$,
-	apply Lemma~\ref{lem:equiv} to obtain an equivalent interleaving $I'$ satisfying the theorem condition.
+	apply \Cref{lem:equiv} to obtain an equivalent interleaving $I'$ satisfying the theorem condition.
 	The resulting $S'$ can then be MCed without ever simulating HTM rollbacks.
@@
 
-Next comes the issue of memory accesses within transactions with regard to data-race analysis (\sect{\ref{sec:background-datarace}}).
-Theorem~\ref{thm:atom} provides that the body of all transactions may be executed atomically within the MC environment.
+Next comes the issue of memory accesses within transactions with regard to data-race analysis (\cref{sec:background-datarace}).
+\Cref{thm:atom} provides that the body of all transactions may be executed atomically within the MC environment.
 While they may interleave between other non-transactional sequences,
@@
 Let $\tiiat{\mu},\tjjat{\nu}$ be a pair of memory accesses to the same address, at least one a write,
-in some transactional execution $I$ normalized under Lemma~\ref{lem:equiv}.
+in some transactional execution $I$ normalized under \Cref{lem:equiv}.
 Then let $\mathsf{lockify}_m(\tkkat{L})$ denote a function over instructions in $I$,
@@
 the execution with the boundaries of all successful transactions replaced by an abstract global lock.
-Lemma~\ref{lem:equiv} guarantees mutual exclusion of $m$.
+\Cref{lem:equiv} guarantees mutual exclusion of $m$.
 \vspace{1em}
@@
 				then they cannot both be in successful transactions,
-				or else Lemma~\ref{lem:equiv} normalization would provide
+				or else \Cref{lem:equiv} normalization would provide
 				the corresponding HTM end and begin for $\tiiat{\epsilon}$ and $\tjjat{\chi}$ respectively.
@@
 and the {\tt 410TESTS} or {\tt STUDENTTESTS} line (respectively) of {\tt config-incomplete.mk} be edited to add the test name,
-before running {\tt p2-setup.sh} (\sect{\ref{sec:landslide-setup}})
+before running {\tt p2-setup.sh} (\cref{sec:landslide-setup})
 % TODO(implementation): Provide an obfuscated refp2?
@@
 		(always returning {\tt \_XABORT\_RETRY} as the failure code)
-		(\revisionminor{\sect{\ref{sec:tm-design-failure}}}, \sect{\ref{sec:txn-failure}}),
-		and treatment of transactional regions during data-race analysis (\sect{\ref{sec:txn-datarace}}).
+		(\revisionminor{\cref{sec:tm-design-failure}}, \cref{sec:txn-failure}),
+		and treatment of transactional regions during data-race analysis (\cref{sec:txn-datarace}).
 	\item {\tt -A} (for ``xAbort codes'')
-		enables multiple xabort failure codes (\sect{\ref{sec:txn-failure}}).
+		enables multiple xabort failure codes (\cref{sec:txn-failure}).
 		Requires {\tt -X}.
@@
 		enables state space reduction based on independences between individual {\tt xbegin} results
-		(\sect{\ref{sec:tm-retrysets}}).
+		(\cref{sec:tm-retrysets}).
 		Requires {\tt -X} and {\em not} {\tt -A}.
@@
 (serving an analogous purpose as the {\tt all\_explored} flag for thread scheduling).
-The state space estimators (\sect{\ref{sec:landslide-estimate}})
+The state space estimators (\cref{sec:landslide-estimate})
 check the length of these lists, in addition to DPOR's tagged threads,
@@
 depends on currently-unsupported {\tt xbegin} nesting, which I discuss further in
-\sect{\ref{sec:tm-warpzone-nested}}.
+\cref{sec:tm-warpzone-nested}.
 
 \subsubsection{Injecting abort codes}
-When traversing the state space (\sect{\ref{sec:landslide-statespace}}),
-in addition to performing DPOR to select non-independent thread interleavings (\sect{\ref{sec:landslide-dpor}}),
+When traversing the state space (\cref{sec:landslide-statespace}),
+in addition to performing DPOR to select non-independent thread interleavings (\cref{sec:landslide-dpor}),
 the abort codes under each {\tt xbegin} preemption point are also considered ``marked'' paths which must be tested.
@@
 (analogous to {\tt mutex\_trylock()}),
-Landslide's scheduler sets the {\tt user\_txn} action flag for that thread (\sect{\ref{sec:landslide-scheduler}}),
+Landslide's scheduler sets the {\tt user\_txn} action flag for that thread (\cref{sec:landslide-scheduler}),
 and if using Pure Happens-Before,
-applies \textsc{FT acquire} (\sect{\ref{sec:landslide-phb}}) using a dummy lock address to represent the abstract global lock.
+applies \textsc{FT acquire} (\cref{sec:landslide-phb}) using a dummy lock address to represent the abstract global lock.
 When a thread reaches {\tt \_xend()},
@@
 will fail to identify and prune equivalent thread interleavings.
-Figure~\ref{fig:retry-set-example} shows a minimal example program which exhibits this problem.
+\Cref{fig:retry-set-example} shows a minimal example program which exhibits this problem.
 In this program, each thread's transactional path conflicts with the other thread's abort path,
@@
 
-While Figure~\ref{fig:retry-set-example}'s example may seem contrived
+While \Cref{fig:retry-set-example}'s example may seem contrived
 (what program's transactions would just give up and do no work if aborted?),
@@
 Also, the evaluation will later show
-(\sect{\ref{sec:tm-eval-retrysets}})
+(\cref{sec:tm-eval-retrysets})
 that even programs with fully conflicting success/abort paths
@@
 an experimental feature named after prior work's analogous {\em sleep sets}
-(\cite{partial-order-methods,dpor,optimal-dpor}, \sect{\ref{sec:landslide-sleepsets}}).
-Whenever DPOR tags a new branch for exploration (\sect{\ref{sec:landslide-explore}}),
+(\cite{partial-order-methods,dpor,optimal-dpor}, \cref{sec:landslide-sleepsets}).
+Whenever DPOR tags a new branch for exploration (\cref{sec:landslide-explore}),
 if either or both sides of a memory conflict were part of a transaction,
@@
 to accompany that branch.
-Like the {\tt xabort\_codes} lists, the state space estimators (\sect{\ref{sec:landslide-estimate}})
+Like the {\tt xabort\_codes} lists, the state space estimators (\cref{sec:landslide-estimate})
 check the number of retry sets when counting the number of marked children.
@@
 	Before the search ordering update to Landslide's normal sleep sets implementation
-	(\sect{\ref{sec:landslide-sleepsets}}),
+	(\cref{sec:landslide-sleepsets}),
 	I observed false-positive retry-set-blocked deadlocks fairly often;
@@
 %The {\tt landslide} repository's {\tt retry-sets} branch implements this feature,
-Upcoming in the evaluation, Figure~\ref{fig:retry-reduction} will visualize the reduction achieved in two test cases.
-
-% In contrast to Landslide's implementation of the sleep set reduction (\sect{\ref{sec:landslide-explore}}),
+Upcoming in the evaluation, \Cref{fig:retry-reduction} will visualize the reduction achieved in two test cases.
+
+% In contrast to Landslide's implementation of the sleep set reduction (\cref{sec:landslide-explore}),
 % blah blah which is implicit and works in retrospect by pruning at tag-sibling time, which figures out what would have been in a sleep set by analyzing the past independences and explored siblings,
@@
 Accordingly, there is no baseline against which to compare Landslide's performance.
-Likewise, since Landslide's HTM semantics emulation relies on the equivalences proved in \sect{\ref{sec:tm-design}},
+Likewise, since Landslide's HTM semantics emulation relies on the equivalences proved in \cref{sec:tm-design},
 I did not actually implement an HTM-style speculative-execution-and-rollback simulation mode.
@@
 	%\item How quickly does Landslide verify correct transactional programs of varying sizes?
-	\item How does Iterative Deepening's (\sect{\ref{sec:quicksand-id}})
-		performance compare to Maximal State Space mode (\sect{\ref{sec:landslide-quicksand-options}})?%
+	\item How does Iterative Deepening's (\cref{sec:quicksand-id})
+		performance compare to Maximal State Space mode (\cref{sec:landslide-quicksand-options})?%
 		\footnote{Not necessarily related to HTM,
-		but the latter was implemented well after Chapter~\ref{chap:quicksand}'s conference paper was published,
+		but the latter was implemented well after \Cref{chap:quicksand}'s conference paper was published,
 		so this was the most convenient test suite to evaluate it on.}
@@
 	\begin{itemize}
-		\item {\tt htm1}: The bug from Figure~\ref{fig:htm-example}/\ref{fig:htm-example-reproduced}.
+		\item {\tt htm1}: The bug from \Cref{fig:htm-example,fig:htm-example-reproduced}.
 			%(assertion failure on {\tt count}).
-		\item {\tt htm2}: The fixed version as in Figure~\ref{fig:htm-fixed}.
+		\item {\tt htm2}: The fixed version as in \Cref{fig:htm-fixed}.
 		\item {\tt counter}: Microbenchmark version of {\tt htm2} which replaces the complex locking failure path with an atomic {\tt xadd}, from \cite{tm-benchmark-cmu}.
@@
 			% TODO: make sure figure number reference is right
-		\item {\tt fig63}: Generalized version of Figure~\ref{fig:retry-set-example},
+		\item {\tt fig63}: Generalized version of \Cref{fig:retry-set-example},
 			contrived to induce as much reduction as possible from the retry sets optimization.
@@
 this still includes abandoned smaller jobs that were run to saturate the set of data-race preemption points
-(Chapter~\ref{chap:quicksand}).
+(\Cref{chap:quicksand}).
 To minimize variance in CPU-time measurements,
@@
 implemented since then:
-the sleep sets optimization for DPOR (\sect{\ref{sec:landslide-sleepsets}}, commits 0447666 and 588687c),
+the sleep sets optimization for DPOR (\cref{sec:landslide-sleepsets}, commits 0447666 and 588687c),
 the {\tt thrlib\_function} and {\tt TRUSTED\_THR\_JOIN} directives
 to mark internal thread library logic as trusted
-(\sect{\ref{sec:landslide-config-landslide}}, commits 64a02e4 and a50d4ea),
+(\cref{sec:landslide-config-landslide}, commits 64a02e4 and a50d4ea),
 and fixing a soundness bug in which Landslide could neglect to inject transaction failures
@@
 		{Landslide's bug-finding performance on various test configurations.
-		Iterative Deepening (\sect{\ref{sec:quicksand-id}}), optimized for fast bug-finding,
-		is compared against Maximal State Space mode (\sect{\ref{sec:landslide-quicksand-options}}),
+		Iterative Deepening (\cref{sec:quicksand-id}), optimized for fast bug-finding,
+		is compared against Maximal State Space mode (\cref{sec:landslide-quicksand-options}),
 		optimized for fast verification.
@@
 		* marks the winning measurements between each series.
-		Lastly, state space estimation (\sect{\ref{sec:landslide-estimate}})
+		Lastly, state space estimation (\cref{sec:landslide-estimate})
 		%though approximate at best,
@@
 
-Table~\ref{tab:tm-buges}
+\Cref{tab:tm-buges}
 presents the bug-finding results.
-I configured Landslide to run Quicksand's Iterative Deepening algorithm on 8 cores, %(\sect{\ref{sec:quicksand-id}}),
+I configured Landslide to run Quicksand's Iterative Deepening algorithm on 8 cores, %(\cref{sec:quicksand-id}),
 shown left,
-as well as to prioritize the maximal state space, %(\sect{\ref{sec:landslide-quicksand-options}}),
+as well as to prioritize the maximal state space, %(\cref{sec:landslide-quicksand-options}),
 shown right,
@@
 despite prohibitive exponential explosion in total state space size.
-This corroborates the results from Chapter~\ref{chap:quicksand},
+This corroborates the results from \Cref{chap:quicksand},
 extending its good news to the world of HTM.
@@
 	exposed a previously-unknown bug in the transactional AVL tree.
-	Figure~\ref{fig:avlbug} shows the root cause, essentially the {\tt htm1} bug in disguise.
+	\Cref{fig:avlbug} shows the root cause, essentially the {\tt htm1} bug in disguise.
 	This manifested
@@
 	which the test case detects with {\tt \_xtest()}.
-	Figure~\ref{fig:spinlockbug} shows the root cause:
+	\Cref{fig:spinlockbug} shows the root cause:
 	the {\tt isfree()} routine (corresponding to the AVL's {\tt \_retry})
@@
 	is a reasonable performance property for a user to want to verify,
-	so I consider Landslide confirming the bug (and later verifying its absence, in \sect{\ref{sec:tm-verif}})
+	so I consider Landslide confirming the bug (and later verifying its absence, in \cref{sec:tm-verif})
 	a positive result anyway.
@@
 Note that in the AVL tree bug,
-the code's author was the very same person who proposed the protocol in Figure~\ref{fig:htm-fixed},
+the code's author was the very same person who proposed the protocol in \Cref{fig:htm-fixed},
 yet still got it wrong once, having to write it out by hand throughout both data structures.
@@
 then trusted in turn when checking their client programs \cite{dbug-phdthesis}.
-\sect{\ref{sec:tm-verif}} explores this further.
+\cref{sec:tm-verif} explores this further.
 
@@
 does not necessarily correlate with better performance in terms of CPU-time.
-Comparing Table~\ref{tab:tm-buges}'s trends
-to the break-even point in Quicksand's evaluation (\sect{\ref{sec:quicksand-eval}}),
+Comparing \Cref{tab:tm-buges}'s trends
+to the break-even point in Quicksand's evaluation (\cref{sec:quicksand-eval}),
 most of these tests are too small for its approach to pay off,
 with {\tt swapbug} and {\tt avl\_insert} as its notable wins.
-While plenty more wins were observed in \sect{\ref{sec:quicksand-eval}},
+While plenty more wins were observed in \cref{sec:quicksand-eval},
 %an upper bound also exists:
@@
 For the test cases with no bugs found, I sought to provide Landslide's verification guarantee
-(\sect{\ref{sec:quicksand-soundness}})
+(\cref{sec:quicksand-soundness})
 for up to as many threads and test iterations as possible under a reasonable time limit.
@@
 %were used in Table~\ref{tab:tm-buges},
-are shown in Table~\ref{tab:tm-verifs} in the ``Baseline DPOR'' column.
+are shown in \Cref{tab:tm-verifs} in the ``Baseline DPOR'' column.
 %Of course, because stateless model checking is a perpetual existence
@@
 		(i.e., injecting only {\tt \_XABORT\_RETRY}).
-		``Retry sets'' skips equivalent success path and/or retry abort reorderings (\sect{\ref{sec:tm-retrysets}});
+		``Retry sets'' skips equivalent success path and/or retry abort reorderings (\cref{sec:tm-retrysets});
 		``STM'' suppresses retry aborts and dynamically detects when to inject conflict aborts and so on
-		(\sect{\ref{sec:txn-abort-codes}}).
+		(\cref{sec:txn-abort-codes}).
 		State space estimates measured after a timeout of 10 hours (and include those 10 hours in the predicted total).
@@
 verifying up to $(K, N)$ represents a guarantee that,
-even repeating Figure~\ref{fig:htm-fixed}'s atomicity protocol $N$ times
+even repeating \Cref{fig:htm-fixed}'s atomicity protocol $N$ times
 in any scheduling sequence or combination of transaction aborts,
@@
 %
-Table~\ref{tab:tm-test-case-evence} confers a sense of the complexity of this evaluation's test suite,
+\Cref{tab:tm-test-case-evence} confers a sense of the complexity of this evaluation's test suite,
 with the final column showing the approximate maximum number of preemption points
@@
 	%while they do use the P2 thread API, %in the first place,
-	(the P2 thread API was annotated as trusted (\sect{\ref{sec:landslide-config-landslide}})
+	(the P2 thread API was annotated as trusted (\cref{sec:landslide-config-landslide})
 	and so does not contribute to state space size).
@@
 % (in the author's fancy, it may as well drop exponentially in the same way as the state spaces grow)
-(for example, none of the bugs in \sect{\ref{sec:tm-eval-bugs}} required any higher parameters than (2,2) to expose).
+(for example, none of the bugs in \cref{sec:tm-eval-bugs} required any higher parameters than (2,2) to expose).
 % and by analogy with ICB, none of the bugs in quicksand evaluation took more than a preemption bound of 3 to expose
@@
 
-Firstly, the middle column of Table~\ref{tab:tm-verifs}
-shows the impact of Landslide's experimental retry set reduction (\sect{\ref{sec:tm-retrysets}}).
+Firstly, the middle column of \Cref{tab:tm-verifs}
+shows the impact of Landslide's experimental retry set reduction (\cref{sec:tm-retrysets}).
 Despite its conservative implementation,
@@
 using transactional paths that conflict only with aborts and not with each other, and vice versa.
-In {\tt fig63}(2,1), corresponding to Figure~\ref{fig:retry-set-example},
+In {\tt fig63}(2,1), corresponding to \Cref{fig:retry-set-example},
 retry-set-enabled DPOR correctly prunes down to the optimal 6 interleavings,
 while the baseline treats it identically to the fully-conflicting {\tt counter}(2,1).
-Figure~\ref{fig:retry-reduction} depicts the difference between the state spaces explored by the two approaches.
+\Cref{fig:retry-reduction} depicts the difference between the state spaces explored by the two approaches.
 
@@
 \footnote{
-Not pictured in Figure~\ref{fig:retry-reduction} is the symmetric subtree of branches 5-6 in the right half of the state space,
+Not pictured in \Cref{fig:retry-reduction} is the symmetric subtree of branches 5-6 in the right half of the state space,
 which would occur after branch 10, reordering the blue thread before the pink's data race.
 Such would be equivalent to branches 2 and 4,
-and is pruned by the normal sleep set algorithm (\sect{\ref{sec:landslide-sleepsets}}, {\tt equiv\_already\_explored()})
+and is pruned by the normal sleep set algorithm (\cref{sec:landslide-sleepsets}, {\tt equiv\_already\_explored()})
 even in baseline DPOR, with no need for retry sets.
@@
 suppresses retry aborts
-(\sect{\ref{sec:txn-abort-codes}}),
+(\cref{sec:txn-abort-codes}),
 which must be checked at every transactional preemption point,
@@
 abort, success and abort, abort may (in fact, must) be skipped.
-The final column in Table~\ref{tab:tm-verifs} shows the result of STM semantics verification,
+The final column in \Cref{tab:tm-verifs} shows the result of STM semantics verification,
 which always results in at least 2x reduction compared to the baseline,
@@
 Visual inspection of the AVL tree and separate-chaining map implementations \cite{tm-benchmark-cmu},
-after correcting the former's atomicity bug (Figure~\ref{fig:avlbug}),
+after correcting the former's atomicity bug (\Cref{fig:avlbug}),
 reveals that every use of HTM followed exactly the same pattern:
@@
 even when the client's accesses are independent.
-Figure~\ref{fig:htm-lock} shows their core logic.
+\Cref{fig:htm-lock} shows their core logic.
 Finally, I parameterized them over how the lock was implemented:
 a real-world {\tt spinlock} implementation from \cite{spinlock-rtm-github},
-{\tt spin\_fixed}, the same with the performance bug from \sect{\ref{sec:tm-eval-bugs}} fixed,
+{\tt spin\_fixed}, the same with the performance bug from \cref{sec:tm-eval-bugs} fixed,
 and {\tt mutex}, using Landslide-annotated P2 mutexes (as the AVL and map implementations do).
@@
 
-Table~\ref{tab:tm-verifs2} shows the new resulting levels of verification
+\Cref{tab:tm-verifs2} shows the new resulting levels of verification
 Landslide reached before the same 10-hour timeout.
 Provided that one trusts the {\tt lock} tests correctly check the desired properties,
-and that open-coding hadn't introduced any new bugs (such as Figure~\ref{fig:avlbug}'s),
+and that open-coding hadn't introduced any new bugs (such as \Cref{fig:avlbug}'s),
 the benefit is clear:
@@
 	\end{center}
-	\caption[Continuation of Table~\ref{tab:tm-verifs}, demonstrating abstraction reduction.]
-		{Continuation of Table~\ref{tab:tm-verifs},
+	\caption[Continuation of \Cref{tab:tm-verifs}, demonstrating abstraction reduction.]
+		{Continuation of \Cref{tab:tm-verifs},
 		% XXX: if you put this citation, it will cause "(document)" to appear in the list of back-references
@@
 firstly,
-the impact that fixing Figure~\ref{fig:spinlockbug}'s performance bug
+the impact that fixing \Cref{fig:spinlockbug}'s performance bug
 (changing a read+write to a read only)
@@
 the spinlock's wait loop is open-coded, and Landslide must fall back on its costlier heuristic synchronization detection
-(\sect{\ref{sec:landslide-blocking}}).
+(\cref{sec:landslide-blocking}).
 In this way (and also, of course, because {\tt mutex}'s state spaces are smaller overall),
@@
 
-For all the reduction retry sets demonstrated in Table~\ref{tab:tm-verifs},
+For all the reduction retry sets demonstrated in \Cref{tab:tm-verifs},
 some inefficiencies remain in its strategy.
@@
 For now, its use is prohibited in conjunction with other state-space-affecting features
-such as ICB (\sect{\ref{sec:landslide-icb}})
+such as ICB (\cref{sec:landslide-icb})
 as well as multiple abort codes.
@@
 having carefully scrutinized its behaviour
-while constructing Figure~\ref{fig:retry-set-example}
+while constructing \Cref{fig:retry-set-example}
 and from inspecting state spaces arising from larger test parameters as well; %while implementing it;
@@
 
-In \sect{\ref{sec:tm-eval-stm}} I showed that state spaces could be reduced even further than with retry sets
+In \cref{sec:tm-eval-stm} I showed that state spaces could be reduced even further than with retry sets
 by assuming an HTM interface which abstracts away {\tt \_XABORT\_RETRY} behind a loop.
@@
 observable under HTM.
-As an example, note in Table~\ref{tab:tm-verifs}
+As an example, note in \Cref{tab:tm-verifs}
 how STM semantics reduced % oops
@@
 and give up on {\tt \_XABORT\_RETRY} by switching to STM mode on-the-fly,
-much like Landslide's heuristic synchronization detection does for yield loops (\sect{\ref{sec:landslide-blocking}}).
+much like Landslide's heuristic synchronization detection does for yield loops (\cref{sec:landslide-blocking}).
 
@@
 separately against their APIs
-with abstraction reduction (\sect{\ref{sec:tm-abstraction}}),
+with abstraction reduction (\cref{sec:tm-abstraction}),
 then check the rest of the program with (for example) traditional mutexes that can nest safely.
@@
 
-Section~\ref{sec:tm-design}'s formalization of thread interleavings does not account for read/write reorderings
+\cref{sec:tm-design}'s formalization of thread interleavings does not account for read/write reorderings
 possible on relaxed consistency architectures \cite{memory-consistency-models}.
 In fact,
-even after \cite{tm-benchmark-cmu}'s proposed fix to the atomicity protocol in Figure~\ref{fig:htm-fixed},
+even after \cite{tm-benchmark-cmu}'s proposed fix to the atomicity protocol in \Cref{fig:htm-fixed},
 it is still incorrect on Total Store Order (TSO) architectures such as x86,
@@
 rather than using a separate flag.
-In Figure~\ref{fig:htm-fixed}, a release barrier before line A is also necessary under PSO.
+In \Cref{fig:htm-fixed}, a release barrier before line A is also necessary under PSO.
 % On Partial Store Order (PSO) architectures,

### 2018-12-04 thesis/intro.tex
write a real intro

@@
 
+\revision{Pick up any recent concurrency-related research paper, %from the last few decades,
+and chances are its very first sentence will tell you
+%how notoriously difficult concurrency bugs are compared to sequential ones
+that concurrency bugs are notoriously difficult to find, reproduce, diagnose, and fix
+% ,
+%and/or thereafter verify fixed,
+%compared to sequential ones,
+%on account of the multitude of thread interleavings one must test
+\cite{nidhugg,optimal-dpor,coredet,sealing,quicksand,landslide,this-thesis,concuerror,
+cspec, % 2nd sentence in cspec >_>
+bpor,
+parrot, % 2nd sentence in parrot
+peregrine,
+bugtm, % 1st and 2nd sentence
+satcheck,
+racerx,
+ifrit, % 1st para, approx
+datacollider,
+fasttrack,
+mcr-tsopso,
+mcr,
+macemc,
+contessa,
+lvish,
+dthreads,
+taxdc,
+learning-from-mistakes, % sentence 3ish
+% TODO start up again on page 161 (AVIO: Detecting atomicity violations via access interleaving invariants.)
+}.
+%compared to sequential programming.
+%
+%Indeed, just as conditional statements and loops introduce additional possible program behaviours
+%depending on different input,
+%concurrent execution introduces additional possible behaviours even under a fixed input.
+%Unlike input nondeterminism,
+%concurrency nondeterminism typically manifests at the whims of chance,
+%making bug-finding and verification especially challenging.
+%
+%What is a concurrency bug, and
+How could a single programming technique be so troublesome
+as to inspire multiple decades of verification research?
+Allow me to begin this thesis by answering this question
+in layperson's terms,
+using two parables.
+
+My favourite way to explain concurrency to a non-technical audience
+%(e.g., family members)
+is by analogy with unlocking car doors.
+In many %makes and
+models of cars, the door handle and locking mechanism are intertwined
+in such a way that the door cannot unlock while its handle is being pulled~\cite{car-doors}.
+Most car users, then,
+%are familiar with
+have suffered the experience of
+trying to open the passenger door
+at exactly the same moment as the driver turns the key in the lock:
+%not only does the door not open, but
+it fails to unlock, and
+they must beseech the driver to turn the key again!
+If they had pulled the handle any later, it would have opened as normal;
+any earlier, and it would have been as though not having the key at all,
+but would at least have unlocked properly at the first turn of the key later.
+This violation of expected behaviour -- that the key need not be turned twice --
+is a {\em bug},
+and its dependence on the two travelers' actions {\em interleaving}
+%with the driver's actions
+in a particular way
+makes it a {\em concurrency bug}.
+This example involves only three events (pulling the handle, releasing the handle, and turning the key),
+but what of a system with hundreds or thousands of times more?
+Unfortunately, the number of possible interleavings grows {\em exponentially} with the number of events.
+
+%Exponential explosion is easily understood through the tale of Sissa Ibn D\`{a}hir
+The Islamic historian Ibn Khallik\={a}n told perhaps the first cautionary tale of exponential explosion
+\cite{ibn-khallikan,khallikān1868ibn}.
+Sissa Ibn D\`{a}hir,
+credited with the invention of chaturanga (the precursor to modern chess),
+was invited by King Shihr\^{a}m to request any reward he desired.
+Sissa requested that a grain of wheat be placed on the first square of a chessboard,
+two in the second, and so on, doubling the number of grains in each previous square until all 64 be filled.
+The king at first laughed, thinking it a pittance,
+realizing only later that to fulfill this would require
+more than all the wheat in all the cities on Earth.
+Modern programmers will recognize this sum as $2^{64}-1$,
+the largest value representable by an unsigned 64-bit integer,
+or approximately 18 billion billion.
+The number of interleavings in a concurrent system grows at a similarly intractable rate
+(although not necessarily exactly doubling each time),
+with the number of chessboard squares filled with grain corresponding to the number of events in a program's execution.
+And Sissa is the concurrent programmer,
+who always writes software large enough to be far beyond the reach of any exhaustive verification strategy.
+}
+
 \section{Motivation}
 
-% "Concurrency is notoriously difficult to get right." [....]
-% dejafu A Concurrency Testing Library for Haskell (2016 ver.)
-% parrot?
-% TODO more
-
-% TODO: god.........
-Modern computer architectures have turned to increasing CPU core count, rather than clock speed, to improve processing power \cite{mooreslaw}.
 To take advantage of multiple cores for performance, programmers must write software to execute {\em concurrently} --

### 2018-12-04 thesis/landslide.tex
write a real intro

@@
 The following annotations are specified in {\tt pebsim/config.landslide} akin to the static configuration options
-described in \cref{sec:landslide-directly}..
+described in \cref{sec:landslide-directly}.
 These specify the names of kernel functions, global variables, default values, and so on

### 2018-12-05 thesis/landslide.tex
llamas and studence

@@
 		\begin{itemize}
-			\item {\tt ADDR} shall be the code address (in hex) of the racing address,
+			\llitem {\tt ADDR} shall be the code address (in hex) of the racing address,
 			{\em before} the execution of which a preemption will be issued.
-			\item {\tt TID} indicates a thread ID required to be running for this data race.
+			\llitem {\tt TID} indicates a thread ID required to be running for this data race.
 				To specify data race PPs across all threads at once, set {\tt FILTER\_DRS\_BY\_TID=0} (see next section).
-			\item {\tt LAST\_CALL} indicates a code address required to be the site of the last {\tt call} instruction executed
+			\llitem {\tt LAST\_CALL} indicates a code address required to be the site of the last {\tt call} instruction executed
 				(similar to specifying a stack trace, but using a full stack trace here degrades performance too much),
@@
 				For further discussion see \cref{sec:quicksand-pps}.
-			\item {\tt CURRENT\_SYSCALL} indicates the system call number if a user-space data race comes from within a kernel system call which accesses user memory (Pebbles only).
+			\llitem {\tt CURRENT\_SYSCALL} indicates the system call number if a user-space data race comes from within a kernel system call which accesses user memory (Pebbles only).
 				Usually 0 (i.e., not in kernel code) but {\tt deschedule}'s system call number is common as well.
@@
 
-\begin{itemize}
+\begin{enumerate}
 \item {\bf Search algorithm options}
@@
 \end{itemize}
-\end{itemize}
+\end{enumerate}
 
@@
 		\begin{itemize}
-			\item If the last pushed return address is at offset 0 into the body of its containing function,
+			\llitem If the last pushed return address is at offset 0 into the body of its containing function,
 				Landslide will find the next pushed return address at {\tt esp+0}.
-			\item If as above but the function is the page fault handler, at {\tt esp+4}.
-			\item If the return address is at offset 1 and the previous instruction was {\tt push ebp},
+			\llitem If as above but the function is the page fault handler, at {\tt esp+4}.
+			\llitem If the return address is at offset 1 and the previous instruction was {\tt push ebp},
 				Landslide will find the next pushed return address at {\tt esp+4}.
-			\item If the return address is a multiple of 4 offset
+			\llitem If the return address is a multiple of 4 offset
 				and all previous instructions are of the form {\tt mov m32,r32},
@@
 		\begin{itemize}
-			\item track invocations of {\tt timer\_sleep()} and {\tt list\_insert\_ordered()}
+			\llitem track invocations of {\tt timer\_sleep()} and {\tt list\_insert\_ordered()}
 				to infer when a thread is sleeping rather than blocked automatically,
 				rather than relying on the {\tt tell\_landslide\_sleeping()} annotation.
-			\item allow {\tt sema\_up()} to reenter itself,
+			\llitem allow {\tt sema\_up()} to reenter itself,
 				which may happen when an IDE interrupt is taken when interrupts are re-enabled at the end of said function.
-			\item handle interrupt disabling/enabling as an abstract global lock for the purposes of happens-before analysis.
+			\llitem handle interrupt disabling/enabling as an abstract global lock for the purposes of happens-before analysis.
 		\end{itemize}
@@
 
-\begin{itemize}
+\begin{enumerate}
 	\item Reads and writes ({\tt memory.c})
@@
 			\begin{itemize}
-				\item {\tt kern\_mutex\_\{,try\}locking\_done()} cases of {\tt kern\_update\_state\_\allowbreak{}machine()}
-				\item {\tt user\_mutex\_\{,try\}lock\_exiting()} cases of {\tt user\_update\_state\_\allowbreak{}machine()}
-				\item {\tt cli} case of {\tt kern\_update\_state\_machine()} (Pintos only)
-				\item {\tt cli}/{\tt sti} lock handoff case in {\tt sched\_update()} (Pebbles only)
+				\llitem {\tt kern\_mutex\_\{,try\}locking\_done()} cases of {\tt kern\_update\_state\_\allowbreak{}machine()}
+				\llitem {\tt user\_mutex\_\{,try\}lock\_exiting()} cases of {\tt user\_update\_state\_\allowbreak{}machine()}
+				\llitem {\tt cli} case of {\tt kern\_update\_state\_machine()} (Pintos only)
+				\llitem {\tt cli}/{\tt sti} lock handoff case in {\tt sched\_update()} (Pebbles only)
 			\end{itemize}
@@
 			\begin{itemize}
-				\item {\tt kern\_mutex\_unlocking()} case of {\tt kern\_update\_state\_machine()}
-				\item {\tt user\_mutex\_unlock\_entering()} case of {\tt user\_update\_state\_machine()}
-				\item {\tt sti} case of {\tt kern\_update\_state\_machine()} (Pintos only)
-				\item {\tt cli}/{\tt sti} lock handoff case in {\tt sched\_update()} (Pebbles only)
+				\llitem {\tt kern\_mutex\_unlocking()} case of {\tt kern\_update\_state\_machine()}
+				\llitem {\tt user\_mutex\_unlock\_entering()} case of {\tt user\_update\_state\_machine()}
+				\llitem {\tt sti} case of {\tt kern\_update\_state\_machine()} (Pintos only)
+				\llitem {\tt cli}/{\tt sti} lock handoff case in {\tt sched\_update()} (Pebbles only)
 			\end{itemize}
@@
 	\end{itemize}
-\end{itemize}
+\end{enumerate}
 

### 2018-12-05 thesis/overview.tex
llamas and studence

@@
 concurrency bugs can arise from the interaction of any subset of kernel components with each other.
-The most common bugs studence face arise from the interaction of some component with itself (e.g., concurrent invocations of {\tt new\_pages}/{\tt remove\_pages} in the same process),
+The most common bugs students face arise from the interaction of some component with itself (e.g., concurrent invocations of {\tt new\_pages}/{\tt remove\_pages} in the same process),
 or from the interaction between an exiting thread and some other thread trying to communicate with it ({\tt vanish} versus, well, anything else, really).

### 2018-12-05 thesis/intro.tex
revise intro a bit

@@
 %how notoriously difficult concurrency bugs are compared to sequential ones
-that concurrency bugs are notoriously difficult to find, reproduce, diagnose, and fix
+that concurrency bugs are notoriously difficult to find, reproduce, diagnose, and/or fix
 % ,
@@
 \cite{nidhugg,optimal-dpor,coredet,sealing,quicksand,landslide,this-thesis,concuerror,
-cspec, % 2nd sentence in cspec >_>
+%cspec, % 2nd sentence in cspec >_>
 bpor,
-parrot, % 2nd sentence in parrot
+%parrot, % 2nd sentence in parrot
 peregrine,
-bugtm, % 1st and 2nd sentence
+%bugtm, % 1st and 2nd sentence
 satcheck,
 racerx,
-ifrit, % 1st para, approx
+%ifrit, % 1st para, approx
 datacollider,
@@
 contessa,
-lvish,
+%lvish, % "it must become easier"
 dthreads,
 taxdc,
-learning-from-mistakes, % sentence 3ish
-% TODO start up again on page 161 (AVIO: Detecting atomicity violations via access interleaving invariants.)
-}.
+%learning-from-mistakes, % sentence 3ish
+avio,
+symbiosis,
+chess-icb,
+chess,
+recordreplaydrs,
+kendo, % first WHOLE PARAGRAPH geez dude
+%hybriddatarace, % 3rd sentence
+parallel-dpor,
+%eraser, % 3rd sentence 2nd paragraph
+%racefuzzer, % 2nd sentence
+%tsan, % 2nd sentence
+sully-thesis,
+%compcerttso, % 2nd sentence
+modist,
+maple,
+simtester}.
 %compared to sequential programming.
@@
 with the number of chessboard squares filled with grain corresponding to the number of events in a program's execution.
-And Sissa is the concurrent programmer,
-who always writes software large enough to be far beyond the reach of any exhaustive verification strategy.
+A concurrency bug, then, is a single poisoned grain of wheat,
+which must be rooted out from the stock before feeding the townspeople.
+And Sissa himself is the programmer,
+whose software is invariably large enough to be far beyond the reach of any exhaustive verification strategy.
 }

### 2018-12-05 thesis/education.tex
more llama

@@
 	\begin{itemize}
-		\item Adds {\tt tell\_landslide\_thread\_on\_rq()}
+		\llitem Adds {\tt tell\_landslide\_thread\_on\_rq()}
 			and {\tt tell\_landslide\_thread\_off\_rq()}
@@
 			which in turn uses {\tt get\_rq\_addr()} and {\tt READY\_LIST\_LENGTH} described above.
-		\item Modifies the existing {\tt priority-sema} and {\tt alarm-simultaneous} tests to be more Landslide-friendly.
-		\item Inserts the {\tt tell\_landslide\_sched\_init\_done()},
+		\llitem Modifies the existing {\tt priority-sema} and {\tt alarm-simultaneous} tests to be more Landslide-friendly.
+		\llitem Inserts the {\tt tell\_landslide\_sched\_init\_done()},
 			{\tt tell\_landslide\_vanishing()},

### 2018-12-05 thesis/landslide.tex
et cetera

@@
 	\item {\tt -d TRACEDIR}: directory for resulting bug traces (default current directory)
-	\item {\tt -v}: verbose mode (issues output for each executed interleaving by each instance of landslide, makes progress reports more detailed, etc)
+	\item {\tt -v}: verbose mode (issues output for each executed interleaving by each instance of landslide, makes progress reports more detailed, et cetera)
 	\item {\tt -l}: leave Landslide log files from completed state spaces even when no bug was found (deleted automatically by default)
@@
 	\item {\tt PINTOS\_KERNEL=0} - configure Landslide for Pebbles (0) or Pintos (1) kernel architecture. Normally set automatically by the setup scripts.
-	\item {\tt TESTING\_USERSPACE=1} - configure Landslide whether to test (i.e., focus preemption points, memory analysis, etc. on) the userspace or kernelspace code.
+	\item {\tt TESTING\_USERSPACE=1} - configure Landslide whether to test (i.e., focus preemption points, memory analysis, et cetera on) the userspace or kernelspace code.
 	\item {\tt CURRENT\_THREAD\_LIVES\_ON\_RQ=0} - Landslide infers the list of runnable threads from the {\tt tell\_landslide\_on\_rq()} and {\tt off\_rq()} annotations (described below).
@@
 		indicating the point after which Landslide should begin analysis.
-		Any threads already initialized before this point (init, idle, etc) should be specified with {\tt starting\_threads} (previous subsection).
+		Any threads already initialized before this point (init, idle, et cetera) should be specified with {\tt starting\_threads} (previous subsection).
 	\item {\tt tell\_landslide\_forking()} - to be called whenever a new thread is created,
@@
 		stack trace of current thread at the moment of preemption,
-		depth in the tree, parent node pointer, etc.;
+		depth in the tree, parent node pointer, et cetera;
 	\item Snapshots of the current state of the scheduler (\cref{sec:landslide-scheduler})
@@
 Both rely on the helper function {\tt mem\_translate()} for virtual address resolution,
-which at present supports only the normal x86 32-bit addressing mode (no PAE, long mode, etc.).
+which at present supports only the normal x86 32-bit addressing mode (no PAE, long mode, et cetera).
 
@@
 The {\tt data\_race} command specifies an instruction pointer value to identify as a data race preemption point.
-The point can optionally be qualified by a thread ID, most recent system call number, etc.,
+The point can optionally be qualified by a thread ID, most recent system call number, et cetera,
 as described in \cref{sec:landslide-dynamicconfig},
@@
 \footnote{I describe $T_j$ as a thread ID here, rather than as a thread transition,
-because the nature of the transition (its memory accesses, the subsequent state, etc)
+because the nature of the transition (its memory accesses, the subsequent state, et cetera)
 is unknown until actually executed.}
@@
 In cases where {\tt yield} itself is not involved,
-Landslide also counts the number of atomic instructions ({\tt xchg}, {\tt xadd}, {\tt cmpxchg}, etc),
+Landslide also counts the number of atomic instructions ({\tt xchg}, {\tt xadd}, {\tt cmpxchg}, et cetera),
 and marks the thread blocked if it exceeds 100 such ({\tt TOO\_MANY\_XCHGS\_TIGHT\_LOOP}).%

### 2018-12-05 thesis/overview.tex
et cetera

@@
 and allows the testing tool to reduce the possible space of interleavings
-(or the set of false positive data-race candidates reported, etc.;
+(or the set of false positive data-race candidates reported, et cetera;
 \revisionminor{see \cref{sec:landslide-config-landslide}})
@@
 \end{itemize}
-Both Pebbles and Pintos basecodes offer a standard C library including {\tt malloc}, string-formatting, printing, etc.
+Both Pebbles and Pintos basecodes offer a standard C library including {\tt malloc}, string-formatting, printing, et cetera.
 

### 2018-12-05 thesis/landslide.tex
explain the landslide approach

@@
 and Landslide uses its access to the simulator's internal state to inspect and manipulate the memory and thread scheduling of the program as it executes.
+\revision{\Cref{fig:landslide-architecture} visualizes Landslide in relation to its execution environment,
+showing how it communicates with each of its surrounding and/or simulated programs.
+
+\begin{figure}[h]
+	\begin{center}
+		\includegraphics[width=0.875\textwidth]{landslide-new.pdf}
+	\end{center}
+	\caption{Landslide's execution environment when testing 15-410 student projects.}
+	\label{fig:landslide-architecture}
+\end{figure}
+
+\subsubsection{Design}
+
+From an implementation point of view,
+Landslide's main ``execution loop'' is simply the simulated CPU's own fetch-decode-execute loop:
+each time it simulates an instruction or memory access,
+it invokes Landslide through the simulator's module interface.
+More abstractly speaking, the general sequence of events during a Landslide test runs as follows.
+
+\begin{enumerate}
+	\item Instrument the test program by inspecting its binary,
+		learning the addresses of important functions, global variables, et cetera
+		(\cref{sec:landslide-glue}).
+	\item Execute the program under simulation.
+		\begin{enumerate}
+			\item At each instruction:
+				\begin{enumerate}
+					\item Update the {\em scheduler},
+						a state machine depicting the runnable and/or blocked
+						threads currently existing in the simulated program,
+						as well as a set of action flags to track what each thread is up to,
+						such as waiting to lock a mutex at a given memory address
+						(\cref{sec:landslide-scheduler}).
+					\item Record reads and writes to shared memory (\cref{sec:landslide-memory}).
+					\item Check whether the current program state constitutes a bug,
+						and if so,
+						emit a preemption trace (\cref{sec:landslide-foundabug})
+						and halt Landslide's execution.
+						Bug-detection predicates range from
+						simply checking if the program tripped an assert
+						or accessed invalid memory,
+						to checking the above-mentioned memory accesses against the current
+						heap allocation state for use-after-frees (\cref{sec:landslide-valgrind-mode}),
+						to querying the scheduler for deadlocks (\cref{sec:landslide-fp-deadlock}),
+						to checking heuristically for infinite loops and livelock
+						(\cref{sec:landslide-infloop}).
+					\item Check whether the current program state is a preemption point
+						(\cref{sec:landslide-pps}).
+					\item Query the scheduler to detect when the test has completed successfully,
+						i.e., all threads have finished executing normally
+						(\cref{sec:landslide-scheduler-statemachine}).
+				\end{enumerate}
+			\item At each preemption point, identified as in 2(a)iv above:
+				\begin{enumerate}
+					\item Check the set of memory accesses since the previous preemption point,
+						recorded as in 2(a)ii above,
+						for conflicts with other threads
+						(\cref{sec:landslide-shm}),
+						and further for data races (\cref{sec:landslide-datarace}).
+					\item Select which thread should run next (\cref{sec:landslide-arbiter}),
+						Heuristically detecting yield-blocked threads
+						may inform this decision (\cref{sec:landslide-blocking-yield}).
+					\item Checkpoint the execution state in case future executions
+						should wish to rewind and try a different thread from the one first chosen here
+						(\cref{sec:landslide-save}, \cref{sec:landslide-timetravel}).
+						% may be informed by heuristix/yield blocking
+					\item Force the chosen thread to run by injecting timer interrupts
+						(\cref{sec:landslide-interrupce},
+						\cref{sec:landslide-scheduler-interrupce}).
+				\end{enumerate}
+				Note that Landslide maintains the invariant that each transition between two preemption points
+				consist of instructions executed by exactly one thread;
+				i.e., every thread switch must be punctuated by a preemption point.
+				% preemption point one thread per invariant...
+		\end{enumerate}
+	\item At the end of each execution, identified as in 2(a)v above:
+		\begin{enumerate}
+			\item Analyze the set of memory conflicts,
+				computed as in 2(b)i above,
+				using Dynamic Partial Order Reduction (DPOR)
+				to decide which preemption point to backtrack to and
+				% once there
+				%% once then % lol
+				which new thread to run
+				(\cref{sec:landslide-dpor}).
+				This may be constrained by Iterative Context Bounding (\cref{sec:landslide-icb}).
+			\item Estimate the structure of the resulting state space
+				to predict overall runtime and total interleavings needed to test
+				(\cref{sec:landslide-estimate}).
+			\item If DPOR selected an alternate interleaving to explore,
+				backtrack the simulation state to that preemption point
+				(\cref{sec:landslide-timetravel})
+				force the new thread to run
+				(\cref{sec:landslide-interrupce}, \cref{sec:landslide-scheduler-interrupce}),
+				and repeat from step 2.
+				%If DPOR produced no results,
+				Otherwise, halt Landslide's execution,
+				declaring the explored state space free of bugs.
+		\end{enumerate}
+\end{enumerate}
+}
+
+\subsubsection{Implementation}
+
 As of this thesis's writing, Landslide supports the use of two possible simulators:
@@
 and is also roughly 3x faster.
+
+\subsubsection{Disclaimer}
 
@@
 \subsubsection{State machine}
+\label{sec:landslide-scheduler-statemachine}
 
@@
 and locksets, action flags, and runqueue membership are updated accordingly.
+\revision{Landslide also queries the scheduler state after it updates every instruction,
+via {\tt test\_update\_state()} ({\tt test.c}),
+to check the existence and/or runnability of all the system's threads
+and determine whether or not the test case has completed execution.}
 
 \subsubsection{Interrupt injection}
+\label{sec:landslide-scheduler-interrupce}
 
@@
 \subsubsection{Memory conflicts}
+\label{sec:landslide-shm}
 
@@
 \subsection{Bug-finding output}
+\label{sec:landslide-foundabug}
 
@@
 \subsubsection{Infinite loop detection}
+\label{sec:landslide-infloop}
 
@@
 \subsubsection{Yield-loop detection}
+\label{sec:landslide-blocking-yield}
 
@@
 \subsubsection{False-positive deadlock avoidance}
+\label{sec:landslide-fp-deadlock}
 

### 2018-12-05 thesis/education.tex
favourably

@@
 students reported being able both to diagnose and to verify as fixed the vast majority of Landslide's reported bugs,
-compared Landslide favorably to stress testing,
+compared Landslide favourably to stress testing,
 and found the experience worthwhile and worth recommending.

### 2018-12-05 thesis/landslide.tex
handle most remaining todos in landslide

@@
 		(\cref{sec:landslide-glue}).
+		Further interpret test-specific annotations
+		as informed by
+		configuration options (\cref{sec:landslide-directly}),
+		test case annotations (\cref{sec:landslide-testcases}),
+		and
+		kernel annotations (\cref{sec:landslide-kernel-anno}).
 	\item Execute the program under simulation.
@@
 						and if so,
-						emit a preemption trace (\cref{sec:landslide-foundabug})
+						emit a preemption trace
+						(\cref{sec:landslide-foundabug}, \cref{sec:landslide-bugreport})
 						and halt Landslide's execution.
@@
 						(\cref{sec:landslide-save}, \cref{sec:landslide-timetravel}).
-						% may be informed by heuristix/yield blocking
 					\item Force the chosen thread to run by injecting timer interrupts
@@
 
-% TODO: add some backup linkeroos
 The Bochs port of Landslide is likewise open-source
@@
 The HEAD commit at the time of writing is 8127151.
+In the event of a GitHub outage or the above link otherwise not working,
+a backup of the repository
+% checked out at the above commit
+may be found at
+% TODO: make there actually be smth at this commit
+\url{https://www.cs.cmu.edu/~410/landslide.tar.bz2}.
+%
 The Simics port uses Simics's proprietary API and is hence unlicensed and available upon request for educational use only.
@@
 This section describes the features of Landslide the average student user should expect to interact with.
-% TODO: make more specific section reference
-Separate user guides also exist, described in \Cref{chap:education}.
+Separate user guides also exist, described in \cref{sec:education-pebbles-recruiting}.
 
@@
 The preferred method of invoking Landslide is through Quicksand, the Iterative Deepening wrapper program which has all of \Cref{chap:quicksand} to itself.
-% TODO: add quicksand symlink
-This is done via the {\tt ./landslide} script in the top-level directory, which:
+This is done via the {\tt ./quicksand} script in the top-level directory, which:
 \begin{itemize}
@@
 	\item {\tt -C}: enable ``control experiment'' mode, i.e., run only 1 instance of Landslide, with all (non-data-race) preemption points enabled in advance
-		% TODO: put a section reference here
 	\item {\tt -I}: enable Iterative Context Bounding (requires {\tt -C}, although future work may relax this restriction);
@@
 	\item {\tt -4}: support Pebbles architecture (enabled automatically when either {\tt p2-setup.sh} or {\tt psu-setup.sh} is run)
-		% TODO: put a section reference
-	\item {\tt -e ETAFACTOR}: configure heuristic state space ETA deferring factor (described in detail in {\tt id/option.c})
-	\item {\tt -E ETATHRESH}: configure heuristic threshold of state space progress for judging ETA stability (described in detail in {\tt id/option.c})
+	\item {\tt -e ETAFACTOR}: configure heuristic state space ETA deferring factor (\cref{sec:quicksand-id}; described in detail in {\tt id/option.c})
+	\item {\tt -E ETATHRESH}: configure heuristic threshold of state space progress for judging ETA stability (\cref{sec:quicksand-id}; described in detail in {\tt id/option.c})
 \end{itemize}
@@
 				From personal experience I found this option rather useless and recommend always supplying 0.
-				% TODO: fix this section ref
 				For further discussion see \cref{sec:quicksand-pps}.
@@
 \section{Kernel annotations}
+\label{sec:landslide-kernel-anno}
 
@@
 but a CMU 15-410 student who wishes to use Landslide on her kernel project shall need to brave forth hereupon.
+An alternate step-by-step guide is also available at
+\url{http://www.contrib.andrew.cmu.edu/~bblum/landslide-guide-kernel.pdf}.
 
@@
 second, selecting which preemption point to backtrack to after completing an execution
-and which thread to ``have switched to'' %
-% TODO: post-revisions; uncomment; change " %" above to "%"
-%\footnote{willan on-having switched to \cite{hhgg-reu}}
+and which thread to ``have switched to''
+\footnote{\revision{willan on-having switched to \cite{hhgg-reu}}}
 instead, in {\tt explore.c},
@@
 (choosing independently at each preemption point, resulting in an overall unpredictable exploration order).
-% TODO: decide whether you're gonna put those estimate graphs in here
-%% I
 It's not exposed to {\tt config.landslide} but rather the user must edit it in {\tt arbiter.c} directly,
@@
 The {\tt data\_race} command specifies an instruction pointer value to identify as a data race preemption point.
-The point can optionally be qualified by a thread ID, most recent system call number, et cetera,
+It can optionally be qualified by a thread ID, most recent system call number, et cetera,
 as described in \cref{sec:landslide-dynamicconfig},
@@
 
-In Preempt-Everywhere mode, instead Landslide marks all shared memory accesses
-as long as they are not either part of the mutex implementation or part of the running thread's stack frame.
+In Preempt-Everywhere mode, Landslide instead marks all shared memory accesses
+as long as they are not part of either the mutex implementation or the running thread's stack frame.
 The {\tt data\_race} command is ignored and {\tt suspected\_data\_race()}
@@
 (In the case of a tie, differing only by which thread to run, it chooses arbitrarily.)
-All other new interleavings are marked to explore later in their corresponding {\tt struct hax},
+All other new interleavings are marked to explore later, %in their corresponding {\tt struct hax},
 and automatically included in the result of any future iterations of DPOR until they are tested.
@@
 DPOR consumes from this set in depth-first order
-(note the reversed order of $\bigstar$1 and $\bigstar$2)
+(note the reversed order of $\bigstar$1 and \revisionminor{$\bigstar$3})
 to avoid storing in memory any representation of exponentially-sized subtrees outside of the current branch.
-Note also that in $\bigstar$2,
+Note also that in \revisionminor{$\bigstar$3},
 the reordered \dporTB{{\tt tmp2~<-~x}} is not directly part of the memory conflict which marked it,
@@
 \Cref{fig:dpor-example-1} shows the outcome
-(with the previously-marked $\bigstar$2, now re-labeled as $\bigstar$3, having its subtree abbreviated).
+%(with the previously-marked $\bigstar$2, now re-labeled as $\bigstar$3, having its subtree abbreviated). % old ver
+\revision{(with the new $\bigstar$2 appearing in $\bigstar$1's subtree, to be explored before $\bigstar$3)}.
 
@@
 
-% TODO: talk abt free re malloc false positives
-
 \subsubsection{Pure Happens-Before}
@@
 As simple as it would be to state ``{\bf foreach} $S \in {\mathsf{sort}_P(U)}$'' in pseudocode,
-implementing such an ordering would be much less straightforward. %
-% TODO: post committee review
-%\footnote{Translating conference papers' pseudocode into usable implementations is often quite difficult \cite{itg2}.}
+implementing such an ordering would be much less straightforward.
+\footnote{\revision{Many conference papers' pseudocode algorithms
+often fail to be straightforward to translate into usable implementations~\cite[Figure 5]{itg2}.}}
 

### 2018-12-06 thesis/landslide.tex
adjust placement of sleepsets figure
landslide summary

@@
 
-\subsubsection{Design}
+\subsubsection{Overview}
 
@@
 	\item {\tt splice\_pre\_vanish\_trace} ({\tt stack.c})
-		will, when a {\tt vanish}ing thread has already deleted its Simics symbol table object,
+		will, when a {\tt vanish}ing thread has already
+		\revisionminor{unmapped all of its user memory,}
+		%deleted its Simics symbol table object,
 		splice in a saved ``pre-vanish'' stack trace
@@
 
-Ultimately, the second half of the state space is pruned symmetrically.
-In general, the number of ways to interleave two threads executing $N$ and $M$ events each is given by ${N+M \choose N}$%
-\footnote{Generalizing to $K$ threads, and simplifying to $N$ events each,
-this formula becomes $\frac{n!k!}{n!^k}$.};
-in this case, the original state space's size was ${3+3 \choose 3} = 20$.
-DPOR's reduction is characterized by replacing $N$ and $M$ with the number of {\em conflicting} events only;
-in this case, ignoring all {\tt tmpN++} reorderings and testing only ${2+2 \choose 2} = 6$ branches.
-
-\subsubsection{Sleep set reduction}
-\label{sec:landslide-sleepsets}
-
-In the presence of non-conflicting transitions as well as conflicting ones,
-DPOR's approach as described so far can still end up testing equivalent interleavings.
-As the presence of more equivalence arrows in \Cref{fig:dpor-example-2} hints,
-its reduced subset state space still contains redundancy,
-arising from the fact that one pair of those \revisionminor{four} events is two reads, and hence not actually in conflict.
-Visual inspection shows that $\bigstar$4, while locally justified
-in trying to reorder \dporTA{{\tt tmp1~<-~x}} before \dporTB{{\tt x~<-~tmp2}},
-effectively serves % oops
-only to reorder it with \dporTB{{\tt tmp2~<-~x}}
-relative to the first symmetric subtree ($\bigstar$1).
-In other words,
-even though DPOR marked each new branch with the intent only to reorder conflicting accesses,
-$\bigstar$1 and $\bigstar$4 contained interleavings equivalent up to independent reorderings anyway.
-\Cref{fig:sleepsets} summarizes the relevant interleavings to highlight one such equivalence.
-
-\begin{figure}[h]
+% TODO: check figure placement
+% should be right below dpor-example-2, on the previous page
+% i had to put ! not just ht to get it to show up correctly
+\begin{figure}[ht!]
 	\begin{center}
@@
 
+Ultimately, the second half of the state space is pruned symmetrically.
+In general, the number of ways to interleave two threads executing $N$ and $M$ events each is given by ${N+M \choose N}$%
+\footnote{Generalizing to $K$ threads, and simplifying to $N$ events each,
+this formula becomes $\frac{n!k!}{n!^k}$.};
+in this case, the original state space's size was ${3+3 \choose 3} = 20$.
+DPOR's reduction is characterized by replacing $N$ and $M$ with the number of {\em conflicting} events only;
+in this case, ignoring all {\tt tmpN++} reorderings and testing only ${2+2 \choose 2} = 6$ branches.
+
+\subsubsection{Sleep set reduction}
+\label{sec:landslide-sleepsets}
+
+In the presence of non-conflicting transitions as well as conflicting ones,
+DPOR's approach as described so far can still end up testing equivalent interleavings.
+As the presence of more equivalence arrows in \Cref{fig:dpor-example-2} hints,
+its reduced subset state space still contains redundancy,
+arising from the fact that one pair of those \revisionminor{four} events is two reads, and hence not actually in conflict.
+Visual inspection shows that $\bigstar$4, while locally justified
+in trying to reorder \dporTA{{\tt tmp1~<-~x}} before \dporTB{{\tt x~<-~tmp2}},
+effectively serves % oops
+only to reorder it with \dporTB{{\tt tmp2~<-~x}}
+relative to the first symmetric subtree ($\bigstar$1).
+In other words,
+even though DPOR marked each new branch with the intent only to reorder conflicting accesses,
+$\bigstar$1 and $\bigstar$4 contained interleavings equivalent up to independent reorderings anyway.
+\Cref{fig:sleepsets} summarizes the relevant interleavings to highlight one such equivalence.
+
 Intuitively speaking,
@@
 This check is implemented in {\tt try\_avoid\_fp\_deadlock()} in {\tt arbiter.c}.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{Summary}
+
+\revision{This chapter has presented Landslide,
+a stateless model checker for Pebbles and Pintos kernels and Pebbles userspace thread libraries.
+For the most part, Landslide's contribution to the world of concurrency testing is as a feat of engineering
+which combines many existing techniques from prior work in a single implementation.
+These include:
+
+\begin{itemize}
+	\item Uses Dynamic Partial Order Reduction (DPOR) \cite{dpor}
+		to prune equivalent interleavings while exploring the state space (\cref{sec:landslide-dpor}).
+	\item Uses its simulator's memory tracing integration \cite{simics,bochs}
+		to power data race analysis \cite{djit,fasttrack} (\cref{sec:landslide-datarace}).
+		%and DPOR \cite{dpor} (\cref{sec:landslide-dpor}).
+	\item Supports Preemption Sealing \cite{sealing} to configure the set of active preemption points
+		and hence keep state space sizes manageable (\cref{sec:landslide-pps}).
+	\item Extends DPOR with Recursive and Weighted Backtrack Estimation \cite{estimation}
+		to predict overall ultimate state space size and completion time (\cref{sec:landslide-estimate}).
+	\item Extends DPOR (at the user's option) with Iterative Context Bounding \cite{chess-icb}
+		to heuristically uncover bugs faster should they exist,
+		at the expense of overall completion time (\cref{sec:landslide-icb}).
+\end{itemize}
+
+However, over the years Landslide has also grown some convenience and/or accessibility features
+that could be considered contributions in their own right.
+
+\begin{itemize}
+	\item Extends Preemption Sealing for finer-grained control
+		over whether individual functions and/or global variables
+		should be considered for, or excluded from,
+		DPOR and/or data-race analysis
+		(\cref{sec:landslide-config-landslide};
+		{\tt thrlib\_function} and {\tt ignore\_dr\_function}, respectively).
+		This allows a test case author to precisely target the exact subset of code
+		she is interested in testing,
+		avoiding unnecessary state space inflation from trusted code.
+	\item Adds a mutex testing mode to configure whether the
+		data race analysis should consider lock' internal accesses
+		(\cref{sec:landslide-dynamicconfig}; {\tt TESTING\_MUTEXES}).
+		This allows the user to verify that her lock implementation
+		correctly provides mutual exclusion
+		(upcoming in \cref{sec:education-pebbles-tests}; {\tt mutex\_test}),
+		before relying on them as trusted code in subsequent tests.
+	\item Implements the sleep set extension to DPOR \cite{partial-order-methods,dpor,optimal-dpor},
+		which identifies and skips further equivalences based on which memory accesses
+		DPOR intended to reorder,
+		%, rather than just switching threads and seeing what happens
+		in a way that is slightly less optimal compared to prior work,
+		but much simpler to implement
+		and requires less state at runtime (\cref{sec:landslide-sleepsets}).
+	\item Heuristically identifies infinite loops and livelocks
+		by leveraging knowledge of how long the test program should be expected to run
+		obtained during previous executions,
+		uniquely suited to stateless model checking (\cref{sec:landslide-infloop}).
+	\item Provides automatic code instrumentation facilities,
+		beyond just the simple task of identifying important function boundaries,
+		that attempt to understand a wide variety of even the most unusual student implementations
+		(\cref{sec:landslide-pebblesspecifics},
+		\cref{sec:landslide-pintosspecifics}).
+		This includes an advanced heuristic for detecting ad-hoc synchronization loops
+		that Landslide should treat as blocking, %a thread,
+		even if it does not trigger the automatically-inserted blocking annotations
+		(\cref{sec:landslide-blocking-yield}),
+		and a further heuristic to ensure that any inaccuracy on the part of the first heuristic
+		does not result in false-positive bug reports (\cref{sec:landslide-fp-deadlock}).
+\end{itemize}
+
+The following three chapters, \Cref{chap:quicksand,chap:education,chap:tm},
+will each build upon Landslide to present the thesis's main research contributions.}

### 2018-12-06 thesis/intro.tex
minor intro edits

@@
 more than all the wheat in all the cities on Earth.
-Modern programmers will recognize this sum as $2^{64}-1$,
+Modern programmers will recognize this sum as $2^{64}$-$1$,
 the largest value representable by an unsigned 64-bit integer,
@@
 	%\item {\bf Pintos:} \Cref{chap:410} discusses my evaluation of Landslide in the University of Chicago's and Berkeley's OS class environments using the Pintos kernel, corresponding to contribution 2 above.
-	\item {\bf Transactional Memory:} \Cref{chap:tm} presents my extension of Landslide's concurrency model to handle transactional concurrency and the evaluation thereof, corresponding to contribution 3 above.
+	\item {\bf Transactions:} \Cref{chap:tm} presents my extension of Landslide's concurrency model to handle transactional concurrency and the evaluation thereof, corresponding to contribution 3 above.
 	\item {\bf Related Work:} \Cref{chap:relatedwork} honors my neighbours and ancestors in research spirit.

### 2018-12-06 thesis/thesis.tex
fix arabic text style and RTL issue
fix arabic citation with help from ppl who know the language
add an english comment about japanese acks line
ack stanley

@@
 \setCJKmainfont{HeiseiMinStd-W5}[Path = ./fonts/]
+%\newfontfamily\arabicfont{Droid Sans Arabic} % doesn't respond to Script = Arabic :(
+%\newfontfamily\arabicfont{LateefGR-Regular}[Path = ./fonts/] % doesn't respond to Script = Arabic :(
+\newfontfamily\arabicfont{Amiri-Regular}[Path = ./fonts/, Script = Arabic]
 
@@
 %having surrounded me pretty much since starting undergrad,
-nurtured my love of teaching
-well before taking the class myself and well after graduation.
+has nurtured my love of teaching
+since well before I took the class myself.
+%and well after graduation.
 %As skilled as the instructor is,
-15-410 could not be the world's best operating systems class
+15-410 could not be the world's best OS class
 without %TAs
-course staff %(staves)
-who value thorough grading, the Socratic method,
-and empathy to struggling students as much
-from year to year as the crew I have known.
+course staff's %(staves)
+unwavering dedication to
+thorough grading, the Socratic method,
+and empathy to struggling students.
+%as much
+%from year to year as the crew I have known.
 It was a blessing to serve, and later to do research, among you all.
+Thanks especially to Stanley Zhang,
+%the 2018 TA who took on the mantle of ensuring
+the 2018 TA who who worked with me to ensure
+15-410 could keep grading projects with Landslide even after I leave.
 Thanks to Wind River for generously
-providing CMU with Simics educational licenses.
+providing
+%CMU with
+Simics educational licenses.
 %I am also grateful
-Thanks also to Adam Blank and Miriam Zimmerman
+Thanks also to Adam Blank %and Miriam Zimmerman % nah, she said it was just adam
 for founding and organizing the class Great Practical Ideas,
@@
 pushing me to refine my presentation skills and to network with strangers (now familiar faces) from industry.
-I also thank those %industry
-attendees from each of the PDL Consortium member companies
+%I also thank those %industry
+I also thank the industry attendees
+%from each of the PDL Consortium member companies
 for their interest %, even in more theoretical work,
-and for the enlightening conversations resulting therefrom.
+%and for the enlightening conversations resulting therefrom.
+and enlightening conversations.
 %Thanks to the anonymous OOPSLA reviewer who went the extra mile to read my supplementary proofs
@@
 
+% I wrote the below myself, but constructed it carefully to not just convey my feelings accurately
+% but also produce intelligible output through google translate, for the sake of 日本語のできない読者の方々。
+% (The translation is not perfect, like 達成感 means more my personal sense of accomplishment,
+% rather than sounding objective (and consequently, kinda pompous) as it does in google's ouput, but it ain't bad.)
+% Of course, since google translate is powered by neural nets and is under active research and development,
+% I copy the current output here in case it changes in the future.
+%
+%%%% I would like to express my sincere gratitude to my fellow learners and native speakers
+%%%% who have studied Japanese, talked and taught me variously.
+%%%% I think that learning a foreign language is the second most successful achievement of my life.
+%
+% My own (loose) translation would be more like,
+% I am forever grateful to the friends I made through Japanese,
+% both my companions in language learning and the native speakers I was able to meet and befriend to begin with,
+% who studied together with me, carried on conversations with me,
+% and taught me a great variety of things.
+% I consider learning a foreign language (*nuance of ongoing process rather than completed achievement)
+% to be the second greatest sense of accomplishment I have felt in my life.
 日本語を

### 2018-12-06 thesis/education.tex
summary for educ

@@
 \label{chap:education}
+% TODO: look at this spacing hack again on fresh eyes and see if it rly improves matters or not
+\vspace{1em}
 
@@
 
+% fix spacing between paragraphs, since sect 5.1 header starts next page
+\vspace{3em}
 Concurrency is taught in as many different ways as there are
@@
 
-Human subjects research is inherently messy.
-Each of the previous section's approaches to evaluating Landslide's educational value
-was accompanied by some drawback which prevented it from being perfectly objective science,
-but many of them presented tentatively positive results nonetheless,
-\revisionminor{and relatively very little negative feedback such as false positives}.
-Landslide helped many students find and fix many bugs (\cref{sec:education-eval-bugfinding}),
-but making a direct comparison to stress testing, the prior state of the art, is not straightforward.
-%and not statistically significant,
-Immediate improvement in students' project grades was observed
-(\cref{sec:education-eval-grades}),
-although statistical significance was lost when attempting to account for selection bias;
-and impact on grades alone is a very narrow measure of pedagogical value anyway.
-Landslide's debugging power was also found to be statistically significant
-for the {\tt thr\_exit\_join} bug in particular.
-Students responded overwhelmingly positively in the survey (\cref{sec:education-eval-survey}),
-although it is easy to imagine students being equally happy with a ``debugging tool'' that just tells them all the answers.
-
-Nevertheless, I believe each of these partial results taken together
-paint an overall picture of success:
-students fixed their own bugs {\em and} were happy about it;
-students were able to ask for help rather than be deterred by inevitable technical difficulties as far as I know;
-students provided intelligent feedback suggesting they truly understood the debugging process;
-% well not the student who said "traces should point to where the bug is" but eh can't win em all c.c
-et cetera.
-%
-The rest of this section will discuss the study's limitations and provide some perspective for the future.
+This section will discuss the study's limitations and offer some perspective for the future.
 
@@
 	\item \revision{{\em In addition to finding bugs, did you manage to fully verify your code under
-		any of the tests by letting Landslide complete all state spaces before reaching the specified time out?}
-		This would provide a sense of how thoroughly students understood the underlying research technique,
+		any tests by letting Landslide complete all state spaces before reaching the specified time out?}
+		This would measure how thoroughly students understood the underlying research technique,
 		and serve as a follow-up to the ``how many bug fixes verified'' question
@@
 \subsection{Future educational use}
+\label{sec:education-future-use}
 
@@
 could easily be used on other thread-library-like programming projects at any university.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{Summary}
+
+% report on participation %s
+
+\revision{This chapter has introduced Landslide as an educational tool
+in the undergraduate operating systems classes at CMU, U. Chicago, and PSU.
+I have made the following contributions:
+
+\begin{itemize}
+	\item A lecture, polished over seven semesters,
+		for introducing stateless model checking as a concurrency testing approach
+		in high-level undergraduate classes
+		(with many slides admittedly specific to Landslide as they may be)
+		(\cref{sec:education-pebbles-recruiting}).
+	\item A system for automatically instrumenting student thread library (CMU, PSU) and kernel (U. Chicago)
+		project implementations to be compatible with stateless model checking
+		(\cref{sec:education-pebbles-instrumentation},
+		\cref{sec:education-pintos-instrumentation}).
+	\item An experimental procedure for evaluating debugging and verification tools
+		as student-accessible teaching aids,
+		including a survey for measuring the more intangible human experience
+		(\cref{sec:education-survey-pebbles},
+		\cref{sec:education-survey-pintos}).
+	\item A large evaluation spanning 9 semesters of operating systems classes
+		(\cref{sec:education-eval-bugfinding}),
+		2 of which were held at other universities besides CMU,
+		in which 145 students used Landslide,
+		a majority of which found bugs (\cref{sec:education-eval-bugfinding}),
+		and a vast majority of which expressed via the survey that they had a good experience
+		(\cref{sec:education-eval-survey}).
+	\item A thorough statistical analysis of Landslide's effect on student submission quality
+		(\cref{sec:education-eval-grades}),
+		evaluating both impact on ultimate project grades
+		and student tendency to submit any of four well-known case study bugs.
+		Ultimately, while positive trends were observed,
+		this study concluded that a statistically-significant conclusion could not be drawn
+		while simultaneously compensating for several obvious sources of possible bias.
+		Nevertheless, this level of statistical rigor is not yet widespread in computer science research,
+		so I would hope this study may set a precedent that help mature the field.
+	\item Concrete action taken to establish Landslide
+		as a permanent fixture of 15-410's grading process at CMU,
+		student-oriented user guides should the course staff wish it to remain available as a debugging tool,
+		as well as general recommendations for any Pintos-based OS class
+		at other universities to use it for grading as well
+		(\cref{sec:education-future-use}).
+\end{itemize}
+}
+
+Human subjects research is inherently messy.
+Each %of the previous section's
+\revisionminor{individual}
+approaches to evaluating Landslide's educational value
+was accompanied by some drawback which prevented it from being perfectly objective science,
+but many of them presented tentatively positive results nonetheless,
+\revisionminor{and relatively very little negative feedback such as false positives}.
+Landslide helped many students find and fix many bugs (\cref{sec:education-eval-bugfinding}),
+but making a direct comparison to stress testing, the prior state of the art, is not straightforward.
+%and not statistically significant,
+Immediate improvement in students' project grades was observed
+(\cref{sec:education-eval-grades}),
+although statistical significance was lost when attempting to account for selection bias;
+and impact on grades alone is a very narrow measure of pedagogical value anyway.
+Landslide's debugging power was also found to be statistically significant
+for the {\tt thr\_exit\_join} bug in particular.
+Students responded overwhelmingly positively in the survey (\cref{sec:education-eval-survey}),
+although it is easy to imagine students being equally happy with a ``debugging tool'' that just tells them all the answers.
+%
+Nevertheless, I believe each of these partial results taken together
+paint an overall picture of success:
+students fixed their own bugs {\em and} were happy about it,
+students were able to ask for help rather than be deterred by inevitable technical difficulties, %as far as I know;
+and students provided intelligent feedback suggesting they truly understood the debugging process.
+% well not the student who said "traces should point to where the bug is" but eh can't win em all c.c
+%et cetera.

### 2018-12-07 thesis/tm.tex
llamas and studence
et cetera
write intro/motivation for htm

@@
 
-The example TSX program from \Cref{fig:htm-example} (\cref{sec:overview-tm})
-is reproduced here for the reader's convenience.
+% or "Contribution"
+\revision{{\bf Motivation.}
+For the most part,
+existing stateless model checkers
+model interrupt-driven thread switching as the only source of nondeterminism in a program's execution.
+In other words, assuming a fixed test input,%
+\footnote{Input nondeterminism is another research field,
+addressed by symbolic execution \cite{symbolic}; see \cref{related:symbolic}.}
+the only way multiple behaviours can arise is by scheduling different threads at each preemption point.
+Only recently have checkers emerged that extend this model by
+incorporating other nondeterminism sources:
+store buffer reordering arising from relaxed memory models
+\cite{cdschecker,tsopso,rcmc}
+and event-driven application programming models
+\cite{r4,eventracer},
+both of which are especially relevant in mobile programming.
+Under TM, {\sf begin}'s return value is also nondeterministic,
+and moreover may depend on execution events from the future in a parallel universe
+(i.e., subsequent memory conflicts detected and reverted),
+and no model checker yet exists which can model this.
+In this chapter I will develop a concurrency model to incorporate abort nondeterminism
+(\cref{sec:tm-design}),
+including two equivalence proofs to help cope with the consequent state space explosion
+(\cref{sec:tm-design-failure} and \cref{sec:tm-design-locks}),
+extend Landslide to support testing TM programs under a variety of execution semantics
+(\cref{sec:tm-implementation}),
+and demonstrate its practicality
+%\cref{sec:tm-eval}
+both by finding new bugs
+(\cref{sec:tm-eval-bugs})
+and by verifying correctness
+(\cref{sec:tm-verif})
+in a variety of microbenchmarks and real-world HTM programs.}
+
+\revisionminor{The following sections will make heavy reference
+to the example TSX program from \Cref{fig:htm-example} (\cref{sec:overview-tm})
+while defining the new concurrency model,
+so I reproduce it here in \Cref{fig:htm-example-reproduced} for the reader's convenience.}
 
@@
 
+\revision{
 \begin{figure}[h]
@@
 	\end{center}
-	\caption[Impossible hypothetical interleaving within transactions of \Cref{fig:htm-example-reproduced}.]
-		{Impossible hypothetical interleaving within transactions of \Cref{fig:htm-example-reproduced}.
+	\caption[Impossible hypothetical interleaving within transactions of \Cref{fig:htm-example-assemblified}.]
+		{Impossible hypothetical interleaving within transactions of \Cref{fig:htm-example-assemblified}.
 		The effects of $\tiat{2a-2b}$ are never globally visible, as its transasction must abort,
@@
 \end{figure}
+}
 
@@
 	\]
-	with $b<n$, $j \ne i$, $k \ne i$, etc., either:
+	with $b<n$, $j \ne i$, $k \ne i$, et cetera, either:
 	\begin{enumerate}
@@
 
-\begin{itemize}
+\begin{enumerate}
 	\item Unit tests and microbenchmarks
@@
 		and {\tt mutex} (replaces the spinlock with a Landslide-annotated P2 mutex to reduce state space size).
-\end{itemize}
+\end{enumerate}
 

### 2018-12-07 thesis/landslide.tex
fix some figure placements, footnote stuffs, cref stuffs

@@
 
-Use of {\tt tell\_landslide()} annotations
+Use of the {\tt tell\_landslide()} annotations
 \revisionminor{(\cref{sec:tell-landslide})}
 is not necessary,
-although {\tt tell\_landslide\_\allowbreak{}preempt()} and {\tt tell\_landslide\_dump\_stack()}
+although {\tt tell\_\allowbreak{}landslide\_preempt()} and {\tt tell\_landslide\_dump\_stack()}
 may optionally be used at the user's convenience.
@@
 Ultimately, the second half of the state space is pruned symmetrically.
-In general, the number of ways to interleave two threads executing $N$ and $M$ events each is given by ${N+M \choose N}$%
+In general, the number of ways to interleave two threads executing $N$ and $M$ events each is given by ${N+M \choose N}$;%
 \footnote{Generalizing to $K$ threads, and simplifying to $N$ events each,
-this formula becomes $\frac{n!k!}{n!^k}$.};
+this formula becomes $\frac{n!k!}{n!^k}$.}
 in this case, the original state space's size was ${3+3 \choose 3} = 20$.
@@
 %I summarize their respective implementations here.
-\Cref{chap:warpzone-heuristics} will discuss their limitations and some opportunities for future improvement.
+\cref{sec:warpzone-heuristics} will discuss their limitations and some opportunities for future improvement.
 
@@
 but Landslide has the benefit of knowledge from past iterations to inform its sense of how long the program ``should'' run.
-
-\begin{figure}[h]
-	\begin{center}
-		\begin{tabular}{cc}
-			\includegraphics[width=0.28\textwidth]{heuristic-livelock.pdf}
-			&
-			\includegraphics[width=0.28\textwidth]{heuristic-tightloop.pdf}
-			\\
-			(a) Infinite loop around preemption points.
-			&
-			(b) Stuck between preemption points.
-		\end{tabular}
-	\end{center}
-	\caption{Detecting infinite loops heuristically by comparison to past interleavings.}
-	\label{fig:heuristic-loop}
-\end{figure}
 
@@
 -- a program with such $N$ would probably have an impossibly large state space anyway.
+
+\begin{figure}[t]
+	\begin{center}
+		\begin{tabular}{cc}
+			\includegraphics[width=0.28\textwidth]{heuristic-livelock.pdf}
+			&
+			\includegraphics[width=0.28\textwidth]{heuristic-tightloop.pdf}
+			\\
+			(a) Infinite loop around preemption points.
+			&
+			(b) Stuck between preemption points.
+		\end{tabular}
+	\end{center}
+	\caption{Detecting infinite loops heuristically by comparison to past interleavings.}
+	\label{fig:heuristic-loop}
+\end{figure}
 

### 2018-12-07 thesis/thesis.tex
fix some figure placements, footnote stuffs, cref stuffs
name formatting acks

@@
 \usepackage{microtype}
-\usepackage[nameinlink]{cleveref}
+% "nosort" causes no sort AND no compress of 3+ section references in one;
+% it's "sort" to sort but NOT compress; nothing to.. compress but not sort, i guess? not the greatest naming
+\usepackage[nameinlink,nosort]{cleveref}
 
@@
 Kartik Subramanian,
-Kellie Medlin,
+Kellie ``K2'' Medlin,
 Laura Abbott,
@@
 Margaret Meyerhofer,
-Michael ``rntz'' Arntzenius,
+Michael A\texttt{rntz}enius,
 Michael J. ``Sully'' Sullivan,

### 2018-12-07 thesis/education.tex
fix some figure placements, footnote stuffs, cref stuffs
assorted small prose

@@
 These I analyzed by hand to determine how many distinct bugs were reported
-(as multiple preemption traces may refer to the same bug; inspecting their code if necessary),
+\revisionminor{(inspecting their code if necessary,
+as multiple traces may refer to the same bug),}
 how many were deterministic
-(i.e., Landslide reported them on the very first interleaving tested, with no need for artificial preemptions),
+(i.e., reported on the very first interleaving tested, with no need for artificial preemptions),
 how many were concurrency bugs,
@@
 % TODO: should i report the T.AD test statistic? i don't even know what it means lol but it's what the p value like, applies to..
-\begin{figure}[t]
+\begin{figure}[h]
 	\begin{center}
@@
 % TODO check figure placement
-\begin{table}[t]
+\begin{table}[h]
 	\begin{center}

### 2018-12-07 thesis/landslide.tex
assorted small prose

@@
 implementing such an ordering would be much less straightforward.
-\footnote{\revision{Many conference papers' pseudocode algorithms
+\footnote{\revision{The pseudocode algorithms found in many conference papers
 often fail to be straightforward to translate into usable implementations~\cite[Figure 5]{itg2}.}}
@@
 	\item Uses its simulator's memory tracing integration \cite{simics,bochs}
-		to power data race analysis \cite{djit,fasttrack} (\cref{sec:landslide-datarace}).
+		to power data race analysis \cite{djit,fasttrack} (\cref{sec:landslide-datarace}),
+		as well as DPOR.
 		%and DPOR \cite{dpor} (\cref{sec:landslide-dpor}).
@@
 
-However, over the years Landslide has also grown some convenience and/or accessibility features
+However, over the years Landslide has also grown some convenience, accessibility, and or optimization features
 that could be considered contributions in their own right.

### 2018-12-07 thesis/education.tex
handle many todos in educ

@@
 \begin{table}[t]
-	% TODO
 	\begin{center}
@@
 However, the distribution difference being smaller in P3 than in P2
-% TODO: is there a good way to treat this more formally?
-% TODO: run a combined A-D test for each of the two control pairs
 further suggests, albeit informally,
@@
 
-% TODO: should i report the T.AD test statistic? i don't even know what it means lol but it's what the p value like, applies to..
 \begin{figure}[h]
@@
 (Landslide, of course, replaces this {\tt misbehav}iour with model checking).
-% TODO: i feel tihs sentence should go somewhere else, like after analysis, not in description
 Hence, despite being a subtle concurrency bug,
@@
 by checking only the first thread interleaving with Landslide
-% TODO - more specific section reference - quicksand det/nondet drpps eval
-(similarly to \cref{sec:quicksand-eval})
+(similarly to \cref{sec:quicksand-eval-nondets})
 and assuming the students would find and fix any such ``deterministic'' use-after-free bugs before the deadline.
@@
 The survey responses were very positive:
-% TODO: corroborate this with snapshot data?
+% corroborate this with snapshot data? -- haha, never
 students reported being able both to diagnose and to verify as fixed the vast majority of Landslide's reported bugs,
@@
 which probably speaks more to their enthusiasm than to a careful attempt to maximize expected number of bugs found
-% TODO: put a section reference to future work resource exhaustion / malloc failure injection discussion
-(even the author themself recognizes stress testing's advantages for certain types of bugs, such as resource exhaustion).
+(even the author themself recognizes stress testing's advantages for certain types of bugs,
+such as resource exhaustion; see \cref{sec:warpzone-verif}).
 Nevertheless, the responses' overall bias to spending at least half the CPU-time on Landslide
@@
 compared to PSU's relative tempering of P2's difficulty to make it more accessible (\cref{sec:pebbles}),
-% TODO: represent u c in this sentents?
 these trends seem to correlate with the different levels of preparation each course's students had,
@@
 
-% TODO
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-% TODO - u can verify the bug just by taking survey graph #1 and comparing to landslide snapshots (sep. by groups who answered survey)
+% u can verify the bug just by taking survey graph #1 and comparing to landslide snapshots (sep. by groups who answered survey)
+% ^ haha, never
 % anything else?
@@
 
-% TODO: is this ok to report without IRB approval? ask dave
 I took no special measures to compensate for bias in gender, race, or being non-natively English-speaking

### 2018-12-07 thesis/tm.tex
handle some todos in txn

@@
 In other words, assuming a fixed test input,%
-\footnote{Input nondeterminism is another research field,
-addressed by symbolic execution \cite{symbolic}; see \cref{related:symbolic}.}
+\footnote{Symbolic execution \cite{symbolic}, a neighbouring research area,
+addresses input nondeterminism; see \cref{related:symbolic}.}
 the only way multiple behaviours can arise is by scheduling different threads at each preemption point.
@@
 	satisfying case 1.
-	Otherwise, by DPOR's definition of transition dependence (\cite{dpor}, \cref{sec:landslide-dpor-hb}),
+	Otherwise, by DPOR's definition of transition dependence (\cite{dpor}; see also \cref{sec:landslide-dpor-hb}),
 	$\tiiat{\beta_{b+1}}\dots\tiiat{\beta_n}$ is independent with the transitions of \tjj and \tkk,
@@
 		\item {\tt swapbug}: {\tt swap} modified to introduce circular locking in the failure path. %(deadlock).
-			% TODO: make sure figure number reference is right
 		\item {\tt fig63}: Generalized version of \Cref{fig:retry-set-example},
@@
 
-			% TODO: check these again, will 2,4 even complete on mutex
 			{\tt lock}({\tt spinlock})	& 1	& 2	& 4	& 28	\\ % 2,2
 			{\tt lock}({\tt spin\_fixed})	& 1	& 2	& 4	& 28	\\ % 2,2 or 4,1
-			{\tt lock}({\tt mutex})		& 1	& 2	& 2	& 30	\\ % 2,3 gives 30; 2,4 would give 40
+			{\tt lock}({\tt mutex})		& 1	& 2	& 2	& 40	\\ % 2,4
 			% not putting lock_fast because it like, doesn't have a "state space" per se

### 2018-12-08 thesis/related.tex
write intro/motivation for htm
weak atomicity description, proof, & implementation, wee

@@
 % TODO: citation as noun - global scan
-\revision{\cite{mc-tm-with-spin} used SPIN \cite{spin} to model check McRT STM \cite{mcrt-stm},
+\revision{McRT STM \cite{mcrt-stm},
 an STM implementation built by Intel,
-verifying}
-up to 2 threads running 1 transaction each with up to 3 memory accesses.
+was checked using SPIN \cite{spin}, verifying
+up to 2 threads running 1 transaction each with up to 3 memory accesses \cite{mc-tm-with-spin}.}
 This kind of verification,
@@
 
-An article from relatively early in the timeline of TSX \cite{htm-subtleties}
+%An article from relatively early in the timeline of TSX
+\revisionminor{The paper which originally defined weak and strong atomicity~\cite{htm-subtleties}
+also}
 warns of several false equivalence pitfalls when converting conventionally-locking code to use transactions,
@@
 \subsubsection{Symbolic execution}
+\label{related:symbolic}
 

### 2018-12-08 thesis/tm.tex
weak atomicity description, proof, & implementation, wee

@@
 
+\subsubsection{Implementations}
+
 % do they track changes in pthread-style TLS? or?
-{\bf Implementations.}
 Software TM implementations (STM) typically function as a library,
 tracking staged memory accesses in local memory,
-and aborting whenever a conflict is detected between two transactions' tracked accesses \cite{stm-pldi06}.
+and aborting whenever a conflict is detected between two transactions' tracked accesses \cite{stm-pldi06,mcrt-stm,swisstm}.
 Hardware TM implementations (HTM) use processor-level hardware support,
@@
 
-{\bf Terminology.}
-The world of hardware transactional memory is home to several more confusing acronyms in addition to ``HTM''.
+\subsubsection{Terminology}
+
+The world of transactional memory is home to several more confusing acronyms similar to ``HTM''.
 Transactional Synchronization Extensions (TSX)
@@
 
+\revision{Transactional atomicity comes in two flavours,
+{\em strong} and {\em weak atomicity}~\cite{htm-subtleties}.
+%
+HTM implementations use the processor's cache coherence protocol to detect conflicts,
+and hence any conflicting memory access from another CPU will cause an abort,
+regardless of whether that access was itself transactional.
+Hence, HTM transactions are {\em strongly atomic}, i.e., appear indivisible to all other code.
+%
+STM implementations rely on the programmer to use its interface to access any shared memory
+she intends to protect with its transactions,
+which cannot detect conflicts with non-transactional accesses that bypass it.
+Hence, STM transactions are {\em weakly atomic}, i.e.,
+its transactions will abort only from conflicts with other transactions,
+while other code may interleave with them freely.
+%
+The manner in which non-transactional code may interleave with STM transactions
+further depends on whether the STM uses
+{\em eager versioning},
+wherein transactional updates are immediately visible to non-transactional code,
+or
+{\em lazy versioning},
+wherein such updates are visible only after commit
+\cite[\S{}2]{stm-weak-atomicity}.
+Unless explicitly noted as weak atomicity,
+this chapter assumes strong atomicity,
+and ``STM semantics'' refers to a strongly-atomic HTM program emulating STM with retry loops
+(see \cref{sec:txn-abort-codes}).
+Treatment of weak atomicity in
+\cref{sec:tm-weak-atomicity} and \cref{sec:tm-impl-weak-atomicity}
+assumes eager versioning.}
+
 % or "Contribution"
-\revision{{\bf Motivation.}
-For the most part,
+\subsubsection{Motivation}
+
+\revision{For the most part,
 existing stateless model checkers
@@
 (\cref{sec:tm-design-failure} and \cref{sec:tm-design-locks}),
+extend the proofs to account for weak atomicity (\cref{sec:tm-weak-atomicity}),
 extend Landslide to support testing TM programs under a variety of execution semantics
@@
 
-Then, without loss of generality (for any number of other threads \tjj/\tkk, and for any number of thread switches away from \tii during the transaction):
+Then, without loss of generality (for any number of other threads \tjj/\tkk,
+\revisionminor{WLOG presented here as two threads,}
+and for any number of thread switches away from \tii during the transaction,
+\revisionminor{WLOG presented here as one switch}):
 \vspace{1em}
@@
 Finally, let $I' = \exists m. \mathsf{lockify}_m(I)$,
-the execution with the boundaries of all successful transactions replaced by an abstract global lock.
+the execution with the boundaries of all successful transactions replaced by an abstract global lock $m$.
 \Cref{lem:equiv} guarantees mutual exclusion of $m$.
@@
 \begin{theorem}[Transactions are a Global Lock]
+	\label{thm:global-lock}
 	$\tiiat{\mu},\tjjat{\nu}$ is a data race in $I$ iff it is a data race in $I'$.
@@
 \begin{proof}
-I prove one case for each variant definiton for data races supported in Landslide \cite{quicksand}.
+I prove one case \revisionminor{for each of the two variant definitons of data races that Landslide supports
+(\cref{sec:landslide-datarace})}.
+%\cite{quicksand}. % wtf no
 For each, I state below what it means to race in an execution with synchronizing HTM instructions.
@@
 
+\subsection{Weak atomicity}
+\label{sec:tm-weak-atomicity}
+
+\newcommand\weakhilight[1]{{\ul{#1}}\xspace}
+
+\revision{Prior work has highlighted the difference between
+{\em strong} and {\em weak atomicity}
+in TM systems
+\cite{htm-subtleties,stm-weak-atomicity}.
+Thus far we have assumed transactions will abort from any intervening memory conflict,
+whether or not that access is itself transactional.
+This corresponds to strong atomicity, the execution semantics provided by TSX.
+% \cite{bugtm}. % sec 3.1 of that paper, not sure of an earlier citation that also states it in these exact terms
+However, under weakly-atomic STM systems,
+transactions do not abort when
+a non-transactional conflicting access interleaves within them,
+and condition \#1 of the execution semantics from \cref{sec:tm-design-failure} does not hold.
+To model weak atomicity, we must weaken \Cref{lem:equiv} as follows (updated parts underlined for emphasis).
+\vspace{1em}
+
+\begin{lemma}[Equivalence of Aborts \weakhilight{under Weak Atomicity}]
+	\label{lem:equiv-weak}
+	For any interleaving prefix
+	\[
+	\begin{tabular}{c}
+		$\tiiat{\alpha},\tiiat{\beta_1}\dots\tiiat{\beta_b},$\\
+		$\tjjat{\gamma_1}\dots\tjjat{\gamma_j},$ \\
+		$\tkkat{\kappa_1}\dots\tkkat{\kappa_k},$ \\
+		$\tiiat{\beta_{b+1}}$ % $\dots\tiiat{\beta_{n-1}}$ -- excluded bc might abort
+	\end{tabular}
+	\]
+	with $b<n$, $j \ne i$, $k \ne i$, et cetera, either:
+	\begin{enumerate}
+		\item \weakhilight{All events in
+			$\tjjat{\gamma_1}\dots\tjjat{\gamma_j},
+			 \tkkat{\kappa_1}\dots\tkkat{\kappa_k}$
+			 % ehh, bad parallel structure with below regarding the verb 'exists'
+			are non-transactional, and the original interleaving itself exists, or}
+		\item $\tiiat{\alpha},\tjjat{\gamma_1}\dots\tjjat{\gamma_j},\tkkat{\kappa_1}\dots\tkkat{\kappa_k},\tiiat{\phi_1}\dots$
+			(conflicting case), or
+		\item $\tiiat{\alpha},\tiiat{\beta_1}\dots\tiiat{\beta_b}\dots\tiiat{\beta_n},\tjjat{\gamma_1}\dots\tjjat{\gamma_j},\tkkat{\kappa_1}\dots\tkkat{\kappa_k}$
+			(independent case)
+	\end{enumerate}
+	exists and is observationally equivalent.
+\end{lemma}
+
+\begin{proof}
+	As before, \weakhilight{except first case on
+	whether any intervening operations by $\tjj$/$\tkk$ are transactional.%
+	\footnote{If $\tjj$/$\tkk$ already have a transaction active,
+	their {\sf begin} must have occurred before $\tiiat{\alpha}$,
+	and the theorem can apply starting from the earlier {\tt begin} with the thread IDs swapped.}
+	If not, $\tii$ may ultimately proceed with $\tiiat{\beta_{b+1}}$ without aborting,
+	and the new case \#1 applies.}
+\end{proof}
+
+\Cref{thm:atom} must be weakened correspondingly.
+\vspace{1em}
+
+\begin{theorem}[\weakhilight{Weak} Atomicity of Transactions]
+	\label{thm:atom-weak}
+	For any state space $S$ of a transactionally-concurrent program,
+	an equivalent state space exists in which all transactions are either executed atomically
+	\weakhilight{with respect to other transactions only},
+	or aborted immediately.
+\end{theorem}
+
+\begin{proof}
+	As before.
+\end{proof}
+
+Conveniently, in \cref{sec:tm-design-locks}'s formulation of
+$I' = \exists m. \mathsf{lockify}_m(I)$,
+\Cref{thm:atom-weak} still guarantees transaction-transaction atomicity,
+so mutual exclusion of $m$ is preserved,
+%(a global lock does not protect against interleaving by arbitrary other code that never takes the lock either)
+and
+%the proof of
+\Cref{thm:global-lock} applies unmodified.
+\cref{sec:tm-impl-weak-atomicity}
+discusses the implementation consequences of this weakening.}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-Support for TSX programs in Landslide is implemented in four parts, broadly speaking:
+Support for TSX programs in Landslide is implemented in \revisionminor{five} parts, broadly speaking:
 the user interface,
@@
 \subsection{User Interface}
+\label{sec:tm-interface}
 
@@
 		Requires {\tt -X} and {\em not} {\tt -A}.
+	\item \revision{{\tt -W} (for ``Weak atomicity'')
+		enables weak atomicity, %for transactions,
+		allowing non-transactional code to interleave within transactions,
+		corresponding to \cref{sec:tm-weak-atomicity}.
+		Requires {\tt -S}.}
 \end{itemize}
@@
 the vector clocks are simply checked as usual.
+
+\subsection{Weak atomicity}
+\label{sec:tm-impl-weak-atomicity}
+
+\revision{If the user supplies the {\tt -W} option (\cref{sec:tm-interface}),
+Landslide will emulate STM's weak atomicity semantics (\cref{sec:tm-weak-atomicity}).
+%to allow non-transactional code
+%to interleave between a {\tt \_xbegin()}/{\tt \_xend()} pair.
+Concretely, it achieves this by
+reporting both halves of data races that occur between transactional and non-transactional code (\cref{sec:landslide-datarace})
+(previously, only the latter half would be reported),
+allowing preemption points to be identified during transactions (\cref{sec:landslide-pps}),
+and allowing system calls during transactions to proceed rather than forcing a {\tt CAPACITY} abort
+(as programmers may now need to nest other synchronization within transactions to protect against non-transactional code).
+%and traversing the interleaving arbitrarily far back from a {\tt \_xend()} to find the appropriate preemption
+%(previously, the xbegin would be at the immediately previous PP)
+Landslide requires the other STM options ({\tt -A -S}) also to be enabled to emulate weak atomicity,
+because I know of no existing HTM system which implements it.
+As one might expect, it inflates state spaces considerably, % TODO: measure it with htm2 or some such?
+and moreover,
+every single HTM benchmark in the upcoming evaluation,
+implemented to rely on strong atomicity,
+exhibits bugs under weak atomicity.
+Commit 68401fc implements this mode.}
 

### 2018-12-08 thesis/tm.tex
abbreviate weak atomicity proof substantially, no ugly underlining

@@
 
-\newcommand\weakhilight[1]{{\ul{#1}}\xspace}
-
 \revision{Prior work has highlighted the difference between
 {\em strong} and {\em weak atomicity}
-in TM systems
+of transactions
 \cite{htm-subtleties,stm-weak-atomicity}.
@@
 % \cite{bugtm}. % sec 3.1 of that paper, not sure of an earlier citation that also states it in these exact terms
-However, under weakly-atomic STM systems,
-transactions do not abort when
-a non-transactional conflicting access interleaves within them,
+However, weakly-atomic STM systems
+do not abort transactions when
+a non-transactional access conflicts,
 and condition \#1 of the execution semantics from \cref{sec:tm-design-failure} does not hold.
-To model weak atomicity, we must weaken \Cref{lem:equiv} as follows (updated parts underlined for emphasis).
+To model weak atomicity, \Cref{lem:equiv} must be extended with a third possibility:
+%as follows (updated parts underlined for emphasis).
 \vspace{1em}
 
-\begin{lemma}[Equivalence of Aborts \weakhilight{under Weak Atomicity}]
+\begin{lemma}[Weak Equivalence of Aborts]
 	\label{lem:equiv-weak}
-	For any interleaving prefix
-	\[
-	\begin{tabular}{c}
-		$\tiiat{\alpha},\tiiat{\beta_1}\dots\tiiat{\beta_b},$\\
-		$\tjjat{\gamma_1}\dots\tjjat{\gamma_j},$ \\
-		$\tkkat{\kappa_1}\dots\tkkat{\kappa_k},$ \\
-		$\tiiat{\beta_{b+1}}$ % $\dots\tiiat{\beta_{n-1}}$ -- excluded bc might abort
-	\end{tabular}
-	\]
-	with $b<n$, $j \ne i$, $k \ne i$, et cetera, either:
+	For any interleaving prefix such as defined in \Cref{lem:equiv}, either one of the original two conditions hold, or:
 	\begin{enumerate}
-		\item \weakhilight{All events in
+		\setcounter{enumi}{2}
+		\item All events in
 			$\tjjat{\gamma_1}\dots\tjjat{\gamma_j},
@@
 			 % ehh, bad parallel structure with below regarding the verb 'exists'
-			are non-transactional, and the original interleaving itself exists, or}
-		\item $\tiiat{\alpha},\tjjat{\gamma_1}\dots\tjjat{\gamma_j},\tkkat{\kappa_1}\dots\tkkat{\kappa_k},\tiiat{\phi_1}\dots$
-			(conflicting case), or
-		\item $\tiiat{\alpha},\tiiat{\beta_1}\dots\tiiat{\beta_b}\dots\tiiat{\beta_n},\tjjat{\gamma_1}\dots\tjjat{\gamma_j},\tkkat{\kappa_1}\dots\tkkat{\kappa_k}$
-			(independent case)
+			are non-transactional, and the original interleaving itself is a legal execution.
 	\end{enumerate}
-	exists and is observationally equivalent.
 \end{lemma}
@@
 \begin{proof}
-	As before, \weakhilight{except first case on
+	As before, except first case on
 	whether any intervening operations by $\tjj$/$\tkk$ are transactional.%
-	\footnote{If $\tjj$/$\tkk$ already have a transaction active,
+	\footnote{If $\tjj$ or $\tkk$ already have a transaction active,
 	their {\sf begin} must have occurred before $\tiiat{\alpha}$,
 	and the theorem can apply starting from the earlier {\tt begin} with the thread IDs swapped.}
-	If not, $\tii$ may ultimately proceed with $\tiiat{\beta_{b+1}}$ without aborting,
-	and the new case \#1 applies.}
+	If so, either the original case \#1 or \#2 applies.
+	If not, $\tii$ may ultimately proceed with $\tiiat{\beta_{b+1}}$
+	(its preempted next transactional instruction)
+	without aborting,
+	and the new case \#3 applies instead.
 \end{proof}
@@
 
-\begin{theorem}[\weakhilight{Weak} Atomicity of Transactions]
+\begin{theorem}[Weak Atomicity of Transactions]
 	\label{thm:atom-weak}
 	For any state space $S$ of a transactionally-concurrent program,
-	an equivalent state space exists in which all transactions are either executed atomically
-	\weakhilight{with respect to other transactions only},
+	an equivalent state space exists in which all transactions are either executed
+	{\bf atomically
+	with respect to other transactions only},
 	or aborted immediately.
@@
 
-Conveniently, in \cref{sec:tm-design-locks}'s formulation of
+Conveniently,
+\Cref{thm:atom-weak} still guarantees transaction-transaction atomicity,
+so mutual exclusion of $m$ is preserved
+in \cref{sec:tm-design-locks}'s formulation of
 $I' = \exists m. \mathsf{lockify}_m(I)$,
-\Cref{thm:atom-weak} still guarantees transaction-transaction atomicity,
-so mutual exclusion of $m$ is preserved,
 %(a global lock does not protect against interleaving by arbitrary other code that never takes the lock either)

### 2018-12-08 thesis/tm.tex
tweak a bunch of smalle things in tm; confirm/adjust figure placement

@@
 
-\subsection{Modeling Transaction Failure}
+\subsection{Modeling transaction failure}
 \label{sec:tm-design-failure}
@@
 
-\subsection{Failure Injection}
+\subsection{Failure injection}
 \label{sec:txn-failure}
@@
 
-Up to four abort codes are considered depending on the testing options listed in the previous section.
+Up to four abort codes are considered depending on \cref{sec:tm-interface}'s testing options.
 All abort codes are simultaneously added to both lists,
@@
 		which the specification allows to abort for any reason.%
-		\footnote{As far as I know, the reason for syscall aborts is both hardware- and kernel-dependent,
+		\footnote{\revision{As far as I know, the reason for syscall aborts is both hardware- and kernel-dependent,
 		and in fact nondeterministic;
@@
 		For simplicity of both implementation and client programs,
-		Landslide always distinguishes system call aborts with {\tt \_XABORT\_CAPACITY}.}
+		Landslide always distinguishes system call aborts with {\tt \_XABORT\_CAPACITY}.}}
 		See below for discussion of other cache capacity concerns.}
@@
 
-\subsection{Data Race Analysis}
+\subsection{Data race analysis}
 \label{sec:txn-datarace}
@@
 
-The notation {\tt testname}$(K,N)$ will denote a test configuration of $K$ threads, each running $N$ iterations of the test logic.
+The notation {\tt testname}$(K,N)$ %will denote
+denotes a test configuration of $K$ threads, each running $N$ iterations of the test logic.
 All tests were run on an 8-core 2.7GHz Core i7 with 32 GB RAM.
@@
 Landslide also found two bugs in the ``real-world'' transactional algorithms I tested.
-\begin{itemize}
-	\item
-	{\bf Atomicity violation.}
-	{\tt avl\_insert} with any parameters higher than (2,1)
-	exposed a previously-unknown bug in the transactional AVL tree.
-	\Cref{fig:avlbug} shows the root cause, essentially the {\tt htm1} bug in disguise.
-	This manifested
-	alternately
-	% landslide-trace-1533850116.527842.html
-	% landslide-trace-1533850236.873032.html
-	as a segfault (for test parameters (2,2) and (3,1))
-	and
-	% landslide-trace-1533854371.201589.html
-	as a consistency-check assertion failure (for test parameters (2,3)).
-	The presence of {\tt while (\_retry);} makes the necessary preemption window extremely small
-	(between it and {\tt \_xbegin()}),
-	making the bug extremely unlikely to manifest under stress testing,
-	but Landslide is blind to such matters of chance.
-
-	As a matter of full disclosure,
-	I noted that the loop does not affect the test's possible behaviours,
-	only its {\em likely} ones,
-	and so removed it to make the test more Landslide-friendly.
-	%keep the state space size manageable.
-	%Lest any reader think I hacked the test after already knowing about the bug,
-	To dispel any doubt about bias or test hacking,
-	I confirmed that Landslide still finds the bug with the spin loop unmodified,
-	on (3,1) in the same 53 interleavings,
-	although it suffers resource exhaustion
-	% from dealing with the loop
-	on parameters of (2,2) or greater.
-	% specifically, it runs out of file descriptors, since on bochs you need 1 fd per PP,
-	% and the branch just gets 1000+ PPs deep
-	\item
-	{\bf Spurious spinlock abort.}
-	{\tt lock\_fast} discovered a spurious transactional-path write conflict
-	in the {\tt spinlock} HTM-lock implementation.%
-	\footnote{{\tt lock\_fast}'s unusual (9,9) parameter shows that this state space size is constant: %in $(K,N)$:
-	DPOR will always either deem all thread transitions independent and end exploration immediately,
-	or the test's assertion will trip as soon as the first conflict is found.}
-	This ``performance bug'' causes the lock to suffer slow-path spin-locking
-	even in cases where the user's thread transitions are completely independent
-	(for example, locking the root of an AVL tree,
-	then traversing in different directions to make disjoint modifications),
-	which the test case detects with {\tt \_xtest()}.
-	\Cref{fig:spinlockbug} shows the root cause:
-	the {\tt isfree()} routine (corresponding to the AVL's {\tt \_retry})
-	used an atomic compare-and-swap that would always write to memory even without modifying it.
-	I corrected this in the {\tt spin\_fixed} implementation by replacing it
-	with a normal read
-	(being used only in the transactional path, no barriers are required to protect it).
-	A cursory search on Github found one user of this code,
-	a transactional LevelDB implementation \cite{htm-leveldb-github},
-	whose author had also noticed and corrected this problem in the same way.
-
-	As another matter of full disclosure,
-	I noticed this bug through manual inspection
-	while adapting the spinlock's client code to be Landslide-friendly,
-	then wrote {\tt lock\_fast} specifically to target this behaviour,
-	so unlike {\tt avl\_insert}, it does not count as Landslide finding a previously-unknown bug.
-	However, I feel in retrospect that how and when
-	an HTM-backed concurrency abstraction
-	will fall into its slow path
-	is a reasonable performance property for a user to want to verify,
-	so I consider Landslide confirming the bug (and later verifying its absence, in \cref{sec:tm-verif})
-	a positive result anyway.
-\end{itemize}
 
@@
 
-\begin{figure}[h]
+\begin{itemize}
+	\item
+	{\bf Atomicity violation.}
+	{\tt avl\_insert} with any parameters higher than (2,1)
+	exposed a previously-unknown bug in the transactional AVL tree.
+	\Cref{fig:avlbug} shows the root cause, essentially the {\tt htm1} bug in disguise.
+	This manifested
+	alternately
+	% landslide-trace-1533850116.527842.html
+	% landslide-trace-1533850236.873032.html
+	as a segfault (for test parameters (2,2) and (3,1))
+	and
+	% landslide-trace-1533854371.201589.html
+	as a consistency-check assertion failure (for test parameters (2,3)).
+	The presence of {\tt while (\_retry);} makes the necessary preemption window extremely small
+	(between it and {\tt \_xbegin()}),
+	making the bug extremely unlikely to manifest under stress testing,
+	but Landslide is blind to such matters of chance.
+
+	As a matter of full disclosure,
+	I noted that the loop does not affect the test's possible behaviours,
+	only its {\em likely} ones,
+	and so removed it to make the test more Landslide-friendly.
+	%keep the state space size manageable.
+	%Lest any reader think I hacked the test after already knowing about the bug,
+	To dispel any doubt about bias or test hacking,
+	I confirmed that Landslide still finds the bug with the spin loop unmodified,
+	on (3,1) in the same 53 interleavings,
+	although it suffers resource exhaustion
+	% from dealing with the loop
+	on parameters of (2,2) or greater.
+	% specifically, it runs out of file descriptors, since on bochs you need 1 fd per PP,
+	% and the branch just gets 1000+ PPs deep
+	\item
+	{\bf Spurious spinlock abort.}
+	{\tt lock\_fast} discovered a spurious transactional-path write conflict
+	in the {\tt spinlock} HTM-lock implementation.%
+	\footnote{{\tt lock\_fast}'s unusual (9,9) parameter shows that this state space size is constant: %in $(K,N)$:
+	DPOR will always either deem all thread transitions independent and end exploration immediately,
+	or the test's assertion will trip as soon as the first conflict is found.}
+	This ``performance bug'' causes the lock to suffer slow-path spin-locking
+	even in cases where the user's thread transitions are completely independent
+	(for example, locking the root of an AVL tree,
+	then traversing in different directions to make disjoint modifications).
+	The test case detects this with {\tt \_xtest()}.
+	\Cref{fig:spinlockbug} shows the root cause:
+	the {\tt isfree()} routine (corresponding to the AVL's {\tt \_retry})
+	used an atomic compare-and-swap that would always write to memory even without modifying it.
+	I corrected this in
+	% the
+	{\tt spin\_fixed}
+	%implementation
+	by replacing it
+	with a normal read
+	(being used only in the transactional path, no barriers are required to protect it).
+	A cursory search on Github found one user of this code,
+	a transactional LevelDB implementation \cite{htm-leveldb-github},
+	whose author had also noticed and corrected this problem in the same way.
+
+	As another matter of full disclosure,
+	I noticed this bug through manual inspection
+	while adapting the spinlock's client code to be Landslide-friendly,
+	then wrote {\tt lock\_fast} specifically to target this behaviour,
+	so unlike {\tt avl\_insert}, it does not count as Landslide finding a previously-unknown bug.
+	However, I feel in retrospect that how and when
+	an HTM-backed concurrency abstraction
+	will fall into its slow path
+	is a reasonable performance property for a user to want to verify,
+	so I consider Landslide confirming the bug (and later verifying its absence, in \cref{sec:tm-verif})
+	a positive result anyway.
+\end{itemize}
+
+\begin{figure}[t]
 	\begin{center}
@@
 
-\begin{table}[h]
+% TODO: check figure placement
+\begin{table}[t]
 	\begin{center}

### 2018-12-09 thesis/related.tex
fix double cite of symbolic execution paper

@@
 which seeks good coverage of possible thread execution paths under schedule nondeterminism,
-another popular testing approach is symbolic execution \cite{symbolic-execution},
+another popular testing approach is symbolic execution \cite{symbolic},
 which seeks good coverage of possible flow control paths under input nondeterminism.

### 2018-12-09 thesis/education.tex
spell numbers

@@
 		\cref{sec:education-survey-pintos}).
-	\item A large evaluation spanning 9 semesters of operating systems classes
+	\item A large evaluation spanning nine semesters of operating systems classes
 		(\cref{sec:education-eval-bugfinding}),
-		2 of which were held at other universities besides CMU,
+		two of which were held at other universities besides CMU,
 		in which 145 students used Landslide,

### 2018-12-09 thesis/related.tex
rephrase & rearrange a few things in related

@@
 I would distinguish it by its ability to find shared memory preemption points via dynamic tracing,
-rather than relying on user annotations or imprecise compiler instrumentation
-as other tools do.
+rather than relying on user annotations or imprecise compiler instrumentation.
+%as other tools do.
 Compared to all other tools I know of,
@@
 the Sisyphean rock of stateless MC.
-The notion that some interleavings of concurrent threads could lead to indistinguishable program states and be therefore redundant,
+The notion that some interleavings
+%of concurrent threads
+could lead to indistinguishable program states and be therefore redundant,
 known as {\em partial order reduction} (POR),
@@
 Pintos \cite{pintos} has recently emerged as the most popular educational kernel
-(by count of top CS schools in the United States who teach by it);
+(by count of top CS schools in the United States who use it);
 it trades off the prevalence of its concurrency challenges to cover various OS topics more broadly,
@@
 and also had difficulty understanding single failing executions.
-In fact, the study explicitly recommended tool support for testing many thread interleavings automatically (model checking)
+In fact, the study explicitly recommended tool support
+for testing many interleavings automatically (model checking)
 and for execution traces to communicate sequences of important events (preemption traces),
@@
 (for example, the {\tt atomic\_*} tests from \cref{sec:education-pebbles-tests})
-that could ultimately lead to a higher solve rate on Landslide's bug reports during P2 (\cref{sec:education-eval-bugs-cmu}).
+that could ultimately lead to a higher solve rate on Landslide's bug reports during P2~(\cref{sec:education-eval-bugs-cmu}).
+%
+\revision{The Deadlock Empire~\cite{deadlockempire} is an educational introduction to concurrent programming and debugging,
+presented as an online puzzle-hunt of sorts, backed up by a fantasy-themed narrative.
+Each puzzle requires the user to manually step through the execution of multiple threads
+to find buggy interleavings.
+%which in some cases requires the use of an ``expand'' button that
+Though itself lacking in scientific analysis of its educational power,
+it demonstrates the kind of user-friendly features and overall approach
+that many students expressed desire for in their survey responses (\cref{sec:education-eval-survey}).
+Future work could extend Landslide's preemption traces to include similar interactivity.}
 
@@
 which is counter to 15-410's educational goals.
-
-\revision{The Deadlock Empire~\cite{deadlockempire} is an educational introduction to concurrent programming and debugging,
-presented as an online puzzle-hunt of sorts, backed up by a fantasy-themed narrative.
-Each puzzle requires the user to manually step through the execution of multiple threads
-to find buggy interleavings.
-%which in some cases requires the use of an ``expand'' button that
-Though itself lacking in scientific analysis of its educational power,
-it demonstrates the kind of user-friendly features and overall approach
-that many students expressed desire for in their survey responses (\cref{sec:education-eval-survey}).
-Future work could extend Landslide's preemption traces to include similar interactivity.}
 

### 2018-12-09 thesis/related.tex
cite rust mutex/rwlock stuff

@@
 %
-Rust \cite{rust-language}
+Rust %\cite{rust-language}
 presents a type system with more explicit memory management,
@@
 It proposes a borrow-check analysis to ensure memory and type safety
-despite mutable references,
-and a trait system to ensure no shared state between threads by default.
+despite mutable references \cite[\S{}4.2,\S{}10.3]{rust-book},
+and a trait system to ensure no shared state between threads by default \cite[\S{}10.2,\S{}16.4]{rust-book}.
 Its concurrency libraries then offer interfaces which relax this restriction,
@@
 using the type system to enforce sound locking discipline across such accesses,
-again preserving type soundness and data-race freedom.%
-\footnote{The author themself contributed the original design for this latter feature.}
+again preserving type soundness and data-race freedom \cite[\S{}16.2]{rust-book}.%
+\footnote{The author themself contributed the original design for this latter feature \cite{rust-rwarc}.}
 I know of no existing model checker for Rust as of yet.

### 2018-12-09 thesis/overview.tex
clean up some citation issues

@@
 \subsubsection{Shared state}
-When a programming language offers multithreaded parallelism but forbids access to any shared state between threads \cite{rust-language},
+When a programming language offers multithreaded parallelism but forbids access to any shared state between threads \cite{rust-book},
 the simultaneity of threads is largely irrelevant to the program's behaviour.

### 2018-12-09 thesis/thesis.tex
weak atomicity description, proof, & implementation, wee
clean up some citation issues

@@
 \PassOptionsToPackage{hyphens}{url}
-\usepackage[backref,pageanchor=true,plainpages=false, pdfpagelabels, bookmarks,bookmarksnumbered,
+\usepackage[backref=page,pageanchor=true,plainpages=false, pdfpagelabels, bookmarks,bookmarksnumbered,
 colorlinks=true,

### 2018-12-10 thesis/tm.tex
summarize tm chapater

@@
 Landslide requires the other STM options ({\tt -A -S}) also to be enabled to emulate weak atomicity,
-because I know of no existing HTM system which implements it.
+partly because I know of no existing HTM system which implements it,
+but mostly to keep the implementation simple.
 As one might expect, it inflates state spaces considerably, % TODO: measure it with htm2 or some such?
@@
 In this program, each thread's transactional path conflicts with the other thread's abort path,
-while the two transactional paths (disjoint memory accesses) and the two abort paths (reads only) are pairwise independent.
+while the two transactional paths (disjoint memory accesses) and the two abort paths (reads only) are
+independent with each other.
 
@@
 
-% TODO: is this true? at all? htm2 is bigger in STM only because it's got a separate expliit abort
 Likewise, the cases of {\tt htm2}(2,3), (2,4), and (4,1),
 in which retry sets achieved better reduction than STM mode,
-show that the latter does not necessarily subsume the former,
-and that combining the two could in theory achieve further reduction still.
+\revisionminor{suggest} that the latter does not necessarily subsume the former,
+and that combining the two could in theory achieve further reduction still,
 However, {\tt xbegin} results other than {\tt \_XABORT\_RETRY}
@@
 ),
-so how to soundly prune either success paths or retry aborts
-while other abort codes are in play
-remains an open problem.
+so \revisionminor{it remains an open problem}
+to soundly prune either success paths or retry aborts
+while other abort codes are in play,
+\revisionminor{and preserve the reduction achieved with the baseline configuration}.
 
@@
 provides an example of how to use compiler intrinsics to emit the necessary barriers.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{Summary}
+
+\revision{This chapter has extended the concurrency model of stateless model checking to support transactional memory,
+and proposed several new reduction strategies for coping with the concomitant % hi william
+state space explosion.
+I have made the following contributions:
+
+\begin{itemize}
+	\item Three equivalence proofs,
+		demonstrating that model checkers can soundly emulate strongly-atomic HTM execution semantics by
+		treating transactions as globally-uninterruptible atomic sections
+		and injecting failure without simulating rollback
+		(\cref{sec:tm-design-failure})
+		and applying conventional data race analysis with all transactions treated as a global lock
+		(\cref{sec:tm-design-locks}),
+		and can emulate weakly-atomic STM systems by relaxing the above atomicity to apply to other transactions only
+		(\cref{sec:tm-weak-atomicity}).
+	\item An implementation of both forms of transactional memory in Landslide,
+		with further options to configure the variety of abort codes injected
+		(\cref{sec:tm-implementation}).
+	\item A suite of transactional benchmark programs,
+		designed to produce interesting state spaces of varying size, bugginess, and reducibility
+		(\cref{sec:tm-eval-exp-setup}).
+		This test suite is more appropriate for evaluating model checkers
+		% get rekt STAMP :triumph:
+		than prior work, which to date has focused on testing the raw performance of TM backend implementations,
+		with little focus on the interactions of abort paths.
+	\item Two previously-unknown bugs found in transactional programs from prior work
+		(\cref{sec:tm-eval-bugs}).
+	\item Retry set reduction,
+		a new state space reduction algorithm
+		(\cref{sec:tm-retrysets})
+		which the evaluation shows identifies and skips equivalent interleavings
+		compared to standard DPOR, even when transactional and abort paths are fully mutually conflicting
+		(\cref{sec:tm-eval-retrysets}).
+	\item Demonstration of two further heuristic reduction strategies on HTM programs,
+		abort code reduction (\cref{sec:tm-eval-stm})
+		and abstraction reduction (\cref{sec:tm-abstraction})
+		which rely on human intuition to ensure their soundness
+		to tackle even larger state spaces still.
+	\item Verifications of \Cref{fig:htm-fixed}'s atomicity protocol %(with retry set reduction)
+		and the {\tt lock}({\tt mutex}) transactional lock %(with abort code reduction)
+		up to 2 threads, 4 iterations and 4 threads, 1 iteration for each
+		(\cref{sec:tm-verif}).
+		%$(K,N)$=$(2,4)$ and $(4,1)$ for each.
+		% verified htm atomicity protocol (htm2) and mutex (lock mutex) up to (2,4), (4,1)
+\end{itemize}
+
+As concurrent systems grow in complexity to meet modern performance demands,
+they grow ever more fraught with opportunity for subtle bugs,
+and the programmer likewise faces ever more difficulty in verifying her programs.
+Following in the footsteps of related work for relaxed memory concurrency,
+this chapter has enabled stateless model checking to address yet another new technique,
+and concludes this thesis's contributions to the field.
+The upcoming chapters will offer a tour of related research problems,
+solved and unsolved,
+practical and philosophical,
+past and future.
+
+}

### 2018-12-10 thesis/landslide.tex
minor changes landslide

@@
 and which thread to ``have switched to''
-\footnote{\revision{willan on-having switched to \cite{hhgg-reu}}}
+\footnote{\revisionminor{willan on-having switched to \cite{hhgg-reu}}}
 instead, in {\tt explore.c},
@@
 implementing such an ordering would be much less straightforward.
-\footnote{\revision{The pseudocode algorithms found in many conference papers
+\footnote{\revisionminor{The pseudocode algorithms found in many conference papers
 often fail to be straightforward to translate into usable implementations~\cite[Figure 5]{itg2}.}}
@@
 	\item Adds a mutex testing mode to configure whether the
-		data race analysis should consider lock' internal accesses
+		data race analysis should consider locks' internal accesses
 		(\cref{sec:landslide-dynamicconfig}; {\tt TESTING\_MUTEXES}).

### 2018-12-10 thesis/intro.tex
fix intro

@@
 
-To take advantage of multiple cores for performance, programmers must write software to execute {\em concurrently} --
-using multiple {\em threads} which execute multiple parts of a program's logic simultaneously.
-However, when threads access the same shared data, they may interleave in unexpected ways which change the outcome of their execution.
+To take advantage of multiple cores for performance, programmers must write software to execute concurrently,
+i.e.,
+using multiple threads to run several parts of a program's logic simultaneously.
+However, when threads access the same shared data,
+they may interleave in unexpected ways which change the outcome of their execution,
+\revisionminor{just as when multiple travelers interact with car doors at the same time.}
 When an unexpected interleaving produces undesirable program behaviour,
 for example, by corrupting shared data structures,
+\revisionminor{or by leaving the door locked as in the car example,}
 we call it a {\em concurrency bug}.
-Concurrency bugs are notoriously hard for programmers to find and debug
-because the specific thread interleaving required to trigger them arises at random during normal execution,
+\revisionminor{The specific interleaving required to expose such a bug}
+arises at random during normal execution,
 and often with very low probability.
-%Concurrency bugs are notoriously hard to find and reproduce because they only appear in specific thread interleavings, which arise at random during normal program execution.
-% TODO(LAYMAN): give example of trying to open car door at same time as friend turns key to unlock it.
-
+%
 Most commonly, a programmer searches for concurrency bugs in her code by running it many times (in parallel, in serial, or both),
-hoping that eventually, it will run according to the particular interleaving required to expose a hypothetical bug.
+hoping that eventually,
+\revisionminor{she will chance upon such an interleaving should one exist.}
+%it will run according to the particular interleaving required to expose a hypothetical bug.
 This technique, known as {\em stress testing}, is unreliable,
@@
 Nevertheless, stress testing remains popular because of how easily a programmer can use it:
-she simply wraps her program in a loop, sets it to run overnight, and kills it if her patience runs out before it finds a bug.
+she simply wraps her program in a loop, sets it to run overnight, and \revisionminor{interrupts} it if her patience runs out before it finds a bug.
 
@@
 or to verify their absence,
-which provides more reliable coverage, progress, and verification than stress testing.
-A stateless model checker tests a program by forcing it to execute a new unique thread interleaving on each iteration of the test,
-capturing and controlling the randomness in a finite state space of all possible interleavings.
-
-Unfortunately, the size of these state spaces is exponentially proportional to the size of the tested program.
-% TODO(LAYMAN): explain exponential explosion by relating the parable of grains of rice on a chessboard.
-For even moderately-sized programs, there may be more possible ways to interleave every thread's every instruction
+which provides more reliable coverage, \revisionminor{reproducibility}, and verification than stress testing.
+A stateless model checker tests programs by forcing them to execute a new unique thread interleaving on each iteration,
+%of the test,
+capturing and controlling the randomness in a finite {\em state space} of all possible interleavings.
+%
+%Unfortunately,
+%the size of these state spaces is exponentially proportional to the size of the tested program.
+\revisionminor{However, attempting to exhaustively check the entirety of such state spaces is akin to Sissa's reward:}
+for even moderately-sized programs, there may be more possible ways to interleave every thread's every instruction
 than particles in the universe.
@@
 so \revisionminor{a checker must} be told how to identify and control the execution of each thread.
-Some expert concurrency research wizards may be willing to add manual annotations to their code,
-but required manual effort is a serious downside for anyone with a looming deadline,
-and especially so for students who are still learning basic concurrency principles.
+\revision{Many undergraduate computer science curricula
+culminate in project-oriented systems classes
+\cite{thrlib,kspec,pintos}
+in which
+students implement these very types of programs,
+but are left to their own devices when it comes time to root out bugs.}
+Some expert concurrency research wizards may be willing to add manual annotations to their code
+\revisionminor{for the sake of verification},
+but requiring manual effort is a serious downside for anyone with a looming deadline,
+and especially so for students who are still learning basic concurrency principles \revisionminor{in the first place}.
 %We should not expect programmers to add effortful manual annotations to their code,
@@
 
+\revision{Finally,
+in the struggle to meet ever-increasing performance demands,
+hardware and software alike has grown more and more complicated features for fine-grained concurrency control.
+Recent stateless model checking research is already beginning
+to address some such cases,
+allowing for new sources of nondeterminism beyond simply the ability to interleave threads arbitrarily.
+For example,
+relaxed memory consistency allows for multicore systems to
+reorder accesses to main memory \cite{memory-consistency-models},
+which introduces store buffer nondeterminism in addition to thread nondeterminism \cite{tsopso},
+and event-driven programming models,
+useful especially in mobile applications to improve responsiveness and reduce power consumption,
+allow for threads to reenter themselves via event handlers \cite{r4}.
+Transactional memory,
+which allows the programmer to specify arbitrary atomic execution sequences,
+has been supported by software libraries for decades \cite{transactional-memory},
+but only recently have processor manufacturers introduced hardware-backed transactions~\cite{htm-haswell},
+replete with new complexity,
+and no verification tool which can
+%model the complex concurrent interactions that arise in programs that use them.
+accurately model the concurrency of programs that use them yet exists.}
+
 \section{Contributions}
 
-This thesis will solve both problems discussed above.
-My thesis statement is as follows:
+\revision{This thesis will address each of the problems introduced above,
+%establishing stateless model checking
+%as a testing technique that
+%you might come away from this actually wanting to use,
+%instead of as a mere curiosity appearing in obscure conference papers.
+establishing stateless model checking
+as a modern verification technique
+that can meet realistic human needs.
+My thesis statement is as follows:}
 
@@
 
-I have built Landslide \cite{landslide}, a stateless model checker for thread libraries and kernels,
-and I have developed some techniques for automatically choosing the best thread interleavings to test
-and for automatically instrumenting operating system kernels in an educational setting.
-This thesis will comprise three major contributions:
+\revision{The foundation of this work is Landslide, a stateless model checker
+%for thread libraries and kernels that
+I have built over the last seven years, first debuting in my M.S. thesis~\cite{landslide}.
+Since then, I have extended it with a wide array of features and algorithms,
+some from prior work and some of my own devising,
+some which work behind the scenes and others which rely on human feedback to be effective,
+in pursuit of this thesis statement.
+
+The first half of the statement will serve as the overarching theme of this work:
+that as impossible a problem as perfect formal verification may be,
+one may still hope to achieve meaningful test results
+using an algorithm that knows its own limits,
+and compensates for them with heuristics informed by user's own human intuition.
+The second half of the statement can be broken down into three parts,
+coping with exponential explosion,
+helping students,
+and addressing modern concurrency models,
+which correspond to the three major contributions of this thesis, as follows:}
 
 \begin{enumerate}
-	\item {\bf Meaningful state spaces (\Cref{chap:quicksand}).}
-		I will present {\em Iterative Deepening}, a new algorithm for navigating the trade-off in how many preemption points to test at once.
-		Iterative Deepening incorporates state space estimation \cite{estimation} to decide on-the-fly whether each state space is worth pursuing, and uses data race analysis \cite{tsan} to find new preemption point candidates based on a program's dynamic behaviour.
-		This section will include a large evaluation of the technique, comparing its performance to three \revisionminor{prior-work} approaches across 600+ unique tests.
-		I will show that Iterative Deepening of preemption points outperforms prior work in terms \revisionminor{of both} finding bugs quickly and completely verifying correctness when no bug exists.
+	\item \revisionminor{{\bf Data-race preemption points and Iterative Deepening (\Cref{chap:quicksand}).}
+		I will present Quicksand, a stateless model checking execution framework
+		that manages multiple simultaneous Landslide instances
+		to automatically cope with exponential explosion.
+		Quicksand is powered by {\em Iterative Deepening},}
+		a new algorithm for navigating the trade-off in how many preemption points to test at once.
+		Iterative Deepening incorporates state space estimation \cite{estimation}
+		to decide on-the-fly whether each state space is worth pursuing,
+		and uses data race analysis \cite{tsan,fasttrack}
+		to find new preemption point candidates based on a program's dynamic behaviour.
+		\revision{This chapter will include a soundness proof,
+		showing that testing only synchronization and data-race preemption points
+		still constitutes a full formal verification of the test,
+		and}
+		a large evaluation, comparing its performance to three \revisionminor{prior-work} approaches
+		across 600+ unique tests.
+		I will show that \revisionminor{Quicksand}
+		outperforms prior work in terms \revisionminor{of both} finding bugs quickly
+		and verifying correctness when no bug exists.
 	\item {\bf Educational use (\Cref{chap:education}).}
-		For the past five semesters, I have offered a fully-automated version of Landslide to students in 15-410, CMU's undergraduate Operating System Design and Implementation class \cite{kspec,thrlib}, for use as a debugging aid during the thread library project.
-		Recently I have also extended Landslide to handle Pintos kernel projects from other universities \cite{pintos}.
-		In the two most recent semesters, I collaborated with Operating Systems course staff at two such schools, the University of Chicago and University of California at Berkeley,
-		to provide debugging feedback to their students.
-
+		For the past \revisionminor{seven} semesters,
+		I have offered a fully-automated version of Landslide to students in 15-410,
+		CMU's undergraduate Operating System Design and Implementation class,
+		for use as a debugging aid during the thread library project \cite{thrlib}.
+		%Recently
+		I have also extended Landslide to handle Pintos kernel projects from other universities \cite{pintos}.
+		In the two most recent semesters, I collaborated with Operating Systems course staff at the University of Chicago,
+		\revisionminor{which uses Pintos,
+		and at The Pennsylvania State University,
+		which has recently adopted CMU's thread library project,}
+		to provide
+		%debugging feedback to their students.
+		\revisionminor{their students an opportunity to benefit from Landslide as well.}
 		At all three universities I then collected statistics on the numbers and types of bugs found,
@@
 		This section will present the study's results
-		to evaluate the suitability of stateless model checking in an educational setting.
-	\item {\bf Hardware Transactional Memory (HTM) (\Cref{chap:tm}).}
-		HTM is a relatively new concurrent programming technique \cite{htm-haswell,htm-experience}
-		which is not yet addressed by modern model checkers.
-		I have extended Landslide's concurrency model to support
-		HTM's execution semantics and several advanced features, % abort codes p much
-		%both hardware (HTM) and software (STM) variants of TM,
-		and tested several ``real-world'' HTM programs and benchmarks.
-		This section will discuss the theoretical techniques I used to model the new form of concurrency,
-		present associated correctness proofs of my approach,
-		and show the verification results.
+		\revisionminor{and prove that stateless model checking is suitable for use} in an educational setting.
+	\item {\bf Hardware transactional memory (HTM) (\Cref{chap:tm}).}
+		\revision{I will introduce a new concurrency model
+		for stateless model checkers to emulate HTM's execution semantics
+		in terms of existing concurrency primitives.
+		This model is accompanied by a soundness proof
+		which allows checkers to avoid simulating conflict abort rollbacks,
+		reducing the state space substantially,
+		while preserving the formal verification guarantee.
+		I have implemented a new transactional testing mode in Landslide,
+		with support for the many different abort reasons HTM introduces,
+		optional weak atomicity semantics for simulating software TM instead,
+		and a ``retry set'' reduction algorithm to identify and prune new types of equivalences.
+		The evaluation includes several real-world HTM programs and benchmarks,
+		and shows that Landslide can both find bugs and verify correctness
+		using the new model with reasonable performance.}
 \end{enumerate}
 
-\section{Organization}
+\section{Meta}
 
-The rest of this dissertation is organized as follows.
-
-\begin{itemize}
-	\item {\bf Background:} \Cref{chap:background} will present the requisite background material on concurrent programming, stateless model checking, and the various types of programs targeted by Landslide.
-	\item {\bf Landslide:} \Cref{chap:landslide} explains the design and implementation of Landslide
-		%the core of my stateless model checking framework,
-		and all the special features it's been equipped with over the years.
-		It is the foundation upon which all three contributions above build.
-	\item {\bf Quicksand:} \Cref{chap:quicksand} presents the Iterative Deepening framework which more intelligently chooses which state spaces to test, corresponding to contribution 1 above.
-	\item {\bf Education:} \Cref{chap:education} discusses my evaluation of Landslide
-		in CMU's 15-410 class environment using the Pebbles kernel,
-		and in the University of Chicago's and Berkeley's OS class environments using the Pintos kernel,
-		corresponding to contribution 2 above.
-	%\item {\bf Pebbles:} \Cref{chap:410} discusses my evaluation of Landslide in CMU's 15-410 class environment using the Pebbles kernel, corresponding to contribution 2 above.
-	%\item {\bf Pintos:} \Cref{chap:410} discusses my evaluation of Landslide in the University of Chicago's and Berkeley's OS class environments using the Pintos kernel, corresponding to contribution 2 above.
-	\item {\bf Transactions:} \Cref{chap:tm} presents my extension of Landslide's concurrency model to handle transactional concurrency and the evaluation thereof, corresponding to contribution 3 above.
-	\item {\bf Related Work:} \Cref{chap:relatedwork} honors my neighbours and ancestors in research spirit.
-	\item {\bf Future Work:} \Cref{chap:warpzone} lists stateless model checking's remaining shortcomings and how new research might address them.
-	\item {\bf Conclusion:} \Cref{chap:conclusion} provides some thoughts on the future of the field.
-\end{itemize}
-
-\section{Notes on reading this thesis}
+\revision{Dear reader, whether you have come to read this whole document,
+to pick and choose whichever sections interest you,
+or just to skim the figures and footnotes looking for easter eggs,
+it is truly an honor to have you here.
+Before we get underway,
+allow me to present some notes on reading this thesis
+that may soften the blow of these 200+ pages.}
 
@@
 The impersonal ``the programmer'' will be referred to as she/her
-to challenge readers' unconscious bias about gender in computer science,
-as well as to highlight her role as the intended user, separate from the underlying research.
-%to distinguish her from the thesis audience concerned with the underlying research,
+to highlight her role as the intended user, separate from the underlying research,
+as well as to challenge readers' unconscious bias about gender in computer science.
 Gender-neutral pronouns will be used to refer to
@@
 as well as
-on the author themself.
+on the author themself,
+who would also prefer them for any third-person reference by you.
 
@@
 
+\section{Organization}
+
+The rest of this dissertation is organized as follows.
+
+\begin{itemize}
+	\item {\bf Background:} \Cref{chap:background} will present the requisite background material on concurrent programming, stateless model checking, and the various types of programs targeted by Landslide,
+		\revisionminor{concluding with a glossary of terminology for the reader's convenient reference.}
+		%(\cref{sec:glossary})
+	\item {\bf Landslide:} \Cref{chap:landslide} explains the design and implementation of Landslide
+		%the core of my stateless model checking framework,
+		and all the special features it's been equipped with over the years.
+		It is the foundation upon which all three contributions above build.
+	\item {\bf Quicksand:} \Cref{chap:quicksand} introduces
+		\revisionminor{Quicksand,
+		a wrapper framework for Landslide
+		which incorporates data-race analysis at run-time to find new preemption points
+		and uses the new Iterative Deepening algorithm to}
+		more intelligently choose which state spaces to test, corresponding to contribution 1 above.
+	\item {\bf Education:} \Cref{chap:education} discusses my evaluation of Landslide
+		in CMU's
+		\revisionminor{and The Pennsylvania State University's
+		class environment using thread libraries based on the Pebbles kernel},
+		and in the University of Chicago's
+		OS class environment using the Pintos kernel,
+		corresponding to contribution 2 above.
+	\item {\bf Transactions:} \Cref{chap:tm} presents my extension of Landslide's concurrency model to handle transactional concurrency and the evaluation thereof, corresponding to contribution 3 above.
+	\item {\bf Related Work:} \Cref{chap:relatedwork} honors my neighbours and ancestors in research spirit.
+	\item {\bf Future Work:} \Cref{chap:warpzone} identifies stateless model checking's remaining shortcomings and how new research might address them.
+	\item {\bf Conclusion:} \Cref{chap:conclusion} offers some thoughts on the future of the field.
+\end{itemize}
+
 \newpage

### 2018-12-10 thesis/overview.tex
polish background chapter; was pretty stale after being copied from the proposal

@@
 
-This chapter introduces the necessary background material on concurrency, stateless model checking, data-race analysis, and the relevant undergraduate operating systems classes.
+\revisionminor{This chapter introduces the necessary background material on concurrency
+(\cref{sec:overview-concurrency}),
+stateless model checking,
+(\cref{sec:overview-stateless}),
+data-race analysis
+(\cref{sec:background-datarace}),
+and the relevant undergraduate operating systems classes
+(\cref{sec:overview-edu}).
+It concludes with a glossary of terminology
+(\cref{sec:glossary})
+that I hope the reader may find convenient as a quick reference
+while reading later chapters as well.}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
 \section{Concurrency}
-
-\subsection{The Basics}
+\label{sec:overview-concurrency}
+
+\revision{This section presents an elementary introduction to
+%the fundamentals of
+concurrent programming.
+I have taken care to make it accessible to readers with even only basic programming experience,
+familiar with flow control, functions, memory, and objects at most,
+although I cannot avoid the inherent difficulty of multiple layers of abstraction building upon one another.
+For an explanation to be ``elementary''
+means that it requires very little be known ahead of time to approach it,
+not necessarily that it be easy
+\cite{lost-lecture}.
+%\footnote{``Elementary'' does not mean easy to understand.
+%``Elementary'' means that very little is required to know ahead of time in order to understand it,
+%except to have an infinite amount of intelligence
+%\cite{lost-lecture}.}
+}
+
+\subsection{The basics}
 
@@
 
-\subsection{Transactional Memory}
+\subsection{Transactional memory}
 \label{sec:overview-tm}
@@
 are represented as a directed acyclic graph (DAG) instead of as a tree.
-This approach is called \revisionminor{{\em stateful} model checking \cite{spin}}.
+This approach is called \revisionminor{{\em stateful} model checking.} %\cite{spin}.
 This thesis focuses on {\em stateless} model checking (and execution trees, not DAGs),
 which instead analyzes the sequence of execution events to avoid a prohibitive memory footprint.
-Henceforth ``stateless model checking'' will be abbreviated simply as ``model checking'' for brevity.
+Henceforth ``stateless model checking'' will often be abbreviated simply as ``model checking'' for brevity.
 Also, the term ``state space'' was originally coined to refer to the stateful approach's emphasis on the DAG's nodes
@@
 
-{\bf Abstractly speaking:}
-Let {\em independent transitions} denote a pair of executions of two threads, each from one preemption point to the next,
+%{\bf Abstractly speaking:}
+\revisionminor{To define it abstractly, let}
+{\em independent transitions} denote a pair of executions of two threads, each from one preemption point to the next,
 in which there are no read/write or write/write access pairs to the same memory between threads.
@@
 
-{\bf Concretely speaking:}
-\Cref{fig:dpor} highlights part of an execution tree where the execution ordering of threads 1 and 2 are swapped,
-and each interleaving has a respective ``subtree'' (i.e., possible interleavings given the fixed execution prefix leading up to it).
-The specifics of execution before the thread 1/thread 2 sequence,
-other possible threads to run instead of threads 1 or 2,
+%{\bf Concretely speaking:}
+\revisionminor{For a friendlier, more concrete example, consider \Cref{fig:dpor}, which}
+highlights part of an execution tree where the execution ordering of threads 1 and 2 are swapped,
+and each interleaving has a respective subtree (i.e., possible interleavings given the fixed execution prefix leading up to it).
+\revisionminor{Any events executed}
+before the thread 1/thread 2 sequence,
+any other runnable threads besides these two, %threads 1 or 2,
 and what logic the program executes in those subtrees
@@
 	\begin{center}
-	\includegraphics[width=0.6\textwidth]{dpor-new-colors.pdf}
+	\includegraphics[width=0.55\textwidth]{dpor-new-colors.pdf}
 	\end{center}
@@
 
-\section{Data Race Analysis}
+\section{Data race analysis}
 \label{sec:background-datarace}
@@
 \end{itemize}
-Because any combination of system calls or fault handlers can be invoked by user programs simultaneously,
+Because user programs can invoke any combination of system calls or fault handlers simultaneously,
 concurrency bugs can arise from the interaction of any subset of kernel components with each other.
@@
 
-The Pintos kernel architecture \cite{pintos} is used at several universities, including Berkeley, Stanford, and the University of Chicago.
-The Pintos basecode implements a rudimentary kernel, consisting of a context switcher, round-robin scheduler, locking primitives, and program loader,
+The Pintos kernel architecture \cite{pintos} is used at several universities, including the University of Chicago, Berkeley, and Stanford.
+Its basecode implements a rudimentary kernel, consisting of a context switcher, round-robin scheduler, locking primitives, and program loader,
 upon which students add more features in several projects.

### 2018-12-10 thesis/thesis.tex
thick changebars
ack joan karen deb and catherine; tighten up surrounding prose to trim requisite two lines

@@
 %\newcommand\revision[1]{#1\xspace}
-\newcommand\revision[1]{\cbstart{}#1\cbend\xspace}
+\newcommand\revision[1]{\cbstart[0.5em]{}#1\cbend\xspace}
 %\newcommand\revisionminor[1]{#1\xspace}
@@
 \begin{acknowledgments}
-\revision{I honor here the many people who supported me during these 7 years and made this journey possible.
+\revision{I honor here the many who %made this journey of 7 long years possible.
+supported me during these 7 long years and made this journey possible.
 
@@
 
-{Many others have collaborated with me directly on the research itself,
+{Many have collaborated with me directly on the research itself,
 whom I thank here in approximate chapter order.
-Michael J. Sullivan and Nathaniel Wesley Filardo were Landslide's first users besides myself back in its prototype days.
+Michael J. Sullivan and Nathaniel Wesley Filardo were Landslide's first other users
+%besides myself
+back in its prototype days.
 Joshua Wise, Miriam Zimmerman, and Glenn Willen helped revise my explanation of DPOR in \Cref{chap:landslide}.
-Professor Brandon Lucia provided valuable comments and revisions to \Cref{chap:quicksand}'s content
+Professor Brandon Lucia provided valuable revisions to \Cref{chap:quicksand}'s content
 that ultimately led to its publication at OOPSLA 2016.
@@
 graciously allowed me to use their students as guinea pigs
-and helped greatly to improve Landslide's stability and robustness for use beyond CMU's walls,
+and helped to improve Landslide's
+%stability and
+robustness for use beyond CMU's walls,
 greatly enriching \Cref{chap:education}'s evaluation.
 David Blum and Skye Toor provided
-invaluable advice on statistical analysis
-that helped me evaluate the student
+invaluable advice towards the statistical analysis
+used therein.
+%that helped me evaluate the student
+%grade distributions therein.
+Mario Dehesa-Azuara generously contributed the transactional
 % justify last line bc of manual page break
@@
 \noindent
-grade distributions therein.
-Mario Dehesa-Azuara generously contributed the
-transactional memory data structures
+memory data structures
 that make up about half of \Cref{chap:tm}'s evaluation suite.
@@
 and enlightening conversations.
+Joan Digney and Karen Lindenfelser
+make a lot of gears turn behind the scenes at PDL,
+and Deb Cavlovich and Catherine Copetas
+do likewise for CSD as a whole.
 %Thanks to the anonymous OOPSLA reviewer who went the extra mile to read my supplementary proofs
@@
 %(the lack of any stronger criticism a silent vote of confidence for their overall correctness).
-Thanks to Nicholas D. Matsakis for the mentorship and collaboration during my two summers as a Rust research intern at Mozilla,
+Thanks to Nicholas D. Matsakis for the mentorship %and collaboration
+during my two summers as a Rust research intern at Mozilla,
 and to the rest of the Rust community as well for their %extremely important
 work
-bringing better concurrent programs to the world from a different angle.
-I thank everyone outside my %immediate
-research circle who
+bringing better concurrent programs into the world. %from a different angle.
+And to everyone at large who
+%outside my %immediate
+%research circle who
 %whether at the retreat or otherwise,
-insisted how cool my research was for directly helping students --
-every such reminder helped me keep my head up during darker times.
+insisted how cool it was that my research helps students:
+thank you for helping me keep my head up during darker times.
 

### 2018-12-10 thesis/related.tex
rephrase some things in related to make citations not begin on a fresh line, as dave wishes

@@
 The problem of relaxed memory nondeterminism alone has inspired the creation of several new tools in the past few years.
-Relacy \cite{relacy}, a header-only C++ model checking library for checking synchronization primitives,
+Relacy \cite{relacy}, a header-only C++ model checking library for synchronization primitives,
 was the first to broach this field,
@@
 %(according to later citing papers) % cdschecker
-does not fully model all possible relaxed memory behaviours.
+does not fully model all memory reorderingss.
 %designed for verifying synchronization primitives,
@@
 beyond the shared-memory-threading model outlined in \cref{sec:landslide-dpor}.
-TransDPOR \cite{transdpor} provides extra domain-specific reduction for message-passing actor programs
+TransDPOR \cite{transdpor} provides additional domain-specific reduction for message-passing actor programs
 by exploiting the fact that the dependency relation is transitive in the absence of shared state.
-The $R^4$ algorithm \cite{r4} (corresponding to the R4 checker mentioned above)
+The $R^4$ algorithm \cite{r4}, used by the R4 checker mentioned above,
 extends DPOR to event-driven programs by separating the notion of enabled events from that of multiple threads.
@@
 by using hardware breakpoints and random sampling to achieve high performance.
-% TODO: read some more recent confs for data race shit since thesprop
-% probably like weak memory c.c
 
@@
 This suggests motivating students with Landslide's verification guarantee,
-% TODO fix this explanation its too roundabout
 although it is tricky to avoid accientally encouraging % oops
@@
 Transactional memory (TM), first introduced in 1993 \cite{transactional-memory},
-% TODO expand this to talk about some stm implementations?
-has received renewed attention in recent years with the announcement of Intel's Haswell architecture \cite{htm-haswell},
+has received renewed attention in recent years
+%with the announcement of
+since the launch of Intel's Haswell architecture \cite{htm-haswell},
 which supports hardware transactions (HTM) using new x86 instructions.
@@
 A subsequent paper \cite{applying-tm-bugs}
-found a majority of bugs in their study to be easily fixable with hand-written transactions,
+found a majority of bugs to be easily fixable with hand-written transactions,
 while others remained out of scope due to blocking {\tt cond\_wait()} operations and the like;
-more recently, the tool BugTM \cite{bugtm} aims to deploy such repairs in production code fully automatically.
+more recently, the tool BugTM~\cite{bugtm} aims to deploy such repairs in production code fully automatically.
 However, these studies all optimize for empirical correctness at best,
@@
 %
-Rust %\cite{rust-language}
-presents a type system with more explicit memory management,
+Rust's
+%\cite{rust-language}
+type system supports more explicit memory management,
 in-place mutation,
@@
 Analogous to stateless MC,
-which seeks good coverage of possible thread execution paths under schedule nondeterminism,
+which tests many possible thread execution paths under schedule nondeterminism,
 another popular testing approach is symbolic execution \cite{symbolic},
-which seeks good coverage of possible flow control paths under input nondeterminism.
+which tests many possible flow control paths under input nondeterminism.
 Symbolic executors

### 2018-12-10 thesis/education.tex
fix start-of-line citations and citation-as-noun in other chapaters

@@
 whether it offers pedagogical merit in the hands of students and/or TAs is a separate question.
-And while my M.S. thesis~\cite{landslide} showed that students
+And while I showed in my M.S. thesis \cite{landslide} that students
 could annotate P3 Pebbles kernels and thence use Landslide to debug them,

### 2018-12-10 thesis/overview.tex
fix start-of-line citations and citation-as-noun in other chapaters

@@
 \item
-\cite{hybriddatarace} introduces the {\em Limited Happens-Before} relation,
-which will report such potential races
+The {\em Limited Happens-Before} relation \cite{hybriddatarace}
+reports such potential races
 by considering only blocking operations like {\tt cond\_wait} to enforce the order.

### 2018-12-10 thesis/tm.tex
decaps some section names
fix start-of-line citations and citation-as-noun in other chapaters

@@
 
-\section{Concurrency Model}
+\section{Concurrency model}
 \label{sec:tm-design}
@@
 
-Intel's official TSX documentation summarizes its interface and behaviour in prose \cite{intel-tsx-overview}.
+Intel's official TSX documentation \cite{intel-tsx-overview}
+summarizes its interface and behaviour in prose.
 Recent work has used proof assistants to formalize some of the execution semantics
@@
 
-\subsection{Memory Access Analysis}
+\subsection{Memory access analysis}
 \label{sec:tm-design-locks}
@@
 
-\subsection{User Interface}
+\subsection{User interface}
 \label{sec:tm-interface}
@@
 it could find or verify the absence of such bugs.
-Visual inspection of \cite{tm-benchmark-cmu}'s HTM data structures found no barriers used in this implementation pattern;
+Visual inspection of the transactional AVL tree and hashmap
+\cite{tm-benchmark-cmu} found no barriers used in this implementation pattern;
 I would urge any reader interested in using those to add them in by hand first.

### 2018-12-10 thesis/intro.tex
explain the joke of landslide - pebbles, in intro

@@
 %for thread libraries and kernels that
-I have built over the last seven years, first debuting in my M.S. thesis~\cite{landslide}.
-Since then, I have extended it with a wide array of features and algorithms,
-some from prior work and some of my own devising,
+I have built over the last seven years, first debuting in my M.S. thesis~\cite{landslide},
+so named because it tests the stability of Pebbles programs students write in 15-410 at CMU.
+Since then, I have extended it with many features and algorithms,
+%some from prior work and some of my own devising,
 some which work behind the scenes and others which rely on human feedback to be effective,

### 2018-12-11 thesis/thesis.tex
ack nisei

@@
 I thank especially the other Stimhack moderators,
-the members of {\sf \#rainbow-coalition},
+%the members of
+{\sf \#rainbow-coalition},
+Nisei,
 and the local Pittsburgh crew
 for their support, encouragement,
-and mutual exchange of inclusive values in these turbulent yet hopeful times.
+and mutual exchange of inclusive ideals in these turbulent yet hopeful times.
 %and

### 2018-12-11 thesis/thesis.tex
revise the concl per daves wishes

@@
 the undergraduate operating systems student,
-as its target audience.
-Along the way, it justified its techniques' applicability to problems beyond CMU's walls as well
-by showing positive results at other universities
-and by addressing new concurrency models developed by industry in just the last decade.
+as its target audience,
+\revision{ultimately helping 73\% of participating students to find bugs in their code,
+56\% to fix any such bugs,
+and 34\% to fix all such bugs,
+before their project deadline.}
+Along the way, I justified \revisionminor{Landslide's} applicability to problems beyond CMU's walls as well
+by showing positive results at other universities,
+and by addressing a new concurrency model developed by \revisionminor{the hardware} industry in just the last decade.
 % also by comparing the preemption bounds of quicksand eval's bugs to prior work in ICB
@@
 has also been fulfilled.
+
+\revision{Let us revisit the first half of the thesis statement,
+which in \cref{sec:thestatement} I mentioned would serve as the work's overarching theme:
+that by
+{\em combining theoretically-founded automatic reduction techniques and user-informed heuristic ones,}
+stateless model checking can meet human testing needs.
+How have each of the major contributions of this work embodied this theme?
+%Let me also opine briefly on which of this work's contributions constitute lasting lessons
+%that any future research building hereupon should heed.
+Let me also briefly address:
+What lasting lessons have we learned from each,
+which any future research building hereupon would be wise to heed?
+
+\begin{itemize}
+	\item In \Cref{chap:quicksand},
+		the soundness of data-race-driven Iterative Deepening
+		combined with Quicksand's reliance on state space estimation
+		to effectively balance formal verification with heuristic bug-finding speed
+		better than any prior work.
+		% 未来は、
+		Even as modern programming languages such as Rust \cite{rust-book}
+		threaten to obsolete the data-race preemption point proof
+		by statically forbidding data races to begin with,
+		Iterative Deepening's general approach remains relevant alongside ICB \cite{chess-icb},
+		each providing a different tradeoff between bug-finding and completion time,
+		and future model checkers should offer a mixture of different kinds of partial verification results
+		depending on the user's wishes.
+	\item In \Cref{chap:education},
+		the behind-the-scenes yield-loop blocking algorithm
+		(itself admittedly heuristic)
+		allowed Landslide to automatically instrument all student implementations
+		with very few false positives,
+		and exposing partial results such as data race reports
+		was shown to sometimes help students reach deeper understanding.
+		% 未来は、
+		Although high-level undergraduate operating systems courses represent
+		a very small fraction of all the world's programmers,
+		let alone all those in higher education,
+		Landslide's adventures in 15-410 argue that computer science classes
+		should more intentionally teach principled debugging techniques and tools
+		rather than let students struggle to reach correctness independently. %on their own.
+		Also, any future debugging tool should heed my students' survey feedback
+		and take friendly user interface design more seriously than I did with Landslide.
+	\item In \Cref{chap:tm},
+		the equivalence of HTM's interface to failure injection and global locks
+		allowed Landslide to avoid both performance overhead and state space explosion when simulating aborts,
+		and the heuristic use of abort code reduction and abstraction reduction
+		allowed verifying algorithms and data structures up to thread/iteration counts previously infeasible.
+		% retry set reduction, obv; it also applies to failure injection as well
+		% even when synch is complicated, defining abstraction boundaries and verifying your shit separately
+		% is ,,good
+		% - even if you have to be careful that you're "proving the right thing"
+		% 未来は、
+		Emulating complex execution semantics with simpler, already-understood primitives,
+		and the retry set reduction,
+		both allowing substantial yet sound state space reduction,
+		are the obvious take-aways for any future model checker wishing to tackle HTM.
+		I expect the latter should easily apply to other opportunities for failure-injection beyond transactions as well,
+		such as memory allocation, filesystems, and distributed systems.
+		More in general,
+		the verification section demonstrated the value of compartmentalizing one's verification efforts,
+		checking modules separately against a common, well-understood interface
+		to reach yet further up the exponential cliff.
+\end{itemize}
+}
 
@@
 % - manual inspection of data races
+\revision{Here I have demonstrated that stateless model checking can realistically fulfill this ideal,
+and I envision a future of concurrency verification where % when
+human and program may cooperate to debug and verify even more complex software than ever before.}
 
@@
 understanding and respecting their limitations just as they complement ours.
-This thesis has been a profoundly meaningful journey for me and I hope you enjoyed reading it.
+%This thesis has been a profoundly meaningful journey for me and I hope you enjoyed reading it.
 

### 2018-12-11 thesis/thesis.tex
slight revise concl

@@
 and I envision a future of concurrency verification where % when
-human and program may cooperate to debug and verify even more complex software than ever before.}
+human and program may cooperate to debug and verify more complex software than ever before.}
 

### 2018-12-12 thesis/quicksand.tex
fill in lots of eval stuff, start filling in verif table, discuss abstraction reduction
revise tm, cleanup todos in bgnd/landslide, fix cross-chapter section references, etc -- sending tm chapter off to committee here
write a ton of relwork
more future - verification etc, some more related works too
future - generalize from k,n; combine test case and verif sects
use cleveref
write quicksand design, beginnings of proof stuff

@@
+% TODO: scan for abbrs like PP, MC, SSS-MC
 \chapter{Quicksand}
@@
 \inspirationalquote{
-% original - TODO, put in
+% original - on second thought, i'm not actually gonna put this in
+% the nuance in japanese is more desperate, more disorganized, and less generally heroic-seeming
+% i think the english translation actually works better for an inspirational quote
 %{\footnotesize 嫌な事も悲しい事もあったけど、守りたい物だってたくさんこの世界にはあったから。} \\
@@
 %{Kaname Madoka, Mahou Shoujo Madoka{\raisebox{0.1em}{$\scriptstyle \bigstar$}}Magica}
+% TODO: make the star italic
+% TODO: see https://tex.stackexchange.com/questions/63179/shear-transform-a-box/63188#63188
 {Kaname Madoka, Puella Magi Madoka{\raisebox{0.1em}{$\scriptstyle \bigstar$}}Magica}
@@
 
-% TODO
+% TODO: intro
 
-\section{Preemption points} % ??? -- label needed from reference to LAST_CALL in landslide.tex
+% TODO: mention that all source files in this chapter are relative to id/ subdir in the repo
+
+The contributions of this chapter were published as
+{\em Stateless Model Checking with Data-Race Preemption Points}
+in OOPSLA 2016 \cite{quicksand}.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{Preemption points}
 \label{sec:quicksand-pps}
+
+% reference figure (nb color scheme tree)
+% its pps were drawn on every instruction
+% for larger test cases, this is not practical
+% consider the example program of so and so (show mx/yield prog)
+% here is a resulting state space
+% this needs a DR PP to FAB
+% qs example (take from thesis slides)
+
+% TODO make sure to define maximal and mminnimal
+
+% TODO: define "data race" (pure hb) vs "data race candidate" (limited); also do scane of this chapater for conisstencey
+
+In related work,
+Portend \cite{portend} has already suggested combining data race analysis with preemption-driven artificial scheduling,
+although it obtains its data-race candidates from a stand-alone, single-pass analysis.
+In order to make sure we identify every one of a program's data races that could possibly arise under the given test case,
+we must check many different interleavings to begin with,
+perform the Portend approach for every data race we find,
+which may in turn uncover more data races
+(hidden in flow control paths reachable only through interleavings of the first race, perhaps),
+and then continue model checking the multiple data races together
+in a sort of bidirectional feedback loop between the two algorithms.
+I will refer to data races that require model checking to expose to begin with as
+{\em nondeterministic data races}. % TODO: rename, this name is awful
+In the next section,
+\cref{sec:quicksand-id} will show how I achieve this in Quicksand, and
+\cref{sec:quicksand-soundness} will justify the technique's formal verification power.
+In the evaluation (\cref{sec:quicksand-eval}),
+\cref{sec:quicksand-eval-nondets} in particular will justify the need
+for this feedback loop between model checking and data race analysis
+by showing that many bugs require nondeterministic data races to expose.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{Iterative Deepening}
+\label{sec:quicksand-id}
+
+To address the problem of choosing meaningful preemption points,
+I have developed an algorithm called {\em Iterative Deepening},
+implemented in a wrapper program specific to Landslide called {\em Quicksand}.
+Named after the analogous technique in chess artificial intelligence \cite{iterative-deepening-chess-ai},
+Iterative Deepening
+is a search strategy,
+for exponentially-sized state spaces in general,
+which
+makes progressively deeper searches of the state space until the CPU budget is exhausted.
+In this context, the depth roughly corresponds to the subset of preemption points used.
+Hence, Quicksand
+schedules multiple Landslide instances in parallel to
+test many different subsets of the available preemption points,
+
+For the remainder of the chapter, I will use Iterative Deepening to refer to the algorithm in the abstract,
+which could in principle apply to any stateless model checking domain,
+and Quicksand to refer to the specific implementation,
+which relies on data race analysis and specific heuristics to optimize its testing approach
+for kernels and thread libraries.
+I will also henceforth refer to each unique set of preemption points as a {\em job}.
+
+%\subsection{Design}
+
+% TODO: put fig:id here, explain generally multiple state spaces
+
+Note that Iterative Deepening is a {\em wrapper} algorithm around stateless MC.
+A model checker is still used to test each state space, and other reduction techniques such as DPOR
+are still applicable in each.
+Moreover, because Iterative Deepening treats the set of preemption points as mutable,
+it can add new preemption points reactively based on any runtime analysis.
+Here
+will focus on run-time data race analysis~\cite{tsan,fasttrack} as the mechanism for finding new preemption candidates.
+The next section (\cref{sec:quicksand-soundness})
+will prove that in fact,
+in addition to statically-known synchronization preemption points,
+this suffices to provide at least as strong verification guarantees as any other possible preemption point set.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\subsection{Changing state spaces}
+
+To introduce the Iterative Deepening algorithm,
+I will first show a simple approach for handling new PPs in the absence of any CPU budget restriction.
+
+Given unlimited CPU time for testing,
+one would always want to switch to the new maximal state space whenever adding a new preemption point.
+The maximal state space is guaranteed to subsume all execution sequences reachable in any subset state space,
+so considering any incomplete subset of the known PPs would be duplicate work.
+\Cref{alg:algorithm0} demonstrates this na\"ive approach.
+It is seeded with the set of all statically-known synchronization API PPs,
+and invoked whenever a new data-race candidate is found.
+%
+The upcoming proofs in \cref{sec:quicksand-soundness},
+being concerned with the verification guarantee provided when the search may complete within the CPU budget,
+are based on this simple version of Iterative Deepening.
+The user may also wish to configure her testing tool to prefer this approach, at her discretion,
+such as when she believes all bugs have been fixed and wants a verification as fast as possible;
+% TODO: clarify sec reference to -M flag
+\cref{sec:quicksand-implementation} discusses this execution mode further.
+
+\newcommand\AllPPs{\ensuremath{\mathcal{A}}}
+\newcommand\PendingJobs{\ensuremath{\mathcal{P}}}
+\newcommand\SuspendedJobs{\ensuremath{\mathcal{S}}}
+\newcommand\GetETA[1]{ETA(#1)}
+\newcommand\GetPPSet[1]{PPSet(#1)}
+
+\begin{algorithm}[h]
+	\SetKwInOut{Input}{Input}
+	\Input{$j$, the currently-running job}
+	\Input{\AllPPs, the set of all known preemption points} %, sorted by decreasing heuristic priority}
+	\eIf{$\exists p \in \AllPPs . p \not\in \GetPPSet{j}$}{
+		return NewJob(\AllPPs) // New maximal state space
+	}{
+		return $j$ // $j$ is still maximal
+	}
+	\caption{Na\"ive Iterative Deepening method}
+	\label{alg:algorithm0}
+\end{algorithm}
+
+However, in many tests of even modestly-sized programs,
+full verification is not feasible,
+and focusing on the maximal state space alone is likely to be fruitless.
+%
+Hence, Iterative Deepening also allows for prioritizing subset jobs
+based on number of preemption points, ETA, and whether data race candidates are included among their preemption points.
+It relies on state-space estimation \cite{estimation}
+to predict which jobs are likely to complete within a reasonable time,
+before actually testing a large fraction of interleavings for each.
+The overall goal is to decide automatically when to defer testing a state space,
+so an inexpert user can provide only her total CPU budget as a test parameter,
+and to enable completing appropriately-sized jobs within that budget.
+Quicksand seeks to maximize completed state spaces,
+as each one serves as a guarantee that all interleavings possible with its PPs were tested;
+\cref{sec:quicksand-discussion} discusses some limitations of this approach.
+The next three subsections will show how to schedule these smaller jobs
+based on their preemption points and ETAs.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\subsection{Initial preemption points}
+
+Iterative Deepening must be seeded with a set of initial state spaces,
+which can be any number of subsets of the statically-available preemption points
+that prior work model checkers would use.
+The upcoming soundness proof relies on the maximal state space being included among these for verification's sake,
+but to optimize for finding bugs faster,
+implementations may wish to simultaneously to try testing subsets thereof.
+
+For testing user-space code, Quicksand begins with the four jobs from \Cref{fig:id}:
+$\{yield\}$,
+$\{yield,lock\}$,
+$\{yield,unlock\}$,
+and $\{yield,lock,unlock\}$,
+By extension, these also induce preemptions on any other primitives
+which use
+internal locks,
+such as condition variables or semaphores.
+Preempting on voluntary switches such as {\tt yield} is always necessary to maintain
+Landslide's invariant that only one thread runs between consecutive preemption points,
+so the {\tt yield} preemption point is always implicitly enabled.
+
+For kernel-level testing, interrupt-disabling is analogous to locking,
+so preemptions must also be introduced
+just before a disable-interrupt opcode (on x86, {\tt cli})
+and just after interrupts are re-enabled (on x86, {\tt sti}).
+During data-race detection, {\tt cli} and {\tt sti} are treated as a single global lock
+(note that {\tt cli}'d memory accesses can still race with others that have interrupts on).%
+\footnote{Some kernels disable preemption without disabling interrupts,
+which can be communicated to the MC using manual annotations,
+and must be treated similarly.
+This also assumes uni-processor scheduling; for SMP kernels, replace {\tt cli}/{\tt sti} with spinlocks.}
+Quicksand is configured to begin with
+$\{yield\}$,
+$\{yield,lock\}$,
+$\{yield,unlock\}$,
+$\{yield,cli\}$,
+$\{yield,sti\}$,
+and $\{yield,lock,$ $unlock,cli,sti\}$.
+As a heuristic, it doesn't test every intermediate subset such as $\{lock,sti\}$,
+which would result in $2^p$ jobs right off the bat,
+although this could potentially be improved in future work (\cref{sec:quicksand-discussion}).
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\subsection{Data-race preemption points}
+
+As discussed in \cref{sec:quicksand-pps},
+%runtime data race detection (\cref{sec:background-datarace})
+%may find candidate unsynchronized memory conflicts that should be investigated further.
+data races may beget new interleavings not reachable by preempting on synchronization API boundaries alone.
+Because each data race indicates an access pair that can interleave at instruction granularity,
+different program behaviour may arise if the threads are preempted just before the racing instructions,
+some of which the programmer may not have even expected, i.e., be bugs,
+and it is logical to apply model checking to find or verify the absence of such bugs.
+%it is logical to re-execute the test and issue preemptions just before those instructions
+%to test alternate thread interleavings.
+
+With Iterative Deepening,
+this is a simple matter of creating a new state space
+with an additional preemption point enabled on the racing instructions by each thread,
+as shown in \Cref{alg:handledatarace}.
+I will term these {\em data-race preemption points},
+and they will form the foundation of Quicksand's contribution.
+Note that even though a data race may involve two different instructions,
+$\alpha$ and $\beta$, Quicksand's strategy is to add new state spaces with only one new racing instruction at a time.
+Rather than adding a single large state space,
+configured to preempt on both involved instructions,
+i.e., $AB =$ \GetPPSet{$j_0$} $\cup$ $\alpha$ $\cup$ $\beta$,
+it prefers to add multiple smaller jobs which have a higher chance of completing in time, i.e.,
+$A =$ \GetPPSet{$j_0$} $\cup$ $\alpha$ and
+$B =$ \GetPPSet{$j_0$} $\cup$ $\beta$.
+If $A$ and $B$ are bug-free, they will in turn add $AB$ later during their own execution.
+The condition on line~1 ensures that we avoid duplicating any state spaces with multiple data-race preemption points;
+for example, $AB$ is reachable by multiple paths through its different subsets $A$ and $B$,
+but should of course be tested only once.
+
+\newcommand\AllJobs{\ensuremath{\mathcal{J}}}
+\begin{algorithm}[t]
+	\SetKwInOut{Input}{Input}
+	\Input{$j_0$, the currently-running job}
+	\Input{\AllJobs, the set of all existing (or completed) jobs}
+	\Input{$\alpha$, an instruction reported by the MC as part of a racing access pair}
+	\If{$\forall j \in \AllJobs,$
+	\GetPPSet{$j_0$} $\cup$ $\alpha$
+	$\not\subseteq$
+	\GetPPSet{$j$}
+	}{
+		AddNewJob(\GetPPSet{$j_0$} $\cup$ $\alpha$, HeuristicPriority($\alpha$)) \\
+	}
+	\If{$\forall j \in \AllJobs,$ \GetPPSet{$j$} $\neq \{yield, \alpha\}$}{
+		AddNewJob($\{yield, \alpha\}$, HeuristicPriority($\alpha$))
+	}
+	\caption{Adding new jobs with data-race PPs.}
+	\label{alg:handledatarace}
+\end{algorithm}
+
+Furthermore,
+Iterative Deepening allows not always strictly increasing the number of preemption points
+whenever a new data race is identified.
+For each instruction involved in a data race, Quicksand adds two new jobs:
+a ``small'' job to preempt on that instruction only (line~5),
+and a ``big'' job to preempt on that instruction as well as each preemption point used by the reporting job (line~2).
+%
+Hence,
+each {\em pair} of racing accesses will spawn four new jobs.
+\Cref{fig:new-dr-jobs} visualizes the resulting overall workflow in Quicksand,
+including the four such jobs resulting from one data race report.%
+\footnote{As an optimization,
+though the big jobs should be expected to uncover more data races and in turn
+produce even bigger jobs still,
+small jobs should be forbidden from ``reproducing'',
+as their purpose is only fast heuristic bug-finding rather than exhaustive coverage;
+see {\tt handle\_data\_race()} in {\tt messaging.c}.}
+%
+The rationale of spawning multiple jobs is that one cannot know in advance which will be most fruitful:
+while the big job risks not completing in time,
+the small job risks missing the data race entirely if the original PPs were required to expose it.
+In practice, I have observed many bugs found quickly by these small jobs,
+and many other bugs missed by the small jobs found eventually by the big jobs.
+This phenomenon motivates Iterative Deepening to prioritize jobs at run-time.
+
+\begin{figure}[t]
+        %\includegraphics[width=0.48\textwidth]{dr-jobs-v2.pdf}
+        asdfsaf
+        % TODO: put the figure form yr snake fight
+	\caption[Quicksand incorporates data race reports as new preemption points at run-time.]
+		{Quicksand incorporates data race reports as new preemption points at run-time,
+		by managing the exploration of multiple state spaces,
+		communicating with each MC instance to receive ETAs, data race candidates, and bug reports.
+                %When an access pair is reported as a data race candidate,
+                When a data race is reported,
+                a new preemption point is added for each involved access,
+                and new jobs are added for later testing,
+                corresponding to different combinations of those with the existing preemption points.}
+        \label{fig:new-dr-jobs}
+\end{figure}
+
+The new state spaces may expose a failure, in which case Iterative Deepening must stop and report a data-race bug,
+or complete successfully, indicating a {\em benign} (i.e., false-positive) data race.
+They may also uncover a new data-race candidate entirely in some alternate interleaving,
+in which case we may iteratively advance to a superset state space which will preempt at both racing access pairs.
+Being constrained by a CPU budget,
+Iterative Deepening may time out before completing a data race's associated state space,
+in which case the data race remains neither confirmed nor refuted.
+%Depending on how much burden the implementation wants to impose on the user,
+%it may then report it as a
+%report a potential false positive that the user must handle
+In such cases, Quicksand elects to impose some burden on the user
+by reporting it as a potential false positive
+and recommend that she investigate it by hand to judge for herself whether it be a bug.
+% TODO: fix this to point to partial verif
+% TODO (also) make sure that all other quicksand-discussion references get updated if appropriate
+\cref{sec:quicksand-discussion} will discuss future opportunities for improving
+debugging output in cases of such {\em partial verification}.
+However, experience shows that this interactivity pays off:
+in the next chapter's educational user study (\cref{sec:education-eval}),
+one student reported during the survey that they used this recommendation,
+combined with their own intuition,
+to find a bug that Quicksand was not able to find alone (\cref{sec:education-reasons-worthwhile}).
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\subsection{Choosing the best job}
+
+With a limited CPU budget, many larger tests are likely to be fail to complete in time,
+and smaller tests are likely to be more fruitful at finding bugs quickly.
+A model checker's state space estimation (\cref{sec:landslide-estimate})
+can provide a hint to select between these jobs.
+How to handle jobs whose ETAs are too high for the given CPU budget
+is the heart of Iterative Deepening,
+and is listed formally in \Cref{alg:shouldworkblock}.%
+\footnote{
+Though its worst-case performance is $O(mn)$ in the
+%number of pending and suspended jobs,
+sizes of $\mathcal{P}$ and $\mathcal{S}$,
+in practice the non-constant portion beyond line~4 runs very infrequently
+and is negligible compared to the exponentially-sized state spaces.}.
+
+\begin{algorithm}[t]
+	\SetKwInOut{Input}{Input}
+	%\textbf{Function} GetBestJob($j_0$, PendingJobs, SuspendedJobs): \\
+	\Input{$j_0$, the currently-running job}
+	\Input{\PendingJobs, the list of pending jobs, sorted by decreasing heuristic priority}
+	\Input{\SuspendedJobs, the list of already-suspended jobs, sorted by increasing ETA}
+	\Input{$T$, the remaining time in the CPU budget}
+	\If{\GetETA{$j_0$} $<$ HeuristicETAFactor $\times$ $T$}{
+		return $j_0$ // Common case: job is expected to finish.
+	}
+	\ForEach{job $j_P \in$ \PendingJobs}{
+		// Don't run a pending job if a subset of it is already suspended; its ETA would be at least as bad. \\
+		\If {$\forall j_S \in$ \SuspendedJobs, \GetPPSet{$j_S$} $\not\subset$ \GetPPSet{$j_P$}}{
+			return $j_P$
+		}
+	}
+	%// no pending jobs; maybe resume a suspended job \\
+	\ForEach{job $j_S \in$ \SuspendedJobs}{
+		\If{\GetPPSet{$j_0$} $\not\subset$ \GetPPSet{$j_S$}
+			$\land$
+			\GetETA{$j_0$} $>$ \GetETA{$j_S$}}{
+			// If a subset of $j_S$ is also suspended, don't run the larger one first. \\
+			\If{$\forall j_{S2} \in$ \SuspendedJobs, \GetPPSet{$j_{S2}$} $\not\subset$ \GetPPSet{$j_S$}}{
+				return $j_S$
+			}
+		}
+	}
+	return $j_0$ // \GetETA{$j_0$} was bad, but no other $j$ was better.
+	\caption{Suspending exploration of a state space in favor of a potentially smaller one.}
+	\label{alg:shouldworkblock}
+\end{algorithm}
+
+Its main feature is understanding that if \GetPPSet{$j_1$} $\subset$ \GetPPSet{$j_2$},
+and $j_1$ is suspended,
+then $j_2$'s state space is guaranteed to be strictly larger, so $j_2$ will take at least as long.
+Hence, as long as $j_1$ is suspended on account of being too big,
+$j_2$ should not be tested either,
+unless $j_1$ is later resumed and its ETA improves over time after further execution.
+%reveals that it might finish in time after all.
+Similarly, whenever a job finds a bug, all pending superset jobs may safely be cancelled,
+as they are guaranteed to contain the same program behaviour, and likely to simply find the same bug again.
+%
+Implementation-wise,
+Quicksand receives an updated estimate from each Landslide instance whenever it finishes executing a new interleaving,
+and separates them accordingly
+into a set of {\em suspended} jobs,
+i.e., partially-explored state spaces with high ETAs,
+and a set of {\em pending} jobs,
+i.e., untested ones with unknown ETAs.
+When Landslide reports an ETA too high for some job,
+it is compared with other pending and suspended jobs to find another one more likely to complete in time.%
+\footnote{Note that when Quicksand is configured to use multiple CPUs,
+simultaneously-running jobs are not considered among the set of possible jbos to switch to,
+so if there are fewer total jobs with ETA lower than the time budget than the allowed parallelism factor,
+some CPUs may end up speculatively running large jobs
+in hopes that the ETA turns out to be an overestimate.}
+
+Iterative Deepening also accounts for the inherent inaccuracy of ETA estimates.
+Line~1 heuristically scales up the time remaining to avoid suspending jobs too aggressively
+in case their ETAs are actually overestimated.
+Lines~12-15 account for the
+%bizarre
+possibility that among two suspended jobs,
+%given two jobs,
+%%$j_1,j_2$,
+\GetPPSet{$j_1$} $\subset$ \GetPPSet{$j_2$}
+but
+\GetETA{$j_1$} $>$ \GetETA{$j_2$}.
+This may seem surprising,
+but can often arise because estimates tend to get more accurate over time,
+and $j_1$ perhaps ran much longer, on account of being overall smaller,
+before becoming suspended.
+In such scenarios,
+the algorithm heuristically assumes the smaller job's ETA is more accurate,
+in order to avoid repeatedly resuming larger jobs briefly only to find that their ETAs keep getting worse and worse
+(in other words, it lets us avoid thrashing in Quicksand). % ¯\_(ツ)_/¯
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\subsection{Heuristics}
+
+As predicting the ETAs of state spaces of unknown size
+and using that plus size of a set of preemption points as a proxy for how likely a job is to find bugs or complete
+is a fundamentally messy process,
+it is appropriate to equip the algorithm with some heuristics informed by experience.
+\Cref{alg:shouldworkblock} allows the option to heuristically scale a job's ETA
+when comparing it to the overall time budget,
+which can compensate for any inaccuracy by the estimator.
+Quicksand uses a scaling factor defaulting to 2,
+chosen based on experiments from prior work \cite{estimation}.
+%though we allow changing it via the command line.
+It also includes a heuristic to
+never suspend jobs before they pass a certain threshold of interleavings tested,
+with a default of 32,
+informed by my personal experience that ETAs require around that much progress into the state space
+before they stabilize (at least relative to each other on similar state spaces,
+not necessarily relative to the ultimate true size).%
+\footnote{These two heuristics are configurable with the
+{\tt -e} and {\tt -E} command-line options, respectively,
+as discussed in \cref{sec:landslide-quicksand-options}.}
+
+Landslide classifies data-race candidates as {\em both-order} or {\em single-order},
+as defined in prior work \cite{portend},
+based on whether it observed the racing instructions ordered in both possible sequences or only one
+in the original state space, respectively.
+Single-order candidates are more likely to be false positives (\cref{sec:background-datarace}),
+although preempting during the access itself is necessary to say for sure.
+Hence, Quicskand add preemption points for both types of candidates,
+and heuristically prioritizes jobs with both-order data races
+over those with only single-order data races.
+The HeuristicPriority($\alpha$) call in \Cref{alg:handledatarace} corresponds to this strategy.
+For single-order races, Quicksand does not initially add a PP for the later access at all:
+if preempting on the first access is capable of reordering the race,
+it will be updated to both-order in the new state space, and the second preemption point will be added then.
+\cref{sec:warpzone-heuristics} will discuss opportunities for future work to expand
+these heuristics with more nuanced search strategies still.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\subsection{Reallocation false positives}
+\label{sec:quicksand-id-realloc}
+
+Finally, I identified a particular class of false positive data-race candidates
+under the Limited Happens-Before analysis (\cref{sec:background-datarace})
+in which the associated memory was recycled by re-allocation between the two accesses,
+and claim that it is safe to completely disregard them when considering where to add new preemption points.
+Figure~\ref{fig:recycle} shows a common code pattern and interleaving which can expose such behaviour.
+If the {\tt malloc} on line~4 returns the same address passed to {\tt free} on line~2,
+then lines~1 and 7 will be flagged as a potential data race.
+I term this a {\em reallocation false-positive data-race candidiate}.
+To the human eye, this is obviously a false positive:
+reordering lines~4-7 before lines~1-2 will cause {\tt malloc} to return a different region of allocated memory,
+in turn causing {\tt x} and {\tt y} to no longer collide.
+In studying a similar pattern, the Eraser tool from prior work \cite{eraser}
+found that Thread 2's logic usually corresponds to an initialization pattern,
+but for generality I have added an arbitrary {\tt publish} action to the example on line~6.
+
+% TODO: fix syntax hilight
+% TODO check figure placement
+% TODO check figure camption
+\begin{figure}[t]
+	\begin{center}
+	\begin{tabular}{rll}
+		& \multicolumn{2}{c}{\texttt{struct x \{ int foo; int baz; \} *x;}} \\
+		& \multicolumn{2}{c}{\texttt{struct y \{ int bar; \} *y;~~~~~~~~~~}} \\
+		\\
+		& {\bf Thread 1} & {\bf Thread 2} \\
+		1 & \texttt{\hilight{brickred}{x->foo = ...;}} & \\
+		2 & \texttt{\hilight{olivegreen}{free}(x);} \\
+		3 & & \texttt{\hilight{commentblue}{// x's memory reallocated}} \\
+		4 & & \texttt{y~=~\hilight{olivegreen}{malloc}(sizeof *y);} \\
+		5 & & \texttt{\hilight{commentblue}{// ...initialize...}}\\
+		6 & & \texttt{publish(y);} \\
+		7 & & \texttt{\hilight{brickred}{y->bar = ...;}} \\
+	\end{tabular}
+	\end{center}
+	\caption{A common execution pattern with {\tt malloc()} that produces false positive data race candidates.}
+	\label{fig:recycle}
+\end{figure}
+
+As long as the allocation heap is properly synchronized,
+a Pure Happens-Before analysis should identify a happens-before edge
+between line~2's {\tt free} and line~4's {\tt malloc},
+and it would report no race.
+However, the upcoming evaluation will show that Limited Happens-Before retains some advantages over Pure
+(\cref{sec:quicksand-eval-bugs}),
+so it is useful to be able to automatically suppress data-race candidates
+that are certain to end up being false positives when reordered.
+
+The ability to disregard reallocation false positives is unique to Iterative Deepening.
+When limited to a single test execution, suppressing any data race candidate matching this pattern is unsound.
+Consider the more unusual program in Figure~\ref{fig:recycle-bug},
+in which the memory is recycled the same way, but the racing access's address is not tied to {\tt malloc}'s return value.
+Here, reordering lines~6-7 before line~3 will allow {\tt x} and {\tt x2} to race.
+Such collisions could instead be avoided with a hacked allocator which never recycles memory,
+but this could unacceptably impact performance in {\tt malloc}-heavy tests.
+Moreover,
+in the context of Iterative Deepening, being able to discard certain data race candidates
+allows Quicksand to skip exploring some entire state spaces,
+and hence run fewer Landslides overall;
+this is analogous to DPOR's ability to skip equivalent interleavings within a single Landslide instance.
+Upcoming in the evaluation, \cref{sec:quicksand-eval-verif}'s \Cref{tab:drstatistics}
+will show how many redundant state spaces Quicksand is able to prune with this technique.
+
+% TODO smae as above
+\begin{figure}[t]
+	\begin{tabular}{rll}
+		& {\bf Thread 1} & {\bf Thread 2} \\
+		1 & \texttt{publish(x);} & \\
+		2 & \texttt{\hilight{brickred}{x->foo = ...;}} & \\
+		3 & \texttt{\hilight{olivegreen}{free}(x);} \\
+		4 & & \texttt{x2 = get\_published\_x();} \\
+		5 & & \texttt{\hilight{commentblue}{// x's memory recycled}} \\
+		6 & & \texttt{y~=~\hilight{olivegreen}{malloc}(sizeof *y);} \\
+		7 & & \texttt{\hilight{brickred}{x2->foo = ...;}} \\
+	\end{tabular}
+	\caption{If a single-pass \revision{Limited HB analysis}~discarded candidates matching the malloc-recycle pattern,
+it would miss the bug in this adversarial program.}
+	\label{fig:recycle-bug}
+\end{figure}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{Soundness}
+\label{sec:quicksand-soundness}
+
+Adding new data-race preemption points in a feedback loop can uncover bugs
+not previously reachable by preempting on synchronization APIs alone,
+as some prior model checkers do \cite{dbug-ssv},
+but how does it compare to the other extreme end of the trade-off,
+that is,
+committing in advance to preempt on every single shared memory access \cite{inspect}?
+It turns out,
+assuming sufficient CPU budget,
+Iterative Deepening can in principle expose every possible program behaviour that
+even that latter approach can find,
+providing an equally strong verification guarantee.
+This section presents a proof of this claim (\cref{sec:quicksand-convergence}),
+as well as a supplementary proof (\cref{sec:quicksand-realloc})
+of the soundness of pruning reallocation false positives discussed previously (\cref{sec:quicksand-id-realloc}).
+
+I am presenting these proofs as they appeared in the OOPSLA paper \cite{quicksand}:
+written as sketches in informal prose,
+to optimize for rapidly conveying an intuition for why it works
+rather than to justify every internal step within the proof structure.%
+\footnote{Also because I am already over 200 pages in this thesis, all told.}
+% footnote yeah, i'm writing this last
+% footnote of footnote yeah, the committee never read this part
+A more rigorous treatment
+is available in the tech report which accompanied the conference paper \cite{quicksand-soundness}.
+
+{\bf Assumptions.}
+The proofs are built on a DPOR definition which assumes sequentially-consistent memory hardware.
+All algorithms involved are assumed to operate on a machine model of a single globally-consistent execution trace,
+which fundamentally cannot account for memory reordering nondeterminism.
+\cref{sec:quicksand-discussion} discusses this limitation further;
+for existing work on combining DPOR with relaxed memory, I refer the reader to \cite{tsopso}.
+They also assume the Limited Happens-Before definition
+for the data-race analysis.
+I leave the case for Pure Happens-Before to future work,
+although if I may appeal to intuition,
+it requires only to show that for any data-race candidate
+Limited Happens-Before reports in a given execution,
+that Pure Happens-Before does not,
+either it will be a false positive,
+or the latter will find it in an alternate execution within the same state space,
+or the latter will find a different data race that ultimately leads to a bigger state space in which the first one may be found.
+
+\subsection{Convergence}
+\label{sec:quicksand-convergence}
+
+\subsubsection{Equivalence}
+
+% TODO: decide if this shouldn't be reordered befor econvergence?
+\subsubsection{Saturation}
+
+\subsection{Reallocation false positives}
+\label{sec:quicksand-realloc}
+
+% todo rename "malloc recycle" to "reallocation false positives"
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{Implementation}
+\label{sec:quicksand-implementation}
+
+% TODO: add -M flag discussion
+
+% TODO: cant swap x100
+
+% TODO: put a code snippet of what ap rogress report looks like (take from some htm log)
+
+% this section developers guide to quicksand
+
+% job
+% messaging
+% work
+% option
+% pp
+
+% cool tricks
+% cant swap x100 % cite [Inf09] ?
+
+% how is free re malloc implemented in landslide?
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 \label{sec:quicksand-eval}
+
+\subsection{Finding bugs}
+\label{sec:quicksand-eval-bugs}
+
+\subsection{Verification}
+\label{sec:quicksand-eval-verif}
+
+\subsection{Nondeterministic data races}
+\label{sec:quicksand-eval-nondets}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{Discussion}
+\label{sec:quicksand-discussion}
+
+% identifying & skipping redundancies in smaller ID state spaces via memoization (same applies to ICB!)
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{Summary}
+
+quicksand is gr9

### 2018-12-12 thesis/quicksand.tex
copy in iterative deepening proof

@@
 % TODO: define "data race" (pure hb) vs "data race candidate" (limited); also do scane of this chapater for conisstencey
+
+% TODO: use this somehow
+Committing in advance to
+preempt on every instruction is certain to expose any possible bug under the given test case,
+but invites massive state space explosion.
+Even as DPOR helps to skip equivalent interleavings of non-conflicting transitions,
+DPOR itself is $O(n^2)$ in the number of preemption points in a single execution,
+which is not compatible with such an approach.
+Accordingly, stateless model checkers must find more efficient ways to be able to uncover bugs such as these.
 
@@
 To introduce the Iterative Deepening algorithm,
-I will first show a simple approach for handling new PPs in the absence of any CPU budget restriction.
+I will first show a simple approach for handling new preemption points in the absence of any CPU budget restriction.
 
@@
 The maximal state space is guaranteed to subsume all execution sequences reachable in any subset state space,
-so considering any incomplete subset of the known PPs would be duplicate work.
+so considering any incomplete subset of the known preemption points would be duplicate work.
 \Cref{alg:algorithm0} demonstrates this na\"ive approach.
-It is seeded with the set of all statically-known synchronization API PPs,
+It is seeded with the set of all statically-known synchronization API preemption points,
 and invoked whenever a new data-race candidate is found.
@@
 Quicksand seeks to maximize completed state spaces,
-as each one serves as a guarantee that all interleavings possible with its PPs were tested;
+as each one serves as a guarantee that all possible interleavings therein were tested;
 \cref{sec:quicksand-discussion} discusses some limitations of this approach.
@@
 \subsection{Initial preemption points}
+\label{sec:quicksand-initial-pps}
 
@@
 	}
-	\caption{Adding new jobs with data-race PPs.}
+	\caption{Adding new jobs with data-race preemption points.}
 	\label{alg:handledatarace}
@@
 while the big job risks not completing in time,
-the small job risks missing the data race entirely if the original PPs were required to expose it.
+the small job risks missing the data race entirely if the original preemption points were required to expose it.
 In practice, I have observed many bugs found quickly by these small jobs,
@@
 The HeuristicPriority($\alpha$) call in \Cref{alg:handledatarace} corresponds to this strategy.
-For single-order races, Quicksand does not initially add a PP for the later access at all:
+For single-order races, Quicksand does not initially add a preemption point for the later access at all:
 if preempting on the first access is capable of reordering the race,
@@
 that is,
-committing in advance to preempt on every single shared memory access \cite{inspect}?
+committing in advance to preempt on every single shared memory access \cite{spin,inspect}?
 It turns out,
@@
 
-\subsection{Convergence}
+\subsection{Convergence to total verification}
 \label{sec:quicksand-convergence}
 
+The proof of Iterative Deepening's soundness is in two parts.
+In the first part, I prove that for any possible interleaving
+one could execute with preemptions anywhere,
+an equivalent interleaving must exist using only data-race and synchronization preemption points.
+In the second, I prove that starting from synchronization preemption points only,
+Iterative Deepening must eventually reach a state space containing such an interleaving,
+no matter how many data races are involved.
+
+\newcommand\ppnext[1]{\ensuremath{\mathsf{next}(#1)}}
+\newcommand\ppinstr[1]{\ensuremath{\mathsf{instr}(#1)}}
+\newcommand\ppothers[1]{\ensuremath{\mathsf{others}(#1)}}
+
+Given a preemption point $p$,
+let $\ppnext{p}$ denote the next transition after $p$ executed by the thread which ran immediately before $p$,
+let $\ppinstr{p}$ denote the first instruction of $\ppnext{p}$,
+and let $\ppothers{p}$ denote the transitions by other threads between $p$ and $\ppnext{p}$.
+
+% TODO: decide if these subsection headers are good, or, not
 \subsubsection{Equivalence}
 
-% TODO: decide if this shouldn't be reordered befor econvergence?
+\begin{lemma}[Equivalence of non-data-race preemption points]
+	For any thread interleaving possible by preempting on any instruction,
+	there exists an equivalent interleaving which uses only data-race and synchronization API preemption points.
+        \label{lem:relevant}
+\end{lemma}
+
+\begin{proof}
+Let $p$ be the first preemption point in the given interleaving
+such that $\ppinstr{p}$ is not a data race with $\ppothers{p}$ nor is a synchronization API boundary.
+Because $\ppinstr{p}$ is not a synchronization boundary,
+no lock can be held during $\ppothers{p}$ that was also held by the first thread across $p$.
+Hence, because $\ppinstr{p}$ is not a data race, it cannot be a shared memory conflict with $\ppothers{p}$ at all.
+Let $i$ be the first instruction among $\ppnext{p}$ which is such a conflict, or a synchronization boundary.
+If $i$ is a shared memory conflict, it must be a data race, for the same reasoning as above.
+Modify the input interleaving by reordering $\ppinstr{p}$ until $i$, not including $i$, to before $\ppothers{p}$.
+% TODO: fix this handwavey citation to at least point to something that says what soundness of dpor means
+By the soundness of DPOR \cite{dpor}, this is equivalent to the input interleaving.
+In other words, $p$ has been transformed into $p'$ such that $\ppnext{p'} = i$,
+which is a data race or synchronization boundary.
+All other preemption points in the input trace can be inductively converted in the same manner.
+\end{proof}
+
 \subsubsection{Saturation}
 
-\subsection{Reallocation false positives}
+For Iterative Deepening to ``eventually'' reach a certain state space,
+all data-race preemption points involved must be {\em reachable} during the test.
+
+\begin{definition}[Reachability]
+	A data race candidate, and its associated preemption point(s),
+	are reachable if it will be identified by a model checker
+	configured to preempt only on already-reachable preemption points.
+\end{definition}
+
+Initially, the statically-available synchronization API preemption points (\cref{sec:quicksand-initial-pps})
+are reachable.
+Reachability of data-race preemption points is transitive.
+
+\begin{lemma}[Saturation of data races]
+        Given any interleaving comprising only data-race and synchronization API preemption points,
+        all involved preemption points are reachable.
+        \label{lem:saturation}
+\end{lemma}
+
+\begin{proof}
+Induct on the preemption points according to the order of their preemptions during an execution sequence.
+Given that the interleaving prefix preceding some point $p$ is reachable,
+either $p$ must be reachable,
+or a new data race among $\ppothers{p}$ will be newly reachable. %, not previously reachable, is now reachable.
+The latter condition suffices because in a finitely-sized codebase,
+there must be finitely many unique racing instruction pairs.
+%, so induction on the number of new data-race preemption points among $\ppothers{p}$ will make $p$ itself reachable.
+
+First, $p$ must be ``coalesced'' away, as well as any other not-yet-reachable points in $\ppothers{p}$.
+Consider the alternate interleaving in which the first thread executes past $p$ until the first already-reachable point,
+then the other threads among $\ppothers{p}$ execute the same way.
+This interleaving's preemption points are all reachable,
+so a state space $\mathcal{S}$ containing it will be tested.
+
+If $p$ is a not-yet-reachable data-race preemption point,
+it must be possible for some other thread to execute a data-racing instruction with $\ppinstr{p}$.
+If this conflict was observed in the state space containing our coalesced interleaving, $p$ is reached.
+Otherwise, appeal to the soundness property of DPOR:
+%(\cref{sec:landslide-dpor}): % FIXME ugh, this section reference is so handwavy/imprecise
+If a program behaviour is possible by interleaving threads at the boundaries of the given transitions,
+it will be tested in the containing state space.
+By contrapositive, to expose this behaviour,
+one or more preemptions must occur in the middle of some transition, rather than at the boundaries.
+
+Let us now see by contradiction that there cannot be {\em multiple} data-race preemption points
+which must all be enabled before either data race can be identified;
+i.e., a circular dependency.
+Assume there does not exist a single transition $t_1 \in \mathcal{S}$
+which alone can be split into $\{t_1',t_1''\}$ by a point $q$,
+such that another thread's concurrent transition $t_2$ conflicts with $t_1''$.
+By the soundness of DPOR, because all $t_2$s are independent with $t_1''$, $\mathcal{S} \equiv \mathcal{S} \cup q$.
+Replacing $\mathcal{S}$ with $\mathcal{S} \cup q$ in the above assumption shows that no {\em pair} of new $q$s would expose new program behaviour, and inductively, no set of $q$s of any size, which contradicts the previous paragraph.
+
+Hence, a single new not-yet-reachable data race is reachable in $\mathcal{S}$. Hence $p$ will be reached.
+\end{proof}
+
+\subsubsection{Convergence}
+
+I name the overall soundness property ``convergence''
+in reference to the way it must eventually arrive,
+after potentially many iterated state spaces,
+at full verification strength.
+
+\begin{theorem}[Convergence]
+	If a bug can be exposed by any thread interleaving possible
+	by preempting on all instructions during a specific test,
+	Iterative Deepening will eventually test an equivalent interleaving which exposes the same bug.
+        \label{thm:convergence}
+\end{theorem}
+
+\begin{proof}
+	For any possible interleaving,
+	Lemma \ref{lem:relevant} provides an equivalent one with only data-race and synchronization preemption points,
+	and Lemma \ref{lem:saturation} proves all involved preemption points are reachable.
+	Hence, Iterative Deepening will eventually test a state space containing the equivalent buggy interleaving.
+\end{proof}
+
+And thus Iterative Deepening is sound. % :)
+
+\subsection{Suppressing reallocation false positives}
 \label{sec:quicksand-realloc}

### 2018-12-12 thesis/quicksand.tex
fix stray revision cmd in quicksand pasta

@@
 	\end{tabular}
-	\caption{If a single-pass \revision{Limited HB analysis}~discarded candidates matching the malloc-recycle pattern,
+	\caption{If a single-pass Limited HB analysis discarded candidates matching the malloc-recycle pattern,
 it would miss the bug in this adversarial program.}

### 2018-12-12 thesis/quicksand.tex
reallocation false positives

@@
 and just after interrupts are re-enabled (on x86, {\tt sti}).
-During data-race detection, {\tt cli} and {\tt sti} are treated as a single global lock
+During data race analysis, {\tt cli} and {\tt sti} are treated as a single global lock
 (note that {\tt cli}'d memory accesses can still race with others that have interrupts on).%
@@
 that are certain to end up being false positives when reordered.
+Such collisions could instead be avoided with a hacked allocator which never recycles memory,
+but this could unacceptably impact performance in {\tt malloc}-heavy tests.
 
@@
 Here, reordering lines~6-7 before line~3 will allow {\tt x} and {\tt x2} to race.
-Such collisions could instead be avoided with a hacked allocator which never recycles memory,
-but this could unacceptably impact performance in {\tt malloc}-heavy tests.
+Discarding the data race report as a false positive after checking just this one execution
+would overlook such a bug,
+but Iterative Deepening is guaranteed to explore the alternate interleaving,
+in which the true data race will show up without {\tt free()} and {\tt malloc()} interposing,
+so it is safe to suppress at first, as I will prove in \cref{sec:quicksand-realloc}.
 Moreover,
@@
 or the latter will find it in an alternate execution within the same state space,
-or the latter will find a different data race that ultimately leads to a bigger state space in which the first one may be found.
+or the latter will find a different data race that ultimately leads to a bigger state space in which the first one may be found,
+much like a generalization of \cref{sec:quicksand-realloc}.
 
@@
 Given that the interleaving prefix preceding some point $p$ is reachable,
-either $p$ must be reachable,
-or a new data race among $\ppothers{p}$ will be newly reachable. %, not previously reachable, is now reachable.
+the proof goal is that either $p$ be reachable,
+or a new data race among $\ppothers{p}$, not previously reachable, be newly reachable.
 The latter condition suffices because in a finitely-sized codebase,
@@
 
-And thus Iterative Deepening is sound. % :)
+% And thus Iterative Deepening is sound. % :)
+This spells QED so we are done \cite{vargomax}. % even more :)))
 
@@
 
-% todo rename "malloc recycle" to "reallocation false positives"
+Next I prove that \cref{sec:quicksand-id-realloc}'s optimization
+of discarding reallocation false positives is sound under Iterative Deepening.
+
+\begin{theorem}[Soundness of eliminating reallocation data-race candidates]
+        If a reallocation candidate is not a false positive,
+DPOR will reorder threads such that
+%DPOR will test an alternate thread interleaving in which
+either
+the accesses can race without fitting the reallocation pattern,
+or a use-after-free bug will be reported immediately.
+\end{theorem}
+
+\begin{proof}
+Any such program must contain an access $a_1$ by one thread T1,
+followed by a {\tt free} and a {\tt malloc} possibly by either thread,
+followed by an access $a_2$ by the other thread T2,
+not depending on the result of the middle malloc.
+Without loss of generality, fix T1 to perform the {\tt free} and T2 the subsequent {\tt malloc},
+as shown in \Cref{fig:recycle-bug}.
+The other cases are similar,
+although note that if T2 performs the {\tt free},
+and the {\tt malloc} is reordered before it,
+T2's final access will be a use-after-free immediately.
+Let us also assume the only way for the program to get pointers to heap memory is through {\tt malloc};
+hence, there must also be some ``publish'' action $p$ by T1 which communicates the address to T2.
+Because this is a true potential data race,
+$p$ must occur before $a_1$, as $a_2$ cannot be reordered before $p$.
+
+The proof goal is that a preemption point be identified during T1 between $p$ and $a_1$.
+The publish action must involve some thread communication,
+whether through a shared data structure or message-passing API.
+If locking or message-passing is used, the set of static synchronization preemption points
+(\cref{sec:quicksand-initial-pps})
+suffices to provide one.
+Otherwise, $p$ (and the corresponding read by T2) will be a potential data race,
+although that may itself be another reallocation candidate.
+In this case, apply induction on the chain of pointers, if any, leading to the shared address containing $p$:
+in the base case, $p$ is communicated via global data or message-passing,
+and in the inductive step, DPOR will reorder threads sufficiently to identify a preemption point on $p$.
+% the below feels kinda shaky :\ don't remember exactly how this works
+% but if someone doubts, they can check out the TR
+Note that this induction may result in several possible intermediate preemption points,
+each requiring a new state space to be tested,
+of course, \Cref{thm:convergence} guarantees this under Iterative Deepening.
+Hence there will be a preemption point between $p$ and $a_1$ no matter the mode of communication.
+
+With this preemption point,
+DPOR will reorder $a_2$ before $a_1$, while not changing $a_2$'s location.
+As T2's {\tt malloc} now occurs before T1's {\tt free}, it will allocate different memory.
+Hence $a_1$ and $a_2$ %will be in the same allocation;
+can race without appearing to fall under the reallocation pattern.
+\end{proof}
+
+Note that this proof does not require PPs on
+the internal lock of {\tt malloc} or {\tt free},
+which is an ideal candidate to ignore via {\tt without\_function} (\cref{sec:landslide-pps})
+to reduce state space size.
+Future work may generalize this proof structure to not rely on specific knowledge
+of {\tt malloc()}'s and {\tt free()}'s behaviour,
+but instead to require only any intervening synchronization event,
+thereby extending the overall soundness proof to accomodate Pure Happens-Before as well as Limited Happens-Before.
+The experiments in future chapters of this thesis will assume that this holds.
 
@@
 % how is free re malloc implemented in landslide?
+
+Landslide includes a generation counter in its heap allocation tracking (\cref{sec:landslide-memory})
+to recognize the reallocation pattern during data race analysis (\cref{sec:quicksand-id-realloc}).
+Each heap allocation is given a unique ID,
+and when evaluating whether two heap accesses can race,
+the IDs of their containing blocks must match
+({\tt was\_freed\_remalloced()} in {\tt memory.c}).
 

### 2018-12-12 thesis/future.tex
handle many todos in educ
3 more citations to dunk on machine learning in warp zone

@@
 \section{Verification}
+\label{sec:warpzone-verif}
 
@@
 Machine learning has become popular recently as a magic cure-all for many programming problems,
-but is also notorious for amplifying any biases in its training set in often unpredictable ways
-\cite{conceptnet-bias,amazon-ai}.
-In the domain of concurrency testing, this would translate to being able to find only types of bugs
-already seen in the training data, abandoning any novel bug patterns to further obscurity.
+but is also notorious for amplifying any biases in its training set
+(or in the minds of its developers)
+in unpredictable ways
+that can harm people of color \cite{conceptnet-bias,gender-shades},
+women \cite{gender-bias-ai,amazon-ai},
+and transgender people \cite{misgendering-machines}.
+In the domain of concurrency testing,
+%careless application of
+such bias would be unlikely to result in interpersonal discrimination,
+but
+%I believe
+it would translate to
+%being able to find
+finding only types of bugs
+already seen in the training data,
+abandoning any novel bug patterns to further obscurity.
 Granted,
@@
 but when a bug is not found
-these strategies are also transparent enough that the user can understand the nature of the partial verification.
+these strategies also enable the user to understand the nature of the partial verification.
 %
@@
 Nevertheless,
-any overall search heuristic for concurrency testing
-must be designed to support the human user's agency first and foremost,
+I firmly believe that using machine learning as a bug-finding heuristic
+without compensating extremely carefully for its inherent biases
+would run counter to this thesis's central tenets of transparency and user agency.
+Any overall search heuristic
+%for concurrency testing
+must be designed to support the human user first and foremost,
 that she may readily understand partial verification results

### 2018-12-13 thesis/quicksand.tex
quicksand impl

@@
-% TODO: scan for abbrs like PP, MC, SSS-MC
+% TODO: scan for abbrs like PP, MC, SSS-MC - well mc is ok but rephrase if necessary anyway
 \chapter{Quicksand}
@@
 % here is a resulting state space
-% this needs a DR PP to FAB
+% this needs a DR preemption point to FAB
 % qs example (take from thesis slides)
@@
 \subsection{Data-race preemption points}
+\label{sec:quicksand-dr-pps}
 
@@
 \subsection{Choosing the best job}
+\label{sec:quicksand-choosing}
 
@@
 \subsection{Heuristics}
+\label{sec:quicksand-heuristics}
 
@@
 Replacing $\mathcal{S}$ with $\mathcal{S} \cup q$ in the above assumption shows that no {\em pair} of new $q$s would expose new program behaviour, and inductively, no set of $q$s of any size, which contradicts the previous paragraph.
-
+%
 Hence, a single new not-yet-reachable data race is reachable in $\mathcal{S}$. Hence $p$ will be reached.
@@
 
-% And thus Iterative Deepening is sound. % :)
-This spells QED so we are done \cite{vargomax}. % even more :)))
+And thus Iterative Deepening is sound. % :)
 
@@
 
-Note that this proof does not require PPs on
+This spells QED so we are done \cite{vargomax}. % even more :)))
+Note that this proof does not rely on the existence of preemption points on
 the internal lock of {\tt malloc} or {\tt free},
@@
 
-% TODO: add -M flag discussion
-
-% TODO: cant swap x100
-
-% TODO: put a code snippet of what ap rogress report looks like (take from some htm log)
-
-% this section developers guide to quicksand
-
-% job
-% messaging
-% work
-% option
-% pp
-
-% cool tricks
-% cant swap x100 % cite [Inf09] ?
-
-% how is free re malloc implemented in landslide?
-
-Landslide includes a generation counter in its heap allocation tracking (\cref{sec:landslide-memory})
-to recognize the reallocation pattern during data race analysis (\cref{sec:quicksand-id-realloc}).
+Quicksand is an independent program that wraps the execution of several stateless model checker instances.
+It is expected that Landslide be this checker,
+but any other tool which implements the same messaging interface would be compatible as well.
+The implementation is roughly 3000 lines of C.
+All source files mentioned in this section live in the {\tt id/} subdirectory of the Landslide repository,
+with the exception of the Landslide extensions listed last.
+As \Cref{chap:landslide} was in some sense a developer's guide to Landslide,
+this section will serve similarly for Quicksand.
+
+\subsection{User interface}
+
+The available command-line options for configuring its
+CPU-time budget, exploration modes, and so on
+are listed in \cref{sec:landslide-quicksand-options}.
+Additionally, Quicksand periodically issues a progress report
+at fixed intervals to inform the user on the completion, bug-finding, and/or estimated progress of each job.
+\Cref{fig:quicksand-progress} shows an example.
+I highlight a few notable features of the jobs therein
+to serve as a concrete example that may cement the reader's intuition
+of \cref{sec:quicksand-id}'s more abstract algorithm descriptions:
+\begin{itemize}
+	\item Jobs 0, 1, 2, and 3 are the initially-seeded state spaces (\cref{sec:quicksand-initial-pps}).
+	\item Job 2 reports a bug found, and shows the filename of the HTML preemption trace
+		(\cref{sec:landslide-bugreport}, \Cref{fig:bugreport})
+		which the user should examine to diagnose it.%
+		\footnote{I actually cheated by copy/pasting this job alone from a different run of Quicksand;
+		the other jobs come from a test with the bug already fixed,
+		in order that exploration progress far enough to defer some jobs for the sake of example.
+		In a real execution, jobs 3, 5, 7, 8, 9, and 11 would be cancelled.}
+	\item Jobs 4 and 6 are the ``small'' jobs added to test the two data races in isolation;
+		5 and 7 are the corresponding ``large'' jobs (\cref{sec:quicksand-dr-pps}).
+	\item Job 11 is the maximal state space, containing all synchronization preemption points
+		and both currently-known data races.
+	\item The intermediate jobs 8, 9, and 10 have been suspended for having ETAs
+		at least twice as large as the provided CPU budget (1 hour),
+		according to the ETA scaling factor heuristic (\cref{sec:quicksand-heuristics}).
+	\item Note that job 11's ETA is currently lower than 8's, 9's, and 10's, despite being a strict superset of each.
+		This corresponds to the surprising ETA inversion situation discussed in \cref{sec:quicksand-choosing}:
+		Quicksand simply hasn't made as much progress into job 11 (compare their percentage estimates rather than ETAs)
+		for its ETA to be accurate enough yet.
+\end{itemize}
+Future work could extend these progress reports to be more interactive,
+allowing the user to reprioritize state spaces at her whim
+or to disable certain data-race preemption points after checking them by hand,
+as discussed in \cref{sec:future-friendly}.
+
+% TODO: syn hi
+\begin{figure}[t]
+	\begin{center}
+		\footnotesize
+	\begin{tabular}{l}
+			% version from buggy run
+		%\texttt{==== PROGRESS REPORT ====} \\
+		%\texttt{total time elapsed: 40s} \\
+		%\texttt{[JOB 0] COMPLETE (4 interleavings tested; 7s elapsed)} \\
+		%\texttt{       PPs: \{ \}} \\
+		%\texttt{[JOB 1] Running (10.677083\%; ETA 2m 15s)} \\
+		%\texttt{       Log: id/ls-output.log.ujrk9U -- PPs: \{ 'mutex\_lock' \}} \\
+		%\texttt{[JOB 2] BUG FOUND: landslide-trace-1544661430.29.html (51 interleavings tested; 1 preemptions)} \\
+		%\texttt{       Log of lprintf()s: id/ls-setup.log.7UMGLr} \\
+		%\texttt{       Log: id/ls-output.log.3Ol8rG -- PPs: \{ 'mutex\_unlock' \}} \\
+		%\texttt{[JOB 3] CANCELLED} \\
+		%\texttt{       PPs: \{ 'mutex\_lock' 'mutex\_unlock' \}} \\
+		%\texttt{[JOB 5] COMPLETE (4 interleavings tested; 7s elapsed)} \\
+		%\texttt{       PPs: \{ 'data race 5@ 0x1000ec5' \}} \\
+		%\texttt{[JOB 13] Setting up...} \\
+		%\texttt{       PPs: \{ 'data race 6@ 0x1000ec5' \}} \\
+		%\texttt{[JOB 14] Pending...} \\
+		%\texttt{       PPs: \{ 'mutex\_unlock' 'data race 6@ 0x1000ec5' \}} \\
+		%\texttt{=========================} \\
+
+		\texttt{==== PROGRESS REPORT ====} \\
+		\texttt{total time elapsed: 2m 52s} \\
+		\texttt{[JOB 0] COMPLETE (4 interleavings tested; 7s elapsed)} \\
+		\texttt{~~~~PPs: \{ \}} \\
+		\texttt{[JOB 1] Running (21.932870\%; ETA 8m 14s)} \\
+		\texttt{~~~~ PPs: \{ 'mutex\_lock' \}} \\
+		% version that came from the actual test
+		%\texttt{[JOB 2] Running (46.373457\%; ETA 4m 35s)} \\
+		%\texttt{~~~~ PPs: \{ 'mutex\_unlock' \}} \\
+		\texttt{[JOB 2] BUG FOUND: landslide-trace-1544661430.29.html (51 interleavings tested)} \\
+		%\texttt{~~~~Log of lprintf()s: id/ls-setup.log.7UMGLr} \\
+		\texttt{~~~~ PPs: \{ 'mutex\_unlock' \}} \\
+		\texttt{[JOB 3] Running (9.852431\%; ETA 24m 32s)} \\
+		\texttt{~~~~ PPs: \{ 'mutex\_lock' 'mutex\_unlock' \}} \\
+		\texttt{[JOB 4] COMPLETE (6 interleavings tested; 9s elapsed)} \\
+		\texttt{~~~~PPs: \{ 'data race @ 0x102917' \}} \\
+		\texttt{[JOB 5] Running (3.710938\%; ETA 1h 2m 14s)} \\
+		\texttt{~~~~ PPs: \{ 'mutex\_lock' 'mutex\_unlock' 'data race @ 0x102917' \}} \\
+		\texttt{[JOB 6] COMPLETE (4 interleavings tested; 8s elapsed)} \\
+		\texttt{~~~~PPs: \{ 'data race @ 0x1000ecf' \}} \\
+		\texttt{[JOB 7] Running (6.119792\%; ETA 33m 14s)} \\
+		\texttt{~~~~ PPs: \{ 'mutex\_lock' 'mutex\_unlock' 'data race @ 0x1000ecf' \}} \\
+		\texttt{[JOB 11] Running (3.670247\%; ETA 50m 16s)} \\
+		\texttt{~~~~ PPs: \{ 'mutex\_lock' 'mutex\_unlock' 'data race @ 0x102917' 'data race @ 0x1000ecf' \}} \\
+		\texttt{[JOB 8] Deferred... (33.340567\%; ETA 2h 6m 3s)} \\
+		\texttt{~~~~ PPs: \{ 'mutex\_unlock' 'data race @ 0x102917' \}} \\
+		\texttt{[JOB 9] Deferred... (34.466226\%; ETA 2h 35m 37s)} \\
+		\texttt{~~~~ PPs: \{ 'mutex\_unlock' 'data race @ 0x1000ecf' \}} \\
+		\texttt{[JOB 10] Deferred... (11.113790\%; ETA 4h 20m 31s)} \\
+		\texttt{~~~~ PPs: \{ 'mutex\_lock' 'data race @ 0x102917' \}} \\
+		% no, i deleted all of them pretending the other data races dont exist
+		%\texttt{And 63 more pending jobs should time allow.} \\
+		\texttt{=========================} \\
+	\end{tabular}
+	\end{center}
+	\caption{Example Quicksand progress report showing the various possible job states.}
+	\label{fig:quicksand-progress}
+\end{figure}
+
+Besides the progress reports, Quicksand also prints a notice for each new data race that Landslide detects,
+like so, corresponding to the above progress report, for example:
+\begin{center}
+	%{\tt Found a racy access at 0x0100006f in txn (410user/progs/htm2.c:47)}
+	\begin{tabular}{l}
+	{\tt Found a racy access at 0x00102917 in deschedule <unknown>} \\
+	{\tt Found a racy access at 0x01000ecf in cond\_signal (cond.c:101)}
+	\end{tabular}
+\end{center}
+If using Limited Happens-Before instead of Pure,
+it prints ``potentially-racy access'' instead.
+This is implemented in {\tt pp\_new()} in {\tt pp.c}.
+If the CPU budget runs out and Quicksand must stop exploring prematurely
+(or the user's patience runs out and she interrupts it with ctrl-C),
+it prints a final report of all data races it was not able to finish classifying as either buggy or benign,
+and urges the user to finish checking them with visual inspection
+({\tt print\_live\_data\_race\_pps()} in {\tt pp.c}).
+It is this feature which one respondent in the student user survey (\cref{sec:education-reasons-worthwhile})
+credited for finding an extra bug.
+
+If the verbose option ({\tt -v}) is supplied,
+Quicksand will also print one line per interleaving tested by all its Landslide instances,
+showing the number of branches tested so far, the estimated percent progress, and the ETA,
+as shown earlier in \cref{sec:landslide-estimate}.
+This produces a lot more output, and can make progress reports hard to read as they scroll off the screen quickly,
+but the author personally finds this mode less disorienting than ten seconds of pure silence between each progress report.
+Of course, future work could improve this with a GUI, or at least a split-screen console view.
+It will also cause the progress reports to report more detailed information,
+such as which preemption points are nondeterministic data-races (\cref{sec:quicksand-pps})
+and number of reallocation false positives suppressed (\cref{sec:quicksand-id-realloc}).
+
+\subsection{Model checker interface}
+\label{sec:quicksand-impl-mc}
+
+The interface with the model checker has two parts.
+First, when starting each job, Quicksand creates a configuration file declaring which preemption points to use,
+plus other test-case-specific options such as
+which preemption points to suppress (\cref{sec:landslide-pps}),
+especially those arising from the {\tt malloc} lock (\cref{sec:quicksand-realloc}),
+which functions DPOR and data-race analysis should treat as trusted code (\cref{sec:landslide-config-landslide}),
+whether to enable mutex-testing mode (\cref{sec:landslide-dynamicconfig}),
+transactional memory (\cref{sec:tm-implementation}),
+and so on.
+This is done by {\tt run\_job()} in {\tt job.c}.
+
+Then, a dedicated Quicksand thread
+({\tt start\_job()} in {\tt job.c})
+communicates with the MC process via message-passing
+over a FIFO pipe % not gonna cite io.c c.c
+({\tt talk\_to\_child()} in {\tt messaging.c}).
+Landslide messages after testing each interleaving to report updated progress and ETAs
+as well as whenever a new data-race candidate or bug is found.
+Quicksand in turn replies whether to resume/suspend (due to too high ETA) or quit (due to timeout)
+({\tt handle\_should\_continue()} in {\tt messaging.c}).
+It suspends jobs simply by making Landslide wait on a message-passing reply.
+Should Quicksand later re-schedule a suspended job, it sends a message to continue,
+resuming Landslide right where it left off;
+otherwise, it wakes it up only after time runs out, causing it to exit immediately.
+The message-passing format is defined at the top of {\tt messaging.c},
+and a matching definition appears in Landslide's source file of the same name.
+
+\subsection{Architecture}
+
+Quicksand's overall architecture is a thread pool of workers,
+one for each CPU it was configured to use with {\tt -c} (\cref{sec:landslide-quicksand-options}).
+These threads do not correspond directly to each active instance of Landslide,
+as some may be deferred;
+rather, each worker thread links up temporarily to a job thread, whose duties the previous paragraph describes,
+and process them as the overall work-queue of state spaces to be tested demands.
+Following is a brief description of each of Quicksand's major modules.
+
+\begin{itemize}
+	\item {\bf Job management} ({\tt job.c}):
+		Contains the lifecycle code for job threads,
+		generation of Landslide configuration files,
+		and Linux process management code to launch child Landslide instances ({\tt run\_job()}).
+	\item {\bf Messaging} ({\tt messaging.c}):
+		Manages communication with child Landslides ({\tt talk\_to\_child()},
+		implementing certain aspects of Iterative Deepening,
+		creating new jobs in response to data race reports ({\tt handle\_data\_race()})
+		and deferring too-big jobs in reponse to ETA updates  ({\tt handle\_estimate()}).
+	\item {\bf Preemption point registry} ({\tt pp.c}):
+		Tracks the set of known synchronization primitives (initialized by main)
+		and data races ({\tt pp\_new()}),
+		including set comparison routines ({\tt pp\_subset()})
+		and computing a job's priority based on the types of included preemption points ({\tt unexplored\_priority()}).
+	\item {\bf Workqueue} ({\tt work.c}):
+		Implements the per-CPU worker threads,
+		including the check for whether to switch priority from one job to another
+		(\Cref{alg:shouldworkblock}, {\tt should\_work\_block()} and {\tt get\_job()}),
+		as well as managing the shared workqueue of jobs overall ({\tt workqueue\_thread()}).
+		Also implements the fixed-interval progress reporting ({\tt progress\_report\_thread()}).
+	\item {\bf Bugs} ({\tt bug.c}):
+		Tracks a list of found bugs for main to repeat at program exit,
+		and implements the check for superset state spaces to be cancelled if a subset already found a bug
+		({\tt bug\_already\_found()}).
+	\item {\bf Options} ({\tt option.c}):
+		Processes command-line options,
+		checking for legality of various combinations of exploration modes.
+		New options may be added with the convenient, if scarily-implemented,
+		macros {\tt DEF\_CMDLINE\_FLAG()} and {\tt DEF\_CMDLINE\_OPTION()}.
+	\item {\bf Main} ({\tt main.c}):
+		Initializes the default state spaces,
+		waits for worker threads to terminate after either completion or time-out,
+		and issues a final list of bug reports, data race reports, or congratulations as appropriate.
+\end{itemize}
+
+Finally, because in very large tests,
+the number of suspended Landslide instances may grow without abatement,
+Quicksand checks every progress report interval whether the memory footprint of these inactive Landslides
+pose a threat to the system's total memory.
+Implemented in {\tt cant\_swap()} \cite{cant-stop} in {\tt work.c},
+it checks if the system's memory usage exceeds a fixed percentage of its total
+({\tt RAM\_USAGE\_DANGERZONE}, default 90),
+and if so,
+abandons a fixed percentage of suspended Landslides
+({\tt KILL\_DEFERRED\_JOBS}, default 50).
+%
+Generally, the currently-running Landslide instances should never threaten to hit swap,
+as there can only be as many of those as CPUs,
+but this also accounts for memory used by other processes,
+so this is not guaranteed to avoid swapping on a heavily-stressed system
+(such as running multiple Quicksands at once).
+Naturally, if Quicksand ever needs to invoke this protocol,
+any hope at a total verification is compromised.
+
+\subsection{Exploration modes}
+\label{sec:quicksand-impl-modes}
+
+In addition to Iterative Deepening,
+which Quicksand defaults to if no options are given to specify otherwise,
+Quicksand also supports several alternate exploration strategies, as follows.
+
+\begin{itemize}
+	\item {\bf Single state space, basic DPOR} ({\tt -C}):
+		Runs a single instance of Landslide configured to preempt on all statically-known synchronization PPs.
+		Corresponds to dBug's approach \cite{dbug-ssv}.
+		Never adds any new preemption points based on data race reports.
+	\item {\bf Single state space, ICB} ({\tt -I}):
+		Runs a single instance of Landslide with preemption points as above,
+		but running Iterative Context Bounding with BPOR (\cref{sec:landslide-icb})
+		instead of plain DPOR.
+		Corresponds to Chess's approach \cite{chess}.
+		Requires either {\tt -C} or {\tt -M} (see below).
+	\item {\bf Single state space, preempt-everywhere} ({\tt -0}, i.e., numeral zero):
+		Runs a single instance of Landslide as above,
+		but preempting on every shared memory access, not just synchronization.
+		Corresponds to the approach of SPIN \cite{spin} and Inspect \cite{inspect}.
+		May optionally be combined with {\tt -I}.
+	\item {\bf Maximal state space mode} ({\tt -M})
+		Runs the na\"{i}ve version of Iterative Deepening shown in \Cref{alg:algorithm0},
+		i.e.,
+		immediately abandons any state space whenever a superset of it exists.
+		This results in always testing the maximal state space only, with no inherent parallelism,
+		and optimizes for the fastest verification when the user has reason to believe no bugs will exist.
+		No prior work implements this approach.
+		Note that this mode was implemented after \cite{quicksand}'s publication,
+		and I will feature it in the evaluation of transactional memory
+		(\cref{sec:tm-eval}) rather than in this chapter.
+\end{itemize}
+
+Note that Quicksand restricts ICB to be usable only in modes when it is running only one Landslide at a time.
+ICB is itself a heuristic search ordering strategy to uncover bugs faster,
+so while technically easy to run Iterative Deepening with all jobs thereunder running ICB,
+that would suffer both approaches' repeated work compounded.
+\cref{sec:warpzone-heuristics} discusses integrating the two approaches to hopefully reap the benefits of both.
+However, maximal state space mode does support ICB,
+as the former focuses on verification only,
+but if the result is a time-out,
+the user may find an ICB-style preemption-bounded partial verification useful.
+
+\subsection{Landslide extensions}
+\label{sec:quicksand-impl-landslide}
+
+I have added several features to Landslide specifically for use under Quicksand.
+Source files mentioned in this subsection live under the usual Landslide source directory.
+
+The other end of the messaging protocol (\cref{sec:quicksand-impl-mc})
+is implemented in {\tt messaging.c}.
+When Quicksand suspends Landslide,
+it detects how much time it spent asleep,
+and corrects for that amount during its next ETA computation ({\tt fudge\_time()} in {\tt estimate.c}).
+Landslide's data race analysis also includes a heuristic to avoid reporting ``too suspicious''
+data-race candidates which it believes arise from the initialization pattern \cite{eraser}:
+if a conflicting access pair is single-order (\cref{sec:quicksand-heuristics})
+and also arose during a known synchronization API's {\tt init()} or {\tt destroy()} function,
+Landslide will not message it to Quicksand,
+at least not until it is reclassified as both-order.
+
+To recognize the reallocation pattern
+discussed in \cref{sec:quicksand-id-realloc}
+during data race analysis,
+Landslide includes a generation counter in its heap allocation tracking (\cref{sec:landslide-memory}).
 Each heap allocation is given a unique ID,
@@
 the IDs of their containing blocks must match
-({\tt was\_freed\_remalloced()} in {\tt memory.c}).
+({\tt was\_freed\_remalloced()} in {\tt memory.c}),
+in addition to the other requirements of Happens-Before.
+
+Preempt-everywhere mode (\cref{sec:quicksand-impl-modes})
+imposes a heavy burden on Landslide on account of the sheer number of preemption points involved.
+First of all, because there are separate tracing entrypoints for memory accesses and instructions ({\tt instrument.c}),
+it cannot simply invoke the checkpointing routine (\cref{sec:landslide-timetravel}) immediately.
+Also, we must still exclude thread-local and kernel (if testing userspace) or user (if testing kernelspace) accesses.
+Rather,
+the memory analysis (\cref{sec:landslide-memory}) invokes
+{\tt maybe\_preempt\_here()} in {\tt pp.c}
+for every access it would ordinary record for DPOR.
+If the access is outside of the current stack frame,
+and not part of the mutexes (unless {\tt TESTING\_MUTEXES}),
+this sets a scheduler action flag {\tt preempt\_for\_shm\_here}
+which makes preemption point identification treat it the same as a data race (\cref{sec:landslide-pps}).
+{\tt check\_withins()} is also modified to never switch to allowlist mode.
+Finally,
+Landslide increases its heuristic constant for infinite loop detection (\cref{sec:landslide-infloop})
+on the first interleaving from 4000 to $2^{20}$,
+to account for the increased orders of magnitude in state space size.
 

### 2018-12-19 thesis/quicksand.tex
quicksand expt setup, test suite

@@
 \cref{sec:quicksand-soundness} will justify the technique's formal verification power.
-In the evaluation (\cref{sec:quicksand-eval}),
+In the evaluation, %(\cref{sec:quicksand-eval}),
 \cref{sec:quicksand-eval-nondets} in particular will justify the need
@@
 Iterative Deepening
-is a search strategy,
-for exponentially-sized state spaces in general,
+is a search strategy
+for exponentially-sized state spaces, in general,
 which
@@
 
-Note that Iterative Deepening is a {\em wrapper} algorithm around stateless MC.
+Note that Iterative Deepening is a wrapper algorithm around stateless MC.
 A model checker is still used to test each state space, and other reduction techniques such as DPOR
@@
 The maximal state space is guaranteed to subsume all execution sequences reachable in any subset state space,
-so considering any incomplete subset of the known preemption points would be duplicate work.
+so considering any incomplete subset of the known preemption points would be redundant work.
 \Cref{alg:algorithm0} demonstrates this na\"ive approach.
@@
         % TODO: put the figure form yr snake fight
-	\caption[Quicksand incorporates data race reports as new preemption points at run-time.]
-		{Quicksand incorporates data race reports as new preemption points at run-time,
+	\caption[Quicksand incorporates data races as new preemption points at run-time.]
+		{Quicksand incorporates data races as new preemption points at run-time,
 		by managing the exploration of multiple state spaces,
@@
 	\end{center}
+	% TODO: short ver
 	\caption{A common execution pattern with {\tt malloc()} that produces false positive data race candidates.}
@@
 	\end{tabular}
+	% TODO: shortver
 	\caption{If a single-pass Limited HB analysis discarded candidates matching the malloc-recycle pattern,
@@
 
+% TODO: check spcaing between defs/lemmas/etc and surrounding text
 \begin{definition}[Reachability]
@@
 		instead of plain DPOR.
-		Corresponds to Chess's approach \cite{chess}.
+		Corresponds to CHESS's approach \cite{chess}.
 		Requires either {\tt -C} or {\tt -M} (see below).
-	\item {\bf Single state space, preempt-everywhere} ({\tt -0}, i.e., numeral zero):
+	\item {\bf Single state space, preempt-everywhere} ({\tt -0}):
 		Runs a single instance of Landslide as above,
 		but preempting on every shared memory access, not just synchronization.
-		Corresponds to the approach of SPIN \cite{spin} and Inspect \cite{inspect}.
-		May optionally be combined with {\tt -I}.
-	\item {\bf Maximal state space mode} ({\tt -M})
+		Corresponds to the approach of SPIN \cite{spin} and Inspect \cite{inspect};
+		CHESS supports this mode as well with optional compiler instrumentation.
+		Requires {\tt -C}; may optionally be combined with {\tt -I}.
+	\item {\bf Maximal state space mode} ({\tt -M}):
 		Runs the na\"{i}ve version of Iterative Deepening shown in \Cref{alg:algorithm0},
@@
 on the first interleaving from 4000 to $2^{20}$,
-to account for the increased orders of magnitude in state space size.
+to account for the increased orders of magnitude in preemptible events.
 
@@
 
+% TODO: introd this
+
+I pose the following evaluation questions.
+
+\begin{enumerate}
+	\item Does integrating data-race preemption points improve the accuracy of model checking?
+		\begin{enumerate}
+			\item Do data-race preemption points expose new bugs that couldn't be found with
+				synchronization ones alone?
+			\item Do data-race preemption points expose the same bugs
+				the preempt-everywhere approach could find, faster?
+			\item Does Quicksand provide more full verifications %of bug-free programs
+				more quickly than the preempt-everywhere approach?
+			% not related to new phrasing of Q1, and also the answer is "kinda sorta not really sometimes" anyway
+			%\item Does Iterative Deepening find bugs faster in subset state spaces,
+			%	even without data-race preemption points?
+			\item How does the choice between Pure and Limited Happens-Before
+				affect bug-finding and verification performance?
+		\end{enumerate}
+	\item Does testing alternate interleavings with model checking improve the accuracy of data race analysis?
+		\begin{enumerate}
+			\item Does Quicksand avoid false positives compared to single-execution Limited Happens-Before?
+			\item Does Quicksand find data-race bugs that single-execution Pure or Limited Happens-Before would miss?
+		\end{enumerate}
+\end{enumerate}
+
+\subsubsection{Test suite}
+
+The test suite consists of 79 P2 student thread libraries (\cref{sec:pebbles}),
+submitted in 15-410 during the Spring 2014, Fall 2014, and Spring 2015 semesters,%
+\footnote{The 2014 semesters were before \Cref{chap:education}'s user study experiments,
+and for Spring 2015 (the first semester thereof),
+students who used Landslide during the project were excluded from this dataset.}
+and 78 Pintos student kernels (\cref{sec:overview-pintos}),
+submitted in Berkeley's CS162 and U. Chicago's CMSC 23000 during Spring 2015.
+The P2s in this dataset average 1807 lines of C and x86 assembly code,
+%, with a standard deviation of 489.5, % i don't know where hte pintos stdev went and i don't wanna recalc it
+and the Pintoses average 718 lines (by {\tt diff} to the provided basecode),
+for a total of 198,772 lines of code tested for this evaluation.
+
+I chose P2s and Pintoses for this test suite because of the relative ease of generating hundreds of unique state spaces,
+varied in size and correctness, and with a diverse set of bug types.
+In addition to concurrency bugs,
+many of the codebases exhibited {\em deterministic} bugs (i.e., encountered on the first interleaving tested),
+which I fixed by hand before running these tests to ensure that every bug in this study required meaningful work by the model checker.
+While many prior work stateless model checking papers
+% TODO: cite some
+publish studies of single-digit or low-double-digit numbers of
+bugs found in ``real-world'' programs, often confirmed by the upstream developers as severe,
+to motivate stateless model checking to be used in production settings,
+I believe this approach to be too anecdotal for comparing several model checking strategies {\em against each other},
+and opt for this approach instead for better statistical significance.%
+\footnote{Not to mention -- as I couldn't say in a conference paper, but can say now --
+that extending Landslide to support native Linux programs,
+complete with filesystem and network nondeterminism,
+would have been an engineering burden beyond my ability to do alone and still graduate on time.}
+
+\newcommand\mxtest{\texttt{mx\_test}}
+\newcommand\tej{\texttt{join\_test}}
+\newcommand\bct{\texttt{bcast\_test}}
+\newcommand\paraguay{\texttt{signal\_test}}
+\newcommand\paradise{\texttt{sem\_test}}
+\newcommand\rwldgr{\texttt{rwlock\_test}}
+\newcommand\prisema{\texttt{sched\_test}}
+\newcommand\waitsimple{\texttt{wait\_test}}
+\newcommand\alarmsimul{\texttt{alarm\_test}}
+
+I tested P2s with 6 multithreaded programs:
+\mxtest, for locking algorithm correctness,
+\tej, a test of thread lifecycle,
+\bct~and \paraguay~for condition variables,
+\paradise~for semaphores,
+% nb: there used to be a "f u latex" comment here with a \\ after dgr, prob for margins
+and \rwldgr for r/w locks.
+These are the same tests I distributed Landslide with in \Cref{chap:education};
+\cref{sec:education-pebbles-tests} describes them in further detail.
+For \mxtest, \paradise, and \paraguay,
+I used the {\tt without\_function} command to exclude {\tt thr\_create}, {\tt thr\_exit}, and {\tt thr\_join} preemption points,
+and for \mxtest I enabled {\tt TESTING\_MUTEXES} (\cref{sec:landslide-dynamicconfig}).
+I tested Pintoses with 3 programs from the class's provided test suite:
+\prisema, a test of the kernel scheduling algorithm,
+\alarmsimul, for the timer sleep routine,
+and \waitsimple, for process lifecycle system calls.
+These are a subset of those used in \Cref{chap:education}; see \cref{sec:education-pintos-tests}.
+Some of the Pintoses were partially implemented,
+so each test could only be run on a subset of the 78 submissions;
+see the ``Number tested'' column in \Cref{tab:drbugs}.
+For all tests,
+I also excluded preemption points on {\tt malloc}'s internal lock using {\tt without\_function}.
+In total, the evaluation comprises 629 unique tests (i.e., pairs of a test program and a Pintos or P2),
+at least 181 of which will be seen to expose bugs.
+
+\subsubsection{Experimental setup}
+
+To evaluate the benefits of data-race preemption points and Iterative Deepening separately,
+I ran the test suite under Quicksand in three different experimental configurations,
+each of which was given a 1-hour budget and 10 CPUs for each test.
+
+\begin{itemize}
+	\item {\bf QS-Limited-HB}: Quicksand with Landslide configured to use Limited Happens-Before for its data race analysis ({\tt -H}),
+	\item {\bf QS-Pure-HB}: Quicksand using Pure Happens-Before instead ({\tt -V}), and
+	\item {\bf QS-Sync-Only}: Quicksand with initial preemption points only \cref{sec:quicksand-initial-pps},
+		but never adding new ones from reported data races.
+\end{itemize}
+
+I represented the MC State of the Art%
+\footnote{The author's DJ name.}
+with three configurations of stand-alone Landslide on the same test suite,
+corresponding to the search strategies discussed in \cref{sec:quicksand-impl-modes} and \cref{sec:quicksand-impl-landslide}.
+
+\begin{itemize}
+	\item {\bf SSS-MC-DPOR}: Single state space mode ({\tt -C})
+		using the maximal preemption point set from \cref{sec:quicksand-initial-pps},
+		explored with DPOR (\cref{sec:landslide-dpor}),
+		%From prior work, dBug \cite{dbug-ssv} implements this approach, among others. %% nb -- already discussed above
+        \item {\bf SSS-MC-ICB}: With preemption points as above,
+		but instead using ICB \cite{chess-icb} with BPOR \cite{bpor} to find bugs faster
+		({\tt -I}, \cref{sec:landslide-icb}), and
+		%From prior work, CHESS \cite{chess} implements this approach.
+        \item {\bf SSS-MC-Shared-Mem}:
+		Using ICB+BPOR, configured to preempt on any shared memory access ({\tt -0}).
+                (decided at runtime, excluding threads' accesses to their own stacks),
+                which in principle includes all possible data races.
+		%From prior work, Inspect \cite{inspect} implements this approach,
+		%and CHESS also supports this mode with optional compiler instrumentation.
+\end{itemize}
+
+Prior work has shown how to parallelize DPOR of a single state space across multiple processors \cite{parallel-dpor},
+but it remains an open research problem how to extend the algorithm to ICB.
+Hence, I optimistically gave control experiments a linear speedup of 10 hours per test with 1 CPU;
+i.e., assuming it could parallelize with 100\% efficiency.
+To match this,
+Quicksand reports the CPU-time spent in addition to the wall-clock time for a resource-fair comparison,
+although Quicksand's inherent parallelism is a convenient benefit unto itself.
+All tests ran on 12-core 3.2 GHz Xeon W3670 machines with 12GB of RAM.
+
 \subsection{Finding bugs}

### 2018-12-19 thesis/education.tex
write quicksand design, beginnings of proof stuff
daves final revisions to educ

@@
 
-Since the Spring 2015 semester I have given a guest lecture in 15-410
+%Since the Spring 2015 semester
+From the Spring 2015 semester through Spring 2018,
+I have given a guest lecture in 15-410
 to recruit students to participate in the user study.
@@
 The most recent lecture slides are available on the course website at
-\url{http://www.cs.cmu.edu/~410-s18/lectures/L14_Landslide.pdf},
+\url{https://www.cs.cmu.edu/~410-s18/lectures/L14_Landslide.pdf},
 and all semesters' editions at
@@
 (CMU version)
-and
+and at % these "at"s can optionally be adjusted to make url line breaks better
 \url{http://www.contrib.andrew.cmu.edu/~bblum/landslide-guide-psu.pdf}
 (PSU version)
-and
+and at
 \url{https://github.com/bblum/talks/tree/master/irb}
@@
 		Copied directly from the P2 test suite;
-		tests for mutually-exclusive and deadlock-free {\tt rwlock\_downgrade()}
-		\revisionminor{(a function which atomically converts a lock held in exclusive write mode to shared read mode)}.
-		Written by me (as a TA).
+		tests R/W locks for mutually-exclusive and deadlock-free {\tt rwlock\_downgrade()},
+		\revisionminor{which should atomically convert a lock held in exclusive write mode to shared read mode}.
+		Written by me (during my time as a TA).
 	\item {\tt thr\_exit\_join}:
@@
 %\url{https://github.com/Berkeley-CS162/group0/tree/master/pintos/src/tests}.
-The fifth test is available at
-\url{https://github.com/bblum/landslide/blob/master/pebsim/pintos/priority-donate-multiple.c}.
+The fifth test is available in the Landslide repository at {\tt pebsim/pintos/priority-donate-multiple.c}.
+%\url{https://github.com/bblum/landslide/blob/master/pebsim/pintos/priority-donate-multiple.c}.
 
@@
 \section{Evaluation}
+\label{sec:education-eval}
 
@@
 
-\begin{itemize}
+\vspace{0.5em}
+\begin{enumerate}
 	\item How many bugs, and of what severities, does Landslide help students find and fix before submitting their code?
@@
 	\item How well does Landslide apply to operating systems projects outside of CMU?
-\end{itemize}
-\vspace{1em}
+\end{enumerate}
+\vspace{0.5em}
 
@@
 % should you be using the combined/grouped AD test instead?
-\footnote{I choose not to correct for multiplicity among these 4 comparsons \revisionminor{as} I do in the next section
+\footnote{I choose not to correct for multiplicity among these four comparsons \revisionminor{as} I do in the next section
 because each tests a different hypothesis.
@@
 		% fisher.test(rbind(c(11,10),c(0,16)))$p.value --> 0.0006143355
-		Exit UAF  & 11	& 10	& 0	& 16	& $+52.4\%$	& 0.00061	& 0.0125 \\
+		Exit UAF  & 11	& 10	& 0	& 16	& $+52.4$\%	& 0.00061	& 0.0125 \\
 		% I spent more hours than I should have looking at this landslide-user-submitted-bug group.
@@
 		% but didn't rerun mutex_test to check their update.
-		Mutex     & 17	& 1	& 19	& 0	& $-05.6\%$	& 0.49 & 0.0125 \\
-		Paraguay  & 20	& 1	& 14	& 2	& $+07.7\%$	& 0.57	& 0.0125 \\
+		Mutex     & 17	& 1	& 19	& 0	& $-05.6$\%	& 0.49 & 0.0125 \\
+		Paraguay  & 20	& 1	& 14	& 2	& $+07.7$\%	& 0.57	& 0.0125 \\
 		% nb this is any downgrade related ding not just dgr tested ones
@@
 		% nice p value
-		Downgrade & 10	& 3	& 19	& 4	& $-05.7\%$	& 0.69	& 0.0125 \\ % TODO verify these buggers by eye
+		Downgrade & 10	& 3	& 19	& 4	& $-05.7$\%	& 0.69	& 0.0125 \\ % TODO verify these buggers by eye
 		\hline
-		\bf Total & 58	& 15	& 52	& 22	& $+09.2\%$	& 0.26 & 0.05 \\
+		\bf Total & 58	& 15	& 52	& 22	& $+09.2$\%	& 0.26 & 0.05 \\
 	\end{tabular}
@@
 \subsubsection{Reasons worthwhile}
+\label{sec:education-reasons-worthwhile}
 
@@
 		in high-level undergraduate classes
-		(with many slides admittedly specific to Landslide as they may be)
+		(with many slides admittedly specific to Landslide)
 		(\cref{sec:education-pebbles-recruiting}).

### 2018-12-19 thesis/intro.tex
slight revise concl
daves final revisions to intro

@@
 but would at least have unlocked properly at the first turn of the key later.
-This violation of expected behaviour -- that the key need not be turned twice --
-is a {\em bug},
+One would reasonably expect to never have to turn the key more than once,
+so this violation of expected behaviour is a {\em bug},
+% more ambiguous version pointed out by garth
+%This violation of expected behaviour -- that the key need not be turned twice -- is a {\em bug},
 and its dependence on the two travelers' actions {\em interleaving}
@@
 \section{Contributions}
+\label{sec:thestatement}
 
@@
 My thesis statement is as follows:}
+
+\vspace{0.75em}
 
@@
 using an algorithm that knows its own limits,
-and compensates for them with heuristics informed by user's own human intuition.
-The second half of the statement can be broken down into three parts,
+and compensates for them with heuristics informed by the user's own human intuition.
+The second half of the statement can be broken down into three parts:
 coping with exponential explosion,
 helping students,
-and addressing modern concurrency models,
+and addressing modern concurrency models;
 which correspond to the three major contributions of this thesis, as follows:}
@@
 		and at The Pennsylvania State University,
-		which has recently adopted CMU's thread library project,}
+		which recently adopted CMU's thread library project,}
 		to provide
@@
 although some of the research being theoretical and several layers of abstractions deep,
-I cannot promise all easy reading.
+I cannot promise easy reading to all.
 \Cref{chap:background}, Background, provides what I hope are friendly concrete examples
@@
 who would also prefer them for any third-person reference by you.
+
+\subsubsection{Fonts}
+
+The body of this thesis is set in Bitstream Charter,
+the monospace font is Inconsolata,
+the epigraph font is Alexa, %Std.
+the Japanese font is Heisei Mincho, %Std W5.
+and
+the Arabic font is Amiri. %Regular.
+%The SIGBOVIK font is Birbaslo.
 

### 2018-12-19 thesis/landslide.tex
quicksand impl
proofread text wrapping, etc in landslide

@@
 						emit a preemption trace
-						(\cref{sec:landslide-foundabug}, \cref{sec:landslide-bugreport})
+						(\cref{sec:landslide-bugreport},
+						\cref{sec:landslide-foundabug})
 						and halt Landslide's execution.
@@
 						and further for data races (\cref{sec:landslide-datarace}).
-					\item Select which thread should run next (\cref{sec:landslide-arbiter}),
+					\item Select which thread should run next (\cref{sec:landslide-arbiter}).
 						Heuristically detecting yield-blocked threads
@@
 					\item Force the chosen thread to run by injecting timer interrupts
-						(\cref{sec:landslide-interrupce},
-						\cref{sec:landslide-scheduler-interrupce}).
+						(\cref{sec:landslide-scheduler-interrupce},
+						\cref{sec:landslide-interrupce}).
 				\end{enumerate}
@@
 				force the new thread to run
-				(\cref{sec:landslide-interrupce}, \cref{sec:landslide-scheduler-interrupce}),
+				(\cref{sec:landslide-scheduler-interrupce},
+				\cref{sec:landslide-interrupce}),
 				and repeat from step 2.
@@
 		(\cref{sec:landslide-icb})
-	\item {\tt -0}: enable Preempt-Everywhere mode (\cref{sec:quicksand-eval}, requires {\tt -C})
+	\item {\tt -0}: enable Preempt-Everywhere mode
+		(\cref{sec:quicksand-impl-modes}, \cref{sec:quicksand-impl-landslide}, requires {\tt -C})
 	\item {\tt -M}: enable Maximal State Space mode,
@@
 		in timer interrupts injected during userspace execution.
-	\item {\tt splice\_pre\_vanish\_trace} ({\tt stack.c})
+	\item {\tt splice\_pre\_vanish\_trace()} ({\tt stack.c})
 		will, when a {\tt vanish}ing thread has already
@@
 		(saved previously in {\tt sched\_update\_kern\_state\_machine()})
-		so the user can see the userspace execution sequence preceding the {\tt vanish} invocation.
-	\item {\tt stack\_trace} ({\tt stack.c})
+		so that the user can see the userspace execution sequence preceding the {\tt vanish} invocation.
+	\item {\tt stack\_trace()} ({\tt stack.c})
 		will, when {\tt ebp} crosses from kernel- to userspace across a system call boundary
@@
 	\item User or kernel {\tt mutex\_lock()} or {\tt mutex\_unlock()} call
-	\item Custom kernel preemption point using {\tt tell\_landslide\_preempt()}
-		(relic of \cite{landslide}, largely obsoleted by data-race preemption pooints)
+	\item Custom preemption point requested by the user with {\tt tell\_landslide\_preempt()}
+		(relic of \cite{landslide}, largely obsoleted by data-race preemption points (\cref{sec:quicksand-dr-pps}))
 \end{itemize}
@@
 Landslide detects such ad-hoc {\em yield-loop blocking}
-by keeping a counter for each thread to track how many {\tt yield}s it has invoked since the last ``interesting'' activity.
+by keeping a counter for each thread to track how many {\tt yield}s it has invoked since the last interesting activity.
 % nb. not system calls, not even d/mr, as some student mutexes may use those and you don't want to count them

### 2018-12-19 thesis/overview.tex
daves final revisions to overview, as well as some of myown tweaks

@@
 \subsubsection{Shared state}
-When a programming language offers multithreaded parallelism but forbids access to any shared state between threads \cite{rust-book},
+When a programming language offers multithreaded parallelism but forbids access to any shared state between threads
+\cite{erlang,rust-book},
 the simultaneity of threads is largely irrelevant to the program's behaviour.
@@
 			\begin{tabular}{l}
-				\texttt{\ctype{int} x;} \\
+				\texttt{\ctype{int} x = \const{0};} \\
 				\texttt{\ctype{void} \call{count}() \{} \\
@@
 \begin{tabular}{rll}
+	\\
 	& \multicolumn{2}{c}{\texttt{\ctype{int} x = \const{0}; \ctype{bool} y = \const{false}; \ctype{mutex\_t} mx;}} \\
@@
 The other three (a stack-crawling backtrace utility, a bare-metal game with device drivers, and a small extension to the P3 kernel) are not of concern in this thesis.
-The course's prerequisite is 15-213, Introduction to Computer Systems \cite{csapp,sigcse01:CSaPP}.
+The course's prerequisite is 15-213, Introduction to Computer Systems \cite{sigcse01:CSaPP,csapp}.
 Both P2 and P3 are built using the {\em Pebbles} system call specification, outlined in \Cref{tab:syscalls}.
@@
 Students may use any x86 atomic instruction(s) they desire, such as {\tt xchg}, {\tt xadd}, or {\tt cmpxchg}, and/or the {\tt deschedule}/ {\tt make\_runnable} system calls offered by the reference kernel.
-	\item Condition variables, with the interface {\tt cond\_wait(cvp, mp)}, {\tt cond\_signal} {\tt (cvp)}, and {\tt cond\_broadcast(cvp)}.
-{\tt cond\_wait} blocks the invoking thread, ``simultaneously'' releasing a mutex which protects some associated state (atomically, with respect to other calls to signal or broadcast under that mutex).
-		{\tt cond\_signal} and {\tt cond\_broadcast} wake one or all waiting threads.
+	\item Condition variables, with the interface {\tt cond\_wait(cvp, mp)}, {\tt cond\_signal(cvp)}, and {\tt cond\_broadcast(cvp)}.
+{\tt cond\_wait()} blocks the invoking thread, ``simultaneously'' releasing a mutex which protects some associated state (atomically, with respect to other calls under the same mutex).
+		{\tt cond\_signal()} and {\tt cond\_broadcast()} wake one or all waiting threads.
 		Students must use the {\tt deschedule} and {\tt make\_runnable} system calls to implement blocking (busy-waiting is forbidden), and typically include an internal mutex to protect the condition variable's state as well.
@@
 		Students typically implement semaphores using mutexes and condition variables, not using atomic instructions or system calls directly.
-	\item Reader-writer locks (rwlocks), with the interface {\tt rwlock\_lock(rwp, mode)} and {\tt rwlock\_unlock(rwp)}. {\tt mode} may be either {\tt RWLOCK\_READ} or {\tt RWLOCK\_\allowbreak{}WRITE}.
+	\item Reader-writer locks (also R/W-locks or rwlocks), with the interface {\tt rwlock\_lock(rwp, mode)} and {\tt rwlock\_unlock(rwp)}.
+{\tt mode} may be either {\tt RWLOCK\_READ} or {\tt RWLOCK\_\allowbreak{}WRITE}.
 		Behaves as mutexes, but multiple readers may access the critical section simultaneously.
@@
 This requires less implementation complexity than the more featureful
-\revisionminor{{\tt rfork} model, of Plan 9 \cite{Pike90plan9}, or the {\tt clone} model, of Linux}.
+\revisionminor{{\tt rfork} model of Plan 9 \cite{Pike90plan9}, or the {\tt clone} model of Linux}.
 
@@
 		\revisionminor{Short for ``application programming interface'' (which nobody ever says).}
-		The set of function signatures which comprise the interface to a body of code.
+		A set of function signatures which comprise the interface to a body of code.
 		For example, the standard {\em mutex} API consists of
@@
 		must technically also be atomicity violations.
-		Also, this thesis is focused on identifying failures, not root causes,
+		Also, this thesis is focused on identifying failures,
+		leaving root cause diagnosis to the user,
+		%not root causes,
 		so will simply avoid the term entirely.
@@
 		An execution state among a set of N {\em threads},
-		where each thread is blocked waiting for an event that can be caused only by another thread in the set
+		where each is blocked waiting for an event that can be caused only by another thread in the set
 		\cite{deadlock}.
@@
 	\item {\bf pthreads.}
-		Thread library for Linux programs provided by POSIX.
+		POSIX thread library for Linux programs. %provided by POSIX.
 		\revision{The modern pthreads model traces its lineage back to
-		Mach's C threads \cite{mach-c-threads}
-		and DEC's Systems Research Center threads \cite{src-threads},
+		DEC's Systems Research Center threads \cite{src-threads}
+		and Mach's C threads \cite{mach-c-threads},
 		and is now specified by ISO/IEC/IEEE \cite{iso-posix}.}

### 2018-12-19 thesis/tm.tex
tm map basicer mutex 2,4
finish the last tm test
daves final revisions to tm

@@
 by which programmers may attempt an arbitrary sequence of shared memory accesses,
-which will either be committed (made visible to other processors/threads) atomically,
+which will either be entirely committed (i.e., made visible to other processors/threads) atomically,
 such that no intermediate state modification is ever visible,
 or, in the case of a conflict which would prevent such,
-discarded with an error code returned to allow the programmer to \revisionminor{invoke} a synchronized backup path.
+entirely discarded with an error code returned to allow the programmer to \revisionminor{invoke} a synchronized backup path.
 %
@@
 		with any other access to the same memory since the transaction started,
-		they are instead discarded and execution state reverts to the {\sf begin} with an error code as described above.
+		they are instead discarded and execution state reverts to the {\sf begin} point
+		with an error code as described above.
 	\item {\sf \revisionminor{abort}} explicitly aborts a transaction,
@@
 incorporating other nondeterminism sources:
-store buffer reordering arising from relaxed memory models
+store-buffer reordering arising from relaxed memory models
 \cite{cdschecker,tsopso,rcmc}
@@
 		{Impossible hypothetical interleaving within transactions of \Cref{fig:htm-example-assemblified}.
-		The effects of $\tiat{2a-2b}$ are never globally visible, as its transasction must abort,
+		The effects of $\tiat{2a-2b}$ are never globally visible, as its transaction must abort,
 		and moreover $\tiat{2c-3}$ will not execute at all.}
@@
 
-\Cref{thm:atom} must be weakened correspondingly.
+\Cref{thm:atom} must be weakened correspondingly, yielding \Cref{thm:atom-weak}.
 \vspace{1em}
@@
 			& 2,4 & \cpu{217.98} & \ints{4984} \\
+			& 2,5 & \cpu{3417.86} & \ints{76787} \\
 			& 3,1 & \cpu{8.26} & \ints{129} \\
 			& 3,2 & \cpu{1403.46} & \ints{30653} \\
-			& 3,3 & \ETAdag{qqq} & \ETAdag{qqq} \\
+			& 3,3 & \ETAdag{11d 22h} & \ETAdag{23136172} \\
+			% [JOB 0] progress: 764124/23136172 brs (3.302724%), ETA 11d 11h 45m 28s (elapsed 9h 59m 59s)
 			& 4,1 & \cpu{199.96} & \ints{4488} \\
-			& 4,2 & \ETAdag{qqq} & \ETAdag{qqq} \\
+			& 4,2 & \ETAdag{41d 0h} & \ETAdag{66520074} \\
+			% [JOB 0] progress: 702872/66520074 brs (1.056631%), ETA 40d 13h 44m 41s (elapsed 9h 59m 59s)
 			\cline{1-4}
@@
 			& 2,3 & \cpu{12946.21} & \ints{244691} \\
-			& 2,4 & \ETAdag{qqq} & \ETAdag{qqq} \\
+			& 2,4 & \ETAdag{2d 17h} & \ETAdag{1130184} \\
+			% [JOB 3] progress: 572848/1130184 brs (50.686221%), ETA 2d 6h 52m 20s (elapsed 9h 59m 55s)
 			& 3,1 & \cpu{637.81} & \ints{12707} \\
@@
 			% [JOB 3] progress: 601402/1778661 brs (33.812059%), ETA 23h 33m 16s (elapsed 9h 59m 55s)
-			\\
 		\end{tabular}
@@
 \begin{itemize}
-	\item Three equivalence proofs,
-		demonstrating that model checkers can soundly emulate strongly-atomic HTM execution semantics by
-		treating transactions as globally-uninterruptible atomic sections
+	\item An approach for model checkers to emulate strongly-atomic HTM
+		by treating transactions as globally-uninterruptible atomic sections
 		and injecting failure without simulating rollback
@@
 		(\cref{sec:tm-design-locks}),
-		and can emulate weakly-atomic STM systems by relaxing the above atomicity to apply to other transactions only
-		(\cref{sec:tm-weak-atomicity}).
+		plus an extension to also emulate weakly-atomic STM systems
+		by relaxing the above atomicity to apply to other transactions only
+		(\cref{sec:tm-weak-atomicity}),
+		each section with a corresponding soundness proof of equivalence to existing systems' semantics.
 	\item An implementation of both forms of transactional memory in Landslide,

### 2018-12-19 thesis/future.tex
tweak wording of dunk on machine learning in future

@@
 in unpredictable ways
-that can harm people of color \cite{conceptnet-bias,gender-shades},
+that have been shown to harm people of color \cite{conceptnet-bias,gender-shades},
 women \cite{gender-bias-ai,amazon-ai},

### 2018-12-19 thesis/thesis.tex
daves final revisions to acks and concl, change month, funding statemence

@@
 \author{Ben Blum}
-\date{October 2018}
+\date{December 2018}
 \Year{2018}
@@
 
-\support{}
+\support{This research was sponsored by the U.S. Army Research Office under grant
+number W911NF0910273 and by Intel ISTC-CC.
+The views and conclusions contained in this document are those of the author
+and should not be interpreted as representing the official policies,
+either expressed or implied, of any sponsoring institution, the U.S. government, or any other entity.}
 \disclaimer{}
@@
 despite my often stubborn refusal to leave my own little comfort zone.
-So, even when it took me sending 10+ email reminders to get your attention,
+So, despite the times I needed to send 10+ email reminders to get your attention:
 truly,
@@
 Professor Brandon Lucia provided valuable revisions to \Cref{chap:quicksand}'s content
-that ultimately led to its publication at OOPSLA 2016.
+	that ultimately led to its publication at \mbox{OOPSLA}~2016.
 Michael J. Sullivan double-checked the proofs therein.
@@
 Sol Boucher, and Michael J. Sullivan
-helped me rehearse and polish the snake fight talk.
+helped me rehearse and polish the snake-fight talk.
 
@@
 during my two summers as a Rust research intern at Mozilla,
-and to the rest of the Rust community as well for their %extremely important
-work
+and to the rest of the Rust community as well for
+%their extremely important work
 bringing better concurrent programs into the world. %from a different angle.
@@
 Rob Simmons,
-Tom Murphy 7,
+Tom 7,
 and
@@
 My parents, Eve and David Blum, and my grandparents, Margie Granach and Elsie Blum,
-provided bottomless emotional support throughout my education.
+provided bottomless emotional support throughout my career as a student. %education.
 I appreciate beyond words for the privileged life they provided
@@
 From those who stayed in Pittsburgh %with me
-to those who %moved away but
+to those who moved away but % can comment 'moved away but'
 kept in touch,
@@
 		combined with Quicksand's reliance on state space estimation
-		to effectively balance formal verification with heuristic bug-finding speed
+		effectively balanced formal verification with heuristic bug-finding speed
 		better than any prior work.
@@
 	\item In \Cref{chap:education},
-		the behind-the-scenes yield-loop blocking algorithm
-		(itself admittedly heuristic)
-		allowed Landslide to automatically instrument all student implementations
-		with very few false positives,
+		Landslide's behind-the-scenes yield-loop blocking algorithm
+		%(itself admittedly heuristic)
+		allowed it to automatically instrument all student implementations
+		with no false positives
+		(what false positives did exist were minor unrelated Landslide bugs,
+		all fixed quickly whenever reported),
 		and exposing partial results such as data race reports
@@
 		Although high-level undergraduate operating systems courses represent
-		a very small fraction of all the world's programmers,
-		let alone all those in higher education,
+		a very small fraction of all programmers in higher education,
+		let alone all those in the world,
 		Landslide's adventures in 15-410 argue that computer science classes
@@
 		allowed Landslide to avoid both performance overhead and state space explosion when simulating aborts,
-		and the heuristic use of abort code reduction and abstraction reduction
+		and heuristic use of abort code reduction and abstraction reduction
 		allowed verifying algorithms and data structures up to thread/iteration counts previously infeasible.

### 2018-12-20 thesis/landslide.tex
quicksand eval

@@
 %C files live in
-{\tt work/modules/landslide/} (Simics implemenation) or {\tt src/bochs-2.6.8/instrument/landslide/} (Bochs implementation)
+{\tt work/modules/landslide/} (Simics implementation) or {\tt src/bochs-2.6.8/instrument/landslide/} (Bochs implementation)
 relative to the repository root.
@@
 \subsubsection{Memory conflicts}
+\label{sec:landslide-dpor-conflix}
 

### 2018-12-20 thesis/quicksand.tex
quicksand eval

@@
 in addition to the other requirements of Happens-Before.
+If the generations do not match,
+Landslide sets the {\tt free\_re\_malloc} flag in the messaging protocol to Quicksand.
+If the race is later observed in a reordering which avoids the reallocation pattern
+(such as in \Cref{fig:recycle-bug}),
+Landslide will report it as normal,
+and Quicksand will promote it to a normal preemption point in the registry
+({\tt pp\_new()} in {\tt pp.c}, ``{\tt for realsies}'' case).
+Also included in this message is a flag to indicate whether a data race was found
+nondeterministically (i.e., not on the first interleaving), such as described in \cref{sec:quicksand-pps}.
 
@@
 			\item Does Quicksand avoid false positives compared to single-execution Limited Happens-Before?
-			\item Does Quicksand find data-race bugs that single-execution Pure or Limited Happens-Before would miss?
+			\item Does Quicksand find data-race bugs that single-execution Pure Happens-Before or Limited Happens-Before alone would miss?
 		\end{enumerate}
@@
 
-\newcommand\mxtest{\texttt{mx\_test}}
-\newcommand\tej{\texttt{join\_test}}
-\newcommand\bct{\texttt{bcast\_test}}
-\newcommand\paraguay{\texttt{signal\_test}}
-\newcommand\paradise{\texttt{sem\_test}}
-\newcommand\rwldgr{\texttt{rwlock\_test}}
-\newcommand\prisema{\texttt{sched\_test}}
-\newcommand\waitsimple{\texttt{wait\_test}}
-\newcommand\alarmsimul{\texttt{alarm\_test}}
-
-I tested P2s with 6 multithreaded programs:
+\newcommand\mxtest{\texttt{mutex\_test}\xspace}
+\newcommand\tej{\texttt{thr\_exit\_join}\xspace}
+\newcommand\bct{\texttt{broadcast\_test}\xspace}
+\newcommand\paraguay{\texttt{paraguay}\xspace}
+\newcommand\paradise{\texttt{paradise\_lost}\xspace}
+\newcommand\rwldgr{\texttt{rwlock\_downgrade\_read\_test}\xspace}
+\newcommand\prisema{\texttt{priority-sema}\xspace}
+\newcommand\waitsimple{\texttt{wait-test}\xspace}
+\newcommand\alarmsimul{\texttt{alarm-simultaneous}\xspace}
+
+I tested P2s with six multithreaded programs:
 \mxtest, for locking algorithm correctness,
 \tej, a test of thread lifecycle,
-\bct~and \paraguay~for condition variables,
-\paradise~for semaphores,
+\bct and \paraguay for condition variables,
+\paradise for semaphores,
 % nb: there used to be a "f u latex" comment here with a \\ after dgr, prob for margins
-and \rwldgr for r/w locks.
+and \rwldgr for R/W locks.
 These are the same tests I distributed Landslide with in \Cref{chap:education};
@@
 For \mxtest, \paradise, and \paraguay,
-I used the {\tt without\_function} command to exclude {\tt thr\_create}, {\tt thr\_exit}, and {\tt thr\_join} preemption points,
+I used the {\tt without\_function} command to exclude
+{\tt thr\_create()}, {\tt thr\_exit()}, and {\tt thr\_join()} preemption points,
 and for \mxtest I enabled {\tt TESTING\_MUTEXES} (\cref{sec:landslide-dynamicconfig}).
-I tested Pintoses with 3 programs from the class's provided test suite:
+I tested Pintoses with three programs from the class's provided test suite:
 \prisema, a test of the kernel scheduling algorithm,
@@
 For all tests,
-I also excluded preemption points on {\tt malloc}'s internal lock using {\tt without\_function}.
+I also excluded preemption points on {\tt malloc()}'s internal lock using {\tt without\_function}.
 In total, the evaluation comprises 629 unique tests (i.e., pairs of a test program and a Pintos or P2),
@@
 \subsubsection{Experimental setup}
+\label{sec:quicksand-expt-setup}
 
@@
 	\item {\bf QS-Pure-HB}: Quicksand using Pure Happens-Before instead ({\tt -V}), and
-	\item {\bf QS-Sync-Only}: Quicksand with initial preemption points only \cref{sec:quicksand-initial-pps},
+	\item {\bf QS-Sync-Only}: Quicksand with initial preemption points only,
+		as described in \cref{sec:quicksand-initial-pps},
 		but never adding new ones from reported data races.
@@
         \item {\bf SSS-MC-Shared-Mem}:
-		Using ICB+BPOR, configured to preempt on any shared memory access ({\tt -0}).
+		Using ICB+BPOR, configured to preempt on any shared memory access ({\tt -0})
                 (decided at runtime, excluding threads' accesses to their own stacks),
@@
 but it remains an open research problem how to extend the algorithm to ICB.
-Hence, I optimistically gave control experiments a linear speedup of 10 hours per test with 1 CPU;
+Hence, I optimistically gave all control experiments a linear speedup of 10 hours per test with 1 CPU;
 i.e., assuming it could parallelize with 100\% efficiency.
 To match this,
-Quicksand reports the CPU-time spent in addition to the wall-clock time for a resource-fair comparison,
-although Quicksand's inherent parallelism is a convenient benefit unto itself.
+Quicksand reports both the CPU-time and wall-clock time spent during its execution.
+Comparing CPU-time leads to a more fair comparison,
+although Quicksand's inherent parallelism, which only a wall-clock time comparison would show,
+is also a convenient benefit unto itself.
 All tests ran on 12-core 3.2 GHz Xeon W3670 machines with 12GB of RAM.
 
-\subsection{Finding bugs}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\subsection{Bug-finding}
 \label{sec:quicksand-eval-bugs}
+
+\Cref{fig:dowefindbugsfaster}
+plots the bug-finding performance of Quicksand's three experimental trials
+against the three control approaches
+in a cumulative distribution of total bugs found against elapsed time.
+The farthest-right point on each series indicates in how many total test cases that trial found a bug
+after the 10 CPU-hour timeout.
+\Cref{fig:dowefindbugsfaster}(a)
+is the main, resource-fair comparison by CPU-time,
+while \Cref{fig:dowefindbugsfaster}(b) shows
+the wall-clock time until each bug was found.
+
+\begin{figure}[p]
+	\begin{tabular}{c}
+        \includegraphics[width=0.71\textwidth]{../proposal/dowefindbugsfaster-v2.pdf} \\
+		\begin{tabular}{p{\textwidth}}
+                (a) Bugs found by Quicksand versus control experiments, measured in CPU time.
+                %as a function of elapsed CPU time.
+                Overall,
+                a more resource-fair comparison than (b),
+                although Quicksand's start-up overhead is exaggerated, as the SSS-MC tests are not parallelized. \\
+		\end{tabular}
+                \\
+                \\
+        \includegraphics[width=0.71\textwidth]{../proposal/dowefindbugsfaster-wallclock-v2.pdf} \\
+                %(b) Bugs found by elapsed wall-clock time.
+		\begin{tabular}{p{\textwidth}}
+                (b) Bugs found by Quicksand versus control experiments, measured in wall-clock time.
+                Quicksand is parallelized tenfold; the vertical line indicates its 1 hour limit. \\
+		\end{tabular}
+        \end{tabular}
+	\caption[Bug-finding performance comparison
+	between Quicksand and prior work.]
+	{Bug-finding performance comparison between
+        by several configurations of Quicksand and the single-state-space approaches.
+	Quicksand finds 125\% as many bugs
+        with data-race PPs
+	at the 10-hour mark, compared to the best prior work approach.}
+        \label{fig:dowefindbugsfaster}
+\end{figure}
+
+\subsubsection{Performance comparison}
+
+Compared to SSS-MC-ICB (the fastest among the control experiments),
+Quicksand finds more bugs within any fixed CPU budget greater than 200 seconds.
+In other words, draw a vertical line at $x=N$ for any $N>200$ to represent timing out each test after $y$ seconds elapsed,
+and Quicksand's bug total will exceed that of ICB.
+SSS-MC-Shared-Mem initially suffers a substantial performance penalty from the sheer number of preemption points it must analyze,
+but ultimately outstrips SSS-MC-ICB, which fundamentally cannot find data-race bugs,
+after 135 CPU-minutes with its 100th bug found,
+ultimately finishing the 10 CPU-hours as the best prior work approach in the long term.
+Compared to SSS-MC-Shared-Mem, Quicksand's Limited HB version
+finishes with 125\% as many bugs in total.
+
+Regarding Quicksand's tenfold parallelism,
+before the break-even point at 200 seconds,
+it lags behind SSS-MC-ICB due to the additional start-up overhead
+of testing many state spaces at once even though the easy bugs may be found extremely quickly in any of them.
+However, converting ICB's early CPU-time advantage
+into faster wall-clock performance remains an open research problem \cite{parallel-dpor}.
+\Cref{fig:dowefindbugsfaster}(b) gives Quicksand full credit for its inherent parallelism,
+which ICB cannot yet practically match:
+with a processor allocation of 10 CPUs, it outperforms all prior work approaches
+for any fixed budget of wall-clock time
+(i.e., comparing across a vertical line at $x=N$ for any $N$).
+% After 1 hour of wall-clock time, tenfold Quicksand performs 158\% as well as SSS-MC-ICB. % totally meaningless
+
+The QS-Sync-Only experiment tests whether Iterative Deepening would be effective
+even for model checking domains without data races,
+such as distributed systems \cite{macemc,modist,samc,dbug-retreat,concuerror}
+and programming languages whose type systems statically reject concurrent mutable shared state
+\cite{erlang,haskell,rust-book}.
+When Quicksand ignores all data-race candidates,
+its results are competitive with SSS-MC-DPOR, although SSS-MC-ICB outperforms it slightly.
+This is unsurprising: the seed subsets of preemption points
+that QS-Sync-Only is limited to
+(\cref{sec:quicksand-initial-pps})
+are much less flexible than ICB's preemption strategy.
+This result suggests that in future work,
+Quicksand should consider using ICB in parallel with its default configuration when it finds no data-race candidates to test.
+I discuss this possibility further in \cref{sec:warpzone-heuristics}.
+
+On the other hand,
+comparing QS-Limited-HB to SSS-MC-Shared-Mem
+shows that Iterative Deepening thoroughly outperforms ICB when shared-memory preemptions come into play.
+%We attribute this to the fact that
+Statically configuring a preemption point for every shared memory access in advance
+produces orders of magnitude more points than
+waiting for an access to be identified as part of a data race at runtime.
+%
+In principle, DPOR and ICB+BPOR should suffice to identify and prune any equivalent thread interleavings
+arising from extraneous preemption points on non-conflicting accesses.
+However, in practice,
+%we found that
+the sheer number of accesses during each new execution
+(often thousands)
+added significant performance overhead to
+%the MC when computing DPOR and backtracking.
+DPOR's $O(n^2)$ memory independence computation (\cref{sec:landslide-dpor-conflix}),
+as well as the $O(n)$ overhead of checkpointing the execution state at each preemption point (\cref{sec:landslide-timetravel}).
+Iterative Deepening avoids this overhead by waiting until runtime
+to identify fewer, more relevant preemption points dynamically,
+and is hence more suitable for model checking when data races are involved.
+
+\subsubsection{Types of bugs}
+
+\Cref{tab:drbugs} provides more detail on each of the bugs shown in \Cref{fig:dowefindbugsfaster},
+broken down by test case.
+The left half shows the number found by each experimental approach,
+with the totals of each column corresponding to the values at $x=10$ hours in \Cref{fig:dowefindbugsfaster}.
+In \mxtest, which checks the lock implementation for correctly providing mutual exclusion
+(rather than trusting its correctness, as all other tests do),
+SSS-MC-ICB and SSS-MC-DPOR
+found dramatically fewer bugs (just 1).
+Prior work has proposed {\em abstraction reduction} \cite{dbug-phdthesis},
+in which verifying correctness properties of synchronization primitives
+allows subsequently trusting them in other tests which use them to mitigate state space explosion;
+\cref{sec:tm-abstraction} will explore this technique further.
+By contrast, QS-Limited-HB found 10 mutex bugs, and SSS-MC-Shared-Mem found 12.
+In the scope of this chapter,
+this serves as strong evidence that new low-level synchronization code must be verified with data-race preemption points,
+whether combined with Iterative Deepening or ICB.
+
+\begin{table}[t]
+	\begin{center}
+	\small
+	\begin{tabular}{r|c||c|c||c|c|c}
+		%\multicolumn{2}{c||}{} & \multicolumn{5}{c||}{\bf {Total bugs}} & \multicolumn{4}{c||}{\bf {Data-race bugs}} \\
+		% ---
+		& {\bf Num.} & \multicolumn{2}{c||}{\bf Quicksand} & \multicolumn{3}{c}{\bf {Single-state-space MC}}
+		%& \multicolumn{2}{c|}{\bf {Limited HB}} & \multicolumn{2}{c||}{\bf {Pure HB}} &
+		%{\bf Mutual} & {\bf Avg. tested}
+		\\
+		% ---
+		{\bf Test} & {\bf tested} & {\bf LHB} & {\bf PHB} & {\bf ICB} & {\bf DPOR} & {\bf ShMem}
+		%& {\bf All} & {\bf Nondet.} & {\bf {All}} & {\bf {Nondet.}}
+		%& {\bf timeouts} & {\bf subset SSes}
+		\\
+		% ---
+		\hline
+		%         #test qs-lhb qs-phb icb dpor every dronly nondets(LHB) dronly(PHB) nondets(LHB) mutual-TO comp.SSes
+		{\tt broadcast\_test}    & 79  & 8   & 8   & 5   & 6   & 7   \\ % & 2 & 1 & {2} & {1} & 7    & 112.3 \\
+		{\tt thr\_exit\_join}    & 79  & 23  & 20  & 13  & 13  & 14  \\ % & 11& 4 & {7} & {3} & {12} & {69.7} \\
+		{\tt mutex\_test}        & 79  & 10  & 9   & 1   & 1   & 12  \\ % & 9 & 1 & {8} & {1} & 0    & -     \\
+		{\tt paradise\_lost}     & 79  & 17  & 16  & 12  & 11  & 12  \\ % & 7 & 3 & {6} & {2} & {50} & {77.4} \\
+		{\tt paraguay}           & 79  & 10  & 8   & 5   & 5   & 11  \\ % & 6 & 1 & {3} & {2} & 45   & {59.6} \\
+		{\tt rwlock\_downgrade}  & 79  & 27  & 26  & 25  & 23  & 28  \\ % & 4 & 1 & {3} & {0} & {44} & {86.3} \\
+		\hline
+		{\tt priority-sema}      & 59  & 7   & 7   & 1   & 1   & 8   \\ % & 6 & 4 & {6} & {6} & 2    & 13.0  \\
+		{\tt alarm-simultaneous} & 44  & 21  & 12  & 16  & 5   & 29  \\ % & 17& 1 & {7} & {6} & {17} & {7.8}  \\
+		{\tt wait-simple}        & 52  & 30  & 26  & 24  & 23  & 1   \\ % & 7 & 2 & {2} & {0} & {15} & {33.8} \\
+		\hline
+		{\bf Total} & 629 & 153 & 132 & 102 & 88  & 122 \\ % 69& 15& {44}& {21}& {192}& {65.8} \\
+	\end{tabular}
+	\end{center}
+	\caption[Summary of bugs found by each test program.]
+	{Summary of bugs
+	%and data races
+	found by each test program.
+	QS-LHB and QS-PHB are Quicksand; ICB/DPOR/ShMem are the controls (\cref{sec:quicksand-expt-setup}).
+	}
+	\label{tab:drbugs}
+\end{table}
+
+To ensure that the corpus of P2 and Pintos bugs gives an unbiased comparison between Quicksand and ICB,
+I also counted the preemption bounds at which ICB found each of its bugs,
+i.e., the minimum number of involuntary thread switches each bug required to expose.
+Table~\ref{tab:icb-bounds} shows the distribution of these bounds,
+which is consistent with the results of \cite[Table 2]{chess-icb},
+reproduced in the rightmost column
+(obtained under a different test suite, of course, of only 5 programs).
+This shows no bias towards bugs that would be harder for ICB to find.
+In fact, this evaluation's preemption bound distribution is {\em more} heavily biased towards fewer preemptions,
+suggesting that if anything,
+my test suite is even friendlier still to ICB than that of prior work.
+
+\begin{table}[h]
+	\begin{center}
+		\small
+	\begin{tabular}{r||c|c||c}
+		{\bf Bound} & {\bf SSS-MC-ICB} & {\bf SSS-MC-Shared-Mem} & Prior work \cite{chess-icb} \\
+		\hline
+		0         & 2     & 1     & 3 \\
+		1         & 82    & 86    & 7 \\
+		2         & 16    & 32    & 5 \\
+		3         & 2     & 3     & 1 \\
+		4+        & 0     & 0     & 0 \\
+		\hline
+		\bf Total & 102   & 122   & 16 \\
+	\end{tabular}
+	\end{center}
+	\caption[Distribution preemption bounds among bugs found by ICB.]
+	{Distribution of preemption bounds among bugs found by ICB control experiments.
+	Bound 0 means the bug was found by switching threads only on {\tt yield} calls.}
+	\label{tab:icb-bounds}
+\end{table}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
-\subsection{Nondeterministic data races}
+% TODO: put some txt here
+
+\subsubsection{Full verification}
+
+\Cref{fig:totalverif} plots the cumulative distribution of total verifications provided by each approach,
+in the same style of graph as \Cref{fig:dowefindbugsfaster}.
+For 167 of the 629 tests,
+QS-Pure-HB was able to reach and complete the maximal state space with no bugs found,
+hence providing the total verification guarantee justified by the proofs in \cref{sec:quicksand-soundness}.
+QS-Limited-HB completed a verification for 153 of 629 tests,
+slightly slower on account of Limited Happens-Before's higher false positive rate.
+%(which Quicksand must confirm each time by running an extra instance of Landslide). % explained below
+The next best approach for verifications was SSS-MC-Shared-Mem, which completed its search in only 39 cases.
+
+\begin{figure}[t]
+	\begin{center}
+	\includegraphics[width=0.71\textwidth]{../proposal/totalverifs-v2.pdf}
+	\end{center}
+	\caption[Verification performance comparison
+	between Quicksand and prior work.]
+	{Verification performance comparison between
+	Quicksand and single-state-space approaches.
+	Among the latter, only SSS-MC-Shared-Mem is theoretically capable of verifying any test with data races;
+	the others' series include only tests with no data races whatsoever,
+	in which case synchronization preemption points alone suffice for a full verification.}
+	\label{fig:totalverif}
+\end{figure}
+
+Ultimately, using Limited Happens-Before for finding data-race candidates allowed Quicksand to find more bugs,
+while Pure Happens-Before allowed for reaching full verification faster.
+I attribute this trade-off to the fact that
+Limited Happens-Before need not wait to test many alternate thread interleavings before
+finding a data-race candidate to begin with;
+%a potential data-race candidate is confirmed;
+rather, it can add new jobs to start testing potential races immediately.%
+\footnote{Upcoming, \Cref{tab:drstatistix} will corroborate this conclusion:
+the difference is most dramatic in {\tt alarm-\allowbreak{}simultaneous},
+the test where Quicksand struggled most to finish even small subset jobs.}
+%
+On the other hand, Limited Happens-Before can get overwhelmed by too many false positives,
+needing to refute such candidates by testing new state spaces for each one,
+while Pure Happens-Before can refute false positives {\em en passant}
+by testing alternate interleavings in its original state spaces.
+%In both cases, the performance differences seem to become significant only after about 10 minutes of CPU time,
+%which suggests... I don't actually know!
+This suggests that
+%each approach has merit, and that
+model checkers which incorporate data-race analysis should implement both modes
+and offer the user to choose based on their desired and/or expected testing outcome.
+
+The only testing modes which are theoretically capable of verifying any test with data races
+were QS-Pure-HB, QS-Limited-HB, and SSS-MC-Shared-Mem (i.e., preempt-everywhere mode).
+When QS-Sync-Only, SSS-MC-DPOR, and SSS-MC-ICB
+complete their respective maximal state spaces (i.e., all synchronization preemption points),
+that constitutes a full verification only in the case where no data races were identified at all,
+meaning Iterative Deepening would search no deeper than that anyway.
+Therefore, in \Cref{fig:totalverif},
+the data series for these latter three configurations
+represent only completed tests with no data races.
+Even though SSS-MC-Shared-Mem tends to hang out in the same neighbourhood as them,
+note that SSS-MC-Shared-Mem is still steadily increasing in verifications provided at the 10-hour cutoff
+(let alone the Quicksand ones),
+while the other three seem to reach a plateau of around 20-30 tests relatively soon.
+
+Overall, including data-race preemption points increases verification capacity by 4.25x.
+Assuming sequentially-consistent hardware,
+QS-Pure-HB classified many true data races as benign,
+while the SSS-MC-ICB approach could at best report such races to the user.
+This graph's results show that code written in a natural environment by inexpert users (students)
+generally does not obey the sort of strict coding discipline necessary
+for a model checker to make simplifying assumptions such as ``no data races'',
+justifying this chapter's claim that data-race preemption points are essential to model checking.
+
+\subsubsection{Partial verification}
+
+When a model checking job times out,
+the user would more likely prefer a summary of what parts of the test were verified
+rather than to write off all the CPU time spent as wasted.
+Prior work has proposed Probabilistic Concurrency Testing %(PCT)
+\cite{randomized-scheduler},
+which attempts to quantify the probability
+that a bug may remain in some untested interleaving after a time-out,
+although is opaque to the user about which parts of her code were actually tested.
+Meanwhile, ICB proposes to inform the user
+of the maximum number of preemptions that were used to test any individual interleaving,
+under the assurance that most bugs are likely to be found with fewer preemptions (\cref{tab:icb-bounds}).
+Quicksand instead reports which subsets of preemption points resulted in state spaces that did complete in time,
+in hopes that the user can supplement such a result with her own intuition
+by inspecting the code corresponding to the preemption points not tested (especially data races).
+From prior work, Preemption Sealing \cite{sealing}
+has argued the value of similar {\em compositional testing} when full verification is intractable,
+deferring to the user's expertise to judge the value of each subset of preemption points verified.
+\Cref{tab:partialverifs} shows Quicksand's partial verification results on timed-out tests.
+
+\begin{table}[h]
+	\begin{center}
+	\small
+	\begin{tabular}{r|c||c|c}
+		% ---
+		& {\bf Num.} & {\bf Mutual} & {\bf Avg. tested} \\
+		{\bf Test} & {\bf tested} & {\bf timeouts} & {\bf subset SSes} \\
+		% ---
+		\hline
+		%                        #test mutual-TO comp.SSes
+		{\tt broadcast\_test}    & 79  & 7    & 112.3 \\
+		{\tt thr\_exit\_join}    & 79  & {12} & {69.7} \\
+		{\tt mutex\_test}        & 79  & 0    & -     \\
+		{\tt paradise\_lost}     & 79  & {50} & {77.4} \\
+		{\tt paraguay}           & 79  & 45   & {59.6} \\
+		{\tt rwlock\_downgrade}  & 79  & {44} & {86.3} \\
+		\hline
+		{\tt priority-sema}      & 59  & 2    & 13.0  \\
+		{\tt alarm-simultaneous} & 44  & {17} & {7.8}  \\
+		{\tt wait-simple}        & 52  & {15} & {33.8} \\
+		\hline
+		{\bf Total}              & 629 & {192}& {65.8} \\
+	\end{tabular}
+	\end{center}
+	\caption[Summary of partial verification results on timed-out tests.]
+	{Summary of partial verification results on timed-out tests.
+	``Mutual timeouts'' counts how often both QS-Limited-HB and SSS-MC-ICB
+	(the best bug-finding approach from each group)
+	timed out with no bug found.
+	Among those, ``Average tested subset SSes'' counts how many partial verifications
+	QS-Limited-HB provided on average for each test. %(\sect{\ref{sec:eval-sssmc}}). % ???
+	}
+	\label{tab:partialverifs}
+\end{table}
+
+On 229 tests, SSS-MC-ICB timed out after 10 hours with no bugs found. % thesis note - pretty sure this *isn't* preempt-everywhere
+Among these tests, QS-Limited-HB % thesis note -- i think that's what past-me meant. they just wrote, 'quicksand' but.
+found bugs in 37.
+The other 192 represent cases where neither Quicksand nor ICB were able to provide a conclusive result either way.
+\footnote{Thesis note: SSS-MC-Shared-Mem was added in a subsequent revision to \cite{quicksand}
+from when this analysis was conducted, at which time SSS-MC-ICB was the best-performing approach among control experiments.
+Nevertheless, mutual timeouts among all six testing approaches constituted roughly one third of the test suite.}
+For these 192,
+I show the number of state spaces Quicksand was able to complete in the
+``Average tested subset SSes'' column.
+% of Table~\ref{tab:drbugs}. % thesis - now split into sep table
+%While obviously not as strong as full verification,
+These completions guarantee that, if the test program could expose a bug,
+it would depend on a data race not discovered yet,
+or be reachable only under a superset combination of preemption points not yet reached.
+
+On the other hand, attempting to maximize the raw number of completed state spaces
+may not be the most user-friendly way to present partial verification results.
+For example, included in these numbers are small state spaces which are subsets of other state spaces also completed;
+the user need not examine both subset and superset separately.
+Future work could conduct user studies to determine the most effective way to present these partial results
+from a software development perspective,
+such as described in \cref{sec:future-friendly}.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\subsection{Data race analysis}
 \label{sec:quicksand-eval-nondets}
+
+Beyond finding new bugs and completing full verifications with data-race PPs,
+I evaluated Quicksand's performance for classifying data-race candidates in two ways:
+its ability to check nondeterministic data races not reachable under a single-pass analysis (\cref{sec:quicksand-pps})
+and its ability to suppress reallocation false positives (\cref{sec:quicksand-id-realloc}).
+\Cref{tab:drstatistix} presents the results for this section.
+
+\begin{table}[t]
+	\begin{center}
+	\footnotesize
+	\begin{tabular}{r|c||c|c|c|c||c|c|c||c}
+		\multicolumn{2}{c||}{} & \multicolumn{4}{c||}{\bf {Data-race bugs}}
+		& \multicolumn{3}{c||}{\bf {Verifications}} \\
+		% ---
+		& {\bf Num.}
+		%& \multicolumn{2}{c|}{\bf Quicksand} & \multicolumn{3}{c||}{\bf {State-of-the-art}}
+		& \multicolumn{2}{c|}{\bf {Limited HB}}
+		& \multicolumn{2}{c||}{\bf {Pure HB}}
+		& \multicolumn{3}{c||}{\bf {Pure HB}}
+		& \multicolumn{1}{c}{\bf {Realloc.}}
+		\\
+		% ---
+		{\bf Test} & {\bf tested}
+		%& {\bf LHB} & {\bf PHB} & {\bf ICB} & {\bf DPOR} & {\bf ShMem}
+		& {\bf All} & {\bf N.D.} & {\bf {All}} & {\bf {N.D.}}
+		& {\bf DR PPs} & {\bf Benign} & {\bf Untested} & {\bf FPs} \\
+		% ---
+		\hline
+		%                      #test dronly nond dronly nond
+		%                                                   drpps benign untest FRMs
+		{\tt broadcast\_test}    & 79  & 2 & 1 & {2} & {1} & 655  & 97  & 150 & 52  \\
+		{\tt thr\_exit\_join}    & 79  & 11& 4 & {7} & {3} & 566  & 68  & 249 & 338 \\
+		{\tt mutex\_test}        & 79  & 9 & 1 & {8} & {1} & 911  & 127 & 44  & 7   \\
+		{\tt paradise\_lost}     & 79  & 7 & 3 & {6} & {2} & 783  & 2   & 414 & 166 \\
+		{\tt paraguay}           & 79  & 6 & 1 & {3} & {2} & 936  & 9   & 510 & 180 \\
+		{\tt rwlock\_downgrade}  & 79  & 4 & 1 & {3} & {0} & 543  & 1   & 310 & 156 \\
+		\hline
+		{\tt priority-sema}      & 59  & 6 & 4 & {6} & {6} & 65   & 51  & 3   & 0   \\
+		{\tt alarm-simultaneous} & 44  & 17& 1 & {7} & {6} & 35   & 0   & 29  & 35  \\
+		{\tt wait-simple}        & 52  & 7 & 2 & {2} & {0} & 71   & 1   & 28  & 31  \\
+		\hline
+		{\bf Total}              & 629 & 69& 15& {44}& {21}& 4565 & 356 & 1737& 965 \\
+	\end{tabular}
+	\end{center}
+	\caption[Data race statistics among Quicksand experiments.]
+	{Data race statistics among Quicksand experiments.
+	``Data-race bugs'' counts, among Quicksand's bugs, how many required data-race preemption points to expose;
+	among those, the ``N.D.'' (``nondeterministic'') columns show how many candidates
+	required model checking to identify
+	in the first place
+	(\cref{sec:quicksand-eval-nondets}).
+	%
+	``Total DR PPs'' counts how many unique data-racing instructions
+	QS-Pure-HB identified among tests where it found no bugs.
+	Among those, ``Benign DRs'' counts how many were refuted as non-failing,
+	while ``Untested DR PPs'' counts how many could not be checked in the time limit.
+	Finally, ``Realloc. FPs'' counts how many reallocation false positives QS-Limited-HB suppressed.
+	}
+	\label{tab:drstatistix}
+\end{table}
+
+\subsubsection{Nondeterministic data races}
+
+Some memory accesses may be hidden in a control flow path that requires a nondeterministic preemption to be executed
+(\cref{sec:quicksand-pps}).
+In such cases, a single-pass dynamic data-race detector
+might not achieve the coverage necessary
+to identify a racing access pair as a candidate at all,
+let alone check the resulting behaviour with such as Landslide.
+I instrumented Landslide to report these to Quicksand
+and counted how many such led to Quicksand finding new bugs when used as preemption points.
+Such bugs could be considered {\em false negatives} of the single-pass approach.
+The left half of \Cref{tab:drstatistix}
+breaks down the types of bugs
+found in each test case,
+showing both the total number of data-race bugs
+and the number among those that required such nondeterministic data races to expose.
+% I denote these data race reports as {\em nondeterministic},
+% while those that could be found on the first interleaving .....
+%
+To ensure a fair comparison, I disabled Quicksand's reallocation false positive suppression
+(\cref{sec:quicksand-id-realloc}, itself evaluated in the next section)
+for this experiment.
+This prevents Landslide from suppressing an observed reallocation data-race candidate on the first interleaving,
+which would falsely classify it as nondeterministic,
+even though a single-pass would not (indeed, should not) suppress such candidates.
+
+\Cref{fig:dr-falsenegs}
+%compares the types of data-race candidates necessary to expose each data-race bug in the test suite.
+visualizes the difference between single-pass and model-checking-enabled data race analysis.
+The first and third series represent the bugs found using preemption points from single-pass data-race candidates only,
+% not entirely true, as portend could be given data-race traces from an MC
+%but they don't do it in their paper, so i feel comfortable making this claim
+i.e., the state-of-the-art approach used by RaceFuzzer \cite{racefuzzer} and Portend \cite{portend}.
+The second and fourth series show all data-race bugs Quicksand found,
+which includes the former type as well as new bugs involving nondeterministic races.
+QS-Limited-HB found a nice 69 data-race bugs in total, % nice
+15 of which %could not be found with single-pass data-race candidates alone.
+required nondeterministic data-race preemption points to expose.
+QS-Pure-HB is even more dependent thereupon, %on nondeterministic data-race PPs,
+%requiring nondeterministic data-race PPs
+requiring them in 21 cases among its 44 total data-race bugs.
+Moreover, although the frequency
+of these nondeterministic races varies across the different test cases
+(for example, almost all in {\tt broadcast\_test} were nondeterministic; almost none in {\tt mutex\_test}),
+they are still at least present in all tests,
+meaning it is not just an issue of writing ``better'' test cases to avoid them.
+
+\begin{figure}[t]
+	\begin{center}
+	\includegraphics[width=0.85\textwidth]{../proposal/nondets.pdf} % TODO: get some gf colors on here
+	\end{center}
+	\caption[Nondeterministic data-race preemption points are required to find some bugs.]
+	{Nondeterministic data-race preemption points are required to find some bugs.
+	Requiring stateless model checking integration to identify to begin with,
+	incorporating these races as new preemption points
+	allowed Quicksand to find
+	% lhb: 127.77%
+	% phb: 191.30%
+	128\% (Limited HB) to 191\% (Pure HB) as many data-race bugs
+	compared to using single-pass candidates alone.
+	}
+	\label{fig:dr-falsenegs}
+\end{figure}
+
+Note that I am not comparing how much testing time is required before identifying the data-race candidates involved in each bug.
+Single-pass data races can all be found after a single program execution,
+while Quicksand may potentially take up to all 10 CPU-hours before identifying a nondeterministic data race.
+However, prior work data-race tools \cite{tsan,fasttrack},
+being not integrated with a model checker,
+are not intended to discover new candidates under subsequent runs.
+Running a single-pass data-race tool repeatedly for 10 CPU-hours could potentially uncover some nondeterministic candidates,
+but stress testing's comparative problem with achieving reliable coverage is already well-understood
+\cite{chess-icb,gambit},
+so I hope the reader will consider this experiment enough evidence for model checking already.
+Likewise, replay-based tools such as RaceFuzzer \cite{racefuzzer} and Portend \cite{portend}
+depend upon the data-race detector to provide an execution trace leading to each candidate.
+This result suggests that
+such tools could benefit from a similar feedback loop as is used in Iterative Deepening,
+for example, to discover new transitively-reachable data races while testing initial ones,
+even if full verification not necessarily be their goal.
+
+\subsubsection{Reallocation false positive suppression}
+
+In \cref{sec:quicksand-realloc} I showed the soundness of
+suppressing data race reports between two heap accesses when the surrounding memory was re-allocated in between.
+\Cref{tab:drstatistix}'s ``Realloc FPs'' column shows the total number of such data-race candidates for each test program,
+totaling 965 across all tests.
+% NB. Calculated this by grepping QS-LHB id-logs for "for realsies" (see quicksand pp.c).
+Among these, only 64 were observed to avoid the reallocation in an alternate interleaving,
+thereupon being promoted to real data-race preemption points.
+\cref{sec:quicksand-realloc}'s proof guarantees the safety of pruning all state spaces resulting from the 901 others.
+%
+Among the 64 true data-races, %which initially fit the malloc-recycle pattern,
+none exposed a new bug when used as a preemption point.
+This suggests that for other data-race tools,
+suppressing reallocation candidates may be a productive heuristic,
+even if unsound without Iterative Deepening.
+However, Quicksand was able to correctly identify the 64 violations of that heuristic (among 26 distinct tests),
+and fall back to classifying them with DPOR.
 

### 2018-12-20 thesis/quicksand.tex
quicksand discussion ; also split dowefindbugsfaster into 2 figures

@@
 
-\Cref{fig:dowefindbugsfaster}
+\Cref{fig:dowefindbugsfaster-cpu}
 plots the bug-finding performance of Quicksand's three experimental trials
 against the three control approaches
-in a cumulative distribution of total bugs found against elapsed time.
+in a cumulative distribution of total bugs found against elapsed CPU-time.
 The farthest-right point on each series indicates in how many total test cases that trial found a bug
 after the 10 CPU-hour timeout.
-\Cref{fig:dowefindbugsfaster}(a)
-is the main, resource-fair comparison by CPU-time,
-while \Cref{fig:dowefindbugsfaster}(b) shows
-the wall-clock time until each bug was found.
-
-\begin{figure}[p]
-	\begin{tabular}{c}
-        \includegraphics[width=0.71\textwidth]{../proposal/dowefindbugsfaster-v2.pdf} \\
-		\begin{tabular}{p{\textwidth}}
-                (a) Bugs found by Quicksand versus control experiments, measured in CPU time.
-                %as a function of elapsed CPU time.
-                Overall,
-                a more resource-fair comparison than (b),
-                although Quicksand's start-up overhead is exaggerated, as the SSS-MC tests are not parallelized. \\
-		\end{tabular}
-                \\
-                \\
-        \includegraphics[width=0.71\textwidth]{../proposal/dowefindbugsfaster-wallclock-v2.pdf} \\
-                %(b) Bugs found by elapsed wall-clock time.
-		\begin{tabular}{p{\textwidth}}
-                (b) Bugs found by Quicksand versus control experiments, measured in wall-clock time.
-                Quicksand is parallelized tenfold; the vertical line indicates its 1 hour limit. \\
-		\end{tabular}
-        \end{tabular}
-	\caption[Bug-finding performance comparison
-	between Quicksand and prior work.]
-	{Bug-finding performance comparison between
-        by several configurations of Quicksand and the single-state-space approaches.
-	Quicksand finds 125\% as many bugs
-        with data-race PPs
-	at the 10-hour mark, compared to the best prior work approach.}
-        \label{fig:dowefindbugsfaster}
+\Cref{fig:dowefindbugsfaster-wall} shows
+the same experiments, measured by wall-clock time until each bug was found instead.
+
+\begin{figure}[h]
+	\begin{center}
+        \includegraphics[width=0.9\textwidth]{../proposal/dowefindbugsfaster-v2.pdf} \\
+	\end{center}
+	\caption[Quicksand's bug-finding performance measured in CPU time.]
+		{Quicksand's bug-finding performance measured in CPU time.
+		Quicksand finds 125\% as many bugs with data-race preemption points at the 10-hour mark,
+		compared to the best prior work approach.
+		Quicksand's startup overhead is exaggerated, as the control experiments are not parallelized.}
+        \label{fig:dowefindbugsfaster-cpu}
 \end{figure}
+% TODO: fix placement of the legend in this graph so that it's not awful page back and forthing
+\begin{figure}[h]
+	\begin{center}
+        \includegraphics[width=0.9\textwidth]{../proposal/dowefindbugsfaster-wallclock-v2.pdf} \\
+	\end{center}
+	\caption[Quicksand's bug-finding performance measured in wall-clock time.]
+		{Quicksand's bug-finding performance measured in wall-clock time.
+		Quicksand is parallelized tenfold; the vertical line indicates its 1 hour limit.}
+        \label{fig:dowefindbugsfaster-wall}
+\end{figure}
+
+%\begin{figure}[p]
+%	\begin{tabular}{c}
+%        \includegraphics[width=0.71\textwidth]{../proposal/dowefindbugsfaster-v2.pdf} \\
+%		\begin{tabular}{p{\textwidth}}
+%                (a) Bugs found by Quicksand versus control experiments, measured in CPU time.
+%                %as a function of elapsed CPU time.
+%                Overall,
+%                a more resource-fair comparison than (b),
+%                although Quicksand's start-up overhead is exaggerated, as the SSS-MC tests are not parallelized. \\
+%		\end{tabular}
+%                \\
+%                \\
+%        \includegraphics[width=0.71\textwidth]{../proposal/dowefindbugsfaster-wallclock-v2.pdf} \\
+%                %(b) Bugs found by elapsed wall-clock time.
+%		\begin{tabular}{p{\textwidth}}
+%                (b) Bugs found by Quicksand versus control experiments, measured in wall-clock time.
+%                Quicksand is parallelized tenfold; the vertical line indicates its 1 hour limit. \\
+%		\end{tabular}
+%        \end{tabular}
+%	\caption[Bug-finding performance comparison
+%	between Quicksand and prior work.]
+%	{Bug-finding performance comparison between
+%        by several configurations of Quicksand and the single-state-space approaches.
+%	Quicksand finds 125\% as many bugs
+%        with data-race PPs
+%	at the 10-hour mark, compared to the best prior work approach.}
+%        \label{fig:dowefindbugsfaster}
+%\end{figure}
 
@@
 into faster wall-clock performance remains an open research problem \cite{parallel-dpor}.
-\Cref{fig:dowefindbugsfaster}(b) gives Quicksand full credit for its inherent parallelism,
+\Cref{fig:dowefindbugsfaster-wall} gives Quicksand full credit for its inherent parallelism,
 which ICB cannot yet practically match:
@@
 
-\Cref{tab:drbugs} provides more detail on each of the bugs shown in \Cref{fig:dowefindbugsfaster},
+\Cref{tab:drbugs} provides more detail on each of the bugs shown in \Cref{fig:dowefindbugsfaster-cpu},
 broken down by test case.
 The left half shows the number found by each experimental approach,
-with the totals of each column corresponding to the values at $x=10$ hours in \Cref{fig:dowefindbugsfaster}.
+with the totals of each column corresponding to the values at $x=10$ hours in \Cref{fig:dowefindbugsfaster-cpu}.
 In \mxtest, which checks the lock implementation for correctly providing mutual exclusion
@@
 	\small
-	\begin{tabular}{r|c||c|c||c|c|c}
+	\begin{tabular}{r|c||c|c|c|c|c}
 		%\multicolumn{2}{c||}{} & \multicolumn{5}{c||}{\bf {Total bugs}} & \multicolumn{4}{c||}{\bf {Data-race bugs}} \\
 		% ---
-		& {\bf Num.} & \multicolumn{2}{c||}{\bf Quicksand} & \multicolumn{3}{c}{\bf {Single-state-space MC}}
+		& {\bf Num.} & \multicolumn{2}{c|}{\bf Quicksand} & \multicolumn{3}{c}{\bf {Single-state-space MC}}
 		%& \multicolumn{2}{c|}{\bf {Limited HB}} & \multicolumn{2}{c||}{\bf {Pure HB}} &
@@
 	\begin{tabular}{r||c|c||c}
-		{\bf Bound} & {\bf SSS-MC-ICB} & {\bf SSS-MC-Shared-Mem} & Prior work \cite{chess-icb} \\
+		{\bf Bound} & {\bf SSS-MC-ICB} & {\bf SSS-MC-Shared-Mem} & {\bf Prior work ICB} \cite{chess-icb} \\
 		\hline
@@
 \Cref{fig:totalverif} plots the cumulative distribution of total verifications provided by each approach,
-in the same style of graph as \Cref{fig:dowefindbugsfaster}.
+in the same style of graph as \Cref{fig:dowefindbugsfaster-cpu} and \Cref{fig:dowefindbugsfaster-wall}.
 For 167 of the 629 tests,
@@
 	\begin{center}
-	\includegraphics[width=0.71\textwidth]{../proposal/totalverifs-v2.pdf}
+	\includegraphics[width=0.9\textwidth]{../proposal/totalverifs-v2.pdf}
 	\end{center}
@@
 \subsubsection{Partial verification}
+\label{sec:quicksand-eval-partial}
 
@@
 rather than to write off all the CPU time spent as wasted.
-Prior work has proposed Probabilistic Concurrency Testing %(PCT)
-\cite{randomized-scheduler},
-which attempts to quantify the probability
-that a bug may remain in some untested interleaving after a time-out,
-although is opaque to the user about which parts of her code were actually tested.
-Meanwhile, ICB proposes to inform the user
-of the maximum number of preemptions that were used to test any individual interleaving,
-under the assurance that most bugs are likely to be found with fewer preemptions (\cref{tab:icb-bounds}).
-Quicksand instead reports which subsets of preemption points resulted in state spaces that did complete in time,
+To this end,
+Quicksand
+%instead
+reports which subsets of preemption points resulted in state spaces that did complete in time,
 in hopes that the user can supplement such a result with her own intuition
@@
 	(the best bug-finding approach from each group)
-	timed out with no bug found.
+	timed out. %with no bug found.
 	Among those, ``Average tested subset SSes'' counts how many partial verifications
@@
 or be reachable only under a superset combination of preemption points not yet reached.
-
-On the other hand, attempting to maximize the raw number of completed state spaces
-may not be the most user-friendly way to present partial verification results.
-For example, included in these numbers are small state spaces which are subsets of other state spaces also completed;
-the user need not examine both subset and superset separately.
-Future work could conduct user studies to determine the most effective way to present these partial results
-from a software development perspective,
-such as described in \cref{sec:future-friendly}.
 
@@
 	QS-Pure-HB identified among tests where it found no bugs.
-	Among those, ``Benign DRs'' counts how many were refuted as non-failing,
-	while ``Untested DR PPs'' counts how many could not be checked in the time limit.
+	Among those, ``Benign'' counts how many were refuted as non-failing,
+	while ``Untested'' counts how many could not be checked in the time limit.
 	Finally, ``Realloc. FPs'' counts how many reallocation false positives QS-Limited-HB suppressed.
@@
 	\begin{center}
-	\includegraphics[width=0.85\textwidth]{../proposal/nondets.pdf} % TODO: get some gf colors on here
+	\includegraphics[width=0.9\textwidth]{../proposal/nondets.pdf} % TODO: get some gf colors on here
 	\end{center}
@@
 
-% identifying & skipping redundancies in smaller ID state spaces via memoization (same applies to ICB!)
+This section will list the current limitations of Quicksand and Iterative Deepening
+and discuss opportunities for future improvement.
+
+\subsubsection{Avoiding redundant work}
+
+When Quicksand extends a small state space with more preemption points,
+the new state space is guaranteed to test a superset of interleavings compared to the old one.
+Because Quicksand prioritizes completing small state spaces before their descendants,
+the superset state spaces we run later will repeat each branch of their already-completed subsets,
+and any interleaving which does not preempt threads on any of the new preemption points will be repeated work.
+%
+% lol i never did this
+%We measured the proportion of repeated work among completed state spaces across our test suite;
+%on average, {\bf \large 999\%} of the interleavings in each test were repeated, with some tests as high as {\bf \large 9999\%}.
+This may make Quicksand slower than the single-state-space approach to find certain bugs,
+for example, if both {\tt mutex\_lock()} and {\tt mutex\_unlock()} preemption points
+together expose a bug, but not either alone.
+Predicting whether an upcoming interleaving has already been tested is not straightforward,
+but future implementations
+of Iterative Deepening and/or ICB
+could incorporate cross-job memoization
+to prune some or all such repeated work.
+
+Similarly, when pursuing total verification,
+if the state space resulting from preempting on every instruction
+(or equivalently, the maximal state space, thanks to \cref{sec:quicksand-soundness})
+could be completed in time,
+a model checker which immediately jumped to that state space,
+abandoning all smaller subsets would certainly achieve verification faster.
+Quicksand's maximal state space mode ({\tt -M}, see \cref{sec:quicksand-impl-modes})
+can strike a middle ground between Iterative Deepening and single-state-space preempt-everywhere,%
+\footnote{Thesis note: {\tt -M} mode was implemented after the conference paper's publication \cite{quicksand},
+and will be evaluated alongside transactional memory later in \cref{sec:tm-eval}.}
+but future implementations of Iterative Deepening could prioritize the maximal state space more flexibly still.
+For example, pinning its job to one of the available processors regardless of the status of any smaller jobs
+would avoid getting too flooded with smaller jobs to even begin the maximal job before time runs out.
+When full verification is infeasible,
+completing even an intermediate-sized job would allow immediately pruning all subset jobs thereof,
+perhaps using a form of binary search (on the preemption point set size) to find an appropriately-sized intermediate job.
+
+\subsubsection{Preemption point subsets}
+
+Quicksand was able to partially guarantee safety for some preemption points
+in 93\% of tests with too-large maximal state spaces (\cref{sec:quicksand-eval-partial}).
+However, in 6 cases, no more than the minimal state space could be verified,
+and in 18 others, no state spaces were completed at all.
+Larger state spaces often result from finer-grained locking,
+which can indicate a more intricate concurrent algorithm or an unnecessarily complicated design (or both).
+Such programs may require even more rigorous verification than a program with a single global lock,
+making them important to consider for future work.
+While Quicksand uses {\tt within\_function} (\cref{sec:quicksand-impl-mc})
+{\em statically} to restrict where preemption points could arise in advance of the test,
+future
+Iterative Deepening
+implementations could use this mechanism to {\em dynamically} subset preemption points further,
+making partial verification of larger tests possible,
+potentially even involving the user with interface options
+to enable and disable preemption points of her choosing at run-time.
+\cref{sec:warpzone-heuristics} discusses this possibility further.
+
+\subsubsection{Static data race analysis}
+
+In \cref{sec:quicksand-eval-bugs}, I evaluated the state-of-the-art approach's ability to find data-race-induced failures
+by configuring a static predicate to preempt on any non-stack memory access
+({\tt -0}, see \cref{sec:quicksand-impl-modes}).
+This introduced hundreds of new preemption points on each new test execution,
+with a prohibitive performance impact.
+While this performance could be improved by
+relaxing the preemption strategy,
+instead using a static or single-pass analysis to find data-race candidate PPs in advance \cite{portend},
+that would sacrifice soundness of the verification guarantee, as I showed in \cref{sec:quicksand-eval-nondets}.
+However, Quicksand itself could employ static data-race analysis such as RacerX \cite{racerx},
+or single-pass dynamic analysis such as ThreadSanitizer \cite{tsan} in future work.
+% These tools tend to err on the side of false positives,
+% but also include .....
+Any data-race candidates identified in advance could heuristically be included in Quicksand's initial seed PP sets
+(\cref{sec:quicksand-initial-pps}),
+enabling it to focus on the most suspicious races immediately,
+rather than waiting for them to be identified after potentially many iterations of model checking.
+
+\subsubsection{Partial verification}
+
+Quicksand is not the first concurrency tester to provide a partial verification guarantee
+when it times out on too-large tests.
+Probabilistic Concurrency Testing %(PCT)
+\cite{randomized-scheduler}
+proposes to use random exploration of the state space and quantify the probability
+that a bug may remain in some untested interleaving after a time-out,
+although it is opaque to the user about which parts of her code were actually tested.
+Meanwhile, ICB proposes to inform the user
+of the maximum number of preemptions that were used to test any individual interleaving,
+under the assurance that most bugs are likely to be found with fewer preemptions (\Cref{tab:icb-bounds}).
+These guarantees could each be useful to developers in different scenarios,
+and future work could combine the two approaches to provide both at once.
+Iterative Deepening offers a clear benefit to the expert user
+via the {\tt within\_function} command,
+which enables her to restrict a test's scope to only the modules of a codebase she wishes to test.
+
+Likewise, when full verification is not computationally feasible,
+some jobs with data-race preemption points will inevitably time out,
+and Quicksand cannot guarantee those races are
+false positives or
+benign, even though no bug was found.
+In the ``Untested DR PPs'' column of \Cref{tab:drstatistix},
+I show how many such candidates Quicksand (with Pure Happens-Before) could not verify in each test,
+ultimately totaling 38\% of all data races in tests which timed out.
+In prior work,
+Portend introduced the {\em k-witness harmless} metric \cite{portend}
+for heuristically classifying the likelihood that each data race lead to a failure or be benign.
+Quicksand could incorporate this metric to guide the user's attention
+to the unverified data races most likely to be worth her time.
+
+In \cref{sec:quicksand-eval-partial} I presented partial verification results
+measured in tens or hundreds of subset state spaces completed on average per test case.
+However,
+%On the other hand,
+attempting to maximize the raw number of completed state spaces
+is not necessarily the most user-friendly way to present partial verifications.
+For starters, those numbers included small state spaces which were subsets of other state spaces also completed;
+the user need not examine both subset and superset separately to understand what was tested.
+Future work should at least perform basic set comparisons
+to present only the non-redundant state spaces completed when time runs out.
+For a further research challenge,
+user studies could help to determine the most effective interface for presenting these partial results
+from a software development perspective,
+which I discuss further in \cref{sec:future-friendly}.
 

### 2018-12-20 thesis/tm.tex
quicksand discussion ; also split dowefindbugsfaster into 2 figures

@@
 
-In this section I review some of the evaluation's results in a broader context,
+This section will review %some of
+the evaluation's results in a broader context,
 list the current limitations of Landslide's implementation,

### 2018-12-20 thesis/quicksand.tex
tweaks to qs chapter in advance of sending out incmoplete ver.

@@
 
-Note that Iterative Deepening is a wrapper algorithm around stateless MC.
+Note that Iterative Deepening is a wrapper algorithm around stateless model checking.
 A model checker is still used to test each state space, and other reduction techniques such as DPOR
@@
 as each one serves as a guarantee that all possible interleavings therein were tested;
-\cref{sec:quicksand-discussion} discusses some limitations of this approach.
+\cref{sec:quicksand-discussion-partial} discusses some limitations of this approach.
 The next three subsections will show how to schedule these smaller jobs
@@
 \footnote{Some kernels disable preemption without disabling interrupts,
-which can be communicated to the MC using manual annotations,
+which can be communicated to the model checker using manual annotations,
 and must be treated similarly.
@@
 which would result in $2^p$ jobs right off the bat,
-although this could potentially be improved in future work (\cref{sec:quicksand-discussion}).
+although this could potentially be improved in future work (\cref{sec:quicksand-discussion-subsets}).
 
@@
 \begin{figure}[t]
-        %\includegraphics[width=0.48\textwidth]{dr-jobs-v2.pdf}
-        asdfsaf
-        % TODO: put the figure form yr snake fight
+	\begin{center}
+        \includegraphics[width=0.8\textwidth]{dr-jobs-v2.pdf}
+	\end{center}
 	\caption[Quicksand incorporates data races as new preemption points at run-time.]
@@
 		by managing the exploration of multiple state spaces,
-		communicating with each MC instance to receive ETAs, data race candidates, and bug reports.
+		communicating with each Landslide instance to receive ETAs, data races, and bug reports.
                 %When an access pair is reported as a data race candidate,
@@
 and recommend that she investigate it by hand to judge for herself whether it be a bug.
-% TODO: fix this to point to partial verif
-% TODO (also) make sure that all other quicksand-discussion references get updated if appropriate
-\cref{sec:quicksand-discussion} will discuss future opportunities for improving
+\cref{sec:quicksand-discussion-partial} will discuss future opportunities for improving
 debugging output in cases of such {\em partial verification}.
@@
 the algorithm heuristically assumes the smaller job's ETA is more accurate,
-in order to avoid repeatedly resuming larger jobs briefly only to find that their ETAs keep getting worse and worse
-(in other words, it lets us avoid thrashing in Quicksand). % ¯\_(ツ)_/¯
+in order to avoid repeatedly resuming larger jobs briefly only to find that their ETAs keep getting worse and worse.%
+\footnote{In other words, it lets us avoid thrashing in Quicksand.} % ¯\_(ツ)_/¯
 
@@
 a Pure Happens-Before analysis should identify a happens-before edge
-between line~2's {\tt free} and line~4's {\tt malloc},
-and it would report no race.
+between line~2's {\tt free()} and line~4's {\tt malloc()},
+and report no race.
 However, the upcoming evaluation will show that Limited Happens-Before retains some advantages over Pure
 (\cref{sec:quicksand-eval-bugs}),
-so it is useful to be able to automatically suppress data-race candidates
+so it is useful to
+%be able to
+automatically suppress data-race candidates
 that are certain to end up being false positives when reordered.
 Such collisions could instead be avoided with a hacked allocator which never recycles memory,
-but this could unacceptably impact performance in {\tt malloc}-heavy tests.
+but that could unacceptably impact performance in {\tt malloc()}-heavy tests.
 
@@
 this is analogous to DPOR's ability to skip equivalent interleavings within a single Landslide instance.
-Upcoming in the evaluation, \cref{sec:quicksand-eval-verif}'s \Cref{tab:drstatistics}
+Upcoming in the evaluation, \cref{sec:quicksand-eval-verif}'s \Cref{tab:drstatistix}
 will show how many redundant state spaces Quicksand is able to prune with this technique.
@@
 \begin{figure}[t]
+	\begin{center}
 	\begin{tabular}{rll}
@@
 	\end{tabular}
+	\end{center}
 	% TODO: shortver
@@
 which fundamentally cannot account for memory reordering nondeterminism.
-\cref{sec:quicksand-discussion} discusses this limitation further;
-for existing work on combining DPOR with relaxed memory, I refer the reader to \cite{tsopso}.
+%\cref{sec:quicksand-discussion} discusses this limitation further; % that's a lie..
+For existing work on combining DPOR with relaxed memory, I refer the reader to \cite{tsopso}.
 They also assume the Limited Happens-Before definition
@@
 	\item {\bf Single state space, basic DPOR} ({\tt -C}):
-		Runs a single instance of Landslide configured to preempt on all statically-known synchronization PPs.
+		Runs a single instance of Landslide configured to preempt
+		on all statically-known synchronization preemption points.
 		Corresponds to dBug's approach \cite{dbug-ssv}.
@@
 %	Quicksand finds 125\% as many bugs
-%        with data-race PPs
+%        with data-race preemption points
 %	at the 10-hour mark, compared to the best prior work approach.}
@@
 
-Beyond finding new bugs and completing full verifications with data-race PPs,
+Beyond finding new bugs and completing full verifications with data-race preemption points,
 I evaluated Quicksand's performance for classifying data-race candidates in two ways:
@@
 required nondeterministic data-race preemption points to expose.
-QS-Pure-HB is even more dependent thereupon, %on nondeterministic data-race PPs,
-%requiring nondeterministic data-race PPs
+QS-Pure-HB is even more dependent thereupon, %on nondeterministic data-race preemption points,
+%requiring nondeterministic data-race preemption points
 requiring them in 21 cases among its 44 total data-race bugs.
@@
 
-\subsubsection{Avoiding redundant work}
+\subsection{Avoiding redundant work}
 
@@
 
-\subsubsection{Preemption point subsets}
+\subsection{Preemption point subsets}
+\label{sec:quicksand-discussion-subsets}
 
@@
 relaxing the preemption strategy,
-instead using a static or single-pass analysis to find data-race candidate PPs in advance \cite{portend},
+instead using a static or single-pass analysis to find data-race candidates in advance \cite{portend},
 that would sacrifice soundness of the verification guarantee, as I showed in \cref{sec:quicksand-eval-nondets}.
@@
 % but also include .....
-Any data-race candidates identified in advance could heuristically be included in Quicksand's initial seed PP sets
+Any data-race candidates identified in advance could heuristically be included in Quicksand's initial seed preemption point sets
 (\cref{sec:quicksand-initial-pps}),
@@
 
-\subsubsection{Partial verification}
-
-Quicksand is not the first concurrency tester to provide a partial verification guarantee
-when it times out on too-large tests.
-Probabilistic Concurrency Testing %(PCT)
-\cite{randomized-scheduler}
-proposes to use random exploration of the state space and quantify the probability
-that a bug may remain in some untested interleaving after a time-out,
-although it is opaque to the user about which parts of her code were actually tested.
-Meanwhile, ICB proposes to inform the user
-of the maximum number of preemptions that were used to test any individual interleaving,
-under the assurance that most bugs are likely to be found with fewer preemptions (\Cref{tab:icb-bounds}).
-These guarantees could each be useful to developers in different scenarios,
-and future work could combine the two approaches to provide both at once.
-Iterative Deepening offers a clear benefit to the expert user
-via the {\tt within\_function} command,
-which enables her to restrict a test's scope to only the modules of a codebase she wishes to test.
-
-Likewise, when full verification is not computationally feasible,
+\subsection{Partial verification}
+\label{sec:quicksand-discussion-partial}
+
+% Likewise,
+When full verification is not computationally feasible,
 some jobs with data-race preemption points will inevitably time out,
@@
 to the unverified data races most likely to be worth her time.
-
+%
 In \cref{sec:quicksand-eval-partial} I presented partial verification results
@@
 
+Quicksand is not the first concurrency tester to provide a partial verification guarantee
+when it times out on too-large tests.
+Probabilistic Concurrency Testing (PCT)
+\cite{randomized-scheduler}
+proposes to use random exploration of the state space and quantify the probability
+that a bug may remain in some untested interleaving after a time-out,
+eschewing DPOR's depth-first search model to
+instead sample broad cross-sections of large state spaces.
+However, it proposes no alternate reduction strategy, making full verification impractical,
+and furthermore is opaque to the user about which parts of her code were actually tested.
+Meanwhile, ICB proposes to inform the user
+of the maximum number of preemptions used to test any individual interleaving,
+under the assurance that most bugs are likely to be found with fewer preemptions (\Cref{tab:icb-bounds}).
+Iterative Deepening
+%offers a clear benefit to the expert user
+%via the {\tt within\_function} command,
+%which enables her
+allows the expert user
+to restrict a test's scope via the {\tt within\_function} command to only the modules of a codebase she wishes to test.
+These guarantees could each be useful to developers in different scenarios,
+and future work could combine the three approaches to provide all benefits at once,
+for example, using ETAs to heuristically decide when to switch between DPOR, ICB, and/or PCT in large state spaces,
+as discussed further in \cref{sec:warpzone-heuristics}.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-quicksand is gr9
+quicksand is gr8

### 2018-12-20 thesis/overview.tex
few tweaks across chapters, use less subsubsection

@@
 These so-called {\em preemption points} may be identified by any combination of human intuition and machine analysis.
-Typical preemption points include the boundaries of synchronization APIs (e.g., {\tt mutex\_lock}) or accesses to shared variables.
+Typical preemption points include synchronization API boundaries (e.g., {\tt mutex\_lock()})
+or accesses to shared variables.
 Considering that at each preemption point multiple threads exist as options to run next,
@@
 Hence, in a program with $n$ preemption points and $k$ threads available to run at each, the state space size is $O(n^k)$.
-Nevertheless, to fully test all of a program's possible behaviours, we must check the executions corresponding to every branch of the tree.
+Nevertheless, to fully test all possible behaviours of a program, we must check the executions corresponding to every branch of the tree.
 Addressing the scaling problem in this exponential relation is the central research problem for all model checkers,
@@
 
-Some model checkers explicitly store the set of visited program states as a means of identifying equivalent interleavings \cite{spin}.
+Some model checkers,
+such as SPIN \cite{spin},
+explicitly store the set of visited program states as a means of identifying equivalent interleavings.
 From the perspective of such tools, state spaces such as these wherein equivalent states may be reached by multiple paths
@@
 	\item Condition variables, with the interface {\tt cond\_wait(cvp, mp)}, {\tt cond\_signal(cvp)}, and {\tt cond\_broadcast(cvp)}.
-{\tt cond\_wait()} blocks the invoking thread, ``simultaneously'' releasing a mutex which protects some associated state (atomically, with respect to other calls under the same mutex).
+{\tt cond\_wait()} blocks the invoking thread, ``simultaneously'' releasing a mutex which protects some associated state (atomically, with respect to other calls under that mutex).
 		{\tt cond\_signal()} and {\tt cond\_broadcast()} wake one or all waiting threads.

### 2018-12-20 thesis/related.tex
clarify model checking wrt tla+
few tweaks across chapters, use less subsubsection

@@
 
-\section{Stateless Model Checking}
+\section{Stateless model checking}
 
@@
 
-\subsubsection{Tools}
+\subsection{Tools}
 
@@
 %using explicit state tracking,
-rather than attempting to exhaustively exercise every thread interleaving directly.
+rather than attempting to exhaustively exercise every thread interleaving directly
+(see \cref{sec:glossary}).
 %
@@
 
-\subsubsection{Algorithms}
+\subsection{Algorithms}
 \label{sec:related-algs}
@@
 (whether as predicted by automatic estimation \cite{estimation} or by a human's wild guess),
-various heuristic exploration strategies have been proposed.
-Preemption Sealing \cite{sealing} allows programmers to manually exclude preemption points
+various heuristic strategies have been proposed.
+Preemption Sealing~\cite{sealing} allows programmers to manually exclude preemption points
 arising from trusted source code modules;
@@
 which is empirically more likely to expose bugs sooner should they exist;
-BPOR \cite{bpor} extends DPOR to preserve soundness
+BPOR~\cite{bpor} extends DPOR to preserve soundness
 thereunder. % what a word
@@
 focusing on preemption point subsets rather than context switch bounding.
-\revision{Maple \cite{maple} proposed a heuristic metric for measuring the amount of ``concurrency coverage'',
+\revision{Maple~\cite{maple} proposed a heuristic metric for measuring the amount of ``concurrency coverage'',
 analogous to line-by-line coverage for sequential programs,
@@
 and independence of values read/written.}
-DeMeter \cite{demeter} adapted abstraction reduction to distributed systems verification under the name Dynamic Interface Reduction,
+DeMeter~\cite{demeter} adapted abstraction reduction to distributed systems verification under the name Dynamic Interface Reduction,
 while dBug \cite{dbug-phdthesis} applied abstraction reduction to synchronization primitives,
@@
 which is counter to 15-410's educational goals.
-
-% TODO: more papers to look up:
-% Tool Design and Student Testing Behavior in an Introductory Java Course
-% Testing Strategies for the Automated Grading of Student Programs
 

### 2018-12-21 thesis/quicksand.tex
shearbox the star in madoka magica
quicksand write intro and about pps

@@
 \label{chap:quicksand}
+
+% https://tex.stackexchange.com/questions/186746/define-shearbox-with-rotatebox-and-scalebox
+\newcommand\xshearbox[2]{%
+  \FPeval{\sheark}{(root(2,(#1)*(#1)+4)+#1)/2}\FPeval{\shearl}{1/\sheark}%
+  \FPeval{\sheara}{arctan(-\sheark)*180/pi}\FPeval{\shearb}{90+\sheara}%
+  \rotatebox{\shearb}{\scalebox{\sheark}[\shearl]{{\rotatebox{\sheara}{\smash{#2}}}}}%
+}
+
 \inspirationalquote{
@@
 %{Kaname Madoka, Mahou Shoujo Madoka{\raisebox{0.1em}{$\scriptstyle \bigstar$}}Magica}
-% TODO: make the star italic
-% TODO: see https://tex.stackexchange.com/questions/63179/shear-transform-a-box/63188#63188
-{Kaname Madoka, Puella Magi Madoka{\raisebox{0.1em}{$\scriptstyle \bigstar$}}Magica}
+{Kaname Madoka, Puella Magi Madoka{\raisebox{0.1em}{\scalebox{0.65}{\xshearbox{0.25}{$\bigstar$}}}}Magica}
 %\inspirationalquote{Always, somewhere, someone is fighting for you.
@@
 
-% TODO: intro
-
-% TODO: mention that all source files in this chapter are relative to id/ subdir in the repo
+\vspace{2em}
+\qrevision{There is a fundamental disconnect between existing stateless model checkers
+and human users
+when it comes to testing concurrent code meaningfully within a fixed CPU budget.
+Existing tools
+%are configured to
+test systems according to a fixed preemption strategy,
+leading to runtime dependent entirely on the complexity of the test program,
+which may range from minutes to tens of thousands of years.
+Meanwhile, users approach testing with a finite amount of patience,
+%usually in the rough order of magnitude of a day:
+usually not varying from one test cycle to another as their code changes and evolves:
+students frantically testing last-minute changes facing a project deadline
+will likely wait no longer than an hour for test results,
+while a company preparing its product for production deployment % its, not their. companies arent people :triumph:
+may spend upwards of weeks on rigorous stability testing.
+Regardless of the use case,
+a stateless model checker committing in advance to test whichever single state space arises from its fixed strategy
+is certain to either under- or over-shoot its user's needs.
+A model checker which preempts the system too often will fail to complete the test in time,
+and one which preempts infrequently enough to complete with time to spare will leave the user wondering if it overlooked any bugs.
+
+This chapter presents Quicksand,
+an execution framework for model checking to manage this trade-off at run-time.
+Given a fixed CPU budget,
+representing the user's patience for testing,
+Quicksand dynamically alters its preemption strategy based on data race analysis
+(\cite{tsan,fasttrack}, \cref{sec:landslide-datarace})
+and optimizes the size of state spaces on-the-fly,
+guided by state space estimation
+(\cite{estimation}, \cref{sec:landslide-estimate}),
+to best match that budget.
+I will discuss the trade-off inherent in number of preemption points used (\cref{sec:quicksand-pps}),
+introduce {\em Iterative Deepening},
+the algorithm that Quicksand uses to automatically navigate that trade-off
+(\cref{sec:quicksand-id}),
+prove its soundness relative to the more expensive full verification approach
+(\cref{sec:quicksand-soundness}),
+and present a large evaluation of Quicksand against several
+state-of-the-art approaches % implemented under Landslide
+in which Quicksand performs best on both bug-finding and verification
+(\cref{sec:quicksand-eval}).
+% justifying the claim that model checkers should navigate said tradeoff at runtime
+}
 
@@
 
-% reference figure (nb color scheme tree)
-% its pps were drawn on every instruction
-% for larger test cases, this is not practical
-% consider the example program of so and so (show mx/yield prog)
-% here is a resulting state space
-% this needs a DR preemption point to FAB
-% qs example (take from thesis slides)
-
-% TODO make sure to define maximal and mminnimal
-
-% TODO: define "data race" (pure hb) vs "data race candidate" (limited); also do scane of this chapater for conisstencey
-
-% TODO: use this somehow
+\qrevision{When configuring a model checker's preemption strategy,
+or indeed,
+choosing a model checker to begin with,
+the resulting state space is {\em parameterized} by the set of preemption points.
+%
+In the first example of \Cref{fig:tree}
+I constructed the state space by expanding both threads' {\tt x++;} operations into three pseudo-assembly instructions,
+then designating every instruction as a possible preemption point,
+yielding ${3+3 \choose 3} = 20$ total interleavings.
+Later, \cref{sec:landslide-dpor} showed DPOR prune 16 of those as equivalent,
+although the reduced state space size is still combinatorial
+in the number of {\em conflicting} events, rather than the total number.
+For larger tests, %with hundreds or thousands of conflicting events in any given execution,
+committing in advance to test all possible interleavings quickly becomes impractical.
+Accordingly, many existing model checkers opt for preempting on only a subset of execution events,
+such as synchronization API boundaries.
+
+\subsection{Types of preemption points} % TODO: find better name for sec'n
+
+Consider the new example program in \Cref{fig:pps-example}(a),
+in which one thread protects its accesses to {\tt count} with a mutex,
+while the other protects its accesses with atomic increment instructions.
+Assuming {\tt count} is only ever incremented, never decremented,
+the assertion in Thread 2 expects both of its preceding increments to be visible,
+no matter how many other threads come incrementing % incrementしてくる
+{\tt count} simultaneously.
+However, this assumes any other accesses to {\tt count} use the same protection mechanism,
+i.e., {\tt atomic\_xadd()},
+but since Thread 1 uses a mutex (which Thread 2 never touches),
+the threads can interleave to cause the assert to fail,
+as shown in \Cref{fig:pps-example}(b).
+%
+\Cref{fig:pps-statespace} shows the resulting state space
+supposing the model checker preempts only on synchronization APIs.%
+\footnote{Most modern C/C++ programs invoke atomic memory instructions by using compiler intrinsics,
+which could themselves be instrumented as a known synchronization API.
+However, not all programs are guaranteed to use well-understood interfaces;
+in fact, in the 15-410 class projects to be tested in the upcoming evaluation (\cref{sec:quicksand-eval-suite}),
+students are encouraged to roll their own atomics to get more experience writing x86 assembly.
+For the sake of this example, I leave {\tt atomic\_xadd()} uninstrumented.}
+However, \Cref{fig:pps-example}(b)'s interleaving involves
+preempting Thread 1 between its load and store of {\tt count},
+which is not a known synchronization call, but rather a data race.
+Hence, none of the interleavings in \Cref{fig:pps-statespace}'s state space
+will expose the failure;
+a {\em data-race preemption point} is required to find this bug.
+
+\begin{figure}[t]
+	\begin{center}
+		\begin{tabular}{p{0.8\textwidth}}
+			\begin{center}
+			\begin{tabular}{ll}
+				\multicolumn{2}{c}{Initially {\tt int count = \const{0}; mutex\_t m;}} \\
+				\\
+				{\bf Thread 1} & {\bf Thread 2} \\
+				\hline
+				\texttt{\hilight{darkorange}{mutex\_lock}(\&m);} & \texttt{atomic\_xadd(\&count, \const{1});} \\
+				\texttt{count++;}                                & \texttt{\hilight{olivegreen}{yield}();} \\
+				\texttt{\hilight{darkblue}{mutex\_unlock}(\&m);} & \texttt{atomic\_xadd(\&count, \const{1});} \\
+				\texttt{assert(count >= \const{1});}      & \texttt{assert(count >= \const{2});} \\
+			\end{tabular}
+			\end{center}
+			\\
+			(a) Example program with synchronization API calls highlighted.
+			\\
+			\\
+			\begin{center}
+			\begin{tabular}{ll}
+				{\bf Thread 1} & {\bf Thread 2} \\
+				\hline
+				\texttt{\hilight{darkorange}{mutex\_lock}(\&m);} \\
+				\texttt{int tmp = count;} \\
+							& \texttt{atomic\_xadd(\&count, \const{1});} \\
+							& \texttt{\hilight{olivegreen}{yield}();} \\
+							& \texttt{atomic\_xadd(\&count, \const{1});} \\
+				\texttt{count = tmp + 1;} \\
+							& \texttt{\hilight{assertfail}{assert(count >= 2);}} \\
+				\texttt{\hilight{darkblue}{mutex\_unlock}(\&m);} \\
+				\texttt{assert(count >= \const{1});}
+			\end{tabular}
+			\end{center}
+			\\
+			(b) Interleaving of (a) in which Thread 2's assertion fails,
+			as the single increment of Thread 1 overwrites both those of Thread 2.
+		\end{tabular}
+	\end{center}
+	\caption[Example bug requiring data-race preemption points to expose.]
+	{Example bug requiring data-race preemption points to expose.
+	Because the two threads use different modes of synchronization to protect their respective accesses to {\tt count},
+	preempting on synchronization calls alone is insufficient to expose the bug.
+	Rather, Thread 1 must be preempted between its non-atomic load and store of {\tt count}.
+	}
+	\label{fig:pps-example}
+\end{figure}
+
+\begin{figure}[ht!] % TODO: check placement
+	\begin{center}
+		% FIXME: fix it to be a real tree
+		\includegraphics[width=0.9\textwidth]{../oopsla/tree-maximal-only.pdf}
+	\end{center}
+	\caption[State space of \Cref{fig:pps-example} with synchronization preemption points only.]
+	{State space of \Cref{fig:pps-example} with synchronization preemption points only.
+	Note that none of these interleavings preempt Thread 1 at the necessary place
+	to trigger Thread 2's assertion failure..}
+	\label{fig:pps-statespace}
+\end{figure}
+}
+
+\qrevisionminor{How should a model checker know to instrument this particular data race for preemption
+in order to find the assertion failure lurking underneath?}
 Committing in advance to
-preempt on every instruction is certain to expose any possible bug under the given test case,
+preempt on every instruction is certain to include this data race,
 but invites massive state space explosion.
@@
 Accordingly, stateless model checkers must find more efficient ways to be able to uncover bugs such as these.
-
+%
 In related work,
-Portend \cite{portend} has already suggested combining data race analysis with preemption-driven artificial scheduling,
+Portend~\cite{portend} proposed to combine data race analysis with preemption-driven artificial scheduling,
 although it obtains its data-race candidates from a stand-alone, single-pass analysis.
-In order to make sure we identify every one of a program's data races that could possibly arise under the given test case,
-we must check many different interleavings to begin with,
-perform the Portend approach for every data race we find,
+In order to
+%make sure we
+identify every
+%one of a program's
+data race that could possibly arise under the given test case,
+a model checker must check many different interleavings to begin with,
+perform the Portend approach for every data race it finds,
 which may in turn uncover more data races
 (hidden in flow control paths reachable only through interleavings of the first race, perhaps),
-and then continue model checking the multiple data races together
-in a sort of bidirectional feedback loop between the two algorithms.
-I will refer to data races that require model checking to expose to begin with as
-{\em nondeterministic data races}. % TODO: rename, this name is awful
-In the next section,
+and then continue model checking those multiple races together
+in a
+%sort of
+bidirectional feedback loop between the two algorithms.
+Upcoming,
 \cref{sec:quicksand-id} will show how I achieve this in Quicksand, and
@@
 In the evaluation, %(\cref{sec:quicksand-eval}),
-\cref{sec:quicksand-eval-nondets} in particular will justify the need
-for this feedback loop between model checking and data race analysis
-by showing that many bugs require nondeterministic data races to expose.
+\qrevisionminor{\cref{sec:quicksand-eval-bugs} and \cref{sec:quicksand-eval-verif}
+will show that Quicksand strikes a healthy balance between fast bug-finding and full verification,
+and
+\cref{sec:quicksand-eval-nondets}
+%in particular
+will justify the need
+for such a feedback loop %between model checking and data race analysis.
+by showing that many
+data races require model checking to reliably detect.
+%bugs require nondeterministic data races to expose.
+}
+
+\subsection{Terminology}
+
+% TODO: add section refs to this para
+\qrevision{For the rest of this chapter, I will use the following terminology as shorthand for the concepts explained above.
+
+\begin{itemize}
+	\item {\bf Single-state-space model checking} refers to the state-of-the-art model checking strategy,
+		i.e., approaching each test with preemption points fixed in advance.
+	\item {\bf Maximal state space} refers to the set of thread interleavings possible
+		by preempting on all currently-known preemption points,
+		whether synchronization or data-race,
+		i.e., the singular state space tested by {\em single-state-space model checking}.
+	\item {\bf Minimal state space} indicates the opposite:
+		those thread interleavings requiring no more than Landslide's mandatory preemptions on voluntary context switches.
+	\item {\bf Data-race bug}
+		will refer to a concrete failure, such as \Cref{fig:pps-example}'s assertion failure above, whereas
+	\item {\bf Data race} refers to the racing access pair itself (\cref{sec:landslide-datarace}).
+		% TODO: scan this chapter for consistency of dis
+	\item {\bf Data-race candidate} shall refer specifically to
+		potentially-racing accesses identified by Limited Happens-Before, %(\cref{sec:landslide-lhb}), % sub sub sec
+		when disambiguation with {\em data races} is necessary.
+	\item {\bf Benign data races} are those that, when reordered in an alternate interleaving,
+		do not lead to a {\em data-race bug}, while
+	\item {\bf False positives} are {\em data-race candidates} that,
+		upon trying to reorder them,
+		turn out not to exist in the alternate interleaving at all,
+		such as in \Cref{fig:hb-example}(b).
+	\item {\bf Nondeterministic data race}
+		will refer to data races that cannot be exposed on the first thread interleaving,
+		but require model checking to expose to begin with.
+\end{itemize}
+}
 
@@
 \subsubsection{Test suite}
+\label{sec:quicksand-eval-suite}
 

### 2018-12-21 thesis/quicksand.tex
show value of count in fig pp example

@@
 			\begin{center}
-			\begin{tabular}{ll}
-				{\bf Thread 1} & {\bf Thread 2} \\
+			\begin{tabular}{llc}
+				{\bf Thread 1} & {\bf Thread 2} & {\bf Value of {\tt count}} \\
 				\hline
-				\texttt{\hilight{darkorange}{mutex\_lock}(\&m);} \\
-				\texttt{int tmp = count;} \\
-							& \texttt{atomic\_xadd(\&count, \const{1});} \\
-							& \texttt{\hilight{olivegreen}{yield}();} \\
-							& \texttt{atomic\_xadd(\&count, \const{1});} \\
-				\texttt{count = tmp + 1;} \\
-							& \texttt{\hilight{assertfail}{assert(count >= 2);}} \\
-				\texttt{\hilight{darkblue}{mutex\_unlock}(\&m);} \\
-				\texttt{assert(count >= \const{1});}
+				\texttt{\hilight{darkorange}{mutex\_lock}(\&m);} & & 0 \\
+				\texttt{int tmp = count;} & & 0 \\
+							& \texttt{atomic\_xadd(\&count, \const{1});} & 1 \\
+							& \texttt{\hilight{olivegreen}{yield}();} & 1 \\
+							& \texttt{atomic\_xadd(\&count, \const{1});} & 2 \\
+				\texttt{count = tmp + 1;} & & 1 \\
+							& \texttt{\hilight{assertfail}{assert(count >= 2);}} & 1 \\
+				\texttt{\hilight{darkblue}{mutex\_unlock}(\&m);} & & 1 \\
+				\texttt{assert(count >= \const{1});} & & 1 \\
 			\end{tabular}

### 2018-12-22 thesis/quicksand.tex
write more blurbs for quicksand: eval, verif intros; expt bias; summary

@@
 
-% TODO: introd this
-
-I pose the following evaluation questions.
+\qrevision{In Quicksand, Iterative Deepening and data race analysis are intimately connected:
+the former relies on the later to supply it with new preemption points, thereby refining its search for new concurrent behaviours,
+while the latter relies on the former to thoroughly check all possible interleavings around its reported memory accesses and classify them as buggy or benign.
+Despite this synergy, which is necessary for total verification soundness (\cref{sec:quicksand-soundness}),
+each of these two techniques is a contribution in its own right when it comes to bug-finding performance.
+Hence, this evaluation will measure not only the combined approach's full verification power,
+but also the bug-finding performance of each technique separately,
+as compared to state-of-the-art single-state-space ICB and DPOR.
+I pose the following evaluation questions.}
 
@@
 
-\subsubsection{Test suite}
-\label{sec:quicksand-eval-suite}
+\subsection{Experimental setup}
+\label{sec:quicksand-expt-setup}
 
@@
 would have been an engineering burden beyond my ability to do alone and still graduate on time.}
+
+\subsubsection{Test cases}
+\label{sec:quicksand-eval-suite}
 
@@
 
-\subsubsection{Experimental setup}
-\label{sec:quicksand-expt-setup}
+\subsubsection{Model checker configuration}
+\label{sec:quicksand-expt-trials}
 
@@
 	found by each test program.
-	QS-LHB and QS-PHB are Quicksand; ICB/DPOR/ShMem are the controls (\cref{sec:quicksand-expt-setup}).
+	QS-LHB and QS-PHB are Quicksand; ICB/DPOR/ShMem are the controls (\cref{sec:quicksand-expt-trials}).
 	}
@@
 
-% TODO: put some txt here
+\qrevision{The previous section showed that Quicksand's suite of bug-finding heuristics,
+built around Iterative Deepening,
+outperform the best single-state-space approaches,
+even after correcting
+% in a way fair to the state of the art; indeed, {\em unfair} to landslide, by assuming perfect 10x speedup for control expts
+for its inherent parallelism.
+This section will hold Quicksand to its promise to uphold the other side of the trade-off as well:
+that it reach full verification on correct tests reasonably quickly.}
 
@@
 while the other three seem to reach a plateau of around 20-30 tests relatively soon.
+
+\qrevision{A single-state-space model checker could rely on the user
+to properly synchronize all reported data races,
+in accordance with the philosophy that even non-failing races should count as bugs
+\cite{miscompile-benign,data-races-are-evil},
+ultimately improving the number of tests it can verify with no data races.
+However, RacerX \cite{racerx} showed that overwhelming the user with warnings about non-failing behaviours
+jeopardizes their patience for the tool,
+which motivates Quicksand to follow in the footsteps of Portend \cite{portend} instead.}
 
@@
 
+\subsection{Experimental bias}
+
+\qrevision{The evaluation design (\cref{sec:quicksand-expt-trials})
+contains two major shortcomings
+that, to my surprise, no conference reviewer, audience member, or colleague ever called me out on.
+Firstly, the single-state-space preempt-everywhere
+strategy was conducted only with ICB enabled (SSS-MC-ShMem).
+This resulted in reasonable bug-finding performance,
+ultimately reaching 80\% as many bugs found as Quicksand's best (QS-Limited-HB) in \Cref{fig:dowefindbugsfaster-cpu}.
+However, ICB tends to repeat work across multiple preemption bound iterations, %hurts its overall completion time,
+as evidenced between SSS-MC-ICB and SSS-MC-DPOR in \Cref{fig:totalverif}.
+Correspondingly, a version of SSS-MC-ShMem configured to use traditional DPOR without ICB would
+effectively begin with a preemption bound of infinity
+and perhaps be more competitive with Quicksand on verifications.
+%by avoiding ICB's tendency to repeat interleavings. %across iterations.
+Of course, this would trade off against its bug-finding performance,
+but an expert user could compensate by deciding between the two modes
+depending whether she thinks the test is more likely to be buggy or correct.
+Granting such user involvement,
+Quicksand's maximal state space mode (\cref{sec:quicksand-impl-modes})
+would correspondingly verify more tests than QS-Pure-HB, %within those 10 CPU-hours,
+especially if it were given 10 wall-clock hours on 1 CPU rather than 1 on 10.
+Future work could also improve both Quicksand's and ICB's ability to identify and skip redundant work
+across their respective iterations,
+as discussed below.
+
+Secondly, and more fundamentally, representing state-of-the-art approaches by reimplementing them in one's own tool is fraught.
+The comparison between Iterative Deepening and ICB in
+\cref{sec:quicksand-eval-bugs} and \cref{sec:quicksand-eval-verif}
+could possibly be conflated by Landslide's implementation of ICB and/or DPOR being slower
+on account of the simulated execution environment (\cref{sec:landslide-architecture}).
+A more rigorously scientific comparison would extend a prior work model checker, such as CHESS \cite{chess},
+to support dynamically-configured data-race preemption points,
+and evaluate Quicksand with it versus its own ICB implementation as well,
+to isolate any such conflating factors.
+Concurrently with these results' publication in OOPSLA \cite{quicksand},
+more advanced state space reduction algorithms were proposed, such as Maximal Causality Reduction (MCR) \cite{mcr}.
+It is not immediately obvious that MCR's benefit would be orthogonal with Iterative Deepening;
+i.e., the benefit of Quicksand with a MCR-enabled model checker might be reduced
+compared to the benefit shown in \cref{sec:quicksand-eval-bugs} and \cref{sec:quicksand-eval-verif}.
+Future comparisons of model-checking strategies should strive to reach these higher bars of scientific rigor.}
+
 \subsection{Avoiding redundant work}
@@
 
-quicksand is gr8
+\qrevision{In order to supplant conventional stress testing,
+which, despite its
+%tendency to overlook subtle and severe nondeterministic bugs,
+%lack of any formal coverage guarantees,
+inability to reliably expose, reproduce, or verify absent bugs in any finite amount of testing time,
+remains a popular choice for concurrency programmers of all skill levels,
+stateless model checking must meet users' needs regarding realistic testing budgets.
+This chapter has presented Quicksand, which automatically navigates the trade-off
+between fast bug-finding and formal verification depending on the size of the test.
+My contributions have been as follows.
+
+\begin{itemize}
+	\item Iterative Deepening (\cref{sec:quicksand-id}),
+		an algorithm for model checkers to simultaneously test multiple state spaces,
+		incorporating new preemption points identified with dynamic analysis on the fly.
+	\item A proof of convergence (\cref{sec:quicksand-convergence}),
+		showing that for any verification obtained under even the most extreme preemption strategy,
+		Iterative Deepening with data race analysis provides an equivalently strong one,
+		with far fewer preemption points necessary.
+	\item A technique for suppressing certain false positive data-race candidates
+		under Limited Happens-Before (\cref{sec:landslide-lhb})
+		by identifying intervening {\tt malloc()} and {\tt free()} calls
+		(\cref{sec:quicksand-id-realloc}),
+		and a corresponding soundness proof when this technique is used under Iterative Deepening
+		(\cref{sec:quicksand-realloc}).
+	\item Quicksand (\cref{sec:quicksand-implementation}),
+		an Iterative Deepening implementation which incorporates several heuristics for prioritizing
+		which state spaces are most likely to uncover bugs % this claim is not scientifically justified
+		or, should no bugs exist therein,
+		which ones are most likely to complete within a user-specified fixed CPU budget,
+		as informed by state space estimation (\cref{sec:landslide-estimate}).
+	\item A 629-test evaluation of Quicksand
+		against several prior state-of-the-art model checking approaches implemented in Landslide
+		(\cref{sec:quicksand-eval}),
+		which showed that Quicksand provides both faster bug-finding (\cref{sec:quicksand-eval-bugs})
+		and more full verifications (\cref{sec:quicksand-eval-verif}),
+		delivering ``the best of both worlds'' as promised,
+		and also demonstrated the need for a bidirectional feedback loop
+		between model checking and data race analysis
+		(\cref{sec:quicksand-eval-nondets}).
+\end{itemize}
+
+The next chapter will tell of my experience and results deploying Landslide in an educational setting,
+equipped with Quicksand to allow even inexperienced student users to benefit from stateless model checking
+with little to no manual configuration burden.}

### 2018-12-22 thesis/quicksand.tex
edit qs and tm stuff for consistency, clarity, typoity, etc

@@
 (\cite{tsan,fasttrack}, \cref{sec:landslide-datarace})
-and optimizes the size of state spaces on-the-fly,
+and optimizes the size of state spaces on the fly,
 guided by state space estimation
@@
 In the first example of \Cref{fig:tree}
-I constructed the state space by expanding both threads' {\tt x++;} operations into three pseudo-assembly instructions,
+I constructed the state space by expanding both threads' {\tt x++} operations into three pseudo-assembly instructions,
 then designating every instruction as a possible preemption point,
@@
 	Note that none of these interleavings preempt Thread 1 at the necessary place
-	to trigger Thread 2's assertion failure..}
+	to trigger Thread 2's assertion failure.}
 	\label{fig:pps-statespace}
@@
 
-Note that Iterative Deepening is a wrapper algorithm around stateless model checking.
+%Note that
+Iterative Deepening is a wrapper algorithm around stateless model checking.
 A model checker is still used to test each state space, and other reduction techniques such as DPOR
@@
 it can add new preemption points reactively based on any runtime analysis.
-Here
+This chapter
 will focus on run-time data race analysis~\cite{tsan,fasttrack} as the mechanism for finding new preemption candidates.
@@
 
+% TODO: fix this section, the way it lists the pps is fucked, figure reference prob dnagling
 Iterative Deepening must be seeded with a set of initial state spaces,
@@
 
-For testing user-space code, Quicksand begins with the four jobs from \Cref{fig:id}:
+For testing user-space code, Quicksand begins with the four
+% TODO: decide
+jobs from \Cref{fig:id}:
+%possible combinations of preemption points from \Cref{fig:pps-statespace}:
 $\{yield\}$,
@@
 \newcommand\AllJobs{\ensuremath{\mathcal{J}}}
-\begin{algorithm}[t]
+\begin{algorithm}[h]
 	\SetKwInOut{Input}{Input}
@@
 	\begin{center}
-        \includegraphics[width=0.8\textwidth]{dr-jobs-v2.pdf}
+        \includegraphics[width=0.9\textwidth]{dr-jobs-v2.pdf}
 	\end{center}
 	\caption[Quicksand incorporates data races as new preemption points at run-time.]
-		{Quicksand incorporates data races as new preemption points at run-time,
+		{Quicksand incorporates data races as new preemption points at run-time
 		by managing the exploration of multiple state spaces,
@@
                 When a data race is reported,
-                a new preemption point is added for each involved access,
+                a new preemption point is added for each involved memory access,
                 and new jobs are added for later testing,
-                corresponding to different combinations of those with the existing preemption points.}
+                corresponding to different combinations thereof with the existing preemption points.}
         \label{fig:new-dr-jobs}
@@
 
-I am presenting these proofs as they appeared in the OOPSLA paper \cite{quicksand}:
+I present these proofs as they appeared in the OOPSLA paper \cite{quicksand}:
 written as sketches in informal prose,
@@
 		in order that exploration progress far enough to defer some jobs for the sake of example.
-		In a real execution, jobs 3, 5, 7, 8, 9, and 11 would be cancelled.}
+		In a real execution, the superset jobs 3, 5, 7, 8, 9, and 11 would be cancelled.}
 	\item Jobs 4 and 6 are the ``small'' jobs added to test the two data races in isolation;

### 2018-12-22 thesis/tm.tex
few tweaks across chapters, use less subsubsection
edit qs and tm stuff for consistency, clarity, typoity, etc

@@
 
-\subsubsection{Experimental setup}
+\subsection{Experimental setup}
 \label{sec:tm-eval-exp-setup}
@@
 		\begin{tabular}{l}
-		\texttt{\flow{while} (\_retry);} \\
+		\texttt{\flow{while} (\_retry) \flow{continue};} \\ % in the actual code it's "while (retry);"; i'm adding continue for clarity
 		\texttt{\flow{if} (\call{\_xbegin}() == \const{SUCCESS}) \{} \\
@@
 	as a consistency-check assertion failure (for test parameters (2,3)).
-	The presence of {\tt while (\_retry);} makes the necessary preemption window extremely small
+	The presence of {\tt while (\_retry) continue} makes the necessary preemption window extremely small
 	(between it and {\tt \_xbegin()}),
@@
 
-\subsubsection{Retry set optimality}
+\subsection{Retry set optimality}
 
@@
 
-\subsubsection{STM reduction soundness}
+\subsection{STM reduction soundness}
 
@@
 
-\subsubsection{Nested transactions}
+\subsection{Nested transactions}
 \label{sec:tm-warpzone-nested}
@@
 
-\subsubsection{Relaxed memory orderings}
+\subsection{Relaxed memory orderings}
 \label{sec:tm-warpzone-relaxed}

### 2018-12-22 thesis/tm.tex
tm: document weak atomicity proof assumption of homogeneity. also tweak some text later on

@@
 \end{proof}
+
+\qrevision{This proof assumes that all operations $\tiiat{\beta}$
+%within $\tii$'s
+are in fact executed transactionally.
+In STM systems, the user must annotate all shared memory accesses she wishes to be protected,
+so an incorrectly-annotated program could mix transactional and non-transactional operations
+in $\tii$.
+To emulate the interactions thereof with other threads would require fully simulating failure rollback after all,
+which I must leave to future work.
+I believe this assumption to be justified for the sake of any future weakly-atomic HTM systems,
+in which the user would not have the mixed blessing of such fine-grained annotations.}
 
@@
 
-While prior work has focused on verifying implementations of transactional memory themselves
+While prior work has focused on verifying transactional memory implementations themselves
 \cite{specifying-verifying-tm,tm-correctness,tm-completeness,mc-tm-with-spin},
@@
 % mumble smth new concurrency model
-Accordingly, there is no baseline against which to compare Landslide's performance.
-Likewise, since Landslide's HTM semantics emulation relies on the equivalences proved in \cref{sec:tm-design},
+Accordingly, there is no baseline against which to compare its performance.
+Likewise, since Landslide's HTM emulation relies on the equivalences proved in \cref{sec:tm-design},
 I did not actually implement an HTM-style speculative-execution-and-rollback simulation mode.
@@
 		\item {\tt map\_basic}: Separate-chaining hashmap concurrent insertion test \cite{tm-benchmark-cmu}.
-		\item {\tt map\_basicer}: {\tt map\_basic} modified with a larger initial size to skip the resizing step.
+		\item {\tt map\_basicer}: {\tt map\_basic} configured with a larger initial map size to skip the resizing step.
 		\item {\tt avl\_mutex}: {\tt avl\_fixed} with transactional sections simplified by abstraction into a mutex.
@@
 	\end{itemize}
-		These are each parameterized over implementations {\tt spinlock} (from \cite{spinlock-rtm-github}),
+		These are each parameterized over the implementations {\tt spinlock} (from \cite{spinlock-rtm-github}),
 		{\tt spin\_\allowbreak{}fixed} (spoilers!!),
@@
 	rather than racing pairs
-	(the other half of a pair might well be within a transaction, which cannot be preempted on).
+	(the other half of a pair might well be within a transaction, which cannot be preempted within).
 	}

### 2018-12-22 thesis/quicksand.tex
rename quicksand motivation/pps section

@@
 to best match that budget.
-I will discuss the trade-off inherent in number of preemption points used (\cref{sec:quicksand-pps}),
+I will discuss the trade-off inherent in number of preemption points used
+(\cref{sec:quicksand-motivation}),
 introduce {\em Iterative Deepening},
@@
 
-\section{Preemption points}
-\label{sec:quicksand-pps}
+\section{Motivation}
+\label{sec:quicksand-motivation}
 
@@
 
-\subsection{Types of preemption points} % TODO: find better name for sec'n
+\subsection{Preemption points}
+\label{sec:quicksand-pps}
 
@@
 \subsection{Terminology}
+\label{sec:quicksand-terminology}
 

### 2018-12-22 thesis/education.tex
data-race

@@
 and, as mentioned previously (\cref{sec:education-pebbles-tests}),
-checks for data race access pairs inside the mutex implementation itself.
+checks for data-race access pairs inside the mutex implementation itself.
 As students have free rein to design their mutex internals,
@@
 		This is likely due to exceeding Linux's process limit and/or exhausting the class's official VM's memory,
-		which can arise when many data race candidates cause Quicksand to spawn many Landslide jobs,
+		which can arise when many data-race candidates cause Quicksand to spawn many Landslide jobs,
 		and (as on {\tt paraguay}) each with very large state spaces that must defer.
@@
 annotating the type of each preemption point (yield, mutex, or data race),
-and/or indicating the adversarial memory access for each data race preemption point.
+and/or indicating the adversarial memory access for each data-race preemption point.
 

### 2018-12-22 thesis/intro.tex
data-race

@@
 		to decide on-the-fly whether each state space is worth pursuing,
-		and uses data race analysis \cite{tsan,fasttrack}
+		and uses data-race analysis \cite{tsan,fasttrack}
 		to find new preemption point candidates based on a program's dynamic behaviour.

### 2018-12-22 thesis/landslide.tex
write more blurbs for quicksand: eval, verif intros; expt bias; summary
data-race

@@
 			\llitem {\tt TID} indicates a thread ID required to be running for this data race.
-				To specify data race PPs across all threads at once, set {\tt FILTER\_DRS\_BY\_TID=0} (see next section).
+				To specify data-race preemption points across all threads at once, set {\tt FILTER\_DRS\_BY\_TID=0} (see next section).
 			\llitem {\tt LAST\_CALL} indicates a code address required to be the site of the last {\tt call} instruction executed
@@
 	\item {\tt TESTING\_MUTEXES=0} - configure ``mutex testing'' mode (1),
-		in which the data race analysis will not consider a mutex's implementation to be protected by the mutex itself.
+		in which the data-race analysis will not consider a mutex's implementation to be protected by the mutex itself.
 		In other words, the mutex's internal memory accesses will be flagged as data races,
@@
 		Typical threads to use this for are init and idle.
-	\item {\tt ignore\_sym NAME SIZE} - specifies a global variable {\tt NAME} of a given {\tt SIZE} in bytes whose memory accesses should be ignored for the purposes of DPOR and data race analysis.
+	\item {\tt ignore\_sym NAME SIZE} - specifies a global variable {\tt NAME} of a given {\tt SIZE} in bytes whose memory accesses should be ignored for the purposes of DPOR and data-race analysis.
 		Typical symbols to use this for are the console or heap mutex.
-	\item {\tt sched\_func NAME} - specifies a function whose memory accesses should all be ignored for the purposes of DPOR and data race analysis.
+	\item {\tt sched\_func NAME} - specifies a function whose memory accesses should all be ignored for the purposes of DPOR and data-race analysis.
 		Typical functions to use this for are the timer handler and context switcher.
@@
 		{\tt USERSPACE} should be 0 or 1 to denote a kernel-space or user-space function respectively.
-	\item {\tt thrlib\_function NAME} - specifies a function whose memory accesses should be ignored both by the data race analysis and by DPOR.
+	\item {\tt thrlib\_function NAME} - specifies a function whose memory accesses should be ignored both by the data-race analysis and by DPOR.
 		This is recommended for marking trusted-correct thread library code when testing multithreaded client code thereof,
@@
 
-The following annotations enable Landslide to track locksets~\cite{eraser} for data race analysis.
+The following annotations enable Landslide to track locksets~\cite{eraser} for data-race analysis.
 If not provided, it will be as if Landslide assumes no guarantees about mutual exclusion or happens-before,
@@
 	\item {\tt tell\_landslide\_mutex\_locking(void *mutex\_addr)} - indicates the beginning of the lock routine for
-		whatever synchronization API Landslide should treat as the primitive for data race detection.
+		whatever synchronization API Landslide should treat as the primitive for data-race detection.
 		In Pintos this is the {\tt sema\_*()} function family; in Pebbles they may be called anything.
@@
 \section{Architecture}
+\label{sec:landslide-architecture}
 
@@
 for DPOR (\cref{sec:landslide-dpor})
-and data race analysis (\cref{sec:landslide-datarace}).
+and data-race analysis (\cref{sec:landslide-datarace}).
 For userspace tests, it also tracks which virtual address space ({\tt cr3}) belongs to the test program
@@
 
-The {\tt data\_race} command specifies an instruction pointer value to identify as a data race preemption point.
+The {\tt data\_race} command specifies an instruction pointer value to identify as a data-race preemption point.
 It can optionally be qualified by a thread ID, most recent system call number, et cetera,
@@
 Whenever a conflict is identified,
-\revisionminor{Landslide} also invokes the data race analysis (\cref{sec:landslide-datarace}).
+\revisionminor{Landslide} also invokes the data-race analysis (\cref{sec:landslide-datarace}).
 It checks for {\tt free()}/access conflicts as well as access pairs,
@@
 	\item Uses its simulator's memory tracing integration \cite{simics,bochs}
-		to power data race analysis \cite{djit,fasttrack} (\cref{sec:landslide-datarace}),
+		to power data-race analysis \cite{djit,fasttrack} (\cref{sec:landslide-datarace}),
 		as well as DPOR.
@@
 	\item Adds a mutex testing mode to configure whether the
-		data race analysis should consider locks' internal accesses
+		data-race analysis should consider locks' internal accesses
 		(\cref{sec:landslide-dynamicconfig}; {\tt TESTING\_MUTEXES}).

### 2018-12-22 thesis/overview.tex
data-race

@@
 
-A data race analysis may be either {\em static} (inspecting source code) \cite{racerx} or {\em dynamic} (tracking individual accesses arising at run-time) \cite{tsan}.
+A data-race analysis may be either {\em static} (inspecting source code) \cite{racerx}
+or {\em dynamic} (tracking individual accesses arising at run-time) \cite{tsan}.
 This paper focuses exclusively on dynamic analysis,
@@
 	\item {\bf Iterative Deepening.}
-		The {\em heuristic} search ordering strategy and framework for incorporating {\em data race preemption points}
+		The {\em heuristic} search ordering strategy and framework for incorporating {\em data-race preemption points}
 		used by {\em Quicksand}.
@@
 		even if there were false negatives
-		(e.g., applying {\em DPOR} without using {\em data race preemption points}).
+		(e.g., applying {\em DPOR} without using {\em data-race preemption points}).
 		Used by analogy with logic, in which said of an inference system,

### 2018-12-22 thesis/quicksand.tex
data-race

@@
 representing the user's patience for testing,
-Quicksand dynamically alters its preemption strategy based on data race analysis
+Quicksand dynamically alters its preemption strategy based on data-race analysis
 (\cite{tsan,fasttrack}, \cref{sec:landslide-datarace})
@@
 In related work,
-Portend~\cite{portend} proposed to combine data race analysis with preemption-driven artificial scheduling,
+Portend~\cite{portend} proposed to combine data-race analysis with preemption-driven artificial scheduling,
 although it obtains its data-race candidates from a stand-alone, single-pass analysis.
@@
 will justify the need
-for such a feedback loop %between model checking and data race analysis.
+for such a feedback loop %between model checking and data-race analysis.
 by showing that many
@@
 and Quicksand to refer to the specific implementation,
-which relies on data race analysis and specific heuristics to optimize its testing approach
+which relies on data-race analysis and specific heuristics to optimize its testing approach
 for kernels and thread libraries.
@@
 This chapter
-will focus on run-time data race analysis~\cite{tsan,fasttrack} as the mechanism for finding new preemption candidates.
+will focus on run-time data-race analysis~\cite{tsan,fasttrack} as the mechanism for finding new preemption candidates.
 The next section (\cref{sec:quicksand-soundness})
@@
 Hence, Iterative Deepening also allows for prioritizing subset jobs
-based on number of preemption points, ETA, and whether data race candidates are included among their preemption points.
+based on number of preemption points, ETA, and whether data-race candidates are included among their preemption points.
 It relies on state-space estimation \cite{estimation}
@@
 and just after interrupts are re-enabled (on x86, {\tt sti}).
-During data race analysis, {\tt cli} and {\tt sti} are treated as a single global lock
+During data-race analysis, {\tt cli} and {\tt sti} are treated as a single global lock
 (note that {\tt cli}'d memory accesses can still race with others that have interrupts on).%
@@
 As discussed in \cref{sec:quicksand-pps},
-%runtime data race detection (\cref{sec:background-datarace})
+%runtime data-race detection (\cref{sec:background-datarace})
 %may find candidate unsynchronized memory conflicts that should be investigated further.
@@
 		communicating with each Landslide instance to receive ETAs, data races, and bug reports.
-                %When an access pair is reported as a data race candidate,
+                %When an access pair is reported as a data-race candidate,
                 When a data race is reported,
@@
 	% TODO: short ver
-	\caption{A common execution pattern with {\tt malloc()} that produces false positive data race candidates.}
+	\caption{A common execution pattern with {\tt malloc()} that produces false positive data-race candidates.}
 	\label{fig:recycle}
@@
 The ability to disregard reallocation false positives is unique to Iterative Deepening.
-When limited to a single test execution, suppressing any data race candidate matching this pattern is unsound.
+When limited to a single test execution, suppressing any data-race candidate matching this pattern is unsound.
 Consider the more unusual program in Figure~\ref{fig:recycle-bug},
@@
 Moreover,
-in the context of Iterative Deepening, being able to discard certain data race candidates
+in the context of Iterative Deepening, being able to discard certain data-race candidates
 allows Quicksand to skip exploring some entire state spaces,
@@
 \begin{definition}[Reachability]
-	A data race candidate, and its associated preemption point(s),
+	A data-race candidate, and its associated preemption point(s),
 	are reachable if it will be identified by a model checker
@@
 It will also cause the progress reports to report more detailed information,
-such as which preemption points are nondeterministic data-races (\cref{sec:quicksand-pps})
+such as which preemption points are nondeterministic data races (\cref{sec:quicksand-pps})
 and number of reallocation false positives suppressed (\cref{sec:quicksand-id-realloc}).
@@
 and corrects for that amount during its next ETA computation ({\tt fudge\_time()} in {\tt estimate.c}).
-Landslide's data race analysis also includes a heuristic to avoid reporting ``too suspicious''
+Landslide's data-race analysis also includes a heuristic to avoid reporting ``too suspicious''
 data-race candidates which it believes arise from the initialization pattern \cite{eraser}:
@@
 discussed in \cref{sec:quicksand-id-realloc}
-during data race analysis,
+during data-race analysis,
 Landslide includes a generation counter in its heap allocation tracking (\cref{sec:landslide-memory}).
@@
 
-\qrevision{In Quicksand, Iterative Deepening and data race analysis are intimately connected:
+\qrevision{In Quicksand, Iterative Deepening and data-race analysis are intimately connected:
 the former relies on the later to supply it with new preemption points, thereby refining its search for new concurrent behaviours,
@@
 		\end{enumerate}
-	\item Does testing alternate interleavings with model checking improve the accuracy of data race analysis?
+	\item Does testing alternate interleavings with model checking improve the accuracy of data-race analysis?
 		\begin{enumerate}
@@
 \begin{itemize}
-	\item {\bf QS-Limited-HB}: Quicksand with Landslide configured to use Limited Happens-Before for its data race analysis ({\tt -H}),
+	\item {\bf QS-Limited-HB}: Quicksand with Landslide configured to use Limited Happens-Before for its data-race analysis ({\tt -H}),
 	\item {\bf QS-Pure-HB}: Quicksand using Pure Happens-Before instead ({\tt -V}), and
@@
 %compares the types of data-race candidates necessary to expose each data-race bug in the test suite.
-visualizes the difference between single-pass and model-checking-enabled data race analysis.
+visualizes the difference between single-pass and model-checking-enabled data-race analysis.
 The first and third series represent the bugs found using preemption points from single-pass data-race candidates only,
@@
 %
-Among the 64 true data-races, %which initially fit the malloc-recycle pattern,
+Among the 64 true data races, %which initially fit the malloc-recycle pattern,
 none exposed a new bug when used as a preemption point.
@@
 
-\subsubsection{Static data race analysis}
+\subsubsection{Static data-race analysis}
 
@@
 		showing that for any verification obtained under even the most extreme preemption strategy,
-		Iterative Deepening with data race analysis provides an equivalently strong one,
+		Iterative Deepening with data-race analysis provides an equivalently strong one,
 		with far fewer preemption points necessary.
@@
 		and also demonstrated the need for a bidirectional feedback loop
-		between model checking and data race analysis
+		between model checking and data-race analysis
 		(\cref{sec:quicksand-eval-nondets}).

### 2018-12-22 thesis/tm.tex
tweak the position of tm abstraction verifs table, in the wake of new latex layout from the previous commit's paragraph
data-race

@@
 
-\Cref{tab:tm-verifs2} shows the new resulting levels of verification
-Landslide reached before the same 10-hour timeout.
-Provided that one trusts the {\tt lock} tests correctly check the desired properties,
-and that open-coding hadn't introduced any new bugs (such as \Cref{fig:avlbug}'s),
-the benefit is clear:
-the data structure tests' state spaces become much more tractable,
-%the data structure tests suddenly reach unprecedented thread/iteration counts with ease,
-% 'so on' here includes free/re-malloc mem conflicts; see comment in avl_mutex.cpp
-their state space growth now defined only by internal conflicts from tree rebalancing, map resizing, and so on.
-In total,
-summing the testing times of {\tt lock}({\tt mutex})$(K,N)$ and {\tt avl\_mutex}$(K,N)$
-produces the same verification as {\tt avl\_fixed}$(K,N)$ far more cheaply.
-Furthermore, {\tt lock}'s verification can be reused,
-whereas {\tt avl\_fixed} and {\tt map\_basic} effectively duplicated the mutex verification between them.
-
-\begin{table}[t]
+\begin{table}[hb!] % ugh... this placement directive... (be on same page as above figure)
 	\begin{center}
+		\begin{tabular}{cc}
 		\footnotesize
-		\begin{tabular}{cc}
 		\begin{tabular}{cc||r|r}
@@
 			&
+		\footnotesize
 			\begin{tabular}{cc||r|r}
@@
 		by verifying HTM mutex implementations separately.
-		Tested with STM semantics, as both lock implementations include a retry loop.}
+		Tested with STM semantics, as both lock implementations include a retry loop.
+		\\
+		% this above extra vspace necessary so that this table and fig6.9 get a page all to themselves,
+		% without latex trying to sneak two lines of body prose in at the bottom or between them
+		}
 	\label{tab:tm-verifs2}
 \end{table}
+
+\Cref{tab:tm-verifs2} shows the new resulting levels of verification
+Landslide reached before the same 10-hour timeout.
+Provided that one trusts the {\tt lock} tests correctly check the desired properties,
+and that open-coding hadn't introduced any new bugs (such as \Cref{fig:avlbug}'s),
+the benefit is clear:
+the data structure tests' state spaces become much more tractable,
+%the data structure tests suddenly reach unprecedented thread/iteration counts with ease,
+% 'so on' here includes free/re-malloc mem conflicts; see comment in avl_mutex.cpp
+their state space growth now defined only by internal conflicts from tree rebalancing, map resizing, and so on.
+In total,
+summing the testing times of {\tt lock}({\tt mutex})$(K,N)$ and {\tt avl\_mutex}$(K,N)$
+produces the same verification as {\tt avl\_fixed}$(K,N)$ far more cheaply.
+Furthermore, {\tt lock}'s verification can be reused,
+whereas {\tt avl\_fixed} and {\tt map\_basic} effectively duplicated the mutex verification between them.
 
@@
 		(\cref{sec:tm-design-failure})
-		and applying conventional data race analysis with all transactions treated as a global lock
+		and applying conventional data-race analysis with all transactions treated as a global lock
 		(\cref{sec:tm-design-locks}),

### 2018-12-22 thesis/related.tex
Data-race (haha)

@@
 
-\section{Data race analysis}
+\section{Data-race analysis}
 \label{sec:related-data-race}
 
-Data race analysis, originating with the lockset-only analysis of Eraser \cite{eraser},
+Data-race analysis, originating with the lockset-only analysis of Eraser \cite{eraser},
 has since grown into a mature field in its own right,
 which Landslide more borrows as building blocks for its own methods rather than contributing new techniques to.
-Data race detectors are largely distinguished by their particular flavour of the Happens-Before (HB) relation,
+Data-race detectors are largely distinguished by their particular flavour of the Happens-Before (HB) relation,
 as discussed in \cref{sec:background-hb}.

### 2018-12-23 thesis/intro.tex
やはり data race analysis

--- 2018-12-19 thesis/intro.tex
@@
 		a wrapper framework for Landslide
-		which incorporates data-race analysis at run-time to find new preemption points
+		which incorporates data race analysis at run-time to find new preemption points
 		and uses the new Iterative Deepening algorithm to}

### 2018-12-23 thesis/landslide.tex
Data-race (haha)
やはり data race analysis

--- 2018-12-20 thesis/landslide.tex
@@
 		(and according to my experience, \revisionminor{much} faster as well).
-	\item {\tt -H}: use Limited Happens-Before for data-race analysis (\cref{sec:background-hb})
+	\item {\tt -H}: use Limited Happens-Before for data race analysis (\cref{sec:background-hb})
 		(default for Pebbles kernelspace mode)
-	\item {\tt -V}: use vector-clock-based Pure Happens-Before for data-race analysis (\cref{sec:background-hb})
+	\item {\tt -V}: use vector-clock-based Pure Happens-Before for data race analysis (\cref{sec:background-hb})
 		(default for P2/PSU userspace and Pintos modes)
@@
 		should be considered for, or excluded from,
-		DPOR and/or data-race analysis
+		DPOR and/or data race analysis
 		(\cref{sec:landslide-config-landslide};

### 2018-12-23 thesis/overview.tex
Data-race (haha)
やはり data race analysis

@@
 (\cref{sec:overview-stateless}),
-data-race analysis
+data race analysis
 (\cref{sec:background-datarace}),
@@
 and allows the testing tool to reduce the possible space of interleavings
-(or the set of false positive data-race candidates reported, et cetera;
+(or the set of false positive data race candidates reported, et cetera;
 \revisionminor{see \cref{sec:landslide-config-landslide}})
@@
 	\end{center}
-	\caption[Data-race analyses may be prone to either {\em false negatives} or {\em false positives}.]
-		{{Data-race analyses may be prone to either {\em false negatives} or {\em false positives}.
+	\caption[Data race analyses may be prone to either {\em false negatives} or {\em false positives}.]
+		{{Data race analyses may be prone to either {\em false negatives} or {\em false positives}.
 	Applying Happens-Before to program (a) will miss the potential race possible between A1/A2 in an alternate interleaving,
@@
 
-A data-race analysis may be either {\em static} (inspecting source code) \cite{racerx}
+A data race analysis may be either {\em static} (inspecting source code) \cite{racerx}
 or {\em dynamic} (tracking individual accesses arising at run-time) \cite{tsan}.
@@
 %, we use the Limited Happens-Before relation for our analysis.
-%The justification for this is that, while stand-alone data-race analyses must avoid inundating the user with false alarms \cite{racerx},
-%my work incorporates data-race analysis in an internal feedback loop, and reports only directly observed failures to the user.
+%The justification for this is that, while stand-alone data race analyses must avoid inundating the user with false alarms \cite{racerx},
+%my work incorporates data race analysis in an internal feedback loop, and reports only directly observed failures to the user.
 %Hence, I accept some overhead from false positives for the sake of more thorough testing.
@@
 		have misinterpreted its title to believe % oops
-		that Landslide be only a data-race detector,
+		that Landslide be only a data race detector,
 		rather than a model checker.}

### 2018-12-23 thesis/quicksand.tex
Data-race (haha)
やはり data race analysis

--- 2018-12-22 thesis/quicksand.tex
@@
 Portend~\cite{portend} proposed to combine data race analysis with preemption-driven artificial scheduling,
-although it obtains its data-race candidates from a stand-alone, single-pass analysis.
+although it obtains its data race candidates from a stand-alone, single-pass analysis.
 In order to
@@
 		% TODO: scan this chapter for consistency of dis
-	\item {\bf Data-race candidate} shall refer specifically to
+	\item {\bf Data race candidate} shall refer specifically to
 		potentially-racing accesses identified by Limited Happens-Before, %(\cref{sec:landslide-lhb}), % sub sub sec
@@
 		do not lead to a {\em data-race bug}, while
-	\item {\bf False positives} are {\em data-race candidates} that,
+	\item {\bf False positives} are {\em data race candidates} that,
 		upon trying to reorder them,
@@
 It is seeded with the set of all statically-known synchronization API preemption points,
-and invoked whenever a new data-race candidate is found.
+and invoked whenever a new data race candidate is found.
 %
@@
 or complete successfully, indicating a {\em benign} (i.e., false-positive) data race.
-They may also uncover a new data-race candidate entirely in some alternate interleaving,
+They may also uncover a new data race candidate entirely in some alternate interleaving,
 in which case we may iteratively advance to a superset state space which will preempt at both racing access pairs.
@@
 
-Landslide classifies data-race candidates as {\em both-order} or {\em single-order},
+Landslide classifies data race candidates as {\em both-order} or {\em single-order},
 as defined in prior work \cite{portend},
@@
 
-Finally, I identified a particular class of false positive data-race candidates
+Finally, I identified a particular class of false positive data race candidates
 under the Limited Happens-Before analysis (\cref{sec:background-datarace})
@@
 then lines~1 and 7 will be flagged as a potential data race.
-I term this a {\em reallocation false-positive data-race candidiate}.
+I term this a {\em reallocation false-positive data race candidiate}.
 To the human eye, this is obviously a false positive:
@@
 %be able to
-automatically suppress data-race candidates
+automatically suppress data race candidates
 that are certain to end up being false positives when reordered.
@@
 They also assume the Limited Happens-Before definition
-for the data-race analysis.
+for the data race analysis.
 I leave the case for Pure Happens-Before to future work,
 although if I may appeal to intuition,
-it requires only to show that for any data-race candidate
+it requires only to show that for any data race candidate
 Limited Happens-Before reports in a given execution,
@@
 
-\begin{theorem}[Soundness of eliminating reallocation data-race candidates]
+\begin{theorem}[Soundness of eliminating reallocation data race candidates]
         If a reallocation candidate is not a false positive,
@@
 especially those arising from the {\tt malloc} lock (\cref{sec:quicksand-realloc}),
-which functions DPOR and data-race analysis should treat as trusted code (\cref{sec:landslide-config-landslide}),
+which functions DPOR and data race analysis should treat as trusted code (\cref{sec:landslide-config-landslide}),
 whether to enable mutex-testing mode (\cref{sec:landslide-dynamicconfig}),
@@
 Landslide messages after testing each interleaving to report updated progress and ETAs
-as well as whenever a new data-race candidate or bug is found.
+as well as whenever a new data race candidate or bug is found.
 Quicksand in turn replies whether to resume/suspend (due to too high ETA) or quit (due to timeout)
@@
 Landslide's data race analysis also includes a heuristic to avoid reporting ``too suspicious''
-data-race candidates which it believes arise from the initialization pattern \cite{eraser}:
+data race candidates which it believes arise from the initialization pattern \cite{eraser}:
 if a conflicting access pair is single-order (\cref{sec:quicksand-heuristics})
@@
 \cite{erlang,haskell,rust-book}.
-When Quicksand ignores all data-race candidates,
+When Quicksand ignores all data race candidates,
 its results are competitive with SSS-MC-DPOR, although SSS-MC-ICB outperforms it slightly.
@@
 This result suggests that in future work,
-Quicksand should consider using ICB in parallel with its default configuration when it finds no data-race candidates to test.
+Quicksand should consider using ICB in parallel with its default configuration when it finds no data race candidates to test.
 I discuss this possibility further in \cref{sec:warpzone-heuristics}.
@@
 
-Ultimately, using Limited Happens-Before for finding data-race candidates allowed Quicksand to find more bugs,
+Ultimately, using Limited Happens-Before for finding data race candidates allowed Quicksand to find more bugs,
 while Pure Happens-Before allowed for reaching full verification faster.
@@
 Limited Happens-Before need not wait to test many alternate thread interleavings before
-finding a data-race candidate to begin with;
-%a potential data-race candidate is confirmed;
+finding a data race candidate to begin with;
+%a potential data race candidate is confirmed;
 rather, it can add new jobs to start testing potential races immediately.%
@@
 %each approach has merit, and that
-model checkers which incorporate data-race analysis should implement both modes
+model checkers which incorporate data race analysis should implement both modes
 and offer the user to choose based on their desired and/or expected testing outcome.
@@
 Beyond finding new bugs and completing full verifications with data-race preemption points,
-I evaluated Quicksand's performance for classifying data-race candidates in two ways:
+I evaluated Quicksand's performance for classifying data race candidates in two ways:
 its ability to check nondeterministic data races not reachable under a single-pass analysis (\cref{sec:quicksand-pps})
@@
 (\cref{sec:quicksand-pps}).
-In such cases, a single-pass dynamic data-race detector
+In such cases, a single-pass dynamic data race detector
 might not achieve the coverage necessary
@@
 for this experiment.
-This prevents Landslide from suppressing an observed reallocation data-race candidate on the first interleaving,
+This prevents Landslide from suppressing an observed reallocation data race candidate on the first interleaving,
 which would falsely classify it as nondeterministic,
@@
 \Cref{fig:dr-falsenegs}
-%compares the types of data-race candidates necessary to expose each data-race bug in the test suite.
+%compares the types of data race candidates necessary to expose each data-race bug in the test suite.
 visualizes the difference between single-pass and model-checking-enabled data race analysis.
-The first and third series represent the bugs found using preemption points from single-pass data-race candidates only,
-% not entirely true, as portend could be given data-race traces from an MC
+The first and third series represent the bugs found using preemption points from single-pass data race candidates only,
+% not entirely true, as portend could be given data race traces from an MC
 %but they don't do it in their paper, so i feel comfortable making this claim
@@
 QS-Limited-HB found a nice 69 data-race bugs in total, % nice
-15 of which %could not be found with single-pass data-race candidates alone.
+15 of which %could not be found with single-pass data race candidates alone.
 required nondeterministic data-race preemption points to expose.
@@
 
-Note that I am not comparing how much testing time is required before identifying the data-race candidates involved in each bug.
+Note that I am not comparing how much testing time is required before identifying the data race candidates involved in each bug.
 Single-pass data races can all be found after a single program execution,
 while Quicksand may potentially take up to all 10 CPU-hours before identifying a nondeterministic data race.
-However, prior work data-race tools \cite{tsan,fasttrack},
+However, prior work data race tools \cite{tsan,fasttrack},
 being not integrated with a model checker,
 are not intended to discover new candidates under subsequent runs.
-Running a single-pass data-race tool repeatedly for 10 CPU-hours could potentially uncover some nondeterministic candidates,
+Running a single-pass data race tool repeatedly for 10 CPU-hours could potentially uncover some nondeterministic candidates,
 but stress testing's comparative problem with achieving reliable coverage is already well-understood
@@
 Likewise, replay-based tools such as RaceFuzzer \cite{racefuzzer} and Portend \cite{portend}
-depend upon the data-race detector to provide an execution trace leading to each candidate.
+depend upon the data race detector to provide an execution trace leading to each candidate.
 This result suggests that
@@
 suppressing data race reports between two heap accesses when the surrounding memory was re-allocated in between.
-\Cref{tab:drstatistix}'s ``Realloc FPs'' column shows the total number of such data-race candidates for each test program,
+\Cref{tab:drstatistix}'s ``Realloc FPs'' column shows the total number of such data race candidates for each test program,
 totaling 965 across all tests.
@@
 %
-Among the 64 true data-races, %which initially fit the malloc-recycle pattern,
+Among the 64 true data races, %which initially fit the malloc-recycle pattern,
 none exposed a new bug when used as a preemption point.
-This suggests that for other data-race tools,
+This suggests that for other data race tools,
 suppressing reallocation candidates may be a productive heuristic,
@@
 relaxing the preemption strategy,
-instead using a static or single-pass analysis to find data-race candidates in advance \cite{portend},
+instead using a static or single-pass analysis to find data race candidates in advance \cite{portend},
 that would sacrifice soundness of the verification guarantee, as I showed in \cref{sec:quicksand-eval-nondets}.
-However, Quicksand itself could employ static data-race analysis such as RacerX \cite{racerx},
+However, Quicksand itself could employ static data race analysis such as RacerX \cite{racerx},
 or single-pass dynamic analysis such as ThreadSanitizer \cite{tsan} in future work.
@@
 % but also include .....
-Any data-race candidates identified in advance could heuristically be included in Quicksand's initial seed preemption point sets
+Any data race candidates identified in advance could heuristically be included in Quicksand's initial seed preemption point sets
 (\cref{sec:quicksand-initial-pps}),
@@
 		with far fewer preemption points necessary.
-	\item A technique for suppressing certain false positive data-race candidates
+	\item A technique for suppressing certain false positive data race candidates % TODO: hm, should false-positive be hyph?
 		under Limited Happens-Before (\cref{sec:landslide-lhb})

### 2018-12-23 thesis/related.tex
やはり data race analysis

--- 2018-12-20 thesis/related.tex
@@
 Since these foundational algorithms,
-many more recent works have contributed to make data-race analysis more precise, more performant,
+many more recent works have contributed to make data race analysis more precise, more performant,
 and/or more domain-specific.
@@
 This has motivated replay analysis
-to classify data-race candidates by their impact on program behaviour
-by extending single-pass data-race analysis to many thread interleavings.
+to classify data race candidates by their impact on program behaviour
+by extending single-pass data race analysis to many thread interleavings.
 It was first introduced in \cite{recordreplaydrs},
@@
 Portend \cite{portend} is closest in spirit to Quicksand:
-it tests alternate executions based on single-pass data-race candidates to classify them in a taxonomy of likely severity,
+it tests alternate executions based on single-pass data race candidates to classify them in a taxonomy of likely severity,
 including non-failing races which nevertheless cause nondeterministic output
@@
 which \cref{sec:quicksand-eval-nondets} showed is necessary to find certain bugs.
-Quicksand builds on Portend's approach by introducing a feedback loop between the data-race analysis and model checking,
+Quicksand builds on Portend's approach by introducing a feedback loop between the data race analysis and model checking,
 which results in a stronger verification property
@@
 % Eytani et al.~\cite{towards-a-framework} present a promising framework for testing concurrent programs,
-% which can incorporate model checking as well as static analysis, resource exhaustion, data-race analysis, and coverage analysis.
+% which can incorporate model checking as well as static analysis, resource exhaustion, data race analysis, and coverage analysis.
 % However, it lacks an evaluation, and makes mention of its educational value only in its future work remarks.

### 2018-12-23 thesis/tm.tex
Data-race (haha)
やはり data race analysis

--- 2018-12-22 thesis/tm.tex
@@
 simulating transaction aborts as immediate failure injections,
-and treating transaction atomicity as a global mutex during data-race analysis;
+and treating transaction atomicity as a global mutex during data race analysis;
 and provide corresponding equivalence proofs.
@@
 
-Next comes the issue of memory accesses within transactions with regard to data-race analysis (\cref{sec:background-datarace}).
+Next comes the issue of memory accesses within transactions with regard to data race analysis (\cref{sec:background-datarace}).
 \Cref{thm:atom} provides that the body of all transactions may be executed atomically within the MC environment.
@@
 I claim this level of atomicity is equivalent to that provided by a global lock,
-and hence abstracting it as such in Landslide's data-race analysis is sound.
+and hence abstracting it as such in Landslide's data race analysis is sound.
 
@@
 \end{itemize}
-Hence, data-race analysis is sound when transaction boundaries are replaced by an abstract global lock.
+Hence, data race analysis is sound when transaction boundaries are replaced by an abstract global lock.
 \end{proof}
@@
 		(\revisionminor{\cref{sec:tm-design-failure}}, \cref{sec:txn-failure}),
-		and treatment of transactional regions during data-race analysis (\cref{sec:txn-datarace}).
+		and treatment of transactional regions during data race analysis (\cref{sec:txn-datarace}).
 	\item {\tt -A} (for ``xAbort codes'')

### 2018-12-24 thesis/intro.tex
assorted minor proofreads

@@
 Pronoun use will vary between more specific and more ambiguous to convey additional nuance.
-The singular ``I'' is associated with my own research contributions,
+The singular ``I''
+%is associated with
+connotes
+my own research contributions,
 while the royal ``we'' should be taken to include the reader,
@@
 as well as
-on the author themself,
-who would also prefer them for any third-person reference by you.
+on the author themself.
+%who would also prefer them for any third-person reference by you.
 

### 2018-12-24 thesis/landslide.tex
assorted minor proofreads

@@
 		(usually 0; note that many kernels use a nesting depth counter where any positive value corresponds to disabled).%
-		\footnote{These two options replace the deprecated {\tt kern\_ready\_for\_timer\_interrupt()} annotation from {\tt student.c} described in \cite{landslide} \sect{6.2.3}.}
+		\footnote{These two options replace the deprecated {\tt kern\_ready\_for\_timer\_interrupt()} annotation from {\tt student.c} described in \cite[\sect{6.2.3}]{landslide}.}
 	\item {\tt PATHOS\_SYSCALL\_IRET\_DISTANCE=VALUE} - indicate how much stack space is used by the reference kernel's system call wrappers.
@@
 In the Bochs version, function names and line numbers are handled separately:
-Bochs is patched with a new API function named {\tt bx\_dbg\_symbolic\_address\_landslide}%
-\footnote{does the same thing as the existing {\tt bx\_dbg\_symbolic\_address}, but with a better type signature}
+Bochs is patched with a new API function named {\tt bx\_dbg\_symbolic\_address\_landslide()}%
+\footnote{Does the same thing as the existing {\tt bx\_dbg\_symbolic\_address()}, but with a better type signature.}
 which provides function names and hexadecimal offsets;
-while for line numbers, {\tt pebsim/pintos/build.sh}%
-\footnote{Line numbers in Bochs for Pebbles/P2s are not supported yet; see {\tt p2-setup.sh} for the work-in-progress.}
+while for line numbers, {\tt pebsim/build.sh} (and {\tt pebsim/pintos/\allowbreak{}build.sh} for Pintos)
+%\footnote{Line numbers in Bochs for Pebbles/P2s are not supported yet; see {\tt p2-setup.sh} for the work-in-progress.}
 generates a header file {\tt line\_numbers.h}
@@
 
-The stack tracer is implemented in {\tt stack.c}.
-It does the standard approach of following the base pointer chain
+The stack tracer is implemented in {\tt stack.c}
+using the standard approach of following the base pointer chain
 (not supporting code compiled with {\tt -fomit-frame-pointer} by doing anything clever
-like understanding how much stack frame is allocated for each function),
+like understanding how much stack frame each function allocates),
 and printing symbol table information for the pushed return addresses at the top of each frame.
-
+%
 However, it also offers several special-case features
@@
 \begin{itemize}
-	\item If no {\tt within\_function} commands are given, operate in ``denylist mode'':
+	\item If no {\tt within\_function} commands are given, operate in ``denylist'' mode:
 		the preemption point is by default valid as long as no {\tt without\_function} calls reject it.
-		Otherwise, operate in ``allowlist mode'':
+		Otherwise, operate in ``allowlist'' mode:
 		the preemption point is by default rejected unless at least one {\tt within\_function} directive matches.
@@
 
-\subsubsection{Run-time (Recursive) estimation}
+\subsubsection{Runtime (Recursive) estimation}
 

### 2018-12-24 thesis/quicksand.tex
assorted minor proofreads

@@
 yielding ${3+3 \choose 3} = 20$ total interleavings.
-Later, \cref{sec:landslide-dpor} showed DPOR prune 16 of those as equivalent,
+Later, \cref{sec:landslide-dpor} showed that DPOR prunes 16 of those as equivalent,
 although the reduced state space size is still combinatorial

### 2018-12-24 thesis/quicksand.tex
proofread and fix formatting of many things in qs chapter

@@
 
-% TODO: add section refs to this para
 \qrevision{For the rest of this chapter, I will use the following terminology as shorthand for the concepts explained above.
@@
 	\item {\bf Data race} refers to the racing access pair itself (\cref{sec:landslide-datarace}).
-		% TODO: scan this chapter for consistency of dis
 	\item {\bf Data race candidate} shall refer specifically to
@@
 such as when she believes all bugs have been fixed and wants a verification as fast as possible;
-% TODO: clarify sec reference to -M flag
-\cref{sec:quicksand-implementation} discusses this execution mode further.
+\cref{sec:quicksand-impl-modes} discusses this execution mode further.
 
@@
 and claim that it is safe to completely disregard them when considering where to add new preemption points.
-Figure~\ref{fig:recycle} shows a common code pattern and interleaving which can expose such behaviour.
+\Cref{fig:recycle} shows a common code pattern and interleaving which can expose such behaviour.
 If the {\tt malloc} on line~4 returns the same address passed to {\tt free} on line~2,
@@
 When limited to a single test execution, suppressing any data race candidate matching this pattern is unsound.
-Consider the more unusual program in Figure~\ref{fig:recycle-bug},
+Consider the more unusual program in \Cref{fig:recycle-bug},
 in which the memory is recycled the same way, but the racing access's address is not tied to {\tt malloc}'s return value.
@@
 
+\subsubsection{Equivalence}
+
 \newcommand\ppnext[1]{\ensuremath{\mathsf{next}(#1)}}
@@
 
-% TODO: decide if these subsection headers are good, or, not
-\subsubsection{Equivalence}
-
+\vspace{0.5em}
 \begin{lemma}[Equivalence of non-data-race preemption points]
@@
 Modify the input interleaving by reordering $\ppinstr{p}$ until $i$, not including $i$, to before $\ppothers{p}$.
-% TODO: fix this handwavey citation to at least point to something that says what soundness of dpor means
-By the soundness of DPOR \cite{dpor}, this is equivalent to the input interleaving.
+By the soundness of DPOR (\cref{sec:landslide-dpor}; \cite{dpor}), this is equivalent to the input interleaving.
 In other words, $p$ has been transformed into $p'$ such that $\ppnext{p'} = i$,
@@
 
-% TODO: check spcaing between defs/lemmas/etc and surrounding text
+\vspace{0.5em}
 \begin{definition}[Reachability]
@@
 \end{definition}
+\vspace{0.5em}
 
@@
 
+\vspace{0.5em}
 \begin{lemma}[Saturation of data races]
@@
 
+\vspace{0.5em}
 \begin{theorem}[Convergence]
@@
 
+\vspace{0.5em}
 \begin{theorem}[Soundness of eliminating reallocation data race candidates]
@@
 ({\tt start\_job()} in {\tt job.c})
-communicates with the MC process via message-passing
+communicates with its corresponding model checker process via message-passing
 over a FIFO pipe % not gonna cite io.c c.c
-({\tt talk\_to\_child()} in {\tt messaging.c}).
+({\tt talk\_to\_\allowbreak{}child()} in {\tt messaging.c}).
 Landslide messages after testing each interleaving to report updated progress and ETAs
@@
 	\item {\bf Messaging} ({\tt messaging.c}):
-		Manages communication with child Landslides ({\tt talk\_to\_child()},
+		Manages communication with child Landslides ({\tt talk\_\allowbreak{}to\_child()},
 		implementing certain aspects of Iterative Deepening,
@@
 		including the check for whether to switch priority from one job to another
-		(\Cref{alg:shouldworkblock}, {\tt should\_work\_block()} and {\tt get\_job()}),
+		(\Cref{alg:shouldworkblock}, {\tt should\_work\_\allowbreak{}block()} and {\tt get\_job()}),
 		as well as managing the shared workqueue of jobs overall ({\tt workqueue\_thread()}).
@@
 		checking for legality of various combinations of exploration modes.
-		New options may be added with the convenient, if scarily-implemented,
+		New options may be added with the convenient
+		%, if scarily-implemented,
 		macros {\tt DEF\_CMDLINE\_FLAG()} and {\tt DEF\_CMDLINE\_OPTION()}.
@@
 
-Note that Quicksand restricts ICB to be usable only in modes when it is running only one Landslide at a time.
+Quicksand restricts ICB to be usable only in modes when it runs only one Landslide at a time.
 ICB is itself a heuristic search ordering strategy to uncover bugs faster,
@@
 However, maximal state space mode does support ICB,
-as the former focuses on verification only,
+as it focuses on verification only,
 but if the result is a time-out,
@@
 I chose P2s and Pintoses for this test suite because of the relative ease of generating hundreds of unique state spaces,
-varied in size and correctness, and with a diverse set of bug types.
-In addition to concurrency bugs,
+varied in size and correctness, and with a diverse set of bug types.%
+\footnote{In addition to concurrency bugs,
 many of the codebases exhibited {\em deterministic} bugs (i.e., encountered on the first interleaving tested),
-which I fixed by hand before running these tests to ensure that every bug in this study required meaningful work by the model checker.
+which I fixed by hand before running these tests
+to ensure that every bug in this study required meaningful work by the model checker.}
 While many prior work stateless model checking papers
-% TODO: cite some
+% TODO: find one that tests like... apache. to back up your claim of realest of the real world
+\cite{chess-icb,optimal-dpor,mcr,rcmc}
+% bpor could go here but it's already known bugs benchmarks, less the 'real worl' approach
+% wow, ZKW15 is 121 programs! nice job!!
 publish studies of single-digit or low-double-digit numbers of
-bugs found in ``real-world'' programs, often confirmed by the upstream developers as severe,
+bugs found in ``real-world'' programs,
+sometimes reported to and confirmed by the upstream developers, %as severe,
 to motivate stateless model checking to be used in production settings,
@@
 i.e., the minimum number of involuntary thread switches each bug required to expose.
-Table~\ref{tab:icb-bounds} shows the distribution of these bounds,
+\Cref{tab:icb-bounds} shows the distribution of these bounds,
 which is consistent with the results of \cite[Table 2]{chess-icb},
@@
 
-\begin{table}[h]
+% TODO: check placement
+\begin{table}[t]
 	\begin{center}
@@
 ``Average tested subset SSes'' column.
-% of Table~\ref{tab:drbugs}. % thesis - now split into sep table
+% of \Cref{tab:drbugs}. % thesis - now split into sep table
 %While obviously not as strong as full verification,
@@
 
-Note that I am not comparing how much testing time is required before identifying the data race candidates involved in each bug.
-Single-pass data races can all be found after a single program execution,
-while Quicksand may potentially take up to all 10 CPU-hours before identifying a nondeterministic data race.
+Note that I do not compare how much testing time is required before identifying the data races %candidates
+involved in each bug.
+While single-pass data races are all found after a single test execution,
+Quicksand may potentially take up to all 10 CPU-hours before identifying a nondeterministic data race.
 However, prior work data race tools \cite{tsan,fasttrack},
@@
 even if unsound without Iterative Deepening.
-However, Quicksand was able to correctly identify the 64 violations of that heuristic (among 26 distinct tests),
+However, Quicksand was able to correctly identify the 64 violations of that heuristic, among 26 distinct tests,
 and fall back to classifying them with DPOR.

### 2018-12-24 thesis/quicksand.tex
more qs proofreads; own up to the "fake" state spaces

@@
 each {\em pair} of racing accesses will spawn four new jobs.
-\Cref{fig:new-dr-jobs} visualizes the resulting overall workflow in Quicksand,
+\Cref{fig:new-dr-jobs} depicts the resulting overall workflow in Quicksand,
 including the four such jobs resulting from one data race report.%
@@
 as their purpose is only fast heuristic bug-finding rather than exhaustive coverage;
-see {\tt handle\_data\_race()} in {\tt messaging.c}.}
+see {\tt handle\_data\_race()} in {\tt messaging.c}.}%
+\footnote{\qrevision{For visual simplicity,
+fake state spaces are shown here %for each job
+to convey only relative size differences, %among them,
+but not the internal asymmetric structure inherent to interleavings of multiple threads
+(compare to \Cref{fig:pps-statespace}).}}
 %
@@
 \footnote{
-Though its worst-case performance is $O(mn)$ in the
-%number of pending and suspended jobs,
-sizes of $\mathcal{P}$ and $\mathcal{S}$,
+Though its worst-case performance is $O(|\mathcal{P}|\times|\mathcal{S}|)$,
+%in the
+%%number of pending and suspended jobs,
+%sizes of $\mathcal{P}$ and $\mathcal{S}$,
 in practice the non-constant portion beyond line~4 runs very infrequently
-and is negligible compared to the exponentially-sized state spaces.}.
+and is negligible compared to the exponentially-sized state spaces themselves.}
 
@@
 in order to avoid repeatedly resuming larger jobs briefly only to find that their ETAs keep getting worse and worse.%
-\footnote{In other words, it lets us avoid thrashing in Quicksand.} % ¯\_(ツ)_/¯
+\footnote{In order to avoid thrashing in Quicksand.} % ¯\_(ツ)_/¯
 

### 2018-12-24 thesis/quicksand.tex
add line nrs to pp example program

@@
 			\begin{center}
-			\begin{tabular}{ll}
-				\multicolumn{2}{c}{Initially {\tt int count = \const{0}; mutex\_t m;}} \\
+			\begin{tabular}{rlcrl}
+				\multicolumn{5}{c}{Initially {\tt int count = \const{0}; mutex\_t m;}} \\
 				\\
-				{\bf Thread 1} & {\bf Thread 2} \\
+				& {\bf Thread 1} & & & {\bf Thread 2} \\
 				\hline
-				\texttt{\hilight{darkorange}{mutex\_lock}(\&m);} & \texttt{atomic\_xadd(\&count, \const{1});} \\
-				\texttt{count++;}                                & \texttt{\hilight{olivegreen}{yield}();} \\
-				\texttt{\hilight{darkblue}{mutex\_unlock}(\&m);} & \texttt{atomic\_xadd(\&count, \const{1});} \\
-				\texttt{assert(count >= \const{1});}      & \texttt{assert(count >= \const{2});} \\
+				11 & \texttt{\hilight{darkorange}{mutex\_lock}(\&m);} & &
+						21 & \texttt{atomic\_xadd(\&count, \const{1});} \\
+				12 & \texttt{count++;}                                & &
+						22 & \texttt{\hilight{olivegreen}{yield}();} \\
+				13 & \texttt{\hilight{darkblue}{mutex\_unlock}(\&m);} & &
+						23 & \texttt{atomic\_xadd(\&count, \const{1});} \\
+				14 & \texttt{assert(count >= \const{1});}             & &
+						24 & \texttt{assert(count >= \const{2});} \\
 			\end{tabular}

### 2018-12-24 thesis/quicksand.tex
fix algorithm KwIn stuff

@@
 
-\newcommand\AllPPs{\ensuremath{\mathcal{A}}}
-\newcommand\PendingJobs{\ensuremath{\mathcal{P}}}
-\newcommand\SuspendedJobs{\ensuremath{\mathcal{S}}}
-\newcommand\GetETA[1]{ETA(#1)}
-\newcommand\GetPPSet[1]{PPSet(#1)}
-
-\begin{algorithm}[h]
-	\SetKwInOut{Input}{Input}
-	\Input{$j$, the currently-running job}
-	\Input{\AllPPs, the set of all known preemption points} %, sorted by decreasing heuristic priority}
-	\eIf{$\exists p \in \AllPPs . p \not\in \GetPPSet{j}$}{
-		return NewJob(\AllPPs) // New maximal state space
-	}{
-		return $j$ // $j$ is still maximal
-	}
-	\caption{Na\"ive Iterative Deepening method}
-	\label{alg:algorithm0}
-\end{algorithm}
-
 However, in many tests of even modestly-sized programs,
@@
 based on their preemption points and ETAs.
+
+\newcommand\AllPPs{\ensuremath{\mathcal{A}}}
+\newcommand\PendingJobs{\ensuremath{\mathcal{P}}}
+\newcommand\SuspendedJobs{\ensuremath{\mathcal{S}}}
+\newcommand\GetETA[1]{ETA(#1)}
+\newcommand\GetPPSet[1]{PPSet(#1)}
+
+\begin{algorithm}[t]
+	\KwIn{$j$, the currently-running job}
+	\KwIn{\AllPPs, the set of all known preemption points} %, sorted by decreasing heuristic priority}
+	\eIf{$\exists p \in \AllPPs . p \not\in \GetPPSet{j}$}{
+		return NewJob(\AllPPs) // New maximal state space; switch to it
+	}{
+		return $j$ // $j$ is still maximal
+	}
+	\caption{Na\"ive Iterative Deepening method}
+	\label{alg:algorithm0}
+\end{algorithm}
 
@@
 \begin{algorithm}[h]
-	\SetKwInOut{Input}{Input}
-	\Input{$j_0$, the currently-running job}
-	\Input{\AllJobs, the set of all existing (or completed) jobs}
-	\Input{$\alpha$, an instruction reported by the MC as part of a racing access pair}
+	\KwIn{$j_0$, the currently-running job}
+	\KwIn{\AllJobs, the set of all existing (or completed) jobs}
+	\KwIn{$\alpha$, an instruction reported by the model checker as part of a data race}
 	\If{$\forall j \in \AllJobs,$
@@
 \begin{algorithm}[t]
-	\SetKwInOut{Input}{Input}
 	%\textbf{Function} GetBestJob($j_0$, PendingJobs, SuspendedJobs): \\
-	\Input{$j_0$, the currently-running job}
-	\Input{\PendingJobs, the list of pending jobs, sorted by decreasing heuristic priority}
-	\Input{\SuspendedJobs, the list of already-suspended jobs, sorted by increasing ETA}
-	\Input{$T$, the remaining time in the CPU budget}
+	\KwIn{$j_0$, the currently-running job}
+	\KwIn{\PendingJobs, the list of pending jobs, sorted by decreasing heuristic priority}
+	\KwIn{\SuspendedJobs, the list of already-suspended jobs, sorted by increasing ETA}
+	\KwIn{$T$, the remaining time in the CPU budget}
 	\If{\GetETA{$j_0$} $<$ HeuristicETAFactor $\times$ $T$}{
@@
 	return $j_0$ // \GetETA{$j_0$} was bad, but no other $j$ was better.
-	\caption{Suspending exploration of a state space in favor of a potentially smaller one.}
+	\caption{Suspending exploration of a job in favor of a potentially smaller one.}
 	\label{alg:shouldworkblock}

### 2018-12-24 thesis/quicksand.tex
more revs to quicksand; tweak some figures, phrasings, etc

@@
 		when disambiguation with {\em data races} is necessary.
+	\item {\bf Data-race preemption point}
+		denotes a custom model checker configuration,
+		issued after finding a {\em data race},
+		requesting it to preempt each involved thread just before its racing memory access
+		(\cref{sec:quicksand-dr-pps}).
 	\item {\bf Benign data races} are those that, when reordered in an alternate interleaving,
@@
 
-Given unlimited CPU time for testing,
-one would always want to switch to the new maximal state space whenever adding a new preemption point.
+Given unlimited CPU time, %for testing,
+%one would always want to
+switching to the new maximal state space whenever adding a new preemption point
+\qrevisionminor{would be the quickest way to reach full verification}.
 The maximal state space is guaranteed to subsume all execution sequences reachable in any subset state space,
@@
 For testing user-space code, Quicksand begins with the four
-% TODO: decide
-jobs from \Cref{fig:id}:
-%possible combinations of preemption points from \Cref{fig:pps-statespace}:
+%jobs from \Cref{fig:id}:
+\qrevisionminor{possible combinations of preemption points from \Cref{fig:pps-statespace}:}
 $\{yield\}$,
@@
 $\{yield,unlock\}$,
-and $\{yield,lock,unlock\}$,
+and $\{yield,lock,$ $unlock\}$, % FIXME argh
 By extension, these also induce preemptions on any other primitives
@@
 as shown in \Cref{alg:handledatarace}.
-I will term these {\em data-race preemption points},
-and they will form the foundation of Quicksand's contribution.
+%I will term these {\em data-race preemption points},
+%and they will
+These {\em data-race preemption points} form the foundation of Quicksand's contribution.
 Note that even though a data race may involve two different instructions,
@@
 % TODO check figure camption
-\begin{figure}[t]
+\begin{figure}[h]
 	\begin{center}
@@
 	\end{center}
-	% TODO: short ver
-	\caption{A common execution pattern with {\tt malloc()} that produces false positive data race candidates.}
+	\caption[Reallocation false positive pattern.]
+	{Reallocation false positive pattern arising from
+	{\tt malloc()} returning the same memory that was just {\tt free()}d.}
 	\label{fig:recycle}
@@
 % TODO smae as above
-\begin{figure}[t]
+\begin{figure}[h]
 	\begin{center}
@@
 	\end{center}
-	% TODO: shortver
-	\caption{If a single-pass Limited HB analysis discarded candidates matching the malloc-recycle pattern,
+	\caption[Reallocation-pattern data race that hides a true bug.]
+	{Reallocation-pattern data race that hides a true bug.
+	If a single-pass Limited Happens-Before analysis discarded all data race candidates
+	with intervening reallocations,
 	it would miss the bug in this adversarial program.}

### 2018-12-24 thesis/quicksand.tex
two tiny tweaks

@@
 All source files mentioned in this section live in the {\tt id/} subdirectory of the Landslide repository,
-with the exception of the Landslide extensions listed last.
+with the exception of the Landslide extensions (listed last).
 As \Cref{chap:landslide} was in some sense a developer's guide to Landslide,

### 2018-12-24 thesis/thesis.tex
upd change bar config for qs
shearbox the star in madoka magica
quicksand write intro and about pps
add inco, birb lite fonts; sigbovik easter egg; fix the issue where numbers in urls show up in charter instead of monospace
two tiny tweaks

@@
 \usepackage[xetex]{xechangebar}
+\usepackage[nomessages]{fp} % for xshearbox, see quicksand.tex
 
@@
 \newfontfamily\arabicfont{Amiri-Regular}[Path = ./fonts/, Script = Arabic]
+\setmonofont{Inconsolata-Regular}[Path = ./fonts/]
+\newfontfamily\birbaslo{BirbasloText}[Path = ./fonts/]
+
+% https://tex.stackexchange.com/questions/99770/problem-with-digits-in-urls-when-using-mathspec-and-hyperref/99774#99774
+% make numbers render in monospace in urls, while render in main font in math mode
+\makeatletter
+     \DeclareMathSymbol{0}{\mathalpha}{\eu@DigitsArabic@symfont}{`0}
+     \DeclareMathSymbol{1}{\mathalpha}{\eu@DigitsArabic@symfont}{`1}
+     \DeclareMathSymbol{2}{\mathalpha}{\eu@DigitsArabic@symfont}{`2}
+     \DeclareMathSymbol{3}{\mathalpha}{\eu@DigitsArabic@symfont}{`3}
+     \DeclareMathSymbol{4}{\mathalpha}{\eu@DigitsArabic@symfont}{`4}
+     \DeclareMathSymbol{5}{\mathalpha}{\eu@DigitsArabic@symfont}{`5}
+     \DeclareMathSymbol{6}{\mathalpha}{\eu@DigitsArabic@symfont}{`6}
+     \DeclareMathSymbol{7}{\mathalpha}{\eu@DigitsArabic@symfont}{`7}
+     \DeclareMathSymbol{8}{\mathalpha}{\eu@DigitsArabic@symfont}{`8}
+     \DeclareMathSymbol{9}{\mathalpha}{\eu@DigitsArabic@symfont}{`9}
+\makeatother
 
@@
 
-%\newcommand\revision[1]{#1\xspace}
-\newcommand\revision[1]{\cbstart[0.5em]{}#1\cbend\xspace}
-%\newcommand\revisionminor[1]{#1\xspace}
-\newcommand\revisionminor[1]{\cbstart{}#1\cbend\xspace}
+\newcommand\revision[1]{#1\xspace}
+%\newcommand\revision[1]{\cbstart[0.5em]{}#1\cbend\xspace}
+\newcommand\revisionminor[1]{#1\xspace}
+%\newcommand\revisionminor[1]{\cbstart{}#1\cbend\xspace}
+% quicksand chapter revisions only
+\newcommand\qrevision[1]{\cbstart[0.5em]{}#1\cbend\xspace}
+\newcommand\qrevisionminor[1]{\cbstart{}#1\cbend\xspace}
 
@@
 \definecolor{commentblue}{RGB}{0,0,192}
+% colors copied back from snake fight for quicksand example codez
+\definecolor{darkorange}{RGB}{192,128,0}
+\definecolor{darkblue}{RGB}{0,97,192}
+\definecolor{assertfail}{RGB}{224,0,56} % was 'pinkish' in snakefight
 % colors from sigbovik for tsx/etc
@@
 and 34\% to fix all such bugs,
-before their project deadline.}
+before their project deadlines.}
 Along the way, I justified \revisionminor{Landslide's} applicability to problems beyond CMU's walls as well

### 2018-12-24 thesis/overview.tex
assorted minor proofreads
consistency of function ()s in overview

@@
 	\begin{tabular}{cc}
-		\begin{tabular}{p{0.45\textwidth}p{0.5\textwidth}}
-			{\small
+		\begin{tabular}{p{0.44\textwidth}p{0.495\textwidth}}
+			{
 			\begin{tabular}{l}
@@
 			&
-			{\small \begin{tabular}{ll}
+			{\begin{tabular}{ll}
 				{\bf Thread 1} & {\bf Thread 2} \\
@@
 
-\section{Stateless Model Checking}
+\section{Stateless model checking}
 \label{sec:overview-stateless} % ugh this label name
@@
 \end{itemize}
-Other than {\tt thr\_init} (which is necessarily single-threaded), several concurrency errors between any two (or all three) of these functions are very common in student submissions.
+Other than {\tt thr\_init()} (which is necessarily single-threaded),
+concurrency errors between any two (or all three) of these functions are very common in student submissions.
 
@@
 \begin{itemize}
-	\item Semaphores (the basic concurrency primitive, implemented using direct scheduler calls): {\tt sema\_up}, {\tt sema\_down}, {\tt sema\_try\_down};
-	\item Locks (which wrap a semaphore initialized to 1), {\tt lock\_acquire}, {\tt lock\_\allowbreak{}release}, {\tt lock\_try\_acquire};
-	\item Condition variables (also implemented using scheduler calls): {\tt cond\_wait}, {\tt cond\_\allowbreak{}signal}, {\tt cond\_broadcast}, with the same semantics as Pebbles P2 condvars;
-	\item Basic round-robin scheduling facilities: {\tt thread\_block} (a kernel-level analogue to Pebbles's {\tt deschedule}), {\tt thread\_yield}
-	\item Kernel thread lifecycle management, {\tt thread\_create} and {\tt thread\_exit}, including stack space memory management;
+	\item Semaphores (the basic concurrency primitive, implemented using direct scheduler calls): {\tt sema\_up()}, {\tt sema\_down()}, {\tt sema\_try\_down()};
+	\item Locks (which wrap a 1-initialized semaphore), {\tt lock\_acquire()}, {\tt lock\_\allowbreak{}release()}, {\tt lock\_try\_acquire()};
+	\item Condition variables (also implemented using scheduler calls): {\tt cond\_wait()}, {\tt cond\_\allowbreak{}signal()}, {\tt cond\_broadcast()}, with the same semantics as Pebbles P2 condvars;
+	\item Basic round-robin scheduling facilities: {\tt thread\_block()} (a kernel-level analogue to Pebbles's {\tt deschedule}), {\tt thread\_yield()}
+	\item Kernel thread lifecycle management, {\tt thread\_create()} and {\tt thread\_exit()}, including stack space memory management;
 	\item Interrupt and fault handlers;
-	\item A page allocator, {\tt palloc\_get\_page}, {\tt palloc\_get\_multiple}, {\tt palloc\_free\_page}, and {\tt palloc\_free\_multiple}
+	\item A page allocator, {\tt palloc\_get\_page()}, {\tt palloc\_get\_multiple()}, {\tt palloc\_free\_page()}, and {\tt palloc\_free\_multiple()}.
 \end{itemize}

### 2018-12-25 thesis/quicksand.tex
finish quicksand chapter

@@
 	\begin{center}
-		\begin{tabular}{p{0.8\textwidth}}
+		\begin{tabular}{p{0.9\textwidth}}
 			\begin{center}
@@
 			\\
+			\begin{center}
 			(a) Example program with synchronization API calls highlighted.
-			\\
+			\end{center}
 			\\
 			\begin{center}
-			\begin{tabular}{llc}
-				{\bf Thread 1} & {\bf Thread 2} & {\bf Value of {\tt count}} \\
+			\begin{tabular}{llllc}
+				& {\bf Thread 1} & & {\bf Thread 2} & {\bf Value of {\tt count}} \\
 				\hline
-				\texttt{\hilight{darkorange}{mutex\_lock}(\&m);} & & 0 \\
-				\texttt{int tmp = count;} & & 0 \\
-							& \texttt{atomic\_xadd(\&count, \const{1});} & 1 \\
-							& \texttt{\hilight{olivegreen}{yield}();} & 1 \\
-							& \texttt{atomic\_xadd(\&count, \const{1});} & 2 \\
-				\texttt{count = tmp + 1;} & & 1 \\
-							& \texttt{\hilight{assertfail}{assert(count >= 2);}} & 1 \\
-				\texttt{\hilight{darkblue}{mutex\_unlock}(\&m);} & & 1 \\
-				\texttt{assert(count >= \const{1});} & & 1 \\
+				11 & \texttt{\hilight{darkorange}{mutex\_lock}(\&m);} & & & 0 \\
+				12a & \texttt{int tmp = count;} & & & 0 \\
+							& & 21 & \texttt{atomic\_xadd(\&count, \const{1});} & 1 \\
+							& & 22 & \texttt{\hilight{olivegreen}{yield}();} & 1 \\
+							& & 23 & \texttt{atomic\_xadd(\&count, \const{1});} & 2 \\
+				12b & \texttt{count = tmp + 1;} & & & 1 \\
+				13 & \texttt{\hilight{darkblue}{mutex\_unlock}(\&m);} & & & 1 \\
+				14 & \texttt{assert(count >= \const{1});} & & & 1 \\
+							& & 24 & \texttt{\hilight{assertfail}{assert(count >= 2);}} & 1 \\
 			\end{tabular}
@@
 			\\
-			(b) Interleaving of (a) in which Thread 2's assertion fails,
-			as the single increment of Thread 1 overwrites both those of Thread 2.
+			%(b) Interleaving of (a) in which Thread 2's assertion fails,
+			%as the single increment of Thread 1 overwrites both those of Thread 2.
+			(b) Buggy interleaving of (a) in which
+			the single increment of Thread 1 overwrites both those of Thread 2.
+			Note the data-race preemption at 12a-12b.
 		\end{tabular}
@@
 	\begin{center}
-		% FIXME: fix it to be a real tree
-		\includegraphics[width=0.9\textwidth]{../oopsla/tree-maximal-only.pdf}
+		\includegraphics[width=0.98\textwidth]{dr-statespace.pdf}
 	\end{center}
@@
 	{State space of \Cref{fig:pps-example} with synchronization preemption points only.
-	Note that none of these interleavings preempt Thread 1 at the necessary place
+	Note that none of these interleavings preempt Thread 1 at the necessary place (between 12a and 12b)
 	to trigger Thread 2's assertion failure.}
@@
 
-%\subsection{Design}
-
-% TODO: put fig:id here, explain generally multiple state spaces
-
 %Note that
@@
 A model checker is still used to test each state space, and other reduction techniques such as DPOR
+(\cref{sec:landslide-dpor})
 are still applicable in each.
@@
 
-% TODO: fix this section, the way it lists the pps is fucked, figure reference prob dnagling
 Iterative Deepening must be seeded with a set of initial state spaces,
@@
 so the {\tt yield} preemption point is always implicitly enabled.
-
+%
 For kernel-level testing, interrupt-disabling is analogous to locking,
@@
 As a heuristic, it doesn't test every intermediate subset such as $\{lock,sti\}$,
-which would result in $2^p$ jobs right off the bat,
+which would result in $2^p$ jobs, %right off the bat,
 although this could potentially be improved in future work (\cref{sec:quicksand-discussion-subsets}).
@@
 	\begin{center}
-        \includegraphics[width=0.9\textwidth]{dr-jobs-v2.pdf}
+        \includegraphics[width=0.9\textwidth]{dr-jobs-v3.pdf}
 	\end{center}
@@
 
-% TODO: fix syntax hilight
-% TODO check figure placement
-% TODO check figure camption
 \begin{figure}[h]
@@
 	\begin{tabular}{rll}
-		& \multicolumn{2}{c}{\texttt{struct x \{ int foo; int baz; \} *x;}} \\
-		& \multicolumn{2}{c}{\texttt{struct y \{ int bar; \} *y;~~~~~~~~~~}} \\
+		& \multicolumn{2}{c}{\texttt{\ctype{struct x} \{ \ctype{int} foo; \ctype{int} baz; \} *x;}} \\
+		& \multicolumn{2}{c}{\texttt{\ctype{struct y} \{ \ctype{int} bar; \} *y;~~~~~~~~~}} \\
 		\\
 		& {\bf Thread 1} & {\bf Thread 2} \\
-		1 & \texttt{\hilight{brickred}{x->foo = ...;}} & \\
-		2 & \texttt{\hilight{olivegreen}{free}(x);} \\
-		3 & & \texttt{\hilight{commentblue}{// x's memory reallocated}} \\
-		4 & & \texttt{y~=~\hilight{olivegreen}{malloc}(sizeof *y);} \\
-		5 & & \texttt{\hilight{commentblue}{// ...initialize...}}\\
-		6 & & \texttt{publish(y);} \\
-		7 & & \texttt{\hilight{brickred}{y->bar = ...;}} \\
+		1 & \texttt{\hilight{assertfail}{x->foo = ...;}} & \\
+		2 & \texttt{\call{free}(x);} \\
+		3 & & \texttt{\ccomment{// x's memory reallocated}} \\
+		4 & & \texttt{y~=~\call{malloc}(\flow{sizeof} *y);} \\
+		5 & & \texttt{\ccomment{// ...initialize...}}\\
+		6 & & \texttt{\call{publish}(y);} \\
+		7 & & \texttt{\hilight{assertfail}{y->bar = ...;}} \\
 	\end{tabular}
@@
 
-% TODO smae as above
 \begin{figure}[h]
@@
 		& {\bf Thread 1} & {\bf Thread 2} \\
-		1 & \texttt{publish(x);} & \\
-		2 & \texttt{\hilight{brickred}{x->foo = ...;}} & \\
-		3 & \texttt{\hilight{olivegreen}{free}(x);} \\
-		4 & & \texttt{x2 = get\_published\_x();} \\
-		5 & & \texttt{\hilight{commentblue}{// x's memory recycled}} \\
-		6 & & \texttt{y~=~\hilight{olivegreen}{malloc}(sizeof *y);} \\
-		7 & & \texttt{\hilight{brickred}{x2->foo = ...;}} \\
+		1 & \texttt{\call{publish}(x);} & \\
+		2 & \texttt{\hilight{assertfail}{x->foo = ...;}} & \\
+		3 & \texttt{\call{free}(x);} \\
+		4 & & \texttt{x2 = \call{get\_published\_x}();} \\
+		5 & & \texttt{\ccomment{// x's memory recycled}} \\
+		6 & & \texttt{y~=~\call{malloc}(\flow{sizeof} *y);} \\
+		7 & & \texttt{\hilight{assertfail}{x2->foo = ...;}} \\
 	\end{tabular}
@@
 rather than to justify every internal step within the proof structure.%
-\footnote{Also because I am already over 200 pages in this thesis, all told.}
+\footnote{Also because this thesis is long enough already.}
 % footnote yeah, i'm writing this last
@@
 Any such program must contain an access $a_1$ by one thread T1,
-followed by a {\tt free} and a {\tt malloc} possibly by either thread,
+followed by a {\tt free()} and a {\tt malloc()} possibly by either thread,
 followed by an access $a_2$ by the other thread T2,
 not depending on the result of the middle malloc.
-Without loss of generality, fix T1 to perform the {\tt free} and T2 the subsequent {\tt malloc},
+Without loss of generality, fix T1 to perform the {\tt free()} and T2 the subsequent {\tt malloc()},
 as shown in \Cref{fig:recycle-bug}.
 The other cases are similar,
-although note that if T2 performs the {\tt free},
-and the {\tt malloc} is reordered before it,
+although note that if T2 performs the {\tt free()},
+and the {\tt malloc()} is reordered before it,
 T2's final access will be a use-after-free immediately.
-Let us also assume the only way for the program to get pointers to heap memory is through {\tt malloc};
+Let us also assume the only way for the program to get pointers to heap memory is through {\tt malloc()};
 hence, there must also be some ``publish'' action $p$ by T1 which communicates the address to T2.
@@
 DPOR will reorder $a_2$ before $a_1$, while not changing $a_2$'s location.
-As T2's {\tt malloc} now occurs before T1's {\tt free}, it will allocate different memory.
+As T2's {\tt malloc()} now occurs before T1's {\tt free()}, it will allocate different memory.
 Hence $a_1$ and $a_2$ %will be in the same allocation;
@@
 Note that this proof does not rely on the existence of preemption points on
-the internal lock of {\tt malloc} or {\tt free},
+the internal lock of {\tt malloc()} or {\tt free()},
 which is an ideal candidate to ignore via {\tt without\_function} (\cref{sec:landslide-pps})
@@
 of \cref{sec:quicksand-id}'s more abstract algorithm descriptions:
-\begin{itemize}
-	\item Jobs 0, 1, 2, and 3 are the initially-seeded state spaces (\cref{sec:quicksand-initial-pps}).
-	\item Job 2 reports a bug found, and shows the filename of the HTML preemption trace
-		(\cref{sec:landslide-bugreport}, \Cref{fig:bugreport})
-		which the user should examine to diagnose it.%
-		\footnote{I actually cheated by copy/pasting this job alone from a different run of Quicksand;
-		the other jobs come from a test with the bug already fixed,
-		in order that exploration progress far enough to defer some jobs for the sake of example.
-		In a real execution, the superset jobs 3, 5, 7, 8, 9, and 11 would be cancelled.}
-	\item Jobs 4 and 6 are the ``small'' jobs added to test the two data races in isolation;
-		5 and 7 are the corresponding ``large'' jobs (\cref{sec:quicksand-dr-pps}).
-	\item Job 11 is the maximal state space, containing all synchronization preemption points
-		and both currently-known data races.
-	\item The intermediate jobs 8, 9, and 10 have been suspended for having ETAs
-		at least twice as large as the provided CPU budget (1 hour),
-		according to the ETA scaling factor heuristic (\cref{sec:quicksand-heuristics}).
-	\item Note that job 11's ETA is currently lower than 8's, 9's, and 10's, despite being a strict superset of each.
-		This corresponds to the surprising ETA inversion situation discussed in \cref{sec:quicksand-choosing}:
-		Quicksand simply hasn't made as much progress into job 11 (compare their percentage estimates rather than ETAs)
-		for its ETA to be accurate enough yet.
-\end{itemize}
-Future work could extend these progress reports to be more interactive,
-allowing the user to reprioritize state spaces at her whim
-or to disable certain data-race preemption points after checking them by hand,
-as discussed in \cref{sec:future-friendly}.
-
-% TODO: syn hi
+
 \begin{figure}[t]
@@
 
-		\texttt{==== PROGRESS REPORT ====} \\
-		\texttt{total time elapsed: 2m 52s} \\
-		\texttt{[JOB 0] COMPLETE (4 interleavings tested; 7s elapsed)} \\
-		\texttt{~~~~PPs: \{ \}} \\
-		\texttt{[JOB 1] Running (21.932870\%; ETA 8m 14s)} \\
-		\texttt{~~~~ PPs: \{ 'mutex\_lock' \}} \\
+		\texttt{\ccomment{==================================== PROGRESS REPORT ====================================}} \\
+		\texttt{\flow{total time elapsed: 2m 52s}} \\
+		\texttt{\ccomment{[JOB 0]} \ctype{COMPLETE (4 interleavings tested; 7s elapsed)}} \\
+		\texttt{~~~~\ccomment{PPs: \{ \}}} \\
+		\texttt{\ccomment{[JOB 1]} \flow{Running (21.932870\%; ETA 8m 14s)}} \\
+		\texttt{~~~~ \ccomment{PPs: \{ 'mutex\_lock' \}}} \\
 		% version that came from the actual test
-		%\texttt{[JOB 2] Running (46.373457\%; ETA 4m 35s)} \\
-		%\texttt{~~~~ PPs: \{ 'mutex\_unlock' \}} \\
-		\texttt{[JOB 2] BUG FOUND: landslide-trace-1544661430.29.html (51 interleavings tested)} \\
+		%\texttt{\ccomment{[JOB 2]} Running (46.373457\%; ETA 4m 35s)} \\
+		%\texttt{~~~~ \ccomment{PPs: \{ 'mutex\_unlock' \}}} \\
+		\texttt{\ccomment{[JOB 2]} \hilight{assertfail}{BUG FOUND: landslide-trace-1544661430.29.html (51 interleavings tested)}} \\
 		%\texttt{~~~~Log of lprintf()s: id/ls-setup.log.7UMGLr} \\
-		\texttt{~~~~ PPs: \{ 'mutex\_unlock' \}} \\
-		\texttt{[JOB 3] Running (9.852431\%; ETA 24m 32s)} \\
-		\texttt{~~~~ PPs: \{ 'mutex\_lock' 'mutex\_unlock' \}} \\
-		\texttt{[JOB 4] COMPLETE (6 interleavings tested; 9s elapsed)} \\
-		\texttt{~~~~PPs: \{ 'data race @ 0x102917' \}} \\
-		\texttt{[JOB 5] Running (3.710938\%; ETA 1h 2m 14s)} \\
-		\texttt{~~~~ PPs: \{ 'mutex\_lock' 'mutex\_unlock' 'data race @ 0x102917' \}} \\
-		\texttt{[JOB 6] COMPLETE (4 interleavings tested; 8s elapsed)} \\
-		\texttt{~~~~PPs: \{ 'data race @ 0x1000ecf' \}} \\
-		\texttt{[JOB 7] Running (6.119792\%; ETA 33m 14s)} \\
-		\texttt{~~~~ PPs: \{ 'mutex\_lock' 'mutex\_unlock' 'data race @ 0x1000ecf' \}} \\
-		\texttt{[JOB 11] Running (3.670247\%; ETA 50m 16s)} \\
-		\texttt{~~~~ PPs: \{ 'mutex\_lock' 'mutex\_unlock' 'data race @ 0x102917' 'data race @ 0x1000ecf' \}} \\
-		\texttt{[JOB 8] Deferred... (33.340567\%; ETA 2h 6m 3s)} \\
-		\texttt{~~~~ PPs: \{ 'mutex\_unlock' 'data race @ 0x102917' \}} \\
-		\texttt{[JOB 9] Deferred... (34.466226\%; ETA 2h 35m 37s)} \\
-		\texttt{~~~~ PPs: \{ 'mutex\_unlock' 'data race @ 0x1000ecf' \}} \\
-		\texttt{[JOB 10] Deferred... (11.113790\%; ETA 4h 20m 31s)} \\
-		\texttt{~~~~ PPs: \{ 'mutex\_lock' 'data race @ 0x102917' \}} \\
+		\texttt{~~~~ \ccomment{PPs: \{ 'mutex\_unlock' \}}} \\
+		\texttt{\ccomment{[JOB 3]} \flow{Running (9.852431\%; ETA 24m 32s)}} \\
+		\texttt{~~~~ \ccomment{PPs: \{ 'mutex\_lock' 'mutex\_unlock' \}}} \\
+		\texttt{\ccomment{[JOB 4]} \ctype{COMPLETE (6 interleavings tested; 9s elapsed)}} \\
+		\texttt{~~~~\ccomment{PPs: \{ 'data race @ 0x102917' \}}} \\
+		\texttt{\ccomment{[JOB 5]} \flow{Running (3.710938\%; ETA 1h 2m 14s)}} \\
+		\texttt{~~~~ \ccomment{PPs: \{ 'mutex\_lock' 'mutex\_unlock' 'data race @ 0x102917' \}}} \\
+		\texttt{\ccomment{[JOB 6]} \ctype{COMPLETE (4 interleavings tested; 8s elapsed)}} \\
+		\texttt{~~~~\ccomment{PPs: \{ 'data race @ 0x1000ecf' \}}} \\
+		\texttt{\ccomment{[JOB 7]} \flow{Running (6.119792\%; ETA 33m 14s)}} \\
+		\texttt{~~~~ \ccomment{PPs: \{ 'mutex\_lock' 'mutex\_unlock' 'data race @ 0x1000ecf' \}}} \\
+		\texttt{\ccomment{[JOB 11]} \flow{Running (3.670247\%; ETA 50m 16s)}} \\
+		\texttt{~~~~ \ccomment{PPs: \{ 'mutex\_lock' 'mutex\_unlock' 'data race @ 0x102917' 'data race @ 0x1000ecf' \}}} \\
+		\texttt{\ccomment{[JOB 8]} \call{Deferred... (33.340567\%; ETA 2h 6m 3s)}} \\
+		\texttt{~~~~ \ccomment{PPs: \{ 'mutex\_unlock' 'data race @ 0x102917' \}}} \\
+		\texttt{\ccomment{[JOB 9]} \call{Deferred... (34.466226\%; ETA 2h 35m 37s)}} \\
+		\texttt{~~~~ \ccomment{PPs: \{ 'mutex\_unlock' 'data race @ 0x1000ecf' \}}} \\
+		\texttt{\ccomment{[JOB 10]} \call{Deferred... (11.113790\%; ETA 4h 20m 31s)}} \\
+		\texttt{~~~~ \ccomment{PPs: \{ 'mutex\_lock' 'data race @ 0x102917' \}}} \\
 		% no, i deleted all of them pretending the other data races dont exist
 		%\texttt{And 63 more pending jobs should time allow.} \\
-		\texttt{=========================} \\
+		\texttt{\ccomment{=========================================================================================}} \\
 	\end{tabular}
@@
 \end{figure}
+
+\begin{itemize}
+	\item Jobs 0, 1, 2, and 3 are the initially-seeded state spaces (\cref{sec:quicksand-initial-pps}).
+	\item Job 2 reports a bug found, and shows the filename of the HTML preemption trace
+		(\cref{sec:landslide-bugreport}, \Cref{fig:bugreport})
+		which the user should examine to diagnose it.%
+		\footnote{I actually cheated by copy/pasting this job alone from a different run of Quicksand;
+		the other jobs come from a test with the bug already fixed,
+		in order that exploration progress far enough to defer some jobs for the sake of example.
+		In a real execution, the superset jobs 3, 5, 7, 8, 9, and 11 would be cancelled.}
+	\item Jobs 4 and 6 are the ``small'' jobs added to test the two data races in isolation;
+		5 and 7 are the corresponding ``large'' jobs (\cref{sec:quicksand-dr-pps}).
+	\item Job 11 is the maximal state space, containing all synchronization preemption points
+		and both currently-known data races.
+	\item The intermediate jobs 8, 9, and 10 have been suspended for having ETAs
+		at least twice as large as the provided CPU budget (1 hour),
+		according to the ETA scaling factor heuristic (\cref{sec:quicksand-heuristics}).
+	\item Note that job 11's ETA is currently lower than 8's, 9's, and 10's, despite being a strict superset of each.
+		This corresponds to the ETA inversion situation discussed in \cref{sec:quicksand-choosing}:
+		Quicksand simply hasn't made as much progress into job 11 (compare their percentage estimates rather than ETAs)
+		for its ETA to be accurate enough yet.
+\end{itemize}
+Future work could extend these progress reports to be more interactive,
+allowing the user to reprioritize state spaces at her whim
+or to disable certain data-race preemption points after checking them by hand,
+as discussed in \cref{sec:future-friendly}.
 
@@
 	\begin{center}
-        \includegraphics[width=0.9\textwidth]{../proposal/dowefindbugsfaster-v2.pdf} \\
+        \includegraphics[width=0.96\textwidth]{../proposal/dowefindbugsfaster-v2.pdf} \\
 	\end{center}
@@
 	\begin{center}
-        \includegraphics[width=0.9\textwidth]{../proposal/dowefindbugsfaster-wallclock-v2.pdf} \\
+        \includegraphics[width=0.96\textwidth]{../proposal/dowefindbugsfaster-wallclock-v2.pdf} \\
 	\end{center}
@@
 	\begin{center}
-	\includegraphics[width=0.9\textwidth]{../proposal/totalverifs-v2.pdf}
+	\includegraphics[width=0.96\textwidth]{../proposal/totalverifs-v2.pdf}
 	\end{center}
@@
 The other 192 represent cases where neither Quicksand nor ICB were able to provide a conclusive result either way.
-\footnote{Thesis note: SSS-MC-Shared-Mem was added in a subsequent revision to \cite{quicksand}
-from when this analysis was conducted, at which time SSS-MC-ICB was the best-performing approach among control experiments.
+\footnote{Thesis note: SSS-MC-Shared-Mem was added
+%in a subsequent revision to \cite{quicksand}
+%from when this analysis was conducted
+subsequently to this analysis's publication in \cite{quicksand},
+at which time SSS-MC-ICB was the best-performing approach among control experiments.
 Nevertheless, mutual timeouts among all six testing approaches constituted roughly one third of the test suite.}
@@
 	\begin{center}
-	\includegraphics[width=0.9\textwidth]{../proposal/nondets.pdf} % TODO: get some gf colors on here
+		\includegraphics[width=0.88\textwidth]{nondets.pdf}
 	\end{center}
@@
 \cref{sec:quicksand-eval-bugs} and \cref{sec:quicksand-eval-verif}
-could possibly be conflated by Landslide's implementation of ICB and/or DPOR being slower
+could possibly have been conflated by Landslide's implementation of ICB and/or DPOR being slower
 on account of the simulated execution environment (\cref{sec:landslide-architecture}).
@@
 Concurrently with these results' publication in OOPSLA \cite{quicksand},
-more advanced state space reduction algorithms were proposed, such as Maximal Causality Reduction (MCR) \cite{mcr}.
-It is not immediately obvious that MCR's benefit would be orthogonal with Iterative Deepening;
-i.e., the benefit of Quicksand with a MCR-enabled model checker might be reduced
+more advanced state space reduction algorithms were proposed,
+such as
+%Maximal Causality Reduction (MCR)
+MCR \cite{mcr} and RCMC \cite{rcmc} (\cref{sec:related-smc}).
+It is not immediately obvious that these techniques' benefits would be orthogonal with Iterative Deepening;
+in other words, the benefit of Quicksand with a MCR- or RCMC-enabled model checker might be reduced
 compared to the benefit shown in \cref{sec:quicksand-eval-bugs} and \cref{sec:quicksand-eval-verif}.
-Future comparisons of model-checking strategies should strive to reach these higher bars of scientific rigor.}
+Future model checkers' evaluations should strive to fairly represent these and other latest algorithms in their comparisons.}
+%Future comparisons of model-checking strategies should strive to reach these higher bars of scientific rigor.}
 
@@
 		with far fewer preemption points necessary.
-	\item A technique for suppressing certain false positive data race candidates % TODO: hm, should false-positive be hyph?
+	\item A technique for suppressing certain false positive data race candidates
 		under Limited Happens-Before (\cref{sec:landslide-lhb})

### 2018-12-26 thesis/education.tex
やはり data race analysis
ensure parens on malloc()/free()

@@
 		This is likely due to exceeding Linux's process limit and/or exhausting the class's official VM's memory,
-		which can arise when many data-race candidates cause Quicksand to spawn many Landslide jobs,
+		which can arise when many data race candidates cause Quicksand to spawn many Landslide jobs,
 		and (as on {\tt paraguay}) each with very large state spaces that must defer.
@@
 	\item Landslide issued false invalid heap access reports to one student
-		who had been using {\tt new\_pages} to allocate thread stacks ``very close'' to the {\tt malloc} heap.
+		who had been using {\tt new\_pages} to allocate thread stacks ``very close'' to the {\tt malloc()} heap.
 		They reported this during the study and I fixed it promptly for them in commit 4a26da7 (Bochs repository only),

### 2018-12-26 thesis/landslide.tex
ensure parens on malloc()/free()

@@
 			This could be solved in future work with a new annotation.}
-	\item {\tt ALLOW\_REENTRANT\_MALLOC\_FREE=0} - allow two threads to be in {\tt malloc}, {\tt free}, or so on simultaneously without declaring it a bug.%
+	\item {\tt ALLOW\_REENTRANT\_MALLOC\_FREE=0} - allow two threads to be in {\tt malloc()}, {\tt free()}, or so on simultaneously without declaring it a bug.%
 		\footnote{Used in Pintos, where those functions lock/unlock the heap mutex themselves rather than relying on a wrapper function to do so before invoking them.}
@@
 {\tt mem\_update()} serves as the main entrypoint for tracking heap allocations.
-It's called every instruction to check for the boundaries of the {\tt malloc} library,
+It's called every instruction to check for the boundaries of the {\tt malloc()} library,
 and behaves in a similar way to the scheduler state machine described above.
@@
 	\item {\tt memory.c} (various functions)
-		will handle page allocations from the {\tt palloc} family of functions in a separate memory heap,
-		allowing the usual allocator {\tt malloc} to allocate and free from {\tt palloc}'ed memory,
+		will handle page allocations from the {\tt palloc()} family of functions in a separate memory heap,
+		allowing the usual allocator {\tt malloc()} to allocate and free from {\tt palloc()}ed memory,
 		and checking both allocation heaps when checking for use-after-frees.
@@
 	\item {\tt getfunc.sh} defines several functions commonly used by {\tt build.sh} and {\tt definegen.sh} to extract function or global variable addresses from the program binary.
-	\item {\tt symbols.sh} defines the names of kernel functions that can be instrumented automatically without a corresponding manual annotation (e.g., {\tt malloc} and friends, the names of the {\tt tell\_landslide} family, various library helpers such as {\tt panic}).
+	\item {\tt symbols.sh} defines the names of kernel functions that can be instrumented automatically without a corresponding manual annotation (e.g., {\tt malloc()} and friends, the names of the {\tt tell\_landslide} family, various library helpers such as {\tt panic()}).
 	\item {\tt build.sh} ensures the build of Landslide is up-to-date,
@@
 
-		Rejects stack trace 1 (last matching directive is to exclude {\tt malloc}),
-		accepts stack trace 2 (last matching directive is to include {\tt mutex\_lock})
+		Rejects stack trace 1 (last matching directive is to exclude {\tt malloc()}),
+		accepts stack trace 2 (last matching directive is to include {\tt mutex\_lock()}).
 	\item
@@
 		Accepts both (last matching directive is to accept {\tt mutex\_lock()},
-		regardless of {\tt main()})
+		regardless of {\tt main()}).
 \end{itemize}

### 2018-12-26 thesis/overview.tex
ensure parens on malloc()/free()

@@
 \end{itemize}
-Both Pebbles and Pintos basecodes offer a standard C library including {\tt malloc}, string-formatting, printing, et cetera.
+Both Pebbles and Pintos basecodes offer a standard C library including {\tt malloc()}, string-formatting, printing, et cetera.
 
@@
 
-Some schools further offer a virtual memory project, extending the provided VM with a frame table and supplemental page table and fault handler \cite{standford-cs140,uchicago-cs230}, or supplemental HTTP server and {\tt malloc} assignments \cite{berkeley-cs162}.
+Some schools further offer a virtual memory project, extending the provided VM with a frame table and supplemental page table and fault handler \cite{standford-cs140,uchicago-cs230}, or supplemental HTTP server and {\tt malloc()} assignments \cite{berkeley-cs162}.
 \revisionminor{This thesis is not concerned with these assignments.

### 2018-12-26 thesis/quicksand.tex
ensure parens on malloc()/free()

@@
 \Cref{fig:recycle} shows a common code pattern and interleaving which can expose such behaviour.
-If the {\tt malloc} on line~4 returns the same address passed to {\tt free} on line~2,
+If the {\tt malloc()} on line~4 returns the same address passed to {\tt free()} on line~2,
 then lines~1 and 7 will be flagged as a potential data race.
@@
 To the human eye, this is obviously a false positive:
-reordering lines~4-7 before lines~1-2 will cause {\tt malloc} to return a different region of allocated memory,
+reordering lines~4-7 before lines~1-2 will cause {\tt malloc()} to return a different region of allocated memory,
 in turn causing {\tt x} and {\tt y} to no longer collide.
@@
 Consider the more unusual program in \Cref{fig:recycle-bug},
-in which the memory is recycled the same way, but the racing access's address is not tied to {\tt malloc}'s return value.
+in which the memory is recycled the same way, but the racing access's address is not tied to {\tt malloc()}'s return value.
 Here, reordering lines~6-7 before line~3 will allow {\tt x} and {\tt x2} to race.
@@
 which preemption points to suppress (\cref{sec:landslide-pps}),
-especially those arising from the {\tt malloc} lock (\cref{sec:quicksand-realloc}),
+especially those arising from the {\tt malloc()} lock (\cref{sec:quicksand-realloc}),
 which functions DPOR and data race analysis should treat as trusted code (\cref{sec:landslide-config-landslide}),

### 2018-12-26 thesis/education.tex
cpu-time or cpu time

@@
 such as resource exhaustion; see \cref{sec:warpzone-verif}).
-Nevertheless, the responses' overall bias to spending at least half the CPU-time on Landslide
+Nevertheless, the responses' overall bias to spending at least half the CPU time on Landslide
 shows clearly that the students found the experience worthwhile.

### 2018-12-26 thesis/quicksand.tex
cpu-time or cpu time

@@
 
-Given unlimited CPU time, %for testing,
+Given unlimited testing time, %for testing,
 %one would always want to
@@
 To match this,
-Quicksand reports both the CPU-time and wall-clock time spent during its execution.
-Comparing CPU-time leads to a more fair comparison,
+Quicksand reports both the CPU time and wall-clock time spent during its execution.
+Comparing CPU time leads to a more fair comparison,
 although Quicksand's inherent parallelism, which only a wall-clock time comparison would show,
@@
 against the three control approaches
-in a cumulative distribution of total bugs found against elapsed CPU-time.
+in a cumulative distribution of total bugs found against elapsed CPU time.
 The farthest-right point on each series indicates in how many total test cases that trial found a bug

### 2018-12-26 thesis/tm.tex
cpu-time or cpu time

@@
 All tests were run on an 8-core 2.7GHz Core i7 with 32 GB RAM.
-Reported CPU-times include time spent on all state spaces
+Reported CPU times include time spent on all state spaces
 Quicksand saw fit to run,
@@
 		optimized for fast verification.
-		For each, I list the CPU-time and wall-clock time elapsed,
+		For each, I list the CPU time and wall-clock time elapsed,
 		plus the number of interleavings tested in the ultimately buggy state space
@@
 Quicksand's ability to find bugs in fewer distinct interleavings (i.e., overall smaller state spaces)
-does not necessarily correlate with better performance in terms of CPU-time.
+does not necessarily correlate with better performance in terms of CPU time.
 Comparing \Cref{tab:tm-buges}'s trends

### 2018-12-26 thesis/landslide.tex
nobe

@@
 %(or state space, or preemption point log, however calling it suits you)
-is stored as a chain of preemption point nodes named {\tt struct hax} defined in {\tt tree.h}.
+is stored as a chain of preemption point nodes named {\tt struct nobe} defined in {\tt tree.h}.
 Although the state space of possible interleavings is exponentially-sized,
@@
 (for the test cases used in this thesis, typically 20-1000).
-Each {\tt hax} stores the following information:
+Each {\tt nobe} stores the following information:
 
@@
 each of whose state is kept generally read-only,
-and all modifications must go through {\tt modify\_hax()} ({\tt timetravel.h}) using a modification callback,
+and all modifications must go through {\tt modify\_nobe()} ({\tt timetravel.h}) using a modification callback,
 which internally casts away the {\tt const}, performs the requested modification,
@@
 Whenever {\tt arbiter\_interested()} returns true,
-Landslide creates a new {\tt struct hax} in the execution tree (\cref{sec:landslide-save}),
+Landslide creates a new {\tt struct nobe} in the execution tree (\cref{sec:landslide-save}),
 creates a checkpoint (\cref{sec:landslide-timetravel}),
@@
 is unknown until actually executed.}
-Landslide's implementation represents $S$ as a list of {\tt struct hax}es, defined in {\tt tree.h},
+Landslide's implementation represents $S$ as a list of {\tt struct nobe}s, defined in {\tt tree.h},
 each one representing a preemption point, or intermediate state between two transitions.
@@
 Therefore, to avoid repeating interleavings,
-Landslide need only store at each {\tt struct hax} a list of threads
+Landslide need only store at each {\tt struct nobe} a list of threads
 such that their corresponding subtree $U$ is fully explored,
 and can omit any non-constant-size information about the contents of that $U$
-({\tt struct hax\_child}).
+({\tt struct nobe\_child}).
 Hence the memory cost of Landslide's DPOR implementation is $O(nk)$,
@@
 (In the case of a tie, differing only by which thread to run, it chooses arbitrarily.)
-All other new interleavings are marked to explore later, %in their corresponding {\tt struct hax},
+All other new interleavings are marked to explore later, %in their corresponding {\tt struct nobe},
 and automatically included in the result of any future iterations of DPOR until they are tested.

### 2018-12-26 thesis/tm.tex
nobe

@@
 
-Each preemption point ({\tt struct hax} as defined in {\tt tree.h})
+Each preemption point ({\tt struct nobe} as defined in {\tt tree.h})
 includes three new fields.
@@
 ``saturating'' the retry sets to fall back to no pruning
-({\tt update\_hax\_abort\_set()} and
-{\tt update\_hax\_abandon\_abort\_set()}, respectively).
+({\tt update\_nobe\_abort\_set()} and
+{\tt update\_nobe\_abandon\_abort\_set()}, respectively).
 

### 2018-12-26 thesis/landslide.tex
footnote abt no-memory return-value yield-blocking loops

@@
 whatever condition the blocked thread was waiting for
-will show up in the conflicts between it and whichever thread fulfills it.
+will show up in the conflicts between it and whichever thread fulfills it.%
+\footnote{A yield-blocking loop may depend on another thread's execution with no memory conflicts at all
+simply by checking the return value to observe whether the yield was successful.
+In principle, Landslide should check for such loops whenever the yielded-to thread's runnability changes;
+at present, this is not implemented,
+and instead leans on the false positive deadlock detector to wake the blocking thread up.}
 Hence, every memory conflict detected during DPOR is also checked against any

### 2018-12-26 thesis/education.tex
mechanics proofreading - function names, footnote placement, figure optmz, syn hi

@@
 		%defines the length of that array in a macro to be used by {\tt is\_runqueue()}
-		\revisionminor{emits macros to configure {\tt is\_runqueue()}'s behaviour to handle
+		\revisionminor{emits macros to configure {\tt is\_runqueue()} to handle
 		certain common alternate approaches}
 		(part of the patch described below).
-		Either way defines a function {\tt get\_rq\_addr()} to return the address of the %(first)
+		Either way,
+		defines a function {\tt get\_rq\_addr()} to return the address of the %(first)
 		list.
@@
 		Unmodified from the basecode's version, for {\em userprog}.
-		Userspace process {\tt exec()}s a child process, which immediately exits, and {\tt wait()}s on it.
+		Userspace process {\tt exec}s a child process, which immediately exits, and {\tt wait}s on it.
 	\item {\tt wait-twice}:
@@
 		this root cause manifested in heap errors in \revisionminor{three} separate locations.
-		Lastly, the heap checking alone found a deterministic use-after-free in a third group's {\tt exec()}.
+		Lastly, the heap checking alone found a deterministic use-after-free in a third group's {\tt process\_execute()}.
 \end{enumerate}
@@
 }
-I deem the difference in grade distributions significant when $p < 0.05$%
+I deem the difference in grade distributions significant when $p < 0.05$,%
 % FIXME: is this legit?
@@
 because each tests a different hypothesis.
-},
+}
 although as in all $p$-value calculations,
@@
 	\item Improve explanation of data races [in the user guide, perhaps].
-	\item Happy with it as-is [\revisionminor{reported by three} students]
+	\item Happy with it as-is [\revisionminor{reported by three} students].
 		% wow, latex -- if you put [] outside of the revision, everything is awful forever
-	\item \revisionminor{[Seven non-respondents]}
+	\item \revisionminor{[Seven non-respondents.]}
 \end{enumerate}

### 2018-12-26 thesis/future.tex
mechanics proofreading - function names, footnote placement, figure optmz, syn hi

@@
 took until implementing retry set reduction to discover,
-well after publishing the first experimental results under it in \cite{sigbovik-htm}.
+well after publishing the first experimental results in \cite{sigbovik-htm}.
 As the concurrency model becomes more complex,

### 2018-12-26 thesis/quicksand.tex
mechanics proofreading - function names, footnote placement, figure optmz, syn hi

@@
 	\begin{center}
-		\begin{tabular}{p{0.9\textwidth}}
+		\begin{tabular}{p{0.91\textwidth}} % dont hyphenate 'overwrite'
 			\begin{center}
@@
 found bugs in 37.
-The other 192 represent cases where neither Quicksand nor ICB were able to provide a conclusive result either way.
+The other 192 represent cases where neither Quicksand nor ICB were able to provide a conclusive result either way.%
 \footnote{Thesis note: SSS-MC-Shared-Mem was added
@@
 	\begin{center}
-		\includegraphics[width=0.88\textwidth]{nondets.pdf}
+		\includegraphics[width=0.87\textwidth]{nondets.pdf}
 	\end{center}
@@
 	% phb: 191.30%
-	128\% (Limited HB) to 191\% (Pure HB) as many data-race bugs
+	128\% (Limited Happens-Before) to 191\% (Pure Happens-Before) as many data-race bugs
 	compared to using single-pass candidates alone.

### 2018-12-26 thesis/tm.tex
mechanics proofreading - function names, footnote placement, figure optmz, syn hi

@@
 			$2a$ & \texttt{~~~~temp <- x;} \\
-			$2b$ & \texttt{~~~~temp <- temp + 1;} \\
+			$2b$ & \texttt{~~~~temp <- temp + \const{1};} \\
 			$2c$ & \texttt{~~~~x <- temp;} \\
@@
 			$6a$ & \texttt{~~~~temp <- x;} \\
-			$6b$ & \texttt{~~~~temp <- temp + 1;} \\
+			$6b$ & \texttt{~~~~temp <- temp + \const{1};} \\
 			$6c$ & \texttt{~~~~x <- temp;} \\
@@
 			\texttt{temp <- x;} & \\
-			\texttt{temp <- temp + 1;} & \\
+			\texttt{temp <- temp + \const{1};} & \\
 			& \texttt{status = \call{\_xbegin}()} \\
@@
 		\begin{tabular}{l}
-			\texttt{\ctype{static int} num\_in\_section = 0;} \\
+			\texttt{\ctype{static int} num\_in\_section = \const{0};} \\
 			\texttt{\flow{for} (\ctype{int} i = \const{0}; i < \const{NITERS}; i++) \{} \\

### 2018-12-26 thesis/landslide.tex
nobe -> pp for function names

@@
 each of whose state is kept generally read-only,
-and all modifications must go through {\tt modify\_nobe()} ({\tt timetravel.h}) using a modification callback,
+and all modifications must go through {\tt modify\_pp()} ({\tt timetravel.h}) using a modification callback,
 which internally casts away the {\tt const}, performs the requested modification,

### 2018-12-26 thesis/tm.tex
nobe -> pp for function names

@@
 ``saturating'' the retry sets to fall back to no pruning
-({\tt update\_nobe\_abort\_set()} and
-{\tt update\_nobe\_abandon\_abort\_set()}, respectively).
+({\tt update\_pp\_abort\_set()} and
+{\tt update\_pp\_abandon\_abort\_set()}, respectively).
 

### 2018-12-26 thesis/quicksand.tex
figure captions:

@@
 	\end{center}
-	\caption[Nondeterministic data-race preemption points are required to find some bugs.]
-	{Nondeterministic data-race preemption points are required to find some bugs.
-	Requiring stateless model checking integration to identify to begin with,
-	incorporating these races as new preemption points
+	\caption[Some bugs require nondeterministic data-race preemption points to expose.]
+	{Some bugs require nondeterministic data-race preemption points to expose.
+	Incorporating these data races,
+	which are not observed under single-pass analysis,
+	%which stateless model checking integration to identify to begin with,
+	as new preemption points
 	allowed Quicksand to find

### 2018-12-26 thesis/tm.tex
figure captions:

@@
 	\end{center}
-	\caption[Variant of the program in \Cref{fig:htm-example-reproduced}.]
-		{Variant of the program in \Cref{fig:htm-example-reproduced},
+	\caption[More correct version of the program in \Cref{fig:htm-example-reproduced}.]
+		{More correct version of the program in \Cref{fig:htm-example-reproduced},
 		with additional synchronization to protect the failure path from the transactional path.

### 2018-12-26 thesis/tm.tex
change location of LFTs from 410user to landslide-friendly-tests

@@
 which matches GCC's interface exactly,
-located in {\tt 410user/inc/htm.h}
-under {\tt pebsim/p2-basecode/}.
-The interface is implemented in {\tt 410user/libhtm/htm.c},
+located in {\tt landslide-friendly-tests/inc/htm.h}
+under the subdirectory {\tt pebsim/p2-basecode/}.
+The interface is implemented in {\tt landslide-friendly-tests/\allowbreak{}libhtm/htm.c},
 perhaps surprisingly, as totally empty functions;
@@
 on a (hopefully) correct P2 as usual.
-Several example HTM tests are provided as {\tt 410user/progs/htm*.c}.
+Several example HTM tests are provided as {\tt landslide-friendly-tests/htm*.c}.
 
@@
 		to avoid circular causality
-		(see {\tt 410user/progs/htm\_causality.c} for rationale).
+		(see {\tt landslide-friendly-tests/htm\_causality.c} for rationale).
 		Should the transaction happen to be executed first,
@@
 and conflict aborts cannot occur before their conflicting transaction to begin with%
-\footnote{See {\tt 410user/progs/htm\_causality.c}; that was a fun realization to have
+\footnote{See {\tt landslide-friendly-tests/htm\_causality.c}; that was a fun realization to have
 already halfway into implementing retry sets the wrong way at first.}%
@@
 I would urge any reader interested in using those to add them in by hand first.
-The test case {\tt lock}({\tt mutex}) ({\tt 410user/progs/htm\_mutex.c} in the repository)
+The test case {\tt lock}({\tt mutex}) ({\tt landslide-friendly-tests/htm\_mutex.c} in the repository)
 provides an example of how to use compiler intrinsics to emit the necessary barriers.

### 2018-12-26 thesis/landslide.tex
consistency of preempt-everywhere and maximal state space modes decapitalized

@@
 		(\cref{sec:landslide-icb})
-	\item {\tt -0}: enable Preempt-Everywhere mode
+	\item {\tt -0}: enable preempt-everywhere mode
 		(\cref{sec:quicksand-impl-modes}, \cref{sec:quicksand-impl-landslide}, requires {\tt -C})
-	\item {\tt -M}: enable Maximal State Space mode,
+	\item {\tt -M}: enable maximal state space mode,
 		which prioritizes the maximal state space to optimize for fast verification,
@@
 		corresponds to {\tt -I} in \cref{sec:landslide-quicksand-options}.
-	\item {\tt PREEMPT\_EVERYWHERE=0} - enable Preempt-Everywhere mode (\cref{sec:quicksand-eval});
+	\item {\tt PREEMPT\_EVERYWHERE=0} - enable preempt-everywhere mode (\cref{sec:quicksand-eval});
 		corresponds to {\tt -0} in \cref{sec:landslide-quicksand-options}.
@@
 
-In Preempt-Everywhere mode, Landslide instead marks all shared memory accesses
+In preempt-everywhere mode, Landslide instead marks all shared memory accesses
 as long as they are not part of either the mutex implementation or the running thread's stack frame.

### 2018-12-26 thesis/thesis.tex
turn off changebars
update acks

@@
 \usepackage{xeCJK}
-\usepackage[xetex]{xechangebar}
+%\usepackage[xetex]{xechangebar}
 \usepackage[nomessages]{fp} % for xshearbox, see quicksand.tex
@@
 % quicksand chapter revisions only
-\newcommand\qrevision[1]{\cbstart[0.5em]{}#1\cbend\xspace}
-\newcommand\qrevisionminor[1]{\cbstart{}#1\cbend\xspace}
+%\newcommand\qrevision[1]{\cbstart[0.5em]{}#1\cbend\xspace}
+%\newcommand\qrevisionminor[1]{\cbstart{}#1\cbend\xspace}
+\newcommand\qrevision[1]{#1\xspace}
+\newcommand\qrevisionminor[1]{#1\xspace}
 
@@
 Ji\v{r}\'{i} \v{S}im\v{s}a, another former student of Garth's,
-first sparked my interest in the field,
-and then gave invaluable support during my first few years.
+%first sparked my interest in the field,
+%and then
+gave invaluable support during my first few years.
 Ji\v{r}\'{i}'s mentorship
-oriented me in the world of research and helped me find my bearings as a grad student,
+oriented me in the world of research,
+%and helped me find my bearings as a grad student,
 from teaching me how DPOR works
@@
 Michael J. Sullivan and Nathaniel Wesley Filardo were Landslide's first other users
-%besides myself
+besides myself
 back in its prototype days.
@@
 and helped to improve Landslide's
-%stability and
+stability and
 robustness for use beyond CMU's walls,
@@
 David Blum and Skye Toor provided
-invaluable advice towards the statistical analysis
-used therein.
-%that helped me evaluate the student
-%grade distributions therein.
+invaluable advice on statistical analysis
+that helped me evaluate the student
+grade distributions therein.
 Mario Dehesa-Azuara generously contributed the transactional
@@
 memory data structures
-that make up about half of \Cref{chap:tm}'s evaluation suite.
+that make up about half of \Cref{chap:tm}'s evaluation
+suite.
 %
@@
 Chris Martens,
-Jason Reed,
+jcreed,
 Jim McCann,
@@
 and
-William Lovas:
+William ``wjl'' Lovas:
 you each have been role models to me
-perhaps more than you know.
+far more than you know.
 
@@
 Looking back on my 11 years at CMU,
-I am especially grateful for the friendships made early on %in my undergrad years
+I am especially grateful for the early friendships %made early on %in my undergrad years
 that continued well past first graduation.
 From those who stayed in Pittsburgh %with me
-to those who moved away but % can comment 'moved away but'
+to those who %moved away but % can comment 'moved away but'
 kept in touch,
-from my regular tabletop, board, and/or dance gaming partners
+from regular tabletop, board, and/or dance gaming partners
 to those who %inexhaustibly
-welcomed me with company and
-a couch to sleep on during travel,
+%welcomed me with
+offered
+company and
+a couch to sleep on while I traveled, %during travel,
 I thank
 %-- and in any order other than alphabetical here would be far too cruel --
-8 Gianfortoni,
+8~Gianfortoni,
 Adam Blank,
@@
 Andrew Krieger,
-Car Bauer,
+car bauer,
 Carolyn Sawyer,
-Elizabeth Kemp,
+Elizabeth ``Kempy'' Kemp,
 Elly Fong-Jones,
-Emily Forney,
-Emily Leathers,
+E.~Forney,
+Emily~J Leathers,
 Emma Cating-Subramanian,
 Eric Faust,
-Gabe Routh,
+Gabriel Mengde Routh,
 Greg Hanneman,
-Hazel Vird\'{o},
+hazel vird\'{o},
 Jack Ferris,
 Jake Lengyel,
-Jason Deng,
-Josh Keller,
+Jason ``Wyrm'' Deng,
+Joshua Keller,
 Joshua Wise,
@@
 Julia Tuttle,
-Kartik Subramanian,
+Kartik Null Cating-Subramanian,
 Kellie ``K2'' Medlin,
 Laura Abbott,
-Maija Mednieks,
+Maija E. Mednieks,
 Margaret Meyerhofer,
 Michael A\texttt{rntz}enius,
-Michael J. ``Sully'' Sullivan,
+Michael ``Sully'' Sullivan,
 Miriam Zimmerman,
@@
 Richard Ha,
-Rory Glenn,
+rglenn,
 Ryan Pearl,
+Todd Eisenberger,
 and
-Todd Eisenberger
+Zachary McCord
 for their wonderful friendship.
@@
 Cassie Orr,
+Charlie McGuffey,
 Dan Guzek,
@@
 Grant Wu,
-Jenny Lin,
+Isaac Grosof,
+jennylin,
 Karl Schulz,
@@
 Lea Albaugh,
-Marlena Abraham,
+Marlena N. Abraham,
 Matt Stanec,
 Matthew Maurer,
+Priya Donti,
+Roie Levin,
+Ryan Kavanagh,
 Sarah R. Allen,
 Sol Boucher,
+Sophia Wu,
 Stefan Muller,
@@
 Lastly,
-the Android: Netrunner community has become such an important part of my life
-over the last half decade or so.
+a shout out to the Android: Netrunner community.
+%has become such an important part of my life
+%over the last half decade or so.
 %for the better part of my time as a grad student.

### 2018-12-26 thesis/education.tex
change location of LFTs from 410user to landslide-friendly-tests
handle some todos, dont handle others

@@
 
-\noindent The tests can all be viewed at \url{https://github.com/bblum/landslide/tree/master/pebsim/p2-basecode/410user/progs}.
+\noindent The tests can all be viewed at
+%\url{https://github.com/bblum/landslide/tree/master/pebsim/p2-basecode/410user/progs}.
+\url{https://github.com/bblum/landslide/tree/master/pebsim/p2-basecode/landslide-friendly-tests}.
 
@@
 		% nice p value
-		Downgrade & 10	& 3	& 19	& 4	& $-05.7$\%	& 0.69	& 0.0125 \\ % TODO verify these buggers by eye
+		Downgrade & 10	& 3	& 19	& 4	& $-05.7$\%	& 0.69	& 0.0125 \\
 		\hline
@@
 their code, introducing the bug, without testing it again thereafter.
-% TODO: some reference to future work about focused testing on small code diffs? talked with some industry ppl about this sometimes
+% not-to-do: some reference to future work about focused testing on small code diffs? talked with some industry ppl about this sometimes
 

### 2018-12-26 thesis/tm.tex
handle some todos, dont handle others

@@
 but mostly to keep the implementation simple.
-As one might expect, it inflates state spaces considerably, % TODO: measure it with htm2 or some such?
+As one might expect, it inflates state spaces considerably,
 and moreover,

### 2018-12-26 thesis/education.tex
update HEAD commit and latest links

@@
 The PSU version of the lecture
-%was given in Spring 2018, and
+was given in Spring 2018, and
 is available at
-\url{http://www.contrib.andrew.cmu.edu/~bblum/psu-lecture.pdf}
+\url{https://www.cs.cmu.edu/~410/landslide/psu-lecture.pdf},
 as well as under the github link above.
@@
 which upon completion linked them to the Landslide user guide,
-which is available at
-\url{http://www.contrib.andrew.cmu.edu/~bblum/landslide-guide-p2.pdf}
+which is available online at
+\url{https://www.cs.cmu.edu/~410/landslide/landslide-guide-p2.pdf}
 (CMU version)
 and at % these "at"s can optionally be adjusted to make url line breaks better
-\url{http://www.contrib.andrew.cmu.edu/~bblum/landslide-guide-psu.pdf}
+\url{https://www.cs.cmu.edu/~410/landslide/landslide-guide-psu.pdf}
 (PSU version)

### 2018-12-26 thesis/landslide.tex
update HEAD commit and latest links

@@
 and available at \url{https://github.com/bblum/landslide}.
-% TODO: refresh
-The HEAD commit at the time of writing is 8127151.
+The HEAD commit at the time of writing is 5e4082a.
 In the event of a GitHub outage or the above link otherwise not working,
 a backup of the repository
-% checked out at the above commit
+checked out at that commit
 may be found at
-% TODO: make there actually be smth at this commit
-\url{https://www.cs.cmu.edu/~410/landslide.tar.bz2}.
+\url{https://www.cs.cmu.edu/~410/landslide/landslide.tar.bz2}.
 %
@@
 An alternate step-by-step guide is also available at
-\url{http://www.contrib.andrew.cmu.edu/~bblum/landslide-guide-kernel.pdf}.
+\url{https://www.cs.cmu.edu/~410/landslide/landslide-guide-kernel.pdf}.
 

### 2018-12-26 thesis/quicksand.tex
update HEAD commit and latest links

@@
 
-\vspace{2em}
+\vspace{1em}
 \qrevision{There is a fundamental disconnect between existing stateless model checkers
@@
 
-The contributions of this chapter were published as
-{\em Stateless Model Checking with Data-Race Preemption Points}
-in OOPSLA 2016 \cite{quicksand}.
+The main contributions of this chapter were published in
+the 2016 ACM SIGPLAN International Conference on Object-Oriented Programming, Systems, Languages, and Applications
+(OOPSLA '16)
+as
+{\em Stateless Model Checking with Data-Race Preemption Points}~\cite{quicksand}.
 

### 2018-12-26 thesis/education.tex
more revs

@@
 	\item {\tt alarm-simultaneous}:
-		Modified to be Landslide-friendly from the basecode, for {\em threads}.
+		Modified to be Landslide-friendly from the basecode, also for {\em threads}.
 		Creates two child threads which each invoke {\tt timer\_sleep()} for a different amount of time.
@@
 failed to confidently reject even the null hypothesis.}
-However, the distribution difference being smaller in P3 than in P2
-further suggests, albeit informally,
+However, the difference being smaller in P3 than in P2
+suggests, albeit informally,
 that the larger impact in P2 is not attributable {\em only} to selection bias,

### 2018-12-26 thesis/overview.tex
cpu-time or cpu time
handle some todos, dont handle others
more revs

@@
 		The primary challenge of this exercise is ensuring the aforementioned atomicity between {\tt cond\_wait}'s unlock and deschedule, with respect to the rest of the interface.
-	\item Semaphores, with the interface {\tt sem\_wait(sp)} and {\tt sem\_signal(sp)}
-		(sometimes called \revisionminor{$P$ and $V$ in other literature,
+	\item Semaphores, with the interface {\tt sem\_wait(sp)} and {\tt sem\_\allowbreak{}signal(sp)}
+		(sometimes also called \revisionminor{$P$ and $V$ in other literature,
 		from Dutch,
@@
 		Students typically implement semaphores using mutexes and condition variables, not using atomic instructions or system calls directly.
-	\item Reader-writer locks (also R/W-locks or rwlocks), with the interface {\tt rwlock\_lock(rwp, mode)} and {\tt rwlock\_unlock(rwp)}.
+	\item Reader-writer locks (also called R/W-locks or rwlocks), with the interface {\tt rwlock\_\allowbreak{}lock(rwp, mode)} and {\tt rwlock\_unlock(rwp)}.
 {\tt mode} may be either {\tt RWLOCK\_READ} or {\tt RWLOCK\_\allowbreak{}WRITE}.
@@
 		Under {\em total store order} (TSO),
-		implemented for example on x86,
+		which x86 implements,
 		one CPU's writes may appear reordered after its subsequent loads, from the perspective of another CPU.
@@
 		By contrast, {\em sequential consistency} (SC) permits no such reorderings.
-		% TODO: check that quicksand discussion shouldn't perhaps get a more subsection'ed sect reference
-		For further discussion see \cref{sec:quicksand-discussion} and \cref{sec:tm-warpzone-relaxed}.}
+		For further discussion see \cref{sec:quicksand-soundness}, \cref{sec:tm-warpzone-relaxed}, and \cref{sec:related-other-verif}.}
 	\item {\bf Schedule.}
@@
 		which provides the property that if a bug exists under the given test case,
-		given arbitrary CPU time,
+		given arbitrary testing time,
 		it will eventually be found;

### 2018-12-26 thesis/thesis.tex
more acks

@@
 movie watching,
-and/or as more permanent Pittsburgh residents that I am extremely torn up about moving away from:
+and/or as more permanent Pittsburgh residents that I am extremely
+sad to move away from:
+%torn up about moving away from:
 %leaving behind (for now at least):
@@
 Brendan McShane,
-Brian Saghy,
+Brian E. Saghy,
 Carlo Angiuli,
